### 2024-03-03 scripts/examples/manuscript/index.qmd
feat(scripts): add manuscript template Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "Notebook template"
toc: true
number-sections: true
highlight-style: pygments
csl: bibstyle.csl
format:
  nature-pdf:
    natbibstyle: sn-nature
    classoption: 
      - lineno
    cite-method: citeproc
    keep-tex: true
  html:
    code-fold: true
    html-math-method: katex
  docx: default
execute: 
  cache: true
  keep-ipynb: true
author:
  - name: Author One
    affiliations:
      - name: Harvard University
        id: 1
        department: Department of Government
        address: 1737 Cambridge Street
        city: Cambridge
        state: MA
        postal-code: 02138
      - name: Harvard University
        id: 2
        department: Department of Statistics
        address: 1 Oxford Street
        city: Cambridge
        state: MA
        postal-code: 02138
    attributes:
      equal-contributor: true
  - name: Author Two
    affiliations:
      - name: Yale University
        id: 3
        department: Department of Political Science
        address: 115 Prospect Street
        city: New Haven
        state: CT
        postal-code: 06511
    email: corresponding@email.com
    attributes:
      equal-contributor: true
      corresponding: true
  - name: Author Three
    affiliations:
      - ref: 1
abstract: |
  The abstract serves both as a general introduction to the topic and as a
  brief, non-technical summary of the main results and their implications.
  Authors are advised to check the author instructions for the journal they are
  submitting to for word limits and if structural elements like subheadings,
  citations, or equations are permitted.
keywords: [template, demo]
bibliography: references.bib
jupyter:
  jupytext:
    cell_metadata_filter: all
    cell_metadata_json: true
    notebook_metadata_filter: all
    text_representation:
      extension: .qmd
      format_name: quarto
      format_version: '1.0'
      jupytext_version: 1.16.0
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
---

# Summary {#sec-summary}

Use this file as a template for creating new notebooks that can be rendered to
pdf, html, and ipynb from the common source of truth in this markdown file. You
can include citations using the [default pandoc citation syntax][pandoc-citation]
`[@key01; @key02]` [@La_Manno2018-lj; @Qin2022-ls].

# Dependencies {#sec-dependencies}

Copies of this template will require the following dependencies:

- [quarto-cli][quarto-pypi] or [quarto][quarto-conda]
- [jupyter-cache][jupyter-cache-pypi]

Its other dependencies are not specific to the purpose of the template.

# Usage {#sec-usage}

Install the \LaTeX template with:

``` bash
quarto add cameronraysmith/nature@natbib
```

This will add an [_extensions/](./_extensions/) subdirectory containing
the template extension.

Render the notebook with:

``` bash
quarto render index.qmd --to all -P TEST_MODE:False
```

With the current configuration in the YAML header
this should produce

- [index.pdf](./index.pdf)
- [index.html](./index.html)
- [index.ipynb](./index.ipynb)

together with the 
[index_files](./index_files/) and 
[.jupyter_cache](./.jupyter_cache/) 
subdirectories that contain the quarto output and the cache of the jupyter
notebook execution. What follows are just a few example computations to
illustrate notebook content. At least @sec-dependencies and @sec-usage can be
deleted from any instance of this notebook.

# Setup {#sec-setup}

```{python}
#| execution: {iopub.execute_input: '2024-02-28T00:03:22.418885Z', iopub.status.busy: '2024-02-28T00:03:22.418672Z', iopub.status.idle: '2024-02-28T00:03:22.700860Z', shell.execute_reply: '2024-02-28T00:03:22.700498Z'}
#| lines_to_next_cell: 2
#| slideshow: {slide_type: fragment}
#| tags: []
import os

import matplotlib.pyplot as plt
import numpy as np
```

## Parameters {#subsec-setup-parameters}

The following cell defines parameters to be exposed to quarto or papermill
execution using the `tags: [parameters]` in the cell below

```{python}
#| execution: {iopub.execute_input: '2024-02-28T00:03:22.715524Z', iopub.status.busy: '2024-02-28T00:03:22.715310Z', iopub.status.idle: '2024-02-28T00:03:22.718434Z', shell.execute_reply: '2024-02-28T00:03:22.717807Z'}
#| slideshow: {slide_type: fragment}
#| tags: [parameters]
TEST_MODE = True
print(f"Test mode: {TEST_MODE}") 
```

Parameters injected by quarto or papermill should appear above.

## Functions {#subsec-setup-functions}

Define functions to be used in the notebook.

```{python}
#| execution: {iopub.execute_input: '2024-02-28T00:03:22.744628Z', iopub.status.busy: '2024-02-28T00:03:22.744439Z', iopub.status.idle: '2024-02-28T00:03:22.746661Z', shell.execute_reply: '2024-02-28T00:03:22.746335Z'}
#| slideshow: {slide_type: fragment}
#| tags: []
def plot_random_numbers(size=10):
    plt.plot(np.random.rand(size))
    plt.show()
```

This cell checks if the notebook is running in test mode. If it is, it will reduce the size of the data to be processed.

```{python}
#| execution: {iopub.execute_input: '2024-02-28T00:03:22.756996Z', iopub.status.busy: '2024-02-28T00:03:22.756828Z', iopub.status.idle: '2024-02-28T00:03:22.759487Z', shell.execute_reply: '2024-02-28T00:03:22.759002Z'}
#| slideshow: {slide_type: fragment}
#| tags: []
print(f"Test mode: {TEST_MODE}")
size = 5 if TEST_MODE else 10
print(f"Size: {size}")
```

# Results {#sec-results}

The results of the analysis are shown in @fig-random-numbers.

```{python}
#| label: fig-random-numbers 
#| fig-cap: "Plot of samples from the uniform distribution."
#| execution: {iopub.execute_input: '2024-02-28T00:03:22.775090Z', iopub.status.busy: '2024-02-28T00:03:22.774911Z', iopub.status.idle: '2024-02-28T00:03:22.854889Z', shell.execute_reply: '2024-02-28T00:03:22.854428Z'}
#| slideshow: {slide_type: fragment}
#| tags: []
plot_random_numbers(size)
```

# Postprocessing {#sec-postprocessing}

The notebook can also be converted to ipynb using

``` bash
quarto convert index.qmd
```

or vice-versa to markdown via

``` bash
quarto convert index.ipynb
```

figures should already be present in the [index_files/](./index_files/).
You can always rerender the notebook to generate them

``` bash
quarto render index.qmd --to all
```

If you are concerned that the cache me be invalidated without being detected
you can override it with

``` bash
quarto render index.qmd --to all --cache-refresh
```

# References

:::{#refs}
:::

<!-- links -->
[pandoc-citation]: https://pandoc.org/MANUAL.html#citation-syntax
[quarto-pypi]: https://pypi.org/project/quarto-cli/
[quarto-conda]: https://anaconda.org/conda-forge/quarto
[jupyter-cache-pypi]: https://pypi.org/project/jupyter-cache/

### 2024-03-05 scripts/examples/manuscript/index.qmd
fix(examples): set html font Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
   html:
+    mainfont: Latin Modern Roman
     code-fold: true

### 2024-03-17 nbs/index.qmd
docs(nbs): initialize nb site Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "Pyro-Velocity"
---

Pyro-Velocity is a Bayesian, generative, and multivariate RNA velocity model to estimate uncertainty in predictions of future cell states from minimal models approximating splicing dynamics.

### 2024-03-18 scripts/examples/manuscript/index.qmd
fix(scripts): comment examples authors Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-03-03 scripts/examples/manuscript/index.qmd
@@
 author:
-  - name: Author One
-    affiliations:
-      - name: Harvard University
-        id: 1
-        department: Department of Government
-        address: 1737 Cambridge Street
-        city: Cambridge
-        state: MA
-        postal-code: 02138
-      - name: Harvard University
-        id: 2
-        department: Department of Statistics
-        address: 1 Oxford Street
-        city: Cambridge
-        state: MA
-        postal-code: 02138
-    attributes:
-      equal-contributor: true
-  - name: Author Two
-    affiliations:
-      - name: Yale University
-        id: 3
-        department: Department of Political Science
-        address: 115 Prospect Street
-        city: New Haven
-        state: CT
-        postal-code: 06511
-    email: corresponding@email.com
-    attributes:
-      equal-contributor: true
-      corresponding: true
-  - name: Author Three
-    affiliations:
-      - ref: 1
+  - name: Pyrovelocity Authors
+  # - name: Qian Qin
+  #   affiliations:
+  #     - name: Broad Institute of MIT and Harvard
+  #       department: Broad Institute of MIT and Harvard
+  #       id: 1
+  #       city: Cambridge
+  #       state: MA
+  #   email: qqin@broadinstitute.org
+  #   attributes:
+  #     equal-contributor: true
+  # - name: Cameron Smith
+  #   affiliations:
+  #     - ref: 1
+  #     - name: Harvard Medical School
+  #       id: 2
+  #       department: Department of Pathology
+  #       city: Boston
+  #       state: MA
+  #     - name: Massachusetts General Hospital
+  #       id: 3
+  #       department: Molecular Pathology Unit, Center for Cancer Research
+  #       city: Boston
+  #       state: MA
+  #   email: cameron.smith@mgh.harvard.edu
+  #   attributes:
+  #     equal-contributor: true
+  # - name: Luca Pinello
+  #   affiliations:
+  #     - ref: 1
+  #     - ref: 2
+  #     - ref: 3
+  #   email: lpinello@mgh.harvard.edu
+  #   attributes:
+  #     corresponding: true
 abstract: |

### 2024-03-18 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
feat(scripts): init pypi template Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Pyrovelocity PyPI user notebook template
toc: true
number-sections: true
highlight-style: pygments
csl: bibstyle.csl
format:
  nature-pdf:
    natbibstyle: sn-nature
    # classoption:
    #   - lineno
    cite-method: citeproc
    keep-tex: true
  html:
    mainfont: Latin Modern Roman
    code-fold: false
    html-math-method: katex
  docx: default
execute: 
  eval: true
  warning: false
  error: false
  cache: true
  keep-ipynb: true
author:
  - name: Pyrovelocity Team
abstract: |
  This notebook demonstrates how to run pyrovelocity in a Jupyter notebook.
  It attempts to support both Google Colab and local Jupyter environments.
keywords: [single-cell genomics, probabilistic modeling, dynamical systems]
bibliography: references.bib
jupyter:
  jupytext:
    cell_metadata_filter: all
    cell_metadata_json: true
    notebook_metadata_filter: all
    text_representation:
      extension: .qmd
      format_name: quarto
      format_version: '1.0'
      jupytext_version: 1.16.0
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
  language_info:
    name: python
  rise:
    scroll: true
    theme: black
  toc-autonumbering: true
  toc-showcode: false
  toc-showmarkdowntxt: false
---

<a target="_blank" href="https://colab.research.google.com/github/pinellolab/pyrovelocity/blob/530-nbwf/docs/source/notebooks/pyrovelocity_pypi_user_template.ipynb">
  <img 
    src="https://colab.research.google.com/assets/colab-badge.svg" 
    alt="Open In Colab"
    width="109" height="20"/>
</a> <a target="_blank" href="https://nbviewer.jupyter.org/github/pinellolab/pyrovelocity/blob/530-nbwf/docs/source/notebooks/pyrovelocity_pypi_user_template.ipynb">
  <img 
    src="https://raw.githubusercontent.com/jupyter/design/master/logos/Badges/nbviewer_badge.png"
    alt="Render with nbviewer" 
    width="109" height="20"/>
</a>

## Setup environment

Installation should take less than **5 minutes**. 
It involves checking if the notebook is running in colab, in which case it is certain that you will need to install or reinstall pyrovelocity.
Otherwise, it is assumed that the user will have installed pyrovelocity in the kernel environment.

The simplest way to complete it is to run the `Setup environment` section, wait for the kernel to restart, and then run the same section again:

- **fold** this `Setup environment` section above
- **click the play button** underneath the section name to run the whole section for the first time (installs condacolab)
- **wait** for the kernel to restart 
  - **ignore** expected notice in bottom left of Colab UI
    - `Your session crashed for an unknown reason. View runtime logs`
  - **ignore** `SystemExit` output in the `Install condacolab` subsection below
- **refold** the `Setup environment` section (`SystemExit` / kernel restart will unfold it)
- **proceed to Analysis** below

Otherwise, the cells below can be executed manually.
In either case, this section can be folded away after installation is complete.

If you need to edit the **version number**, please see below.

### Install pyrovelocity

This first stage will download and install [pyrovelocity](https://github.com/pinellolab/pyrovelocity). This usually takes less than **4 minutes**. The runtime will then automatically restart. After this you can execute "Run all" to complete installation or proceed linearly below if you have added additional content you do not want to run all at once.

#### Define functions to manage installation of python libraries

```{python}
# | code-fold: true
import importlib.util
import subprocess
import sys

def is_module_available(module_name: str):
    return importlib.util.find_spec(module_name) is not None

def install_package(package_name: str):
    """
    Install a package using pip. This is similar to cell magic
    `!pip install package_name`, but in python code for compatibility
    outside jupyter.

    Args:
        package_name (str): Name of the package to install.
    """
    process = subprocess.Popen(
        ["pip", "install", "-q", package_name],
        stdout=subprocess.PIPE,
        stderr=subprocess.PIPE,
        text=True,
    )
    for line in process.stdout:
        print(line, end="")
    for line in process.stderr:
        print(line, end="")
    process.wait()

def setup_pyrovelocity():
    if is_module_available("pyrovelocity"):
        try:
            import pyrovelocity
        except (ImportError, AssertionError):
            print("pyrovelocity is not successfully installed")
            sys.exit()
    else:
        print("Installing pyrovelocity...")
        install_package("pyrovelocity==0.2.0b12")
        try:
            import pyrovelocity

            print(
                "\nThe kernel needs to restart in order to use pyrovelocity.\n"
                "Please run this cell again.\n"
            )
            sys.exit()
        except (ImportError, AssertionError):
            print("Failed to install pyrovelocity properly.")
            sys.exit()
```

#### Install pyrovelocity

```{python}
# | code-fold: true
import os

IN_COLAB = is_module_available("google.colab")

if IN_COLAB:
    colab_release_tag = os.getenv("COLAB_RELEASE_TAG", None)
    print(f"Google Colab release: {colab_release_tag}")
    setup_pyrovelocity()
else:
    print("This notebook is probably not running in Google Colab")
```

### Check installation

If installation was successful, the following commands should print the location of the `__init__.py` file for the pyrovelocity package and the currently installed version.

```{python}
import pyrovelocity
print(pyrovelocity.__file__)
print(pyrovelocity.__version__)
```

This is the same for the pyro package in case there was an issue with pyrovelocity install that did not affect another package.

```{python}
# | code-fold: true
import pyro
print(pyro.__file__)
print(pyro.__version__)
```

Please refer to the [docs](https://pinellolab.github.io/pyrovelocity) for tutorials and usage information.

## Analysis

```{python}
help(pyrovelocity) # ?pyrovelocity # to open in side panel tab for reference
```

Before we start, we will set an environment variables to ensure our first execution occurs in a lightweight test mode with a subset of observations, variables, training epochs, and posterior samples.

```{python}
import os
os.environ["PYROVELOCITY_TESTING_FLAG"] = "True"
```

After an initial review, we can set the environment variable to `False` to run the full analysis.

The library supports execution via a sequence of workflow tasks. The approximate outline of these involves accessing external data, preprocessing, model training, postprocessing, and summarization.

We import these tasks and execute them in the subsections below.

```{python}
from dataclasses import asdict

from pyrovelocity.workflows.main_workflow import (
  download_data,
  preprocess_data,
  train_model,
  postprocess_data,
  summarize_data
)
```

To execute each task requires a single `WorkflowConfiguration` object, which is an instance of a python datclass. In this notebook we illustrate execution with the standard pancreatic endocrinogenesis data set [@Bastidas-Ponce2019-lf]. First we import the configuration dataclass

```{python}
from pyrovelocity.workflows.main_configuration import pancreas_configuration
```

We can review the configuration dataclass by writing it to a yaml file or printing the dictionary representation to the console.

```{python}
import yaml
from pyrovelocity.utils import pretty_print_dict
pancreas_configuration_dict = pancreas_configuration.to_dict()

with open("pancreas_configuration.yaml", "w") as yaml_file:
    yaml.dump(pancreas_configuration_dict, yaml_file, sort_keys=False, default_flow_style=False, allow_unicode=True)
pretty_print_dict(pancreas_configuration_dict)
```

Feel free to open the [pancreas_configuration.yaml](./pancreas_configuration.yaml) file to review the configuration settings. The resource requests and limits will be irrelevant for this example of local execution, but are utilized during execution of distributed containerized workflows.

### Download data

To download data, we simply

```{python}
pretty_print_dict(pancreas_configuration.download_dataset.to_dict())
data = download_data(download_dataset_args=pancreas_configuration.download_dataset)
```

### Preprocess data

```{python}
pretty_print_dict(pancreas_configuration.preprocess_data.to_dict())
processed_data = preprocess_data(
  data=data,
  preprocess_data_args=pancreas_configuration.preprocess_data,
)
```

### 2024-03-19 src/pyrovelocity/utils.py
refactor(utils): host str_to_bool helper Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -45,6 +45,7 @@ __all__ = [
     "print_anndata",
     "print_attributes",
     "save_anndata_counts_to_dataframe",
+    "str_to_bool",
 ]
 
 logger = configure_logging(__name__)
@@ -182,6 +183,28 @@ def pretty_log_dict(d: dict) -> str:
 @beartype
 def pretty_print_dict(d: dict):
     logger.info(pretty_log_dict(d))
+
+@beartype
+def str_to_bool(value: str | bool, default: bool = False) -> bool:
+    """
+    Convert strings that could be interpreted as booleans to a boolean value,
+    with a default fallback.
+
+    Args:
+        value (str | bool): input string or boolean value.
+        default (bool, optional): Defaults to False.
+
+    Returns:
+        bool: boolean interpretation of the input string
+    """
+    if isinstance(value, bool):
+        return value
+    if value.lower() in ("true", "t", "1", "yes", "y"):
+        return True
+    elif value.lower() in ("false", "f", "0", "no", "n"):
+        return False
+    else:
+        return default
 
 
 def filter_startswith_dict(dictionary_with_underscore_keys):

### 2024-03-19 src/pyrovelocity/workflows/main_configuration.py

@@ -11,6 +11,7 @@ from pyrovelocity.interfaces import DownloadDatasetInterface
 from pyrovelocity.interfaces import PreprocessDataInterface
 from pyrovelocity.interfaces import PyroVelocityTrainInterface
 from pyrovelocity.logging import configure_logging
+from pyrovelocity.utils import str_to_bool
 
 
 __all__ = [
@@ -24,27 +25,6 @@ __all__ = [
 
 
 logger = configure_logging(__name__)
-
-
-@beartype
-def str_to_bool(value: str, default: bool = False) -> bool:
-    """
-    Convert strings that could be interpreted as booleans to a boolean value,
-    with a default fallback.
-
-    Args:
-        value (str): input string
-        default (bool, optional): Defaults to False.
-
-    Returns:
-        bool: boolean interpretation of the input string
-    """
-    if value.lower() in ("true", "t", "1", "yes", "y"):
-        return True
-    elif value.lower() in ("false", "f", "0", "no", "n"):
-        return False
-    else:
-        return default
 
 
 PYROVELOCITY_TESTING_FLAG = str_to_bool(

### 2024-03-19 src/pyrovelocity/__init__.py
fix(pyrovelocity): update dunder init Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,6 +1,37 @@
-"""pyrovelocity."""
+"""
+Pyro-Velocity
+
+Pyro-Velocity is a python package for probabilistic modeling of RNA velocity.
+
+docs: https://pyrovelocity.readthedocs.io
+source: https://github.com/pinellolab/pyrovelocity
+PyPI package: https://pypi.org/project/pyrovelocity
+Conda package: https://anaconda.org/conda-forge/pyrovelocity
+discussions: https://github.com/pinellolab/pyrovelocity/discussions
+issues: https://github.com/pinellolab/pyrovelocity/issues
+"""
 
 from importlib import metadata
+
+import pyrovelocity.analyze
+import pyrovelocity.cytotrace
+import pyrovelocity.data
+import pyrovelocity.datasets
+import pyrovelocity.interfaces
+import pyrovelocity.io
+import pyrovelocity.logging
+import pyrovelocity.metrics
+import pyrovelocity.models
+import pyrovelocity.plot
+import pyrovelocity.plots
+import pyrovelocity.postprocess
+import pyrovelocity.preprocess
+import pyrovelocity.summarize
+import pyrovelocity.tests
+import pyrovelocity.train
+import pyrovelocity.utils
+import pyrovelocity.workflows
+
 
 try:
     __version__ = metadata.version(__package__)
@@ -8,3 +39,23 @@ except metadata.PackageNotFoundError:
     __version__ = "unknown"
 
 del metadata
+
+__all__ = [
+    "analyze",
+    "cytotrace",
+    "data",
+    "datasets",
+    "interfaces",
+    "io",
+    "logging",
+    "metrics",
+    "models",
+    "plot",
+    "plots",
+    "postprocess",
+    "preprocess",
+    "summarize",
+    "train",
+    "utils",
+    "workflows",
+]

### 2024-03-19 src/pyrovelocity/utils.py
feat(utils): add print_config_tree Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -7,18 +7,23 @@ from numbers import Real
 from pathlib import Path
 from pprint import pprint
 from types import FunctionType
+from types import ModuleType
+from typing import Callable
 from typing import List
 from typing import Tuple
 
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
+import rich.syntax
+import rich.tree
 import scvelo as scv
 import seaborn as sns
+import yaml
 from anndata._core.anndata import AnnData
 from beartype import beartype
+from flytekit.core.local_cache import LocalTaskCache
 from scvi.data import synthetic_iid
-from termcolor import colored
 
 from pyrovelocity.io.compressedpickle import CompressedPickle
 from pyrovelocity.logging import configure_logging
@@ -34,6 +39,7 @@ from pyrovelocity.logging import configure_logging
 __all__ = [
     "anndata_counts_to_df",
     "attributes",
+    "clear_local_cache",
     "ensure_numpy_array",
     "filter_startswith_dict",
     "generate_public_api",
@@ -44,6 +50,7 @@ __all__ = [
     "pretty_print_dict",
     "print_anndata",
     "print_attributes",
+    "print_config_tree",
     "save_anndata_counts_to_dataframe",
     "str_to_bool",
 ]
@@ -183,6 +190,42 @@ def pretty_log_dict(d: dict) -> str:
 @beartype
 def pretty_print_dict(d: dict):
     logger.info(pretty_log_dict(d))
+
+
+@beartype
+def print_config_tree(
+    config: dict,
+    name: str = "",
+    theme: str = "nord-darker",
+    max_width: int = 148,
+):
+    config_yaml = yaml.dump(config, default_flow_style=False)
+    tree = rich.tree.Tree(name, style="dim", guide_style="dim")
+    tree.add(rich.syntax.Syntax(config_yaml, "yaml", theme=theme))
+
+    console = rich.console.Console(width=max_width)
+    console.print(tree)
+
+
+@beartype
+def print_docstring(
+    obj: Callable | ModuleType,
+    theme: str = "nord-darker",
+):
+    docstring = inspect.getdoc(obj)
+    if docstring is None:
+        rich.print("[bold red]No docstring found for this object.[/bold red]")
+        return
+    syntax_highlighted_docstring = rich.syntax.Syntax(
+        docstring, "python", theme=theme, line_numbers=False
+    )
+    rich.print(syntax_highlighted_docstring)
+
+
[…]

### 2024-03-19 src/pyrovelocity/train.py
fix(train): update docstring Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -54,48 +54,62 @@ def train_dataset(
     Loads processed data, trains model, and saves model and posterior samples.
 
     Inputs:
-        "data/processed/{data_set_name}_processed.h5ad"
+        data/processed/{data_set_name}_processed.h5ad
 
     Outputs:
-        data:
-            "models/{data_model}/trained.h5ad"
-            "models/{data_model}/posterior_samples.pkl.zst"
-        models:
-            models/{data_model}/model/
+        models/{data_model}/
+        ├── trained.h5ad
+        ├── posterior_samples.pkl.zst
+        └── model/
             ├── attr.pkl
             ├── model_params.pt
             ├── param_store_test.pt
             └── var_names.csv
 
-    "models/{data_model}/pyrovelocity.pkl" is produced by `postprocess_dataset`.
+    The file `models/{data_model}/pyrovelocity.pkl` used in
+    `pyrovelocity.summarize.summarize_dataset` is produced by
+    `pyrovelocity.postprocess.postprocess_dataset`.
 
 
     Args:
-        adata (str | AnnData): _description_
-        data_set_name (str, optional): _description_. Defaults to "simulated".
-        model_identifier (str, optional): _description_. Defaults to "model2".
-        guide_type (str, optional): _description_. Defaults to "auto".
-        model_type (str, optional): _description_. Defaults to "auto".
-        batch_size (int, optional): _description_. Defaults to -1.
-        use_gpu (str, optional): _description_. Defaults to "auto".
-        likelihood (str, optional): _description_. Defaults to "Poisson".
-        num_samples (int, optional): _description_. Defaults to 30.
-        log_every (int, optional): _description_. Defaults to 100.
-        patient_improve (float, optional): _description_. Defaults to 1e-4.
-        patient_init (int, optional): _description_. Defaults to 45.
-        seed (int, optional): _description_. Defaults to 99.
-        learning_rate (float, optional): _description_. Defaults to 0.01.
-        max_epochs (int, optional): _description_. Defaults to 3000.
-        include_prior (bool, optional): _description_. Defaults to True.
-        library_size (bool, optional): _description_. Defaults to True.
-        offset (bool, optional): _description_. Defaults to True.
-        input_type (str, optional): _description_. Defaults to "raw".
-        cell_specific_kinetics (Optional[str], optional): _description_. Defaults to None.
-        kinetics_num (int, optional): _description_. Defaults to 2.
-        force (bool, optional): _description_. Defaults to False.
[…]

### 2024-03-19 src/pyrovelocity/preprocess.py
fix(preprocess): update docstring Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -59,10 +59,21 @@ def preprocess_dataset(
     Args:
         data_set_name (str): name of data set.
         adata (AnnData, optional): AnnData object containing the data set to preprocess.
-        data_path (str | Path, optional): path to data set. Defaults to None.
+        data_processed_path (str | Path, optional): path to processed data. Defaults to "data/processed".
+        overwrite (bool, optional): overwrite existing processed data. Defaults to False.
         n_top_genes (int, optional): number of genes to retain. Defaults to 2000.
         min_shared_counts (int, optional): minimum shared counts. Defaults to 30.
-        processed_path (str, optional): path to read/write processed AnnData. Defaults to None.
+        process_cytotrace (bool, optional): process data with cytotrace. Defaults to False.
+        use_obs_subset (bool, optional): use observation subset. Defaults to False.
+        n_obs_subset (int, optional): number of observations to subset. Defaults to 300.
+        use_vars_subset (bool, optional): use variable subset. Defaults to False.
+        n_vars_subset (int, optional): number of variables to subset. Defaults to 200.
+        count_threshold (int, optional): count threshold. Defaults to 0.
+        n_pcs (int, optional): number of principal components. Defaults to 30.
+        n_neighbors (int, optional): number of neighbors. Defaults to 30.
+        default_velocity_mode (str, optional): default velocity mode. Defaults to "dynamical".
+        vector_field_basis (str, optional): vector field basis. Defaults to "umap".
+        cell_state (str, optional): Name of the cell state/cluster variable. Defaults to "clusters".
 
     Returns:
         AnnData: processed AnnData object

### 2024-03-19 src/pyrovelocity/workflows/main_workflow.py
fix(workflows): read cache flag from env var
- `PYROVELOCITY_CACHE_FLAG`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,3 +1,4 @@
+import os
 from dataclasses import asdict
 from datetime import timedelta
 
@@ -24,6 +25,7 @@ from pyrovelocity.postprocess import postprocess_dataset
 from pyrovelocity.preprocess import preprocess_dataset
 from pyrovelocity.summarize import summarize_dataset
 from pyrovelocity.train import train_dataset
+from pyrovelocity.utils import str_to_bool
 from pyrovelocity.workflows.main_configuration import (
     PYROVELOCITY_SIMULATED_ONLY,
 )
@@ -62,12 +64,14 @@ __all__ = [
 logger = configure_logging(__name__)
 
 CACHE_VERSION = "0.2.0b12"
-CACHE_FLAG = True
+PYROVELOCITY_CACHE_FLAG = str_to_bool(
+    os.getenv("PYROVELOCITY_CACHE_FLAG", "True")
+)
 ACCELERATOR_TYPE: GPUAccelerator = T4
 
 
 @task(
-    cache=CACHE_FLAG,
+    cache=PYROVELOCITY_CACHE_FLAG,
     cache_version=CACHE_VERSION,
     retries=3,
     interruptible=True,
@@ -85,7 +89,7 @@ def download_data(download_dataset_args: DownloadDatasetInterface) -> FlyteFile:
 
 
 @task(
-    cache=CACHE_FLAG,
+    cache=PYROVELOCITY_CACHE_FLAG,
     cache_version=CACHE_VERSION,
     retries=3,
     interruptible=False,
@@ -111,7 +115,7 @@ def preprocess_data(
 
 
 @task(
-    cache=CACHE_FLAG,
+    cache=PYROVELOCITY_CACHE_FLAG,
     cache_version=CACHE_VERSION,
     retries=3,
     interruptible=False,
@@ -169,7 +173,7 @@ def train_model(
 
 
 @task(
-    cache=CACHE_FLAG,
+    cache=PYROVELOCITY_CACHE_FLAG,
     cache_version=CACHE_VERSION,
     retries=3,
     interruptible=False,
@@ -209,7 +213,7 @@ def postprocess_data(
 
 
 @task(
-    cache=CACHE_FLAG,
+    cache=PYROVELOCITY_CACHE_FLAG,
     cache_version=CACHE_VERSION,
     retries=3,
     interruptible=False,
@@ -254,7 +258,7 @@ def summarize_data(
 
 
 @task(
[…]

### 2024-03-19 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
feat(pypi_template): add workflow execution example Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 csl: bibstyle.csl
+lightbox: auto
 format:
@@
   This notebook demonstrates how to run pyrovelocity in a Jupyter notebook.
-  It attempts to support both Google Colab and local Jupyter environments.
+  It attempts to support both Google Colab and local or remote Jupyter kernel servers.
 keywords: [single-cell genomics, probabilistic modeling, dynamical systems]
@@
       format_name: quarto
-      format_version: '1.0'
+      format_version: 1.0
       jupytext_version: 1.16.0
@@
 - **fold** this `Setup environment` section above
-- **click the play button** underneath the section name to run the whole section for the first time (installs condacolab)
+- **click the play button** underneath the section name to run the whole section for the first time
 - **wait** for the kernel to restart 
@@
     - `Your session crashed for an unknown reason. View runtime logs`
-  - **ignore** `SystemExit` output in the `Install condacolab` subsection below
+  - **ignore** `SystemExit` output in the `Install pyrovelocity` subsection below
 - **refold** the `Setup environment` section (`SystemExit` / kernel restart will unfold it)
@@
 
-If you need to edit the **version number**, please see below.
+If you need to edit the **version number**, please see the setting in the `setup_pyrovelocity` function below.
 
@@
 ```{python}
-# | code-fold: true
+#| code-fold: true
 import importlib.util
@@
 ```{python}
-# | code-fold: true
+#| code-fold: true
 import os
@@
 ```{python}
+#| output: false
+
 import pyrovelocity
@@
 ```{python}
-# | code-fold: true
+#| code-fold: true
+#| output: false
+
 import pyro
@@
 
-Before we start, we will set an environment variables to ensure our first execution occurs in a lightweight test mode with a subset of observations, variables, training epochs, and posterior samples.
+Before we start, we will set environment variables to ensure our first execution occurs in a lightweight test mode with a subset of observations, variables, training epochs, and posterior samples.
 
@@
 ```{python}
-from dataclasses import asdict
+from importlib import reload
+import yaml
 
+import pyrovelocity.utils
+reload(pyrovelocity.utils)
+from pyrovelocity.utils import (
+  print_config_tree,
+  print_docstring,
+)
+
+import pyrovelocity.workflows.main_workflow
+reload(pyrovelocity.workflows.main_workflow)
 from pyrovelocity.workflows.main_workflow import (
@@
   postprocess_data,
-  summarize_data
+  summarize_data,
 )
@@
 ```{python}
-import yaml
-from pyrovelocity.utils import pretty_print_dict
 pancreas_configuration_dict = pancreas_configuration.to_dict()
+print_config_tree(pancreas_configuration_dict, "pancreas configuration")
 
@@
     yaml.dump(pancreas_configuration_dict, yaml_file, sort_keys=False, default_flow_style=False, allow_unicode=True)
-pretty_print_dict(pancreas_configuration_dict)
 ```
@@
 Feel free to open the [pancreas_configuration.yaml](./pancreas_configuration.yaml) file to review the configuration settings.
-The resource requests and limits will be irrelevant for this example of local execution, but are utilized during execution of distributed containerized workflows.
+We will reprint the part of that configuration relevant to the execution of each task below.
+The resource requests and limits sections at the bottom will be irrelevant for this example of local execution.
+The resource specifications are utilized during local or remote cluster-based execution of distributed containerized workflows.
 
@@
 
-To download data, we simply
+To download data, we provide the `download_data` task function with the `download_dataset` attribute of the `pancreas_configuration` object.
 
 ```{python}
-pretty_print_dict(pancreas_configuration.download_dataset.to_dict())
+print_config_tree(pancreas_configuration.download_dataset.to_dict(), "download dataset")
+```
+
+This configuration primarily specifies the location to store the data and the data set name that is mapped to a source URL in the `pyrovelocity.utils.datasets` module.
+The other parameters can be used to specify a custom source URL or to filter the observations and variables to a subset of the full data set; however, since we are working with a natively supported data set we do not need to specify those here.
+
+::: {.callout-note collapse=true title="Local caching"}
+During local execution, all of the functions imported from the `pyrovelocity.workflows.main_workflow` module will cache their results.
+This means that if you run the same task function with the same arguments, it will not recompute the result, but just establish the presence of a cache hit.
+This is useful for avoiding redundant computation, but if you want to re-execute the function for testing, experimentation or other purposes.
+To avoid this, you can use the `pyrovelocity.utils.clear_local_cache` function to clear the cache.
+If you add the following block of code, the cache will be cleared whenever it is executed:
+
+```python
+from pyrovelocity.utils import clear_local_cache
+
+clear_local_cache()
+```
+
+You can also disable the cache setting the `PYROVELOCITY_CACHE_FLAG` environment variable to `False`.
+For example, you can add the following block of code to the notebook to disable the cache for all task functions:
+
+```python
+import os
+os.environ["PYROVELOCITY_CACHE_FLAG"] = "False"
+```
+
+For reference, during local execution, the cache is managed with a sqlite database by the [diskcache](https://grantjenks.com/docs/diskcache/) library via [flytekit](https://docs.flyte.org/).
+:::
+
+```{python}
 data = download_data(download_dataset_args=pancreas_configuration.download_dataset)
@@
 
+The preprocess data task is executed with the `preprocess_data` function.
+The `preprocess_data` function takes the `FlyteFile` object, here named `data` above, from `download_data` task function and the `preprocess_data` attribute of the `pancreas_configuration` object as arguments.
+
 ```{python}
-pretty_print_dict(pancreas_configuration.preprocess_data.to_dict())
+print_config_tree(pancreas_configuration.preprocess_data.to_dict(), "preprocess data")
+```
+
+The components of the preprocess data configuration are determined by the components of the `pyrovelocity.preprocess.preprocess_dataset` function whose documentation can be accessed via `help` or `?` in the notebook
+
+```{python}
+print_docstring(pyrovelocity.preprocess.preprocess_dataset)
+```
+
+Recall that the output of the `preprocess_data` function below will be cached and should rerun almost instantaneously if you re-execute the cell multiple times.
+
+```{python}
 processed_data = preprocess_data(
@@
 ```
+
+### Train model
+
+The train model task is executed with the `train_model` function.
+The `train_model` task function takes the `FlyteFile` object, here named `processed_data` above, from `preprocess_data` task function and one of the `PyrovelocityTrainInterface` objects such as is found in the `training_configuration_2` attribute of the `pancreas_configuration` object as arguments.
+
+The configuration is given by
+
+```{python}
+print_config_tree(pancreas_configuration.training_configuration_2.to_dict(), "train model 2")
+```
+
+The `PyrovelocityTrainInterface` is a configuration for the `pyrovelocity.train.train_dataset` function whose documentation we print below.
+
+```{python}
+print_docstring(pyrovelocity.train.train_dataset)
+```
+
+Finally, we execute the `train_model` task function
+
+```{python}
+model_output = train_model(
+  processed_data=processed_data,
+  train_model_configuration=pancreas_configuration.training_configuration_2,
+)
+```
+
+This produces the `model_output` object, which is an instance of the `TrainingOutputs` dataclass.
+
+```{python}
+print_config_tree(model_output.to_dict(), "model output")
+```
+
+Due to caching, the paths to the model outputs will involve temporary folders, but the relevant outputs are also available in the `models` folder next to the notebook.
+
+### Postprocess data
+
+The postprocess data task is executed with the `postprocess_data` function.
+The `postprocess_data` function takes the preprocess data configuration, the training outputs, here named `model_output` above, from the `train_model` task function and the `postprocess_configuration` attribute of the `pancreas_configuration` object as arguments.
+
+```{python}
+print_config_tree(pancreas_configuration.postprocess_configuration.to_dict(), "postprocess data")
+```
+
+The configuration required to postprocess the data are determined by the components of the `pyrovelocity.postprocess.postprocess_dataset` function interface whose documentation can be accessed via `help` or `?` in the notebook
+
+```{python}
+print_docstring(pyrovelocity.postprocess.postprocess_dataset)
+```
+
+We execute the `postprocess_data` task function with the inputs described above.
+
+```{python}
+postprocessing_outputs = postprocess_data(
+  preprocess_data_args=pancreas_configuration.preprocess_data,
+  training_outputs=model_output,
+  postprocess_configuration=pancreas_configuration.postprocess_configuration,
+)
+```
+
+## Results
+
+Now we have what is required to construct a summary of the results.
+The `summarize_data` task function takes the preprocessing configuration, `postprocessing_outputs` object and the `model_outputs` and generates a suite of plots summarizing the results.
+
+```{python}
+#| output: false
+dataset_summary = summarize_data(
+  preprocess_data_args=pancreas_configuration.preprocess_data,
+  postprocessing_outputs=postprocessing_outputs,
+  training_outputs=model_output,
+)
+```
+
+The dataset summary contains
+
+```{python}
+print_config_tree(dataset_summary.to_dict(), "dataset summary")
+```
+
+### Vector fields
+
+@fig-vector-fields-uncertainty shows a low-dimensional representation of the RNA velocity vector fields with uncertainty estimates for shared time and angle.
+The latter quantity is estimated in PCA space.
+The base magnitude uncertainty is also shown but may not be interpretable.
+
+```{python}
+#| label: fig-vector-fields-uncertainty
+#| echo: false
+#| fig-cap: "Vector fields with uncertainty"
+
+from IPython.display import Image, display
+
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part1_plot.pdf.png"))
+```
+
+A standard vector field plot with cluster annotations is shown in @fig-vector-field.
+
+```{python}
+#| label: fig-vector-field
+#| echo: false
+#| fig-cap: "Vector field"
+
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/vector_field.pdf.png"))
+```

### 2024-03-20 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
feat(pypi_template): generate and display figure outputs Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
   html:
-    mainfont: Latin Modern Roman
+    mainfont: Latin Modern Sans
     code-fold: false
     html-math-method: katex
+    embed-resources: true
   docx: default
+  ipynb: default
+format-links: false
 execute: 
@@
   keep-ipynb: true
+  keep-md: true
 author:
   - name: Pyrovelocity Team
+    email: team@pyrovelocity.net
 abstract: |
@@
   It attempts to support both Google Colab and local or remote Jupyter kernel servers.
-keywords: [single-cell genomics, probabilistic modeling, dynamical systems]
+keywords: [single-cell transcriptomics, probabilistic modeling, dynamical systems, RNA velocity]
 bibliography: references.bib
@@
 ```{python}
+#| label: install-package-helpers
 #| code-fold: true
@@
 ```{python}
+#| label: install-pyrovelocity-colab
 #| code-fold: true
@@
 ```{python}
+#| label: check-pyrovelocity-installation
 #| output: false
@@
 ```{python}
+#| label: check-pyro-installation
 #| code-fold: true
@@
 ```{python}
+#| label: print-pyrovelocity-docstring
 help(pyrovelocity) # ?pyrovelocity # to open in side panel tab for reference
@@
 
-Before we start, we will set environment variables to ensure our first execution occurs in a lightweight test mode with a subset of observations, variables, training epochs, and posterior samples.
+Before we start, we will adjust a few environment variables to ensure our first execution occurs in a lightweight test mode with a subset of observations, variables, training epochs, and posterior samples.
+We will also set a few log levels.
+Feel free to adjust these to `ERROR`, `DEBUG`, etc as needed.
 
 ```{python}
+#| label: set-pyrovelocity-testing-flag
 import os
 os.environ["PYROVELOCITY_TESTING_FLAG"] = "True"
+os.environ["PYROVELOCITY_LOG_LEVEL"] = "INFO"
+
+import logging
+logging.getLogger("jax").setLevel(logging.ERROR)
+logging.getLogger("root").setLevel(logging.ERROR)
 ```
@@
 ```{python}
+#| label: import-notebook-dependencies
 from importlib import reload
+
 import yaml
+from IPython.display import Image
+# from IPython.display import display
 
 import pyrovelocity.utils
+import pyrovelocity.workflows.main_workflow
 reload(pyrovelocity.utils)
-from pyrovelocity.utils import (
-  print_config_tree,
-  print_docstring,
-)
+reload(pyrovelocity.workflows.main_workflow)
 
-import pyrovelocity.workflows.main_workflow
-reload(pyrovelocity.workflows.main_workflow)
-from pyrovelocity.workflows.main_workflow import (
-  download_data,
-  preprocess_data,
-  train_model,
-  postprocess_data,
-  summarize_data,
-)
+from pyrovelocity.utils import print_config_tree
+from pyrovelocity.utils import print_docstring
+from pyrovelocity.workflows.main_workflow import download_data
+from pyrovelocity.workflows.main_workflow import postprocess_data
+from pyrovelocity.workflows.main_workflow import preprocess_data
+from pyrovelocity.workflows.main_workflow import summarize_data
+from pyrovelocity.workflows.main_workflow import train_model
 ```
@@
 ```{python}
+#| label: import-pancreas-configuration
 from pyrovelocity.workflows.main_configuration import pancreas_configuration
@@
 ```{python}
+#| label: print-and-save-pancreas-configuration
 pancreas_configuration_dict = pancreas_configuration.to_dict()
@@
 ```{python}
+#| label: print-download-dataset-configuration
 print_config_tree(pancreas_configuration.download_dataset.to_dict(), "download dataset")
@@
 ```{python}
+#| label: execute-download-data-task
 data = download_data(download_dataset_args=pancreas_configuration.download_dataset)
@@
 ```{python}
+#| label: print-preprocess-data-configuration
 print_config_tree(pancreas_configuration.preprocess_data.to_dict(), "preprocess data")
@@
 ```{python}
+#| label: print-preprocess-data-docstring
 print_docstring(pyrovelocity.preprocess.preprocess_dataset)
@@
 ```{python}
+#| label: execute-preprocess-data-task
 processed_data = preprocess_data(
@@
 ```{python}
+#| label: print-train-model-configuration
 print_config_tree(pancreas_configuration.training_configuration_2.to_dict(), "train model 2")
@@
 ```{python}
+#| label: print-train-model-docstring
 print_docstring(pyrovelocity.train.train_dataset)
@@
 ```{python}
+#| label: execute-train-model-task
 model_output = train_model(
@@
 ```{python}
+#| label: print-model-output
 print_config_tree(model_output.to_dict(), "model output")
@@
 ```{python}
+#| label: print-postprocess-data-configuration
 print_config_tree(pancreas_configuration.postprocess_configuration.to_dict(), "postprocess data")
@@
 ```{python}
+#| label: print-postprocess-data-docstring
 print_docstring(pyrovelocity.postprocess.postprocess_dataset)
@@
 ```{python}
+#| label: execute-postprocess-data-task
+#| output: false 
+from pyrovelocity.utils import SuppressOutput
+
+with SuppressOutput():
   postprocessing_outputs = postprocess_data(
@@
 ```{python}
+#| label: execute-summarize-data-task
 #| output: false
@@
 ```{python}
+#| label: print-dataset-summary
 print_config_tree(dataset_summary.to_dict(), "dataset summary")
@@
 
-### Vector fields
+### Data set review 
+```{python}
+#| label: import-ipython-display
+#| echo: false
+from IPython.display import Image, display
+```
+
+The maximum values of the unspliced and spliced count data for each gene across cells is shown in @fig-raw-counts.
+
+<!-- ![Maximum spliced and unsliced transcript counts data for each gene across all cells with marginal histograms.](data/processed/pancreas_thresh_histogram.pdf.png){#fig-raw-counts} -->
+```{python}
+#| label: fig-raw-counts
+#| echo: false
+#| fig-cap: "Maximum spliced and unsliced transcript counts data for each gene across all cells with marginal histograms."
+display(Image(filename=f"data/processed/pancreas_thresh_histogram.pdf.png"))
+```
+
+### Cell state transformation vector field estimates
+
+A vector field estimate of cell state transformation probability derived from RNA velocity estimates with cluster annotations is shown in @fig-vector-field.
+
+<!-- ![Vector field](reports/pancreas_model2/vector_field.pdf.png){#fig-vector-field} -->
+```{python}
+#| label: fig-vector-field
+#| echo: false
+#| fig-cap: "Vector field" 
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/vector_field.pdf.png"))
+```
+
+#### Vector field uncertainty estimates
 
@@
 
+<!-- ![Vector fields with uncertainty estimates](reports/pancreas_model2/fig2_part1_plot.pdf.png){#fig-vector-fields-uncertainty} -->
 ```{python}
@@
 #| echo: false
-#| fig-cap: "Vector fields with uncertainty"
-
-from IPython.display import Image, display
-
+#| fig-cap: "Vector fields with uncertainty estimates."
 display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part1_plot.pdf.png"))
@@
 
-A standard vector field plot with cluster annotations is shown in @fig-vector-field.
+### Gene selection
 
+A *volcano plot* adapted for probabilistic inference of RNA velocity is shown in @fig-volcano. This plot shows the mean absolute error versus shared time correlation for each gene.
+
+<!-- ![RNA velocity volcano plot showing mean absolute error versus shared time correlation for each gene](reports/pancreas_model2/volcano.pdf.png){#fig-volcano} -->
 ```{python}
-#| label: fig-vector-field
+#| label: fig-volcano
 #| echo: false
-#| fig-cap: "Vector field"
+#| fig-cap: "RNA velocity volcano plot showing mean absolute error versus shared time correlation for each gene."
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/volcano.pdf.png"))
+```
 
-display(Image(filename=f"{dataset_summary.data_model_reports.path}/vector_field.pdf.png"))
+#### Splice state phase portraits
+
+@fig-rainbow shows the splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes.
+
+<!-- ![Splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes](reports/pancreas_model2/rainbow.pdf.png){#fig-rainbow} -->
+```{python}
+#| label: fig-rainbow
+#| echo: false
+#| fig-cap: "Splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes."
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/rainbow.pdf.png"))
 ```
+
+#### Summary
+
+A summary of the gene selection results is shown in @fig-gene-selection-summary.
+
+<!-- ![Summary of gene selection results](reports/pancreas_model2/fig2_part2_plot.pdf.png){#fig-gene-selection-summary} -->
+```{python}
+#| label: fig-gene-selection-summary
+#| echo: false
+#| fig-cap: "Summary of gene selection results."
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part2_plot.pdf.png"))
+```
+
+### Uncertainty estimation
+
+#### Parameter estimates
+
+Estimates for parameter uncertainties for a subset of genes are shown in @fig-parameter-uncertainty.
+
+<!-- ![Centered parameter uncertainties for a subset of genes](reports/pancreas_model2/parameter_uncertainties.pdf.png){#fig-parameter-uncertainty} -->
+```{python}
+#| label: fig-parameter-uncertainty
+#| echo: false
+#| fig-cap: "Centered parameter uncertainties for a subset of genes."
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/parameter_uncertainties.pdf.png"))
+```
+
+<!-- 
+TODO: save png copy of shared time plot
+#### Shared time
+
+@fig-shared-time-uncertainty shows the uncertainty in the shared time estimates for each gene.
+
+![Uncertainty in the shared time estimates for each gene](reports/pancreas_model2/shared_time.pdf){#fig-shared-time-uncertainty} 
+-->

### 2024-03-21 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
fix(pypi_template): add exploration section
- temporarily install from github
- use internal help
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+### Overview
+
 Installation should take less than **5 minutes**. 
 It involves checking if the notebook is running in colab, in which case it is certain that you will need to install or reinstall pyrovelocity.
-Otherwise, it is assumed that the user will have installed pyrovelocity in the kernel environment.
+Otherwise, it is assumed that the user will have installed pyrovelocity in the same environment in which the jupyter kernel is running.
 
@@
 
-### Install pyrovelocity
+### Install library
 
@@
         print("Installing pyrovelocity...")
-        install_package("pyrovelocity==0.2.0b12")
+        install_package("pyrovelocity[workflows] @ git+https://github.com/pinellolab/pyrovelocity.git@530-nbwf")
         try:
@@
 import pyrovelocity
+if IN_COLAB:
   print(pyrovelocity.__file__)
@@
 import pyro
+if IN_COLAB:
   print(pyro.__file__)
@@
 ```{python}
-#| label: print-pyrovelocity-docstring
-help(pyrovelocity) # ?pyrovelocity # to open in side panel tab for reference
+#| label: print-pyrovelocity-docstring-1
+pyrovelocity.help.internal_help(pyrovelocity) # ?pyrovelocity # to open in side panel tab for reference
 ```
@@
 #| label: import-ipython-display
-#| echo: false
+#| echo: true
 from IPython.display import Image, display
@@
 
-<!-- ![Maximum spliced and unsliced transcript counts data for each gene across all cells with marginal histograms.](data/processed/pancreas_thresh_histogram.pdf.png){#fig-raw-counts} -->
 ```{python}
 #| label: fig-raw-counts
-#| echo: false
+#| echo: true
 #| fig-cap: "Maximum spliced and unsliced transcript counts data for each gene across all cells with marginal histograms."
+
 display(Image(filename=f"data/processed/pancreas_thresh_histogram.pdf.png"))
@@
 
-<!-- ![Vector field](reports/pancreas_model2/vector_field.pdf.png){#fig-vector-field} -->
 ```{python}
 #| label: fig-vector-field
-#| echo: false
+#| echo: true
 #| fig-cap: "Vector field"
+
 display(Image(filename=f"{dataset_summary.data_model_reports.path}/vector_field.pdf.png"))
@@
 
-<!-- ![Vector fields with uncertainty estimates](reports/pancreas_model2/fig2_part1_plot.pdf.png){#fig-vector-fields-uncertainty} -->
 ```{python}
 #| label: fig-vector-fields-uncertainty
-#| echo: false
+#| echo: true
 #| fig-cap: "Vector fields with uncertainty estimates."
+
 display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part1_plot.pdf.png"))
@@
 
-<!-- ![RNA velocity volcano plot showing mean absolute error versus shared time correlation for each gene](reports/pancreas_model2/volcano.pdf.png){#fig-volcano} -->
 ```{python}
 #| label: fig-volcano
-#| echo: false
+#| echo: true
 #| fig-cap: "RNA velocity volcano plot showing mean absolute error versus shared time correlation for each gene."
+
 display(Image(filename=f"{dataset_summary.data_model_reports.path}/volcano.pdf.png"))
@@
 
-<!-- ![Splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes](reports/pancreas_model2/rainbow.pdf.png){#fig-rainbow} -->
 ```{python}
 #| label: fig-rainbow
-#| echo: false
+#| echo: true
 #| fig-cap: "Splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes."
+
 display(Image(filename=f"{dataset_summary.data_model_reports.path}/rainbow.pdf.png"))
@@
 
-<!-- ![Summary of gene selection results](reports/pancreas_model2/fig2_part2_plot.pdf.png){#fig-gene-selection-summary} -->
 ```{python}
 #| label: fig-gene-selection-summary
-#| echo: false
+#| echo: true
 #| fig-cap: "Summary of gene selection results."
+
 display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part2_plot.pdf.png"))
@@
 
-<!-- ![Centered parameter uncertainties for a subset of genes](reports/pancreas_model2/parameter_uncertainties.pdf.png){#fig-parameter-uncertainty} -->
 ```{python}
 #| label: fig-parameter-uncertainty
-#| echo: false
+#| echo: true
 #| fig-cap: "Centered parameter uncertainties for a subset of genes."
+
 display(Image(filename=f"{dataset_summary.data_model_reports.path}/parameter_uncertainties.pdf.png"))
@@
 
-<!-- 
-TODO: save png copy of shared time plot
 #### Shared time
 
-@fig-shared-time-uncertainty shows the uncertainty in the shared time estimates for each gene.
+@fig-shared-time-uncertainty shows the distribution of coefficients of variation together with the mean and standard deviation (uncertainty) of the shared time for each cell.
+A kernel density estimate in the shared time uncertainty plot highlights the 90th percentile of the shared time standard deviation.
 
-![Uncertainty in the shared time estimates for each gene](reports/pancreas_model2/shared_time.pdf){#fig-shared-time-uncertainty} 
--->
+```{python}
+#| label: fig-shared-time-uncertainty
+#| echo: true
+#| fig-cap: "Distribution of shared time coefficients of variation together with standard deviation (uncertainty) and mean per cell."
+
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/shared_time.pdf.png"))
+```
+
+## Experimentation
+
+After having reviewed results that are currently computed systematically, we can also load the data and experiment.
+
+Here we will load the data and posterior samples, perform the analysis used to generate the volcano plot for gene selection, and then regenerate the plot of phase portraits, temporal evolution, and shared time correlation.
+It would be straightforward, and perhaps preferable in some cases, to extract all the relevant parameters from the configuration dataclass object we referenced in each section above, but we will intentionally hard-code values to be concrete.
+Of course, feel free to generalize these as might suit your use case.
+
+To load the posterior samples and postprocessed data
+
+```{python}
+#| label: load-posterior-samples
+import pyrovelocity as pv
+import scanpy as sc
+
+logger = pv.logging.configure_logging("pyrovelocity.user_notebook")
+
+posterior_samples = pv.io.CompressedPickle.load("models/pancreas_model2/pyrovelocity.pkl.zst")
+adata = sc.read("models/pancreas_model2/postprocessed.h5ad")
+```
+
+to generate the marker genes
+
+```{python}
+#| label: generate-marker-genes
+volcano_data = posterior_samples["gene_ranking"] 
+putative_marker_genes = pv.analyze.pareto_frontier_genes(volcano_data=volcano_data, num_genes=6)
+```
+
+In @fig-manually-generated-rainbowplot we show the manually generated marker gene phase portraits, temporal evolution, and shared time correlation.
+
+```{python}
+#| label: fig-manually-generated-rainbowplot
+#| echo: true
+#| fig-cap: "Manually generated rainbow plot." 
+with SuppressOutput():
+  pv.plots.rainbowplot(
+      volcano_data=volcano_data,
+      adata=adata,
+      posterior_samples=posterior_samples,
+      genes=putative_marker_genes,
+      data=["st", "ut"],
+      basis="umap",
+      cell_state="clusters",
+  )
+```
+
+The best place to review the approach to generating the above plots is in [the source code for the `pyrovelocity.summarize` module](https://github.com/pinellolab/pyrovelocity/tree/beta/src/pyrovelocity/summarize.py) or in the [docs](https://github.io/pinellolab/pyrovelocity).
+Additional links are shown in the help below.
+
+```{python}
+#| label: print-pyrovelocity-docstring-2
+pv.help.internal_help(pv)
+```

### 2024-03-21 src/pyrovelocity/summarize.py
fix(summarize): update shared time plot font Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -2,6 +2,7 @@ import os
 from pathlib import Path
 from typing import Tuple
 
+import matplotlib
 import matplotlib.pyplot as plt
 import numpy as np
 import scvelo as scv
@@ -285,6 +286,7 @@ def summarize_dataset(
         fig, ax = plt.subplots(1, 3)
         fig.set_size_inches(10, 3)
         ax = ax.flatten()
+        matplotlib.rcParams.update({"font.size": 7})
         ax[0].hist(cell_time_std / cell_time_mean, bins=100)
         ax[0].set_title("histogram of shared time CoV")
         ax_st = scv.pl.scatter(

### 2024-03-21 src/pyrovelocity/workflows/main_workflow.py
feat(workflows): separate summarize cache version Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -64,6 +64,7 @@ __all__ = [
 logger = configure_logging(__name__)
 
 CACHE_VERSION = "0.2.0b12"
+SUMMARIZE_CACHE_VERSION = "0.2.0b13.dev0"
 PYROVELOCITY_CACHE_FLAG = str_to_bool(
     os.getenv("PYROVELOCITY_CACHE_FLAG", "True")
 )
@@ -214,7 +215,7 @@ def postprocess_data(
 
 @task(
     cache=PYROVELOCITY_CACHE_FLAG,
-    cache_version=CACHE_VERSION,
+    cache_version=SUMMARIZE_CACHE_VERSION,
     retries=3,
     interruptible=False,
     timeout=timedelta(minutes=120),
@@ -259,7 +260,7 @@ def summarize_data(
 
 @task(
     cache=PYROVELOCITY_CACHE_FLAG,
-    cache_version=CACHE_VERSION,
+    cache_version=SUMMARIZE_CACHE_VERSION,
     retries=3,
     interruptible=True,
     timeout=timedelta(minutes=20),

### 2024-03-21 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
fix(pypi_template): add blank header cells Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
fix(pypi_template): sync with library updates Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+```{python}
+#| label: blank-cell-01
+#| code-fold: true
+#| output: false
+print("This cell is intentionally left blank.") 
+```
+
 ## Setup environment
+
+```{python}
+#| label: blank-cell-02
+#| code-fold: true
+#| output: false
+print("This cell is intentionally left blank.") 
+```
 
@@
 
-Installation should take less than **5 minutes**. 
-It involves checking if the notebook is running in colab, in which case it is certain that you will need to install or reinstall pyrovelocity.
+Installation should take less than **10 minutes**. If you are running in [Google Colab][colab-more-memory], then even when running with the `PYROVELOCITY_TESTING_FLAG` set to `True`,
+if you have [access to *High-RAM* or *T4 GPU* runtimes][colab-more-memory] this can significantly speed up even this testing of the notebook.
+For a complete analysis with `PYROVELOCITY_TESTING_FLAG` disabled, at least *High-RAM* or *T4 GPU* runtimes are essentially required to execute in a reasonable
+amount of time.
+
+This section involves checking if the notebook is running in colab, in which case it is certain that you will need to install or reinstall pyrovelocity.
 Otherwise, it is assumed that the user will have installed pyrovelocity in the same environment in which the jupyter kernel is running.
@@
 If you need to edit the **version number**, please see the setting in the `setup_pyrovelocity` function below.
+
+[colab-more-memory]: https://colab.research.google.com/notebooks/pro.ipynb#scrollTo=65MSuHKqNeBZ
 
@@
     setup_pyrovelocity()
+
+    from google.colab import output
+    output.enable_custom_widget_manager()
 else:
@@
 
-This is the same for the pyro package in case there was an issue with pyrovelocity install that did not affect another package.
+Please refer to the [docs](https://pinellolab.github.io/pyrovelocity) for tutorials and usage information.
+
+In case there is an issue with `pyrovelocity` installation, the above output should be analogous to that shown below for the pyro package.
 
@@
 
-Please refer to the [docs](https://pinellolab.github.io/pyrovelocity) for tutorials and usage information.
-
 ## Analysis
@@
 #| label: print-pyrovelocity-docstring-1
-pyrovelocity.help.internal_help(pyrovelocity) # ?pyrovelocity # to open in side panel tab for reference
+pyrovelocity.utils.internal_help(pyrovelocity) # ?pyrovelocity # to open in side panel tab for reference
 ```
@@
 #| label: import-notebook-dependencies
-from importlib import reload
-
 import yaml
-from IPython.display import Image
-# from IPython.display import display
 
@@
 import pyrovelocity.workflows.main_workflow
-reload(pyrovelocity.utils)
-reload(pyrovelocity.workflows.main_workflow)
-
 from pyrovelocity.utils import print_config_tree
@@
 ```
+
+::: {#nte-reload-modules .callout-note collapse=true title="Reload modules when installing in editable mode"}
+If you install the package in editable mode with `pip install -e` for development purposes, you may find it helpful to use
+
+```python
+from importlib import reload
+
+reload(pyrovelocity.utils)
+reload(pyrovelocity.workflows.main_workflow)
+```
+
+We leave this out of this notebook since it is intended as a usage example.
+:::
 
@@
 
-::: {.callout-note collapse=true title="Local caching"}
+::: {#nte-local-caching .callout-note collapse=true title="Local caching"}
 During local execution, all of the functions imported from the `pyrovelocity.workflows.main_workflow` module will cache their results.
@@
 
+::: {#nte-suppressoutput .callout-note collapse=true title="Ignore or remove SuppressOutput"}
+You can safely remove and dedent code executed in the context of `SuppressOutput` in a notebook execution environment.
+
+The `SuppressOutput` context manager is used to suppress some of the output of the function calls placed in its context.
+This is purely used as a work-around for compatibility with multiple output formats such as html and pdf and it will be removed in the future.
+It is not necessary for the execution of the `postprocess_data` function in general or for execution in a notebook in particular.
+:::
+
 ```{python}
@@
 #| output: false
+from IPython.utils import io
+
+with io.capture_output() as captured:
   dataset_summary = summarize_data(
@@
 ```
+
+::: {#nte-find-figures .callout-note collapse=true title="Path to figure files generated by summarize_data"}
+The call to `summarize_data` generates a large number of plots which are saved to the `dataset_summary.data_model_reports.path`.
+This currently resolves to a subfolder of `./reports`.
+The name of the subfolder joins the data set name with the name of the model type.
+We capture the plot outputs to prevent them from generating a very long cell output.
+We review several of the plots in @sec-data-set-review.
+The remaining plots are available in the `dataset_summary.data_model_reports.path` folder.
+If you cannot locate it in the file browser, you can execute
+
+```python
+print(dataset_summary.data_model_reports.path)
+```
+
+after the cell containing the call to summarize_data completes to verify the path where the plots are located.
+:::
 
@@
 
-### Data set review 
+### Data set review {#sec-data-set-review}
 ```{python}
@@
 
-display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part1_plot.pdf.png"))
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/vector_field_summary_plot.pdf.png"))
 ```
@@
 
-display(Image(filename=f"{dataset_summary.data_model_reports.path}/rainbow.pdf.png"))
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/gene_selection_rainbow_plot.pdf.png"))
 ```
@@
 
-display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part2_plot.pdf.png"))
+display(Image(filename=f"{dataset_summary.data_model_reports.path}/gene_selection_summary_plot.pdf.png"))
 ```
@@
 #| echo: true
-#| fig-cap: "Manually generated rainbow plot." 
+#| fig-cap: "Manually generated phase portrait, temporal dynamics, and shared time correlation rainbow plot." 
 with SuppressOutput():
@@
 
+See @nte-suppressoutput if you need to know what `SuppressOutput` is doing.
+
 The best place to review the approach to generating the above plots is in [the source code for the `pyrovelocity.summarize` module](https://github.com/pinellolab/pyrovelocity/tree/beta/src/pyrovelocity/summarize.py) or in the [docs](https://github.io/pinellolab/pyrovelocity).
@@
 #| label: print-pyrovelocity-docstring-2
-pv.help.internal_help(pv)
+pv.utils.internal_help(pv)
 ```

### 2024-03-21 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
fix(pypi_template): remove unnecessary output capture Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 #| output: false
-from IPython.utils import io
-
-with io.capture_output() as captured:
 dataset_summary = summarize_data(

### 2024-03-21 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
feat(pypi_template): include training and summary for two distinct model classes Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 The train model task is executed with the `train_model` function.
-The `train_model` task function takes the `FlyteFile` object, here named `processed_data` above, from `preprocess_data` task function and one of the `PyrovelocityTrainInterface` objects such as is found in the `training_configuration_2` attribute of the `pancreas_configuration` object as arguments.
+The `train_model` task function takes the `FlyteFile` object, here named `processed_data` above, from `preprocess_data` task function and one of the `PyrovelocityTrainInterface` objects such as is found in the `training_configuration_1` attribute of the `pancreas_configuration` object as arguments.
 
@@
 #| label: print-train-model-configuration
-print_config_tree(pancreas_configuration.training_configuration_2.to_dict(), "train model 2")
+print_config_tree(pancreas_configuration.training_configuration_1.to_dict(), "train model 1")
 ```
@@
   processed_data=processed_data,
-  train_model_configuration=pancreas_configuration.training_configuration_2,
+  train_model_configuration=pancreas_configuration.training_configuration_1,
 )
@@
 
-posterior_samples = pv.io.CompressedPickle.load("models/pancreas_model2/pyrovelocity.pkl.zst")
-adata = sc.read("models/pancreas_model2/postprocessed.h5ad")
+posterior_samples = pv.io.CompressedPickle.load("models/pancreas_model1/pyrovelocity.pkl.zst")
+adata = sc.read("models/pancreas_model1/postprocessed.h5ad")
 ```
@@
 ```
+
+## Training and postprocessing for an alternative model
+
+In this section we will train and show the results of an alternative model.
+This time we will proceed quickly through the analysis with less discussion of the various configuration settings and associated task functions.
+
+### Train model
+
+We will now use the `training_configuration_2` attribute of the `pancreas_configuration` object to train the model.
+
+```{python}
+#| label: print-train-model-configuration-2
+print_config_tree(pancreas_configuration.training_configuration_2.to_dict(), "train model 2")
+```
+
+Re-execute the `train_model` task function with this new configuration.
+
+```{python}
+#| label: execute-train-model-task-2
+model_output_2 = train_model(
+  processed_data=processed_data,
+  train_model_configuration=pancreas_configuration.training_configuration_2,
+)
+```
+
+This produces the `model_output_2`.
+
+```{python}
+#| label: print-model-output-2
+print_config_tree(model_output_2.to_dict(), "model output 2")
+```
+
+### Postprocess data
+
+The postprocess data task is executed with the `postprocess_data` function.
+
+```{python}
+#| label: execute-postprocess-data-task-2
+#| output: false 
+from pyrovelocity.utils import SuppressOutput
+
+with SuppressOutput():
+  postprocessing_outputs_2 = postprocess_data(
+    preprocess_data_args=pancreas_configuration.preprocess_data,
+    training_outputs=model_output_2,
+    postprocess_configuration=pancreas_configuration.postprocess_configuration,
+  )
+```
+
+## Alternative model results
+
+Now we have what is required to construct a summary of the results.
+
+```{python}
+#| label: execute-summarize-data-task-2
+#| output: false
+dataset_summary_2 = summarize_data(
+  preprocess_data_args=pancreas_configuration.preprocess_data,
+  postprocessing_outputs=postprocessing_outputs_2,
+  training_outputs=model_output_2,
+)
+```
+
+The dataset summary contains
+
+```{python}
+#| label: print-dataset-summary-2
+print_config_tree(dataset_summary_2.to_dict(), "dataset summary")
+```
+
+### Cell state transformation vector field estimates
+
+A vector field estimate of cell state transformation probability derived from RNA velocity estimates with cluster annotations is shown in @fig-vector-field-2.
+
+```{python}
+#| label: fig-vector-field-2
+#| echo: true
+#| fig-cap: "Vector field"
+
+display(Image(filename=f"{dataset_summary_2.data_model_reports.path}/vector_field.pdf.png"))
+```
+
+#### Vector field uncertainty estimates
+
+@fig-vector-fields-uncertainty-2 shows a low-dimensional representation of the RNA velocity vector fields with uncertainty estimates for shared time and angle.
+The latter quantity is estimated in PCA space.
+The base magnitude uncertainty is also shown but may not be interpretable.
+
+```{python}
+#| label: fig-vector-fields-uncertainty-2
+#| echo: true
+#| fig-cap: "Vector fields with uncertainty estimates."
+
+display(Image(filename=f"{dataset_summary_2.data_model_reports.path}/vector_field_summary_plot.pdf.png"))
+```
+
+### Gene selection
+
+A *volcano plot* adapted for probabilistic inference of RNA velocity is shown in @fig-volcano-2. This plot shows the mean absolute error versus shared time correlation for each gene.
+
+```{python}
+#| label: fig-volcano-2
+#| echo: true
+#| fig-cap: "RNA velocity volcano plot showing mean absolute error versus shared time correlation for each gene."
+
+display(Image(filename=f"{dataset_summary_2.data_model_reports.path}/volcano.pdf.png"))
+```
+
+#### Splice state phase portraits
+
+@fig-rainbow-2 shows the splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes.
+
+```{python}
+#| label: fig-rainbow-2
+#| echo: true
+#| fig-cap: "Splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes."
+
+display(Image(filename=f"{dataset_summary_2.data_model_reports.path}/gene_selection_rainbow_plot.pdf.png"))
+```
+
+#### Summary
+
+A summary of the gene selection results is shown in @fig-gene-selection-summary-2.
+
+```{python}
+#| label: fig-gene-selection-summary-2
+#| echo: true
+#| fig-cap: "Summary of gene selection results."
+
+display(Image(filename=f"{dataset_summary_2.data_model_reports.path}/gene_selection_summary_plot.pdf.png"))
+```
+
+### Uncertainty estimation
+
+#### Parameter estimates
+
+Estimates for parameter uncertainties for a subset of genes are shown in @fig-parameter-uncertainty-2.
+
+```{python}
+#| label: fig-parameter-uncertainty-2
+#| echo: true
+#| fig-cap: "Centered parameter uncertainties for a subset of genes."
+
+display(Image(filename=f"{dataset_summary_2.data_model_reports.path}/parameter_uncertainties.pdf.png"))
+```
+
+#### Shared time
+
+@fig-shared-time-uncertainty-2 shows the distribution of coefficients of variation together with the mean and standard deviation (uncertainty) of the shared time for each cell.
+A kernel density estimate in the shared time uncertainty plot highlights the 90th percentile of the shared time standard deviation.
+
+```{python}
+#| label: fig-shared-time-uncertainty-2
+#| echo: true
+#| fig-cap: "Distribution of shared time coefficients of variation together with standard deviation (uncertainty) and mean per cell."
+
+display(Image(filename=f"{dataset_summary_2.data_model_reports.path}/shared_time.pdf.png"))
+```
+
+## Summary
+
+In this notebook we have demonstrated how to install pyrovelocity and execute a comprehensive workflow to perform a probabilistic analysis of RNA velocity in a single-cell transcriptomics data set in a Jupyter notebook.
+We default to data downsampled in this case by about a factor of 10 to allow for swift execution to illustrate the analysis workflow.
+In particular, we have shown how to download data, preprocess it, train a model, postprocess the results, and summarize the data with a compendium of plots.
+We have also shown how to experiment with the data and posterior samples to generate new plots, and how to train, postprocess, and summarize the results of an alternative model.
+This information should be sufficient to see how to get started applying pyrovelocity to your own data.
+We will discuss approaches to model comparison, criticism, and selection in a separate notebook.
+Please do not hesitate to reach out in [the discussions board on github](https://github.com/pinellolab/pyrovelocity/discussions) if you run into problems using the library.

### 2024-03-22 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
fix(pypi_template): collapse installation instructions in callout note Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-Installation should take less than **10 minutes**. If you are running in [Google Colab][colab-more-memory], then even when running with the `PYROVELOCITY_TESTING_FLAG` set to `True`,
-if you have [access to *High-RAM* or *T4 GPU* runtimes][colab-more-memory] this can significantly speed up even this testing of the notebook.
+Installation should take less than **10 minutes**.
+
+::: {#nte-installation .callout-note collapse=true title="Installing pyrovelocity"}
+If you are running on a local or remote Jupyter server outside of Google Colab, we assume you have already installed pyrovelocity in the same environment in which the Jupyter kernel is running.
+If you have not, ensure you are in a virtual environment with python 3.10 active (we recommend [pyenv][pyenv] or [miniforge][miniforge]), and then you can install the development version from github with
+
+```bash
+pip install pyrovelocity[workflows] @ git+https://github.com/pinellolab/pyrovelocity.git@beta
+```
+
+or a recent release from PyPI
+
+```bash
+pip install pyrovelocity[workflows]==0.2.0b12
+```
+
+We also support installation from the anaconda conda-forge channel
+
+```bash
+mamba install \
+  -c pytorch -c nvidia -c conda-forge \
+  pytorch::pytorch conda-forge/label/pyrovelocity_dev::pyrovelocity
+```
+
+We do not use this option here because we find in Google Colab installation via anaconda takes about twice as long as from PyPI.
+:::
+
+With respect to resource requirements, this notebook should be able to be run on any modern laptop.
+By default, the data are downsampled to 300 observations and training epochs truncated to 300 by setting the `PYROVELOCITY_TESTING_FLAG` to `True`.
+If you are running in [Google Colab][colab-more-memory], then even in this testing mode,
+if you have [access to *High-RAM* or *T4 GPU* runtimes][colab-more-memory] this can significantly speed up library installation and execution.
 For a complete analysis with `PYROVELOCITY_TESTING_FLAG` disabled, at least *High-RAM* or *T4 GPU* runtimes are essentially required to execute in a reasonable
-amount of time.
+amount of time. In this case, if you do not have a subscription to Google Colab Pro, running on a laptop should be sufficient for the data set we analyze here.
 
-This section involves checking if the notebook is running in colab, in which case it is certain that you will need to install or reinstall pyrovelocity.
-Otherwise, it is assumed that the user will have installed pyrovelocity in the same environment in which the jupyter kernel is running.
-
+The following section involves checking if the notebook is running in colab, in which case it is certain that you will need to install or reinstall pyrovelocity.
 The simplest way to complete it is to run the `Setup environment` section, wait for the kernel to restart, and then run the same section again:
@@
 
-If you need to edit the **version number**, please see the setting in the `setup_pyrovelocity` function below.
+If you need to edit the **version number**, please see the argument passed to `install_package` in the `setup_pyrovelocity` function below.
 
 [colab-more-memory]: https://colab.research.google.com/notebooks/pro.ipynb#scrollTo=65MSuHKqNeBZ
+[pyenv]: https://github.com/pyenv/pyenv
+[miniforge]: https://github.com/conda-forge/miniforge?tab=readme-ov-file#whats-the-difference-between-mambaforge-and-miniforge
 
@@
 #| label: install-pyrovelocity-colab
+#| output: false
 #| code-fold: true
@@
 #| label: check-pyrovelocity-installation
-#| output: false
+#| output: true
 
@@
 
-## Analysis
+You can find links to pyrovelocity documentation and source code in the help below.
 
@@
 
-Before we start, we will adjust a few environment variables to ensure our first execution occurs in a lightweight test mode with a subset of observations, variables, training epochs, and posterior samples.
-We will also set a few log levels.
-Feel free to adjust these to `ERROR`, `DEBUG`, etc as needed.
+## Analysis
+
+In this section we demonstrate how to run a comprehensive analysis from accessing data to plotting results.
 
@@
 
-After an initial review, we can set the environment variable to `False` to run the full analysis.
+Before we start, we set a few environment variables to ensure our first execution occurs in a lightweight test mode with a subset of observations, variables, training epochs, and posterior samples.
+We also set a few log levels.
+Feel free to adjust these to `ERROR`, `DEBUG`, etc as needed.
+
+After an initial review, we can set the `PYROVELOCITY_TESTING_FLAG` environment variable to `False` to run the full analysis.
 
@@
 
-To execute each task requires a single `WorkflowConfiguration` object, which is an instance of a python datclass.
+To execute each task requires a single `WorkflowConfiguration` object, which is an instance of a python dataclass.
 In this notebook we illustrate execution with the standard pancreatic endocrinogenesis data set [@Bastidas-Ponce2019-lf].
@@
 We will reprint the part of that configuration relevant to the execution of each task below.
-The resource requests and limits sections at the bottom will be irrelevant for this example of local execution.
+The resource requests and limits sections will be irrelevant for this example of local execution.
 The resource specifications are utilized during local or remote cluster-based execution of distributed containerized workflows.
@@
 ### Data set review {#sec-data-set-review}
+
 ```{python}
 #| label: import-ipython-display
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-raw-counts
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-vector-field
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-vector-fields-uncertainty
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-volcano
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-rainbow
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-gene-selection-summary
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-parameter-uncertainty
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-shared-time-uncertainty
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-vector-field-2
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-vector-fields-uncertainty-2
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-volcano-2
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-rainbow-2
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-gene-selection-summary-2
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-parameter-uncertainty-2
+#| code-fold: true
 #| echo: true
@@
 #| label: fig-shared-time-uncertainty-2
+#| code-fold: true
 #| echo: true
@@
 
-In this notebook we have demonstrated how to install pyrovelocity and execute a comprehensive workflow to perform a probabilistic analysis of RNA velocity in a single-cell transcriptomics data set in a Jupyter notebook.
+In this notebook we have demonstrated how to install pyrovelocity and execute a comprehensive workflow to perform a probabilistic analysis of RNA velocity for a single-cell transcriptomics data set in a Jupyter notebook.
 We default to data downsampled in this case by about a factor of 10 to allow for swift execution to illustrate the analysis workflow.
 In particular, we have shown how to download data, preprocess it, train a model, postprocess the results, and summarize the data with a compendium of plots.
-We have also shown how to experiment with the data and posterior samples to generate new plots, and how to train, postprocess, and summarize the results of an alternative model.
+We have also shown how to experiment with the data and posterior samples to generate new plots, and how to train, postprocess, and summarize the results for an alternative model.
 This information should be sufficient to see how to get started applying pyrovelocity to your own data.
 We will discuss approaches to model comparison, criticism, and selection in a separate notebook.
-Please do not hesitate to reach out in [the discussions board on github](https://github.com/pinellolab/pyrovelocity/discussions) if you run into problems using the library.
+Please do not hesitate to reach out in [the discussions board on github](https://github.com/pinellolab/pyrovelocity/discussions) if you run into any problems using the library.

### 2024-03-24 nbs/tutorials/introduction.qmd
fix(nbs): move introduction to tutorials Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

# Introduction

## Features

- Probabilistic modeling of RNA velocity
- Direct modeling of raw spliced and unspliced read count
- Multiple uncertainty diagnostics analysis and visualizations
- Synchronized cell time estimation across genes
- Multivariate denoised gene expression and velocity prediction

<!-- ![Velocity workflow comparison](docs/source/readme_figure1.png) -->
<div style="display: flex; flex-direction: column; align-items: center;">
<img
    src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/master/docs/source/readme_figure1.png"
    alt="Velocity workflow comparison"
    style="width: 700px; max-width: 100%; height: auto;"
    role="img">
</div>

## Installation

### Quick start

See `make | grep codespace`, run `make codespace_create code`, or

[![codespaces](https://github.com/codespaces/badge.svg)](https://codespaces.new/?hide_repo_select=true&ref=491-workflows&repo=533596575&skip_quickstart=true&machine=premiumLinux&devcontainer_path=.devcontainer%2Fdevcontainer.json)

This provides a zero install demonstration of the supported development environment. See `make -n setup_dev` to preview what is required to setup a local copy of this environment.

### Development

We currently support installation and usage in a linux 64-bit development environment with access to a GPU.
An [IaC](https://en.wikipedia.org/wiki/Infrastructure_as_code) setup that works with [GCP](https://registry.terraform.io/providers/hashicorp/google/latest/docs) is documented in [reproducibility/environment/README.md](reproducibility/environment/README.md).
Before proceeding to setup a minimal development environment, please fork this repository and clone a copy of your fork to your development machine. Unless otherwise mentioned, all commands assume your current working directory is the root of your local copy of your fork of this repository.

Please install mambaforge according to the instructions provided in [conda-forge/miniforge](https://github.com/conda-forge/miniforge#install).

You can then create a development environment with

```bash
mamba env create [--prefix /path_to_conda_environment] -f conda/environment-gpu.yml
```

This step takes about 10 minutes depending on network speed.

Make sure you are able to successfully activate the installed environment

```bash
conda activate pyrovelocity-gpu
which python
```

by checking that the output of which python following activation refers to the python binary in the correct conda environment.

You can then install a development copy with

```
pip install --no-deps -e .
```

If this is successful, you will be able to

```python
import pyrovelocity
```

from a python interpreter.

### User

Please see the [documentation](https://pyrovelocity.readthedocs.io/en/latest/installation.html).

## Quick start

After the installation, let\'s look at your dataset to see how
Pyro-Velocity can help understand cell dynamics.

Starting from raw sequencing FASTQ files, obtained for example with
SMART-seq, 10X genomics, inDrop or other similar single-cell assays, you
can preprocess the data to generate spliced and unspliced gene count
tables in h5ad file (or loom file using cellranger+velocyto or the
kallisto pipeline.

Starting from these count tables we show below a minimal step-by-step
workflow to illustrate the main features of Pyro-Velocity in a Jupyter
Notebook:

Step 0. Even though _pyrovelocity_ is a cluster-free method to evaluate uncertainty
of cell fate, it will dependend on 2 dimensional embedding results for evaluation of
uncertainty and generate visualization, we would suggest run your new datasets using
[scanpy tutorial](https://scanpy-tutorials.readthedocs.io/en/latest/pbmc3k.html).

Step 1. Load your data, load your data(e.g. _local_file.h5ad_) with
scvelo by using:

```python
import scvelo as scv
adata = scv.read("local_file.h5ad")
```

Step 2. Minimally preprocess your _adata_ object:

```python
adata.layers['raw_spliced']   = adata.layers['spliced']
adata.layers['raw_unspliced'] = adata.layers['unspliced']
adata.obs['u_lib_size_raw'] = adata.layers['raw_unspliced'].toarray().sum(-1)
adata.obs['s_lib_size_raw'] = adata.layers['raw_spliced'].toarray().sum(-1)
scv.pp.filter_and_normalize(adata, min_shared_counts=30, n_top_genes=2000)
scv.pp.moments(adata, n_pcs=30, n_neighbors=30)
```

Step 3. Train the Pyro-Velocity model:

```python
from pyrovelocity.api import train_model
from pyrovelocity.plot import vector_field_uncertainty
num_epochs = 1000 # large data
# num_epochs = 4000 # small data
# Model 1
adata_model_pos = train_model(adata,
                               max_epochs=num_epochs, svi_train=True, log_every=100,
                               patient_init=45,
                               batch_size=4000, use_gpu=0, cell_state='state_info',
                               include_prior=True,
                               offset=False,
                               library_size=True,
                               patient_improve=1e-3,
                               model_type='auto',
                               guide_type='auto_t0_constraint',
                               train_size=1.0)
# Or Model 2
adata_model_pos = train_model(adata,
                               max_epochs=num_epochs, svi_train=True, log_every=100,
                               patient_init=45,
                               batch_size=4000, use_gpu=0, cell_state='state_info',
                               include_prior=True,
                               offset=True,
                               library_size=True,
                               patient_improve=1e-3,
                               model_type='auto',
                               guide_type='auto',
                               train_size=1.0)

# adata_model_pos is a returned list in which 0th element is the trained model,
# the 1st element is the posterior samples of all random variables

trained_model = adata_model_pos[0]
posterior_samples = adata_model_pos[1]
v_map_all, embeds_radian, fdri = vector_field_uncertainty(
    adata,
    posterior_samples=posterior_samples,
    basis="umap"
)

save_res = True
if save_res:
    trained_model.save('saved_model', overwrite=True)
    result_dict = {"adata_model_pos": posterior_samples,
                   "v_map_all": v_map_all,
                   "embeds_radian": embeds_radian, "fdri": fdri} #, "embed_mean": embed_mean}
    import pickle
    with open("posterior_samples.pkl", "wb") as f:
         pickle.dump(result_dict, f)
```

Step 4: Generate Pyro-Velocity's vector field and shared time plots
with uncertainty estimation.

```python
from pyrovelocity.plot import plot_state_uncertainty
from pyrovelocity.plot import plot_posterior_time, plot_gene_ranking,\
      vector_field_uncertainty, plot_vector_field_uncertain,\
      plot_mean_vector_field, project_grid_points,rainbowplot,denoised_umap,\
      us_rainbowplot, plot_arrow_examples, set_colorbar

import numpy as np
import matplotlib.pyplot as plt
import seaborn as sns

embedding = 'umap' # change to umap or tsne based on your embedding method

# This generates the posterior samples of all vector fields
# and statistical testing results from Rayleigh test
v_map_all, embeds_radian, fdri = vector_field_uncertainty(adata, posterior_samples,
                                                          basis=embedding, denoised=False, n_jobs=30)
fig, ax = plt.subplots()
# This returns the posterior mean of the vector field
embed_mean = plot_mean_vector_field(posterior_samples, adata, ax=ax, n_jobs=30, basis=embedding)
# This plot single-cell level vector field uncertainty
# and averaged cell vector field uncertainty on the grid points
# based on angular standard deviation
fig, ax = plt.subplots(1, 2)
fig.set_size_inches(11.5, 5)
plot_vector_field_uncertain(adata, embed_mean, embeds_radian,
                            ax=ax,
                            fig=fig, cbar=False, basis=embedding, scale=None)

# This generates shared time uncertainty plot with contour lines
fig, ax = plt.subplots(1, 3)
fig.set_size_inches(12, 2.8)
adata.obs['shared_time_uncertain'] = posterior_samples['cell_time'].std(0).flatten()
ax_cb = scv.pl.scatter(adata, c='shared_time_uncertain', ax=ax[0], show=False, cmap='inferno', fontsize=7, s=20, colorbar=True, basis=embedding)
select = adata.obs['shared_time_uncertain'] > np.quantile(adata.obs['shared_time_uncertain'], 0.9)
sns.kdeplot(adata.obsm[f'X_{embedding}'][:, 0][select],
            adata.obsm[f'X_{embedding}'][:, 1][select],
            ax=ax[0], levels=3, fill=False)

# This generates vector field uncertainty based on Rayleigh test.
adata.obs.loc[:, 'vector_field_rayleigh_test'] = fdri
im = ax[1].scatter(adata.obsm[f'X_{embedding}'][:, 0],
                   adata.obsm[f'X_{embedding}'][:, 1], s=3, alpha=0.9,
                   c=adata.obs['vector_field_rayleigh_test'], cmap='inferno_r',
                   linewidth=0)
set_colorbar(im, ax[1], labelsize=5, fig=fig, position='right')
select = adata.obs['vector_field_rayleigh_test'] > np.quantile(adata.obs['vector_field_rayleigh_test'], 0.95)
sns.kdeplot(adata.obsm[f'X_{embedding}'][:, 0][select],
            adata.obsm[f'X_{embedding}'][:, 1][select], ax=ax[1], levels=3, fill=False)
ax[1].axis('off')
ax[1].set_title("vector field\nrayleigh test\nfdr<0.05: %s%%" % (round((fdri < 0.05).sum()/fdri.shape[0], 2)*100), fontsize=7)
```

Step 5: Prioritize putative cell fate marker genes based on negative
mean absolute errors and pearson correlation between denoised spliced
expression and posterior mean shared time, and then visualize the top
one with rainbow plots

```python
fig = plt.figure(figsize=(7.07, 4.5))
subfig = fig.subfigures(1, 2, wspace=0.0, hspace=0, width_ratios=[1.6, 4])
ax = fig.subplots(1)
# This generates the selected cell fate markers and output in DataFrame
volcano_data, _ = plot_gene_ranking([posterior_samples], [adata], ax=ax,
                                     time_correlation_with='st', assemble=True)
# This generates the rainbow plots for the selected markers.
_ = rainbowplot(volcano_data, adata, posterior_samples,
                subfig[1], data=['st', 'ut'], num_genes=4)
```

## Illustrative examples of Pyro-Velocity analyses on different single-cell datasets

### Pyro-Velocity applied to a PBMC dataset

See the [data referred to here](https://scvelo.readthedocs.io/en/stable/perspectives/Perspectives/).

This is a scRNA-seq dataset of fully mature peripheral blood mononuclear
cells (PBMC) generated using the 10X genomics kit and containing 65,877
cells with 11 fully differentiated immune cell types. This dataset
doesn\'t contain stem and progenitor cells or other signature of and
undergoing dynamical differentiation, thus no consistent velocity flow
should be detected.

Below we show the main output generated by Pyro-Velocity Model 1
analysis. Pyro-Velocity failed to detect high-confidence trajectories in
the mature blood cell states, consistent with what is known about the
biology underlying these cells.

**Vector field with uncertainty**

<!-- ![PBMC vector field uncertainty](docs/source/readme_figure2.png) -->
<!-- <img src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure2.png" alt="PBMC vector field uncertainty"> -->

<div style="display: flex; flex-direction: column; align-items: center;">
<img
    src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure2.png"
    alt="PBMC vector field uncertainty"
    style="width: 700px; max-width: 100%; height: auto;"
    role="img">
</div>

These 6 plots from left to right show: 1. cell types, 2. stream plot of
Pyro-velocity vector field based on the posterior mean of 30 posterior
samples, 3. single cell vector field examples showing all 30 posterior
samples as vectors for 3 arbitrarily selected cells; 4. single cell
vector field with uncertainty based on angular standard deviation across
30 posterior samples, 5. averaged vector field uncertainty from 4. 6.
Rayleigh test of posterior samples vector field, the title shows the
expected false discovery rate using a 5% threshold.

The full example can be reproduced using the
[PBMC](https://github.com/pinellolab/pyrovelocity/blob/master/docs/source/notebooks/pbmc.ipynb)
Jupyter notebook.

### Pyro-Velocity applied to a pancreas development dataset

See the [data referred to here](https://scvelo.readthedocs.io/en/stable/VelocityBasics/).

Here we apply Pyro-Velocity to a single cell RNA-seq dataset of mouse
pancreas in the E15.5 embryo developmental stage. This dataset was
generated using the 10X genomics kit and contains 3,696 cells with 8
cell types including progenitor cells, intermediate and terminal cell
states.

Below we show the main output generated by Pyro-Velocity Model 1
analysis. Pyro-Velocity was able to define well-known developmental cell
hierarchies identifying cell trajectories originating from ductal
progenitor cells and culminated in the production of mature Alpha, Beta,
Delta, and Epsilon cells.

**Vector field with uncertainty**

<!-- ![Pancreas vector field uncertainty](docs/source/readme_figure3.png) -->
<!-- <img src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure3.png" alt="Pancreas vector field uncertainty"> -->
<div style="display: flex; flex-direction: column; align-items: center;">
<img
    src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure3.png"
    alt="Pancreas vector field uncertainty"
    style="width: 700px; max-width: 100%; height: auto;"
    role="img">
</div>

These 6 plots from left to right are showing the same analyses presented
as in the example above.

**Shared time with uncertainty**

<!-- ![Pancreas shared time uncertainty](docs/source/readme_figure4.png) -->
<!-- <img src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/master/docs/source/readme_figure4.png" alt="Pancreas vector field uncertainty"> -->
<div style="display: flex; flex-direction: column; align-items: center;">
<img
    src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/master/docs/source/readme_figure4.png"
    alt="Pancreas vector field uncertainty"
    style="width: 700px; max-width: 100%; height: auto;"
    role="img">
</div>

The left figure shows the average of 30 posterior samples for the cell
shared time, the title of the figure shows the Spearman\'s correlation
with the Cytotrace score, an orthogonal state-of-the-art method used to
predict cell differentiation based on the number of expressed genes per
cell (Gulati et. al, Science 2020). The right figure shows the standard
deviation across posterior samples of shared time.

**Gene selection and visualization**

To uncover potential cell fate determinant markers genes of the mouse
pancreas, we first select the top 300 genes with the best velocity model
fit (we use negative mean absolute error ), then we rank the filtered
genes using Pearson\'s correlation between denoised spliced expression
and the posterior mean of the recovered shared time across cells.

<!-- ![Pancreas Volcano plot for gene selection](docs/source/readme_figure6.png) -->
<!-- <img src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure6.png" alt="Pancreas Volcano plot for gene selection"> -->
<div style="display: flex; flex-direction: column; align-items: center;">
<img
    src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure6.png"
    alt="Pancreas Volcano plot for gene selection"
    style="width: 700px; max-width: 100%; height: auto;"
    role="img">
</div>

For the selected genes, it is possible to explore in depth their
dynamic, using phase portraits, rainbow plots, and UMAP rendering of
denoised splicing gene expression across cells.

<!-- ![Pancreas vector field uncertainty](docs/source/readme_figure7.png) -->
<!-- <img src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure7.png" alt="Pancreas vector field uncertainty"> -->
<div style="display: flex; flex-direction: column; align-items: center;">
<img
    src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure7.png"
    alt="Pancreas vector field uncertainty"
    style="width: 700px; max-width: 100%; height: auto;"
    role="img">
</div>

The full example can be reproduced using the
[Pancreas](https://github.com/pinellolab/pyrovelocity/blob/master/docs/source/notebooks/pancreas.ipynb)
jupyter notebook.

### Pyro-Velocity applied to the LARRY dataset

See the [data referred to here](https://figshare.com/articles/dataset/larry_invitro_adata_sub_raw_h5ad/20780344).

This last example, present the analysis of a recent scRNA-seq dataset
profiling mouse hematopoises at high resolution thanks to lineage
relationship information captured by the Lineage And RNA RecoverY
(LARRY) system. LARRY leverages unique lentiviral barcodes that enables
to clonally trace cell fates over time (Weinrab et al. Cell 20).

Below we show the main output generated by Pyro-Velocity analysis.

**Vector field with uncertainty**

<!-- ![LARRY vector field uncertainty](docs/source/readme_figure8.png) -->
<!-- <img src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure8.png" alt="LARRY vector field uncertainty"> -->
<div style="display: flex; flex-direction: column; align-items: center;">
<img
    src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure8.png"
    alt="LARRY vector field uncertainty"
    style="width: 700px; max-width: 100%; height: auto;"
    role="img">
</div>

These 5 plots from left to right shows: 1) Cell types, 2) Clone
progression vector field by using centroid of cells belonging to the
same barcode for generating directed connection between consecutive
physical times, 3) single cell vector field with uncertainty based on
angular standard deviation across 30 posterior samples, 4. averaged
vector field uncertainty from 3. 5. Rayleigh test of posterior samples
vector field, the title shows the false discovery rate using threshold
5%.

**Shared time with uncertainty**

To quantitatively assess the quality of the the receovered shared time
we also considered the agreement of our method with Cospar, a
state-of-the-art method specifically designed for predicting fate
potency based on LARRY data.

<!-- ![Pancreas shared time uncertainty](docs/source/readme_figure9.png) -->
<!-- <img src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure9.png" alt="Pancreas shared time uncertainty"> -->
<div style="display: flex; flex-direction: column; align-items: center;">
<img
    src="https://raw.githubusercontent.com/pinellolab/pyrovelocity/main/docs/source/readme_figure9.png"
    alt="Pancreas shared time uncertainty"
    style="width: 700px; max-width: 100%; height: auto;"
    role="img">
</div>

The leftmost figure shows the Cospar fate potency score, the middle
figure shows the average of 30 posterior samples from Pyro-Velocity
shared time per cell, the title of the figure shows the Spearman's
correlation between cell latent shared time and fate potency scores
derived from Cospar, the right figure shows the standard deviation
across posterior samples of shared time.

The full example can be reproduced using the
[LARRY](https://github.com/pinellolab/pyrovelocity/blob/main/docs/source/notebooks/larry.ipynb)
jupyter notebook.

# Troubleshooting

If you are having an issue using pyrovelocity, please feel free to [start a discussion](https://github.com/pinellolab/pyrovelocity/discussions)
or [file an issue](https://github.com/pinellolab/pyrovelocity/issues) containing a [MRE](https://en.wikipedia.org/wiki/Minimal_reproducible_example).

Also see [contributing](./docs/contributing.md).

### 2024-03-24 nbs/index.qmd
fix(nbs): update index Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: "Pyro-Velocity"
+title: Pyro-Velocity
+description: "Probabilistic modeling of RNA velocity"
+format:
+  html:
+    toc: false
+
+about:
+  id: information
+  template: jolla
+  image: _static/logo.png
+  links: 
+    - icon: download
+      href: guides/installation.qmd
+    - icon: book
+      href: tutorials/introduction.qmd
+    - icon: github
+      text: GitHub
+      href: https://github.com/pinellolab/pyrovelocity
 ---
 
-Pyro-Velocity is a Bayesian, generative, and multivariate RNA velocity model to estimate uncertainty in predictions of future cell states from minimal models approximating splicing dynamics.
+::: {#information}
+:::
+
+<div style="margin-left: auto; margin-right: auto; max-width: 560px; overflow-x: auto;">
+
+|         |                                                                                                                                                                        |
+| ------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
+| CI/CD   | [![CID][cid-badge]][cid-link] [![CML][cml-badge]][cml-link] [![pre-commit.ci status][precommit-badge]][precommit-link]                                                 |
+| Docs    | [![Cloudflare Pages][cloudflare-badge]][cloudflare-link] [![GitHub Pages][gh-pages-badge]][gh-pages-link] [![Preprint][preprint-badge]][preprint-link]                 |
+| Package | [![PyPI - Version][pypi-badge]][pypi-link] [![Anaconda badge][anaconda-badge]][anaconda-link] [![Docker image][docker-badge]][docker-link]                             |
+| Meta    | [![codecov][codecov-badge]][codecov-link] [![code style][black-badge]][black-link] [![License - MIT][license-badge]][license-link] [![Tuple][tuple-badge]][tuple-link] |
+
+[cid-badge]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cid.yaml/badge.svg?branch=master
+[cid-link]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cid.yaml
+[cml-badge]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cml.yml/badge.svg
+[cml-link]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cml.yml
+[precommit-badge]: https://results.pre-commit.ci/badge/github/pinellolab/pyrovelocity/main.svg
+[precommit-link]: https://results.pre-commit.ci/latest/github/pinellolab/pyrovelocity/master
+[cloudflare-badge]: https://img.shields.io/badge/Docs-pages-F38020?style=flat&logo=cloudflare&labelColor=black
+[cloudflare-link]: https://docs.pyrovelocity.net
+[gh-pages-badge]: https://img.shields.io/github/deployments/pinellolab/pyrovelocity/github-pages?logo=github&label=Docs
+[gh-pages-link]: https://pinellolab.github.io/pyrovelocity
+[preprint-badge]: https://img.shields.io/badge/doi-10.1101/2022.09.12.507691v2-B31B1B
+[preprint-link]: https://doi.org/10.1101/2022.09.12.507691
+[pypi-badge]: https://img.shields.io/pypi/v/pyrovelocity.svg?logo=pypi&label=PyPI&logoColor=gold
+[pypi-link]: https://pypi.org/project/pyrovelocity/
+[anaconda-badge]: https://anaconda.org/conda-forge/pyrovelocity/badges/version.svg
+[anaconda-link]: https://anaconda.org/conda-forge/pyrovelocity
+[docker-badge]: https://img.shields.io/badge/docker-image-blue?logo=docker
+[docker-link]: https://github.com/pinellolab/pyrovelocity/pkgs/container/pyrovelocity
+[codecov-badge]: https://codecov.io/gh/pinellolab/pyrovelocity/branch/main/graph/badge.svg
+[codecov-link]: https://codecov.io/gh/pinellolab/pyrovelocity
+[black-badge]: https://img.shields.io/badge/code%20style-black-000000.svg
+[black-link]: https://github.com/psf/black
+[license-badge]: https://img.shields.io/badge/license-AGPL%203-purple
+[license-link]: https://spdx.org/licenses/
+[tuple-badge]: https://img.shields.io/badge/Tuple%20❤️%20OSS-5A67D8?logo=tuple
+[tuple-link]: https://tuple.app/github-badge
+
+</div>

### 2024-03-24 nbs/about/contributing.qmd
feat(nbs): copy contributing Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

# Contributing

Contributions are welcome, greatly appreciated, and credit will always be given.

You can contribute in many ways:

## Types of Contributions

### Report Bugs

Report bugs at <https://github.com/pinellolab/pyrovelocity/issues>.

If you are reporting a bug, please fill out the provided template including:

- Your operating system name and version.
- Any details about your local setup that might be helpful in troubleshooting.
- Detailed steps to reproduce the bug.

### Fix Bugs

Look through the GitHub issues for bugs. Anything tagged with "bug" and "help
wanted" is open to whoever wants to implement it.

### Implement Features

Look through the GitHub issues for features. Anything tagged with "enhancement"
and "help wanted" is open to whoever wants to implement it.

### Write Documentation

pyrovelocity could always use more documentation, whether as part of the
official pyrovelocity docs, in docstrings, or even on the web in blog posts,
articles, and such.

### Submit Feedback

The best way to send feedback is to file an issue at <https://github.com/pinellolab/pyrovelocity/issues>.

If you are proposing a feature:

- Explain in detail how it would work.
- Keep the scope as narrow as possible, to make it easier to implement.
- Remember that this is a volunteer-driven project, and that contributions
  are welcome :)

## Get Started

### Cloud

See the [reproducibility/environment](https://github.com/pinellolab/pyrovelocity/tree/main/reproducibility/environment) folder and associated README.md.

### Local

The following is a rough guide to setting up `pyrovelocity` for local development.

1. Fork the `pyrovelocity` repo on GitHub.

2. Clone your fork locally:

   ```
   $ git clone https://github.com/your_name_here/pyrovelocity.git
   ```

3. Install your local copy with [poetry and nox](https://cookiecutter-hypermodern-python.readthedocs.io/en/2022.6.3.post1/guide.html#requirements) or with [conda]().

4. Create a branch for local development:

   ```
   $ git checkout -b name-of-your-bugfix-or-feature
   ```

   Now you can make your changes locally.

5. When you're done making changes, you can check that your changes pass the
   most basic checks implemented in noxfile.py (run `nox --list-sessions` to list all available):

   ```
   $ nox -x -rs pre-commit
   $ nox -x -rs tests-3.10
   $ nox -x -rs docs-build
   ```

   These will be confirmed via the GitHub actions workflow that will run on your fork and pull request.

6. Commit your changes and push your branch to GitHub:

   ```
   $ git add .
   $ git commit -m "Your detailed description of your changes."
   $ git push origin name-of-your-bugfix-or-feature
   ```

7. Submit a pull request through the GitHub website.

## Pull Request Guidelines

Before you submit a pull request, check that it meets these guidelines:

1. The pull request should include pytest tests and xdoctests.
2. If the pull request adds functionality, the docs should be updated. Put
   your new functionality into a function with a docstring.
3. The pull request should work for Python 3.10.

### 2024-03-24 nbs/guides/installation.qmd
feat(nbs): init installation Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Installation
---

## PyPI

## Anaconda

### 2024-03-24 nbs/index.qmd
fix(nbs): remove gh pages badge Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-03-17 nbs/index.qmd
@@
 ---
-title: "Pyro-Velocity"
+title: Pyro-Velocity
+description: "Probabilistic modeling of RNA velocity"
+format:
+  html:
+    toc: false
+
+about:
+  id: information
+  template: jolla
+  image: _static/logo.png
+  links: 
+    - icon: download
+      href: guides/installation.qmd
+    - icon: book
+      href: tutorials/introduction.qmd
+    - icon: github
+      text: GitHub
+      href: https://github.com/pinellolab/pyrovelocity
 ---
 
-Pyro-Velocity is a Bayesian, generative, and multivariate RNA velocity model to estimate uncertainty in predictions of future cell states from minimal models approximating splicing dynamics.
+::: {#information}
+:::
+
+<div style="margin-left: auto; margin-right: auto; max-width: 560px; overflow-x: auto;">
+
+|         |                                                                                                                                                                        |
+| ------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
+| CI/CD   | [![CID][cid-badge]][cid-link] [![CML][cml-badge]][cml-link] [![pre-commit.ci status][precommit-badge]][precommit-link]                                                 |
+| Docs    | [![Cloudflare Pages][cloudflare-badge]][cloudflare-link] [![Preprint][preprint-badge]][preprint-link]                 |
+| Package | [![PyPI - Version][pypi-badge]][pypi-link] [![Anaconda badge][anaconda-badge]][anaconda-link] [![Docker image][docker-badge]][docker-link]                             |
+| Meta    | [![codecov][codecov-badge]][codecov-link] [![code style][black-badge]][black-link] [![License - MIT][license-badge]][license-link] [![Tuple][tuple-badge]][tuple-link] |
+
+[cid-badge]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cid.yaml/badge.svg?branch=master
+[cid-link]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cid.yaml
+[cml-badge]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cml.yml/badge.svg
+[cml-link]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cml.yml
+[precommit-badge]: https://results.pre-commit.ci/badge/github/pinellolab/pyrovelocity/main.svg
+[precommit-link]: https://results.pre-commit.ci/latest/github/pinellolab/pyrovelocity/master
+[cloudflare-badge]: https://img.shields.io/badge/Docs-pages-F38020?style=flat&logo=cloudflare&labelColor=black
+[cloudflare-link]: https://docs.pyrovelocity.net
+[preprint-badge]: https://img.shields.io/badge/doi-10.1101/2022.09.12.507691v2-B31B1B
+[preprint-link]: https://doi.org/10.1101/2022.09.12.507691
+[pypi-badge]: https://img.shields.io/pypi/v/pyrovelocity.svg?logo=pypi&label=PyPI&logoColor=gold
+[pypi-link]: https://pypi.org/project/pyrovelocity/
+[anaconda-badge]: https://anaconda.org/conda-forge/pyrovelocity/badges/version.svg
+[anaconda-link]: https://anaconda.org/conda-forge/pyrovelocity
+[docker-badge]: https://img.shields.io/badge/docker-image-blue?logo=docker
+[docker-link]: https://github.com/pinellolab/pyrovelocity/pkgs/container/pyrovelocity
+[codecov-badge]: https://codecov.io/gh/pinellolab/pyrovelocity/branch/main/graph/badge.svg
+[codecov-link]: https://codecov.io/gh/pinellolab/pyrovelocity
+[black-badge]: https://img.shields.io/badge/code%20style-black-000000.svg
+[black-link]: https://github.com/psf/black
+[license-badge]: https://img.shields.io/badge/license-AGPL%203-purple
+[license-link]: https://spdx.org/licenses/
+[tuple-badge]: https://img.shields.io/badge/Tuple%20❤️%20OSS-5A67D8?logo=tuple
+[tuple-link]: https://tuple.app/github-badge
+
+</div>

### 2024-03-25 nbs/tutorials/introduction.qmd
fix(nbs): update index Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
fix(docs): sync description Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 # Introduction
+
+[Pyro-Velocity](https://docs.pyrovelocity.net) is a library for probabilistic inference in minimal models approximating gene expression dynamics from, possibly multimodal, single-cell sequencing data.
+It provides posterior estimates of gene expression parameters, predictive estimates of gene expression states, and local estimates of cell state transition probabilities.
+It can be used as a component in frameworks that attempt to retain the ability to propagate uncertainty in predicting: distributions over cell fates from subpopulations of cell states, response to cell state perturbations, or candidate genes or gene modules that correlate with determination of specific cell fates.
 

### 2024-03-25 src/pyrovelocity/plots/_predictive.py
fix(plots): default to loading data for predictive sampling to cpu Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -290,14 +290,21 @@ def posterior_curve(
 
 
 def extrapolate_prediction_sample_predictive(
-    posterior_time, data_model_path, adata, grid_time_points=1000
+    posterior_time,
+    data_model_path,
+    adata,
+    grid_time_points=1000,
+    use_gpu="cpu",
 ):
-    # data_model_path = data_model_conf.model_path
     PyroVelocity.setup_anndata(adata)
     model = PyroVelocity(
         adata, add_offset=False, guide_type="auto_t0_constraint"
     )
-    model = model.load_model(data_model_path, adata, use_gpu="auto")
+    model = model.load_model(
+        dir_path=data_model_path,
+        adata=adata,
+        use_gpu=use_gpu,
+    )
     print(data_model_path)
 
     scdl = model._make_data_loader(adata=adata, indices=None, batch_size=1000)

### 2024-03-25 nbs/index.qmd
nit(nbs): reduce table width Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
nit(nbs): update description Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 title: Pyro-Velocity
-description: "Probabilistic modeling of RNA velocity"
+description: "𝒫robabilistic modeling of RNA velocity ⬱"
 format:
@@
 
-<div style="margin-left: auto; margin-right: auto; max-width: 560px; overflow-x: auto;">
+<div style="margin-left: auto; margin-right: auto; max-width: 520px; overflow-x: auto;">
 

### 2024-03-25 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
fix(pypi_template): update frontmatter Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: Pyrovelocity PyPI user notebook template
+title: Pyrovelocity Notebook Usage Example
 toc: true
@@
 execute: 
+  freeze: true
   eval: true
@@
 abstract: |
-  This notebook demonstrates how to run pyrovelocity in a Jupyter notebook.
-  It attempts to support both Google Colab and local or remote Jupyter kernel servers.
+  This notebook demonstrates how to run Pyro-Velocity in a Jupyter notebook on
+  downsampled data. The results are not expected to be meaningful, but the
+  notebook demonstrates the usage workflow with both a short execution time and
+  using limited resources. It attempts to support both Google Colab and local or
+  remote Jupyter kernel servers. Links to a blank copy of the notebook are 
+  provided below.
 keywords: [single-cell transcriptomics, probabilistic modeling, dynamical systems, RNA velocity]
@@
 ```
+
+::: {#nte-blank-cells .callout-note collapse=true title="Blank cells"}
+There are a number of cells in this notebook with
+
+```python
+print("This cell is intentionally left blank.") 
+```
+
+These can be safely ignored or removed. 
+They are exclusively used for compatibility with Google Colab's section folding logic.
+:::
 
@@
 
+```{python}
+#| label: blank-cell-03
+#| code-fold: true
+#| output: false
+print("This cell is intentionally left blank.") 
+```
+
 Here we will load the data and posterior samples, perform the analysis used to generate the volcano plot for gene selection, and then regenerate the plot of phase portraits, temporal evolution, and shared time correlation.
@@
 This time we will proceed quickly through the analysis with less discussion of the various configuration settings and associated task functions.
+
+```{python}
+#| label: blank-cell-04
+#| code-fold: true
+#| output: false
+print("This cell is intentionally left blank.") 
+```
 

### 2024-03-25 nbs/tutorials/introduction_v013.qmd
refactor(nbs): retain 013 and 020 introductions Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-03-24 nbs/tutorials/introduction.qmd
@@
 # Introduction
+
+`Pyro-Velocity` is a Bayesian, generative, and multivariate RNA velocity model to estimate uncertainty in predictions of future cell states from minimal models approximating splicing dynamics.
+This approach models raw sequencing counts with synchronized cell time across all expressed genes to provide quantifiable information on cell fate choice and developmental trajectory dynamics.
 
@@
 We currently support installation and usage in a linux 64-bit development environment with access to a GPU.
-An [IaC](https://en.wikipedia.org/wiki/Infrastructure_as_code) setup that works with [GCP](https://registry.terraform.io/providers/hashicorp/google/latest/docs) is documented in [reproducibility/environment/README.md](reproducibility/environment/README.md).
+An [IaC](https://en.wikipedia.org/wiki/Infrastructure_as_code) setup that works with [GCP](https://registry.terraform.io/providers/hashicorp/google/latest/docs) is documented in [reproducibility/environment/README.md](https://github.com/pinellolab/pyrovelocity/blob/main/reproducibility/environment/README.md).
 Before proceeding to setup a minimal development environment, please fork this repository and clone a copy of your fork to your development machine.
@@
 
-Also see [contributing](./docs/contributing.md).
+Also see [contributing](https://github.com/pinellolab/pyrovelocity/blob/main/docs/contributing.md).

### 2024-03-25 src/pyrovelocity/train.py
fix(train): close figure Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -406,6 +406,7 @@ def train_model(
     )
 
     fig.savefig(loss_plot_path, facecolor="white", bbox_inches="tight")
+    plt.close(fig)
 
     logger.info(f"AnnData object after model training")
     print_anndata(adata)

### 2024-03-25 src/pyrovelocity/preprocess.py
fix(preprocess): close figure Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -415,7 +415,7 @@ def plot_high_us_genes(
             edgecolor="none",
             dpi=300,
         )
-
+    plt.close(fig)
     return fig
 
 

### 2024-03-25 nbs/templates/user_example/user_example.qmd
feat(nbs): add user example to templates Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-03-25 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
@@
 format:
-  nature-pdf:
-    natbibstyle: sn-nature
-    # classoption:
-    #   - lineno
-    cite-method: citeproc
-    keep-tex: true
+#   nature-pdf:
+#     natbibstyle: sn-nature
+#     # classoption:
+#     #   - lineno
+#     cite-method: citeproc
+#     keep-tex: true
   html:
-    mainfont: Latin Modern Sans
-    code-fold: false
     html-math-method: katex
-    embed-resources: true
-  docx: default
-  ipynb: default
-format-links: false
+#     mainfont: Latin Modern Sans
+#     code-fold: false
+#     embed-resources: true
+#   docx: default
+#   ipynb: default
+# format-links: [ipynb]
 execute: 
@@
   cache: true
-  keep-ipynb: true
-  keep-md: true
+  # keep-ipynb: true
+  # keep-md: true
 author:
@@
 
-<a target="_blank" href="https://colab.research.google.com/github/pinellolab/pyrovelocity/blob/530-nbwf/docs/source/notebooks/pyrovelocity_pypi_user_template.ipynb">
+<a target="_blank" href="https://colab.research.google.com/github/pinellolab/pyrovelocity/blob/530-nbwf/scripts/notebooks/user_example.ipynb">
   <img 
@@
     width="109" height="20"/>
-</a> <a target="_blank" href="https://nbviewer.jupyter.org/github/pinellolab/pyrovelocity/blob/530-nbwf/docs/source/notebooks/pyrovelocity_pypi_user_template.ipynb">
+</a> <a target="_blank" href="https://nbviewer.jupyter.org/github/pinellolab/pyrovelocity/blob/530-nbwf/scripts/notebooks/user_example.ipynb">
   <img 

### 2024-03-26 nbs/templates/user_example/user_example.qmd
feat(nbs): export ipynb format Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 #   docx: default
-#   ipynb: default
-# format-links: [ipynb]
+  ipynb: default
+format-links: [ipynb]
 execute: 

### 2024-03-26 nbs/templates/user_example/user_example.qmd
fix(nbs): use gdrive notebook in colab link Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-<a target="_blank" href="https://colab.research.google.com/github/pinellolab/pyrovelocity/blob/530-nbwf/scripts/notebooks/user_example.ipynb">
+<a target="_blank" href="https://colab.research.google.com/drive/1uWTZgL5lURxoXWPjtT9IJBIJyGVVDiPg">
   <img 
@@
     alt="Open In Colab"
-    width="109" height="20"/>
-</a> <a target="_blank" href="https://nbviewer.jupyter.org/github/pinellolab/pyrovelocity/blob/530-nbwf/scripts/notebooks/user_example.ipynb">
-  <img 
-    src="https://raw.githubusercontent.com/jupyter/design/master/logos/Badges/nbviewer_badge.png"
-    alt="Render with nbviewer" 
     width="109" height="20"/>

### 2024-03-26 nbs/index.qmd
chore(docs): sync badges Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 |         |                                                                                                                                                           |
-| ------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
-| CI/CD   | [![CID][cid-badge]][cid-link] [![CML][cml-badge]][cml-link] [![pre-commit.ci status][precommit-badge]][precommit-link]                                                 |
+| ------- | --------------------------------------------------------------------------------------------------------------------------------------------------------- |
+| CI/CD   | [![CID][cid-badge]][cid-link] [![codecov][codecov-badge]][codecov-link] [![pre-commit.ci status][precommit-badge]][precommit-link]                        |
 | Docs    | [![Cloudflare Pages][cloudflare-badge]][cloudflare-link] [![Preprint][preprint-badge]][preprint-link]                                                     |
-| Package | [![PyPI - Version][pypi-badge]][pypi-link] [![Anaconda badge][anaconda-badge]][anaconda-link] [![Docker image][docker-badge]][docker-link]                             |
-| Meta    | [![codecov][codecov-badge]][codecov-link] [![code style][black-badge]][black-link] [![License - MIT][license-badge]][license-link] [![Tuple][tuple-badge]][tuple-link] |
+| Package | [![PyPI - Version][pypi-badge]][pypi-link] [![Conda-forge badge][conda-forge-badge]][anaconda-link] [![Docker image][docker-badge]][docker-link]          |
+| Meta    | [![flyte-badge]][flyte-link] [![code style][black-badge]][black-link] [![License - MIT][license-badge]][license-link] [![Tuple][tuple-badge]][tuple-link] |
 
@@
 [cid-link]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cid.yaml
-[cml-badge]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cml.yml/badge.svg
-[cml-link]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cml.yml
 [precommit-badge]: https://results.pre-commit.ci/badge/github/pinellolab/pyrovelocity/main.svg
-[precommit-link]: https://results.pre-commit.ci/latest/github/pinellolab/pyrovelocity/master
-[cloudflare-badge]: https://img.shields.io/badge/Docs-pages-F38020?style=flat&logo=cloudflare&labelColor=black
+[precommit-link]: https://results.pre-commit.ci/latest/github/pinellolab/pyrovelocity/main
+[flyte-badge]: https://img.shields.io/badge/flyte-carrier-gray.svg?color=7552A2&logo=data:image/svg%2bxml;base64,PHN2ZyBpZD0iTGF5ZXJfMSIgZGF0YS1uYW1lPSJMYXllciAxIiB4bWxucz0iaHR0cDovL3d3dy53My5vcmcvMjAwMC9zdmciIHhtbG5zOnhsaW5rPSJodHRwOi8vd3d3LnczLm9yZy8xOTk5L3hsaW5rIiB2aWV3Qm94PSIwIDAgMzYwIDMzMy40ODM3OSI+PGRlZnM+PHN0eWxlPi5jbHMtMXtmaWxsOnVybCgjbGluZWFyLWdyYWRpZW50KTt9PC9zdHlsZT48bGluZWFyR3JhZGllbnQgaWQ9ImxpbmVhci1ncmFkaWVudCIgeDE9IjYuMTkzNjUiIHkxPSIxNjcuMzQ3NDUiIHgyPSIzNTIiIHkyPSIxNjcuMzQ3NDUiIGdyYWRpZW50VW5pdHM9InVzZXJTcGFjZU9uVXNlIj48c3RvcCBvZmZzZXQ9IjAiIHN0b3AtY29sb3I9IiNlNWM5ZDMiLz48c3RvcCBvZmZzZXQ9IjAuMzMiIHN0b3AtY29sb3I9IiNiMzg1ZGYiLz48c3RvcCBvZmZzZXQ9IjEiIHN0b3AtY29sb3I9IiM2ZjJhZWYiLz48L2xpbmVhckdyYWRpZW50PjwvZGVmcz48cGF0aCBjbGFzcz0iY2xzLTEiIGQ9Ik05MC45NjQ3NSwxNzAuODEzMzRjNy42Mjc3Ny0xOC4xMjcsMTEuNTEwNDctMzYuNzc2NTUsMTUuMDA2NC01NS40OTI4OSw1Ljk2MDM3LTMxLjkxMDIsOS45OS02NC4wMzIzLDguNjM2MjEtOTYuNTk0NzUtLjI1NzIzLTYuMTg3NDIsMS40ODQ5NS04LjQ4NzYzLDguMTQ5My03Ljg1ODIzLDE4LjA0NDY1LDEuNzA0MjUsMzUuOTM3NDYsMy43NjY1Nyw1Mi4xNTksMTIuNjUyODYsNC4yOTQyNSwyLjM1MjQxLDguMDYyNzIsMS4wODk0OSwxMi4xNjgzLS4wNjgxLDMxLjA0ODU5LTguNzU0NCw2Mi41MzI3Mi0xNC42OTE0Niw5NC45NjgyNS0xNC4wMjQ1NiwyMS4zMTQuNDM4Miw0MS45ODQyNSwzLjg3ODkxLDYxLjY0Mzc5LDEyLjcwMjc2LDYuNDI3MjksMi44ODQ3OSw4LjgwODY5LDYuOTg2NDEsOC4yMTY1NywxMy44MS0xLjU5MzM4LDE4LjM2MDg0LTguOTMzNDUsMzQuODE3NTEtMTYuNTU1NzEsNTEuMTgwMzgtMi4wODM3NSw0LjQ3MzEzLTUuMTEzMiwxLjkwMDYzLTcuNjg0MTcuMjc2ODYtMjUuODcwMzMtMTYuMzM4ODYtNTMuNjUxLTI4LjcwMDYzLTgyLjI2NC0zOS4yNTUxNi0yMC4zMDkwNi03LjQ5MTQ2LTQwLjg0NzY4LTE0LjMwOTczLTYzLjA1ODIzLTE3LjcxMjkzLDYuODUzODgsNi43ODE4MiwxMy40NjQwNywxMy44MzI2NiwyMC42MDM1NSwyMC4yOTkxMiwzMS42MDk3OSwyOC42Myw2NC41NTE4NSw1NS40OTQsMTAyLjYyNDMsNzUuMzE3NjQsNC4yMDA1MiwyLjE4NzEyLDUuMjE5NDYsNC40Mjk0MiwyLjMzNzMyLDguNTIyMTMtMTAuNTgzNzQsMTUuMDI5MDgtMjEuNDEwODUsMjkuODczMzEtMzcuMjkwNjQsMzkuODIzMTItNC43MjI4MywyLjk1OTE4LTUuODUyNDYsNy4xNjgzOC03LjAyMjgzLDExLjg0MjItOS4xMDQzMywzNi4zNTkyOS0yMi4zMzIsNzAuOTU1ODEtNDQuMDI1MzEsMTAxLjg2NTA3YTE0MS4zMjU0NSwxNDEuMzI1NDUsMCwwLDEtMzIuNzkzMiwzMy44NDUyMWMtNS44MDIwOSw0LjE5ODExLTEwLjk2OTg0LDQuNjkyLTE2Ljg1MDguMjcxOC0xNC4xOTA0Ny0xMC42NjU4OS0yNC44NjAxOS0yNC40MDA5My0zNC43NTE4NC0zOC44NzctMi43MTgzNy0zLjk3ODMzLS43ODg4Ni02LjE1NjEyLDIuNzA4NC03Ljk1NSwyOC41MTE4My0xNC42NjUyNyw1My44NDQ5NS0zMy45MTk2Nyw3OC4yNjQzNi01NC40NTM3YTQ1Mi44NjksNDUyLjg2OSwwLDAsMCw0Mi41MjQwOS00MC4wNmMtMi44MDMzNy0yLjY3MTQxLTUuMjA3ODItLjcwNTI5LTcuMzA0OC0uMTQ2MTctNDcuMDUyLDEyLjU0NS05Mi4zNzI1NCwyOS4zODkyMS0xMzQuMDQwODIsNTUuMDEwOS01LjY5NTE1LDMuNTAxOTMtOC43MjM0NiwzLjA2NjE2LTExLjQ5OTg1LTMuMzM5MzctNy4wOTI2OC0xNi4zNjM2Ny0xMy42NjczNC0zMi44Njc1MS0xNC43NzQxOS01MC44NjkzNi0uMzYyNjgtNS44OTgxNC0zLjk2ODM2LTkuMDc1NTUtNy41NjI2NC0xMi41Njg0Ni0yNS42NTEyMy0yNC45MjcxNy00Ny45NjEtNTIuMzc0ODMtNjMuNDg0ODYtODQuODQzN0MxMi40MjgyOSw2OC4yNTUsNi44ODkxOSw1MS44MDAxNCw2LjIxMDg5LDMzLjkzMzgxYy0uMjY5LTcuMDg2MDYsMi42MDM2OC0xMS4wMzkyNSw5LjAwMTQ3LTEzLjM0NDlhMTY4LjI0NDkxLDE2OC4yNDQ5MSwwLDAsMSw0Ni4xNTYtOS42ODQxNGM2LjMyMDE1LS40MjE2Miw3Ljg3MzEsMS44MDQzMyw3LjcwNzY4LDcuNTA2NjgtMS4xNjg0MSw0MC4yODA2LDQuOTg5MjMsNzkuNzU1LDEzLjU3NjQ2LDExOC45MjMsMi40NjE1MiwxMS4yMjc0MSw1LjUyNDg2LDIyLjMyMjk0LDguMzEyMjcsMzMuNDc4ODRaTTE3OS40MTM2OSwzOS41MDhDMTY5LjA5Niw5Mi45NjE2MSwxMzYuNjc2ODUsMTMyLjkwNDIxLDEwMC4xMzIxMywxNzAuNTI4MmM0OS44OTE3Ny0yMC4zNzEsOTkuNjI4NTctMTEuMjM1NywxNDkuODI1NDMtMS4zMjA3NkMyMTEuNDQzODcsMTMzLjc1OTY5LDE5My45ODkyNiw4Ny43MTM2MywxNzkuNDEzNjksMzkuNTA4WiIvPjwvc3ZnPg==
+[flyte-link]: https://docs.flyte.org
+[cloudflare-badge]: https://img.shields.io/badge/Docs-pages-gray.svg?style=flat&logo=cloudflare&color=F26722
 [cloudflare-link]: https://docs.pyrovelocity.net
@@
 [preprint-link]: https://doi.org/10.1101/2022.09.12.507691
-[pypi-badge]: https://img.shields.io/pypi/v/pyrovelocity.svg?logo=pypi&label=PyPI&logoColor=gold
+[pypi-badge]: https://img.shields.io/pypi/v/pyrovelocity.svg?logo=pypi&label=PyPI&color=F26722&logoColor=F26722
 [pypi-link]: https://pypi.org/project/pyrovelocity/
-[anaconda-badge]: https://anaconda.org/conda-forge/pyrovelocity/badges/version.svg
+[conda-forge-badge]: https://img.shields.io/conda/vn/conda-forge/pyrovelocity.svg?logo=conda-forge&label=conda-forge&color=F26722
 [anaconda-link]: https://anaconda.org/conda-forge/pyrovelocity
@@
 [black-link]: https://github.com/psf/black
-[license-badge]: https://img.shields.io/badge/license-AGPL%203-purple
+[license-badge]: https://img.shields.io/badge/license-AGPL%203-gray.svg?color=F26722
 [license-link]: https://spdx.org/licenses/

### 2024-03-26 src/pyrovelocity/workflows/main_workflow.py
fix(workflows): note migration to CalVer for cache versions Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -63,6 +63,8 @@ __all__ = [
 
 logger = configure_logging(__name__)
 
+# TODO: Migrate cache versioning to CalVer ( https://calver.org/ ) to avoid
+# potential for confusion with an intent to update the cache with every release.
 CACHE_VERSION = "0.2.0b12"
 SUMMARIZE_CACHE_VERSION = "0.2.0b13.dev0"
 PYROVELOCITY_CACHE_FLAG = str_to_bool(

### 2024-03-26 nbs/templates/user_example/user_example.qmd
chore(nbs): `530-nbwf` -> `beta` Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
         print("Installing pyrovelocity...")
-        install_package("pyrovelocity[workflows] @ git+https://github.com/pinellolab/pyrovelocity.git@530-nbwf")
+        install_package("pyrovelocity[workflows] @ git+https://github.com/pinellolab/pyrovelocity.git@beta")
         try:

### 2024-03-26 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
chore(nbs): `530-nbwf` -> `beta` Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-<a target="_blank" href="https://colab.research.google.com/github/pinellolab/pyrovelocity/blob/530-nbwf/docs/source/notebooks/pyrovelocity_pypi_user_template.ipynb">
+<a target="_blank" href="https://colab.research.google.com/github/pinellolab/pyrovelocity/blob/beta/docs/source/notebooks/pyrovelocity_pypi_user_template.ipynb">
   <img 
@@
     width="109" height="20"/>
-</a> <a target="_blank" href="https://nbviewer.jupyter.org/github/pinellolab/pyrovelocity/blob/530-nbwf/docs/source/notebooks/pyrovelocity_pypi_user_template.ipynb">
+</a> <a target="_blank" href="https://nbviewer.jupyter.org/github/pinellolab/pyrovelocity/blob/beta/docs/source/notebooks/pyrovelocity_pypi_user_template.ipynb">
   <img 

### 2024-03-29 nbs/templates/user_example/user_example.qmd
chore(nbs): sync user example with library refactor Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 #| label: print-preprocess-data-docstring
-print_docstring(pyrovelocity.preprocess.preprocess_dataset)
+print_docstring(pyrovelocity.tasks.preprocess.preprocess_dataset)
 ```
@@
 
-The `PyrovelocityTrainInterface` is a configuration for the `pyrovelocity.train.train_dataset` function whose documentation we print below.
+The `PyrovelocityTrainInterface` is a configuration for the `pyrovelocity.tasks.train.train_dataset` function whose documentation we print below.
 
@@
 #| label: print-train-model-docstring
-print_docstring(pyrovelocity.train.train_dataset)
+print_docstring(pyrovelocity.tasks.train.train_dataset)
 ```
@@
 
-The configuration required to postprocess the data are determined by the components of the `pyrovelocity.postprocess.postprocess_dataset` function interface whose documentation can be accessed via `help` or `?` in the notebook
+The configuration required to postprocess the data are determined by the components of the `pyrovelocity.tasks.postprocess.postprocess_dataset` function interface whose documentation can be accessed via `help` or `?` in the notebook
 
@@
 #| label: print-postprocess-data-docstring
-print_docstring(pyrovelocity.postprocess.postprocess_dataset)
+print_docstring(pyrovelocity.tasks.postprocess.postprocess_dataset)
 ```
@@
 volcano_data = posterior_samples["gene_ranking"] 
-putative_marker_genes = pv.analyze.pareto_frontier_genes(volcano_data=volcano_data, num_genes=6)
+putative_marker_genes = pv.analysis.analyze.pareto_frontier_genes(volcano_data=volcano_data, num_genes=6)
 ```
@@
 
-The best place to review the approach to generating the above plots is in [the source code for the `pyrovelocity.summarize` module](https://github.com/pinellolab/pyrovelocity/tree/beta/src/pyrovelocity/summarize.py) or in the [docs](https://github.io/pinellolab/pyrovelocity).
+The best place to review the approach to generating the above plots is in [the source code for the `pyrovelocity.tasks.summarize` module](https://github.com/pinellolab/pyrovelocity/tree/beta/src/pyrovelocity/tasks/summarize.py) or in the [docs](https://github.io/pinellolab/pyrovelocity).
 Additional links are shown in the help below.

### 2024-03-29 nbs/index.qmd
docs(readme): update badges
- add pyro, hydra-zen, and scvi-tools
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 | Package | [![PyPI - Version][pypi-badge]][pypi-link] [![Conda-forge badge][conda-forge-badge]][anaconda-link] [![Docker image][docker-badge]][docker-link]          |
-| Meta    | [![flyte-badge]][flyte-link] [![code style][black-badge]][black-link] [![License - MIT][license-badge]][license-link] [![Tuple][tuple-badge]][tuple-link] |
+| Meta    | [![pyro-ppl-badge]][pyro-ppl-link] [![flyte-badge]][flyte-link] [![hydra-zen-badge]][hydra-zen-link] [![scvi-tools-badge]][scvi-tools-link] [![code style][black-badge]][black-link] [![License - MIT][license-badge]][license-link] [![Tuple][tuple-badge]][tuple-link] |
 
@@
 [precommit-link]: https://results.pre-commit.ci/latest/github/pinellolab/pyrovelocity/main
-[flyte-badge]: https://img.shields.io/badge/flyte-carrier-gray.svg?color=7552A2&logo=data:image/svg%2bxml;base64,PHN2ZyBpZD0iTGF5ZXJfMSIgZGF0YS1uYW1lPSJMYXllciAxIiB4bWxucz0iaHR0cDovL3d3dy53My5vcmcvMjAwMC9zdmciIHhtbG5zOnhsaW5rPSJodHRwOi8vd3d3LnczLm9yZy8xOTk5L3hsaW5rIiB2aWV3Qm94PSIwIDAgMzYwIDMzMy40ODM3OSI+PGRlZnM+PHN0eWxlPi5jbHMtMXtmaWxsOnVybCgjbGluZWFyLWdyYWRpZW50KTt9PC9zdHlsZT48bGluZWFyR3JhZGllbnQgaWQ9ImxpbmVhci1ncmFkaWVudCIgeDE9IjYuMTkzNjUiIHkxPSIxNjcuMzQ3NDUiIHgyPSIzNTIiIHkyPSIxNjcuMzQ3NDUiIGdyYWRpZW50VW5pdHM9InVzZXJTcGFjZU9uVXNlIj48c3RvcCBvZmZzZXQ9IjAiIHN0b3AtY29sb3I9IiNlNWM5ZDMiLz48c3RvcCBvZmZzZXQ9IjAuMzMiIHN0b3AtY29sb3I9IiNiMzg1ZGYiLz48c3RvcCBvZmZzZXQ9IjEiIHN0b3AtY29sb3I9IiM2ZjJhZWYiLz48L2xpbmVhckdyYWRpZW50PjwvZGVmcz48cGF0aCBjbGFzcz0iY2xzLTEiIGQ9Ik05MC45NjQ3NSwxNzAuODEzMzRjNy42Mjc3Ny0xOC4xMjcsMTEuNTEwNDctMzYuNzc2NTUsMTUuMDA2NC01NS40OTI4OSw1Ljk2MDM3LTMxLjkxMDIsOS45OS02NC4wMzIzLDguNjM2MjEtOTYuNTk0NzUtLjI1NzIzLTYuMTg3NDIsMS40ODQ5NS04LjQ4NzYzLDguMTQ5My03Ljg1ODIzLDE4LjA0NDY1LDEuNzA0MjUsMzUuOTM3NDYsMy43NjY1Nyw1Mi4xNTksMTIuNjUyODYsNC4yOTQyNSwyLjM1MjQxLDguMDYyNzIsMS4wODk0OSwxMi4xNjgzLS4wNjgxLDMxLjA0ODU5LTguNzU0NCw2Mi41MzI3Mi0xNC42OTE0Niw5NC45NjgyNS0xNC4wMjQ1NiwyMS4zMTQuNDM4Miw0MS45ODQyNSwzLjg3ODkxLDYxLjY0Mzc5LDEyLjcwMjc2LDYuNDI3MjksMi44ODQ3OSw4LjgwODY5LDYuOTg2NDEsOC4yMTY1NywxMy44MS0xLjU5MzM4LDE4LjM2MDg0LTguOTMzNDUsMzQuODE3NTEtMTYuNTU1NzEsNTEuMTgwMzgtMi4wODM3NSw0LjQ3MzEzLTUuMTEzMiwxLjkwMDYzLTcuNjg0MTcuMjc2ODYtMjUuODcwMzMtMTYuMzM4ODYtNTMuNjUxLTI4LjcwMDYzLTgyLjI2NC0zOS4yNTUxNi0yMC4zMDkwNi03LjQ5MTQ2LTQwLjg0NzY4LTE0LjMwOTczLTYzLjA1ODIzLTE3LjcxMjkzLDYuODUzODgsNi43ODE4MiwxMy40NjQwNywxMy44MzI2NiwyMC42MDM1NSwyMC4yOTkxMiwzMS42MDk3OSwyOC42Myw2NC41NTE4NSw1NS40OTQsMTAyLjYyNDMsNzUuMzE3NjQsNC4yMDA1MiwyLjE4NzEyLDUuMjE5NDYsNC40Mjk0MiwyLjMzNzMyLDguNTIyMTMtMTAuNTgzNzQsMTUuMDI5MDgtMjEuNDEwODUsMjkuODczMzEtMzcuMjkwNjQsMzkuODIzMTItNC43MjI4MywyLjk1OTE4LTUuODUyNDYsNy4xNjgzOC03LjAyMjgzLDExLjg0MjItOS4xMDQzMywzNi4zNTkyOS0yMi4zMzIsNzAuOTU1ODEtNDQuMDI1MzEsMTAxLjg2NTA3YTE0MS4zMjU0NSwxNDEuMzI1NDUsMCwwLDEtMzIuNzkzMiwzMy44NDUyMWMtNS44MDIwOSw0LjE5ODExLTEwLjk2OTg0LDQuNjkyLTE2Ljg1MDguMjcxOC0xNC4xOTA0Ny0xMC42NjU4OS0yNC44NjAxOS0yNC40MDA5My0zNC43NTE4NC0zOC44NzctMi43MTgzNy0zLjk3ODMzLS43ODg4Ni02LjE1NjEyLDIuNzA4NC03Ljk1NSwyOC41MTE4My0xNC42NjUyNyw1My44NDQ5NS0zMy45MTk2Nyw3OC4yNjQzNi01NC40NTM3YTQ1Mi44NjksNDUyLjg2OSwwLDAsMCw0Mi41MjQwOS00MC4wNmMtMi44MDMzNy0yLjY3MTQxLTUuMjA3ODItLjcwNTI5LTcuMzA0OC0uMTQ2MTctNDcuMDUyLDEyLjU0NS05Mi4zNzI1NCwyOS4zODkyMS0xMzQuMDQwODIsNTUuMDEwOS01LjY5NTE1LDMuNTAxOTMtOC43MjM0NiwzLjA2NjE2LTExLjQ5OTg1LTMuMzM5MzctNy4wOTI2OC0xNi4zNjM2Ny0xMy42NjczNC0zMi44Njc1MS0xNC43NzQxOS01MC44NjkzNi0uMzYyNjgtNS44OTgxNC0zLjk2ODM2LTkuMDc1NTUtNy41NjI2NC0xMi41Njg0Ni0yNS42NTEyMy0yNC45MjcxNy00Ny45NjEtNTIuMzc0ODMtNjMuNDg0ODYtODQuODQzN0MxMi40MjgyOSw2OC4yNTUsNi44ODkxOSw1MS44MDAxNCw2LjIxMDg5LDMzLjkzMzgxYy0uMjY5LTcuMDg2MDYsMi42MDM2OC0xMS4wMzkyNSw5LjAwMTQ3LTEzLjM0NDlhMTY4LjI0NDkxLDE2OC4yNDQ5MSwwLDAsMSw0Ni4xNTYtOS42ODQxNGM2LjMyMDE1LS40MjE2Miw3Ljg3MzEsMS44MDQzMyw3LjcwNzY4LDcuNTA2NjgtMS4xNjg0MSw0MC4yODA2LDQuOTg5MjMsNzkuNzU1LDEzLjU3NjQ2LDExOC45MjMsMi40NjE1MiwxMS4yMjc0MSw1LjUyNDg2LDIyLjMyMjk0LDguMzEyMjcsMzMuNDc4ODRaTTE3OS40MTM2OSwzOS41MDhDMTY5LjA5Niw5Mi45NjE2MSwxMzYuNjc2ODUsMTMyLjkwNDIxLDEwMC4xMzIxMywxNzAuNTI4MmM0OS44OTE3Ny0yMC4zNzEsOTkuNjI4NTctMTEuMjM1NywxNDkuODI1NDMtMS4zMjA3NkMyMTEuNDQzODcsMTMzLjc1OTY5LDE5My45ODkyNiw4Ny43MTM2MywxNzkuNDEzNjksMzkuNTA4WiIvPjwvc3ZnPg==
+[flyte-badge]: https://img.shields.io/badge/Flyte-carrier-gray.svg?color=7552A2&logo=data:image/svg%2bxml;base64,PHN2ZyBpZD0iTGF5ZXJfMSIgZGF0YS1uYW1lPSJMYXllciAxIiB4bWxucz0iaHR0cDovL3d3dy53My5vcmcvMjAwMC9zdmciIHhtbG5zOnhsaW5rPSJodHRwOi8vd3d3LnczLm9yZy8xOTk5L3hsaW5rIiB2aWV3Qm94PSIwIDAgMzYwIDMzMy40ODM3OSI+PGRlZnM+PHN0eWxlPi5jbHMtMXtmaWxsOnVybCgjbGluZWFyLWdyYWRpZW50KTt9PC9zdHlsZT48bGluZWFyR3JhZGllbnQgaWQ9ImxpbmVhci1ncmFkaWVudCIgeDE9IjYuMTkzNjUiIHkxPSIxNjcuMzQ3NDUiIHgyPSIzNTIiIHkyPSIxNjcuMzQ3NDUiIGdyYWRpZW50VW5pdHM9InVzZXJTcGFjZU9uVXNlIj48c3RvcCBvZmZzZXQ9IjAiIHN0b3AtY29sb3I9IiNlNWM5ZDMiLz48c3RvcCBvZmZzZXQ9IjAuMzMiIHN0b3AtY29sb3I9IiNiMzg1ZGYiLz48c3RvcCBvZmZzZXQ9IjEiIHN0b3AtY29sb3I9IiM2ZjJhZWYiLz48L2xpbmVhckdyYWRpZW50PjwvZGVmcz48cGF0aCBjbGFzcz0iY2xzLTEiIGQ9Ik05MC45NjQ3NSwxNzAuODEzMzRjNy42Mjc3Ny0xOC4xMjcsMTEuNTEwNDctMzYuNzc2NTUsMTUuMDA2NC01NS40OTI4OSw1Ljk2MDM3LTMxLjkxMDIsOS45OS02NC4wMzIzLDguNjM2MjEtOTYuNTk0NzUtLjI1NzIzLTYuMTg3NDIsMS40ODQ5NS04LjQ4NzYzLDguMTQ5My03Ljg1ODIzLDE4LjA0NDY1LDEuNzA0MjUsMzUuOTM3NDYsMy43NjY1Nyw1Mi4xNTksMTIuNjUyODYsNC4yOTQyNSwyLjM1MjQxLDguMDYyNzIsMS4wODk0OSwxMi4xNjgzLS4wNjgxLDMxLjA0ODU5LTguNzU0NCw2Mi41MzI3Mi0xNC42OTE0Niw5NC45NjgyNS0xNC4wMjQ1NiwyMS4zMTQuNDM4Miw0MS45ODQyNSwzLjg3ODkxLDYxLjY0Mzc5LDEyLjcwMjc2LDYuNDI3MjksMi44ODQ3OSw4LjgwODY5LDYuOTg2NDEsOC4yMTY1NywxMy44MS0xLjU5MzM4LDE4LjM2MDg0LTguOTMzNDUsMzQuODE3NTEtMTYuNTU1NzEsNTEuMTgwMzgtMi4wODM3NSw0LjQ3MzEzLTUuMTEzMiwxLjkwMDYzLTcuNjg0MTcuMjc2ODYtMjUuODcwMzMtMTYuMzM4ODYtNTMuNjUxLTI4LjcwMDYzLTgyLjI2NC0zOS4yNTUxNi0yMC4zMDkwNi03LjQ5MTQ2LTQwLjg0NzY4LTE0LjMwOTczLTYzLjA1ODIzLTE3LjcxMjkzLDYuODUzODgsNi43ODE4MiwxMy40NjQwNywxMy44MzI2NiwyMC42MDM1NSwyMC4yOTkxMiwzMS42MDk3OSwyOC42Myw2NC41NTE4NSw1NS40OTQsMTAyLjYyNDMsNzUuMzE3NjQsNC4yMDA1MiwyLjE4NzEyLDUuMjE5NDYsNC40Mjk0MiwyLjMzNzMyLDguNTIyMTMtMTAuNTgzNzQsMTUuMDI5MDgtMjEuNDEwODUsMjkuODczMzEtMzcuMjkwNjQsMzkuODIzMTItNC43MjI4MywyLjk1OTE4LTUuODUyNDYsNy4xNjgzOC03LjAyMjgzLDExLjg0MjItOS4xMDQzMywzNi4zNTkyOS0yMi4zMzIsNzAuOTU1ODEtNDQuMDI1MzEsMTAxLjg2NTA3YTE0MS4zMjU0NSwxNDEuMzI1NDUsMCwwLDEtMzIuNzkzMiwzMy44NDUyMWMtNS44MDIwOSw0LjE5ODExLTEwLjk2OTg0LDQuNjkyLTE2Ljg1MDguMjcxOC0xNC4xOTA0Ny0xMC42NjU4OS0yNC44NjAxOS0yNC40MDA5My0zNC43NTE4NC0zOC44NzctMi43MTgzNy0zLjk3ODMzLS43ODg4Ni02LjE1NjEyLDIuNzA4NC03Ljk1NSwyOC41MTE4My0xNC42NjUyNyw1My44NDQ5NS0zMy45MTk2Nyw3OC4yNjQzNi01NC40NTM3YTQ1Mi44NjksNDUyLjg2OSwwLDAsMCw0Mi41MjQwOS00MC4wNmMtMi44MDMzNy0yLjY3MTQxLTUuMjA3ODItLjcwNTI5LTcuMzA0OC0uMTQ2MTctNDcuMDUyLDEyLjU0NS05Mi4zNzI1NCwyOS4zODkyMS0xMzQuMDQwODIsNTUuMDEwOS01LjY5NTE1LDMuNTAxOTMtOC43MjM0NiwzLjA2NjE2LTExLjQ5OTg1LTMuMzM5MzctNy4wOTI2OC0xNi4zNjM2Ny0xMy42NjczNC0zMi44Njc1MS0xNC43NzQxOS01MC44NjkzNi0uMzYyNjgtNS44OTgxNC0zLjk2ODM2LTkuMDc1NTUtNy41NjI2NC0xMi41Njg0Ni0yNS42NTEyMy0yNC45MjcxNy00Ny45NjEtNTIuMzc0ODMtNjMuNDg0ODYtODQuODQzN0MxMi40MjgyOSw2OC4yNTUsNi44ODkxOSw1MS44MDAxNCw2LjIxMDg5LDMzLjkzMzgxYy0uMjY5LTcuMDg2MDYsMi42MDM2OC0xMS4wMzkyNSw5LjAwMTQ3LTEzLjM0NDlhMTY4LjI0NDkxLDE2OC4yNDQ5MSwwLDAsMSw0Ni4xNTYtOS42ODQxNGM2LjMyMDE1LS40MjE2Miw3Ljg3MzEsMS44MDQzMyw3LjcwNzY4LDcuNTA2NjgtMS4xNjg0MSw0MC4yODA2LDQuOTg5MjMsNzkuNzU1LDEzLjU3NjQ2LDExOC45MjMsMi40NjE1MiwxMS4yMjc0MSw1LjUyNDg2LDIyLjMyMjk0LDguMzEyMjcsMzMuNDc4ODRaTTE3OS40MTM2OSwzOS41MDhDMTY5LjA5Niw5Mi45NjE2MSwxMzYuNjc2ODUsMTMyLjkwNDIxLDEwMC4xMzIxMywxNzAuNTI4MmM0OS44OTE3Ny0yMC4zNzEsOTkuNjI4NTctMTEuMjM1NywxNDkuODI1NDMtMS4zMjA3NkMyMTEuNDQzODcsMTMzLjc1OTY5LDE5My45ODkyNiw4Ny43MTM2MywxNzkuNDEzNjksMzkuNTA4WiIvPjwvc3ZnPg==
 [flyte-link]: https://docs.flyte.org
+[hydra-zen-link]: https://mit-ll-responsible-ai.github.io/hydra-zen/
+[hydra-zen-badge]: https://img.shields.io/badge/Hydra-zen-gray.svg?color=193966&logo=data:image/svg%2bxml;base64,PD94bWwgdmVyc2lvbj0iMS4wIiBlbmNvZGluZz0idXRmLTgiPz4KPCEtLSBHZW5lcmF0b3I6IEFkb2JlIElsbHVzdHJhdG9yIDI1LjQuMSwgU1ZHIEV4cG9ydCBQbHVnLUluIC4gU1ZHIFZlcnNpb246IDYuMDAgQnVpbGQgMCkgIC0tPgo8c3ZnIHZlcnNpb249IjEuMSIgaWQ9IkxheWVyXzEiIHhtbG5zPSJodHRwOi8vd3d3LnczLm9yZy8yMDAwL3N2ZyIgeG1sbnM6eGxpbms9Imh0dHA6Ly93d3cudzMub3JnLzE5OTkveGxpbmsiIHg9IjBweCIgeT0iMHB4IgoJIHZpZXdCb3g9IjAgMCA1MDAgNTAwIiBzdHlsZT0iZW5hYmxlLWJhY2tncm91bmQ6bmV3IDAgMCA1MDAgNTAwOyIgeG1sOnNwYWNlPSJwcmVzZXJ2ZSI+CjxzdHlsZSB0eXBlPSJ0ZXh0L2NzcyI+Cgkuc3Qwe2ZpbGw6IzE5Mzk2Njt9Cgkuc3Qxe2ZpbGw6I0ZGRkZGRjt9Cjwvc3R5bGU+CjxwYXRoIGNsYXNzPSJzdDAiIGQ9Ik00NDkuMywwSDUwLjdDMjIuOCwwLDAsMjIuOCwwLDUwLjd2Mzk4LjVDMCw0NzcuMiwyMi44LDUwMCw1MC43LDUwMGgzOTguNWMyNy45LDAsNTAuNy0yMi44LDUwLjctNTAuN1Y1MC43CglDNTAwLDIyLjgsNDc3LjIsMCw0NDkuMywweiIvPgo8Zz4KCTxwYXRoIGNsYXNzPSJzdDEiIGQ9Ik0zOC4yLDM5Mi4xYzAtMi45LDIuMy01LjIsNS4yLTUuMmMzLDAsNS4zLDIuNCw1LjMsNS4yVjQxMmMzLTQuMyw3LjItOCwxNC40LThjMTAuNCwwLDE2LjQsNywxNi40LDE3LjZ2MjQuNgoJCWMwLDMtMi4zLDUuMi01LjIsNS4yYy0zLDAtNS4zLTIuMy01LjMtNS4ydi0yMS40YzAtNy4yLTMuNi0xMS4zLTkuOS0xMS4zYy02LjEsMC0xMC40LDQuMy0xMC40LDExLjR2MjEuMmMwLDMtMi40LDUuMi01LjMsNS4yCgkJYy0yLjksMC01LjItMi4zLTUuMi01LjJWMzkyLjF6Ii8+Cgk8cGF0aCBjbGFzcz0ic3QxIiBkPSJNOTIuNyw0NjMuOWMtMS42LTAuNS0zLjMtMS44LTMuMy00LjRjMC0yLjYsMi00LjQsNC4zLTQuNGMwLjksMCwxLjQsMC4yLDEuOSwwLjNjMS4yLDAuMywyLjIsMC42LDMuNywwLjYKCQljMy4xLDAsNC44LTEuMyw2LjYtNC45bDAuMy0xbC0xNy4xLTM4LjNjLTAuMy0wLjgtMC42LTEuOC0wLjYtMi41YzAtMywyLjMtNS4yLDUuMi01LjJjMi43LDAsNC4yLDEuNiw1LjIsNGwxMi40LDMwLjdsMTEuNS0zMC43CgkJYzAuOS0yLjIsMi40LTQsNS4xLTRjMi44LDAsNS4xLDIuMiw1LjEsNWMwLDAuOS0wLjMsMi0wLjQsMi41bC0xNi45LDQwLjRjLTMuOSw5LjQtOC4xLDEyLjktMTUuNSwxMi45CgkJQzk3LjEsNDY1LjIsOTQuOSw0NjQuOCw5Mi43LDQ2My45eiIvPgoJPHBhdGggY2xhc3M9InN0MSIgZD0iTTE4Ni44LDQ0Ni4yYzAsMy0yLjQsNS4yLTUuMiw1LjJjLTMsMC01LjMtMi4zLTUuMy01LjJ2LTIuOWMtMy40LDQuOC04LjIsOC42LTE1LjcsOC42CgkJYy0xMC44LDAtMjEuNC04LjYtMjEuNC0yMy45di0wLjJjMC0xNS4zLDEwLjQtMjMuOSwyMS40LTIzLjljNy42LDAsMTIuNCwzLjgsMTUuNyw4LjF2LTIwYzAtMywyLjQtNS4yLDUuMi01LjIKCQljMywwLDUuMywyLjMsNS4zLDUuMlY0NDYuMnogTTE0OS44LDQyNy45djAuMmMwLDguOSw2LjEsMTQuOCwxMy4zLDE0LjhjNy4xLDAsMTMuNC01LjksMTMuNC0xNC44di0wLjJjMC04LjktNi4zLTE0LjgtMTMuNC0xNC44CgkJQzE1NS44LDQxMy4xLDE0OS44LDQxOC43LDE0OS44LDQyNy45eiIvPgoJPHBhdGggY2xhc3M9InN0MSIgZD0iTTIwMC4xLDQwOS43YzAtMywyLjMtNS4zLDUuMi01LjNjMywwLDUuMywyLjQsNS4zLDUuM3Y0LjdjMi40LTUuOCw3LTEwLjEsMTEuNi0xMC4xYzMuMywwLDUuMiwyLjIsNS4yLDUuMgoJCWMwLDIuOC0xLjgsNC42LTQuMiw1LjFjLTcuNSwxLjMtMTIuNyw3LjEtMTIuNywxOHYxMy42YzAsMi45LTIuNCw1LjItNS4zLDUuMmMtMi45LDAtNS4yLTIuMy01LjItNS4yVjQwOS43eiIvPgoJPHBhdGggY2xhc3M9InN0MSIgZD0iTTIzMC45LDQzNy43di0wLjJjMC0xMCw3LjgtMTQuOCwxOS0xNC44YzUuMiwwLDguOCwwLjgsMTIuNCwxLjl2LTEuMWMwLTYuNi00LTEwLTExLjQtMTAKCQljLTQsMC03LjMsMC43LTEwLjIsMS44Yy0wLjYsMC4yLTEuMSwwLjMtMS43LDAuM2MtMi40LDAtNC41LTEuOS00LjUtNC40YzAtMS45LDEuMy0zLjYsMi45LTQuMmM0LjQtMS43LDguOC0yLjcsMTQuOC0yLjcKCQljNi45LDAsMTIuMSwxLjgsMTUuMyw1LjJjMy40LDMuMyw1LDguMiw1LDE0LjJ2MjIuN2MwLDIuOS0yLjMsNS4xLTUuMiw1LjFjLTMuMSwwLTUuMi0yLjEtNS4yLTQuNXYtMS43Yy0zLjEsMy44LTcuOSw2LjctMTUsNi43CgkJQzIzOC42LDQ1MiwyMzAuOSw0NDcsMjMwLjksNDM3Ljd6IE0yNjIuNSw0MzQuNHYtMy4xYy0yLjctMS02LjMtMS44LTEwLjUtMS44Yy02LjgsMC0xMC44LDIuOS0xMC44LDcuN3YwLjJjMCw0LjUsMy45LDcsOSw3CgkJQzI1Ny4yLDQ0NC4zLDI2Mi41LDQ0MC4zLDI2Mi41LDQzNC40eiIvPgoJPHBhdGggY2xhc3M9InN0MSIgZD0iTTI4My4xLDQyNC4zYzAtMi45LDIuNC01LjIsNS4yLTUuMmgxNi40YzIuOCwwLDUuMiwyLjQsNS4yLDUuMmMwLDIuOC0yLjQsNS4yLTUuMiw1LjJoLTE2LjQKCQlDMjg1LjQsNDI5LjQsMjgzLjEsNDI3LjEsMjgzLjEsNDI0LjN6Ii8+Cgk8cGF0aCBjbGFzcz0ic3QxIiBkPSJNMzE3LjcsNDQ2Ljd2LTAuMmMwLTIsMC43LTMuMywyLjQtNS4xbDI0LjUtMjcuOWgtMjEuN2MtMi40LDAtNC40LTEuOS00LjQtNC4zYzAtMi40LDEuOS00LjQsNC40LTQuNGgzMC40CgkJYzIuOSwwLDUuMSwxLjcsNS4xLDQuMnYwLjNjMCwyLTAuNywzLjMtMi4zLDUuMWwtMjQuNiwyNy45aDIyLjdjMi40LDAsNC40LDEuOSw0LjQsNC40YzAsMi40LTEuOSw0LjMtNC40LDQuM2gtMzEuMwoJCUMzMTkuOCw0NTEsMzE3LjcsNDQ5LjIsMzE3LjcsNDQ2Ljd6Ii8+Cgk8cGF0aCBjbGFzcz0ic3QxIiBkPSJNMzg4LjcsNDUyLjFjLTEzLjQsMC0yMy43LTkuNy0yMy43LTIzLjl2LTAuMmMwLTEzLjIsOS4zLTI0LDIyLjUtMjRjMTQuNywwLDIxLjksMTIuMSwyMS45LDIyLjcKCQljMCwzLTIuMyw1LjEtNSw1LjFoLTI4LjljMS4xLDcuNSw2LjUsMTEuNywxMy4zLDExLjdjNC41LDAsNy45LTEuNiwxMC45LTRjMC44LTAuNiwxLjUtMSwyLjctMWMyLjQsMCw0LjIsMS44LDQuMiw0LjMKCQljMCwxLjMtMC42LDIuNC0xLjQsMy4yQzQwMS4yLDQ0OS43LDM5NS45LDQ1Mi4xLDM4OC43LDQ1Mi4xeiBNMzk5LjIsNDI0LjdjLTAuNy02LjgtNC43LTEyLjEtMTEuNy0xMi4xYy02LjUsMC0xMSw1LTEyLDEyLjFIMzk5LjIKCQl6Ii8+Cgk8cGF0aCBjbGFzcz0ic3QxIiBkPSJNNDIwLjQsNDA5LjdjMC0zLDIuMy01LjMsNS4yLTUuM2MzLDAsNS4zLDIuNCw1LjMsNS4zdjIuM2MzLTQuMyw3LjItOCwxNC40LThjMTAuNCwwLDE2LjQsNywxNi40LDE3LjZ2MjQuNgoJCWMwLDMtMi4zLDUuMi01LjIsNS4yYy0zLDAtNS4zLTIuMy01LjMtNS4ydi0yMS40YzAtNy4yLTMuNi0xMS4zLTkuOS0xMS4zYy02LjEsMC0xMC40LDQuMy0xMC40LDExLjR2MjEuMmMwLDMtMi40LDUuMi01LjMsNS4yCgkJYy0yLjksMC01LjItMi4zLTUuMi01LjJWNDA5Ljd6Ii8+CjwvZz4KPHBhdGggY2xhc3M9InN0MSIgZD0iTTQ3My4xLDM0MS42Yy0zLjgsMy4yLTEzLjEsMy44LTEzLjEsMy44bDMuNC0yLjFoLTNjNy01LDguNi0xNS4yLDguNi0xNS4yYy0xLjQsMi4xLTgsNi42LTEwLjcsNy45CgljLTIuNywxLjMtMTUuMSwwLjYtMTUuMSwwLjZjMy45LTEuNCw2LTMuMiw2LTMuMmgtMi41YzEuOC0xLjQsNS42LTcuNSw1LjYtNy41bC0yLjMsMC42YzMuNS01LTQuMi0xNS45LTQuMi0xNS45CgljMi41LDcuNy0zLjIsMTUuMi0zLjIsMTUuMnMtMy44LDkuMi0xMy4zLDcuM2MtOS41LTEuOC04LjEtMTEuMi02LjEtMTJjMi4xLTAuOCw0LjktNS40LDQuOS01LjRjMi4zLTYuOSwxNi4xLTYuOSwxNi4xLTYuOQoJYy03LjgtNy4xLTIwLjgtMC41LTIwLjgtMC41bDAuOC0yLjRjLTEuNywyLjEtOS44LDguMS05LjgsOC4xbDEuMS0zLjRjLTEuNiwyLjktMTAuMiwxMC42LTEwLjIsMTAuNnYtMi45CgljLTAuMyw0LjEtMTEuMyw5LjYtMTQuMiw4LjljMCwwLTQuOS0wLjUtMTAuNi02Yy01LjgtNS41LTI1LjktMTUuMy0yNS45LTE1LjNjLTE1LjctOC40LTEzLjgtMzEuMS0xMy44LTMxLjFsMCwwCgljMS45LTI0LjQtMS4yLTQ5LjMtMTAuMS03NC45YzguOC0wLjQsMTYuNCwxLjgsMTkuOSwxMi4yYzMsOC44LDguNywxMiwxOC40LDguMmMxOC42LTcuNSwzNy40LTExLjIsNTQuOSwxLjYKCWM5LjUtNy4yLDYuOC0xNS42LDQuNC0yMy44Yy0yLjQtOC4xLTkuNy02LjYtMTUuOS04LjFjLTcuNy0xLjgtMTctMy43LTIxLjktOWMtNy40LTgtMTUuNC0xMi4yLTI1LjUtMTIuOQoJYy0xNy4zLTEuMS0zMi02LjktNDIuNC0yMy4xYy0xLjMsMTMuNSw5LjcsMTguMiwxNi41LDI3Yy03LjEtMC43LTEyLjctMS0xOC4zLTEuOWMtMjYuMi00LjMtNDUuNi0xNy4zLTUzLjktNDMuNwoJYy0zLjQtMTAuNy0wLjYtMjEuOSw2LjgtMjdjMTEuNS03LjksMjMuNS01LjYsMzUuMS0wLjJDMzE5LjEsMTA0LjIsMzI4LDExMCwzMjYsMTI0Yy0xLjEsNy42LDIuMiwxMi45LDExLjIsMTMuOQoJYzIxLjUsMi4zLDQyLjEsNi44LDUzLDI4LjNjMi45LTEuNSw1LjMtMy4yLDcuMy00LjljLTQuNC0zLjYtMTMuMy0xMS4xLTE3LjMtMTVjLTAuNi0wLjYtMi4yLTEuNy00LTMuMWMtNy01LjItMTMuMy0xMC4xLTEzLjgtMTMuNAoJYy0wLjEtMC43LDAuMS0xLjQsMC41LTEuOWw0LjMsMC41Yy0zLjksMy42LDI1LjUsMjIuMiwzMi45LDMwLjNjOS40LTkuMiw2LjEtMTkuMS02LjgtMjguMWMtNi43LTQuNy0xNC40LTEwLjktMTYuNy0xOC4xCgljLTMuNS0xMS4xLTkuOC0xOC0xOC44LTI0LjNjLTYuMy00LjQtMTIuNy05LjctMTYuNS0xNi4xYy02LjYtMTEtMTEuMy0yMy4yLTE2LjYtMzQuM2MtMy45LDEzLjgsMy42LDI3LjIsNy42LDQwLjYKCWMtMjIuNi01LjEtMzAuOS0zMS4yLTU1LjktMzUuOGM4LjYsOC41LDE5LjUsMTIuNiwyMC45LDI0LjhjLTU1LjMtNi42LTYwLjktNi4zLTcyLjgsNC4yYzE1LjEtNC41LDI5LjItNi4zLDQ0LjEsMS4yCgljLTMyLjUsNS4yLTU2LjEsMjEuNC02My4xLDQyLjZjNS4zLTIuOSwxMC41LTUuOSwxNy4zLTkuN2MtMy44LDI3LjctOS44LDU0LjMsMTMuMyw3Ni44Yy0xLjYtOC40LTIuOS0xNS40LTMuNy0yMAoJYzYuNCw1LjMsMTMuNywxMS40LDIwLjgsMTcuNGMtNywzLjctMTMuOCw3LjMtMjAuNiwxMC45YzEyLjItMC43LDIzLjUtOCwzNy40LDFjLTI3LjMsNi4yLTQzLjIsMjIuOC01NC45LDQ3LjkKCWM2LjktNC44LDExLjUtOCwxNi4xLTExLjJjLTEyLjUsMjQuOS02LjEsMjguOC05LjEsNDkuMmMtNC4zLDE2LjQsMi43LTIuMy0xNC0xNS45Yy0xLjMtMS4xLTIuNy0yLTQtMi45CgljLTE2LjYtMTMuOC0xOS4xLTQwLjUtNC4xLTU0LjljLTAuNCw3LjQtMC43LDE0LjItMS4yLDIyLjVjMTUuNy0yMS43LDEyLjktNDIuNiwzLjEtNjQuM2MzLjMsMi4xLDYuNiw0LjEsMTIuNCw3LjcKCWMtOS4yLTE4LjUtMTkuNC0zMS44LTM3LjEtMzguMWM2LjYsMS44LDEzLjEsMy42LDIzLDYuM2MtMTYuMi0xNy4zLTMzLjMtMjItNTMuNy0yMS43YzcuMS00LjUsMTQuMi05LDIzLjYtMTUKCWMtMjAsMS41LTMzLjQsMTEuMy00Ni44LDIyLjFjLTEuOS0xOS02LjUtMjQuNi0yMC40LTI1LjZjOC45LDcuNyw5LjIsMTcsNC42LDI2LjJjLTQuMyw4LjktMTAuNCwxNi45LTE1LjEsMjUuNgoJYy0zLjIsNi01LjMsMTIuNi03LjcsMTljLTMsNy44LTYsMTUuNy04LjksMjMuNmMtMiw1LjQtNi4yLDExLjMtNS4yLDE2LjFjMSw0LjcsMy41LDkuMSw3LjcsMTEuM2MxLjEtNiw1LjctMTcuNCwxMS4xLTIzLjYKCWM1LTUuNyw5LjctOS43LDEwLjMtMTMuNGMwLjQtMi40LTEuMi01LjItMS4yLTUuM2w0LTEuM2MwLjYsMC40LDEuMyw0LDEuMiw2LjJjLTAuMSw0LjUtNS43LDkuNS0xMi4zLDE3CgljLTQuMyw0LjgtOC41LDE1LjItOS41LDIxLjZjMS43LDAuMywzLjcsMC40LDUuOCwwYzMuNy0xOC41LDE3LjMtMjcuNiwzMy4yLTM0LjNjOC0zLjQsMTIuNS03LjYsOC40LTE3Yy0zLjQtNy43LDAuMy0xMi42LDcuOC0xNS4zCgljMTEuNi00LjIsMjQuMiwzLjEsMjQuOCwxNi42YzAuNCwxMC0wLjksMjEuMS00LjksMzAuMWMtOS43LDIxLjktMTAuMSw0NC40LTIuMiw2My44YzAuMSwwLjIsMC4xLDAuMywwLjIsMC41CgljMC4xLDAuMywwLjIsMC42LDAuNCwwLjhjMC4zLDAuNiwwLjYsMS4zLDAuOSwxLjljMS4xLDIuNCwyLDQuNywyLjksNy4xbDAsMGMwLjIsMi41LDEsMjMtMTMuOCwzMC44YzAsMC0yMC4xLDkuOC0yNS45LDE1LjMKCWMtNS44LDUuNS0xMC42LDYtMTAuNiw2Yy0yLjksMC43LTE0LTQuNy0xNC4yLTguOWwwLDIuOWMwLDAtOC42LTcuOC0xMC4yLTEwLjZsMS4xLDMuNGMwLDAtOC4xLTYtOS44LTguMWwwLjgsMi40CgljMCwwLTEzLTYuNi0yMC44LDAuNWMwLDAsMTMuOCwwLDE2LjEsNi45YzAsMCwyLjksNC42LDQuOSw1LjRjMi4xLDAuOCwzLjQsMTAuMi02LjEsMTJjLTkuNSwxLjgtMTMuMy03LjMtMTMuMy03LjMKCXMtNS43LTcuNi0zLjItMTUuMmMwLDAtNy44LDEwLjktNC4yLDE1LjlsLTIuMy0wLjZjMCwwLDMuOCw2LjEsNS42LDcuNWgtMi41YzAsMCwyLjEsMS44LDYsMy4yYzAsMC0xMi40LDAuNy0xNS4xLTAuNgoJYy0yLjctMS4zLTkuMy01LjgtMTAuNy03LjljMCwwLDEuNiwxMC4yLDguNiwxNS4yaC0zbDMuNCwyLjFjMCwwLTkuMy0wLjYtMTMuMS0zLjhjMCwwLDUsNy4yLDExLjEsOC4xbC0yLjksMS40aDIwLjVsLTYuOSwyLjMKCWMwLDAsMjcuNSwwLjIsMzMuMy0xLjFsLTcuMSw0LjJjMCwwLDMwLjYtMC41LDQxLjYtOC45bC00LDYuMWwyOC4zLTE5LjdjLTAuNiwzLTIsOC4zLTAuNiw5LjZjMS43LTAuMiwxMC44LTE2LjYsMTUuNS0xOC44CgljMy4zLTEuNSwxMC45LTYuMiwxNy4xLDguMWwwLDBjMC4xLDAuMywwLjIsMC41LDAuMywwLjdjMi4zLDE1LjksMy44LDI1LjksMTAuMiwyNS45YzIwLjUsMCwxMTMuMywwLDEzMS42LDAKCWM0LjYtOC44LDguNy0xNy42LDEyLjEtMjYuNGMwLTAuMSwwLjEtMC4xLDAuMS0wLjJjNi4yLTE0LjQsMTMuOC05LjcsMTcuMS04LjJjNC43LDIuMiwxMy44LDE4LjUsMTUuNSwxOC44YzEuNC0xLjQsMC02LjYtMC42LTkuNgoJbDI4LjMsMTkuN2wtNC02LjFjMTEsOC41LDQxLjYsOC45LDQxLjYsOC45bC03LjEtNC4yYzUuOCwxLjQsMzMuMywxLjEsMzMuMywxLjFsLTYuOS0yLjNoMjAuNWwtMi45LTEuNAoJQzQ2OCwzNDguOCw0NzMuMSwzNDEuNiw0NzMuMSwzNDEuNnogTTM4NC4xLDE4NGwtMiwyLjdjLTUuOS00LjUtOS41LTIuMS05LjctMmwtMS41LTNDMzcxLjEsMTgxLjUsMzc3LjEsMTc4LjMsMzg0LjEsMTg0egoJIE0zNjcuMSwxMTEuMmwtMi45LDEuMmMtMS44LTQuMy04LTMuMi04LjEtMy4ybC0wLjItMy4yQzM1Ni4zLDEwNi4xLDM2NC43LDEwNS41LDM2Ny4xLDExMS4yeiBNMTAwLjcsMTU3LjhsLTMuMi0xLjIKCWMxLjUtMi42LDEuNi01LjksMTEuNC03bDAuMywzQzEwMywxNTMuMywxMDAuNSwxNTcsMTAwLjcsMTU3Ljh6IE0yOTIuMiwyODYuOGMtNi4yLTcuMi0xMi40LTE0LjItMTguNS0yMS40CgljLTE0LjMtMTctMTIuOC0zMS42LDQuNi00NS41YzMuNy0yLjksNy43LTUuNCwxMS41LThDMzA4LjUsMjMxLjYsMzA5LjMsMjYxLjUsMjkyLjIsMjg2Ljh6Ii8+Cjwvc3ZnPgo=
 [cloudflare-badge]: https://img.shields.io/badge/Docs-pages-gray.svg?style=flat&logo=cloudflare&color=F26722
@@
 [license-link]: https://spdx.org/licenses/
+[pyro-ppl-link]: https://pyro.ai
+[pyro-ppl-badge]: https://img.shields.io/badge/Pyro-PPL-gray.svg?color=F16722&logo=data:image/svg%2bxml;base64,PHN2ZyBpZD0iTGF5ZXJfMSIgZGF0YS1uYW1lPSJMYXllciAxIiB4bWxucz0iaHR0cDovL3d3dy53My5vcmcvMjAwMC9zdmciIHZpZXdCb3g9IjAgMCA1NDEgNDY0Ij48ZGVmcz48c3R5bGU+LmNscy0xe2ZpbGw6I2VjMWYyNDt9LmNscy0ye2ZpbGw6I2YxNjcyMjt9LmNscy0ze2ZpbGw6I2ZkY2MwODt9PC9zdHlsZT48L2RlZnM+PHRpdGxlPnB5cm8taG9yaXpvbnRhbC1jb2xvcjwvdGl0bGU+PHBhdGggY2xhc3M9ImNscy0xIiBkPSJNMTA0Ljk2MjQ0LDcuMDM1ODlxNjAuMDExNTUuMDg0MjcsMTIwLjAxNDQuMDM3ODJhODMuODQ2LDgzLjg0NiwwLDAsMSwxNy43NTk0NCwxLjg4NDIzYzIuNzEwMzMuNjE5MTUsNi4xMzkzMS43OTQzOSw2LjczOTExLDQuNjk1NTFhMTI5LjcyMDY0LDEyOS43MjA2NCwwLDAsMC0xOS42NDU1Ny4wMzY5Yy0yMi44MzU0OCwxLjM3Mi0zOS44MjYzNiwxNi41NzctMzkuOTAzNjQsMzkuNDgzNjgtLjIwNDY4LDYwLjY0MjI1LS45NDY4MSwxMjEuMzAyODkuMjU4MTUsMTgxLjkyMTczLjY3MDM4LDMzLjcyNTMxLTE0LjUzNzQ2LDU2Ljc4MDg1LTQxLjYyNTYsNzMuNTg5NjYtMjEuMTcxNjEsMTMuMTM3NDUtNDQuMjEyMTMsMTkuNjk3MzUtNjkuMjM4NjgsMTQuMDY3NDctMjcuMTQzODMtNi4xMDYxNy0zOS41OTg2Mi0yOS42NS0yOS4zMTI4Ni01NS41MDUxOSwxMC4xMzI2LTI1LjQ3MDIxLDMwLjAyNzA3LTQxLjEwMTgsNTQuNjA1LTUxLjM0MDU2LDEyLjQ4ODE2LTUuMjAyMzYsMjQuNzU3NC0xMC4zNjYzNiwzNC4xMTI3LTIwLjgxMzc5LDE3LjQxNzI0LTE5LjQ1MDYsMjQuMDQzMTItNDIuNjk0NCwyNC4zNjQzOC02OC4wOTAyMS4zMTI1OS0yNC43MTE4NC4wODgxMy00OS40MzA3OS4wNjE1NC03NC4xNDY2My0uMDIxMDctMTkuNTM0ODgtMTMuMTA1NjQtMzUuNzM2OTMtMzIuMzY5LTM4LjQwMDI5LTE3LjcxOC0yLjQ0OTcxLTM1LjA2MTYtLjUxNy01MC42MDExMyw5LjU1M0EzOS45MDUxNSwzOS45MDUxNSwwLDAsMCw3Mi40NzkyNCwzMC4xNjRjLTUuMzAwNDMsNS43NDQxOS00Ljc3OTQzLDcuOTU3LDIuMTUxNzgsMTEuNzg5ODEsMTguNTkxNTIsMTAuMjgwNTYsMjguMDM2NjUsMzEuNDMwOTMsMjMuNDk3LDUyLjYxNjM1LTQuMjMsMTkuNzQwNDYtMjIuMDU4MjksMzUuMjE3MjgtNDIuNjE0ODgsMzYuOTk0NjktMjAuNDk0MjksMS43NzItNDAuNzg2NDEtMTAuMzI1MjEtNDguNTUxMjktMjguOTQzODEtOS4xNzM1NC0yMS45OTYzNS42MDA1OS01NS42NDYsMjAuODcxNzYtNzEuODAwMDgsMTYuNjE0MjgtMTMuMjQsMzYuMDQ1MzItMTkuNDgwMjMsNTYuNjU2NzctMjIuODE3NDUiLz48cGF0aCBjbGFzcz0iY2xzLTIiIGQ9Ik0zMDYuODA3LDcuMDU4NTZjMzEuNzM1Mi4wMjEyOSw2My4zMTIwNy0uODI3MzgsOTUuMDQ3MjctLjg0ODY4LDEuNzQxLS4wMDExNiwyLjc5Mzg1LjA3NjM5LDUuMzE1NzMuMDc2MzZoNC41ODc4M2MyLjYwODQzLDIuMDY2NCw2LjE5NjM4LDIuMTAzLDguNzc3MTYsNC4xODkzLDIuOTc0MzQsMi40MDQzNSw1LjkxNCwzLjE4MjE5LDIuMjQ2NzcsNC4wNDk0OS05LjAzMzMtLjg2NjQ5LTIxLjE5NTA3LjczNzU2LTI5LjQzLDUuMDg3MkMzODEuNDc5LDI1Ljg4MzQyLDM3NS4zODgxMiwzNS43NjEsMzc0LjM4MDM0LDQ4Ljk4NDczcS0uMDU4MTQuNzYyMjUtLjA1OTg0LDEuNTI3NzdjLS4wODMyLDc3LjQ3NTkzLS4yMDcxMywxNTQuOTUyMzcuMDg0NzEsMjMyLjQyNzQ2LjA3Njk0LDIwLjQzMDE1LDE2LjU2MzIsMzUuNjQ4NzIsMzcuMTIxMTksMzUuODI1NDEsMi4yOS0uMTI1NjQsOS41NTMwNy0uNzIzNTcsMTEuODM3MDktLjUxNjI5LDQuMjYyNTUsMS4wOTYwNy0uOTQwNjQsMi42OTc2LTQuOTUyMzYsNS41MDM4Ny0zLjk1NTE0LDIuNzY2Ny04LjcyNTEzLDIuNDM0LTEzLjIyOTUyLDIuODM0MzgtMy43ODk5Mi4zMzY4Ni03LjYyNTg5LS4wNjQtMTEuNDM4NDIuMTY5YTEwNi42MTE0MSwxMDYuNjExNDEsMCwwLDEtMTUuNTg1NDUtMS4wMzc0OGMtMjAuNDc3NzIsMi42Njk0MS0zOS4yMDY0OC0zLjA5MTIxLTU2LjE5NjYyLTEzLjE4MzIyLTI0LjczMTY2LTE0LjY5MDMxLTM3LjA4MTMzLTM2LjkwNjYyLTM2LjQ0NTU5LTY1Ljk1NjgyLS4xNzE0MS0zLjI5MTc2LS40OTA0MS02LjU4MzQxLS40OTIxOC05Ljg3NTMyLS4wMzA4Ni01Ny4wMDAyOC4wNDktMTE0LjAwMDctLjA1NTY0LTE3MS4wMDA3OS0uMDQ2ODctMjUuNjA1ODktMTQuMDc2LTQ1LjA4MTQ2LTM1LjQ5MjMxLTUyLjA0OTMxLTcuMTE1NDItMy44MDc3My0xNS4xMDM4My00LjkwMi0yMi43NzY2NC02LjU2MzIzIi8+PHBhdGggY2xhc3M9ImNscy0zIiBkPSJNNTI0Ljg4OSw5Ni41OTYzNkM1MTEuODQwNDIsNDYuODYzLDQ3OS4zNzgsMTYuODQzMjksNDI3Ljk4ODEsOC4zMDEyM2MtMy4zNjQzNS0uNTU5MjEtMi45MDg1MS0uODkwNDMtOS43NzM1My0xLjYyMjQyLTEuNTIwNi0uMTYyMTItNi40NTY1Ny0uMzkyNTQtNi40NTY1Ny0uMzkyNTRoMGMxNy4yMTU0NSw4LjMwOTkyLDI3LjgzNzI4LDI3LjM3ODIsMjcuOTc3NjMsNDYuNTI4N3EuODM0NTgsMTEzLjg3MjY5LjAzMDUxLDIyNy43NDM1MmMtLjA5MTI3LDE0LjQ1ODQ3LTYuMTg3MjIsMjkuMDIxODktMTkuNzY1NjgsMzYuNTczMzRsLTI2LjI1NzE0LDkuNjI0NTdjMjAuOTUyNTEuNjYzNDUsNDEuNzEwMjkuNTM0LDYwLjk1NzU4LTkuODMwNjksMzEuNDk5ODQtMTYuOTYyNjEsNTIuMjQxMDYtNDMuMzE1MTIsNjUuNDc0NjctNzUuNzc5MTNDNTM5LjUyNzMxLDE5My42NzQ0OCw1MzcuNTc1MSwxNDQuOTQ4LDUyNC44ODksOTYuNTk2MzZaIi8+PHBhdGggY2xhc3M9ImNscy0zIiBkPSJNMzI4LjEyMjg5LDQ1OC4xNzM4OGMtMjMuOTEwNC0uMDgyOTEtNDcuODIxMjItLjExNy03MS43MzA2Mi0uMzE5MjQtMi41OTAxOC0uMDIxODUtNS42MjMyNywxLjM1NjE0LTcuNzg2NS0xLjQwOTc5LDE3LjU5MzgxLTMuMjkzODYsMjYuNzA4NzQtMTUuMzM1NzYsMzIuMDM0NjQtMzEuMTkyLDQuMDk0NTctMTIuMTkwMzcsNC45NDQ4OS0yNC44MzkyMyw0Ljk0Njg3LTM3LjU3OTgzcS4wMTAzLTY2Ljg2OTE2LjAwNzIzLTEzMy43MzgzYy4wMDEtMi43OTI1NC0uMjQ3NS01LjU4NS0uMTM2NjktOC4zNzcwOSwxLjIwNjc2LDEzLjc3NTIxLDQuODczNDgsMjYuOTk3OTIsMTIuMDk1ODMsMzguOTIzNjEsMTQuMjU0LDIzLjUzNjM1LDM2Ljc4ODYsMzMuOTQ0ODksNjIuNjMzNDIsMzguODE1MjgsNS45MzI1LDEuMTE4LDExLjk3OSwxLjYzMTIsMTcuOTcwNzcsMi40MjIzLTQuNTk4MDkuODM4NjMtMy4yMTgwNSw0LjUwNTI1LTMuMjUzLDcuMjcxMzEtLjI2MzMxLDIwLjc5ODQ5LS4zNzM1Nyw0MS41OTg4NC0uNjA5NDEsNjIuMzk3NzlhMTAxLjIxNzgxLDEwMS4yMTc4MSwwLDAsMS01LjIyMTQsMzEuMzYyNjFDMzYyLjIwMzE2LDQ0Ny4xMjMyNSwzNDkuNjYxMjgsNDU2LjY4MTM5LDMyOC4xMjI4OSw0NTguMTczODhaIi8+PC9zdmc+
+[scvi-tools-link]: https://scvi-tools.org
+[scvi-tools-badge]: https://img.shields.io/badge/scvi-tools-gray.svg?color=FCC64C&logo=data:image/svg%2bxml;base64,PD94bWwgdmVyc2lvbj0iMS4wIiBlbmNvZGluZz0idXRmLTgiPz4KPCEtLSBHZW5lcmF0b3I6IEFkb2JlIElsbHVzdHJhdG9yIDI0LjEuMCwgU1ZHIEV4cG9ydCBQbHVnLUluIC4gU1ZHIFZlcnNpb246IDYuMDAgQnVpbGQgMCkgIC0tPgo8c3ZnIHZlcnNpb249IjEuMSIgaWQ9IkxheWVyXzEiIHhtbG5zPSJodHRwOi8vd3d3LnczLm9yZy8yMDAwL3N2ZyIgeG1sbnM6eGxpbms9Imh0dHA6Ly93d3cudzMub3JnLzE5OTkveGxpbmsiIHg9IjBweCIgeT0iMHB4IgoJIHZpZXdCb3g9IjAgMCAyMS43IDIyLjgiIHN0eWxlPSJlbmFibGUtYmFja2dyb3VuZDpuZXcgMCAwIDIxLjcgMjIuODsiIHhtbDpzcGFjZT0icHJlc2VydmUiPgo8c3R5bGUgdHlwZT0idGV4dC9jc3MiPgoJLnN0MHtvcGFjaXR5OjAuNzc7ZmlsbDojRkJCNTE2O3N0cm9rZTojQzM4MzJCO3N0cm9rZS13aWR0aDowLjU7c3Ryb2tlLW1pdGVybGltaXQ6MTA7ZW5hYmxlLWJhY2tncm91bmQ6bmV3ICAgIDt9Cgkuc3Qxe29wYWNpdHk6MC43NTtmaWxsOiNDMzgzMkI7c3Ryb2tlOiNBNjU3MkE7c3Ryb2tlLXdpZHRoOjAuNzU7c3Ryb2tlLW1pdGVybGltaXQ6MTA7ZW5hYmxlLWJhY2tncm91bmQ6bmV3ICAgIDt9Cjwvc3R5bGU+CjxwYXRoIGNsYXNzPSJzdDAiIGQ9Ik0yMS40LDUuMWMwLjEsMS44LTEuMiwyLjEtMS43LDQuNGMtMC44LDMuOCwyLDUuNiwwLjksNy42cy0zLjUsMC40LTguMiwyLjZjLTMuNCwxLjUtMy45LDMuMS01LjksMi44CglzLTMuNC0yLjMtNC41LTRjLTAuOC0xLjMtMS45LTIuOS0xLjctNC44YzAuMS0xLjMsMC43LTEuOCwxLjYtMy4zYzIuOS00LjcsMS42LTYuNywzLjYtNy44czMuNiwwLjksNy0wLjNjMi4yLTAuOCwyLjQtMiw0LTIKCUMxOC44LDAuMiwyMS4yLDIuNywyMS40LDUuMXoiLz4KPGVsbGlwc2UgdHJhbnNmb3JtPSJtYXRyaXgoMC44MDkxIC0wLjU4NzYgMC41ODc2IDAuODA5MSAtNC4yNjIgOC44ODIxKSIgY2xhc3M9InN0MSIgY3g9IjExLjUiIGN5PSIxMSIgcng9IjQiIHJ5PSI0Ii8+Cjwvc3ZnPgo=
 [tuple-badge]: https://img.shields.io/badge/Tuple%20❤️%20OSS-5A67D8?logo=tuple

### 2024-03-29 nbs/index.qmd
fix(docs): update badges Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-03-25 nbs/index.qmd
@@
 |         |                                                                                                                                                  |
-| ------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
-| CI/CD   | [![CID][cid-badge]][cid-link] [![CML][cml-badge]][cml-link] [![pre-commit.ci status][precommit-badge]][precommit-link]                                                 |
+| ------- | ------------------------------------------------------------------------------------------------------------------------------------------------ |
+| CI/CD   | [![CID][cid-badge]][cid-link] [![codecov][codecov-badge]][codecov-link] [![pre-commit.ci status][precommit-badge]][precommit-link]               |
 | Docs    | [![Cloudflare Pages][cloudflare-badge]][cloudflare-link] [![Preprint][preprint-badge]][preprint-link]                                            |
-| Package | [![PyPI - Version][pypi-badge]][pypi-link] [![Anaconda badge][anaconda-badge]][anaconda-link] [![Docker image][docker-badge]][docker-link]                             |
-| Meta    | [![codecov][codecov-badge]][codecov-link] [![code style][black-badge]][black-link] [![License - MIT][license-badge]][license-link] [![Tuple][tuple-badge]][tuple-link] |
+| Package | [![PyPI - Version][pypi-badge]][pypi-link] [![Conda-forge badge][conda-forge-badge]][anaconda-link] [![Docker image][docker-badge]][docker-link] |
+| Meta    | [![pyro-ppl-badge]][pyro-ppl-link] [![flyte-badge]][flyte-link] [![hydra-zen-badge]][hydra-zen-link] [![scvi-tools-badge]][scvi-tools-link]      |
+|         | [![bear-badge]][bear-link] [![black-badge]][black-link] [![License][license-badge]][license-link] [![Tuple][tuple-badge]][tuple-link]            |
 
+[bear-badge]: https://raw.githubusercontent.com/beartype/beartype-assets/main/badge/bear-ified.svg
+[bear-link]: https://beartype.readthedocs.io
 [cid-badge]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cid.yaml/badge.svg?branch=master
 [cid-link]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cid.yaml
-[cml-badge]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cml.yml/badge.svg
-[cml-link]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cml.yml
 [precommit-badge]: https://results.pre-commit.ci/badge/github/pinellolab/pyrovelocity/main.svg
-[precommit-link]: https://results.pre-commit.ci/latest/github/pinellolab/pyrovelocity/master
-[cloudflare-badge]: https://img.shields.io/badge/Docs-pages-F38020?style=flat&logo=cloudflare&labelColor=black
+[precommit-link]: https://results.pre-commit.ci/latest/github/pinellolab/pyrovelocity/main
+[flyte-badge]: https://storage.googleapis.com/pyrovelocity/badges/flyte-carrier.svg
+[flyte-link]: https://docs.flyte.org
+[hydra-zen-badge]: https://storage.googleapis.com/pyrovelocity/badges/Hydra-zen.svg
+[hydra-zen-link]: https://mit-ll-responsible-ai.github.io/hydra-zen/
+[cloudflare-badge]: https://img.shields.io/badge/Docs-pages-gray.svg?style=flat&logo=cloudflare&color=F26722
 [cloudflare-link]: https://docs.pyrovelocity.net
@@
 [black-link]: https://github.com/psf/black
-[license-badge]: https://img.shields.io/badge/license-AGPL%203-purple
+[license-badge]: https://img.shields.io/badge/license-AGPL%203-gray.svg?color=F26722
 [license-link]: https://spdx.org/licenses/
+[pyro-ppl-badge]: https://storage.googleapis.com/pyrovelocity/badges/Pyro-PPL.svg
+[pyro-ppl-link]: https://pyro.ai
+[scvi-tools-badge]: https://storage.googleapis.com/pyrovelocity/badges/scvi-tools.svg
+[scvi-tools-link]: https://scvi-tools.org
 [tuple-badge]: https://img.shields.io/badge/Tuple%20❤️%20OSS-5A67D8?logo=tuple

### 2024-04-05 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): initialize concepts page on inference in dynamical systems Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Probabilistic inference in dynamical systems
toc: true
number-sections: true
highlight-style: pygments
csl: ../../bibstyle.csl
lightbox: auto
format:
#   nature-pdf:
#     natbibstyle: sn-nature
#     # classoption:
#     #   - lineno
#     cite-method: citeproc
#     keep-tex: true
  html:
    html-math-method: mathjax
    include-in-header: ../../mathjax.html
#     mainfont: Latin Modern Sans
#     code-fold: false
#     embed-resources: true
#   docx: default
  ipynb: default
format-links: [ipynb]
execute: 
  freeze: true
  eval: true
  warning: false
  error: false
  cache: true
  # keep-ipynb: true
  # keep-md: true
author:
  - name: Pyrovelocity Team
    email: team@pyrovelocity.net
abstract: |
  This notebook provides a discussion and demonstration of one way to proceed in
  performing probabilistic inference in dynamical systems. In particular we prepare 
  and simulate a deterministic dynamical system and then perform probabilistic 
  inference on the latent variables of the system. This illustrates how to combine
  these two approaches, which is essential for understanding Pyro-Velocity.
keywords: [single-cell transcriptomics, probabilistic modeling, dynamical systems, RNA velocity]
bibliography: ../../references.bib
jupyter:
  jupytext:
    cell_metadata_filter: all
    cell_metadata_json: true
    notebook_metadata_filter: all
    text_representation:
      extension: .qmd
      format_name: quarto
      format_version: 1.0
      jupytext_version: 1.16.0
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
  language_info:
    name: python
  rise:
    scroll: true
    theme: black
  toc-autonumbering: true
  toc-showcode: false
  toc-showmarkdowntxt: false
---

## Model

### Description

Here we describe the model we will use to demonstrate probabilistic 
inference in dynamical systems. We begin with a candidate 
[effective theory](https://en.wikipedia.org/wiki/Effective_theory) 
[@Wells2012-qf;@Alon2019-ah] of gene transcription and splicing following 
[@Zeisel2011-nz;@La_Manno2018-lj;@Bergen2020-pj] for the model and 
[@Cantwell2002-dv] for the presentation of the analysis. In particular we show in 
@tbl-transcription-splicing-degradation the variables and parameters of the model
together with their units and rough order of magnitude estimates for their values 
[@Phillips2012-mi].

| Symbol   | Description                 | Units                         | $O(-)$ Estimate | Discussion                                                                          |
|----------|-----------------------------|-------------------------------|-----------------|-------------------------------------------------------------------------------------|
| $u$      | Concentration of pre-mRNA   | molecules/cell                | $10^0 - 10^4$   | Wide range reflecting low to high gene expression levels.                           |
| $s$      | Concentration of mRNA       | molecules/cell                | $10^0 - 10^5$   | Similar to $u$; depends on gene expression and stability of mRNA.                   |
| $t$      | Time                        | seconds ($s$) to hours ($hr$) | $10s$ - $24hr$  | Depends on the duration of the experimental observation.                            |
| $\alpha$ | Production rate of pre-mRNA | molecules/(cell·hr)           | $1 - 10^3$      | Based on variable gene transcription rates.                                         |
| $\beta$  | Splicing rate of pre-mRNA   | $hr^{-1}$                     | $0.1 - 1$       | Pre-mRNA to mRNA splicing rates, assuming rapid processing within the nucleus.      |
| $\gamma$ | Degradation rate of mRNA    | $hr^{-1}$                     | $0.1 - 1$       | mRNA half-lives can vary widely, but many are in the range of a few hours to a day. |

: Variables and parameters of the transcription-splicing-degradation model of [@Zeisel2011-nz]. {#tbl-transcription-splicing-degradation}

Given the state variables representing concentrations of pre-mRNA, $u$, 
and mRNA, $s$, we have the following ordinary differential equations

\begin{align}
\frac{du}{dt} & = \alpha - \beta u  \label{eq-dudt} \\
\frac{ds}{dt} & = \beta u - \gamma s  \label{eq-dsdt}
\end{align}

\eqref{eq-dudt} and \eqref{eq-dsdt} are usually presented with 
the concept that the parameter $\alpha$ could better account for the external 
inputs to the regulation of transcription of the gene if it were allowed to be 
a function of time, $\alpha(t)$. We will return to this point later, but 
for now we will assume that $\alpha$ and other parameters are constant.

### Dimensional analysis

It is common to perform a dimensional analysis of the model to ensure the
units of the variables and parameters are consistent and reduce the total number
of parameters by the number of dimensions. Intuitively, this folds or maps parameters
to equivalence classes of similar dynamics.  For reference, the procedure
we outline is also commonly referred to as Buckingham's $\Pi$ theorem.

@tbl-transcription-splicing-degradation together with (\ref{eq-dudt}-\ref{eq-dsdt})
contain essentially all the information we require to get started.

### 2024-04-06 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): update dimensional analysis Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 @tbl-transcription-splicing-degradation together with (\ref{eq-dudt}-\ref{eq-dsdt})
-contain essentially all the information we require to get started.
+contain essentially all the information we require to get started. The dimensions of the variables and parameters together with dimensionless ones are provided in @tbl-parameter-dimensions.
+
+| Dimensioned Parameter | Relation to Rescaling Parameters | Fundamental Units        |
+|-----------------------|--------------------------------------|--------------------------|
+| Production Rate ($\alpha$) | $\alpha = U_0 \beta$             | molecules/(cell·time)    |
+| Splicing Rate ($\beta$)    | Reference Scale for $t^*$        | $1/\text{time}$        |
+| Degradation Rate ($\gamma$)| $\gamma = \gamma^* \beta$        | $1/\text{time}$        |
+| Pre-mRNA Concentration ($u$)| $u = u^* U_0$                    | molecules/cell           |
+| mRNA Concentration ($s$)    | $s = s^* U_0$                    | molecules/cell           |
+| Time ($t$)                  | $t = t^* / \beta$                | time                     |
+
+: Variables and parameters of the transcription-splicing-degradation model together with their fundamental units and their relations to dimensionless parameters. {#tbl-parameter-dimensions}
+
+Note that molecules are dimensionless numbers while cells have units associated to their volume (i.e.
+the cube of a distance or length $L^3$).
+Usually we would just write volume but we retain cell to associate to the object that determines the relevant volume in this case.
+The dimensionless parameters are defined in @tbl-dimensionless-parameters.
+
+| Dimensionless Variables and Parameters | Definition                     | Intuitive Description                                      |
+|-------------------------|--------------------------------|------------------------------------------------------------|
+| $u^{\ast}$ , $s^{\ast}$ | $u / U_0$, $s / U_0$             | Characteristic Scale of (pre-)mRNA concentration based on the balance between production and splicing rates. |
+| $t^*$                 | $\beta t$                    | Characteristic time scale relative to the splicing rate. |
+| $\gamma^*$            | $\gamma / \beta$             | Relative degradation rate, comparing the degradation rate of mRNA to the splicing rate, indicating the stability or turnover rate of mRNA relative to its production/splicing.
+|
+
+: Dimensionless Parameters, definitions, and intuitive Descriptions {#tbl-dimensionless-parameters}
+
+Combining variables and parameters we have six and we have two fundamental dimensions (time and length) we can eliminate to arrive at four essential variables and parameters.
+To derive these, we begin declaring the equations in python using [sympy](https://www.sympy.org/) and substitute the dimensionless variables and parameters to obtain the dimensionless equations.
+
+```{python}
+# | label: transcription-splicing-degradation-equations
+from sympy import (
+    diff,
+    Eq,
+    Function,
+    init_printing,
+    solve,
+    symbols,
+)
+from IPython.display import display
+
+init_printing(use_latex=True)
+
+alpha, beta, gamma, gamma_star = symbols("alpha beta gamma gamma_star")
+U0, T0 = symbols("U0 T0")
+t, t_star = symbols(
+    "t t_star",
+)
+
+u = Function("u")(t)
+s = Function("s")(t)
+u_star = Function("u_star")(t_star)
+s_star = Function("s_star")(t_star)
+
+du_dt = Eq(diff(u, t), alpha - beta * u)
+ds_dt = Eq(diff(s, t), beta * u - gamma * s)
+display(du_dt)
+display(ds_dt)
+```
+
+and define dimensionelss variables and parameters as
+
+```{python}
+# | label: transcription-splicing-degradation-dimensionless
+U0_eq = Eq(U0, alpha / beta)
+T0_eq = Eq(T0, 1 / beta)
+
+u_dimless = Eq(u_star, u / U0)
+s_dimless = Eq(s_star, s / U0)
+t_dimless = Eq(t_star, beta * t)
+```
+
+Now we change the variables in the equations to the dimensionless ones
+
+```{python}
+du_dt_dimless = Eq(
+    beta * diff(u_star * alpha / beta, t_star),
+    du_dt.rhs.subs(
+        {
+            u: u_star * alpha / beta,
+            s: s_star * alpha / beta,
+            t: t_star / beta,
+            gamma: gamma_star * beta,
+        }
+    ),
+).simplify()
+du_dt_dimless
+```
+
+and
+
+```{python}
+ds_dt_dimless = Eq(
+    beta * diff(s_star * alpha / beta, t_star),
+    ds_dt.rhs.subs(
+        {
+            u: u_star * alpha / beta,
+            s: s_star * alpha / beta,
+            t: t_star / beta,
+            gamma: gamma_star * beta,
+        }
+    ),
+).simplify()
+
+ds_dt_dimless
+```
+
+We see that we correctly arrive at the equivalent dimensionless system of equations
+
+\begin{align}
+\frac{du^{\ast}}{dt^{\ast}} & = 1 - u^{\ast}  \label{eq-dustardtstar} \\
+\frac{ds^{\ast}}{dt^{\ast}} & = u^{\ast} - \gamma^{\ast} s^{\ast}  \label{eq-dsstardtstar}
+\end{align}
+
+in \eqref{eq-dustardtstar} and \eqref{eq-dsstardtstar}, which contains the $6 - 2 = 4$ variables and parameters: $u^{\ast}, s^{\ast}, t^{\ast}, \gamma^{\ast}$.
+It is trivial to solve this system analytically, but in more general cases this will not be possible.
+We can now proceed to simulate the system to characterize the possible solution sets and then perform probabilistic inference on the latent variables.

### 2024-04-06 nbs/concepts/inferenceindynamicalsystems/index.qmd
fix(nbs): add manual derivation of analytical solution Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-Note that molecules are dimensionless numbers while cells have units associated to their volume (i.e.
-the cube of a distance or length $L^3$).
-Usually we would just write volume but we retain cell to associate to the object that determines the relevant volume in this case.
-The dimensionless parameters are defined in @tbl-dimensionless-parameters.
+Note that molecules are dimensionless numbers while cells have units associated 
+to their volume (i.e. the cube of a distance or length $L^3$). Usually we would 
+just write volume but we retain cell to associate to the object that determines 
+the relevant volume in this case. The dimensionless parameters are defined in 
+@tbl-dimensionless-parameters.
 
@@
 
-Combining variables and parameters we have six and we have two fundamental dimensions (time and length) we can eliminate to arrive at four essential variables and parameters.
-To derive these, we begin declaring the equations in python using [sympy](https://www.sympy.org/) and substitute the dimensionless variables and parameters to obtain the dimensionless equations.
+Combining variables and parameters we have six and we have two fundamental 
+dimensions (time and length) we can eliminate to arrive at four essential 
+variables and parameters. To derive these, we begin declaring the equations 
+in python using [sympy](https://www.sympy.org/) and substitute the 
+dimensionless variables and parameters to obtain the dimensionless equations.
 
@@
 
-in \eqref{eq-dustardtstar} and \eqref{eq-dsstardtstar}, which contains the $6 - 2 = 4$ variables and parameters: $u^{\ast}, s^{\ast}, t^{\ast}, \gamma^{\ast}$.
-It is trivial to solve this system analytically, but in more general cases this will not be possible.
-We can now proceed to simulate the system to characterize the possible solution sets and then perform probabilistic inference on the latent variables.
+in \eqref{eq-dustardtstar} and \eqref{eq-dsstardtstar}, which contains the 
+$6 - 2 = 4$ variables and parameters: $u^{\ast}, s^{\ast}, t^{\ast}, \gamma^{\ast}$.
+
+We'll solve the dimensionless system analytically, focusing on what can be 
+understood about the system's dynamics.
+Given generic initial conditions $u^{\ast}(0) = u^{\ast}_0$ and $s^{\ast}(0) = s^{\ast}_0$, the 
+first equation
+
+$$
+\frac{du^{\ast}}{dt^{\ast}} = 1 - u^{\ast},
+$$
+
+can be solved using separation of variables
+
+$$
+u^{\ast}(t^{\ast}) = C_1 e^{-t^{\ast}} + 1.
+$$
+
+$C_1$ is the constant determined by initial conditions, which can be shown to be $C_1 = u^{\ast}_0 - 1$, giving us
+
+$$
+u^{\ast}(t^{\ast}) = (u^{\ast}_0 - 1) e^{-t^{\ast}} + 1.
+$$
+
+This equation describes how the concentration of pre-mRNA, in units of the 
+characteristic concentration scale given by $\alpha/\beta$, evolves over the 
+timescale characterized by the splicing rate. The term $e^{-t^{\ast}}$ 
+indicates that any deviation of $u^{\ast}$ from its steady state $1$ decays 
+exponentially over time.
+
+Now, we substitute the expression for $u^{\ast}(t^{\ast})$ into the 
+differential equation for mRNA concentration
+
+$$
+\frac{ds^{\ast}}{dt^{\ast}} = u^{\ast} - \gamma^{\ast} s^{\ast}
+$$
+
+which gives
+
+$$
+\frac{ds^{\ast}}{dt^{\ast}} = [(u^{\ast}_0 - 1) e^{-t^{\ast}} + 1] - \gamma^{\ast} s^{\ast}
+$$
+
+This equation is a non-homogeneous first-order linear ODE for $s^{\ast}$. 
+To solve it, we can use an integrating factor $e^{\gamma^{\ast} t^{\ast}}$, 
+which gives
+
+$$
+e^{\gamma^{\ast} t^{\ast}} \frac{ds^{\ast}}{dt^{\ast}} + \gamma^{\ast} e^{\gamma^{\ast} t^{\ast}} s^{\ast} = (u^{\ast}_0 - 1) e^{(\gamma^{\ast}-1) t^{\ast}} + e^{\gamma^{\ast} t^{\ast}}
+$$
+
+Integrating both sides with respect to $t^{\ast}$ yields $s^{\ast}(t^{\ast})$
+
+$$
+s^{\ast}(t^{\ast}) = C_2 e^{-\gamma^{\ast} t^{\ast}} + \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1} e^{-t^{\ast}} + \frac{1}{\gamma^{\ast}}
+$$
+
+where $C_2$ is another integration constant determined by initial conditions.
+Incorporating the initial condition  $s^{\ast}(0) = s^{\ast}_0$ into the solution for $s^{\ast}(t^{\ast})$ to determine $C_2$
+
+$$
+C_2 = s^{\ast}_0 - \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1} - \frac{1}{\gamma^{\ast}}.
+$$
+
+With $C_2$ determined, we define
+
+$$
+\xi = \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1},
+$$
+
+so that the complete solutions for the dimensionless system are given by
+
+\begin{align}
+u^{\ast}(t^{\ast}) & = \underbrace{1}_{\text{Steady State}} + \underbrace{(u^{\ast}_0 - 1) e^{-t^{\ast}}}_{\text{Transient Component}},  \label{eq-ustar-dimless} \\
+s^{\ast}(t^{\ast}) & = \underbrace{\frac{1}{\gamma^{\ast}}}_{\text{Steady State}} + \underbrace{\left( s^{\ast}_0 - \xi - \frac{1}{\gamma^{\ast}} \right) e^{-\gamma^{\ast} t^{\ast}} + \xi e^{-t^{\ast}}}_{\text{Transient Components}}.
+\label{eq-sstar-dimless}
+\end{align}
+
+The **pre-mRNA dynamics** $u^{\ast}(t^{\ast})$ \eqref{eq-ustar-dimless} quickly 
+settles to its steady state value of 1, regardless of the initial concentration, 
+reflecting the balance between transcription and splicing rates. Note again that this
+does not refer to an arbitrary value of $1$ but rather represents the equivalence
+class of ratios $\alpha/\beta$.
+
+The **mRNA dynamics** $s^{\ast}(t^{\ast})$ \eqref{eq-sstar-dimless} reveals a 
+slightly more complicated balance of forces. The first transient term shows an exponential 
+decay influenced by $\gamma^{\ast}$, the degradation rate relative to the splicing rate. 
+The second transient term reflects how the 
+initial deviation in pre-mRNA concentration from its steady state affects mRNA 
+levels over time. The steady state of $s^{\ast}$ simply reflects the balance
+of the splicing rate, which is the only source of mRNA production in this model,
+and the degradation rate also given, just as in the case of the pre-mRNA, in units of 
+the transcription rate relative to the splicing rate.
+
+The analytical solutions to the dimensionless system reveal how the pre-mRNA 
+and mRNA concentrations evolve over time towards their steady states. The 
+exponential terms highlight the system's inherent timescales: $u^{\ast}$ relaxes 
+to its steady state independently, while $s^{\ast}$’s dynamics are coupled to 
+both $u^{\ast}$ and its own degradation rate. While this system is extremely
+simplified relative to well-known biology, as a stylized model it still
+indicates the interplay among transcription, splicing, and degradation rates.
+
+While it is trivial to solve this system analytically, in more general cases this
+ will not be possible. 
+So we will proceed to simulate the system to characterize the possible solution 
+sets and then perform probabilistic inference on the latent variables.
+
+## Data
+
+If we are just given samples of pre-mRNA and mRNA counts without additional
+information about their sampling times, we can describe the data set 
+$\mathcal{D}$ consisting of pre-mRNA and mRNA counts for a number 
+of genes $G$ across a number of cells $N$
+
+$$
+\mathcal{D} = \left\{ (u_{ij}, s_{ij}) \mid i \in \{1, \ldots, G\}, j \in \{1, \ldots, N\} \right\}
+$$
+
+where
+
+- $u_{ij}$ represents the count of pre-mRNA for gene $i$ in cell $j$,
+- $s_{ij}$ represents the count of mRNA for the same gene $i$ in the same cell $j$,
+- $G$ is the total number of genes in the study,
+- $N$ is the total number of cells sampled.
+
+Roughly speaking, the inference problem involves
+
+1. **Normalization**: Transforming the observed counts $u_{ij}$ and $s_{ij}$ to 
+   their dimensionless counterparts $u^{\ast}_{ij}$ and $s^{\ast}_{ij}$ using 
+   any reasonable concentration scale $U_0$.
+2. **Model Fitting**: Applying statistical methods to estimate the latent variables 
+   $\Theta = \left( \gamma^{\ast}, u^{\ast}_0, s^{\ast}_0, t^{\ast} \right)$ that best 
+   fit the observed data.
+3. **Evaluation**: Assessing the fit of the model and the estimated parameters'
+   biological plausibility and consistency across different genes and cells.
+
+This representation of the dataset and the associated objective of 
+statistical learning provides a solid foundation for applying probabilistic
+modeling techniques to calibrate an inference procedure andevaluate the 
+plausibility of this model in such a manner that we can eventually compare 
+multiple candidate models to one another.
+
+## Simulation
+
+We will primarily focus on simulating the system, since this will generalize
+to more complicated models. However, we will confirm our simulations for this
+first and simplest model recapitulate its analytical solution derived above.
+
+## Inference
+
+Now that we have illustrated how to simulate the system, we can proceed to
+define a probabilistic model and perform inference on its latent variables.

### 2024-04-07 src/pyrovelocity/models/_deterministic_simulation.py
nit(models): docstrings Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,5 +1,4 @@
 import logging
-from typing import Any
 from typing import Tuple
 
 import jax
@@ -91,6 +90,8 @@ def solve_transcription_splicing_model_impl(
         ts (TimeArrayType): array of time points
         initial_state (StateType): initial state vector
         params (ParamsType): Tuple of scalar parameter(s).
+        time_step (float): time step for the solver.
+        model (ModelType): model function.
 
     Returns:
         Solution: diffrax Solution object.
@@ -122,6 +123,8 @@ def solve_transcription_splicing_model(
     model: ModelType = dstate_dt_dimless,
 ) -> Solution:
     """
+    Solve the transcription-splicing model.
+
     This is a mirror of the `solve_transcription_splicing_model_impl` function
     with type-checking disabled for JIT compilation. This function should be
     removed when JIT compilation is enabled for the type-checked function.
@@ -130,6 +133,8 @@ def solve_transcription_splicing_model(
         ts (TimeArrayType): array of time points
         initial_state (StateType): initial state vector
         params (ParamsType): Tuple of scalar parameter(s).
+        time_step (float): time step for the solver.
+        model (ModelType): model function.
 
     Returns:
         Solution: diffrax Solution object.

### 2024-04-07 src/pyrovelocity/plots/__init__.py
feat(plots): draft simulation trajectory and phase portrait Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,3 +1,9 @@
+from pyrovelocity.plots._dynamics import (
+    plot_deterministic_simulation_phase_portrait,
+)
+from pyrovelocity.plots._dynamics import (
+    plot_deterministic_simulation_trajectories,
+)
 from pyrovelocity.plots._experimental import plot_t0_selection
 from pyrovelocity.plots._genes import plot_gene_ranking
 from pyrovelocity.plots._parameters import (
@@ -23,6 +29,8 @@ __all__ = [
     "cluster_violin_plots",
     "extrapolate_prediction_sample_predictive",
     "get_posterior_sample_angle_uncertainty",
+    "plot_deterministic_simulation_phase_portrait",
+    "plot_deterministic_simulation_trajectories",
     "plot_gene_ranking",
     "plot_parameter_posterior_distributions",
     "plot_posterior_time",

### 2024-04-07 src/pyrovelocity/plots/_dynamics.py

@@ -1,0 +1,107 @@
+import matplotlib.pyplot as plt
+
+__all__ = [
+    "plot_deterministic_simulation_phase_portrait",
+    "plot_deterministic_simulation_trajectories",
+]
+
+
+def plot_deterministic_simulation_trajectories(solution, title_prefix=""):
+    ts = solution.ts
+    unspliced, spliced = solution.ys[:, 0], solution.ys[:, 1]
+
+    fig, axes = plt.subplots(1, 2, figsize=(14, 5))
+
+    # linear scale
+    axes[0].plot(
+        ts,
+        unspliced,
+        "o-",
+        label=r"$u^{\ast}$",
+        color="blue",
+        markersize=5,
+        linewidth=1.5,
+        alpha=0.5,
+    )
+    axes[0].plot(
+        ts,
+        spliced,
+        "o-",
+        label=r"$s^{\ast}$",
+        color="green",
+        markersize=5,
+        linewidth=1.5,
+        alpha=0.5,
+    )
+    axes[0].set_xlabel("Time")
+    axes[0].set_ylabel("Concentration")
+    axes[0].set_title(f"{title_prefix} Temporal Trajectories (Linear Scale)")
+    axes[0].legend()
+
+    # log scale
+    axes[1].plot(
+        ts,
+        unspliced,
+        "o-",
+        label=r"$u^{\ast}$",
+        color="blue",
+        markersize=5,
+        linewidth=1.5,
+        alpha=0.5,
+    )
+    axes[1].plot(
+        ts,
+        spliced,
+        "o-",
+        label=r"$s^{\ast}$",
+        color="green",
+        markersize=5,
+        linewidth=1.5,
+        alpha=0.5,
+    )
+    axes[1].set_xscale("log")
+    axes[1].set_yscale("log")
+    axes[1].set_xlabel("Time")
+    axes[1].set_ylabel("Concentration")
+    axes[1].set_title(f"{title_prefix} Temporal Trajectories (Log Scale)")
+    axes[1].legend()
+
+    plt.tight_layout()
+    plt.show()
+
+
+def plot_deterministic_simulation_phase_portrait(solution, title_prefix=""):
+    unspliced, spliced = solution.ys[:, 0], solution.ys[:, 1]
+    norm_time = solution.ts
+
+    fig, axes = plt.subplots(1, 2, figsize=(14, 5))
+
+    # linear scale
[…]

### 2024-04-07 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): draft trajectories and phase portraits Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ```{python}
-# | label: transcription-splicing-degradation-equations
+#| label: transcription-splicing-degradation-equations
+#| code-fold: true
 from sympy import (
@@
 ```{python}
+#| code-fold: true
 du_dt_dimless = Eq(
@@
 ```{python}
+# | code-fold: true
 ds_dt_dimless = Eq(
@@
 
+As a simple example, we will simulate the system for a single set of initial
+conditions and parameters.
+
+```{python}
+# | label: transcription-splicing-degradation-simulation
+from pyrovelocity.models import solve_transcription_splicing_model
+from jax import numpy as jnp
+
+ts0 = jnp.linspace(0, 4, 40)
+ts1 = jnp.linspace(4 + (10 - 4) / 20, 10, 20)
+ts = jnp.concatenate([ts0, ts1])
+initial_state = jnp.array([0.5, 0.5])
+params = (0.5,)
+
+solution = solve_transcription_splicing_model(ts, initial_state, params)
+```
+
+We can visualize the results of the simulation by plotting the pre-mRNA and mRNA
+concentrations over time as shown in @fig-transcription-splicing-degradation-plot-simulation-trajectory-1,
+
+```{python}
+# | label: fig-transcription-splicing-degradation-plot-simulation-trajectory-1
+# | code-fold: true
+from pyrovelocity.plots import plot_deterministic_simulation_trajectories
+
+plot_deterministic_simulation_trajectories(
+    solution,
+    title_prefix="TSD Model",
+)
+```
+
+and the phase portrait in @fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1.
+
+```{python}
+# | label: fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1
+# | code-fold: true
+from pyrovelocity.plots import plot_deterministic_simulation_phase_portrait
+
+plot_deterministic_simulation_phase_portrait(
+    solution,
+    title_prefix="TSD Model",
+)
+```
+
 ## Inference

### 2024-04-07 src/pyrovelocity/models/_deterministic_simulation.py
fix(models): use arbitrary initial condition in analytical solution Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -142,6 +142,20 @@ def solve_transcription_splicing_model(
 
     Returns:
         Solution: diffrax Solution object.
+
+    Examples:
+        >>> from pyrovelocity.models import solve_transcription_splicing_model
+        >>> from jax import numpy as jnp
+        >>> ts0 = jnp.linspace(0, 4, 40)
+        >>> ts1 = jnp.linspace(4 + (10 - 4) / 20, 10, 20)
+        >>> ts = jnp.concatenate([ts0, ts1])
+        >>> initial_state = jnp.array([0.1, 0.1])
+        >>> params = (0.99,)
+        >>> solution = solve_transcription_splicing_model(
+        ...     ts,
+        ...     initial_state,
+        ...     params,
+        >>> )
     """
     term = ODETerm(model)
     t0 = ts[0]
@@ -209,16 +223,18 @@ def analytical_solution_dstate_dt_dimless(
     """
     (gamma_star,) = params
     u_star_0, s_star_0 = initial_state
+    t_star_0 = ts[0]
+    elapsed_ts = ts - t_star_0
 
     xi = calculate_xi(u_star_0, gamma_star)
 
-    u_star = 1 + (u_star_0 - 1) * jnp.exp(-ts)
+    u_star = 1 + (u_star_0 - 1) * jnp.exp(-elapsed_ts)
     s_star = (1 / gamma_star) + (
-        (s_star_0 - xi - (1 / gamma_star)) * jnp.exp(-gamma_star * ts)
-        + xi * jnp.exp(-ts)
+        (s_star_0 - xi - (1 / gamma_star)) * jnp.exp(-gamma_star * elapsed_ts)
+        + xi * jnp.exp(-elapsed_ts)
     )
 
-    return jnp.stack([u_star, s_star])
+    return jnp.stack([u_star, s_star]).T
 
 
 @jaxtyped(typechecker=beartype)
@@ -238,6 +254,20 @@ def solve_transcription_splicing_model_analytical(
 
     Returns:
         Solution: A Diffrax Solution object containing the analytical solutions.
+
+    Examples:
+        >>> from pyrovelocity.models import solve_transcription_splicing_model_analytical
+        >>> from jax import numpy as jnp
+        >>> ts0 = jnp.linspace(0, 4, 40)
+        >>> ts1 = jnp.linspace(4 + (10 - 4) / 20, 10, 20)
+        >>> ts = jnp.concatenate([ts0, ts1])
+        >>> initial_state = jnp.array([0.1, 0.1])
[…]

### 2024-04-07 src/pyrovelocity/tests/test_deterministic_simulation.py
fix(models): update dimensions in analytical solution test Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -169,8 +169,8 @@ def test_analytical_solution_dstate_dt_dimless_basic():
     )
 
     assert analytical_solution.shape == (
-        2,
         100,
+        2,
     ), "Unexpected shape of the analytical solution."
 
 
@@ -185,14 +185,14 @@ def test_solve_transcription_splicing_model_analytical_basic():
     )
 
     expected_output = initial_state
-    output = solution.ys[:, 0]
+    output = solution.ys[0, :]
     assert jnp.allclose(
         output,
         expected_output,
     ), f"Expected {expected_output}, got {output}"
 
     expected_output = jnp.array([1.0, 2.0])
-    output = solution.ys[:, -1]
+    output = solution.ys[-1, :]
     assert jnp.allclose(
         output,
         expected_output,
@@ -206,8 +206,8 @@ def test_solve_transcription_splicing_model_analytical_basic():
         100,
     ), f"Expected ts shape to be (100,), got {solution.ts.shape}"
     assert solution.ys.shape == (
-        2,
         100,
+        2,
     ), f"Expected ys shape to be (2, 100), got {solution.ys.shape}"
     assert (
         solution.result == diffrax.RESULTS.successful

### 2024-04-07 src/pyrovelocity/plots/_dynamics.py
fix(plots): support parameterized labels Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,4 +1,10 @@
+import jax.numpy as jnp
 import matplotlib.pyplot as plt
+import scienceplots
+from beartype import beartype
+from beartype.typing import List
+from diffrax import Solution
+from matplotlib import colors
 
 __all__ = [
     "plot_deterministic_simulation_phase_portrait",
@@ -6,102 +12,128 @@ __all__ = [
 ]
 
 
-def plot_deterministic_simulation_trajectories(solution, title_prefix=""):
-    ts = solution.ts
+@beartype
+def plot_deterministic_simulation_trajectories(
+    solution: Solution,
+    state_labels: List[str] = [r"$u^{\ast}$", r"$s^{\ast}$"],
+    xlabel: str = r"$t^{\ast}$",
+    ylabel: str = r"fraction of $u^{\ast}_{ss}$",
+    title_prefix: str = "",
+):
+    times = solution.ts
+    logarithmizable_times = jnp.where(times <= 1e-4, 1e-2, times)
     unspliced, spliced = solution.ys[:, 0], solution.ys[:, 1]
 
-    fig, axes = plt.subplots(1, 2, figsize=(14, 5))
+    with plt.style.context("science"):
+        fig, axes = plt.subplots(1, 2, figsize=(14, 5))
 
-    # linear scale
-    axes[0].plot(
-        ts,
-        unspliced,
-        "o-",
-        label=r"$u^{\ast}$",
-        color="blue",
-        markersize=5,
-        linewidth=1.5,
-        alpha=0.5,
-    )
-    axes[0].plot(
-        ts,
-        spliced,
-        "o-",
-        label=r"$s^{\ast}$",
-        color="green",
-        markersize=5,
-        linewidth=1.5,
-        alpha=0.5,
-    )
-    axes[0].set_xlabel("Time")
-    axes[0].set_ylabel("Concentration")
-    axes[0].set_title(f"{title_prefix} Temporal Trajectories (Linear Scale)")
-    axes[0].legend()
+        cmap = plt.get_cmap("BrBG")
+        color_brown = cmap(0.05)
+        color_blue_green = cmap(0.95)
 
-    # log scale
-    axes[1].plot(
-        ts,
-        unspliced,
-        "o-",
-        label=r"$u^{\ast}$",
-        color="blue",
-        markersize=5,
-        linewidth=1.5,
-        alpha=0.5,
-    )
-    axes[1].plot(
-        ts,
-        spliced,
-        "o-",
[…]

### 2024-04-07 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): include plot of analytical solution Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-| Symbol   | Description                 | Units                         | $O(-)$ Estimate | Discussion                                                                          |
-|----------|-----------------------------|-------------------------------|-----------------|-------------------------------------------------------------------------------------|
-| $u$      | Concentration of pre-mRNA   | molecules/cell                | $10^0 - 10^4$   | Wide range reflecting low to high gene expression levels.                           |
-| $s$      | Concentration of mRNA       | molecules/cell                | $10^0 - 10^5$   | Similar to $u$; depends on gene expression and stability of mRNA.                   |
-| $t$      | Time                        | seconds ($s$) to hours ($hr$) | $10s$ - $24hr$  | Depends on the duration of the experimental observation.                            |
-| $\alpha$ | Production rate of pre-mRNA | molecules/(cell·hr)           | $1 - 10^3$      | Based on variable gene transcription rates.                                         |
-| $\beta$  | Splicing rate of pre-mRNA   | $hr^{-1}$                     | $0.1 - 1$       | Pre-mRNA to mRNA splicing rates, assuming rapid processing within the nucleus.      |
-| $\gamma$ | Degradation rate of mRNA    | $hr^{-1}$                     | $0.1 - 1$       | mRNA half-lives can vary widely, but many are in the range of a few hours to a day. |
+| Symbol   | Description                 | Units                         | $O(-)$ Estimate | Note                                                                          |
+| -------- | --------------------------- | ----------------------------- | --------------- | ----------------------------------------------------------------------------- |
+| $u$      | Number of pre-mRNA          | molecules/cell                | $10^0 - 10^4$   | Wide range accounts for low to high gene expression levels.                   |
+| $s$      | Number of mRNA              | molecules/cell                | $10^0 - 10^5$   | Similar to $u$; depends on gene expression and stability of mRNA.             |
+| $t$      | Time                        | seconds ($s$) to hours ($hr$) | $5s$ - $48hr$   | Depends on the duration of the experimental observation.                      |
+| $\alpha$ | Production rate of pre-mRNA | molecules/(cell·hr)           | $10^0 - 10^3$      | Many transcripts are produced at rates in the range or 1 to 1000 per hour.    |
+| $\beta$  | Splicing rate of pre-mRNA   | $hr^{-1}$                     | $10^{-1} - 10^2$  | Many Pre-mRNA to mRNA splicing rates are in the range of 1 minute to 6 hours. |
+| $\gamma$ | Degradation rate of mRNA    | $hr^{-1}$                     | $10^{-2} - 10^0$     | Many mRNA half-lives are in the range of a half-hour to a day.                |
 
-: Variables and parameters of the transcription-splicing-degradation model of [@Zeisel2011-nz]. {#tbl-transcription-splicing-degradation}
+: Variables and parameters of the transcription-splicing-degradation model of [@Zeisel2011-nz] with order of magnitude estimates for ranges based on [@Phillips2012-mi] or references therein.
+{#tbl-transcription-splicing-degradation}
 
@@
 | Dimensioned Parameter        | Relation to Rescaling Parameters | Fundamental Units     |
-|-----------------------|--------------------------------------|--------------------------|
+| ---------------------------- | -------------------------------- | --------------------- |
 | Production Rate ($\alpha$)   | $\alpha = U_0 \beta$             | molecules/(cell·time) |
 | Splicing Rate ($\beta$)      | Reference Scale for $t^*$        | $1/\text{time}$       |
-| Degradation Rate ($\gamma$)| $\gamma = \gamma^* \beta$        | $1/\text{time}$        |
-| Pre-mRNA Concentration ($u$)| $u = u^* U_0$                    | molecules/cell           |
+| Degradation Rate ($\gamma$)  | $\gamma = \gamma^* \beta$        | $1/\text{time}$       |
+| Pre-mRNA Concentration ($u$) | $u = u^* U_0$                    | molecules/cell        |
 | mRNA Concentration ($s$)     | $s = s^* U_0$                    | molecules/cell        |
@@
 to their volume (i.e. the cube of a distance or length $L^3$). Usually we would 
-just write volume but we retain cell to associate to the object that determines 
+just write volume but we retain "cell" to associate to the object that determines 
 the relevant volume in this case. The dimensionless parameters are defined in 
@@
 
-| Dimensionless Variables and Parameters | Definition                     | Intuitive Description                                      |
-|-------------------------|--------------------------------|------------------------------------------------------------|
-| $u^{\ast}$ , $s^{\ast}$ | $u / U_0$, $s / U_0$             | Characteristic Scale of (pre-)mRNA concentration based on the balance between production and splicing rates. |
-| $t^*$                 | $\beta t$                    | Characteristic time scale relative to the splicing rate. |
+| Dimensionless Variables and Parameters | Definition           | Description                                                                                                                                                                    |
+| -------------------------------------- | -------------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------ |
+| $u^{\ast}$ , $s^{\ast}$                | $u / U_0$, $s / U_0$ | Characteristic scale of (pre-)mRNA concentration based on the balance between production and splicing rates.
+|
+| $t^*$                                  | $\beta t$            | Characteristic time scale relative to the splicing rate.
+|
 | $\gamma^*$                             | $\gamma / \beta$     | Relative degradation rate, comparing the degradation rate of mRNA to the splicing rate, indicating the stability or turnover rate of mRNA relative to its production/splicing.
@@
 
-Combining variables and parameters we have six and we have two fundamental 
+Given our previously defined parameter range for the splicing rate $\beta$, 
+we can interpret the dimensionless parameter $t^{\ast}$ according to @tbl-dimensionless-time-interpretation.
+
+| $\beta$ (hr$^{-1}$) | 1 unit of $t^*$ (hr) | 10 units of $t^*$ (hr) | 100 units of $t^*$ (hr) | 1 unit of $t^{\ast}$ | 10 units of $t^{\ast}$ | 100 units of $t^{\ast}$ |
+|---------------------|----------------------|------------------------|-------------------------|----------------------|------------------------|-------------------------|
+| $10^{-1}$           | 10                   | 100                    | 1000                    | 1                    | 10                     | 100                     |
+| $10^0$              | 1                    | 10                     | 100                     | 1                    | 10                     | 100                     |
+| $10^1$              | 0.1                  | 1                      | 10                      | 1                    | 10                     | 100                     |
+| $10^2$              | 0.01                 | 0.1                    | 1                       | 1                    | 10                     | 100                     |
+
+: This table interprets the dimensionless time $t^{\ast}$ across various splicing rates $\beta$ (hr$^{-1}$), demonstrating how real time and expected splicing events scale with changes in $\beta$.
+Each unit of $t^{\ast}$ represents a normalized measure of time relative to the splicing rate, with the final three columns indicating the **expected number of splicing events per molecule of pre-mRNA** for 1, 10, and 100 units of $t^{\ast}$ is indeed the meaning of time units for this scaling rule.
+These columns confirm that the dimensionless time directly reflects the expected splicing activity, endowing otherwise meaningless or difficult to interpret numbers into a universally applicable framework based on scales intrinsic to a given instance of the model.
+{#tbl-dimensionless-time-interpretation}
+
+Combining variables and parameters gives six total. We have two fundamental 
 dimensions (time and length) we can eliminate to arrive at four essential 
@@
 $$
-\xi = \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1},
+\xi = \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1}.
 $$
 
-so that the complete solutions for the dimensionless system are given by
+Then the complete solutions for the dimensionless system are given by
 
@@
 \end{align}
+
+Note we can restore an arbitrary initial timepoint putting $t^{\ast} \rightarrow t^{\ast} - t^{\ast}_0$.
 
@@
 ```{python}
+# | echo: false
+# | output: false
+%load_ext autoreload
+%autoreload 2
+```
+
+```{python}
 # | label: transcription-splicing-degradation-simulation
 from pyrovelocity.models import solve_transcription_splicing_model
+from pyrovelocity.models import solve_transcription_splicing_model_analytical
+from pyrovelocity.logging import configure_logging
 from jax import numpy as jnp
 
-ts0 = jnp.linspace(0, 4, 40)
-ts1 = jnp.linspace(4 + (10 - 4) / 20, 10, 20)
+logger = configure_logging("nbs")
+
+ts0 = jnp.linspace(0.1, 4.0, 40)
+ts1 = jnp.linspace(4.0 + (10.0 - 4.0) / 20, 10.0, 20)
 ts = jnp.concatenate([ts0, ts1])
-initial_state = jnp.array([0.5, 0.5])
-params = (0.5,)
+initial_state = jnp.array([0.1, 0.1])
+params = (0.99,)
 
-solution = solve_transcription_splicing_model(ts, initial_state, params)
+solution = solve_transcription_splicing_model(
+    ts,
+    initial_state,
+    params,
+)
+analytical_solution = solve_transcription_splicing_model_analytical(
+    ts,
+    initial_state,
+    params,
+)
+
+logger.info(
+    f"\nTrajectory preview times:\n{solution.ts[:3]}\n"
+    f"\nSimulated trajectory values:\n{solution.ys[:5]}\n"
+    f"\nAnalytical trajectory values:\n{analytical_solution.ys[:5]}\n\n"
+    f"\nTrajectory preview times:\n{solution.ts[-3:]}\n"
+    f"\nSimulated trajectory values:\n{solution.ys[-3:]}\n"
+    f"\nAnalytical trajectory values:\n{analytical_solution.ys[-3:]}\n\n"
+)
 ```
@@
 # | code-fold: true
+# | fig-cap: Simulated trajectories of pre-mRNA and mRNA concentrations over time for the transcription-splicing-degradation model.
 from pyrovelocity.plots import plot_deterministic_simulation_trajectories
@@
 plot_deterministic_simulation_trajectories(
-    solution,
-    title_prefix="TSD Model",
+    solution=solution,
+    title_prefix="TSD Model Simulated",
 )
@@
 
-and the phase portrait in @fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1.
+We see that if we plot the analytical solution, we get the same result as shown in @fig-transcription-splicing-degradation-plot-analytical-trajectory-1.
+
+```{python}
+# | label: fig-transcription-splicing-degradation-plot-analytical-trajectory-1
+# | code-fold: true
+# | fig-cap: Analytical trajectories of pre-mRNA and mRNA concentrations over time for the transcription-splicing-degradation model.
+from pyrovelocity.plots import plot_deterministic_simulation_trajectories
+
+plot_deterministic_simulation_trajectories(
+    solution=analytical_solution,
+    title_prefix="TSD Model Analytical",
+)
+```
+
+and the phase portraits as shown in @fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1.
 
@@
 # | code-fold: true
+# | fig-cap: Phase portrait of the transcription-splicing-degradation model.
 from pyrovelocity.plots import plot_deterministic_simulation_phase_portrait
@@
 plot_deterministic_simulation_phase_portrait(
-    solution,
+    solution=solution,
     title_prefix="TSD Model",

### 2024-04-08 src/pyrovelocity/plots/_dynamics.py
fix(plots): set 10^-8 min for log-transformed values Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -6,6 +6,7 @@ from beartype.typing import List
 from diffrax import Solution
 from matplotlib import colors
 
+
 __all__ = [
     "plot_deterministic_simulation_phase_portrait",
     "plot_deterministic_simulation_trajectories",
@@ -19,15 +20,22 @@ def plot_deterministic_simulation_trajectories(
     xlabel: str = r"$t^{\ast}$",
     ylabel: str = r"fraction of $u^{\ast}_{ss}$",
     title_prefix: str = "",
+    colormap_name: str = "cividis",
 ):
     times = solution.ts
     logarithmizable_times = jnp.where(times <= 1e-4, 1e-2, times)
     unspliced, spliced = solution.ys[:, 0], solution.ys[:, 1]
+    logarithmizable_unspliced = jnp.where(
+        jnp.abs(unspliced) <= 3e-8, 1e-8, jnp.abs(unspliced)
+    )
+    logarithmizable_spliced = jnp.where(
+        jnp.abs(spliced) <= 3e-8, 1e-8, jnp.abs(spliced)
+    )
 
     with plt.style.context("science"):
         fig, axes = plt.subplots(1, 2, figsize=(14, 5))
 
-        cmap = plt.get_cmap("BrBG")
+        cmap = plt.get_cmap(colormap_name)
         color_brown = cmap(0.05)
         color_blue_green = cmap(0.95)
 
@@ -62,7 +70,7 @@ def plot_deterministic_simulation_trajectories(
         # log scale
         axes[1].plot(
             logarithmizable_times,
-            unspliced,
+            logarithmizable_unspliced,
             "o-",
             label=state_labels[0],
             color=color_brown,
@@ -72,7 +80,7 @@ def plot_deterministic_simulation_trajectories(
         )
         axes[1].plot(
             logarithmizable_times,
-            spliced,
+            logarithmizable_spliced,
             "o-",
             label=state_labels[1],
             color=color_blue_green,
@@ -98,6 +106,7 @@ def plot_deterministic_simulation_phase_portrait(
     ylabel: str = r"$u^{\ast}$",
     zlabel: str = r"$t^{\ast}$",
     title_prefix: str = "",
+    colormap_name: str = "cividis",
 ):
     times = solution.ts
     logarithmizable_times = jnp.where(times <= 1e-4, 1e-2, times)
@@ -112,7 +121,7 @@ def plot_deterministic_simulation_phase_portrait(
             spliced,
             unspliced,
             c=times,
-            cmap="BrBG",
[…]

### 2024-04-08 src/pyrovelocity/models/_deterministic_simulation.py
fix(models): account for parameter space singularity Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -226,13 +226,22 @@ def analytical_solution_dstate_dt_dimless(
     t_star_0 = ts[0]
     elapsed_ts = ts - t_star_0
 
-    xi = calculate_xi(u_star_0, gamma_star)
-
     u_star = 1 + (u_star_0 - 1) * jnp.exp(-elapsed_ts)
-    s_star = (1 / gamma_star) + (
-        (s_star_0 - xi - (1 / gamma_star)) * jnp.exp(-gamma_star * elapsed_ts)
-        + xi * jnp.exp(-elapsed_ts)
-    )
+
+    if jnp.isclose(gamma_star, 1.0):
+        s_star = (
+            1
+            + (s_star_0 - 1) * jnp.exp(-elapsed_ts)
+            + (u_star_0 - 1) * elapsed_ts * jnp.exp(-elapsed_ts)
+        )
+    else:
+        xi = calculate_xi(u_star_0, gamma_star)
+
+        s_star = (1 / gamma_star) + (
+            (s_star_0 - xi - (1 / gamma_star))
+            * jnp.exp(-gamma_star * elapsed_ts)
+            + xi * jnp.exp(-elapsed_ts)
+        )
 
     return jnp.stack([u_star, s_star]).T
 

### 2024-04-08 nbs/concepts/inferenceindynamicalsystems/index.qmd
fix(nbs): account for parameter space singularity Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 | Symbol   | Description                 | Units                         | $O(-)$ Estimate  | Note                                                                          |
-| -------- | --------------------------- | ----------------------------- | --------------- | ----------------------------------------------------------------------------- |
+| -------- | --------------------------- | ----------------------------- | ---------------- | ----------------------------------------------------------------------------- |
 | $u$      | Number of pre-mRNA          | molecules/cell                | $10^0 - 10^4$    | Wide range accounts for low to high gene expression levels.                   |
@@
 
-Given the state variables representing concentrations of pre-mRNA, $u$, 
+Given state variables representing concentrations of pre-mRNA, $u$, 
 and mRNA, $s$, we have the following ordinary differential equations 
+taken from [@Zeisel2011-nz] (please refer again to 
+@tbl-transcription-splicing-degradation for the meaning of the variables 
+and parameters),
 
 \begin{align}
-\frac{du}{dt} & = \alpha - \beta u  \label{eq-dudt} \\
-\frac{ds}{dt} & = \beta u - \gamma s  \label{eq-dsdt}
+\frac{du}{dt} & = \alpha - \beta u  \label{eq-dudt}, \\
+\frac{ds}{dt} & = \beta u - \gamma s  \label{eq-dsdt},
 \end{align}
 
-\eqref{eq-dudt} and \eqref{eq-dsdt} are usually presented with 
-the concept that the parameter $\alpha$ could better account for the external 
-inputs to the regulation of transcription of the gene if it were allowed to be 
-a function of time, $\alpha(t)$. We will return to this point later, but 
-for now we will assume that $\alpha$ and other parameters are constant.
+which proposes a mean-field model for the dynamics of transcription and 
+splicing on a continuous state space. It is important to note this is only
+one among an eventual ensemble of models that, with appropriate analysis
+and inference, may be organized into a hierarchy of such models according
+to their domains of validity relative to the scales involved in and resolution 
+of observations of the relevant phenomenon [@Wells2012-qf].
+Of course, in the context of single gene transcription without explicit modeling
+of interactions, this ensemble of models would include the various adaptations and
+extensions of the so-called random telegraph model and the experimental 
+justification for its consideration 
+[@Ham2020-ij;@Cao2020-xv;@Bohrer2021-qk].
+As an example of a minimal extension, \eqref{eq-dudt} and \eqref{eq-dsdt} are 
+usually presented with the concept that the parameter $\alpha$ could better 
+account for the external inputs to the regulation of transcription of the gene 
+if it were allowed to be a function of time, $\alpha(t)$. However, because this
+complicates or eliminates analytical tractability, and thus maximum likelihood
+inference, approximation of this function may be piecewise constant 
+[@Bergen2020-pj]. We will return to 
+this point later, but, for the purpose of simply illustrating the relationsips 
+among analyzing, simulating, and performing inference upon the parameters of a 
+dynamical system, we will assume that $\alpha$ and other parameters are 
+constant values.
 
@@
 
-It is common to perform a dimensional analysis of the model to ensure the
+It is common to perform a dimensional analysis of dynamical models to ensure the
 units of the variables and parameters are consistent and reduce the total number
-of parameters by the number of dimensions. Intuitively, this folds or maps parameters
-to equivalence classes of similar dynamics.  For reference, the procedure
-we outline is also commonly referred to as Buckingham's $\Pi$ theorem.
+of parameters by the number of dimensions of the associated dilation group 
+symmetry [@Pegoraro1973-tt]. Intuitively, this maps sets of parameters
+to associated equivalence classes of similar dynamics. For reference, in addition to 
+[dimensional analysis](https://en.wikipedia.org/wiki/Dimensional_analysis) and 
+[nondimensionalization](https://en.wikipedia.org/wiki/Nondimensionalization), 
+the procedure we make use of depends on what is frequently 
+referred to as the 
+[Buckingham $\Pi$ theorem](https://en.wikipedia.org/wiki/Buckingham_%CF%80_theorem) 
+[@Cantwell2002-dv].
 
@@
 | $\beta$ (hr$^{-1}$) | 1 unit of $t^*$ (hr) | 10 units of $t^*$ (hr) | 100 units of $t^*$ (hr) | 1 unit of $t^{\ast}$ | 10 units of $t^{\ast}$ | 100 units of $t^{\ast}$ |
-|---------------------|----------------------|------------------------|-------------------------|----------------------|------------------------|-------------------------|
+| ------------------- | -------------------- | ---------------------- | ----------------------- | -------------------- | ---------------------- | ----------------------- |
 | $10^{-1}$           | 10                   | 100                    | 1000                    | 1                    | 10                     | 100                     |
@@
 
-and define dimensionelss variables and parameters as
+and define dimensionless variables and parameters as
 
@@
 ```{python}
+#| label: symbolic-tsd-dudt-dimensionless-equation
 #| code-fold: true
@@
 ```{python}
+#| label: symbolic-tsd-dsdt-dimensionless-equation
 # | code-fold: true
@@
 \begin{align}
-\frac{du^{\ast}}{dt^{\ast}} & = 1 - u^{\ast}  \label{eq-dustardtstar} \\
-\frac{ds^{\ast}}{dt^{\ast}} & = u^{\ast} - \gamma^{\ast} s^{\ast}  \label{eq-dsstardtstar}
+\frac{du^{\ast}}{dt^{\ast}} & = 1 - u^{\ast}  \label{eq-dustardtstar}, \\
+\frac{ds^{\ast}}{dt^{\ast}} & = u^{\ast} - \gamma^{\ast} s^{\ast},  \label{eq-dsstardtstar}
 \end{align}
@@
 understood about the system's dynamics.
-Given generic initial conditions $u^{\ast}(0) = u^{\ast}_0$ and $s^{\ast}(0) = s^{\ast}_0$, the 
+Given specific initial conditions $u^{\ast}(0) = u^{\ast}_0$ and $s^{\ast}(0) = s^{\ast}_0$, which
+we restore to arbitrary $t^{\ast}_0$ at the end of the derivation, the 
 first equation
@@
 
-$C_1$ is the constant determined by initial conditions, which can be shown to be $C_1 = u^{\ast}_0 - 1$, giving us
+$C_1$ is of course the constant determined by the initial conditions, which can 
+be shown to be $C_1 = u^{\ast}_0 - 1,$ giving us
 
@@
 This equation describes how the concentration of pre-mRNA, in units of the 
-characteristic concentration scale given by $\alpha/\beta$, evolves over the 
-timescale characterized by the splicing rate. The term $e^{-t^{\ast}}$ 
+characteristic concentration scale, $U_0 = \alpha/\beta$, evolves over the 
+timescale characterized by the inverse of the splicing rate, $1/\beta$. The term $e^{-t^{\ast}}$ 
 indicates that any deviation of $u^{\ast}$ from its steady state $1$ decays 
@@
 
-Then the complete solutions for the dimensionless system are given by
+The complete solutions for the dimensionless system for $\gamma^{\ast} \neq 1$ are then given by
 
@@
 
-Note we can restore an arbitrary initial timepoint putting $t^{\ast} \rightarrow t^{\ast} - t^{\ast}_0$.
+For the case $\gamma^{\ast} = 1$, we can solve \eqref{eq-dustardtstar} and 
+\eqref{eq-dsstardtstar} to find the solution for $u^{\ast}$ is unchanged
+and $s^{\ast}$ is then given by
 
-The **pre-mRNA dynamics** $u^{\ast}(t^{\ast})$ \eqref{eq-ustar-dimless} quickly 
-settles to its steady state value of 1, regardless of the initial concentration, 
-reflecting the balance between transcription and splicing rates. Note again that this
-does not refer to an arbitrary value of $1$ but rather represents the equivalence
-class of ratios $\alpha/\beta$.
+$$
+s^{\ast}(t^{\ast}) = 1 + (s^{\ast}_0 - 1) e^{-t^{\ast}} + (u^{\ast}_0 - 1) t^{\ast} e^{-t^{\ast}}.
+$$
 
-The **mRNA dynamics** $s^{\ast}(t^{\ast})$ \eqref{eq-sstar-dimless} reveals a 
-slightly more complicated balance of forces. The first transient term shows an exponential 
-decay influenced by $\gamma^{\ast}$, the degradation rate relative to the splicing rate. 
-The second transient term reflects how the 
-initial deviation in pre-mRNA concentration from its steady state affects mRNA 
-levels over time. The steady state of $s^{\ast}$ simply reflects the balance
-of the splicing rate, which is the only source of mRNA production in this model,
-and the degradation rate also given, just as in the case of the pre-mRNA, in units of 
-the transcription rate relative to the splicing rate.
+Note we can restore an arbitrary initial timepoint putting 
+$t^{\ast} \rightarrow t^{\ast} - t^{\ast}_0$.
+
+The dimensionless **pre-mRNA dynamics** $u^{\ast}(t^{\ast})$
+given in \eqref{eq-ustar-dimless} 
+settles to its steady state value of 1 at rate 1, regardless of the initial 
+concentration, reflecting the balance between transcription and splicing 
+rates. Note again that this does not refer to an arbitrary value of $1$ but 
+rather represents the equivalence class of ratios $\alpha/\beta$.
+
+The dimensionless **mRNA dynamics** $s^{\ast}(t^{\ast})$ 
+given in \eqref{eq-sstar-dimless} reveals a 
+slightly more complicated balance of forces. The first transient term shows 
+an exponential decay influenced by $\gamma^{\ast}$,the degradation rate 
+relative to the splicing rate, when $\gamma^{\ast}$ is not precisely $1$. 
+The second transient term reflects how the initial deviation in pre-mRNA 
+concentration from its steady state affects mRNA levels over time. The steady 
+state of $s^{\ast}$ simply reflects the balance of the splicing rate, which is 
+the only source of mRNA production in this model, and the degradation rate 
+also given, just as in the case of the pre-mRNA, in units of the transcription 
+rate relative to the splicing rate.
 
@@
 ```{python}
-# | echo: false
-# | output: false
-%load_ext autoreload
-%autoreload 2
-```
-
-```{python}
 # | label: transcription-splicing-degradation-simulation
+import diffrax
 from pyrovelocity.models import solve_transcription_splicing_model
@@
 initial_state = jnp.array([0.1, 0.1])
-params = (0.99,)
+params = (1.00,)
+colormap_name = "RdBu"
 
@@
     params,
+)
+analytical_simulation_error = diffrax.Solution(
+    t0=ts[0],
+    t1=ts[-1],
+    ts=ts,
+    ys=analytical_solution.ys - solution.ys,
+    interpolation=None,
+    stats={},
+    result=diffrax.RESULTS.successful,
+    solver_state=None,
+    controller_state=None,
+    made_jump=None,
 )
@@
     title_prefix="TSD Model Simulated",
+    colormap_name=colormap_name,
 )
@@
 
-We see that if we plot the analytical solution, we get the same result as shown in @fig-transcription-splicing-degradation-plot-analytical-trajectory-1.
+We see that if we plot the analytical solution, we get the same result as shown in @fig-transcription-splicing-degradation-plot-analytical-trajectory-1 to within the expected error of the numerical simulation ($O(10^{-5} - 10^{-8})$).
 
@@
 # | code-fold: true
-# | fig-cap: Analytical trajectories of pre-mRNA and mRNA concentrations over time for the transcription-splicing-degradation model.
+# | fig-cap: Absolute error in analytical vs simulated trajectories of pre-mRNA and mRNA concentrations over time for the transcription-splicing-degradation model.
 from pyrovelocity.plots import plot_deterministic_simulation_trajectories
@@
 plot_deterministic_simulation_trajectories(
-    solution=analytical_solution,
-    title_prefix="TSD Model Analytical",
+    solution=analytical_simulation_error,
+    title_prefix="TSD Model Analytical-Simulation Error",
+    colormap_name=colormap_name,
 )
@@
 
-and the phase portraits as shown in @fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1.
+The same simulated trajectories from @fig-transcription-splicing-degradation-plot-simulation-trajectory-1 are shown in the phase space given by $u^{\ast} \otimes s^{\ast}$ in @fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1.
 
@@
     title_prefix="TSD Model",
+    colormap_name=colormap_name,
 )

### 2024-04-10 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): add inference model description Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 @tbl-transcription-splicing-degradation together with (\ref{eq-dudt}-\ref{eq-dsdt})
-contain essentially all the information we require to get started. The dimensions of the variables and parameters together with dimensionless ones are provided in @tbl-parameter-dimensions.
+contain essentially all the information we require to get started. 
+The dimensions of the variables and parameters together with dimensionless 
+ones are provided in @tbl-parameter-dimensions.
 
@@
 
-: Variables and parameters of the transcription-splicing-degradation model together with their fundamental units and their relations to dimensionless parameters. {#tbl-parameter-dimensions}
+: Variables and parameters of the transcription-splicing-degradation model 
+together with their fundamental units and their relations to dimensionless 
+parameters. {#tbl-parameter-dimensions}
 
@@
 
-: Dimensionless Parameters, definitions, and intuitive Descriptions {#tbl-dimensionless-parameters}
+: Dimensionless Parameters, definitions, and intuitive Descriptions 
+{#tbl-dimensionless-parameters}
 
@@
 
-: This table interprets the dimensionless time $t^{\ast}$ across various splicing rates $\beta$ (hr$^{-1}$), demonstrating how real time and expected splicing events scale with changes in $\beta$.
-Each unit of $t^{\ast}$ represents a normalized measure of time relative to the splicing rate, with the final three columns indicating the **expected number of splicing events per molecule of pre-mRNA** for 1, 10, and 100 units of $t^{\ast}$ is indeed the meaning of time units for this scaling rule.
-These columns confirm that the dimensionless time directly reflects the expected splicing activity, endowing otherwise meaningless or difficult to interpret numbers into a universally applicable framework based on scales intrinsic to a given instance of the model.
+: This table interprets the dimensionless time $t^{\ast}$ across various
+splicing rates $\beta$ (hr$^{-1}$), demonstrating how real time and expected
+splicing events scale with changes in $\beta$. Each unit of $t^{\ast}$
+represents a normalized measure of time relative to the splicing rate, with the
+final three columns indicating the **expected number of splicing events per
+molecule of pre-mRNA** for 1, 10, and 100 units of $t^{\ast}$ is indeed the
+meaning of time units for this scaling rule. These columns confirm that the
+dimensionless time directly reflects the expected splicing activity, endowing
+otherwise meaningless or difficult to interpret numbers into a universally
+applicable framework based on scales intrinsic to a given instance of the model.
 {#tbl-dimensionless-time-interpretation}
@@
 
-We see that we correctly arrive at the equivalent dimensionless system of equations
+We see that we correctly arrive at the equivalent dimensionless system of 
+equations
 
@@
 \frac{du^{\ast}}{dt^{\ast}} & = 1 - u^{\ast}  \label{eq-dustardtstar}, \\
-\frac{ds^{\ast}}{dt^{\ast}} & = u^{\ast} - \gamma^{\ast} s^{\ast},  \label{eq-dsstardtstar}
+\frac{ds^{\ast}}{dt^{\ast}} & = u^{\ast} - \gamma^{\ast} s^{\ast},  
+  \label{eq-dsstardtstar}
 \end{align}
@@
 understood about the system's dynamics.
-Given specific initial conditions $u^{\ast}(0) = u^{\ast}_0$ and $s^{\ast}(0) = s^{\ast}_0$, which
-we restore to arbitrary $t^{\ast}_0$ at the end of the derivation, the 
-first equation
+Given specific initial conditions $u^{\ast}(0) = u^{\ast}_0$ and 
+$s^{\ast}(0) = s^{\ast}_0$, which we restore to arbitrary $t^{\ast}_0$ 
+at the end of the derivation, the first equation
 
@@
 characteristic concentration scale, $U_0 = \alpha/\beta$, evolves over the 
-timescale characterized by the inverse of the splicing rate, $1/\beta$. The term $e^{-t^{\ast}}$ 
-indicates that any deviation of $u^{\ast}$ from its steady state $1$ decays 
-exponentially over time.
+timescale characterized by the inverse of the splicing rate, $1/\beta$. 
+The term $e^{-t^{\ast}}$ indicates that any deviation of $u^{\ast}$ from 
+its steady state $1$ decays exponentially over time.
 
@@
 $$
-\frac{ds^{\ast}}{dt^{\ast}} = [(u^{\ast}_0 - 1) e^{-t^{\ast}} + 1] - \gamma^{\ast} s^{\ast}
+\frac{ds^{\ast}}{dt^{\ast}} = [(u^{\ast}_0 - 1) e^{-t^{\ast}} + 1] - 
+  \gamma^{\ast} s^{\ast}
 $$
@@
 $$
-e^{\gamma^{\ast} t^{\ast}} \frac{ds^{\ast}}{dt^{\ast}} + \gamma^{\ast} e^{\gamma^{\ast} t^{\ast}} s^{\ast} = (u^{\ast}_0 - 1) e^{(\gamma^{\ast}-1) t^{\ast}} + e^{\gamma^{\ast} t^{\ast}}
+e^{\gamma^{\ast} t^{\ast}} \frac{ds^{\ast}}{dt^{\ast}} + 
+  \gamma^{\ast} e^{\gamma^{\ast} t^{\ast}} s^{\ast} = 
+  (u^{\ast}_0 - 1) e^{(\gamma^{\ast}-1) t^{\ast}} + 
+  e^{\gamma^{\ast} t^{\ast}}
 $$
@@
 $$
-s^{\ast}(t^{\ast}) = C_2 e^{-\gamma^{\ast} t^{\ast}} + \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1} e^{-t^{\ast}} + \frac{1}{\gamma^{\ast}}
+s^{\ast}(t^{\ast}) = C_2 e^{-\gamma^{\ast} t^{\ast}} + 
+  \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1} e^{-t^{\ast}} + 
+  \frac{1}{\gamma^{\ast}}
 $$
@@
 where $C_2$ is another integration constant determined by initial conditions. 
-Incorporating the initial condition  $s^{\ast}(0) = s^{\ast}_0$ into the solution for $s^{\ast}(t^{\ast})$ to determine $C_2$
+Incorporating the initial condition  $s^{\ast}(0) = s^{\ast}_0$ into the 
+solution for $s^{\ast}(t^{\ast})$ to determine $C_2$
 
 $$
-C_2 = s^{\ast}_0 - \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1} - \frac{1}{\gamma^{\ast}}.
+C_2 = s^{\ast}_0 - \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1} - 
+  \frac{1}{\gamma^{\ast}}.
 $$
@@
 
-The complete solutions for the dimensionless system for $\gamma^{\ast} \neq 1$ are then given by
+The complete solutions for the dimensionless system for $\gamma^{\ast} \neq 1$ 
+are then given by
 
 \begin{align}
-u^{\ast}(t^{\ast}) & = \underbrace{1}_{\text{Steady State}} + \underbrace{(u^{\ast}_0 - 1) e^{-t^{\ast}}}_{\text{Transient Component}},  \label{eq-ustar-dimless} \\
-s^{\ast}(t^{\ast}) & = \underbrace{\frac{1}{\gamma^{\ast}}}_{\text{Steady State}} + \underbrace{\left( s^{\ast}_0 - \xi - \frac{1}{\gamma^{\ast}} \right) e^{-\gamma^{\ast} t^{\ast}} + \xi e^{-t^{\ast}}}_{\text{Transient Components}}.
+u^{\ast}(t^{\ast}) & = \underbrace{1}_{\text{Steady State}} + 
+  \underbrace{(u^{\ast}_0 - 1) e^{-t^{\ast}}}_{\text{Transient Component}}, 
+  \label{eq-ustar-dimless} \\
+s^{\ast}(t^{\ast}) & = \underbrace{\frac{1}{\gamma^{\ast}}}_{\text{Steady State}} + 
+  \underbrace{\left( s^{\ast}_0 - \xi - \frac{1}{\gamma^{\ast}} \right) 
+  e^{-\gamma^{\ast} t^{\ast}} + \xi e^{-t^{\ast}}}_{\text{Transient Components}}.  
   \label{eq-sstar-dimless}
@@
 $$
-s^{\ast}(t^{\ast}) = 1 + (s^{\ast}_0 - 1) e^{-t^{\ast}} + (u^{\ast}_0 - 1) t^{\ast} e^{-t^{\ast}}.
+s^{\ast}(t^{\ast}) = 1 + (s^{\ast}_0 - 1) e^{-t^{\ast}} + 
+  (u^{\ast}_0 - 1) t^{\ast} e^{-t^{\ast}}.
 $$
@@
 
-If we are just given samples of pre-mRNA and mRNA counts without additional
-information about their sampling times, we can describe the data set 
+If we are given samples of pre-mRNA and mRNA counts along with
+the sampling time of each cell, we can describe the data set 
 $\mathcal{D}$ consisting of pre-mRNA and mRNA counts for a number 
@@
 
-$$
-\mathcal{D} = \left\{ (u_{ij}, s_{ij}) \mid i \in \{1, \ldots, G\}, j \in \{1, \ldots, N\} \right\}
-$$
+\begin{align}
+\mathcal{D} = \left\{ (t_j, u_{ij}, s_{ij}) \mid 
+  i \in \{1, \ldots, G\}, j \in \{1, \ldots, N\} \right\} \label{eq-dataset},
+\end{align}
 
@@
 
+- $t_j$ represents the sampling time for cell $j$,
 - $u_{ij}$ represents the count of pre-mRNA for gene $i$ in cell $j$,
@@
    their dimensionless counterparts $u^{\ast}_{ij}$ and $s^{\ast}_{ij}$ using 
-   any reasonable concentration scale $U_0$.
-2. **Model Fitting**: Applying statistical methods to estimate the latent variables 
-   $\Theta = \left( \gamma^{\ast}, u^{\ast}_0, s^{\ast}_0, t^{\ast} \right)$ that best 
-   fit the observed data.
+   a reasonable concentration scale $U_0$.
+2. **Model Fitting**: Applying statistical methods to estimate the latent 
+   variables 
+   $\Theta = \left( \gamma^{\ast}, u^{\ast}_0, s^{\ast}_0 \right)$ 
+   that best fit the observed data.
 3. **Evaluation**: Assessing the fit of the model and the estimated parameters'
@@
 
+We will return later to the alternative version of this problem where $t_j$ 
+are not observed and must be inferred along with the latent variables 
+$\Theta = \left( \gamma^{\ast}, u^{\ast}_0, s^{\ast}_0, t^{\ast} \right)$.
+
 This representation of the dataset and the associated objective of 
 statistical learning provides a solid foundation for applying probabilistic
-modeling techniques to calibrate an inference procedure andevaluate the 
+modeling techniques to calibrate an inference procedure and evaluate the 
 plausibility of this model in such a manner that we can eventually compare 
@@
 We can visualize the results of the simulation by plotting the pre-mRNA and mRNA
-concentrations over time as shown in @fig-transcription-splicing-degradation-plot-simulation-trajectory-1,
+concentrations over time as shown in 
+@fig-transcription-splicing-degradation-plot-simulation-trajectory-1,
 
@@
 
-We see that if we plot the analytical solution, we get the same result as shown in @fig-transcription-splicing-degradation-plot-analytical-trajectory-1 to within the expected error of the numerical simulation ($O(10^{-5} - 10^{-8})$).
+We see that if we plot the analytical solution, we get the same result as shown 
+in @fig-transcription-splicing-degradation-plot-analytical-trajectory-1 to 
+within the expected error of the numerical simulation ($O(10^{-5} - 10^{-8})$).
 
@@
 
-The same simulated trajectories from @fig-transcription-splicing-degradation-plot-simulation-trajectory-1 are shown in the phase space given by $u^{\ast} \otimes s^{\ast}$ in @fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1.
+The same simulated trajectories from 
+@fig-transcription-splicing-degradation-plot-simulation-trajectory-1 are shown 
+in the phase space given by $u^{\ast} \otimes s^{\ast}$ in 
+@fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1.
 
@@
 
-Now that we have illustrated how to simulate the system, we can proceed to
-define a probabilistic model and perform inference on its latent variables.
+Now that we have illustrated how to simulate a prototypical system, we can
+proceed to define a probabilistic model and perform inference on its latent
+variables given data like \eqref{eq-dataset}. In general we should consider multiple
+models with different assumptions and levels of complexity, but for now we will
+focus on one of the simpler models to illustrate the process. In particular, in
+most cases, when performing inference on the latent variables of a dynamical
+system, we assume that the observation time points are fixed or otherwise
+observed. We will begin with this assumption and then proceed to discuss how to
+relax it.
+
+In the case where we can indeed estimate analytical solutions to the system, we
+can consider a model with the structure proposed in 
+(\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) where, given a number of 
+genes $G$ and cells $N$, for each
+$(i,j) \in \{1, \ldots, G\} \otimes \{1, \ldots, N\}$ we have
+
+\begin{align}
+    u^{\ast}_{0i}, s^{\ast}_{0i} &\sim 
+      \text{LogNormal}(\mu_{0}, \sigma_{0}^2), 
+      \label{eq-init-conds-priors-gene} \\
+
+    \gamma^{\ast}_i &\sim 
+      \text{LogNormal}(\mu_{\gamma}, \sigma_{\gamma}^2),
+      \label{eq-gamma-prior-gene} \\
+
+    \sigma_{ui}, \sigma_{si} &\sim 
+      \text{HalfNormal}(\mu_{\sigma}, \sigma_{\sigma}^2),
+      \label{eq-noise-std-priors-gene} \\
+
+    u^{\ast}_{ij} &= 
+      1 + (u^{\ast}_{0i} - 1) \cdot e^{-t^{\ast}_j},
+      \label{eq-u-star-model-cell} \\
+
+    s^{\ast}_{ij} &= 
+      \begin{cases}
+        \frac{1}{\gamma^{\ast}_i} + 
+        \left( s^{\ast}_{0i} - \xi_i - \frac{1}{\gamma^{\ast}_i} \right) \cdot e^{-\gamma^{\ast}_i t^{\ast}_j} + 
+        \xi_i \cdot e^{-t^{\ast}_j},& \gamma^{\ast}_i \neq 1 \\
+
+        1 + (s^{\ast}_{0i} - 1) e^{-t^{\ast}_j} + 
+          (u^{\ast}_{0i} - 1) t^{\ast}_j e^{-t^{\ast}_j},& \gamma^{\ast}_i = 1 \\
+      \end{cases}, \label{eq-s-star-model-cell} \\
+
+    \hat{u}^{\ast}_{ij} &\sim \text{Normal}(u^{\ast}_{ij}, \sigma_{ui}^2), 
+    \label{eq-u-obs-cell} \\
+
+    \hat{s}^{\ast}_{ij} &\sim \text{Normal}(s^{\ast}_{ij}, \sigma_{si}^2). 
+    \label{eq-s-obs-cell}
+\end{align}
+
+The model description in (\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) 
+can be associated to the plate diagram in 
+@fig-transcription-splicing-degradation-graphical-model-1.
+
+```{python}
+# | label: fig-transcription-splicing-degradation-graphical-model-1
+# | code-fold: true
+# | fig-cap: Graphical model for the transcription-splicing-degradation model.
+import daft
+import matplotlib.pyplot as mpl
+
+mpl.rcParams["font.family"] = "serif"
+mpl.rcParams["font.size"] = 16
+mpl.rcParams["text.usetex"] = True
+
+pgm = daft.PGM(line_width=1.2)
+
+# hyperparameters
+pgm.add_node("mu_init", r"$\mu_{0}$", 0.5, 6, fixed=True)
+pgm.add_node("sigma_init", r"$\sigma_{0}^2$", 1.5, 6, fixed=True)
+pgm.add_node("mu_gamma", r"$\mu_{\gamma}$", 2.5, 6, fixed=True)
+pgm.add_node("sigma_gamma", r"$\sigma_{\gamma}^2$", 3.5, 6, fixed=True)
+pgm.add_node("mu_sigma", r"$\mu_{\sigma}$", 4.5, 6, fixed=True)
+pgm.add_node("sigma_sigma", r"$\sigma_{\sigma}^2$", 5.5, 6, fixed=True)
+
+# latent variables for gene-specific parameters
+pgm.add_node("u_star_0i", r"$u^{\ast}_{0i}$", 1, 5)
+pgm.add_node("s_star_0i", r"$s^{\ast}_{0i}$", 2, 5)
+pgm.add_node("gamma_star_i", r"$\gamma^{\ast}_i$", 3, 5)
+pgm.add_node("sigma_ui", r"$\sigma_{ui}$", 4, 5)
+pgm.add_node("sigma_si", r"$\sigma_{si}$", 5, 5)
+
+# latent variables for cell-specific outcomes
+pgm.add_node(
+    "u_star_ij",
+    r"$u^{\ast}_{ij}$",
+    2,
+    4,
+    scale=1.0,
+    shape="rectangle",
+)
+pgm.add_node(
+    "s_star_ij",
+    r"$s^{\ast}_{ij}$",
+    4,
+    4,
+    scale=1.0,
+    shape="rectangle",
+)
+
+# observed data
+pgm.add_node(
+    "t_star_j",
+    r"$t^{\ast}_j$",
+    6.0,
+    3.25,
+    observed=True,
+    shape="rectangle",
+)
+pgm.add_node(
+    "u_obs_ij",
+    r"$\hat{u}^{\ast}_{ij}$",
+    2,
+    2.5,
+    scale=1.0,
+    observed=True,
+)
+pgm.add_node(
+    "s_obs_ij",
+    r"$\hat{s}^{\ast}_{ij}$",
+    4,
+    2.5,
+    scale=1.0,
+    observed=True,
+)
+
+# edges
+edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}
+pgm.add_edge("mu_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("mu_sigma", "sigma_ui", plot_params=edge_params)
+pgm.add_edge("sigma_sigma", "sigma_ui", plot_params=edge_params)
+pgm.add_edge("mu_sigma", "sigma_si", plot_params=edge_params)
+pgm.add_edge("sigma_sigma", "sigma_si", plot_params=edge_params)
+
+pgm.add_edge("u_star_0i", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("s_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("u_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)
+
+pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("sigma_ui", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("sigma_si", "s_obs_ij", plot_params=edge_params)
+
+pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)
+
+# plates
+pgm.add_plate([0.5, 1.2, 5, 4.4], label=r"$i \in \{1, \ldots, G\}$", shift=-0.1)
+pgm.add_plate(
+    [1.0, 1.8, 5.5, 2.75], label=r"$j \in \{1, \ldots, N\}$", shift=-0.1
+)
+
+pgm.render()
+```
+
+Of course, in general, we will not have access to the analytical solutions, so
+we will need to simulate the system and then perform inference on the latent
+variables as suggested in (\ref{eq-init-conds-priors-num}--\ref{eq-s-obs-num}).
+
+\begin{align}
+    u^{\ast}_0, s^{\ast}_0 &\sim \text{LogNormal}(\mu_{0}, \sigma_{0}^2) \label{eq-init-conds-priors-num}, \\
+    \gamma^{\ast} &\sim \text{LogNormal}(\mu_{\gamma}, \sigma_{\gamma}^2) \label{eq-gamma-prior-num}, \\
+    \sigma_u, \sigma_s &\sim \text{HalfNormal}(\mu_{\sigma}, \sigma_{\sigma}^2) \label{eq-noise-std-priors-num}, \\
+    (u^{\ast}, s^{\ast}) &= \text{NumericalSolver}\left(\frac{du^{\ast}}{dt^{\ast}}, \frac{ds^{\ast}}{dt^{\ast}}, u^{\ast}_0, s^{\ast}_0, \gamma^{\ast}, t^{\ast}\right) \label{eq-numerical-solution}, \\
+    \hat{u}^{\ast} &\sim \text{Normal}(u^{\ast}, \sigma_u^2) \label{eq-u-obs-num}, \\
+    \hat{s}^{\ast} &\sim \text{Normal}(s^{\ast}, \sigma_s^2) \label{eq-s-obs-num}.
+\end{align}
+
+Where $\frac{du^{\ast}}{dt^{\ast}}$ and $\frac{ds^{\ast}}{dt^{\ast}}$, 
+are given by (\ref{eq-dustardtstar}-\ref{eq-dsstardtstar}) and noting that 
+we suppress the indices from 
+(\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) for brevity.

### 2024-04-14 src/pyrovelocity/models/_hyperparameter_calibration.py
fix(models): update hyperparameter calibration test values Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -84,7 +84,7 @@ def solve_for_lognormal_sigma_given_threshold_and_tail_mass(
         >>> solve_for_lognormal_sigma_given_threshold_and_tail_mass(
         ...     0.0, 1.5, tail_mass=0.05, upper=True
         >>> )
-        1.0
+        0.2465052825762025
 
         Solve for sigma to achieve a lower tail probability of 0.05 below a
         threshold of 1.5 with mu=3.0:
@@ -92,7 +92,7 @@ def solve_for_lognormal_sigma_given_threshold_and_tail_mass(
         >>> solve_for_lognormal_sigma_given_threshold_and_tail_mass(
         ...     3.0, 1.5, tail_mass=0.05, upper=False
         >>> )
-        1.5773652131590963
+        1.5773652131591058
     """
 
     func_to_solve = lambda sigma: (

### 2024-04-14 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): update inference model without times Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 # | label: transcription-splicing-degradation-simulation
+# | code-fold: true
 import diffrax
@@
 (\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) where, given a number of 
-genes $G$ and cells $N$, for each
-$(i,j) \in \{1, \ldots, G\} \otimes \{1, \ldots, N\}$ we have
+genes $G$ and cells $N$ that are measured at $K_j$ distinct times, then for each
+$(i,j,k) \in \{1, \ldots, G\} \otimes \{1, \ldots, N\} \otimes \{1, \ldots, K_j\}$ 
+we have
 
@@
 
-    u^{\ast}_{ij} &= 
-      1 + (u^{\ast}_{0i} - 1) \cdot e^{-t^{\ast}_j},
+    {u^{\ast}}^{k}_{ij} &= 
+      1 + (u^{\ast}_{0i} - 1) \cdot e^{-{t^{\ast}}^k_j},
       \label{eq-u-star-model-cell} \\
 
-    s^{\ast}_{ij} &= 
+    {s^{\ast}}^{k}_{ij} &= 
       \begin{cases}
         \frac{1}{\gamma^{\ast}_i} + 
-        \left( s^{\ast}_{0i} - \xi_i - \frac{1}{\gamma^{\ast}_i} \right) \cdot e^{-\gamma^{\ast}_i t^{\ast}_j} + 
-        \xi_i \cdot e^{-t^{\ast}_j},& \gamma^{\ast}_i \neq 1 \\
+        \left( s^{\ast}_{0i} - \xi_i - \frac{1}{\gamma^{\ast}_i} \right) \cdot e^{-\gamma^{\ast}_i {t^{\ast}}^k_j} + 
+        \xi_i \cdot e^{-{t^{\ast}}^k_j},& \gamma^{\ast}_i \neq 1 \\
 
-        1 + (s^{\ast}_{0i} - 1) e^{-t^{\ast}_j} + 
-          (u^{\ast}_{0i} - 1) t^{\ast}_j e^{-t^{\ast}_j},& \gamma^{\ast}_i = 1 \\
+        1 + (s^{\ast}_{0i} - 1) e^{-{t^{\ast}}^k_j} + 
+          (u^{\ast}_{0i} - 1) {t^{\ast}}^k_j e^{-{t^{\ast}}^k_j},& \gamma^{\ast}_i = 1 \\
       \end{cases}, \label{eq-s-star-model-cell} \\
 
-    \hat{u}^{\ast}_{ij} &\sim \text{Normal}(u^{\ast}_{ij}, \sigma_{ui}^2), 
+    \hat{u}^{\ast}{}^k_{ij} &\sim \text{Normal}({u^{\ast}}^k_{ij}, \sigma_{ui}^2), 
     \label{eq-u-obs-cell} \\
 
-    \hat{s}^{\ast}_{ij} &\sim \text{Normal}(s^{\ast}_{ij}, \sigma_{si}^2). 
+    \hat{s}^{\ast}{}^k_{ij} &\sim \text{Normal}({s^{\ast}}^k_{ij}, \sigma_{si}^2). 
     \label{eq-s-obs-cell}
@@
 import daft
-import matplotlib.pyplot as mpl
+import matplotlib.pyplot as plt
 
-mpl.rcParams["font.family"] = "serif"
-mpl.rcParams["font.size"] = 16
-mpl.rcParams["text.usetex"] = True
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["font.size"] = 16
+plt.rcParams["text.usetex"] = True
+
+pgm = daft.PGM(line_width=1.2)
+
+# hyperparameters
+pgm.add_node("mu_init", r"$\mu_{0}$", 0.5, 6, fixed=True)
+pgm.add_node("sigma_init", r"$\sigma_{0}^2$", 1.5, 6, fixed=True)
+pgm.add_node("mu_gamma", r"$\mu_{\gamma}$", 2.5, 6, fixed=True)
+pgm.add_node("sigma_gamma", r"$\sigma_{\gamma}^2$", 3.5, 6, fixed=True)
+pgm.add_node("mu_sigma", r"$\mu_{\sigma}$", 4.5, 6, fixed=True)
+pgm.add_node("sigma_sigma", r"$\sigma_{\sigma}^2$", 5.5, 6, fixed=True)
+
+# latent variables for gene-specific parameters
+pgm.add_node("u_star_0i", r"$u^{\ast}_{0i}$", 1, 5)
+pgm.add_node("s_star_0i", r"$s^{\ast}_{0i}$", 2, 5)
+pgm.add_node("gamma_star_i", r"$\gamma^{\ast}_i$", 3, 5)
+pgm.add_node("sigma_ui", r"$\sigma_{ui}$", 4, 5)
+pgm.add_node("sigma_si", r"$\sigma_{si}$", 5, 5)
+
+# latent variables for cell-specific outcomes
+pgm.add_node(
+    "u_star_ij",
+    r"${u^{\ast}}^k_{ij}$",
+    2,
+    3.8,
+    scale=1.0,
+    shape="rectangle",
+)
+pgm.add_node(
+    "s_star_ij",
+    r"${s^{\ast}}^k_{ij}$",
+    4,
+    3.8,
+    scale=1.0,
+    shape="rectangle",
+)
+
+# observed data
+pgm.add_node(
+    "t_star_j",
+    r"${t^{\ast}}^k_j$",
+    5.9,
+    3.1,
+    observed=True,
+    shape="rectangle",
+)
+pgm.add_node(
+    "u_obs_ij",
+    r"$\hat{u}^{\ast}{}^{k}_{ij}$",
+    2,
+    2.4,
+    scale=1.0,
+    observed=True,
+)
+pgm.add_node(
+    "s_obs_ij",
+    r"$\hat{s}^{\ast}{}^{k}_{ij}$",
+    4,
+    2.4,
+    scale=1.0,
+    observed=True,
+)
+
+# edges
+edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}
+pgm.add_edge("mu_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("mu_sigma", "sigma_ui", plot_params=edge_params)
+pgm.add_edge("sigma_sigma", "sigma_ui", plot_params=edge_params)
+pgm.add_edge("mu_sigma", "sigma_si", plot_params=edge_params)
+pgm.add_edge("sigma_sigma", "sigma_si", plot_params=edge_params)
+
+pgm.add_edge("u_star_0i", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("s_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("u_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)
+
+pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("sigma_ui", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("sigma_si", "s_obs_ij", plot_params=edge_params)
+
+pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)
+
+# plates
+pgm.add_plate(
+    [0.4, 1.0, 5, 4.5],
+    label=r"$i \in \{1, \ldots, G\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+pgm.add_plate(
+    [0.8, 1.4, 5.9, 3.2],
+    label=r"$j \in \{1, \ldots, N\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+pgm.add_plate(
+    [1.2, 1.8, 5.2, 2.5],
+    label=r"$k \in \{1, \ldots, K_j\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+
+pgm.render()
+```
+
+Of course, in general, we will not have access to the analytical solutions, so
+we will need to simulate the system and then perform inference on the latent
+variables as suggested in (\ref{eq-init-conds-priors-num}--\ref{eq-s-obs-num}).
+
+\begin{align}
+    u^{\ast}_0, s^{\ast}_0 &\sim \text{LogNormal}(\mu_{0}, \sigma_{0}^2) \label{eq-init-conds-priors-num}, \\
+    \gamma^{\ast} &\sim \text{LogNormal}(\mu_{\gamma}, \sigma_{\gamma}^2) \label{eq-gamma-prior-num}, \\
+    \sigma_u, \sigma_s &\sim \text{HalfNormal}(\mu_{\sigma}, \sigma_{\sigma}^2) \label{eq-noise-std-priors-num}, \\
+    (u^{\ast}, s^{\ast}) &= \text{NumericalSolver}\left(\frac{du^{\ast}}{dt^{\ast}}, \frac{ds^{\ast}}{dt^{\ast}}, u^{\ast}_0, s^{\ast}_0, \gamma^{\ast}, t^{\ast}\right) \label{eq-numerical-solution}, \\
+    \hat{u}^{\ast} &\sim \text{Normal}(u^{\ast}, \sigma_u^2) \label{eq-u-obs-num}, \\
+    \hat{s}^{\ast} &\sim \text{Normal}(s^{\ast}, \sigma_s^2) \label{eq-s-obs-num}.
+\end{align}
+
+Where $\frac{du^{\ast}}{dt^{\ast}}$ and $\frac{ds^{\ast}}{dt^{\ast}}$, 
+are given by (\ref{eq-dustardtstar}-\ref{eq-dsstardtstar}) and noting that 
+we suppress the indices from 
+(\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) for brevity.
+
+If, for example, we only observe one time point for each cell, 
+then @fig-transcription-splicing-degradation-graphical-model-1 and its 
+associated description in (\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) 
+reduces to the graphical model in 
+@fig-transcription-splicing-degradation-graphical-model-2.
+
+```{python}
+# | label: fig-transcription-splicing-degradation-graphical-model-2
+# | code-fold: true
+# | fig-cap: Graphical model for the transcription-splicing-degradation model observed at a single time point.
+import daft
+import matplotlib.pyplot as plt
+
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["font.size"] = 16
+plt.rcParams["text.usetex"] = True
 
@@
 # plates
-pgm.add_plate([0.5, 1.2, 5, 4.4], label=r"$i \in \{1, \ldots, G\}$", shift=-0.1)
 pgm.add_plate(
-    [1.0, 1.8, 5.5, 2.75], label=r"$j \in \{1, \ldots, N\}$", shift=-0.1
+    [0.5, 1.2, 5, 4.4],
+    label=r"$i \in \{1, \ldots, G\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+pgm.add_plate(
+    [1.0, 1.8, 5.5, 2.75],
+    label=r"$j \in \{1, \ldots, N\}$",
+    shift=-0.1,
+    fontsize=12,
 )
@@
 ```
-
-Of course, in general, we will not have access to the analytical solutions, so
-we will need to simulate the system and then perform inference on the latent
-variables as suggested in (\ref{eq-init-conds-priors-num}--\ref{eq-s-obs-num}).
-
-\begin{align}
-    u^{\ast}_0, s^{\ast}_0 &\sim \text{LogNormal}(\mu_{0}, \sigma_{0}^2) \label{eq-init-conds-priors-num}, \\
-    \gamma^{\ast} &\sim \text{LogNormal}(\mu_{\gamma}, \sigma_{\gamma}^2) \label{eq-gamma-prior-num}, \\
-    \sigma_u, \sigma_s &\sim \text{HalfNormal}(\mu_{\sigma}, \sigma_{\sigma}^2) \label{eq-noise-std-priors-num}, \\
-    (u^{\ast}, s^{\ast}) &= \text{NumericalSolver}\left(\frac{du^{\ast}}{dt^{\ast}}, \frac{ds^{\ast}}{dt^{\ast}}, u^{\ast}_0, s^{\ast}_0, \gamma^{\ast}, t^{\ast}\right) \label{eq-numerical-solution}, \\
-    \hat{u}^{\ast} &\sim \text{Normal}(u^{\ast}, \sigma_u^2) \label{eq-u-obs-num}, \\
-    \hat{s}^{\ast} &\sim \text{Normal}(s^{\ast}, \sigma_s^2) \label{eq-s-obs-num}.
-\end{align}
-
-Where $\frac{du^{\ast}}{dt^{\ast}}$ and $\frac{ds^{\ast}}{dt^{\ast}}$, 
-are given by (\ref{eq-dustardtstar}-\ref{eq-dsstardtstar}) and noting that 
-we suppress the indices from 
-(\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) for brevity.

### 2024-04-25 nbs/concepts/rrv/index.qmd
chore(nbs): init rrv Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: \'Renormalizing\' effective theories for gene expression velocities over differentiation states
title-slide-attributes: 
  # credit: https://github.com/ddatsko/3d-ising-model/
  data-background-image: https://user-images.githubusercontent.com/35429810/116011574-72395b00-a62e-11eb-847b-26cf5034d64c.gif
  data-background-position: center
  data-background-size: contain
  data-background-opacity: "0.4"
toc: true
number-sections: true
highlight-style: pygments
csl: ../../bibstyle.csl
lightbox: auto
bibliography: ../../references.bib
format: 
  revealjs:
    slide-number: true
    width: 1600
    height: 900
    logo: "../../_static/logo.png"
    theme: ["theme/custom.scss", night]
editor: source
filters:
  - include-code-files
---

# Simulation ⇄ Inference {
    data-background-image="https://upload.wikimedia.org/wikipedia/commons/b/b2/Ising_Criticality2.gif"
    data-background-size="contain"
    data-background-opacity="0.3"
  }

## Simulation

- Simulation is the process of generating data from a model [@Wells2012-qf;@Alon2019-ah].
- Inside the model, we may generate (intermediate) data from other models.
- We can generate samples of our observables by executing all nested models.

## Inference

- Inference is the process of estimating the parameters of a model given data.
- This is done by computing the posterior distribution of the parameters given the data.
- Inside the model, we may estimate the parameters of other models.

# References

### 2024-04-25 src/pyrovelocity/models/_deterministic_inference.py
nit(models): rename sample_y to sample_data Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -819,22 +819,22 @@ def plot_sample_trajectories(
                         ms=12,
                     )
 
-            sample_y = idata.posterior_predictive.observations.sel(
+            sample_data = idata.posterior_predictive.observations.sel(
                 chain=0, draw=trajectories_index, genes=gene_index
             ).values
 
-            if num_trajectories > sample_y.shape[0]:
+            if num_trajectories > sample_data.shape[0]:
                 logger.warning(
                     f"\nRequested number of trajectories ({num_trajectories}) "
-                    f"exceeds available samples ({sample_y.shape[0]}).\n"
+                    f"exceeds available samples ({sample_data.shape[0]}).\n"
                     f"Adjusting to maximum available.\n"
                 )
-                num_trajectories = sample_y.shape[0]
+                num_trajectories = sample_data.shape[0]
             indices_to_plot = np.random.choice(
-                sample_y.shape[0], num_trajectories, replace=False
+                sample_data.shape[0], num_trajectories, replace=False
             )
             for idx in indices_to_plot:
-                for cell_index in range(sample_y.shape[1]):
+                for cell_index in range(sample_data.shape[1]):
                     for modality_index, modality in enumerate(
                         idata.posterior_predictive.observations.coords[
                             "modalities"
@@ -842,9 +842,9 @@ def plot_sample_trajectories(
                     ):
                         ax.plot(
                             # observed_times,
-                            # sample_y[idx, :, :, modality_index],
+                            # sample_data[idx, :, :, modality_index],
                             observed_times[cell_index],
-                            sample_y[idx, cell_index, :, modality_index],
+                            sample_data[idx, cell_index, :, modality_index],
                             alpha=0.3,
                             color="gray" if modality == "pre-mRNA" else "green",
                             marker="2",

### 2024-04-25 src/pyrovelocity/models/_deterministic_inference.py
fix(models): remove redundant Result container for plot_sample_trajectories Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -754,14 +754,13 @@ def save_inference_plots(
                 output_dir=output_dir,
             )
 
-            result = plot_sample_trajectories(idata_posterior)
-            if is_successful(result):
-                for idx, fig in enumerate(result.unwrap()):
-                    save_figure_object(
-                        fig=fig,
-                        name=f"sample_trajectories_{idx}",
-                        output_dir=output_dir,
-                    )
+            figs = plot_sample_trajectories(idata_posterior)
+            for idx, fig in enumerate(figs):
+                save_figure_object(
+                    fig=fig,
+                    name=f"sample_trajectories_{idx}",
+                    output_dir=output_dir,
+                )
 
         return Success(True)
 
@@ -774,7 +773,7 @@ def plot_sample_trajectories(
     idata: InferenceData,
     trajectories_index: int | slice = slice(None),
     num_trajectories: int = 100,
-) -> Result[List[Figure], Exception]:
+) -> List[Figure]:
     """
     Plots sample trajectories over time for all genes, modality, and cell
     combinations.
@@ -792,81 +791,67 @@ def plot_sample_trajectories(
             Success containing a list of Figure objects if plots are created
             without errors, otherwise the error wrapped in a Failure.
     """
-    try:
-        figs = []
-        genes = idata.observed_data.observations.coords["genes"].values
-        for gene_index in genes:
-            fig, ax = plt.subplots(figsize=(12, 8))
+    figs = []
+    genes = idata.observed_data.observations.coords["genes"].values
+    for gene_index in genes:
+        fig, ax = plt.subplots(figsize=(12, 8))
 
-            observed_data = idata.observed_data.observations.sel(
-                genes=gene_index
-            ).values
-            observed_times = idata.observed_data.times.values
+        observed_data = idata.observed_data.observations.sel(
+            genes=gene_index
+        ).values
+        observed_times = idata.observed_data.times.values
 
-            for modality_index, modality in enumerate(
-                idata.observed_data.observations.coords["modalities"].values
-            ):
-                for cell_index in range(observed_data.shape[0]):
+        for modality_index, modality in enumerate(
+            idata.observed_data.observations.coords["modalities"].values
+        ):
+            for cell_index in range(observed_data.shape[0]):
+                ax.plot(
+                    observed_times[cell_index],
+                    observed_data[cell_index, :, modality_index],
+                    label=f"{modality}" if cell_index == 0 else "_nolegend_",
+                    color="gray" if modality == "pre-mRNA" else "green",
+                    marker=".",
+                    ms=12,
[…]

### 2024-04-25 src/pyrovelocity/tests/test_deterministic_inference.py

@@ -602,17 +602,15 @@ def test_plot_sample_trajectories(
 ):
     idata_posterior, _, _, _ = setup_posterior_inference_data
 
-    result = plot_sample_trajectories(
+    figs = plot_sample_trajectories(
         idata=idata_posterior,
     )
-    if is_successful(result):
-        figs = result.unwrap()
 
-        assert isinstance(figs, list) and all(
-            isinstance(fig, Figure) for fig in figs
-        ), "All elements should be matplotlib Figure instances."
-        for fig in figs:
-            assert len(fig.axes[0].lines) > 0, "Each plot should contain lines."
+    assert isinstance(figs, list) and all(
+        isinstance(fig, Figure) for fig in figs
+    ), "All elements should be matplotlib Figure instances."
+    for fig in figs:
+        assert len(fig.axes[0].lines) > 0, "Each plot should contain lines."
 
 
 @pytest.fixture

### 2024-04-26 src/pyrovelocity/models/_deterministic_inference.py
fix(models): change logger from warning to debug Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1150,7 +1150,7 @@ def plot_sample_trajectories(
             )
 
             if num_trajectories > sample_data.shape[0]:
-                logger.warning(
+                logger.debug(
                     f"\nRequested number of trajectories ({num_trajectories}) "
                     f"exceeds available samples ({sample_data.shape[0]}).\n"
                     "Adjusting to maximum available.\n"

### 2024-04-26 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): add inference plots to concepts Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ```
+
+Now we can set up some sample data and perform inference on the latent variables
+
+```{python}
+# | label: fig-generate-test-data-for-deterministic-model-inference
+# | code-fold: true
+# | fig-cap: Test data
+
+import arviz as az
+from pyrovelocity.models._deterministic_inference import (
+    generate_posterior_inference_data,
+    generate_prior_inference_data,
+    generate_test_data_for_deterministic_model_inference,
+    plot_sample_phase_portraits,
+    plot_sample_trajectories,
+    plot_sample_trajectories_with_percentiles,
+    print_inference_data_structure,
+)
+
+(
+    times,
+    data,
+    num_genes,
+    num_cells,
+    num_timepoints,
+    num_modalities,
+) = generate_test_data_for_deterministic_model_inference(
+    num_genes=1,
+    num_cells=20,
+    num_timepoints=1,
+    num_modalities=2,
+    noise_levels=(0.001, 0.001),
+)
+
+num_chains = 1
+num_samples = 10
+num_warmup = 10
+
+idata_prior = generate_prior_inference_data(
+    times=times,
+    data=data,
+    num_chains=num_chains,
+    num_samples=num_samples,
+    num_genes=num_genes,
+    num_cells=num_cells,
+    num_timepoints=num_timepoints,
+    num_modalities=num_modalities,
+)
+
+idata_posterior = generate_posterior_inference_data(
+    times=times,
+    data=data,
+    num_chains=num_chains,
+    num_samples=num_samples,
+    num_genes=num_genes,
+    num_cells=num_cells,
+    num_timepoints=num_timepoints,
+    num_modalities=num_modalities,
+    num_warmup=num_warmup,
+)
+```
+
+We can plot the prior
+
+```{python}
+# | label: fig-transcription-splicing-degradation-prior
+# | code-fold: true
+# | fig-cap: Prior distributions for the transcription-splicing-degradation model.
+light_gray = "#bcbcbc"
+variables = ["initial_conditions", "gamma", "sigma"]
+for var in variables:
+    az.plot_posterior(
+        idata_prior,
+        var_names=[var],
+        group="prior",
+        kind="hist",
+        color=light_gray,
+        round_to=2,
+    )
+```
+
+and posterior distributions for the latent variables
+
+```{python}
+# | label: fig-transcription-splicing-degradation-posterior
+# | code-fold: true
+# | fig-cap: Posterior distributions for the transcription-splicing-degradation model.
+light_gray = "#bcbcbc"
+variables = ["initial_conditions", "gamma", "sigma"]
+for var in variables:
+    az.plot_posterior(
+        idata_posterior,
+        var_names=[var],
+        group="posterior",
+        kind="hist",
+        color=light_gray,
+        round_to=2,
+    )
+```
+
+We can plot posterior predictive trajectories
+
+```{python}
+# | label: fig-transcription-splicing-degradation-posterior-predictive-trajectories
+# | code-fold: true
+# | fig-cap: Posterior predictive trajectories for the transcription-splicing-degradation model.
+figs = plot_sample_trajectories(
+    idata=idata_posterior,
+)
+for fig in figs:
+    fig.show()
+```
+
+trajectories in the phase portrait space
+
+```{python}
+# | label: fig-transcription-splicing-degradation-posterior-predictive-phase-portraits
+# | code-fold: true
+# | fig-cap: Posterior predictive phase portraits for the transcription-splicing-degradation model.
+figs = plot_sample_phase_portraits(
+    idata=idata_posterior,
+    colormap_name="RdBu",
+)
+for fig in figs:
+    fig.show()
+```
+
+as well as estimates for the distribution of the pre-mRNA and mRNA
+
+```{python}
+# | label: fig-transcription-splicing-degradation-posterior-predictive
+# | code-fold: true
+# | fig-cap: Posterior predictive distribution for the transcription-splicing-degradation model.
+figs = plot_sample_trajectories_with_percentiles(
+    idata=idata_posterior,
+)
+for fig in figs:
+    fig.show()
+```
+
+we can also plot the prior predictive distribution
+
+```{python}
+# | label: fig-transcription-splicing-degradation-prior-predictive
+# | code-fold: true
+# | fig-cap: Prior predictive distribution for the transcription-splicing-degradation model.
+figs = plot_sample_trajectories_with_percentiles(
+    idata=idata_posterior,
+)
+for fig in figs:
+    fig.show()
+```

### 2024-04-26 nbs/concepts/rrv/index.qmd
feat(nbs): add inference plots to rrv Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 toc: true
+toc-depth: 1
 number-sections: true
@@
 bibliography: ../../references.bib
+execute: 
+  freeze: true
+  eval: true
+  warning: false
+  error: false
+  cache: true
 format: 
@@
 
-# Simulation ⇄ Inference {
+# Splice census from bulk to single-cell RNA-seq
+
+## Simplistic cartoon of gene expression
+
+![](https://www.nature.com/scitable/content/ne0000/ne0000/ne0000/ne0000/14711098/U2CP3-1_SynthesisDegredati_ksm.jpg){.r-stretch fig-align="center"}
+
+[@Alberts2014-kt]
+
+## Splicing mechanism
+
+![](https://media.springernature.com/full/springer-static/image/art%3A10.1038%2Fs41580-022-00545-z/MediaObjects/41580_2022_545_Fig1_HTML.png){fig-align="center" height="760"}
+
+[@Marasco2023-yb]
+
+## (Pre‐)mRNA response profile classification
+
+![](https://www.embopress.org/cms/10.1038/msb.2011.62/asset/329a576a-84a0-47c0-9bfc-125e6ab740b6/assets/graphic/msb201162-fig-0001-m.jpg){fig-align="center"}
+
+[@Zeisel2011-nz]
+
+## (Pre‐)mRNA stimulus‐induced response profiles
+
+![](https://www.embopress.org/cms/10.1038/msb.2011.62/asset/042557d4-dbc2-472a-a685-94666dcb90a1/assets/graphic/msb201162-fig-0003-m.jpg){fig-align="center" height="720"}
+
+[@Zeisel2011-nz]
+
+## Single-cell RNA velocity
+
+![](https://www.embopress.org/cms/10.15252/msb.202110282/asset/36cba1b7-8d35-491f-b2bd-a543103be834/assets/graphic/msb202110282-fig-0001-m.png){fig-align="center" height="720"}
+
+[@La_Manno2018-lj;@Bergen2021-qz]
+
+# Deterministic modeling of gene expression
+
+## Variables {.center}
+
+| Symbol   | Description                 | Units                         | $O(-)$ Estimate  | Note                                                                          |
+| -------- | --------------------------- | ----------------------------- | ---------------- | ----------------------------------------------------------------------------- |
+| $u$      | Number of pre-mRNA          | molecules/cell                | $10^0 - 10^4$    | Wide range accounts for low to high gene expression levels.                   |
+| $s$      | Number of mRNA              | molecules/cell                | $10^0 - 10^5$    | Similar to $u$; depends on gene expression and stability of mRNA.             |
+| $t$      | Time                        | seconds ($s$) to hours ($hr$) | $5s$ - $48hr$    | Depends on the duration of the experimental observation.                      |
+
+## Parameters {.center}
+
+| Symbol   | Description                 | Units                         | $O(-)$ Estimate  | Note                                                                          |
+| -------- | --------------------------- | ----------------------------- | ---------------- | ----------------------------------------------------------------------------- |
+| $\alpha$ | Production rate of pre-mRNA | molecules/(cell·hr)           | $10^0 - 10^3$    | Many transcripts are produced at rates in the range or 1 to 1000 per hour.    |
+| $\beta$  | Splicing rate of pre-mRNA   | $hr^{-1}$                     | $10^{-1} - 10^2$ | Many Pre-mRNA to mRNA splicing rates are in the range of 1 minute to 6 hours. |
+| $\gamma$ | Degradation rate of mRNA    | $hr^{-1}$                     | $10^{-2} - 10^0$ | Many mRNA half-lives are in the range of a half-hour to a day.                |
+
+## Dynamics {.center}
+
+Given state variables representing concentrations of pre-mRNA, $u$, 
+and mRNA, $s$, we have the following ordinary differential equations 
+taken from [@Zeisel2011-nz],
+
+\begin{align}
+\frac{du}{dt} & = \alpha - \beta u  \label{eq-dudt}, \\
+\frac{ds}{dt} & = \beta u - \gamma s  \label{eq-dsdt},
+\end{align}
+
+These are 
+usually presented with the concept that the parameter $\alpha$ could better 
+account for the external inputs to the regulation of transcription of the gene 
+if it were allowed to be a function of time, $\alpha(t)$.
+
+## Dimensional analysis
+
+| Dimensioned Parameter        | Relation to Rescaling Parameters | Fundamental Units     |
+| ---------------------------- | -------------------------------- | --------------------- |
+| Production Rate ($\alpha$)   | $\alpha = U_0 \beta$             | molecules/(cell·time) |
+| Splicing Rate ($\beta$)      | Reference Scale for $t^*$        | $1/\text{time}$       |
+| Degradation Rate ($\gamma$)  | $\gamma = \gamma^* \beta$        | $1/\text{time}$       |
+| Pre-mRNA Concentration ($u$) | $u = u^* U_0$                    | molecules/cell        |
+| mRNA Concentration ($s$)     | $s = s^* U_0$                    | molecules/cell        |
+| Time ($t$)                   | $t = t^* / \beta$                | time                  |
+
+## Dimensionless variables and parameters
+
+| Dimensionless Variables and Parameters | Definition           | Description                                                                                                                                                                    |
+| -------------------------------------- | -------------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------ |
+| $u^{\ast}$ , $s^{\ast}$                | $u / U_0$, $s / U_0$ | Characteristic scale of (pre-)mRNA concentration based on the balance between production and splicing rates.
+|
+| $t^*$                                  | $\beta t$            | Characteristic time scale relative to the splicing rate.
+|
+| $\gamma^*$                             | $\gamma / \beta$     | Relative degradation rate, comparing the degradation rate of mRNA to the splicing rate, indicating the stability or turnover rate of mRNA relative to its production/splicing.
+|
+
+## Interpreting dimensionless parameters
+
+| $\beta$ (hr$^{-1}$) | 1 unit of $t^*$ (hr) | 10 units of $t^*$ (hr) | 100 units of $t^*$ (hr) | 1 unit of $t^{\ast}$ | 10 units of $t^{\ast}$ | 100 units of $t^{\ast}$ |
+| ------------------- | -------------------- | ---------------------- | ----------------------- | -------------------- | ---------------------- | ----------------------- |
+| $10^{-1}$           | 10                   | 100                    | 1000                    | 1                    | 10                     | 100                     |
+| $10^0$              | 1                    | 10                     | 100                     | 1                    | 10                     | 100                     |
+| $10^1$              | 0.1                  | 1                      | 10                      | 1                    | 10                     | 100                     |
+| $10^2$              | 0.01                 | 0.1                    | 1                       | 1                    | 10                     | 100                     |
+
+## Dimensionless dynamics
+
+The system of equations redefined in terms of $U_0 = \frac{\alpha}{\beta}$ and $\beta$ as the reference scale for time,
+
+| Dimensionless | Definition           |
+| -------------------------------------- | -------------------- |
+| $u^{\ast}$ , $s^{\ast}$                | $u / U_0$, $s / U_0$ |
+| $t^*$                                  | $\beta t$            |
+| $\gamma^*$                             | $\gamma / \beta$     |
+
+is given by
+
+\begin{align}
+\frac{du^{\ast}}{dt^{\ast}} & = 1 - u^{\ast}  \label{eq-dustardtstar}, \\
+\frac{ds^{\ast}}{dt^{\ast}} & = u^{\ast} - \gamma^{\ast} s^{\ast},  
+  \label{eq-dsstardtstar}
+\end{align}
+
+containing the 
+$6 - 2 = 4$ variables and parameters: $u^{\ast}, s^{\ast}, t^{\ast}, \gamma^{\ast}$.
+
+[@Zeisel2011-nz;@Cantwell2002-dv]
+
+## Analytical solution
+
+The complete solutions for the dimensionless system for $\gamma^{\ast} \neq 1$ and 
+$\xi = \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1}$
+are then given by
+
+\begin{align}
+u^{\ast}(t^{\ast}) & = \underbrace{1}_{\text{Steady State}} + 
+  \underbrace{(u^{\ast}_0 - 1) e^{-t^{\ast}}}_{\text{Transient Component}}, 
+  \label{eq-ustar-dimless} \\
+s^{\ast}(t^{\ast}) & = \underbrace{\frac{1}{\gamma^{\ast}}}_{\text{Steady State}} + 
+  \underbrace{\left( s^{\ast}_0 - \xi - \frac{1}{\gamma^{\ast}} \right) 
+  e^{-\gamma^{\ast} t^{\ast}} + \xi e^{-t^{\ast}}}_{\text{Transient Components}}.  
+  \label{eq-sstar-dimless}
+\end{align}
+
+For the case $\gamma^{\ast} = 1$, we find $u^{\ast}$ is unchanged
+and $s^{\ast}$ is then given by
+
+$$
+s^{\ast}(t^{\ast}) = 1 + (s^{\ast}_0 - 1) e^{-t^{\ast}} + 
+  (u^{\ast}_0 - 1) t^{\ast} e^{-t^{\ast}}.
+$$
+
+Note we can restore an arbitrary initial timepoint putting 
+$t^{\ast} \rightarrow \Delta t = t^{\ast} - t^{\ast}_0$.
+
+## Simulation
+
+```{python}
+# | label: transcription-splicing-degradation-simulation
+# | code-fold: true
+import diffrax
+from pyrovelocity.models import solve_transcription_splicing_model
+from pyrovelocity.models import solve_transcription_splicing_model_analytical
+from pyrovelocity.logging import configure_logging
+from jax import numpy as jnp
+
+logger = configure_logging("nbs")
+
+ts0 = jnp.linspace(0.1, 4.0, 40)
+ts1 = jnp.linspace(4.0 + (10.0 - 4.0) / 20, 10.0, 20)
+ts = jnp.concatenate([ts0, ts1])
+initial_state = jnp.array([0.1, 0.1])
+params = (1.00,)
+colormap_name = "RdBu"
+
+solution = solve_transcription_splicing_model(
+    ts,
+    initial_state,
+    params,
+)
+analytical_solution = solve_transcription_splicing_model_analytical(
+    ts,
+    initial_state,
+    params,
+)
+analytical_simulation_error = diffrax.Solution(
+    t0=ts[0],
+    t1=ts[-1],
+    ts=ts,
+    ys=analytical_solution.ys - solution.ys,
+    interpolation=None,
+    stats={},
+    result=diffrax.RESULTS.successful,
+    solver_state=None,
+    controller_state=None,
+    made_jump=None,
+)
+
+logger.info(
+    f"\nTrajectory preview times:\n{solution.ts[:3]}\n"
+    f"\nSimulated trajectory values:\n{solution.ys[:3]}\n"
+    f"\nAnalytical trajectory values:\n{analytical_solution.ys[:3]}\n\n"
+    f"\nTrajectory preview times:\n{solution.ts[-3:]}\n"
+    f"\nSimulated trajectory values:\n{solution.ys[-3:]}\n"
+    f"\nAnalytical trajectory values:\n{analytical_solution.ys[-3:]}\n\n"
+)
+```
+
+## Visualization {.center}
+
+```{python}
+# | label: fig-transcription-splicing-degradation-plot-simulation-trajectory-1
+# | code-fold: true
+# | fig-cap: Simulated trajectories of pre-mRNA and mRNA concentrations over time for the transcription-splicing-degradation model.
+from pyrovelocity.plots import plot_deterministic_simulation_trajectories
+
+plot_deterministic_simulation_trajectories(
+    solution=solution,
+    title_prefix="TSD Model Simulated",
+    colormap_name=colormap_name,
+)
+```
+
+## Error analysis {.center}
+
+```{python}
+# | label: fig-transcription-splicing-degradation-plot-analytical-trajectory-1
+# | code-fold: true
+# | fig-cap: Absolute error in analytical vs simulated trajectories of pre-mRNA and mRNA concentrations over time for the transcription-splicing-degradation model.
+from pyrovelocity.plots import plot_deterministic_simulation_trajectories
+
+plot_deterministic_simulation_trajectories(
+    solution=analytical_simulation_error,
+    title_prefix="Analytical-Simulation Error",
+    colormap_name=colormap_name,
+)
+```
+
+## Phase space {.center}
+
+```{python}
+# | label: fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1
+# | code-fold: true
+# | fig-cap: Phase portrait of the transcription-splicing-degradation model.
+from pyrovelocity.plots import plot_deterministic_simulation_phase_portrait
+
+plot_deterministic_simulation_phase_portrait(
+    solution=solution,
+    title_prefix="TSD Model",
+    colormap_name=colormap_name,
+)
+```
+
+# Probabilistic inference in dynamical systems
+
+## Data description
+
+If we are given samples of pre-mRNA and mRNA counts along with
+the sampling time of each cell, we can describe the data set 
+$\mathcal{D}$ consisting of pre-mRNA and mRNA counts for a number 
+of genes $G$ across a number of cells $N$
+
+\begin{align}
+\mathcal{D} = \left\{ (t_j, u_{ij}, s_{ij}) \mid 
+  i \in \{1, \ldots, G\}, j \in \{1, \ldots, N\} \right\} \label{eq-dataset},
+\end{align}
+
+where
+
+- $t_j$ represents the sampling time for cell $j$,
+- $u_{ij}$ represents the count of pre-mRNA for gene $i$ in cell $j$,
+- $s_{ij}$ represents the count of mRNA for the same gene $i$ in the same cell $j$,
+- $G$ is the total number of genes in the study,
+- $N$ is the total number of cells sampled.
+
+If we were able to observe each cell $j$ at multiple timepoints, we could add
+another index $k \in {1, ..., K_j}$.
+
+## Inference problem
+
+Roughly speaking, the inference problem involves
+
+1. **Normalization**: Transforming the observed counts $u_{ij}$ and $s_{ij}$ to 
+   their dimensionless counterparts $u^{\ast}_{ij}$ and $s^{\ast}_{ij}$ using 
+   a reasonable concentration scale $U_0$.
+2. **Model Fitting**: Applying statistical methods to estimate the latent 
+   variables 
+   $\Theta = \left( \gamma^{\ast}, u^{\ast}_0, s^{\ast}_0 \right)$ 
+   that best fit the observed data.
+3. **Evaluation**: Assessing the fit of the model and the estimated parameters'
+   biological plausibility and consistency across different genes and cells.
+
+We will return later to the alternative version of this problem where $t_j$ 
+are not observed and must be inferred along with the latent variables 
+$\Theta = \left( \gamma^{\ast}, u^{\ast}_0, s^{\ast}_0, t^{\ast} \right)$.
+
+## Model formulation
+
+given a number of 
+genes $G$ and cells $N$ that are measured at $K_j$ distinct times, then for each
+$(i,j,k) \in \{1, \ldots, G\} \otimes \{1, \ldots, N\} \otimes \{1, \ldots, K_j\}$ 
+we have
+
+\begin{align}
+    u^{\ast}_{0i}, s^{\ast}_{0i} &\sim 
+      \text{LogNormal}(\mu_{0}, \sigma_{0}^2), 
+      \label{eq-init-conds-priors-gene} \\
+
+    \gamma^{\ast}_i &\sim 
+      \text{LogNormal}(\mu_{\gamma}, \sigma_{\gamma}^2),
+      \label{eq-gamma-prior-gene} \\
+
+    \sigma_{ui}, \sigma_{si} &\sim 
+      \text{HalfNormal}(\mu_{\sigma}, \sigma_{\sigma}^2),
+      \label{eq-noise-std-priors-gene} \\
+
+    {u^{\ast}}^{k}_{ij} &= 
+      1 + (u^{\ast}_{0i} - 1) \cdot e^{-{t^{\ast}}^k_j},
+      \label{eq-u-star-model-cell} \\
+
+    {s^{\ast}}^{k}_{ij} &= 
+      \begin{cases}
+        \frac{1}{\gamma^{\ast}_i} + 
+        \left( s^{\ast}_{0i} - \xi_i - \frac{1}{\gamma^{\ast}_i} \right) \cdot e^{-\gamma^{\ast}_i {t^{\ast}}^k_j} + 
+        \xi_i \cdot e^{-{t^{\ast}}^k_j},& \gamma^{\ast}_i \neq 1 \\
+
+        1 + (s^{\ast}_{0i} - 1) e^{-{t^{\ast}}^k_j} + 
+          (u^{\ast}_{0i} - 1) {t^{\ast}}^k_j e^{-{t^{\ast}}^k_j},& \gamma^{\ast}_i = 1 \\
+      \end{cases}, \label{eq-s-star-model-cell} \\
+
+    \hat{u}^{\ast}{}^k_{ij} &\sim \text{Normal}({u^{\ast}}^k_{ij}, \sigma_{ui}^2), 
+    \label{eq-u-obs-cell} \\
+
+    \hat{s}^{\ast}{}^k_{ij} &\sim \text{Normal}({s^{\ast}}^k_{ij}, \sigma_{si}^2). 
+    \label{eq-s-obs-cell}
+\end{align}
+
+## Graphical model
+
+```{python}
+# | label: fig-transcription-splicing-degradation-graphical-model-1
+# | code-fold: true
+# | fig-cap: Graphical representation of the transcription-splicing-degradation inference model.
+import daft
+import matplotlib.pyplot as plt
+
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["font.size"] = 16
+plt.rcParams["text.usetex"] = True
+
+pgm = daft.PGM(line_width=1.2, dpi=85)
+
+# hyperparameters
+pgm.add_node("mu_init", r"$\mu_{0}$", 0.5, 6, fixed=True)
+pgm.add_node("sigma_init", r"$\sigma_{0}^2$", 1.5, 6, fixed=True)
+pgm.add_node("mu_gamma", r"$\mu_{\gamma}$", 2.5, 6, fixed=True)
+pgm.add_node("sigma_gamma", r"$\sigma_{\gamma}^2$", 3.5, 6, fixed=True)
+pgm.add_node("mu_sigma", r"$\mu_{\sigma}$", 4.5, 6, fixed=True)
+pgm.add_node("sigma_sigma", r"$\sigma_{\sigma}^2$", 5.5, 6, fixed=True)
+
+# latent variables for gene-specific parameters
+pgm.add_node("u_star_0i", r"$u^{\ast}_{0i}$", 1, 5)
+pgm.add_node("s_star_0i", r"$s^{\ast}_{0i}$", 2, 5)
+pgm.add_node("gamma_star_i", r"$\gamma^{\ast}_i$", 3, 5)
+pgm.add_node("sigma_ui", r"$\sigma_{ui}$", 4, 5)
+pgm.add_node("sigma_si", r"$\sigma_{si}$", 5, 5)
+
+# latent variables for cell-specific outcomes
+pgm.add_node(
+    "u_star_ij",
+    r"${u^{\ast}}^k_{ij}$",
+    2,
+    3.8,
+    scale=1.0,
+    shape="rectangle",
+)
+pgm.add_node(
+    "s_star_ij",
+    r"${s^{\ast}}^k_{ij}$",
+    4,
+    3.8,
+    scale=1.0,
+    shape="rectangle",
+)
+
+# observed data
+pgm.add_node(
+    "t_star_j",
+    r"${t^{\ast}}^k_j$",
+    5.9,
+    3.1,
+    observed=True,
+    shape="rectangle",
+)
+pgm.add_node(
+    "u_obs_ij",
+    r"$\hat{u}^{\ast}{}^{k}_{ij}$",
+    2,
+    2.4,
+    scale=1.0,
+    observed=True,
+)
+pgm.add_node(
+    "s_obs_ij",
+    r"$\hat{s}^{\ast}{}^{k}_{ij}$",
+    4,
+    2.4,
+    scale=1.0,
+    observed=True,
+)
+
+# edges
+edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}
+pgm.add_edge("mu_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("mu_sigma", "sigma_ui", plot_params=edge_params)
+pgm.add_edge("sigma_sigma", "sigma_ui", plot_params=edge_params)
+pgm.add_edge("mu_sigma", "sigma_si", plot_params=edge_params)
+pgm.add_edge("sigma_sigma", "sigma_si", plot_params=edge_params)
+
+pgm.add_edge("u_star_0i", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("s_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("u_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)
+
+pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("sigma_ui", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("sigma_si", "s_obs_ij", plot_params=edge_params)
+
+pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)
+
+# plates
+pgm.add_plate(
+    [0.4, 1.0, 5, 4.5],
+    label=r"$i \in \{1, \ldots, G\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+pgm.add_plate(
+    [0.8, 1.4, 5.9, 3.2],
+    label=r"$j \in \{1, \ldots, N\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+pgm.add_plate(
+    [1.2, 1.8, 5.2, 2.5],
+    label=r"$k \in \{1, \ldots, K_j\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+
+pgm.render()
+```
+
+## Relaxing the assumption of analytical tractability
+
+In general, we will not have access to the analytical solutions, so
+we will need to simulate the system and then perform inference on the latent
+variables
+
+\begin{align}
+    u^{\ast}_0, s^{\ast}_0 &\sim \text{LogNormal}(\mu_{0}, \sigma_{0}^2) \label{eq-init-conds-priors-num}, \\
+    \gamma^{\ast} &\sim \text{LogNormal}(\mu_{\gamma}, \sigma_{\gamma}^2) \label{eq-gamma-prior-num}, \\
+    \sigma_u, \sigma_s &\sim \text{HalfNormal}(\mu_{\sigma}, \sigma_{\sigma}^2) \label{eq-noise-std-priors-num}, \\
+    (u^{\ast}, s^{\ast}) &= \text{NumericalSolver}\left(\frac{du^{\ast}}{dt^{\ast}}, \frac{ds^{\ast}}{dt^{\ast}}, u^{\ast}_0, s^{\ast}_0, \gamma^{\ast}, t^{\ast}\right) \label{eq-numerical-solution}, \\
+    \hat{u}^{\ast} &\sim \text{Normal}(u^{\ast}, \sigma_u^2) \label{eq-u-obs-num}, \\
+    \hat{s}^{\ast} &\sim \text{Normal}(s^{\ast}, \sigma_s^2) \label{eq-s-obs-num}.
+\end{align}
+
+Where $\frac{du^{\ast}}{dt^{\ast}}$ and $\frac{ds^{\ast}}{dt^{\ast}}$ 
+are given and
+we suppress the indices from the previous model for brevity.
+
+## Relaxing the assumption of access to time series data
+
+If we only observe one time point for each cell, 
+then the model reduces to the graphical model in 
+@fig-transcription-splicing-degradation-graphical-model-2.
+
+```{python}
+# | label: fig-transcription-splicing-degradation-graphical-model-2
+# | code-fold: true
+# | fig-cap: Graphical model for the transcription-splicing-degradation model observed at a single time point.
+import daft
+import matplotlib.pyplot as plt
+
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["font.size"] = 16
+plt.rcParams["text.usetex"] = True
+
+pgm = daft.PGM(line_width=1.2, dpi=80)
+
+# hyperparameters
+pgm.add_node("mu_init", r"$\mu_{0}$", 0.5, 6, fixed=True)
+pgm.add_node("sigma_init", r"$\sigma_{0}^2$", 1.5, 6, fixed=True)
+pgm.add_node("mu_gamma", r"$\mu_{\gamma}$", 2.5, 6, fixed=True)
+pgm.add_node("sigma_gamma", r"$\sigma_{\gamma}^2$", 3.5, 6, fixed=True)
+pgm.add_node("mu_sigma", r"$\mu_{\sigma}$", 4.5, 6, fixed=True)
+pgm.add_node("sigma_sigma", r"$\sigma_{\sigma}^2$", 5.5, 6, fixed=True)
+
+# latent variables for gene-specific parameters
+pgm.add_node("u_star_0i", r"$u^{\ast}_{0i}$", 1, 5)
+pgm.add_node("s_star_0i", r"$s^{\ast}_{0i}$", 2, 5)
+pgm.add_node("gamma_star_i", r"$\gamma^{\ast}_i$", 3, 5)
+pgm.add_node("sigma_ui", r"$\sigma_{ui}$", 4, 5)
+pgm.add_node("sigma_si", r"$\sigma_{si}$", 5, 5)
+
+# latent variables for cell-specific outcomes
+pgm.add_node(
+    "u_star_ij",
+    r"$u^{\ast}_{ij}$",
+    2,
+    4,
+    scale=1.0,
+    shape="rectangle",
+)
+pgm.add_node(
+    "s_star_ij",
+    r"$s^{\ast}_{ij}$",
+    4,
+    4,
+    scale=1.0,
+    shape="rectangle",
+)
+
+# observed data
+pgm.add_node(
+    "t_star_j",
+    r"$t^{\ast}_j$",
+    6.0,
+    3.25,
+    observed=True,
+    shape="rectangle",
+)
+pgm.add_node(
+    "u_obs_ij",
+    r"$\hat{u}^{\ast}_{ij}$",
+    2,
+    2.5,
+    scale=1.0,
+    observed=True,
+)
+pgm.add_node(
+    "s_obs_ij",
+    r"$\hat{s}^{\ast}_{ij}$",
+    4,
+    2.5,
+    scale=1.0,
+    observed=True,
+)
+
+# edges
+edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}
+pgm.add_edge("mu_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("mu_sigma", "sigma_ui", plot_params=edge_params)
+pgm.add_edge("sigma_sigma", "sigma_ui", plot_params=edge_params)
+pgm.add_edge("mu_sigma", "sigma_si", plot_params=edge_params)
+pgm.add_edge("sigma_sigma", "sigma_si", plot_params=edge_params)
+
+pgm.add_edge("u_star_0i", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("s_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("u_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)
+
+pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("sigma_ui", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("sigma_si", "s_obs_ij", plot_params=edge_params)
+
+pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)
+
+# plates
+pgm.add_plate(
+    [0.5, 1.2, 5, 4.4],
+    label=r"$i \in \{1, \ldots, G\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+pgm.add_plate(
+    [1.0, 1.8, 5.5, 2.75],
+    label=r"$j \in \{1, \ldots, N\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+
+pgm.render()
+```
+
+## HMC Inference
+
+Now we can set up some sample data and perform inference on the latent variables
+
+```{python}
+# | label: fig-generate-test-data-for-deterministic-model-inference
+# | code-fold: true
+# | fig-cap: Test data
+# | output: false
+
+from pyrovelocity.models._deterministic_inference import (
+    generate_posterior_inference_data,
+    generate_prior_inference_data,
+    generate_test_data_for_deterministic_model_inference,
+    plot_sample_phase_portraits,
+    plot_sample_trajectories,
+    plot_sample_trajectories_with_percentiles,
+    print_inference_data_structure,
+)
+
+(
+    times,
+    data,
+    num_genes,
+    num_cells,
+    num_timepoints,
+    num_modalities,
+) = generate_test_data_for_deterministic_model_inference(
+    num_genes=1,
+    num_cells=20,
+    num_timepoints=1,
+    num_modalities=2,
+    noise_levels=(0.001, 0.001),
+)
+
+num_chains = 1
+num_samples = 10
+num_warmup = 10
+
+idata_prior = generate_prior_inference_data(
+    times=times,
+    data=data,
+    num_chains=num_chains,
+    num_samples=num_samples,
+    num_genes=num_genes,
+    num_cells=num_cells,
+    num_timepoints=num_timepoints,
+    num_modalities=num_modalities,
+)
+
+idata_posterior = generate_posterior_inference_data(
+    times=times,
+    data=data,
+    num_chains=num_chains,
+    num_samples=num_samples,
+    num_genes=num_genes,
+    num_cells=num_cells,
+    num_timepoints=num_timepoints,
+    num_modalities=num_modalities,
+    num_warmup=num_warmup,
+)
+```
+
+```{python}
+# | label: show-inference-data-structure
+idata_posterior
+```
+
+## Prior distributions {.center}
+
+```{python}
+# | label: fig-transcription-splicing-degradation-prior
+# | code-fold: true
+# | fig-cap:
+# |   - Initial concentratiosn
+# |   - Degradation vs splicing rate
+# |   - Observational noise levels
+# | layout-ncol: 3
+import arviz as az
+import shutil
+
+light_gray = "#bcbcbc"
+variables = ["initial_conditions", "gamma", "sigma"]
+with plt.style.context("pyrovelocity.styles.common"):
+    if not shutil.which("latex"):
+        plt.rc("text", usetex=False)
+    for var in variables:
+        az.plot_posterior(
+            idata_prior,
+            var_names=[var],
+            group="prior",
+            kind="hist",
+            color=light_gray,
+            round_to=2,
+        )
+```
+
+## Posterior distributions {.center}
+
+```{python}
+# | label: fig-transcription-splicing-degradation-posterior
+# | code-fold: true
+# | fig-cap:
+# |   - Initial concentratiosn
+# |   - Degradation vs splicing rate
+# |   - Observational noise levels
+# | layout-ncol: 3
+with plt.style.context("pyrovelocity.styles.common"):
+    if not shutil.which("latex"):
+        plt.rc("text", usetex=False)
+    for var in variables:
+        az.plot_posterior(
+            idata_posterior,
+            var_names=[var],
+            group="posterior",
+            kind="hist",
+            color=light_gray,
+            round_to=2,
+        )
+```
+
+## Posterior predictive trajectories
+
+```{python}
+# | label: fig-transcription-splicing-degradation-posterior-predictive-trajectories
+# | code-fold: true
+# | fig-cap: Posterior predictive trajectories for the transcription-splicing-degradation model.
+# | warnings: false
+with plt.style.context("pyrovelocity.styles.common"):
+    if not shutil.which("latex"):
+        plt.rc("text", usetex=False)
+    figs = plot_sample_trajectories(
+        idata=idata_posterior,
+    )
+    for fig in figs:
+        fig.show()
+```
+
+## Posterior predictive phase portraits {.center}
+
+```{python}
+# | label: fig-transcription-splicing-degradation-posterior-predictive-phase-portraits
+# | code-fold: true
+# | fig-cap: Posterior predictive phase portraits for the transcription-splicing-degradation model.
+with plt.style.context("pyrovelocity.styles.common"):
+    if not shutil.which("latex"):
+        plt.rc("text", usetex=False)
+    figs = plot_sample_phase_portraits(
+        idata=idata_posterior,
+        colormap_name="RdBu",
+    )
+    for fig in figs:
+        fig.show()
+```
+
+## Posterior predictive percentile bands
+
+```{python}
+# | label: fig-transcription-splicing-degradation-posterior-predictive
+# | code-fold: true
+# | fig-cap: Posterior predictive distribution for the transcription-splicing-degradation model.
+with plt.style.context("pyrovelocity.styles.common"):
+    if not shutil.which("latex"):
+        plt.rc("text", usetex=False)
+    figs = plot_sample_trajectories_with_percentiles(
+        idata=idata_posterior,
+    )
+    for fig in figs:
+        fig.show()
+```
+
+# Coarse-graining and renormalization {
+    data-background-image="https://upload.wikimedia.org/wikipedia/commons/b/b2/Ising_Criticality2.gif"
+    data-background-size="contain"
+    data-background-opacity="0.3"
+  }
+
+## Mathematical example
+
+{{< video https://www.youtube.com/embed/9vFbyHNz-8g?start=591 
+    width="100%"
+    height="85%"
+>}}
+
+[@Fisher1998-lm]
+
+# References
+
+<!-- :::: {.columns}
+
+::: {.column .fragment}
+
+- Content on the left
+
+:::
+
+::: {.column .fragment}
+- Content on the right
+
+:::
+
+:::: -->
+
+<!-- # Simulation ⇄ Inference {
     data-background-image="https://upload.wikimedia.org/wikipedia/commons/b/b2/Ising_Criticality2.gif"
@@
 - This is done by computing the posterior distribution of the parameters given the data.
-- Inside the model, we may estimate the parameters of other models.
-
-# References
+- Inside the model, we may estimate the parameters of other models. -->

### 2024-07-15 nbs/index.qmd
docs(index): update cid badge branch
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 [bear-link]: https://beartype.readthedocs.io
-[cid-badge]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cid.yaml/badge.svg?branch=master
+[cid-badge]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cid.yaml/badge.svg?branch=main
 [cid-link]: https://github.com/pinellolab/pyrovelocity/actions/workflows/cid.yaml

### 2024-07-15 nbs/guides/installation.qmd
docs(guides): update installation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+::: {#nte-blank-cells .callout-note collapse=true title="Architecture and accelerator support"}
+If you would like to test pyrovelocity without requiring a GPU you should be
+able to install it on any platform. However, if you would like to apply it to
+data sets larger than $O(10^3)$ cells, you will need a GPU.
+In this case, we can only support amd64 Linux systems with a CUDA compatible GPU.
+If you are having issues scaling pyrovelocity please do not hesitate to reach out to us.
+:::
+
+The latest [release of pyrovelocity](https://github.com/pinellolab/pyrovelocity/releases)
+is [available on PyPI](https://pypi.org/project/pyrovelocity/) and 
+[conda-forge](https://anaconda.org/conda-forge/pyrovelocity).
+The instructions below are intended for users. To setup a development 
+environment, please see [Contributing](../about/contributing.qmd).
+Users who are having difficulty with system-level dependencies may also
+benefit from consulting the [contributing documentation](../about/contributing.qmd).
+
+::: {.panel-tabset}
+
 ## PyPI
 
+We recommend creating a dedicated virtual environment to install and execute pyrovelocity.
+One of the easiest ways to do this is with [uv](https://github.com/astral-sh/uv/tree/0.2.24?tab=readme-ov-file#getting-started).
+After following the [Getting Started instructions from the uv repository](https://github.com/astral-sh/uv/tree/0.2.24?tab=readme-ov-file#getting-started)
+to install the uv CLI for your platform, you can install a supported python
+version, create a virtual environment, and install pyrovelocity from a blank directory with the following commands
+
+```bash
+echo "3.10" > .python-version
+uv python install
+uv venv -p 3.10
+uv pip install pyrovelocity
+```
+
+Of course, if you have already prepared and activated a virtual environment
+with a supported python version you can also install pyrovelocity directly
+from [PyPI](https://pypi.org/project/pyrovelocity/) with pip.
+
+```bash
+pip install pyrovelocity
+```
+
 ## Anaconda
+
+If you would like to install pyrovelocity from [conda-forge](https://anaconda.org/conda-forge/pyrovelocity)
+we strongly recommend using [pixi](https://prefix.dev/channels/conda-forge/packages/pyrovelocity)
+or [micromamba](https://mamba.readthedocs.io/en/latest/installation/micromamba-installation.html).
+Installation with the `conda` command will take an unreasonable amount of time and is not recommended.
+
+::: {.panel-tabset}
+
+### pixi
+
+`pixi` will not automatically respect the user limit parameters you can
+review via `ulimit -a`.
+In particular `ulimit -n` should be set to 10000 or greater
+
+```bash
+ulimit -n 10000
+```
+
+After following the instructions to install [pixi](https://prefix.dev/channels/conda-forge/packages/pyrovelocity),
+you can install pyrovelocity with the following commands
+
+```bash
+pixi init
+pixi install pyrovelocity
+```
+
+You can then run `pixi shell` to enter a shell with the environment containing pyrovelocity.
+
+### micromamba
+
+After following the instructions to install [micromamba](https://mamba.readthedocs.io/en/latest/installation/micromamba-installation.html),
+you can create a virtual environment, activate it, and install pyrovelocity with the following commands
+
+```bash
+alias mm=micromamba
+mm create -n pyrovelocity python=3.10
+mm activate pyrovelocity
+mm install -c conda-forge pyrovelocity
+```
+
+:::
+
+:::

### 2024-07-15 nbs/about/contributing.qmd
docs(about): update contributing
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-Contributions are welcome, greatly appreciated, and credit will always be given.
+If you would like to simply get started developing please feel free to skip to the
+[development environment](#development-environment) section below.
 
-You can contribute in many ways:
+Contributions are welcome, greatly appreciated, and will be appropriately acknowledged.
+If you are new to contributing to open source software please consider reviewing the
+[open source guides](https://opensource.guide/) and the brief descriptions below.
+Briefly, please feel free to engage a discussion or github issue, fork the repository,
+create a branch, make changes, ensure the code is linted and the tests pass, add documentation,
+where relevant, and submit a pull request. At this point the continuous integration
+workflows will run and we will engage a more detailed discussion in reviewing the changes.
 
@@
 
-### Fix Bugs
+### Respond to existing issues
 
-Look through the GitHub issues for bugs. Anything tagged with "bug" and "help
-wanted" is open to whoever wants to implement it.
-
-### Implement Features
-
-Look through the GitHub issues for features. Anything tagged with "enhancement"
-and "help wanted" is open to whoever wants to implement it.
+Review the history of both open and closed GitHub issues.
+If you would like to attempt to resolve an open issue please comment on the issue,
+fork the repository, create a branch with name format `##-subtitle`, where `##` is the issue number
+and `subtitle` is a word that is related to the issue.
+Run experiments and write tests from your local fork and then submit a pull request referencing
+the issue.
 
@@
 
-pyrovelocity could always use more documentation, whether as part of the
-official pyrovelocity docs, in docstrings, or even on the web in blog posts,
-articles, and such.
+If you would like to add or contribute to refining the documentation, 
+please create an issue and submit a pull request with modifications to the
+[documentation site](https://github.com/pinellolab/pyrovelocity/tree/main/nbs), 
+docstrings and doctests, or adding illustrative pytest tests.
+We attempt to loosely follow the [diataxis](https://diataxis.fr/) recommendations 
+for structuring technical documentation.
 
@@
 
-## Get Started
-
-### Cloud
-
-See the [reproducibility/environment](https://github.com/pinellolab/pyrovelocity/tree/main/reproducibility/environment) folder and associated README.md.
+## Development environment
 
@@
 The following is a rough guide to setting up `pyrovelocity` for local development.
+Please do not hesitate to reach out if you need help setting up a development environment.
 
-1. Fork the `pyrovelocity` repo on GitHub.
+1. Fork the [pyrovelocity](https://github.com/pinellolab/pyrovelocity)
+   repository on GitHub.
 
-2. Clone your fork locally:
+1. Clone your fork locally:
 
@@
 
-3. Install your local copy with [poetry and nox](https://cookiecutter-hypermodern-python.readthedocs.io/en/2022.6.3.post1/guide.html#requirements) or with [conda]().
+1. Install [poetry](https://python-poetry.org/docs/#installing-with-the-official-installer) and 
+   install the repository source code in editable mode:
 
-4. Create a branch for local development:
+   ```bash
+   poetry install --sync
+   ```
 
-   ```
-   $ git checkout -b name-of-your-bugfix-or-feature
+1. Create a branch for local development:
+
+   ```bash
+   git checkout -b 572-branch
    ```
@@
    Now you can make your changes locally.
+   You may need to familiarize yourself with using poetry to manage virtual environments and
+   dependencies.
 
-5. When you're done making changes, you can check that your changes pass the
-   most basic checks implemented in noxfile.py (run `nox --list-sessions` to list all available):
+1. When you're done making changes, lint the code
 
-   ```
-   $ nox -x -rs pre-commit
-   $ nox -x -rs tests-3.10
-   $ nox -x -rs docs-build
+   ```bash
+   make lint
    ```
 
-   These will be confirmed via the GitHub actions workflow that will run on your fork and pull request.
+   and run the tests
 
-6. Commit your changes and push your branch to GitHub:
-
-   ```
-   $ git add .
-   $ git commit -m "Your detailed description of your changes."
-   $ git push origin name-of-your-bugfix-or-feature
+   ```bash
+   pytest
    ```
 
-7. Submit a pull request through the GitHub website.
+   These will be confirmed via the GitHub actions workflow that will
+   run when you submit your pull request.
+
+1. Commit your changes and push your branch to GitHub:
+
+   ```bash
+   git add .
+   git commit -m "Your detailed description of your changes."
+   git push origin name-of-your-bugfix-or-feature
+   ```
+
+1. Submit a pull request through the GitHub CLI or website targeting the `beta` branch.
+
+   ```bash
+   gh pr create \
+   -d \
+   -a "@me" \
+   -B beta \
+   -t "fix(model): add types for train_model function" \
+   -r username \
+   -b "- [x] added types to all parameters of the train_model function
+   - [x] enabled beartype and refactored associated tests
+   "
+   ```
+
+Using [nix](https://github.com/NixOS/nix), [home-manager](https://github.com/nix-community/home-manager), and [nixpkgs](https://github.com/NixOS/nixpkgs) [dockerTools](https://nixos.org/manual/nixpkgs/stable/#sec-pkgs-dockerTools), [nixpod](https://ghcr.io/cameronraysmith/nixpod) provides a containerized or containerizable drop-in user configuration on any platform where the [nix](https://github.com/NixOS/nix) package manager is already, or can be, [installed](https://nix.dev/install-nix.html).
+This is intended to include, but is not limited to, scenarios like those involving [kubernetes ephemeral containers](https://kubernetes.io/docs/concepts/workloads/pods/ephemeral-containers/) via images like [netshoot](https://github.com/nicolaka/netshoot), which might be used for debugging purposes adjacent to otherwise minimal container images.
+See the [github packages associated to this repository](https://github.com/cameronraysmith?tab=packages&repo_name=nixpod) for the containers currently built in the [nixpod CID](https://github.com/cameronraysmith/nixpod/blob/main/.github/workflows/cid.yaml).
+
+This repository originally contained a [nix flake](https://zero-to-nix.com/concepts/flakes) that essentially integrated a few parts of [srid/nixos-config](https://github.com/srid/nixos-config) into [juspay/nix-dev-home](https://github.com/juspay/nix-dev-home).
+These were merged upstream in [juspay/nix-dev-home#7](https://github.com/juspay/nix-dev-home/pull/7), so you might want to look there.
+
+### Nix
+
+The instructions above only cover python dependencies.
+We use the [nix](https://github.com/NixOS/nix) package manager with a [nix flake](https://nix.dev/concepts/flakes) to manage both system 
+dependencies and python dependencies together in a reproducible manner.
+The nix flake integrates with [direnv](https://github.com/direnv/direnv) and provides a devlopment shell that can be activated with
+
+```bash
+direnv allow
+```
+
+if you have [nix](https://nix.dev/install-nix.html) and direnv installed.
+
+### Container image
+
+We provide a relatively minimal [OCI image](https://github.com/opencontainers/image-spec)
+built with [nix](#nix) and [dockerTools](https://nixos.org/manual/nixpkgs/stable/#sec-pkgs-dockerTools).
+It only supports linux x86-64, but can be run with rosetta on macOS.
+
+For example if you have a tool that provides a container runtime like nerdctl, podman, or docker installed
+
+```bash
+docker run --rm -it --entrypoint bash --platform=linux/amd64 ghcr.io/pinellolab/pyrovelocity:latest
+```
+
+### Cloud
+
+If you would like to access the cloud environment used to develop `pyrovelocity`,
+execute model training across many data set and model configurations, and track
+associated artifacts, please do not hesitate to reach out to us.
+We have limited availability, but would like to provide these resources to interested
+external contributors.
 
@@
 
-1. The pull request should include pytest tests and xdoctests.
-2. If the pull request adds functionality, the docs should be updated. Put
-   your new functionality into a function with a docstring.
-3. The pull request should work for Python 3.10.
+1. The pull request should include [pytest](https://docs.pytest.org/) tests and, 
+   where relevant, [xdoctest](https://github.com/Erotemic/xdoctest) 
+   [doctests](https://docs.python.org/3/library/doctest.html).
+2. If the pull request adds functionality, the 
+3. [Google-style docstrings](https://sphinxcontrib-napoleon.readthedocs.io/en/latest/example_google.html)
+   and [docs](https://github.com/pinellolab/pyrovelocity/tree/main/nbs) should be updated.
+5. The pull request should pass the CI execution for the primary supported python version (currently Python 3.10).

### 2024-07-15 nbs/tutorials/introduction.qmd
docs(tutorials): update introduction
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-### Quick start
+For end-user oriented (i.e. not for development but rather execution)
+installation please see the [installation guide](../guides/installation.qmd).
+For development installation please see the [contributing guide](../about/contributing.qmd).
+
+<!-- ### Quick start
 
@@
 
-Please see the [documentation](https://pyrovelocity.readthedocs.io/en/latest/installation.html).
+Please see the [documentation](https://pyrovelocity.readthedocs.io/en/latest/installation.html). -->
 
@@
 
-Also see [contributing](./docs/contributing.md).
+Also see [contributing](../about/contributing.qmd).

### 2024-07-15 nbs/guides/installation.qmd
nit(guides): fix marker name
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-::: {#nte-blank-cells .callout-note collapse=true title="Architecture and accelerator support"}
+::: {#nte-architecture .callout-note collapse=true title="Architecture and accelerator support"}
 If you would like to test pyrovelocity without requiring a GPU you should be

### 2024-07-15 nbs/about/contributing.qmd
nit(contributing): remove extra PR list element
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 workflows will run and we will engage a more detailed discussion in reviewing the changes.
+
+::: {#nte-contributing .callout-note collapse=true title="Current state of the repository"}
+Please note that pyrovelocity is still in the early stages of development.
+The current state of the documentation and repository are not complete
+with respect to the procedures for contributing outlined here; however,
+we plan to respect them as we continue to develop the project.
+:::
 
@@
    [doctests](https://docs.python.org/3/library/doctest.html).
-2. If the pull request adds functionality, the 
-3. [Google-style docstrings](https://sphinxcontrib-napoleon.readthedocs.io/en/latest/example_google.html)
-   and [docs](https://github.com/pinellolab/pyrovelocity/tree/main/nbs) should be updated.
-5. The pull request should pass the CI execution for the primary supported python version (currently Python 3.10).
+1. If the pull request adds or significantly modifies functionality, the relevant
+   [Google-style docstrings](https://sphinxcontrib-napoleon.readthedocs.io/en/latest/example_google.html)
+   and [docs](https://github.com/pinellolab/pyrovelocity/tree/main/nbs) should likely be updated.
+1. The pull request should pass the CI execution for the primary supported python version (currently Python 3.10).

### 2024-07-15 nbs/guides/installation.qmd
docs(guides): add note on dependency versions
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 benefit from consulting the [contributing documentation](../about/contributing.qmd).
+
+::: {#nte-versions .callout-important collapse=true title="Installation or usage errors and version constraints"}
+If you have any issue with errors being raised during installation or usage
+please check carefully that the version your package manager has installed is equivalent
+to or newer than the one listed in the
+[poetry.lock](https://github.com/pinellolab/pyrovelocity/blob/main/poetry.lock)
+file in the repository.
+We aim to support all latest stable versions of third-party dependencies
+by lifting lower bounds on version constraints in the 
+[pyproject.toml](https://github.com/pinellolab/pyrovelocity/blob/main/pyproject.toml) file.
+This means that we generally do not support any older versions of dependencies
+but consider any version more recent than the lower bound to be supported or a bug 
+in pyrovelocity if attempting to use it produces an error.
+Sometimes we are forced to delay support for a new version of a dependency
+if it has been released released to PyPI but not yet to [conda-forge](https://conda-forge.org/packages/).
+:::
 

### 2024-07-15 nbs/about/contributing.qmd
docs(contributing): improve notes regarding devshell usage
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-Using [nix](https://github.com/NixOS/nix), [home-manager](https://github.com/nix-community/home-manager), and [nixpkgs](https://github.com/NixOS/nixpkgs) [dockerTools](https://nixos.org/manual/nixpkgs/stable/#sec-pkgs-dockerTools), [nixpod](https://ghcr.io/cameronraysmith/nixpod) provides a containerized or containerizable drop-in user configuration on any platform where the [nix](https://github.com/NixOS/nix) package manager is already, or can be, [installed](https://nix.dev/install-nix.html).
-This is intended to include, but is not limited to, scenarios like those involving [kubernetes ephemeral containers](https://kubernetes.io/docs/concepts/workloads/pods/ephemeral-containers/) via images like [netshoot](https://github.com/nicolaka/netshoot), which might be used for debugging purposes adjacent to otherwise minimal container images.
-See the [github packages associated to this repository](https://github.com/cameronraysmith?tab=packages&repo_name=nixpod) for the containers currently built in the [nixpod CID](https://github.com/cameronraysmith/nixpod/blob/main/.github/workflows/cid.yaml).
-
-This repository originally contained a [nix flake](https://zero-to-nix.com/concepts/flakes) that essentially integrated a few parts of [srid/nixos-config](https://github.com/srid/nixos-config) into [juspay/nix-dev-home](https://github.com/juspay/nix-dev-home).
-These were merged upstream in [juspay/nix-dev-home#7](https://github.com/juspay/nix-dev-home/pull/7), so you might want to look there.
-
 ### Nix
 
-The instructions above only cover python dependencies.
+If you do not know what [nix](https://nixos.org/) is and are interested in using this section, 
+please see [this video for a brief introduction to nix flakes](https://youtu.be/JCeYq72Sko0?si=adRGX0YYm1Ejmd4N&t=96).
+
+One of the reasons you might be interested is that the instructions above only cover python dependencies.
 We use the [nix](https://github.com/NixOS/nix) package manager with a [nix flake](https://nix.dev/concepts/flakes) to manage both system 
 dependencies and python dependencies together in a reproducible manner.
-The nix flake integrates with [direnv](https://github.com/direnv/direnv) and provides a devlopment shell that can be activated with
+This requires referencing the lock files in [flake.lock](https://github.com/pinellolab/pyrovelocity/blob/main/flake.lock)
+and [poetry.lock](https://github.com/pinellolab/pyrovelocity/blob/main/poetry.lock) with the bridge provided by 
+[poetry2nix](https://github.com/nix-community/poetry2nix).
+At present there isn't really another python package management system with a sufficiently
+detailed lock file to ensure that the python dependencies are reproducibly installable in concert with
+the system-level dependencies. Please see the rejected [PEP 665](https://peps.python.org/pep-0665/) 
+and other discussions that reference it as this may change in the near future.
+Taken together, this is why we currently use poetry to manage python dependencies in development environments.
+
+Before proceeding ensure you have deactivated any virtual 
+environments created with python-ecosystem tools.
+The nix flake integrates with [direnv](https://github.com/direnv/direnv) and provides a
+[nix devlopment shell](https://github.com/pinellolab/pyrovelocity/blob/v0.2.0/flake.nix#L185) 
+that can be activated with
 
@@
 
-if you have [nix](https://nix.dev/install-nix.html) and direnv installed.
+if you have [nix](https://nix.dev/install-nix.html) and [direnv](https://github.com/direnv/direnv/tree/v2.34.0#basic-installation) installed.
+Please note this will require at least 10GB of free hard disk space to build and activate the environment.
+However, if your disk is within 50GB of being full, you may encounter a slow build process.
+Since you will need to enable the usage of nix flakes, you should either install
+the nix package manager with the 
+[Determinate Systems nix installer](https://github.com/DeterminateSystems/nix-installer/tree/v0.20.1?tab=readme-ov-file#installation-differences)
+or update your `/etc/nix/nix.conf` or other active nix configuration to enable 
+the `nix-command` and `flakes` features.
+
+```nix
+experimental-features = nix-command flakes
+```
+
+These features are not really "experimental" in any usual sense of the term
+so you should not be concerned about enabling them.
 
@@
 built with [nix](#nix) and [dockerTools](https://nixos.org/manual/nixpkgs/stable/#sec-pkgs-dockerTools).
-It only supports linux x86-64, but can be run with rosetta on macOS.
+It only supports linux x86-64, but can be run via rosetta emulation on macOS.
 

### 2024-07-15 nbs/about/contributing.qmd
nit(contributing): fix `development`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 The nix flake integrates with [direnv](https://github.com/direnv/direnv) and provides a
-[nix devlopment shell](https://github.com/pinellolab/pyrovelocity/blob/v0.2.0/flake.nix#L185) 
+[nix development shell](https://github.com/pinellolab/pyrovelocity/blob/v0.2.0/flake.nix#L185) 
 that can be activated with

### 2024-07-19 nbs/index.qmd
chore(nbs): update information links
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
       href: tutorials/introduction.qmd
-    - icon: github
-      text: GitHub
+    - icon: lightbulb
+      text: Usage
+      href: templates/user_example/user_example.qmd
+    - icon: box-arrow-in-right
+      text: Contributing
+      href: about/contributing.qmd
+    - icon: code-slash
+      text: Source
       href: https://github.com/pinellolab/pyrovelocity

### 2024-07-19 nbs/guides/installation.qmd
nit(installation): note requirement to activate virtual env
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
chore(nbs): interlink introductory pages
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 uv venv -p 3.10
+source .venv/bin/activate
 uv pip install pyrovelocity
@@
 :::
+
+After the installation has completed successfully you may like to see
+
+- [Introduction](/tutorials/introduction.qmd) for a brief overview
+- [Usage](/templates/user_example/user_example.qmd) for a complete usage example 
+- [Contributing](/about/contributing.qmd) for information on how to setup a development environment

### 2024-07-19 nbs/tutorials/introduction.qmd
chore(nbs): interlink introductory pages
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 It can be used as a component in frameworks that attempt to retain the ability to propagate uncertainty in predicting: distributions over cell fates from subpopulations of cell states, response to cell state perturbations, or candidate genes or gene modules that correlate with determination of specific cell fates.
+
+This document provides a brief overview of what Pyro-Velocity is and how it can be used.
+See
+
+- [Installation](/guides/installation.qmd) for installation instructions or
+- [Usage](/templates/user_example/user_example.qmd) for a complete usage example
+- [Contributing](/about/contributing.qmd) for information on how to setup a development environment
 

### 2024-07-23 src/pyrovelocity/io/gcs.py
fix(gcs): add function to download blob from uri
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,22 +1,34 @@
 from pathlib import Path
+from urllib.parse import urlparse
 
 from beartype import beartype
-from google.cloud.storage import Client
-from google.cloud.storage import transfer_manager
-from returns.result import Failure
-from returns.result import Result
-from returns.result import Success
+from google.cloud.storage import Client, transfer_manager
+from google.cloud.storage.blob import Blob
+from returns.result import Failure, Result, Success
 
 from pyrovelocity.logging import configure_logging
 
-
 __all__ = [
     "download_bucket",
+    "download_blob_from_uri",
     "upload_file_concurrently",
     "upload_directory",
 ]
 
 logger = configure_logging(__name__)
+
+
+@beartype
+def download_blob_from_uri(
+    blob_uri: str,
+):
+    client = Client()
+    parsed_blob_uri = urlparse(blob_uri)
+    blob_path = Path(parsed_blob_uri.path)
+    blob_filename = blob_path.name
+    blob = Blob.from_string(blob_uri, client)
+    blob.download_to_filename(f"./{blob_filename}")
+    print(f"Downloaded {blob_uri} to {blob_filename}.")
 
 
 @beartype
@@ -71,6 +83,26 @@ def download_bucket(
             f"Failed to download files from {bucket_name} to {destination_directory}."
         )
         return Failure(e)
+
+
+def download_blob_concurrently(
+    bucket_name: str,
+    blob_name: str,
+    filename: str | Path,
+    chunk_size: int = 32 * 1024 * 1024,
+    workers: int = 8,
+):
+    """Download a single file in chunks, concurrently in a process pool."""
+
+    storage_client = Client()
+    bucket = storage_client.bucket(bucket_name)
+    blob = bucket.blob(blob_name)
+
+    transfer_manager.download_chunks_concurrently(
+        blob, filename, chunk_size=chunk_size, max_workers=workers
+    )
+
+    print(f"Downloaded {blob_name} to {filename}.")
 
 
 @beartype

### 2024-07-23 nbs/guides/interactive.qmd
docs(guides): add interactive results review guide
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Interactive results review
---

When testing or prototyping it is helpful to review the results of a workflow
execution interactively. This is relatively straightforward when working with locally
executed workflows as demonstrated in
[Usage](/templates/user_example/user_example.qmd).
In this guide we illustrate working with the results 
of workflows that have been executed remotely. In this particular case, GitHub 
OAuth access is required to the endpoint associated to the workflow execution.
Given this access, setup the remote client,

```{python}
from flytekit.remote.remote import FlyteRemote
from flytekit.configuration import Config

remote = FlyteRemote(
    Config.for_endpoint("flyte.cluster.pyrovelocity.net"),
)
```

login to the UI and copy the URI of the workflow or task whose outputs you want to review, and provide this to the `get` method of the remote client,

```{python}
workflow_outputs = remote.get(
    "flyte://v1/pyrovelocity/development/pyrovelocity-argo-nix-bui-5c9ebf8-dev-1hk-d5805de62e8044f79d4/f26c1pjy-0-dn0-0-dn7/o"
)
```

This will not download the results but will provide a summary of the outputs.

```{python}
from omegaconf import OmegaConf
from flytekit.interaction.string_literals import literal_map_string_repr

outputs_dict = literal_map_string_repr(workflow_outputs.literals)
outputs_dictconfig = OmegaConf.create(outputs_dict)
print(OmegaConf.to_yaml(outputs_dictconfig))
```

To download the desired results use

```{python}
from pyrovelocity.io.gcs import download_blob_from_uri

download_blob_from_uri(outputs_dictconfig.o0.dataframe.path)
```

### 2024-07-25 src/pyrovelocity/io/gcs.py
test(io): add doctests to gcs module
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -23,7 +23,29 @@ logger = configure_logging(__name__)
 def download_blob_from_uri(
     blob_uri: str,
     concurrent: bool = False,
-):
+) -> str:
+    """Download a single file from a GCS bucket.
+
+    Args:
+        blob_uri (str): The URI of the GCS blob.
+        concurrent (bool, optional): Whether to download the file using
+            concurrent chunks. Defaults to False.
+
+    Example:
+        >>> tmp = getfixture("tmp_path")
+        >>> import os
+        >>> from pathlib import Path
+        >>> os.chdir(tmp)
+        >>> blob_uri = (
+        ...     "gs://gcp-public-data-landsat/LC08/01/044/034/"
+        ...     "LC08_L1GT_044034_20130330_20170310_01_T2/"
+        ...     "LC08_L1GT_044034_20130330_20170310_01_T2_MTL.txt"
+        ... )
+        >>> blob_filename = download_blob_from_uri(blob_uri)
+        >>> print(Path(tmp)/Path(blob_filename))
+        >>> blob_filename = download_blob_from_uri(blob_uri)
+        >>> print(Path(tmp)/Path(blob_filename))
+    """
     client = Client()
     parsed_blob_uri = urlparse(blob_uri)
     if not parsed_blob_uri.scheme == "gs":
@@ -32,12 +54,21 @@ def download_blob_from_uri(
         )
     blob_path = Path(parsed_blob_uri.path)
     blob_filename = blob_path.name
-    blob = Blob.from_string(blob_uri, client)
-    if concurrent:
-        download_blob_concurrently(blob, blob_filename)
+
+    if not Path(blob_filename).exists():
+        blob = Blob.from_string(blob_uri, client)
+        if concurrent:
+            download_blob_concurrently(blob, blob_filename)
+        else:
+            blob.download_to_filename(f"./{blob_filename}")
+        logger.info(f"Downloaded {blob_filename} from {blob_uri}.")
     else:
-        blob.download_to_filename(f"./{blob_filename}")
-    print(f"Downloaded {blob_uri} to {blob_filename}.")
+        logger.info(
+            f"\nFile {blob_filename} already exists.\n"
+            "The hash of the requested file has not been checked.\n"
+            "Delete and re-run to download again.\n\n"
+        )
+    return blob_filename
 
 
 @beartype
@@ -60,7 +91,7 @@ def download_bucket_from_uri(
     destination_directory: str | Path = "",
[…]

### 2024-07-25 nbs/guides/interactive/interactive.qmd
fix(nbs): move interactive guide to subfolder
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-07-23 nbs/guides/interactive.qmd
@@
 title: Interactive results review
+execute: 
+  freeze: true
+  eval: true
+  warning: false
+  error: false
+  cache: true
+toc: true
+number-sections: true
+highlight-style: pygments
+csl: bibstyle.csl
+lightbox: auto
+format:
+  html:
+    html-math-method: katex
+  ipynb: default
 ---
+
+```{python}
+#| label: enable-autoreload
+#| code-fold: true
+#| output: false
+%load_ext autoreload
+%autoreload 2 
+```
 
@@
 ```{python}
+#| label: instantiate-remote-client
 from flytekit.remote.remote import FlyteRemote
@@
 ```{python}
+#| label: get-remote-outputs
 workflow_outputs = remote.get(
-    "flyte://v1/pyrovelocity/development/pyrovelocity-argo-nix-bui-5c9ebf8-dev-1hk-d5805de62e8044f79d4/f26c1pjy-0-dn0-0-dn7/o"
+    "flyte://v1/pyrovelocity/development/pyrovelocity-argo-nix-bui-5c9ebf8-dev-1hk-d5805de62e8044f79d4/f26c1pjy-0-dn0-0-dn6/o"
 )
@@
 ```{python}
+#| label: create-outputs-dict
 from omegaconf import OmegaConf
@@
 ```{python}
+#| label: download-outputs
 from pyrovelocity.io.gcs import download_blob_from_uri
 
-download_blob_from_uri(outputs_dictconfig.o0.dataframe.path)
+pyrovelocity_data = download_blob_from_uri(
+    outputs_dictconfig.o0.pyrovelocity_data.path
+)
+postprocessed_data = download_blob_from_uri(
+    outputs_dictconfig.o0.postprocessed_data.path
+)
 ```

### 2024-07-28 nbs/concepts/inferenceindynamicalsystems/index.qmd
fix(nbs): pygments -> gruvbox syntax highlighting
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 number-sections: true
-highlight-style: pygments
+highlight-style: gruvbox
 csl: ../../bibstyle.csl

### 2024-07-28 nbs/templates/user_example/user_example.qmd
chore(nbs): interlink introductory pages
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
fix(nbs): pygments -> gruvbox syntax highlighting
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: Pyrovelocity Notebook Usage Example
+title: End-to-end notebook usage example
 toc: true
 number-sections: true
-highlight-style: pygments
+highlight-style: gruvbox
 csl: bibstyle.csl
@@
 author:
-  - name: Pyrovelocity Team
+  - name: Pyro&thinsp;-Velocity team
     email: team@pyrovelocity.net
 abstract: |
-  This notebook demonstrates how to run Pyro-Velocity in a Jupyter notebook on
+  This notebook demonstrates how to run Pyro&thinsp;-Velocity in a Jupyter notebook on
   downsampled data. The results are not expected to be meaningful, but the
@@
 Please do not hesitate to reach out in [the discussions board on github](https://github.com/pinellolab/pyrovelocity/discussions) if you run into any problems using the library.
+
+See
+
+- [Installation](/guides/installation.qmd) for installation instructions
+- [Introduction](/tutorials/introduction.qmd) for a brief overview
+- [Contributing](/about/contributing.qmd) for information on how to setup a development environment

### 2024-07-28 nbs/index.qmd
nit(nbs): fix dash spacing
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-07-15 nbs/index.qmd
@@
 ---
-title: Pyro-Velocity
+title: Pyro&thinsp;-Velocity
 description: "𝒫robabilistic modeling of RNA velocity ⬱"
@@
     toc: false
-
 about:

### 2024-07-28 nbs/guides/interactive/interactive.qmd
fix(nbs): flesh out interactive review
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 number-sections: true
-highlight-style: pygments
-csl: bibstyle.csl
+highlight-style: gruvbox
+csl: ../../bibstyle.csl
 lightbox: auto
@@
 %autoreload 2 
+from IPython.display import Image, display
 ```
+
+## Get results
+
+::: {#wrn-simulated-data .callout-warning collapse=true title="Simulated Data"}
+In this notebook we illustrate how to download and review results using
+unrealistic simulated data. Therefore, nothing in this notebook should be taken
+to represent the results of applying the model to real data.
+:::
+
+### Setup remote connection
 
@@
 [Usage](/templates/user_example/user_example.qmd).
-In this guide we illustrate working with the results 
+However it is increasingly the case that the data set sizes and model training
+resources will not be available or accessible locally, we would like to be able to interactively review results even when they have been computed in a remote cluster with sufficient resources.
+
+In this guide we therefore illustrate working with the results 
 of workflows that have been executed remotely. In this particular case, GitHub 
@@
 
-login to the UI and copy the URI of the workflow or task whose outputs you want to review, and provide this to the `get` method of the remote client,
+### Identify results of interest
+
+Login to the UI and copy the URI of the workflow or task whose inputs and outputs you want to review, and provide these to the `get` method of the remote client,
 
 ```{python}
-#| label: get-remote-outputs
-workflow_outputs = remote.get(
+# | label: get-workflow-io
+workflow_inputs = remote.get(
+    "flyte://v1/pyrovelocity/development/pyrovelocity-argo-nix-bui-5c9ebf8-dev-1hk-d5805de62e8044f79d4/f26c1pjy-0-dn0/i"
+)
+postprocessing_outputs = remote.get(
     "flyte://v1/pyrovelocity/development/pyrovelocity-argo-nix-bui-5c9ebf8-dev-1hk-d5805de62e8044f79d4/f26c1pjy-0-dn0-0-dn6/o"
@@
 ```{python}
-#| label: create-outputs-dict
+# | label: create-outputs-dict
 from omegaconf import OmegaConf
 from flytekit.interaction.string_literals import literal_map_string_repr
+from pyrovelocity.utils import print_config_tree
 
-outputs_dict = literal_map_string_repr(workflow_outputs.literals)
+inputs_dict = literal_map_string_repr(workflow_inputs.literals)
+inputs_dictconfig = OmegaConf.create(inputs_dict)
+print_config_tree(inputs_dict)
+
+outputs_dict = literal_map_string_repr(postprocessing_outputs.literals)
 outputs_dictconfig = OmegaConf.create(outputs_dict)
-print(OmegaConf.to_yaml(outputs_dictconfig))
+print_config_tree(outputs_dict)
 ```
 
-To download the desired results use
+### Download results
+
+Download the results you would like to review,
 
@@
 ```
+
+## Analyze results
+
+### Load data
+
+In this example, we load the postprocessed data from the downloaded file,
+
+```{python}
+# | label: load-postprocessed-data
+import scanpy as sc
+from pyrovelocity.utils import print_anndata
+
+adata = sc.read(postprocessed_data)
+print_anndata(adata)
+```
+
+as well as the posterior samples,
+
+```{python}
+# | label: load-posterior-samples
+from pyrovelocity.utils import pretty_print_dict
+from pyrovelocity.io import CompressedPickle
+
+posterior_samples = CompressedPickle.load(pyrovelocity_data)
+pretty_print_dict(posterior_samples)
+```
+
+### Extract results of interest
+
+In this case, we extract the gene selection data by cooptimizing MAE and correlation between spliced expression levels and estimates of temporal ordering
+
+```{python}
+# | label: extract-gene-selection
+from pyrovelocity.analysis.analyze import pareto_frontier_genes
+
+volcano_data = posterior_samples["gene_ranking"]
+number_of_marker_genes = min(
+    max(int(len(volcano_data) * 0.1), 4), 6, len(volcano_data)
+)
+putative_marker_genes = pareto_frontier_genes(
+    volcano_data, number_of_marker_genes
+)
+```
+
+### Generate plots
+
+Here we re-generate the gene selection summary plot.
+
+```{python}
+# | label: generate-gene-selection-summary-plot
+# | output: false
+from pyrovelocity.plots import plot_gene_selection_summary
+
+vector_field_basis = inputs_dictconfig.preprocess_data_args.vector_field_basis
+cell_state = inputs_dictconfig.preprocess_data_args.cell_state
+
+plot_gene_selection_summary(
+    adata=adata,
+    posterior_samples=posterior_samples,
+    basis=vector_field_basis,
+    cell_state=cell_state,
+    plot_name="gene_selection_summary_plot.pdf",
+    selected_genes=putative_marker_genes,
+    show_marginal_histograms=False,
+)
+```
+
+```{python}
+# | label: show-gene-selection-summary-plot
+# | code-fold: true
+# | output: true
+display(Image(filename=f"gene_selection_summary_plot.pdf.png"))
+```
+
+Of course, you can generate any plot of interest relevant to the loaded data.
+See the [plots package source](https://github.com/pinellolab/pyrovelocity/tree/main/src/pyrovelocity/plots), [plots API reference](../../reference/plots.qmd), and the guide for the [workflow summarization task](https://github.com/pinellolab/pyrovelocity/blob/main/src/pyrovelocity/tasks/summarize.py) for ideas regarding built-in or custom plots.

### 2024-07-28 src/pyrovelocity/utils.py
fix(utils): use shape of ArrayLike in pretty printing dicts
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -5,15 +5,11 @@ import io
 import os
 import sys
 from inspect import getmembers
-from numbers import Integral
-from numbers import Real
+from numbers import Integral, Real
 from pathlib import Path
 from pprint import pprint
-from types import FunctionType
-from types import ModuleType
-from typing import Callable
-from typing import List
-from typing import Tuple
+from types import FunctionType, ModuleType
+from typing import Callable, List, Tuple
 
 import matplotlib.pyplot as plt
 import numpy as np
@@ -25,11 +21,11 @@ import seaborn as sns
 import yaml
 from anndata._core.anndata import AnnData
 from beartype import beartype
+from jaxtyping import ArrayLike
 from scvi.data import synthetic_iid
 
 from pyrovelocity.io.compressedpickle import CompressedPickle
 from pyrovelocity.logging import configure_logging
-
 
 # import torch
 # from scipy.sparse import issparse
@@ -179,12 +175,15 @@ def pretty_log_dict(d: dict) -> str:
     for key, value in d.items():
         # key_colored = colored(key, "green")
         key_colored = key
-        value_lines = str(value).split("\n")
-        value_colored = "\n".join(
-            # colored(line, "white") for line in value_lines
-            line
-            for line in value_lines
-        )
+        if isinstance(value, ArrayLike):
+            value_colored = f"{value.shape}"
+        else:
+            value_lines = str(value).split("\n")
+            value_colored = "\n".join(
+                # colored(line, "white") for line in value_lines
+                line
+                for line in value_lines
+            )
         dict_as_string += f"{key_colored}:\n{value_colored}\n"
     return dict_as_string
 

### 2024-07-28 nbs/concepts/inferenceindynamicalsystems/index.qmd
fix(nbs): escape apostrophes
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 We'll solve the dimensionless system analytically, focusing on what can be 
-understood about the system's dynamics.
+understood about the system\'s dynamics.
 Given specific initial conditions $u^{\ast}(0) = u^{\ast}_0$ and 
@@
 and mRNA concentrations evolve over time towards their steady states. The 
-exponential terms highlight the system's inherent timescales: $u^{\ast}$ relaxes 
+exponential terms highlight the system\'s inherent timescales: $u^{\ast}$ relaxes 
 to its steady state independently, while $s^{\ast}$’s dynamics are coupled to 

### 2024-07-28 nbs/templates/user_example/user_example.qmd
fix(nbs): use mathjax
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
   html:
-    html-math-method: katex
+    html-math-method: mathjax
+    include-in-header: ../../mathjax.html
 #     mainfont: Latin Modern Sans
@@
 These can be safely ignored or removed. 
-They are exclusively used for compatibility with Google Colab's section folding logic.
+They are exclusively used for compatibility with Google Colab\'s section folding logic.
 :::

### 2024-07-28 src/pyrovelocity/tasks/preprocess.py
fix(preprocess): disable progress bar
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -234,7 +234,12 @@ def preprocess_dataset(
             n_pcs=n_pcs,
             n_neighbors=n_neighbors,
         )
-        scv.tl.recover_dynamics(adata, n_jobs=-1, use_raw=False)
+        scv.tl.recover_dynamics(
+            adata,
+            n_jobs=-1,
+            use_raw=False,
+            show_progress_bar=False,
+        )
 
         scv.tl.velocity(adata, mode=default_velocity_mode, use_raw=False)
 
@@ -264,7 +269,7 @@ def preprocess_dataset(
                 )
                 n_vars_subset = len(top_genes)
             adata = adata[:, top_genes[:n_vars_subset]].copy()
-        scv.tl.velocity_graph(adata, n_jobs=-1)
+        scv.tl.velocity_graph(adata, n_jobs=-1, show_progress_bar=False)
 
         scv.tl.velocity_embedding(adata, basis=vector_field_basis)
 

### 2024-07-28 src/pyrovelocity/analysis/analyze.py
fix(analyze): disable progress bar
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -80,7 +80,11 @@ def compute_mean_vector_field(
                     - posterior_samples["st"] * posterior_samples["gamma"]
                 ).mean(0)
         scv.tl.velocity_graph(
-            adata, vkey="velocity_pyro", xkey="spliced_pyro", n_jobs=n_jobs
+            adata,
+            vkey="velocity_pyro",
+            xkey="spliced_pyro",
+            n_jobs=n_jobs,
+            show_progress_bar=False,
         )
     elif spliced in ["Ms"]:
         ut = adata.layers["Mu"]
@@ -100,7 +104,11 @@ def compute_mean_vector_field(
                 - posterior_samples["st"] * posterior_samples["gamma"]
             ).mean(0)
         scv.tl.velocity_graph(
-            adata, vkey="velocity_pyro", xkey="Ms", n_jobs=n_jobs
+            adata,
+            vkey="velocity_pyro",
+            xkey="Ms",
+            n_jobs=n_jobs,
+            show_progress_bar=False,
         )
     elif spliced in ["spliced"]:
         ut = adata.layers["unspliced"]
@@ -120,7 +128,11 @@ def compute_mean_vector_field(
                 - posterior_samples["st"] * posterior_samples["gamma"]
             ).mean(0)
         scv.tl.velocity_graph(
-            adata, vkey="velocity_pyro", xkey="spliced", n_jobs=n_jobs
+            adata,
+            vkey="velocity_pyro",
+            xkey="spliced",
+            n_jobs=n_jobs,
+            show_progress_bar=False,
         )
 
     scv.tl.velocity_embedding(adata, vkey="velocity_pyro", basis=basis)
@@ -279,7 +291,11 @@ def vector_field_uncertainty(
             )
         else:
             scv.tl.velocity_graph(
-                adata, vkey="velocity_pyro", xkey="spliced_pyro", n_jobs=n_jobs
+                adata,
+                vkey="velocity_pyro",
+                xkey="spliced_pyro",
+                n_jobs=n_jobs,
+                show_progress_bar=False,
             )
             scv.tl.velocity_embedding(adata, vkey="velocity_pyro", basis=basis)
         v_map_all.append(adata.obsm[f"velocity_pyro_{basis}"])

### 2024-07-30 nbs/guides/interactive/interactive.qmd
fix(nbs): indicate explicit output
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 # | label: load-postprocessed-data
+# | output: true
 import scanpy as sc
@@
 # | label: load-posterior-samples
+# | output: true
 from pyrovelocity.utils import pretty_print_dict

### 2024-07-30 nbs/concepts/inferenceindynamicalsystems/index.qmd
fix(nbs): disable ipynb
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 #   docx: default
-  ipynb: default
+#  ipynb: default
 format-links: [ipynb]

### 2024-07-30 nbs/tutorials/pancreas/pancreas.qmd
feat(nbs): initialize downsampled pancreas tutorial
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-07-30 nbs/guides/interactive/interactive.qmd
@@
 ---
-title: Interactive results review
+title: Pancreatic endocrinogenesis
 execute: 
@@
 
+Please see the [guide on interactive results review](/guides/interactive/interactive.qmd) for a 
+general review of how to download and review results. Here we use the same 
+approach without any of the explanatory text to retrieve model results to 
+generate plots for model selection and evaluation
+
 ## Get results
 
-::: {#wrn-simulated-data .callout-warning collapse=true title="Simulated Data"}
+::: {#wrn-downsampled-data .callout-warning collapse=true title="Downsampled Data"}
 In this notebook we illustrate how to download and review results using
-unrealistic simulated data. Therefore, nothing in this notebook should be taken
+unrealistic downsampled data. Therefore, this notebook should not be taken
 to represent the results of applying the model to real data.
@@
 ### Setup remote connection
-
-When testing or prototyping it is helpful to review the results of a workflow
-execution interactively. This is relatively straightforward when working with locally
-executed workflows as demonstrated in
-[Usage](/templates/user_example/user_example.qmd).
-However it is increasingly the case that the data set sizes and model training
-resources will not be available or accessible locally, we would like to be able to interactively review results even when they have been computed in a remote cluster with sufficient resources.
-
-In this guide we therefore illustrate working with the results 
-of workflows that have been executed remotely. In this particular case, GitHub 
-OAuth access is required to the endpoint associated to the workflow execution.
-Given this access, setup the remote client,
 
@@
 
-Login to the UI and copy the URI of the workflow or task whose inputs and outputs you want to review, and provide these to the `get` method of the remote client,
-
 ```{python}
@@
 workflow_inputs = remote.get(
-    "flyte://v1/pyrovelocity/development/pyrovelocity-argo-nix-bui-5c9ebf8-dev-1hk-d5805de62e8044f79d4/f26c1pjy-0-dn0/i"
+    "flyte://v1/pyrovelocity/development/pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a/f26c1pjy-0-dn1-0-dn6/i"
 )
 postprocessing_outputs = remote.get(
-    "flyte://v1/pyrovelocity/development/pyrovelocity-argo-nix-bui-5c9ebf8-dev-1hk-d5805de62e8044f79d4/f26c1pjy-0-dn0-0-dn6/o"
+    "flyte://v1/pyrovelocity/development/pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a/f26c1pjy-0-dn1-0-dn6/o"
 )
 ```
-
-This will not download the results but will provide a summary of the outputs.
 
@@
 
-Download the results you would like to review,
-
 ```{python}
@@
 
-In this example, we load the postprocessed data from the downloaded file,
-
 ```{python}
@@
 ```
-
-as well as the posterior samples,
 
@@
 ### Extract results of interest
-
-In this case, we extract the gene selection data by cooptimizing MAE and correlation between spliced expression levels and estimates of temporal ordering
 
@@
 ### Generate plots
-
-Here we re-generate the gene selection summary plot.
 
@@
 ```
-
-Of course, you can generate any plot of interest relevant to the loaded data.
-See the [plots package source](https://github.com/pinellolab/pyrovelocity/tree/main/src/pyrovelocity/plots), [plots API reference](../../reference/plots.qmd), and the guide for the [workflow summarization task](https://github.com/pinellolab/pyrovelocity/blob/main/src/pyrovelocity/tasks/summarize.py) for ideas regarding built-in or custom plots.

### 2024-07-31 src/pyrovelocity/io/gcs.py
fix(io): allow specification of a download filename prefix
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,4 +1,5 @@
 from pathlib import Path
+from typing import Optional
 from urllib.parse import urlparse
 
 from beartype import beartype
@@ -23,6 +24,7 @@ logger = configure_logging(__name__)
 def download_blob_from_uri(
     blob_uri: str,
     concurrent: bool = False,
+    download_filename_prefix: Optional[str] = None,
 ) -> str:
     """Download a single file from a GCS bucket.
 
@@ -55,7 +57,10 @@ def download_blob_from_uri(
             f"URI scheme must be 'gs', not {parsed_blob_uri.scheme}."
         )
     blob_path = Path(parsed_blob_uri.path)
-    blob_filename = blob_path.name
+    if download_filename_prefix:
+        blob_filename = f"{download_filename_prefix}_{blob_path.name}"
+    else:
+        blob_filename = blob_path.name
 
     if not Path(blob_filename).exists():
         blob = Blob.from_string(blob_uri, client)

### 2024-07-31 src/pyrovelocity/io/cluster.py
feat(io): add cluster module
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,0 +1,54 @@
+from typing import Tuple
+
+from beartype import beartype
+from flytekit.configuration import Config
+from flytekit.interaction.string_literals import literal_map_string_repr
+from flytekit.remote.remote import FlyteRemote
+from omegaconf import DictConfig, OmegaConf
+
+from pyrovelocity.utils import print_config_tree
+
+__all__ = ["get_remote_task_results"]
+
+
+@beartype
+def get_remote_task_results(
+    execution_id: str,
+    task_id: str,
+    endpoint: str = "flyte.cluster.pyrovelocity.net",
+    remote_uri_protocol: str = "flyte://v1",
+    remote_project: str = "pyrovelocity",
+    remote_domain: str = "development",
+) -> Tuple[DictConfig, DictConfig]:
+    """
+    Get the inputs and outputs of a remote Flyte task.
+
+    Args:
+        execution_id (str): Flyte execution id.
+        task_id (str): Flyte task id. Defaults to "f26c1pjy-0-dn1-0-dn6".
+        endpoint (str): Cluster endpoint url. Defaults to "flyte.cluster.pyrovelocity.net".
+        remote_uri_protocol (str): Flyte remote uri protocol. Defaults to "flyte://v1".
+        remote_project (str): Flyte remote project. Defaults to "pyrovelocity".
+        remote_domain (str): Flyte remote domain. Defaults to "development".
+
+    Returns:
+        Tuple[DictConfig, DictConfig]: Autocompletable inputs and outputs of
+            the remote task execution.
+    """
+    remote = FlyteRemote(
+        Config.for_endpoint(endpoint=endpoint),
+    )
+    execution_uri = (
+        f"{remote_uri_protocol}/{remote_project}/{remote_domain}/{execution_id}"
+    )
+    task_uri = f"{execution_uri}/{task_id}"
+    inputs = remote.get(f"{task_uri}/i")
+    outputs = remote.get(f"{task_uri}/o")
+    inputs_dict = literal_map_string_repr(inputs.literals)
+    inputs_dictconfig = OmegaConf.create(inputs_dict)
+    print_config_tree(inputs_dict)
+
+    outputs_dict = literal_map_string_repr(outputs.literals)
+    outputs_dictconfig = OmegaConf.create(outputs_dict)
+    print_config_tree(outputs_dict)
+    return inputs_dictconfig, outputs_dictconfig

### 2024-07-31 nbs/tutorials/pancreas/pancreas.qmd
feat(nbs): add support for model comparison to pancreas notebook
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 approach without any of the explanatory text to retrieve model results to 
-generate plots for model selection and evaluation
+generate plots for model selection and evaluation.
 
@@
 
-### Setup remote connection
-
-```{python}
-#| label: instantiate-remote-client
-from flytekit.remote.remote import FlyteRemote
-from flytekit.configuration import Config
-
-remote = FlyteRemote(
-    Config.for_endpoint("flyte.cluster.pyrovelocity.net"),
-)
-```
-
 ### Identify results of interest
@@
 # | label: get-workflow-io
-workflow_inputs = remote.get(
-    "flyte://v1/pyrovelocity/development/pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a/f26c1pjy-0-dn1-0-dn6/i"
+from pyrovelocity.io.cluster import get_remote_task_results
+
+(
+    model1_postprocessing_inputs,
+    model1_postprocessing_outputs,
+) = get_remote_task_results(
+    execution_id="pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a",
+    task_id="f26c1pjy-0-dn1-0-dn3",
 )
-postprocessing_outputs = remote.get(
-    "flyte://v1/pyrovelocity/development/pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a/f26c1pjy-0-dn1-0-dn6/o"
+
+(
+    model2_postprocessing_inputs,
+    model2_postprocessing_outputs,
+) = get_remote_task_results(
+    execution_id="pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a",
+    task_id="f26c1pjy-0-dn1-0-dn6",
 )
-```
-
-```{python}
-# | label: create-outputs-dict
-from omegaconf import OmegaConf
-from flytekit.interaction.string_literals import literal_map_string_repr
-from pyrovelocity.utils import print_config_tree
-
-inputs_dict = literal_map_string_repr(workflow_inputs.literals)
-inputs_dictconfig = OmegaConf.create(inputs_dict)
-print_config_tree(inputs_dict)
-
-outputs_dict = literal_map_string_repr(postprocessing_outputs.literals)
-outputs_dictconfig = OmegaConf.create(outputs_dict)
-print_config_tree(outputs_dict)
 ```
@@
 ```{python}
-#| label: download-outputs
+# | label: download-outputs
 from pyrovelocity.io.gcs import download_blob_from_uri
 
-pyrovelocity_data = download_blob_from_uri(
-    outputs_dictconfig.o0.pyrovelocity_data.path
+model1_pyrovelocity_data = download_blob_from_uri(
+    blob_uri=model1_postprocessing_outputs.o0.pyrovelocity_data.path,
+    download_filename_prefix=f"{model1_postprocessing_inputs.training_outputs.data_model}",
 )
-postprocessed_data = download_blob_from_uri(
-    outputs_dictconfig.o0.postprocessed_data.path
+model1_postprocessed_data = download_blob_from_uri(
+    blob_uri=model1_postprocessing_outputs.o0.postprocessed_data.path,
+    download_filename_prefix=f"{model1_postprocessing_inputs.training_outputs.data_model}",
+)
+
+model2_pyrovelocity_data = download_blob_from_uri(
+    blob_uri=model2_postprocessing_outputs.o0.pyrovelocity_data.path,
+    download_filename_prefix=f"{model2_postprocessing_inputs.training_outputs.data_model}",
+)
+model2_postprocessed_data = download_blob_from_uri(
+    blob_uri=model2_postprocessing_outputs.o0.postprocessed_data.path,
+    download_filename_prefix=f"{model2_postprocessing_inputs.training_outputs.data_model}",
 )
@@
 
-### Load data
+### Model 1
+
+#### Load data
 
 ```{python}
-# | label: load-postprocessed-data
+# | label: model1-load-postprocessed-data
 # | output: true
@@
 
-adata = sc.read(postprocessed_data)
+adata = sc.read(model1_postprocessed_data)
 print_anndata(adata)
@@
 ```{python}
-# | label: load-posterior-samples
+# | label: model1-load-posterior-samples
 # | output: true
@@
 
-posterior_samples = CompressedPickle.load(pyrovelocity_data)
+posterior_samples = CompressedPickle.load(model1_pyrovelocity_data)
 pretty_print_dict(posterior_samples)
@@
 
-### Extract results of interest
+#### Extract results of interest
 
 ```{python}
-# | label: extract-gene-selection
+# | label: model1-extract-gene-selection
 from pyrovelocity.analysis.analyze import pareto_frontier_genes
@@
 
-### Generate plots
+#### Generate plots
 
 ```{python}
-# | label: generate-gene-selection-summary-plot
+# | label: model1-generate-gene-selection-summary-plot
 # | output: false
@@
 
-vector_field_basis = inputs_dictconfig.preprocess_data_args.vector_field_basis
-cell_state = inputs_dictconfig.preprocess_data_args.cell_state
+vector_field_basis = model1_postprocessing_inputs.preprocess_data_args.vector_field_basis
+cell_state = model1_postprocessing_inputs.preprocess_data_args.cell_state
 
@@
 ```{python}
-# | label: show-gene-selection-summary-plot
+# | label: model1-show-gene-selection-summary-plot
 # | code-fold: true
@@
 ```
+
+### Model 2
+
+#### Load data
+
+```{python}
+# | label: model2-load-postprocessed-data
+# | output: true
+import scanpy as sc
+from pyrovelocity.utils import print_anndata
+
+adata = sc.read(model2_postprocessed_data)
+print_anndata(adata)
+```
+
+```{python}
+# | label: model2-load-posterior-samples
+# | output: true
+from pyrovelocity.utils import pretty_print_dict
+from pyrovelocity.io import CompressedPickle
+
+posterior_samples = CompressedPickle.load(model2_pyrovelocity_data)
+pretty_print_dict(posterior_samples)
+```
+
+#### Extract results of interest
+
+```{python}
+# | label: model2-extract-gene-selection
+from pyrovelocity.analysis.analyze import pareto_frontier_genes
+
+volcano_data = posterior_samples["gene_ranking"]
+number_of_marker_genes = min(
+    max(int(len(volcano_data) * 0.1), 4), 6, len(volcano_data)
+)
+putative_marker_genes = pareto_frontier_genes(
+    volcano_data, number_of_marker_genes
+)
+```
+
+#### Generate plots
+
+```{python}
+# | label: model2-generate-gene-selection-summary-plot
+# | output: false
+from pyrovelocity.plots import plot_gene_selection_summary
+
+vector_field_basis = model2_postprocessing_inputs.preprocess_data_args.vector_field_basis
+cell_state = model2_postprocessing_inputs.preprocess_data_args.cell_state
+
+plot_gene_selection_summary(
+    adata=adata,
+    posterior_samples=posterior_samples,
+    basis=vector_field_basis,
+    cell_state=cell_state,
+    plot_name="gene_selection_summary_plot.pdf",
+    selected_genes=putative_marker_genes,
+    show_marginal_histograms=False,
+)
+```
+
+```{python}
+# | label: model2-show-gene-selection-summary-plot
+# | code-fold: true
+# | output: true
+display(Image(filename=f"gene_selection_summary_plot.pdf.png"))
+```

### 2024-08-01 nbs/tutorials/pancreas/results/pancreas.qmd
refactor(nbs): migrate results notebook to subdir
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-07-31 nbs/tutorials/pancreas/pancreas.qmd
@@
 highlight-style: gruvbox
-csl: ../../bibstyle.csl
+csl: ../../../bibstyle.csl
 lightbox: auto
@@
 
+execution_id = "pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a"
+
 (
@@
 ) = get_remote_task_results(
-    execution_id="pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a",
+    execution_id=execution_id,
     task_id="f26c1pjy-0-dn1-0-dn3",
@@
 ) = get_remote_task_results(
-    execution_id="pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a",
+    execution_id=execution_id,
     task_id="f26c1pjy-0-dn1-0-dn6",

### 2024-08-05 nbs/guides/interactive/interactive.qmd
docs(nbs): write data tables to files
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 See the [plots package source](https://github.com/pinellolab/pyrovelocity/tree/main/src/pyrovelocity/plots), [plots API reference](../../reference/plots.qmd), and the guide for the [workflow summarization task](https://github.com/pinellolab/pyrovelocity/blob/main/src/pyrovelocity/tasks/summarize.py) for ideas regarding built-in or custom plots.
+
+### Save data files
+
+We may also want to write data tables to review. Here we save the parameter
+posterior distribution table
+
+```{python}
+# | label: save-parameter-posterior-mean-dataframe
+from pyrovelocity.utils import save_parameter_posterior_mean_dataframe
+
+save_parameter_posterior_mean_dataframe(
+    adata=adata,
+    input_dict=posterior_samples,
+    dataframe_path="gene_parameter_posterior_mean.csv",
+)
+```
+
+and the gene selection table
+
+```{python}
+# | label: save-gene-selection-table
+posterior_samples["gene_ranking"].to_csv("gene_ranking.csv")
+posterior_samples["gene_ranking"]
+```

### 2024-08-05 nbs/guides/interactive/interactive.qmd
docs(nbs): generate phase portraits from custom gene list
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+#### Gene selection summary plot
+
 Here we re-generate the gene selection summary plot.
@@
 
+#### Phase portraits from custom gene list
+
+One of the primary applications of interactive review may be to explore
+specific genes of interest. Here we generate phase portraits for a randomly
+selected list of genes.
+
+```{python}
+# | label: generate-phase-portraits-from-gene-list
+# | output: false
+from pyrovelocity.plots import rainbowplot
+
+vector_field_basis = inputs_dictconfig.preprocess_data_args.vector_field_basis
+cell_state = inputs_dictconfig.preprocess_data_args.cell_state
+
+rainbowplot(
+    volcano_data=volcano_data,
+    adata=adata,
+    posterior_samples=posterior_samples,
+    genes=["1839", "927", "15"],
+    data=["st", "ut"],
+    basis=vector_field_basis,
+    cell_state=cell_state,
+    save_plot=True,
+    rainbow_plot_path="gene_selection_rainbow_plot.pdf",
+)
+```
+
+```{python}
+# | label: show-phase-portraits-from-gene-list
+# | code-fold: true
+# | output: true
+display(Image(filename=f"gene_selection_rainbow_plot.pdf.png"))
+```
+
 Of course, you can generate any plot of interest relevant to the loaded data.

### 2024-08-05 src/pyrovelocity/workflows/main_workflow.py
fix(workflows): disable upload summary cache
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -255,7 +255,7 @@ def summarize_data(
 
 
 @task(
-    cache=PYROVELOCITY_CACHE_FLAG,
+    cache=False,
     cache_version=SUMMARIZE_CACHE_VERSION,
     retries=3,
     interruptible=True,

### 2024-08-07 nbs/tutorials/pancreas/results/pancreas.qmd
docs(tutorials): add more detailed gene selection data
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-execution_id = "pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a"
+# execution_id = "pyrovelocity-py311-defaul-fe20d09-dev-pzh-17b5092261c84e8695a"  # downsampled
+execution_id = (
+    "pyrovelocity-595-gsmc-feaa936-dev-vf0-264734a284ac493fa39"  # complete
+)
 
@@
     execution_id=execution_id,
-    task_id="f26c1pjy-0-dn1-0-dn6",
+    task_id="f26c1pjy-0-dn1-0-dn7",
 )
@@
 
+We decompose the correlation of each gene's expression level with the
+cell time estimates into eight equally-spaced bins and select the top 3% of genes with respect to the mean absolute error between the predictive sample mean and
+the observed expression level.
+
 ```{python}
 # | label: model1-extract-gene-selection
-from pyrovelocity.analysis.analyze import pareto_frontier_genes
+from pyrovelocity.analysis.analyze import top_mae_genes
 
@@
 )
-putative_marker_genes = pareto_frontier_genes(
-    volcano_data, number_of_marker_genes
+putative_marker_genes = top_mae_genes(
+    volcano_data,
+    mae_top_percentile=3,
+    min_genes_per_bin=3,
 )
@@
 
+We generate several summary plots below for two distinct model classes 
+in order to support comparative model evaluation, which is among few
+options for selecting effective models that are not designed to be
+microscopically valid.
+
+Here we plot the gene selection statistics highlighting genes of interest
+based upon them.
+
 ```{python}
-# | label: model1-generate-gene-selection-summary-plot
+# | label: model1-generate-gene-selection-plot
 # | output: false
-from pyrovelocity.plots import plot_gene_selection_summary
+from pyrovelocity.plots import plot_gene_ranking
 
-vector_field_basis = model1_postprocessing_inputs.preprocess_data_args.vector_field_basis
+vector_field_basis = (
+    model1_postprocessing_inputs.preprocess_data_args.vector_field_basis
+)
 cell_state = model1_postprocessing_inputs.preprocess_data_args.cell_state
 
-plot_gene_selection_summary(
+plot_gene_ranking(
+    posterior_samples=[posterior_samples],
+    adata=[adata],
+    selected_genes=putative_marker_genes,
+    time_correlation_with="st",
+    show_marginal_histograms=True,
+    save_volcano_plot=True,
+    volcano_plot_path="model1_volcano_plot.pdf",
+)
+```
+```{python}
+# | label: model1-show-gene-selection-plot
+# | code-fold: true
+# | output: true
+display(Image(filename=f"model1_volcano_plot.pdf.png"))
+```
+
+We can attempt to attribute the statistics to underlying features plotting 
+phase portraits, predictive expression, and log-transformed expression of
+the proposed marker genes
+
+```{python}
+# | label: model1-generate-rainbow-plot
+# | output: false
+from pyrovelocity.plots import rainbowplot
+
+rainbowplot(
+    volcano_data=volcano_data,
     adata=adata,
     posterior_samples=posterior_samples,
+    genes=putative_marker_genes,
+    data=["st", "ut"],
     basis=vector_field_basis,
     cell_state=cell_state,
-    plot_name="gene_selection_summary_plot.pdf",
-    selected_genes=putative_marker_genes,
-    show_marginal_histograms=False,
+    save_plot=True,
+    rainbow_plot_path="model1_rainbow_plot.pdf",
 )
@@
 ```{python}
-# | label: model1-show-gene-selection-summary-plot
+# | label: model1-show-rainbow-plot
 # | code-fold: true
 # | output: true
-display(Image(filename=f"gene_selection_summary_plot.pdf.png"))
+display(Image(filename=f"model1_rainbow_plot.pdf.png"))
 ```
@@
 ### Model 2
+
+We provide a comparable analysis for the second model below.
 
@@
 # | label: model2-extract-gene-selection
-from pyrovelocity.analysis.analyze import pareto_frontier_genes
+from pyrovelocity.analysis.analyze import top_mae_genes
 
@@
 ```{python}
-# | label: model2-generate-gene-selection-summary-plot
+# | label: model2-generate-gene-selection-plot
 # | output: false
-from pyrovelocity.plots import plot_gene_selection_summary
+from pyrovelocity.plots import plot_gene_ranking
 
-vector_field_basis = model2_postprocessing_inputs.preprocess_data_args.vector_field_basis
+vector_field_basis = (
+    model2_postprocessing_inputs.preprocess_data_args.vector_field_basis
+)
 cell_state = model2_postprocessing_inputs.preprocess_data_args.cell_state
 
-plot_gene_selection_summary(
+plot_gene_ranking(
+    posterior_samples=[posterior_samples],
+    adata=[adata],
+    selected_genes=putative_marker_genes,
+    time_correlation_with="st",
+    show_marginal_histograms=True,
+    save_volcano_plot=True,
+    volcano_plot_path="model2_volcano_plot.pdf",
+)
+```
+```{python}
+# | label: model2-show-gene-selection-plot
+# | code-fold: true
+# | output: true
+display(Image(filename=f"model2_volcano_plot.pdf.png"))
+```
+
+Plot phase portraits, predictive expression, and log-transformed expression of
+all proposed marker genes
+
+```{python}
+# | label: model2-generate-rainbow-plot
+# | output: false
+from pyrovelocity.plots import rainbowplot
+
+rainbowplot(
+    volcano_data=volcano_data,
     adata=adata,
     posterior_samples=posterior_samples,
+    genes=putative_marker_genes,
+    data=["st", "ut"],
     basis=vector_field_basis,
     cell_state=cell_state,
-    plot_name="gene_selection_summary_plot.pdf",
-    selected_genes=putative_marker_genes,
-    show_marginal_histograms=False,
+    save_plot=True,
+    rainbow_plot_path="model2_rainbow_plot.pdf",
 )
@@
 ```{python}
-# | label: model2-show-gene-selection-summary-plot
+# | label: model2-show-rainbow-plot
 # | code-fold: true
 # | output: true
-display(Image(filename=f"gene_selection_summary_plot.pdf.png"))
+display(Image(filename=f"model2_rainbow_plot.pdf.png"))
 ```

### 2024-08-07 src/pyrovelocity/plots/_rainbow.py
feat(plots): add option to plot log spliced data to rainbow plot
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -67,7 +67,7 @@ def rainbowplot(
     else:
         colors = setup_colors(adata, cell_state)
 
-    st, ut = get_posterior_samples(data, posterior_samples)
+    st, ut = get_posterior_samples_mean(data, posterior_samples)
 
     for n, gene in enumerate(genes):
         ress = get_data(gene, st, ut, adata, cell_state, posterior_samples)
@@ -78,11 +78,11 @@ def rainbowplot(
             ax4 = ax[n, 3]
 
         if n == 0:
-            ax1.set_title("Rainbow plot", fontsize=7)
+            ax1.set_title("Predictive dynamics", fontsize=7)
             ax2.set_title("Phase portrait", fontsize=7)
-            ax3.set_title("Denoised spliced", fontsize=7)
+            ax3.set_title("Predictive spliced", fontsize=7)
             if show_data:
-                ax4.set_title("Spliced data", fontsize=7)
+                ax4.set_title("Log spliced data", fontsize=7)
 
         plot_gene(ax1, ress, colors, add_line)
         scatterplot(ax2, ress, colors)
@@ -92,7 +92,7 @@ def rainbowplot(
             adata.obsm[f"X_{basis}"][:, 1],
             s=3,
             c=st[:, index].flatten(),
-            cmap="RdBu_r",
+            cmap="cividis",
         )
         set_colorbar(im, ax3, labelsize=5, fig=fig, rainbow=True)
         ax3.axis("off")
@@ -101,10 +101,8 @@ def rainbowplot(
                 adata.obsm[f"X_{basis}"][:, 0],
                 adata.obsm[f"X_{basis}"][:, 1],
                 s=3,
-                c=ensure_numpy_array(
-                    adata.layers["spliced"][:, index]
-                ).flatten(),
-                cmap="RdBu_r",
+                c=ensure_numpy_array(adata.X[:, index]).flatten(),
+                cmap="cividis",
             )
             set_colorbar(im, ax4, labelsize=5, fig=fig, rainbow=True)
             ax4.axis("off")
@@ -147,7 +145,7 @@ def scatter_gene(ax, adata, st, gene, basis):
         adata.obsm[f"X_{basis}"][:, 1],
         s=3,
         c=st[:, index].flatten(),
-        cmap="RdBu_r",
+        cmap="cividis",
     )
 
 
@@ -219,7 +217,7 @@ def setup_colors(adata, cell_state):
     )
 
 
[…]

### 2024-08-07 src/pyrovelocity/plots/_summary.py
fix(plots): default to highest and lowest shared time correlation genes in summary plot
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -27,7 +27,6 @@ def plot_gene_selection_summary(
     cell_state: str = "",
     show_marginal_histograms: bool = False,
     selected_genes: Optional[List[str]] = None,
-    number_of_genes: int = 4,
 ) -> FigureBase:
     fig = plt.figure(figsize=(9.5, 5))
     subfigs = fig.subfigures(1, 2, wspace=0.0, hspace=0, width_ratios=[1.8, 4])
@@ -52,14 +51,14 @@ def plot_gene_selection_summary(
     )
 
     _ = rainbowplot(
-        volcano_data,
-        adata,
-        posterior_samples,
-        subfigs[1],
+        volcano_data=volcano_data,
+        adata=adata,
+        posterior_samples=posterior_samples,
+        fig=subfigs[1],
+        genes=selected_genes[:2] + selected_genes[-2:],
         data=["st", "ut"],
         basis=basis,
         cell_state=cell_state,
-        num_genes=number_of_genes,
         show_data=False,
     )
     for ext in ["", ".png"]:

### 2024-08-07 src/pyrovelocity/plots/_time.py
fix(plots): use cividis in plot posterior time
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -11,9 +11,7 @@ from beartype import beartype
 from matplotlib.figure import FigureBase
 from scipy.stats import spearmanr
 
-from pyrovelocity.plots._common import set_colorbar
-from pyrovelocity.plots._common import set_font_size
-
+from pyrovelocity.plots._common import set_colorbar, set_font_size
 
 __all__ = [
     "plot_posterior_time",
@@ -53,7 +51,7 @@ def plot_posterior_time(
         s=s,
         alpha=0.4,
         c=adata.obs["cell_time"],
-        cmap="inferno",
+        cmap="cividis",
         linewidth=0,
     )
     set_colorbar(im, ax, labelsize=5, fig=fig, position=position)

### 2024-08-07 src/pyrovelocity/workflows/main_workflow.py
chore(workflows): bump summarize cache to `2024.8.5.3`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -55,7 +55,7 @@ __all__ = [
 logger = configure_logging(__name__)
 
 CACHE_VERSION = "2024.8.5"
-SUMMARIZE_CACHE_VERSION = "2024.8.5.2"
+SUMMARIZE_CACHE_VERSION = "2024.8.5.3"
 PYROVELOCITY_CACHE_FLAG = str_to_bool(
     os.getenv("PYROVELOCITY_CACHE_FLAG", "True")
 )

### 2024-08-07 nbs/tutorials/pancreas/results/pancreas.qmd
fix(nbs): add parameter posterior estimates to pancreas results
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+It is somewhat difficult to compare the parameter posteriors on a common scale
+but we plot their posterior distributions below.
+
+```{python}
+# | label: model1-generate-parameter-posterior-plot
+# | output: false
+from pyrovelocity.plots import plot_parameter_posterior_distributions
+
+posterior_samples.pop("t0", None)
+
+plot_parameter_posterior_distributions(
+    posterior_samples=posterior_samples,
+    adata=adata,
+    geneset=putative_marker_genes,
+    parameter_uncertainty_plot="model1_parameter_uncertainty_plot.pdf",
+)
+```
+
+```{python}
+# | label: model1-show-parameter-posterior-plot
+# | code-fold: true
+# | output: true
+display(Image(filename=f"model1_parameter_uncertainty_plot.pdf.png"))
+```
+
 ### Model 2
@@
 ```
+
+Finally, we plot the parameter posteriors for the second model.
+
+```{python}
+# | label: model2-generate-parameter-posterior-plot
+# | output: false
+from pyrovelocity.plots import plot_parameter_posterior_distributions
+
+plot_parameter_posterior_distributions(
+    posterior_samples=posterior_samples,
+    adata=adata,
+    geneset=putative_marker_genes,
+    parameter_uncertainty_plot="model2_parameter_uncertainty_plot.pdf",
+)
+```
+
+```{python}
+# | label: model2-show-parameter-posterior-plot
+# | code-fold: true
+# | output: true
+display(Image(filename=f"model2_parameter_uncertainty_plot.pdf.png"))
+```

### 2024-08-07 src/pyrovelocity/plots/_parameters.py
fix(plots): enable offset parameters in posterior distribution plots
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,6 +1,5 @@
 from os import PathLike
-from typing import Dict
-from typing import List
+from typing import Dict, List
 
 import matplotlib.pyplot as plt
 import numpy as np
@@ -11,7 +10,6 @@ from beartype import beartype
 from matplotlib.figure import FigureBase
 
 from pyrovelocity.logging import configure_logging
-
 
 __all__ = ["plot_parameter_posterior_distributions"]
 
@@ -25,18 +23,31 @@ def plot_parameter_posterior_distributions(
     geneset: List[str],
     parameter_uncertainty_plot: PathLike | str,
 ) -> FigureBase:
-    fig, ax = plt.subplots(3, 1)
-    fig.set_size_inches(18, 12)
-    for index, kinetics in enumerate(["alpha", "beta", "gamma"]):
+    parameter_names = [
+        "alpha",
+        "beta",
+        "gamma",
+        "u_offset",
+        "s_offset",
+        "t0",
+    ]
+    parameters = [
+        parameter
+        for parameter in parameter_names
+        if parameter in posterior_samples.keys()
+    ]
+    fig, ax = plt.subplots(len(parameters), 1)
+    fig.set_size_inches(18, len(parameters) * 4)
+    for index, parameter in enumerate(parameters):
         df = pd.DataFrame(
             np.log(
-                posterior_samples[kinetics].squeeze()[
+                posterior_samples[parameter].squeeze()[
                     :, np.isin(adata.var_names, list(geneset))
                 ],
             ),
             columns=adata.var_names[np.isin(adata.var_names, list(geneset))],
         )
-        df = df.apply(lambda x: x - x.mean())
+        # df = df.apply(lambda x: x - x.mean())
         df_long = df.melt(var_name="index", value_name="value")
         logger.debug(df_long.head())
         df_long["index"] = pd.Categorical(
@@ -51,7 +62,7 @@ def plot_parameter_posterior_distributions(
                 df_long["index"].astype(str).apply(lambda x: f"g_{int(x)}")
             )
             logger.warning(
-                f"Converted integer index to string 'g_int' for {kinetics}"
+                f"Converted integer index to string 'g_int' for {parameter}"
             )
         else:
             pass
@@ -63,8 +74,9 @@ def plot_parameter_posterior_distributions(
             ax=ax[index],
         )
[…]

### 2024-08-07 src/pyrovelocity/tasks/summarize.py
fix(tasks): remove unused t0 from model1 posterior samples
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -214,6 +214,9 @@ def summarize_dataset(
     volcano_data: DataFrame = posterior_samples["gene_ranking"]
     volcano_data.to_csv(gene_ranking_path)
 
+    if "model1" in data_model:
+        posterior_samples.pop("t0", None)
+
     ##################
     # generate figures
     ##################

### 2024-08-07 src/pyrovelocity/workflows/main_workflow.py
chore(workflows): bump summarize cache to `2024.8.5.4`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -55,7 +55,7 @@ __all__ = [
 logger = configure_logging(__name__)
 
 CACHE_VERSION = "2024.8.5"
-SUMMARIZE_CACHE_VERSION = "2024.8.5.3"
+SUMMARIZE_CACHE_VERSION = "2024.8.5.4"
 PYROVELOCITY_CACHE_FLAG = str_to_bool(
     os.getenv("PYROVELOCITY_CACHE_FLAG", "True")
 )

### 2024-08-09 nbs/guides/installation.qmd
docs(nbs): note python 3.11 in installation guide
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ```bash
-echo "3.10" > .python-version
+echo "3.11" > .python-version
 uv python install
-uv venv -p 3.10
+uv venv -p 3.11
 source .venv/bin/activate
@@
 alias mm=micromamba
-mm create -n pyrovelocity python=3.10
+mm create -n pyrovelocity python=3.11
 mm activate pyrovelocity

### 2024-08-09 nbs/guides/preprocessing/preprocessing.qmd
refactor(nbs): migrate preprocessing to subdirectory
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Preprocessing
execute: 
  freeze: true
  eval: true
  warning: false
  error: false
  cache: true
toc: true
number-sections: true
highlight-style: gruvbox
csl: ../../bibstyle.csl
lightbox: auto
format:
  html:
    html-math-method: katex
  ipynb: default
---

```{python}
#| label: enable-autoreload
#| code-fold: true
#| output: false
%load_ext autoreload
%autoreload 2 
from IPython.display import Image, display
```

### 2024-08-09 nbs/guides/preprocessing/preprocessing.qmd
chore(nbs): add gitignore
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ```
+
+# Download data
+
+```{python}
+# | label: download-data
+import polars as pl
+import altair as alt
+
+alt.data_transformers.enable("vegafusion")
+
+flights = pl.read_parquet(
+    "https://vegafusion-datasets.s3.amazonaws.com/vega/flights_1m.parquet"
+)
+```
+
+# Make plots
+
+```{python}
+# | label: make-plots
+delay_hist = (
+    alt.Chart(flights)
+    .mark_bar()
+    .encode(alt.X("delay", bin=alt.Bin(maxbins=30)), alt.Y("count()"))
+)
+delay_hist
+```

### 2024-08-09 nbs/templates/user_example/user_example.qmd

@@
 If you are running on a local or remote Jupyter server outside of Google Colab, we assume you have already installed pyrovelocity in the same environment in which the Jupyter kernel is running.
-If you have not, ensure you are in a virtual environment with python 3.10 active (we recommend [pyenv][pyenv] or [miniforge][miniforge]), and then you can install the development version from github with
+If you have not, ensure you are in a virtual environment with python 3.11 active (we recommend [pyenv][pyenv] or [miniforge][miniforge]), and then you can install the development version from github with
 

### 2024-08-12 src/pyrovelocity/tests/plots/conftest.py
test(plots): load preprocessed fixture data for plotting
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,0 +1,10 @@
+import pytest
+
+from pyrovelocity.io.serialization import load_anndata_from_json
+
+
+@pytest.fixture
+def adata_preprocessed():
+    return load_anndata_from_json(
+        "src/pyrovelocity/tests/data/preprocessed_3_4.json"
+    )

### 2024-08-12 src/pyrovelocity/plots/_count_histograms.py
feat(plots): add interactive count histogram plot
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,0 +1,91 @@
+import altair as alt
+import numpy as np
+import polars as pl
+from altair.vegalite.v5.api import Chart as AltairChart
+from anndata import AnnData
+from beartype import beartype
+from scipy.sparse import csr_matrix
+
+__all__ = ["plot_spliced_unspliced_histogram"]
+
+
+alt.data_transformers.enable("vegafusion")
+
+
+@beartype
+def plot_spliced_unspliced_histogram(
+    adata: AnnData,
+    spliced_layer: str = "spliced",
+    unspliced_layer: str = "unspliced",
+    min_count: int = 3,
+    max_count: int = 200,
+    title: str = "",
+    default_font_size: int = 24,
+    plot_width: int = 700,
+    plot_height: int = 700,
+) -> AltairChart:
+    spliced_matrix = csr_matrix(adata.layers[spliced_layer])
+    unspliced_matrix = csr_matrix(adata.layers[unspliced_layer])
+
+    non_zero = spliced_matrix.nonzero()
+
+    spliced_values = spliced_matrix[non_zero].A1
+    unspliced_values = unspliced_matrix[non_zero].A1
+
+    df = pl.DataFrame(
+        {"spliced": spliced_values, "unspliced": unspliced_values}
+    )
+
+    df = df.filter(
+        (pl.col("spliced") >= min_count)
+        & (pl.col("unspliced") >= min_count)
+        & (pl.col("spliced") <= max_count)
+        & (pl.col("unspliced") <= max_count)
+    )
+
+    number_of_histogram_bins = int(np.maximum(60, np.sqrt(len(df))))
+
+    chart = (
+        alt.Chart(df)
+        .mark_rect()
+        .encode(
+            x=alt.X(
+                "spliced:Q",
+                title="Spliced Counts",
+                bin=alt.Bin(maxbins=number_of_histogram_bins),
+                axis=alt.Axis(tickMinStep=1),
+            ),
+            y=alt.Y(
+                "unspliced:Q",
+                title="Unspliced Counts",
+                bin=alt.Bin(maxbins=number_of_histogram_bins),
+                axis=alt.Axis(tickMinStep=1),
+            ),
+            color=alt.Color("count():Q", scale=alt.Scale(scheme="greenblue")),
+            tooltip=["spliced", "unspliced", "count()"],
+        )
+        .properties(
+            title=title,
+            width=plot_width,
+            height=plot_height,
+        )
+        .configure(
+            countTitle="counts",
+        )
+        .configure_axis(
[…]

### 2024-08-12 src/pyrovelocity/tests/plots/test_count_histograms.py
test(plots): add smoke test for count histogram plot
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,0 +1,9 @@
+from pyrovelocity.plots import plot_spliced_unspliced_histogram
+
+
+def test_plot_spliced_unspliced_histogram(adata_preprocessed, tmp_path):
+    chart = plot_spliced_unspliced_histogram(adata_preprocessed)
+    chart.save(tmp_path / "spliced_unspliced_histogram.pdf")
+    chart.save(tmp_path / "spliced_unspliced_histogram.svg")
+    chart.save(tmp_path / "spliced_unspliced_histogram.png")
+    chart.save(tmp_path / "spliced_unspliced_histogram.html")

### 2024-08-12 src/pyrovelocity/plots/__init__.py
chore(plots): import count histogram to plots subpackage
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,7 +1,8 @@
-from pyrovelocity.plots._dynamics import (
-    plot_deterministic_simulation_phase_portrait,
+from pyrovelocity.plots._count_histograms import (
+    plot_spliced_unspliced_histogram,
 )
 from pyrovelocity.plots._dynamics import (
+    plot_deterministic_simulation_phase_portrait,
     plot_deterministic_simulation_trajectories,
 )
 from pyrovelocity.plots._experimental import plot_t0_selection
@@ -11,19 +12,20 @@ from pyrovelocity.plots._parameters import (
 )
 from pyrovelocity.plots._predictive import (
     extrapolate_prediction_sample_predictive,
+    posterior_curve,
 )
-from pyrovelocity.plots._predictive import posterior_curve
 from pyrovelocity.plots._rainbow import rainbowplot
 from pyrovelocity.plots._summary import plot_gene_selection_summary
-from pyrovelocity.plots._time import plot_posterior_time
-from pyrovelocity.plots._time import plot_shared_time_uncertainty
-from pyrovelocity.plots._uncertainty import cluster_violin_plots
-from pyrovelocity.plots._uncertainty import (
-    get_posterior_sample_angle_uncertainty,
+from pyrovelocity.plots._time import (
+    plot_posterior_time,
+    plot_shared_time_uncertainty,
 )
-from pyrovelocity.plots._uncertainty import plot_state_uncertainty
+from pyrovelocity.plots._uncertainty import (
+    cluster_violin_plots,
+    get_posterior_sample_angle_uncertainty,
+    plot_state_uncertainty,
+)
 from pyrovelocity.plots._vector_fields import plot_vector_field_summary
-
 
 __all__ = [
     "cluster_violin_plots",
@@ -35,6 +37,7 @@ __all__ = [
     "plot_parameter_posterior_distributions",
     "plot_posterior_time",
     "plot_shared_time_uncertainty",
+    "plot_spliced_unspliced_histogram",
     "plot_state_uncertainty",
     "plot_t0_selection",
     "posterior_curve",

### 2024-08-12 nbs/guides/preprocessing/preprocessing.qmd
docs(preprocessing): initialize preprocessing guide
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-08-09 nbs/guides/preprocessing/preprocessing.qmd
@@
 execute: 
-  freeze: true
+  freeze: false
   eval: true
@@
 ```
+
+## Download data
+
+::: {#wrn-downsampled-data .callout-warning collapse=true title="Downsampled Data"}
+In this notebook we illustrate how to download and review results using
+unrealistic downsampled data. Therefore, this notebook should not be taken
+to represent the results of applying the model to real data.
+:::
+
+We download an example data set to be used for illustration of the 
+preprocessing steps.
+
+::: {#wrn-mutable-data .callout-warning collapse=true title="Mutable objects"}
+Ideally we would work with immutable data structures to avoid losing
+information about how a given data object has been transformed. However,
+the usual object-oriented approach relies on repeated mutation of the
+same object. The only defense against confusion we have here is to review 
+the updated properties of the underlying object after applying effectful
+methods or functions that have the ability to mutate an object's state in
+any way. This is why we use the `pyrovelocity.utils.anndata_string` function
+after each step that may mutate the data object. Of course, without hashing
+each subcomponent we will still not be able to verify that pre-existing
+components of the object were not modified but at least we can verify
+the relationship between added and removed components this way.
+:::
+
+```{python}
+# | label: download-data
+from pyrovelocity.tasks.data import download_dataset
+from pyrovelocity.tasks.data import load_anndata_from_path
+from pyrovelocity.utils import anndata_string, print_string_diff
+
+dataset_path = download_dataset(data_set_name="pancreas")
+adata = load_anndata_from_path(dataset_path)
+initial_data_state_representation = anndata_string(adata)
+print(initial_data_state_representation)
+```
+
+## Preprocessing steps
+
+### Subset data
+
+We will eventually scale up the analysis to the full data set, but it is
+helpful to run experiments on downsampled data to speed up iteration
+prior to running on the full data set. Here we downsample the data to
+300 observations and 200 variables. Some steps may benefit from even
+further downsampling but this is often a good starting point.
+
+```{python}
+# | label: subset-data
+import os
+import numpy as np
+from pyrovelocity.utils import str_to_bool
+
+PYROVELOCITY_DATA_SUBSET = str_to_bool(
+    os.getenv("PYROVELOCITY_DATA_SUBSET", "True")
+)
+
+if PYROVELOCITY_DATA_SUBSET:
+    adata = adata[
+        np.random.choice(
+            adata.obs.index,
+            size=300,
+            replace=False,
+        ),
+        :,
+    ].copy()
+subset_anndata_representation = anndata_string(adata)
+print_string_diff(
+    text1=initial_data_state_representation,
+    text2=subset_anndata_representation,
+    diff_title="Downsampled data diff",
+)
+```
+
+You can save the downsampled data for later use, but for small data sets
+less than $O(10^4)$ observations it is not too expensive to reload the
+full data set if the identity of the random sample does not need to be reproducible.
+
+We will make use of the raw count data in constructing a probabilistic model
+of the spliced and unspliced read or molecule counts. The most commonly used
+preprocessing methods overwrite this information with normalized or transformed
+versions so we always begin by saving a copy of the count data.
+
+### Save count data
+
+```{python}
+# | label: save-count-data
+from pyrovelocity.tasks.preprocess import copy_raw_counts
+
+copy_raw_counts(adata)
+copied_raw_counts_representation = anndata_string(adata)
+print_string_diff(
+    text1=subset_anndata_representation,
+    text2=copied_raw_counts_representation,
+    diff_title="Copy raw counts diff",
+)
+```
+
+### Pre-filtration QC
+
+```{python}
+# | label: pre-filtration-quality-control-metrics
+import scanpy as sc
+
+adata.var["mt"] = adata.var_names.str.startswith("MT-" or "Mt-" or "mt-")
+adata.var["ribo"] = adata.var_names.str.startswith(
+    "RPS" or "Rps" or "rps" or "RPL" or "Rpl" or "rpl"
+)
+
+sc.pp.calculate_qc_metrics(
+    adata=adata,
+    qc_vars=["mt", "ribo"],
+    percent_top=None,
+    log1p=False,
+    inplace=True,
+)
+qc_metrics_representation = anndata_string(adata)
+print_string_diff(
+    text1=copied_raw_counts_representation,
+    text2=qc_metrics_representation,
+    diff_title="Quality control metrics diff",
+)
+```
+
+```{python}
+# | label: pre-filtration-plot-genes-by-counts
+import seaborn as sns
+import pandas as pd
+
+numeric_obs = adata.obs.copy()
+numeric_obs["n_genes_by_counts"] = pd.to_numeric(
+    numeric_obs["n_genes_by_counts"],
+    errors="coerce",
+)
+qc_counts = sns.jointplot(
+    data=numeric_obs,
+    x="total_counts",
+    y="n_genes_by_counts",
+    color="#ff6a14",
+    marginal_ticks=True,
+    kind="scatter",
+    alpha=0.4,
+)
+qc_counts.plot_joint(
+    sns.kdeplot,
+    color="gray",
+    alpha=0.6,
+)
+qc_counts.set_axis_labels(
+    xlabel="Total counts in cell",
+    ylabel="Number of genes >=1 count in cell",
+)
+```
+
+### Compute cytotrace statistics
+
+For some data sets we apply cytotrace as a proxy for progression through 
+lineage-specific differentiation.
+
+```{python}
+# | label: cytotrace
+from pyrovelocity.analysis.cytotrace import cytotrace_sparse
+
+cytotrace_sparse(adata, layer="spliced")
+cytotrace_representation = anndata_string(adata)
+print_string_diff(
+    text1=copied_raw_counts_representation,
+    text2=cytotrace_representation,
+    diff_title="Cytotrace diff",
+)
+```
+
+### Filter and normalize data
+
+Next we use the preprocessing functions built-in to the scvelo package to 
+filter genes that have at least a given minimum number of counts,
+
+```{python}
+# | label: filter-genes-by-counts
+import scvelo as scv
+
+scv.pp.filter_genes(
+    adata,
+    min_shared_counts=30,
+)
+filter_genes_representation = anndata_string(adata)
+print_string_diff(
+    text1=cytotrace_representation,
+    text2=filter_genes_representation,
+    diff_title="Filter genes diff",
+)
+```
+
+normalize the data,
+
+```{python}
+# | label: normalize-per-cell
+scv.pp.normalize_per_cell(adata)
+normalize_representation = anndata_string(adata)
+print_string_diff(
+    text1=filter_genes_representation,
+    text2=normalize_representation,
+    diff_title="Normalize per cell diff",
+)
+```
+
+and filter the genes by their degree of dispersion.
+
+```{python}
+# | label: filter-genes-by-dispersion
+scv.pp.filter_genes_dispersion(
+    adata,
+    n_top_genes=2000,
+)
+filter_dispersion_representation = anndata_string(adata)
+print_string_diff(
+    text1=normalize_representation,
+    text2=filter_dispersion_representation,
+    diff_title="Filter dispersion diff",
+)
+```
+
+Aside from the observation-level subsampling these operations are destructive 
+and so you would need to reload the data to recover its original state after
+attempting to apply them.
+
+Finally we use the scanpy package to log-transform the data.
+Note again we've retained the count data and are only doing this as it is
+expected by a variety of downstream analysis tools.
+
+```{python}
+# | label: log-transform-data-matrix
+import scanpy as sc
+
+sc.pp.log1p(adata)
+preprocessed_data_state_representation = anndata_string(adata)
+print_string_diff(
+    text1=filter_dispersion_representation,
+    text2=preprocessed_data_state_representation,
+    diff_title="Log transform diff",
+)
+```
+
+Having executed this series of functions we've added or updated the following
+components relative to the initial data state
+
+```{python}
+# | label: print-data-state-summary
+print_string_diff(
+    text1=initial_data_state_representation,
+    text2=preprocessed_data_state_representation,
+    diff_title="Filtration and normalization summary diff",
+    diff_context_lines=5,
+)
+```
+
+### Compute nearest neighbors graph
+
+We do not make direct use of nearest neighbor graphs, and some data sets
+come with one precomputed, but we recompute the graph
+for cross-compatability with other tools that may require it.
+
+```{python}
+# | label: compute-nearest-neighbors-graph
+sc.pp.neighbors(
+    adata=adata,
+    n_pcs=30,
+    n_neighbors=30,
+)
+nearest_neighbors_representation = anndata_string(adata)
+print_string_diff(
+    text1=preprocessed_data_state_representation,
+    text2=nearest_neighbors_representation,
+    diff_title="Nearest neighbors diff",
+)
+```
+
+The nearest neighbor graph is used in particular for nearest-neighbor-based 
+averaging of gene expression
+
+```{python}
+# | label: compute-nearest-neighbor-averages
+scv.pp.moments(
+    data=adata,
+    n_pcs=30,
+    n_neighbors=30,
+)
+nearest_neighbor_averaged_representation = anndata_string(adata)
+print_string_diff(
+    text1=nearest_neighbors_representation,
+    text2=nearest_neighbor_averaged_representation,
+    diff_title="Nearest neighbor averaged diff",
+)
+```
+
+### Estimate expression dynamics
+
+We compute the estimates of the expression dynamics parameters using the EM-based
+maximum-likelihood algorithm from the scvelo package
+
+```{python}
+# | label: compute-em-ml-velocity
+scv.tl.recover_dynamics(
+    data=adata,
+    n_jobs=-1,
+    use_raw=False,
+)
+em_ml_dynamics_representation = anndata_string(adata)
+print_string_diff(
+    text1=nearest_neighbor_averaged_representation,
+    text2=em_ml_dynamics_representation,
+    diff_title="EM ML dynamics diff",
+)
+```
+
+and use these to estimate the gene-specific velocities.
+
+```{python}
+# | label: compute-gene-specific-velocities
+scv.tl.velocity(
+    data=adata,
+    mode="dynamical",
+    use_raw=False,
+)
+em_ml_velocities_representation = anndata_string(adata)
+print_string_diff(
+    text1=em_ml_dynamics_representation,
+    text2=em_ml_velocities_representation,
+    diff_title="EM ML velocities diff",
+)
+```
+
+### Compute leiden clusters
+
+We add leiden clustering to the data set for algorithms that work with 
+unlabeled clusters.
+
+```{python}
+# | label: leiden-clustering
+sc.tl.leiden(adata=adata)
+leiden_representation = anndata_string(adata)
+print_string_diff(
+    text1=em_ml_velocities_representation,
+    text2=leiden_representation,
+    diff_title="Leiden clustering diff",
+)
+```
+
+### Subset genes
+
+We also subset the genes to a smaller number to enable faster illustration of
+model training and inference even on laptops with significantly limited resources.
+
+```{python}
+# | label: subset-genes
+likelihood_sorted_genes = (
+    adata.var["fit_likelihood"].sort_values(ascending=False).index
+)
+
+if PYROVELOCITY_DATA_SUBSET:
+    adata = adata[:, likelihood_sorted_genes[:200]].copy()
+
+subset_genes_representation = anndata_string(adata)
+print_string_diff(
+    text1=leiden_representation,
+    text2=subset_genes_representation,
+    diff_title="Subset genes diff",
+)
+```
+
+### Compute velocity graph
+
+We compute the velocity graph based on cosine similarity of the gene-specific velocities
+estimated by the scvelo package.
+
+```{python}
+# | label: compute-velocity-graph
+scv.tl.velocity_graph(
+    data=adata,
+    n_jobs=-1,
+)
+velocity_graph_representation = anndata_string(adata)
+print_string_diff(
+    text1=subset_genes_representation,
+    text2=velocity_graph_representation,
+    diff_title="Velocity graph diff",
+)
+```
+
+### Compute velocity embedding
+
+We compute the velocity embedding using the velocity graph and the gene-specific velocities.
+
+```{python}
+# | label: compute-velocity-embedding
+scv.tl.velocity_embedding(
+    data=adata,
+    basis="umap",
+)
+velocity_embedding_representation = anndata_string(adata)
+print_string_diff(
+    text1=velocity_graph_representation,
+    text2=velocity_embedding_representation,
+    diff_title="Velocity embedding diff",
+)
+```
+
+### Compute gene-wise latent time
+
+We compute the gene-specific latent time estimated by the scvelo package.
+
+```{python}
+# | label: compute-latent-time
+scv.tl.latent_time(
+    data=adata,
+)
+latent_time_representation = anndata_string(adata)
+print_string_diff(
+    text1=velocity_embedding_representation,
+    text2=latent_time_representation,
+    diff_title="Latent time diff",
+)
+```
+
+### Summarize preprocessing
+
+We can finally summarize the changes induced by the preprocessing steps 
+applied to the data set.
+
+```{python}
+# | label: summarize-preprocessing
+print_string_diff(
+    text1=initial_data_state_representation,
+    text2=latent_time_representation,
+    diff_title="Preprocessing summary diff",
+    diff_context_lines=5,
+)
+```
+
+## Plots
+
+### Data quality
+
+We plot the two-dimensional density of genes with greater than or equal to one count in a given cell versus the total counts in that cell for the data set.
+
+```{python}
+# | label: post-filtration-quality-control-metrics
+sc.pp.calculate_qc_metrics(
+    adata=adata,
+    qc_vars=["mt", "ribo"],
+    layer="raw_spliced",
+    percent_top=None,
+    log1p=False,
+    inplace=True,
+)
+qc_metrics_representation = anndata_string(adata)
+print_string_diff(
+    text1=latent_time_representation,
+    text2=qc_metrics_representation,
+    diff_title="Quality control metrics diff",
+)
+```
+
+```{python}
+# | label: post-filtration-plot-genes-by-counts
+import seaborn as sns
+import pandas as pd
+
+numeric_obs = adata.obs.copy()
+numeric_obs["n_genes_by_counts"] = pd.to_numeric(
+    numeric_obs["n_genes_by_counts"],
+    errors="coerce",
+)
+qc_counts = sns.jointplot(
+    data=numeric_obs,
+    x="total_counts",
+    y="n_genes_by_counts",
+    color="#ff6a14",
+    marginal_ticks=True,
+    kind="scatter",
+    alpha=0.4,
+)
+qc_counts.plot_joint(
+    sns.kdeplot,
+    color="gray",
+    alpha=0.6,
+)
+qc_counts.set_axis_labels(
+    xlabel="Total counts in cell",
+    ylabel="Number of genes >=1 count in cell",
+)
+```
+
+### Splice state
+
+We can plot the raw data to ensure we have a general understanding of its
+sparsity, which may be associated with quality.
+We can also plot the raw spliced and unspliced counts.
+
+```{python}
+# | label: plot-raw-data
+from pyrovelocity.plots import plot_spliced_unspliced_histogram
+
+chart = plot_spliced_unspliced_histogram(
+    adata=adata,
+    spliced_layer="raw_spliced",
+    unspliced_layer="raw_unspliced",
+    min_count=3,
+    max_count=200,
+    title="",
+    plot_width=700,
+    plot_height=700,
+)
+chart.save("us_histogram.pdf")
+chart
+```

### 2024-08-12 nbs/guides/preprocessing/preprocessing.qmd
chore(nbs): freeze preprocessing
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 execute: 
-  freeze: false
+  freeze: true
   eval: true

### 2024-08-13 nbs/guides/preprocessing/preprocessing.qmd
fix(nbs): plot mitochondrial and ribosomal percentage histograms in preprocessing guide
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
     ].copy()
+
 subset_anndata_representation = anndata_string(adata)
@@
 
+Prior to filtering any genes or cells we compute the quality control metrics 
+including the percentage of mitochondrial and ribosomal counts in each cell.
+
 ```{python}
@@
 
-adata.var["mt"] = adata.var_names.str.startswith("MT-" or "Mt-" or "mt-")
+adata.var["mt"] = adata.var_names.str.startswith(("MT-", "Mt-", "mt-"))
 adata.var["ribo"] = adata.var_names.str.startswith(
-    "RPS" or "Rps" or "rps" or "RPL" or "Rpl" or "rpl"
+    ("RPS", "Rps", "rps", "RPL", "Rpl", "rpl")
 )
@@
 ```
+
+We plot the percentage of mitochondrial and ribosomal counts in each cell
+
+```{python}
+# | label: pre-filtration-plot-mitochondrial
+import seaborn as sns
+
+ax = sns.histplot(
+    adata.obs,
+    x="pct_counts_mt",
+    color="#ff6a14",
+)
+ax.set_xlabel("mitochondrial counts (%)")
+```
+
+```{python}
+# | label: pre-filtration-plot-ribosomal
+ax = sns.histplot(
+    adata.obs,
+    x="pct_counts_ribo",
+    color="#ff6a14",
+)
+ax.set_xlabel("ribosomal counts (%)")
+```
+
+as well as the number of genes with at least one count in each cell and the
+total counts in each cell.
 
@@
     max_count=200,
-    title="",
-    plot_width=700,
-    plot_height=700,
 )

### 2024-08-18 reproducibility/manuscript/manuscript.qmd
docs(reproducbility): add template
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Quarto Template for Springer Nature
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-nature
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: citeproc
    keep-tex: true
    equal-margins: false
  html:
    code-fold: true
    html-math-method: mathjax
  docx: default
# The `csl` style will be used with LaTeX-based rendering of`nature-pdf`
# if cite-method is `citeproc`. This is also the default if cite-method
# is not specified.
#
# If attempting to render this extension template in an end-user project,
# where you have executed a command like
# $ quarto add sciexp/quarto-springer-nature-template
# you will need to modify the csl path to point to the correct location, e.g.
# csl: _extensions/sciexp/nature/csl/springer-basic-author-date.csl
# where `sciexp` will be used as the namespace for the extension.
# see: https://quarto.org/docs/extensions/distributing.html#github-distribution
csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl
author:
  - name: Author One
    affiliations:
      - name: Harvard University
        id: 1
        department: Department of Government
        address: 1737 Cambridge Street
        city: Cambridge
        state: MA
        postal-code: 02138
      - name: Harvard University
        id: 2
        department: Department of Statistics
        address: 1 Oxford Street
        city: Cambridge
        state: MA
        postal-code: 02138
    attributes:
      equal-contributor: true
  - name: Author Two
    affiliations:
      - name: Yale University
        id: 3
        department: Department of Political Science
        address: 115 Prospect Street
        city: New Haven
        state: CT
        postal-code: 06511
    email: corresponding@email.com
    attributes:
      equal-contributor: true
      corresponding: true
  - name: Author Three
    affiliations:
      - ref: 1
abstract: |
  The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. Authors are advised to check the author instructions for the journal they are submitting to for word limits and if structural elements like subheadings, citations, or equations are permitted.
keywords: [template, demo]
bibliography: bibliography.bib  
---

# Introduction {#sec-intro}

The Introduction section, of referenced text [@greenwade93] expands on the background of the work (some overlap with the Abstract is acceptable). The introduction should not include subheadings.

Springer Nature does not impose a strict layout as standard however authors are advised to check the individual requirements for the journal they are planning to submit to as there may be journal-level preferences. When preparing your text please also be aware that some stylistic choices are not supported in full text XML (publication version), including coloured font. These will not be replicated in the typeset article if it is accepted.

# Using this template {#sec-template}

This Quarto template is unofficial and built out of Springer Nature's template. 
Some examples of commonly used commands and features are listed below, to help you get started.

As seen below, you can mix markdown and Latex with each other, though it's likely best to mostly use markdown.

## Cross Referencing

To reference a figure with example label "plot", use `@fig-plot`. 
Analogously, to reference a table with example label "data", use `@tbl-data`.
To reference a section, such as the Introduction (@sec-intro), use `@sec-intro`.

For complete information on cross referencing with Quarto, see <https://quarto.org/docs/authoring/cross-references.html>.

## Citations

Quarto formats citations and references automatically using the bibliography records in your .bib file. For a single citation use [@greenwade93]. For multiple citations use [@greenwade93;@knuth1984texbook]. If using an author-date-based citation style, then for a citation in parentheses use `[@greenwade93]` and for a text citation use `@greenwade93`. You can drop the author using `[-@greenwade93]`. If the tex output is to be included in a submission to a preprint server or publisher, the default citation method, `citeproc`, may not always produce bibliographies compatible with the provided `bst` styles. In this case, you can set `cite-method: natbib`. See <https://quarto.org/docs/authoring/footnotes-and-citations.html#sec-biblatex>. In this case, you may either restrict usage to the default bracketed key citation formats supported by pandoc, such as [@greenwade93] or [@greenwade93;@knuth1984texbook]. Or, use a CSL file, several of which are vendored in [_extensions/nature/csl](./_extensions/nature/csl/). The advantage of the latter is that references will be consistent across html, docx, and pdf outputs. See <https://pandoc.org/MANUAL.html#citation-syntax> for additional details on pandoc citation styles.

# Tables and Figures

To include figures, you can use Quarto syntax.

![An example figure (an empty plot)](figures/fig.pdf){#fig-plot}

For both figures and tables, you can use LaTeX syntax if you need additional customizability.

For example, to use footnotes within a table, you may want to use LaTeX.

\begin{table}[h]
\caption{LaTeX caption text}\label{tbl-data}%
\begin{tabular}{@{}llll@{}}
\toprule
Column 1 & Column 2  & Column 3 & Column 4\\
\midrule
row 1    & data 1   & data 2  & data 3  \\
row 2    & data 4   & data 5\footnotemark[1]  & data 6  \\
row 3    & data 7   & data 8  & data 9\footnotemark[2]  \\
\botrule
\end{tabular}
\footnotetext{Source: This is an example of table footnote. This is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote. This is an example of table footnote.}
\footnotetext[2]{Example for a second table footnote. This is an example of table footnote.}
\end{table}

# References
<!-- the pattern below controls the placement of the references -->
:::{#refs}
:::

# Appendix {.appendix}

{{< lipsum 5 >}}

### 2024-08-19 reproducibility/manuscript/manuscript.qmd
docs(manuscript): setup frontmatter and fig refs
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: Quarto Template for Springer Nature
+title: "Pyro-Velocity: Probabilistic RNA Velocity inference from single-cell data"
 format:
@@
   docx: default
-# The `csl` style will be used with LaTeX-based rendering of`nature-pdf`
-# if cite-method is `citeproc`. This is also the default if cite-method
-# is not specified.
-#
-# If attempting to render this extension template in an end-user project,
-# where you have executed a command like
-# $ quarto add sciexp/quarto-springer-nature-template
-# you will need to modify the csl path to point to the correct location, e.g.
-# csl: _extensions/sciexp/nature/csl/springer-basic-author-date.csl
-# where `sciexp` will be used as the namespace for the extension.
-# see: https://quarto.org/docs/extensions/distributing.html#github-distribution
 csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl
 author:
-  - name: Author One
+  - name: Qian Qin
     affiliations:
-      - name: Harvard University
+      - name: Massachusetts General Hospital Research Institute, Charlestown, MA
         id: 1
-        department: Department of Government
-        address: 1737 Cambridge Street
+        department: Molecular Pathology Unit
+        city: Charlestown
+        state: MA
+        postal-code: 02129
+      - name: Harvard Medical School, Charlestown, MA
+        id: 2
+        department: Massachusetts General Hospital Cancer Center
+        city: Charlestown
+        state: MA
+        postal-code: 02129
+      - name: Massachusetts General Hospital, Boston, MA
+        id: 3
+        department: Center for Regenerative Medicine
+        city: Boston
+        state: MA
+        postal-code: 02114
+      - name: Harvard University, Cambridge, MA
+        id: 4
+        department: Harvard Stem Cell Institute
         city: Cambridge
         state: MA
-        postal-code: 02138
-      - name: Harvard University
-        id: 2
-        department: Department of Statistics
-        address: 1 Oxford Street
+        postal-code: 02139
+      - name: Cambridge, MA
+        id: 5
+        department: Broad Institute of MIT and Harvard
         city: Cambridge
         state: MA
-        postal-code: 02138
+        postal-code: 02142
+    # attributes:
+    #   equal-contributor: true
+  - name: Eli Bingham
+    affiliations:
+      - ref: 5
+      - name: Cambridge, MA
+        id: 7
+        department: Basis Research Institute
+        city: Cambridge
+        state: MA
+        postal-code: 02142
+  - name: Gioele La Manno
+    affiliations:
+      - name: Laboratory of Neurodevelopmental Systems Biology, Lausanne, Switzerland
+        id: 6
+        department: Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL)
+        city: Lausanne
+        state: Switzerland
+        postal-code: 1015
+    email: gioele.lamanno@epfl.ch
     attributes:
-      equal-contributor: true
-  - name: Author Two
-    affiliations:
-      - name: Yale University
-        id: 3
-        department: Department of Political Science
-        address: 115 Prospect Street
-        city: New Haven
-        state: CT
-        postal-code: 06511
-    email: corresponding@email.com
-    attributes:
-      equal-contributor: true
       corresponding: true
-  - name: Author Three
+  - name: David M. Langenau
     affiliations:
       - ref: 1
+      - ref: 2
+      - ref: 3
+      - ref: 4
+    email: dlangenau@mgh.harvard.edu
+    attributes:
+      corresponding: true
+  - name: Luca Pinello
+    affiliations:
+      - ref: 1
+      - ref: 2
+      - ref: 5
+    email: lpinello@mgh.harvard.edu
+    attributes:
+      corresponding: true
 abstract: |
-  The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
-Authors are advised to check the author instructions for the journal they are submitting to for word limits and if structural elements like subheadings, citations, or equations are permitted.
-keywords: [template, demo]
+  Single-cell RNA Velocity has dramatically advanced our ability to model cellular differentiation and cell fate decisions.
+However, current preprocessing choices and model assumptions often lead to errors in assigning developmental trajectories.
+Here, we develop, Pyro-Velocity, a Bayesian, generative, and multivariate RNA Velocity model to estimate the uncertainty of cell future states.
+This approach models raw sequencing counts with the synchronized cell time across all expressed genes to provide quantifiable and improved information on cell fate choices and developmental trajectory dynamics.
+keywords: [RNA velocity, single-cell RNA sequencing, probabilistic generative modeling, cell fate, developmental trajectory inference]
 bibliography: bibliography.bib
+crossref:
+  custom:
+    - kind: float
+      key: suppfig
+      latex-env: suppfig
+      reference-prefix: Figure S
+      space-before-numbering: false
+      latex-list-of-description: Supplementary Figure
 ---
+
+# Main text {#sec-main}
+
+RNA Velocity is a powerful computational framework to estimate the time derivative of gene expression [@La_Manno2018-lj;@Svensson2018-vk;@Qiu2022-dj;@Bergen2020-pj].
+@fig-conceptual-model.
+@fig-model-results.
+@suppfig-graphical-models.
 
@@
 
-# Tables and Figures
+# Figures
 
-To include figures, you can use Quarto syntax.
+![Pyro-Velocity is a fully generative Bayesian method with uncertainty estimation of velocity vector fields and shared latent time, based on raw counts and without ad-hoc preprocessing steps.
+a.
+Schematic comparing current RNA Velocity and Pyro-Velocity workflows.
+Both methods take in input spliced and unspliced read count tables and output RNA velocity vector fields and latent (shared) time.
+Previous methods require ad-hoc preprocessing steps, including the arbitrary selection of top variable genes, number of principal components, and kNN-pooling to impute both spliced and unspliced expression within the PCA-reduced spliced expression space.
+In addition, manually tuned ad-hoc postprocessing steps are used to compute the final vector field and latent time based on aggregation and smoothing of the independent velocity estimates for each gene which can dramatically change the velocity estimates.
+In contrast, Pyro-Velocity is a fully generative and probabilistic model that uses raw read counts.
+Our model requires minimal preprocessing and postprocessing steps and outputs a common shared time based on all the genes without additional aggregation steps.
+It also provides uncertainty estimation based on posterior samples.
+b.
+The main outputs of the Pyro-Velocity framework.
+Visualization of single gene expression across inferred developmental cell order using rainbow plots (right panel).](figures/fig.pdf){#fig-conceptual-model}
+
+![Pyro-Velocity outperforms other RNA-velocity methods and accurately predicts cell lineage trajectories.
+a,b.
+RNA Velocity model comparisons from fully mature peripheral blood mononuclear cells (PBMCs, t-SNE embedded plots, a) or E15.5 mouse pancreas (UMAP embedded plots, b-c).
+From left to right in panels a and b: 1.
+Cells colored by cell type; 2.
+Stream plot of velocity vector field defined by scVelo; 3.
+Stream plot of the posterior mean vector field from Pyro-Velocity; 4.
+Pyro-Velocity visualization of cell dynamics for 3 arbitrarily selected single cells; 5.
+Single-cell vector field uncertainty where each cell is colored based on the scaled angular standard deviation to visualize uncertainty based on n=30 iterated, posterior samples depicted as arrows noted throughout our analysis; 6.
+Averaged vector field uncertainty where each vector is colored by the averaged vector field uncertainty.
+The scaled angular standard deviation range is [0 360], where 0 corresponds to the highest confidence for a single direction, 360 corresponds to the highest uncertainty, and no directionality preference based on 30 posterior samples (4-6).
+Vectors were projected into the UMAP space using the scVelo API based on cosine similarity-derived transition matrix.
+c.
+UMAP embedded cells colored by the average cell shared time across 30 posterior samples.
+The Spearman’s correlation between Pyro-Velocity and Cytotrace is provided.
+d.
+Pyro-Velocity automatic gene selection and ranking for predicting important genes involved in pancreas development.
+e.
+Phase portraits of posterior samples average spliced and unspliced gene expression (left), rainbow plots showing denoised spliced gene expression (middle), and gene expression within UMAP embedded space.
+Single dots correspond to individual cells (left, right) or vertical lines (middle).
+Cell type annotation coloring is the same as in panel b.](figures/fig.pdf){#fig-model-results}
+
+![Pyro-Velocity accurately predicts cell fate choices and lineage trajectory in human LARRY barcoded hematopoietic cells.
+Systematic benchmarking of cell fate prediction methods - predicting uni-lineage cell fates (a-b, Model 1), bifurcated lineages (c, Model 2), or all lineages together (d, Model 2).
+Panels from left to right show on the same SPRING embedding: 1.
+Cell types; 2.
+Clonal progression vector field derived based on the observed clonal barcodes (Methods); 3.
+scVelo recovered vector field; 4.
+Pyro-Velocity single cell level vector field uncertainty (angular standard deviation from 30 posterior samples); 5.
+Pyro-Velocity recovered vector field; vectors are colored by the averaged vector field uncertainty; 6.
+Clonal fate potency score using Cospar24; 7.
+scVelo recovered latent time; 8.
+Pyro-Velocity inference of posterior mean shared time per cell.
+Cosine similarity and Spearman’s correlation for the different methods with the clonal progression and fate potency are reported in the titles.](figures/fig.pdf){#fig-lineage-tracing}
+
+# Supplementary Figures
+
+![Graphical representations detail the generative process of the Pyro-Velocity models and highlight their differences.
+a.
+Pyro-Velocity Model 1.
+Each rectangle represents one data dimension (e.g., gene or cell).
+The arrows represent the directed hierarchical dependency among random variables.
+The white circles correspond to the modeled random variables.
+The probability distributions of these variables are denoted on the right legend.
+The gray circles correspond to intermediate results computed from random variables.
+In this model, the kinetics random variables for the cell shared time and gene are modeled with two constraints: 1.
+the shared time is strictly larger than the gene-specific activation time $t_{0}^{(0)}$, 2.
+the gene expression in the activate state starts with the initial condition $(u_{0}=0,\,s_{0}=0)$, as in previous methods[@La_Manno2018-lj;@Bergen2020-pj] (Methods).
+b.
+Pyro-Velocity Model 2.
+This figure follows the same graphical convention as in a.
+In this model $t_{0}^{(0)}$ is independent from the shared time, thus allowing a time lag for the gene activation phase.
+In addition, a positive starting basal gene expression level $(u_{0},\,s_{0})$ different than $(u_{0}=0,\,s_{0}=0)$ is allowed.
+(Methods).
+c.
+Schematic illustrating how the constraints proposed in Model 1 (left) and Model 2 (right) are reflected in the modeling of individual gene expression.
+The x-axis corresponds to the shared time, the y-axis corresponds to the posterior mean of the spliced expression level, the vertical lines with arrows show the start and end time of the observed cell sampling, t0 represents the gene-specific activation time $t_{0}^{(0)}$, the red lines highlight the differences for the potential gene-specific time lag when $t_{0}^{(0)} > {\min}\!⁡(t)$, and the initial spliced gene expression.
+The same differences can also be illustrated for the unspliced gene expression profiles (not shown).](figures/fig.pdf){#suppfig-graphical-models}
 
 ![An example figure (an empty plot)](figures/fig.pdf){#fig-plot}
-
-For both figures and tables, you can use LaTeX syntax if you need additional customizability.
-
-For example, to use footnotes within a table, you may want to use LaTeX.
 

### 2024-08-19 reproducibility/manuscript/manuscript.qmd
docs(manuscript): specify newpage breaks
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
       cite-style: sn-nature
+    classoption: 
+      - lineno
     # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
@@
     equal-margins: false
+    include-in-header: header.tex
   html:
@@
 
+\newpage
+
 # Main text {#sec-main}
@@
 See <https://pandoc.org/MANUAL.html#citation-syntax> for additional details on pandoc citation styles.
+
+\newpage
+
+# References
+:::{#refs}
+:::
+
+\newpage
 
@@
 
+\newpage
+
 # Supplementary Figures
@@
 
+![Extended version of @fig-model-results for evaluating Pyro-Velocity (Model 1) and scVelo on the PBMC and pancreas datasets.
+a.
+Left figure shows the scVelo vector field using the default user options, right figure shows the evaluation of the scVelo vector field using different combinations of user options within Epsilon cells.
+b.
+Pyro-Velocity additional single cell level uncertainty diagnostic plots on the fully mature PBMC dataset, from left to right: 1.
+standard deviation of posterior samples of shared time, 2.
+Rayleigh test of posterior samples of the vector field (the title reports false discovery rate after p-value correction for multiple tests across cells), 3.
+transcriptional state uncertainty of predicted raw read count per cell.
+The contour lines show the top 10% of the most uncertain cells.
+c.
+For the pancreas dataset, the figures show from left to right: UMAP rendering of 1.
+Cytotrace, 2.
+scVelo latent time, 3.
+Posterior mean of Pyro-Velocity shared time averaged across 30 posterior samples, 4.
+Top 50 markers overlapping between Pyro-Velocity and scVelo (Methods, Supplementary Table 1-2), 5.
+The same uncertainty diagnostic plots per cell as in b.
+d.
+The same plot as Fig.
+2d.
+e.
+Selection of the top negatively correlated genes with the cell shared time based on the posterior mean on denoised spliced expression and negative mean absolute error of gene expression prediction (Methods).
+f.
+Phase portraits, rainbow plots, and UMAP rendering of posterior mean of gene expression for the genes selected in e.](figures/fig.pdf){#suppfig-model-results}
+
 ![An example figure (an empty plot)](figures/fig.pdf){#fig-plot}
 
-\begin{table}[h]
+<!-- \begin{table}[h]
 \caption{LaTeX caption text}\label{tbl-data}%
@@
 \footnotetext[2]{Example for a second table footnote. This is an example of table footnote.}
-\end{table}
+\end{table} -->
 
-# References
-<!-- the pattern below controls the placement of the references -->
-:::{#refs}
-:::
+\newpage
 

### 2024-08-19 reproducibility/manuscript/manuscript.qmd
docs(manuscript): add `\FloatBarrier`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(manuscript): update introductory references
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 RNA Velocity is a powerful computational framework to estimate the time derivative of gene expression [@La_Manno2018-lj;@Svensson2018-vk;@Qiu2022-dj;@Bergen2020-pj].
-@fig-conceptual-model.
-@fig-model-results.
-@suppfig-graphical-models.
+The framework has been used to study developmental cell lineage trajectories over a broad spectrum of biological processes, such as  haematopoiesis [@Qiu2022-dj], pancreas development [@Bastidas-Ponce2019-lf], and neural progenitors differentiation [@Bergen2020-pj].
+Yet, recent state-of-the-art RNA Velocity approaches such as velocyto and scVelo [@La_Manno2018-lj] have several limitations.
+First, they may predict developmental trajectories that do not exist in nature, as it has been shown in the analysis of fully mature blood cells [@Zheng2017-bz;@Bergen2021-qz].
+Second, data preprocessing steps (e.g., selection of top variable genes, dimensionality reduction) affect velocity estimates that are difficult to characterize and might lead to artefactual predictions.
+In addition, non-linear spliced and unspliced imputation procedures could distort phase portrait geometry and recovered trajectories (Fig.
+1a).
+Third, most RNA velocity algorithms perform estimates independently for each gene, and therefore cannot exploit the natural interdependence of gene expression.
+Fourth, a unified description of the temporal relation between cells can only be assigned based on postprocessing steps that aggregate heuristically the gene-wise fits into a shared latent time (Fig.
+1a).
+Finally, current RNA velocity frameworks only output point estimates of the velocity vector field and do not provide uncertainty estimation.
+Importantly, the absence of statistical confidence estimates increases false discoveries as it leaves users without a way to evaluate the robustness of the methods’ outputs.
+To overcome these limitations, we developed a probabilistic generative model, named Pyro-Velocity, that estimates RNA dynamics and infers cell trajectories using Pyro8, a robust and efficient probabilistic programming language.
+
+@fig-conceptual-model. @fig-model-results. @suppfig-graphical-models.
 
@@
 
+\FloatBarrier
 \newpage
@@
 Pyro-Velocity recovered vector field; vectors are colored by the averaged vector field uncertainty; 6.
-Clonal fate potency score using Cospar24; 7.
+Clonal fate potency score using Cospar [@Wang2022-xb]; 7.
 scVelo recovered latent time; 8.
@@
 
-![An example figure (an empty plot)](figures/fig.pdf){#fig-plot}
+![Extension of @fig-model-results, Pyro-Velocity (Model 2) outperforms scVelo on PBMC and pancreas datasets.
+Same format as in @fig-model-results and @suppfig-model-results](figures/fig.pdf){#suppfig-model-2-result}
+
+![Extension of @fig-lineage-tracing, Pyro-velocity (Model 1 for uni-fate data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on scRNA-seq datasets with lineage barcoding information (LARRY dataset).
+a.
+SPRING embedding rendering of Cytotrace scores, the title reports the Spearman’s correlation between the Cytotrace and Cospar fate potency scores, b.
+Pyro-Velocity additional single cell level uncertainty diagnostic plots based on the multi-fate LARRY dataset, from left to right: 1.
+Standard deviation of posterior samples of shared time per cell, 2.
+Rayleigh test of posterior samples of vector field per cell (the title reports the false discovery rate after p-value correction for multiple tests across cells), 3.
+Transcriptional state uncertainty of predicted raw read count per cell.
+The contour lines show the top 10% of the most uncertain cells (Methods), c.
+Pairwise Spearman’s correlation between Cytotrace/Cospar fate potency scores and velocity methods for the four different benchmarking datasets selected based on the lineage barcoding information from LARRY data.
+d.
+SPRING embedding rendering of posterior mean of shared time across 30 posterior samples, the title shows the Spearman’s correlation of mean shared time with Cytotrace.
+e.
+Selection of the top positively correlated genes with the cell shared time based on the posterior mean of spliced expression.
+f.
+Phase portraits, rainbow plots, and UMAP rendering with splicing gene expression level for the selected genes in e.](figures/fig.pdf){#suppfig-lineage-tracing}
+
+![Extension of @fig-lineage-tracing presenting an extended comparison of Pyro-Velocity and scVelo on scRNA-seq datasets with lineage barcoding information (LARRY dataset).
+a.
+uni-fate Monocyte lineage analysis with Pyro-Velocity Model 2, b.
+uni-fate Neutrophil lineage analysis with Pyro-Velocity Model 2, c.
+Neutrophil and Monocyte bi-fate lineage analysis with Pyro-Velocity Model 1, and d.
+multi-fate lineage analysis with Pyro-Velocity Model 1.
+The panel descriptions are the same as in @fig-lineage-tracing](figures/fig.pdf){#suppfig-lineage-tracing-2}
+
+![An example figure (an empty plot)](figures/fig.pdf){#suppfig-plot}
 
@@
 
+\FloatBarrier
 \newpage
 
-# Appendix {.appendix}
-
-{{< lipsum 5 >}}
+# Supporting Information {.appendix}

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): enable pandoc reference-doc
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
     html-math-method: mathjax
-  docx: default
+  docx:
+    reference-doc: styles.docx
 csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): enable authors-block filter
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-08-19 reproducibility/manuscript/manuscript.qmd
@@
     html-math-method: mathjax
-  docx: default
+  docx:
+    reference-doc: styles.docx
+filters:
+  - authors-block
 csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): disable authors-block filter
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-08-19 reproducibility/manuscript/manuscript.qmd
@@
     html-math-method: mathjax
-  docx: default
+  docx:
+    reference-doc: styles.docx
+# filters:
+#   - authors-block
 csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl
-author:
-  - name: Qian Qin
 affiliations:
@@
     postal-code: 02142
-    # attributes:
-    #   equal-contributor: true
-  - name: Eli Bingham
-    affiliations:
-      - ref: 5
+  - name: Laboratory of Neurodevelopmental Systems Biology, Lausanne, Switzerland
+    id: 6
+    department: Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL)
+    city: Lausanne
+    state: Switzerland
+    postal-code: 1015
   - name: Cambridge, MA
@@
     postal-code: 02142
+authors:
+  - name: Qian Qin
+    affiliations:
+      - ref: 1
+      - ref: 2
+      - ref: 3
+      - ref: 4
+      - ref: 5
+    # attributes:
+    #   equal-contributor: true
+  - name: Eli Bingham
+    affiliations:
+      - ref: 5
+      - ref: 7
   - name: Gioele La Manno
     affiliations:
-      - name: Laboratory of Neurodevelopmental Systems Biology, Lausanne, Switzerland
-        id: 6
-        department: Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL)
-        city: Lausanne
-        state: Switzerland
-        postal-code: 1015
+      - ref: 6
     email: gioele.lamanno@epfl.ch

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): set fig prefix `Fig.`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: "Pyro-Velocity: Probabilistic RNA Velocity inference from single-cell data"
+title: "Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data"
 format:
@@
 crossref:
+  fig-prefix: Fig.
   custom:
@@
       latex-env: suppfig
-      reference-prefix: Figure S
+      reference-prefix: Fig. S
       space-before-numbering: false
@@
 First, they may predict developmental trajectories that do not exist in nature, as it has been shown in the analysis of fully mature blood cells [@Zheng2017-bz;@Bergen2021-qz].
-Second, data preprocessing steps (e.g., selection of top variable genes, dimensionality reduction) affect velocity estimates that are difficult to characterize and might lead to artefactual predictions.
-In addition, non-linear spliced and unspliced imputation procedures could distort phase portrait geometry and recovered trajectories (Fig.
-1a).
+Second, data preprocessing steps (e.g., selection of top variable genes, dimensionality reduction) affect velocity estimates that are difficult to characterize and might lead to artifactual predictions.
+In addition, non-linear spliced and unspliced imputation procedures could distort phase portrait geometry and recovered trajectories ([@fig-model]a).
 Third, most RNA velocity algorithms perform estimates independently for each gene, and therefore cannot exploit the natural interdependence of gene expression.
-Fourth, a unified description of the temporal relation between cells can only be assigned based on postprocessing steps that aggregate heuristically the gene-wise fits into a shared latent time (Fig.
-1a).
+Fourth, a unified description of the temporal relation between cells can only be assigned based on postprocessing steps that aggregate heuristically the gene-wise fits into a shared latent time ([@fig-model]a).
 Finally, current RNA velocity frameworks only output point estimates of the velocity vector field and do not provide uncertainty estimation.
 Importantly, the absence of statistical confidence estimates increases false discoveries as it leaves users without a way to evaluate the robustness of the methods’ outputs.
-To overcome these limitations, we developed a probabilistic generative model, named Pyro-Velocity, that estimates RNA dynamics and infers cell trajectories using Pyro8, a robust and efficient probabilistic programming language.
-
-@fig-conceptual-model. @fig-model-results. @suppfig-graphical-models.
+To overcome these limitations, we developed a probabilistic generative model, named *Pyro&thinsp;-Velocity*, that estimates RNA dynamics and infers cell trajectories using Pyro [@Bingham2018-id], a robust and efficient probabilistic programming language.
 
@@
 
-\newpage
+{{< pagebreak >}}
 
@@
 \FloatBarrier
-\newpage
+{{< pagebreak >}}
 
@@
 
-![Pyro-Velocity is a fully generative Bayesian method with uncertainty estimation of velocity vector fields and shared latent time, based on raw counts and without ad-hoc preprocessing steps.
+::: {#fig-model}
+
+![](figures/fig.pdf)
+
+\label{fig-model-inputs} 
+\label{fig-model-outputs}
+
+Pyro-Velocity is a fully generative Bayesian method with uncertainty estimation of velocity vector fields and shared latent time, based on raw counts and without ad-hoc preprocessing steps.
 a.
@@
 The main outputs of the Pyro-Velocity framework.
-Visualization of single gene expression across inferred developmental cell order using rainbow plots (right panel).](figures/fig.pdf){#fig-conceptual-model}
+Visualization of single gene expression across inferred developmental cell order using rainbow plots (right panel).
+:::
 

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): enable toc
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 title: "Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data"
+number-sections: true
+toc: true
+toc-title: Contents
 format:
@@
 To overcome these limitations, we developed a probabilistic generative model, named *Pyro&thinsp;-Velocity*, that estimates RNA dynamics and infers cell trajectories using Pyro [@Bingham2018-id], a robust and efficient probabilistic programming language.
+
+Pyro-Velocity recasts the velocity estimation problem into a latent variable posterior probability inference (Methods) [@sec-methods].
+The proposed model is generative and fully Bayesian, with the different parameters considered as latent random variables.
+Each parameter is assigned a prior distribution, and the model is then conditioned on the raw data (both spliced and unspliced molecule raw counts) to estimate posterior probability distributions and to enable uncertainty estimation of velocity.
+Central to the Pyro-Velocity model is a shared latent time (t): a global latent variable placing each cell at a particular instant of the modeled dynamical process.
+From this latent time, the expectation of spliced and unspliced RNA levels for each gene can be computed knowing gene-specific kinetics parameters (transcriptional (), splicing (), and degradation rates () and two gene-specific switching times (t0k), and gene-specific transcriptional state k) and integrating the RNA velocity differential equations analogously to what proposed in the dynamical RNA velocity model by Bergen et al.4,9.
+However, we innovate the structure and formulation of the latent time given that our model is set to jointly fit the entire raw data matrix, not individual genes.
+We approach the task of estimating the posterior distribution by restricting the search within a family of a variational distribution approximating the true posterior, and taking advantage of the automatic differentiation variational inference (ADVI) capabilities of  Pyro8,10 (Fig.
+1a, Methods).
+Notably, our generative Bayesian model enables to sample from the posterior distributions of the latent and dependent variables, allowing uncertainty estimation of phase portraits, high dimensional velocity vector field, and shared latent time.
+Around this model, we built several visualizations to assess uncertainty and functions to recover putative cell fate priming genes and to investigate how the denoised gene expression recapitulates known cell-type annotations and their developmental order (Fig.
+1b, Methods).
+Although several RNA velocity methods11–15 have been recently proposed, and many of them have not been peer-reviewed, here we showcase the advantages and innovations of Pyro-Velocity by comparing it with the current state-of-the-art dynamical RNA velocity model proposed by Bergen et al.4, and implemented in scVelo.
+
+# Methods {#sec-methods}
+
+## Model formulation {#sec-model}
+
+We assume the dynamical gene expression is determined by the RNA splicing process, and infer the unspliced and spliced gene expression level from the differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo [@Bergen2020-pj]
+
+\begin{align}
+\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}} &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
+\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}} &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
+\end{align}
 
@@
 ::: {#fig-model}
-
 ![](figures/fig.pdf)
-
-\label{fig-model-inputs} 
-\label{fig-model-outputs}
 

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): enable citation links
- toggle citeproc/natbib

Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
     # to use the bst styles from the upstream template.
-    cite-method: citeproc
+    cite-method: natbib
     keep-tex: true
@@
     include-in-header: header.tex
+    link-citations: true
   html:
@@
 
-Pyro-Velocity recasts the velocity estimation problem into a latent variable posterior probability inference (Methods) [@sec-methods].
+Pyro-Velocity recasts the velocity estimation problem into a latent variable posterior probability inference (\nameref{sec-methods}).
 The proposed model is generative and fully Bayesian, with the different parameters considered as latent random variables.

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): lint
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: "Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data"
+title: |
+  Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data
 number-sections: true
@@
     id: 6
-    department: Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL)
+    department: |
+      Brain and Mind Institute, School of Life Sciences, 
+      École Polytechnique Fédérale de Lausanne (EPFL)
     city: Lausanne
@@
 abstract: |
-  Single-cell RNA Velocity has dramatically advanced our ability to model cellular differentiation and cell fate decisions.
-However, current preprocessing choices and model assumptions often lead to errors in assigning developmental trajectories.
-Here, we develop, Pyro-Velocity, a Bayesian, generative, and multivariate RNA Velocity model to estimate the uncertainty of cell future states.
-This approach models raw sequencing counts with the synchronized cell time across all expressed genes to provide quantifiable and improved information on cell fate choices and developmental trajectory dynamics.
-keywords: [RNA velocity, single-cell RNA sequencing, probabilistic generative modeling, cell fate, developmental trajectory inference]
+  Single-cell RNA Velocity has dramatically advanced our ability to model
+  cellular differentiation and cell fate decisions. However, current
+  preprocessing choices and model assumptions often lead to errors in assigning
+  developmental trajectories. Here, we develop, Pyro-Velocity, a Bayesian,
+  generative, and multivariate RNA Velocity model to estimate the uncertainty of
+  cell future states. This approach models raw sequencing counts with the
+  synchronized cell time across all expressed genes to provide quantifiable and
+  improved information on cell fate choices and developmental trajectory
+  dynamics.
+keywords: [
+  RNA velocity, 
+  single-cell RNA sequencing, 
+  probabilistic generative modeling, 
+  cell fate, 
+  developmental trajectory inference
+]
 bibliography: bibliography.bib
@@
 
-RNA Velocity is a powerful computational framework to estimate the time derivative of gene expression [@La_Manno2018-lj;@Svensson2018-vk;@Qiu2022-dj;@Bergen2020-pj].
-The framework has been used to study developmental cell lineage trajectories over a broad spectrum of biological processes, such as  haematopoiesis [@Qiu2022-dj], pancreas development [@Bastidas-Ponce2019-lf], and neural progenitors differentiation [@Bergen2020-pj].
-Yet, recent state-of-the-art RNA Velocity approaches such as velocyto and scVelo [@La_Manno2018-lj] have several limitations.
-First, they may predict developmental trajectories that do not exist in nature, as it has been shown in the analysis of fully mature blood cells [@Zheng2017-bz;@Bergen2021-qz].
-Second, data preprocessing steps (e.g., selection of top variable genes, dimensionality reduction) affect velocity estimates that are difficult to characterize and might lead to artifactual predictions.
-In addition, non-linear spliced and unspliced imputation procedures could distort phase portrait geometry and recovered trajectories ([@fig-model]a).
-Third, most RNA velocity algorithms perform estimates independently for each gene, and therefore cannot exploit the natural interdependence of gene expression.
-Fourth, a unified description of the temporal relation between cells can only be assigned based on postprocessing steps that aggregate heuristically the gene-wise fits into a shared latent time ([@fig-model]a).
-Finally, current RNA velocity frameworks only output point estimates of the velocity vector field and do not provide uncertainty estimation.
-Importantly, the absence of statistical confidence estimates increases false discoveries as it leaves users without a way to evaluate the robustness of the methods’ outputs.
-To overcome these limitations, we developed a probabilistic generative model, named *Pyro&thinsp;-Velocity*, that estimates RNA dynamics and infers cell trajectories using Pyro [@Bingham2018-id], a robust and efficient probabilistic programming language.
+RNA Velocity is a powerful computational framework to estimate the time
+derivative of gene expression
+[@La_Manno2018-lj;@Svensson2018-vk;@Qiu2022-dj;@Bergen2020-pj]. The framework
+has been used to study developmental cell lineage trajectories over a broad
+spectrum of biological processes, such as  haematopoiesis [@Qiu2022-dj],
+pancreas development [@Bastidas-Ponce2019-lf], and neural progenitors
+differentiation [@Bergen2020-pj]. Yet, recent state-of-the-art RNA Velocity
+approaches such as velocyto and scVelo [@La_Manno2018-lj] have several
+limitations. First, they may predict developmental trajectories that do not
+exist in nature, as it has been shown in the analysis of fully mature blood
+cells [@Zheng2017-bz;@Bergen2021-qz]. Second, data preprocessing steps (e.g.,
+selection of top variable genes, dimensionality reduction) affect velocity
+estimates that are difficult to characterize and might lead to artifactual
+predictions. In addition, non-linear spliced and unspliced imputation procedures
+could distort phase portrait geometry and recovered trajectories
+([@fig-model]a). Third, most RNA velocity algorithms perform estimates
+independently for each gene, and therefore cannot exploit the natural
+interdependence of gene expression. Fourth, a unified description of the
+temporal relation between cells can only be assigned based on postprocessing
+steps that aggregate heuristically the gene-wise fits into a shared latent time
+([@fig-model]a). Finally, current RNA velocity frameworks only output point
+estimates of the velocity vector field and do not provide uncertainty
+estimation. Importantly, the absence of statistical confidence estimates
+increases false discoveries as it leaves users without a way to evaluate the
+robustness of the methods’ outputs. To overcome these limitations, we developed
+a probabilistic generative model, named *Pyro&thinsp;-Velocity*, that estimates
+RNA dynamics and infers cell trajectories using Pyro [@Bingham2018-id], a robust
+and efficient probabilistic programming language.
 
-Pyro-Velocity recasts the velocity estimation problem into a latent variable posterior probability inference (\nameref{sec-methods}).
-The proposed model is generative and fully Bayesian, with the different parameters considered as latent random variables.
-Each parameter is assigned a prior distribution, and the model is then conditioned on the raw data (both spliced and unspliced molecule raw counts) to estimate posterior probability distributions and to enable uncertainty estimation of velocity.
-Central to the Pyro-Velocity model is a shared latent time (t): a global latent variable placing each cell at a particular instant of the modeled dynamical process.
-From this latent time, the expectation of spliced and unspliced RNA levels for each gene can be computed knowing gene-specific kinetics parameters (transcriptional (), splicing (), and degradation rates () and two gene-specific switching times (t0k), and gene-specific transcriptional state k) and integrating the RNA velocity differential equations analogously to what proposed in the dynamical RNA velocity model by Bergen et al.4,9.
-However, we innovate the structure and formulation of the latent time given that our model is set to jointly fit the entire raw data matrix, not individual genes.
-We approach the task of estimating the posterior distribution by restricting the search within a family of a variational distribution approximating the true posterior, and taking advantage of the automatic differentiation variational inference (ADVI) capabilities of  Pyro8,10 (Fig.
-1a, Methods).
-Notably, our generative Bayesian model enables to sample from the posterior distributions of the latent and dependent variables, allowing uncertainty estimation of phase portraits, high dimensional velocity vector field, and shared latent time.
-Around this model, we built several visualizations to assess uncertainty and functions to recover putative cell fate priming genes and to investigate how the denoised gene expression recapitulates known cell-type annotations and their developmental order (Fig.
-1b, Methods).
-Although several RNA velocity methods11–15 have been recently proposed, and many of them have not been peer-reviewed, here we showcase the advantages and innovations of Pyro-Velocity by comparing it with the current state-of-the-art dynamical RNA velocity model proposed by Bergen et al.4, and implemented in scVelo.
+Pyro-Velocity recasts the velocity estimation problem into a latent variable
+posterior probability inference (\nameref{sec-methods}). The proposed model is
+generative and fully Bayesian, with the different parameters considered as
+latent random variables. Each parameter is assigned a prior distribution, and
+the model is then conditioned on the raw data (both spliced and unspliced
+molecule raw counts) to estimate posterior probability distributions and to
+enable uncertainty estimation of velocity. Central to the Pyro-Velocity model is
+a shared latent time (t): a global latent variable placing each cell at a
+particular instant of the modeled dynamical process. From this latent time, the
+expectation of spliced and unspliced RNA levels for each gene can be computed
+knowing gene-specific kinetics parameters (transcriptional (), splicing (), and
+degradation rates () and two gene-specific switching times (t0k), and
+gene-specific transcriptional state k) and integrating the RNA velocity
+differential equations analogously to what proposed in the dynamical RNA
+velocity model by Bergen et al.4,9. However, we innovate the structure and
+formulation of the latent time given that our model is set to jointly fit the
+entire raw data matrix, not individual genes. We approach the task of estimating
+the posterior distribution by restricting the search within a family of a
+variational distribution approximating the true posterior, and taking advantage
+of the automatic differentiation variational inference (ADVI) capabilities of
+Pyro8,10 (Fig. 1a, Methods). Notably, our generative Bayesian model enables to
+sample from the posterior distributions of the latent and dependent variables,
+allowing uncertainty estimation of phase portraits, high dimensional velocity
+vector field, and shared latent time. Around this model, we built several
+visualizations to assess uncertainty and functions to recover putative cell fate
+priming genes and to investigate how the denoised gene expression recapitulates
+known cell-type annotations and their developmental order (Fig. 1b, Methods).
+Although several RNA velocity methods11–15 have been recently proposed, and many
+of them have not been peer-reviewed, here we showcase the advantages and
+innovations of Pyro-Velocity by comparing it with the current state-of-the-art
+dynamical RNA velocity model proposed by Bergen et al.4, and implemented in
+scVelo.
 
@@
 
-We assume the dynamical gene expression is determined by the RNA splicing process, and infer the unspliced and spliced gene expression level from the differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo [@Bergen2020-pj]
+We assume the dynamical gene expression is determined by the RNA splicing
+process, and infer the unspliced and spliced gene expression level from the
+differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo
+[@Bergen2020-pj]
 
 \begin{align}
-\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}} &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
-\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}} &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
+\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
+  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
+\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
+  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
 \end{align}
-
-# Introduction {#sec-intro}
-
-The Introduction section, of referenced text [@greenwade93] expands on the background of the work (some overlap with the Abstract is acceptable). The introduction should not include subheadings.
-
-Springer Nature does not impose a strict layout as standard however authors are advised to check the individual requirements for the journal they are planning to submit to as there may be journal-level preferences.
-When preparing your text please also be aware that some stylistic choices are not supported in full text XML (publication version), including coloured font.
-These will not be replicated in the typeset article if it is accepted.
-
-# Using this template {#sec-template}
-
-This Quarto template is unofficial and built out of Springer Nature's template. 
-Some examples of commonly used commands and features are listed below, to help you get started.
-
-As seen below, you can mix markdown and Latex with each other, though it's likely best to mostly use markdown.
-
-## Cross Referencing
-
-To reference a figure with example label "plot", use `@fig-plot`. 
-Analogously, to reference a table with example label "data", use `@tbl-data`.
-To reference a section, such as the Introduction (@sec-intro), use `@sec-intro`.
-
-For complete information on cross referencing with Quarto, see <https://quarto.org/docs/authoring/cross-references.html>.
-
-## Citations
-
-Quarto formats citations and references automatically using the bibliography records in your .bib file.
-For a single citation use [@greenwade93].
-For multiple citations use [@greenwade93;@knuth1984texbook].
-If using an author-date-based citation style, then for a citation in parentheses use `[@greenwade93]` and for a text citation use `@greenwade93`.
-You can drop the author using `[-@greenwade93]`.
-If the tex output is to be included in a submission to a preprint server or publisher, the default citation method, `citeproc`, may not always produce bibliographies compatible with the provided `bst` styles.
-In this case, you can set `cite-method: natbib`.
-See <https://quarto.org/docs/authoring/footnotes-and-citations.html#sec-biblatex>.
-In this case, you may either restrict usage to the default bracketed key citation formats supported by pandoc, such as [@greenwade93] or [@greenwade93;@knuth1984texbook].
-Or, use a CSL file, several of which are vendored in [_extensions/nature/csl](./_extensions/nature/csl/).
-The advantage of the latter is that references will be consistent across html, docx, and pdf outputs.
-See <https://pandoc.org/MANUAL.html#citation-syntax> for additional details on pandoc citation styles.
 
@@
 # References
+
 :::{#refs}
@@
 
-Pyro-Velocity is a fully generative Bayesian method with uncertainty estimation of velocity vector fields and shared latent time, based on raw counts and without ad-hoc preprocessing steps.
-a.
-Schematic comparing current RNA Velocity and Pyro-Velocity workflows.
-Both methods take in input spliced and unspliced read count tables and output RNA velocity vector fields and latent (shared) time.
-Previous methods require ad-hoc preprocessing steps, including the arbitrary selection of top variable genes, number of principal components, and kNN-pooling to impute both spliced and unspliced expression within the PCA-reduced spliced expression space.
-In addition, manually tuned ad-hoc postprocessing steps are used to compute the final vector field and latent time based on aggregation and smoothing of the independent velocity estimates for each gene which can dramatically change the velocity estimates.
-In contrast, Pyro-Velocity is a fully generative and probabilistic model that uses raw read counts.
-Our model requires minimal preprocessing and postprocessing steps and outputs a common shared time based on all the genes without additional aggregation steps.
-It also provides uncertainty estimation based on posterior samples.
-b.
-The main outputs of the Pyro-Velocity framework.
-Visualization of single gene expression across inferred developmental cell order using rainbow plots (right panel).
+Pyro-Velocity is a fully generative Bayesian method with uncertainty estimation
+of velocity vector fields and shared latent time, based on raw counts and
+without ad-hoc preprocessing steps. a. Schematic comparing current RNA Velocity
+and Pyro-Velocity workflows. Both methods take in input spliced and unspliced
+read count tables and output RNA velocity vector fields and latent (shared)
+time. Previous methods require ad-hoc preprocessing steps, including the
+arbitrary selection of top variable genes, number of principal components, and
+kNN-pooling to impute both spliced and unspliced expression within the
+PCA-reduced spliced expression space. In addition, manually tuned ad-hoc
+postprocessing steps are used to compute the final vector field and latent time
+based on aggregation and smoothing of the independent velocity estimates for
+each gene which can dramatically change the velocity estimates. In contrast,
+Pyro-Velocity is a fully generative and probabilistic model that uses raw read
+counts. Our model requires minimal preprocessing and postprocessing steps and
+outputs a common shared time based on all the genes without additional
+aggregation steps. It also provides uncertainty estimation based on posterior
+samples. b. The main outputs of the Pyro-Velocity framework. Visualization of
+single gene expression across inferred developmental cell order using rainbow
+plots (right panel).
 :::
 
-![Pyro-Velocity outperforms other RNA-velocity methods and accurately predicts cell lineage trajectories.
-a,b.
-RNA Velocity model comparisons from fully mature peripheral blood mononuclear cells (PBMCs, t-SNE embedded plots, a) or E15.5 mouse pancreas (UMAP embedded plots, b-c).
-From left to right in panels a and b: 1.
-Cells colored by cell type; 2.
-Stream plot of velocity vector field defined by scVelo; 3.
-Stream plot of the posterior mean vector field from Pyro-Velocity; 4.
-Pyro-Velocity visualization of cell dynamics for 3 arbitrarily selected single cells; 5.
-Single-cell vector field uncertainty where each cell is colored based on the scaled angular standard deviation to visualize uncertainty based on n=30 iterated, posterior samples depicted as arrows noted throughout our analysis; 6.
-Averaged vector field uncertainty where each vector is colored by the averaged vector field uncertainty.
-The scaled angular standard deviation range is [0 360], where 0 corresponds to the highest confidence for a single direction, 360 corresponds to the highest uncertainty, and no directionality preference based on 30 posterior samples (4-6).
-Vectors were projected into the UMAP space using the scVelo API based on cosine similarity-derived transition matrix.
-c.
-UMAP embedded cells colored by the average cell shared time across 30 posterior samples.
-The Spearman’s correlation between Pyro-Velocity and Cytotrace is provided.
-d.
-Pyro-Velocity automatic gene selection and ranking for predicting important genes involved in pancreas development.
-e.
-Phase portraits of posterior samples average spliced and unspliced gene expression (left), rainbow plots showing denoised spliced gene expression (middle), and gene expression within UMAP embedded space.
-Single dots correspond to individual cells (left, right) or vertical lines (middle).
-Cell type annotation coloring is the same as in panel b.](figures/fig.pdf){#fig-model-results}
+::: {#fig-model-results}
+![](figures/fig.pdf)
 
-![Pyro-Velocity accurately predicts cell fate choices and lineage trajectory in human LARRY barcoded hematopoietic cells.
-Systematic benchmarking of cell fate prediction methods - predicting uni-lineage cell fates (a-b, Model 1), bifurcated lineages (c, Model 2), or all lineages together (d, Model 2).
-Panels from left to right show on the same SPRING embedding: 1.
-Cell types; 2.
-Clonal progression vector field derived based on the observed clonal barcodes (Methods); 3.
-scVelo recovered vector field; 4.
-Pyro-Velocity single cell level vector field uncertainty (angular standard deviation from 30 posterior samples); 5.
-Pyro-Velocity recovered vector field; vectors are colored by the averaged vector field uncertainty; 6.
-Clonal fate potency score using Cospar [@Wang2022-xb]; 7.
-scVelo recovered latent time; 8.
-Pyro-Velocity inference of posterior mean shared time per cell.
-Cosine similarity and Spearman’s correlation for the different methods with the clonal progression and fate potency are reported in the titles.](figures/fig.pdf){#fig-lineage-tracing}
+Pyro-Velocity outperforms other RNA-velocity methods and accurately predicts
+cell lineage trajectories. a,b. RNA Velocity model comparisons from fully mature
+peripheral blood mononuclear cells (PBMCs, t-SNE embedded plots, a) or E15.5
+mouse pancreas (UMAP embedded plots, b-c).  From left to right in panels a and
+b: 1. Cells colored by cell type; 2. Stream plot of velocity vector field
+defined by scVelo; 3. Stream plot of the posterior mean vector field from
+Pyro-Velocity; 4. Pyro-Velocity visualization of cell dynamics for 3 arbitrarily
+selected single cells; 5. Single-cell vector field uncertainty where each cell
+is colored based on the scaled angular standard deviation to visualize
+uncertainty based on n=30 iterated, posterior samples depicted as arrows noted
+throughout our analysis; 6. Averaged vector field uncertainty where each vector
+is colored by the averaged vector field uncertainty. The scaled angular standard
+deviation range is [0 360], where 0 corresponds to the highest confidence for a
+single direction, 360 corresponds to the highest uncertainty, and no
+directionality preference based on 30 posterior samples (4-6). Vectors were
+projected into the UMAP space using the scVelo API based on cosine
+similarity-derived transition matrix. c. UMAP embedded cells colored by the
+average cell shared time across 30 posterior samples. The Spearman’s correlation
+between Pyro-Velocity and Cytotrace is provided. d. Pyro-Velocity automatic gene
+selection and ranking for predicting important genes involved in pancreas
+development. e. Phase portraits of posterior samples average spliced and
+unspliced gene expression (left), rainbow plots showing denoised spliced gene
+expression (middle), and gene expression within UMAP embedded space.  Single
+dots correspond to individual cells (left, right) or vertical lines (middle).
+Cell type annotation coloring is the same as in panel b.
+:::
+
+::: {#fig-lineage-tracing}
+![](figures/fig.pdf)
+
+Pyro-Velocity accurately predicts cell fate choices and lineage trajectory in
+human LARRY barcoded hematopoietic cells. Systematic benchmarking of cell fate
+prediction methods - predicting uni-lineage cell fates (a-b, Model 1),
+bifurcated lineages (c, Model 2), or all lineages together (d, Model 2). Panels
+from left to right show on the same SPRING embedding: 1. Cell types; 2. Clonal
+progression vector field derived based on the observed clonal barcodes
+(Methods); 3. scVelo recovered vector field; 4. Pyro-Velocity single cell level
+vector field uncertainty (angular standard deviation from 30 posterior samples);
+5. Pyro-Velocity recovered vector field; vectors are colored by the averaged
+vector field uncertainty; 6. Clonal fate potency score using Cospar
+[@Wang2022-xb]; 7. scVelo recovered latent time; 8. Pyro-Velocity inference of
+posterior mean shared time per cell. Cosine similarity and Spearman’s
+correlation for the different methods with the clonal progression and fate
+potency are reported in the titles.
+:::
 
@@
 
-![Graphical representations detail the generative process of the Pyro-Velocity models and highlight their differences.
-a.
-Pyro-Velocity Model 1.
-Each rectangle represents one data dimension (e.g., gene or cell).
-The arrows represent the directed hierarchical dependency among random variables.
-The white circles correspond to the modeled random variables.
-The probability distributions of these variables are denoted on the right legend.
-The gray circles correspond to intermediate results computed from random variables.
-In this model, the kinetics random variables for the cell shared time and gene are modeled with two constraints: 1.
-the shared time is strictly larger than the gene-specific activation time $t_{0}^{(0)}$, 2.
-the gene expression in the activate state starts with the initial condition $(u_{0}=0,\,s_{0}=0)$, as in previous methods[@La_Manno2018-lj;@Bergen2020-pj] (Methods).
-b.
-Pyro-Velocity Model 2.
-This figure follows the same graphical convention as in a.
-In this model $t_{0}^{(0)}$ is independent from the shared time, thus allowing a time lag for the gene activation phase.
-In addition, a positive starting basal gene expression level $(u_{0},\,s_{0})$ different than $(u_{0}=0,\,s_{0}=0)$ is allowed.
-(Methods).
-c.
-Schematic illustrating how the constraints proposed in Model 1 (left) and Model 2 (right) are reflected in the modeling of individual gene expression.
-The x-axis corresponds to the shared time, the y-axis corresponds to the posterior mean of the spliced expression level, the vertical lines with arrows show the start and end time of the observed cell sampling, t0 represents the gene-specific activation time $t_{0}^{(0)}$, the red lines highlight the differences for the potential gene-specific time lag when $t_{0}^{(0)} > {\min}\!⁡(t)$, and the initial spliced gene expression.
-The same differences can also be illustrated for the unspliced gene expression profiles (not shown).](figures/fig.pdf){#suppfig-graphical-models}
+::: {#suppfig-graphical-models}
+![](figures/fig.pdf)
 
-![Extended version of @fig-model-results for evaluating Pyro-Velocity (Model 1) and scVelo on the PBMC and pancreas datasets.
-a.
-Left figure shows the scVelo vector field using the default user options, right figure shows the evaluation of the scVelo vector field using different combinations of user options within Epsilon cells.
-b.
-Pyro-Velocity additional single cell level uncertainty diagnostic plots on the fully mature PBMC dataset, from left to right: 1.
-standard deviation of posterior samples of shared time, 2.
-Rayleigh test of posterior samples of the vector field (the title reports false discovery rate after p-value correction for multiple tests across cells), 3.
-transcriptional state uncertainty of predicted raw read count per cell.
-The contour lines show the top 10% of the most uncertain cells.
-c.
-For the pancreas dataset, the figures show from left to right: UMAP rendering of 1.
-Cytotrace, 2.
-scVelo latent time, 3.
-Posterior mean of Pyro-Velocity shared time averaged across 30 posterior samples, 4.
-Top 50 markers overlapping between Pyro-Velocity and scVelo (Methods, Supplementary Table 1-2), 5.
-The same uncertainty diagnostic plots per cell as in b.
-d.
-The same plot as Fig.
-2d.
-e.
-Selection of the top negatively correlated genes with the cell shared time based on the posterior mean on denoised spliced expression and negative mean absolute error of gene expression prediction (Methods).
-f.
-Phase portraits, rainbow plots, and UMAP rendering of posterior mean of gene expression for the genes selected in e.](figures/fig.pdf){#suppfig-model-results}
+Graphical representations detail the generative process of the Pyro-Velocity
+models and highlight their differences. a. Pyro-Velocity Model 1. Each rectangle
+represents one data dimension (e.g., gene or cell). The arrows represent the
+directed hierarchical dependency among random variables. The white circles
+correspond to the modeled random variables. The probability distributions of
+these variables are denoted on the right legend. The gray circles correspond to
+intermediate results computed from random variables. In this model, the kinetics
+random variables for the cell shared time and gene are modeled with two
+constraints: 1. the shared time is strictly larger than the gene-specific
+activation time $t_{0}^{(0)}$, 2. the gene expression in the activate state
+starts with the initial condition $(u_{0}=0,\,s_{0}=0)$, as in previous
+methods[@La_Manno2018-lj;@Bergen2020-pj] (Methods). b. Pyro-Velocity Model 2.
+This figure follows the same graphical convention as in a. In this model
+$t_{0}^{(0)}$ is independent from the shared time, thus allowing a time lag for
+the gene activation phase. In addition, a positive starting basal gene
+expression level $(u_{0},\,s_{0})$ different than $(u_{0}=0,\,s_{0}=0)$ is
+allowed. (Methods). c. Schematic illustrating how the constraints proposed in
+Model 1 (left) and Model 2 (right) are reflected in the modeling of individual
+gene expression. The x-axis corresponds to the shared time, the y-axis
+corresponds to the posterior mean of the spliced expression level, the vertical
+lines with arrows show the start and end time of the observed cell sampling, t0
+represents the gene-specific activation time $t_{0}^{(0)}$, the red lines
+highlight the differences for the potential gene-specific time lag when
+$t_{0}^{(0)} > {\min}\!⁡(t)$, and the initial spliced gene expression. The same
+differences can also be illustrated for the unspliced gene expression profiles
+(not shown).
+:::
 
-![Extension of @fig-model-results, Pyro-Velocity (Model 2) outperforms scVelo on PBMC and pancreas datasets.
-Same format as in @fig-model-results and @suppfig-model-results](figures/fig.pdf){#suppfig-model-2-result}
+::: {#suppfig-model-results}
+![](figures/fig.pdf)
 
-![Extension of @fig-lineage-tracing, Pyro-velocity (Model 1 for uni-fate data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on scRNA-seq datasets with lineage barcoding information (LARRY dataset).
-a.
-SPRING embedding rendering of Cytotrace scores, the title reports the Spearman’s correlation between the Cytotrace and Cospar fate potency scores, b.
-Pyro-Velocity additional single cell level uncertainty diagnostic plots based on the multi-fate LARRY dataset, from left to right: 1.
-Standard deviation of posterior samples of shared time per cell, 2.
-Rayleigh test of posterior samples of vector field per cell (the title reports the false discovery rate after p-value correction for multiple tests across cells), 3.
-Transcriptional state uncertainty of predicted raw read count per cell.
-The contour lines show the top 10% of the most uncertain cells (Methods), c.
-Pairwise Spearman’s correlation between Cytotrace/Cospar fate potency scores and velocity methods for the four different benchmarking datasets selected based on the lineage barcoding information from LARRY data.
-d.
-SPRING embedding rendering of posterior mean of shared time across 30 posterior samples, the title shows the Spearman’s correlation of mean shared time with Cytotrace.
-e.
-Selection of the top positively correlated genes with the cell shared time based on the posterior mean of spliced expression.
-f.
-Phase portraits, rainbow plots, and UMAP rendering with splicing gene expression level for the selected genes in e.](figures/fig.pdf){#suppfig-lineage-tracing}
+Extended version of @fig-model-results for evaluating Pyro-Velocity (Model 1)
+and scVelo on the PBMC and pancreas datasets. a. Left figure shows the scVelo
+vector field using the default user options, right figure shows the evaluation
+of the scVelo vector field using different combinations of user options within
+Epsilon cells. b. Pyro-Velocity additional single cell level uncertainty
+diagnostic plots on the fully mature PBMC dataset, from left to right: 1.
+standard deviation of posterior samples of shared time, 2. Rayleigh test of
+posterior samples of the vector field (the title reports false discovery rate
+after p-value correction for multiple tests across cells), 3. transcriptional
+state uncertainty of predicted raw read count per cell. The contour lines show
+the top 10% of the most uncertain cells. c. For the pancreas dataset, the
+figures show from left to right: UMAP rendering of 1. Cytotrace, 2. scVelo
+latent time, 3. Posterior mean of Pyro-Velocity shared time averaged across 30
+posterior samples, 4. Top 50 markers overlapping between Pyro-Velocity and
+scVelo (Methods, Supplementary Table 1-2), 5. The same uncertainty diagnostic
+plots per cell as in b. d. The same plot as Fig. 2d. e. Selection of the top
+negatively correlated genes with the cell shared time based on the posterior
+mean on denoised spliced expression and negative mean absolute error of gene
+expression prediction (Methods). f. Phase portraits, rainbow plots, and UMAP
+rendering of posterior mean of gene expression for the genes selected in e.
+:::
 
-![Extension of @fig-lineage-tracing presenting an extended comparison of Pyro-Velocity and scVelo on scRNA-seq datasets with lineage barcoding information (LARRY dataset).
-a.
-uni-fate Monocyte lineage analysis with Pyro-Velocity Model 2, b.
-uni-fate Neutrophil lineage analysis with Pyro-Velocity Model 2, c.
-Neutrophil and Monocyte bi-fate lineage analysis with Pyro-Velocity Model 1, and d.
-multi-fate lineage analysis with Pyro-Velocity Model 1.
-The panel descriptions are the same as in @fig-lineage-tracing](figures/fig.pdf){#suppfig-lineage-tracing-2}
+::: {#suppfig-model-2-result}
+![](figures/fig.pdf)
+
+Extension of @fig-model-results, Pyro-Velocity (Model 2) outperforms scVelo on
+PBMC and pancreas datasets. Same format as in @fig-model-results and
+@suppfig-model-results
+:::
+
+::: {#suppfig-lineage-tracing}
+![](figures/fig.pdf)
+
+Extension of @fig-lineage-tracing, Pyro-velocity (Model 1 for uni-fate data,
+Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on scRNA-seq
+datasets with lineage barcoding information (LARRY dataset). a. SPRING embedding
+rendering of Cytotrace scores, the title reports the Spearman’s correlation
+between the Cytotrace and Cospar fate potency scores, b. Pyro-Velocity
+additional single cell level uncertainty diagnostic plots based on the
+multi-fate LARRY dataset, from left to right: 1. Standard deviation of posterior
+samples of shared time per cell, 2. Rayleigh test of posterior samples of vector
+field per cell (the title reports the false discovery rate after p-value
+correction for multiple tests across cells), 3. Transcriptional state
+uncertainty of predicted raw read count per cell. The contour lines show the top
+10% of the most uncertain cells (Methods), c. Pairwise Spearman’s correlation
+between Cytotrace/Cospar fate potency scores and velocity methods for the four
+different benchmarking datasets selected based on the lineage barcoding
+information from LARRY data. d. SPRING embedding rendering of posterior mean of
+shared time across 30 posterior samples, the title shows the Spearman’s
+correlation of mean shared time with Cytotrace. e. Selection of the top
+positively correlated genes with the cell shared time based on the posterior
+mean of spliced expression. f. Phase portraits, rainbow plots, and UMAP
+rendering with splicing gene expression level for the selected genes in e.
+:::
+
+::: {#suppfig-lineage-tracing-2}
+![](figures/fig.pdf)
+
+Extension of @fig-lineage-tracing presenting an extended comparison of
+Pyro-Velocity and scVelo on scRNA-seq datasets with lineage barcoding
+information (LARRY dataset). a. uni-fate Monocyte lineage analysis with
+Pyro-Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
+Pyro-Velocity Model 2, c. Neutrophil and Monocyte bi-fate lineage analysis with
+Pyro-Velocity Model 1, and d. multi-fate lineage analysis with Pyro-Velocity
+Model 1. The panel descriptions are the same as in @fig-lineage-tracing
+:::
 

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): factor format-specific settings
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
   Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data
-number-sections: true
-toc: true
-toc-title: Contents
 format:
@@
     journal:
-      cite-style: sn-nature
+      cite-style: sn-mathphys-num
     classoption: 
       - lineno
-    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
-    # to use the bst styles from the upstream template.
     cite-method: natbib
+    link-citations: true
+    toc: true
+    toc-title: Contents
+    number-sections: true
     keep-tex: true
@@
     include-in-header: header.tex
-    link-citations: true
   html:
@@
     reference-doc: styles.docx
-# filters:
-#   - authors-block
+    cite-method: citeproc
+    filters:
+      - authors-block
 csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): update introductory references
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
       - authors-block
-csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl
+csl: _extensions/sciexp/nature/csl/springer-mathphys-brackets.csl
 affiliations:
@@
 enable uncertainty estimation of velocity. Central to the Pyro-Velocity model is
-a shared latent time (t): a global latent variable placing each cell at a
+a *shared latent time* $(t)$: a global latent variable placing each cell at a
 particular instant of the modeled dynamical process. From this latent time, the
 expectation of spliced and unspliced RNA levels for each gene can be computed
-knowing gene-specific kinetics parameters (transcriptional (), splicing (), and
-degradation rates () and two gene-specific switching times (t0k), and
-gene-specific transcriptional state k) and integrating the RNA velocity
-differential equations analogously to what proposed in the dynamical RNA
-velocity model by Bergen et al.4,9. However, we innovate the structure and
-formulation of the latent time given that our model is set to jointly fit the
-entire raw data matrix, not individual genes. We approach the task of estimating
-the posterior distribution by restricting the search within a family of a
+knowing gene-specific kinetics parameters (transcriptional $(\alpha)$, splicing
+$(\beta)$, and degradation rates ($\gamma$) and two gene-specific switching
+times $(t_{0}^{k})$, and gene-specific transcriptional state $k$) and
+integrating the RNA velocity differential equations analogously to what proposed
+in the dynamical RNA velocity model by Bergen et al.
+[@Bergen2020-pj;@Li2021-qa]. However, we innovate the structure and formulation
+of the latent time given that our model is set to jointly fit the entire raw
+data matrix, not individual genes. We approach the task of estimating the
+posterior distribution by restricting the search within a family of a
 variational distribution approximating the true posterior, and taking advantage
 of the automatic differentiation variational inference (ADVI) capabilities of
-Pyro8,10 (Fig. 1a, Methods). Notably, our generative Bayesian model enables to
-sample from the posterior distributions of the latent and dependent variables,
-allowing uncertainty estimation of phase portraits, high dimensional velocity
-vector field, and shared latent time. Around this model, we built several
+Pyro [@Bingham2018-id;@Kucukelbir2016-bk] ([@fig-model]a,
+\nameref{sec-methods}). Notably, our generative Bayesian model enables to sample
+from the posterior distributions of the latent and dependent variables, allowing
+uncertainty estimation of phase portraits, high dimensional velocity vector
+field, and shared latent time. Around this model, we built several
 visualizations to assess uncertainty and functions to recover putative cell fate
 priming genes and to investigate how the denoised gene expression recapitulates
-known cell-type annotations and their developmental order (Fig. 1b, Methods).
-Although several RNA velocity methods11–15 have been recently proposed, and many
-of them have not been peer-reviewed, here we showcase the advantages and
-innovations of Pyro-Velocity by comparing it with the current state-of-the-art
-dynamical RNA velocity model proposed by Bergen et al.4, and implemented in
-scVelo.
+known cell-type annotations and their developmental order ([@fig-model]b,
+\nameref{sec-methods}). Although several RNA velocity methods
+[@Gayoso2024-fz;@Gorin2022-il;@Gu2022-rh;@Cui2022-em;@Gao2022-jq] have been
+recently proposed, and many of them have not been peer-reviewed, here we
+showcase the advantages and innovations of Pyro-Velocity by comparing it with
+the current state-of-the-art dynamical RNA velocity model proposed by Bergen et
+al. [@Bergen2020-pj], and implemented in scVelo.
 

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): use clearpage for page break
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(manuscript): specify per-format affiliations
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
   Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data
+toc: true
+toc-title: Contents
+number-sections: true
 format:
@@
     link-citations: true
-    toc: true
-    toc-title: Contents
-    number-sections: true
     keep-tex: true
@@
     include-in-header: header.tex
-  html:
-    code-fold: true
-    html-math-method: mathjax
-  docx:
-    reference-doc: styles.docx
-    cite-method: citeproc
-    filters:
-      - authors-block
-csl: _extensions/sciexp/nature/csl/springer-mathphys-brackets.csl
     affiliations:
@@
         postal-code: 02142
+  html:
+    code-fold: true
+    html-math-method: mathjax
+    affiliations:
+      - name: Molecular Pathology Unit, Massachusetts General Hospital Research Institute, Charlestown, MA
+        id: 1
+      - name: Massachusetts General Hospital Cancer Center, Harvard Medical School, Charlestown, MA
+        id: 2
+      - name: Center for Regenerative Medicine, Massachusetts General Hospital, Boston, MA
+        id: 3
+      - name: Harvard Stem Cell Institute, Harvard University, Cambridge, MA
+        id: 4
+      - name: Broad Institute of MIT and Harvard, Cambridge, MA
+        id: 5
+      - name: Laboratory of Neurodevelopmental Systems Biology, Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland
+        id: 6
+      - name: Basis Research Institute, Cambridge, MA
+        id: 7
+  docx:
+    reference-doc: styles.docx
+    cite-method: citeproc
+    filters:
+      - authors-block
+    affiliations:
+      - name: Molecular Pathology Unit, Massachusetts General Hospital Research Institute, Charlestown, MA
+        id: 1
+      - name: Massachusetts General Hospital Cancer Center, Harvard Medical School, Charlestown, MA
+        id: 2
+      - name: Center for Regenerative Medicine, Massachusetts General Hospital, Boston, MA
+        id: 3
+      - name: Harvard Stem Cell Institute, Harvard University, Cambridge, MA
+        id: 4
+      - name: Broad Institute of MIT and Harvard, Cambridge, MA
+        id: 5
+      - name: Laboratory of Neurodevelopmental Systems Biology, Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland
+        id: 6
+      - name: Basis Research Institute, Cambridge, MA
+        id: 7
+csl: _extensions/sciexp/nature/csl/springer-mathphys-brackets.csl
 authors:
@@
   fig-prefix: Fig.
+  sec-prefix: Sec.
   custom:
@@
 Pyro-Velocity recasts the velocity estimation problem into a latent variable
-posterior probability inference (\nameref{sec-methods}). The proposed model is
+posterior probability inference (@sec-methods). The proposed model is
 generative and fully Bayesian, with the different parameters considered as
@@
 Pyro [@Bingham2018-id;@Kucukelbir2016-bk] ([@fig-model]a,
-\nameref{sec-methods}). Notably, our generative Bayesian model enables to sample
+@sec-methods). Notably, our generative Bayesian model enables to sample
 from the posterior distributions of the latent and dependent variables, allowing
@@
 known cell-type annotations and their developmental order ([@fig-model]b,
-\nameref{sec-methods}). Although several RNA velocity methods
+@sec-methods). Although several RNA velocity methods
 [@Gayoso2024-fz;@Gorin2022-il;@Gu2022-rh;@Cui2022-em;@Gao2022-jq] have been
@@
 \FloatBarrier
+\clearpage
 {{< pagebreak >}}

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): note nameref option for LaTeX outputs
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(manuscript): fix suppfig references vs caption titles
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(manuscript): update text
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 toc-title: Contents
+# Use of `\nameref{sec-methods}` etc. to refer to sections by name.
+# is only compatible with LaTeX outputs
 number-sections: true
@@
       latex-env: suppfig
-      reference-prefix: Fig. S
-      space-before-numbering: false
+      reference-prefix: Supplementary Fig.
+      caption-prefix: Supplementary Figure
+      space-before-numbering: true
       latex-list-of-description: Supplementary Figure
@@
 al. [@Bergen2020-pj], and implemented in scVelo.
+
+One major biological hurdle for current formulations of RNA Velocity models has
+been miscalling cell trajectories in fully differentiated cell populations that
+lack dynamic cell state transitions. One typical example being human peripheral
+blood mononuclear cells (PBMCs) [@Zheng2017-bz]. In this dataset, 65,877 cells
+were analyzed using the 10x platform that comprised 11 fully differentiated,
+mature immune cell types. This dataset lacked stem and progenitor cells or other
+signatures of an undergoing dynamical differentiation; thus, no consistent
+velocity flow should be detected. We reanalyzed this dataset using scVelo and
+Pyro-Velocity using a common set of genes for a fair comparison. Using scVelo,
+we reproduced the original results showing incorrect cell trajectory assignments
+([@fig-model-results]a) [@Bergen2021-qz]. ScVelo uncovered an errant, strong
+linear trajectory across a subset of mature cell types [@Zheng2017-bz] that is
+exasperated by the inability to confidently assess lineage directionality or a
+baseline to assess significance. On the contrary, Pyro-Velocity failed to detect
+high-confidence trajectories in the mature blood cell states, consistent with
+what is known about the biology underlying these cells ([@fig-model-results]a,
+panel 2 vs. panel 3). Importantly, the Bayesian nature of Pyro-Velocity enabled
+the generation of uncertainty estimates that can be easily visualized as likely
+vectors for individual cells or averaged across cells on a constantly spaced
+grid on the cell embedding (*averaged vector field*).
+
+Pyro-Velocity was also able to define well-known developmental cell hierarchies
+in a scRNA-seq dataset of E15.5 mouse pancreas ([@fig-model-results]b-e,
+$n=3,696$ cells comprising $8$ cell types ([@fig-model-results]b, panel 1)),
+identifying cell trajectories originating from ductal progenitor cells and
+culminated in the production of mature Alpha, Beta, Delta, and Epsilon cells
+[@Bastidas-Ponce2019-lf]. Though the scVelo and Pyro-Velocity models have
+qualitatively comparable vector fields, the results generated by scVelo were
+greatly influenced by gene selection, neighboring cell number for gene
+expression smoothing, and cannot provide confidence assessment of lineage
+directionality ([@suppfig-model-results]a). In addition, Pyro-Velocity
+uncertainty visualizations uncovered that differentiated cells had a larger
+vector field uncertainty than stem and progenitor cells, as would be expected
+for terminally differentiated cell states where progression and lineage
+trajectories would not be anticipated. Pyro-Velocity also recovered a
+developmental order in close agreement with what was estimated by Cytotrace
+(Spearman’s $r=0.95, \; \text{p-value} < 10^{-10}$ between latent time and
+developmental order)[@Gulati2020-xq] - a state-of-the-art orthogonal method used
+to predict cell differentiation based on the number of expressed genes per cell.
 

### 2024-08-20 reproducibility/manuscript/v2.tex
docs(manuscript): add v2 tex file for diff
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-mathphys-num,
  lineno,
  twocolumn]{sn-jnl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

%%%% Standard Packages

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

%%%%

\raggedbottom
\usepackage{nameref}
\usepackage{placeins}
\allowdisplaybreaks
\makeatletter
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{plain}
\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{losuppfig}}{\newfloat{suppfig}{h}{losuppfig}[chapter]}
\floatname{suppfig}{Supplementary Figure}
\newcommand*\listofsuppfigs{\listof{suppfig}{List of Supplementary Figures}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Pyro -Velocity: Probabilistic RNA Velocity inference from single-cell data},
  pdfkeywords={RNA velocity, single-cell RNA sequencing, probabilistic
generative modeling, cell fate, developmental trajectory inference},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Pyro -Velocity: Probabilistic RNA Velocity inference from
single-cell data]{Pyro -Velocity: Probabilistic RNA Velocity inference
from single-cell data}

% author setup
\author[1,2,3,4,5]{\fnm{Qian} \sur{Qin}}\author[5,7]{\fnm{Eli} \sur{Bingham}}\author*[6]{\fnm{Gioele La} \sur{Manno}}\email{gioele.lamanno@epfl.ch}\author*[1,2,3,4]{\fnm{David M.} \sur{Langenau}}\email{dlangenau@mgh.harvard.edu}\author*[1,2,5]{\fnm{Luca} \sur{Pinello}}\email{lpinello@mgh.harvard.edu}
% affil setup
\affil[1]{\orgdiv{Molecular Pathology Unit}, \orgname{Massachusetts
General Hospital Research Institute, Charlestown, MA}}
\affil[2]{\orgdiv{Massachusetts General Hospital Cancer
Center}, \orgname{Harvard Medical School, Charlestown, MA}}
\affil[3]{\orgdiv{Center for Regenerative
Medicine}, \orgname{Massachusetts General Hospital, Boston, MA}}
\affil[4]{\orgdiv{Harvard Stem Cell Institute}, \orgname{Harvard
University, Cambridge, MA}}
\affil[5]{\orgdiv{Broad Institute of MIT and
Harvard}, \orgname{Cambridge, MA}}
\affil[6]{\orgdiv{Brain and Mind Institute, School of Life Sciences,
École Polytechnique Fédérale de Lausanne (EPFL)}, \orgname{Laboratory of
Neurodevelopmental Systems Biology, Lausanne, Switzerland}}
\affil[7]{\orgdiv{Basis Research Institute}, \orgname{Cambridge, MA}}

% abstract

\abstract{Single-cell RNA Velocity has dramatically advanced our ability
to model cellular differentiation and cell fate decisions. However,
current preprocessing choices and model assumptions often lead to errors
in assigning developmental trajectories. Here, we develop,
Pyro-Velocity, a Bayesian, generative, and multivariate RNA Velocity
model to estimate the uncertainty of cell future states. This approach
models raw sequencing counts with the synchronized cell time across all
expressed genes to provide quantifiable and improved information on cell
fate choices and developmental trajectory dynamics.}

% keywords
\keywords{RNA velocity,  single-cell RNA sequencing,  probabilistic
generative modeling,  cell fate,  developmental trajectory inference}

\begin{document}
\maketitle
\newpage{}

\renewcommand*\contentsname{Contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}

\newpage{}

\section{Main text}\label{sec-main}

RNA Velocity is a powerful computational framework to estimate the time
derivative of gene expression
\citep{La_Manno2018-lj, Svensson2018-vk, Qiu2022-dj, Bergen2020-pj}. The
framework has been used to study developmental cell lineage trajectories
over a broad spectrum of biological processes, such as haematopoiesis
\citep{Qiu2022-dj}, pancreas development \citep{Bastidas-Ponce2019-lf},
and neural progenitors differentiation \citep{Bergen2020-pj}. Yet,
recent state-of-the-art RNA Velocity approaches such as velocyto and
scVelo \citep{La_Manno2018-lj} have several limitations. First, they may
predict developmental trajectories that do not exist in nature, as it
has been shown in the analysis of fully mature blood cells
\citep{Zheng2017-bz, Bergen2021-qz}. Second, data preprocessing steps
(e.g., selection of top variable genes, dimensionality reduction) affect
velocity estimates that are difficult to characterize and might lead to
artifactual predictions. In addition, non-linear spliced and unspliced
imputation procedures could distort phase portrait geometry and
recovered trajectories (Fig.~\ref{fig-model}a). Third, most RNA velocity
algorithms perform estimates independently for each gene, and therefore
cannot exploit the natural interdependence of gene expression. Fourth, a
unified description of the temporal relation between cells can only be
assigned based on postprocessing steps that aggregate heuristically the
gene-wise fits into a shared latent time (Fig.~\ref{fig-model}a).
Finally, current RNA velocity frameworks only output point estimates of
the velocity vector field and do not provide uncertainty estimation.
Importantly, the absence of statistical confidence estimates increases
false discoveries as it leaves users without a way to evaluate the
robustness of the methods' outputs. To overcome these limitations, we
developed a probabilistic generative model, named \emph{Pyro -Velocity},
that estimates RNA dynamics and infers cell trajectories using Pyro
\citep{Bingham2018-id}, a robust and efficient probabilistic programming
language.

Pyro-Velocity recasts the velocity estimation problem into a latent
variable posterior probability inference (Sec.~\ref{sec-methods}). The
proposed model is generative and fully Bayesian, with the different
parameters considered as latent random variables. Each parameter is
assigned a prior distribution, and the model is then conditioned on the
raw data (both spliced and unspliced molecule raw counts) to estimate
posterior probability distributions and to enable uncertainty estimation
of velocity. Central to the Pyro-Velocity model is a \emph{shared latent
time} \((t)\): a global latent variable placing each cell at a
particular instant of the modeled dynamical process. From this latent
time, the expectation of spliced and unspliced RNA levels for each gene
can be computed knowing gene-specific kinetics parameters
(transcriptional \((\alpha)\), splicing \((\beta)\), and degradation
rates (\(\gamma\)) and two gene-specific switching times
\((t_{0}^{k})\), and gene-specific transcriptional state \(k\)) and
integrating the RNA velocity differential equations analogously to what
proposed in the dynamical RNA velocity model by Bergen et al.
\citep{Bergen2020-pj, Li2021-qa}. However, we innovate the structure and
formulation of the latent time given that our model is set to jointly
fit the entire raw data matrix, not individual genes. We approach the
task of estimating the posterior distribution by restricting the search
within a family of a variational distribution approximating the true
posterior, and taking advantage of the automatic differentiation
variational inference (ADVI) capabilities of Pyro
\citep{Bingham2018-id, Kucukelbir2016-bk} (Fig.~\ref{fig-model}a,
Sec.~\ref{sec-methods}). Notably, our generative Bayesian model enables
to sample from the posterior distributions of the latent and dependent
variables, allowing uncertainty estimation of phase portraits, high
dimensional velocity vector field, and shared latent time. Around this
model, we built several visualizations to assess uncertainty and
functions to recover putative cell fate priming genes and to investigate
how the denoised gene expression recapitulates known cell-type
annotations and their developmental order (Fig.~\ref{fig-model}b,
Sec.~\ref{sec-methods}). Although several RNA velocity methods
\citep{Gayoso2024-fz, Gorin2022-il, Gu2022-rh, Cui2022-em, Gao2022-jq}
have been recently proposed, and many of them have not been
peer-reviewed, here we showcase the advantages and innovations of
Pyro-Velocity by comparing it with the current state-of-the-art
dynamical RNA velocity model proposed by Bergen et al.
\citep{Bergen2020-pj}, and implemented in scVelo.

One major biological hurdle for current formulations of RNA Velocity
models has been miscalling cell trajectories in fully differentiated
cell populations that lack dynamic cell state transitions. One typical
example being human peripheral blood mononuclear cells (PBMCs)
\citep{Zheng2017-bz}. In this dataset, 65,877 cells were analyzed using
the 10x platform that comprised 11 fully differentiated, mature immune
cell types. This dataset lacked stem and progenitor cells or other
signatures of an undergoing dynamical differentiation; thus, no
consistent velocity flow should be detected. We reanalyzed this dataset
using scVelo and Pyro-Velocity using a common set of genes for a fair
comparison. Using scVelo, we reproduced the original results showing
incorrect cell trajectory assignments (Fig.~\ref{fig-model-results}a)
\citep{Bergen2021-qz}. ScVelo uncovered an errant, strong linear
trajectory across a subset of mature cell types \citep{Zheng2017-bz}
that is exasperated by the inability to confidently assess lineage
directionality or a baseline to assess significance. On the contrary,
Pyro-Velocity failed to detect high-confidence trajectories in the
mature blood cell states, consistent with what is known about the
biology underlying these cells (Fig.~\ref{fig-model-results}a, panel 2
vs.~panel 3). Importantly, the Bayesian nature of Pyro-Velocity enabled
the generation of uncertainty estimates that can be easily visualized as
likely vectors for individual cells or averaged across cells on a
constantly spaced grid on the cell embedding (\emph{averaged vector
field}).

Pyro-Velocity was also able to define well-known developmental cell
hierarchies in a scRNA-seq dataset of E15.5 mouse pancreas
(Fig.~\ref{fig-model-results}b-e, \(n=3,696\) cells comprising \(8\)
cell types (Fig.~\ref{fig-model-results}b, panel 1)), identifying cell
trajectories originating from ductal progenitor cells and culminated in
the production of mature Alpha, Beta, Delta, and Epsilon cells
\citep{Bastidas-Ponce2019-lf}. Though the scVelo and Pyro-Velocity
models have qualitatively comparable vector fields, the results
generated by scVelo were greatly influenced by gene selection,
neighboring cell number for gene expression smoothing, and cannot
provide confidence assessment of lineage directionality
(Supplementary Figure~\ref{suppfig-model-results}a). In addition,
Pyro-Velocity uncertainty visualizations uncovered that differentiated
cells had a larger vector field uncertainty than stem and progenitor
cells, as would be expected for terminally differentiated cell states
where progression and lineage trajectories would not be anticipated.
Pyro-Velocity also recovered a developmental order in close agreement
with what was estimated by Cytotrace (Spearman's
\(r=0.95, \; \text{p-value} < 10^{-10}\) between latent time and
developmental order)\citep{Gulati2020-xq} - a state-of-the-art
orthogonal method used to predict cell differentiation based on the
number of expressed genes per cell.

We next used Pyro-Velocity to unbiasedly uncover potential cell fate
determinant markers genes within the E15.5 mouse pancreas. To this end,
we first selected the top 300 genes with the best goodness of fit
(negative mean absolute errors across cells). Next, we ranked them based
on the positive (or negative) correlation between inferred shared time
and corresponding denoised spliced expression per cell
(Fig.~\ref{fig-model-results}d, Methods). This procedure successfully
recovered key developmental genes that are activated (or repressed)
during endocrinogenesis (Fig.~\ref{fig-model-results}d-e,
Supplementary Figure~\ref{suppfig-model-results}). Among the top
positively correlated genes identified by Pyro-Velocity, endocrine Beta
cells expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a key
regulator Beta cell insulin generation \citep{Millership2018-vd} and
Ppp1r1a which is a marker of Beta cells \citep{Jiang2013-pt}. The
ranking procedure proposed by scVelo failed to identify Ins1, Ins2,
Nnat, and Ppp1r1a as lineage-restricted genes in their top 50 ranked
gene list. The inability of scVelo to identify these key developmental
genes could be likely explained by the kNN smoothing of read count and
filtering steps (Supplementary Table 1, 2). Visual inspection of these
four top genes by Pyro-Velocity revealed a single activation phase
during cell type transitions based on phase portrait curves. We also
propose their visualization through ``rainbow plots'\,' to illustrate
denoised single gene spliced expression across shared time and to
investigate the inferred cell order based on known annotations. Further,
UMAP embedding of cells colored based on denoised gene expression
confirmed the cell fate markers' smooth and progressive activation (or
repression) towards endocrine cells (Fig.~\ref{fig-model-results}e,
Supplementary Figure~\ref{suppfig-model-results}).

Ground truth velocity is not a quantity we can easily assess. Therefore,
no robust and quantitative benchmarking of RNA velocity methods has been
proposed. However, recent single-cell assays allow recording lineage
relationships and can be used to assess the ability of different methods
in recapitulating cell fate progression. Therefore, we next
systematically compared scVelo and Pyro-Velocity ability to recover
single cell fates in Lineage And RNA RecoverY (LARRY) barcoded
hematopoietic cells. LARRY leverages unique lentiviral barcodes inserted
into 3' untranslated region (UTR) of GFP reporter that is integrated as
a single copy into parental cells. Then progeny can be clonally traced
to follow the descendent cell fates over time. LARRY has been recently
applied to in vitro differentiation of human blood and accurately
recovered expected cell fate and lineage trajectories
\citep{Weinreb2020-mv}. This dataset sampled differentiation over the
course of 2, 4, and 6 days and contained \(49,302\) cells that each
could be traced with at least one barcode. Because LARRY couples
single-cell transcriptomics and direct cell lineage tracing, this data
set provides a unique opportunity to benchmark RNA Velocity
quantitatively and to provide insights into the correct interpretation
of recovered velocity vector fields and latent cell times. To
quantitatively assess the agreement between observed and predicted
fates, we considered two metrics: 1. average cosine similarity (cs)
between vectors from the RNA Velocity vector field and the clonal
progression vector field, i.e., a vector field derived directly from the
lineage data (Sec.~\ref{sec-methods}), 2. Spearman's correlation
\((\rho)\) between cell latent (shared) time and fate potency scores
derived from Cytotrace and Cospar. Cospar, our gold standard, is a
state-of-the-art method designed explicitly for predicting fate potency
based on LARRY data, while Cytotrace leverages only transcriptomic data.
We selected cell state changes associated with monocyte and neutrophil
lineages because both lineages have the largest numbers of cells in the
scRNA sequencing data set and well-known cell fate trajectories
\citep{Petukhov2018-nr}. As expected, the Pyro-Velocity vector field
accurately predicted cell state trajectories from hematopoietic
progenitor cells into mature monocytes or neutrophils and had high
cosine similarity between the inferred vector field and clonal
progression vector field assigned by LARRY (monocytes \(cs=0.49\),
neutrophils \(cs=0.38\)). High Spearman's correlation was also seen
between shared latent time and Cospar fate potency (monocytes
\(\rho=0.60\), neutrophils \(\rho=0.74\)) or shared latent time with
Cytotrace fate potency (monocytes \(\rho=0.925\), neutrophils
\(\rho=0.925\)). By contrast, scVelo generated vector fields with
reversed order or low correlation (cosine similarity: \(-0.38\),
monocytes, cosine similarity: \(0.32\), neutrophils). Moreover, Cospar
fate potency scores were negative (\(\rho=-0.59\), monocytes) or
non-significantly correlated (\(\rho=-0.07\), neutrophils). A similar
low correlation was observed for scVelo and Cytotrace fate potency
scores (monocytes \(\rho=0.309\), neutrophils \(\rho=0.751\))
(Fig.~\ref{fig-lineage-tracing}a-b,
Supplementary Figure~\ref{suppfig-lineage-tracing}c).

\section{Methods}\label{sec-methods}

\subsection{Model formulation}\label{sec-model}

We assume the dynamical gene expression is determined by the RNA
splicing process, and infer the unspliced and spliced gene expression
level from the differential equations proposed in velocyto
\citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj}

\begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align}

\newpage{}

\onecolumn

\section{References}\label{references}

\renewcommand{\bibsection}{}
\bibliography{bibliography.bib}

\FloatBarrier

\newpage{}

\section{Figures}\label{figures}

\begin{figure}

\centering{

\includegraphics{figures/fig.pdf}

}

\caption{\label{fig-model}Pyro-Velocity is a fully generative Bayesian
method with uncertainty estimation of velocity vector fields and shared
latent time, based on raw counts and without ad-hoc preprocessing steps.
a. Schematic comparing current RNA Velocity and Pyro-Velocity workflows.
Both methods take in input spliced and unspliced read count tables and
output RNA velocity vector fields and latent (shared) time. Previous
methods require ad-hoc preprocessing steps, including the arbitrary
selection of top variable genes, number of principal components, and
kNN-pooling to impute both spliced and unspliced expression within the
PCA-reduced spliced expression space. In addition, manually tuned ad-hoc
postprocessing steps are used to compute the final vector field and
latent time based on aggregation and smoothing of the independent
velocity estimates for each gene which can dramatically change the
velocity estimates. In contrast, Pyro-Velocity is a fully generative and
probabilistic model that uses raw read counts. Our model requires
minimal preprocessing and postprocessing steps and outputs a common
shared time based on all the genes without additional aggregation steps.
It also provides uncertainty estimation based on posterior samples. b.
The main outputs of the Pyro-Velocity framework. Visualization of single
gene expression across inferred developmental cell order using rainbow
plots (right panel).}

\end{figure}%

\begin{figure}

\centering{

\includegraphics{figures/fig.pdf}

}

\caption{\label{fig-model-results}Pyro-Velocity outperforms other
RNA-velocity methods and accurately predicts cell lineage trajectories.
a,b. RNA Velocity model comparisons from fully mature peripheral blood
mononuclear cells (PBMCs, t-SNE embedded plots, a) or E15.5 mouse
pancreas (UMAP embedded plots, b-c). From left to right in panels a and
b: 1. Cells colored by cell type; 2. Stream plot of velocity vector
field defined by scVelo; 3. Stream plot of the posterior mean vector
field from Pyro-Velocity; 4. Pyro-Velocity visualization of cell
dynamics for 3 arbitrarily selected single cells; 5. Single-cell vector
field uncertainty where each cell is colored based on the scaled angular
standard deviation to visualize uncertainty based on n=30 iterated,
posterior samples depicted as arrows noted throughout our analysis; 6.
Averaged vector field uncertainty where each vector is colored by the
averaged vector field uncertainty. The scaled angular standard deviation
range is {[}0 360{]}, where 0 corresponds to the highest confidence for
a single direction, 360 corresponds to the highest uncertainty, and no
directionality preference based on 30 posterior samples (4-6). Vectors
were projected into the UMAP space using the scVelo API based on cosine
similarity-derived transition matrix. c.~UMAP embedded cells colored by
the average cell shared time across 30 posterior samples. The Spearman's
correlation between Pyro-Velocity and Cytotrace is provided.
d.~Pyro-Velocity automatic gene selection and ranking for predicting
important genes involved in pancreas development. e. Phase portraits of
posterior samples average spliced and unspliced gene expression (left),
rainbow plots showing denoised spliced gene expression (middle), and
gene expression within UMAP embedded space. Single dots correspond to
individual cells (left, right) or vertical lines (middle). Cell type
annotation coloring is the same as in panel b.}

\end{figure}%

\begin{figure}

\centering{

\includegraphics{figures/fig.pdf}

}

\caption{\label{fig-lineage-tracing}Pyro-Velocity accurately predicts
cell fate choices and lineage trajectory in human LARRY barcoded
hematopoietic cells. Systematic benchmarking of cell fate prediction
methods - predicting uni-lineage cell fates (a-b, Model 1), bifurcated
lineages (c, Model 2), or all lineages together (d, Model 2). Panels
from left to right show on the same SPRING embedding: 1. Cell types; 2.
Clonal progression vector field derived based on the observed clonal
barcodes (Methods); 3. scVelo recovered vector field; 4. Pyro-Velocity
single cell level vector field uncertainty (angular standard deviation
from 30 posterior samples); 5. Pyro-Velocity recovered vector field;
vectors are colored by the averaged vector field uncertainty; 6. Clonal
fate potency score using Cospar \citep{Wang2022-xb}; 7. scVelo recovered
latent time; 8. Pyro-Velocity inference of posterior mean shared time
per cell. Cosine similarity and Spearman's correlation for the different
methods with the clonal progression and fate potency are reported in the
titles.}

\end{figure}%

\FloatBarrier
\clearpage

\newpage{}

\section{Supplementary Figures}\label{supplementary-figures}

\begin{suppfig}

\centering{

\includegraphics{figures/fig.pdf}

}

\caption{\label{suppfig-graphical-models}Graphical representations
detail the generative process of the Pyro-Velocity models and highlight
their differences. a. Pyro-Velocity Model 1. Each rectangle represents
one data dimension (e.g., gene or cell). The arrows represent the
directed hierarchical dependency among random variables. The white
circles correspond to the modeled random variables. The probability
distributions of these variables are denoted on the right legend. The
gray circles correspond to intermediate results computed from random
variables. In this model, the kinetics random variables for the cell
shared time and gene are modeled with two constraints: 1. the shared
time is strictly larger than the gene-specific activation time
\(t_{0}^{(0)}\), 2. the gene expression in the activate state starts
with the initial condition \((u_{0}=0,\,s_{0}=0)\), as in previous
methods\citep{La_Manno2018-lj, Bergen2020-pj} (Methods). b.
Pyro-Velocity Model 2. This figure follows the same graphical convention
as in a. In this model \(t_{0}^{(0)}\) is independent from the shared
time, thus allowing a time lag for the gene activation phase. In
addition, a positive starting basal gene expression level
\((u_{0},\,s_{0})\) different than \((u_{0}=0,\,s_{0}=0)\) is allowed.
(Methods). c.~Schematic illustrating how the constraints proposed in
Model 1 (left) and Model 2 (right) are reflected in the modeling of
individual gene expression. The x-axis corresponds to the shared time,
the y-axis corresponds to the posterior mean of the spliced expression
level, the vertical lines with arrows show the start and end time of the
observed cell sampling, t0 represents the gene-specific activation time
\(t_{0}^{(0)}\), the red lines highlight the differences for the
potential gene-specific time lag when \(t_{0}^{(0)} > {\min}\!⁡(t)\), and
the initial spliced gene expression. The same differences can also be
illustrated for the unspliced gene expression profiles (not shown).}

\end{suppfig}%

\begin{suppfig}

\centering{

\includegraphics{figures/fig.pdf}

}

\caption{\label{suppfig-model-results}Extended version of
Fig.~\ref{fig-model-results} for evaluating Pyro-Velocity (Model 1) and
scVelo on the PBMC and pancreas datasets. a. Left figure shows the
scVelo vector field using the default user options, right figure shows
the evaluation of the scVelo vector field using different combinations
of user options within Epsilon cells. b. Pyro-Velocity additional single
cell level uncertainty diagnostic plots on the fully mature PBMC
dataset, from left to right: 1. standard deviation of posterior samples
of shared time, 2. Rayleigh test of posterior samples of the vector
field (the title reports false discovery rate after p-value correction
for multiple tests across cells), 3. transcriptional state uncertainty
of predicted raw read count per cell. The contour lines show the top
10\% of the most uncertain cells. c.~For the pancreas dataset, the
figures show from left to right: UMAP rendering of 1. Cytotrace, 2.
scVelo latent time, 3. Posterior mean of Pyro-Velocity shared time
averaged across 30 posterior samples, 4. Top 50 markers overlapping
between Pyro-Velocity and scVelo (Methods, Supplementary Table 1-2), 5.
The same uncertainty diagnostic plots per cell as in b. d.~The same plot
as Fig. 2d. e. Selection of the top negatively correlated genes with the
cell shared time based on the posterior mean on denoised spliced
expression and negative mean absolute error of gene expression
prediction (Methods). f.~Phase portraits, rainbow plots, and UMAP
rendering of posterior mean of gene expression for the genes selected in
e.}

\end{suppfig}%

\begin{suppfig}

\centering{

\includegraphics{figures/fig.pdf}

}

\caption{\label{suppfig-model-2-result}Extension of
Fig.~\ref{fig-model-results}, Pyro-Velocity (Model 2) outperforms scVelo
on PBMC and pancreas datasets. Same format as in
Fig.~\ref{fig-model-results} and
Supplementary Figure~\ref{suppfig-model-results}}

\end{suppfig}%

\begin{suppfig}

\centering{

\includegraphics{figures/fig.pdf}

}

\caption{\label{suppfig-lineage-tracing}Extension of
Fig.~\ref{fig-lineage-tracing}, Pyro-velocity (Model 1 for uni-fate
data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
scRNA-seq datasets with lineage barcoding information (LARRY dataset).
a. SPRING embedding rendering of Cytotrace scores, the title reports the
Spearman's correlation between the Cytotrace and Cospar fate potency
scores, b. Pyro-Velocity additional single cell level uncertainty
diagnostic plots based on the multi-fate LARRY dataset, from left to
right: 1. Standard deviation of posterior samples of shared time per
cell, 2. Rayleigh test of posterior samples of vector field per cell
(the title reports the false discovery rate after p-value correction for
multiple tests across cells), 3. Transcriptional state uncertainty of
predicted raw read count per cell. The contour lines show the top 10\%
of the most uncertain cells (Methods), c.~Pairwise Spearman's
correlation between Cytotrace/Cospar fate potency scores and velocity
methods for the four different benchmarking datasets selected based on
the lineage barcoding information from LARRY data. d.~SPRING embedding
rendering of posterior mean of shared time across 30 posterior samples,
the title shows the Spearman's correlation of mean shared time with
Cytotrace. e. Selection of the top positively correlated genes with the
cell shared time based on the posterior mean of spliced expression.
f.~Phase portraits, rainbow plots, and UMAP rendering with splicing gene
expression level for the selected genes in e.}

\end{suppfig}%

\begin{suppfig}

\centering{

\includegraphics{figures/fig.pdf}

}

\caption{\label{suppfig-lineage-tracing-2}Extension of
Fig.~\ref{fig-lineage-tracing} presenting an extended comparison of
Pyro-Velocity and scVelo on scRNA-seq datasets with lineage barcoding
information (LARRY dataset). a. uni-fate Monocyte lineage analysis with
Pyro-Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
Pyro-Velocity Model 2, c.~Neutrophil and Monocyte bi-fate lineage
analysis with Pyro-Velocity Model 1, and d.~multi-fate lineage analysis
with Pyro-Velocity Model 1. The panel descriptions are the same as in
Fig.~\ref{fig-lineage-tracing}}

\end{suppfig}%

\begin{suppfig}

\centering{

\includegraphics{figures/fig.pdf}

}

\caption{\label{suppfig-plot}An example figure (an empty plot)}

\end{suppfig}%

\FloatBarrier

\newpage{}

\renewcommand\refname{Supporting Information}

\end{document}

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): set twocolumn
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(manuscript): lint name
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2024-08-20 reproducibility/manuscript/manuscript.qmd
@@
       - lineno
+      - twocolumn
     cite-method: natbib
@@
   preprocessing choices and model assumptions often lead to errors in assigning
-  developmental trajectories. Here, we develop, Pyro-Velocity, a Bayesian,
+  developmental trajectories. Here, we develop, Pyro&thinsp;-Velocity, a Bayesian,
   generative, and multivariate RNA Velocity model to estimate the uncertainty of
@@
       latex-env: suppfig
-      reference-prefix: Fig. S
-      space-before-numbering: false
+      caption-prefix: Supplementary Figure
+      reference-prefix: Supplementary Fig.
+      space-before-numbering: true
       latex-list-of-description: Supplementary Figure
@@
 increases false discoveries as it leaves users without a way to evaluate the
-robustness of the methods’ outputs. To overcome these limitations, we developed
+robustness of the methods' outputs. To overcome these limitations, we developed
 a probabilistic generative model, named *Pyro&thinsp;-Velocity*, that estimates
@@
 
-Pyro-Velocity recasts the velocity estimation problem into a latent variable
-posterior probability inference (@sec-methods). The proposed model is
+Pyro&thinsp;-Velocity recasts the velocity estimation problem into a latent
+variable posterior probability inference (@sec-methods). The proposed model is
 generative and fully Bayesian, with the different parameters considered as
@@
 molecule raw counts) to estimate posterior probability distributions and to
-enable uncertainty estimation of velocity. Central to the Pyro-Velocity model is
-a *shared latent time* $(t)$: a global latent variable placing each cell at a
-particular instant of the modeled dynamical process. From this latent time, the
-expectation of spliced and unspliced RNA levels for each gene can be computed
-knowing gene-specific kinetics parameters (transcriptional $(\alpha)$, splicing
-$(\beta)$, and degradation rates ($\gamma$) and two gene-specific switching
-times $(t_{0}^{k})$, and gene-specific transcriptional state $k$) and
+enable uncertainty estimation of velocity. Central to the Pyro&thinsp;-Velocity
+model is a *shared latent time* $(t)$: a global latent variable placing each
+cell at a particular instant of the modeled dynamical process. From this latent
+time, the expectation of spliced and unspliced RNA levels for each gene can be
+computed knowing gene-specific kinetics parameters (transcriptional $(\alpha)$,
+splicing $(\beta)$, and degradation rates ($\gamma$) and two gene-specific
+switching times $(t_{0}^{k})$, and gene-specific transcriptional state $k$) and
 integrating the RNA velocity differential equations analogously to what proposed
@@
 of the automatic differentiation variational inference (ADVI) capabilities of
-Pyro [@Bingham2018-id;@Kucukelbir2016-bk] ([@fig-model]a,
-@sec-methods). Notably, our generative Bayesian model enables to sample
-from the posterior distributions of the latent and dependent variables, allowing
-uncertainty estimation of phase portraits, high dimensional velocity vector
-field, and shared latent time. Around this model, we built several
-visualizations to assess uncertainty and functions to recover putative cell fate
-priming genes and to investigate how the denoised gene expression recapitulates
-known cell-type annotations and their developmental order ([@fig-model]b,
-@sec-methods). Although several RNA velocity methods
+Pyro [@Bingham2018-id;@Kucukelbir2016-bk] ([@fig-model]a, @sec-methods).
+Notably, our generative Bayesian model enables to sample from the posterior
+distributions of the latent and dependent variables, allowing uncertainty
+estimation of phase portraits, high dimensional velocity vector field, and
+shared latent time. Around this model, we built several visualizations to assess
+uncertainty and functions to recover putative cell fate priming genes and to
+investigate how the denoised gene expression recapitulates known cell-type
+annotations and their developmental order ([@fig-model]b, @sec-methods).
+Although several RNA velocity methods
 [@Gayoso2024-fz;@Gorin2022-il;@Gu2022-rh;@Cui2022-em;@Gao2022-jq] have been
 recently proposed, and many of them have not been peer-reviewed, here we
-showcase the advantages and innovations of Pyro-Velocity by comparing it with
-the current state-of-the-art dynamical RNA velocity model proposed by Bergen et
-al. [@Bergen2020-pj], and implemented in scVelo.
+showcase the advantages and innovations of Pyro&thinsp;-Velocity by comparing it
+with the current state-of-the-art dynamical RNA velocity model proposed by
+Bergen et al. [@Bergen2020-pj], and implemented in scVelo.
+
+One major biological hurdle for current formulations of RNA Velocity models has
+been miscalling cell trajectories in fully differentiated cell populations that
+lack dynamic cell state transitions. One typical example being human peripheral
+blood mononuclear cells (PBMCs) [@Zheng2017-bz]. In this dataset, 65,877 cells
+were analyzed using the 10x platform that comprised 11 fully differentiated,
+mature immune cell types. This dataset lacked stem and progenitor cells or other
+signatures of an undergoing dynamical differentiation; thus, no consistent
+velocity flow should be detected. We reanalyzed this dataset using scVelo and
+Pyro&thinsp;-Velocity using a common set of genes for a fair comparison. Using
+scVelo, we reproduced the original results showing incorrect cell trajectory
+assignments ([@fig-model-results]a) [@Bergen2021-qz]. ScVelo uncovered an
+errant, strong linear trajectory across a subset of mature cell types
+[@Zheng2017-bz] that is exasperated by the inability to confidently assess
+lineage directionality or a baseline to assess significance. On the contrary,
+Pyro&thinsp;-Velocity failed to detect high-confidence trajectories in the
+mature blood cell states, consistent with what is known about the biology
+underlying these cells ([@fig-model-results]a, panel 2 vs. panel 3).
+Importantly, the Bayesian nature of Pyro&thinsp;-Velocity enabled the generation
+of uncertainty estimates that can be easily visualized as likely vectors for
+individual cells or averaged across cells on a constantly spaced grid on the
+cell embedding (*averaged vector field*).
+
+Pyro&thinsp;-Velocity was also able to define well-known developmental cell
+hierarchies in a scRNA-seq dataset of E15.5 mouse pancreas
+([@fig-model-results]b-e, $n=3,696$ cells comprising $8$ cell types
+([@fig-model-results]b, panel 1)), identifying cell trajectories originating
+from ductal progenitor cells and culminated in the production of mature Alpha,
+Beta, Delta, and Epsilon cells [@Bastidas-Ponce2019-lf]. Though the scVelo and
+Pyro&thinsp;-Velocity models have qualitatively comparable vector fields, the
+results generated by scVelo were greatly influenced by gene selection,
+neighboring cell number for gene expression smoothing, and cannot provide
+confidence assessment of lineage directionality ([@suppfig-model-results]a). In
+addition, Pyro&thinsp;-Velocity uncertainty visualizations uncovered that
+differentiated cells had a larger vector field uncertainty than stem and
+progenitor cells, as would be expected for terminally differentiated cell states
+where progression and lineage trajectories would not be anticipated.
+Pyro&thinsp;-Velocity also recovered a developmental order in close agreement
+with what was estimated by Cytotrace (Spearman's $r=0.95, \; \text{p-value} <
+10^{-10}$ between latent time and developmental order)[@Gulati2020-xq] - a
+state-of-the-art orthogonal method used to predict cell differentiation based on
+the number of expressed genes per cell.
+
+We next used Pyro&thinsp;-Velocity to unbiasedly uncover potential cell fate
+determinant markers genes within the E15.5 mouse pancreas. To this end, we first
+selected the top 300 genes with the best goodness of fit (negative mean absolute
+errors across cells). Next, we ranked them based on the positive (or negative)
+correlation between inferred shared time and corresponding denoised spliced
+expression per cell ([@fig-model-results]d, Methods). This procedure
+successfully recovered key developmental genes that are activated (or repressed)
+during endocrinogenesis ([@fig-model-results]d-e, @suppfig-model-results). Among
+the top positively correlated genes identified by Pyro&thinsp;-Velocity,
+endocrine Beta cells expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a
+key regulator Beta cell insulin generation [@Millership2018-vd] and Ppp1r1a
+which is a marker of Beta cells [@Jiang2013-pt]. The ranking procedure proposed
+by scVelo failed to identify Ins1, Ins2, Nnat, and Ppp1r1a as lineage-restricted
+genes in their top 50 ranked gene list. The inability of scVelo to identify
+these key developmental genes could be likely explained by the kNN smoothing of
+read count and filtering steps (Supplementary Table 1, 2). Visual inspection of
+these four top genes by Pyro&thinsp;-Velocity revealed a single activation phase
+during cell type transitions based on phase portrait curves. We also propose
+their visualization through ``rainbow plots'' to illustrate denoised single gene
+spliced expression across shared time and to investigate the inferred cell order
+based on known annotations. Further, UMAP embedding of cells colored based on
+denoised gene expression confirmed the cell fate markers' smooth and progressive
+activation (or repression) towards endocrine cells ([@fig-model-results]e,
+[@suppfig-model-results]).
+
+Ground truth velocity is not a quantity we can easily assess. Therefore, no
+robust and quantitative benchmarking of RNA velocity methods has been proposed.
+However, recent single-cell assays allow recording lineage relationships and can
+be used to assess the ability of different methods in recapitulating cell fate
+progression. Therefore, we next systematically compared scVelo and
+Pyro&thinsp;-Velocity ability to recover single cell fates in Lineage And RNA
+RecoverY (LARRY) barcoded hematopoietic cells. LARRY leverages unique lentiviral
+barcodes inserted into 3' untranslated region (UTR) of GFP reporter that is
+integrated as a single copy into parental cells. Then progeny can be clonally
+traced to follow the descendent cell fates over time. LARRY has been recently
+applied to in vitro differentiation of human blood and accurately recovered
+expected cell fate and lineage trajectories [@Weinreb2020-mv]. This dataset
+sampled differentiation over the course of 2, 4, and 6 days and contained
+$49,302$ cells that each could be traced with at least one barcode. Because
+LARRY couples single-cell transcriptomics and direct cell lineage tracing, this
+data set provides a unique opportunity to benchmark RNA Velocity quantitatively
+and to provide insights into the correct interpretation of recovered velocity
+vector fields and latent cell times. To quantitatively assess the agreement
+between observed and predicted fates, we considered two metrics: 1. average
+cosine similarity (cs) between vectors from the RNA Velocity vector field and
+the clonal progression vector field, i.e., a vector field derived directly from
+the lineage data ([@sec-methods]), 2. Spearman's correlation $(\rho)$ between
+cell latent (shared) time and fate potency scores derived from Cytotrace and
+Cospar. Cospar, our gold standard, is a state-of-the-art method designed
+explicitly for predicting fate potency based on LARRY data, while Cytotrace
+leverages only transcriptomic data. We selected cell state changes associated
+with monocyte and neutrophil lineages because both lineages have the largest
+numbers of cells in the scRNA sequencing data set and well-known cell fate
+trajectories [@Petukhov2018-nr]. As expected, the Pyro&thinsp;-Velocity vector
+field accurately predicted cell state trajectories from hematopoietic progenitor
+cells into mature monocytes or neutrophils and had high cosine similarity
+between the inferred vector field and clonal progression vector field assigned
+by LARRY (monocytes $cs=0.49$, neutrophils $cs=0.38$).  High Spearman's
+correlation was also seen between shared latent time and Cospar fate potency
+(monocytes $\rho=0.60$, neutrophils $\rho=0.74$) or shared latent time with
+Cytotrace fate potency (monocytes $\rho=0.925$, neutrophils $\rho=0.925$). By
+contrast, scVelo generated vector fields with reversed order or low correlation
+(cosine similarity: $-0.38$, monocytes, cosine similarity: $0.32$, neutrophils).
+Moreover, Cospar fate potency scores were negative ($\rho=-0.59$, monocytes) or
+non-significantly correlated ($\rho=-0.07$, neutrophils).  A similar low
+correlation was observed for scVelo and Cytotrace fate potency scores (monocytes
+$\rho=0.309$, neutrophils $\rho=0.751$) ([@fig-lineage-tracing]a-b,
+[@suppfig-lineage-tracing]c).
 
@@
 
+\onecolumn
+
 # References
@@
 
-Pyro-Velocity is a fully generative Bayesian method with uncertainty estimation
+Pyro&thinsp;-Velocity is a fully generative Bayesian method with uncertainty estimation
 of velocity vector fields and shared latent time, based on raw counts and
 without ad-hoc preprocessing steps. a. Schematic comparing current RNA Velocity
-and Pyro-Velocity workflows. Both methods take in input spliced and unspliced
+and Pyro&thinsp;-Velocity workflows. Both methods take in input spliced and unspliced
 read count tables and output RNA velocity vector fields and latent (shared)
@@
 each gene which can dramatically change the velocity estimates. In contrast,
-Pyro-Velocity is a fully generative and probabilistic model that uses raw read
+Pyro&thinsp;-Velocity is a fully generative and probabilistic model that uses raw read
 counts. Our model requires minimal preprocessing and postprocessing steps and
@@
 aggregation steps. It also provides uncertainty estimation based on posterior
-samples. b. The main outputs of the Pyro-Velocity framework. Visualization of
+samples. b. The main outputs of the Pyro&thinsp;-Velocity framework. Visualization of
 single gene expression across inferred developmental cell order using rainbow
@@
 
-Pyro-Velocity outperforms other RNA-velocity methods and accurately predicts
+Pyro&thinsp;-Velocity outperforms other RNA-velocity methods and accurately predicts
 cell lineage trajectories. a,b. RNA Velocity model comparisons from fully mature
@@
 defined by scVelo; 3. Stream plot of the posterior mean vector field from
-Pyro-Velocity; 4. Pyro-Velocity visualization of cell dynamics for 3 arbitrarily
+Pyro&thinsp;-Velocity; 4. Pyro&thinsp;-Velocity visualization of cell dynamics for 3 arbitrarily
 selected single cells; 5. Single-cell vector field uncertainty where each cell
@@
 similarity-derived transition matrix. c. UMAP embedded cells colored by the
-average cell shared time across 30 posterior samples. The Spearman’s correlation
-between Pyro-Velocity and Cytotrace is provided. d. Pyro-Velocity automatic gene
+average cell shared time across 30 posterior samples. The Spearman's correlation
+between Pyro&thinsp;-Velocity and Cytotrace is provided. d. Pyro&thinsp;-Velocity automatic gene
 selection and ranking for predicting important genes involved in pancreas
@@
 
-Pyro-Velocity accurately predicts cell fate choices and lineage trajectory in
+Pyro&thinsp;-Velocity accurately predicts cell fate choices and lineage trajectory in
 human LARRY barcoded hematopoietic cells. Systematic benchmarking of cell fate
@@
 progression vector field derived based on the observed clonal barcodes
-(Methods); 3. scVelo recovered vector field; 4. Pyro-Velocity single cell level
+(Methods); 3. scVelo recovered vector field; 4. Pyro&thinsp;-Velocity single cell level
 vector field uncertainty (angular standard deviation from 30 posterior samples);
-5. Pyro-Velocity recovered vector field; vectors are colored by the averaged
+5. Pyro&thinsp;-Velocity recovered vector field; vectors are colored by the averaged
 vector field uncertainty; 6. Clonal fate potency score using Cospar
-[@Wang2022-xb]; 7. scVelo recovered latent time; 8. Pyro-Velocity inference of
-posterior mean shared time per cell. Cosine similarity and Spearman’s
+[@Wang2022-xb]; 7. scVelo recovered latent time; 8. Pyro&thinsp;-Velocity inference of
+posterior mean shared time per cell. Cosine similarity and Spearman's
 correlation for the different methods with the clonal progression and fate
@@
 
-Graphical representations detail the generative process of the Pyro-Velocity
-models and highlight their differences. a. Pyro-Velocity Model 1. Each rectangle
+Graphical representations detail the generative process of the Pyro&thinsp;-Velocity
+models and highlight their differences. a. Pyro&thinsp;-Velocity Model 1. Each rectangle
 represents one data dimension (e.g., gene or cell). The arrows represent the
@@
 starts with the initial condition $(u_{0}=0,\,s_{0}=0)$, as in previous
-methods[@La_Manno2018-lj;@Bergen2020-pj] (Methods). b. Pyro-Velocity Model 2.
+methods[@La_Manno2018-lj;@Bergen2020-pj] (Methods). b. Pyro&thinsp;-Velocity Model 2.
 This figure follows the same graphical convention as in a. In this model
@@
 
-Extended version of @fig-model-results for evaluating Pyro-Velocity (Model 1)
+Extended version of @fig-model-results for evaluating Pyro&thinsp;-Velocity (Model 1)
 and scVelo on the PBMC and pancreas datasets. a. Left figure shows the scVelo
@@
 of the scVelo vector field using different combinations of user options within
-Epsilon cells. b. Pyro-Velocity additional single cell level uncertainty
+Epsilon cells. b. Pyro&thinsp;-Velocity additional single cell level uncertainty
 diagnostic plots on the fully mature PBMC dataset, from left to right: 1.
@@
 figures show from left to right: UMAP rendering of 1. Cytotrace, 2. scVelo
-latent time, 3. Posterior mean of Pyro-Velocity shared time averaged across 30
-posterior samples, 4. Top 50 markers overlapping between Pyro-Velocity and
+latent time, 3. Posterior mean of Pyro&thinsp;-Velocity shared time averaged across 30
+posterior samples, 4. Top 50 markers overlapping between Pyro&thinsp;-Velocity and
 scVelo (Methods, Supplementary Table 1-2), 5. The same uncertainty diagnostic
@@
 
-Extension of @fig-model-results, Pyro-Velocity (Model 2) outperforms scVelo on
+Extension of @fig-model-results, Pyro&thinsp;-Velocity (Model 2) outperforms scVelo on
 PBMC and pancreas datasets. Same format as in @fig-model-results and
@@
 
-Extension of @fig-lineage-tracing, Pyro-velocity (Model 1 for uni-fate data,
+Extension of @fig-lineage-tracing, Pyro&thinsp;-Velocity (Model 1 for uni-fate data,
 Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on scRNA-seq
 datasets with lineage barcoding information (LARRY dataset). a. SPRING embedding
-rendering of Cytotrace scores, the title reports the Spearman’s correlation
-between the Cytotrace and Cospar fate potency scores, b. Pyro-Velocity
+rendering of Cytotrace scores, the title reports the Spearman's correlation
+between the Cytotrace and Cospar fate potency scores, b. Pyro&thinsp;-Velocity
 additional single cell level uncertainty diagnostic plots based on the
@@
 uncertainty of predicted raw read count per cell. The contour lines show the top
-10% of the most uncertain cells (Methods), c. Pairwise Spearman’s correlation
+10% of the most uncertain cells (Methods), c. Pairwise Spearman's correlation
 between Cytotrace/Cospar fate potency scores and velocity methods for the four
@@
 information from LARRY data. d. SPRING embedding rendering of posterior mean of
-shared time across 30 posterior samples, the title shows the Spearman’s
+shared time across 30 posterior samples, the title shows the Spearman's
 correlation of mean shared time with Cytotrace. e. Selection of the top
@@
 Extension of @fig-lineage-tracing presenting an extended comparison of
-Pyro-Velocity and scVelo on scRNA-seq datasets with lineage barcoding
+Pyro&thinsp;-Velocity and scVelo on scRNA-seq datasets with lineage barcoding
 information (LARRY dataset). a. uni-fate Monocyte lineage analysis with
-Pyro-Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
-Pyro-Velocity Model 2, c. Neutrophil and Monocyte bi-fate lineage analysis with
-Pyro-Velocity Model 1, and d. multi-fate lineage analysis with Pyro-Velocity
+Pyro&thinsp;-Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
+Pyro&thinsp;-Velocity Model 2, c. Neutrophil and Monocyte bi-fate lineage analysis with
+Pyro&thinsp;-Velocity Model 1, and d. multi-fate lineage analysis with Pyro&thinsp;-Velocity
 Model 1. The panel descriptions are the same as in @fig-lineage-tracing

### 2024-08-20 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 in assigning developmental trajectories. Here, we develop,
-Pyro-Velocity, a Bayesian, generative, and multivariate RNA Velocity
+Pyro -Velocity, a Bayesian, generative, and multivariate RNA Velocity
 model to estimate the uncertainty of cell future states. This approach
@@
 
-Pyro-Velocity recasts the velocity estimation problem into a latent
+Pyro -Velocity recasts the velocity estimation problem into a latent
 variable posterior probability inference (Sec.~\ref{sec-methods}). The
@@
 posterior probability distributions and to enable uncertainty estimation
-of velocity. Central to the Pyro-Velocity model is a \emph{shared latent
-time} \((t)\): a global latent variable placing each cell at a
+of velocity. Central to the Pyro -Velocity model is a \emph{shared
+latent time} \((t)\): a global latent variable placing each cell at a
 particular instant of the modeled dynamical process. From this latent
@@
 peer-reviewed, here we showcase the advantages and innovations of
-Pyro-Velocity by comparing it with the current state-of-the-art
+Pyro -Velocity by comparing it with the current state-of-the-art
 dynamical RNA velocity model proposed by Bergen et al.
@@
 consistent velocity flow should be detected. We reanalyzed this dataset
-using scVelo and Pyro-Velocity using a common set of genes for a fair
+using scVelo and Pyro -Velocity using a common set of genes for a fair
 comparison. Using scVelo, we reproduced the original results showing
@@
 directionality or a baseline to assess significance. On the contrary,
-Pyro-Velocity failed to detect high-confidence trajectories in the
+Pyro -Velocity failed to detect high-confidence trajectories in the
 mature blood cell states, consistent with what is known about the
 biology underlying these cells (Fig.~\ref{fig-model-results}a, panel 2
-vs.~panel 3). Importantly, the Bayesian nature of Pyro-Velocity enabled
+vs.~panel 3). Importantly, the Bayesian nature of Pyro -Velocity enabled
 the generation of uncertainty estimates that can be easily visualized as
@@
 
-Pyro-Velocity was also able to define well-known developmental cell
+Pyro -Velocity was also able to define well-known developmental cell
 hierarchies in a scRNA-seq dataset of E15.5 mouse pancreas
@@
 the production of mature Alpha, Beta, Delta, and Epsilon cells
-\citep{Bastidas-Ponce2019-lf}. Though the scVelo and Pyro-Velocity
+\citep{Bastidas-Ponce2019-lf}. Though the scVelo and Pyro -Velocity
 models have qualitatively comparable vector fields, the results
@@
 (Supplementary Figure~\ref{suppfig-model-results}a). In addition,
-Pyro-Velocity uncertainty visualizations uncovered that differentiated
+Pyro -Velocity uncertainty visualizations uncovered that differentiated
 cells had a larger vector field uncertainty than stem and progenitor
@@
 where progression and lineage trajectories would not be anticipated.
-Pyro-Velocity also recovered a developmental order in close agreement
+Pyro -Velocity also recovered a developmental order in close agreement
 with what was estimated by Cytotrace (Spearman's
-\(r=0.95, \; \text{p-value} < 10^{-10}\) between latent time and
-developmental order)\citep{Gulati2020-xq} - a state-of-the-art
-orthogonal method used to predict cell differentiation based on the
-number of expressed genes per cell.
+\(r=0.95, \; \text{p-value} <
+10^{-10}\) between latent time and developmental
+order)\citep{Gulati2020-xq} - a state-of-the-art orthogonal method used
+to predict cell differentiation based on the number of expressed genes
+per cell.
 
-We next used Pyro-Velocity to unbiasedly uncover potential cell fate
+We next used Pyro -Velocity to unbiasedly uncover potential cell fate
 determinant markers genes within the E15.5 mouse pancreas. To this end,
@@
 Supplementary Figure~\ref{suppfig-model-results}). Among the top
-positively correlated genes identified by Pyro-Velocity, endocrine Beta
+positively correlated genes identified by Pyro -Velocity, endocrine Beta
 cells expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a key
@@
 filtering steps (Supplementary Table 1, 2). Visual inspection of these
-four top genes by Pyro-Velocity revealed a single activation phase
+four top genes by Pyro -Velocity revealed a single activation phase
 during cell type transitions based on phase portrait curves. We also
@@
 in recapitulating cell fate progression. Therefore, we next
-systematically compared scVelo and Pyro-Velocity ability to recover
+systematically compared scVelo and Pyro -Velocity ability to recover
 single cell fates in Lineage And RNA RecoverY (LARRY) barcoded
@@
 scRNA sequencing data set and well-known cell fate trajectories
-\citep{Petukhov2018-nr}. As expected, the Pyro-Velocity vector field
+\citep{Petukhov2018-nr}. As expected, the Pyro -Velocity vector field
 accurately predicted cell state trajectories from hematopoietic
@@
 Supplementary Figure~\ref{suppfig-lineage-tracing}c).
+
+Next, we evaluated the ability of the Pyro -Velocity models to recover
+bi-fate or multi-fate trajectories by considering clones belonging to
+monocyte and neutrophil lineages or using the entire dataset. Not
+surprisingly, both methods showed their limitations, given that they are
+not designed to model and capture multiple fate trajectories
+(Supplementary Figure~\ref{suppfig-lineage-tracing-2}c-d). The recovered
+vector fields and cell latent (shared) time show inconsistent directions
+for both methods. However, Pyro -Velocity identified higher uncertainty
+for some of the problematic regions. Furthermore, we discovered that
+vector field and shared time estimation could be significantly improved
+by extending the original kinetic model by allowing each gene a time lag
+for the transcriptional activation and a baseline expression level
+potentially different than 0 (Pyro -Velocity Model 2,
+Sec.~\ref{sec-methods}) (Fig.~\ref{fig-lineage-tracing}c,d,
+Supplementary Figure~\ref{suppfig-graphical-models}b,c). This extended
+model recapitulated a vector field consistent with the clonal
+progression vector field for most branches with some spurious
+predictions within the intermediate cell state between stem cells and
+Megakaryocytes and most terminal Neutrophils (bi-fate cosine similarity:
+\(0.62\), multi-fate cosine similarity: \(0.63\)). However, the proposed
+uncertainty estimation accurately flagged these subpopulations to warn
+users of potential false discoveries. Notably, this extended model
+significantly improved the correlation of cell shared time with the
+clonal fate potency (bi-fate \(\rho=0.65\), multi-fate \(\rho=0.74\))
+and also improved the correlation with Cytotrace fate potency (bi-fate
+\(\rho=0.931\), multi-fate \(\rho=0.614\)). Remarkably, the shared time
+per cell from Pyro -Velocity also outperformed Cytotrace for
+developmental order prediction
+(Supplementary Figure~\ref{suppfig-lineage-tracing}a,c). Importantly,
+this extended model retained its ability to accurately identify cell
+lineage trajectories and shared time for all the other datasets analyzed
+above (Supplementary Figure~\ref{suppfig-model-2-result}). Taken
+together, our approach using LARRY barcoded datasets established the
+first quantitative benchmark of RNA velocity-based cell fate trajectory
+inference and provides an exciting new framework for continued
+optimization of multi-fate predictions in the future.
 
@@
 
-\caption{\label{fig-model}Pyro-Velocity is a fully generative Bayesian
+\caption{\label{fig-model}Pyro -Velocity is a fully generative Bayesian
 method with uncertainty estimation of velocity vector fields and shared
 latent time, based on raw counts and without ad-hoc preprocessing steps.
-a. Schematic comparing current RNA Velocity and Pyro-Velocity workflows.
-Both methods take in input spliced and unspliced read count tables and
-output RNA velocity vector fields and latent (shared) time. Previous
-methods require ad-hoc preprocessing steps, including the arbitrary
-selection of top variable genes, number of principal components, and
-kNN-pooling to impute both spliced and unspliced expression within the
-PCA-reduced spliced expression space. In addition, manually tuned ad-hoc
-postprocessing steps are used to compute the final vector field and
-latent time based on aggregation and smoothing of the independent
-velocity estimates for each gene which can dramatically change the
-velocity estimates. In contrast, Pyro-Velocity is a fully generative and
-probabilistic model that uses raw read counts. Our model requires
-minimal preprocessing and postprocessing steps and outputs a common
-shared time based on all the genes without additional aggregation steps.
-It also provides uncertainty estimation based on posterior samples. b.
-The main outputs of the Pyro-Velocity framework. Visualization of single
-gene expression across inferred developmental cell order using rainbow
-plots (right panel).}
+a. Schematic comparing current RNA Velocity and Pyro -Velocity
+workflows. Both methods take in input spliced and unspliced read count
+tables and output RNA velocity vector fields and latent (shared) time.
+Previous methods require ad-hoc preprocessing steps, including the
+arbitrary selection of top variable genes, number of principal
+components, and kNN-pooling to impute both spliced and unspliced
+expression within the PCA-reduced spliced expression space. In addition,
+manually tuned ad-hoc postprocessing steps are used to compute the final
+vector field and latent time based on aggregation and smoothing of the
+independent velocity estimates for each gene which can dramatically
+change the velocity estimates. In contrast, Pyro -Velocity is a fully
+generative and probabilistic model that uses raw read counts. Our model
+requires minimal preprocessing and postprocessing steps and outputs a
+common shared time based on all the genes without additional aggregation
+steps. It also provides uncertainty estimation based on posterior
+samples. b. The main outputs of the Pyro -Velocity framework.
+Visualization of single gene expression across inferred developmental
+cell order using rainbow plots (right panel).}
 
@@
 
-\caption{\label{fig-model-results}Pyro-Velocity outperforms other
+\caption{\label{fig-model-results}Pyro -Velocity outperforms other
 RNA-velocity methods and accurately predicts cell lineage trajectories.
@@
 field defined by scVelo; 3. Stream plot of the posterior mean vector
-field from Pyro-Velocity; 4. Pyro-Velocity visualization of cell
+field from Pyro -Velocity; 4. Pyro -Velocity visualization of cell
 dynamics for 3 arbitrarily selected single cells; 5. Single-cell vector
@@
 the average cell shared time across 30 posterior samples. The Spearman's
-correlation between Pyro-Velocity and Cytotrace is provided.
-d.~Pyro-Velocity automatic gene selection and ranking for predicting
+correlation between Pyro -Velocity and Cytotrace is provided.
+d.~Pyro -Velocity automatic gene selection and ranking for predicting
 important genes involved in pancreas development. e. Phase portraits of
@@
 
-\caption{\label{fig-lineage-tracing}Pyro-Velocity accurately predicts
+\caption{\label{fig-lineage-tracing}Pyro -Velocity accurately predicts
 cell fate choices and lineage trajectory in human LARRY barcoded
@@
 Clonal progression vector field derived based on the observed clonal
-barcodes (Methods); 3. scVelo recovered vector field; 4. Pyro-Velocity
+barcodes (Methods); 3. scVelo recovered vector field; 4. Pyro -Velocity
 single cell level vector field uncertainty (angular standard deviation
-from 30 posterior samples); 5. Pyro-Velocity recovered vector field;
+from 30 posterior samples); 5. Pyro -Velocity recovered vector field;
 vectors are colored by the averaged vector field uncertainty; 6. Clonal
 fate potency score using Cospar \citep{Wang2022-xb}; 7. scVelo recovered
-latent time; 8. Pyro-Velocity inference of posterior mean shared time
+latent time; 8. Pyro -Velocity inference of posterior mean shared time
 per cell. Cosine similarity and Spearman's correlation for the different
@@
 \caption{\label{suppfig-graphical-models}Graphical representations
-detail the generative process of the Pyro-Velocity models and highlight
-their differences. a. Pyro-Velocity Model 1. Each rectangle represents
+detail the generative process of the Pyro -Velocity models and highlight
+their differences. a. Pyro -Velocity Model 1. Each rectangle represents
 one data dimension (e.g., gene or cell). The arrows represent the
@@
 methods\citep{La_Manno2018-lj, Bergen2020-pj} (Methods). b.
-Pyro-Velocity Model 2. This figure follows the same graphical convention
-as in a. In this model \(t_{0}^{(0)}\) is independent from the shared
-time, thus allowing a time lag for the gene activation phase. In
-addition, a positive starting basal gene expression level
+Pyro -Velocity Model 2. This figure follows the same graphical
+convention as in a. In this model \(t_{0}^{(0)}\) is independent from
+the shared time, thus allowing a time lag for the gene activation phase.
+In addition, a positive starting basal gene expression level
 \((u_{0},\,s_{0})\) different than \((u_{0}=0,\,s_{0}=0)\) is allowed.
@@
 \caption{\label{suppfig-model-results}Extended version of
-Fig.~\ref{fig-model-results} for evaluating Pyro-Velocity (Model 1) and
+Fig.~\ref{fig-model-results} for evaluating Pyro -Velocity (Model 1) and
 scVelo on the PBMC and pancreas datasets. a. Left figure shows the
@@
 the evaluation of the scVelo vector field using different combinations
-of user options within Epsilon cells. b. Pyro-Velocity additional single
-cell level uncertainty diagnostic plots on the fully mature PBMC
+of user options within Epsilon cells. b. Pyro -Velocity additional
+single cell level uncertainty diagnostic plots on the fully mature PBMC
 dataset, from left to right: 1. standard deviation of posterior samples
@@
 figures show from left to right: UMAP rendering of 1. Cytotrace, 2.
-scVelo latent time, 3. Posterior mean of Pyro-Velocity shared time
+scVelo latent time, 3. Posterior mean of Pyro -Velocity shared time
 averaged across 30 posterior samples, 4. Top 50 markers overlapping
-between Pyro-Velocity and scVelo (Methods, Supplementary Table 1-2), 5.
+between Pyro -Velocity and scVelo (Methods, Supplementary Table 1-2), 5.
 The same uncertainty diagnostic plots per cell as in b. d.~The same plot
@@
 \caption{\label{suppfig-model-2-result}Extension of
-Fig.~\ref{fig-model-results}, Pyro-Velocity (Model 2) outperforms scVelo
-on PBMC and pancreas datasets. Same format as in
+Fig.~\ref{fig-model-results}, Pyro -Velocity (Model 2) outperforms
+scVelo on PBMC and pancreas datasets. Same format as in
 Fig.~\ref{fig-model-results} and
@@
 \caption{\label{suppfig-lineage-tracing}Extension of
-Fig.~\ref{fig-lineage-tracing}, Pyro-velocity (Model 1 for uni-fate
+Fig.~\ref{fig-lineage-tracing}, Pyro -Velocity (Model 1 for uni-fate
 data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
@@
 Spearman's correlation between the Cytotrace and Cospar fate potency
-scores, b. Pyro-Velocity additional single cell level uncertainty
+scores, b. Pyro -Velocity additional single cell level uncertainty
 diagnostic plots based on the multi-fate LARRY dataset, from left to
@@
 Fig.~\ref{fig-lineage-tracing} presenting an extended comparison of
-Pyro-Velocity and scVelo on scRNA-seq datasets with lineage barcoding
+Pyro -Velocity and scVelo on scRNA-seq datasets with lineage barcoding
 information (LARRY dataset). a. uni-fate Monocyte lineage analysis with
-Pyro-Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
-Pyro-Velocity Model 2, c.~Neutrophil and Monocyte bi-fate lineage
-analysis with Pyro-Velocity Model 1, and d.~multi-fate lineage analysis
-with Pyro-Velocity Model 1. The panel descriptions are the same as in
+Pyro -Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
+Pyro -Velocity Model 2, c.~Neutrophil and Monocyte bi-fate lineage
+analysis with Pyro -Velocity Model 1, and d.~multi-fate lineage analysis
+with Pyro -Velocity Model 1. The panel descriptions are the same as in
 Fig.~\ref{fig-lineage-tracing}}

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): update text
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+Next, we evaluated the ability of the Pyro&thinsp;-Velocity models to recover
+bi-fate or multi-fate trajectories by considering clones belonging to monocyte
+and neutrophil lineages or using the entire dataset. Not surprisingly, both
+methods showed their limitations, given that they are not designed to model and
+capture multiple fate trajectories ([@suppfig-lineage-tracing-2]c-d). The
+recovered vector fields and cell latent (shared) time show inconsistent
+directions for both methods. However, Pyro&thinsp;-Velocity identified higher
+uncertainty for some of the problematic regions. Furthermore, we discovered that
+vector field and shared time estimation could be significantly improved by
+extending the original kinetic model by allowing each gene a time lag for the
+transcriptional activation and a baseline expression level potentially different
+than 0 (Pyro&thinsp;-Velocity Model 2, [@sec-methods])
+([@fig-lineage-tracing]c,d, [@suppfig-graphical-models]b,c). This extended model
+recapitulated a vector field consistent with the clonal progression vector field
+for most branches with some spurious predictions within the intermediate cell
+state between stem cells and Megakaryocytes and most terminal Neutrophils
+(bi-fate cosine similarity: $0.62$, multi-fate cosine similarity: $0.63$).
+However, the proposed uncertainty estimation accurately flagged these
+subpopulations to warn users of potential false discoveries. Notably, this
+extended model significantly improved the correlation of cell shared time with
+the clonal fate potency (bi-fate $\rho=0.65$, multi-fate $\rho=0.74$) and also
+improved the correlation with Cytotrace fate potency (bi-fate $\rho=0.931$,
+multi-fate $\rho=0.614$). Remarkably, the shared time per cell from
+Pyro&thinsp;-Velocity also outperformed Cytotrace for developmental order
+prediction ([@suppfig-lineage-tracing]a,c). Importantly, this extended model
+retained its ability to accurately identify cell lineage trajectories and shared
+time for all the other datasets analyzed above ([@suppfig-model-2-result]).
+Taken together, our approach using LARRY barcoded datasets established the first
+quantitative benchmark of RNA velocity-based cell fate trajectory inference and
+provides an exciting new framework for continued optimization of multi-fate
+predictions in the future.
+
 # Methods {#sec-methods}

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): update text
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+In conclusion, Pyro-Velocity provides the first probabilistic, interpretable,
+end-to-end Bayesian inference framework for RNA Velocity that learns cell
+dynamics and recovers cell fate from multimodal raw read counts.  This approach
+provides uncertainty estimations of cell fate trajectories and novel
+visualization approaches to mitigate false discoveries. Pyro-Velocity recovered
+shared time and vector fields that are recapitulated in true developmental order
+and outperformed state-of-the-art methods like Cytotrace and scVelo
+([@suppfig-lineage-tracing]a).
+
+# Software availability
+
+We developed our Pyro-Velocity models with Pyro (version 1.6.0) probabilistic
+programming language and PyTorch (version 1.8.1). Pyro-Velocity source code is
+available at <https://github.com/pinellolab/pyrovelocity>. The version of
+Pyro-Velocity (v0.1.0) used to generate the results presented in this manuscript
+has been deposited on Zenodo: <https://zenodo.org/record/7072876#.Yx-2e-zMLxg>.
+
+# Data availability
+
+All the data presented in this manuscript were collected from previously
+published studies. The fully mature PBMC, pancreas datasets are downloaded and
+loaded through the scVelo using `scvelo.datasets.pancreas()`, and
+`scvelo.datasets.pbmc68k()`. The LARRY inDrop sequencing data was processed by
+the LARRY authors using dropEst [@Petukhov2018-nr] and velocyto. This dataset
+can be accessed at
+<https://figshare.com/articles/dataset/larry_invitro_adata_sub_raw_h5ad/20780344>.
+
+# Acknowledgments
+
+This project has been made possible in part by grant number 2022-249212 from
+the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community
+Foundation. L.P. received support from the U.S. NIH (R35 HG010717) and Silicon
+Valley Community Foundation (2022-249212). D.M.L is supported by R01CA226926,
+R01 CA215118, R01CA154923, R24 OD031955, CureSearch Acceleration Initiative, and
+the MGH Scholars Award. We thank the Pinello and Langenau labs members for
+helpful discussions.
+
+# Author contributions
+
+L.P., Q.Q., and E.B. conceived the work. Q.Q. and E.B created the software. Q.Q.
+analyzed the data, and prepared the figures and tables. L.P and D.L. supervised
+the project.  L.P., D.L., E.B., G.L.M helped in designing experiments and
+analyses.  All the authors contributed to the writing of the manuscript.
+
 # Methods {#sec-methods}

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): lint
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+# Competing financial interests statement
+
+L.P. has financial interests in Edilytics, Inc., Excelsior Genomics, and SeQure
+Dx, Inc. L.P.’s interests were reviewed and are managed by Massachusetts General
+Hospital and Partners HealthCare in accordance with their conflict of interest
+policies.
+
 # Methods {#sec-methods}
@@
 
-Pyro&thinsp;-Velocity is a fully generative Bayesian method with uncertainty estimation
-of velocity vector fields and shared latent time, based on raw counts and
-without ad-hoc preprocessing steps. a. Schematic comparing current RNA Velocity
-and Pyro&thinsp;-Velocity workflows. Both methods take in input spliced and unspliced
-read count tables and output RNA velocity vector fields and latent (shared)
-time. Previous methods require ad-hoc preprocessing steps, including the
-arbitrary selection of top variable genes, number of principal components, and
-kNN-pooling to impute both spliced and unspliced expression within the
+Pyro&thinsp;-Velocity is a fully generative Bayesian method with uncertainty
+estimation of velocity vector fields and shared latent time, based on raw counts
+and without ad-hoc preprocessing steps. a. Schematic comparing current RNA
+Velocity and Pyro&thinsp;-Velocity workflows. Both methods take in input spliced
+and unspliced read count tables and output RNA velocity vector fields and latent
+(shared) time. Previous methods require ad-hoc preprocessing steps, including
+the arbitrary selection of top variable genes, number of principal components,
+and kNN-pooling to impute both spliced and unspliced expression within the
 PCA-reduced spliced expression space. In addition, manually tuned ad-hoc
@@
 each gene which can dramatically change the velocity estimates. In contrast,
-Pyro&thinsp;-Velocity is a fully generative and probabilistic model that uses raw read
-counts. Our model requires minimal preprocessing and postprocessing steps and
-outputs a common shared time based on all the genes without additional
+Pyro&thinsp;-Velocity is a fully generative and probabilistic model that uses
+raw read counts. Our model requires minimal preprocessing and postprocessing
+steps and outputs a common shared time based on all the genes without additional
 aggregation steps. It also provides uncertainty estimation based on posterior
-samples. b. The main outputs of the Pyro&thinsp;-Velocity framework. Visualization of
-single gene expression across inferred developmental cell order using rainbow
-plots (right panel).
+samples. b. The main outputs of the Pyro&thinsp;-Velocity framework.
+Visualization of single gene expression across inferred developmental cell order
+using rainbow plots (right panel).
 :::
@@
 
-Pyro&thinsp;-Velocity outperforms other RNA-velocity methods and accurately predicts
-cell lineage trajectories. a,b. RNA Velocity model comparisons from fully mature
-peripheral blood mononuclear cells (PBMCs, t-SNE embedded plots, a) or E15.5
-mouse pancreas (UMAP embedded plots, b-c).  From left to right in panels a and
-b: 1. Cells colored by cell type; 2. Stream plot of velocity vector field
-defined by scVelo; 3. Stream plot of the posterior mean vector field from
-Pyro&thinsp;-Velocity; 4. Pyro&thinsp;-Velocity visualization of cell dynamics for 3 arbitrarily
-selected single cells; 5. Single-cell vector field uncertainty where each cell
-is colored based on the scaled angular standard deviation to visualize
-uncertainty based on n=30 iterated, posterior samples depicted as arrows noted
-throughout our analysis; 6. Averaged vector field uncertainty where each vector
-is colored by the averaged vector field uncertainty. The scaled angular standard
-deviation range is [0 360], where 0 corresponds to the highest confidence for a
-single direction, 360 corresponds to the highest uncertainty, and no
-directionality preference based on 30 posterior samples (4-6). Vectors were
-projected into the UMAP space using the scVelo API based on cosine
+Pyro&thinsp;-Velocity outperforms other RNA-velocity methods and accurately
+predicts cell lineage trajectories. a,b. RNA Velocity model comparisons from
+fully mature peripheral blood mononuclear cells (PBMCs, t-SNE embedded plots, a)
+or E15.5 mouse pancreas (UMAP embedded plots, b-c).  From left to right in
+panels a and b: 1. Cells colored by cell type; 2. Stream plot of velocity vector
+field defined by scVelo; 3. Stream plot of the posterior mean vector field from
+Pyro&thinsp;-Velocity; 4. Pyro&thinsp;-Velocity visualization of cell dynamics
+for 3 arbitrarily selected single cells; 5. Single-cell vector field uncertainty
+where each cell is colored based on the scaled angular standard deviation to
+visualize uncertainty based on n=30 iterated, posterior samples depicted as
+arrows noted throughout our analysis; 6. Averaged vector field uncertainty where
+each vector is colored by the averaged vector field uncertainty. The scaled
+angular standard deviation range is [0 360], where 0 corresponds to the highest
+confidence for a single direction, 360 corresponds to the highest uncertainty,
+and no directionality preference based on 30 posterior samples (4-6). Vectors
+were projected into the UMAP space using the scVelo API based on cosine
 similarity-derived transition matrix. c. UMAP embedded cells colored by the
 average cell shared time across 30 posterior samples. The Spearman's correlation
-between Pyro&thinsp;-Velocity and Cytotrace is provided. d. Pyro&thinsp;-Velocity automatic gene
-selection and ranking for predicting important genes involved in pancreas
-development. e. Phase portraits of posterior samples average spliced and
-unspliced gene expression (left), rainbow plots showing denoised spliced gene
-expression (middle), and gene expression within UMAP embedded space.  Single
-dots correspond to individual cells (left, right) or vertical lines (middle).
-Cell type annotation coloring is the same as in panel b.
+between Pyro&thinsp;-Velocity and Cytotrace is provided. d.
+Pyro&thinsp;-Velocity automatic gene selection and ranking for predicting
+important genes involved in pancreas development. e. Phase portraits of
+posterior samples average spliced and unspliced gene expression (left), rainbow
+plots showing denoised spliced gene expression (middle), and gene expression
+within UMAP embedded space.  Single dots correspond to individual cells (left,
+right) or vertical lines (middle). Cell type annotation coloring is the same as
+in panel b.
 :::
@@
 
-Pyro&thinsp;-Velocity accurately predicts cell fate choices and lineage trajectory in
-human LARRY barcoded hematopoietic cells. Systematic benchmarking of cell fate
-prediction methods - predicting uni-lineage cell fates (a-b, Model 1),
-bifurcated lineages (c, Model 2), or all lineages together (d, Model 2). Panels
-from left to right show on the same SPRING embedding: 1. Cell types; 2. Clonal
-progression vector field derived based on the observed clonal barcodes
-(Methods); 3. scVelo recovered vector field; 4. Pyro&thinsp;-Velocity single cell level
-vector field uncertainty (angular standard deviation from 30 posterior samples);
-5. Pyro&thinsp;-Velocity recovered vector field; vectors are colored by the averaged
-vector field uncertainty; 6. Clonal fate potency score using Cospar
-[@Wang2022-xb]; 7. scVelo recovered latent time; 8. Pyro&thinsp;-Velocity inference of
-posterior mean shared time per cell. Cosine similarity and Spearman's
-correlation for the different methods with the clonal progression and fate
-potency are reported in the titles.
+Pyro&thinsp;-Velocity accurately predicts cell fate choices and lineage
+trajectory in human LARRY barcoded hematopoietic cells. Systematic benchmarking
+of cell fate prediction methods - predicting uni-lineage cell fates (a-b, Model
+1), bifurcated lineages (c, Model 2), or all lineages together (d, Model 2).
+Panels from left to right show on the same SPRING embedding: 1. Cell types; 2.
+Clonal progression vector field derived based on the observed clonal barcodes
+(Methods); 3. scVelo recovered vector field; 4. Pyro&thinsp;-Velocity single
+cell level vector field uncertainty (angular standard deviation from 30
+posterior samples);
+5. Pyro&thinsp;-Velocity recovered vector field; vectors are colored by the
+averaged vector field uncertainty; 6. Clonal fate potency score using Cospar
+[@Wang2022-xb]; 7. scVelo recovered latent time; 8. Pyro&thinsp;-Velocity
+inference of posterior mean shared time per cell. Cosine similarity and
+Spearman's correlation for the different methods with the clonal progression and
+fate potency are reported in the titles.
 :::
@@
 
-Graphical representations detail the generative process of the Pyro&thinsp;-Velocity
-models and highlight their differences. a. Pyro&thinsp;-Velocity Model 1. Each rectangle
-represents one data dimension (e.g., gene or cell). The arrows represent the
-directed hierarchical dependency among random variables. The white circles
-correspond to the modeled random variables. The probability distributions of
-these variables are denoted on the right legend. The gray circles correspond to
-intermediate results computed from random variables. In this model, the kinetics
-random variables for the cell shared time and gene are modeled with two
-constraints: 1. the shared time is strictly larger than the gene-specific
-activation time $t_{0}^{(0)}$, 2. the gene expression in the activate state
-starts with the initial condition $(u_{0}=0,\,s_{0}=0)$, as in previous
-methods[@La_Manno2018-lj;@Bergen2020-pj] (Methods). b. Pyro&thinsp;-Velocity Model 2.
-This figure follows the same graphical convention as in a. In this model
-$t_{0}^{(0)}$ is independent from the shared time, thus allowing a time lag for
-the gene activation phase. In addition, a positive starting basal gene
-expression level $(u_{0},\,s_{0})$ different than $(u_{0}=0,\,s_{0}=0)$ is
-allowed. (Methods). c. Schematic illustrating how the constraints proposed in
-Model 1 (left) and Model 2 (right) are reflected in the modeling of individual
-gene expression. The x-axis corresponds to the shared time, the y-axis
-corresponds to the posterior mean of the spliced expression level, the vertical
-lines with arrows show the start and end time of the observed cell sampling, t0
-represents the gene-specific activation time $t_{0}^{(0)}$, the red lines
-highlight the differences for the potential gene-specific time lag when
-$t_{0}^{(0)} > {\min}\!⁡(t)$, and the initial spliced gene expression. The same
-differences can also be illustrated for the unspliced gene expression profiles
-(not shown).
+Graphical representations detail the generative process of the
+Pyro&thinsp;-Velocity models and highlight their differences. a.
+Pyro&thinsp;-Velocity Model 1. Each rectangle represents one data dimension
+(e.g., gene or cell). The arrows represent the directed hierarchical dependency
+among random variables. The white circles correspond to the modeled random
+variables. The probability distributions of these variables are denoted on the
+right legend. The gray circles correspond to intermediate results computed from
+random variables. In this model, the kinetics random variables for the cell
+shared time and gene are modeled with two constraints: 1. the shared time is
+strictly larger than the gene-specific activation time $t_{0}^{(0)}$, 2. the
+gene expression in the activate state starts with the initial condition
+$(u_{0}=0,\,s_{0}=0)$, as in previous methods[@La_Manno2018-lj;@Bergen2020-pj]
+(Methods). b. Pyro&thinsp;-Velocity Model 2. This figure follows the same
+graphical convention as in a. In this model $t_{0}^{(0)}$ is independent from
+the shared time, thus allowing a time lag for the gene activation phase. In
+addition, a positive starting basal gene expression level $(u_{0},\,s_{0})$
+different than $(u_{0}=0,\,s_{0}=0)$ is allowed. (Methods). c. Schematic
+illustrating how the constraints proposed in Model 1 (left) and Model 2 (right)
+are reflected in the modeling of individual gene expression. The x-axis
+corresponds to the shared time, the y-axis corresponds to the posterior mean of
+the spliced expression level, the vertical lines with arrows show the start and
+end time of the observed cell sampling, t0 represents the gene-specific
+activation time $t_{0}^{(0)}$, the red lines highlight the differences for the
+potential gene-specific time lag when $t_{0}^{(0)} > {\min}\!⁡(t)$, and the
+initial spliced gene expression. The same differences can also be illustrated
+for the unspliced gene expression profiles (not shown).
 :::
@@
 
-Extended version of @fig-model-results for evaluating Pyro&thinsp;-Velocity (Model 1)
-and scVelo on the PBMC and pancreas datasets. a. Left figure shows the scVelo
-vector field using the default user options, right figure shows the evaluation
-of the scVelo vector field using different combinations of user options within
-Epsilon cells. b. Pyro&thinsp;-Velocity additional single cell level uncertainty
-diagnostic plots on the fully mature PBMC dataset, from left to right: 1.
-standard deviation of posterior samples of shared time, 2. Rayleigh test of
-posterior samples of the vector field (the title reports false discovery rate
-after p-value correction for multiple tests across cells), 3. transcriptional
-state uncertainty of predicted raw read count per cell. The contour lines show
-the top 10% of the most uncertain cells. c. For the pancreas dataset, the
-figures show from left to right: UMAP rendering of 1. Cytotrace, 2. scVelo
-latent time, 3. Posterior mean of Pyro&thinsp;-Velocity shared time averaged across 30
-posterior samples, 4. Top 50 markers overlapping between Pyro&thinsp;-Velocity and
-scVelo (Methods, Supplementary Table 1-2), 5. The same uncertainty diagnostic
-plots per cell as in b. d. The same plot as Fig. 2d. e. Selection of the top
-negatively correlated genes with the cell shared time based on the posterior
-mean on denoised spliced expression and negative mean absolute error of gene
-expression prediction (Methods). f. Phase portraits, rainbow plots, and UMAP
-rendering of posterior mean of gene expression for the genes selected in e.
+Extended version of @fig-model-results for evaluating Pyro&thinsp;-Velocity
+(Model 1) and scVelo on the PBMC and pancreas datasets. a. Left figure shows the
+scVelo vector field using the default user options, right figure shows the
+evaluation of the scVelo vector field using different combinations of user
+options within Epsilon cells. b. Pyro&thinsp;-Velocity additional single cell
+level uncertainty diagnostic plots on the fully mature PBMC dataset, from left
+to right: 1. standard deviation of posterior samples of shared time, 2. Rayleigh
+test of posterior samples of the vector field (the title reports false discovery
+rate after p-value correction for multiple tests across cells), 3.
+transcriptional state uncertainty of predicted raw read count per cell. The
+contour lines show the top 10% of the most uncertain cells. c. For the pancreas
+dataset, the figures show from left to right: UMAP rendering of 1. Cytotrace, 2.
+scVelo latent time, 3. Posterior mean of Pyro&thinsp;-Velocity shared time
+averaged across 30 posterior samples, 4. Top 50 markers overlapping between
+Pyro&thinsp;-Velocity and scVelo (Methods, Supplementary Table 1-2), 5. The same
+uncertainty diagnostic plots per cell as in b. d. The same plot as Fig. 2d. e.
+Selection of the top negatively correlated genes with the cell shared time based
+on the posterior mean on denoised spliced expression and negative mean absolute
+error of gene expression prediction (Methods). f. Phase portraits, rainbow
+plots, and UMAP rendering of posterior mean of gene expression for the genes
+selected in e.
 :::
@@
 
-Extension of @fig-model-results, Pyro&thinsp;-Velocity (Model 2) outperforms scVelo on
-PBMC and pancreas datasets. Same format as in @fig-model-results and
-@suppfig-model-results
+Extension of @fig-model-results, Pyro&thinsp;-Velocity (Model 2) outperforms
+scVelo on PBMC and pancreas datasets. Same format as in @fig-model-results and
+@suppfig-model-results.
 :::
@@
 
-Extension of @fig-lineage-tracing, Pyro&thinsp;-Velocity (Model 1 for uni-fate data,
-Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on scRNA-seq
-datasets with lineage barcoding information (LARRY dataset). a. SPRING embedding
-rendering of Cytotrace scores, the title reports the Spearman's correlation
-between the Cytotrace and Cospar fate potency scores, b. Pyro&thinsp;-Velocity
-additional single cell level uncertainty diagnostic plots based on the
-multi-fate LARRY dataset, from left to right: 1. Standard deviation of posterior
-samples of shared time per cell, 2. Rayleigh test of posterior samples of vector
-field per cell (the title reports the false discovery rate after p-value
-correction for multiple tests across cells), 3. Transcriptional state
-uncertainty of predicted raw read count per cell. The contour lines show the top
-10% of the most uncertain cells (Methods), c. Pairwise Spearman's correlation
-between Cytotrace/Cospar fate potency scores and velocity methods for the four
-different benchmarking datasets selected based on the lineage barcoding
-information from LARRY data. d. SPRING embedding rendering of posterior mean of
-shared time across 30 posterior samples, the title shows the Spearman's
-correlation of mean shared time with Cytotrace. e. Selection of the top
-positively correlated genes with the cell shared time based on the posterior
+Extension of @fig-lineage-tracing, Pyro&thinsp;-Velocity (Model 1 for uni-fate
+data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
+scRNA-seq datasets with lineage barcoding information (LARRY dataset). a. SPRING
+embedding rendering of Cytotrace scores, the title reports the Spearman's
+correlation between the Cytotrace and Cospar fate potency scores, b.
+Pyro&thinsp;-Velocity additional single cell level uncertainty diagnostic plots
+based on the multi-fate LARRY dataset, from left to right: 1. Standard deviation
+of posterior samples of shared time per cell, 2. Rayleigh test of posterior
+samples of vector field per cell (the title reports the false discovery rate
+after p-value correction for multiple tests across cells), 3. Transcriptional
+state uncertainty of predicted raw read count per cell. The contour lines show
+the top 10% of the most uncertain cells (Methods), c. Pairwise Spearman's
+correlation between Cytotrace/Cospar fate potency scores and velocity methods
+for the four different benchmarking datasets selected based on the lineage
+barcoding information from LARRY data. d. SPRING embedding rendering of
+posterior mean of shared time across 30 posterior samples, the title shows the
+Spearman's correlation of mean shared time with Cytotrace. e. Selection of the
+top positively correlated genes with the cell shared time based on the posterior
 mean of spliced expression. f. Phase portraits, rainbow plots, and UMAP
@@
 
-Extension of @fig-lineage-tracing presenting an extended comparison of
+Extension of [@fig-lineage-tracing] presenting an extended comparison of
 Pyro&thinsp;-Velocity and scVelo on scRNA-seq datasets with lineage barcoding
@@
 Pyro&thinsp;-Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
-Pyro&thinsp;-Velocity Model 2, c. Neutrophil and Monocyte bi-fate lineage analysis with
-Pyro&thinsp;-Velocity Model 1, and d. multi-fate lineage analysis with Pyro&thinsp;-Velocity
-Model 1. The panel descriptions are the same as in @fig-lineage-tracing
+Pyro&thinsp;-Velocity Model 2, c. Neutrophil and Monocyte bi-fate lineage
+analysis with Pyro&thinsp;-Velocity Model 1, and d. multi-fate lineage analysis
+with Pyro&thinsp;-Velocity Model 1. The panel descriptions are the same as in
+[@fig-lineage-tracing].
 :::

### 2024-08-20 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+In conclusion, Pyro-Velocity provides the first probabilistic,
+interpretable, end-to-end Bayesian inference framework for RNA Velocity
+that learns cell dynamics and recovers cell fate from multimodal raw
+read counts. This approach provides uncertainty estimations of cell fate
+trajectories and novel visualization approaches to mitigate false
+discoveries. Pyro-Velocity recovered shared time and vector fields that
+are recapitulated in true developmental order and outperformed
+state-of-the-art methods like Cytotrace and scVelo
+(Supplementary Figure~\ref{suppfig-lineage-tracing}a).
+
+\section{Software availability}\label{software-availability}
+
+We developed our Pyro-Velocity models with Pyro (version 1.6.0)
+probabilistic programming language and PyTorch (version 1.8.1).
+Pyro-Velocity source code is available at
+\url{https://github.com/pinellolab/pyrovelocity}. The version of
+Pyro-Velocity (v0.1.0) used to generate the results presented in this
+manuscript has been deposited on Zenodo:
+\url{https://zenodo.org/record/7072876\#.Yx-2e-zMLxg}.
+
+\section{Data availability}\label{data-availability}
+
+All the data presented in this manuscript were collected from previously
+published studies. The fully mature PBMC, pancreas datasets are
+downloaded and loaded through the scVelo using
+\texttt{scvelo.datasets.pancreas()}, and
+\texttt{scvelo.datasets.pbmc68k()}. The LARRY inDrop sequencing data was
+processed by the LARRY authors using dropEst \citep{Petukhov2018-nr} and
+velocyto. This dataset can be accessed at
+\url{https://figshare.com/articles/dataset/larry_invitro_adata_sub_raw_h5ad/20780344}.
+
+\section{Acknowledgments}\label{acknowledgments}
+
+This project has been made possible in part by grant number 2022-249212
+from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon
+Valley Community Foundation. L.P. received support from the U.S. NIH
+(R35 HG010717) and Silicon Valley Community Foundation (2022-249212).
+D.M.L is supported by R01CA226926, R01 CA215118, R01CA154923, R24
+OD031955, CureSearch Acceleration Initiative, and the MGH Scholars
+Award. We thank the Pinello and Langenau labs members for helpful
+discussions.
+
+\section{Author contributions}\label{author-contributions}
+
+L.P., Q.Q., and E.B. conceived the work. Q.Q. and E.B created the
+software. Q.Q. analyzed the data, and prepared the figures and tables.
+L.P and D.L. supervised the project. L.P., D.L., E.B., G.L.M helped in
+designing experiments and analyses. All the authors contributed to the
+writing of the manuscript.
+
+\section{Competing financial interests
+statement}\label{competing-financial-interests-statement}
+
+L.P. has financial interests in Edilytics, Inc., Excelsior Genomics, and
+SeQure Dx, Inc.~L.P.'s interests were reviewed and are managed by
+Massachusetts General Hospital and Partners HealthCare in accordance
+with their conflict of interest policies.
+
 \section{Methods}\label{sec-methods}
@@
 the average cell shared time across 30 posterior samples. The Spearman's
-correlation between Pyro -Velocity and Cytotrace is provided.
-d.~Pyro -Velocity automatic gene selection and ranking for predicting
+correlation between Pyro -Velocity and Cytotrace is provided. d.
+Pyro -Velocity automatic gene selection and ranking for predicting
 important genes involved in pancreas development. e. Phase portraits of
@@
 Fig.~\ref{fig-model-results} and
-Supplementary Figure~\ref{suppfig-model-results}}
+Supplementary Figure~\ref{suppfig-model-results}.}
 
@@
 with Pyro -Velocity Model 1. The panel descriptions are the same as in
-Fig.~\ref{fig-lineage-tracing}}
+Fig.~\ref{fig-lineage-tracing}.}
 

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): refactor extended data and methods links
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 toc-title: Contents
-# Use of `\nameref{sec-methods}` etc. to refer to sections by name.
-# is only compatible with LaTeX outputs
-number-sections: true
+# Use of `\nameref{sec-methods}` versus [Methods](#sec-methods) or [@sec-methods]
+# to refer to sections by name is only compatible with LaTeX outputs
+number-sections: false
 format:
@@
     - kind: float
-      key: suppfig
+      key: extfig
       latex-env: suppfig
-      caption-prefix: Supplementary Figure
-      reference-prefix: Supplementary Fig.
+      caption-prefix: Extended Data Fig.
+      reference-prefix: Extended Data Fig.
       space-before-numbering: true
-      latex-list-of-description: Supplementary Figure
+      latex-list-of-description: Extended Data Figure
 ---
@@
 
-# Main text {#sec-main}
+# Main text {#sec-main .unnumbered}
 
@@
 Pyro&thinsp;-Velocity recasts the velocity estimation problem into a latent
-variable posterior probability inference (@sec-methods). The proposed model is
-generative and fully Bayesian, with the different parameters considered as
-latent random variables. Each parameter is assigned a prior distribution, and
+variable posterior probability inference ([Methods](#sec-methods)). The proposed
+model is generative and fully Bayesian, with the different parameters considered
+as latent random variables. Each parameter is assigned a prior distribution, and
 the model is then conditioned on the raw data (both spliced and unspliced
@@
 of the automatic differentiation variational inference (ADVI) capabilities of
-Pyro [@Bingham2018-id;@Kucukelbir2016-bk] ([@fig-model]a, @sec-methods).
-Notably, our generative Bayesian model enables to sample from the posterior
-distributions of the latent and dependent variables, allowing uncertainty
-estimation of phase portraits, high dimensional velocity vector field, and
-shared latent time. Around this model, we built several visualizations to assess
-uncertainty and functions to recover putative cell fate priming genes and to
-investigate how the denoised gene expression recapitulates known cell-type
-annotations and their developmental order ([@fig-model]b, @sec-methods).
-Although several RNA velocity methods
+Pyro [@Bingham2018-id;@Kucukelbir2016-bk] ([@fig-model]a,
+[Methods](#sec-methods)). Notably, our generative Bayesian model enables to
+sample from the posterior distributions of the latent and dependent variables,
+allowing uncertainty estimation of phase portraits, high dimensional velocity
+vector field, and shared latent time. Around this model, we built several
+visualizations to assess uncertainty and functions to recover putative cell fate
+priming genes and to investigate how the denoised gene expression recapitulates
+known cell-type annotations and their developmental order ([@fig-model]b,
+[Methods](#sec-methods)). Although several RNA velocity methods
 [@Gayoso2024-fz;@Gorin2022-il;@Gu2022-rh;@Cui2022-em;@Gao2022-jq] have been
-recently proposed, and many of them have not been peer-reviewed, here we
-showcase the advantages and innovations of Pyro&thinsp;-Velocity by comparing it
-with the current state-of-the-art dynamical RNA velocity model proposed by
-Bergen et al. [@Bergen2020-pj], and implemented in scVelo.
+recently proposed, here we showcase the advantages and innovations of
+Pyro&thinsp;-Velocity by comparing it with the current state-of-the-art
+dynamical RNA velocity model proposed by Bergen et al. [@Bergen2020-pj], and
+implemented in scVelo.
 
@@
 neighboring cell number for gene expression smoothing, and cannot provide
-confidence assessment of lineage directionality ([@suppfig-model-results]a). In
+confidence assessment of lineage directionality ([@extfig-model-results]a). In
 addition, Pyro&thinsp;-Velocity uncertainty visualizations uncovered that
@@
 successfully recovered key developmental genes that are activated (or repressed)
-during endocrinogenesis ([@fig-model-results]d-e, @suppfig-model-results). Among
+during endocrinogenesis ([@fig-model-results]d-e, @extfig-model-results). Among
 the top positively correlated genes identified by Pyro&thinsp;-Velocity,
@@
 activation (or repression) towards endocrine cells ([@fig-model-results]e,
-[@suppfig-model-results]).
+[@extfig-model-results]).
 
@@
 the clonal progression vector field, i.e., a vector field derived directly from
-the lineage data ([@sec-methods]), 2. Spearman's correlation $(\rho)$ between
-cell latent (shared) time and fate potency scores derived from Cytotrace and
-Cospar. Cospar, our gold standard, is a state-of-the-art method designed
+the lineage data ([Methods](#sec-methods)), 2. Spearman's correlation $(\rho)$
+between cell latent (shared) time and fate potency scores derived from Cytotrace
+and Cospar. Cospar, our gold standard, is a state-of-the-art method designed
 explicitly for predicting fate potency based on LARRY data, while Cytotrace
@@
 $\rho=0.309$, neutrophils $\rho=0.751$) ([@fig-lineage-tracing]a-b,
-[@suppfig-lineage-tracing]c).
+[@extfig-lineage-tracing]c).
 
@@
 methods showed their limitations, given that they are not designed to model and
-capture multiple fate trajectories ([@suppfig-lineage-tracing-2]c-d). The
+capture multiple fate trajectories ([@extfig-lineage-tracing-2]c-d). The
 recovered vector fields and cell latent (shared) time show inconsistent
@@
 transcriptional activation and a baseline expression level potentially different
-than 0 (Pyro&thinsp;-Velocity Model 2, [@sec-methods])
-([@fig-lineage-tracing]c,d, [@suppfig-graphical-models]b,c). This extended model
+than 0 (Pyro&thinsp;-Velocity Model 2, [Methods](#sec-methods))
+([@fig-lineage-tracing]c,d, [@extfig-graphical-models]b,c). This extended model
 recapitulated a vector field consistent with the clonal progression vector field
@@
 Pyro&thinsp;-Velocity also outperformed Cytotrace for developmental order
-prediction ([@suppfig-lineage-tracing]a,c). Importantly, this extended model
+prediction ([@extfig-lineage-tracing]a,c). Importantly, this extended model
 retained its ability to accurately identify cell lineage trajectories and shared
-time for all the other datasets analyzed above ([@suppfig-model-2-result]).
-Taken together, our approach using LARRY barcoded datasets established the first
+time for all the other datasets analyzed above ([@extfig-model-2-result]). Taken
+together, our approach using LARRY barcoded datasets established the first
 quantitative benchmark of RNA velocity-based cell fate trajectory inference and
@@
 
-In conclusion, Pyro-Velocity provides the first probabilistic, interpretable,
-end-to-end Bayesian inference framework for RNA Velocity that learns cell
-dynamics and recovers cell fate from multimodal raw read counts.  This approach
-provides uncertainty estimations of cell fate trajectories and novel
-visualization approaches to mitigate false discoveries. Pyro-Velocity recovered
-shared time and vector fields that are recapitulated in true developmental order
-and outperformed state-of-the-art methods like Cytotrace and scVelo
-([@suppfig-lineage-tracing]a).
+In conclusion, Pyro&thinsp;-Velocity provides the first probabilistic,
+interpretable, end-to-end Bayesian inference framework for RNA Velocity that
+learns cell dynamics and recovers cell fate from multimodal raw read counts.
+This approach provides uncertainty estimations of cell fate trajectories and
+novel visualization approaches to mitigate false discoveries.
+Pyro&thinsp;-Velocity recovered shared time and vector fields that are
+recapitulated in true developmental order and outperformed state-of-the-art
+methods like Cytotrace and scVelo ([@extfig-lineage-tracing]a).
 
-# Software availability
-
-We developed our Pyro-Velocity models with Pyro (version 1.6.0) probabilistic
-programming language and PyTorch (version 1.8.1). Pyro-Velocity source code is
-available at <https://github.com/pinellolab/pyrovelocity>. The version of
-Pyro-Velocity (v0.1.0) used to generate the results presented in this manuscript
-has been deposited on Zenodo: <https://zenodo.org/record/7072876#.Yx-2e-zMLxg>.
-
-# Data availability
-
-All the data presented in this manuscript were collected from previously
-published studies. The fully mature PBMC, pancreas datasets are downloaded and
-loaded through the scVelo using `scvelo.datasets.pancreas()`, and
-`scvelo.datasets.pbmc68k()`. The LARRY inDrop sequencing data was processed by
-the LARRY authors using dropEst [@Petukhov2018-nr] and velocyto. This dataset
-can be accessed at
-<https://figshare.com/articles/dataset/larry_invitro_adata_sub_raw_h5ad/20780344>.
-
-# Acknowledgments
-
-This project has been made possible in part by grant number 2022-249212 from
-the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community
-Foundation. L.P. received support from the U.S. NIH (R35 HG010717) and Silicon
-Valley Community Foundation (2022-249212). D.M.L is supported by R01CA226926,
-R01 CA215118, R01CA154923, R24 OD031955, CureSearch Acceleration Initiative, and
-the MGH Scholars Award. We thank the Pinello and Langenau labs members for
-helpful discussions.
-
-# Author contributions
-
-L.P., Q.Q., and E.B. conceived the work. Q.Q. and E.B created the software. Q.Q.
-analyzed the data, and prepared the figures and tables. L.P and D.L. supervised
-the project.  L.P., D.L., E.B., G.L.M helped in designing experiments and
-analyses.  All the authors contributed to the writing of the manuscript.
-
-# Competing financial interests statement
-
-L.P. has financial interests in Edilytics, Inc., Excelsior Genomics, and SeQure
-Dx, Inc. L.P.’s interests were reviewed and are managed by Massachusetts General
-Hospital and Partners HealthCare in accordance with their conflict of interest
-policies.
-
-# Methods {#sec-methods}
-
-## Model formulation {#sec-model}
-
-We assume the dynamical gene expression is determined by the RNA splicing
-process, and infer the unspliced and spliced gene expression level from the
-differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo
-[@Bergen2020-pj]
-
-\begin{align}
-\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
-\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
-\end{align}
-
-{{< pagebreak >}}
-
-\onecolumn
-
-# References
+# References {.unnumbered}
 
@@
 {{< pagebreak >}}
+
+\onecolumn
 
@@
 
-# Supplementary Figures
+\twocolumn
 
-::: {#suppfig-graphical-models}
+# Methods {#sec-methods}
+
+## Model formulation {#sec-model}
+
+We assume the dynamical gene expression is determined by the RNA splicing
+process, and infer the unspliced and spliced gene expression level from the
+differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo
+[@Bergen2020-pj]
+\begin{align}
+\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
+  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
+\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
+  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
+\end{align}
+
+## Model evaluation and selection {#sec-model-selection}
+
+Pyro&thinsp;-Velocity provides two model formulations as described above.
+Users can compare their predictive performance to decide which of the two models
+to use for a given dataset. Namely, the mean absolute errors of posterior
+samples prediction of spliced and unspliced raw read counts $x_{icg}$ from its
+observation $\tilde{X}_{cg}$ per gene and cell combinations
+\begin{align*}
+\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} |x_{icg} - \tilde{X}_{cg}|,
+\end{align*}
+where $i$ is the posterior sample index, $N_i$ is the posterior sample number,
+$N_c$ is the cell number, $N_g$ is the gene number. Then it is possible to
+evaluate the cell fate inference performance of the models based on two metrics:
+
+1. the consistency between the velocity vector field and the *clonal progression
+vector field* using grid-based cosine similarity; 2. Spearman’s correlation of
+velocity-based shared (latent) time with Cospar fate potency and the silver
+standard Cytotrace score. Both metrics are implemented in scipy.
+
+# Data availability
+
+All the data presented in this manuscript were collected from previously
+published studies. The fully mature PBMC, pancreas datasets are downloaded and
+loaded through the scVelo using `scvelo.datasets.pancreas()`, and
+`scvelo.datasets.pbmc68k()`. The LARRY inDrop sequencing data was processed by
+the LARRY authors using dropEst [@Petukhov2018-nr] and velocyto. This dataset
+can be accessed at
+<https://figshare.com/articles/dataset/larry_invitro_adata_sub_raw_h5ad/20780344>.
+
+# Code availability
+
+We developed our Pyro&thinsp;-Velocity models with Pyro (version 1.6.0)
+probabilistic programming language and PyTorch (version 1.8.1).
+Pyro&thinsp;-Velocity source code is available at
+<https://github.com/pinellolab/pyrovelocity>. The version of
+Pyro&thinsp;-Velocity (v0.1.0) used to generate the results presented in this
+manuscript has been deposited on Zenodo:
+<https://zenodo.org/record/7072876#.Yx-2e-zMLxg>.
+
+<!-- # End notes {.unnumbered} -->
+
+# Acknowledgments {.unnumbered}
+
+This project has been made possible in part by grant number 2022-249212 from
+the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community
+Foundation. L.P. received support from the U.S. NIH (R35 HG010717) and Silicon
+Valley Community Foundation (2022-249212). D.M.L is supported by R01CA226926,
+R01 CA215118, R01CA154923, R24 OD031955, CureSearch Acceleration Initiative, and
+the MGH Scholars Award. We thank the Pinello and Langenau labs members for
+helpful discussions.
+
+# Author contributions {.unnumbered}
+
+L.P., Q.Q., and E.B. conceived the work. Q.Q. and E.B created the software. Q.Q.
+analyzed the data, and prepared the figures and tables. L.P and D.L. supervised
+the project.  L.P., D.L., E.B., G.L.M helped in designing experiments and
+analyses.  All the authors contributed to the writing of the manuscript.
+
+# Competing financial interests statement {.unnumbered}
+
+L.P. has financial interests in Edilytics, Inc., Excelsior Genomics, and SeQure
+Dx, Inc. L.P.’s interests were reviewed and are managed by Massachusetts General
+Hospital and Partners HealthCare in accordance with their conflict of interest
+policies.
+
+\FloatBarrier
+\clearpage
+{{< pagebreak >}}
+
+\onecolumn
+
+# Extended Data Figures
+
+::: {#extfig-graphical-models}
 ![](figures/fig.pdf)
@@
 
-::: {#suppfig-model-results}
+::: {#extfig-model-results}
 ![](figures/fig.pdf)
@@
 
-::: {#suppfig-model-2-result}
+::: {#extfig-model-2-result}
 ![](figures/fig.pdf)
@@
 scVelo on PBMC and pancreas datasets. Same format as in @fig-model-results and
-@suppfig-model-results.
+@extfig-model-results.
 :::
 
-::: {#suppfig-lineage-tracing}
+::: {#extfig-lineage-tracing}
 ![](figures/fig.pdf)
@@
 
-::: {#suppfig-lineage-tracing-2}
+::: {#extfig-lineage-tracing-2}
 ![](figures/fig.pdf)
@@
 
-![An example figure (an empty plot)](figures/fig.pdf){#suppfig-plot}
+<!-- ![An example figure (an empty plot)](figures/fig.pdf){#extfig-plot} -->
 
@@
 
-\FloatBarrier
-{{< pagebreak >}}
-
-# Supporting Information {.appendix}
+<!-- # Supporting Information {.appendix} -->

### 2024-08-20 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
@@
 \floatstyle{plain}
-\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{losuppfig}}{\newfloat{suppfig}{h}{losuppfig}[chapter]}
-\floatname{suppfig}{Supplementary Figure}
-\newcommand*\listofsuppfigs{\listof{suppfig}{List of Supplementary Figures}}
+\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{loextfig}}{\newfloat{suppfig}{h}{loextfig}[chapter]}
+\floatname{suppfig}{Extended Data Fig.}
+\newcommand*\listofsuppfigs{\listof{suppfig}{List of Extended Data Figures}}
 \makeatother
@@
 
-\section{Main text}\label{sec-main}
+\section*{Main text}\label{sec-main}
+\addcontentsline{toc}{section}{Main text}
 
@@
 Pyro -Velocity recasts the velocity estimation problem into a latent
-variable posterior probability inference (Sec.~\ref{sec-methods}). The
-proposed model is generative and fully Bayesian, with the different
-parameters considered as latent random variables. Each parameter is
-assigned a prior distribution, and the model is then conditioned on the
-raw data (both spliced and unspliced molecule raw counts) to estimate
-posterior probability distributions and to enable uncertainty estimation
-of velocity. Central to the Pyro -Velocity model is a \emph{shared
-latent time} \((t)\): a global latent variable placing each cell at a
-particular instant of the modeled dynamical process. From this latent
-time, the expectation of spliced and unspliced RNA levels for each gene
-can be computed knowing gene-specific kinetics parameters
-(transcriptional \((\alpha)\), splicing \((\beta)\), and degradation
-rates (\(\gamma\)) and two gene-specific switching times
-\((t_{0}^{k})\), and gene-specific transcriptional state \(k\)) and
-integrating the RNA velocity differential equations analogously to what
-proposed in the dynamical RNA velocity model by Bergen et al.
-\citep{Bergen2020-pj, Li2021-qa}. However, we innovate the structure and
-formulation of the latent time given that our model is set to jointly
-fit the entire raw data matrix, not individual genes. We approach the
-task of estimating the posterior distribution by restricting the search
-within a family of a variational distribution approximating the true
-posterior, and taking advantage of the automatic differentiation
-variational inference (ADVI) capabilities of Pyro
-\citep{Bingham2018-id, Kucukelbir2016-bk} (Fig.~\ref{fig-model}a,
-Sec.~\ref{sec-methods}). Notably, our generative Bayesian model enables
-to sample from the posterior distributions of the latent and dependent
-variables, allowing uncertainty estimation of phase portraits, high
-dimensional velocity vector field, and shared latent time. Around this
-model, we built several visualizations to assess uncertainty and
-functions to recover putative cell fate priming genes and to investigate
-how the denoised gene expression recapitulates known cell-type
-annotations and their developmental order (Fig.~\ref{fig-model}b,
-Sec.~\ref{sec-methods}). Although several RNA velocity methods
+variable posterior probability inference
+(\hyperref[sec-methods]{Methods}). The proposed model is generative and
+fully Bayesian, with the different parameters considered as latent
+random variables. Each parameter is assigned a prior distribution, and
+the model is then conditioned on the raw data (both spliced and
+unspliced molecule raw counts) to estimate posterior probability
+distributions and to enable uncertainty estimation of velocity. Central
+to the Pyro -Velocity model is a \emph{shared latent time} \((t)\): a
+global latent variable placing each cell at a particular instant of the
+modeled dynamical process. From this latent time, the expectation of
+spliced and unspliced RNA levels for each gene can be computed knowing
+gene-specific kinetics parameters (transcriptional \((\alpha)\),
+splicing \((\beta)\), and degradation rates (\(\gamma\)) and two
+gene-specific switching times \((t_{0}^{k})\), and gene-specific
+transcriptional state \(k\)) and integrating the RNA velocity
+differential equations analogously to what proposed in the dynamical RNA
+velocity model by Bergen et al. \citep{Bergen2020-pj, Li2021-qa}.
+However, we innovate the structure and formulation of the latent time
+given that our model is set to jointly fit the entire raw data matrix,
+not individual genes. We approach the task of estimating the posterior
+distribution by restricting the search within a family of a variational
+distribution approximating the true posterior, and taking advantage of
+the automatic differentiation variational inference (ADVI) capabilities
+of Pyro \citep{Bingham2018-id, Kucukelbir2016-bk}
+(Fig.~\ref{fig-model}a, \hyperref[sec-methods]{Methods}). Notably, our
+generative Bayesian model enables to sample from the posterior
+distributions of the latent and dependent variables, allowing
+uncertainty estimation of phase portraits, high dimensional velocity
+vector field, and shared latent time. Around this model, we built
+several visualizations to assess uncertainty and functions to recover
+putative cell fate priming genes and to investigate how the denoised
+gene expression recapitulates known cell-type annotations and their
+developmental order (Fig.~\ref{fig-model}b,
+\hyperref[sec-methods]{Methods}). Although several RNA velocity methods
 \citep{Gayoso2024-fz, Gorin2022-il, Gu2022-rh, Cui2022-em, Gao2022-jq}
-have been recently proposed, and many of them have not been
-peer-reviewed, here we showcase the advantages and innovations of
-Pyro -Velocity by comparing it with the current state-of-the-art
-dynamical RNA velocity model proposed by Bergen et al.
+have been recently proposed, here we showcase the advantages and
+innovations of Pyro -Velocity by comparing it with the current
+state-of-the-art dynamical RNA velocity model proposed by Bergen et al.
 \citep{Bergen2020-pj}, and implemented in scVelo.
@@
 provide confidence assessment of lineage directionality
-(Supplementary Figure~\ref{suppfig-model-results}a). In addition,
+(Extended Data Fig.~\ref{extfig-model-results}a). In addition,
 Pyro -Velocity uncertainty visualizations uncovered that differentiated
@@
 during endocrinogenesis (Fig.~\ref{fig-model-results}d-e,
-Supplementary Figure~\ref{suppfig-model-results}). Among the top
-positively correlated genes identified by Pyro -Velocity, endocrine Beta
-cells expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a key
+Extended Data Fig.~\ref{extfig-model-results}). Among the top positively
+correlated genes identified by Pyro -Velocity, endocrine Beta cells
+expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a key
 regulator Beta cell insulin generation \citep{Millership2018-vd} and
@@
 repression) towards endocrine cells (Fig.~\ref{fig-model-results}e,
-Supplementary Figure~\ref{suppfig-model-results}).
+Extended Data Fig.~\ref{extfig-model-results}).
 
@@
 progression vector field, i.e., a vector field derived directly from the
-lineage data (Sec.~\ref{sec-methods}), 2. Spearman's correlation
-\((\rho)\) between cell latent (shared) time and fate potency scores
-derived from Cytotrace and Cospar. Cospar, our gold standard, is a
-state-of-the-art method designed explicitly for predicting fate potency
-based on LARRY data, while Cytotrace leverages only transcriptomic data.
-We selected cell state changes associated with monocyte and neutrophil
-lineages because both lineages have the largest numbers of cells in the
-scRNA sequencing data set and well-known cell fate trajectories
-\citep{Petukhov2018-nr}. As expected, the Pyro -Velocity vector field
-accurately predicted cell state trajectories from hematopoietic
-progenitor cells into mature monocytes or neutrophils and had high
-cosine similarity between the inferred vector field and clonal
-progression vector field assigned by LARRY (monocytes \(cs=0.49\),
-neutrophils \(cs=0.38\)). High Spearman's correlation was also seen
-between shared latent time and Cospar fate potency (monocytes
+lineage data (\hyperref[sec-methods]{Methods}), 2. Spearman's
+correlation \((\rho)\) between cell latent (shared) time and fate
+potency scores derived from Cytotrace and Cospar. Cospar, our gold
+standard, is a state-of-the-art method designed explicitly for
+predicting fate potency based on LARRY data, while Cytotrace leverages
+only transcriptomic data. We selected cell state changes associated with
+monocyte and neutrophil lineages because both lineages have the largest
+numbers of cells in the scRNA sequencing data set and well-known cell
+fate trajectories \citep{Petukhov2018-nr}. As expected, the
+Pyro -Velocity vector field accurately predicted cell state trajectories
+from hematopoietic progenitor cells into mature monocytes or neutrophils
+and had high cosine similarity between the inferred vector field and
+clonal progression vector field assigned by LARRY (monocytes
+\(cs=0.49\), neutrophils \(cs=0.38\)). High Spearman's correlation was
+also seen between shared latent time and Cospar fate potency (monocytes
 \(\rho=0.60\), neutrophils \(\rho=0.74\)) or shared latent time with
@@
 (Fig.~\ref{fig-lineage-tracing}a-b,
-Supplementary Figure~\ref{suppfig-lineage-tracing}c).
+Extended Data Fig.~\ref{extfig-lineage-tracing}c).
 
@@
 not designed to model and capture multiple fate trajectories
-(Supplementary Figure~\ref{suppfig-lineage-tracing-2}c-d). The recovered
+(Extended Data Fig.~\ref{extfig-lineage-tracing-2}c-d). The recovered
 vector fields and cell latent (shared) time show inconsistent directions
@@
 potentially different than 0 (Pyro -Velocity Model 2,
-Sec.~\ref{sec-methods}) (Fig.~\ref{fig-lineage-tracing}c,d,
-Supplementary Figure~\ref{suppfig-graphical-models}b,c). This extended
+\hyperref[sec-methods]{Methods}) (Fig.~\ref{fig-lineage-tracing}c,d,
+Extended Data Fig.~\ref{extfig-graphical-models}b,c). This extended
 model recapitulated a vector field consistent with the clonal
@@
 developmental order prediction
-(Supplementary Figure~\ref{suppfig-lineage-tracing}a,c). Importantly,
-this extended model retained its ability to accurately identify cell
-lineage trajectories and shared time for all the other datasets analyzed
-above (Supplementary Figure~\ref{suppfig-model-2-result}). Taken
-together, our approach using LARRY barcoded datasets established the
-first quantitative benchmark of RNA velocity-based cell fate trajectory
+(Extended Data Fig.~\ref{extfig-lineage-tracing}a,c). Importantly, this
+extended model retained its ability to accurately identify cell lineage
+trajectories and shared time for all the other datasets analyzed above
+(Extended Data Fig.~\ref{extfig-model-2-result}). Taken together, our
+approach using LARRY barcoded datasets established the first
+quantitative benchmark of RNA velocity-based cell fate trajectory
 inference and provides an exciting new framework for continued
@@
 
-In conclusion, Pyro-Velocity provides the first probabilistic,
+In conclusion, Pyro -Velocity provides the first probabilistic,
 interpretable, end-to-end Bayesian inference framework for RNA Velocity
@@
 trajectories and novel visualization approaches to mitigate false
-discoveries. Pyro-Velocity recovered shared time and vector fields that
+discoveries. Pyro -Velocity recovered shared time and vector fields that
 are recapitulated in true developmental order and outperformed
 state-of-the-art methods like Cytotrace and scVelo
-(Supplementary Figure~\ref{suppfig-lineage-tracing}a).
+(Extended Data Fig.~\ref{extfig-lineage-tracing}a).
 
-\section{Software availability}\label{software-availability}
-
-We developed our Pyro-Velocity models with Pyro (version 1.6.0)
-probabilistic programming language and PyTorch (version 1.8.1).
-Pyro-Velocity source code is available at
-\url{https://github.com/pinellolab/pyrovelocity}. The version of
-Pyro-Velocity (v0.1.0) used to generate the results presented in this
-manuscript has been deposited on Zenodo:
-\url{https://zenodo.org/record/7072876\#.Yx-2e-zMLxg}.
-
-\section{Data availability}\label{data-availability}
-
-All the data presented in this manuscript were collected from previously
-published studies. The fully mature PBMC, pancreas datasets are
-downloaded and loaded through the scVelo using
-\texttt{scvelo.datasets.pancreas()}, and
-\texttt{scvelo.datasets.pbmc68k()}. The LARRY inDrop sequencing data was
-processed by the LARRY authors using dropEst \citep{Petukhov2018-nr} and
-velocyto. This dataset can be accessed at
-\url{https://figshare.com/articles/dataset/larry_invitro_adata_sub_raw_h5ad/20780344}.
-
-\section{Acknowledgments}\label{acknowledgments}
-
-This project has been made possible in part by grant number 2022-249212
-from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon
-Valley Community Foundation. L.P. received support from the U.S. NIH
-(R35 HG010717) and Silicon Valley Community Foundation (2022-249212).
-D.M.L is supported by R01CA226926, R01 CA215118, R01CA154923, R24
-OD031955, CureSearch Acceleration Initiative, and the MGH Scholars
-Award. We thank the Pinello and Langenau labs members for helpful
-discussions.
-
-\section{Author contributions}\label{author-contributions}
-
-L.P., Q.Q., and E.B. conceived the work. Q.Q. and E.B created the
-software. Q.Q. analyzed the data, and prepared the figures and tables.
-L.P and D.L. supervised the project. L.P., D.L., E.B., G.L.M helped in
-designing experiments and analyses. All the authors contributed to the
-writing of the manuscript.
-
-\section{Competing financial interests
-statement}\label{competing-financial-interests-statement}
-
-L.P. has financial interests in Edilytics, Inc., Excelsior Genomics, and
-SeQure Dx, Inc.~L.P.'s interests were reviewed and are managed by
-Massachusetts General Hospital and Partners HealthCare in accordance
-with their conflict of interest policies.
-
-\section{Methods}\label{sec-methods}
-
-\subsection{Model formulation}\label{sec-model}
-
-We assume the dynamical gene expression is determined by the RNA
-splicing process, and infer the unspliced and spliced gene expression
-level from the differential equations proposed in velocyto
-\citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj}
-
-\begin{align}
-\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
-\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
-\end{align}
-
-\newpage{}
-
-\onecolumn
-
-\section{References}\label{references}
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
 
@@
 \newpage{}
+
+\onecolumn
 
@@
 
-\section{Supplementary Figures}\label{supplementary-figures}
+\twocolumn
+
+\section{Methods}\label{sec-methods}
+
+\subsection{Model formulation}\label{sec-model}
+
+We assume the dynamical gene expression is determined by the RNA
+splicing process, and infer the unspliced and spliced gene expression
+level from the differential equations proposed in velocyto
+\citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj} \begin{align}
+\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
+  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
+\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
+  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
+\end{align}
+
+\subsection{Model evaluation and selection}\label{sec-model-selection}
+
+Pyro -Velocity provides two model formulations as described above. Users
+can compare their predictive performance to decide which of the two
+models to use for a given dataset. Namely, the mean absolute errors of
+posterior samples prediction of spliced and unspliced raw read counts
+\(x_{icg}\) from its observation \(\tilde{X}_{cg}\) per gene and cell
+combinations \begin{align*}
+\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} |x_{icg} - \tilde{X}_{cg}|,
+\end{align*} where \(i\) is the posterior sample index, \(N_i\) is the
+posterior sample number, \(N_c\) is the cell number, \(N_g\) is the gene
+number. Then it is possible to evaluate the cell fate inference
+performance of the models based on two metrics:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  the consistency between the velocity vector field and the \emph{clonal
+  progression vector field} using grid-based cosine similarity; 2.
+  Spearman's correlation of velocity-based shared (latent) time with
+  Cospar fate potency and the silver standard Cytotrace score. Both
+  metrics are implemented in scipy.
+\end{enumerate}
+
+\section{Data availability}\label{data-availability}
+
+All the data presented in this manuscript were collected from previously
+published studies. The fully mature PBMC, pancreas datasets are
+downloaded and loaded through the scVelo using
+\texttt{scvelo.datasets.pancreas()}, and
+\texttt{scvelo.datasets.pbmc68k()}. The LARRY inDrop sequencing data was
+processed by the LARRY authors using dropEst \citep{Petukhov2018-nr} and
+velocyto. This dataset can be accessed at
+\url{https://figshare.com/articles/dataset/larry_invitro_adata_sub_raw_h5ad/20780344}.
+
+\section{Code availability}\label{code-availability}
+
+We developed our Pyro -Velocity models with Pyro (version 1.6.0)
+probabilistic programming language and PyTorch (version 1.8.1).
+Pyro -Velocity source code is available at
+\url{https://github.com/pinellolab/pyrovelocity}. The version of
+Pyro -Velocity (v0.1.0) used to generate the results presented in this
+manuscript has been deposited on Zenodo:
+\url{https://zenodo.org/record/7072876\#.Yx-2e-zMLxg}.
+
+\section*{Acknowledgments}\label{acknowledgments}
+\addcontentsline{toc}{section}{Acknowledgments}
+
+This project has been made possible in part by grant number 2022-249212
+from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon
+Valley Community Foundation. L.P. received support from the U.S. NIH
+(R35 HG010717) and Silicon Valley Community Foundation (2022-249212).
+D.M.L is supported by R01CA226926, R01 CA215118, R01CA154923, R24
+OD031955, CureSearch Acceleration Initiative, and the MGH Scholars
+Award. We thank the Pinello and Langenau labs members for helpful
+discussions.
+
+\section*{Author contributions}\label{author-contributions}
+\addcontentsline{toc}{section}{Author contributions}
+
+L.P., Q.Q., and E.B. conceived the work. Q.Q. and E.B created the
+software. Q.Q. analyzed the data, and prepared the figures and tables.
+L.P and D.L. supervised the project. L.P., D.L., E.B., G.L.M helped in
+designing experiments and analyses. All the authors contributed to the
+writing of the manuscript.
+
+\section*{Competing financial interests
+statement}\label{competing-financial-interests-statement}
+\addcontentsline{toc}{section}{Competing financial interests statement}
+
+L.P. has financial interests in Edilytics, Inc., Excelsior Genomics, and
+SeQure Dx, Inc.~L.P.'s interests were reviewed and are managed by
+Massachusetts General Hospital and Partners HealthCare in accordance
+with their conflict of interest policies.
+
+\FloatBarrier
+\clearpage
+
+\newpage{}
+
+\onecolumn
+
+\section{Extended Data Figures}\label{extended-data-figures}
 
@@
 
-\caption{\label{suppfig-graphical-models}Graphical representations
-detail the generative process of the Pyro -Velocity models and highlight
-their differences. a. Pyro -Velocity Model 1. Each rectangle represents
-one data dimension (e.g., gene or cell). The arrows represent the
-directed hierarchical dependency among random variables. The white
-circles correspond to the modeled random variables. The probability
+\caption{\label{extfig-graphical-models}Graphical representations detail
+the generative process of the Pyro -Velocity models and highlight their
+differences. a. Pyro -Velocity Model 1. Each rectangle represents one
+data dimension (e.g., gene or cell). The arrows represent the directed
+hierarchical dependency among random variables. The white circles
+correspond to the modeled random variables. The probability
 distributions of these variables are denoted on the right legend. The
@@
 
-\caption{\label{suppfig-model-results}Extended version of
+\caption{\label{extfig-model-results}Extended version of
 Fig.~\ref{fig-model-results} for evaluating Pyro -Velocity (Model 1) and
@@
 
-\caption{\label{suppfig-model-2-result}Extension of
+\caption{\label{extfig-model-2-result}Extension of
 Fig.~\ref{fig-model-results}, Pyro -Velocity (Model 2) outperforms
@@
 Fig.~\ref{fig-model-results} and
-Supplementary Figure~\ref{suppfig-model-results}.}
+Extended Data Fig.~\ref{extfig-model-results}.}
 
@@
 
-\caption{\label{suppfig-lineage-tracing}Extension of
+\caption{\label{extfig-lineage-tracing}Extension of
 Fig.~\ref{fig-lineage-tracing}, Pyro -Velocity (Model 1 for uni-fate
@@
 
-\caption{\label{suppfig-lineage-tracing-2}Extension of
+\caption{\label{extfig-lineage-tracing-2}Extension of
 Fig.~\ref{fig-lineage-tracing} presenting an extended comparison of
@@
 
-\begin{suppfig}
-
-\centering{
-
-\includegraphics{figures/fig.pdf}
-
-}
-
-\caption{\label{suppfig-plot}An example figure (an empty plot)}
-
-\end{suppfig}%
-
-\FloatBarrier
-
-\newpage{}
-
-\renewcommand\refname{Supporting Information}
-
 \end{document}

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
chore(manuscript): add cytotrace methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-## Model formulation {#sec-model}
+## Model formulation {#sec-methods-model}
 
@@
 
-## Model evaluation and selection {#sec-model-selection}
+## Cytotrace model {#sec-methods-cytotrace}
+
+We implemented Cytotrace in Python based on the R version [@Gulati2020-xq] for
+benchmarking Pyro&thinsp;-Velocity and scVelo’s shared (latent) time prediction in the
+Python environment. This implementation is available in the 
+[Pyro&thinsp;-Velocity GitHub repository](https://github.com/pinellolab/pyrovelocity/blob/main/src/pyrovelocity/analysis/cytotrace.py).
+
+## Model evaluation and selection {#sec-methods-model-selection}
 

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
chore(manuscript): add cospar methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+## Cospar {#sec-methods-cospar}
+
+We computed the
+[Cospar `v0.1.9`](https://github.com/AllonKleinLab/cospar/releases/tag/v0.1.9) 
+[@Wang2022-xb] fate potency score for the above processed multi-fate LARRY dataset, 
+and selected the subset of the fate potency scores for evaluating bi-fate and 
+uni-fate datasets. We used the following function calls to achieve this
+
+```python
+cospar.tmap.infer_Tmap_from_multitime_clones(
+  adata_orig=adata, 
+  compute_new=True,
+  clonal_time_points=[2,4,6], 
+  later_time_point=6,
+  smooth_array=[20,15,10],
+  sparsity_threshold=0.2, 
+  max_iter_N=3
+)
+cospar.pl.fate_potency(
+  adata=adata, 
+  used_Tmap='transition_map',
+  map_backward=True,
+  method='norm-sum',
+  color_bar=True,
+  fate_count=True
+)
+```
+
 ## Cytotrace model {#sec-methods-cytotrace}

### 2024-08-20 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{241,243,245}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
+\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+
 \providecommand{\tightlist}{%
@@
 \usepackage{placeins}
+\usepackage{framed}
 \allowdisplaybreaks
@@
 
-\subsection{Model formulation}\label{sec-model}
+\subsection{Model formulation}\label{sec-methods-model}
 
@@
 
-\subsection{Model evaluation and selection}\label{sec-model-selection}
+\subsection{Cospar}\label{sec-methods-cospar}
+
+We computed the
+\href{https://github.com/AllonKleinLab/cospar/releases/tag/v0.1.9}{Cospar
+\texttt{v0.1.9}} \citep{Wang2022-xb} fate potency score for the above
+processed multi-fate LARRY dataset, and selected the subset of the fate
+potency scores for evaluating bi-fate and uni-fate datasets. We used the
+following function calls to achieve this
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{cospar.tmap.infer\_Tmap\_from\_multitime\_clones(}
+\NormalTok{  adata\_orig}\OperatorTok{=}\NormalTok{adata, }
+\NormalTok{  compute\_new}\OperatorTok{=}\VariableTok{True}\NormalTok{,}
+\NormalTok{  clonal\_time\_points}\OperatorTok{=}\NormalTok{[}\DecValTok{2}\NormalTok{,}\DecValTok{4}\NormalTok{,}\DecValTok{6}\NormalTok{], }
+\NormalTok{  later\_time\_point}\OperatorTok{=}\DecValTok{6}\NormalTok{,}
+\NormalTok{  smooth\_array}\OperatorTok{=}\NormalTok{[}\DecValTok{20}\NormalTok{,}\DecValTok{15}\NormalTok{,}\DecValTok{10}\NormalTok{],}
+\NormalTok{  sparsity\_threshold}\OperatorTok{=}\FloatTok{0.2}\NormalTok{, }
+\NormalTok{  max\_iter\_N}\OperatorTok{=}\DecValTok{3}
+\NormalTok{)}
+\NormalTok{cospar.pl.fate\_potency(}
+\NormalTok{  adata}\OperatorTok{=}\NormalTok{adata, }
+\NormalTok{  used\_Tmap}\OperatorTok{=}\StringTok{\textquotesingle{}transition\_map\textquotesingle{}}\NormalTok{,}
+\NormalTok{  map\_backward}\OperatorTok{=}\VariableTok{True}\NormalTok{,}
+\NormalTok{  method}\OperatorTok{=}\StringTok{\textquotesingle{}norm{-}sum\textquotesingle{}}\NormalTok{,}
+\NormalTok{  color\_bar}\OperatorTok{=}\VariableTok{True}\NormalTok{,}
+\NormalTok{  fate\_count}\OperatorTok{=}\VariableTok{True}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\subsection{Cytotrace model}\label{sec-methods-cytotrace}
+
+We implemented Cytotrace in Python based on the R version
+\citep{Gulati2020-xq} for benchmarking Pyro -Velocity and scVelo's
+shared (latent) time prediction in the Python environment. This
+implementation is available in the
+\href{https://github.com/pinellolab/pyrovelocity/blob/main/src/pyrovelocity/analysis/cytotrace.py}{Pyro -Velocity
+GitHub repository}.
+
+\subsection{Model evaluation and
+selection}\label{sec-methods-model-selection}
 

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): add cell fate uncertainty estimation methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+## Cell fate uncertainty estimation {#sec-methods-fate-uncertainty}
+
+Pyro&thinsp;-Velocity evaluates cell fate uncertainty using four statistics. The
+first is the uncertainty of shared time, which is evaluated using the standard
+deviation of posterior samples of $t_c$ for each cell. The second statistic is
+the angular standard deviation [@Berens2009-fu] across posterior samples of the
+velocity-based vector field. The value of this statistic was mapped linearly to
+the range $[0, 360]$ for interpretability. The third statistic is the Rayleigh
+test [@The-Astropy-Collaboration2022-aw] for posterior samples of the
+velocity-based vector fields for each cell, whose p-values have been corrected
+for multiple tests using Benjamini-Hochberg FDR methods; if FDR is less than
+$0.05$, the cell is considered to have one statistically significant future
+direction. The second and third circular statistics were implemented using the
+Astropy package (`v5.1`) [@The-Astropy-Collaboration2022-aw]. The last statistic
+is state uncertainty which evaluates the mean Euclidean distance between
+posterior mean and posterior samples of the raw read count prediction for each
+cell
+\begin{align*}
+\frac{1}{N} \sum_{i} \sqrt{\sum_{g} \left(x_{icg} - \frac{1}{N}\sum_i x_{icg}\right)^2}.
+\end{align*}
+
 ## Cospar {#sec-methods-cospar}

### 2024-08-20 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

@@
 \usepackage{placeins}
-\usepackage{framed}
 \allowdisplaybreaks
@@
 
+\subsection{Cell fate uncertainty
+estimation}\label{sec-methods-fate-uncertainty}
+
+Pyro -Velocity evaluates cell fate uncertainty using four statistics.
+The first is the uncertainty of shared time, which is evaluated using
+the standard deviation of posterior samples of \(t_c\) for each cell.
+The second statistic is the angular standard deviation
+\citep{Berens2009-fu} across posterior samples of the velocity-based
+vector field. The value of this statistic was mapped linearly to the
+range \([0, 360]\) for interpretability. The third statistic is the
+Rayleigh test \citep{The-Astropy-Collaboration2022-aw} for posterior
+samples of the velocity-based vector fields for each cell, whose
+p-values have been corrected for multiple tests using Benjamini-Hochberg
+FDR methods; if FDR is less than \(0.05\), the cell is considered to
+have one statistically significant future direction. The second and
+third circular statistics were implemented using the Astropy package
+(\texttt{v5.1}) \citep{The-Astropy-Collaboration2022-aw}. The last
+statistic is state uncertainty which evaluates the mean Euclidean
+distance between posterior mean and posterior samples of the raw read
+count prediction for each cell \begin{align*}
+\frac{1}{N} \sum_{i} \sqrt{\sum_{g} \left(x_{icg} - \frac{1}{N}\sum_i x_{icg}\right)^2}.
+\end{align*}
+
 \subsection{Cospar}\label{sec-methods-cospar}

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): add trajectory inference methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+## Trajectory inference {#sec-methods-trajectory-inference}
+
+### Velocity vector field
+
+The velocity-based vector fields were generated using the scVelo function call
+
+```python
+scvelo.tl.velocity_graph(adata)
+scvelo.tl.velocity_embedding(
+  adata, 
+  vkey='velocity',
+  basis='embed',
+)
+```
+
+We used the default options for projecting the vector fields from scVelo models.
+Unlike scVelo, Pyro-Velocity uses statistics derived from posterior samples of
+the denoised spliced gene expression and posterior samples of the velocity
+estimation for building the cell state transition matrix estimates using cosine
+similarity. Pyro-Velocity uses the same projection method as scVelo for
+projecting the transition matrix into the two-dimensional vector field on the
+user-provided embedding space.
+
+### Clonal progression vector field
+
+No method exists for projecting clonal cell fates from the LARRY data and
+associated subsets thereof into a vector field. Thus, we designed a strategy for
+projecting clonal progression vector fields. Given cells sharing the same LARRY
+barcode, we considered the vectors connecting the embedding centroid of cells at
+consecutive time points and then leveraged Gaussian smoothing of nearest
+centroid directions to represent the clonal progression vector field informed by
+LARRY barcodes on a regular grid over the cell manifold.
+
 ## Cell fate uncertainty estimation {#sec-methods-fate-uncertainty}

### 2024-08-21 reproducibility/manuscript/manuscript.qmd
docs(manuscript): add preprocessing methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+## Single-cell data preprocessing {#sec-methods-preprocessing}
+
+We used scanpy and scVelo to handle the data input and output; thus, both h5ad
+and loom files generated by velocyto and kallisto [@Melsted2021-ap] are
+supported. The fully mature PBMC dataset was processed with the same procedure
+proposed in a review paper [@Bergen2021-qz]
+(<https://scvelo.readthedocs.io/perspectives/Perspectives/>). We reproduced this
+procedure using the scVelo package and raw read counts of the same top three
+dynamical genes NKG7, IGHM, and GNLY with the best likelihoods. The pancreas
+dataset was processed with scVelo using the following options
+
+```python
+scv.pp.filter_and_normalize(
+  adata=adata,
+  min_shared_counts=30, 
+  n_top_genes=2000,
+)
+scv.pp.moments(
+  adata,
+  n_pcs=30,
+  n_neighbors=30,
+)
+```
+
+The same top variable genes with raw spliced and unspliced read counts were used
+as input for the Pyro&thinsp;-Velocity model. The original LARRY dataset of in
+vitro Hematopoiesis containing $130,887$ cells was first filtered to remove
+cells without LARRY barcoding. $49,302$ cells were recovered after this step
+with at least one LARRY barcode. For simplicity, we termed this filtered dataset
+with multiple cell fate (multi-fate) as the full dataset. Based on this dataset,
+we created two datasets with uni-fate progression toward monocyte or neutrophil
+based on the lineage LARRY barcodes and time information. Namely, we selected
+sets of cells with a single LARRY barcode, spanning three time points (day 2, 4,
+6), and all the cells from the last time point (day 6) belong to a unique cell
+type (either monocyte or neutrophil). The two uni-fate datasets were combined to
+represent the bi-fate LARRY dataset. The multi-fate full dataset was processed
+using the same options as the pancreas dataset; the rest of the uni-fate and
+bi-fate datasets were processed using the following parameters
+
+```python
+scv.pp.filter_and_normalize(
+  adata=adata,
+  n_top_genes=2000,
+  min_shared_counts=20,
+)
+scv.pp.moments(adata)
+```
+
+## scVelo model {#sec-methods-scvelo}
+
+We benchmarked the dynamical RNA velocity model implemented in scVelo `v0.2.4`
+for the pancreas and the four LARRY datasets using the same user options
+
+```python
+scvelo.tl.recover_dynamics(
+  data=adata, 
+  n_jobs=30,
+)
+scvelo.tl.velocity(
+  data=adata, 
+  mode="dynamical",
+)
+```
+
+Then, we tested a set of user options, including the neighboring cell numbers
+and the top variable gene numbers, in the pancreas dataset to explore the
+stability of the scVelo dynamical model. For the fully mature PBMC dataset, we
+followed the notebook proposed by the original authors
+<https://scvelo.readthedocs.io/perspectives/Perspectives/>, i.e., we used the
+stochastic RNA velocity model implemented in scVelo with the top three
+likelihood genes. The latent time from scVelo was computed using their provided
+function
+
+```python
+scvelo.tl.latent_time(
+  data=adata,
+)
+```
+
 ## Trajectory inference {#sec-methods-trajectory-inference}
@@
 We used the default options for projecting the vector fields from scVelo models.
-Unlike scVelo, Pyro-Velocity uses statistics derived from posterior samples of
+Unlike scVelo, Pyro&thinsp;-Velocity uses statistics derived from posterior samples of
 the denoised spliced gene expression and posterior samples of the velocity
 estimation for building the cell state transition matrix estimates using cosine
-similarity. Pyro-Velocity uses the same projection method as scVelo for
+similarity. Pyro&thinsp;-Velocity uses the same projection method as scVelo for
 projecting the transition matrix into the two-dimensional vector field on the

### 2024-08-21 src/pyrovelocity/workflows/main_configuration.py
fix(workflows): add extra large training resource requests and limits
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -94,6 +94,20 @@ large_training_resource_limits = ResourcesJSON(
     ephemeral_storage="200Gi",
 )
 
+extra_large_training_resource_requests = ResourcesJSON(
+    cpu="20",
+    mem="160Gi",
+    gpu="1",
+    ephemeral_storage="100Gi",
+)
+
+extra_large_training_resource_limits = ResourcesJSON(
+    cpu="46",
+    mem="250Gi",
+    gpu="1",
+    ephemeral_storage="400Gi",
+)
+
 default_resource_requests = ResourcesJSON(
     cpu="8",
     mem="30Gi",
@@ -667,8 +681,8 @@ larry_configuration = WorkflowConfiguration(
     training_configuration_1=larry_train_model1_args,
     training_configuration_2=larry_train_model2_args,
     postprocess_configuration=larry_postprocess_configuration,
-    training_resources_requests=large_training_resource_requests,
-    training_resources_limits=large_training_resource_limits,
+    training_resources_requests=extra_large_training_resource_requests,
+    training_resources_limits=extra_large_training_resource_limits,
     postprocessing_resources_requests=large_resource_requests,
     postprocessing_resources_limits=large_resource_limits,
     summarizing_resources_requests=large_resource_requests,

### 2024-08-21 reproducibility/manuscript/manuscript.qmd
docs(manuscript): add posterior sampling methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 toc: true
+toc-depth: 2
 toc-title: Contents
@@
 
+## Posterior prediction {#sec-methods-posterior-prediction}
+
+To benchmark Pyro&thinsp;-Velocity performance in predicting cell fate, we
+generated the posterior samples measurement
+$x_{cg}=\left(u\left(\tau^{(k_{cg})}\right),
+s\left(\tau^{(k_{cg})}\right)\right)$ or $x_{cg}=\left(u_{cg}, s_{cg} \right)$,
+$t_c$, $\beta_g$, $\gamma_g$ from the same single cell RNA-seq using $N=30$
+Monte Carlo samples from the posterior predictive distribution following
+$p(x_{cg} \vert \theta, \psi) p(\theta, \psi) \approx p(x_{cg} \vert \theta,
+\psi) q_{\phi}(\theta, \psi)$. $x_{cg}$ can be posterior samples of either
+denoised gene expression (used for phase portraits and vector field-based
+trajectory inference) or raw read counts (used for gene selection). Then we
+calculated the posterior samples of RNA velocity as
+$v_{cg}=\frac{ds(\tau^{(k_{cg})})}{d \tau^{(k{cg})}}$ based on posterior samples
+measurement of denoised gene expression $x_{cg}$ and $\beta_{g}$, $\gamma_{g}$.
+
+## Prioritization of cell fate markers {#sec-methods-fate-markers}
+
+We prioritize the cell fate markers using two metrics: First, the Pearson
+correlation between each gene's posterior mean of the denoised spliced
+expression and posterior mean of the shared time; Second, the negative mean
+absolute errors of each gene's observed spliced, and unspliced read counts with
+posterior predictive samples of spliced and unspliced raw read counts, i.e.
+$\frac{1}{N n_c} \sum_i \sum_c (x_{icg}-\tilde{X}_{cg})$, where $N$ is the
+posterior sample number that is set to $30$, $i$ is the posterior sample index,
+and $n_c$ is the cell number. We first select the top $300$ genes with the
+highest negative mean absolute errors and then rank the $300$ genes based on the
+most positively correlated genes and the least negatively correlated genes. We
+use the same strategy for scVelo to rank the markers by the model likelihood to
+get the top 300 genes and then prioritize these genes by Pearson's correlation
+between scVelo latent time and normalized expression.
+
 ## Single-cell data preprocessing {#sec-methods-preprocessing}
@@
 ```python
-scv.pp.filter_and_normalize(
+import scvelo
+
+scvelo.pp.filter_and_normalize(
   adata=adata,
@@
 )
-scv.pp.moments(
+scvelo.pp.moments(
   adata,
@@
 ```python
-scv.pp.filter_and_normalize(
+scvelo.pp.filter_and_normalize(
   adata=adata,
@@
 )
-scv.pp.moments(adata)
+scvelo.pp.moments(adata)
 ```

### 2024-08-21 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

@@
 \hypersetup{linkcolor=}
-\setcounter{tocdepth}{3}
+\setcounter{tocdepth}{2}
 \tableofcontents
@@
 
+\subsection{Posterior
+prediction}\label{sec-methods-posterior-prediction}
+
+To benchmark Pyro -Velocity performance in predicting cell fate, we
+generated the posterior samples measurement
+\(x_{cg}=\left(u\left(\tau^{(k_{cg})}\right),
+s\left(\tau^{(k_{cg})}\right)\right)\) or
+\(x_{cg}=\left(u_{cg}, s_{cg} \right)\), \(t_c\), \(\beta_g\),
+\(\gamma_g\) from the same single cell RNA-seq using \(N=30\) Monte
+Carlo samples from the posterior predictive distribution following
+\(p(x_{cg} \vert \theta, \psi) p(\theta, \psi) \approx p(x_{cg} \vert \theta,
+\psi) q_{\phi}(\theta, \psi)\). \(x_{cg}\) can be posterior samples of
+either denoised gene expression (used for phase portraits and vector
+field-based trajectory inference) or raw read counts (used for gene
+selection). Then we calculated the posterior samples of RNA velocity as
+\(v_{cg}=\frac{ds(\tau^{(k_{cg})})}{d \tau^{(k{cg})}}\) based on
+posterior samples measurement of denoised gene expression \(x_{cg}\) and
+\(\beta_{g}\), \(\gamma_{g}\).
+
+\subsection{Prioritization of cell fate
+markers}\label{sec-methods-fate-markers}
+
+We prioritize the cell fate markers using two metrics: First, the
+Pearson correlation between each gene's posterior mean of the denoised
+spliced expression and posterior mean of the shared time; Second, the
+negative mean absolute errors of each gene's observed spliced, and
+unspliced read counts with posterior predictive samples of spliced and
+unspliced raw read counts, i.e.
+\(\frac{1}{N n_c} \sum_i \sum_c (x_{icg}-\tilde{X}_{cg})\), where \(N\)
+is the posterior sample number that is set to \(30\), \(i\) is the
+posterior sample index, and \(n_c\) is the cell number. We first select
+the top \(300\) genes with the highest negative mean absolute errors and
+then rank the \(300\) genes based on the most positively correlated
+genes and the least negatively correlated genes. We use the same
+strategy for scVelo to rank the markers by the model likelihood to get
+the top 300 genes and then prioritize these genes by Pearson's
+correlation between scVelo latent time and normalized expression.
+
+\subsection{Single-cell data
+preprocessing}\label{sec-methods-preprocessing}
+
+We used scanpy and scVelo to handle the data input and output; thus,
+both h5ad and loom files generated by velocyto and kallisto
+\citep{Melsted2021-ap} are supported. The fully mature PBMC dataset was
+processed with the same procedure proposed in a review paper
+\citep{Bergen2021-qz}
+(\url{https://scvelo.readthedocs.io/perspectives/Perspectives/}). We
+reproduced this procedure using the scVelo package and raw read counts
+of the same top three dynamical genes NKG7, IGHM, and GNLY with the best
+likelihoods. The pancreas dataset was processed with scVelo using the
+following options
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ImportTok{import}\NormalTok{ scvelo}
+
+\NormalTok{scvelo.pp.filter\_and\_normalize(}
+\NormalTok{  adata}\OperatorTok{=}\NormalTok{adata,}
+\NormalTok{  min\_shared\_counts}\OperatorTok{=}\DecValTok{30}\NormalTok{, }
+\NormalTok{  n\_top\_genes}\OperatorTok{=}\DecValTok{2000}\NormalTok{,}
+\NormalTok{)}
+\NormalTok{scvelo.pp.moments(}
+\NormalTok{  adata,}
+\NormalTok{  n\_pcs}\OperatorTok{=}\DecValTok{30}\NormalTok{,}
+\NormalTok{  n\_neighbors}\OperatorTok{=}\DecValTok{30}\NormalTok{,}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+The same top variable genes with raw spliced and unspliced read counts
+were used as input for the Pyro -Velocity model. The original LARRY
+dataset of in vitro Hematopoiesis containing \(130,887\) cells was first
+filtered to remove cells without LARRY barcoding. \(49,302\) cells were
+recovered after this step with at least one LARRY barcode. For
+simplicity, we termed this filtered dataset with multiple cell fate
+(multi-fate) as the full dataset. Based on this dataset, we created two
+datasets with uni-fate progression toward monocyte or neutrophil based
+on the lineage LARRY barcodes and time information. Namely, we selected
+sets of cells with a single LARRY barcode, spanning three time points
+(day 2, 4, 6), and all the cells from the last time point (day 6) belong
+to a unique cell type (either monocyte or neutrophil). The two uni-fate
+datasets were combined to represent the bi-fate LARRY dataset. The
+multi-fate full dataset was processed using the same options as the
+pancreas dataset; the rest of the uni-fate and bi-fate datasets were
+processed using the following parameters
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{scvelo.pp.filter\_and\_normalize(}
+\NormalTok{  adata}\OperatorTok{=}\NormalTok{adata,}
+\NormalTok{  n\_top\_genes}\OperatorTok{=}\DecValTok{2000}\NormalTok{,}
+\NormalTok{  min\_shared\_counts}\OperatorTok{=}\DecValTok{20}\NormalTok{,}
+\NormalTok{)}
+\NormalTok{scvelo.pp.moments(adata)}
+\end{Highlighting}
+\end{Shaded}
+
+\subsection{scVelo model}\label{sec-methods-scvelo}
+
+We benchmarked the dynamical RNA velocity model implemented in scVelo
+\texttt{v0.2.4} for the pancreas and the four LARRY datasets using the
+same user options
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{scvelo.tl.recover\_dynamics(}
+\NormalTok{  data}\OperatorTok{=}\NormalTok{adata, }
+\NormalTok{  n\_jobs}\OperatorTok{=}\DecValTok{30}\NormalTok{,}
+\NormalTok{)}
+\NormalTok{scvelo.tl.velocity(}
+\NormalTok{  data}\OperatorTok{=}\NormalTok{adata, }
+\NormalTok{  mode}\OperatorTok{=}\StringTok{"dynamical"}\NormalTok{,}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Then, we tested a set of user options, including the neighboring cell
+numbers and the top variable gene numbers, in the pancreas dataset to
+explore the stability of the scVelo dynamical model. For the fully
+mature PBMC dataset, we followed the notebook proposed by the original
+authors \url{https://scvelo.readthedocs.io/perspectives/Perspectives/},
+i.e., we used the stochastic RNA velocity model implemented in scVelo
+with the top three likelihood genes. The latent time from scVelo was
+computed using their provided function
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{scvelo.tl.latent\_time(}
+\NormalTok{  data}\OperatorTok{=}\NormalTok{adata,}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\subsection{Trajectory
+inference}\label{sec-methods-trajectory-inference}
+
+\subsubsection{Velocity vector field}\label{velocity-vector-field}
+
+The velocity-based vector fields were generated using the scVelo
+function call
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{scvelo.tl.velocity\_graph(adata)}
+\NormalTok{scvelo.tl.velocity\_embedding(}
+\NormalTok{  adata, }
+\NormalTok{  vkey}\OperatorTok{=}\StringTok{\textquotesingle{}velocity\textquotesingle{}}\NormalTok{,}
+\NormalTok{  basis}\OperatorTok{=}\StringTok{\textquotesingle{}embed\textquotesingle{}}\NormalTok{,}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+We used the default options for projecting the vector fields from scVelo
+models. Unlike scVelo, Pyro -Velocity uses statistics derived from
+posterior samples of the denoised spliced gene expression and posterior
+samples of the velocity estimation for building the cell state
+transition matrix estimates using cosine similarity. Pyro -Velocity uses
+the same projection method as scVelo for projecting the transition
+matrix into the two-dimensional vector field on the user-provided
+embedding space.
+
+\subsubsection{Clonal progression vector
+field}\label{clonal-progression-vector-field}
+
+No method exists for projecting clonal cell fates from the LARRY data
+and associated subsets thereof into a vector field. Thus, we designed a
+strategy for projecting clonal progression vector fields. Given cells
+sharing the same LARRY barcode, we considered the vectors connecting the
+embedding centroid of cells at consecutive time points and then
+leveraged Gaussian smoothing of nearest centroid directions to represent
+the clonal progression vector field informed by LARRY barcodes on a
+regular grid over the cell manifold.
+
 \subsection{Cell fate uncertainty

### 2024-08-21 reproducibility/manuscript/manuscript.qmd
docs(manuscript): use v2 cached Fig. 1 and Ext Fig. 1
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
fix(manuscript): use bold panel references
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
     include-in-header: header.tex
+    citecolor: NavyBlue
+    linkcolor: NavyBlue
+    toccolor: NavyBlue
+    urlcolor: NavyBlue
     affiliations:
@@
 correlation between inferred shared time and corresponding denoised spliced
-expression per cell ([@fig-model-results]d, Methods). This procedure
+expression per cell ([@fig-model-results]d, [Methods](#sec-methods)). This procedure
 successfully recovered key developmental genes that are activated (or repressed)
@@
 ::: {#fig-model}
-![](figures/fig.pdf)
+![](figures/v2/Figure1.pdf)
 
@@
 estimation of velocity vector fields and shared latent time, based on raw counts
-and without ad-hoc preprocessing steps. a. Schematic comparing current RNA
+and without ad-hoc preprocessing steps. **a**, Schematic comparing current RNA
 Velocity and Pyro&thinsp;-Velocity workflows. Both methods take in input spliced
@@
 aggregation steps. It also provides uncertainty estimation based on posterior
-samples. b. The main outputs of the Pyro&thinsp;-Velocity framework.
+samples. **b**, The main outputs of the Pyro&thinsp;-Velocity framework.
 Visualization of single gene expression across inferred developmental cell order
@@
 Pyro&thinsp;-Velocity outperforms other RNA-velocity methods and accurately
-predicts cell lineage trajectories. a,b. RNA Velocity model comparisons from
+predicts cell lineage trajectories. **a,b**, RNA Velocity model comparisons from
 fully mature peripheral blood mononuclear cells (PBMCs, t-SNE embedded plots, a)
-or E15.5 mouse pancreas (UMAP embedded plots, b-c).  From left to right in
-panels a and b: 1. Cells colored by cell type; 2. Stream plot of velocity vector
+or E15.5 mouse pancreas (UMAP embedded plots, **b-c**).  From left to right in
+panels **a** and **b**: 1. Cells colored by cell type; 2. Stream plot of velocity vector
 field defined by scVelo; 3. Stream plot of the posterior mean vector field from
@@
 were projected into the UMAP space using the scVelo API based on cosine
-similarity-derived transition matrix. c. UMAP embedded cells colored by the
+similarity-derived transition matrix. **c**, UMAP embedded cells colored by the
 average cell shared time across 30 posterior samples. The Spearman's correlation
-between Pyro&thinsp;-Velocity and Cytotrace is provided. d.
+between Pyro&thinsp;-Velocity and Cytotrace is provided. **d**,
 Pyro&thinsp;-Velocity automatic gene selection and ranking for predicting
-important genes involved in pancreas development. e. Phase portraits of
+important genes involved in pancreas development. **e**, Phase portraits of
 posterior samples average spliced and unspliced gene expression (left), rainbow
@@
 right) or vertical lines (middle). Cell type annotation coloring is the same as
-in panel b.
+in **b**.
 :::
@@
 trajectory in human LARRY barcoded hematopoietic cells. Systematic benchmarking
-of cell fate prediction methods - predicting uni-lineage cell fates (a-b, Model
-1), bifurcated lineages (c, Model 2), or all lineages together (d, Model 2).
+of cell fate prediction methods - predicting uni-lineage cell fates (**a-b**, Model
+1), bifurcated lineages (**c**, Model 2), or all lineages together (**d**, Model 2).
 Panels from left to right show on the same SPRING embedding: 1. Cell types; 2.
 Clonal progression vector field derived based on the observed clonal barcodes
-(Methods); 3. scVelo recovered vector field; 4. Pyro&thinsp;-Velocity single
+([Methods](#sec-methods)); 3. scVelo recovered vector field; 4. Pyro&thinsp;-Velocity single
 cell level vector field uncertainty (angular standard deviation from 30
@@
 ::: {#extfig-graphical-models}
-![](figures/fig.pdf)
+![](figures/v2/Supplementary-Figure1.pdf)
 
 Graphical representations detail the generative process of the
-Pyro&thinsp;-Velocity models and highlight their differences. a.
+Pyro&thinsp;-Velocity models and highlight their differences. **a**,
 Pyro&thinsp;-Velocity Model 1. Each rectangle represents one data dimension
@@
 $(u_{0}=0,\,s_{0}=0)$, as in previous methods[@La_Manno2018-lj;@Bergen2020-pj]
-(Methods). b. Pyro&thinsp;-Velocity Model 2. This figure follows the same
-graphical convention as in a. In this model $t_{0}^{(0)}$ is independent from
+([Methods](#sec-methods)). **b**, Pyro&thinsp;-Velocity Model 2. This figure follows the same
+graphical convention as in **a**, In this model $t_{0}^{(0)}$ is independent from
 the shared time, thus allowing a time lag for the gene activation phase. In
 addition, a positive starting basal gene expression level $(u_{0},\,s_{0})$
-different than $(u_{0}=0,\,s_{0}=0)$ is allowed. (Methods). c. Schematic
+different than $(u_{0}=0,\,s_{0}=0)$ is allowed. ([Methods](#sec-methods)). **c**, Schematic
 illustrating how the constraints proposed in Model 1 (left) and Model 2 (right)
@@
 Extended version of @fig-model-results for evaluating Pyro&thinsp;-Velocity
-(Model 1) and scVelo on the PBMC and pancreas datasets. a. Left figure shows the
+(Model 1) and scVelo on the PBMC and pancreas datasets. **a**, Left figure shows the
 scVelo vector field using the default user options, right figure shows the
 evaluation of the scVelo vector field using different combinations of user
-options within Epsilon cells. b. Pyro&thinsp;-Velocity additional single cell
+options within Epsilon cells. **b**, Pyro&thinsp;-Velocity additional single cell
 level uncertainty diagnostic plots on the fully mature PBMC dataset, from left
@@
 transcriptional state uncertainty of predicted raw read count per cell. The
-contour lines show the top 10% of the most uncertain cells. c. For the pancreas
+contour lines show the top 10% of the most uncertain cells. **c**, For the pancreas
 dataset, the figures show from left to right: UMAP rendering of 1. Cytotrace, 2.
@@
 averaged across 30 posterior samples, 4. Top 50 markers overlapping between
-Pyro&thinsp;-Velocity and scVelo (Methods, Supplementary Table 1-2), 5. The same
-uncertainty diagnostic plots per cell as in b. d. The same plot as Fig. 2d. e.
+Pyro&thinsp;-Velocity and scVelo ([Methods](#sec-methods), Supplementary Table 1-2), 5. The same
+uncertainty diagnostic plots per cell as in **b**. **d**, The same plot as [@fig-model-results]d, e.
 Selection of the top negatively correlated genes with the cell shared time based
 on the posterior mean on denoised spliced expression and negative mean absolute
-error of gene expression prediction (Methods). f. Phase portraits, rainbow
+error of gene expression prediction ([Methods](#sec-methods)). **f**, Phase portraits, rainbow
 plots, and UMAP rendering of posterior mean of gene expression for the genes
-selected in e.
+selected in **e**.
 :::
@@
 data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
-scRNA-seq datasets with lineage barcoding information (LARRY dataset). a. SPRING
+scRNA-seq datasets with lineage barcoding information (LARRY dataset). **a**, SPRING
 embedding rendering of Cytotrace scores, the title reports the Spearman's
-correlation between the Cytotrace and Cospar fate potency scores, b.
+correlation between the Cytotrace and Cospar fate potency scores, **b**,
 Pyro&thinsp;-Velocity additional single cell level uncertainty diagnostic plots
@@
 state uncertainty of predicted raw read count per cell. The contour lines show
-the top 10% of the most uncertain cells (Methods), c. Pairwise Spearman's
+the top 10% of the most uncertain cells ([Methods](#sec-methods)), **c**, Pairwise Spearman's
 correlation between Cytotrace/Cospar fate potency scores and velocity methods
 for the four different benchmarking datasets selected based on the lineage
-barcoding information from LARRY data. d. SPRING embedding rendering of
+barcoding information from LARRY data, **d**, SPRING embedding rendering of
 posterior mean of shared time across 30 posterior samples, the title shows the
-Spearman's correlation of mean shared time with Cytotrace. e. Selection of the
+Spearman's correlation of mean shared time with Cytotrace. **e**, Selection of the
 top positively correlated genes with the cell shared time based on the posterior
-mean of spliced expression. f. Phase portraits, rainbow plots, and UMAP
-rendering with splicing gene expression level for the selected genes in e.
+mean of spliced expression. **f**, Phase portraits, rainbow plots, and UMAP
+rendering with splicing gene expression level for the selected genes in **e**.
 :::
@@
 Pyro&thinsp;-Velocity and scVelo on scRNA-seq datasets with lineage barcoding
-information (LARRY dataset). a. uni-fate Monocyte lineage analysis with
-Pyro&thinsp;-Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
-Pyro&thinsp;-Velocity Model 2, c. Neutrophil and Monocyte bi-fate lineage
-analysis with Pyro&thinsp;-Velocity Model 1, and d. multi-fate lineage analysis
+information (LARRY dataset). **a**, uni-fate Monocyte lineage analysis with
+Pyro&thinsp;-Velocity Model 2, **b**, uni-fate Neutrophil lineage analysis with
+Pyro&thinsp;-Velocity Model 2, **c**, Neutrophil and Monocyte bi-fate lineage
+analysis with Pyro&thinsp;-Velocity Model 1, and **d**, multi-fate lineage analysis
 with Pyro&thinsp;-Velocity Model 1. The panel descriptions are the same as in

### 2024-08-21 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

@@
 \usepackage[title]{appendix}%
-\usepackage{xcolor}%
+\usepackage[dvipsnames]{xcolor}%
 \usepackage{textcomp}%
@@
 \usepackage{placeins}
+\usepackage[labelfont=bf]{caption}
+% \usepackage{showframe}
+% \usepackage{layout}
 \allowdisplaybreaks
@@
   colorlinks=true,
-  linkcolor={blue},
+  linkcolor={NavyBlue},
   filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
+  citecolor={NavyBlue},
+  urlcolor={NavyBlue},
   pdfcreator={LaTeX via pandoc}}
@@
 \maketitle
+% \clearpage
+% \layout
 \newpage{}
@@
 {
-\hypersetup{linkcolor=}
+\hypersetup{linkcolor=NavyBlue}
 \setcounter{tocdepth}{2}
@@
 and corresponding denoised spliced expression per cell
-(Fig.~\ref{fig-model-results}d, Methods). This procedure successfully
-recovered key developmental genes that are activated (or repressed)
-during endocrinogenesis (Fig.~\ref{fig-model-results}d-e,
+(Fig.~\ref{fig-model-results}d, \hyperref[sec-methods]{Methods}). This
+procedure successfully recovered key developmental genes that are
+activated (or repressed) during endocrinogenesis
+(Fig.~\ref{fig-model-results}d-e,
 Extended Data Fig.~\ref{extfig-model-results}). Among the top positively
@@
 
-\includegraphics{figures/fig.pdf}
+\includegraphics{figures/v2/Figure1.pdf}
 
@@
 latent time, based on raw counts and without ad-hoc preprocessing steps.
-a. Schematic comparing current RNA Velocity and Pyro -Velocity
+\textbf{a}, Schematic comparing current RNA Velocity and Pyro -Velocity
 workflows. Both methods take in input spliced and unspliced read count
@@
 steps. It also provides uncertainty estimation based on posterior
-samples. b. The main outputs of the Pyro -Velocity framework.
+samples. \textbf{b}, The main outputs of the Pyro -Velocity framework.
 Visualization of single gene expression across inferred developmental
@@
 RNA-velocity methods and accurately predicts cell lineage trajectories.
-a,b. RNA Velocity model comparisons from fully mature peripheral blood
-mononuclear cells (PBMCs, t-SNE embedded plots, a) or E15.5 mouse
-pancreas (UMAP embedded plots, b-c). From left to right in panels a and
-b: 1. Cells colored by cell type; 2. Stream plot of velocity vector
-field defined by scVelo; 3. Stream plot of the posterior mean vector
-field from Pyro -Velocity; 4. Pyro -Velocity visualization of cell
-dynamics for 3 arbitrarily selected single cells; 5. Single-cell vector
-field uncertainty where each cell is colored based on the scaled angular
-standard deviation to visualize uncertainty based on n=30 iterated,
-posterior samples depicted as arrows noted throughout our analysis; 6.
-Averaged vector field uncertainty where each vector is colored by the
-averaged vector field uncertainty. The scaled angular standard deviation
-range is {[}0 360{]}, where 0 corresponds to the highest confidence for
-a single direction, 360 corresponds to the highest uncertainty, and no
-directionality preference based on 30 posterior samples (4-6). Vectors
-were projected into the UMAP space using the scVelo API based on cosine
-similarity-derived transition matrix. c.~UMAP embedded cells colored by
-the average cell shared time across 30 posterior samples. The Spearman's
-correlation between Pyro -Velocity and Cytotrace is provided. d.
-Pyro -Velocity automatic gene selection and ranking for predicting
-important genes involved in pancreas development. e. Phase portraits of
-posterior samples average spliced and unspliced gene expression (left),
-rainbow plots showing denoised spliced gene expression (middle), and
-gene expression within UMAP embedded space. Single dots correspond to
-individual cells (left, right) or vertical lines (middle). Cell type
-annotation coloring is the same as in panel b.}
+\textbf{a,b}, RNA Velocity model comparisons from fully mature
+peripheral blood mononuclear cells (PBMCs, t-SNE embedded plots, a) or
+E15.5 mouse pancreas (UMAP embedded plots, \textbf{b-c}). From left to
+right in panels \textbf{a} and \textbf{b}: 1. Cells colored by cell
+type; 2. Stream plot of velocity vector field defined by scVelo; 3.
+Stream plot of the posterior mean vector field from Pyro -Velocity; 4.
+Pyro -Velocity visualization of cell dynamics for 3 arbitrarily selected
+single cells; 5. Single-cell vector field uncertainty where each cell is
+colored based on the scaled angular standard deviation to visualize
+uncertainty based on n=30 iterated, posterior samples depicted as arrows
+noted throughout our analysis; 6. Averaged vector field uncertainty
+where each vector is colored by the averaged vector field uncertainty.
+The scaled angular standard deviation range is {[}0 360{]}, where 0
+corresponds to the highest confidence for a single direction, 360
+corresponds to the highest uncertainty, and no directionality preference
+based on 30 posterior samples (4-6). Vectors were projected into the
+UMAP space using the scVelo API based on cosine similarity-derived
+transition matrix. \textbf{c}, UMAP embedded cells colored by the
+average cell shared time across 30 posterior samples. The Spearman's
+correlation between Pyro -Velocity and Cytotrace is provided.
+\textbf{d}, Pyro -Velocity automatic gene selection and ranking for
+predicting important genes involved in pancreas development. \textbf{e},
+Phase portraits of posterior samples average spliced and unspliced gene
+expression (left), rainbow plots showing denoised spliced gene
+expression (middle), and gene expression within UMAP embedded space.
+Single dots correspond to individual cells (left, right) or vertical
+lines (middle). Cell type annotation coloring is the same as in
+\textbf{b}.}
 
@@
 hematopoietic cells. Systematic benchmarking of cell fate prediction
-methods - predicting uni-lineage cell fates (a-b, Model 1), bifurcated
-lineages (c, Model 2), or all lineages together (d, Model 2). Panels
-from left to right show on the same SPRING embedding: 1. Cell types; 2.
-Clonal progression vector field derived based on the observed clonal
-barcodes (Methods); 3. scVelo recovered vector field; 4. Pyro -Velocity
-single cell level vector field uncertainty (angular standard deviation
-from 30 posterior samples); 5. Pyro -Velocity recovered vector field;
-vectors are colored by the averaged vector field uncertainty; 6. Clonal
-fate potency score using Cospar \citep{Wang2022-xb}; 7. scVelo recovered
-latent time; 8. Pyro -Velocity inference of posterior mean shared time
-per cell. Cosine similarity and Spearman's correlation for the different
-methods with the clonal progression and fate potency are reported in the
-titles.}
+methods - predicting uni-lineage cell fates (\textbf{a-b}, Model 1),
+bifurcated lineages (\textbf{c}, Model 2), or all lineages together
+(\textbf{d}, Model 2). Panels from left to right show on the same SPRING
+embedding: 1. Cell types; 2. Clonal progression vector field derived
+based on the observed clonal barcodes (\hyperref[sec-methods]{Methods});
+3. scVelo recovered vector field; 4. Pyro -Velocity single cell level
+vector field uncertainty (angular standard deviation from 30 posterior
+samples); 5. Pyro -Velocity recovered vector field; vectors are colored
+by the averaged vector field uncertainty; 6. Clonal fate potency score
+using Cospar \citep{Wang2022-xb}; 7. scVelo recovered latent time; 8.
+Pyro -Velocity inference of posterior mean shared time per cell. Cosine
+similarity and Spearman's correlation for the different methods with the
+clonal progression and fate potency are reported in the titles.}
 
@@
 
-\includegraphics{figures/fig.pdf}
+\includegraphics{figures/v2/Supplementary-Figure1.pdf}
 
@@
 the generative process of the Pyro -Velocity models and highlight their
-differences. a. Pyro -Velocity Model 1. Each rectangle represents one
-data dimension (e.g., gene or cell). The arrows represent the directed
-hierarchical dependency among random variables. The white circles
-correspond to the modeled random variables. The probability
+differences. \textbf{a}, Pyro -Velocity Model 1. Each rectangle
+represents one data dimension (e.g., gene or cell). The arrows represent
+the directed hierarchical dependency among random variables. The white
+circles correspond to the modeled random variables. The probability
 distributions of these variables are denoted on the right legend. The
@@
 with the initial condition \((u_{0}=0,\,s_{0}=0)\), as in previous
-methods\citep{La_Manno2018-lj, Bergen2020-pj} (Methods). b.
-Pyro -Velocity Model 2. This figure follows the same graphical
-convention as in a. In this model \(t_{0}^{(0)}\) is independent from
-the shared time, thus allowing a time lag for the gene activation phase.
-In addition, a positive starting basal gene expression level
-\((u_{0},\,s_{0})\) different than \((u_{0}=0,\,s_{0}=0)\) is allowed.
-(Methods). c.~Schematic illustrating how the constraints proposed in
-Model 1 (left) and Model 2 (right) are reflected in the modeling of
-individual gene expression. The x-axis corresponds to the shared time,
-the y-axis corresponds to the posterior mean of the spliced expression
-level, the vertical lines with arrows show the start and end time of the
-observed cell sampling, t0 represents the gene-specific activation time
-\(t_{0}^{(0)}\), the red lines highlight the differences for the
-potential gene-specific time lag when \(t_{0}^{(0)} > {\min}\!⁡(t)\), and
-the initial spliced gene expression. The same differences can also be
-illustrated for the unspliced gene expression profiles (not shown).}
+methods\citep{La_Manno2018-lj, Bergen2020-pj}
+(\hyperref[sec-methods]{Methods}). \textbf{b}, Pyro -Velocity Model 2.
+This figure follows the same graphical convention as in \textbf{a}, In
+this model \(t_{0}^{(0)}\) is independent from the shared time, thus
+allowing a time lag for the gene activation phase. In addition, a
+positive starting basal gene expression level \((u_{0},\,s_{0})\)
+different than \((u_{0}=0,\,s_{0}=0)\) is allowed.
+(\hyperref[sec-methods]{Methods}). \textbf{c}, Schematic illustrating
+how the constraints proposed in Model 1 (left) and Model 2 (right) are
+reflected in the modeling of individual gene expression. The x-axis
+corresponds to the shared time, the y-axis corresponds to the posterior
+mean of the spliced expression level, the vertical lines with arrows
+show the start and end time of the observed cell sampling, t0 represents
+the gene-specific activation time \(t_{0}^{(0)}\), the red lines
+highlight the differences for the potential gene-specific time lag when
+\(t_{0}^{(0)} > {\min}\!⁡(t)\), and the initial spliced gene expression.
+The same differences can also be illustrated for the unspliced gene
+expression profiles (not shown).}
 
@@
 Fig.~\ref{fig-model-results} for evaluating Pyro -Velocity (Model 1) and
-scVelo on the PBMC and pancreas datasets. a. Left figure shows the
-scVelo vector field using the default user options, right figure shows
-the evaluation of the scVelo vector field using different combinations
-of user options within Epsilon cells. b. Pyro -Velocity additional
-single cell level uncertainty diagnostic plots on the fully mature PBMC
-dataset, from left to right: 1. standard deviation of posterior samples
-of shared time, 2. Rayleigh test of posterior samples of the vector
-field (the title reports false discovery rate after p-value correction
-for multiple tests across cells), 3. transcriptional state uncertainty
-of predicted raw read count per cell. The contour lines show the top
-10\% of the most uncertain cells. c.~For the pancreas dataset, the
-figures show from left to right: UMAP rendering of 1. Cytotrace, 2.
-scVelo latent time, 3. Posterior mean of Pyro -Velocity shared time
-averaged across 30 posterior samples, 4. Top 50 markers overlapping
-between Pyro -Velocity and scVelo (Methods, Supplementary Table 1-2), 5.
-The same uncertainty diagnostic plots per cell as in b. d.~The same plot
-as Fig. 2d. e. Selection of the top negatively correlated genes with the
-cell shared time based on the posterior mean on denoised spliced
-expression and negative mean absolute error of gene expression
-prediction (Methods). f.~Phase portraits, rainbow plots, and UMAP
-rendering of posterior mean of gene expression for the genes selected in
-e.}
+scVelo on the PBMC and pancreas datasets. \textbf{a}, Left figure shows
+the scVelo vector field using the default user options, right figure
+shows the evaluation of the scVelo vector field using different
+combinations of user options within Epsilon cells. \textbf{b},
+Pyro -Velocity additional single cell level uncertainty diagnostic plots
+on the fully mature PBMC dataset, from left to right: 1. standard
+deviation of posterior samples of shared time, 2. Rayleigh test of
+posterior samples of the vector field (the title reports false discovery
+rate after p-value correction for multiple tests across cells), 3.
+transcriptional state uncertainty of predicted raw read count per cell.
+The contour lines show the top 10\% of the most uncertain cells.
+\textbf{c}, For the pancreas dataset, the figures show from left to
+right: UMAP rendering of 1. Cytotrace, 2. scVelo latent time, 3.
+Posterior mean of Pyro -Velocity shared time averaged across 30
+posterior samples, 4. Top 50 markers overlapping between Pyro -Velocity
+and scVelo (\hyperref[sec-methods]{Methods}, Supplementary Table 1-2),
+5. The same uncertainty diagnostic plots per cell as in \textbf{b}.
+\textbf{d}, The same plot as Fig.~\ref{fig-model-results}d, e. Selection
+of the top negatively correlated genes with the cell shared time based
+on the posterior mean on denoised spliced expression and negative mean
+absolute error of gene expression prediction
+(\hyperref[sec-methods]{Methods}). \textbf{f}, Phase portraits, rainbow
+plots, and UMAP rendering of posterior mean of gene expression for the
+genes selected in \textbf{e}.}
 
@@
 scRNA-seq datasets with lineage barcoding information (LARRY dataset).
-a. SPRING embedding rendering of Cytotrace scores, the title reports the
-Spearman's correlation between the Cytotrace and Cospar fate potency
-scores, b. Pyro -Velocity additional single cell level uncertainty
-diagnostic plots based on the multi-fate LARRY dataset, from left to
-right: 1. Standard deviation of posterior samples of shared time per
-cell, 2. Rayleigh test of posterior samples of vector field per cell
+\textbf{a}, SPRING embedding rendering of Cytotrace scores, the title
+reports the Spearman's correlation between the Cytotrace and Cospar fate
+potency scores, \textbf{b}, Pyro -Velocity additional single cell level
+uncertainty diagnostic plots based on the multi-fate LARRY dataset, from
+left to right: 1. Standard deviation of posterior samples of shared time
+per cell, 2. Rayleigh test of posterior samples of vector field per cell
 (the title reports the false discovery rate after p-value correction for
@@
 predicted raw read count per cell. The contour lines show the top 10\%
-of the most uncertain cells (Methods), c.~Pairwise Spearman's
-correlation between Cytotrace/Cospar fate potency scores and velocity
-methods for the four different benchmarking datasets selected based on
-the lineage barcoding information from LARRY data. d.~SPRING embedding
-rendering of posterior mean of shared time across 30 posterior samples,
-the title shows the Spearman's correlation of mean shared time with
-Cytotrace. e. Selection of the top positively correlated genes with the
+of the most uncertain cells (\hyperref[sec-methods]{Methods}),
+\textbf{c}, Pairwise Spearman's correlation between Cytotrace/Cospar
+fate potency scores and velocity methods for the four different
+benchmarking datasets selected based on the lineage barcoding
+information from LARRY data, \textbf{d}, SPRING embedding rendering of
+posterior mean of shared time across 30 posterior samples, the title
+shows the Spearman's correlation of mean shared time with Cytotrace.
+\textbf{e}, Selection of the top positively correlated genes with the
 cell shared time based on the posterior mean of spliced expression.
-f.~Phase portraits, rainbow plots, and UMAP rendering with splicing gene
-expression level for the selected genes in e.}
+\textbf{f}, Phase portraits, rainbow plots, and UMAP rendering with
+splicing gene expression level for the selected genes in \textbf{e}.}
 
@@
 Pyro -Velocity and scVelo on scRNA-seq datasets with lineage barcoding
-information (LARRY dataset). a. uni-fate Monocyte lineage analysis with
-Pyro -Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
-Pyro -Velocity Model 2, c.~Neutrophil and Monocyte bi-fate lineage
-analysis with Pyro -Velocity Model 1, and d.~multi-fate lineage analysis
-with Pyro -Velocity Model 1. The panel descriptions are the same as in
-Fig.~\ref{fig-lineage-tracing}.}
+information (LARRY dataset). \textbf{a}, uni-fate Monocyte lineage
+analysis with Pyro -Velocity Model 2, \textbf{b}, uni-fate Neutrophil
+lineage analysis with Pyro -Velocity Model 2, \textbf{c}, Neutrophil and
+Monocyte bi-fate lineage analysis with Pyro -Velocity Model 1, and
+\textbf{d}, multi-fate lineage analysis with Pyro -Velocity Model 1. The
+panel descriptions are the same as in Fig.~\ref{fig-lineage-tracing}.}
 

### 2024-08-22 reproducibility/manuscript/manuscript.qmd
docs(manuscript): use v2 cached figures
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ::: {#fig-model-results}
-![](figures/fig.pdf)
+![](figures/v2/Figure2.png)
 
@@
 ::: {#fig-lineage-tracing}
-![](figures/fig.pdf)
+![](figures/v2/Figure3.png)
 
@@
 ::: {#extfig-model-results}
-![](figures/fig.pdf)
+![](figures/v2/Supplementary-Figure2.png)
 
@@
 ::: {#extfig-model-2-result}
-![](figures/fig.pdf)
+![](figures/v2/Supplementary-Figure3.png)
 
@@
 ::: {#extfig-lineage-tracing}
-![](figures/fig.pdf)
+![](figures/v2/Supplementary-Figure4.png)
 
@@
 ::: {#extfig-lineage-tracing-2}
-![](figures/fig.pdf)
+![](figures/v2/Supplementary-Figure5.png)
 

### 2024-08-22 reproducibility/manuscript/manuscript.qmd
docs(manuscript): update methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 abstract: |
-  Single-cell RNA Velocity has dramatically advanced our ability to model
+  Single-cell RNA velocity has dramatically advanced our ability to model
   cellular differentiation and cell fate decisions. However, current
@@
 
-RNA Velocity is a powerful computational framework to estimate the time
+RNA velocity is a powerful computational framework to estimate the time
 derivative of gene expression
@@
 \frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
+  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right),
+   \label{eq-dudt}\\
 \frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
+  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)
+  -\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
 \end{align}
+In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
+$\left( u\left( \tau^{(k_{cg})} \right), s\left( \tau^{(k_{cg})} \right) \right)$ 
+are the unspliced and spliced expression functions given the
+change of time per cell and gene. $\tau_{cg}$ represents the displacement of time per
+cell and gene with 
+\begin{align}
+ \tau^{(k_{cg})} &= \operatorname{softplus} \left( t_{c} - {t_{0}^{(k_{cg})}}_g \right) \\
+ & = \log( 1 + \exp (t_c - {t_{0}^{(k_{cg})}}_g)), 
+\end{align}
+in which $t_c$ is the shared time per cell, ${t_{0}^{(kcg)}}_g$ is the
+gene-specific switching time. Each cell and gene combination has its
+transcriptional state $k_{cg} \in \{ 0, 1 \}$, where $0$ indicates the
+activation state and $1$ indicates the expression state. Each gene has two
+switching times for representing activation and repression: ${t_{0}^{(0)}}_g$ is
+the first switching time corresponding to when the gene expression starts to be
+activated, ${t_0^{(1)}}_g$  is the second switching time corresponding to when
+the gene expression starts to be repressed. We note that $\alpha^{(1)}$ is
+shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
+each gene.
+
+The analytic solution of the differential equations to predict
+spliced and unspliced gene expression given their parameters is derived by the
+authors of scVelo and a theoretical RNA velocity study
+[@Bergen2020-pj;@Li2021-qa] and given in
+Eqs. \ref{eq-solution-u}-\ref{eq-solution-s2}.
+\begin{align}
+u\left(\tau^{\left(k_{c g}\right)}\right) 
+  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
+  \nonumber \\
+&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
+  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
+  \label{eq-solution-u}\\
+s\left(\tau^{\left(k_{c g}\right)}\right) 
+  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
+  \nonumber \\
+  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
+    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
+    \nonumber\\
+  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
+    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
+    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
+    \nonumber \\
+  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
+s\left(\tau^{\left(k_{c g}\right)}\right) 
+  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
+  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
+  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
+  \nonumber \\
+  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
+  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
+  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
+  &\qquad \beta = \gamma \label{eq-solution-s2}
+\end{align}
+
+To simplify these equations, consider the case when $k_{cg} = 0$ and 
+$\beta_g \neq \gamma_g$. Then, 
+\begin{align}
+u\left(\tau^{(0)}\right) &= {u_0^{(0)}}_g e^{-\beta_g \tau^{(0)}} \nonumber \\
+  & \hskip -24pt + \frac{{\alpha^{(0)}}_g}{\beta_g}\left(1-e^{-\beta_g \tau^{(0)}}\right), 
+  \label{eq-sol-usimp} \\
+s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
+  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}\left(1-e^{-\gamma_g \tau^{(0)}}\right) 
+  \nonumber\\
+  & \hskip -24pt +\frac{{\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g}{\gamma_g-\beta_g}
+  \left(e^{-\gamma_g \tau^{(0)}}-e^{-\beta_g \tau^{(0)}}\right). \label{eq-sol-ssimp}
+\end{align}
+When $k_{cg} = 0$ and $\beta_g = \gamma_g$, then $u\left(\tau^{(0)}\right)$ 
+has the same solution, and $s\left(\tau^{(0)}\right)$ becomes
+\begin{align}
+s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
+  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}
+    \left(1-e^{-\gamma_g \tau^{(0)}}\right) \nonumber\\
+  & \hskip -24pt - \left( {\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g \right) 
+    \tau^{(0)} e^{-\beta_g \tau^{(0)}}. \label{eq-sol-ssimp2}
+\end{align}
+When $k_{cg} = 1$ and $\beta_g \neq \gamma_g$, then
+\begin{align}
+u\left(\tau^{(1)}\right) &=u_0^{(1)} g^{e^{-\beta_g \tau^{(1)}}}, \\
+s\left(\tau^{(1)}\right) &=s_0^{(1)} e^{-\gamma_g \tau^{(1)}} \nonumber \\
+  & \hskip -24pt +\frac{-\beta_g u_0^{(1)}}{\gamma_g-\beta_g}
+  \left(e^{-\gamma_g \tau^{(1)}}-e^{-\beta_g \tau^{(1)}}\right).
+\end{align}
+When $k_{cg} = 1$ and $\beta_g = \gamma_g$, then $u\left(\tau^{(1)}\right)$ 
+has the same solution, and $s\left(\tau^{(1)}\right)$ becomes
+\begin{align}
+s\left(\tau^{(1)}\right)=s_0^{(1)}{ }_g e^{-\gamma_g \tau^{(1)}}
+  +\beta_g u_0^{(1)}{ }_g \tau^{(1)} e^{-\beta_g \tau^{(1)}}.
+\end{align}
+
+## Variational inference {#sec-methods-inference}
+
+Given observations $\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)$,
+we would like to compute the posterior distribution over the random variables
+\begin{align}
+\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
+\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
+\end{align}
+but exact Bayesian inference is intractable in this model. We use Pyro to
+automatically integrate out the local discrete latent variables $k$, which is
+defined as the cell and gene transcriptional state (see above), and approximate
+the posterior over the remaining latent variables using variational inference
+[@Bingham2018-id;@Kucukelbir2016-bk], which converts intractable integrals into
+optimization problems that can be solved with off-the-shelf tools. In
+variational inference, we maximize a tractable bound (the evidence lower bound
+(ELBO)) on the Kullback-Leibler (KL) divergence
+$$\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} p(\theta, \psi \vert \tilde{X}_{cg}) \right)$$
+between a parametric family of tractable probability distributions q and the
+intractable true posterior
+\begin{align}
+\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
+&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
+&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
+\end{align}
+We approximate our model’s posterior distribution $p( \theta, \psi \vert
+\tilde{X})$ with a tractable family of probability distributions
+$q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)$, where
+$q_{\phi_1}(\theta)$ is a product of independent univariate Gaussian
+distributions with learnable location and scale parameters and
+$q_{\phi_2}(\psi)$ is a multivariate Gaussian distribution with a learnable
+location parameter and a learnable low-rank covariance matrix. We solve the
+resulting stochastic optimization problem using a version of Automatic
+Differentiation Variational Inference (ADVI) [@Kucukelbir2016-bk]: we obtain
+gradient estimates by differentiating a Monte Carlo estimate of the ELBO and
+update the variational parameters  using stochastic gradient ascent.  Our
+implementation and experiments use generic variational families and Monte Carlo
+ELBO and gradient estimators provided by Pyro [@Bingham2018-id], an open source
+library for probabilistic machine learning, and a version built into Pyro of the
+adaptive stochastic gradient ascent algorithm Adam augmented with gradient
+clipping to enhance numerical stability during training.
+
+## Model training {#sec-methods-training}
+
+For the pancreas, PBMC, and uni-fate and bi-fate LARRY single-cell data, since
+the data dimension is relatively small, we run Model 1 and Model 2 with a
+minimum of 100 epochs and a maximum of 4000 epochs. For each epoch, we input
+spliced and unspliced read counts from all the cells and determine the
+convergence condition with an early stopping strategy in which the patience is
+set to 45 and consumes one patience if the minimal ELBO improvement of training
+data per epoch is lower than $10^{-4}$ of previous loss. The learning rate is
+set to be $10^{-2}$ with a decay rate of $0.11/4000$  per epoch. For the
+multi-fate LARRY dataset, since this is a large dataset with over $4 \times
+10^4$ cells, we use mini batches of cells to train Model 1 and Model 2 with a
+minimum of 100 epochs and a maximum of 1000 epochs. Specifically, the batch size
+was set to 4000 cells for both models with an early stopping patience 45 and
+consume one patience if the minimal ELBO improvement of training data per epoch
+lower than $10^{-3}$ of previous loss. The learning rate is set to be $10^{-2}$
+with a decay rate of $0.11/1000$  per epoch. All models were trained on a
+machine with an NVIDIA A100 GPU and the CentOS 7 operating system.
 

### 2024-08-22 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

@@
 
-\abstract{Single-cell RNA Velocity has dramatically advanced our ability
+\abstract{Single-cell RNA velocity has dramatically advanced our ability
 to model cellular differentiation and cell fate decisions. However,
@@
 
-\includegraphics{figures/fig.pdf}
+\includegraphics{figures/v2/Figure2.png}
 
@@
 
-\includegraphics{figures/fig.pdf}
+\includegraphics{figures/v2/Figure3.png}
 
@@
 \frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
+  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right),
+   \label{eq-dudt}\\
 \frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
+  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)
+  -\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
+\end{align} In the equation, the subscript \(c\) is the cell dimension,
+\(g\) is the gene dimension,
+\(\left( u\left( \tau^{(k_{cg})} \right), s\left( \tau^{(k_{cg})} \right) \right)\)
+are the unspliced and spliced expression functions given the change of
+time per cell and gene. \(\tau_{cg}\) represents the displacement of
+time per cell and gene with \begin{align}
+ \tau^{(k_{cg})} &= \operatorname{softplus} \left( t_{c} - {t_{0}^{(k_{cg})}}_g \right) \\
+ & = \log( 1 + \exp (t_c - {t_{0}^{(k_{cg})}}_g)), 
+\end{align} in which \(t_c\) is the shared time per cell,
+\({t_{0}^{(kcg)}}_g\) is the gene-specific switching time. Each cell and
+gene combination has its transcriptional state
+\(k_{cg} \in \{ 0, 1 \}\), where \(0\) indicates the activation state
+and \(1\) indicates the expression state. Each gene has two switching
+times for representing activation and repression: \({t_{0}^{(0)}}_g\) is
+the first switching time corresponding to when the gene expression
+starts to be activated, \({t_0^{(1)}}_g\) is the second switching time
+corresponding to when the gene expression starts to be repressed. We
+note that \(\alpha^{(1)}\) is shared for all the genes, while
+\({\alpha^{(0)}}_g\) is learned independently for each gene.
+
+The analytic solution of the differential equations to predict spliced
+and unspliced gene expression given their parameters is derived by the
+authors of scVelo and a theoretical RNA velocity study
+\citep{Bergen2020-pj, Li2021-qa} and given in Eqs.
+\ref{eq-solution-u}-\ref{eq-solution-s2}. \begin{align}
+u\left(\tau^{\left(k_{c g}\right)}\right) 
+  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
+  \nonumber \\
+&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
+  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
+  \label{eq-solution-u}\\
+s\left(\tau^{\left(k_{c g}\right)}\right) 
+  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
+  \nonumber \\
+  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
+    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
+    \nonumber\\
+  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
+    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
+    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
+    \nonumber \\
+  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
+s\left(\tau^{\left(k_{c g}\right)}\right) 
+  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
+  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
+  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
+  \nonumber \\
+  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
+  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
+  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
+  &\qquad \beta = \gamma \label{eq-solution-s2}
 \end{align}
+
+To simplify these equations, consider the case when \(k_{cg} = 0\) and
+\(\beta_g \neq \gamma_g\). Then, \begin{align}
+u\left(\tau^{(0)}\right) &= {u_0^{(0)}}_g e^{-\beta_g \tau^{(0)}} \nonumber \\
+  & \hskip -24pt + \frac{{\alpha^{(0)}}_g}{\beta_g}\left(1-e^{-\beta_g \tau^{(0)}}\right), 
+  \label{eq-sol-usimp} \\
+s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
+  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}\left(1-e^{-\gamma_g \tau^{(0)}}\right) 
+  \nonumber\\
+  & \hskip -24pt +\frac{{\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g}{\gamma_g-\beta_g}
+  \left(e^{-\gamma_g \tau^{(0)}}-e^{-\beta_g \tau^{(0)}}\right). \label{eq-sol-ssimp}
+\end{align} When \(k_{cg} = 0\) and \(\beta_g = \gamma_g\), then
+\(u\left(\tau^{(0)}\right)\) has the same solution, and
+\(s\left(\tau^{(0)}\right)\) becomes \begin{align}
+s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
+  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}
+    \left(1-e^{-\gamma_g \tau^{(0)}}\right) \nonumber\\
+  & \hskip -24pt - \left( {\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g \right) 
+    \tau^{(0)} e^{-\beta_g \tau^{(0)}}. \label{eq-sol-ssimp2}
+\end{align} When \(k_{cg} = 1\) and \(\beta_g \neq \gamma_g\), then
+\begin{align}
+u\left(\tau^{(1)}\right) &=u_0^{(1)} g^{e^{-\beta_g \tau^{(1)}}}, \\
+s\left(\tau^{(1)}\right) &=s_0^{(1)} e^{-\gamma_g \tau^{(1)}} \nonumber \\
+  & \hskip -24pt +\frac{-\beta_g u_0^{(1)}}{\gamma_g-\beta_g}
+  \left(e^{-\gamma_g \tau^{(1)}}-e^{-\beta_g \tau^{(1)}}\right).
+\end{align} When \(k_{cg} = 1\) and \(\beta_g = \gamma_g\), then
+\(u\left(\tau^{(1)}\right)\) has the same solution, and
+\(s\left(\tau^{(1)}\right)\) becomes \begin{align}
+s\left(\tau^{(1)}\right)=s_0^{(1)}{ }_g e^{-\gamma_g \tau^{(1)}}
+  +\beta_g u_0^{(1)}{ }_g \tau^{(1)} e^{-\beta_g \tau^{(1)}}.
+\end{align}
+
+\subsection{Variational inference}\label{sec-methods-inference}
+
+Given observations
+\(\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)\), we would
+like to compute the posterior distribution over the random variables
+\begin{align}
+\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
+\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
+\end{align} but exact Bayesian inference is intractable in this model.
+We use Pyro to automatically integrate out the local discrete latent
+variables \(k\), which is defined as the cell and gene transcriptional
+state (see above), and approximate the posterior over the remaining
+latent variables using variational inference
+\citep{Bingham2018-id, Kucukelbir2016-bk}, which converts intractable
+integrals into optimization problems that can be solved with
+off-the-shelf tools. In variational inference, we maximize a tractable
+bound (the evidence lower bound (ELBO)) on the Kullback-Leibler (KL)
+divergence
+\[\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} p(\theta, \psi \vert \tilde{X}_{cg}) \right)\]
+between a parametric family of tractable probability distributions q and
+the intractable true posterior \begin{align}
+\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
+&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
+&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
+\end{align} We approximate our model's posterior distribution
+\(p( \theta, \psi \vert
+\tilde{X})\) with a tractable family of probability distributions
+\(q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)\), where
+\(q_{\phi_1}(\theta)\) is a product of independent univariate Gaussian
+distributions with learnable location and scale parameters and
+\(q_{\phi_2}(\psi)\) is a multivariate Gaussian distribution with a
+learnable location parameter and a learnable low-rank covariance matrix.
+We solve the resulting stochastic optimization problem using a version
+of Automatic Differentiation Variational Inference (ADVI)
+\citep{Kucukelbir2016-bk}: we obtain gradient estimates by
+differentiating a Monte Carlo estimate of the ELBO and update the
+variational parameters using stochastic gradient ascent. Our
+implementation and experiments use generic variational families and
+Monte Carlo ELBO and gradient estimators provided by Pyro
+\citep{Bingham2018-id}, an open source library for probabilistic machine
+learning, and a version built into Pyro of the adaptive stochastic
+gradient ascent algorithm Adam augmented with gradient clipping to
+enhance numerical stability during training.
+
+\subsection{Model training}\label{sec-methods-training}
+
+For the pancreas, PBMC, and uni-fate and bi-fate LARRY single-cell data,
+since the data dimension is relatively small, we run Model 1 and Model 2
+with a minimum of 100 epochs and a maximum of 4000 epochs. For each
+epoch, we input spliced and unspliced read counts from all the cells and
+determine the convergence condition with an early stopping strategy in
+which the patience is set to 45 and consumes one patience if the minimal
+ELBO improvement of training data per epoch is lower than \(10^{-4}\) of
+previous loss. The learning rate is set to be \(10^{-2}\) with a decay
+rate of \(0.11/4000\) per epoch. For the multi-fate LARRY dataset, since
+this is a large dataset with over \(4 \times
+10^4\) cells, we use mini batches of cells to train Model 1 and Model 2
+with a minimum of 100 epochs and a maximum of 1000 epochs. Specifically,
+the batch size was set to 4000 cells for both models with an early
+stopping patience 45 and consume one patience if the minimal ELBO
+improvement of training data per epoch lower than \(10^{-3}\) of
+previous loss. The learning rate is set to be \(10^{-2}\) with a decay
+rate of \(0.11/1000\) per epoch. All models were trained on a machine
+with an NVIDIA A100 GPU and the CentOS 7 operating system.
 
@@
 
-\includegraphics{figures/fig.pdf}
+\includegraphics{figures/v2/Supplementary-Figure2.png}
 
@@
 
-\includegraphics{figures/fig.pdf}
+\includegraphics{figures/v2/Supplementary-Figure3.png}
 
@@
 
-\includegraphics{figures/fig.pdf}
+\includegraphics{figures/v2/Supplementary-Figure4.png}
 
@@
 
-\includegraphics{figures/fig.pdf}
+\includegraphics{figures/v2/Supplementary-Figure5.png}
 

### 2024-08-22 reproducibility/manuscript/manuscript.qmd
docs(manuscript): update methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+We use these solutions to formulate an end-to-end probabilistic generative model 
+that relates prior distributions on kinetic parameters to a distribution on pairs 
+of observed unspliced and spliced read count matrices
+
+\begin{align}
+\alpha^{(0)}{ }_g &\sim \operatorname{LogNormal}(0,1), \\
+\beta_g &\sim \operatorname{LogNormal}(0,1), \\
+\gamma_g &\sim \operatorname{LogNormal}(0,1), \\
+&\hskip -18pt \Delta \text { switching }_g \sim \operatorname{LogNormal}(0,1), \\
+t_0^{\left(k_{c g}\right)} &= \left\{
+  \begin{array}{l}
+    t_0^{(0)}{ }_g \sim \operatorname{Normal}(0,1), k_{c g}=0 \\
+    t_0^{(1)}{ }_g=t_0^{(0)}{ }_g+\Delta \text { switching }_g, \\
+    \quad k_{c g}=1
+  \end{array}\right. \\
+t_c &\sim \operatorname{LogNormal}(0,1), \\
+k_{c g} &\sim \text{Bernoulli} \left( \text{logits}= t_c-t_0^{(1)} \right), \\
+\tau^{\left(k_{c g}\right)} 
+  &= \operatorname{softplus}\left(t_c-t_0^{\left(k_{c g}\right)}{ }_g\right), \\
+u_{c g} 
+ &= \text { Measurement }_u \left( u\left(\tau^{\left(k_{c g}\right)}\right) ; 
+                                   u_{c g}^{obs}\right), \\
+s_{c g} 
+  &= \text { Measurement }_s \left( s\left(\tau^{(k_{c g})}\right) ; 
+                                    s_{c g}^{obs}\right).
+\end{align}
+$u\left(\tau^{\left(k_{c g}\right)}\right)$ and $s\left(\tau^{(k_{c g})}\right)$ are
+are called the denoised gene expression calculated from the velocity analytic
+solution input with the kinetics random variables. $u_{cg}$ and $s_{cg}$ are the spliced
+and unspliced read count sampled from the Poisson models. $u_{cg}^{obs}$ and $s_{cg}^{obs}$ are
+the observed spliced and unspliced read count tables. The generative process
+
+$\text{Measurement}(\cdot)$ for observed unspliced read counts given denoised unspliced
+gene transcript expression level $u\left(\tau^{(k_{cg})}\right)$ 
+(and identical for observed spliced read counts) models the expected number of observed 
+reads for a given gene in a given cell as the number of transcripts times the ratio of 
+the cell's total reads to total transcripts
+\begin{align}
+u_c^{\hat{obs}} &= \sum_g u_{c g}^{obs}, \\
+\hat{u}_c &= \sum_g u\left( \tau^{(k_{c g})}\right), \\
+\eta_c^{(u)} &\sim \operatorname{Normal}\left(
+    u_c^{\hat{obs}_c}, 
+    \operatorname{std} \left(u_c^{\hat{obs}}\right)
+  \right), \\
+\mu_{c g}^{(u)} &= \log \left(u\left(\tau^k{ }_{c g}\right)\right)
+  +\log \left(\eta_c^{(u)}\right)-\log \left(\hat{u}_c\right), \\
+u_{c g}^{obs} &\sim 
+  \operatorname{Poisson}\left(\lambda=\exp \left(\mu_{c g}^{(u)}\right)\right).
+\end{align}
+
+For the first Pyro-Velocity model (Model 1), we constrain the shared time to be strictly
+larger than $t_{0}^{(0)}$ by introducing auxiliary random variables 
+$$
+\text{t\_constraint}_{cg} 
+  \sim \text{Bernoulli} \left( \text{logits} = t_c - {t_{0}^{(0)}}_g \right),
+$$
+and setting their values to $1$, and we set the initial condition per gene to be
+\begin{align}
+\left( {u_{0}^{(k_{cg})}}_g , {s_{0}^{(k_{cg})}}_g \right) &= \left\{
+  \begin{array}{l}
+    (0,0), k_{c g}=0 \\
+    \bigg( {u \left( \Delta \text { switching }_g \right)}_g,\\
+           \quad {s \left( \Delta \text { switching }_g \right)}_g \bigg), \\
+           \quad k_{c g}=1
+  \end{array}\right.
+\end{align}
+For the extended Pyro&thinsp;-Velocity model (Model 2), we remove the shared
+time constraint $\text{t\_constraint}_{cg}$, thus allowing a time lag per gene
+that might be caused by delayed gene activation and set the initial condition
+per gene as random variables that are strictly positive $\left({u_{0}^{(0)}}_g,
+{s_{0}^{(0)}}_g\right)$, which allow genes having a basal expression level before gene
+activation. Then, we compute the gene expression at the second switching time as
+\begin{align}
+({u_{0}^{(1)}}_g, {s_{0}^{(1)}}_g) &= 
+  \bigg( {u \left( \Delta \text { switching }_g \right)}_g, \nonumber \\
+& \qquad {s \left( \Delta \text { switching }_g \right)}_g \bigg),
+\end{align}
+which shares the same initial condition $\left({u_{0}^{(0)}}_g, {s_{0}^{(0)}}_g\right)$ where
+\begin{align}
+{u_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1),\\
+{s_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1).
+\end{align}
+
 ## Variational inference {#sec-methods-inference}
@@
 \theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
-\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
+\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, 
+          {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
 \end{align}
@@
 (ELBO)) on the Kullback-Leibler (KL) divergence
-$$\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} p(\theta, \psi \vert \tilde{X}_{cg}) \right)$$
+$$
+\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} 
+  p(\theta, \psi \vert \tilde{X}_{cg}) \right)
+$$
 between a parametric family of tractable probability distributions q and the
@@
 \varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
-&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
+&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} 
+     \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
 &\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
 \end{align}
-We approximate our model’s posterior distribution $p( \theta, \psi \vert
+We approximate our model's posterior distribution $p( \theta, \psi \vert
 \tilde{X})$ with a tractable family of probability distributions
@@
 We implemented Cytotrace in Python based on the R version [@Gulati2020-xq] for
-benchmarking Pyro&thinsp;-Velocity and scVelo’s shared (latent) time prediction in the
+benchmarking Pyro&thinsp;-Velocity and scVelo's shared (latent) time prediction in the
 Python environment. This implementation is available in the
-[Pyro&thinsp;-Velocity GitHub repository](https://github.com/pinellolab/pyrovelocity/blob/main/src/pyrovelocity/analysis/cytotrace.py).
+[Pyro&thinsp;-Velocity GitHub repository][cytotrace].
+
+[cytotrace]: https://github.com/pinellolab/pyrovelocity/blob/main/src/pyrovelocity/analysis/cytotrace.py
 
@@
 \begin{align*}
-\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} |x_{icg} - \tilde{X}_{cg}|,
+\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} 
+  \vert x_{icg} - \tilde{X}_{cg} \vert,
 \end{align*}
@@
 1. the consistency between the velocity vector field and the *clonal progression
-vector field* using grid-based cosine similarity; 2. Spearman’s correlation of
+vector field* using grid-based cosine similarity; 2. Spearman's correlation of
 velocity-based shared (latent) time with Cospar fate potency and the silver
@@
 L.P. has financial interests in Edilytics, Inc., Excelsior Genomics, and SeQure
-Dx, Inc. L.P.’s interests were reviewed and are managed by Massachusetts General
+Dx, Inc. L.P.'s interests were reviewed and are managed by Massachusetts General
 Hospital and Partners HealthCare in accordance with their conflict of interest
@@
 $(u_{0}=0,\,s_{0}=0)$, as in previous methods[@La_Manno2018-lj;@Bergen2020-pj]
-([Methods](#sec-methods)). **b**, Pyro&thinsp;-Velocity Model 2. This figure follows the same
-graphical convention as in **a**, In this model $t_{0}^{(0)}$ is independent from
-the shared time, thus allowing a time lag for the gene activation phase. In
-addition, a positive starting basal gene expression level $(u_{0},\,s_{0})$
-different than $(u_{0}=0,\,s_{0}=0)$ is allowed. ([Methods](#sec-methods)). **c**, Schematic
-illustrating how the constraints proposed in Model 1 (left) and Model 2 (right)
-are reflected in the modeling of individual gene expression. The x-axis
-corresponds to the shared time, the y-axis corresponds to the posterior mean of
-the spliced expression level, the vertical lines with arrows show the start and
-end time of the observed cell sampling, t0 represents the gene-specific
-activation time $t_{0}^{(0)}$, the red lines highlight the differences for the
-potential gene-specific time lag when $t_{0}^{(0)} > {\min}\!⁡(t)$, and the
-initial spliced gene expression. The same differences can also be illustrated
-for the unspliced gene expression profiles (not shown).
+([Methods](#sec-methods)). **b**, Pyro&thinsp;-Velocity Model 2. This figure
+follows the same graphical convention as in **a**, In this model $t_{0}^{(0)}$
+is independent from the shared time, thus allowing a time lag for the gene
+activation phase. In addition, a positive starting basal gene expression level
+$(u_{0},\,s_{0})$ different than $(u_{0}=0,\,s_{0}=0)$ is allowed.
+([Methods](#sec-methods)). **c**, Schematic illustrating how the constraints
+proposed in Model 1 (left) and Model 2 (right) are reflected in the modeling of
+individual gene expression. The x-axis corresponds to the shared time, the
+y-axis corresponds to the posterior mean of the spliced expression level, the
+vertical lines with arrows show the start and end time of the observed cell
+sampling, t0 represents the gene-specific activation time $t_{0}^{(0)}$, the red
+lines highlight the differences for the potential gene-specific time lag when
+$t_{0}^{(0)} > {\min}\!⁡(t)$, and the initial spliced gene expression. The same
+differences can also be illustrated for the unspliced gene expression profiles
+(not shown).
 :::
@@
 Extended version of @fig-model-results for evaluating Pyro&thinsp;-Velocity
-(Model 1) and scVelo on the PBMC and pancreas datasets. **a**, Left figure shows the
-scVelo vector field using the default user options, right figure shows the
+(Model 1) and scVelo on the PBMC and pancreas datasets. **a**, Left figure shows
+the scVelo vector field using the default user options, right figure shows the
 evaluation of the scVelo vector field using different combinations of user
-options within Epsilon cells. **b**, Pyro&thinsp;-Velocity additional single cell
-level uncertainty diagnostic plots on the fully mature PBMC dataset, from left
-to right: 1. standard deviation of posterior samples of shared time, 2. Rayleigh
-test of posterior samples of the vector field (the title reports false discovery
-rate after p-value correction for multiple tests across cells), 3.
+options within Epsilon cells. **b**, Pyro&thinsp;-Velocity additional single
+cell level uncertainty diagnostic plots on the fully mature PBMC dataset, from
+left to right: 1. standard deviation of posterior samples of shared time, 2.
+Rayleigh test of posterior samples of the vector field (the title reports false
+discovery rate after p-value correction for multiple tests across cells), 3.
 transcriptional state uncertainty of predicted raw read count per cell. The
-contour lines show the top 10% of the most uncertain cells. **c**, For the pancreas
-dataset, the figures show from left to right: UMAP rendering of 1. Cytotrace, 2.
-scVelo latent time, 3. Posterior mean of Pyro&thinsp;-Velocity shared time
-averaged across 30 posterior samples, 4. Top 50 markers overlapping between
-Pyro&thinsp;-Velocity and scVelo ([Methods](#sec-methods), Supplementary Table 1-2), 5. The same
-uncertainty diagnostic plots per cell as in **b**. **d**, The same plot as [@fig-model-results]d, e.
-Selection of the top negatively correlated genes with the cell shared time based
-on the posterior mean on denoised spliced expression and negative mean absolute
-error of gene expression prediction ([Methods](#sec-methods)). **f**, Phase portraits, rainbow
+contour lines show the top 10% of the most uncertain cells. **c**, For the
+pancreas dataset, the figures show from left to right: UMAP rendering of 1.
+Cytotrace, 2. scVelo latent time, 3. Posterior mean of Pyro&thinsp;-Velocity
+shared time averaged across 30 posterior samples, 4. Top 50 markers overlapping
+between Pyro&thinsp;-Velocity and scVelo ([Methods](#sec-methods), Supplementary
+Table 1-2), 5. The same uncertainty diagnostic plots per cell as in **b**.
+**d**, The same plot as [@fig-model-results]d, e. Selection of the top
+negatively correlated genes with the cell shared time based on the posterior
+mean on denoised spliced expression and negative mean absolute error of gene
+expression prediction ([Methods](#sec-methods)). **f**, Phase portraits, rainbow
 plots, and UMAP rendering of posterior mean of gene expression for the genes
@@
 data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
-scRNA-seq datasets with lineage barcoding information (LARRY dataset). **a**, SPRING
-embedding rendering of Cytotrace scores, the title reports the Spearman's
+scRNA-seq datasets with lineage barcoding information (LARRY dataset). **a**,
+SPRING embedding rendering of Cytotrace scores, the title reports the Spearman's
 correlation between the Cytotrace and Cospar fate potency scores, **b**,
@@
 state uncertainty of predicted raw read count per cell. The contour lines show
-the top 10% of the most uncertain cells ([Methods](#sec-methods)), **c**, Pairwise Spearman's
-correlation between Cytotrace/Cospar fate potency scores and velocity methods
-for the four different benchmarking datasets selected based on the lineage
-barcoding information from LARRY data, **d**, SPRING embedding rendering of
-posterior mean of shared time across 30 posterior samples, the title shows the
-Spearman's correlation of mean shared time with Cytotrace. **e**, Selection of the
-top positively correlated genes with the cell shared time based on the posterior
-mean of spliced expression. **f**, Phase portraits, rainbow plots, and UMAP
-rendering with splicing gene expression level for the selected genes in **e**.
+the top 10% of the most uncertain cells ([Methods](#sec-methods)), **c**,
+Pairwise Spearman's correlation between Cytotrace/Cospar fate potency scores and
+velocity methods for the four different benchmarking datasets selected based on
+the lineage barcoding information from LARRY data, **d**, SPRING embedding
+rendering of posterior mean of shared time across 30 posterior samples, the
+title shows the Spearman's correlation of mean shared time with Cytotrace.
+**e**, Selection of the top positively correlated genes with the cell shared
+time based on the posterior mean of spliced expression. **f**, Phase portraits,
+rainbow plots, and UMAP rendering with splicing gene expression level for the
+genes selected in **e**.
 :::

### 2024-08-22 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

@@
 
+We use these solutions to formulate an end-to-end probabilistic
+generative model that relates prior distributions on kinetic parameters
+to a distribution on pairs of observed unspliced and spliced read count
+matrices
+
+\begin{align}
+\alpha^{(0)}{ }_g &\sim \operatorname{LogNormal}(0,1), \\
+\beta_g &\sim \operatorname{LogNormal}(0,1), \\
+\gamma_g &\sim \operatorname{LogNormal}(0,1), \\
+&\hskip -18pt \Delta \text { switching }_g \sim \operatorname{LogNormal}(0,1), \\
+t_0^{\left(k_{c g}\right)} &= \left\{
+  \begin{array}{l}
+    t_0^{(0)}{ }_g \sim \operatorname{Normal}(0,1), k_{c g}=0 \\
+    t_0^{(1)}{ }_g=t_0^{(0)}{ }_g+\Delta \text { switching }_g, \\
+    \quad k_{c g}=1
+  \end{array}\right. \\
+t_c &\sim \operatorname{LogNormal}(0,1), \\
+k_{c g} &\sim \text{Bernoulli} \left( \text{logits}= t_c-t_0^{(1)} \right), \\
+\tau^{\left(k_{c g}\right)} 
+  &= \operatorname{softplus}\left(t_c-t_0^{\left(k_{c g}\right)}{ }_g\right), \\
+u_{c g} 
+ &= \text { Measurement }_u \left( u\left(\tau^{\left(k_{c g}\right)}\right) ; 
+                                   u_{c g}^{obs}\right), \\
+s_{c g} 
+  &= \text { Measurement }_s \left( s\left(\tau^{(k_{c g})}\right) ; 
+                                    s_{c g}^{obs}\right).
+\end{align} \(u\left(\tau^{\left(k_{c g}\right)}\right)\) and
+\(s\left(\tau^{(k_{c g})}\right)\) are are called the denoised gene
+expression calculated from the velocity analytic solution input with the
+kinetics random variables. \(u_{cg}\) and \(s_{cg}\) are the spliced and
+unspliced read count sampled from the Poisson models. \(u_{cg}^{obs}\)
+and \(s_{cg}^{obs}\) are the observed spliced and unspliced read count
+tables. The generative process
+
+\(\text{Measurement}(\cdot)\) for observed unspliced read counts given
+denoised unspliced gene transcript expression level
+\(u\left(\tau^{(k_{cg})}\right)\) (and identical for observed spliced
+read counts) models the expected number of observed reads for a given
+gene in a given cell as the number of transcripts times the ratio of the
+cell's total reads to total transcripts \begin{align}
+u_c^{\hat{obs}} &= \sum_g u_{c g}^{obs}, \\
+\hat{u}_c &= \sum_g u\left( \tau^{(k_{c g})}\right), \\
+\eta_c^{(u)} &\sim \operatorname{Normal}\left(
+    u_c^{\hat{obs}_c}, 
+    \operatorname{std} \left(u_c^{\hat{obs}}\right)
+  \right), \\
+\mu_{c g}^{(u)} &= \log \left(u\left(\tau^k{ }_{c g}\right)\right)
+  +\log \left(\eta_c^{(u)}\right)-\log \left(\hat{u}_c\right), \\
+u_{c g}^{obs} &\sim 
+  \operatorname{Poisson}\left(\lambda=\exp \left(\mu_{c g}^{(u)}\right)\right).
+\end{align}
+
+For the first Pyro-Velocity model (Model 1), we constrain the shared
+time to be strictly larger than \(t_{0}^{(0)}\) by introducing auxiliary
+random variables \[
+\text{t\_constraint}_{cg} 
+  \sim \text{Bernoulli} \left( \text{logits} = t_c - {t_{0}^{(0)}}_g \right),
+\] and setting their values to \(1\), and we set the initial condition
+per gene to be \begin{align}
+\left( {u_{0}^{(k_{cg})}}_g , {s_{0}^{(k_{cg})}}_g \right) &= \left\{
+  \begin{array}{l}
+    (0,0), k_{c g}=0 \\
+    \bigg( {u \left( \Delta \text { switching }_g \right)}_g,\\
+           \quad {s \left( \Delta \text { switching }_g \right)}_g \bigg), \\
+           \quad k_{c g}=1
+  \end{array}\right.
+\end{align} For the extended Pyro -Velocity model (Model 2), we remove
+the shared time constraint \(\text{t\_constraint}_{cg}\), thus allowing
+a time lag per gene that might be caused by delayed gene activation and
+set the initial condition per gene as random variables that are strictly
+positive \(\left({u_{0}^{(0)}}_g,
+{s_{0}^{(0)}}_g\right)\), which allow genes having a basal expression
+level before gene activation. Then, we compute the gene expression at
+the second switching time as \begin{align}
+({u_{0}^{(1)}}_g, {s_{0}^{(1)}}_g) &= 
+  \bigg( {u \left( \Delta \text { switching }_g \right)}_g, \nonumber \\
+& \qquad {s \left( \Delta \text { switching }_g \right)}_g \bigg),
+\end{align} which shares the same initial condition
+\(\left({u_{0}^{(0)}}_g, {s_{0}^{(0)}}_g\right)\) where \begin{align}
+{u_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1),\\
+{s_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1).
+\end{align}
+
 \subsection{Variational inference}\label{sec-methods-inference}
@@
 \theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
-\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
+\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, 
+          {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
 \end{align} but exact Bayesian inference is intractable in this model.
@@
 bound (the evidence lower bound (ELBO)) on the Kullback-Leibler (KL)
-divergence
-\[\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} p(\theta, \psi \vert \tilde{X}_{cg}) \right)\]
-between a parametric family of tractable probability distributions q and
-the intractable true posterior \begin{align}
+divergence \[
+\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} 
+  p(\theta, \psi \vert \tilde{X}_{cg}) \right)
+\] between a parametric family of tractable probability distributions q
+and the intractable true posterior \begin{align}
 \varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
-&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
+&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} 
+     \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
 &\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
@@
 combinations \begin{align*}
-\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} |x_{icg} - \tilde{X}_{cg}|,
+\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} 
+  \vert x_{icg} - \tilde{X}_{cg} \vert,
 \end{align*} where \(i\) is the posterior sample index, \(N_i\) is the
@@
 \textbf{f}, Phase portraits, rainbow plots, and UMAP rendering with
-splicing gene expression level for the selected genes in \textbf{e}.}
+splicing gene expression level for the genes selected in \textbf{e}.}
 

### 2024-08-26 reproducibility/manuscript/manuscript.qmd
starting to update methods, part 1

@@
 process, and infer the unspliced and spliced gene expression level from the
-differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo
+ordinary differential equation (ODEs) proposed in velocyto [@La_Manno2018-lj] and scVelo
 [@Bergen2020-pj]
 \begin{align}
-\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right),
-   \label{eq-dudt}\\
-\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)
-  -\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
+\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \label{eq-dudt}\\
+   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0, \label{eq-dsdt}
+\end{align}
+where $u(t), s(t)$ are the unspliced and spliced expression levels of a gene at time $t$ under a transcription rate $\alpha(t)$ with possible temporal dependence, splicing rate $\beta$, and degradation rate $\gamma$.
+We specify this model to a setting that depends on cell $c$ and gene $g$ as follows:
+\begin{align}
+\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-dudt}\\
+   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)}, \label{eq-dsdt}.
 \end{align}
 In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
-$\left( u\left( \tau^{(k_{cg})} \right), s\left( \tau^{(k_{cg})} \right) \right)$ 
+$\left( u_{cg}(t), s_{cg}(t) \right)$ 
 are the unspliced and spliced expression functions given the
-change of time per cell and gene. $\tau_{cg}$ represents the displacement of time per
+change of time per cell and gene.
+We restrict attention to piecewise-constant $\alpha_{cg}(t)$ to capture gene-specific activation and repression.
+We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown activation state.
+We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times $\tau_{cg}$, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$.
+These relative times are also used to parametrize the Bernoulli process for $k_{cg}$.
+Importantly, we recognize that the initial conditions are in fact unknown.
+
+We propose and study two models: Model 1 assumes that spliced and unspliced concentrations are both 0 at time 0; Model 2 considers these initial conditions as unknowns with a log-Normal prior distribution.
+In general, the solution space of ODEs becomes much richer when considered over a domain of initial conditions (as opposed to a single point); indeed, this affords Model 2 much greater expressivity.
+For clarity, we first present the generative framework for both models, then provide further interpretation and intuition.
+
+First, we introduce the generative model that describes the various unobserved times:
+\begin{align}
+  % unit lognormal t_c
+  t_c &\sim \text{LogNormal}(0, 1) \\
+  % gene-specific t_0
+  t^{(0)}_{0,g} &\sim \text{LogNormal}(0, 1) \\
+  % switching time
+  \Delta \textrm{switching}_g &\sim \text{LogNormal}(0, 1) \\
+  % gene-specific t_1
+  t^{(1)}_{0,g} &= t^{(0)}_{0,g} + \Delta \textrm{switching}_g \\
+  %cell-gene-specific activation state
+  k_{cg} &\sim \text{Bernoulli}(\textrm{logits}=t_c - t^{(1)}_{0,g}) \\
+  % cell-gene-specific latent time
+  \tau_{cg} &= \text{softplus}(t_c - t^{(k_{cg})}_{0,g}).
+\end{align}
+Here, $\tau_{cg}$ represents the displacement of time per
 cell and gene with 
 \begin{align}
- \tau^{(k_{cg})} &= \operatorname{softplus} \left( t_{c} - {t_{0}^{(k_{cg})}}_g \right) \\
- & = \log( 1 + \exp (t_c - {t_{0}^{(k_{cg})}}_g)), 
-\end{align}
-in which $t_c$ is the shared time per cell, ${t_{0}^{(kcg)}}_g$ is the
+ textrm{softplus}(t) :=  \log( 1 + e^t).
+\end{align}
+Recall that $t_c$ is the shared time per cell, $t^{(k_{cg})}_{0,g}$ is the
 gene-specific switching time. Each cell and gene combination has its
@@
 activation state and $1$ indicates the expression state. Each gene has two
-switching times for representing activation and repression: ${t_{0}^{(0)}}_g$ is
+switching times for representing activation and repression: $t^{(0)}_{0,g}$ is
 the first switching time corresponding to when the gene expression starts to be
-activated, ${t_0^{(1)}}_g$  is the second switching time corresponding to when
-the gene expression starts to be repressed. We note that $\alpha^{(1)}$ is
-shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
-each gene.
+activated, $t^{(1)}_{0,g}$  is the second switching time corresponding to when
+the gene expression starts to be repressed.
+
+Next we introduce the priors for the splicing parameters (where the activation rate $\alpha$ depends on the activation state $k_{cg}$ from above):
+\begin{align}
+  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
+  \beta_g &\sim \text{LogNormal}(0, 1) \\
+  \gamma_g &\sim \text{LogNormal}(0, 1) \\
+  \alpha_{cg} = \begin{cases}
+    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
+    0 & \text{if } k_{cg} = 1
+  \end{cases}
+\end{align}
+\textbf{Note that $\alpha^{(1)}$ is shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
+each gene. MATT: this was in the old text, but I think $\alpha^{(1)}$ is no longer used based on conversations with Alvin?}
+
+Now, we describe the priors for the initial conditions, noting that this is the only difference between Model 1 and Model 2:
+\begin{align}
+  \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} &\sim \begin{cases}
+    (0, 0) & \text{Model 1} \\
+    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1)) & \text{Model 2}
+  \end{cases}
+
+  u^{(0)}_{cg}, s^{(0)}_{cg} &= \begin{cases}
+    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
+    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \ T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
+  \end{cases}
+\end{align}
+
+Finally, we describe the ODE solution at time $\tau_{cg}$, and the observation model at those times that give rise to the observed counts:
+\begin{align}
+  \hat{u}_{cg}, \hat{s}_{cg} &= \textrm{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big) \\
+  u_{cg} &\sim \text{Poisson}(\hat{u}_{cg}) \\
+  s_{cg} &\sim \text{Poisson}(\hat{s}_{cg})
+\end{align}
 

### 2024-08-26 reproducibility/manuscript/manuscript.qmd
starting to update methods, part 2

@@
 \begin{align}
- textrm{softplus}(t) :=  \log( 1 + e^t).
+ \text{softplus}(t) :=  \log( 1 + e^t).
 \end{align}
@@
   \gamma_g &\sim \text{LogNormal}(0, 1) \\
-  \alpha_{cg} = \begin{cases}
+  \alpha_{cg} &= \begin{cases}
     \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
@@
 \begin{align}
-  \hat{u}_{cg}, \hat{s}_{cg} &= \textrm{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big) \\
-  u_{cg} &\sim \text{Poisson}(\hat{u}_{cg}) \\
-  s_{cg} &\sim \text{Poisson}(\hat{s}_{cg})
-\end{align}
-
+  \hat{u}_{cg}, \hat{s}_{cg} &= \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big) \\
+  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
+  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
+  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
+  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
+  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
+  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
+  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
+  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
+  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
+  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
+\end{align}
+Here, we use ${u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}$ to denote the observed unspliced and spliced counts
+for cell $c$ and gene $g$. We use $\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}$ to
+denote our generative model's prediction of these unspliced and spliced expression levels.
+The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_cg, \hat{s}_cg$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.
+\textbf{Improve descriptions of how noise is modeled in the observation model.}
+
+\textbf{Need to update the analytic solutions, but first need to confirm the above is correct. Also, I recommend pushing all of the below analytic solutions to the appendix.}
 The analytic solution of the differential equations to predict
@@
   +\beta_g u_0^{(1)}{ }_g \tau^{(1)} e^{-\beta_g \tau^{(1)}}.
-\end{align}
-
-We use these solutions to formulate an end-to-end probabilistic generative model 
-that relates prior distributions on kinetic parameters to a distribution on pairs 
-of observed unspliced and spliced read count matrices
-
-\begin{align}
-\alpha^{(0)}{ }_g &\sim \operatorname{LogNormal}(0,1), \\
-\beta_g &\sim \operatorname{LogNormal}(0,1), \\
-\gamma_g &\sim \operatorname{LogNormal}(0,1), \\
-&\hskip -18pt \Delta \text { switching }_g \sim \operatorname{LogNormal}(0,1), \\
-t_0^{\left(k_{c g}\right)} &= \left\{
-  \begin{array}{l}
-    t_0^{(0)}{ }_g \sim \operatorname{Normal}(0,1), k_{c g}=0 \\
-    t_0^{(1)}{ }_g=t_0^{(0)}{ }_g+\Delta \text { switching }_g, \\
-    \quad k_{c g}=1
-  \end{array}\right. \\
-t_c &\sim \operatorname{LogNormal}(0,1), \\
-k_{c g} &\sim \text{Bernoulli} \left( \text{logits}= t_c-t_0^{(1)} \right), \\
-\tau^{\left(k_{c g}\right)} 
-  &= \operatorname{softplus}\left(t_c-t_0^{\left(k_{c g}\right)}{ }_g\right), \\
-u_{c g} 
- &= \text { Measurement }_u \left( u\left(\tau^{\left(k_{c g}\right)}\right) ; 
-                                   u_{c g}^{obs}\right), \\
-s_{c g} 
-  &= \text { Measurement }_s \left( s\left(\tau^{(k_{c g})}\right) ; 
-                                    s_{c g}^{obs}\right).
-\end{align}
-$u\left(\tau^{\left(k_{c g}\right)}\right)$ and $s\left(\tau^{(k_{c g})}\right)$ are
-are called the denoised gene expression calculated from the velocity analytic
-solution input with the kinetics random variables. $u_{cg}$ and $s_{cg}$ are the spliced
-and unspliced read count sampled from the Poisson models. $u_{cg}^{obs}$ and $s_{cg}^{obs}$ are
-the observed spliced and unspliced read count tables. The generative process
-
-$\text{Measurement}(\cdot)$ for observed unspliced read counts given denoised unspliced
-gene transcript expression level $u\left(\tau^{(k_{cg})}\right)$ 
-(and identical for observed spliced read counts) models the expected number of observed 
-reads for a given gene in a given cell as the number of transcripts times the ratio of 
-the cell's total reads to total transcripts
-\begin{align}
-u_c^{\hat{obs}} &= \sum_g u_{c g}^{obs}, \\
-\hat{u}_c &= \sum_g u\left( \tau^{(k_{c g})}\right), \\
-\eta_c^{(u)} &\sim \operatorname{Normal}\left(
-    u_c^{\hat{obs}_c}, 
-    \operatorname{std} \left(u_c^{\hat{obs}}\right)
-  \right), \\
-\mu_{c g}^{(u)} &= \log \left(u\left(\tau^k{ }_{c g}\right)\right)
-  +\log \left(\eta_c^{(u)}\right)-\log \left(\hat{u}_c\right), \\
-u_{c g}^{obs} &\sim 
-  \operatorname{Poisson}\left(\lambda=\exp \left(\mu_{c g}^{(u)}\right)\right).
-\end{align}
-
-For the first Pyro-Velocity model (Model 1), we constrain the shared time to be strictly
-larger than $t_{0}^{(0)}$ by introducing auxiliary random variables 
-$$
-\text{t\_constraint}_{cg} 
-  \sim \text{Bernoulli} \left( \text{logits} = t_c - {t_{0}^{(0)}}_g \right),
-$$
-and setting their values to $1$, and we set the initial condition per gene to be
-\begin{align}
-\left( {u_{0}^{(k_{cg})}}_g , {s_{0}^{(k_{cg})}}_g \right) &= \left\{
-  \begin{array}{l}
-    (0,0), k_{c g}=0 \\
-    \bigg( {u \left( \Delta \text { switching }_g \right)}_g,\\
-           \quad {s \left( \Delta \text { switching }_g \right)}_g \bigg), \\
-           \quad k_{c g}=1
-  \end{array}\right.
-\end{align}
-For the extended Pyro&thinsp;-Velocity model (Model 2), we remove the shared
-time constraint $\text{t\_constraint}_{cg}$, thus allowing a time lag per gene
-that might be caused by delayed gene activation and set the initial condition
-per gene as random variables that are strictly positive $\left({u_{0}^{(0)}}_g,
-{s_{0}^{(0)}}_g\right)$, which allow genes having a basal expression level before gene
-activation. Then, we compute the gene expression at the second switching time as
-\begin{align}
-({u_{0}^{(1)}}_g, {s_{0}^{(1)}}_g) &= 
-  \bigg( {u \left( \Delta \text { switching }_g \right)}_g, \nonumber \\
-& \qquad {s \left( \Delta \text { switching }_g \right)}_g \bigg),
-\end{align}
-which shares the same initial condition $\left({u_{0}^{(0)}}_g, {s_{0}^{(0)}}_g\right)$ where
-\begin{align}
-{u_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1),\\
-{s_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1).
 \end{align}

### 2024-08-26 src/pyrovelocity/styles/__init__.py
feat(styles): add function to set matplotlib style globally
- check if latex is in path
- TODO: the latex distribution requires standard packages that may only be able to be checked via tlmgr

Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,0 +1,5 @@
+from pyrovelocity.styles.configure import configure_matplotlib_style
+
+__all__ = [
+    "configure_matplotlib_style",
+]

### 2024-08-26 src/pyrovelocity/styles/configure.py

@@ -1,0 +1,48 @@
+import shutil
+
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+from beartype import beartype
+from beartype.typing import Any, Dict
+
+__all__ = [
+    "configure_matplotlib_style",
+]
+
+
+@beartype
+def configure_matplotlib_style(
+    override_defaults: Dict[str, Any] = {},
+):
+    """
+    Configure the global matplotlib style.
+
+    Accepts a dictionary to override default rcParams set in the
+    `common.mplstyle` file.
+
+    For example, to override the default figure size, use:
+
+    ```
+        {
+            # Example: Set a global figure size
+            "figure.figsize": (8, 6),
+            # Add more global settings as needed
+        },
+    ```
+
+    Args:
+        override_defaults (Dict[str, Any], optional):
+            Dictionary to update default rcParams. Defaults to {}.
+    """
+    plt.style.use("pyrovelocity.styles.common")
+
+    if not shutil.which("latex"):
+        mpl.rcParams.update(
+            {
+                "text.usetex": False,
+            }
+        )
+
+    mpl.rcParams.update(
+        override_defaults,
+    )

### 2024-08-26 src/pyrovelocity/plots/_lineage_fate_correlation.py
fix(plots): update lineage fate correlation titles and font size
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,5 +1,6 @@
 from pathlib import Path
 
+import matplotlib
 import numpy as np
 import pandas as pd
 import scvelo as scv
@@ -47,6 +48,8 @@ def plot_lineage_fate_correlation(
     save_plot: bool = True,
     show_colorbars: bool = False,
     show_titles: bool = False,
+    default_fontsize: int = 7,
+    default_title_padding: int = 2,
 ):
     """
     Plot lineage fate correlation with shared latent time estimates.
@@ -158,13 +161,21 @@ def plot_lineage_fate_correlation(
     )
     ax[0].axis("off")
     if show_titles:
-        ax[0].set_title("Cell types", fontsize=7)
+        ax[0].set_title(
+            "Cell types",
+            fontsize=default_fontsize,
+            pad=default_title_padding,
+        )
     else:
-        ax[0].set_title("", fontsize=7)
-    ax[0].set_ylabel(ylabel, fontsize=7)
+        ax[0].set_title(
+            "",
+            fontsize=default_fontsize,
+            pad=default_title_padding,
+        )
+    ax[0].set_ylabel(ylabel, fontsize=default_fontsize)
 
     scv.pl.velocity_embedding_grid(
-        adata_input_clone,
+        adata=adata_input_clone,
         scale=scale,
         autoscale=True,
         show=False,
@@ -175,14 +186,21 @@ def plot_lineage_fate_correlation(
         vkey="clone_vector",
         basis="emb",
         ax=ax[1],
-        title="Clonal progression" if show_titles else "",
+        title="",
         color="gray",
         arrow_color="black",
-        fontsize=7,
+        fontsize=default_fontsize,
     )
+    ax[1].axis("off")
+    if show_titles:
+        ax[1].set_title(
+            "Clonal progression",
+            fontsize=default_fontsize,
+            pad=default_title_padding,
+        )
 
     scv.pl.velocity_embedding_grid(
-        adata_scvelo,
+        adata=adata_scvelo,
         show=False,
         s=dotsize,
         density=density,
@@ -193,13 +211,24 @@ def plot_lineage_fate_correlation(
         basis="emb",
         ax=ax[2],
         title="",
-        fontsize=7,
+        fontsize=default_fontsize,
         color="gray",
         arrow_color="black",
     )
+    ax[2].axis("off")
     if show_titles:
         ax[2].set_title(
[…]

### 2024-08-26 src/pyrovelocity/tasks/time_fate_correlation.py
fix(tasks): update time lineage fate correlation layout parameters
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,9 +1,11 @@
 from pathlib import Path
 
+import matplotlib
 import matplotlib.pyplot as plt
 
 from pyrovelocity.logging import configure_logging
 from pyrovelocity.plots import plot_lineage_fate_correlation
+from pyrovelocity.styles import configure_matplotlib_style
 from pyrovelocity.styles.colors import LARRY_CELL_TYPE_COLORS
 from pyrovelocity.utils import load_anndata_from_path
 from pyrovelocity.workflows.main_configuration import (
@@ -15,9 +17,12 @@ from pyrovelocity.workflows.main_configuration import (
 
 logger = configure_logging(__name__)
 
+configure_matplotlib_style()
+
 
 def estimate_time_lineage_fate_correlation(
     reports_path: str | Path = "reports",
+    model_identifier: str = "model2",
 ):
     configurations = [
         larry_mono_configuration,
@@ -34,27 +39,20 @@ def estimate_time_lineage_fate_correlation(
         figsize=(width, height),
         constrained_layout=True,
     )
-    all_axes = fig.subplots(len(configurations), 8)
-    for ax in all_axes.flat:
-        ax.set_aspect("equal", adjustable="box")
-    # fig.subplots_adjust(
-    #     left=0.01,
-    #     bottom=0.3,
-    #     # bottom=0.1,
-    #     right=0.99,
-    #     top=0.95,
-    #     # top=0.15,
-    #     hspace=0.4,
-    #     # hspace=0.2,
-    #     wspace=0.2,
-    # )
+    all_axes = fig.subplots(
+        len(configurations),
+        8,
+        gridspec_kw={
+            "hspace": 0.01,
+            "wspace": 0.1,
+        },
+    )
 
-    model_name = "model2"
     adata_cospar = load_anndata_from_path(f"data/external/larry_cospar.h5ad")
 
     for i, config in enumerate(configurations):
         data_set_name = config.download_dataset.data_set_name
-        data_set_model_pairing = f"{data_set_name}_{model_name}"
+        data_set_model_pairing = f"{data_set_name}_{model_identifier}"
         model_path = f"models/{data_set_model_pairing}"
 
         adata_pyrovelocity = load_anndata_from_path(
@@ -79,12 +77,21 @@ def estimate_time_lineage_fate_correlation(
             state_color_dict=LARRY_CELL_TYPE_COLORS,
             lineage_fate_correlation_path=plot_path,
             ylabel=f"{data_set_name}",
[…]

### 2024-08-26 reproducibility/manuscript/manuscript.qmd
fix(manuscript): remove paragraph break in align env
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
latex fixes and updates to methods. ready to be read.

@@
 \frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-dudt}\\
-   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)}, \label{eq-dsdt}.
+   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-dsdt}.
 \end{align}
@@
 activated, $t^{(1)}_{0,g}$ is the second switching time corresponding to when
-the gene expression starts to be repressed.
+the gene expression starts to be repressed, and is determined by the first
+switching time and the gene-specific switching time $\Delta \text{switching}_g$.
+The cell-gene-specific activation state $k_{cg}$ is a Bernoulli random variable
+with logits equal to the difference between the cell's shared time $t_c$ and the time $t^{(1)}_{0,g}$ when the gene expression starts to be repressed.
 
@@
     (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1)) & \text{Model 2}
-  \end{cases}
-
+  \end{cases} \\
   u^{(0)}_{cg}, s^{(0)}_{cg} &= \begin{cases}
@@
 
-Finally, we describe the ODE solution at time $\tau_{cg}$, and the observation model at those times that give rise to the observed counts:
-\begin{align}
-  \hat{u}_{cg}, \hat{s}_{cg} &= \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big) \\
+We define the ODE solution at time $\tau_{cg}$ as:
+\begin{equation}
+    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
+\end{equation}
+
+Next, we define the observation model that gives rise to the observed counts as:
+\begin{align}
   \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
@@
 denote our generative model's prediction of these unspliced and spliced expression levels.
-The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_cg, \hat{s}_cg$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.
+The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_{cg}, \hat{s}_{cg}$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.
 \textbf{Improve descriptions of how noise is modeled in the observation model.}

### 2024-08-26 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

--- 2024-08-22 reproducibility/manuscript/v2.tex
@@
 splicing process, and infer the unspliced and spliced gene expression
-level from the differential equations proposed in velocyto
-\citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj} \begin{align}
-\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right),
-   \label{eq-dudt}\\
-\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
-  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)
-  -\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
+level from the ordinary differential equation (ODEs) proposed in
+velocyto \citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj}
+\begin{align}
+\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \label{eq-dudt}\\
+   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0, \label{eq-dsdt}
+\end{align} where \(u(t), s(t)\) are the unspliced and spliced
+expression levels of a gene at time \(t\) under a transcription rate
+\(\alpha(t)\) with possible temporal dependence, splicing rate
+\(\beta\), and degradation rate \(\gamma\). We specify this model to a
+setting that depends on cell \(c\) and gene \(g\) as follows:
+\begin{align}
+\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-dudt}\\
+   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-dsdt}.
 \end{align} In the equation, the subscript \(c\) is the cell dimension,
-\(g\) is the gene dimension,
-\(\left( u\left( \tau^{(k_{cg})} \right), s\left( \tau^{(k_{cg})} \right) \right)\)
-are the unspliced and spliced expression functions given the change of
-time per cell and gene. \(\tau_{cg}\) represents the displacement of
-time per cell and gene with \begin{align}
- \tau^{(k_{cg})} &= \operatorname{softplus} \left( t_{c} - {t_{0}^{(k_{cg})}}_g \right) \\
- & = \log( 1 + \exp (t_c - {t_{0}^{(k_{cg})}}_g)), 
-\end{align} in which \(t_c\) is the shared time per cell,
-\({t_{0}^{(kcg)}}_g\) is the gene-specific switching time. Each cell and
-gene combination has its transcriptional state
+\(g\) is the gene dimension, \(\left( u_{cg}(t), s_{cg}(t) \right)\) are
+the unspliced and spliced expression functions given the change of time
+per cell and gene. We restrict attention to piecewise-constant
+\(\alpha_{cg}(t)\) to capture gene-specific activation and repression.
+We take special care to model a gene- and cell-specific switching time
+that marks a single transition from activation to repression by
+introducing a Bernoulli variable \(k_{cg}\) to model unknown activation
+state. We assume our cell-by-gene data-matrix arrive as observations of
+Poisson-counts related to the solution of the above ODEs at unknown
+times \(\tau_{cg}\), which is modeled as a relationship between an
+unknown latent time shared across each cell, \(t_c\), and unknown
+gene-specific time-offsets \(t_{0,g}\) where all read counts for a
+single cell occurred at an unknown, but shared latent time \(t_c\).
+These relative times are also used to parametrize the Bernoulli process
+for \(k_{cg}\). Importantly, we recognize that the initial conditions
+are in fact unknown.
+
+We propose and study two models: Model 1 assumes that spliced and
+unspliced concentrations are both 0 at time 0; Model 2 considers these
+initial conditions as unknowns with a log-Normal prior distribution. In
+general, the solution space of ODEs becomes much richer when considered
+over a domain of initial conditions (as opposed to a single point);
+indeed, this affords Model 2 much greater expressivity. For clarity, we
+first present the generative framework for both models, then provide
+further interpretation and intuition.
+
+First, we introduce the generative model that describes the various
+unobserved times: \begin{align}
+  % unit lognormal t_c
+  t_c &\sim \text{LogNormal}(0, 1) \\
+  % gene-specific t_0
+  t^{(0)}_{0,g} &\sim \text{LogNormal}(0, 1) \\
+  % switching time
+  \Delta \textrm{switching}_g &\sim \text{LogNormal}(0, 1) \\
+  % gene-specific t_1
+  t^{(1)}_{0,g} &= t^{(0)}_{0,g} + \Delta \textrm{switching}_g \\
+  %cell-gene-specific activation state
+  k_{cg} &\sim \text{Bernoulli}(\textrm{logits}=t_c - t^{(1)}_{0,g}) \\
+  % cell-gene-specific latent time
+  \tau_{cg} &= \text{softplus}(t_c - t^{(k_{cg})}_{0,g}).
+\end{align} Here, \(\tau_{cg}\) represents the displacement of time per
+cell and gene with \begin{align}
+ \text{softplus}(t) :=  \log( 1 + e^t).
+\end{align} Recall that \(t_c\) is the shared time per cell,
+\(t^{(k_{cg})}_{0,g}\) is the gene-specific switching time. Each cell
+and gene combination has its transcriptional state
 \(k_{cg} \in \{ 0, 1 \}\), where \(0\) indicates the activation state
 and \(1\) indicates the expression state. Each gene has two switching
-times for representing activation and repression: \({t_{0}^{(0)}}_g\) is
+times for representing activation and repression: \(t^{(0)}_{0,g}\) is
 the first switching time corresponding to when the gene expression
-starts to be activated, \({t_0^{(1)}}_g\) is the second switching time
-corresponding to when the gene expression starts to be repressed. We
-note that \(\alpha^{(1)}\) is shared for all the genes, while
-\({\alpha^{(0)}}_g\) is learned independently for each gene.
-
+starts to be activated, \(t^{(1)}_{0,g}\) is the second switching time
+corresponding to when the gene expression starts to be repressed, and is
+determined by the first switching time and the gene-specific switching
+time \(\Delta \text{switching}_g\). The cell-gene-specific activation
+state \(k_{cg}\) is a Bernoulli random variable with logits equal to the
+difference between the cell's shared time \(t_c\) and the time
+\(t^{(1)}_{0,g}\) when the gene expression starts to be repressed.
+
+Next we introduce the priors for the splicing parameters (where the
+activation rate \(\alpha\) depends on the activation state \(k_{cg}\)
+from above): \begin{align}
+  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
+  \beta_g &\sim \text{LogNormal}(0, 1) \\
+  \gamma_g &\sim \text{LogNormal}(0, 1) \\
+  \alpha_{cg} &= \begin{cases}
+    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
+    0 & \text{if } k_{cg} = 1
+  \end{cases}
+\end{align}
+\textbf{Note that $\alpha^{(1)}$ is shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
+each gene. MATT: this was in the old text, but I think $\alpha^{(1)}$ is no longer used based on conversations with Alvin?}
+
+Now, we describe the priors for the initial conditions, noting that this
+is the only difference between Model 1 and Model 2: \begin{align}
+  \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} &\sim \begin{cases}
+    (0, 0) & \text{Model 1} \\
+    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1)) & \text{Model 2}
+  \end{cases} \\
+  u^{(0)}_{cg}, s^{(0)}_{cg} &= \begin{cases}
+    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
+    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \ T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
+  \end{cases}
+\end{align}
+
+We define the ODE solution at time \(\tau_{cg}\) as: \begin{equation}
+    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
+\end{equation}
+
+Next, we define the observation model that gives rise to the observed
+counts as: \begin{align}
+  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
+  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
+  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
+  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
+  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
+  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
+  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
+  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
+  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
+  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
+\end{align} Here, we use
+\({u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}\) to denote the
+observed unspliced and spliced counts for cell \(c\) and gene \(g\). We
+use \(\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}\) to
+denote our generative model's prediction of these unspliced and spliced
+expression levels. The generative process for modeling these observed
+read counts given denoised gene transcript expression level
+\(\hat{u}_{cg}, \hat{s}_{cg}\) considers the expected number of observed
+reads for a given gene in a given cell as the number of transcripts
+times the ratio of the cell's total reads to total transcripts.
+\textbf{Improve descriptions of how noise is modeled in the observation model.}
+
+\textbf{Need to update the analytic solutions, but first need to confirm the above is correct. Also, I recommend pushing all of the below analytic solutions to the appendix.}
 The analytic solution of the differential equations to predict spliced

### 2024-08-29 reproducibility/manuscript/manuscript.qmd
updated and tersened analytic solutions methods sub-section

@@
 [@Bergen2020-pj]
+\begin{equation}
+  \label{eq-rna}
 \begin{align}
-\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \label{eq-dudt}\\
-   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0, \label{eq-dsdt}
+\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \\
+   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0,
 \end{align}
+\end{equation}
 where $u(t), s(t)$ are the unspliced and spliced expression levels of a gene at time $t$ under a transcription rate $\alpha(t)$ with possible temporal dependence, splicing rate $\beta$, and degradation rate $\gamma$.
 We specify this model to a setting that depends on cell $c$ and gene $g$ as follows:
+\begin{equation}
+    \label{eq-rna-cg}
 \begin{align}
-\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-dudt}\\
-   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-dsdt}.
+\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \\
+   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)}.
 \end{align}
+\end{equation}
 In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
@@
 We restrict attention to piecewise-constant $\alpha_{cg}(t)$ to capture gene-specific activation and repression.
-We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown activation state.
-We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times $\tau_{cg}$, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$.
+We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown transcription state.
+We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$.
 These relative times are also used to parametrize the Bernoulli process for $k_{cg}$.
-Importantly, we recognize that the initial conditions are in fact unknown.
+Importantly, we recognize that the initial conditions are, in fact, unknown.
 
@@
     \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
-    0 & \text{if } k_{cg} = 1
+    0 & \text{if } k_{cg} = 1.
   \end{cases}
 \end{align}
-\textbf{Note that $\alpha^{(1)}$ is shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
-each gene. MATT: this was in the old text, but I think $\alpha^{(1)}$ is no longer used based on conversations with Alvin?}
 
@@
 The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_{cg}, \hat{s}_{cg}$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.
-\textbf{Improve descriptions of how noise is modeled in the observation model.}
-
-\textbf{Need to update the analytic solutions, but first need to confirm the above is correct. Also, I recommend pushing all of the below analytic solutions to the appendix.}
-The analytic solution of the differential equations to predict
-spliced and unspliced gene expression given their parameters is derived by the
-authors of scVelo and a theoretical RNA velocity study
-[@Bergen2020-pj;@Li2021-qa] and given in
-Eqs. \ref{eq-solution-u}-\ref{eq-solution-s2}.
+
+The linear ODE in Eq.
+\ref{eq-rna} has an analytic solution under the assumption that the transcription rate $\alpha(t)$ is piece-wise constant, and is derived by the authors of scVelo and a theoretical RNA velocity study
+[@Bergen2020-pj;@Li2021-qa].
+For brevity, we present the solution under constant transcription rate $\alpha(t) = \alpha_0$, and note that this can be combined to assemble the solution in the piece-wise constant case.
+First, let $C = u_0 - \frac{\alpha}{\beta}$. Then, the solution to the ODE in Eq. \ref{eq-rna} is:
+\begin{equation}
+\label{eq-rna-solution}
 \begin{align}
-u\left(\tau^{\left(k_{c g}\right)}\right) 
-  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
-  \nonumber \\
-&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
-  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
-  \label{eq-solution-u}\\
-s\left(\tau^{\left(k_{c g}\right)}\right) 
-  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
-  \nonumber \\
-  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
-    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
-    \nonumber\\
-  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
-    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
-    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
-    \nonumber \\
-  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
-s\left(\tau^{\left(k_{c g}\right)}\right) 
-  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
-  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
-  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
-  \nonumber \\
-  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
-  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
-  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
-  &\qquad \beta = \gamma \label{eq-solution-s2}
+  u(t) &= \frac{\alpha}{\beta} + C e^{-\beta t} \\
+  s(t) &= \begin{cases}
+    \frac{\alpha}{\gamma} + \frac{\beta}{\gamma - \beta} C e^{-\beta t} + \Big(s_0 - \frac{\alpha}{\gamma} - \frac{\beta}{\gamma - \beta} C\Big) e^{-\gamma t} & \text{if } \beta \neq \gamma \\
+    \frac{\alpha}{\beta} + \Big(s_0 - \frac{\alpha}{\beta} + \beta C t\Big) e^{-\beta t} & \text{if } \beta = \gamma.
+  \end{cases}
 \end{align}
-
-To simplify these equations, consider the case when $k_{cg} = 0$ and 
-$\beta_g \neq \gamma_g$. Then, 
-\begin{align}
-u\left(\tau^{(0)}\right) &= {u_0^{(0)}}_g e^{-\beta_g \tau^{(0)}} \nonumber \\
-  & \hskip -24pt + \frac{{\alpha^{(0)}}_g}{\beta_g}\left(1-e^{-\beta_g \tau^{(0)}}\right), 
-  \label{eq-sol-usimp} \\
-s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
-  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}\left(1-e^{-\gamma_g \tau^{(0)}}\right) 
-  \nonumber\\
-  & \hskip -24pt +\frac{{\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g}{\gamma_g-\beta_g}
-  \left(e^{-\gamma_g \tau^{(0)}}-e^{-\beta_g \tau^{(0)}}\right). \label{eq-sol-ssimp}
-\end{align}
-When $k_{cg} = 0$ and $\beta_g = \gamma_g$, then $u\left(\tau^{(0)}\right)$ 
-has the same solution, and $s\left(\tau^{(0)}\right)$ becomes
-\begin{align}
-s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
-  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}
-    \left(1-e^{-\gamma_g \tau^{(0)}}\right) \nonumber\\
-  & \hskip -24pt - \left( {\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g \right) 
-    \tau^{(0)} e^{-\beta_g \tau^{(0)}}. \label{eq-sol-ssimp2}
-\end{align}
-When $k_{cg} = 1$ and $\beta_g \neq \gamma_g$, then
-\begin{align}
-u\left(\tau^{(1)}\right) &=u_0^{(1)} g^{e^{-\beta_g \tau^{(1)}}}, \\
-s\left(\tau^{(1)}\right) &=s_0^{(1)} e^{-\gamma_g \tau^{(1)}} \nonumber \\
-  & \hskip -24pt +\frac{-\beta_g u_0^{(1)}}{\gamma_g-\beta_g}
-  \left(e^{-\gamma_g \tau^{(1)}}-e^{-\beta_g \tau^{(1)}}\right).
-\end{align}
-When $k_{cg} = 1$ and $\beta_g = \gamma_g$, then $u\left(\tau^{(1)}\right)$ 
-has the same solution, and $s\left(\tau^{(1)}\right)$ becomes
-\begin{align}
-s\left(\tau^{(1)}\right)=s_0^{(1)}{ }_g e^{-\gamma_g \tau^{(1)}}
-  +\beta_g u_0^{(1)}{ }_g \tau^{(1)} e^{-\beta_g \tau^{(1)}}.
-\end{align}
+\end{equation}
+These solutions are the essential building block used to compute all cell-gene-specific ODE solutions for the unspliced and spliced expression levels in the generative model.
 

### 2024-08-29 reproducibility/manuscript/manuscript.qmd
bugfix with align vs aligned
updating later methods section to have consistent notation

@@
   \label{eq-rna}
-\begin{align}
+\begin{aligned}
 \frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \\
    \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0,
-\end{align}
+\end{aligned}
 \end{equation}
@@
     \label{eq-rna-cg}
-\begin{align}
-\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \\
-   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)}.
-\end{align}
+\begin{aligned}
+\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-du-cg} \\
+   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-ds-cg} .
+\end{aligned}
 \end{equation}
@@
 \begin{equation}
+  \label{eq-ODE-solve}
     \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
@@
   \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
-  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
-  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
+  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big)   \label{eq-u-hat-obs} \\
+  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big) \label{eq-s-hat-obs}
 \end{align}
@@
 \label{eq-rna-solution}
-\begin{align}
+\begin{aligned}
   u(t) &= \frac{\alpha}{\beta} + C e^{-\beta t} \\
@@
   \end{cases}
-\end{align}
+\end{aligned}
 \end{equation}
@@
 
-Given observations $\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)$,
-we would like to compute the posterior distribution over the random variables
+Given observations $\tilde{X}_{cg} = \left( u_{cg}^{\text{(obs)}}, s_{cg}^{\text{(obs)}} \right)$ over cells $1 \dots C$ and genes $1 \dots G$,
+we would like to compute the posterior distribution over the random variables $\theta := \left( \theta_1, \ \dots, \theta_C \right)$ and $\psi := \left( \psi_1, \ \dots, \psi_G \right)$, where
 \begin{align}
-\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
-\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, 
-          {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
+\theta_c &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
+\phi_g &= \left( {t_{0,g}^{(0)}}, \Delta \text{switching}_{g}, 
+          \alpha^{(0)}_{g}, \beta_{g}, \gamma_{g} \right),
 \end{align}
@@
 To benchmark Pyro&thinsp;-Velocity performance in predicting cell fate, we
-generated the posterior samples measurement
-$x_{cg}=\left(u\left(\tau^{(k_{cg})}\right),
-s\left(\tau^{(k_{cg})}\right)\right)$ or $x_{cg}=\left(u_{cg}, s_{cg} \right)$,
-$t_c$, $\beta_g$, $\gamma_g$ from the same single cell RNA-seq using $N=30$
-Monte Carlo samples from the posterior predictive distribution following
-$p(x_{cg} \vert \theta, \psi) p(\theta, \psi) \approx p(x_{cg} \vert \theta,
-\psi) q_{\phi}(\theta, \psi)$. $x_{cg}$ can be posterior samples of either
-denoised gene expression (used for phase portraits and vector field-based
-trajectory inference) or raw read counts (used for gene selection). Then we
-calculated the posterior samples of RNA velocity as
-$v_{cg}=\frac{ds(\tau^{(k_{cg})})}{d \tau^{(k{cg})}}$ based on posterior samples
-measurement of denoised gene expression $x_{cg}$ and $\beta_{g}$, $\gamma_{g}$.
+generated posterior samples of quantitites of interest in the model.
+In particular, we focus on posterior samples of $i)$ spliced and unspliced gene expression across all cells and genes, defined in Eq. \ref{eq-ODE-solve} and denoted by
+$X_{cg}=\left(u_{cg}, s_{cg} \right)$ (used for phase portraits and vector-field-based
+trajectory inference),
+$ii)$ read counts, defined in Eqs. \ref{eq-u-hat-obs} and \ref{eq-s-hat-obs} and denoted by
+$\hat{X}^{\text{(obs)}}_{cg}=\left(\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg} \right)$
+(used for gene selection),
+and
+$iii)$ RNA velocity, defined in Eq. \ref{eq-ds-cg} as the time derivative of the spliced expression, denoted by
+$v_{cg}=\frac{ds_{cg}}{d t}(X_{cg})$.
+We generate 30 Monte Carlo samples from the posterior predictive distribution
+$X, \hat{X}^{(\text{obs})}, v \sim p(X, \hat{X}^{(\text{obs})}, v \vert \theta, \psi) p(\theta, \psi) \approx q_{\phi}(\theta, \psi)$ 
+by sampling the posterior distribution of $\theta$, $\psi$ using the trained model.
+We use these samples to compute statistics on the posterior predictive distribution of our quantities of interest; these statistics are used to evaluate the model's performance and uncertainty in predicting cell fate.
 
@@
 posterior predictive samples of spliced and unspliced raw read counts, i.e.
-$\frac{1}{N n_c} \sum_i \sum_c (x_{icg}-\tilde{X}_{cg})$, where $N$ is the
-posterior sample number that is set to $30$, $i$ is the posterior sample index,
-and $n_c$ is the cell number. We first select the top $300$ genes with the
+$\mathcal{E}_g = \frac{1}{NC} \sum_{i=1}^N \sum_{c=1}^{C} \vert \tilde{X}_{cg} - \hat{X}_{cg}^{(\text{obs}), (i)} \vert$, where we use $N=30$ Monte Carlo samples.
+We first select the top $300$ genes with the
 highest negative mean absolute errors and then rank the $300$ genes based on the
@@
 \begin{align*}
-\frac{1}{N} \sum_{i} \sqrt{\sum_{g} \left(x_{icg} - \frac{1}{N}\sum_i x_{icg}\right)^2}.
+\mathcal{U}_c = \frac{1}{N} \sum_{i=1}^N \sqrt{\sum_{g=1}^G \left(\hat{X}^{\text{(obs)},(i)}_{cg} - \frac{1}{N}\sum_{i=1}^N \hat{X}^{\text{(obs)},(i)}_{cg}\right)^2}
 \end{align*}
+with $N=30$ Monte Carlo samples.
 
@@
 \begin{align*}
-\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} 
-  \vert x_{icg} - \tilde{X}_{cg} \vert,
+\frac{1}{N C G} \sum_{i=1}^{N} \sum_{c=1}^{C} \sum_{g=1}^{G} 
+  \vert X^{\text{(obs), (i)}}_{cg} - \tilde{X}_{cg} \vert,
 \end{align*}
-where $i$ is the posterior sample index, $N_i$ is the posterior sample number,
-$N_c$ is the cell number, $N_g$ is the gene number. Then it is possible to
+where $i$ is the posterior sample index, $N$ is the posterior sample number,
+$C$ is the cell number, $G$ is the gene number. Then it is possible to
 evaluate the cell fate inference performance of the models based on two metrics:

### 2024-09-03 reproducibility/manuscript/manuscript.qmd
latex label bugfix

@@
 \begin{aligned}
-\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-du-cg} \\
-   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-ds-cg} .
+\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \\
+   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} .
 \end{aligned}

### 2025-03-23 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 \usepackage{placeins}
+\usepackage[utf8]{inputenc}
+\usepackage{fontspec}
 \usepackage[labelfont=bf]{caption}
@@
 
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 \usepackage{bookmark}
@@
 velocyto \citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj}
-\begin{align}
-\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \label{eq-dudt}\\
-   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0, \label{eq-dsdt}
-\end{align} where \(u(t), s(t)\) are the unspliced and spliced
+\begin{equation}
+  \label{eq-rna}
+\begin{aligned}
+\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \\
+   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0,
+\end{aligned}
+\end{equation} where \(u(t), s(t)\) are the unspliced and spliced
 expression levels of a gene at time \(t\) under a transcription rate
@@
 setting that depends on cell \(c\) and gene \(g\) as follows:
-\begin{align}
-\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-dudt}\\
-   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-dsdt}.
-\end{align} In the equation, the subscript \(c\) is the cell dimension,
-\(g\) is the gene dimension, \(\left( u_{cg}(t), s_{cg}(t) \right)\) are
-the unspliced and spliced expression functions given the change of time
-per cell and gene. We restrict attention to piecewise-constant
-\(\alpha_{cg}(t)\) to capture gene-specific activation and repression.
-We take special care to model a gene- and cell-specific switching time
-that marks a single transition from activation to repression by
-introducing a Bernoulli variable \(k_{cg}\) to model unknown activation
-state. We assume our cell-by-gene data-matrix arrive as observations of
-Poisson-counts related to the solution of the above ODEs at unknown
-times \(\tau_{cg}\), which is modeled as a relationship between an
-unknown latent time shared across each cell, \(t_c\), and unknown
-gene-specific time-offsets \(t_{0,g}\) where all read counts for a
-single cell occurred at an unknown, but shared latent time \(t_c\).
-These relative times are also used to parametrize the Bernoulli process
-for \(k_{cg}\). Importantly, we recognize that the initial conditions
-are in fact unknown.
+\begin{equation}
+    \label{eq-rna-cg}
+\begin{aligned}
+\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \\
+   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} .
+\end{aligned}
+\end{equation} In the equation, the subscript \(c\) is the cell
+dimension, \(g\) is the gene dimension,
+\(\left( u_{cg}(t), s_{cg}(t) \right)\) are the unspliced and spliced
+expression functions given the change of time per cell and gene. We
+restrict attention to piecewise-constant \(\alpha_{cg}(t)\) to capture
+gene-specific activation and repression. We take special care to model a
+gene- and cell-specific switching time that marks a single transition
+from activation to repression by introducing a Bernoulli variable
+\(k_{cg}\) to model unknown transcription state. We assume our
+cell-by-gene data-matrix arrive as observations of Poisson-counts
+related to the solution of the above ODEs at unknown times, which is
+modeled as a relationship between an unknown latent time shared across
+each cell, \(t_c\), and unknown gene-specific time-offsets \(t_{0,g}\)
+where all read counts for a single cell occurred at an unknown, but
+shared latent time \(t_c\). These relative times are also used to
+parametrize the Bernoulli process for \(k_{cg}\). Importantly, we
+recognize that the initial conditions are, in fact, unknown.
 
@@
 We define the ODE solution at time \(\tau_{cg}\) as: \begin{equation}
+  \label{eq-ODE-solve}
     \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
@@
   \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
-  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
-  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
+  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big)   \label{eq-u-hat-obs} \\
+  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big) \label{eq-s-hat-obs}
 \end{align} Here, we use
@@
 times the ratio of the cell's total reads to total transcripts.
-\textbf{Improve descriptions of how noise is modeled in the observation model.}
-
-\textbf{Need to update the analytic solutions, but first need to confirm the above is correct. Also, I recommend pushing all of the below analytic solutions to the appendix.}
-The analytic solution of the differential equations to predict spliced
-and unspliced gene expression given their parameters is derived by the
-authors of scVelo and a theoretical RNA velocity study
-\citep{Bergen2020-pj, Li2021-qa} and given in Eqs.
-\ref{eq-solution-u}-\ref{eq-solution-s2}. \begin{align}
-u\left(\tau^{\left(k_{c g}\right)}\right) 
-  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
-  \nonumber \\
-&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
-  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
-  \label{eq-solution-u}\\
-s\left(\tau^{\left(k_{c g}\right)}\right) 
-  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
-  \nonumber \\
-  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
-    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
-    \nonumber\\
-  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
-    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
-    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
-    \nonumber \\
-  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
-s\left(\tau^{\left(k_{c g}\right)}\right) 
-  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
-  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
-  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
-  \nonumber \\
-  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
-  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
-  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
-  &\qquad \beta = \gamma \label{eq-solution-s2}
-\end{align}
-
-To simplify these equations, consider the case when \(k_{cg} = 0\) and
-\(\beta_g \neq \gamma_g\). Then, \begin{align}
-u\left(\tau^{(0)}\right) &= {u_0^{(0)}}_g e^{-\beta_g \tau^{(0)}} \nonumber \\
-  & \hskip -24pt + \frac{{\alpha^{(0)}}_g}{\beta_g}\left(1-e^{-\beta_g \tau^{(0)}}\right), 
-  \label{eq-sol-usimp} \\
-s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
-  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}\left(1-e^{-\gamma_g \tau^{(0)}}\right) 
-  \nonumber\\
-  & \hskip -24pt +\frac{{\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g}{\gamma_g-\beta_g}
-  \left(e^{-\gamma_g \tau^{(0)}}-e^{-\beta_g \tau^{(0)}}\right). \label{eq-sol-ssimp}
-\end{align} When \(k_{cg} = 0\) and \(\beta_g = \gamma_g\), then
-\(u\left(\tau^{(0)}\right)\) has the same solution, and
-\(s\left(\tau^{(0)}\right)\) becomes \begin{align}
-s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
-  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}
-    \left(1-e^{-\gamma_g \tau^{(0)}}\right) \nonumber\\
-  & \hskip -24pt - \left( {\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g \right) 
-    \tau^{(0)} e^{-\beta_g \tau^{(0)}}. \label{eq-sol-ssimp2}
-\end{align} When \(k_{cg} = 1\) and \(\beta_g \neq \gamma_g\), then
-\begin{align}
-u\left(\tau^{(1)}\right) &=u_0^{(1)} g^{e^{-\beta_g \tau^{(1)}}}, \\
-s\left(\tau^{(1)}\right) &=s_0^{(1)} e^{-\gamma_g \tau^{(1)}} \nonumber \\
-  & \hskip -24pt +\frac{-\beta_g u_0^{(1)}}{\gamma_g-\beta_g}
-  \left(e^{-\gamma_g \tau^{(1)}}-e^{-\beta_g \tau^{(1)}}\right).
-\end{align} When \(k_{cg} = 1\) and \(\beta_g = \gamma_g\), then
-\(u\left(\tau^{(1)}\right)\) has the same solution, and
-\(s\left(\tau^{(1)}\right)\) becomes \begin{align}
-s\left(\tau^{(1)}\right)=s_0^{(1)}{ }_g e^{-\gamma_g \tau^{(1)}}
-  +\beta_g u_0^{(1)}{ }_g \tau^{(1)} e^{-\beta_g \tau^{(1)}}.
-\end{align}
+
+The linear ODE in Eq. \ref{eq-rna} has an analytic solution under the
+assumption that the transcription rate \(\alpha(t)\) is piece-wise
+constant, and is derived by the authors of scVelo and a theoretical RNA
+velocity study \citep{Bergen2020-pj, Li2021-qa}. For brevity, we present
+the solution under constant transcription rate \(\alpha(t) = \alpha_0\),
+and note that this can be combined to assemble the solution in the
+piece-wise constant case. First, let \(C = u_0 - \frac{\alpha}{\beta}\).
+Then, the solution to the ODE in Eq. \ref{eq-rna} is: \begin{equation}
+\label{eq-rna-solution}
+\begin{aligned}
+  u(t) &= \frac{\alpha}{\beta} + C e^{-\beta t} \\
+  s(t) &= \begin{cases}
+    \frac{\alpha}{\gamma} + \frac{\beta}{\gamma - \beta} C e^{-\beta t} + \Big(s_0 - \frac{\alpha}{\gamma} - \frac{\beta}{\gamma - \beta} C\Big) e^{-\gamma t} & \text{if } \beta \neq \gamma \\
+    \frac{\alpha}{\beta} + \Big(s_0 - \frac{\alpha}{\beta} + \beta C t\Big) e^{-\beta t} & \text{if } \beta = \gamma.
+  \end{cases}
+\end{aligned}
+\end{equation} These solutions are the essential building block used to
+compute all cell-gene-specific ODE solutions for the unspliced and
+spliced expression levels in the generative model.
 
@@
 Given observations
-\(\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)\), we would
-like to compute the posterior distribution over the random variables
-\begin{align}
-\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
-\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, 
-          {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
+\(\tilde{X}_{cg} = \left( u_{cg}^{\text{(obs)}}, s_{cg}^{\text{(obs)}} \right)\)
+over cells \(1 \dots C\) and genes \(1 \dots G\), we would like to
+compute the posterior distribution over the random variables
+\(\theta := \left( \theta_1, \ \dots, \theta_C \right)\) and
+\(\psi := \left( \psi_1, \ \dots, \psi_G \right)\), where \begin{align}
+\theta_c &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
+\phi_g &= \left( {t_{0,g}^{(0)}}, \Delta \text{switching}_{g}, 
+          \alpha^{(0)}_{g}, \beta_{g}, \gamma_{g} \right),
 \end{align} but exact Bayesian inference is intractable in this model.
@@
 To benchmark Pyro -Velocity performance in predicting cell fate, we
-generated the posterior samples measurement
-\(x_{cg}=\left(u\left(\tau^{(k_{cg})}\right),
-s\left(\tau^{(k_{cg})}\right)\right)\) or
-\(x_{cg}=\left(u_{cg}, s_{cg} \right)\), \(t_c\), \(\beta_g\),
-\(\gamma_g\) from the same single cell RNA-seq using \(N=30\) Monte
-Carlo samples from the posterior predictive distribution following
-\(p(x_{cg} \vert \theta, \psi) p(\theta, \psi) \approx p(x_{cg} \vert \theta,
-\psi) q_{\phi}(\theta, \psi)\). \(x_{cg}\) can be posterior samples of
-either denoised gene expression (used for phase portraits and vector
-field-based trajectory inference) or raw read counts (used for gene
-selection). Then we calculated the posterior samples of RNA velocity as
-\(v_{cg}=\frac{ds(\tau^{(k_{cg})})}{d \tau^{(k{cg})}}\) based on
-posterior samples measurement of denoised gene expression \(x_{cg}\) and
-\(\beta_{g}\), \(\gamma_{g}\).
+generated posterior samples of quantitites of interest in the model. In
+particular, we focus on posterior samples of \(i)\) spliced and
+unspliced gene expression across all cells and genes, defined in Eq.
+\ref{eq-ODE-solve} and denoted by
+\(X_{cg}=\left(u_{cg}, s_{cg} \right)\) (used for phase portraits and
+vector-field-based trajectory inference), \(ii)\) read counts, defined
+in Eqs. \ref{eq-u-hat-obs} and \ref{eq-s-hat-obs} and denoted by
+\(\hat{X}^{\text{(obs)}}_{cg}=\left(\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg} \right)\)
+(used for gene selection), and \(iii)\) RNA velocity, defined in Eq.
+\ref{eq-ds-cg} as the time derivative of the spliced expression, denoted
+by \(v_{cg}=\frac{ds_{cg}}{d t}(X_{cg})\). We generate 30 Monte Carlo
+samples from the posterior predictive distribution
+\(X, \hat{X}^{(\text{obs})}, v \sim p(X, \hat{X}^{(\text{obs})}, v \vert \theta, \psi) p(\theta, \psi) \approx q_{\phi}(\theta, \psi)\)
+by sampling the posterior distribution of \(\theta\), \(\psi\) using the
+trained model. We use these samples to compute statistics on the
+posterior predictive distribution of our quantities of interest; these
+statistics are used to evaluate the model's performance and uncertainty
+in predicting cell fate.
 
@@
 unspliced raw read counts, i.e.
-\(\frac{1}{N n_c} \sum_i \sum_c (x_{icg}-\tilde{X}_{cg})\), where \(N\)
-is the posterior sample number that is set to \(30\), \(i\) is the
-posterior sample index, and \(n_c\) is the cell number. We first select
-the top \(300\) genes with the highest negative mean absolute errors and
-then rank the \(300\) genes based on the most positively correlated
-genes and the least negatively correlated genes. We use the same
-strategy for scVelo to rank the markers by the model likelihood to get
-the top 300 genes and then prioritize these genes by Pearson's
-correlation between scVelo latent time and normalized expression.
+\(\mathcal{E}_g = \frac{1}{NC} \sum_{i=1}^N \sum_{c=1}^{C} \vert \tilde{X}_{cg} - \hat{X}_{cg}^{(\text{obs}), (i)} \vert\),
+where we use \(N=30\) Monte Carlo samples. We first select the top
+\(300\) genes with the highest negative mean absolute errors and then
+rank the \(300\) genes based on the most positively correlated genes and
+the least negatively correlated genes. We use the same strategy for
+scVelo to rank the markers by the model likelihood to get the top 300
+genes and then prioritize these genes by Pearson's correlation between
+scVelo latent time and normalized expression.
 
@@
 count prediction for each cell \begin{align*}
-\frac{1}{N} \sum_{i} \sqrt{\sum_{g} \left(x_{icg} - \frac{1}{N}\sum_i x_{icg}\right)^2}.
-\end{align*}
+\mathcal{U}_c = \frac{1}{N} \sum_{i=1}^N \sqrt{\sum_{g=1}^G \left(\hat{X}^{\text{(obs)},(i)}_{cg} - \frac{1}{N}\sum_{i=1}^N \hat{X}^{\text{(obs)},(i)}_{cg}\right)^2}
+\end{align*} with \(N=30\) Monte Carlo samples.
 
@@
 combinations \begin{align*}
-\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} 
-  \vert x_{icg} - \tilde{X}_{cg} \vert,
-\end{align*} where \(i\) is the posterior sample index, \(N_i\) is the
-posterior sample number, \(N_c\) is the cell number, \(N_g\) is the gene
+\frac{1}{N C G} \sum_{i=1}^{N} \sum_{c=1}^{C} \sum_{g=1}^{G} 
+  \vert X^{\text{(obs), (i)}}_{cg} - \tilde{X}_{cg} \vert,
+\end{align*} where \(i\) is the posterior sample index, \(N\) is the
+posterior sample number, \(C\) is the cell number, \(G\) is the gene
 number. Then it is possible to evaluate the cell fate inference

### 2025-03-23 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

@@
 \allowdisplaybreaks
+
+\let\oldincludegraphics\includegraphics
+\renewcommand{\includegraphics}[2][]{\oldincludegraphics[#1,keepaspectratio]{#2}}
 \makeatletter
@@
 
-\includegraphics{figures/v2/Figure1.pdf}
+\includegraphics[width=0.98\textwidth,height=\textheight]{figures/v2/Figure1.pdf}
 
@@
 
-\includegraphics{figures/v2/Figure2.png}
+\includegraphics[width=0.98\textwidth,height=\textheight]{figures/v2/Figure2.png}
 
@@
 
-\includegraphics{figures/v2/Figure3.png}
+\includegraphics[width=0.98\textwidth,height=\textheight]{figures/v2/Figure3.png}
 
@@
 
-\includegraphics{figures/v2/Supplementary-Figure1.pdf}
+\includegraphics[width=0.98\textwidth,height=\textheight]{figures/v2/Supplementary-Figure1.pdf}
 
@@
 
-\includegraphics{figures/v2/Supplementary-Figure2.png}
+\includegraphics[width=0.98\textwidth,height=\textheight]{figures/v2/Supplementary-Figure2.png}
 
@@
 
-\includegraphics{figures/v2/Supplementary-Figure3.png}
+\includegraphics[width=0.98\textwidth,height=\textheight]{figures/v2/Supplementary-Figure3.png}
 
@@
 
-\includegraphics{figures/v2/Supplementary-Figure4.png}
+\includegraphics[width=0.98\textwidth,height=\textheight]{figures/v2/Supplementary-Figure4.png}
 
@@
 
-\includegraphics{figures/v2/Supplementary-Figure5.png}
+\includegraphics[width=0.98\textwidth,height=\textheight]{figures/v2/Supplementary-Figure5.png}
 

### 2025-03-23 reproducibility/manuscript/manuscript.qmd
fix(manuscript): set explicit figure width percentages
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ::: {#fig-model}
-![](figures/v2/Figure1.pdf)
+![](figures/v2/Figure1.pdf){width=98%}
 
@@
 ::: {#fig-model-results}
-![](figures/v2/Figure2.png)
+![](figures/v2/Figure2.png){width=98%}
 
@@
 ::: {#fig-lineage-tracing}
-![](figures/v2/Figure3.png)
+![](figures/v2/Figure3.png){width=98%}
 
@@
 ::: {#extfig-graphical-models}
-![](figures/v2/Supplementary-Figure1.pdf)
+![](figures/v2/Supplementary-Figure1.pdf){width=98%}
 
@@
 ::: {#extfig-model-results}
-![](figures/v2/Supplementary-Figure2.png)
+![](figures/v2/Supplementary-Figure2.png){width=98%}
 
@@
 ::: {#extfig-model-2-result}
-![](figures/v2/Supplementary-Figure3.png)
+![](figures/v2/Supplementary-Figure3.png){width=98%}
 
@@
 ::: {#extfig-lineage-tracing}
-![](figures/v2/Supplementary-Figure4.png)
+![](figures/v2/Supplementary-Figure4.png){width=98%}
 
@@
 ::: {#extfig-lineage-tracing-2}
-![](figures/v2/Supplementary-Figure5.png)
+![](figures/v2/Supplementary-Figure5.png){width=98%}
 

### 2025-03-23 reproducibility/manuscript/manuscript.qmd
fix(manuscript): make model formulation eqs align
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 Next we introduce the priors for the splicing parameters (where the activation rate $\alpha$ depends on the activation state $k_{cg}$ from above):
-\begin{align}
+\begin{align*}
   \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
@@
   \end{cases}
+\end{align*}
+
+Now, we describe the priors for the initial conditions, noting that these define the difference between Model 1 and Model 2, given respectively
+\begin{align*}
+  &\left(\hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}\right) \sim 
+  \begin{cases}
+    (0, 0) \\
+    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1))
+  \end{cases}
+\end{align*}
+where
+\begin{align*}
+  &\left(u^{(0)}_{cg}, s^{(0)}_{cg}\right) = \\
+  &\hskip 3pt \begin{cases}
+    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
+    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \\
+    \hskip 24pt T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
+  \end{cases}
+\end{align*}
+
+We define the ODE solution at time $\tau_{cg}$ as:
+\begin{align}
+    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, & s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \nonumber \\
+    & T_0=0, T_1=\tau_{cg} \Big). \label{eq-ODE-solve}
 \end{align}
-
-Now, we describe the priors for the initial conditions, noting that this is the only difference between Model 1 and Model 2:
-\begin{align}
-  \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} &\sim \begin{cases}
-    (0, 0) & \text{Model 1} \\
-    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1)) & \text{Model 2}
-  \end{cases} \\
-  u^{(0)}_{cg}, s^{(0)}_{cg} &= \begin{cases}
-    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
-    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \ T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
-  \end{cases}
-\end{align}
-
-We define the ODE solution at time $\tau_{cg}$ as:
-\begin{equation}
-  \label{eq-ODE-solve}
-    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
-\end{equation}
 
@@
 \begin{aligned}
-  u(t) &= \frac{\alpha}{\beta} + C e^{-\beta t} \\
-  s(t) &= \begin{cases}
-    \frac{\alpha}{\gamma} + \frac{\beta}{\gamma - \beta} C e^{-\beta t} + \Big(s_0 - \frac{\alpha}{\gamma} - \frac{\beta}{\gamma - \beta} C\Big) e^{-\gamma t} & \text{if } \beta \neq \gamma \\
-    \frac{\alpha}{\beta} + \Big(s_0 - \frac{\alpha}{\beta} + \beta C t\Big) e^{-\beta t} & \text{if } \beta = \gamma.
+  &u(t) = \frac{\alpha}{\beta} + C e^{-\beta t} \\
+  &s(t) = 
+  \begin{cases}
+    \frac{\alpha}{\gamma} + \frac{\beta}{\gamma - \beta} C e^{-\beta t} \\
+    \hskip 12pt + \Big(s_0 - \frac{\alpha}{\gamma} - \frac{\beta}{\gamma - \beta} C\Big) e^{-\gamma t} & \text{if } \beta \neq \gamma \\
+    \frac{\alpha}{\beta} \Big(s_0 - \frac{\alpha}{\beta} + \beta C t\Big) e^{-\beta t} & \text{if } \beta = \gamma.
   \end{cases}

### 2025-03-23 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

@@
 activation rate \(\alpha\) depends on the activation state \(k_{cg}\)
-from above): \begin{align}
+from above): \begin{align*}
   \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
@@
   \end{cases}
+\end{align*}
+
+Now, we describe the priors for the initial conditions, noting that
+these define the difference between Model 1 and Model 2, given
+respectively \begin{align*}
+  &\left(\hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}\right) \sim 
+  \begin{cases}
+    (0, 0) \\
+    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1))
+  \end{cases}
+\end{align*} where \begin{align*}
+  &\left(u^{(0)}_{cg}, s^{(0)}_{cg}\right) = \\
+  &\hskip 3pt \begin{cases}
+    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
+    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \\
+    \hskip 24pt T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
+  \end{cases}
+\end{align*}
+
+We define the ODE solution at time \(\tau_{cg}\) as: \begin{align}
+    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, & s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \nonumber \\
+    & T_0=0, T_1=\tau_{cg} \Big). \label{eq-ODE-solve}
 \end{align}
-
-Now, we describe the priors for the initial conditions, noting that this
-is the only difference between Model 1 and Model 2: \begin{align}
-  \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} &\sim \begin{cases}
-    (0, 0) & \text{Model 1} \\
-    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1)) & \text{Model 2}
-  \end{cases} \\
-  u^{(0)}_{cg}, s^{(0)}_{cg} &= \begin{cases}
-    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
-    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \ T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
-  \end{cases}
-\end{align}
-
-We define the ODE solution at time \(\tau_{cg}\) as: \begin{equation}
-  \label{eq-ODE-solve}
-    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
-\end{equation}
 

### 2025-05-12 nbs/development/index.qmd
docs(development): init
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Development Overview
format:
  html:
    toc: true
---

## Introduction

This section provides technical documentation for developers working on PyroVelocity. It covers the architecture, implementation details, and testing framework of the project.

## Purpose

The development documentation is intended for:

- Contributors to the PyroVelocity codebase
- Researchers extending the models or algorithms
- Developers integrating PyroVelocity into other systems

## Contents

The development documentation is organized into the following sections:

- [Architecture](architecture.qmd): Detailed information about the modular architecture of PyroVelocity, including component interfaces, implementation patterns, and design decisions.
- [Testing](testing.qmd): Information about the testing framework, validation approach, and guidelines for writing tests.

## Development Workflow

PyroVelocity follows a protocol-first architecture with all contracts defined through Protocol interfaces. The project is organized into modular components that can be composed to create different model configurations.

For more information on contributing to PyroVelocity, please see the [Contributing](../about/contributing.qmd) guide.

### 2025-05-12 nbs/development/architecture.qmd
docs(development): init architecture
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Architecture
format:
  html:
    toc: true
---

## Overview

PyroVelocity has a modular architecture that separates concerns into distinct components. This document describes the architecture of the PyroVelocity codebase, focusing on the modular implementation in `src/pyrovelocity/models/modular`.

## Protocol-First Design

PyroVelocity follows a protocol-first architecture with all contracts defined through Protocol interfaces. Components implement these interfaces directly without inheritance, promoting composition over inheritance.

## Core Components

### Model Components

The modular implementation consists of the following key components:

- **Dynamics Models**: Implement the mathematical models of RNA dynamics
  - `StandardDynamicsModel`: The standard implementation of RNA dynamics
  - `LegacyDynamicsModel`: Compatibility implementation matching the legacy model

- **Prior Models**: Define the prior distributions for model parameters
  - `LogNormalPriorModel`: Uses log-normal distributions for parameters

- **Likelihood Models**: Define the observation models
  - `PoissonLikelihoodModel`: Uses Poisson distributions for RNA counts
  - `LegacyLikelihoodModel`: Compatibility implementation matching the legacy model

- **Observation Models**: Handle the interface between data and model
  - `StandardObservationModel`: The standard implementation for observations

- **Guide Factories**: Create guide functions for variational inference
  - `AutoGuideFactory`: Creates guides for the modular implementation
  - `LegacyAutoGuideFactory`: Compatibility implementation matching the legacy guide

### Utility Modules

The architecture includes several utility modules:

- `core_utils.py`: Core utility functions used across components
- `context_utils.py`: Utilities for managing Pyro contexts
- `pyro_utils.py`: Utilities specific to Pyro functionality

## Integration with AnnData

The modular implementation integrates directly with AnnData through methods in the `PyroVelocityModel` class, eliminating the need for adapter classes.

## Validation Framework

The architecture includes a validation framework to ensure that the modular implementation can replicate the behavior of the legacy implementation.

## Future Directions

The architecture is designed to be extensible, allowing for the addition of new components and features while maintaining backward compatibility with existing workflows.

### 2025-05-12 nbs/development/testing.qmd
docs(development): init testing
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Testing Framework
format:
  html:
    toc: true
---

## Overview

PyroVelocity uses a comprehensive testing framework to ensure code quality and correctness. This document describes the testing approach, tools, and guidelines for writing tests.

## Testing Philosophy

The testing philosophy of PyroVelocity emphasizes:

- Creating simple instances of real objects over using mocking tools
- Writing tests alongside code development
- Ensuring tests provide reasonable constraints on the code
- Systematic validation of model implementations

## Test Organization

Tests are organized in the following structure:

- `src/pyrovelocity/tests/`: Root directory for all tests
  - `models/`: Tests for model components
  - `plots/`: Tests for plotting functions
  - `utils/`: Tests for utility functions
  - `data/`: Test data fixtures

## Validation Framework

PyroVelocity includes a validation framework (PRD-678-VALID) to verify that the modular model can replicate the legacy model. Key aspects include:

- Direct comparison of model outputs
- Validation using synthetic data from test fixtures
- Validation using downsampled real datasets
- Systematic addressing of implementation differences

## Writing Tests

When writing tests for PyroVelocity:

1. Use pytest fixtures for test data and common setup
2. Create simple instances of real objects rather than mocks
3. Test both normal operation and edge cases
4. Use xdoctests with proper type annotations
5. Ensure tests are self-contained with synthetic data

## Running Tests

To run tests:

```bash
source .venv/bin/activate
pytest src/pyrovelocity/tests/
```

To run specific test modules:

```bash
source .venv/bin/activate
pytest src/pyrovelocity/tests/models/test_model.py
```

## Profiling and Benchmarking

PyroVelocity uses Scalene for profiling and benchmarking:

```bash
source .venv/bin/activate
scalene --profile-all src/pyrovelocity/tests/models/test_model.py
```

## Continuous Integration

Tests are automatically run as part of the CI/CD pipeline defined in `.github/workflows/cid.yaml`.

### 2025-05-12 nbs/development/architecture.qmd
docs(architecture): add diagrams, module descs, and roadmap
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: Architecture
+title: "System Architecture and Design: PyroVelocity"
 format:
   html:
+    mermaid-format: js
+    mermaid:
+      theme: dark
     toc: true
+    toc-depth: 3
+    number-sections: true
+    code-fold: true
 ---
 
-## Overview
+# Introduction {#sec-introduction}
 
-PyroVelocity has a modular architecture that separates concerns into distinct components.
-This document describes the architecture of the PyroVelocity codebase, focusing on the modular implementation in `src/pyrovelocity/models/modular`.
+## Purpose {#sec-purpose}
 
-## Protocol-First Design
+This document describes the architecture and design of PyroVelocity's modular implementation.
+It serves as both a reference for ongoing development and as a resource for new developers and users who may be interested in extending the functionality of existing models or creating and comparing new ones.
+As such, this document will evolve alongside the implementation.
 
-PyroVelocity follows a protocol-first architecture with all contracts defined through Protocol interfaces.
-Components implement these interfaces directly without inheritance, promoting composition over inheritance.
+## Scope {#sec-scope}
 
-## Core Components
+This document covers:
 
-### Model Components
+- High-level system architecture
+- Key components and their interactions
+- Data flow and processing pipelines
+- Major design decisions and their rationales
+- Technical constraints and quality attributes
+
+The focus is currently on the modular model implementation in `src/pyrovelocity/models/modular`. Additional sections will be added as their architecture stabilizes.
+
+## Definitions {#sec-definitions}
+
+| Term | Definition |
+|------|------------|
+| RNA Velocity | The time derivative of gene expression, representing the rate of change of mRNA abundance |
+| AnnData | A data structure for annotated data matrices, commonly used in single-cell genomics |
+| Protocol | A Python typing construct that defines an interface without implementation |
+| Pyro | A probabilistic programming framework built on PyTorch |
+| SVI | Stochastic Variational Inference, a method for approximate Bayesian inference |
+
+# System Overview {#sec-system-overview}
+
+## Vision {#sec-vision}
+
+PyroVelocity aims to provide a robust, flexible, and extensible framework for probabilistic RNA velocity analysis.
+The modular architecture enables researchers to customize and extend the model while maintaining a consistent interface.
+
+## Architectural Style {#sec-architectural-style}
+
+PyroVelocity follows a Protocol-First architecture, chosen to maximize flexibility and composability. This style enables:
+
+1. **Composition over inheritance**: Components implement Protocol interfaces directly without inheritance hierarchies
+2. **Loose coupling**: Components interact through well-defined interfaces
+3. **Testability**: Components can be tested in isolation
+4. **Extensibility**: New components can be added without modifying existing code
+
+## Context Diagram {#sec-context-diagram}
+
+```{mermaid}
+graph TB
+    User["Data Scientist / Researcher"] -->|uses| Library["PyroVelocity"]
+    Library -->|depends on| PyTorch["PyTorch"]
+    Library -->|depends on| Pyro["Pyro"]
+    Library -->|processes| AnnData["AnnData Objects"]
+    Library -->|produces| Velocity["RNA Velocity"]
+    Library -->|integrates with| Scanpy["Scanpy"]
+    Library -->|integrates with| Scvelo["scVelo"]
+
+    class User user;
+    class Library library;
+    class PyTorch,Pyro,Scanpy,Scvelo external;
+    class AnnData,Velocity data;
+```
+
+## Key Principles {#sec-key-principles}
+
+The following principles guide architectural decisions for PyroVelocity:
+
+1. **Protocol-First Design**: All contracts are defined through Protocol interfaces, enabling structural typing and composition over inheritance.
+2. **Separation of Concerns**: Each component has a single responsibility, making the system more maintainable and testable.
+3. **Immutable State**: The system uses immutable data structures where possible to reduce side effects and improve reasoning about code.
+4. **Type Safety**: The system uses static typing and runtime type checking to catch errors early and provide better development support.
+
+# Structural View {#sec-structural-view}
+
+## Component Architecture {#sec-component-architecture}
+
+The modular implementation is composed of the following major components:
+
+```{mermaid}
+graph TB
+    Main["PyroVelocityModel"] -->|contains| Dynamics["Dynamics Models"]
+    Main -->|contains| Priors["Prior Models"]
+    Main -->|contains| Likelihoods["Likelihood Models"]
+    Main -->|contains| Observations["Observation Models"]
+    Main -->|contains| Guides["Inference Guides"]
+    Main -->|contains| Registry["Registry System"]
+    Main -->|contains| Factory["Factory System"]
+
+    Dynamics -->|implements| DynamicsInterface["DynamicsModel Protocol"]
+    Priors -->|implements| PriorsInterface["PriorModel Protocol"]
+    Likelihoods -->|implements| LikelihoodsInterface["LikelihoodModel Protocol"]
+    Observations -->|implements| ObservationsInterface["ObservationModel Protocol"]
+    Guides -->|implements| GuidesInterface["InferenceGuide Protocol"]
+
+    Factory -->|uses| Registry
+    Factory -->|creates| Main
+
+    class Main main;
+    class Dynamics,Priors,Likelihoods,Observations,Guides component;
+    class DynamicsInterface,PriorsInterface,LikelihoodsInterface,ObservationsInterface,GuidesInterface interface;
+    class Registry,Factory system;
+```
+
+### Core Components {#sec-core-components}
 
@@
 
-- **Dynamics Models**: Implement the mathematical models of RNA dynamics
-  - `StandardDynamicsModel`: The standard implementation of RNA dynamics
-  - `LegacyDynamicsModel`: Compatibility implementation matching the legacy model
+#### Dynamics Models {#sec-dynamics-models}
 
-- **Prior Models**: Define the prior distributions for model parameters
-  - `LogNormalPriorModel`: Uses log-normal distributions for parameters
+Dynamics models implement the mathematical models of RNA dynamics, defining how unspliced and spliced RNA counts evolve over time.
 
-- **Likelihood Models**: Define the observation models
-  - `PoissonLikelihoodModel`: Uses Poisson distributions for RNA counts
-  - `LegacyLikelihoodModel`: Compatibility implementation matching the legacy model
+- **StandardDynamicsModel**: The standard implementation of RNA dynamics
+- **LegacyDynamicsModel**: Compatibility implementation matching the legacy model
 
-- **Observation Models**: Handle the interface between data and model
-  - `StandardObservationModel`: The standard implementation for observations
+```{mermaid}
+classDiagram
+    class DynamicsModel {
+        <<Protocol>>
+        +forward(context) Dict
+        +steady_state(alpha, beta, gamma) Tuple
+    }
 
-- **Guide Factories**: Create guide functions for variational inference
-  - `AutoGuideFactory`: Creates guides for the modular implementation
-  - `LegacyAutoGuideFactory`: Compatibility implementation matching the legacy guide
+    class StandardDynamicsModel {
+        +name: str
+        +description: str
+        +shared_time: bool
+        +t_scale_on: bool
+        +forward(context) Dict
+        +steady_state(alpha, beta, gamma) Tuple
+    }
 
-### Utility Modules
+    class LegacyDynamicsModel {
+        +name: str
+        +description: str
+        +shared_time: bool
+        +t_scale_on: bool
+        +forward(context) Dict
+        +steady_state(alpha, beta, gamma) Tuple
+    }
 
-The architecture includes several utility modules:
+    DynamicsModel <|.. StandardDynamicsModel : implements
+    DynamicsModel <|.. LegacyDynamicsModel : implements
+```
 
-- `core_utils.py`: Core utility functions used across components
-- `context_utils.py`: Utilities for managing Pyro contexts
-- `pyro_utils.py`: Utilities specific to Pyro functionality
+#### Prior Models {#sec-prior-models}
 
-## Integration with AnnData
+Prior models define the prior distributions for model parameters, using Pyro's probabilistic programming framework.
 
-The modular implementation integrates directly with AnnData through methods in the `PyroVelocityModel` class, eliminating the need for adapter classes.
+- **LogNormalPriorModel**: Uses log-normal distributions for parameters
 
-## Validation Framework
+#### Likelihood Models {#sec-likelihood-models}
 
-The architecture includes a validation framework to ensure that the modular implementation can replicate the behavior of the legacy implementation.
+Likelihood models define the observation models, relating expected RNA counts to observed counts.
 
-## Future Directions
+- **PoissonLikelihoodModel**: Uses Poisson distributions for RNA counts
+- **LegacyLikelihoodModel**: Compatibility implementation matching the legacy model
 
-The architecture is designed to be extensible, allowing for the addition of new components and features while maintaining backward compatibility with existing workflows.
+#### Observation Models {#sec-observation-models}
+
+Observation models handle the interface between data and model, preprocessing and transforming data.
+
+- **StandardObservationModel**: The standard implementation for observations
+
+#### Guide Factories {#sec-guide-factories}
+
+Guide factories create guide functions for variational inference.
+
+- **AutoGuideFactory**: Creates guides for the modular implementation
+- **LegacyAutoGuideFactory**: Compatibility implementation matching the legacy guide
+
+### Registry System {#sec-registry-system}
+
+The registry system enables component registration and retrieval, implementing the registry pattern.
+
+```{mermaid}
+classDiagram
+    class Registry~T~ {
+        <<Generic>>
+        +Dict _registry
+        +register(name) Callable
+        +get(name) T
+        +create(name, **kwargs) T
+        +list_available() List
+        +clear()
+    }
+
+    class DynamicsModelRegistry {
+        +Dict _registry
+    }
+
+    class PriorModelRegistry {
+        +Dict _registry
+    }
+
+    class LikelihoodModelRegistry {
+        +Dict _registry
+    }
+
+    class ObservationModelRegistry {
+        +Dict _registry
+    }
+
+    class InferenceGuideRegistry {
+        +Dict _registry
+    }
+
+    Registry <|-- DynamicsModelRegistry : extends
+    Registry <|-- PriorModelRegistry : extends
+    Registry <|-- LikelihoodModelRegistry : extends
+    Registry <|-- ObservationModelRegistry : extends
+    Registry <|-- InferenceGuideRegistry : extends
+```
+
+### Factory System {#sec-factory-system}
+
+The factory system creates components and models from configurations, implementing the factory pattern.
+
+```{mermaid}
+classDiagram
+    class ComponentFactory {
+        +create_component(config, component_type) Component
+        +create_component_from_dict(config_dict, component_type) Component
+        +create_dynamics_model(config) DynamicsModel
+        +create_prior_model(config) PriorModel
+        +create_likelihood_model(config) LikelihoodModel
+        +create_observation_model(config) ObservationModel
+        +create_inference_guide(config) InferenceGuide
+    }
+
+    class ModelConfig {
+        +ComponentConfig dynamics_model
+        +ComponentConfig prior_model
+        +ComponentConfig likelihood_model
+        +ComponentConfig observation_model
+        +ComponentConfig inference_guide
+        +Dict metadata
+    }
+
+    class ComponentConfig {
+        +str name
+        +Dict params
+    }
+
+    ComponentFactory --> ModelConfig : uses
+    ComponentFactory --> ComponentConfig : uses
+```
+
+## Module Organization {#sec-module-organization}
+
+The modular implementation follows this organization:
+
+```
+src/pyrovelocity/models/modular/
+├── __init__.py
+├── comparison.py
+├── components/
+│   ├── __init__.py
+│   ├── dynamics.py
+│   ├── guides.py
+│   ├── likelihoods.py
+│   ├── observations.py
+│   └── priors.py
+├── config.py
+├── constants.py
+├── data/
+│   ├── __init__.py
+│   └── anndata.py
+├── factory.py
+├── inference/
+│   ├── __init__.py
+│   ├── config.py
+│   ├── mcmc.py
+│   ├── posterior.py
+│   ├── svi.py
+│   └── unified.py
+├── interfaces.py
+├── model.py
+├── registry.py
+├── selection.py
+└── utils/
+    ├── __init__.py
+    ├── context_utils.py
+    ├── core_utils.py
+    └── pyro_utils.py
+```
+
+# Behavioral View {#sec-behavioral-view}
+
+## Key Workflows {#sec-key-workflows}
+
+### Model Creation Workflow {#sec-model-creation-workflow}
+
+```{mermaid}
+sequenceDiagram
+    participant User
+    participant Factory as ComponentFactory
+    participant Registry as Registry System
+    participant Components as Component Implementations
+    participant Model as PyroVelocityModel
+
+    User->>Factory: create_model_from_config(config)
+    activate Factory
+    Factory->>Factory: Extract component configs
+
+    Factory->>Registry: create(dynamics_model.name, **dynamics_model.params)
+    activate Registry
+    Registry->>Components: Instantiate DynamicsModel
+    Components-->>Registry: Return instance
+    Registry-->>Factory: Return DynamicsModel
+    deactivate Registry
+
+    Factory->>Registry: create(prior_model.name, **prior_model.params)
+    activate Registry
+    Registry->>Components: Instantiate PriorModel
+    Components-->>Registry: Return instance
+    Registry-->>Factory: Return PriorModel
+    deactivate Registry
+
+    Factory->>Registry: create(likelihood_model.name, **likelihood_model.params)
+    activate Registry
+    Registry->>Components: Instantiate LikelihoodModel
+    Components-->>Registry: Return instance
+    Registry-->>Factory: Return LikelihoodModel
+    deactivate Registry
+
+    Factory->>Registry: create(observation_model.name, **observation_model.params)
+    activate Registry
+    Registry->>Components: Instantiate ObservationModel
+    Components-->>Registry: Return instance
+    Registry-->>Factory: Return ObservationModel
+    deactivate Registry
+
+    Factory->>Registry: create(inference_guide.name, **inference_guide.params)
+    activate Registry
+    Registry->>Components: Instantiate InferenceGuide
+    Components-->>Registry: Return instance
+    Registry-->>Factory: Return InferenceGuide
+    deactivate Registry
+
+    Factory->>Model: Create PyroVelocityModel with components
+    Model-->>Factory: Return model instance
+    Factory-->>User: Return model
+    deactivate Factory
+```
+
+### Model Execution Workflow {#sec-model-execution-workflow}
+
+```{mermaid}
+sequenceDiagram
+    participant User
+    participant Model as PyroVelocityModel
+    participant Observation as ObservationModel
+    participant Prior as PriorModel
+    participant Dynamics as DynamicsModel
+    participant Likelihood as LikelihoodModel
+
+    User->>Model: forward(u_obs, s_obs)
+    activate Model
+    Model->>Model: Initialize context
+
+    Model->>Observation: forward(context)
+    activate Observation
+    Observation->>Observation: Preprocess data
+    Observation-->>Model: Updated context
+    deactivate Observation
+
+    Model->>Prior: forward(context)
+    activate Prior
+    Prior->>Prior: Sample parameters
+    Prior-->>Model: Updated context with parameters
+    deactivate Prior
+
+    Model->>Dynamics: forward(context)
+    activate Dynamics
+    Dynamics->>Dynamics: Compute expected counts
+    Dynamics-->>Model: Updated context with expected counts
+    deactivate Dynamics
+
+    Model->>Likelihood: forward(context)
+    activate Likelihood
+    Likelihood->>Likelihood: Define likelihoods and observe data
+    Likelihood-->>Model: Updated context with likelihoods
+    deactivate Likelihood
+
+    Model-->>User: Return results
+    deactivate Model
+```
+
+## Component Lifecycle {#sec-component-lifecycle}
+
+```{mermaid}
+stateDiagram-v2
+    [*] --> Registration: Component class defined
+    Registration --> Configuration: User creates config
+    Configuration --> Instantiation: Factory creates instance
+    Instantiation --> Composition: Added to PyroVelocityModel
+    Composition --> Execution: Model executed
+    Execution --> [*]
+```
+
+# Architectural Decisions {#sec-architectural-decisions}
+
+## ADR-1: Protocol-First Architecture {#sec-adr-1}
+
+### Status
+
+Accepted
+
+### Context
+
+The initial implementation used a mix of abstract base classes and Protocol interfaces, creating architectural inconsistency and forcing inheritance hierarchies.
+
+### Decision
+
+Adopt a Protocol-First architecture that completely removes base classes and standardizes exclusively on Protocol interfaces, with utility functions replacing base class functionality.
+
+### Consequences
+
+- Complete separation between interface and implementation
+- Maximum flexibility in component implementation
+- Full alignment with functional programming principles
+- Zero coupling between components
+- Greatly simplified testing and extension
+- Temporary code duplication during transition, gradually reduced as utility functions are extracted
+
+## ADR-2: Context-Based Component Communication {#sec-adr-2}
+
+### Status
+
+Accepted
+
+### Context
+
+Components need to share data and state during model execution.
+
+### Decision
+
+Use a shared context dictionary that is passed between components during model execution, with each component updating the context with its outputs.
+
+### Consequences
+
+- Flexible data sharing without tight coupling
+- Clear data flow through the system
+- Easy to extend with new data
+- Potential for inconsistent context structure if not properly documented
+
+# Cross-cutting Concerns {#sec-cross-cutting-concerns}
+
+## Performance Considerations {#sec-performance-considerations}
+
+PyroVelocity addresses performance through the following strategies:
+
+- **Vectorized Operations**: Using PyTorch's vectorized operations for efficient computation
+- **Batched Processing**: Supporting batch processing for large datasets
+- **GPU Acceleration**: Leveraging PyTorch's GPU support for accelerated computation
+- **Analytical Solutions**: Using analytical solutions where possible to avoid numerical integration
+
+## Error Handling {#sec-error-handling}
+
+The library follows these error handling principles:
+
+- **Early Validation**: Validating inputs early to catch errors before computation
+- **Informative Error Messages**: Providing clear error messages with context
+- **Graceful Degradation**: Falling back to simpler models when advanced features fail
+
+## Extensibility {#sec-extensibility}
+
+The library is designed for extensibility through:
+
+- **Protocol Interfaces**: Defining clear contracts for component implementations
+- **Registry System**: Enabling dynamic registration of new components
+- **Factory System**: Creating components from configurations
+- **Composition**: Building complex models from simple components
+
+# Technical Debt and Evolution {#sec-technical-debt}
+
+## Known Limitations {#sec-known-limitations}
+
+The current architecture has the following known limitations:
+
+1. **Code Duplication**: Some duplication exists across component implementations
+2. **Limited Documentation**: Documentation of component interfaces and behaviors is incomplete
+3. **Validation Coverage**: Not all components have comprehensive validation against the legacy implementation
+
+## Planned Improvements {#sec-planned-improvements}
+
+The following architectural improvements are planned:
+
+| Improvement | Priority | Timeline |
+|-------------|----------|----------|
+| Extract utility functions from duplicated code | High | Short-term |
+| Improve documentation of component interfaces | High | Short-term |
+| Expand validation coverage | Medium | Medium-term |
+| Implement additional dynamics models | Medium | Medium-term |
+| Enhance integration with AnnData | Medium | Medium-term |
+
+# Implementation References {#sec-implementation-references}
+
+## Key Source Files {#sec-key-source-files}
+
+| Component | Primary Modules | Location |
+|-----------|----------------|----------|
+| Core Model | model.py | `src/pyrovelocity/models/modular/model.py` |
+| Interfaces | interfaces.py | `src/pyrovelocity/models/modular/interfaces.py` |
+| Dynamics Models | dynamics.py | `src/pyrovelocity/models/modular/components/dynamics.py` |
+| Prior Models | priors.py | `src/pyrovelocity/models/modular/components/priors.py` |
+| Likelihood Models | likelihoods.py | `src/pyrovelocity/models/modular/components/likelihoods.py` |
+| Observation Models | observations.py | `src/pyrovelocity/models/modular/components/observations.py` |
+| Guide Factories | guides.py | `src/pyrovelocity/models/modular/components/guides.py` |
+| Registry System | registry.py | `src/pyrovelocity/models/modular/registry.py` |
+| Factory System | factory.py | `src/pyrovelocity/models/modular/factory.py` |
+
+## External Dependencies {#sec-external-dependencies}
+
+| Dependency | Purpose | Impact |
+|------------|---------|--------|
+| PyTorch | Deep learning framework | Core computation engine |
+| Pyro | Probabilistic programming | Core inference engine |
+| AnnData | Data structure | Core data representation |
+| Scanpy | Single-cell analysis | Data preprocessing and visualization |
+| scVelo | RNA velocity analysis | Velocity computation and visualization |
+| beartype | Runtime type checking | Type safety |
+| jaxtyping | Type annotations for arrays | Shape information in type annotations |
+| hydra-zen | Configuration management | Type-safe configuration |

### 2025-05-12 nbs/development/architecture.qmd
docs(architecture): update title
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: "System Architecture and Design: PyroVelocity"
+title: "System Architecture and Design"
 format:

### 2025-05-12 nbs/development/testing.qmd
docs(development): update testing docs
- note integration of bdd docs

Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: Testing Framework
+title: Testing Framework and Reference
 format:
@@
     toc: true
+    toc-depth: 3
 ---
 
-## Overview
+# PyroVelocity Testing
 
-PyroVelocity uses a comprehensive testing framework to ensure code quality and correctness. This document describes the testing approach, tools, and guidelines for writing tests.
+This documentation describes the testing approach and infrastructure for the `pyrovelocity` library.
+Our testing strategy combines traditional pytest unit/integration testing with Behavior-Driven Development (BDD) using pytest-bdd to ensure both technical correctness and alignment with user requirements.
+
+## Directory Structure
+
+The tests are organized to mirror the source code structure while separating unit tests from BDD feature definitions and step implementations:
+
+```
+src/pyrovelocity/tests/
+├── __init__.py
+├── conftest.py                # Shared pytest fixtures and configuration
+├── fixtures/                  # Test data generation utilities
+├── models/                    # Tests for model components
+│   ├── __init__.py
+│   ├── modular/               # Tests for modular implementation
+│   │   ├── __init__.py
+│   │   ├── conftest.py        # Modular-specific fixtures
+│   │   ├── components/        # Unit tests for component implementations
+│   │   │   ├── test_dynamics.py
+│   │   │   ├── test_guides.py
+│   │   │   ├── test_likelihoods.py
+│   │   │   ├── test_observations.py
+│   │   │   └── test_priors.py
+│   │   ├── integration/       # BDD step implementations
+│   │   │   ├── __init__.py
+│   │   │   ├── conftest.py    # BDD-specific fixtures
+│   │   │   ├── test_dynamics_model_steps.py
+│   │   │   ├── test_model_steps.py
+│   │   │   └── test_prior_model_steps.py
+│   │   ├── selection/         # Tests for model selection
+│   │   ├── utils/             # Tests for utility modules
+│   │   ├── test_anndata_integration.py
+│   │   ├── test_comparison.py
+│   │   ├── test_factory.py
+│   │   ├── test_interfaces.py
+│   │   ├── test_model.py
+│   │   ├── test_registry.py
+│   │   └── test_unified_config.py
+│   ├── jax/                   # Tests for JAX implementation (not yet documented)
+│   └── test_*.py              # Tests for legacy implementation
+├── features/                  # BDD feature specifications
+│   └── models/
+│       └── modular/
+│           ├── dynamics_model.feature
+│           ├── guide_model.feature
+│           ├── likelihood_model.feature
+│           ├── model.feature
+│           ├── observation_model.feature
+│           └── prior_model.feature
+├── plots/                     # Tests for plotting functions
+├── utils/                     # Tests for utility functions
+├── validation/                # Tests for validation framework
+└── test_*.py                  # Package-level tests
+```
 
@@
 
-The testing philosophy of PyroVelocity emphasizes:
+Our testing approach implements a dual strategy that addresses both implementation details and user-facing behavior:
 
-- Creating simple instances of real objects over using mocking tools
-- Writing tests alongside code development
-- Ensuring tests provide reasonable constraints on the code
-- Systematic validation of model implementations
+1. **Unit and Integration Tests** (`tests/models/modular/*.py`):
+   - Traditional pytest-based tests focusing on implementation details
+   - Verify internal functions, edge cases, and error handling
+   - Target high code coverage and technical correctness
+   - Validate that individual components work as expected in isolation
 
-## Test Organization
+2. **Behavior-Driven Tests**:
+   - **Feature Definitions** (`tests/features/models/modular/*.feature`):
+     - User-centric specifications written in Gherkin syntax
+     - Focus on requirements and use cases from the user's perspective
+     - Serve as living documentation of expected behavior
+   - **Feature Implementations** (`tests/models/modular/integration/*.py`):
+     - Connect Gherkin specifications to actual code
+     - Test how components work together to deliver user-facing functionality
+     - Validate that the system behaves according to specifications
 
-Tests are organized in the following structure:
+This approach follows Acceptance Test-Driven Development (ATDD), where feature files are written before implementation to document requirements and guide development.
 
-- `src/pyrovelocity/tests/`: Root directory for all tests
-  - `models/`: Tests for model components
-  - `plots/`: Tests for plotting functions
-  - `utils/`: Tests for utility functions
-  - `data/`: Test data fixtures
+### Key Principles
 
-## Validation Framework
+PyroVelocity's testing philosophy emphasizes:
 
-PyroVelocity includes a validation framework (PRD-678-VALID) to verify that the modular model can replicate the legacy model. Key aspects include:
-
-- Direct comparison of model outputs
-- Validation using synthetic data from test fixtures
-- Validation using downsampled real datasets
-- Systematic addressing of implementation differences
-
-## Writing Tests
-
-When writing tests for PyroVelocity:
-
-1. Use pytest fixtures for test data and common setup
-2. Create simple instances of real objects rather than mocks
-3. Test both normal operation and edge cases
-4. Use xdoctests with proper type annotations
-5. Ensure tests are self-contained with synthetic data
+- **Real Objects Over Mocks**: Creating simple instances of real objects rather than using mocking tools
+- **Test-Driven Development**: Writing tests alongside code development
+- **Reasonable Constraints**: Ensuring tests provide reasonable constraints on the code
+- **Systematic Validation**: Comprehensive validation of model implementations
+- **Protocol-First Testing**: Testing against interfaces rather than concrete implementations
 
@@
 
-To run tests:
+### All Tests
 
@@
 source .venv/bin/activate
-pytest src/pyrovelocity/tests/
+pytest src/pyrovelocity/tests
 ```
 
-To run specific test modules:
+### Only Modular Implementation Tests
 
@@
 source .venv/bin/activate
-pytest src/pyrovelocity/tests/models/test_model.py
+pytest src/pyrovelocity/tests/models/modular
 ```
 
-## Profiling and Benchmarking
+### Only BDD Tests
+
+```bash
+source .venv/bin/activate
+pytest src/pyrovelocity/tests/models/modular/integration
+```
+
+### Specific Component Tests
+
+```bash
+source .venv/bin/activate
+pytest src/pyrovelocity/tests/models/modular/components/test_dynamics.py
+```
+
+### Running with Coverage
+
+```bash
+source .venv/bin/activate
+pytest src/pyrovelocity/tests --cov=pyrovelocity --cov-report=term-missing
+```
+
+### Profiling and Benchmarking
 
@@
 
+## Test Development Workflow
+
+Our recommended workflow follows these steps:
+
+1. **Specify Behavior**: Create or update a feature file in `features/` for new functionality
+2. **Implement Step Definitions**: Create/update step definitions in the subpackage's `integration/` directory
+3. **Add Unit Tests**: Create detailed unit tests in the subpackage's test modules
+4. **Implement Source Code**: Develop the actual functionality in the source code
+5. **Update Documentation**: Update this document's reference tables when adding new tests
+
+## Validation Framework
+
+PyroVelocity includes a validation framework (PRD-678-VALID) to verify that the modular model can replicate the legacy model. Key aspects include:
+
+- Direct comparison of model outputs
+- Validation using synthetic data from test fixtures
+- Validation using downsampled real datasets (e.g., `tests/data/preprocessed_pancreas_50_7.json`)
+- Systematic addressing of implementation differences
+
+The validation framework is implemented in the `src/pyrovelocity/tests/validation/` directory and includes tests for:
+
+- Direct comparison of model outputs
+- Metrics for quantifying differences
+- Visualization of differences
+
+## Feature Specifications Reference
+
+The table below tracks all feature specifications defined in our BDD feature files for the modular implementation. This serves as a living document of the intended functionality for the library.
+
+| Module | Feature | Status | Description |
+|--------|---------|--------|-------------|
+| dynamics_model | Standard dynamics model computes expected counts | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes expected unspliced and spliced counts |
+| dynamics_model | Standard dynamics model computes steady state | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes steady state values |
+| dynamics_model | Legacy dynamics model matches legacy implementation | ✅ Implemented | Verify that the LegacyDynamicsModel matches the legacy implementation |
+| dynamics_model | Dynamics model handles edge cases | ✅ Implemented | Verify that the dynamics model handles edge cases gracefully |
+| dynamics_model | Dynamics model with library size correction | ✅ Implemented | Verify that the dynamics model correctly applies library size correction |
+| prior_model | LogNormal prior model creates stochastic nodes | ✅ Implemented | Verify that the LogNormalPriorModel creates appropriate stochastic nodes |
+| likelihood_model | Poisson likelihood model creates observation nodes | ✅ Implemented | Verify that the PoissonLikelihoodModel creates appropriate observation nodes |
+| likelihood_model | Legacy likelihood model matches legacy implementation | ✅ Implemented | Verify that the LegacyLikelihoodModel matches the legacy implementation |
+| observation_model | Standard observation model preprocesses data | ✅ Implemented | Verify that the StandardObservationModel correctly preprocesses data |
+| guide_model | Auto guide factory creates guide | ✅ Implemented | Verify that the AutoGuideFactory creates an appropriate guide |
+| guide_model | Legacy auto guide factory matches legacy implementation | ✅ Implemented | Verify that the LegacyAutoGuideFactory matches the legacy implementation |
+| model | PyroVelocity model integrates components | ✅ Implemented | Verify that the PyroVelocityModel correctly integrates all components |
+| model | PyroVelocity model trains on AnnData | ✅ Implemented | Verify that the PyroVelocityModel can train on AnnData objects |
+
+### Legend
+- ✅ Implemented: Feature is fully implemented and tested
+- ⚠️ In Progress: Implementation or testing is underway
+- 📝 Specified: Feature is specified but not yet implemented
+- 🔄 Planned: Feature is planned but not fully specified
+
+## Unit Test Coverage Reference
+
+This table tracks the unit test coverage for each module in the modular implementation, helping identify areas that need additional testing.
+
+| Module | Test File | Test Coverage | Test Cases | Notes |
+|--------|-----------|---------------|------------|-------|
+| components/dynamics.py | test_dynamics.py | High | 10+ | Tests for StandardDynamicsModel and LegacyDynamicsModel |
+| components/guides.py | test_guides.py | High | 5+ | Tests for AutoGuideFactory and LegacyAutoGuideFactory |
+| components/likelihoods.py | test_likelihoods.py | High | 5+ | Tests for PoissonLikelihoodModel and LegacyLikelihoodModel |
+| components/observations.py | test_observations.py | High | 5+ | Tests for StandardObservationModel |
+| components/priors.py | test_priors.py | High | 5+ | Tests for LogNormalPriorModel |
+| model.py | test_model.py | High | 10+ | Tests for PyroVelocityModel |
+| factory.py | test_factory.py | High | 10+ | Tests for factory functions |
+| registry.py | test_registry.py | High | 5+ | Tests for component registries |
+| interfaces.py | test_interfaces.py | Medium | 5+ | Tests for Protocol interfaces |
+| data/anndata.py | test_anndata_integration.py | High | 5+ | Tests for AnnData integration |
+| utils/*.py | utils/test_*.py | Medium | 10+ | Tests for utility functions |
+
+## BDD Feature Files Explained
+
+Our feature files use Gherkin syntax to describe expected behavior. Each feature file follows this structure:
+
+```gherkin
+Feature: [Feature Name]
+  As a [type of user]
+  I want to [perform some action]
+  So that [achieve some benefit]
+
+  Background:
+    Given [common setup steps]
+
+  Scenario: [Specific use case name]
+    Given [preconditions]
+    When [actions]
+    Then [expected results]
+```
+
+### Example Feature File
+
+```gherkin
+Feature: Dynamics Model
+  As a computational biologist
+  I want to model RNA velocity dynamics
+  So that I can understand transcriptional dynamics in single-cell data
+
+  Background:
+    Given I have a dynamics model component
+    And I have input data with unspliced and spliced counts
+
+  Scenario Outline: Standard dynamics model computes expected counts
+    Given I have a StandardDynamicsModel
+    When I run the forward method with alpha <alpha>, beta <beta>, and gamma <gamma>
+    Then the model should compute expected unspliced and spliced counts
+    And the expected counts should follow RNA velocity dynamics
+
+    Examples:
+      | alpha | beta | gamma |
+      | 1.0   | 0.5  | 0.2   |
+```
+
+## Step Definitions Structure
+
+Step definition files connect feature files to actual test code. They are organized by component in the `integration/` directory:
+
+```python
+# src/pyrovelocity/tests/models/modular/integration/test_dynamics_model_steps.py
+from importlib.resources import files
+import pytest
+import torch
+from pytest_bdd import scenarios, given, when, then, parsers
+
+# Import the feature file using importlib.resources
+scenarios(str(files("pyrovelocity.tests.features") / "models" / "modular" / "dynamics_model.feature"))
+
+# Import the components
+from pyrovelocity.models.modular.components import (
+    LegacyDynamicsModel,
+    StandardDynamicsModel,
+)
+
+@given("I have a dynamics model component")
+def dynamics_model_component():
+    """Create a generic dynamics model component."""
+    return StandardDynamicsModel()
+
+@when(parsers.parse("I run the forward method with alpha {alpha}, beta {beta}, and gamma {gamma}"), target_fixture="run_forward_method_with_parameters")
+def run_forward_method_with_parameters_fixture(standard_dynamics_model, input_data, alpha, beta, gamma):
+    """Run the forward method with the given parameters."""
+    # Convert string parameters to float
+    alpha_val = float(alpha)
+    beta_val = float(beta)
+    gamma_val = float(gamma)
+
+    # Create context with input data and parameters
+    context = {
+        "u_obs": input_data["u_obs"],
+        "s_obs": input_data["s_obs"],
+        "alpha": torch.tensor([alpha_val] * input_data["n_genes"]),
+        "beta": torch.tensor([beta_val] * input_data["n_genes"]),
+        "gamma": torch.tensor([gamma_val] * input_data["n_genes"]),
+    }
+
+    # Run the forward method
+    result_context = standard_dynamics_model.forward(context)
+
+    # Store the result for later steps
+    return result_context
+```
+
+## Contributing New Tests
+
+When contributing new tests to the project:
+
+1. **Check Existing Coverage**: Review the feature and unit test reference tables
+2. **Update Feature Files**: If implementing a new feature, add or update the relevant feature file
+3. **Follow the Workflow**: Implement step definitions and unit tests as outlined above
+4. **Run Tests Locally**: Ensure all tests pass before submitting a PR
+5. **Update Documentation**: Update the reference tables in this document
+
+## Fixtures and Testing Utilities
+
+PyroVelocity provides a comprehensive set of fixtures for testing, defined in various `conftest.py` files:
+
+### Main Fixtures (`src/pyrovelocity/tests/conftest.py`)
+
+| Fixture Name | Purpose | Scope |
+|--------------|---------|-------|
+| `adata_preprocessed_pancreas_50_7` | Provides preprocessed pancreas dataset | Function |
+| `adata_trained_pancreas_50_7` | Provides trained pancreas dataset | Function |
+| `adata_postprocessed_pancreas_50_7` | Provides postprocessed pancreas dataset | Function |
+| `default_sample_data` | Provides synthetic sample data | Function |
+| `temp_file_path` | Provides a temporary file path | Function |
+| `temp_compressed_pickle_path` | Provides a temporary compressed pickle path | Function |
+| `save_and_load_helper` | Helper for saving and loading data | Function |
+
+### Modular Implementation Fixtures (`src/pyrovelocity/tests/models/modular/conftest.py`)
+
+These fixtures are specific to the modular implementation and provide test instances of various components.
+
+### BDD-Specific Fixtures (`src/pyrovelocity/tests/models/modular/integration/conftest.py`)
+
+| Fixture Name | Purpose | Scope |
+|--------------|---------|-------|
+| `bdd_simple_data` | Provides simple data for BDD testing | Function |
+| `bdd_model_parameters` | Provides model parameters for BDD testing | Function |
+| `bdd_standard_dynamics_model` | Provides a StandardDynamicsModel | Function |
+| `bdd_legacy_dynamics_model` | Provides a LegacyDynamicsModel | Function |
+| `bdd_lognormal_prior_model` | Provides a LogNormalPriorModel | Function |
+| `bdd_poisson_likelihood_model` | Provides a PoissonLikelihoodModel | Function |
+| `bdd_legacy_likelihood_model` | Provides a LegacyLikelihoodModel | Function |
+| `bdd_standard_observation_model` | Provides a StandardObservationModel | Function |
+| `bdd_auto_guide_factory` | Provides an AutoGuideFactory | Function |
+| `bdd_legacy_auto_guide_factory` | Provides a LegacyAutoGuideFactory | Function |
+| `bdd_pyro_velocity_model` | Provides a PyroVelocityModel | Function |
+| `bdd_anndata` | Provides a simple AnnData object | Function |
+| `clear_pyro_param_store` | Clears Pyro's parameter store | Function (autouse) |
+
+## Advanced Testing Topics
+
+### Parameterized Testing
+
+Both unit tests and BDD scenarios can be parameterized:
+
+```python
+# Unit test parameterization
+@pytest.mark.parametrize("alpha,beta,gamma,expected_u_ss,expected_s_ss", [
+    (1.0, 0.5, 0.2, 2.0, 5.0),
+    (2.0, 1.0, 0.5, 2.0, 4.0),
+])
+def test_steady_state(alpha, beta, gamma, expected_u_ss, expected_s_ss):
+    model = StandardDynamicsModel()
+    u_ss, s_ss = model.steady_state(
+        torch.tensor([alpha]),
+        torch.tensor([beta]),
+        torch.tensor([gamma]),
+    )
+    assert torch.allclose(u_ss, torch.tensor([expected_u_ss]))
+    assert torch.allclose(s_ss, torch.tensor([expected_s_ss]))
+```
+
+```gherkin
+# BDD parameterization with Scenario Outline
+Scenario Outline: Standard dynamics model computes steady state
+  Given I have a StandardDynamicsModel
+  When I compute the steady state with alpha <alpha>, beta <beta>, and gamma <gamma>
+  Then the steady state unspliced should equal alpha/beta
+  And the steady state spliced should equal alpha/gamma
+
+  Examples:
+    | alpha | beta | gamma |
+    | 1.0   | 0.5  | 0.2   |
+    | 2.0   | 1.0  | 0.5   |
+```
+
+### Testing with AnnData
+
+PyroVelocity provides fixtures for testing with AnnData objects:
+
+```python
+def test_anndata_integration(bdd_anndata, bdd_pyro_velocity_model):
+    """Test integration with AnnData."""
+    # Prepare AnnData
+    adata = bdd_anndata.copy()
+
+    # Train model
+    bdd_pyro_velocity_model.train(
+        adata,
+        max_epochs=2,
+        batch_size=5,
+        use_gpu=False,
+    )
+
+    # Check that results were stored in AnnData
+    assert "dynamics" in adata.uns
+    assert "model_type" in adata.uns["dynamics"]
+```
+
+### Testing with Pyro
+
+PyroVelocity provides utilities for testing with Pyro:
+
+```python
+def test_pyro_model(clear_pyro_param_store):
+    """Test a Pyro model."""
+    # Define a simple model
+    def model():
+        x = pyro.sample("x", pyro.distributions.Normal(0, 1))
+        return x
+
+    # Define a guide
+    def guide():
+        loc = pyro.param("loc", torch.tensor(0.0))
+        scale = pyro.param("scale", torch.tensor(1.0), constraint=pyro.constraints.positive)
+        pyro.sample("x", pyro.distributions.Normal(loc, scale))
+
+    # Run inference
+    pyro.clear_param_store()
+    svi = pyro.infer.SVI(
+        model=model,
+        guide=guide,
+        optim=pyro.optim.Adam({"lr": 0.01}),
+        loss=pyro.infer.Trace_ELBO(),
+    )
+
+    # Train for a few steps
+    for _ in range(10):
+        loss = svi.step()
+
+    # Check that parameters were updated
+    assert pyro.param("loc").item() != 0.0
+```
+
 ## Continuous Integration
 
-Tests are automatically run as part of the CI/CD pipeline defined in `.github/workflows/cid.yaml`.
+Tests are automatically run as part of the CI/CD pipeline defined in `.github/workflows/cid.yaml`. The CI pipeline includes:
+
+1. **Unit Tests**: Run all unit tests with coverage reporting
+2. **BDD Tests**: Run all BDD tests to verify behavior
+3. **Coverage Enforcement**: Ensure test coverage meets minimum thresholds
+
+## Best Practices for PyroVelocity Testing
+
+1. **Use Protocol Interfaces**: Test against Protocol interfaces rather than concrete implementations
+2. **Create Real Objects**: Prefer creating simple instances of real objects over using mocks
+3. **Test Edge Cases**: Include tests for edge cases and error handling
+4. **Use Fixtures**: Leverage pytest fixtures for common setup and teardown
+5. **Clear Pyro State**: Always clear Pyro's parameter store before and after tests
+6. **Synthetic Data**: Use synthetic data for tests to avoid external dependencies
+7. **Parameterize Tests**: Use parameterization to test multiple cases with the same logic
+8. **Document Tests**: Include docstrings and comments to explain test purpose and logic
+9. **Follow BDD Patterns**: Use Given-When-Then pattern for BDD tests
+10. **Validate Against Legacy**: Validate new implementations against legacy code

### 2025-05-12 src/pyrovelocity/tests/models/modular/integration/test_guide_model_steps.py
test: implement missing step definitions for BDD tests

@@ -1,0 +1,292 @@
+"""
+Step definitions for testing inference guides in PyroVelocity's modular implementation.
+
+This module implements the steps defined in the guide_model.feature file.
+"""
+
+from importlib.resources import files
+
+import pyro
+import pytest
+import torch
+from pyro.infer import autoguide
+from pytest_bdd import given, parsers, scenarios, then, when
+
+# Import the feature file using importlib.resources
+scenarios(str(files("pyrovelocity.tests.features") / "models" / "modular" / "guide_model.feature"))
+
+# Import the components
+from pyrovelocity.models.modular.components import (
+    AutoGuideFactory,
+    LegacyAutoGuideFactory,
+    LegacyDynamicsModel,
+    LegacyLikelihoodModel,
+    LogNormalPriorModel,
+    PoissonLikelihoodModel,
+    StandardDynamicsModel,
+    StandardObservationModel,
+)
+from pyrovelocity.models.modular.model import PyroVelocityModel
+
+
+@given("I have an inference guide component", target_fixture="inference_guide_component")
+def inference_guide_component():
+    """Create a generic inference guide component."""
+    return AutoGuideFactory(guide_type="AutoNormal")
+
+
+@given("I have a PyroVelocity model", target_fixture="pyro_velocity_model")
+def pyro_velocity_model_fixture(bdd_pyro_velocity_model):
+    """Get a PyroVelocity model from the fixture."""
+    return bdd_pyro_velocity_model
+
+
+@given(parsers.parse('I have an AutoGuideFactory with guide_type="{guide_type}"'), target_fixture="inference_guide_component")
+def auto_guide_factory_fixture(guide_type):
+    """Create an AutoGuideFactory with the specified guide type."""
+    return AutoGuideFactory(guide_type=guide_type)
+
+
+@given("I have a LegacyAutoGuideFactory", target_fixture="legacy_auto_guide_factory")
+def legacy_auto_guide_factory_fixture(bdd_legacy_auto_guide_factory):
+    """Get a LegacyAutoGuideFactory from the fixture."""
+    return bdd_legacy_auto_guide_factory
+
+
+@given(parsers.parse("I have an AutoGuideFactory with init_scale={init_scale}"), target_fixture="auto_guide_factory_with_init_scale")
+def auto_guide_factory_with_init_scale_fixture(init_scale):
+    """Create an AutoGuideFactory with the specified init_scale."""
+    # Convert string to float
[…]

### 2025-05-12 src/pyrovelocity/tests/models/modular/integration/test_likelihood_model_steps.py

@@ -1,0 +1,329 @@
+"""
+Step definitions for testing likelihood models in PyroVelocity's modular implementation.
+
+This module implements the steps defined in the likelihood_model.feature file.
+"""
+
+from importlib.resources import files
+
+import pyro
+import pytest
+import torch
+from pytest_bdd import given, parsers, scenarios, then, when
+
+# Import the feature file using importlib.resources
+scenarios(str(files("pyrovelocity.tests.features") / "models" / "modular" / "likelihood_model.feature"))
+
+# Import the components
+from pyrovelocity.models.modular.components import (
+    LegacyLikelihoodModel,
+    PoissonLikelihoodModel,
+)
+
+
+@given("I have a likelihood model component", target_fixture="likelihood_model_component")
+def likelihood_model_component():
+    """Create a generic likelihood model component."""
+    return PoissonLikelihoodModel()
+
+
+@given("I have input data with unspliced and spliced counts", target_fixture="input_data")
+def input_data_fixture(bdd_simple_data):
+    """Get input data from the fixture."""
+    return bdd_simple_data
+
+
+@given("I have expected unspliced and spliced counts", target_fixture="expected_counts")
+def expected_counts_fixture():
+    """Create expected unspliced and spliced counts."""
+    # Create random expected counts for testing
+    torch.manual_seed(42)
+    n_cells = 10
+    n_genes = 5
+
+    # Generate random expected counts
+    u_expected = torch.abs(torch.randn((n_cells, n_genes)))
+    s_expected = torch.abs(torch.randn((n_cells, n_genes)))
+
+    return {
+        "u_expected": u_expected,
+        "s_expected": s_expected,
+    }
+
+
+@given("I have a PoissonLikelihoodModel", target_fixture="poisson_likelihood_model")
+def poisson_likelihood_model_fixture(bdd_poisson_likelihood_model):
+    """Get a PoissonLikelihoodModel from the fixture."""
+    return bdd_poisson_likelihood_model
+
+
+@given("I have a LegacyLikelihoodModel", target_fixture="legacy_likelihood_model")
+def legacy_likelihood_model_fixture(bdd_legacy_likelihood_model):
+    """Get a LegacyLikelihoodModel from the fixture."""
+    return bdd_legacy_likelihood_model
+
+
+@given("I have data with zero counts", target_fixture="zero_count_data")
+def zero_count_data_fixture():
+    """Create data with zero counts."""
[…]

### 2025-05-12 src/pyrovelocity/tests/models/modular/integration/test_observation_model_steps.py

@@ -1,0 +1,338 @@
+"""
+Step definitions for testing observation models in PyroVelocity's modular implementation.
+
+This module implements the steps defined in the observation_model.feature file.
+"""
+
+from importlib.resources import files
+
+# Import necessary modules
+import pyro
+import pytest
+import torch
+from pytest_bdd import given, parsers, scenarios, then, when
+
+# Import the feature file using importlib.resources
+scenarios(str(files("pyrovelocity.tests.features") / "models" / "modular" / "observation_model.feature"))
+
+# Import the components
+from pyrovelocity.models.modular.components import StandardObservationModel
+
+
+@given("I have an observation model component", target_fixture="observation_model_component")
+def observation_model_component():
+    """Create a generic observation model component."""
+    return StandardObservationModel()
+
+
+@given("I have input data with unspliced and spliced counts", target_fixture="input_data")
+def input_data_fixture(bdd_simple_data):
+    """Get input data from the fixture."""
+    return bdd_simple_data
+
+
+@given("I have a StandardObservationModel", target_fixture="standard_observation_model")
+def standard_observation_model_fixture():
+    """Create a StandardObservationModel."""
+    return StandardObservationModel()
+
+
+@given(parsers.parse("I have a StandardObservationModel with use_size_factor={use_size_factor}"), target_fixture="standard_observation_model_with_size_factor")
+def standard_observation_model_with_size_factor_fixture(use_size_factor):
+    """Create a StandardObservationModel with the specified use_size_factor setting."""
+    # Convert string to boolean
+    use_size_factor_bool = use_size_factor.lower() == "true"
+
+    return StandardObservationModel(use_observed_lib_size=use_size_factor_bool)
+
+
+@given("I have a StandardObservationModel with normalization", target_fixture="standard_observation_model_with_normalization")
+def standard_observation_model_with_normalization_fixture():
+    """Create a StandardObservationModel with normalization enabled."""
+    return StandardObservationModel(use_observed_lib_size=True, transform_batch=True)
+
+
+@given("I have data with missing values", target_fixture="data_with_missing_values")
+def data_with_missing_values_fixture():
+    """Create data with missing values (represented as NaNs)."""
+    # Create data with some NaN values
+    torch.manual_seed(42)
+    n_cells = 10
+    n_genes = 5
+
+    # Generate random data
+    u_obs = torch.abs(torch.randn((n_cells, n_genes)))
+    s_obs = torch.abs(torch.randn((n_cells, n_genes)))
+
+    # Set some values to NaN
[…]

### 2025-05-12 nbs/development/testing.qmd
docs(development): add integration tests
- note previously specified and unimplemented integration tests

Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 │   │   │   ├── conftest.py    # BDD-specific fixtures
+│   │   │   ├── test_component_integration.py
 │   │   │   ├── test_dynamics_model_steps.py
+│   │   │   ├── test_guide_model_steps.py
+│   │   │   ├── test_likelihood_model_steps.py
 │   │   │   ├── test_model_steps.py
+│   │   │   ├── test_observation_model_steps.py
 │   │   │   └── test_prior_model_steps.py
@@
 | prior_model | LogNormal prior model creates stochastic nodes | ✅ Implemented | Verify that the LogNormalPriorModel creates appropriate stochastic nodes |
-| likelihood_model | Poisson likelihood model creates observation nodes | ✅ Implemented | Verify that the PoissonLikelihoodModel creates appropriate observation nodes |
+| likelihood_model | Poisson likelihood model defines distributions | ✅ Implemented | Verify that the PoissonLikelihoodModel defines appropriate Poisson distributions |
 | likelihood_model | Legacy likelihood model matches legacy implementation | ✅ Implemented | Verify that the LegacyLikelihoodModel matches the legacy implementation |
-| observation_model | Standard observation model preprocesses data | ✅ Implemented | Verify that the StandardObservationModel correctly preprocesses data |
-| guide_model | Auto guide factory creates guide | ✅ Implemented | Verify that the AutoGuideFactory creates an appropriate guide |
+| likelihood_model | Likelihood model with scaling factors | ✅ Implemented | Verify that the likelihood model correctly incorporates scaling factors |
+| likelihood_model | Likelihood model with zero counts | ✅ Implemented | Verify that the likelihood model handles zero counts gracefully |
+| likelihood_model | Likelihood model with Pyro plates | ✅ Implemented | Verify that the likelihood model correctly uses Pyro plates for batch dimensions |
+| observation_model | Standard observation model transforms data | ✅ Implemented | Verify that the StandardObservationModel correctly transforms input data |
+| observation_model | Observation model computes size factors | ✅ Implemented | Verify that the observation model correctly computes size factors |
+| observation_model | Observation model handles missing values | ✅ Implemented | Verify that the observation model handles missing values gracefully |
+| observation_model | Observation model with normalization | ✅ Implemented | Verify that the observation model correctly normalizes data |
+| observation_model | Observation model preserves data integrity | ✅ Implemented | Verify that the observation model preserves biological signal and doesn't introduce artifacts |
+| guide_model | AutoGuide factory creates guide | ✅ Implemented | Verify that the AutoGuideFactory creates an appropriate guide |
 | guide_model | Legacy auto guide factory matches legacy implementation | ✅ Implemented | Verify that the LegacyAutoGuideFactory matches the legacy implementation |
+| guide_model | Guide with parameter initialization | ✅ Implemented | Verify that the guide correctly initializes parameters with specified scale |
+| guide_model | Guide with different variational families | ✅ Implemented | Verify that the guide supports different variational families |
+| guide_model | Guide with custom initialization | ✅ Implemented | Verify that the guide supports custom initialization functions |
 | model | PyroVelocity model integrates components | ✅ Implemented | Verify that the PyroVelocityModel correctly integrates all components |
@@
 
-### Example Feature File
+### Example Feature Files
+
+PyroVelocity uses feature files to describe the expected behavior of each component. Here are examples from different components:
+
+#### Dynamics Model Example
 
@@
 
+#### Inference Guide Example
+
+```gherkin
+Feature: Inference Guide
+  As a computational biologist
+  I want to define variational distributions for RNA velocity inference
+  So that I can perform efficient approximate Bayesian inference
+
+  Scenario: Guide with different variational families
+    Given I have an AutoGuideFactory with guide_type="<guide_type>"
+    When I create a guide for the model
+    Then the guide should be an instance of <guide_class>
+    And the guide should have the appropriate variational family
+
+    Examples:
+      | guide_type           | guide_class                  |
+      | AutoNormal           | AutoNormal                   |
+      | AutoDiagonalNormal   | AutoDiagonalNormal           |
+      | AutoLowRankMultivariateNormal | AutoLowRankMultivariateNormal |
+      | AutoDelta            | AutoDelta                    |
+```
+
+#### Likelihood Model Example
+
+```gherkin
+Feature: Likelihood Model
+  As a computational biologist
+  I want to define likelihood distributions for RNA velocity observations
+  So that I can relate expected counts to observed counts in single-cell data
+
+  Scenario: Poisson likelihood model defines distributions
+    Given I have a PoissonLikelihoodModel
+    When I run the forward method
+    Then the model should define Poisson distributions for unspliced and spliced counts
+    And the distributions should use the expected counts as rate parameters
+    And the model should register observations with Pyro
+```
+
+#### Observation Model Example
+
+```gherkin
+Feature: Observation Model
+  As a computational biologist
+  I want to preprocess and transform RNA velocity data
+  So that it can be used effectively in the model
+
+  Scenario: Observation model with normalization
+    Given I have a StandardObservationModel with normalization
+    When I run the forward method
+    Then the model should normalize the data
+    And the normalized data should have the expected properties
+    And the context should include the normalized data
+```
+
 ## Step Definitions Structure
 
-Step definition files connect feature files to actual test code. They are organized by component in the `integration/` directory:
+Step definition files connect feature files to actual test code.
+They are organized by component in the `integration/` directory.
+PyroVelocity has implemented step definitions for all the major components:
+
+- `test_dynamics_model_steps.py`: Implements steps for testing dynamics models
+- `test_prior_model_steps.py`: Implements steps for testing prior models
+- `test_likelihood_model_steps.py`: Implements steps for testing likelihood models
+- `test_observation_model_steps.py`: Implements steps for testing observation models
+- `test_guide_model_steps.py`: Implements steps for testing inference guides
+- `test_model_steps.py`: Implements steps for testing the full PyroVelocityModel
+
+Additionally, there's a `test_component_integration.py` file that tests the interactions between different components to ensure they work correctly together.
+
+Here's an example of a step definition file:
 
@@
     return result_context
+```
+
+### Component Integration Tests
+
+In addition to the BDD step definitions, PyroVelocity includes component integration tests that verify the interactions between different components:
+
+```python
+# src/pyrovelocity/tests/models/modular/integration/test_component_integration.py
+class TestComponentIntegration:
+    """Tests for component interactions."""
+
+    def test_dynamics_prior_integration(self):
+        """Test that dynamics model and prior model work together."""
+        # Create component models
+        dynamics_model = StandardDynamicsModel()
+        prior_model = LogNormalPriorModel()
+
+        # Sample parameters from prior
+        n_genes = 5
+        prior_params = prior_model.sample_parameters(n_genes=n_genes)
+
+        # Use parameters with dynamics model
+        u0 = torch.zeros(n_genes)
+        s0 = torch.zeros(n_genes)
+
+        # Simulate dynamics
+        times, u, s = dynamics_model.simulate(
+            u0=u0,
+            s0=s0,
+            alpha=prior_params["alpha"],
+            beta=prior_params["beta"],
+            gamma=prior_params["gamma"],
+            t_max=10.0,
+            n_steps=100,
+        )
+
+    def test_full_model_integration(self, simple_data):
+        """Test that all components work together in the full model."""
+        # Create components
+        dynamics_model = StandardDynamicsModel()
+        prior_model = LogNormalPriorModel()
+        likelihood_model = PoissonLikelihoodModel()
+        observation_model = StandardObservationModel()
+        guide_model = AutoGuideFactory(guide_type="AutoNormal")
+
+        # Create the full model
+        model = PyroVelocityModel(
+            dynamics_model=dynamics_model,
+            prior_model=prior_model,
+            likelihood_model=likelihood_model,
+            observation_model=observation_model,
+            guide_model=guide_model,
+        )
+
+        # Call the model with test data
+        result = model.forward(
+            u_obs=simple_data["u_obs"],
+            s_obs=simple_data["s_obs"]
+        )
 ```

### 2025-05-13 nbs/development/context.qmd
docs(development): init context spec
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "Scientific Context Specification"
format:
  html:
    mermaid-format: js
    mermaid:
      theme: dark
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
---

# Introduction {#sec-introduction}

## Purpose of this Document {#sec-purpose}

This Scientific Context Specification establishes the foundation for the PyroVelocity project by defining the scientific context in which the system operates. It captures the research problem being addressed, stakeholders involved, scientific workflows affected, and domain knowledge essential for understanding the project scope. This document serves as the primary reference for domain scientists, developers, requirements engineers, and other stakeholders to ensure a shared understanding of the project context before detailed requirements are specified.

## Document Conventions {#sec-conventions}

This document uses the following conventions:

- *Italics* are used for terms defined in the glossary
- **Bold** is used for emphasis and key concepts
- Blue text indicates a cross-reference to another section
- Mathematical equations are expressed using LaTeX notation
- Code snippets are shown in monospaced font

## References {#sec-references}

1. La Manno, G., Soldatov, R., Zeisel, A. et al. RNA velocity of single cells. Nature 560, 494–498 (2018). https://doi.org/10.1038/s41586-018-0414-6
2. Bergen, V., Lange, M., Peidli, S. et al. Generalizing RNA velocity to transient cell states through dynamical modeling. Nat Biotechnol 38, 1408–1414 (2020). https://doi.org/10.1038/s41587-020-0591-3
3. Pyro: Deep Universal Probabilistic Programming. https://pyro.ai/
4. AnnData: Annotated Data. https://anndata.readthedocs.io/
5. scanpy: Single-Cell Analysis in Python. https://scanpy.readthedocs.io/

## Document Overview {#sec-overview}

This Scientific Context Specification is organized to progressively build understanding of the PyroVelocity project context. It begins with a definition of project scope, then identifies stakeholders and their concerns. Research goals are articulated to clarify the project's purpose, followed by a value proposition justifying the investment. The scientific domain model captures essential workflows and data objects, while the constraints section outlines limitations that must be considered throughout the project. Finally, a glossary ensures consistent terminology usage across all project artifacts.

# Project Scope {#sec-project-scope}

## Problem Description {#sec-problem-description}

Single-cell RNA sequencing (scRNA-seq) provides a snapshot of gene expression in individual cells, but understanding cellular dynamics and developmental trajectories requires temporal information that is not directly captured in these static measurements. RNA velocity, which measures the rate of change of gene expression, offers a way to infer directional information from these snapshots by leveraging the distinction between unspliced pre-mRNAs and mature spliced mRNAs.

Current methods for RNA velocity analysis have several limitations:

1. **Deterministic Approaches**: Existing methods like scVelo and velocyto use deterministic models that cannot adequately capture the inherent stochasticity in gene expression.

2. **Uncertainty Quantification**: These methods lack robust uncertainty quantification, making it difficult to assess the reliability of velocity estimates and downstream analyses.

3. **Parameter Estimation**: Current approaches often rely on simplistic parameter estimation techniques that may not fully leverage the information in the data.

4. **Cell Fate Prediction**: Existing methods provide limited capabilities for predicting cell fate choices and understanding developmental trajectory dynamics.

PyroVelocity addresses these limitations by developing a Bayesian, generative, and multivariate approach to RNA velocity modeling that provides uncertainty quantification and more robust parameter estimation.

## Research Objectives {#sec-research-objectives}

PyroVelocity aims to develop a probabilistic framework for RNA velocity analysis that:

1. Models raw spliced and unspliced read counts directly, avoiding information loss from normalization
2. Provides uncertainty quantification for velocity estimates and downstream analyses
3. Synchronizes cell time estimation across genes for more coherent trajectory inference
4. Enables multivariate modeling of gene expression and velocity prediction
5. Offers diagnostic analyses and visualizations of model uncertainty
6. Integrates with existing single-cell analysis ecosystems

## Project Boundaries {#sec-project-boundaries}

### In Scope

- Probabilistic modeling of RNA velocity using Pyro/PyTorch
- Parameter inference from single-cell RNA-seq data
- Uncertainty quantification of model predictions
- Integration with existing bioinformatics tools (scanpy, scVelo)
- Visualization of velocity estimates and uncertainty
- Modular implementation with interchangeable components
- JAX/NumPyro implementation for improved performance

### Out of Scope

- Development of new experimental protocols for RNA velocity measurement
- Integration with spatial transcriptomics data
- Multi-cellular tissue modeling
- Clinical application of the models
- Development of a standalone GUI application

# Stakeholder Model {#sec-stakeholder-model}

## Stakeholder Identification {#sec-stakeholder-identification}

| Stakeholder | Role | Responsibilities | Influence Level |
|-------------|------|------------------|-----------------|
| Computational Biologists | Primary Users | Apply the tool to analyze single-cell data | High |
| Bioinformaticians | Users/Developers | Extend the tool, integrate with workflows | High |
| Molecular Biologists | End Users | Interpret results, design follow-up experiments | Medium |
| Single-Cell Genomics Core Facilities | Users | Process data for researchers | Medium |
| Software Developers | Maintainers | Implement and maintain the codebase | High |
| Research Community | Beneficiaries | Use insights from the tool in research | Low |

## User Groups {#sec-user-groups}

| User Group | Description | Expertise Level | Computational Requirements |
|------------|-------------|-----------------|----------------------------|
| Computational Biologists | Researchers with programming experience who analyze single-cell data | High | High-performance computing, GPU access |
| Bioinformaticians | Technical specialists who develop and run analysis pipelines | High | High-performance computing, GPU access |
| Molecular Biologists | Wet-lab researchers who need to interpret results | Low-Medium | Standard workstations with visualization capabilities |
| Data Scientists | Researchers focused on method development | High | High-performance computing, GPU access |

## External Systems {#sec-external-systems}

| System Name | Description | Owned By | Nature of Interface |
|-------------|-------------|----------|---------------------|
| scanpy | Single-cell analysis toolkit | scverse | Python API integration |
| scVelo | RNA velocity analysis toolkit | scverse | Python API integration, data format compatibility |
| AnnData | Annotated data format | scverse | Data structure for input/output |
| Pyro | Probabilistic programming framework | Uber AI/Linux Foundation | Core dependency for model implementation |
| PyTorch | Deep learning framework | Meta AI | Core dependency for tensor operations |
| JAX | Numerical computing library | Google | Alternative backend for NumPyro implementation |

## Stakeholder Concerns {#sec-stakeholder-concerns}

| Stakeholder | Primary Concerns | Acceptance Criteria |
|-------------|------------------|---------------------|
| Computational Biologists | Accuracy, interpretability, integration with existing workflows | Validated against ground truth, clear documentation, compatible with scanpy/scVelo |
| Bioinformaticians | Extensibility, performance, reproducibility | Modular architecture, efficient computation, deterministic results |
| Molecular Biologists | Ease of use, biological interpretability | Clear visualizations, connection to biological processes |
| Software Developers | Code quality, maintainability, test coverage | Clean code, comprehensive tests, CI/CD pipeline |

# Scientific Objectives {#sec-scientific-objectives}

## Research Goals {#sec-research-goals}

| ID | Research Goal | Description | Success Measure | Priority |
|----|---------------|-------------|-----------------|----------|
| RG-01 | Probabilistic RNA Velocity | Develop a Bayesian framework for RNA velocity estimation | Published method with demonstrated advantages | High |
| RG-02 | Uncertainty Quantification | Provide robust uncertainty estimates for velocity and trajectories | Calibrated uncertainty metrics | High |
| RG-03 | Parameter Inference | Infer biologically meaningful parameters from scRNA-seq data | Validated parameter estimates | Medium |
| RG-04 | Cell Fate Prediction | Enable prediction of cell fate choices from velocity estimates | Accurate prediction on benchmark datasets | Medium |
| RG-05 | Multimodal Integration | Support integration of multiple data modalities | Demonstrated improvement with multimodal data | Low |

## Usage Goals {#sec-usage-goals}

| ID | Usage Goal | Related Research Goal | Description | Success Measure | Priority |
|----|------------|----------------------|-------------|-----------------|----------|
| UG-01 | Ease of Use | RG-01 | Provide a user-friendly API for RNA velocity analysis | Adoption by non-expert users | High |
| UG-02 | Visualization | RG-02 | Develop visualizations for uncertainty in velocity estimates | User feedback on interpretability | Medium |
| UG-03 | Integration | RG-01, RG-05 | Integrate with existing single-cell analysis tools | Seamless workflow with scanpy/scVelo | High |
| UG-04 | Reproducibility | RG-03 | Ensure reproducible results across runs and platforms | Deterministic results with fixed seeds | Medium |
| UG-05 | Scalability | RG-01 | Handle large datasets efficiently | Performance benchmarks | Medium |

## System Goals {#sec-system-goals}

| ID | System Goal | Related Usage Goal | Description | Success Measure | Priority |
|----|-------------|-------------------|-------------|-----------------|----------|
| SG-01 | Modular Architecture | UG-01, UG-03 | Implement a flexible, component-based design | Component interchangeability | High |
| SG-02 | Performance Optimization | UG-05 | Optimize computational performance | Benchmark against existing methods | Medium |
| SG-03 | Comprehensive Testing | UG-04 | Implement thorough testing framework | Test coverage metrics | High |
| SG-04 | Documentation | UG-01 | Provide clear, comprehensive documentation | User feedback, documentation coverage | High |
| SG-05 | JAX Implementation | UG-05 | Develop a JAX-based implementation for improved performance | Performance improvement metrics | Medium |

## Goal Relationships {#sec-goal-relationships}

```{mermaid}
graph TD
    RG01[RG-01: Probabilistic RNA Velocity] --> UG01[UG-01: Ease of Use]
    RG01 --> UG03[UG-03: Integration]
    RG02[RG-02: Uncertainty Quantification] --> UG02[UG-02: Visualization]
    RG03[RG-03: Parameter Inference] --> UG04[UG-04: Reproducibility]
    RG01 --> UG05[UG-05: Scalability]
    RG05[RG-05: Multimodal Integration] --> UG03

    UG01 --> SG01[SG-01: Modular Architecture]
    UG01 --> SG04[SG-04: Documentation]
    UG03 --> SG01
    UG04 --> SG03[SG-03: Comprehensive Testing]
    UG05 --> SG02[SG-02: Performance Optimization]
    UG05 --> SG05[SG-05: JAX Implementation]

    class RG01,RG02,RG03,RG05 research;
    class UG01,UG02,UG03,UG04,UG05 usage;
    class SG01,SG02,SG03,SG04,SG05 system;
```

# Value Proposition {#sec-value-proposition}

## Scientific Value {#sec-scientific-value}

PyroVelocity delivers scientific value in the following ways:

1. For computational biologists:
   - More accurate RNA velocity estimates with quantified uncertainty
   - Better understanding of cell fate decisions in development and disease
   - Improved ability to design and interpret single-cell experiments

2. For bioinformaticians:
   - Modular framework for developing and testing new velocity models
   - Integration with existing single-cell analysis workflows
   - Reproducible and reliable velocity analysis

3. For molecular biologists:
   - More reliable identification of key genes driving cell state transitions
   - Better prediction of cell fate choices
   - Improved understanding of temporal dynamics in biological processes

4. For the research community:
   - Open-source software for probabilistic RNA velocity analysis
   - Reproducible computational workflows for gene expression dynamics
   - New insights into the role of gene expression dynamics in cellular decision-making

## Resource Analysis {#sec-resource-analysis}

### Computational Resources

- **High-performance computing**: Required for training models on large datasets
- **GPU computing resources**: Recommended for accelerated training
- **Storage requirements**: 10-100GB for typical single-cell datasets and analysis results
- **Memory requirements**: 16-64GB RAM for large-scale analyses

### Personnel Resources

- **Computational scientists**: Algorithm development and implementation
- **Bioinformaticians**: Integration with existing workflows and testing
- **Domain experts**: Model formulation and result interpretation
- **Software engineers**: Software architecture and maintenance

### Timeline

- **Modular PyTorch/Pyro implementation**: Complete
- **Validation against legacy implementation**: In progress
- **JAX/NumPyro implementation**: In progress
- **Documentation and tutorials**: Ongoing

## Risk Analysis {#sec-risk-analysis}

| Risk ID | Risk Description | Probability | Impact | Mitigation Strategy |
|---------|------------------|-------------|--------|---------------------|
| RK-01 | Model complexity leads to slow performance | Medium | High | Optimize critical components, provide JAX implementation |
| RK-02 | Uncertainty estimates are not well-calibrated | Medium | High | Validate on synthetic data with known ground truth |
| RK-03 | Integration with existing tools is challenging | Low | Medium | Maintain compatibility with AnnData standard |
| RK-04 | Users find probabilistic approach difficult to understand | Medium | Medium | Provide clear documentation and intuitive visualizations |
| RK-05 | Modular architecture increases maintenance burden | Low | Low | Implement comprehensive testing and documentation |

## Success Criteria {#sec-success-criteria}

The project will be considered successful if:

1. The probabilistic model provides more accurate RNA velocity estimates than deterministic methods on benchmark datasets
2. Uncertainty quantification correctly captures true parameter values within confidence intervals
3. The modular implementation successfully replicates the functionality of the legacy implementation
4. The JAX implementation provides significant performance improvements over the PyTorch implementation
5. The software is successfully used by external research groups for RNA velocity analysis
6. The method is published in a peer-reviewed journal

# Scientific Domain Model {#sec-scientific-domain-model}

## Computational Workflow Overview {#sec-computational-workflow-overview}

```{mermaid}
graph TD
    A[Data Preprocessing] --> B[Model Training]
    B --> C[Posterior Sampling]
    C --> D[Velocity Computation]
    D --> E[Uncertainty Quantification]
    E --> F[Visualization and Analysis]

    class A preprocessing;
    class B,C modeling;
    class D,E,F analysis;
```

The PyroVelocity computational workflow consists of six main stages:

1. **Data Preprocessing**: Processing raw single-cell RNA sequencing data into AnnData objects with spliced and unspliced counts
2. **Model Training**: Training the probabilistic model using variational inference
3. **Posterior Sampling**: Generating samples from the posterior distribution of model parameters
4. **Velocity Computation**: Computing RNA velocity from the posterior samples
5. **Uncertainty Quantification**: Quantifying uncertainty in velocity estimates and cell fate predictions
6. **Visualization and Analysis**: Visualizing results and performing downstream analyses

## Scientific Workflows {#sec-scientific-workflows}

### Data Preprocessing Workflow

- **Purpose**: Prepare single-cell RNA-seq data for velocity analysis
- **Inputs**: Raw count matrices or AnnData objects
- **Steps**:
  1. Filter cells and genes based on quality metrics
  2. Normalize data if required
  3. Extract spliced and unspliced counts
  4. Compute quality control metrics
- **Outputs**: Preprocessed AnnData object with spliced and unspliced layers

### Model Training Workflow

- **Purpose**: Train the probabilistic model for RNA velocity
- **Inputs**: Preprocessed AnnData object
- **Steps**:
  1. Initialize model components (dynamics, prior, likelihood, observation, guide)
  2. Set up variational inference
  3. Optimize model parameters using SVI
  4. Monitor convergence and diagnostics
- **Outputs**: Trained model with optimized parameters

### Posterior Analysis Workflow

- **Purpose**: Generate and analyze posterior samples
- **Inputs**: Trained model and preprocessed AnnData object
- **Steps**:
  1. Generate posterior samples for model parameters
  2. Compute RNA velocity from posterior samples
  3. Quantify uncertainty in velocity estimates
  4. Project velocity onto low-dimensional embeddings
- **Outputs**: Velocity estimates with uncertainty quantification

## Scientific Data Objects {#sec-scientific-data-objects}

| Data Object | Description | Attributes | Format | Relationships |
|-------------|-------------|------------|--------|---------------|
| AnnData | Annotated data matrix for single-cell data | Cells (obs), genes (var), matrices (X, layers) | H5AD | Contains spliced/unspliced counts |
| CountMatrix | Matrix of RNA counts | Cells (rows), genes (columns), values (counts) | Dense/sparse array | Component of AnnData |
| ModelParameters | Parameters of the RNA velocity model | alpha, beta, gamma, switching, t | PyTorch tensors | Used by dynamics model |
| PosteriorSamples | Samples from parameter posterior distributions | Parameter samples, chain information | Dictionary of tensors | Generated from trained model |
| VelocityEstimates | RNA velocity vectors for each cell | Velocity magnitude and direction | Array | Computed from posterior samples |

## Domain Algorithms {#sec-domain-algorithms}

### RNA Velocity Dynamics Model

The core of PyroVelocity is a mathematical model of RNA dynamics based on a system of ordinary differential equations:

$$
\begin{align}
\frac{du}{dt} &= \alpha - \beta u \\
\frac{ds}{dt} &= \beta u - \gamma s
\end{align}
$$

Where:
- $u$ represents unspliced mRNA abundance
- $s$ represents spliced mRNA abundance
- $\alpha$ is the transcription rate
- $\beta$ is the splicing rate
- $\gamma$ is the degradation rate

The analytical solution to this system is:

$$
\begin{align}
u(t) &= u_0 e^{-\beta t} + \frac{\alpha}{\beta}(1 - e^{-\beta t}) \\
s(t) &= s_0 e^{-\gamma t} + \frac{\beta u_0}{\gamma - \beta}(e^{-\beta t} - e^{-\gamma t}) + \frac{\alpha \beta}{\gamma \beta}(1 - \frac{\gamma}{\gamma - \beta}e^{-\beta t} + \frac{\beta}{\gamma - \beta}e^{-\gamma t})
\end{align}
$$

This model is implemented in the `StandardDynamicsModel` component of PyroVelocity.

### Variational Inference Algorithm

PyroVelocity uses Stochastic Variational Inference (SVI) for parameter estimation:

1. Define a joint probability model $p(x, z)$ where $x$ are observed data and $z$ are latent variables
2. Approximate the posterior $p(z|x)$ with a variational distribution $q(z)$
3. Minimize the KL divergence between $q(z)$ and $p(z|x)$ by maximizing the ELBO (Evidence Lower Bound)
4. Use stochastic gradient descent for optimization

This is implemented using Pyro's SVI framework in the inference components of PyroVelocity.

## Domain Data Dictionary {#sec-domain-data-dictionary}

| Data Element | Definition | Format | Units | Valid Range |
|--------------|------------|--------|-------|-------------|
| alpha | Transcription rate | Float | Transcripts per unit time | [0, ∞) |
| beta | Splicing rate | Float | Per unit time | [0, ∞) |
| gamma | Degradation rate | Float | Per unit time | [0, ∞) |
| u_obs | Observed unspliced counts | Integer | Counts | [0, ∞) |
| s_obs | Observed spliced counts | Integer | Counts | [0, ∞) |
| ut | True unspliced counts (latent) | Float | Counts | [0, ∞) |
| st | True spliced counts (latent) | Float | Counts | [0, ∞) |
| tau | Cell-specific time | Float | Arbitrary units | [0, ∞) |
| u_scale | Scaling factor for unspliced counts | Float | Dimensionless | (0, ∞) |
| s_scale | Scaling factor for spliced counts | Float | Dimensionless | (0, ∞) |

# Constraints and Rules {#sec-constraints-and-rules}

## Scientific Constraints {#sec-scientific-constraints}

| ID | Constraint | Description | Source | Impact |
|----|------------|-------------|--------|--------|
| SC-01 | RNA Velocity Model Assumptions | The model assumes first-order kinetics for transcription, splicing, and degradation | RNA velocity theory | Limits ability to model complex regulatory dynamics |
| SC-02 | Steady State Assumption | The model assumes cells can reach steady state | RNA velocity theory | May not accurately model highly dynamic processes |
| SC-03 | Parameter Identifiability | Not all parameters may be uniquely identifiable from the data | Statistical theory | Requires careful prior specification and validation |
| SC-04 | Data Quality Requirements | Requires high-quality spliced and unspliced count data | Single-cell sequencing | Limits applicability to certain datasets |

## Technical Constraints {#sec-technical-constraints}

| ID | Constraint | Description | Source | Impact |
|----|------------|-------------|--------|--------|
| TC-01 | AnnData Compatibility | Must maintain compatibility with AnnData format | Ecosystem integration | Influences data structure design |
| TC-02 | Pyro/PyTorch Dependency | Relies on Pyro and PyTorch for probabilistic programming | Implementation choice | Affects performance and deployment options |
| TC-03 | Python Ecosystem | Must integrate with Python-based single-cell analysis tools | Ecosystem integration | Limits language and framework choices |
| TC-04 | Backward Compatibility | Must maintain compatibility with legacy implementation during transition | Project requirement | Constrains design choices |

## Computational Constraints {#sec-computational-constraints}

| ID | Constraint | Description | Source | Impact |
|----|------------|-------------|--------|--------|
| CC-01 | Memory Usage | Large single-cell datasets can exceed available memory | Hardware limitations | Requires efficient memory management |
| CC-02 | Inference Speed | Variational inference can be computationally intensive | Algorithm complexity | Affects user experience and scalability |
| CC-03 | Parallelization | Need for efficient parallelization across cells and genes | Performance requirements | Influences algorithm design |
| CC-04 | GPU Acceleration | Should leverage GPU acceleration when available | Performance requirements | Affects implementation choices |

## Environmental Constraints {#sec-environmental-constraints}

| ID | Constraint | Description | Source | Impact |
|----|------------|-------------|--------|--------|
| EC-01 | Open Source Requirement | Must be released as open-source software | Project policy | Affects licensing and distribution |
| EC-02 | Documentation Standards | Must maintain comprehensive documentation | User needs | Requires ongoing documentation effort |
| EC-03 | Testing Requirements | Must have comprehensive test coverage | Software quality | Requires significant testing effort |
| EC-04 | Integration Requirements | Must integrate with existing single-cell analysis workflows | User needs | Constrains API design |

# Glossary {#sec-glossary}

## Domain Terminology {#sec-domain-terminology}

| Term | Definition | Source |
|------|------------|--------|
| RNA Velocity | The time derivative of gene expression, representing the rate of change of mRNA abundance | La Manno et al., 2018 |
| Spliced mRNA | Mature messenger RNA with introns removed | Molecular biology |
| Unspliced mRNA | Precursor messenger RNA containing introns | Molecular biology |
| Transcription | Process of creating RNA from a DNA template | Molecular biology |
| Splicing | Process of removing introns from pre-mRNA | Molecular biology |
| Degradation | Process of breaking down mRNA | Molecular biology |
| Cell Fate | The developmental trajectory or end state of a cell | Developmental biology |
| Trajectory Inference | Computational methods to order cells along developmental paths | Computational biology |

## Technical Terminology {#sec-technical-terminology}

| Term | Definition | Source |
|------|------------|--------|
| AnnData | Annotated data format for single-cell genomics | AnnData documentation |
| Pyro | Probabilistic programming language built on PyTorch | Pyro documentation |
| Variational Inference | Method for approximating posterior distributions | Statistical learning theory |
| ELBO | Evidence Lower Bound, objective function in variational inference | Statistical learning theory |
| Protocol | Interface definition in Python typing | Python typing documentation |
| Modular Architecture | Design approach based on interchangeable components | Software engineering |
| JAX | Numerical computing library with automatic differentiation | JAX documentation |
| NumPyro | Probabilistic programming library built on JAX | NumPyro documentation |

## Acronyms {#sec-acronyms}

| Acronym | Definition |
|---------|------------|
| RNA | Ribonucleic Acid |
| mRNA | Messenger RNA |
| scRNA-seq | Single-cell RNA sequencing |
| SVI | Stochastic Variational Inference |
| ODE | Ordinary Differential Equation |
| GPU | Graphics Processing Unit |
| API | Application Programming Interface |
| MCMC | Markov Chain Monte Carlo |

### 2025-05-13 nbs/development/context.qmd
docs(development): update context title
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: "Scientific Context Specification"
+title: "Context Specification"
 format:

### 2025-05-13 nbs/development/requirements.qmd
docs(development): init requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "Requirements Specification"
format:
  html:
    mermaid-format: js
    mermaid:
      theme: dark
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
---

# Introduction {#sec-introduction}

## Purpose of this Document {#sec-purpose}

This Requirements Specification defines what PyroVelocity should do from an external (black-box) perspective without specifying how it will be implemented internally. This document builds upon the [Scientific Context Specification](context.qmd) by transforming research needs into specific system requirements. It serves as the primary reference for domain scientists, computational specialists, developers, testers, and stakeholders to ensure a shared understanding of the system's expected behavior conceptually prior to detailed design considerations.

## Document Conventions {#sec-conventions}

This document uses the following conventions:

- *Italics* are used for terms defined in the glossary
- **Bold** is used for emphasis and key concepts
- Blue text indicates a cross-reference to another section
- REQ-XXX identifiers are used for individually traceable requirements
- Mathematical formulas are presented using LaTeX notation
- Code examples are presented in monospaced font
- Computational complexity is expressed using Big O notation

## References {#sec-references}

1. [Scientific Context Specification](context.qmd)
2. [System Architecture and Design](architecture.qmd)
3. La Manno, G., Soldatov, R., Zeisel, A. et al. RNA velocity of single cells. Nature 560, 494–498 (2018). https://doi.org/10.1038/s41586-018-0414-6
4. Bergen, V., Lange, M., Peidli, S. et al. Generalizing RNA velocity to transient cell states through dynamical modeling. Nat Biotechnol 38, 1408–1414 (2020). https://doi.org/10.1038/s41587-020-0591-3
5. Pyro: Deep Universal Probabilistic Programming. https://pyro.ai/
6. AnnData: Annotated Data. https://anndata.readthedocs.io/

## Document Overview {#sec-overview}

This Requirements Specification is organized to progressively define the PyroVelocity system from an external perspective. It begins with a system vision that outlines the overall scope and context, followed by detailed models of usage, computation, and data. The functional requirements organize system capabilities, while quality requirements define non-functional aspects with particular attention to scientific computing needs like accuracy, performance, and reproducibility. System constraints, deployment requirements, and development requirements outline limitations and implementation considerations. The risk assessment identifies potential issues, and the glossary ensures consistent terminology.

# System Vision {#sec-system-vision}

## System Overview {#sec-system-overview}

PyroVelocity is a probabilistic framework for RNA velocity analysis in single-cell genomics. It provides a Bayesian approach to modeling RNA dynamics, enabling robust uncertainty quantification and parameter inference from single-cell RNA sequencing data. The system implements a modular architecture with interchangeable components for dynamics models, prior distributions, likelihood models, observation models, and inference guides.

PyroVelocity aims to overcome limitations of deterministic RNA velocity methods by:

1. Directly modeling raw spliced and unspliced read counts
2. Providing uncertainty quantification for velocity estimates
3. Synchronizing cell time estimation across genes
4. Enabling multivariate modeling of gene expression
5. Offering diagnostic analyses and visualizations of model uncertainty
6. Integrating with existing single-cell analysis ecosystems

The system is currently transitioning from a legacy implementation to a modular PyTorch/Pyro implementation, with a future migration to a JAX/NumPyro implementation for improved performance.

## System Context {#sec-system-context}

PyroVelocity interfaces with the following external entities:

- **Single-cell RNA-seq Data**: Source data containing spliced and unspliced counts
- **AnnData Objects**: Primary data structure for storing and manipulating single-cell data
- **Scanpy/scVelo**: Ecosystem tools for preprocessing and visualization
- **Pyro/PyTorch**: Core probabilistic programming and tensor computation frameworks
- **JAX/NumPyro**: Alternative backend for improved performance (future)

```{mermaid}
graph TB
    User["Data Scientist / Researcher"] -->|uses| Library["PyroVelocity"]
    Library -->|depends on| PyTorch["PyTorch"]
    Library -->|depends on| Pyro["Pyro"]
    Library -->|processes| AnnData["AnnData Objects"]
    Library -->|produces| Velocity["RNA Velocity"]
    Library -->|integrates with| Scanpy["Scanpy"]
    Library -->|integrates with| Scvelo["scVelo"]

    class User user;
    class Library library;
    class PyTorch,Pyro,Scanpy,Scvelo external;
    class AnnData,Velocity data;
```

## Capability Overview {#sec-capability-overview}

| Capability ID | Capability Name | Description | Priority |
|--------------|-----------------|-------------|----------|
| CAP-01 | Probabilistic RNA Velocity | Bayesian modeling of RNA velocity with uncertainty quantification | High |
| CAP-02 | Parameter Inference | Inference of biologically meaningful parameters from scRNA-seq data | High |
| CAP-03 | Uncertainty Visualization | Visualization of uncertainty in velocity estimates and trajectories | High |
| CAP-04 | AnnData Integration | Seamless integration with AnnData objects and the scanpy ecosystem | High |
| CAP-05 | Modular Architecture | Flexible, component-based architecture with interchangeable parts | High |
| CAP-06 | JAX/NumPyro Backend | High-performance implementation using JAX/NumPyro | Medium |
| CAP-07 | Workflow Integration | Integration with workflow systems for reproducible analysis | Medium |

## Usage Scenario Overview {#sec-usage-scenario-overview}

The system supports the following high-level usage scenarios:

- **Data Preprocessing**: Scientists prepare single-cell data for velocity analysis
- **Model Configuration**: Scientists configure model components and parameters
- **Model Training**: Scientists train the model using variational inference
- **Posterior Analysis**: Scientists analyze posterior distributions of parameters
- **Velocity Computation**: Scientists compute RNA velocity with uncertainty
- **Result Visualization**: Scientists visualize velocity and uncertainty
- **Integration with Workflows**: Scientists integrate velocity analysis into larger pipelines

```{mermaid}
graph TD
    A[Data Preprocessing] --> B[Model Configuration]
    B --> C[Model Training]
    C --> D[Posterior Analysis]
    D --> E[Velocity Computation]
    E --> F[Result Visualization]
    F --> G[Integration with Workflows]

    class A,B,C,D,E,F,G scenario;
```

# Usage Model {#sec-usage-model}

## Actors {#sec-actors}

| Actor ID | Actor Name | Description | Related Stakeholder | Primary Goals |
|----------|------------|-------------|---------------------|---------------|
| A-01 | Computational Biologist | Researcher with programming experience who analyzes single-cell data | Computational Biologists | Analyze RNA velocity, infer cell trajectories, quantify uncertainty |
| A-02 | Bioinformatician | Technical specialist who develops and runs analysis pipelines | Bioinformaticians | Integrate RNA velocity into workflows, extend functionality |
| A-03 | Molecular Biologist | Wet-lab researcher who needs to interpret results | Molecular Biologists | Interpret velocity results, design follow-up experiments |
| A-04 | Data Scientist | Researcher focused on method development | Research Community | Develop new models, benchmark against existing methods |
| A-05 | AnnData | Data structure for annotated data matrices | External System | Store and provide access to single-cell data |
| A-06 | Scanpy/scVelo | Single-cell analysis toolkits | External System | Preprocess data, visualize results |

## Usage Scenarios {#sec-usage-scenarios}

| Scenario ID | Scenario Name | Primary Actor | Description | Pre-conditions | Post-conditions |
|-------------|---------------|---------------|-------------|----------------|-----------------|
| US-01 | Data Preprocessing | A-01, A-02 | Prepare single-cell data for velocity analysis | Raw count data available | AnnData object with spliced/unspliced counts |
| US-02 | Model Configuration | A-01, A-04 | Configure model components and parameters | Preprocessed data available | Configured model ready for training |
| US-03 | Model Training | A-01, A-02 | Train the model using variational inference | Configured model and data available | Trained model with optimized parameters |
| US-04 | Posterior Analysis | A-01, A-04 | Analyze posterior distributions of parameters | Trained model available | Parameter distributions and diagnostics |
| US-05 | Velocity Computation | A-01, A-02 | Compute RNA velocity with uncertainty | Posterior samples available | Velocity estimates with uncertainty |
| US-06 | Result Visualization | A-01, A-03 | Visualize velocity and uncertainty | Velocity estimates available | Visualizations for interpretation |
| US-07 | Integration with Workflows | A-02 | Integrate velocity analysis into larger pipelines | Trained model and analysis code available | Integrated workflow |

## Functional Scenarios {#sec-functional-scenarios}

**Scenario ID:** FS-01

**Usage Scenario:** US-03 (Model Training)

**Description:** Training a PyroVelocity model on pancreas development data

**Triggering Event:** User initiates model training

**Steps:**

1. User loads preprocessed AnnData object with spliced/unspliced counts
2. User creates a PyroVelocity model with standard components
3. System initializes model parameters and guide
4. User calls the train method with the AnnData object
5. System performs stochastic variational inference
6. System monitors convergence and reports progress
7. System returns the trained model
8. User saves the trained model for later use

**Alternative Paths:**

* At step 2, if user wants custom components, they configure specific components
* At step 4, if user wants custom training parameters, they provide additional arguments

**Exception Paths:**

* At step 1, if data is not properly formatted, system provides error message
* At step 5, if training fails to converge, system provides diagnostic information

## Common Research Workflows {#sec-common-research-workflows}

**Workflow ID:** WF-01

**Workflow Name:** RNA Velocity Analysis Workflow

**Description:** Complete workflow for RNA velocity analysis from raw data to interpretation

**Participants:** A-01, A-02, A-03

**Workflow Steps:**

1. Data Preprocessing: US-01
2. Model Configuration: US-02
3. Model Training: US-03
4. Posterior Analysis: US-04
5. Velocity Computation: US-05
6. Result Visualization: US-06

**Variations:**

* For large datasets, users may subsample genes or cells before model training
* For complex biological systems, users may use nonlinear dynamics models
* For integration with other analyses, users may export velocity to other tools

# Computational Model {#sec-computational-model}

## Algorithm Specifications {#sec-algorithm-specifications}

**Algorithm ID:** ALG-01

**Algorithm Name:** Standard RNA Velocity Dynamics

**Description:** Standard model of RNA velocity based on first-order kinetics

**Mathematical Formulation:**

```
du/dt = α - βu
ds/dt = βu - γs
```

Where:

- u is the unspliced mRNA abundance
- s is the spliced mRNA abundance
- α is the transcription rate
- β is the splicing rate
- γ is the degradation rate

**Inputs:**

- u0: Initial unspliced RNA counts, tensor[num_cells, num_genes]
- s0: Initial spliced RNA counts, tensor[num_cells, num_genes]
- α: Transcription rates, tensor[num_genes]
- β: Splicing rates, tensor[num_genes]
- γ: Degradation rates, tensor[num_genes]
- τ: Cell-specific time points, tensor[num_cells]

**Outputs:**

- ut: Unspliced RNA counts at time τ, tensor[num_cells, num_genes]
- st: Spliced RNA counts at time τ, tensor[num_cells, num_genes]
- velocity: RNA velocity vectors, tensor[num_cells, num_genes]

**Computational Properties:**

- Time Complexity: O(num_cells * num_genes)
- Space Complexity: O(num_cells * num_genes)
- Numerical Stability: Stable for β ≠ γ, special case handling for β = γ
- Convergence: Converges to steady state (u_ss = α/β, s_ss = α/γ)

**References:**

- La Manno et al. (2018)
- Bergen et al. (2020)

**Algorithm ID:** ALG-02

**Algorithm Name:** Stochastic Variational Inference

**Description:** Approximate Bayesian inference for RNA velocity parameters

**Mathematical Formulation:**

```
ELBO(q) = E_q[log p(x, z)] - E_q[log q(z)]
```

Where:
- p(x, z) is the joint distribution of observed data x and latent variables z
- q(z) is the variational distribution
- ELBO is the Evidence Lower Bound

**Inputs:**

- Model: Probabilistic model p(x, z)
- Guide: Variational distribution q(z)
- Data: Observed data x
- Optimizer: Optimization algorithm
- Num_epochs: Number of training epochs

**Outputs:**

- Optimized guide parameters
- Training diagnostics
- ELBO values

**Computational Properties:**

- Time Complexity: O(num_epochs * num_cells * num_genes)
- Space Complexity: O(num_cells * num_genes)
- Convergence: Depends on model complexity and data characteristics

**References:**

- Pyro documentation
- Hoffman et al. (2013)

## Computational Interfaces {#sec-computational-interfaces}

**Interface ID:** CI-01

**Algorithm:** ALG-01

**Description:** Interface for dynamics models

**Input Parameters:**

| Parameter | Type | Description | Constraints |
|-----------|------|-------------|------------|
| context | Dict[str, Any] | Context dictionary containing model state | Must contain u_obs, s_obs, and parameters |

**Output Parameters:**

| Parameter | Type | Description |
|-----------|------|-------------|
| context | Dict[str, Any] | Updated context with computed values |

**Error Conditions:**

| Error Code | Description | Recovery Action |
|------------|-------------|----------------|
| ValueError | Missing required parameters | Provide missing parameters |
| RuntimeError | Numerical instability | Check parameter values |

**Interface ID:** CI-02

**Algorithm:** ALG-02

**Description:** Interface for model training

**Input Parameters:**

| Parameter | Type | Description | Constraints |
|-----------|------|-------------|------------|
| adata | AnnData | Annotated data object | Must contain spliced/unspliced layers |
| max_epochs | int | Maximum number of training epochs | > 0 |
| batch_size | Optional[int] | Batch size for training | None or > 0 |
| learning_rate | float | Learning rate for optimizer | > 0 |
| use_gpu | bool | Whether to use GPU acceleration | True/False |

**Output Parameters:**

| Parameter | Type | Description |
|-----------|------|-------------|
| model | PyroVelocityModel | Trained model |
| loss_history | List[float] | Training loss history |

**Error Conditions:**

| Error Code | Description | Recovery Action |
|------------|-------------|----------------|
| ValueError | Invalid AnnData format | Preprocess data correctly |
| RuntimeError | Training divergence | Adjust learning rate or model parameters |
| MemoryError | Out of memory | Reduce batch size or use data subsampling |

## Execution Model {#sec-execution-model}

**Execution Model ID:** EM-01

**Description:** Execution model for PyroVelocity

**Execution Environment:** Python environment with PyTorch/Pyro or JAX/NumPyro

**Execution Strategy:**

- Parallelization: Data parallelism across cells and genes
- Distribution: Single-node execution with optional GPU acceleration
- Scheduling: Batch-based processing for large datasets

**Resource Requirements:**

- Compute: Multi-core CPU or CUDA-compatible GPU
- Memory: 16-64GB RAM depending on dataset size
- Storage: 10-100GB for datasets and results
- Network: Minimal (data loading only)

**Scalability Characteristics:**

- Scales linearly with number of cells and genes
- GPU acceleration provides significant speedup for large datasets
- JAX/NumPyro implementation offers improved performance over PyTorch/Pyro

## Performance Parameters {#sec-performance-parameters}

| Algorithm ID | Parameter | Description | Target Value | Measurement Method |
|--------------|-----------|-------------|--------------|-------------------|
| ALG-01 | Execution Time | Time to compute velocity for 10k cells, 2k genes | < 1 minute on GPU | Wall clock time |
| ALG-02 | Training Time | Time to train model on 10k cells, 2k genes | < 30 minutes on GPU | Wall clock time |
| ALG-02 | Memory Usage | Peak memory usage during training | < 16GB | Process monitoring |
| ALG-02 | Convergence Rate | Number of epochs to reach convergence | < 1000 epochs | ELBO monitoring |

# Data Model {#sec-data-model}

## Data Objects {#sec-data-objects}

| Data Object ID | Data Object Name | Description | Related Scientific Data Object | Persistence |
|----------------|------------------|-------------|-------------------------------|------------|
| DO-01 | AnnData | Annotated data matrix for single-cell data | AnnData | Yes |
| DO-02 | CountMatrix | Matrix of RNA counts | CountMatrix | Yes |
| DO-03 | ModelParameters | Parameters of the RNA velocity model | ModelParameters | Yes |
| DO-04 | PosteriorSamples | Samples from parameter posterior distributions | PosteriorSamples | Yes |
| DO-05 | VelocityEstimates | RNA velocity vectors for each cell | VelocityEstimates | Yes |
| DO-06 | ModelState | Immutable state container for model components | N/A | Yes |
| DO-07 | ComponentConfig | Configuration for model components | N/A | Yes |

## Data Object Attributes {#sec-data-object-attributes}

**Data Object:** DO-01 (AnnData)

| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
|--------------|----------------|-------------|------|-------|-------------|---------------|
| ATTR-01 | X | Gene expression matrix | Matrix | Counts | Non-negative | None |
| ATTR-02 | layers["spliced"] | Spliced RNA counts | Matrix | Counts | Non-negative | None |
| ATTR-03 | layers["unspliced"] | Unspliced RNA counts | Matrix | Counts | Non-negative | None |
| ATTR-04 | obs | Cell annotations | DataFrame | Various | None | None |
| ATTR-05 | var | Gene annotations | DataFrame | Various | None | None |
| ATTR-06 | obsm | Cell embeddings | Dict[str, Matrix] | Various | None | None |
| ATTR-07 | uns | Unstructured annotations | Dict | Various | None | None |

**Data Object:** DO-03 (ModelParameters)

| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
|--------------|----------------|-------------|------|-------|-------------|---------------|
| ATTR-08 | alpha | Transcription rates | Tensor | Transcripts per unit time | Non-negative | None |
| ATTR-09 | beta | Splicing rates | Tensor | Per unit time | Non-negative | None |
| ATTR-10 | gamma | Degradation rates | Tensor | Per unit time | Non-negative | None |
| ATTR-11 | switching | Switching times | Tensor | Time | Non-negative | None |
| ATTR-12 | t | Cell-specific times | Tensor | Time | Non-negative | None |
| ATTR-13 | u_scale | Scaling factors for unspliced counts | Tensor | Dimensionless | Positive | 1.0 |
| ATTR-14 | s_scale | Scaling factors for spliced counts | Tensor | Dimensionless | Positive | 1.0 |

## Data Object Relationships {#sec-data-object-relationships}

| Relationship ID | From Object | To Object | Type | Cardinality | Description |
|-----------------|-------------|-----------|------|-------------|-------------|
| REL-01 | AnnData | CountMatrix | Composition | 1:N | AnnData contains multiple count matrices (X, layers) |
| REL-02 | PyroVelocityModel | ModelParameters | Association | 1:1 | Model uses parameters for computation |
| REL-03 | PyroVelocityModel | PosteriorSamples | Association | 1:1 | Model generates posterior samples |
| REL-04 | PosteriorSamples | VelocityEstimates | Derivation | 1:1 | Velocity is computed from posterior samples |
| REL-05 | AnnData | VelocityEstimates | Association | 1:1 | Velocity estimates are stored in AnnData |
| REL-06 | PyroVelocityModel | ModelState | Composition | 1:1 | Model contains state |
| REL-07 | ModelState | ComponentConfig | Composition | 1:N | State contains component configurations |

```{mermaid}
classDiagram
    AnnData *-- CountMatrix : contains
    PyroVelocityModel o-- ModelParameters : uses
    PyroVelocityModel --> PosteriorSamples : generates
    PosteriorSamples --> VelocityEstimates : computes
    AnnData o-- VelocityEstimates : stores
    PyroVelocityModel *-- ModelState : contains
    ModelState *-- ComponentConfig : contains

    class AnnData {
        +Matrix X
        +Dict layers
        +DataFrame obs
        +DataFrame var
    }

    class PyroVelocityModel {
        +DynamicsModel dynamics_model
        +PriorModel prior_model
        +LikelihoodModel likelihood_model
        +ObservationModel observation_model
        +InferenceGuide guide_model
        +ModelState state
    }

    class ModelParameters {
        +Tensor alpha
        +Tensor beta
        +Tensor gamma
        +Tensor switching
        +Tensor t
    }

    class PosteriorSamples {
        +Dict parameter_samples
        +int num_samples
        +int num_chains
    }

    class VelocityEstimates {
        +Matrix velocity
        +Matrix uncertainty
    }

    class ModelState {
        +Dict dynamics_state
        +Dict prior_state
        +Dict likelihood_state
        +Dict observation_state
        +Dict guide_state
    }

    class ComponentConfig {
        +str name
        +Dict params
    }
```

## Data Management Requirements {#sec-data-management-requirements}

| Requirement ID | Requirement Name | Description | Priority |
|----------------|------------------|-------------|----------|
| DM-01 | AnnData Compatibility | System must maintain compatibility with AnnData format | High |
| DM-02 | Serialization | System must support serialization of models and results | High |
| DM-03 | Versioning | System must track model and data versions | Medium |
| DM-04 | Data Provenance | System must track data processing steps | Medium |
| DM-05 | Large Dataset Support | System must support large datasets through chunking or streaming | Medium |

**Data Volume Considerations:**

- Expected data volume: 1-100GB per dataset
- Growth rate: Linear with number of cells and genes
- Retention period: Indefinite for published results

**Data Access Patterns:**

- Random access to cells and genes during training
- Sequential access to posterior samples during analysis
- Batch access to cells during visualization

**Data Quality Requirements:**

- Validation of input data format and content
- Handling of missing values and zeros
- Detection and reporting of data quality issues

# Functional Requirements {#sec-functional-requirements}

## Core Functions {#sec-core-functions}

| Function ID | Function Name | Description | Related Usage Scenarios | Priority |
|-------------|---------------|-------------|-------------------------|----------|
| FN-01 | Data Preprocessing | Prepare single-cell data for velocity analysis | US-01 | High |
| FN-02 | Model Configuration | Configure model components and parameters | US-02 | High |
| FN-03 | Model Training | Train the model using variational inference | US-03 | High |
| FN-04 | Posterior Sampling | Generate samples from posterior distributions | US-04 | High |
| FN-05 | Velocity Computation | Compute RNA velocity with uncertainty | US-05 | High |
| FN-06 | Result Visualization | Visualize velocity and uncertainty | US-06 | High |
| FN-07 | Model Serialization | Save and load trained models | US-03, US-04 | Medium |
| FN-08 | AnnData Integration | Store results in AnnData objects | US-01, US-05, US-06 | High |
| FN-09 | Parameter Analysis | Analyze posterior distributions of parameters | US-04 | Medium |
| FN-10 | Trajectory Inference | Infer cell trajectories from velocity | US-05, US-06 | Medium |

## Function Hierarchy {#sec-function-hierarchy}

```{mermaid}
graph TD
    FN01[FN-01: Data Preprocessing] --> FN02[FN-02: Model Configuration]
    FN02 --> FN03[FN-03: Model Training]
    FN03 --> FN04[FN-04: Posterior Sampling]
    FN04 --> FN05[FN-05: Velocity Computation]
    FN05 --> FN06[FN-06: Result Visualization]
    FN03 --> FN07[FN-07: Model Serialization]
    FN01 --> FN08[FN-08: AnnData Integration]
    FN05 --> FN08
    FN04 --> FN09[FN-09: Parameter Analysis]
    FN05 --> FN10[FN-10: Trajectory Inference]
    FN10 --> FN06

    class FN01,FN02,FN03,FN04,FN05,FN06,FN07,FN08,FN09,FN10 function;
```

## Function Input/Output Specifications {#sec-function-inputoutput-specifications}

**Function:** FN-03 (Model Training)

**Inputs:**

| Input Name | Type | Description | Source | Constraints |
|------------|------|-------------|--------|------------|
| adata | AnnData | Annotated data object | FN-01 | Must contain spliced/unspliced layers |
| model_config | ModelConfig | Model configuration | FN-02 | Valid component configurations |
| training_params | Dict | Training parameters | User | Valid parameter values |

**Outputs:**

| Output Name | Type | Description | Destination | Constraints |
|-------------|------|-------------|------------|------------|
| trained_model | PyroVelocityModel | Trained model | FN-04, FN-05, FN-07 | None |
| loss_history | List[float] | Training loss history | FN-06 | None |
| training_diagnostics | Dict | Training diagnostics | FN-06 | None |

**Function:** FN-05 (Velocity Computation)

**Inputs:**

| Input Name | Type | Description | Source | Constraints |
|------------|------|-------------|--------|------------|
| trained_model | PyroVelocityModel | Trained model | FN-03, FN-07 | None |
| posterior_samples | Dict | Posterior samples | FN-04 | None |
| adata | AnnData | Annotated data object | FN-01 | Must contain spliced/unspliced layers |

**Outputs:**

| Output Name | Type | Description | Destination | Constraints |
|-------------|------|-------------|------------|------------|
| velocity | Matrix | Velocity vectors | FN-06, FN-08, FN-10 | None |
| velocity_uncertainty | Matrix | Uncertainty estimates | FN-06, FN-08 | None |
| velocity_metrics | Dict | Velocity quality metrics | FN-06, FN-08 | None |

## Function Relationships {#sec-function-relationships}

| From Function | To Function | Relationship Type | Description |
|---------------|-------------|-------------------|-------------|
| FN-01 | FN-02 | PROVIDES_DATA_TO | Preprocessed data is used for model configuration |
| FN-02 | FN-03 | PROVIDES_CONFIG_TO | Model configuration is used for training |
| FN-03 | FN-04 | ENABLES | Trained model is used for posterior sampling |
| FN-04 | FN-05 | PROVIDES_DATA_TO | Posterior samples are used for velocity computation |
| FN-05 | FN-06 | PROVIDES_DATA_TO | Velocity estimates are used for visualization |
| FN-03 | FN-07 | PROVIDES_DATA_TO | Trained model is serialized |
| FN-07 | FN-03 | PROVIDES_DATA_TO | Serialized model is loaded for continued training |
| FN-05 | FN-08 | PROVIDES_DATA_TO | Velocity estimates are stored in AnnData |
| FN-04 | FN-09 | PROVIDES_DATA_TO | Posterior samples are used for parameter analysis |
| FN-05 | FN-10 | PROVIDES_DATA_TO | Velocity estimates are used for trajectory inference |

# Quality Requirements {#sec-quality-requirements}

## Accuracy and Precision {#sec-accuracy-and-precision}

| Requirement ID | Requirement Name | Description | Verification Method | Target Value |
|----------------|------------------|-------------|---------------------|--------------|
| ACC-01 | Parameter Estimation Accuracy | Accuracy of inferred parameters on synthetic data | Comparison with ground truth | Within 10% of true values |
| ACC-02 | Velocity Direction Accuracy | Accuracy of velocity direction on benchmark datasets | Comparison with known trajectories | > 80% agreement |
| ACC-03 | Uncertainty Calibration | Calibration of uncertainty estimates | Coverage analysis | 95% confidence intervals contain true values in > 90% of cases |
| ACC-04 | Numerical Stability | Stability of numerical computations | Stress testing with extreme values | No NaN or Inf values in results |

**Validation Approach:**

- Synthetic data with known ground truth
- Benchmark datasets with established trajectories
- Cross-validation with held-out data
- Comparison with deterministic methods

**Error Bounds:**

- Parameter estimation: 95% confidence intervals
- Velocity direction: Angular error < 30 degrees
- Trajectory inference: Pseudotime correlation > 0.8

## Performance Requirements {#sec-performance-requirements}

| Requirement ID | Requirement Name | Description | Measurement Method | Target Value |
|----------------|------------------|-------------|-------------------|-------------|
| PERF-01 | Training Time | Time to train model on standard dataset | Wall clock time | < 30 minutes on GPU |
| PERF-02 | Memory Usage | Peak memory usage during training | Process monitoring | < 16GB for 10k cells |
| PERF-03 | Sampling Speed | Time to generate posterior samples | Wall clock time | < 5 minutes for 1000 samples |
| PERF-04 | Velocity Computation Speed | Time to compute velocity from samples | Wall clock time | < 1 minute for 10k cells |
| PERF-05 | Visualization Speed | Time to generate standard visualizations | Wall clock time | < 30 seconds per plot |

**Scalability Requirements:**

- Linear scaling with number of cells up to 100k cells
- Linear scaling with number of genes up to 5k genes
- Sublinear scaling with number of posterior samples

**Resource Utilization Limits:**

- CPU: < 8 cores at 100% utilization
- GPU: < 16GB VRAM
- Disk: < 100GB for intermediate results

## Reliability Requirements {#sec-reliability-requirements}

| Requirement ID | Requirement Name | Description | Verification Method | Target Value |
|----------------|------------------|-------------|---------------------|-------------|
| REL-01 | Training Convergence | Reliable convergence of training process | Success rate on benchmark datasets | > 95% success rate |
| REL-02 | Numerical Robustness | Robustness to numerical issues | Testing with varied datasets | No failures due to numerical issues |
| REL-03 | Error Handling | Proper handling of errors and edge cases | Exception testing | All errors properly caught and reported |
| REL-04 | State Persistence | Reliable saving and loading of model state | Recovery testing | 100% successful recovery |

**Fault Tolerance Approach:**

- Automatic parameter initialization retry on failure
- Graceful degradation with problematic genes
- Checkpointing during long-running operations
- Validation of inputs before computation

**Data Integrity Requirements:**

- Validation of AnnData objects before processing
- Checksums for serialized models
- Version tracking for models and data

## Usability Requirements {#sec-usability-requirements}

| Requirement ID | Requirement Name | Description | Verification Method | Target Value |
|----------------|------------------|-------------|---------------------|-------------|
| USE-01 | API Simplicity | Simplicity of core API for common tasks | User testing | < 5 lines of code for standard workflow |
| USE-02 | Documentation Quality | Comprehensive and clear documentation | Documentation coverage | 100% API coverage with examples |
| USE-03 | Error Messages | Clear and actionable error messages | User testing | > 90% of errors resolvable from message |
| USE-04 | Visualization Quality | Quality and interpretability of visualizations | User feedback | > 80% positive feedback |
| USE-05 | Integration Ease | Ease of integration with existing workflows | Integration testing | Compatible with scanpy/scVelo workflows |

## Reproducibility Requirements {#sec-reproducibility-requirements}

| Requirement ID | Requirement Name | Description | Verification Method | Target Value |
|----------------|------------------|-------------|---------------------|-------------|
| REP-01 | Seed Control | Control over random seeds for reproducibility | Reproducibility testing | Identical results with same seed |
| REP-02 | Version Tracking | Tracking of software and data versions | Version inspection | All components versioned |
| REP-03 | Parameter Logging | Logging of all parameters and configurations | Log inspection | All parameters recorded |
| REP-04 | Environment Specification | Specification of computational environment | Environment testing | Reproducible environment |
| REP-05 | Workflow Integration | Integration with workflow management systems | Workflow testing | Compatible with standard workflow systems |

**Reproducibility Mechanisms:**

- Explicit random seed control
- Deterministic computation modes
- Configuration serialization
- Comprehensive logging

**Provenance Requirements:**

- Recording of data preprocessing steps
- Tracking of model configurations
- Logging of training parameters
- Documentation of analysis decisions

# System Constraints {#sec-system-constraints}

## Scientific Constraints {#sec-scientific-constraints}

| Constraint ID | Constraint Name | Description | Source | Impact |
|---------------|-----------------|-------------|--------|--------|
| SC-01 | RNA Velocity Model Assumptions | The model assumes first-order kinetics for transcription, splicing, and degradation | RNA velocity theory | Limits ability to model complex regulatory dynamics |
| SC-02 | Steady State Assumption | The model assumes cells can reach steady state | RNA velocity theory | May not accurately model highly dynamic processes |
| SC-03 | Parameter Identifiability | Not all parameters may be uniquely identifiable from the data | Statistical theory | Requires careful prior specification and validation |
| SC-04 | Data Quality Requirements | Requires high-quality spliced and unspliced count data | Single-cell sequencing | Limits applicability to certain datasets |
| SC-05 | Biological Variability | Biological systems exhibit inherent variability | Biology | Requires probabilistic modeling approach |

## Technical Constraints {#sec-technical-constraints}

| Constraint ID | Constraint Name | Description | Source | Impact |
|---------------|-----------------|-------------|--------|--------|
| TC-01 | AnnData Compatibility | Must maintain compatibility with AnnData format | Ecosystem integration | Influences data structure design |
| TC-02 | Pyro/PyTorch Dependency | Relies on Pyro and PyTorch for probabilistic programming | Implementation choice | Affects performance and deployment options |
| TC-03 | Python Ecosystem | Must integrate with Python-based single-cell analysis tools | Ecosystem integration | Limits language and framework choices |
| TC-04 | Backward Compatibility | Must maintain compatibility with legacy implementation during transition | Project requirement | Constrains design choices |
| TC-05 | Protocol-First Architecture | Uses Protocol interfaces for component contracts | Architecture decision | Influences implementation patterns |

## Computational Resource Constraints {#sec-computational-resource-constraints}

| Constraint ID | Constraint Name | Description | Source | Impact |
|---------------|-----------------|-------------|--------|--------|
| RC-01 | Memory Usage | Large single-cell datasets can exceed available memory | Hardware limitations | Requires efficient memory management |
| RC-02 | Inference Speed | Variational inference can be computationally intensive | Algorithm complexity | Affects user experience and scalability |
| RC-03 | Parallelization | Need for efficient parallelization across cells and genes | Performance requirements | Influences algorithm design |
| RC-04 | GPU Acceleration | Should leverage GPU acceleration when available | Performance requirements | Affects implementation choices |
| RC-05 | Deployment Environment | Must run in standard computational biology environments | User environment | Limits dependency choices |

# Deployment Requirements {#sec-deployment-requirements}

## Installation Environment {#sec-installation-environment}

| Requirement ID | Requirement Name | Description | Priority |
|----------------|------------------|-------------|----------|
| DE-01 | Python Compatibility | Compatible with Python 3.8+ | High |
| DE-02 | Package Management | Installable via pip and conda | High |
| DE-03 | Dependency Management | Clear specification of dependencies | High |
| DE-04 | Platform Support | Support for Linux, macOS, and Windows | Medium |
| DE-05 | GPU Support | Optional GPU support with clear instructions | Medium |

**Hardware Requirements:**

- CPU: Multi-core x86_64 processor
- Memory: 16GB+ RAM (32GB+ recommended)
- Storage: 10GB+ free space
- GPU: CUDA-compatible GPU (optional but recommended)

**Software Requirements:**

- Python 3.8+
- PyTorch 1.10+
- Pyro 1.8+
- AnnData 0.8+
- Scanpy 1.9+
- NumPy 1.20+
- SciPy 1.7+

**Network Requirements:**

- Internet connection for installation
- No ongoing network requirements for operation

## Deployment Process Requirements {#sec-deployment-process-requirements}

| Requirement ID | Requirement Name | Description | Priority |
|----------------|------------------|-------------|----------|
| DP-01 | Installation Documentation | Clear installation instructions | High |
| DP-02 | Dependency Resolution | Automatic resolution of dependencies | High |
| DP-03 | Version Compatibility | Clear specification of version compatibility | Medium |
| DP-04 | Installation Verification | Verification of successful installation | Medium |
| DP-05 | Containerization | Support for containerized deployment | Low |

## Integration Requirements {#sec-integration-requirements}

| Requirement ID | Requirement Name | Description | Priority |
|----------------|------------------|-------------|----------|
| INT-01 | AnnData Integration | Seamless integration with AnnData objects | High |
| INT-02 | Scanpy Integration | Compatible with Scanpy workflows | High |
| INT-03 | scVelo Integration | Compatible with scVelo for comparison | High |
| INT-04 | Workflow System Integration | Integration with workflow management systems | Medium |
| INT-05 | Visualization Integration | Integration with standard visualization tools | Medium |

**Integration Points:**

- AnnData for data exchange
- Scanpy for preprocessing and visualization
- scVelo for comparison and validation
- Workflow systems (Snakemake, Nextflow) for pipeline integration
- Visualization tools (Matplotlib, Plotly) for result presentation

**Data Exchange Formats:**

- AnnData (H5AD) for single-cell data
- HDF5 for model serialization
- CSV/TSV for tabular results
- JSON for configuration and metadata
- PDF/PNG for visualizations

# Development Requirements {#sec-development-requirements}

## Development Environment {#sec-development-environment}

| Requirement ID | Requirement Name | Description | Priority |
|----------------|------------------|-------------|----------|
| DEV-01 | Version Control | Git-based version control | High |
| DEV-02 | Issue Tracking | Issue tracking system | High |
| DEV-03 | Continuous Integration | Automated testing and integration | High |
| DEV-04 | Code Quality | Code quality checks and linting | Medium |
| DEV-05 | Development Documentation | Documentation for developers | Medium |

**Development Tools:**

- Git for version control
- GitHub for collaboration and issue tracking
- Poetry for dependency management
- Pytest for testing
- Black for code formatting
- Mypy and Beartype for type checking
- Sphinx/Quarto for documentation

**Version Control Requirements:**

- Feature branch workflow
- Pull request reviews
- Semantic versioning
- Conventional commits

**Continuous Integration/Deployment Requirements:**

- Automated testing on pull requests
- Test coverage reporting
- Documentation building
- Package publishing

## Documentation Requirements {#sec-documentation-requirements}

| Requirement ID | Requirement Name | Description | Priority |
|----------------|------------------|-------------|----------|
| DOC-01 | API Documentation | Comprehensive API documentation | High |
| DOC-02 | User Guide | User guide with examples | High |
| DOC-03 | Tutorials | Step-by-step tutorials | High |
| DOC-04 | Architecture Documentation | Documentation of system architecture | Medium |
| DOC-05 | Development Guide | Guide for contributors | Medium |

**Documentation Types:**

- API reference
- User guide
- Tutorials
- Architecture documentation
- Development guide
- Release notes

**Documentation Format:**

- Quarto for web documentation
- Markdown for GitHub documentation
- Docstrings for API documentation
- Jupyter notebooks for tutorials
- Mermaid diagrams for architecture

## Testing and Validation Requirements {#sec-testing-and-validation-requirements}

| Requirement ID | Requirement Name | Description | Priority |
|----------------|------------------|-------------|----------|
| TEST-01 | Unit Testing | Testing of individual components | High |
| TEST-02 | Integration Testing | Testing of component interactions | High |
| TEST-03 | Validation Testing | Validation against ground truth | High |
| TEST-04 | Performance Testing | Testing of performance characteristics | Medium |
| TEST-05 | Compatibility Testing | Testing of ecosystem compatibility | Medium |

**Testing Approaches:**

- Unit Testing: Pytest for individual components
- Integration Testing: Pytest-BDD for component interactions
- System Testing: End-to-end workflows
- Performance Testing: Benchmarking against reference datasets
- Validation Testing: Comparison with ground truth and legacy implementation

**Test Environment Requirements:**

- Automated test environment
- Synthetic test data
- Benchmark datasets
- Performance monitoring tools
- Continuous integration system

# Risk Assessment {#sec-risk-assessment}

## Technical Risks {#sec-technical-risks}

| Risk ID | Risk Description | Affected Requirements | Probability | Impact | Risk Exposure |
|---------|------------------|----------------------|-------------|--------|---------------|
| TR-01 | Performance issues with large datasets | PERF-01, PERF-02, PERF-03 | Medium | High | High |
| TR-02 | Numerical instability in parameter inference | ACC-01, ACC-03, REL-02 | Medium | High | High |
| TR-03 | Compatibility issues with ecosystem tools | INT-01, INT-02, INT-03 | Low | Medium | Low |
| TR-04 | Difficulty maintaining both PyTorch and JAX implementations | TC-02, TC-04, DEV-03 | Medium | Medium | Medium |
| TR-05 | Dependency conflicts in deployment environments | DE-01, DE-02, DE-03 | Low | Medium | Low |

## Scientific Risks {#sec-scientific-risks}

| Risk ID | Risk Description | Affected Requirements | Probability | Impact | Risk Exposure |
|---------|------------------|----------------------|-------------|--------|---------------|
| SR-01 | Model assumptions not valid for all biological systems | SC-01, SC-02, ACC-02 | High | Medium | High |
| SR-02 | Parameter identifiability issues | SC-03, ACC-01, ACC-03 | Medium | High | High |
| SR-03 | Data quality issues affecting results | SC-04, ACC-02, REL-01 | Medium | High | High |
| SR-04 | Difficulty interpreting uncertainty estimates | USE-04, ACC-03 | Medium | Medium | Medium |
| SR-05 | Challenges in validating results without ground truth | ACC-02, TEST-03 | High | Medium | High |

## Mitigation Strategies {#sec-mitigation-strategies}

| Risk ID | Mitigation Strategy | Responsible Party | Monitoring Approach |
|---------|---------------------|-------------------|---------------------|
| TR-01 | Implement data chunking and streaming for large datasets | Development Team | Performance benchmarking |
| TR-02 | Use robust initialization and regularization techniques | Research Team | Numerical stability testing |
| TR-03 | Maintain compatibility layer and comprehensive integration tests | Development Team | Ecosystem compatibility testing |
| TR-04 | Focus on modular design with shared interfaces | Architecture Team | Code quality metrics |
| TR-05 | Use containerization and environment specification | DevOps Team | Deployment testing |
| SR-01 | Develop multiple model variants for different biological contexts | Research Team | Model comparison on diverse datasets |
| SR-02 | Use informative priors and sensitivity analysis | Research Team | Parameter recovery testing |
| SR-03 | Implement robust preprocessing and quality control | Data Team | Data quality metrics |
| SR-04 | Develop intuitive visualizations for uncertainty | UX Team | User feedback |
| SR-05 | Use synthetic data and cross-validation for validation | Validation Team | Validation metrics |

# Glossary {#sec-glossary}

## Domain Terminology {#sec-domain-terminology}

| Term | Definition | Source |
|------|------------|--------|
| RNA Velocity | The time derivative of gene expression, representing the rate of change of mRNA abundance | La Manno et al., 2018 |
| Spliced mRNA | Mature messenger RNA with introns removed | Molecular biology |
| Unspliced mRNA | Precursor messenger RNA containing introns | Molecular biology |
| Transcription | Process of creating RNA from a DNA template | Molecular biology |
| Splicing | Process of removing introns from pre-mRNA | Molecular biology |
| Degradation | Process of breaking down mRNA | Molecular biology |
| Cell Fate | The developmental trajectory or end state of a cell | Developmental biology |
| Trajectory Inference | Computational methods to order cells along developmental paths | Computational biology |
| Pseudotime | Computational ordering of cells based on their progression through a biological process | Computational biology |

## Technical Terminology {#sec-technical-terminology}

| Term | Definition | Source |
|------|------------|--------|
| AnnData | Annotated data format for single-cell genomics | AnnData documentation |
| Pyro | Probabilistic programming language built on PyTorch | Pyro documentation |
| Variational Inference | Method for approximating posterior distributions | Statistical learning theory |
| ELBO | Evidence Lower Bound, objective function in variational inference | Statistical learning theory |
| Protocol | Interface definition in Python typing | Python typing documentation |
| Modular Architecture | Design approach based on interchangeable components | Software engineering |
| JAX | Numerical computing library with automatic differentiation | JAX documentation |
| NumPyro | Probabilistic programming library built on JAX | NumPyro documentation |
| Railway-Oriented Programming | Programming pattern for explicit error handling | Functional programming |

## Acronyms {#sec-acronyms}

| Acronym | Definition |
|---------|------------|
| RNA | Ribonucleic Acid |
| mRNA | Messenger RNA |
| scRNA-seq | Single-cell RNA sequencing |
| SVI | Stochastic Variational Inference |
| ODE | Ordinary Differential Equation |
| GPU | Graphics Processing Unit |
| API | Application Programming Interface |
| MCMC | Markov Chain Monte Carlo |
| CI/CD | Continuous Integration/Continuous Deployment |
| BDD | Behavior-Driven Development |

### 2025-05-13 nbs/development/architecture.qmd
docs(development): update architecture
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 ---
-title: "System Architecture and Design"
+title: "System Architecture and Design Specification"
 format:
@@
 
-## Purpose {#sec-purpose}
+## Purpose of this Document {#sec-purpose}
 
-This document describes the architecture and design of PyroVelocity's modular implementation.
-It serves as both a reference for ongoing development and as a resource for new developers and users who may be interested in extending the functionality of existing models or creating and comparing new ones.
-As such, this document will evolve alongside the implementation.
+This Scientific Computing System Specification provides a glass-box view of PyroVelocity's architecture, detailing its internal structure, computational methods, and implementation approach.
+This document builds upon the [Scientific Context Specification](context.qmd) and [Requirements Specification](requirements.qmd) by defining how the system realizes the specified requirements through computational algorithms, data structures, and architectural components.
+It serves as the primary reference for domain scientists, computational specialists, developers, and testers to guide the implementation, validation, and deployment of the system.
 
-## Scope {#sec-scope}
+## Document Conventions {#sec-conventions}
 
-This document covers:
+This document uses the following conventions:
 
-- High-level system architecture
-- Key components and their interactions
-- Data flow and processing pipelines
-- Major design decisions and their rationales
-- Technical constraints and quality attributes
+- *Italics* are used for terms defined in the glossary
+- **Bold** is used for emphasis and key concepts
+- Blue text indicates a cross-reference to another section
+- Monospaced font is used for code examples and algorithm pseudocode
+- Mathematical formulas are presented using LaTeX notation
+- Diagrams follow the Mermaid specification
+- Computational complexity is expressed using Big O notation
 
-The focus is currently on the modular model implementation in `src/pyrovelocity/models/modular`. Additional sections will be added as their architecture stabilizes.
+## References {#sec-references}
 
-## Definitions {#sec-definitions}
+1. [Scientific Context Specification](context.qmd)
+2. [Requirements Specification](requirements.qmd)
+3. La Manno, G., Soldatov, R., Zeisel, A. et al. RNA velocity of single cells. Nature 560, 494–498 (2018). https://doi.org/10.1038/s41586-018-0414-6
+4.
+Bergen, V., Lange, M., Peidli, S.
+et al.
+Generalizing RNA velocity to transient cell states through dynamical modeling.
+Nat Biotechnol 38, 1408–1414 (2020).
+https://doi.org/10.1038/s41587-020-0591-3
+5. Pyro: Deep Universal Probabilistic Programming. https://pyro.ai/
+6. AnnData: Annotated Data. https://anndata.readthedocs.io/
 
-| Term | Definition |
-|------|------------|
-| RNA Velocity | The time derivative of gene expression, representing the rate of change of mRNA abundance |
-| AnnData | A data structure for annotated data matrices, commonly used in single-cell genomics |
-| Protocol | A Python typing construct that defines an interface without implementation |
-| Pyro | A probabilistic programming framework built on PyTorch |
-| SVI | Stochastic Variational Inference, a method for approximate Bayesian inference |
+## Document Overview {#sec-overview}
 
-# System Overview {#sec-system-overview}
+This System Specification is organized to progressively define the internal structure and implementation approach of PyroVelocity.
+It begins with an architecture overview that presents the high-level organization, followed by detailed models of the computational implementation, component structure, data processing pipeline, and execution framework.
+These sections define how the system implements the algorithms, data structures, and behaviors specified in the Requirements Specification.
+Additional sections cover reproducibility implementation, scientific validation, and performance optimization.
+The design rationale explains key design decisions and trade-offs made during the system design.
 
-## Vision {#sec-vision}
+# Architecture Overview {#sec-architecture-overview}
 
-PyroVelocity aims to provide a robust, flexible, and extensible framework for probabilistic RNA velocity analysis.
-The modular architecture enables researchers to customize and extend the model while maintaining a consistent interface.
+## Architectural Approach {#sec-architectural-approach}
 
-## Architectural Style {#sec-architectural-style}
+PyroVelocity follows a Protocol-First architecture with a modular component-based design.
+This approach was selected to support the system's quality requirements, particularly flexibility, extensibility, and maintainability.
+The modular design allows different model components to be interchanged while maintaining consistent interfaces, enabling researchers to select the best components for their specific use case.
 
-PyroVelocity follows a Protocol-First architecture, chosen to maximize flexibility and composability. This style enables:
+The system is currently transitioning from a legacy implementation to a modular PyTorch/Pyro implementation, with a future migration to a JAX/NumPyro implementation for improved performance.
+This document focuses primarily on the modular PyTorch/Pyro implementation.
 
-1. **Composition over inheritance**: Components implement Protocol interfaces directly without inheritance hierarchies
-2. **Loose coupling**: Components interact through well-defined interfaces
-3. **Testability**: Components can be tested in isolation
-4. **Extensibility**: New components can be added without modifying existing code
+Key architectural patterns employed include:
 
-## Context Diagram {#sec-context-diagram}
+1. **Protocol-First Design**: All contracts are defined through Protocol interfaces, enabling structural typing and composition over inheritance.
+2. **Component-Based Architecture**: The system is composed of specialized components with clear responsibilities.
+3. **Registry Pattern**: Components are registered by name and retrieved dynamically.
+4. **Factory Pattern**: Factory functions create components and models from configurations.
+5. **Immutable State Pattern**: Model state is stored in immutable containers for functional programming style.
+6. **Railway-Oriented Programming**: Components process data in a pipeline with explicit error handling.
+
+This approach was selected to maximize flexibility and composability, enabling researchers to easily extend the system with new model variants while maintaining a consistent interface.
+
+## Component Overview {#sec-component-overview}
+
+The system consists of the following major components:
 
@@
 graph TB
-    User["Data Scientist / Researcher"] -->|uses| Library["PyroVelocity"]
-    Library -->|depends on| PyTorch["PyTorch"]
-    Library -->|depends on| Pyro["Pyro"]
-    Library -->|processes| AnnData["AnnData Objects"]
-    Library -->|produces| Velocity["RNA Velocity"]
-    Library -->|integrates with| Scanpy["Scanpy"]
-    Library -->|integrates with| Scvelo["scVelo"]
+    Main["PyroVelocityModel"] -->|contains| Dynamics["Dynamics Models"]
+    Main -->|contains| Priors["Prior Models"]
+    Main -->|contains| Likelihoods["Likelihood Models"]
+    Main -->|contains| Observations["Observation Models"]
+    Main -->|contains| Guides["Inference Guides"]
+    Main -->|contains| Registry["Registry System"]
+    Main -->|contains| Factory["Factory System"]
 
-    class User user;
-    class Library library;
-    class PyTorch,Pyro,Scanpy,Scvelo external;
-    class AnnData,Velocity data;
+    Dynamics -->|implements| DynamicsInterface["DynamicsModel Protocol"]
+    Priors -->|implements| PriorsInterface["PriorModel Protocol"]
+    Likelihoods -->|implements| LikelihoodsInterface["LikelihoodModel Protocol"]
+    Observations -->|implements| ObservationsInterface["ObservationModel Protocol"]
+    Guides -->|implements| GuidesInterface["InferenceGuide Protocol"]
+
+    Factory -->|uses| Registry
+    Factory -->|creates| Main
+
+    class Main main;
+    class Dynamics,Priors,Likelihoods,Observations,Guides component;
+    class DynamicsInterface,PriorsInterface,LikelihoodsInterface,ObservationsInterface,GuidesInterface interface;
+    class Registry,Factory system;
 ```
 
-## Key Principles {#sec-key-principles}
+### Core Components {#sec-core-components}
 
-The following principles guide architectural decisions for PyroVelocity:
+1. **PyroVelocityModel**: Main model class that composes specialized component models
+2. **Dynamics Models**: Implement the mathematical models of RNA dynamics
+   - StandardDynamicsModel: Standard implementation of RNA dynamics
+   - LegacyDynamicsModel: Compatibility implementation matching the legacy model
+3. **Prior Models**: Define prior distributions for model parameters
+   - LogNormalPriorModel: Uses log-normal distributions for parameters
+4. **Likelihood Models**: Define observation distributions
+   - PoissonLikelihoodModel: Uses Poisson distributions for RNA counts
+   - LegacyLikelihoodModel: Compatibility implementation matching the legacy model
+5. **Observation Models**: Handle data preprocessing and transformation
+   - StandardObservationModel: Standard implementation for observations
+6. **Inference Guides**: Create guide functions for variational inference
+   - AutoGuideFactory: Creates guides for the modular implementation
+   - LegacyAutoGuideFactory: Compatibility implementation matching the legacy guide
+7. **Registry System**: Enables component registration and retrieval
+8. **Factory System**: Creates components and models from configurations
 
-1. **Protocol-First Design**: All contracts are defined through Protocol interfaces, enabling structural typing and composition over inheritance.
-2. **Separation of Concerns**: Each component has a single responsibility, making the system more maintainable and testable.
-3. **Immutable State**: The system uses immutable data structures where possible to reduce side effects and improve reasoning about code.
-4. **Type Safety**: The system uses static typing and runtime type checking to catch errors early and provide better development support.
+## Major Functions Summary {#sec-major-functions-summary}
+
+| Component | Major Function | Description | Related Core Function |
+|-----------|----------------|-------------|----------------------|
+| PyroVelocityModel | forward | Runs the model forward to compute expected RNA counts | FN-03 |
+| PyroVelocityModel | train | Trains the model using variational inference | FN-03 |
+| PyroVelocityModel | generate_posterior_samples | Generates posterior samples for uncertainty quantification | FN-04 |
+| PyroVelocityModel | compute_velocity | Computes RNA velocity from posterior samples | FN-05 |
+| PyroVelocityModel | store_results_in_anndata | Stores results in AnnData objects | FN-08 |
+| DynamicsModel | forward | Computes expected unspliced and spliced RNA counts | FN-05 |
+| PriorModel | forward | Samples model parameters from prior distributions | FN-03 |
+| LikelihoodModel | forward | Defines likelihood distributions for observed data | FN-03 |
+| ObservationModel | forward | Transforms observed data for model input | FN-01 |
+| InferenceGuide | create_guide | Creates guide functions for variational inference | FN-03 |
+
+## Architecture Decisions {#sec-architecture-decisions}
+
+| Decision ID | Decision | Rationale | Alternatives Considered | Impact |
+|-------------|----------|-----------|-------------------------|--------|
+| AD-01 | Protocol-First Architecture | Enables composition over inheritance, maximum flexibility, and alignment with functional programming principles | Abstract Base Classes | Complete separation between interface and implementation, maximum flexibility, simplified testing |
+| AD-02 | Context-Based Component Communication | Flexible data sharing without tight coupling, clear data flow | Direct method parameters | Easy to extend with new data, potential for inconsistent context structure |
+| AD-03 | Immutable State Pattern | Thread safety, predictable behavior, functional programming style | Mutable state | Simplified reasoning about code, reduced side effects |
+| AD-04 | Registry Pattern | Dynamic component discovery, extensibility | Direct imports | Decoupling between component implementations and usage |
+| AD-05 | Factory Pattern | Encapsulation of creation logic, configuration management | Direct instantiation | Simplified client code, centralized creation logic |
+
+# Computational Implementation {#sec-computational-implementation}
+
+## Mathematical Model Implementation {#sec-mathematical-model-implementation}
+
+**Model ID:** MM-01
+
+**Model Name:** RNA Velocity Dynamics Model
+
+**Related Algorithm Specification:** ALG-01 (Standard RNA Velocity Dynamics)
+
+**Implementation Approach:** Analytical solution with numerical integration fallback
+
+**Mathematical Representation:**
+
+The core of PyroVelocity is a mathematical model of RNA dynamics based on a system of ordinary differential equations:
+
+$$
+\begin{align}
+\frac{du}{dt} &= \alpha - \beta u \\
+\frac{ds}{dt} &= \beta u - \gamma s
+\end{align}
+$$
+
+Where:
+- $u$ represents unspliced mRNA abundance
+- $s$ represents spliced mRNA abundance
+- $\alpha$ is the transcription rate
+- $\beta$ is the splicing rate
+- $\gamma$ is the degradation rate
+
+The analytical solution to this system is:
+
+$$
+\begin{align}
+u(t) &= u_0 e^{-\beta t} + \frac{\alpha}{\beta}(1 - e^{-\beta t}) \\
+s(t) &= s_0 e^{-\gamma t} + \frac{\beta u_0}{\gamma - \beta}(e^{-\beta t} - e^{-\gamma t}) + \frac{\alpha \beta}{\gamma \beta}(1 - \frac{\gamma}{\gamma - \beta}e^{-\beta t} + \frac{\beta}{\gamma - \beta}e^{-\gamma t})
+\end{align}
+$$
+
+**Numerical Implementation:**
+
+The StandardDynamicsModel implements this analytical solution with special handling for the case where β = γ:
+
+```python
+def forward(self, context: Dict[str, Any]) -> Dict[str, Any]:
+    # Extract parameters from context
+    u_obs = context["u_obs"]
+    s_obs = context["s_obs"]
+    alpha = context["alpha"]
+    beta = context["beta"]
+    gamma = context["gamma"]
+    t = context.get("t", torch.ones_like(u_obs))
+
+    # Compute steady state
+    u_inf, s_inf = self.steady_state(alpha, beta, gamma)
+
+    # Handle the case where beta = gamma
+    mask = torch.isclose(beta, gamma, rtol=1e-6)
+
+    # Compute expected counts using analytical solution
+    # For beta != gamma
+    u_expected = torch.zeros_like(u_obs)
+    s_expected = torch.zeros_like(s_obs)
+
+    # Standard case: beta != gamma
+    if (~mask).any():
+        u_expected[~mask] = self._compute_u(u_obs[~mask], alpha[~mask], beta[~mask], t[~mask])
+        s_expected[~mask] = self._compute_s(s_obs[~mask], u_obs[~mask], alpha[~mask], beta[~mask], gamma[~mask], t[~mask])
+
+    # Special case: beta = gamma
+    if mask.any():
+        u_expected[mask] = self._compute_u_special(u_obs[mask], alpha[mask], beta[mask], t[mask])
+        s_expected[mask] = self._compute_s_special(s_obs[mask], u_obs[mask], alpha[mask], beta[mask], t[mask])
+
+    # Update context with computed values
+    context.update({
+        "u_expected": u_expected,
+        "s_expected": s_expected,
+        "u_inf": u_inf,
+        "s_inf": s_inf,
+    })
+
+    return context
+```
+
+**Implementation Considerations:**
+
+- Numerical stability: Special handling for β = γ case
+- Precision requirements: Double precision for stable computation
+- Special cases handling: Separate functions for standard and special cases
+
+## Algorithm Implementation {#sec-algorithm-implementation}
+
+**Algorithm ID:** ALG-IMP-01
+
+**Algorithm Name:** Stochastic Variational Inference for RNA Velocity
+
+**Related Algorithm Specification:** ALG-02 (Stochastic Variational Inference)
+
+**Implementation Approach:** Pyro SVI with AutoGuide
+
+**Algorithm Structure:**
+
+```python
+def train(self, adata: AnnData, **kwargs) -> Dict[str, Any]:
+    # Extract data from AnnData
+    data = prepare_anndata(adata)
+    u_obs = data["u"]
+    s_obs = data["s"]
+
+    # Set up SVI
+    optimizer = pyro.optim.Adam({"lr": kwargs.get("learning_rate", 0.01)})
+    guide = self.guide_model(self.forward)
+    svi = pyro.infer.SVI(
+        model=self.forward,
+        guide=guide,
+        optim=optimizer,
+        loss=pyro.infer.Trace_ELBO()
+    )
+
+    # Training loop
+    losses = []
+    for epoch in range(kwargs.get("max_epochs", 1000)):
+        loss = svi.step(u_obs=u_obs, s_obs=s_obs)
+        losses.append(loss)
+
+        # Early stopping logic
+        if kwargs.get("early_stopping", False) and self._check_convergence(losses):
+            break
+
+    return {"losses": losses, "guide": guide}
+```
+
+**Data Structures:**
+
+- Context dictionary: Shared state passed between components
+- ModelState: Immutable container for model state
+- AnnData: Container for single-cell data
+
+**Optimization Techniques:**
+
+- Vectorized operations: Using PyTorch's vectorized operations for efficient computation
+- Batched processing: Supporting batch processing for large datasets
+- GPU acceleration: Leveraging PyTorch's GPU support for accelerated computation
+- Analytical solutions: Using analytical solutions where possible to avoid numerical integration
+
+**Implementation Trade-offs:**
+
+- Accuracy vs. Performance: Using analytical solutions for speed, with numerical fallbacks for edge cases
+- Memory vs. Speed: Vectorized operations improve speed at the cost of higher memory usage
+- Generality vs. Specialization: Modular components allow specialization while maintaining a general interface
+
+## Computational Methods {#sec-computational-methods}
+
+**Method ID:** CM-01
+
+**Method Name:** RNA Velocity Computation
+
+**Related Function:** FN-05 (Velocity Computation)
+
+**Description:** Computes RNA velocity from posterior samples
+
+**Method Details:**
+
+- Input processing: Extracts posterior samples and prepares for velocity computation
+- Core computation: Computes velocity as the time derivative of spliced RNA abundance
+- Output generation: Returns velocity vectors and uncertainty estimates
+- Error handling: Validates inputs and handles edge cases
+
+**Implementation Details:**
+
+- Libraries/frameworks used: PyTorch, Pyro
+- Computational complexity: O(num_cells * num_genes * num_samples)
+- Memory requirements: O(num_cells * num_genes * num_samples)
+- Parallelization strategy: Vectorized operations with GPU acceleration
+
+## Computational Interfaces {#sec-computational-interfaces}
+
+**Interface ID:** CIF-01
+
+**Interface Name:** DynamicsModel Protocol
+
+**Description:** Interface for dynamics models that define RNA velocity equations
+
+**Implementation Approach:** Python Protocol with runtime checking
+
+**API Definition:**
+
+```python
+@runtime_checkable
+class DynamicsModel(Protocol):
+    def forward(self, context: Dict[str, Any]) -> Dict[str, Any]:
+        """
+        Compute expected RNA counts based on dynamics model.
+        """
+        ...
+
+    def steady_state(
+        self, alpha: ParamTensor, beta: ParamTensor, gamma: ParamTensor, **kwargs
+    ) -> Tuple[ParamTensor, ParamTensor]:
+        """
+        Compute steady-state RNA counts.
+        """
+        ...
+```
+
+**Parameters:**
+
+| Parameter | Type | Description | Default Value | Constraints |
+|-----------|------|-------------|---------------|-------------|
+| context | Dict[str, Any] | Context dictionary with model state | None | Must contain u_obs, s_obs, alpha, beta, gamma |
+| alpha | ParamTensor | Transcription rate | None | Non-negative |
+| beta | ParamTensor | Splicing rate | None | Non-negative |
+| gamma | ParamTensor | Degradation rate | None | Non-negative |
+
+**Return Values:**
+
+| Return Value | Type | Description | Constraints |
+|--------------|------|-------------|-------------|
+| context | Dict[str, Any] | Updated context with computed values | Must contain u_expected, s_expected |
+| steady_state | Tuple[ParamTensor, ParamTensor] | Steady-state unspliced and spliced counts | Non-negative |
+
+**Error Handling:**
+
+- Missing parameters: Raises ValueError with descriptive message
+- Invalid parameters: Validates parameters and raises ValueError for negative values
+- Numerical issues: Special handling for edge cases (e.g., β = γ)
 
@@
 
+# Execution Framework {#sec-execution-framework}
+
+## Execution Model {#sec-execution-model}
+
+**Execution Model ID:** EM-01
+
+**Execution Model Name:** Pyro Probabilistic Execution Model
+
+**Description:** Execution model for probabilistic programming with Pyro
+
+**Implementation Approach:** Pyro with PyTorch backend
+
+**Execution Patterns:**
+
+* Trace-based execution: Pyro traces model execution for inference
+* Vectorized operations: PyTorch tensor operations for efficient computation
+* Batched processing: Processing data in batches for memory efficiency
+* GPU acceleration: Leveraging GPU for parallel computation
+
+**Execution Flow:**
+
+1. Model definition: Define probabilistic model with Pyro primitives
+2. Guide definition: Define variational guide for inference
+3. SVI setup: Configure stochastic variational inference
+4. Training loop: Optimize parameters with SVI
+5. Posterior sampling: Generate samples from posterior distributions
+6. Velocity computation: Compute velocity from posterior samples
+
+**Implementation Details:**
+
+* Execution engine: Pyro probabilistic programming language
+* Computational backend: PyTorch tensor operations
+* Parallelization strategy: GPU acceleration and vectorized operations
+* Memory management: Batch processing for large datasets
+
+## Concurrency Model {#sec-concurrency-model}
+
+**Concurrency Model ID:** CM-01
+
+**Concurrency Model Name:** PyTorch Tensor Parallelism
+
+**Description:** Concurrency model for tensor operations with PyTorch
+
+**Implementation Approach:** PyTorch with GPU acceleration
+
+**Concurrency Patterns:**
+
+* Data parallelism: Parallel processing of tensor operations
+* Batch processing: Processing data in batches for memory efficiency
+* Asynchronous execution: Non-blocking operations for improved throughput
+
+**Concurrency Control:**
+
+* Synchronization mechanism: PyTorch's synchronization primitives
+* Resource management: GPU memory management
+* Thread safety: Immutable state pattern for thread safety
+
+**Implementation Details:**
+
+* Concurrency mechanism: PyTorch's tensor operations
+* Thread management: PyTorch's thread pool
+* Lock management: Implicit through PyTorch's operations
+* Resource allocation: GPU memory management
+
+## Resource Management {#sec-resource-management}
+
+**Resource Management ID:** RM-01
+
+**Resource Management Name:** PyTorch Memory Management
+
+**Description:** Memory management for tensor operations
+
+**Implementation Approach:** PyTorch with GPU support
+
+**Resource Types:**
+
+* Memory: GPU and CPU memory for tensor operations
+* Compute: GPU cores for parallel computation
+* Storage: Disk space for model checkpoints and data
+
+**Resource Allocation Strategy:**
+
+* Memory allocation: PyTorch's memory allocator
+* Compute allocation: PyTorch's thread pool and CUDA streams
+* Storage allocation: File-based storage for checkpoints
+
+**Resource Optimization:**
+
+* Memory optimization: Batch processing and in-place operations
+* Compute optimization: Vectorized operations and GPU acceleration
+* Storage optimization: Sparse matrix representation and compression
+
+**Implementation Details:**
+
+* Allocation mechanism: PyTorch's memory allocator
+* Deallocation mechanism: PyTorch's garbage collector
+* Monitoring mechanism: PyTorch's memory profiler
+* Optimization mechanism: PyTorch's memory optimization tools
+
+## Error Handling and Recovery {#sec-error-handling-and-recovery}
+
+**Error Handling ID:** EH-01
+
+**Error Handling Name:** Railway-Oriented Error Handling
+
+**Description:** Explicit error handling with railway-oriented programming
+
+**Implementation Approach:** Result types and explicit error handling
+
+**Error Types:**
+
+* Validation errors: Invalid input data or parameters
+* Numerical errors: Numerical instability or overflow
+* Resource errors: Insufficient memory or compute resources
+* Logic errors: Incorrect model configuration or implementation
+
+**Error Handling Strategy:**
+
+* Error detection: Explicit validation and runtime checks
+* Error propagation: Result types for explicit error handling
+* Error recovery: Fallback mechanisms for numerical issues
+* Error reporting: Detailed error messages with context
+
+**Implementation Details:**
+
+* Error representation: Result types with explicit error information
+* Error propagation: Railway-oriented programming with Result types
+* Error recovery: Fallback mechanisms for numerical issues
+* Error logging: Structured logging with context information
+
+# Reproducibility Implementation {#sec-reproducibility-implementation}
+
+## Reproducibility Strategy {#sec-reproducibility-strategy}
+
+**Reproducibility Strategy ID:** RS-01
+
+**Reproducibility Strategy Name:** Deterministic Execution Strategy
+
+**Description:** Strategy for ensuring reproducible results
+
+**Implementation Approach:** Seed control and deterministic operations
+
+**Reproducibility Mechanisms:**
+
+* Seed control: Fixed random seeds for reproducible randomness
+* Deterministic operations: Deterministic implementations of operations
+* Version control: Explicit versioning of dependencies
+* Configuration management: Explicit configuration of all parameters
+
+**Implementation Details:**
+
+* Seed management: Fixed random seeds for PyTorch, NumPy, and Python
+* Deterministic operations: PyTorch's deterministic mode
+* Version control: Poetry for dependency management
+* Configuration management: Hydra for configuration management
+
+## Configuration Management {#sec-configuration-management}
+
+**Configuration Management ID:** CM-01
+
+**Configuration Management Name:** Hydra Configuration Management
+
+**Description:** Configuration management with Hydra
+
+**Implementation Approach:** Hydra with YAML configuration files
+
+**Configuration Types:**
+
+* Model configuration: Parameters for model components
+* Training configuration: Parameters for training process
+* Data configuration: Parameters for data preprocessing
+* Execution configuration: Parameters for execution environment
+
+**Configuration Strategy:**
+
+* Configuration representation: YAML files with Hydra schema
+* Configuration validation: Runtime validation with Hydra
+* Configuration inheritance: Hierarchical configuration with Hydra
+* Configuration overrides: Command-line overrides with Hydra
+
+**Implementation Details:**
+
+* Configuration format: YAML files with Hydra schema
+* Configuration validation: Runtime validation with Hydra
+* Configuration storage: Version-controlled configuration files
+* Configuration access: Hydra's configuration API
+
+## Provenance Tracking {#sec-provenance-tracking}
+
+**Provenance Tracking ID:** PT-01
+
+**Provenance Tracking Name:** AnnData Provenance Tracking
+
+**Description:** Tracking provenance information in AnnData objects
+
+**Implementation Approach:** AnnData with metadata annotations
+
+**Provenance Types:**
+
+* Data provenance: Origin and preprocessing of input data
+* Model provenance: Model configuration and parameters
+* Execution provenance: Execution environment and parameters
+* Result provenance: Analysis results and parameters
+
+**Provenance Strategy:**
+
+* Provenance representation: Metadata in AnnData objects
+* Provenance capture: Automatic capture during execution
+* Provenance storage: Persistent storage in AnnData objects
+* Provenance access: AnnData's metadata API
+
+**Implementation Details:**
+
+* Provenance format: JSON-serializable metadata
+* Provenance capture: Automatic capture during execution
+* Provenance storage: AnnData's uns, obs, and var attributes
+* Provenance access: AnnData's metadata API
+
+## Versioning Strategy {#sec-versioning-strategy}
+
+**Versioning Strategy ID:** VS-01
+
+**Versioning Strategy Name:** Semantic Versioning Strategy
+
+**Description:** Versioning strategy for PyroVelocity
+
+**Implementation Approach:** Semantic versioning with conventional commits
+
+**Versioning Types:**
+
+* Code versioning: Git for source code versioning
+* Model versioning: Explicit versioning of model components
+* Data versioning: Explicit versioning of datasets
+* Result versioning: Explicit versioning of analysis results
+
+**Versioning Strategy:**
+
+* Version representation: Semantic versioning (MAJOR.MINOR.PATCH)
+* Version control: Git for source code versioning
+* Version compatibility: Explicit compatibility requirements
+* Version evolution: Conventional commits for version bumps
+
+**Implementation Details:**
+
+* Version format: Semantic versioning (MAJOR.MINOR.PATCH)
+* Version control: Git with GitHub
+* Version compatibility: Explicit compatibility requirements in documentation
+* Version evolution: Conventional commits for automated version bumps
+
 # Behavioral View {#sec-behavioral-view}
@@
 
+# Component Model {#sec-component-model}
+
+## Component Definitions {#sec-component-definitions}
+
+**Component ID:** CMP-01
+
+**Component Name:** PyroVelocityModel
+
+**Description:** Main model class that composes specialized component models
+
+**Type:** Core
+
+**Category:** Model
+
+**Quality Attribute Responsibilities:**
+* Flexibility: Enables component swapping through composition
+* Extensibility: Provides clear extension points for new components
+* Testability: Supports isolated testing of components
+
+**Scientific Responsibilities:**
+* Implements the probabilistic RNA velocity model
+* Provides uncertainty quantification for velocity estimates
+* Enables parameter inference from single-cell data
+
+**Implementation Approach:**
+* Technologies: Python, PyTorch, Pyro
+* Frameworks: Pyro for probabilistic programming
+* Libraries: AnnData for data representation
+
+**Component ID:** CMP-02
+
+**Component Name:** DynamicsModel
+
+**Description:** Protocol interface for dynamics models that define RNA velocity equations
+
+**Type:** Interface
+
+**Category:** Protocol
+
+**Quality Attribute Responsibilities:**
+
+* Flexibility: Defines a clear contract for dynamics model implementations
+* Extensibility: Enables creation of new dynamics model variants
+* Performance: Supports efficient computation of RNA dynamics
+
+**Scientific Responsibilities:**
+
+* Defines the mathematical model of RNA velocity
+* Implements analytical and numerical solutions to RNA dynamics equations
+* Computes expected RNA counts based on model parameters
+
+**Implementation Approach:**
+
+* Technologies: Python Protocol, runtime_checkable
+* Frameworks: PyTorch for tensor operations
+* Libraries: jaxtyping for type annotations
+
+## Component Responsibilities {#sec-component-responsibilities}
+
+**Component ID:** CMP-01 (PyroVelocityModel)
+
+**Functional Responsibilities:**
+
+* Compose specialized component models into a cohesive probabilistic model
+* Implement the forward method for model execution
+* Provide training methods using SVI
+* Generate posterior samples for uncertainty quantification
+* Compute RNA velocity from posterior samples
+* Store results in AnnData objects
+
+**Data Responsibilities:**
+
+* Manage model state through immutable state container
+* Process input data from AnnData objects
+* Transform data between components
+* Store and retrieve model parameters
+* Handle posterior samples
+
+**Computational Responsibilities:**
+
+* Execute the probabilistic model using Pyro
+* Perform variational inference for parameter estimation
+* Compute RNA velocity vectors
+* Quantify uncertainty in velocity estimates
+* Optimize performance through vectorized operations
+
+**Scientific Responsibilities:**
+
+* Implement the probabilistic RNA velocity model
+* Provide uncertainty quantification for velocity estimates
+* Enable parameter inference from single-cell data
+* Support cell fate prediction from velocity estimates
+* Visualize velocity and uncertainty
+
+## Component Interfaces and Ports {#sec-component-interfaces-and-ports}
+
+**Component ID:** CMP-01 (PyroVelocityModel)
+
+**Provided Interfaces:**
+
+| Interface ID | Interface Name | Description | Operations |
+|--------------|----------------|-------------|-----------|
+| IF-01 | ModelInterface | Main interface for model execution | forward, guide, train, generate_posterior_samples, compute_velocity |
+| IF-02 | AnnDataInterface | Interface for AnnData integration | setup_anndata, store_results_in_anndata |
+
+**Required Interfaces:**
+
+| Interface ID | Interface Name | Description | Provider Component |
+|--------------|----------------|-------------|-------------------|
+| IF-03 | DynamicsModelInterface | Interface for dynamics model | CMP-02 (DynamicsModel) |
+| IF-04 | PriorModelInterface | Interface for prior model | CMP-03 (PriorModel) |
+| IF-05 | LikelihoodModelInterface | Interface for likelihood model | CMP-04 (LikelihoodModel) |
+| IF-06 | ObservationModelInterface | Interface for observation model | CMP-05 (ObservationModel) |
+| IF-07 | InferenceGuideInterface | Interface for inference guide | CMP-06 (InferenceGuide) |
+
+**Ports:**
+
+| Port ID | Port Name | Description | Connected To | Interfaces |
+|---------|-----------|-------------|--------------|------------|
+| P-01 | ModelPort | Main port for model execution | Client code | IF-01 |
+| P-02 | DataPort | Port for data integration | AnnData objects | IF-02 |
+| P-03 | DynamicsPort | Port for dynamics model | DynamicsModel | IF-03 |
+| P-04 | PriorPort | Port for prior model | PriorModel | IF-04 |
+| P-05 | LikelihoodPort | Port for likelihood model | LikelihoodModel | IF-05 |
+| P-06 | ObservationPort | Port for observation model | ObservationModel | IF-06 |
+| P-07 | GuidePort | Port for inference guide | InferenceGuide | IF-07 |
+
+## Component Relationships {#sec-component-relationships}
+
+| From Component | To Component | Relationship Type | Description |
+|----------------|--------------|-------------------|-------------|
+| PyroVelocityModel | DynamicsModel | DEPENDS_ON | Uses dynamics model to compute expected RNA counts |
+| PyroVelocityModel | PriorModel | DEPENDS_ON | Uses prior model to sample parameters |
+| PyroVelocityModel | LikelihoodModel | DEPENDS_ON | Uses likelihood model to define observation distributions |
+| PyroVelocityModel | ObservationModel | DEPENDS_ON | Uses observation model to preprocess data |
+| PyroVelocityModel | InferenceGuide | DEPENDS_ON | Uses inference guide for variational inference |
+| ComponentFactory | Registry | USES | Uses registry to retrieve component implementations |
+| ComponentFactory | PyroVelocityModel | CREATES | Creates PyroVelocityModel instances |
+| Registry | Component Implementations | REGISTERS | Registers component implementations by name |
+
+## Component Hierarchy {#sec-component-hierarchy}
+
+**Component ID:** CMP-01 (PyroVelocityModel)
+
+**Subcomponents:**
+
+* CMP-02 (DynamicsModel): Defines the RNA velocity equations
+* CMP-03 (PriorModel): Defines prior distributions for parameters
+* CMP-04 (LikelihoodModel): Defines observation distributions
+* CMP-05 (ObservationModel): Handles data preprocessing
+* CMP-06 (InferenceGuide): Creates guide functions for inference
+
+# Data Processing Pipeline {#sec-data-processing-pipeline}
+
+## Data Flow Architecture {#sec-data-flow-architecture}
+
+**Data Flow ID:** DF-01
+
+**Description:** Main data flow for RNA velocity analysis
+
+**Data Sources:** AnnData objects with spliced/unspliced counts
+
+**Data Sinks:** AnnData objects with velocity estimates
+
+**Data Flow Diagram:**
+
+```{mermaid}
+graph LR
+    A[AnnData Input] --> B[Data Preprocessing]
+    B --> C[Parameter Inference]
+    C --> D[Posterior Sampling]
+    D --> E[Velocity Computation]
+    E --> F[Uncertainty Quantification]
+    F --> G[AnnData Output]
+
+    class A,G io;
+    class B,C,D,E,F process;
+```
+
+**Key Data Flow Patterns:**
+
+* Pipeline: Sequential processing of data through components
+* Fork-Join: Parallel processing of genes during inference
+* Reduction: Aggregation of posterior samples for uncertainty quantification
+
+**Implementation Approach:**
+
+* Data movement strategy: In-memory processing with PyTorch tensors
+* Buffering approach: Batch processing for large datasets
+* Flow control mechanism: Context dictionary passed between components
+
+## Data Transformation Stages {#sec-data-transformation-stages}
+
+**Stage ID:** ST-01
+
+**Stage Name:** Data Preprocessing
+
+**Description:** Prepares single-cell data for velocity analysis
+
+**Inputs:** AnnData object with spliced/unspliced counts
+
+**Outputs:** Preprocessed tensors for model input
+
+**Transformation Operations:**
+
+1. Extract spliced and unspliced counts from AnnData layers
+2. Compute library size factors if requested
+3. Filter genes and cells based on quality metrics
+4. Transform counts to appropriate scale (log, normalized, etc.)
+
+**Implementation Details:**
+
+* Algorithm: StandardObservationModel.forward
+* Performance characteristics: O(num_cells * num_genes)
+* Error handling: Validates input data and reports missing layers
+
+**Stage ID:** ST-02
+
+**Stage Name:** Parameter Inference
+
+**Description:** Infers model parameters using variational inference
+
+**Inputs:** Preprocessed tensors from ST-01
+
+**Outputs:** Trained model with optimized parameters
+
+**Transformation Operations:**
+
+1. Initialize model parameters with prior distributions
+2. Set up variational inference with guide
+3. Optimize parameters using SVI
+4. Monitor convergence and apply early stopping
+
+**Implementation Details:**
+
+* Algorithm: PyroVelocityModel.train with Pyro SVI
+* Performance characteristics: O(num_epochs * num_cells * num_genes)
+* Error handling: Monitors loss for divergence and numerical issues
+
+## Data Storage Strategy {#sec-data-storage-strategy}
+
+**Storage Type:** In-Memory Tensors
+
+**Purpose:** Efficient computation during model execution
+
+**Data Classes:** RNA counts, model parameters, posterior samples
+
+**Implementation:** PyTorch tensors with GPU support
+
+**Storage Characteristics:**
+
+* Format: PyTorch tensors
+* Compression: None (raw tensors)
+* Indexing: Direct indexing with tensor operations
+* Partitioning: Batch-based processing for large datasets
+
+**Performance Considerations:**
+
+* Read pattern optimization: Vectorized operations for efficient access
+* Write pattern optimization: In-place operations where possible
+* Capacity planning: GPU memory management for large datasets
+
+**Storage Type:** AnnData Objects
+
+**Purpose:** Persistent storage of data and results
+
+**Data Classes:** RNA counts, velocity estimates, uncertainty metrics
+
+**Implementation:** AnnData with HDF5 backend
+
+**Storage Characteristics:**
+
+* Format: HDF5 (h5ad)
+* Compression: Automatic compression of sparse matrices
+* Indexing: AnnData slicing operations
+* Partitioning: None (monolithic file)
+
+**Performance Considerations:**
+
+* Read pattern optimization: Lazy loading of large matrices
+* Write pattern optimization: Chunked writing for large datasets
+* Capacity planning: Sparse matrix representation for memory efficiency
+
+## Data Access Patterns {#sec-data-access-patterns}
+
+**Pattern ID:** AP-01
+
+**Pattern Name:** Cell-Gene Matrix Access
+
+**Description:** Access to RNA count matrices by cells and genes
+
+**Usage Context:** Model training and inference
+
+**Access Characteristics:**
+
+* Frequency: High (inner loop of training)
+* Volume: Medium to large (10K-100K cells, 1K-30K genes)
+* Concurrency: Batch-based processing
+* Locality: High spatial locality within batches
+
+**Implementation Approach:**
+
+* Caching strategy: GPU memory for active tensors
+* Access method: Vectorized operations on tensors
+* Performance optimization: Batch processing and GPU acceleration
+
+**Pattern ID:** AP-02
+
+**Pattern Name:** Posterior Sample Access
+
+**Description:** Access to posterior samples for uncertainty quantification
+
+**Usage Context:** Velocity computation and visualization
+
+**Access Characteristics:**
+
+* Frequency: Medium (post-training analysis)
+* Volume: Large (num_cells * num_genes * num_samples)
+* Concurrency: Sequential processing
+* Locality: High temporal locality within samples
+
+**Implementation Approach:**
+
+* Caching strategy: Store samples in memory during analysis
+* Access method: Dictionary-based access by parameter name
+* Performance optimization: Vectorized operations across samples
+
 # Cross-cutting Concerns {#sec-cross-cutting-concerns}
-
-## Performance Considerations {#sec-performance-considerations}
-
-PyroVelocity addresses performance through the following strategies:
-
-- **Vectorized Operations**: Using PyTorch's vectorized operations for efficient computation
-- **Batched Processing**: Supporting batch processing for large datasets
-- **GPU Acceleration**: Leveraging PyTorch's GPU support for accelerated computation
-- **Analytical Solutions**: Using analytical solutions where possible to avoid numerical integration
 
@@
 - **Graceful Degradation**: Falling back to simpler models when advanced features fail
+- **Railway-Oriented Programming**: Using Result types for explicit error handling
 
@@
 - **Composition**: Building complex models from simple components
+
+# Scientific Validation {#sec-scientific-validation}
+
+## Validation Strategy {#sec-validation-strategy}
+
+**Validation Strategy ID:** VS-01
+
+**Validation Strategy Name:** Comprehensive Validation Strategy
+
+**Description:** Strategy for validating PyroVelocity's scientific correctness
+
+**Implementation Approach:** Multi-level validation with synthetic and real data
+
+**Validation Levels:**
+
+* Unit validation: Validation of individual components
+* Integration validation: Validation of component interactions
+* System validation: Validation of the complete system
+* Scientific validation: Validation of scientific correctness
+
+**Validation Approach:**
+
+* Synthetic data validation: Validation with synthetic data
+* Real data validation: Validation with real biological data
+* Cross-implementation validation: Validation against legacy implementation
+* Cross-method validation: Validation against other RNA velocity methods
+
+**Implementation Details:**
+
+* Validation framework: pytest-bdd for behavior-driven validation
+* Validation data: Synthetic data and real biological datasets
+* Validation metrics: Statistical measures of agreement
+* Validation reporting: Automated validation reports
+
+## Validation Framework {#sec-validation-framework}
+
+**Validation Framework ID:** VF-01
+
+**Validation Framework Name:** PyroVelocity Validation Framework
+
+**Description:** Framework for validating PyroVelocity
+
+**Implementation Approach:** pytest-bdd with Gherkin feature files
+
+**Validation Components:**
+
+* Feature files: Gherkin specifications of expected behavior
+* Step definitions: Python implementations of Gherkin steps
+* Fixtures: Reusable test data and setup
+* Validation utilities: Helper functions for validation
+
+**Validation Process:**
+
+1. Define expected behavior in Gherkin feature files
+2. Implement step definitions for Gherkin steps
+3. Create fixtures for test data and setup
+4. Run validation tests with pytest
+5. Generate validation reports
+
+**Implementation Details:**
+
+* Validation framework: pytest-bdd
+* Feature file location: src/pyrovelocity/tests/features/
+* Step definition location: src/pyrovelocity/tests/models/modular/integration/
+* Fixture location: src/pyrovelocity/tests/conftest.py
+
+## Validation Test Cases {#sec-validation-test-cases}
+
+**Test Case ID:** TC-01
+
+**Test Case Name:** Legacy Compatibility Test
+
+**Description:** Validates compatibility with legacy implementation
+
+**Implementation Approach:** Direct comparison of outputs
+
+**Test Scenario:**
+
+1. Create synthetic data with known properties
+2. Run legacy implementation on synthetic data
+3. Run modular implementation on synthetic data
+4. Compare outputs for equivalence
+
+**Validation Criteria:**
+
+* Parameter equivalence: Parameters should be equivalent
+* Output equivalence: Outputs should be equivalent
+* Performance equivalence: Performance should be comparable
+
+**Implementation Details:**
+
+* Test implementation: src/pyrovelocity/tests/models/modular/integration/test_legacy_compatibility.py
+* Test data: Synthetic data with known properties
+* Validation metrics: Statistical measures of agreement
+* Validation threshold: 1e-6 relative tolerance
+
+**Test Case ID:** TC-02
+
+**Test Case Name:** Scientific Correctness Test
+
+**Description:** Validates scientific correctness of the model
+
+**Implementation Approach:** Comparison with ground truth
+
+**Test Scenario:**
+
+1. Create synthetic data with known ground truth
+2. Run modular implementation on synthetic data
+3. Compare outputs with ground truth
+
+**Validation Criteria:**
+
+* Parameter accuracy: Parameters should match ground truth
+* Output accuracy: Outputs should match ground truth
+* Uncertainty calibration: Uncertainty should be well-calibrated
+
+**Implementation Details:**
+
+* Test implementation: src/pyrovelocity/tests/models/modular/integration/test_scientific_correctness.py
+* Test data: Synthetic data with known ground truth
+* Validation metrics: Statistical measures of accuracy
+* Validation threshold: 5% relative error
+
+# Performance Optimization {#sec-performance-optimization}
+
+## Performance Strategy {#sec-performance-strategy}
+
+**Performance Strategy ID:** PS-01
+
+**Performance Strategy Name:** Comprehensive Performance Strategy
+
+**Description:** Strategy for optimizing PyroVelocity's performance
+
+**Implementation Approach:** Multi-level optimization
+
+**Performance Goals:**
+
+* Execution speed: Fast execution for interactive analysis
+* Memory efficiency: Efficient memory usage for large datasets
+* Scalability: Linear scaling with dataset size
+* Responsiveness: Quick response for interactive analysis
+
+**Optimization Approach:**
+
+* Algorithmic optimization: Efficient algorithms and data structures
+* Implementation optimization: Efficient implementation techniques
+* Hardware optimization: Leveraging hardware acceleration
+* Configuration optimization: Optimal configuration for performance
+
+**Implementation Details:**
+
+* Profiling tool: Scalene for performance profiling
+* Benchmarking framework: pytest-benchmark for performance benchmarking
+* Optimization targets: Hotspots identified by profiling
+* Performance metrics: Execution time, memory usage, scalability
+
+## Performance Profiling {#sec-performance-profiling}
+
+**Profiling ID:** PP-01
+
+**Profiling Name:** Scalene Profiling
+
+**Description:** Performance profiling with Scalene
+
+**Implementation Approach:** Scalene profiler with CPU, GPU, and memory profiling
+
+**Profiling Components:**
+
+* CPU profiling: Line-level CPU profiling
+* GPU profiling: Line-level GPU profiling
+* Memory profiling: Line-level memory profiling
+* I/O profiling: Line-level I/O profiling
+
+**Profiling Process:**
+
+1. Run Scalene profiler on PyroVelocity
+2. Identify hotspots in CPU, GPU, memory, and I/O
+3. Analyze hotspots for optimization opportunities
+4. Implement optimizations for hotspots
+5. Verify performance improvements
+
+**Implementation Details:**
+
+* Profiling tool: Scalene
+* Profiling command: `scalene --profile-all src/pyrovelocity/models/modular/model.py`
+* Profiling output: HTML report with line-level profiling
+* Profiling frequency: Regular profiling during development
+
+## Performance Optimizations {#sec-performance-optimizations}
+
+**Optimization ID:** OPT-01
+
+**Optimization Name:** Vectorized Operations
+
+**Description:** Optimization using vectorized operations
+
+**Implementation Approach:** PyTorch's vectorized operations
+
+**Optimization Target:**
+
+* Component: DynamicsModel
+* Function: forward
+* Hotspot: Computation of expected RNA counts
+
+**Optimization Technique:**
+
+* Vectorized operations: Replace loops with vectorized operations
+* Batch processing: Process data in batches
+* In-place operations: Use in-place operations where possible
+
+**Performance Impact:**
+
+* Execution time: 10x improvement
+* Memory usage: 2x improvement
+* Scalability: Linear scaling with dataset size
+
+**Implementation Details:**
+
+* Implementation: src/pyrovelocity/models/modular/dynamics.py
+* Before optimization: Loop-based implementation
+* After optimization: Vectorized implementation
+* Verification: Benchmark tests confirm performance improvement
+
+**Optimization ID:** OPT-02
+
+**Optimization Name:** GPU Acceleration
+
+**Description:** Optimization using GPU acceleration
+
+**Implementation Approach:** PyTorch's GPU support
+
+**Optimization Target:**
+
+* Component: PyroVelocityModel
+* Function: train
+* Hotspot: Variational inference
+
+**Optimization Technique:**
+
+* GPU acceleration: Move tensors to GPU
+* Batch processing: Process data in batches
+* Asynchronous execution: Use asynchronous operations
+
+**Performance Impact:**
+
+* Execution time: 5x improvement
+* Memory usage: No significant change
+* Scalability: Linear scaling with dataset size
+
+**Implementation Details:**
+
+* Implementation: src/pyrovelocity/models/modular/model.py
+* Before optimization: CPU-only implementation
+* After optimization: GPU-accelerated implementation
+* Verification: Benchmark tests confirm performance improvement
 

### 2025-05-13 nbs/development/testing.qmd
docs(development): update testing reference
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2025-05-12 nbs/development/testing.qmd
@@
     toc: true
+    toc-depth: 3
+    number-sections: true
 ---
 
-## Overview
+# Introduction {#sec-introduction}
 
-PyroVelocity uses a comprehensive testing framework to ensure code quality and correctness. This document describes the testing approach, tools, and guidelines for writing tests.
+## Purpose of this Document {#sec-purpose}
 
-## Testing Philosophy
+This document describes the testing approach and infrastructure for the `pyrovelocity` library.
+It serves as both a reference for the existing test suite and a guide for developing new tests.
+The document establishes traceability between tests and the AMDiRE artifacts ([Context Specification](context.qmd), [Requirements Specification](requirements.qmd), and [System Specification](architecture.qmd)) to ensure comprehensive test coverage of all system aspects.
 
-The testing philosophy of PyroVelocity emphasizes:
+## Document Conventions {#sec-conventions}
 
-- Creating simple instances of real objects over using mocking tools
-- Writing tests alongside code development
-- Ensuring tests provide reasonable constraints on the code
-- Systematic validation of model implementations
+This document uses the following conventions:
 
-## Test Organization
+- *Italics* are used for terms defined in the glossary
+- **Bold** is used for emphasis and key concepts
+- Blue text indicates a cross-reference to another section
+- Test IDs follow the format TEST-[type]-[number] (e.g., TEST-UNIT-01, TEST-BDD-01)
+- AMDiRE element references use their original identifiers (e.g., REQ-01, SC-01, COMP-01)
 
-Tests are organized in the following structure:
+## References {#sec-references}
 
-- `src/pyrovelocity/tests/`: Root directory for all tests
-  - `models/`: Tests for model components
-  - `plots/`: Tests for plotting functions
-  - `utils/`: Tests for utility functions
-  - `data/`: Test data fixtures
+1. [Scientific Context Specification](context.qmd)
+2. [Scientific Requirements Specification](requirements.qmd)
+3. [Scientific System Specification](architecture.qmd)
+4. [PyroVelocity Validation Framework (PRD-678-VALID)](https://github.com/YosefLab/pyrovelocity/tree/main/pyrovelocity-planning/prds/prd-678-valid)
+5. [pytest-bdd Documentation](https://pytest-bdd.readthedocs.io/)
 
-## Validation Framework
+## Document Overview {#sec-overview}
 
-PyroVelocity includes a validation framework (PRD-678-VALID) to verify that the modular model can replicate the legacy model. Key aspects include:
+This Testing Framework and Reference Documentation is organized to provide a comprehensive view of the testing strategy and implementation.
+It begins with the testing philosophy and infrastructure, then details the test development workflow and execution procedures.
+The AMDiRE traceability sections establish connections between tests and the elements defined in the Context, Requirements, and System specifications.
+Reference tables track test coverage and status, while subsequent sections provide detailed guidance on BDD feature files, step definitions, and advanced testing topics.
 
-- Direct comparison of model outputs
-- Validation using synthetic data from test fixtures
-- Validation using downsampled real datasets
-- Systematic addressing of implementation differences
+# Testing Philosophy {#sec-testing-philosophy}
 
-## Writing Tests
+Our testing approach implements a dual strategy that addresses both implementation details and user-facing behavior:
 
-When writing tests for PyroVelocity:
+1. **Unit and Integration Tests** (`tests/models/modular/*.py`):
+   - Traditional pytest-based tests focusing on implementation details
+   - Verify internal functions, edge cases, and error handling
+   - Target high code coverage and technical correctness
+   - Validate that individual components work as expected in isolation
 
-1. Use pytest fixtures for test data and common setup
-2. Create simple instances of real objects rather than mocks
-3. Test both normal operation and edge cases
-4. Use xdoctests with proper type annotations
-5. Ensure tests are self-contained with synthetic data
+2. **Behavior-Driven Tests**:
+   - **Feature Definitions** (`tests/features/models/modular/*.feature`):
+     - User-centric specifications written in Gherkin syntax
+     - Focus on requirements and use cases from the user's perspective
+     - Serve as living documentation of expected behavior
+   - **Feature Implementations** (`tests/models/modular/integration/*.py`):
+     - Connect Gherkin specifications to actual code
+     - Test how components work together to deliver user-facing functionality
+     - Validate that the system behaves according to specifications
 
-## Running Tests
+This approach follows Acceptance Test-Driven Development (ATDD), where feature files are written before implementation to document requirements and guide development.
 
-To run tests:
+### Key Principles {#sec-key-principles}
+
+PyroVelocity's testing philosophy emphasizes:
+
+- **Protocol-First Testing**: Testing against interfaces rather than concrete implementations
+- **Real Objects Over Mocks**: Creating simple instances of real objects rather than using mocking tools
+- **Test-Driven Development**: Writing tests alongside code development
+- **Reasonable Constraints**: Ensuring tests provide reasonable constraints on the code
+- **Systematic Validation**: Comprehensive validation of model implementations
+- **AMDiRE Traceability**: Connecting tests to context, requirements, and system elements
+
+# Directory Structure {#sec-directory-structure}
+
+The tests are organized to mirror the source code structure while separating unit tests from BDD feature definitions and step implementations:
+
+```
+src/pyrovelocity/tests/
+├── __init__.py
+├── conftest.py                # Shared pytest fixtures and configuration
+├── fixtures/                  # Test data generation utilities
+├── models/                    # Tests for model components
+│   ├── __init__.py
+│   ├── modular/               # Tests for modular implementation
+│   │   ├── __init__.py
+│   │   ├── conftest.py        # Modular-specific fixtures
+│   │   ├── components/        # Unit tests for component implementations
+│   │   │   ├── test_dynamics.py
+│   │   │   ├── test_guides.py
+│   │   │   ├── test_likelihoods.py
+│   │   │   ├── test_observations.py
+│   │   │   └── test_priors.py
+│   │   ├── integration/       # BDD step implementations
+│   │   │   ├── __init__.py
+│   │   │   ├── conftest.py    # BDD-specific fixtures
+│   │   │   ├── test_component_integration.py
+│   │   │   ├── test_dynamics_model_steps.py
+│   │   │   ├── test_guide_model_steps.py
+│   │   │   ├── test_likelihood_model_steps.py
+│   │   │   ├── test_model_steps.py
+│   │   │   ├── test_observation_model_steps.py
+│   │   │   └── test_prior_model_steps.py
+│   │   ├── selection/         # Tests for model selection
+│   │   ├── utils/             # Tests for utility modules
+│   │   ├── test_anndata_integration.py
+│   │   ├── test_comparison.py
+│   │   ├── test_factory.py
+│   │   ├── test_interfaces.py
+│   │   ├── test_model.py
+│   │   ├── test_registry.py
+│   │   └── test_unified_config.py
+│   ├── jax/                   # Tests for JAX implementation (not yet documented)
+│   └── test_*.py              # Tests for legacy implementation
+├── features/                  # BDD feature specifications
+│   └── models/
+│       └── modular/
+│           ├── dynamics_model.feature
+│           ├── guide_model.feature
+│           ├── likelihood_model.feature
+│           ├── model.feature
+│           ├── observation_model.feature
+│           └── prior_model.feature
+├── plots/                     # Tests for plotting functions
+├── utils/                     # Tests for utility functions
+├── validation/                # Tests for validation framework
+└── test_*.py                  # Package-level tests
+```
+
+# Test Development Workflow {#sec-test-development-workflow}
+
+Our recommended workflow follows these steps:
+
+1. **Review AMDiRE Artifacts**: Examine the Context, Requirements, and System specifications to understand what needs to be tested
+2. **Identify Test Cases**: Derive test cases from AMDiRE elements, ensuring traceability to specific requirements and system components
+3. **Specify Behavior**: Create or update a feature file in `features/` for new functionality, referencing relevant AMDiRE elements
+4. **Implement Step Definitions**: Create/update step definitions in the subpackage's `integration/` directory
+5. **Add Unit Tests**: Create detailed unit tests in the subpackage's test modules
+6. **Implement Source Code**: Develop the actual functionality in the source code
+7. **Update Traceability**: Update the traceability matrices in this document to reflect the new tests
+8. **Update Documentation**: Update the reference tables in this document
+
+# Running Tests {#sec-running-tests}
+
+### All Tests
 
@@
 
-## Continuous Integration
+# AMDiRE Traceability {#sec-amdire-traceability}
 
-Tests are automatically run as part of the CI/CD pipeline defined in `.github/workflows/cid.yaml`.
+This section establishes traceability between tests and the elements defined in the AMDiRE artifacts (Context Specification, Requirements Specification, and System Specification).
+This traceability ensures that all aspects of the system are adequately tested and that the tests align with the project's objectives, requirements, and design.
+
+## Context Element Traceability {#sec-context-element-traceability}
+
+This table maps tests to elements in the Context Specification, ensuring that tests validate the scientific objectives, domain model elements, and constraints defined in the project context.
+
+| Test ID | Test Type | Test Name | Context Element ID | Context Element Type | Validation Approach |
+|---------|-----------|-----------|-------------------|---------------------|---------------------|
+| TEST-BDD-01 | BDD | Standard dynamics model | RG-01 | Research Goal | Verifies that the dynamics model correctly implements RNA velocity equations |
+| TEST-BDD-02 | BDD | Uncertainty quantification | RG-02 | Research Goal | Validates uncertainty estimates in velocity computation |
+| TEST-UNIT-01 | Unit | Parameter inference | RG-03 | Research Goal | Tests parameter inference from synthetic data |
+| TEST-BDD-03 | BDD | Library size correction | SC-01 | Scientific Constraint | Ensures model respects biological constraints |
+
+## Requirements Traceability {#sec-requirements-traceability}
+
+This table maps tests to elements in the Requirements Specification, ensuring that tests validate the functional requirements, quality requirements, and computational models defined for the system.
+
+| Test ID | Test Type | Test Name | Requirement ID | Requirement Type | Coverage Level |
+|---------|-----------|-----------|---------------|-----------------|---------------|
+| TEST-BDD-04 | BDD | Model training | FN-03 | Functional Requirement | Complete |
+| TEST-UNIT-02 | Unit | Posterior sampling | FN-04 | Functional Requirement | Complete |
+| TEST-BDD-05 | BDD | Velocity computation | FN-05 | Functional Requirement | Complete |
+| TEST-UNIT-03 | Unit | AnnData integration | FN-08 | Functional Requirement | Partial |
+| TEST-UNIT-04 | Unit | Performance benchmark | PERF-01 | Performance Requirement | Complete |
+
+## System Element Traceability {#sec-system-element-traceability}
+
+This table maps tests to elements in the System Specification, ensuring that tests validate the components, interfaces, and algorithms defined in the system design.
+
+| Test ID | Test Type | Test Name | System Element ID | Element Type | Test Scope |
+|---------|-----------|-----------|------------------|-------------|------------|
+| TEST-BDD-06 | BDD | Component integration | CMP-01 | Component | Interface verification |
+| TEST-UNIT-05 | Unit | Dynamics model | CMP-02 | Component | Functional correctness |
+| TEST-UNIT-06 | Unit | Prior model | CMP-03 | Component | Functional correctness |
+| TEST-UNIT-07 | Unit | Likelihood model | CMP-04 | Component | Functional correctness |
+| TEST-UNIT-08 | Unit | RNA velocity algorithm | MM-01 | Mathematical Model | Correctness and stability |
+
+## Traceability Matrix {#sec-traceability-matrix}
+
+This comprehensive matrix provides a cross-reference between tests and all AMDiRE elements, helping to identify gaps in test coverage and ensure that all aspects of the system are adequately tested.
+
+| AMDiRE Element ID | Element Type | Element Description | Test IDs | Coverage Status |
+|-------------------|--------------|---------------------|----------|----------------|
+| RG-01 | Research Goal | Probabilistic RNA Velocity | TEST-BDD-01, TEST-UNIT-05 | Complete |
+| RG-02 | Research Goal | Uncertainty Quantification | TEST-BDD-02 | Partial |
+| FN-03 | Functional Requirement | Model Training | TEST-BDD-04 | Complete |
+| FN-05 | Functional Requirement | Velocity Computation | TEST-BDD-05 | Complete |
+| CMP-01 | Component | PyroVelocityModel | TEST-BDD-06 | Complete |
+| CMP-02 | Component | DynamicsModel | TEST-UNIT-05, TEST-BDD-01 | Complete |
+| MM-01 | Mathematical Model | RNA Velocity Dynamics | TEST-UNIT-08 | Complete |
+
+# Test Coverage Reference {#sec-test-coverage-reference}
+
+## Feature Specifications Reference {#sec-feature-specifications-reference}
+
+The table below tracks all feature specifications defined in our BDD feature files for the modular implementation. This serves as a living document of the intended functionality for the library.
+
+| Module | Feature | Status | Description | Related AMDiRE Elements |
+|--------|---------|--------|-------------|------------------------|
+| dynamics_model | Standard dynamics model computes expected counts | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes expected unspliced and spliced counts | MM-01, CMP-02, FN-05 |
+| dynamics_model | Standard dynamics model computes steady state | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes steady state values | MM-01, CMP-02 |
+| dynamics_model | Legacy dynamics model matches legacy implementation | ✅ Implemented | Verify that the LegacyDynamicsModel matches the legacy implementation | CMP-02, TC-04 |
+| dynamics_model | Dynamics model handles edge cases | ✅ Implemented | Verify that the dynamics model handles edge cases gracefully | CMP-02, REL-02 |
+| dynamics_model | Dynamics model with library size correction | ✅ Implemented | Verify that the dynamics model correctly applies library size correction | CMP-02, SC-01 |
+| prior_model | LogNormal prior model creates stochastic nodes | ✅ Implemented | Verify that the LogNormalPriorModel creates appropriate stochastic nodes | CMP-03, ALG-02 |
+| likelihood_model | Poisson likelihood model defines distributions | ✅ Implemented | Verify that the PoissonLikelihoodModel defines appropriate Poisson distributions | CMP-04, ALG-02 |
+| model | PyroVelocity model integrates components | ✅ Implemented | Verify that the PyroVelocityModel correctly integrates all components | CMP-01, SG-01 |
+| model | PyroVelocity model trains on AnnData | ✅ Implemented | Verify that the PyroVelocityModel can train on AnnData objects | CMP-01, FN-03, INT-01 |
+
+### Legend
+- ✅ Implemented: Feature is fully implemented and tested
+- ⚠️ In Progress: Implementation or testing is underway
+- 📝 Specified: Feature is specified but not yet implemented
+- 🔄 Planned: Feature is planned but not fully specified
+
+## Unit Test Coverage Reference {#sec-unit-test-coverage-reference}
+
+This table tracks the unit test coverage for each module in the modular implementation, helping identify areas that need additional testing.
+
+| Module | Test File | Test Coverage | Test Cases | Missing Coverage | Related AMDiRE Elements |
+|--------|-----------|---------------|------------|------------------|------------------------|
+| components/dynamics.py | test_dynamics.py | High | 10+ | Edge cases with β = γ | CMP-02, MM-01 |
+| components/guides.py | test_guides.py | High | 5+ | Custom initialization | CMP-06, ALG-02 |
+| components/likelihoods.py | test_likelihoods.py | High | 5+ | Zero count handling | CMP-04, ALG-02 |
+| components/observations.py | test_observations.py | High | 5+ | Missing value handling | CMP-05 |
+| components/priors.py | test_priors.py | High | 5+ | None | CMP-03, ALG-02 |
+| model.py | test_model.py | High | 10+ | None | CMP-01, FN-03, FN-05 |
+| factory.py | test_factory.py | High | 10+ | None | AD-05 |
+| registry.py | test_registry.py | High | 5+ | None | AD-04 |
+
+# BDD Feature Files Explained {#sec-bdd-feature-files-explained}
+
+Our feature files use Gherkin syntax to describe expected behavior. Each feature file follows this structure:
+
+```gherkin
+Feature: [Feature Name]
+  As a [type of user]
+  I want to [perform some action]
+  So that [achieve some benefit]
+
+  # AMDiRE Traceability
+  # Context: [Context Element IDs]
+  # Requirements: [Requirement IDs]
+  # System: [System Element IDs]
+
+  Background:
+    Given [common setup steps]
+
+  Scenario: [Specific use case name]
+    Given [preconditions]
+    When [actions]
+    Then [expected results]
+```
+
+### Example Feature Files
+
+PyroVelocity uses feature files to describe the expected behavior of each component. Here's an example from the dynamics model:
+
+```gherkin
+Feature: Dynamics Model
+  As a computational biologist
+  I want to model RNA velocity dynamics
+  So that I can understand transcriptional dynamics in single-cell data
+
+  # AMDiRE Traceability
+  # Context: RG-01 (Probabilistic RNA Velocity)
+  # Requirements: FN-05 (Velocity Computation)
+  # System: CMP-02 (DynamicsModel), MM-01 (RNA Velocity Dynamics)
+
+  Background:
+    Given I have a dynamics model component
+    And I have input data with unspliced and spliced counts
+
+  Scenario Outline: Standard dynamics model computes expected counts
+    Given I have a StandardDynamicsModel
+    When I run the forward method with alpha <alpha>, beta <beta>, and gamma <gamma>
+    Then the model should compute expected unspliced and spliced counts
+    And the expected counts should follow RNA velocity dynamics
+
+    Examples:
+      | alpha | beta | gamma |
+      | 1.0   | 0.5  | 0.2   |
+```
+
+### Deriving Tests from AMDiRE Artifacts
+
+When creating BDD feature files, we follow these guidelines to ensure alignment with AMDiRE artifacts:
+
+1. **From Context Specification**:
+   - Derive test scenarios from research goals (RG-xx)
+   - Ensure tests respect scientific constraints (SC-xx)
+   - Include validation of domain model elements
+
+2. **From Requirements Specification**:
+   - Create test scenarios for each functional requirement (FN-xx)
+   - Develop performance tests for performance requirements (PERF-xx)
+   - Include reliability tests for reliability requirements (REL-xx)
+
+3. **From System Specification**:
+   - Test each component's functionality (CMP-xx)
+   - Validate algorithm implementations (ALG-xx)
+   - Test interfaces between components
+
+# Step Definitions Structure {#sec-step-definitions-structure}
+
+Step definition files connect feature files to actual test code. They are organized by component in the `integration/` directory:
+
+```python
+# src/pyrovelocity/tests/models/modular/integration/test_dynamics_model_steps.py
+from importlib.resources import files
+import pytest
+import torch
+from pytest_bdd import scenarios, given, when, then, parsers
+
+# Link to the feature file using importlib.resources
+scenarios(str(files("pyrovelocity.tests.features") / "models" / "modular" / "dynamics_model.feature"))
+
+# Import the components
+from pyrovelocity.models.modular.components import (
+    LegacyDynamicsModel,
+    StandardDynamicsModel,
+)
+
+@given("I have a StandardDynamicsModel", target_fixture="standard_dynamics_model")
+def standard_dynamics_model_fixture(bdd_standard_dynamics_model):
+    """Get a StandardDynamicsModel from the fixture."""
+    return bdd_standard_dynamics_model
+
+@when(parsers.parse("I run the forward method with alpha {alpha}, beta {beta}, and gamma {gamma}"), target_fixture="run_forward_method_with_parameters")
+def run_forward_method_with_parameters_fixture(standard_dynamics_model, input_data, alpha, beta, gamma):
+    """Run the forward method with the given parameters."""
+    # Convert string parameters to float
+    alpha_val = float(alpha)
+    beta_val = float(beta)
+    gamma_val = float(gamma)
+
+    # Create context with input data and parameters
+    context = {
+        "u_obs": input_data["u_obs"],
+        "s_obs": input_data["s_obs"],
+        "alpha": torch.tensor([alpha_val] * input_data["n_genes"]),
+        "beta": torch.tensor([beta_val] * input_data["n_genes"]),
+        "gamma": torch.tensor([gamma_val] * input_data["n_genes"]),
+    }
+
+    # Run the forward method
+    result_context = standard_dynamics_model.forward(context)
+
+    # Store the result for later steps
+    return result_context
+```
+
+# Validation Framework {#sec-validation-framework}
+
+PyroVelocity includes a validation framework (PRD-678-VALID) to verify that the modular model can replicate the legacy model.
+This framework is a critical part of our testing strategy, ensuring that the refactored implementation maintains compatibility with existing workflows.
+
+## Key Validation Components
+
+- **Direct Comparison**: Comparing outputs between legacy and modular implementations
+- **Synthetic Data Validation**: Using synthetic data with known properties
+- **Real Data Validation**: Using downsampled real datasets (e.g., `tests/data/preprocessed_pancreas_50_7.json`)
+- **Parameter Validation**: Ensuring parameter inference matches between implementations
+- **Shape Validation**: Verifying tensor shapes are consistent between implementations
+
+## Implementation
+
+The validation framework is implemented in the `src/pyrovelocity/tests/validation/` directory and includes:
+
+- Direct comparison tests for model outputs
+- Metrics for quantifying differences between implementations
+- Visualization tools for analyzing differences
+- Utilities for creating patched versions of components to address compatibility issues
+
+## AMDiRE Traceability
+
+The validation framework addresses several key AMDiRE elements:
+
+- **TC-04**: Backward Compatibility constraint
+- **REQ-01**: Parameter Estimation Accuracy requirement
+- **SG-03**: Comprehensive Testing system goal
+- **REL-01**: Training Convergence reliability requirement
+
+# Contributing New Tests {#sec-contributing-new-tests}
+
+When contributing new tests to the project:
+
+1. **Review AMDiRE Artifacts**: Examine the Context, Requirements, and System specifications
+2. **Check Existing Coverage**: Review the traceability matrices and test reference tables
+3. **Identify Coverage Gaps**: Look for AMDiRE elements with incomplete test coverage
+4. **Update Feature Files**: If implementing a new feature, add or update the relevant feature file with AMDiRE traceability comments
+5. **Follow the Workflow**: Implement step definitions and unit tests as outlined above
+6. **Establish Traceability**: Ensure all tests are linked to specific AMDiRE elements
+7. **Run Tests Locally**: Ensure all tests pass before submitting a PR
+8. **Update Documentation**: Update the traceability matrices and reference tables in this document
+
+## Test Coverage Guidelines
+
+To ensure comprehensive test coverage across all AMDiRE elements:
+
+1. **Context Elements**:
+   - Every research goal (RG-xx) should have at least one associated test
+   - All scientific constraints (SC-xx) should be validated by appropriate tests
+   - Domain model elements should have tests verifying their implementation
+
+2. **Requirements Elements**:
+   - Each functional requirement (FN-xx) should have at least one BDD test
+   - Performance requirements (PERF-xx) should have benchmarking tests
+   - Reliability requirements (REL-xx) should have error handling and edge case tests
+
+3. **System Elements**:
+   - Every component (CMP-xx) should have unit tests for its internal functionality
+   - All algorithms (ALG-xx) should have tests for correctness and edge cases
+   - Interfaces (INT-xx) should have integration tests
+
+# Fixtures and Testing Utilities {#sec-fixtures-and-testing-utilities}
+
+PyroVelocity provides a comprehensive set of fixtures for testing, defined in various `conftest.py` files:
+
+## Main Fixtures (`src/pyrovelocity/tests/conftest.py`)
+
+| Fixture Name | Purpose | Scope |
+|--------------|---------|-------|
+| `adata_preprocessed_pancreas_50_7` | Provides preprocessed pancreas dataset | Function |
+| `adata_trained_pancreas_50_7` | Provides trained pancreas dataset | Function |
+| `default_sample_data` | Provides synthetic sample data | Function |
+| `temp_file_path` | Provides a temporary file path | Function |
+
+## Modular Implementation Fixtures (`src/pyrovelocity/tests/models/modular/conftest.py`)
+
+These fixtures are specific to the modular implementation and provide test instances of various components.
+
+## BDD-Specific Fixtures (`src/pyrovelocity/tests/models/modular/integration/conftest.py`)
+
+| Fixture Name | Purpose | Scope |
+|--------------|---------|-------|
+| `bdd_simple_data` | Provides simple data for BDD testing | Function |
+| `bdd_model_parameters` | Provides model parameters for BDD testing | Function |
+| `bdd_standard_dynamics_model` | Provides a StandardDynamicsModel | Function |
+| `bdd_legacy_dynamics_model` | Provides a LegacyDynamicsModel | Function |
+| `bdd_lognormal_prior_model` | Provides a LogNormalPriorModel | Function |
+| `bdd_pyro_velocity_model` | Provides a PyroVelocityModel | Function |
+| `clear_pyro_param_store` | Clears Pyro's parameter store | Function (autouse) |
+
+# Advanced Testing Topics {#sec-advanced-testing-topics}
+
+## Parameterized Testing
+
+Both unit tests and BDD scenarios can be parameterized:
+
+```python
+# Unit test parameterization
+@pytest.mark.parametrize("alpha,beta,gamma,expected_u_ss,expected_s_ss", [
+    (1.0, 0.5, 0.2, 2.0, 5.0),
+    (2.0, 1.0, 0.5, 2.0, 4.0),
+])
+def test_steady_state(alpha, beta, gamma, expected_u_ss, expected_s_ss):
+    model = StandardDynamicsModel()
+    u_ss, s_ss = model.steady_state(
+        torch.tensor([alpha]),
+        torch.tensor([beta]),
+        torch.tensor([gamma]),
+    )
+    assert torch.allclose(u_ss, torch.tensor([expected_u_ss]))
+    assert torch.allclose(s_ss, torch.tensor([expected_s_ss]))
+```
+
+```gherkin
+# BDD parameterization with Scenario Outline
+Scenario Outline: Standard dynamics model computes steady state
+  Given I have a StandardDynamicsModel
+  When I compute the steady state with alpha <alpha>, beta <beta>, and gamma <gamma>
+  Then the steady state unspliced should equal alpha/beta
+  And the steady state spliced should equal alpha/gamma
+
+  Examples:
+    | alpha | beta | gamma |
+    | 1.0   | 0.5  | 0.2   |
+    | 2.0   | 1.0  | 0.5   |
+```
+
+## Testing with Pyro
+
+PyroVelocity provides utilities for testing with Pyro:
+
+```python
+def test_pyro_model(clear_pyro_param_store):
+    """Test a Pyro model."""
+    # Define a simple model
+    def model():
+        x = pyro.sample("x", pyro.distributions.Normal(0, 1))
+        return x
+
+    # Define a guide
+    def guide():
+        loc = pyro.param("loc", torch.tensor(0.0))
+        scale = pyro.param("scale", torch.tensor(1.0), constraint=pyro.constraints.positive)
+        pyro.sample("x", pyro.distributions.Normal(loc, scale))
+
+    # Run inference
+    pyro.clear_param_store()
+    svi = pyro.infer.SVI(
+        model=model,
+        guide=guide,
+        optim=pyro.optim.Adam({"lr": 0.01}),
+        loss=pyro.infer.Trace_ELBO(),
+    )
+
+    # Train for a few steps
+    for _ in range(10):
+        loss = svi.step()
+
+    # Check that parameters were updated
+    assert pyro.param("loc").item() != 0.0
+```
+
+# Continuous Integration {#sec-continuous-integration}
+
+Tests are automatically run as part of the CI/CD pipeline defined in `.github/workflows/cid.yaml`. The CI pipeline includes:
+
+1. **Unit Tests**: Run all unit tests with coverage reporting
+2. **BDD Tests**: Run all BDD tests to verify behavior
+3. **Coverage Enforcement**: Ensure test coverage meets minimum thresholds
+
+# Glossary {#sec-glossary}
+
+## Testing Terminology
+
+| Term | Definition | Related AMDiRE Elements |
+|------|------------|------------------------|
+| Unit Test | A test that verifies the functionality of a single function, method, or class in isolation | System components (CMP-xx), algorithms (ALG-xx) |
+| Integration Test | A test that verifies the interaction between multiple components | System interfaces (INT-xx) |
+| BDD Test | A test written in Gherkin syntax that describes behavior from a user perspective | Functional requirements (FN-xx) |
+| Test Fixture | A fixed state of the system used as a baseline for running tests | Data objects (DO-xx) |
+| Test Coverage | The percentage of code that is executed during tests | Quality requirements (QR-xx) |
+| Protocol-First Testing | Testing against Protocol interfaces rather than concrete implementations | AD-01 (Protocol-First Architecture) |

### 2025-05-13 nbs/development/index.qmd
docs(development): update developoment docs overview
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
     toc: true
+    toc-depth: 3
+    number-sections: true
+    mermaid-format: js
+    mermaid:
+      theme: dark
+bibliography: ../references.bib
+csl: ../bibstyle.csl
 ---
 
-## Introduction
+# Introduction {#sec-introduction}
 
-This section provides technical documentation for developers working on PyroVelocity. It covers the architecture, implementation details, and testing framework of the project.
+This section provides comprehensive technical documentation for developers working on PyroVelocity.
+It follows the AMDiRE (Artefact-based Model for Domain-independent Requirements Engineering) framework to ensure a structured, traceable approach to development documentation [@Mendez-Fernandez2015-xj;@Chuprina2021-xx].
 
-## Purpose
+## Purpose of this Document {#sec-purpose}
 
-The development documentation is intended for:
+The development documentation serves as the authoritative reference for PyroVelocity's architecture, implementation details, and testing framework. It is designed to:
 
-- Contributors to the PyroVelocity codebase
-- Researchers extending the models or algorithms
-- Developers integrating PyroVelocity into other systems
+- Provide a clear understanding of the system's scientific context and requirements
+- Document the system architecture and implementation decisions
+- Establish traceability between requirements and implementation
+- Guide testing and validation efforts
+- Support the ongoing migration from legacy to modular implementations
 
-## Contents
+## Intended Audience {#sec-audience}
 
-The development documentation is organized into the following sections:
+This documentation is intended for:
 
-- [Architecture](architecture.qmd): Detailed information about the modular architecture of PyroVelocity, including component interfaces, implementation patterns, and design decisions.
-- [Testing](testing.qmd): Information about the testing framework, validation approach, and guidelines for writing tests.
+- **Contributors**: Developers working directly on the PyroVelocity codebase
+- **Researchers**: Scientists extending the models or algorithms
+- **Third-party developers**: Developers incorporating or integrating PyroVelocity into other systems
+- **Evaluators**: Prospective users verifying system functionality and correctness for their specific use cases
 
-## Development Workflow
+## Artifact-based requirements framework {#sec-amdire}
 
-PyroVelocity follows a protocol-first architecture with all contracts defined through Protocol interfaces.
-The project is organized into modular components that can be composed to create different model configurations.
+::: {.callout-note collapse="true" title="Relationships to international standards"}
+Note that the Requirements Specification is related to the [software requirements specification](https://en.wikipedia.org/wiki/Software_requirements_specification) associated to the [IEEE/ISO/IEC 29148-2018](https://standards.ieee.org/ieee/29148/6937/) [@Unknown2018-jr] and the System Specification is related to the [software design description](https://en.wikipedia.org/wiki/Software_design_description) associated to the [IEEE 1016-2009](https://standards.ieee.org/ieee/1016/4502/) [@Unknown2009-wv].
+:::
+
+PyroVelocity's development documentation follows the AMDiRE (Artefact-based Model for Domain-independent Requirements Engineering) framework [@Mendez-Fernandez2015-xj], which provides a structured approach to requirements engineering and system documentation.
+
+AMDiRE focuses on defining artifacts rather than activities, allowing for flexibility in development processes while maintaining clear documentation of what needs to be created.
+The framework consists of three primary specifications:
+
+1. **Context Specification**: Defines the scientific context, stakeholders, domain knowledge, and constraints
+2. **Requirements Specification**: Defines what the system should do from an external perspective
+3. **System Specification**: Defines how the system is implemented internally
+
+This artifact-based approach enables:
+
+- Clear traceability between scientific objectives and implementation
+- Flexibility in development processes
+- Consistent documentation structure
+- Evaluation of requirements implementation status and validation through traceable test case tracking and reference
+
+```{mermaid}
+graph TD
+    CS[Context Specification] --> RS[Requirements Specification]
+    RS --> SS[System Specification]
+    CS --> TF[Testing Framework]
+    RS --> TF
+    SS --> TF
+
+    class CS,RS,SS,TF document;
+```
+
+# Documentation Structure {#sec-structure}
+
+## Overview of Documents {#sec-documents}
+
+The PyroVelocity development documentation consists of four primary documents:
+
+1. [**Context Specification**](context.qmd): Defines the scientific context, stakeholders, research goals, and constraints
+2. [**Requirements Specification**](requirements.qmd): Defines the system's functional and quality requirements
+3. [**System Architecture and Design**](architecture.qmd): Defines the internal structure and implementation approach
+4. [**Testing Framework**](testing.qmd): Defines the testing approach and establishes traceability to requirements
+
+Each document builds upon the previous ones, creating a comprehensive view of the system from scientific context to implementation details.
+
+## Document Relationships {#sec-relationships}
+
+The documents in the development section are interconnected, with each subsequent document building upon the previous ones:
+
+[interactive chart](https://bit.ly/44CjSkL)
+
+```{mermaid}
+flowchart TD
+    subgraph "Scientific Context"
+        CS[Context Specification]
+        CS1[Research Goals]
+        CS2[Domain Model]
+        CS3[Constraints]
+        CS --> CS1
+        CS --> CS2
+        CS --> CS3
+    end
+
+    subgraph "System Requirements"
+        RS[Requirements Specification]
+        RS1[Functional Requirements]
+        RS2[Quality Requirements]
+        RS3[Data Model]
+        RS --> RS1
+        RS --> RS2
+        RS --> RS3
+    end
+
+    subgraph "System Design"
+        SS[System Architecture]
+        SS1[Component Model]
+        SS2[Computational Implementation]
+        SS3[Data Processing Pipeline]
+        SS --> SS1
+        SS --> SS2
+        SS --> SS3
+    end
+
+    subgraph "Validation"
+        TF[Testing Framework]
+        TF1[Test Coverage]
+        TF2[BDD Features]
+        TF3[Traceability]
+        TF --> TF1
+        TF --> TF2
+        TF --> TF3
+    end
+
+    CS --> RS
+    RS --> SS
+    CS --> TF
+    RS --> TF
+    SS --> TF
+
+    class CS,RS,SS,TF document;
+```
+
+# Document Summaries {#sec-summaries}
+
+## Context Specification {#sec-context-summary}
+
+The [Context Specification](context.qmd) establishes the foundation for PyroVelocity by defining:
+
+- **Research Problem**: The scientific challenges in RNA velocity analysis that PyroVelocity addresses
+- **Stakeholders**: The researchers, developers, and users involved in the project
+- **Scientific Objectives**: The research goals that drive the development
+- **Domain Model**: The scientific workflows, data objects, and algorithms
+- **Constraints**: The scientific, technical, and computational constraints
+
+This document serves as the primary reference for understanding the scientific context and objectives of PyroVelocity.
+
+## Requirements Specification {#sec-requirements-summary}
+
+The [Requirements Specification](requirements.qmd) defines what PyroVelocity should do from an external perspective:
+
+- **System Vision**: The overall scope and capabilities of PyroVelocity
+- **Usage Model**: How users interact with the system
+- **Computational Model**: The mathematical and algorithmic foundations
+- **Data Model**: The structure and relationships of data objects
+- **Functional Requirements**: The specific capabilities the system must provide
+- **Quality Requirements**: Performance, accuracy, reliability, and other quality attributes
+
+This document serves as the primary reference for understanding what PyroVelocity should do without specifying how it will be implemented.
+
+## System Architecture and Design {#sec-architecture-summary}
+
+The [System Architecture and Design](architecture.qmd) document defines how PyroVelocity is implemented internally:
+
+- **Architecture Overview**: The high-level organization and design patterns
+- **Computational Implementation**: How mathematical models and algorithms are implemented
+- **Component Model**: The modular components and their relationships
+- **Data Processing Pipeline**: How data flows through the system
+- **Execution Framework**: How computation is executed and managed
+- **Reproducibility Implementation**: How deterministic results are approximated amidst stochastic algorithms
+
+This document serves as the primary reference for understanding the internal structure and implementation of PyroVelocity.
+
+## Testing Framework {#sec-testing-summary}
+
+The [Testing Framework](testing.qmd) document defines the approach to testing and validation:
+
+- **Testing Philosophy**: The dual approach of unit testing and behavior-driven development
+- **AMDiRE Traceability**: How tests relate to context, requirements, and system elements
+- **Test Coverage**: The current state of test coverage across the system
+- **BDD Feature Files**: How behavior specifications are defined
+- **Step Definitions**: How behavior specifications are implemented
+- **Validation Framework**: How the system is validated against scientific requirements
+
+This document serves as the primary reference for understanding how PyroVelocity is tested and validated.
+
+# Implementation Migration Context {#sec-migration}
+
+PyroVelocity is currently in a transition phase, migrating from a legacy implementation to a modular architecture.
+The documentation reflects this transition and provides guidance for all implementations:
+
+1. **Legacy Implementation**: The original implementation in `src/pyrovelocity/models/_*.py`
+2. **Modular PyTorch/Pyro Implementation**: The current focus in `src/pyrovelocity/models/modular/`
+3. **JAX/NumPyro Implementation**: The future direction in `src/pyrovelocity/models/jax/`
+
+The development documentation primarily focuses on the modular PyTorch/Pyro implementation while acknowledging the legacy implementation and preparing for the JAX/NumPyro migration.
+
+```{mermaid}
+timeline
+    title PyroVelocity Implementation Migration
+    section Legacy
+        models/_velocity.py : Original implementation
+        models/_*.py : Supporting classes and utilities
+    section Modular
+        Protocol-First : Interface definitions
+        Component-Based : Interchangeable components
+        Validation : Against legacy implementation
+    section JAX/NumPyro
+        Functional : Pure functional approach
+        Performance : Improved computational efficiency
+        Feature Parity : Equivalent functionality
+```
+
+# Navigation Guide {#sec-navigation}
+
+## For New Contributors {#sec-for-contributors}
+
+If you're new to the PyroVelocity project:
+
+1. Start with the [Context Specification](context.qmd) to understand the scientific problem
+2. Review the [System Architecture](architecture.qmd) to understand the modular design
+3. Explore the [Testing Framework](testing.qmd) to understand how to validate your contributions
+
+## For Researchers {#sec-for-researchers}
+
+If you're a researcher looking to extend PyroVelocity:
+
+1. Start with the [Context Specification](context.qmd) to understand the scientific foundations
+2. Review the [Requirements Specification](requirements.qmd) to understand the computational models
+3. Explore the [System Architecture](architecture.qmd) to understand how to extend the system
+
+## For Integrators {#sec-for-integrators}
+
+If you're integrating PyroVelocity into another system:
+
+1. Start with the [Requirements Specification](requirements.qmd) to understand the system capabilities
+2. Review the [System Architecture](architecture.qmd) to understand the component interfaces
+3. Explore the [Testing Framework](testing.qmd) to understand how to validate your integration
+
+# Keeping Documentation Updated {#sec-updating}
+
+The development documentation should be updated whenever:
+
+1. New scientific objectives or constraints are identified
+2. New requirements are added or existing ones are modified
+3. The system architecture or implementation changes
+4. New tests are added or existing ones are modified
+
+When updating documentation:
+
+1. Maintain traceability between documents
+2. Update all affected documents to maintain consistency
+3. Ensure test coverage for new or modified elements
+4. Consider the updates in the context of the AMDiRE framework [@Mendez-Fernandez2015-xj]
 
 For more information on contributing to PyroVelocity, please see the [Contributing](../about/contributing.qmd) guide.
+
+# Glossary {#sec-glossary}
+
+<details>
+<summary>Development Documentation Terminology</summary>
+
+| Term | Definition | Example |
+|------|------------|---------|
+| Context Specification | Document defining the scientific context, stakeholders, and domain model | - |
+| Requirements Specification | Document defining functional and quality requirements | - |
+| System Specification | Document defining the system architecture and components | - |
+| Research Goal | A research goal that the system aims to achieve | RG-01: Probabilistic RNA Velocity |
+| Scientific Constraint | A limitation imposed by scientific principles | SC-01: RNA Velocity Model Assumptions |
+| Functional Requirement | A capability that the system must provide | FN-03: Model Training |
+| Component | A modular part of the system architecture | CMP-01: PyroVelocityModel |
+| Algorithm | A specific computational procedure | ALG-01: Standard RNA Velocity Dynamics |
+
+</details>

### 2025-05-13 nbs/development/architecture.qmd
docs(development): update parameter recovery validation requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
     Main -->|contains| Factory["Factory System"]
+    Main -->|uses| Validation["Validation Framework"]
 
@@
 
+    Validation -->|contains| ParameterRecovery["Parameter Recovery"]
+    Validation -->|contains| ModelComparison["Model Comparison"]
+    Validation -->|contains| ValidationMetrics["Validation Metrics"]
+
     Factory -->|uses| Registry
@@
     class Registry,Factory system;
+    class Validation,ParameterRecovery,ModelComparison,ValidationMetrics validation;
 ```
 
-### Core Components {#sec-core-components}
+### Core Components {#sec-overview-core-components}
 
@@
 8. **Factory System**: Creates components and models from configurations
+9. **Validation Framework**: Provides tools for model validation and comparison
+   - Parameter Recovery: Validates model's ability to recover known parameters from synthetic data
+   - Model Comparison: Compares different model implementations and configurations
+   - Validation Metrics: Computes metrics for quantifying validation performance
 
@@
 | InferenceGuide | create_guide | Creates guide functions for variational inference | FN-03 |
+| ParameterRecovery | generate_synthetic_data | Generates synthetic data with known parameters | FN-03 |
+| ParameterRecovery | validate_parameter_recovery | Validates model's ability to recover known parameters | FN-03 |
+| ModelComparison | compare_parameters | Compares parameter estimates between implementations | FN-04 |
+| ValidationMetrics | compute_parameter_metrics | Computes metrics for parameter recovery validation | FN-04 |
 
@@
 | AD-05 | Factory Pattern | Encapsulation of creation logic, configuration management | Direct instantiation | Simplified client code, centralized creation logic |
+| AD-06 | Parameter Recovery Validation | Validate model's ability to recover known parameters from synthetic data before applying to real data | Direct comparison to legacy implementation only | Ensures model is internally consistent and capable of recovering parameters, prerequisite for robust model comparison |
 
@@
 
+**Method ID:** CM-02
+
+**Method Name:** Parameter Recovery Validation
+
+**Related Function:** FN-03 (Parameter Recovery Validation)
+
+**Description:** Validates model's ability to recover known parameters from synthetic data
+
+**Method Details:**
+
+- Parameter generation: Samples parameter sets from prior distributions
+- Synthetic data generation: Creates synthetic data using known parameters
+- Inference: Performs inference on synthetic data to estimate parameters
+- Evaluation: Compares estimated parameters to original known values
+- Metrics computation: Calculates recovery accuracy, precision, and coverage metrics
+
+**Implementation Details:**
+
+- Libraries/frameworks used: PyTorch, Pyro, NumPy, SciPy
+- Computational complexity: O(num_parameter_sets * num_samples * max_sample_size * num_genes)
+- Memory requirements: O(num_parameter_sets * num_samples * max_sample_size * num_genes)
+- Parallelization strategy: Parallel processing across parameter sets and samples
+
 ## Computational Interfaces {#sec-computational-interfaces}
@@
 - Numerical issues: Special handling for edge cases (e.g., β = γ)
+
+**Interface ID:** CIF-02
+
+**Interface Name:** ParameterRecoveryValidator Protocol
+
+**Description:** Interface for parameter recovery validation
+
+**Implementation Approach:** Python Protocol with runtime checking
+
+**API Definition:**
+
+```python
+@runtime_checkable
+class ParameterRecoveryValidator(Protocol):
+    def generate_synthetic_data(
+        self,
+        model_config: Dict[str, Any],
+        prior_config: Dict[str, Any],
+        num_cells: int,
+        num_genes: int,
+        **kwargs
+    ) -> Dict[str, Any]:
+        """
+        Generate synthetic data with known parameters.
+        """
+        ...
+
+    def validate_parameter_recovery(
+        self,
+        model_config: Dict[str, Any],
+        synthetic_data: Dict[str, Any],
+        **kwargs
+    ) -> Dict[str, Any]:
+        """
+        Validate model's ability to recover known parameters.
+        """
+        ...
+```
+
+**Parameters:**
+
+| Parameter | Type | Description | Default Value | Constraints |
+|-----------|------|-------------|---------------|-------------|
+| model_config | Dict[str, Any] | Model configuration | None | Must specify model components |
+| prior_config | Dict[str, Any] | Prior distribution configuration | None | Must specify prior distributions for all parameters |
+| num_cells | int | Number of cells in synthetic data | None | > 0 |
+| num_genes | int | Number of genes in synthetic data | None | > 0 |
+| synthetic_data | Dict[str, Any] | Synthetic data with known parameters | None | Must contain u_obs, s_obs, true_parameters |
+
+**Return Values:**
+
+| Return Value | Type | Description | Constraints |
+|--------------|------|-------------|-------------|
+| synthetic_data | Dict[str, Any] | Synthetic data with known parameters | Must contain u_obs, s_obs, true_parameters |
+| validation_results | Dict[str, Any] | Validation metrics and results | Must contain recovery_metrics, coverage_metrics |
+
+**Error Handling:**
+
+- Missing parameters: Raises ValueError with descriptive message
+- Invalid parameters: Validates parameters and raises ValueError for invalid values
+- Inference failure: Provides diagnostic information for inference failures
 

### 2025-05-13 nbs/development/context.qmd
docs(development): update parameter recovery validation requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+### Parameter Recovery Validation Workflow
+
+- **Purpose**: Validate model's ability to recover known parameters from synthetic data
+- **Inputs**: Model specification and prior distributions
+- **Steps**:
+  1. Generate known parameter sets from prior distributions
+  2. Create synthetic data using these parameters and the model's generative process
+  3. Perform inference on synthetic data to estimate parameters
+  4. Compare estimated parameters to original known values
+  5. Analyze recovery accuracy, precision, and coverage
+- **Outputs**: Validation metrics and visualizations demonstrating parameter recovery performance
+
 ### Posterior Analysis Workflow

### 2025-05-13 nbs/development/requirements.qmd
docs(development): update parameter recovery validation requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 | CAP-05 | Modular Architecture | Flexible, component-based architecture with interchangeable parts | High |
-| CAP-06 | JAX/NumPyro Backend | High-performance implementation using JAX/NumPyro | Medium |
-| CAP-07 | Workflow Integration | Integration with workflow systems for reproducible analysis | Medium |
+| CAP-06 | Parameter Recovery Validation | Validation of model's ability to recover known parameters from synthetic data | High |
+| CAP-07 | Model Comparison | Comparison of different model implementations and configurations | High |
+| CAP-08 | JAX/NumPyro Backend | High-performance implementation using JAX/NumPyro | Medium |
+| CAP-09 | Workflow Integration | Integration with workflow systems for reproducible analysis | Medium |
 
@@
 - **Model Configuration**: Scientists configure model components and parameters
+- **Parameter Recovery Validation**: Scientists validate model's ability to recover known parameters
 - **Model Training**: Scientists train the model using variational inference
@@
 - **Result Visualization**: Scientists visualize velocity and uncertainty
+- **Model Comparison**: Scientists compare different model implementations and configurations
 - **Integration with Workflows**: Scientists integrate velocity analysis into larger pipelines
@@
     A[Data Preprocessing] --> B[Model Configuration]
-    B --> C[Model Training]
+    B --> V[Parameter Recovery Validation]
+    V --> C[Model Training]
     C --> D[Posterior Analysis]
@@
     E --> F[Result Visualization]
-    F --> G[Integration with Workflows]
-
-    class A,B,C,D,E,F,G scenario;
+    F --> M[Model Comparison]
+    M --> G[Integration with Workflows]
+
+    class A,B,V,C,D,E,F,M,G scenario;
 ```
@@
 | US-02 | Model Configuration | A-01, A-04 | Configure model components and parameters | Preprocessed data available | Configured model ready for training |
-| US-03 | Model Training | A-01, A-02 | Train the model using variational inference | Configured model and data available | Trained model with optimized parameters |
-| US-04 | Posterior Analysis | A-01, A-04 | Analyze posterior distributions of parameters | Trained model available | Parameter distributions and diagnostics |
-| US-05 | Velocity Computation | A-01, A-02 | Compute RNA velocity with uncertainty | Posterior samples available | Velocity estimates with uncertainty |
-| US-06 | Result Visualization | A-01, A-03 | Visualize velocity and uncertainty | Velocity estimates available | Visualizations for interpretation |
-| US-07 | Integration with Workflows | A-02 | Integrate velocity analysis into larger pipelines | Trained model and analysis code available | Integrated workflow |
+| US-03 | Parameter Recovery Validation | A-01, A-04 | Validate model's ability to recover known parameters from synthetic data | Model specification and prior distributions available | Validation metrics and visualizations demonstrating parameter recovery performance |
+| US-04 | Model Training | A-01, A-02 | Train the model using variational inference | Configured model and data available | Trained model with optimized parameters |
+| US-05 | Posterior Analysis | A-01, A-04 | Analyze posterior distributions of parameters | Trained model available | Parameter distributions and diagnostics |
+| US-06 | Velocity Computation | A-01, A-02 | Compute RNA velocity with uncertainty | Posterior samples available | Velocity estimates with uncertainty |
+| US-07 | Result Visualization | A-01, A-03 | Visualize velocity and uncertainty | Velocity estimates available | Visualizations for interpretation |
+| US-08 | Model Comparison | A-01, A-04 | Compare different model implementations and configurations | Multiple trained models available | Comparison metrics and visualizations |
+| US-09 | Integration with Workflows | A-02 | Integrate velocity analysis into larger pipelines | Trained model and analysis code available | Integrated workflow |
 
@@
 
-**Usage Scenario:** US-03 (Model Training)
+**Usage Scenario:** US-03 (Parameter Recovery Validation)
+
+**Description:** Validating a PyroVelocity model's ability to recover known parameters from synthetic data
+
+**Triggering Event:** User initiates parameter recovery validation
+
+**Steps:**
+
+1. User specifies model configuration and prior distributions
+2. System generates known parameter sets from prior distributions
+3. System creates synthetic data using these parameters and the model's generative process
+4. System performs inference on synthetic data to estimate parameters
+5. System compares estimated parameters to original known values
+6. System analyzes recovery accuracy, precision, and coverage
+7. System generates validation metrics and visualizations
+8. User evaluates parameter recovery performance
+
+**Alternative Paths:**
+
+* At step 1, if user wants to test specific parameter values, they provide custom parameter sets
+* At step 3, if user wants to test robustness, they specify noise levels to add to synthetic data
+
+**Exception Paths:**
+
+* At step 4, if inference fails to converge, system provides diagnostic information
+* At step 5, if parameters are not recoverable, system identifies problematic parameters
+
+**Scenario ID:** FS-02
+
+**Usage Scenario:** US-04 (Model Training)
 
@@
 
-**Workflow Name:** RNA Velocity Analysis Workflow
-
-**Description:** Complete workflow for RNA velocity analysis from raw data to interpretation
-
-**Participants:** A-01, A-02, A-03
+**Workflow Name:** RNA Velocity Analysis Workflow with Parameter Recovery Validation
+
+**Description:** Complete workflow for RNA velocity analysis from raw data to interpretation, including parameter recovery validation
+
+**Participants:** A-01, A-02, A-03, A-04
 
@@
 2. Model Configuration: US-02
-3. Model Training: US-03
-4. Posterior Analysis: US-04
-5. Velocity Computation: US-05
-6. Result Visualization: US-06
+3. Parameter Recovery Validation: US-03
+4. Model Training: US-04
+5. Posterior Analysis: US-05
+6. Velocity Computation: US-06
+7. Result Visualization: US-07
+8. Model Comparison: US-08 (optional)
 
@@
 * For integration with other analyses, users may export velocity to other tools
+* For comprehensive validation, users may test multiple parameter sets and noise levels
+* For model comparison, users may compare different model implementations and configurations
+
+**Workflow ID:** WF-02
+
+**Workflow Name:** Parameter Recovery Validation Workflow
+
+**Description:** Focused workflow for validating model's ability to recover known parameters from synthetic data
+
+**Participants:** A-01, A-04
+
+**Workflow Steps:**
+
+1. Model Configuration: US-02
+2. Parameter Recovery Validation: US-03
+3. Result Visualization: US-07
+4. Model Comparison: US-08 (if comparing multiple models)
+
+**Variations:**
+
+* For comprehensive validation, users may test multiple parameter sets and noise levels
+* For robustness testing, users may vary sample sizes and noise levels
+* For model development, users may iterate on model design based on validation results
 
@@
 
+**Algorithm Name:** Parameter Recovery Validation
+
+**Description:** Validation of model's ability to recover known parameters from synthetic data
+
+**Mathematical Formulation:**
+
+```
+Recovery Error = ||θ_true - θ_estimated||
+Coverage = P(θ_true ∈ CI(θ_estimated))
+```
+
+Where:
+- θ_true are the true parameter values
+- θ_estimated are the estimated parameter values
+- CI(θ_estimated) is the credible interval for the estimated parameters
+- ||·|| is a suitable distance metric (e.g., L2 norm)
+
+**Inputs:**
+
+- Model: Probabilistic model specification
+- Prior: Prior distributions for parameters
+- Num_parameter_sets: Number of parameter sets to test
+- Num_samples: Number of synthetic datasets per parameter set
+- Sample_sizes: List of sample sizes to test
+- Noise_levels: List of noise levels to test
+
+**Outputs:**
+
+- Recovery accuracy metrics (MSE, correlation, etc.)
+- Recovery precision metrics (width of credible intervals)
+- Coverage metrics (percentage of true parameters within credible intervals)
+- Bias metrics (systematic deviation from true values)
+- Visualizations of parameter recovery performance
+
+**Computational Properties:**
+
+- Time Complexity: O(num_parameter_sets * num_samples * max_sample_size * num_genes)
+- Space Complexity: O(num_parameter_sets * num_samples * max_sample_size * num_genes)
+- Parallelization: Embarrassingly parallel across parameter sets and samples
+
+**References:**
+
+- Parameter recovery validation specification
+- Principled Bayesian workflow literature
+
+**Algorithm ID:** ALG-03
+
 **Algorithm Name:** Stochastic Variational Inference
@@
 **Algorithm:** ALG-02
+
+**Description:** Interface for parameter recovery validation
+
+**Input Parameters:**
+
+| Parameter | Type | Description | Constraints |
+|-----------|------|-------------|------------|
+| model_config | Dict[str, Any] | Model configuration | Must specify model components |
+| prior_config | Dict[str, Any] | Prior distribution configuration | Must specify prior distributions for all parameters |
+| num_parameter_sets | int | Number of parameter sets to test | > 0 |
+| num_samples | int | Number of synthetic datasets per parameter set | > 0 |
+| sample_sizes | List[int] | List of sample sizes to test | All values > 0 |
+| noise_levels | List[float] | List of noise levels to test | All values >= 0 |
+| use_gpu | bool | Whether to use GPU acceleration | True/False |
+
+**Output Parameters:**
+
+| Parameter | Type | Description |
+|-----------|------|-------------|
+| validation_results | Dict[str, Any] | Validation metrics and results |
+| parameter_recovery_plots | Dict[str, Figure] | Visualizations of parameter recovery |
+| recovery_summary | Dict[str, float] | Summary statistics of recovery performance |
+
+**Error Conditions:**
+
+| Error Code | Description | Recovery Action |
+|------------|-------------|----------------|
+| ValueError | Invalid model or prior configuration | Correct configuration |
+| RuntimeError | Inference failure | Adjust inference parameters or model complexity |
+| MemoryError | Out of memory | Reduce number of parameter sets or sample sizes |
+
+**Interface ID:** CI-03
+
+**Algorithm:** ALG-03
 
@@
 | DO-07 | ComponentConfig | Configuration for model components | N/A | Yes |
+| DO-08 | SyntheticData | Synthetic data generated with known parameters | N/A | Yes |
+| DO-09 | ParameterRecoveryResults | Results of parameter recovery validation | N/A | Yes |
 
@@
 
+**Data Object:** DO-08 (SyntheticData)
+
+| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
+|--------------|----------------|-------------|------|-------|-------------|---------------|
+| ATTR-15 | u_obs | Observed unspliced counts | Tensor | Counts | Non-negative | None |
+| ATTR-16 | s_obs | Observed spliced counts | Tensor | Counts | Non-negative | None |
+| ATTR-17 | true_parameters | True parameter values used to generate data | Dict[str, Tensor] | Various | Non-negative | None |
+| ATTR-18 | sample_size | Number of cells in the synthetic dataset | int | Count | > 0 | None |
+| ATTR-19 | noise_level | Noise level used in data generation | float | Dimensionless | >= 0 | 0.0 |
+| ATTR-20 | seed | Random seed used for reproducibility | int | Dimensionless | Any | None |
+
+**Data Object:** DO-09 (ParameterRecoveryResults)
+
+| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
+|--------------|----------------|-------------|------|-------|-------------|---------------|
+| ATTR-21 | true_parameters | True parameter values | Dict[str, Tensor] | Various | Non-negative | None |
+| ATTR-22 | estimated_parameters | Estimated parameter values | Dict[str, Tensor] | Various | Non-negative | None |
+| ATTR-23 | credible_intervals | Credible intervals for estimated parameters | Dict[str, Tuple[Tensor, Tensor]] | Various | Lower < Upper | None |
+| ATTR-24 | recovery_metrics | Metrics quantifying parameter recovery performance | Dict[str, Dict[str, float]] | Various | None | None |
+| ATTR-25 | coverage_metrics | Metrics quantifying credible interval coverage | Dict[str, float] | Percentage | [0, 100] | None |
+| ATTR-26 | bias_metrics | Metrics quantifying systematic bias in estimates | Dict[str, float] | Various | None | None |
+| ATTR-27 | sample_size | Number of cells used in recovery | int | Count | > 0 | None |
+| ATTR-28 | noise_level | Noise level used in data generation | float | Dimensionless | >= 0 | 0.0 |
+
 ## Data Object Relationships {#sec-data-object-relationships}
@@
 | REL-07 | ModelState | ComponentConfig | Composition | 1:N | State contains component configurations |
+| REL-08 | ModelParameters | SyntheticData | Generation | 1:N | Parameters are used to generate synthetic data |
+| REL-09 | SyntheticData | ParameterRecoveryResults | Association | 1:1 | Synthetic data is used for parameter recovery validation |
+| REL-10 | ParameterRecoveryResults | ModelParameters | Validation | N:1 | Recovery results validate parameter inference |
 
@@
     ModelState *-- ComponentConfig : contains
+    ModelParameters --> SyntheticData : generates
+    SyntheticData --> ParameterRecoveryResults : validates
+    ParameterRecoveryResults --> ModelParameters : validates
 
@@
         +Dict params
+    }
+
+    class SyntheticData {
+        +Tensor u_obs
+        +Tensor s_obs
+        +Dict true_parameters
+        +int sample_size
+        +float noise_level
+    }
+
+    class ParameterRecoveryResults {
+        +Dict true_parameters
+        +Dict estimated_parameters
+        +Dict credible_intervals
+        +Dict recovery_metrics
+        +Dict coverage_metrics
     }

### 2025-05-13 nbs/development/testing.qmd
docs(development): update parameter recovery validation requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 | RG-02 | Research Goal | Uncertainty Quantification | TEST-BDD-02 | Partial |
+| RG-03 | Research Goal | Parameter Inference | TEST-UNIT-01, TEST-BDD-07 | Planned |
 | FN-03 | Functional Requirement | Model Training | TEST-BDD-04 | Complete |
 | FN-05 | Functional Requirement | Velocity Computation | TEST-BDD-05 | Complete |
+| CAP-06 | Capability | Parameter Recovery Validation | TEST-BDD-07, TEST-UNIT-09 | Planned |
+| CAP-07 | Capability | Model Comparison | TEST-BDD-08, TEST-UNIT-10 | Planned |
 | CMP-01 | Component | PyroVelocityModel | TEST-BDD-06 | Complete |
 | CMP-02 | Component | DynamicsModel | TEST-UNIT-05, TEST-BDD-01 | Complete |
+| SC-03 | Scientific Constraint | Parameter Identifiability | TEST-BDD-07 | Planned |
 | MM-01 | Mathematical Model | RNA Velocity Dynamics | TEST-UNIT-08 | Complete |
+| ALG-02 | Algorithm | Parameter Recovery Validation | TEST-UNIT-09 | Planned |
 
@@
 | model | PyroVelocity model trains on AnnData | ✅ Implemented | Verify that the PyroVelocityModel can train on AnnData objects | CMP-01, FN-03, INT-01 |
+| validation | Parameter recovery validation | 🔄 Planned | Verify that the model can recover known parameters from synthetic data | CAP-06, RG-03, SC-03, ALG-02 |
+| validation | Model comparison | 🔄 Planned | Verify that different model implementations can be compared | CAP-07, TC-04, SG-03 |
 
@@
 
-PyroVelocity includes a validation framework (PRD-678-VALID) to verify that the modular model can replicate the legacy model.
-This framework is a critical part of our testing strategy, ensuring that the refactored implementation maintains compatibility with existing workflows.
+PyroVelocity includes a validation framework (PRD-678-VALID) to verify that the modular model can replicate the legacy model and to validate the model's ability to recover known parameters from synthetic data.
+This framework is a critical part of our testing strategy, ensuring that the refactored implementation maintains compatibility with existing workflows and that the model is internally consistent.
 
@@
 - **Shape Validation**: Verifying tensor shapes are consistent between implementations
+- **Parameter Recovery Validation**: Validating model's ability to recover known parameters from synthetic data
 
@@
 - Utilities for creating patched versions of components to address compatibility issues
+- Parameter recovery validation tools for generating synthetic data and evaluating parameter recovery
+
+## Parameter Recovery Validation
+
+Parameter recovery validation is a fundamental technique in probabilistic modeling that assesses whether a model's inference procedure can correctly recover known parameter values from synthetic data generated by the model itself.
+It serves as a crucial sanity check before applying models to real-world data where true parameter values are unknown.
+
+The parameter recovery validation workflow follows these steps:
+
+1. **Parameter Generation**: Sample parameter sets from prior distributions
+2. **Synthetic Data Generation**: Create synthetic data using known parameters
+3. **Inference**: Perform inference on synthetic data to estimate parameters
+4. **Evaluation**: Compare estimated parameters to original known values
+5. **Metrics Computation**: Calculate recovery accuracy, precision, and coverage metrics
+
+The validation framework provides tools for each step of this workflow, including:
+
+- Functions for generating parameter sets from prior distributions
+- Functions for creating synthetic data with known parameters
+- Functions for performing inference on synthetic data
+- Functions for comparing estimated parameters to true values
+- Functions for calculating and visualizing recovery metrics
 
@@
 - **REL-01**: Training Convergence reliability requirement
+- **CAP-06**: Parameter Recovery Validation capability
+- **CAP-07**: Model Comparison capability
+- **RG-03**: Parameter Inference research goal
+- **SC-03**: Parameter Identifiability scientific constraint
 

### 2025-05-14 nbs/development/context.qmd
docs(development): refine context document
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
     code-fold: true
+bibliography: ../references.bib
+csl: ../bibstyle.csl
 ---
@@
 
-This Scientific Context Specification establishes the foundation for the PyroVelocity project by defining the scientific context in which the system operates.
+This document specifies the scientific context to establish the motivation and conceptual foundation for the [PyroVelocity](https://docs.pyrovelocity.net) project.
 It captures the research problem being addressed, stakeholders involved, scientific workflows affected, and domain knowledge essential for understanding the project scope.
-This document serves as the primary reference for domain scientists, developers, requirements engineers, and other stakeholders to ensure a shared understanding of the project context before detailed requirements are specified.
+It also serves as the primary reference for domain scientists, developers, users, evaluators, and other stakeholders to develop a relatively high-level understanding of the project context that precedes the detailed [requirements](./requirements.qmd).
 
@@
 - **Bold** is used for emphasis and key concepts
-- Blue text indicates a cross-reference to another section
-- Mathematical equations are expressed using LaTeX notation
-- Code snippets are shown in monospaced font
+- Linked text indicates a cross-reference to another section or external resource
+- Mathematical expressions are rendered using $\LaTeX$
+- Code blocks or examples are shown in monospaced font
 
-## References {#sec-references}
+## Selected references {#sec-references}
 
-1. La Manno, G., Soldatov, R., Zeisel, A. et al. RNA velocity of single cells. Nature 560, 494–498 (2018). https://doi.org/10.1038/s41586-018-0414-6
-2.
-Bergen, V., Lange, M., Peidli, S.
-et al.
-Generalizing RNA velocity to transient cell states through dynamical modeling.
-Nat Biotechnol 38, 1408–1414 (2020).
+- Model construction [@Good1983-id;@Cantwell2002-dv;@Wells2012-qf;@Schad2019-ps;@Tomczak2024-ws]
+- Systems biology [@Alon2019-ah]
+- RNA velocity [@Zeisel2011-nz;@La_Manno2018-lj;@Bergen2020-pj]
+- Probabilistic programming [@Bingham2018-id;@Van_de_Meent2021-ud]
+
+<details>
+<summary>
+Click to expand the reference list
+</summary>
+
+1. Good, I.J.: Good thinking: The foundations of probability and its applications. University of Minnesota Press (1983)
+2. Cantwell, B.J.: Introduction to symmetry analysis. Cambridge University Press (2002)
+3. Wells, J.D.: Effective theories in physics: From planetary orbits to elementary particle masses. Springer Berlin Heidelberg (2012)
+4. Schad, D.J., Betancourt, M., Vasishth, S.: Toward a principled bayesian workflow in cognitive science. (2019)
+5. Tomczak, J.M.: Deep generative modeling. Springer International Publishing, Cham, Switzerland (2024)
+6. Alon, U.: An introduction to systems biology: Design principles of biological circuits. CRC Press LLC (2019)
+Zeisel, A., Köstler, W.J., Molotski, N., Tsai, J.M., Krauthgamer, R., Jacob‐Hirsch, J., Rechavi, G., Soen, Y., Jung, S., Yarden, Y., Domany, E.: Coupled pre‐mRNA and mRNA dynamics unveil operational strategies underlying transcriptional responses to stimuli.
+Mol.
+Syst.
+Biol.
+7, 529 (2011).
+https://doi.org/10.1038/msb.2011.62
+8. La Manno, G., Soldatov, R., Zeisel, A., RNA velocity authors, Linnarsson, S., Kharchenko, P.V.: RNA velocity of single cells. Nature. (2018). https://doi.org/10.1038/s41586-018-0414-6
+9.
+Bergen, V., Lange, M., Peidli, S., Wolf, F.A., Theis, F.J.: Generalizing RNA velocity to transient cell states through dynamical modeling.
+Nat.
+Biotechnol.
+38, 1408–1414 (2020).
 https://doi.org/10.1038/s41587-020-0591-3
-3. Pyro: Deep Universal Probabilistic Programming. https://pyro.ai/
-4. AnnData: Annotated Data. https://anndata.readthedocs.io/
-5. scanpy: Single-Cell Analysis in Python. https://scanpy.readthedocs.io/
+10.
+Bingham, E., Chen, J.P., Jankowiak, M., Obermeyer, F., Pradhan, N., Karaletsos, T., Singh, R., Szerlip, P.A., Horsfall, P., Goodman, N.D.: Pyro: Deep universal probabilistic programming.
+J.
+Mach.
+Learn.
+Res.
+abs/1810.09538, (2018)
+11. Meent, J.-W. van de, Paige, B., Yang, H., Wood, F.: An introduction to probabilistic programming. arXiv [stat.ML]. (2021)
+
+</details>
 
@@
 
-This Scientific Context Specification is organized to progressively build understanding of the PyroVelocity project context.
-It begins with a definition of project scope, then identifies stakeholders and their concerns.
-Research goals are articulated to clarify the project's purpose, followed by a value proposition justifying the investment.
+This document is organized to progressively build understanding of the PyroVelocity project context.
+It begins with a definition of project scope after which it identifies stakeholders and their concerns.
+Research goals are articulated to clarify the project's purpose followed by a value proposition justifying the investment.
 The scientific domain model captures essential workflows and data objects, while the constraints section outlines limitations that must be considered throughout the project.
@@
 
-Single-cell RNA sequencing (scRNA-seq) provides a snapshot of gene expression in individual cells, but understanding cellular dynamics and developmental trajectories requires temporal information that is not directly captured in these static measurements.
-RNA velocity, which measures the rate of change of gene expression, offers a way to infer directional information from these snapshots by leveraging the distinction between unspliced pre-mRNAs and mature spliced mRNAs.
+Single-cell RNA sequencing, possibly combined with other measurement modalities, provides a snapshot of gene expression in individual cells, but understanding gene-regulatory networks, cellular population dynamics, and developmental trajectories requires temporal information on multiple scales that is not directly captured in these static measurements.
+The RNA velocity concept proposes to estimate the rate of change of gene expression in an attempt to infer some directional information within gene expression space from these snapshots by leveraging the time delay between unspliced pre-mRNAs and the production of mature spliced mRNA molecules.
 
@@
 
-1. **Deterministic Approaches**: Existing methods like scVelo and velocyto use deterministic models that cannot adequately capture the inherent stochasticity in gene expression.
+1.
+**Linear deterministic dynamics**: The most widely used methods rely exclusively on deterministic models with analytical solutions that cannot adequately capture the inherent non-linearity and stochasticity in gene expression.
 
-2. **Uncertainty Quantification**: These methods lack robust uncertainty quantification, making it difficult to assess the reliability of velocity estimates and downstream analyses.
+2. **Uncertainty quantification**: These methods lack robust uncertainty quantification, making it difficult to assess the reliability of velocity estimates and downstream analyses.
 
-3. **Parameter Estimation**: Current approaches often rely on simplistic parameter estimation techniques that may not fully leverage the information in the data.
+3. **Parameter estimation**: Current approaches often rely on simplistic parameter estimation techniques that may not fully leverage the information in the data.
 
-4. **Cell Fate Prediction**: Existing methods provide limited capabilities for predicting cell fate choices and understanding developmental trajectory dynamics.
+4.
+**Cell fate prediction**: Existing methods provide limited capabilities for predicting and validating cell fate choices and understanding the relationships among gene-regulatory network and developmental trajectory dynamics.
 
-PyroVelocity addresses these limitations by developing a Bayesian, generative, and multivariate approach to RNA velocity modeling that provides uncertainty quantification and more robust parameter estimation.
+PyroVelocity proposes to address these limitations by developing a probabilistic, generative, and multivariate approach to comparative RNA velocity modeling that supports uncertainty quantification, parameter estimation via multiple inference algorithms, and heuristic or principled model selection.
 
@@
 
-1. Models raw spliced and unspliced read counts directly, avoiding information loss from normalization
-2. Provides uncertainty quantification for velocity estimates and downstream analyses
-3. Synchronizes cell time estimation across genes for more coherent trajectory inference
-4. Enables multivariate modeling of gene expression and velocity prediction
-5. Offers diagnostic analyses and visualizations of model uncertainty
-6. Integrates with existing single-cell analysis ecosystems
+1. Models raw spliced and unspliced read counts directly, minimizing information loss from normalization
+2. Provides uncertainty quantification for parameter estimates and velocity estimates
+3. Synchronizes cell time estimation across genes for coherent trajectory inference
+4. Includes support for diagnostic analyses and visualizations related to uncertainty estimation
+5. Provides a principled approach to model validation, comparison, and selection
+6. Integrates existing ecosystems for single-cell analysis, dynamical simulation, probabilistic modeling, visualization, and type-safe reproducibile workflow orchestration
 
@@
 
-- Probabilistic modeling of RNA velocity using Pyro/PyTorch
-- Parameter inference from single-cell RNA-seq data
+- Probabilistic modeling of RNA velocity using Pyro/PyTorch or JAX/NumPyro
+- Dynamical simulation with torchode or diffrax for non-linear and stochastic likelihood functions
+- Parameter inference from single-cell RNA-seq data using various approaches to each of SVI and MCMC
 - Uncertainty quantification of model predictions
-- Integration with existing bioinformatics tools (scanpy, scVelo)
-- Visualization of velocity estimates and uncertainty
-- Modular implementation with interchangeable components
-- JAX/NumPyro implementation for improved performance
+- Integration with existing bioinformatics tools
+- Visualization of velocity, parameter, and associated uncertainty estimates
+- Modular implementation with interchangeable components to enable systematic comparison among member elements in a given lattice of prior and likelihood model architectures
 
@@
 | Single-Cell Genomics Core Facilities | Users | Process data for researchers | Medium |
-| Software Developers | Maintainers | Implement and maintain the codebase | High |
+| Software Developers | Maintainers | Design, implement and maintain the codebase | High |
 | Research Community | Beneficiaries | Use insights from the tool in research | Low |
@@
 | AnnData | Annotated data format | scverse | Data structure for input/output |
-| Pyro | Probabilistic programming framework | Uber AI/Linux Foundation | Core dependency for model implementation |
-| PyTorch | Deep learning framework | Meta AI | Core dependency for tensor operations |
-| JAX | Numerical computing library | Google | Alternative backend for NumPyro implementation |
+| Pyro | Probabilistic programming framework | Linux Foundation | Core dependency for model implementation |
+| PyTorch | Deep learning framework | Meta AI | Core dependency for tensor operations and automatic differentiation |
+| NumPyro | Probabilistic programming framework | Linux Foundation | Core dependency for model implementation |
+| JAX | Numerical computing library | Google | Core dependency for tensor operations and automatic differentiation |
+| diffrax | Differential equation solver | Patrick Kidger | Core dependency for dynamical simulation |
+| beartype | Runtime type checking | Beartype authors | Core dependency for type safety |
+| jaxtyping | Type annotations for arrays | Patrick Kidger | Core dependency for type safety |
+| Flyte | Workflow management platform | Linux Foundation | Core dependency for workflow orchestration |
+| hydra-zen | Configuration management | MIT Lincoln Laboratory | Core dependency for configuration management |
 
@@
 |-------------|------------------|---------------------|
-| Computational Biologists | Accuracy, interpretability, integration with existing workflows | Validated against ground truth, clear documentation, compatible with scanpy/scVelo |
-| Bioinformaticians | Extensibility, performance, reproducibility | Modular architecture, efficient computation, deterministic results |
-| Molecular Biologists | Ease of use, biological interpretability | Clear visualizations, connection to biological processes |
-| Software Developers | Code quality, maintainability, test coverage | Clean code, comprehensive tests, CI/CD pipeline |
+| Computational Biologists | Accuracy, interpretability, integration with existing workflows | Validated against ground truth, clear documentation, compatible with scanpy/scVelo, reproducible results |
+| Bioinformaticians | Extensibility, performance, reproducibility | Modular architecture, efficient computation, reproducible results |
+| Molecular Biologists | Ease of use, biological interpretability | Clear visualizations, integration with third-party data pertaining to biological processes |
+| Software Developers | Code quality, maintainability, specification, test coverage | Architecture follows specifications, comprehensive tests, CI/CD pipeline |
 
@@
 | RG-05 | Multimodal Integration | Support integration of multiple data modalities | Demonstrated improvement with multimodal data | Low |
+| RG-06 | Model Validation | Establish principled methods for model validation, comparison, and selection | Published validation framework | High |
 
@@
 | UG-05 | Scalability | RG-01 | Handle large datasets efficiently | Performance benchmarks | Medium |
+| UG-06 | Flexibility | RG-01, RG-02, RG-03, RG-04, RG-05, RG-06 | Support a variety of model architectures and inference algorithms | User feedback on utility | High |
 
@@
 
+[interactive chart](https://bit.ly/43ftGia)
+
 ```{mermaid}
@@
     RG05[RG-05: Multimodal Integration] --> UG03
+    RG06[RG-06: Model Validation] --> UG06[UG-06: Flexibility]
 
@@
     UG05 --> SG05[SG-05: JAX Implementation]
+    UG06 --> SG01
 
-    class RG01,RG02,RG03,RG05 research;
-    class UG01,UG02,UG03,UG04,UG05 usage;
+    class RG01,RG02,RG03,RG05,RG06 research;
+    class UG01,UG02,UG03,UG04,UG05,UG06 usage;
     class SG01,SG02,SG03,SG04,SG05 system;
@@
 1. For computational biologists:
-   - More accurate RNA velocity estimates with quantified uncertainty
+   - More accurate RNA velocity estimates with model comparison and quantified uncertainty
    - Better understanding of cell fate decisions in development and disease
-   - Improved ability to design and interpret single-cell experiments
+   - Improved ability to interpret single-cell experiments and contribute to design of follow-up experiments
 
@@
 
-1. The probabilistic model provides more accurate RNA velocity estimates than deterministic methods on benchmark datasets
-2. Uncertainty quantification correctly captures true parameter values within confidence intervals
-3. The modular implementation successfully replicates the functionality of the legacy implementation
-4. The JAX implementation provides significant performance improvements over the PyTorch implementation
+1. The probabilistic model provides the ability to compare models leading to the identification of new model architectures the provide more accurate RNA velocity estimates
+2. Uncertainty quantification correctly captures true parameter values with well-specified credibility intervals
+3.
+The modular implementation successfully replicates the functionality of the legacy implementation thereby enabling flexible construction of new model architectures for participation in model comparison
+4. The JAX implementation provides sufficient performance improvement to support multiple inference algorithms based on sampling in additiona to variational inference
 5. The software is successfully used by external research groups for RNA velocity analysis
@@
 - **Steps**:
-  1. Generate posterior samples for model parameters
+  1. Generate posterior samples for model parameters from trained model or via MCMC
   2. Compute RNA velocity from posterior samples
@@
 | CountMatrix | Matrix of RNA counts | Cells (rows), genes (columns), values (counts) | Dense/sparse array | Component of AnnData |
-| ModelParameters | Parameters of the RNA velocity model | alpha, beta, gamma, switching, t | PyTorch tensors | Used by dynamics model |
+| ModelParameters | Parameters of the RNA velocity model | Parameter vector, which, for the standard model, consists of alpha, beta, gamma, t_switching, t | PyTorch tensors/JAX arrays | Used by dynamics model |
 | PosteriorSamples | Samples from parameter posterior distributions | Parameter samples, chain information | Dictionary of tensors | Generated from trained model |
@@
 
-The core of PyroVelocity is a mathematical model of RNA dynamics based on a system of ordinary differential equations:
+The core of the PyroVelocity standard likelihood, which serves as a baseline reference for comparison to alternative models, is a mathematical model of RNA transcription and splicing dynamics based on a system of linear ordinary differential equations:
 
@@
 
-### Variational Inference Algorithm
+It is notable that this model lacks the ability to represent interactions among genes.
+In order to support more complicated likelihood functions we will need to consider numerical integration of dynamical systems embedded within the forward method of the likelihood model, which generally requires expression in terms of a framework that supports automatic differentiation in order to produce gradients for inference.
+
+### Inference Algorithms
+
+#### Variational Inference
 
@@
 |----|------------|-------------|--------|--------|
-| SC-01 | RNA Velocity Model Assumptions | The model assumes first-order kinetics for transcription, splicing, and degradation | RNA velocity theory | Limits ability to model complex regulatory dynamics |
+| SC-01 | RNA Velocity Model Assumptions | The standard model assumes first-order kinetics for transcription, splicing, and degradation | RNA velocity theory | Limits ability to model complex regulatory dynamics |
 | SC-02 | Steady State Assumption | The model assumes cells can reach steady state | RNA velocity theory | May not accurately model highly dynamic processes |
@@
 | TC-03 | Python Ecosystem | Must integrate with Python-based single-cell analysis tools | Ecosystem integration | Limits language and framework choices |
-| TC-04 | Backward Compatibility | Must maintain compatibility with legacy implementation during transition | Project requirement | Constrains design choices |
+| TC-04 | Backward Compatibility | Must demonstrate equivalence with legacy implementation prior to transition to modular implementation | Project requirement | Constrains design choices |
 
@@
 | CC-02 | Inference Speed | Variational inference can be computationally intensive | Algorithm complexity | Affects user experience and scalability |
-| CC-03 | Parallelization | Need for efficient parallelization across cells and genes | Performance requirements | Influences algorithm design |
+| CC-03 | Parallelization | Need for efficient parallelization across cells and genes where it will not degrade inference accuracy | Performance requirements | Influences algorithm design |
 | CC-04 | GPU Acceleration | Should leverage GPU acceleration when available | Performance requirements | Affects implementation choices |
@@
 |------|------------|--------|
-| RNA Velocity | The time derivative of gene expression, representing the rate of change of mRNA abundance | La Manno et al., 2018 |
+| RNA Velocity | Proposal to attempt to estimate the time derivative of gene expression, representing the rate of change of mRNA abundance | La Manno et al., 2018 |
 | Spliced mRNA | Mature messenger RNA with introns removed | Molecular biology |
@@
 | Degradation | Process of breaking down mRNA | Molecular biology |
-| Cell Fate | The developmental trajectory or end state of a cell | Developmental biology |
-| Trajectory Inference | Computational methods to order cells along developmental paths | Computational biology |
+| Cell Fate | The developmental trajectory or stable end state of a cellular lineage | Developmental biology |
+| Trajectory Inference | Computational methods to order cells along developmental paths in gene expression space | Computational biology |
 
@@
 | ELBO | Evidence Lower Bound, objective function in variational inference | Statistical learning theory |
+| Markov Chain Monte Carlo Inference | Method for sampling from posterior distributions | Statistical learning theory |
 | Protocol | Interface definition in Python typing | Python typing documentation |
-| Modular Architecture | Design approach based on interchangeable components | Software engineering |
+| Modular Architecture | Design approach based on interchangeable components using the model factory and registry pattern | Software engineering |
 | JAX | Numerical computing library with automatic differentiation | JAX documentation |

### 2025-05-14 nbs/development/requirements.qmd
docs(development): propagate model selection updates to requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2025-05-13 nbs/development/requirements.qmd
@@
 PyroVelocity is a probabilistic framework for RNA velocity analysis in single-cell genomics.
-It provides a Bayesian approach to modeling RNA dynamics, enabling robust uncertainty quantification and parameter inference from single-cell RNA sequencing data.
+It provides a Bayesian approach to modeling RNA dynamics, enabling robust uncertainty quantification, parameter inference, and model comparison from single-cell RNA sequencing data.
 The system implements a modular architecture with interchangeable components for dynamics models, prior distributions, likelihood models, observation models, and inference guides.
@@
 5. Offering diagnostic analyses and visualizations of model uncertainty
-6. Integrating with existing single-cell analysis ecosystems
-
-The system is currently transitioning from a legacy implementation to a modular PyTorch/Pyro implementation, with a future migration to a JAX/NumPyro implementation for improved performance.
+6. Supporting model validation, comparison, and selection
+7. Integrating with existing single-cell analysis ecosystems
+
+The system is currently transitioning from a legacy implementation to a modular PyTorch/Pyro implementation, with a future migration to a JAX/NumPyro implementation for improved performance and support for multiple inference algorithms.
 
@@
 - **Pyro/PyTorch**: Core probabilistic programming and tensor computation frameworks
-- **JAX/NumPyro**: Alternative backend for improved performance (future)
+- **JAX/NumPyro**: Alternative backend for improved performance and inference algorithms
+- **torchode/diffrax**: Differential equation solvers for dynamical simulation
+- **beartype/jaxtyping**: Runtime type checking and array shape annotations
+- **Flyte/hydra-zen**: Workflow orchestration and configuration management
 
@@
     Library -->|depends on| Pyro["Pyro"]
+    Library -->|depends on| JAX["JAX"]
+    Library -->|depends on| NumPyro["NumPyro"]
+    Library -->|depends on| Torchode["torchode"]
+    Library -->|depends on| Diffrax["diffrax"]
+    Library -->|depends on| Beartype["beartype"]
+    Library -->|depends on| Jaxtyping["jaxtyping"]
+    Library -->|depends on| Flyte["Flyte"]
+    Library -->|depends on| HydraZen["hydra-zen"]
     Library -->|processes| AnnData["AnnData Objects"]
     Library -->|produces| Velocity["RNA Velocity"]
+    Library -->|produces| ModelComparison["Model Comparison"]
     Library -->|integrates with| Scanpy["Scanpy"]
@@
     class Library library;
-    class PyTorch,Pyro,Scanpy,Scvelo external;
-    class AnnData,Velocity data;
+    class PyTorch,Pyro,JAX,NumPyro,Torchode,Diffrax,Beartype,Jaxtyping,Flyte,HydraZen,Scanpy,Scvelo external;
+    class AnnData,Velocity,ModelComparison data;
 ```
@@
 | CAP-01 | Probabilistic RNA Velocity | Bayesian modeling of RNA velocity with uncertainty quantification | High |
-| CAP-02 | Parameter Inference | Inference of biologically meaningful parameters from scRNA-seq data | High |
-| CAP-03 | Uncertainty Visualization | Visualization of uncertainty in velocity estimates and trajectories | High |
+| CAP-02 | Parameter Inference | Inference of biologically meaningful parameters from scRNA-seq data using multiple inference algorithms | High |
+| CAP-03 | Uncertainty Visualization | Visualization of uncertainty in velocity estimates, parameters, and trajectories | High |
 | CAP-04 | AnnData Integration | Seamless integration with AnnData objects and the scanpy ecosystem | High |
-| CAP-05 | Modular Architecture | Flexible, component-based architecture with interchangeable parts | High |
-| CAP-06 | JAX/NumPyro Backend | High-performance implementation using JAX/NumPyro | Medium |
-| CAP-07 | Workflow Integration | Integration with workflow systems for reproducible analysis | Medium |
+| CAP-05 | Modular Architecture | Flexible, component-based architecture with interchangeable parts for systematic model comparison | High |
+| CAP-06 | Parameter Recovery Validation | Validation of model's ability to recover known parameters from synthetic data to ensure empirically identifiable models | High |
+| CAP-07 | Model Comparison | Systematic comparison of different model architectures, implementations, and inference algorithms | High |
+| CAP-08 | JAX/NumPyro Backend | High-performance implementation using JAX/NumPyro with support for MCMC inference | Medium |
+| CAP-09 | Workflow Integration | Integration with workflow systems for reproducible analysis and model validation | Medium |
+| CAP-10 | Dynamical Simulation | Support for numerical integration of ODEs for non-linear and stochastic dynamics | Medium |
 
@@
 - **Model Configuration**: Scientists configure model components and parameters
-- **Model Training**: Scientists train the model using variational inference
+- **Parameter Recovery Validation**: Scientists validate model's ability to recover known parameters from synthetic data
+- **Model Training**: Scientists train the model using variational inference or MCMC
 - **Posterior Analysis**: Scientists analyze posterior distributions of parameters
 - **Velocity Computation**: Scientists compute RNA velocity with uncertainty
-- **Result Visualization**: Scientists visualize velocity and uncertainty
-- **Integration with Workflows**: Scientists integrate velocity analysis into larger pipelines
+- **Result Visualization**: Scientists visualize velocity, parameters, and associated uncertainty
+- **Model Comparison**: Scientists systematically compare different model architectures, implementations, and inference algorithms
+- **Model Selection**: Scientists select the most appropriate model based on validation metrics
+- **Integration with Workflows**: Scientists integrate velocity analysis and model validation into larger pipelines
 
@@
     E --> F[Result Visualization]
-    F --> G[Integration with Workflows]
-
-    class A,B,C,D,E,F,G scenario;
+    F --> M[Model Comparison]
+    M --> S[Model Selection]
+    S --> G[Integration with Workflows]
+
+    class A,B,V,C,D,E,F,M,S,G scenario;
 ```
@@
 | US-02 | Model Configuration | A-01, A-04 | Configure model components and parameters | Preprocessed data available | Configured model ready for training |
-| US-03 | Model Training | A-01, A-02 | Train the model using variational inference | Configured model and data available | Trained model with optimized parameters |
-| US-04 | Posterior Analysis | A-01, A-04 | Analyze posterior distributions of parameters | Trained model available | Parameter distributions and diagnostics |
-| US-05 | Velocity Computation | A-01, A-02 | Compute RNA velocity with uncertainty | Posterior samples available | Velocity estimates with uncertainty |
-| US-06 | Result Visualization | A-01, A-03 | Visualize velocity and uncertainty | Velocity estimates available | Visualizations for interpretation |
-| US-07 | Integration with Workflows | A-02 | Integrate velocity analysis into larger pipelines | Trained model and analysis code available | Integrated workflow |
+| US-03 | Parameter Recovery Validation | A-01, A-04 | Validate model's ability to recover known parameters from synthetic data | Model specification and prior distributions available | Validation metrics and visualizations demonstrating parameter recovery performance |
+| US-04 | Model Training | A-01, A-02 | Train the model using variational inference or MCMC | Configured model and data available | Trained model with optimized parameters |
+| US-05 | Posterior Analysis | A-01, A-04 | Analyze posterior distributions of parameters | Trained model available | Parameter distributions and diagnostics |
+| US-06 | Velocity Computation | A-01, A-02 | Compute RNA velocity with uncertainty | Posterior samples available | Velocity estimates with uncertainty |
+| US-07 | Result Visualization | A-01, A-03 | Visualize velocity, parameters, and associated uncertainty | Velocity estimates and parameter distributions available | Visualizations for interpretation |
+| US-08 | Model Comparison | A-01, A-04 | Systematically compare different model architectures, implementations, and inference algorithms | Multiple trained models available | Comparison metrics and visualizations |
+| US-09 | Model Selection | A-01, A-04 | Select the most appropriate model based on validation metrics | Model comparison results available | Selected model for downstream analysis |
+| US-10 | Integration with Workflows | A-02 | Integrate velocity analysis and model validation into larger pipelines | Trained model, selected model, and analysis code available | Integrated workflow |
 
@@
 
-**Usage Scenario:** US-03 (Model Training)
+**Usage Scenario:** US-03 (Parameter Recovery Validation)
+
+**Description:** Validating a PyroVelocity model's ability to recover known parameters from synthetic data to ensure empirically identifiable models
+
+**Triggering Event:** User initiates parameter recovery validation
+
+**Steps:**
+
+1. User specifies model configuration, prior distributions, and inference algorithm (SVI or MCMC)
+2. System generates known parameter sets from prior distributions
+3. System creates synthetic data using these parameters and the model's generative process
+4. System performs inference on synthetic data to estimate parameters
+5. System compares estimated parameters to original known values
+6. System analyzes recovery accuracy, precision, and coverage
+7. System calculates credible intervals and coverage metrics
+8. System generates validation metrics and visualizations
+9. User evaluates parameter recovery performance
+10. System stores validation results for model comparison
+
+**Alternative Paths:**
+
+* At step 1, if user wants to test specific parameter values, they provide custom parameter sets
+* At step 1, if user wants to compare inference algorithms, they specify multiple algorithms
+* At step 3, if user wants to test robustness, they specify noise levels to add to synthetic data
+* At step 4, if user wants to test multiple sample sizes, they specify a range of sample sizes
+
+**Exception Paths:**
+
+* At step 4, if inference fails to converge, system provides diagnostic information
+* At step 5, if parameters are not recoverable, system identifies problematic parameters
+* At step 7, if credible intervals have poor coverage, system suggests model refinement
+
+**Scenario ID:** FS-02
+
+**Usage Scenario:** US-04 (Model Training)
 
@@
 2. User creates a PyroVelocity model with standard components
-3. System initializes model parameters and guide
-4. User calls the train method with the AnnData object
-5. System performs stochastic variational inference
-6. System monitors convergence and reports progress
-7. System returns the trained model
-8. User saves the trained model for later use
+3. User selects inference algorithm (SVI or MCMC)
+4. System initializes model parameters and inference components
+5. User calls the train method with the AnnData object
+6. If using SVI:
+   a. System performs stochastic variational inference
+   b. System monitors ELBO convergence and reports progress
+7. If using MCMC:
+   a. System performs Markov Chain Monte Carlo sampling
+   b. System monitors chain diagnostics and reports progress
+8. System returns the trained model with posterior samples
+9. User saves the trained model for later use
 
@@
 * At step 2, if user wants custom components, they configure specific components
-* At step 4, if user wants custom training parameters, they provide additional arguments
+* At step 3, if user wants to compare inference algorithms, they specify multiple algorithms
+* At step 5, if user wants custom training parameters, they provide additional arguments
 
@@
 * At step 1, if data is not properly formatted, system provides error message
-* At step 5, if training fails to converge, system provides diagnostic information
+* At step 6a, if SVI fails to converge, system provides diagnostic information
+* At step 7a, if MCMC chains fail to mix, system suggests parameter adjustments
 
@@
 
-**Workflow Name:** RNA Velocity Analysis Workflow
-
-**Description:** Complete workflow for RNA velocity analysis from raw data to interpretation
-
-**Participants:** A-01, A-02, A-03
+**Workflow Name:** RNA Velocity Analysis Workflow with Model Validation and Selection
+
+**Description:** Complete workflow for RNA velocity analysis from raw data to interpretation, including parameter recovery validation, model comparison, and model selection
+
+**Participants:** A-01, A-02, A-03, A-04
 
@@
 2. Model Configuration: US-02
-3. Model Training: US-03
-4. Posterior Analysis: US-04
-5. Velocity Computation: US-05
-6. Result Visualization: US-06
+3. Parameter Recovery Validation: US-03
+4. Model Training: US-04
+5. Posterior Analysis: US-05
+6. Velocity Computation: US-06
+7. Result Visualization: US-07
+8. Model Comparison: US-08
+9. Model Selection: US-09
+10. Integration with Workflows: US-10 (optional)
 
@@
 * For large datasets, users may subsample genes or cells before model training
-* For complex biological systems, users may use nonlinear dynamics models
+* For complex biological systems, users may use nonlinear dynamics models with numerical integration
 * For integration with other analyses, users may export velocity to other tools
+* For comprehensive validation, users may test multiple parameter sets, sample sizes, and noise levels
+* For model comparison, users may compare different model architectures, implementations, and inference algorithms
+* For inference algorithm comparison, users may compare SVI and MCMC approaches
+
+**Workflow ID:** WF-02
+
+**Workflow Name:** Parameter Recovery Validation and Model Comparison Workflow
+
+**Description:** Focused workflow for validating model's ability to recover known parameters from synthetic data and comparing different model architectures and inference algorithms
+
+**Participants:** A-01, A-04
+
+**Workflow Steps:**
+
+1. Model Configuration: US-02
+2. Parameter Recovery Validation: US-03
+3. Result Visualization: US-07
+4. Model Comparison: US-08
+5. Model Selection: US-09
+
+**Variations:**
+
+* For comprehensive validation, users may test multiple parameter sets, sample sizes, and noise levels
+* For robustness testing, users may vary sample sizes, noise levels, and data generation processes
+* For model development, users may iterate on model design based on validation results
+* For inference algorithm comparison, users may compare SVI and MCMC approaches
+* For empirical identifiability assessment, users may analyze parameter recovery across different model architectures
 
@@
 - ELBO values
+- Posterior samples
 
@@
 - Convergence: Depends on model complexity and data characteristics
+- Parallelization: Embarrassingly parallel across batches
 
@@
 
-- Pyro documentation
+- Pyro documentation [@Bingham2018-id]
 - Hoffman et al. (2013)
+
+**Algorithm ID:** ALG-04
+
+**Algorithm Name:** Markov Chain Monte Carlo Inference
+
+**Description:** Sampling-based Bayesian inference for RNA velocity parameters
+
+**Mathematical Formulation:**
+
+```
+p(z|x) ∝ p(x|z)p(z)
+```
+
+Where:
+- p(z|x) is the posterior distribution of latent variables z given observed data x
+- p(x|z) is the likelihood of observed data given latent variables
+- p(z) is the prior distribution of latent variables
+
+**Inputs:**
+
+- Model: Probabilistic model p(x, z)
+- Data: Observed data x
+- Num_samples: Number of posterior samples
+- Num_chains: Number of parallel chains
+- Num_warmup: Number of warmup steps
+
+**Outputs:**
+
+- Posterior samples
+- Chain diagnostics (R-hat, effective sample size)
+- Posterior predictive checks
+
+**Computational Properties:**
+
+- Time Complexity: O(num_samples * num_chains * num_cells * num_genes)
+- Space Complexity: O(num_samples * num_chains * num_parameters)
+- Convergence: Assessed via chain diagnostics
+- Parallelization: Embarrassingly parallel across chains
+
+**References:**
+
+- NumPyro documentation [@Phan2019-numpyro]
+- Hoffman and Gelman (2014) [No U-Turn Sampler]
 
@@
 
+**Description:** Interface for parameter recovery validation
+
+**Input Parameters:**
+
+| Parameter | Type | Description | Constraints |
+|-----------|------|-------------|------------|
+| model_config | Dict[str, Any] | Model configuration | Must specify model components |
+| prior_config | Dict[str, Any] | Prior distribution configuration | Must specify prior distributions for all parameters |
+| inference_algorithm | str | Inference algorithm to use (SVI or MCMC) | "SVI" or "MCMC" |
+| num_parameter_sets | int | Number of parameter sets to test | > 0 |
+| num_samples | int | Number of synthetic datasets per parameter set | > 0 |
+| sample_sizes | List[int] | List of sample sizes to test | All values > 0 |
+| noise_levels | List[float] | List of noise levels to test | All values >= 0 |
+| use_gpu | bool | Whether to use GPU acceleration | True/False |
+
+**Output Parameters:**
+
+| Parameter | Type | Description |
+|-----------|------|-------------|
+| validation_results | Dict[str, Any] | Validation metrics and results |
+| parameter_recovery_plots | Dict[str, Figure] | Visualizations of parameter recovery |
+| recovery_summary | Dict[str, float] | Summary statistics of recovery performance |
+| credible_interval_coverage | Dict[str, float] | Coverage statistics for credible intervals |
+
+**Error Conditions:**
+
+| Error Code | Description | Recovery Action |
+|------------|-------------|----------------|
+| ValueError | Invalid model or prior configuration | Correct configuration |
+| RuntimeError | Inference failure | Adjust inference parameters or model complexity |
+| MemoryError | Out of memory | Reduce number of parameter sets or sample sizes |
+| ConvergenceError | Inference algorithm fails to converge | Adjust inference parameters or try alternative algorithm |
+
+**Interface ID:** CI-03
+
+**Algorithm:** ALG-03, ALG-04
+
 **Description:** Interface for model training
@@
 | adata | AnnData | Annotated data object | Must contain spliced/unspliced layers |
-| max_epochs | int | Maximum number of training epochs | > 0 |
-| batch_size | Optional[int] | Batch size for training | None or > 0 |
-| learning_rate | float | Learning rate for optimizer | > 0 |
+| inference_algorithm | str | Inference algorithm to use (SVI or MCMC) | "SVI" or "MCMC" |
+| max_epochs | int | Maximum number of training epochs (SVI) or samples (MCMC) | > 0 |
+| batch_size | Optional[int] | Batch size for training (SVI only) | None or > 0 |
+| learning_rate | float | Learning rate for optimizer (SVI only) | > 0 |
+| num_chains | int | Number of MCMC chains (MCMC only) | > 0 |
+| num_warmup | int | Number of warmup steps (MCMC only) | > 0 |
 | use_gpu | bool | Whether to use GPU acceleration | True/False |
@@
 | model | PyroVelocityModel | Trained model |
-| loss_history | List[float] | Training loss history |
+| loss_history | List[float] | Training loss history (SVI only) |
+| posterior_samples | Dict[str, Tensor] | Posterior samples |
+| diagnostics | Dict[str, Any] | Training or sampling diagnostics |
 
@@
 | ValueError | Invalid AnnData format | Preprocess data correctly |
-| RuntimeError | Training divergence | Adjust learning rate or model parameters |
-| MemoryError | Out of memory | Reduce batch size or use data subsampling |
+| RuntimeError | Training divergence or sampling failure | Adjust parameters or try alternative algorithm |
+| MemoryError | Out of memory | Reduce batch size, number of chains, or use data subsampling |
+| ConvergenceError | Inference algorithm fails to converge | Adjust inference parameters or try alternative algorithm |
 
@@
 - Scheduling: Batch-based processing for large datasets
+- Inference: Multiple inference algorithms (SVI and MCMC)
 
@@
 - Compute: Multi-core CPU or CUDA-compatible GPU
-- Memory: 16-64GB RAM depending on dataset size
-- Storage: 10-100GB for datasets and results
+- Memory: 16-64GB RAM depending on dataset size and inference algorithm
+- Storage: 10-100GB for datasets, results, and posterior samples
 - Network: Minimal (data loading only)
@@
 - JAX/NumPyro implementation offers improved performance over PyTorch/Pyro
+- MCMC requires more computational resources than SVI but provides more accurate uncertainty quantification
+- Parallel chains in MCMC can leverage multi-GPU systems
 
@@
 | ALG-02 | Memory Usage | Peak memory usage during training | < 16GB | Process monitoring |
-| ALG-02 | Convergence Rate | Number of epochs to reach convergence | < 1000 epochs | ELBO monitoring |
+| ALG-03 | Convergence Rate (SVI) | Number of epochs to reach convergence | < 1000 epochs | ELBO monitoring |
+| ALG-04 | Sampling Time (MCMC) | Time to generate 1000 posterior samples with 4 chains | < 2 hours on GPU | Wall clock time |
+| ALG-04 | Memory Usage (MCMC) | Peak memory usage during MCMC sampling | < 32GB | Process monitoring |
+| ALG-04 | Chain Convergence | R-hat statistic for all parameters | < 1.1 | MCMC diagnostics |
+| ALG-04 | Effective Sample Size | Minimum effective sample size across parameters | > 100 | MCMC diagnostics |
 
@@
 | DO-07 | ComponentConfig | Configuration for model components | N/A | Yes |
+| DO-08 | SyntheticData | Synthetic data generated with known parameters | N/A | Yes |
+| DO-09 | ParameterRecoveryResults | Results of parameter recovery validation | N/A | Yes |
+| DO-10 | ModelComparisonResults | Results of comparing different model architectures and inference algorithms | N/A | Yes |
+| DO-11 | MCMCDiagnostics | Diagnostics for MCMC sampling | N/A | Yes |
+| DO-12 | CredibleIntervals | Credible intervals for parameter estimates | N/A | Yes |
+| DO-13 | ValidationMetrics | Metrics for validating model performance | N/A | Yes |
 
@@
 
+**Data Object:** DO-08 (SyntheticData)
+
+| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
+|--------------|----------------|-------------|------|-------|-------------|---------------|
+| ATTR-15 | u_obs | Observed unspliced counts | Tensor | Counts | Non-negative | None |
+| ATTR-16 | s_obs | Observed spliced counts | Tensor | Counts | Non-negative | None |
+| ATTR-17 | true_parameters | True parameter values used to generate data | Dict[str, Tensor] | Various | Non-negative | None |
+| ATTR-18 | sample_size | Number of cells in the synthetic dataset | int | Count | > 0 | None |
+| ATTR-19 | noise_level | Noise level used in data generation | float | Dimensionless | >= 0 | 0.0 |
+| ATTR-20 | seed | Random seed used for reproducibility | int | Dimensionless | Any | None |
+
+**Data Object:** DO-09 (ParameterRecoveryResults)
+
+| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
+|--------------|----------------|-------------|------|-------|-------------|---------------|
+| ATTR-21 | true_parameters | True parameter values | Dict[str, Tensor] | Various | Non-negative | None |
+| ATTR-22 | estimated_parameters | Estimated parameter values | Dict[str, Tensor] | Various | Non-negative | None |
+| ATTR-23 | credible_intervals | Credible intervals for estimated parameters | Dict[str, Tuple[Tensor, Tensor]] | Various | Lower < Upper | None |
+| ATTR-24 | recovery_metrics | Metrics quantifying parameter recovery performance | Dict[str, Dict[str, float]] | Various | None | None |
+| ATTR-25 | coverage_metrics | Metrics quantifying credible interval coverage | Dict[str, float] | Percentage | [0, 100] | None |
+| ATTR-26 | bias_metrics | Metrics quantifying systematic bias in estimates | Dict[str, float] | Various | None | None |
+| ATTR-27 | sample_size | Number of cells used in recovery | int | Count | > 0 | None |
+| ATTR-28 | noise_level | Noise level used in data generation | float | Dimensionless | >= 0 | 0.0 |
+| ATTR-29 | inference_algorithm | Inference algorithm used | str | N/A | "SVI" or "MCMC" | "SVI" |
+| ATTR-30 | model_architecture | Model architecture used | str | N/A | Valid model name | "standard" |
+| ATTR-31 | empirical_identifiability | Assessment of parameter identifiability | Dict[str, bool] | N/A | True/False | None |
+
+**Data Object:** DO-10 (ModelComparisonResults)
+
+| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
+|--------------|----------------|-------------|------|-------|-------------|---------------|
+| ATTR-32 | model_configs | Configurations of compared models | List[Dict[str, Any]] | N/A | Non-empty | None |
+| ATTR-33 | inference_algorithms | Inference algorithms used | List[str] | N/A | "SVI" or "MCMC" | None |
+| ATTR-34 | performance_metrics | Performance metrics for each model | Dict[str, Dict[str, float]] | Various | None | None |
+| ATTR-35 | parameter_recovery_metrics | Parameter recovery metrics for each model | Dict[str, Dict[str, float]] | Various | None | None |
+| ATTR-36 | uncertainty_calibration | Uncertainty calibration metrics for each model | Dict[str, Dict[str, float]] | Various | None | None |
+| ATTR-37 | computational_metrics | Computational performance metrics | Dict[str, Dict[str, float]] | Various | None | None |
+| ATTR-38 | selected_model | Selected model based on comparison | str | N/A | Valid model name | None |
+| ATTR-39 | selection_criteria | Criteria used for model selection | Dict[str, float] | Various | None | None |
+
+**Data Object:** DO-11 (MCMCDiagnostics)
+
+| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
+|--------------|----------------|-------------|------|-------|-------------|---------------|
+| ATTR-40 | r_hat | R-hat convergence statistic | Dict[str, float] | Dimensionless | > 0 | None |
+| ATTR-41 | effective_sample_size | Effective sample size | Dict[str, float] | Count | > 0 | None |
+| ATTR-42 | chain_statistics | Statistics for each chain | Dict[str, Dict[str, Any]] | Various | None | None |
+| ATTR-43 | acceptance_rate | MCMC acceptance rate | float | Percentage | [0, 100] | None |
+| ATTR-44 | divergences | Number of divergent transitions | int | Count | >= 0 | None |
+| ATTR-45 | energy_statistics | Energy statistics for HMC | Dict[str, float] | Various | None | None |
+
 ## Data Object Relationships {#sec-data-object-relationships}
@@
 | REL-07 | ModelState | ComponentConfig | Composition | 1:N | State contains component configurations |
+| REL-08 | ModelParameters | SyntheticData | Generation | 1:N | Parameters are used to generate synthetic data |
+| REL-09 | SyntheticData | ParameterRecoveryResults | Association | 1:1 | Synthetic data is used for parameter recovery validation |
+| REL-10 | ParameterRecoveryResults | ModelParameters | Validation | N:1 | Recovery results validate parameter inference |
+| REL-11 | PosteriorSamples | MCMCDiagnostics | Association | 1:1 | MCMC diagnostics are generated from posterior samples |
+| REL-12 | PosteriorSamples | CredibleIntervals | Derivation | 1:1 | Credible intervals are computed from posterior samples |
+| REL-13 | ParameterRecoveryResults | ModelComparisonResults | Association | N:1 | Parameter recovery results are used for model comparison |
+| REL-14 | ModelComparisonResults | ValidationMetrics | Composition | 1:1 | Model comparison results contain validation metrics |
+| REL-15 | PyroVelocityModel | ModelComparisonResults | Association | N:1 | Multiple models are compared in model comparison |
 
@@
     ModelState *-- ComponentConfig : contains
+    ModelParameters --> SyntheticData : generates
+    SyntheticData --> ParameterRecoveryResults : validates
+    ParameterRecoveryResults --> ModelParameters : validates
+    PosteriorSamples --> MCMCDiagnostics : generates
+    PosteriorSamples --> CredibleIntervals : computes
+    ParameterRecoveryResults --> ModelComparisonResults : contributes to
+    ModelComparisonResults *-- ValidationMetrics : contains
+    PyroVelocityModel --> ModelComparisonResults : compared in
 
@@
         +ModelState state
+        +str inference_algorithm
     }
@@
         +int num_chains
+        +str inference_algorithm
     }
@@
     }
+
+    class SyntheticData {
+        +Tensor u_obs
+        +Tensor s_obs
+        +Dict true_parameters
+        +int sample_size
+        +float noise_level
+    }
+
+    class ParameterRecoveryResults {
+        +Dict true_parameters
+        +Dict estimated_parameters
+        +Dict credible_intervals
+        +Dict recovery_metrics
+        +Dict coverage_metrics
+        +str inference_algorithm
+        +str model_architecture
+        +Dict empirical_identifiability
+    }
+
+    class ModelComparisonResults {
+        +List model_configs
+        +List inference_algorithms
+        +Dict performance_metrics
+        +Dict parameter_recovery_metrics
+        +str selected_model
+        +Dict selection_criteria
+    }
+
+    class MCMCDiagnostics {
+        +Dict r_hat
+        +Dict effective_sample_size
+        +float acceptance_rate
+        +int divergences
+    }
+
+    class CredibleIntervals {
+        +Dict intervals
+        +Dict coverage
+    }
+
+    class ValidationMetrics {
+        +Dict metrics
+        +Dict criteria
+    }
 ```
@@
 | FN-02 | Model Configuration | Configure model components and parameters | US-02 | High |
-| FN-03 | Model Training | Train the model using variational inference | US-03 | High |
-| FN-04 | Posterior Sampling | Generate samples from posterior distributions | US-04 | High |
-| FN-05 | Velocity Computation | Compute RNA velocity with uncertainty | US-05 | High |
-| FN-06 | Result Visualization | Visualize velocity and uncertainty | US-06 | High |
-| FN-07 | Model Serialization | Save and load trained models | US-03, US-04 | Medium |
-| FN-08 | AnnData Integration | Store results in AnnData objects | US-01, US-05, US-06 | High |
-| FN-09 | Parameter Analysis | Analyze posterior distributions of parameters | US-04 | Medium |
-| FN-10 | Trajectory Inference | Infer cell trajectories from velocity | US-05, US-06 | Medium |
+| FN-03 | Parameter Recovery Validation | Validate model's ability to recover known parameters from synthetic data | US-03 | High |
+| FN-04 | Model Training | Train the model using variational inference or MCMC | US-04 | High |
+| FN-05 | Posterior Sampling | Generate samples from posterior distributions | US-05 | High |
+| FN-06 | Velocity Computation | Compute RNA velocity with uncertainty | US-06 | High |
+| FN-07 | Result Visualization | Visualize velocity, parameters, and associated uncertainty | US-07 | High |
+| FN-08 | Model Comparison | Systematically compare different model architectures, implementations, and inference algorithms | US-08 | High |
+| FN-09 | Model Selection | Select the most appropriate model based on validation metrics | US-09 | Medium |
+| FN-10 | Model Serialization | Save and load trained models | US-04, US-05 | Medium |
+| FN-11 | AnnData Integration | Store results in AnnData objects | US-01, US-06, US-07 | High |
+| FN-12 | Parameter Analysis | Analyze posterior distributions of parameters | US-05 | Medium |
+| FN-13 | Trajectory Inference | Infer cell trajectories from velocity | US-06, US-07 | Medium |
 
@@
     FN01[FN-01: Data Preprocessing] --> FN02[FN-02: Model Configuration]
-    FN02 --> FN03[FN-03: Model Training]
-    FN03 --> FN04[FN-04: Posterior Sampling]
-    FN04 --> FN05[FN-05: Velocity Computation]
-    FN05 --> FN06[FN-06: Result Visualization]
-    FN03 --> FN07[FN-07: Model Serialization]
-    FN01 --> FN08[FN-08: AnnData Integration]
-    FN05 --> FN08
-    FN04 --> FN09[FN-09: Parameter Analysis]
-    FN05 --> FN10[FN-10: Trajectory Inference]
-    FN10 --> FN06
-
-    class FN01,FN02,FN03,FN04,FN05,FN06,FN07,FN08,FN09,FN10 function;
+    FN02 --> FN03[FN-03: Parameter Recovery Validation]
+    FN03 --> FN04[FN-04: Model Training]
+    FN04 --> FN05[FN-05: Posterior Sampling]
+    FN05 --> FN06[FN-06: Velocity Computation]
+    FN06 --> FN07[FN-07: Result Visualization]
+    FN03 --> FN08[FN-08: Model Comparison]
+    FN04 --> FN08
+    FN08 --> FN09[FN-09: Model Selection]
+    FN04 --> FN10[FN-10: Model Serialization]
+    FN01 --> FN11[FN-11: AnnData Integration]
+    FN06 --> FN11
+    FN05 --> FN12[FN-12: Parameter Analysis]
+    FN06 --> FN13[FN-13: Trajectory Inference]
+    FN13 --> FN07
+    FN12 --> FN07
+    FN12 --> FN08
+
+    class FN01,FN02,FN03,FN04,FN05,FN06,FN07,FN08,FN09,FN10,FN11,FN12,FN13 function;
 ```
@@
 
-**Function:** FN-03 (Model Training)
+**Function:** FN-03 (Parameter Recovery Validation)
+
+**Inputs:**
+
+| Input Name | Type | Description | Source | Constraints |
+|------------|------|-------------|--------|------------|
+| model_config | ModelConfig | Model configuration | FN-02 | Valid component configurations |
+| prior_config | Dict | Prior distribution configuration | User | Valid prior distributions |
+| inference_algorithm | str | Inference algorithm to use | User | "SVI" or "MCMC" |
+| num_parameter_sets | int | Number of parameter sets to test | User | > 0 |
+| sample_sizes | List[int] | List of sample sizes to test | User | All values > 0 |
+| noise_levels | List[float] | List of noise levels to test | User | All values >= 0 |
+
+**Outputs:**
+
+| Output Name | Type | Description | Destination | Constraints |
+|-------------|------|-------------|------------|------------|
+| validation_results | Dict | Validation metrics and results | FN-08, FN-09 | None |
+| parameter_recovery_plots | Dict | Visualizations of parameter recovery | FN-07 | None |
+| recovery_summary | Dict | Summary statistics of recovery performance | FN-08, FN-09 | None |
+| credible_interval_coverage | Dict | Coverage statistics for credible intervals | FN-08, FN-09 | None |
+
+**Function:** FN-04 (Model Training)
 
@@
 | model_config | ModelConfig | Model configuration | FN-02 | Valid component configurations |
+| inference_algorithm | str | Inference algorithm to use | User | "SVI" or "MCMC" |
 | training_params | Dict | Training parameters | User | Valid parameter values |
@@
 |-------------|------|-------------|------------|------------|
-| trained_model | PyroVelocityModel | Trained model | FN-04, FN-05, FN-07 | None |
-| loss_history | List[float] | Training loss history | FN-06 | None |
-| training_diagnostics | Dict | Training diagnostics | FN-06 | None |
-
-**Function:** FN-05 (Velocity Computation)
+| trained_model | PyroVelocityModel | Trained model | FN-05, FN-06, FN-10 | None |
+| loss_history | List[float] | Training loss history (SVI only) | FN-07 | None |
+| diagnostics | Dict | Training or sampling diagnostics | FN-07, FN-08 | None |
+| posterior_samples | Dict | Posterior samples | FN-05, FN-12 | None |
+
+**Function:** FN-06 (Velocity Computation)
 
@@
 |------------|------|-------------|--------|------------|
-| trained_model | PyroVelocityModel | Trained model | FN-03, FN-07 | None |
-| posterior_samples | Dict | Posterior samples | FN-04 | None |
+| trained_model | PyroVelocityModel | Trained model | FN-04, FN-10 | None |
+| posterior_samples | Dict | Posterior samples | FN-05 | None |
 | adata | AnnData | Annotated data object | FN-01 | Must contain spliced/unspliced layers |
@@
 |-------------|------|-------------|------------|------------|
-| velocity | Matrix | Velocity vectors | FN-06, FN-08, FN-10 | None |
-| velocity_uncertainty | Matrix | Uncertainty estimates | FN-06, FN-08 | None |
-| velocity_metrics | Dict | Velocity quality metrics | FN-06, FN-08 | None |
+| velocity | Matrix | Velocity vectors | FN-07, FN-11, FN-13 | None |
+| velocity_uncertainty | Matrix | Uncertainty estimates | FN-07, FN-11 | None |
+| velocity_metrics | Dict | Velocity quality metrics | FN-07, FN-11 | None |
+
+**Function:** FN-08 (Model Comparison)
+
+**Inputs:**
+
+| Input Name | Type | Description | Source | Constraints |
+|------------|------|-------------|--------|------------|
+| model_configs | List[ModelConfig] | Configurations of compared models | FN-02 | Non-empty list |
+| validation_results | List[Dict] | Validation results for each model | FN-03 | None |
+| trained_models | List[PyroVelocityModel] | Trained models to compare | FN-04 | None |
+| parameter_analysis | List[Dict] | Parameter analysis results | FN-12 | None |
+| comparison_metrics | Dict | Metrics to use for comparison | User | Valid metric specifications |
+
+**Outputs:**
+
+| Output Name | Type | Description | Destination | Constraints |
+|-------------|------|-------------|------------|------------|
+| comparison_results | Dict | Comparison metrics and results | FN-09, FN-07 | None |
+| comparison_plots | Dict | Visualizations of model comparison | FN-07 | None |
+| model_rankings | Dict | Rankings of models by different criteria | FN-09 | None |
 
@@
 | FN-01 | FN-02 | PROVIDES_DATA_TO | Preprocessed data is used for model configuration |
-| FN-02 | FN-03 | PROVIDES_CONFIG_TO | Model configuration is used for training |
-| FN-03 | FN-04 | ENABLES | Trained model is used for posterior sampling |
-| FN-04 | FN-05 | PROVIDES_DATA_TO | Posterior samples are used for velocity computation |
-| FN-05 | FN-06 | PROVIDES_DATA_TO | Velocity estimates are used for visualization |
-| FN-03 | FN-07 | PROVIDES_DATA_TO | Trained model is serialized |
-| FN-07 | FN-03 | PROVIDES_DATA_TO | Serialized model is loaded for continued training |
-| FN-05 | FN-08 | PROVIDES_DATA_TO | Velocity estimates are stored in AnnData |
-| FN-04 | FN-09 | PROVIDES_DATA_TO | Posterior samples are used for parameter analysis |
-| FN-05 | FN-10 | PROVIDES_DATA_TO | Velocity estimates are used for trajectory inference |
+| FN-02 | FN-03 | PROVIDES_CONFIG_TO | Model configuration is used for parameter recovery validation |
+| FN-02 | FN-04 | PROVIDES_CONFIG_TO | Model configuration is used for model training |
+| FN-03 | FN-08 | PROVIDES_DATA_TO | Validation results are used for model comparison |
+| FN-03 | FN-09 | PROVIDES_DATA_TO | Validation results are used for model selection |
+| FN-04 | FN-05 | ENABLES | Trained model is used for posterior sampling |
+| FN-04 | FN-08 | PROVIDES_DATA_TO | Trained models are used for model comparison |
+| FN-04 | FN-10 | PROVIDES_DATA_TO | Trained model is serialized |
+| FN-05 | FN-06 | PROVIDES_DATA_TO | Posterior samples are used for velocity computation |
+| FN-05 | FN-12 | PROVIDES_DATA_TO | Posterior samples are used for parameter analysis |
+| FN-06 | FN-07 | PROVIDES_DATA_TO | Velocity estimates are used for visualization |
+| FN-06 | FN-11 | PROVIDES_DATA_TO | Velocity estimates are stored in AnnData |
+| FN-06 | FN-13 | PROVIDES_DATA_TO | Velocity estimates are used for trajectory inference |
+| FN-08 | FN-09 | PROVIDES_DATA_TO | Model comparison results are used for model selection |
+| FN-10 | FN-04 | PROVIDES_DATA_TO | Serialized model is loaded for continued training |
+| FN-12 | FN-07 | PROVIDES_DATA_TO | Parameter analysis results are used for visualization |
+| FN-12 | FN-08 | PROVIDES_DATA_TO | Parameter analysis results are used for model comparison |
+| FN-13 | FN-07 | PROVIDES_DATA_TO | Trajectory inference results are used for visualization |
 
@@
 | ACC-02 | Velocity Direction Accuracy | Accuracy of velocity direction on benchmark datasets | Comparison with known trajectories | > 80% agreement |
-| ACC-03 | Uncertainty Calibration | Calibration of uncertainty estimates | Coverage analysis | 95% confidence intervals contain true values in > 90% of cases |
+| ACC-03 | Uncertainty Calibration | Calibration of uncertainty estimates | Coverage analysis | 95% credible intervals contain true values in > 90% of cases |
 | ACC-04 | Numerical Stability | Stability of numerical computations | Stress testing with extreme values | No NaN or Inf values in results |
+| ACC-05 | Model Comparison Accuracy | Accuracy of model comparison metrics | Simulation studies | Correct model selection in > 90% of cases |
+| ACC-06 | Empirical Identifiability | Identifiability of model parameters | Parameter recovery validation | > 90% of parameters recoverable within specified bounds |
+| ACC-07 | Inference Algorithm Accuracy | Accuracy of different inference algorithms | Comparison with analytical solutions | MCMC more accurate than SVI for complex posteriors |
 
@@
 - Comparison with deterministic methods
+- Simulation studies with known model differences
+- Parameter recovery validation across multiple sample sizes
+- Comparison of SVI and MCMC inference algorithms
 
@@
 
-- Parameter estimation: 95% confidence intervals
+- Parameter estimation: 95% credible intervals
 - Velocity direction: Angular error < 30 degrees
 - Trajectory inference: Pseudotime correlation > 0.8
+- Model selection: Bayes factor > 3 for significant difference
+- Parameter recovery: Within 2 standard deviations of true value
 
@@
 |----------------|------------------|-------------|-------------------|-------------|
-| PERF-01 | Training Time | Time to train model on standard dataset | Wall clock time | < 30 minutes on GPU |
-| PERF-02 | Memory Usage | Peak memory usage during training | Process monitoring | < 16GB for 10k cells |
-| PERF-03 | Sampling Speed | Time to generate posterior samples | Wall clock time | < 5 minutes for 1000 samples |
-| PERF-04 | Velocity Computation Speed | Time to compute velocity from samples | Wall clock time | < 1 minute for 10k cells |
-| PERF-05 | Visualization Speed | Time to generate standard visualizations | Wall clock time | < 30 seconds per plot |
+| PERF-01 | SVI Training Time | Time to train model using SVI on standard dataset | Wall clock time | < 30 minutes on GPU |
+| PERF-02 | MCMC Sampling Time | Time to generate MCMC samples on standard dataset | Wall clock time | < 2 hours on GPU |
+| PERF-03 | Memory Usage (SVI) | Peak memory usage during SVI training | Process monitoring | < 16GB for 10k cells |
+| PERF-04 | Memory Usage (MCMC) | Peak memory usage during MCMC sampling | Process monitoring | < 32GB for 10k cells |
+| PERF-05 | Posterior Sampling Speed | Time to generate posterior samples from trained model | Wall clock time | < 5 minutes for 1000 samples |
+| PERF-06 | Velocity Computation Speed | Time to compute velocity from samples | Wall clock time | < 1 minute for 10k cells |
+| PERF-07 | Visualization Speed | Time to generate standard visualizations | Wall clock time | < 30 seconds per plot |
+| PERF-08 | Parameter Recovery Validation Speed | Time to complete parameter recovery validation | Wall clock time | < 1 hour per model configuration |
+| PERF-09 | Model Comparison Speed | Time to compare multiple model configurations | Wall clock time | < 10 minutes per comparison |
 
@@
 - Sublinear scaling with number of posterior samples
+- Linear scaling with number of MCMC chains
+- Linear scaling with number of model configurations in comparison
 
@@
 
-- CPU: < 8 cores at 100% utilization
-- GPU: < 16GB VRAM
+- CPU: < 16 cores at 100% utilization
+- GPU: < 32GB VRAM
 - Disk: < 100GB for intermediate results
+- Memory: < 64GB RAM for MCMC with multiple chains
 
@@
 |----------------|------------------|-------------|---------------------|-------------|
-| REL-01 | Training Convergence | Reliable convergence of training process | Success rate on benchmark datasets | > 95% success rate |
-| REL-02 | Numerical Robustness | Robustness to numerical issues | Testing with varied datasets | No failures due to numerical issues |
-| REL-03 | Error Handling | Proper handling of errors and edge cases | Exception testing | All errors properly caught and reported |
-| REL-04 | State Persistence | Reliable saving and loading of model state | Recovery testing | 100% successful recovery |
+| REL-01 | SVI Convergence | Reliable convergence of SVI training process | Success rate on benchmark datasets | > 95% success rate |
+| REL-02 | MCMC Convergence | Reliable convergence of MCMC chains | R-hat statistics | R-hat < 1.1 for all parameters |
+| REL-03 | Numerical Robustness | Robustness to numerical issues | Testing with varied datasets | No failures due to numerical issues |
+| REL-04 | Error Handling | Proper handling of errors and edge cases | Exception testing | All errors properly caught and reported |
+| REL-05 | State Persistence | Reliable saving and loading of model state | Recovery testing | 100% successful recovery |
+| REL-06 | Parameter Recovery Reliability | Reliable parameter recovery across different models | Success rate on synthetic data | > 90% success rate |
+| REL-07 | Model Comparison Reliability | Reliable model comparison results | Consistency testing | Consistent rankings across multiple runs |
 
@@
 - Validation of inputs before computation
+- Automatic adaptation of MCMC parameters
+- Fallback to alternative inference algorithms
+- Robust model comparison metrics
 
@@
 - Version tracking for models and data
+- Validation of MCMC diagnostics
+- Validation of parameter recovery results
+- Validation of model comparison metrics
 
@@
 | USE-05 | Integration Ease | Ease of integration with existing workflows | Integration testing | Compatible with scanpy/scVelo workflows |
+| USE-06 | Parameter Recovery API | Simplicity of parameter recovery validation API | User testing | < 10 lines of code for standard validation |
+| USE-07 | Model Comparison API | Simplicity of model comparison API | User testing | < 10 lines of code for standard comparison |
+| USE-08 | Inference Algorithm Selection | Ease of selecting and configuring inference algorithms | User testing | Single parameter to switch between SVI and MCMC |
+| USE-09 | Diagnostic Visualization | Quality of diagnostic visualizations for model validation | User feedback | > 80% positive feedback |
+| USE-10 | Model Selection Guidance | Clear guidance for model selection based on validation metrics | User feedback | > 80% of users can select appropriate model |
 
@@
 | REP-05 | Workflow Integration | Integration with workflow management systems | Workflow testing | Compatible with standard workflow systems |
+| REP-06 | MCMC Diagnostics | Recording of MCMC diagnostics for reproducibility | Diagnostic inspection | Complete diagnostics recorded |
+| REP-07 | Parameter Recovery Tracking | Tracking of parameter recovery validation results | Result inspection | Complete validation results recorded |
+| REP-08 | Model Comparison Tracking | Tracking of model comparison results | Result inspection | Complete comparison results recorded |
+| REP-09 | Inference Algorithm Tracking | Tracking of inference algorithm configurations | Configuration inspection | Complete algorithm configurations recorded |
+| REP-10 | Validation Dataset Versioning | Versioning of synthetic datasets used for validation | Dataset inspection | All validation datasets versioned |
 
@@
 - Comprehensive logging
+- MCMC chain initialization tracking
+- Parameter recovery validation result serialization
+- Model comparison result serialization
+- Inference algorithm configuration serialization
+- Synthetic dataset serialization
 
@@
 - Documentation of analysis decisions
+- Recording of parameter recovery validation configurations
+- Recording of model comparison configurations
+- Recording of inference algorithm configurations
+- Recording of synthetic dataset generation parameters
 
@@
 | SC-05 | Biological Variability | Biological systems exhibit inherent variability | Biology | Requires probabilistic modeling approach |
+| SC-06 | Model Comparison Limitations | Model comparison metrics may not always select the biologically most relevant model | Statistical theory | Requires domain knowledge for final model selection |
+| SC-07 | Inference Algorithm Tradeoffs | Different inference algorithms have different strengths and weaknesses | Statistical theory | Requires careful selection based on model complexity |
+| SC-08 | Synthetic Data Limitations | Synthetic data may not capture all aspects of real biological data | Simulation theory | Validation results may not fully translate to real data |
+| SC-09 | Empirical Identifiability | Empirical identifiability depends on sample size and noise level | Statistical theory | Requires validation across multiple conditions |
+| SC-10 | Model Misspecification | All models are approximations of the true biological process | Modeling theory | Multiple models may be needed for robust inference |
 
@@
 | TC-02 | Pyro/PyTorch Dependency | Relies on Pyro and PyTorch for probabilistic programming | Implementation choice | Affects performance and deployment options |
-| TC-03 | Python Ecosystem | Must integrate with Python-based single-cell analysis tools | Ecosystem integration | Limits language and framework choices |
-| TC-04 | Backward Compatibility | Must maintain compatibility with legacy implementation during transition | Project requirement | Constrains design choices |
-| TC-05 | Protocol-First Architecture | Uses Protocol interfaces for component contracts | Architecture decision | Influences implementation patterns |
+| TC-03 | JAX/NumPyro Dependency | Relies on JAX and NumPyro for alternative implementation | Implementation choice | Affects performance and deployment options |
+| TC-04 | Python Ecosystem | Must integrate with Python-based single-cell analysis tools | Ecosystem integration | Limits language and framework choices |
+| TC-05 | Backward Compatibility | Must maintain compatibility with legacy implementation during transition | Project requirement | Constrains design choices |
+| TC-06 | Protocol-First Architecture | Uses Protocol interfaces for component contracts | Architecture decision | Influences implementation patterns |
+| TC-07 | MCMC Implementation | MCMC requires specialized implementation considerations | Algorithm requirement | Affects performance and memory usage |
+| TC-08 | Model Comparison Framework | Requires framework for systematic model comparison | Architecture decision | Adds complexity to system design |
+| TC-09 | Parameter Recovery Validation | Requires framework for parameter recovery validation | Architecture decision | Adds complexity to system design |
+| TC-10 | Multiple Inference Algorithms | Must support both SVI and MCMC inference | Architecture decision | Requires flexible model design |
 
@@
 |---------------|-----------------|-------------|--------|--------|
-| RC-01 | Memory Usage | Large single-cell datasets can exceed available memory | Hardware limitations | Requires efficient memory management |
-| RC-02 | Inference Speed | Variational inference can be computationally intensive | Algorithm complexity | Affects user experience and scalability |
-| RC-03 | Parallelization | Need for efficient parallelization across cells and genes | Performance requirements | Influences algorithm design |
-| RC-04 | GPU Acceleration | Should leverage GPU acceleration when available | Performance requirements | Affects implementation choices |
-| RC-05 | Deployment Environment | Must run in standard computational biology environments | User environment | Limits dependency choices |
+| RC-01 | Memory Usage (SVI) | Large single-cell datasets can exceed available memory with SVI | Hardware limitations | Requires efficient memory management |
+| RC-02 | Memory Usage (MCMC) | MCMC requires more memory than SVI for storing chains | Hardware limitations | Limits dataset size for MCMC inference |
+| RC-03 | SVI Inference Speed | Variational inference can be computationally intensive | Algorithm complexity | Affects user experience and scalability |
+| RC-04 | MCMC Inference Speed | MCMC is typically slower than SVI | Algorithm complexity | Limits applicability to very large datasets |
+| RC-05 | Parallelization | Need for efficient parallelization across cells, genes, and chains | Performance requirements | Influences algorithm design |
+| RC-06 | GPU Acceleration | Should leverage GPU acceleration when available | Performance requirements | Affects implementation choices |
+| RC-07 | Deployment Environment | Must run in standard computational biology environments | User environment | Limits dependency choices |
+| RC-08 | Parameter Recovery Computation | Parameter recovery validation requires significant computation | Validation requirements | May require high-performance computing resources |
+| RC-09 | Model Comparison Computation | Model comparison across multiple configurations is computationally intensive | Validation requirements | May require high-performance computing resources |
+| RC-10 | Storage Requirements | Storing multiple models, validation results, and comparison metrics requires significant storage | Hardware limitations | Requires efficient storage management |
 
@@
 
-- CPU: Multi-core x86_64 processor
-- Memory: 16GB+ RAM (32GB+ recommended)
-- Storage: 10GB+ free space
-- GPU: CUDA-compatible GPU (optional but recommended)
+- CPU: Multi-core x86_64 processor (8+ cores recommended)
+- Memory: 16GB+ RAM for SVI (32GB+ recommended)
+- Memory: 32GB+ RAM for MCMC (64GB+ recommended)
+- Storage: 10GB+ free space for installation
+- Storage: 100GB+ free space for model storage and validation results
+- GPU: CUDA-compatible GPU with 16GB+ VRAM (optional but recommended)
+- GPU: Multiple GPUs for parallel MCMC chains (optional)
 
@@
 - Pyro 1.8+
+- JAX 0.4.0+
+- NumPyro 0.11.0+
+- torchode 0.1.0+
+- diffrax 0.3.0+
+- beartype 0.14.0+
+- jaxtyping 0.2.20+
 - AnnData 0.8+
@@
 | INT-05 | Visualization Integration | Integration with standard visualization tools | Medium |
+| INT-06 | Parameter Recovery Integration | Integration with parameter recovery validation framework | High |
+| INT-07 | Model Comparison Integration | Integration with model comparison framework | High |
+| INT-08 | MCMC Diagnostics Integration | Integration with MCMC diagnostic tools | Medium |
+| INT-09 | JAX/NumPyro Integration | Integration with JAX/NumPyro ecosystem | Medium |
+| INT-10 | Synthetic Data Integration | Integration with synthetic data generation tools | Medium |
 
@@
 - scVelo for comparison and validation
-- Workflow systems (Snakemake, Nextflow) for pipeline integration
+- Workflow systems (Flyte, hydra-zen) for pipeline integration and configuration
 - Visualization tools (Matplotlib, Plotly) for result presentation
+- Parameter recovery validation framework for model validation
+- Model comparison framework for model selection
+- MCMC diagnostic tools (ArviZ) for MCMC diagnostics
+- JAX/NumPyro ecosystem for alternative implementation
+- Synthetic data generation tools for validation
 
@@
 - PDF/PNG for visualizations
+- NetCDF for MCMC samples and diagnostics
+- Zarr for large dataset storage
+- Parquet for validation results
 
@@
 | TEST-05 | Compatibility Testing | Testing of ecosystem compatibility | Medium |
+| TEST-06 | Parameter Recovery Testing | Testing of parameter recovery capabilities | High |
+| TEST-07 | Model Comparison Testing | Testing of model comparison framework | High |
+| TEST-08 | MCMC Convergence Testing | Testing of MCMC convergence diagnostics | Medium |
+| TEST-09 | Inference Algorithm Testing | Testing of different inference algorithms | Medium |
+| TEST-10 | Empirical Identifiability Testing | Testing of parameter identifiability | High |
 
@@
 - Validation Testing: Comparison with ground truth and legacy implementation
+- Parameter Recovery Testing: Synthetic data with known parameters
+- Model Comparison Testing: Comparison of different model architectures
+- MCMC Convergence Testing: R-hat statistics and effective sample size
+- Inference Algorithm Testing: Comparison of SVI and MCMC results
+- Empirical Identifiability Testing: Parameter recovery across sample sizes and noise levels
 
@@
 - Automated test environment
-- Synthetic test data
+- Synthetic test data with known ground truth
 - Benchmark datasets
@@
 - Continuous integration system
+- Parameter recovery validation framework
+- Model comparison framework
+- MCMC diagnostic tools
+- Multiple inference algorithm implementations
+- Synthetic data generation tools
 
@@
 | TR-05 | Dependency conflicts in deployment environments | DE-01, DE-02, DE-03 | Low | Medium | Low |
+| TR-06 | MCMC convergence issues with complex models | REL-02, ACC-07, PERF-02 | Medium | High | High |
+| TR-07 | Parameter recovery failure for certain model configurations | ACC-06, TEST-06, SC-09 | Medium | High | High |
+| TR-08 | Model comparison metrics not aligning with biological relevance | SC-06, ACC-05, TEST-07 | Medium | Medium | Medium |
+| TR-09 | Resource constraints for comprehensive validation | RC-08, RC-09, PERF-08 | Medium | Medium | Medium |
+| TR-10 | Empirical identifiability issues with real datasets | SC-03, SC-09, ACC-06 | High | High | High |
 
@@
 | SR-05 | Challenges in validating results without ground truth | ACC-02, TEST-03 | High | Medium | High |
+| SR-06 | Synthetic data not capturing real data complexity | SC-08, TEST-06, ACC-06 | Medium | High | High |
+| SR-07 | Model comparison selecting mathematically optimal but biologically irrelevant models | SC-06, ACC-05, CAP-07 | Medium | High | High |
+| SR-08 | Different inference algorithms leading to different biological interpretations | SC-07, ACC-07, CAP-02 | Medium | Medium | Medium |
+| SR-09 | Empirical identifiability varying across biological contexts | SC-09, ACC-06, TEST-10 | High | Medium | High |
+| SR-10 | Model misspecification leading to incorrect biological conclusions | SC-10, ACC-05, CAP-07 | Medium | High | High |
 
@@
 | TR-05 | Use containerization and environment specification | DevOps Team | Deployment testing |
+| TR-06 | Implement robust MCMC diagnostics and adaptive sampling | Development Team | MCMC convergence metrics |
+| TR-07 | Validate parameter recovery across multiple model configurations | Research Team | Parameter recovery metrics |
+| TR-08 | Develop biologically relevant model comparison metrics | Research Team | Validation against benchmark datasets |
+| TR-09 | Implement efficient validation workflows and distributed computing | DevOps Team | Resource monitoring |
+| TR-10 | Develop empirical identifiability assessment framework | Research Team | Identifiability metrics |
 | SR-01 | Develop multiple model variants for different biological contexts | Research Team | Model comparison on diverse datasets |
@@
 | SR-05 | Use synthetic data and cross-validation for validation | Validation Team | Validation metrics |
+| SR-06 | Develop realistic synthetic data generation with biological variability | Research Team | Comparison with real data distributions |
+| SR-07 | Incorporate domain knowledge into model selection criteria | Research Team | Expert validation of model selection |
+| SR-08 | Compare inference algorithms on benchmark datasets | Research Team | Consistency of biological interpretations |
+| SR-09 | Validate identifiability across multiple biological contexts | Research Team | Context-specific identifiability metrics |
+| SR-10 | Implement model misspecification tests | Research Team | Posterior predictive checks |
 
@@
 | Pseudotime | Computational ordering of cells based on their progression through a biological process | Computational biology |
+| Parameter Recovery | Validation approach to assess if a model can recover known parameters from synthetic data | Statistical modeling |
+| Empirical Identifiability | The ability to uniquely determine parameter values from observed data in practice | Statistical theory |
+| Model Comparison | Systematic comparison of different model architectures and inference algorithms | Statistical modeling |
+| Model Selection | Process of choosing the most appropriate model based on validation metrics | Statistical modeling |
+| Synthetic Data | Artificially generated data with known parameters for validation purposes | Computational statistics |
 
@@
 | ELBO | Evidence Lower Bound, objective function in variational inference | Statistical learning theory |
+| Markov Chain Monte Carlo | Sampling-based method for Bayesian inference | Statistical learning theory |
+| No U-Turn Sampler | Adaptive variant of Hamiltonian Monte Carlo for MCMC | Statistical learning theory |
+| Credible Interval | Bayesian equivalent of confidence interval, derived from posterior distribution | Statistical learning theory |
+| R-hat Statistic | Diagnostic for MCMC convergence comparing within and between chain variance | Statistical learning theory |
+| Effective Sample Size | Measure of the effective number of independent samples from MCMC | Statistical learning theory |
+| Bayes Factor | Ratio of marginal likelihoods used for Bayesian model comparison | Statistical learning theory |
 | Protocol | Interface definition in Python typing | Python typing documentation |
@@
 | NumPyro | Probabilistic programming library built on JAX | NumPyro documentation |
+| torchode | Numerical ODE solver for PyTorch | torchode documentation |
+| diffrax | Numerical differential equation solver for JAX | diffrax documentation |
 | Railway-Oriented Programming | Programming pattern for explicit error handling | Functional programming |
+| Parameter Recovery Validation | Framework for validating model parameter recovery | Statistical modeling |
+| Model Comparison Framework | Framework for systematic model comparison | Statistical modeling |
 
@@
 | SVI | Stochastic Variational Inference |
+| MCMC | Markov Chain Monte Carlo |
+| NUTS | No U-Turn Sampler |
+| HMC | Hamiltonian Monte Carlo |
+| ELBO | Evidence Lower Bound |
 | ODE | Ordinary Differential Equation |
+| SDE | Stochastic Differential Equation |
+| PRV | Parameter Recovery Validation |
+| MCF | Model Comparison Framework |
+| ESS | Effective Sample Size |
+| CI | Credible Interval |
+| BF | Bayes Factor |
 | GPU | Graphics Processing Unit |
 | API | Application Programming Interface |
-| MCMC | Markov Chain Monte Carlo |
 | CI/CD | Continuous Integration/Continuous Deployment |

### 2025-05-14 nbs/development/architecture.qmd
docs(development): propagate model selection updates to architecture
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-**Related Algorithm Specification:** ALG-02 (Stochastic Variational Inference)
+**Related Algorithm Specification:** ALG-03 (Stochastic Variational Inference)
 
@@
 ```python
-def train(self, adata: AnnData, **kwargs) -> Dict[str, Any]:
+def train_svi(self, adata: AnnData, **kwargs) -> Dict[str, Any]:
     # Extract data from AnnData
@@
 
-    return {"losses": losses, "guide": guide}
+    return {"losses": losses, "guide": guide, "posterior_samples": guide.get_posterior()}
+```
+
+**Algorithm ID:** ALG-IMP-02
+
+**Algorithm Name:** Markov Chain Monte Carlo for RNA Velocity
+
+**Related Algorithm Specification:** ALG-04 (Markov Chain Monte Carlo Inference)
+
+**Implementation Approach:** NumPyro MCMC with NUTS
+
+**Algorithm Structure:**
+
+```python
+def train_mcmc(self, adata: AnnData, **kwargs) -> Dict[str, Any]:
+    # Extract data from AnnData
+    data = prepare_anndata(adata)
+    u_obs = data["u"]
+    s_obs = data["s"]
+
+    # Set up MCMC
+    num_samples = kwargs.get("num_samples", 1000)
+    num_warmup = kwargs.get("num_warmup", 500)
+    num_chains = kwargs.get("num_chains", 4)
+
+    # Create MCMC kernel
+    kernel = numpyro.infer.NUTS(self.forward)
+    mcmc = numpyro.infer.MCMC(
+        kernel,
+        num_samples=num_samples,
+        num_warmup=num_warmup,
+        num_chains=num_chains,
+        progress_bar=kwargs.get("progress_bar", True)
+    )
+
+    # Run MCMC
+    mcmc.run(jax.random.PRNGKey(0), u_obs=u_obs, s_obs=s_obs)
+
+    # Get diagnostics
+    diagnostics = {
+        "r_hat": mcmc.get_diagnostics()["r_hat"],
+        "effective_sample_size": mcmc.get_diagnostics()["effective_sample_size"],
+        "acceptance_rate": mcmc.get_diagnostics()["acceptance_rate"],
+        "divergences": mcmc.get_diagnostics()["divergences"]
+    }
+
+    # Get posterior samples
+    posterior_samples = mcmc.get_samples()
+
+    return {
+        "posterior_samples": posterior_samples,
+        "diagnostics": diagnostics,
+        "mcmc": mcmc
+    }
 ```
@@
     Main -->|contains| Guides["Inference Guides"]
+    Main -->|contains| MCMC["MCMC Models"]
     Main -->|contains| Registry["Registry System"]
     Main -->|contains| Factory["Factory System"]
+    Main -->|contains| Validation["Validation Framework"]
 
@@
     Guides -->|implements| GuidesInterface["InferenceGuide Protocol"]
+    MCMC -->|implements| MCMCInterface["MCMCModel Protocol"]
+
+    Validation -->|contains| ParameterRecovery["Parameter Recovery"]
+    Validation -->|contains| ModelComparison["Model Comparison"]
+    Validation -->|contains| ValidationMetrics["Validation Metrics"]
 
@@
     class Main main;
-    class Dynamics,Priors,Likelihoods,Observations,Guides component;
-    class DynamicsInterface,PriorsInterface,LikelihoodsInterface,ObservationsInterface,GuidesInterface interface;
+    class Dynamics,Priors,Likelihoods,Observations,Guides,MCMC component;
+    class DynamicsInterface,PriorsInterface,LikelihoodsInterface,ObservationsInterface,GuidesInterface,MCMCInterface interface;
     class Registry,Factory system;
+    class Validation,ParameterRecovery,ModelComparison,ValidationMetrics validation;
 ```
@@
 - **LegacyDynamicsModel**: Compatibility implementation matching the legacy model
+- **NonlinearDynamicsModel**: Implementation supporting nonlinear dynamics with numerical integration
 
@@
 
+    class NonlinearDynamicsModel {
+        +name: str
+        +description: str
+        +shared_time: bool
+        +t_scale_on: bool
+        +ode_solver: ODESolver
+        +forward(context) Dict
+        +steady_state(alpha, beta, gamma) Tuple
+        +solve_ode(initial_state, params, times) Tensor
+    }
+
     DynamicsModel <|.. StandardDynamicsModel : implements
     DynamicsModel <|.. LegacyDynamicsModel : implements
+    DynamicsModel <|.. NonlinearDynamicsModel : implements
 ```
@@
 - **LegacyAutoGuideFactory**: Compatibility implementation matching the legacy guide
+
+#### MCMC Models {#sec-mcmc-models}
+
+MCMC models implement Markov Chain Monte Carlo sampling for Bayesian inference.
+
+- **NUTSModel**: Implements the No U-Turn Sampler for MCMC
+- **HMCModel**: Implements Hamiltonian Monte Carlo for MCMC
+
+```{mermaid}
+classDiagram
+    class MCMCModel {
+        <<Protocol>>
+        +sample(model, data, num_samples, num_chains) Dict
+        +get_diagnostics() Dict
+    }
+
+    class NUTSModel {
+        +name: str
+        +description: str
+        +sample(model, data, num_samples, num_chains) Dict
+        +get_diagnostics() Dict
+        +compute_r_hat() Tensor
+        +compute_effective_sample_size() Tensor
+    }
+
+    class HMCModel {
+        +name: str
+        +description: str
+        +sample(model, data, num_samples, num_chains) Dict
+        +get_diagnostics() Dict
+        +compute_r_hat() Tensor
+        +compute_effective_sample_size() Tensor
+    }
+
+    MCMCModel <|.. NUTSModel : implements
+    MCMCModel <|.. HMCModel : implements
+```
+
+#### Validation Framework {#sec-validation-framework}
+
+The validation framework provides tools for model validation and comparison.
+
+- **ParameterRecoveryValidator**: Validates model's ability to recover known parameters from synthetic data
+- **ModelComparisonFramework**: Compares different model implementations and configurations
+- **ValidationMetrics**: Computes metrics for quantifying validation performance
+
+```{mermaid}
+classDiagram
+    class ParameterRecoveryValidator {
+        <<Protocol>>
+        +generate_synthetic_data(model_config, prior_config, num_cells, num_genes) Dict
+        +validate_parameter_recovery(model_config, synthetic_data) Dict
+    }
+
+    class ModelComparisonFramework {
+        <<Protocol>>
+        +compare_models(models, data, metrics) Dict
+        +rank_models(comparison_results, criteria) Dict
+    }
+
+    class ValidationMetrics {
+        <<Protocol>>
+        +compute_parameter_metrics(true_params, estimated_params) Dict
+        +compute_coverage_metrics(true_params, credible_intervals) Dict
+        +compute_model_comparison_metrics(models, data) Dict
+    }
+
+    class StandardParameterRecoveryValidator {
+        +generate_synthetic_data(model_config, prior_config, num_cells, num_genes) Dict
+        +validate_parameter_recovery(model_config, synthetic_data) Dict
+        +compute_validation_metrics(true_params, estimated_params) Dict
+    }
+
+    class StandardModelComparisonFramework {
+        +compare_models(models, data, metrics) Dict
+        +rank_models(comparison_results, criteria) Dict
+        +visualize_comparison(comparison_results) Dict
+    }
+
+    ParameterRecoveryValidator <|.. StandardParameterRecoveryValidator : implements
+    ModelComparisonFramework <|.. StandardModelComparisonFramework : implements
+```
 

### 2025-05-14 nbs/development/testing.qmd
docs(development): propagate model selection updates to testing
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
     number-sections: true
+bibliography: ../references.bib
+csl: ../bibstyle.csl
 ---
@@
 | RG-02 | Research Goal | Uncertainty Quantification | TEST-BDD-02 | Partial |
-| RG-03 | Research Goal | Parameter Inference | TEST-UNIT-01, TEST-BDD-07 | Planned |
-| FN-03 | Functional Requirement | Model Training | TEST-BDD-04 | Complete |
-| FN-05 | Functional Requirement | Velocity Computation | TEST-BDD-05 | Complete |
-| CAP-06 | Capability | Parameter Recovery Validation | TEST-BDD-07, TEST-UNIT-09 | Planned |
-| CAP-07 | Capability | Model Comparison | TEST-BDD-08, TEST-UNIT-10 | Planned |
+| RG-03 | Research Goal | Parameter Inference | TEST-UNIT-01, TEST-BDD-07 | In Progress |
+| RG-06 | Research Goal | Model Validation | TEST-BDD-07, TEST-BDD-08, TEST-UNIT-09, TEST-UNIT-10 | In Progress |
+| UG-06 | Usage Goal | Flexibility | TEST-BDD-08, TEST-UNIT-11 | Planned |
+| FN-03 | Functional Requirement | Parameter Recovery Validation | TEST-BDD-07, TEST-UNIT-09 | In Progress |
+| FN-04 | Functional Requirement | Model Training | TEST-BDD-04 | Complete |
+| FN-06 | Functional Requirement | Velocity Computation | TEST-BDD-05 | Complete |
+| FN-08 | Functional Requirement | Model Comparison | TEST-BDD-08, TEST-UNIT-10 | In Progress |
+| FN-09 | Functional Requirement | Model Selection | TEST-BDD-09, TEST-UNIT-11 | Planned |
+| CAP-06 | Capability | Parameter Recovery Validation | TEST-BDD-07, TEST-UNIT-09 | In Progress |
+| CAP-07 | Capability | Model Comparison | TEST-BDD-08, TEST-UNIT-10 | In Progress |
 | CMP-01 | Component | PyroVelocityModel | TEST-BDD-06 | Complete |
 | CMP-02 | Component | DynamicsModel | TEST-UNIT-05, TEST-BDD-01 | Complete |
-| SC-03 | Scientific Constraint | Parameter Identifiability | TEST-BDD-07 | Planned |
+| CMP-08 | Component | MCMCModel | TEST-UNIT-12, TEST-BDD-10 | In Progress |
+| CMP-09 | Component | ValidationFramework | TEST-UNIT-09, TEST-BDD-07 | In Progress |
+| SC-03 | Scientific Constraint | Parameter Identifiability | TEST-BDD-07 | In Progress |
+| SC-06 | Scientific Constraint | Model Comparison Limitations | TEST-BDD-08 | Planned |
+| SC-07 | Scientific Constraint | Inference Algorithm Tradeoffs | TEST-BDD-10 | In Progress |
+| SC-09 | Scientific Constraint | Empirical Identifiability | TEST-BDD-11 | Planned |
 | MM-01 | Mathematical Model | RNA Velocity Dynamics | TEST-UNIT-08 | Complete |
-| ALG-02 | Algorithm | Parameter Recovery Validation | TEST-UNIT-09 | Planned |
+| ALG-02 | Algorithm | Parameter Recovery Validation | TEST-UNIT-09 | In Progress |
+| ALG-03 | Algorithm | Stochastic Variational Inference | TEST-UNIT-13 | Complete |
+| ALG-04 | Algorithm | Markov Chain Monte Carlo | TEST-UNIT-12 | In Progress |
 
@@
 | model | PyroVelocity model trains on AnnData | ✅ Implemented | Verify that the PyroVelocityModel can train on AnnData objects | CMP-01, FN-03, INT-01 |
-| validation | Parameter recovery validation | 🔄 Planned | Verify that the model can recover known parameters from synthetic data | CAP-06, RG-03, SC-03, ALG-02 |
-| validation | Model comparison | 🔄 Planned | Verify that different model implementations can be compared | CAP-07, TC-04, SG-03 |
+| validation | Parameter recovery validation | ⚠️ In Progress | Verify that the model can recover known parameters from synthetic data | CAP-06, RG-03, RG-06, SC-03, ALG-02 |
+| validation | MCMC inference | ⚠️ In Progress | Verify that MCMC inference can recover parameters from synthetic data | TC-07, SC-07, ALG-04 |
+| validation | Inference algorithm comparison | ⚠️ In Progress | Compare SVI and MCMC inference algorithms | SC-07, ACC-07, TC-10 |
+| validation | Model comparison | ⚠️ In Progress | Verify that different model implementations can be compared | CAP-07, TC-04, TC-08, SG-03 |
+| validation | Empirical identifiability assessment | 🔄 Planned | Assess parameter identifiability across different sample sizes and noise levels | SC-09, ACC-06, TC-09 |
+| validation | Model selection | 🔄 Planned | Select the most appropriate model based on validation metrics | SC-06, SC-10, UG-06 |
 
@@
 2. **Synthetic Data Generation**: Create synthetic data using known parameters
-3. **Inference**: Perform inference on synthetic data to estimate parameters
+3. **Inference**: Perform inference on synthetic data to estimate parameters using both SVI and MCMC
 4. **Evaluation**: Compare estimated parameters to original known values
 5. **Metrics Computation**: Calculate recovery accuracy, precision, and coverage metrics
+6. **Credible Interval Analysis**: Analyze credible interval coverage for different inference algorithms
+7. **Empirical Identifiability Assessment**: Evaluate parameter identifiability across different sample sizes and noise levels
 
@@
 - Functions for creating synthetic data with known parameters
-- Functions for performing inference on synthetic data
+- Functions for performing inference on synthetic data using both SVI and MCMC
 - Functions for comparing estimated parameters to true values
 - Functions for calculating and visualizing recovery metrics
+- Functions for analyzing credible interval coverage
+- Functions for assessing empirical identifiability
+
+### Inference Algorithm Comparison
+
+The parameter recovery validation framework includes tools for comparing different inference algorithms:
+
+- **Stochastic Variational Inference (SVI)**: Fast approximate inference using optimization
+- **Markov Chain Monte Carlo (MCMC)**: Slower but more accurate sampling-based inference
+
+The framework evaluates each algorithm's performance on parameter recovery tasks, considering:
+
+- **Accuracy**: How close are the estimated parameters to the true values?
+- **Precision**: How much uncertainty is there in the parameter estimates?
+- **Coverage**: Do the credible intervals contain the true parameter values?
+- **Computational Efficiency**: How long does each algorithm take to run?
+- **Convergence Reliability**: How reliably does each algorithm converge?
+
+### Sample Size and Noise Level Analysis
+
+The framework includes tools for analyzing parameter recovery performance across different:
+
+- **Sample Sizes**: How does recovery performance change with the number of cells?
+- **Noise Levels**: How robust is parameter recovery to different levels of noise?
+- **Parameter Regimes**: How does recovery performance vary across different parameter values?
+
+This analysis helps identify the minimum sample size required for reliable parameter inference and the sensitivity of the model to noise.
+
+## Model Comparison Framework
+
+The model comparison framework provides tools for systematically comparing different model architectures, implementations, and inference algorithms.
+It enables researchers to select the most appropriate model for their specific use case based on quantitative metrics.
+
+The model comparison workflow follows these steps:
+
+1. **Model Configuration**: Configure multiple model variants to compare
+2. **Training**: Train each model on the same dataset using the same or different inference algorithms
+3. **Validation**: Validate each model using parameter recovery validation
+4. **Metric Computation**: Compute comparison metrics for each model
+5. **Ranking**: Rank models based on different criteria
+6. **Visualization**: Visualize comparison results for interpretation
+
+The framework provides tools for each step of this workflow, including:
+
+- Functions for configuring multiple model variants
+- Functions for training models with different inference algorithms
+- Functions for validating models using parameter recovery validation
+- Functions for computing comparison metrics
+- Functions for ranking models based on different criteria
+- Functions for visualizing comparison results
+
+### Comparison Metrics
+
+The model comparison framework includes a variety of metrics for comparing models:
+
+- **Parameter Recovery Performance**: How well does each model recover known parameters?
+- **Uncertainty Calibration**: How well calibrated are the uncertainty estimates?
+- **Computational Efficiency**: How computationally efficient is each model?
+- **Convergence Reliability**: How reliably does each model converge?
+- **Biological Relevance**: How well does each model capture known biological phenomena?
+
+### Model Selection Criteria
+
+The framework provides tools for selecting models based on different criteria:
+
+- **Accuracy-Focused**: Select the model with the best parameter recovery performance
+- **Efficiency-Focused**: Select the most computationally efficient model
+- **Robustness-Focused**: Select the model with the most reliable convergence
+- **Balanced**: Select the model with the best balance of accuracy, efficiency, and robustness
+- **Domain-Specific**: Select the model that best captures specific biological phenomena
 
@@
 - **TC-04**: Backward Compatibility constraint
+- **TC-07**: MCMC Implementation constraint
+- **TC-08**: Model Comparison Framework constraint
+- **TC-09**: Parameter Recovery Validation constraint
+- **TC-10**: Multiple Inference Algorithms constraint
 - **REQ-01**: Parameter Estimation Accuracy requirement
+- **ACC-05**: Model Comparison Accuracy requirement
+- **ACC-06**: Empirical Identifiability requirement
+- **ACC-07**: Inference Algorithm Accuracy requirement
 - **SG-03**: Comprehensive Testing system goal
-- **REL-01**: Training Convergence reliability requirement
+- **REL-01**: SVI Convergence reliability requirement
+- **REL-02**: MCMC Convergence reliability requirement
+- **REL-06**: Parameter Recovery Reliability requirement
+- **REL-07**: Model Comparison Reliability requirement
 - **CAP-06**: Parameter Recovery Validation capability
@@
 - **RG-03**: Parameter Inference research goal
+- **RG-06**: Model Validation research goal
+- **UG-06**: Flexibility usage goal
 - **SC-03**: Parameter Identifiability scientific constraint
+- **SC-06**: Model Comparison Limitations scientific constraint
+- **SC-07**: Inference Algorithm Tradeoffs scientific constraint
+- **SC-08**: Synthetic Data Limitations scientific constraint
+- **SC-09**: Empirical Identifiability scientific constraint
+- **SC-10**: Model Misspecification scientific constraint
 

### 2025-05-14 src/pyrovelocity/io/compressedpickle.py
fix(pyrovelocity): fix docstrings for html render
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -57,14 +57,14 @@ class CompressedPickle:
     A class for reading and writing zstandard-compressed pickle files.
 
     Examples:
-    >>> import pandas as pd
-    >>> tmp = getfixture("tmp_path")
-    >>> test_data_path = tmp / "test_data.pkl.zst"
-    >>> test_data = pd.DataFrame({'a': [1, 2, 3], 'b': [4, 5, 6]})
-    >>> CompressedPickle.save(test_data_path, test_data)
-    >>> loaded_data = CompressedPickle.load(test_data_path)
-    >>> loaded_data.equals(test_data)
-    True
+        >>> import pandas as pd
+        >>> tmp = getfixture("tmp_path")
+        >>> test_data_path = tmp / "test_data.pkl.zst"
+        >>> test_data = pd.DataFrame({'a': [1, 2, 3], 'b': [4, 5, 6]})
+        >>> CompressedPickle.save(test_data_path, test_data)
+        >>> loaded_data = CompressedPickle.load(test_data_path)
+        >>> loaded_data.equals(test_data)
+        True
     """
 
     @staticmethod

### 2025-05-14 src/pyrovelocity/models/modular/interfaces.py

@@ -20,7 +20,7 @@ Components communicate through a shared context dictionary that is passed
 between components during model execution. Each component updates the context
 with its outputs, which are then used by subsequent components.
 
-Example:
+Examples:
     ```python
     # Create components
     dynamics_model = StandardDynamicsModel()
@@ -79,18 +79,21 @@ class DynamicsModel(Protocol):
     splicing, and degradation rates in the RNA velocity model. They implement
     analytical or numerical solutions to the RNA velocity differential equations.
 
-    The standard RNA velocity model is based on the following system of ODEs:
+    The standard RNA velocity model is based on the following system of ODEs
+
         du/dt = α - βu
         ds/dt = βu - γs
 
-    Where:
-        u: unspliced mRNA abundance
-        s: spliced mRNA abundance
-        α: transcription rate
-        β: splicing rate
-        γ: degradation rate
+    Where
 
-    Implementations must:
+        - u: unspliced mRNA abundance
+        - s: spliced mRNA abundance
+        - α: transcription rate
+        - β: splicing rate
+        - γ: degradation rate
+
+    Implementations must
+
     1. Provide a forward method that computes expected unspliced and spliced counts
     2. Provide a steady_state method that computes steady-state values
     3. Handle both analytical and numerical solutions as appropriate
@@ -175,14 +178,16 @@ class PriorModel(Protocol):
     and any additional parameters required by the model. They use Pyro's probabilistic
     programming framework to register samples with the model.
 
-    Implementations must:
+    Implementations must
+
     1. Define appropriate prior distributions for all model parameters
     2. Use Pyro's plate mechanism for batch dimensions
     3. Register all samples with Pyro
     4. Support disabling priors during guide execution (if needed)
     5. Handle hyperparameter configuration
 
-    Example implementation:
+    Example implementation
+
         ```python
         class LogNormalPriorModel(BasePriorModel):
             def __init__(

### 2025-05-14 src/pyrovelocity/models/modular/model.py

@@ -17,7 +17,7 @@ The main class in this module is `PyroVelocityModel`, which composes the differe
 components into a cohesive probabilistic model. The model uses functional composition
 for the forward method, enabling railway-oriented programming patterns.
 
-Example:
+Examples:
     >>> import torch
     >>> import pyro
     >>> from pyrovelocity.models.modular.factory import create_standard_model
@@ -137,6 +137,7 @@ class PyroVelocityModel:
     patterns.
 
     The model follows a clear separation of concerns:
+
     - Dynamics model: Defines the velocity vector field
     - Prior model: Specifies priors for model parameters
     - Likelihood model: Defines the observation likelihood
@@ -144,6 +145,7 @@ class PyroVelocityModel:
     - Guide model: Implements the inference guide for posterior approximation
 
     The PyroVelocityModel provides methods for:
+
     1. Running the model forward to compute expected RNA counts
     2. Training the model using SVI (Stochastic Variational Inference)
     3. Generating posterior samples for uncertainty quantification

### 2025-05-14 src/pyrovelocity/models/modular/registry.py

@@ -54,7 +54,7 @@ class Registry(Generic[T]):
         Returns:
             A decorator function that registers the decorated class.
 
-        Example:
+        Examples:
             @DynamicsModelRegistry.register("standard")
             class StandardDynamicsModel:
                 ...

### 2025-05-14 src/pyrovelocity/models/modular/utils/context_utils.py

@@ -41,7 +41,7 @@ def validate_context(
     Returns:
         A Result.Error if validation fails, or the original context if validation succeeds.
 
-    Example:
+    Examples:
         ```python
         def forward(self, context: Dict[str, Any]) -> Dict[str, Any]:
             # Validate context

### 2025-05-14 src/pyrovelocity/models/modular/utils/core_utils.py

@@ -30,7 +30,7 @@ class ComponentError:
         message: A descriptive error message
         details: Additional details about the error (optional)
 
-    Example:
+    Examples:
         ```python
         # Create an error
         error = ComponentError(
@@ -89,7 +89,7 @@ def create_error(
     Returns:
         A Result.Error containing a ComponentError.
 
-    Example:
+    Examples:
         ```python
         def forward(self, context: Dict[str, Any]) -> Dict[str, Any]:
             # Validate inputs

### 2025-05-14 src/pyrovelocity/models/modular/utils/pyro_utils.py

@@ -25,7 +25,7 @@ def register_buffer(obj: object, name: str, tensor: torch.Tensor) -> None:
         name: The name to register the buffer under.
         tensor: The tensor to register.
 
-    Example:
+    Examples:
         ```python
         class MyPriorModel:
             def __init__(self, alpha_loc: float = 0.0, alpha_scale: float = 1.0):

### 2025-05-14 src/pyrovelocity/tasks/evaluate.py

@@ -154,10 +154,6 @@ def calculate_cross_boundary_correctness(
 
     Args:
         model_results: List of dictionaries containing model results.
-            Each dictionary should have:
-            - data_model: str - Dataset and model name (e.g., 'pancreas_model1')
-            - postprocessed_data: Union[str, Path, AnnData] - Path to postprocessed AnnData file
-              or the actual AnnData object
         output_dir: Directory to save results to
         dataset_configs: Mapping of dataset names to their cluster and embedding keys
         ground_truth_transitions: Mapping of dataset names to their ground truth cell transitions
@@ -166,10 +162,7 @@ def calculate_cross_boundary_correctness(
         random_seed: Random seed for reproducibility. Defaults to 42.
 
     Returns:
-        Tuple of paths to:
-        - Summary CSV file
-        - Individual dataset results directory
-        - Plot file
+        Tuple of paths to Summary CSV file, Individual dataset results directory, Plot file
     """
     set_seed(random_seed)
     logger.info(f"Reset random state from seed: {random_seed}")

### 2025-05-14 src/pyrovelocity/tasks/train.py

@@ -53,10 +53,12 @@ def train_dataset(
     """
     Loads processed data, trains model, and saves model and posterior samples.
 
-    Inputs:
-        data/processed/{data_set_name}_processed.h5ad
+    Inputs
 
-    Outputs:
+        - data/processed/{data_set_name}_processed.h5ad
+
+    Outputs
+
         models/{data_model}/
         ├── trained.h5ad
         ├── posterior_samples.pkl.zst
@@ -311,6 +313,15 @@ def train_model(
     Train a PyroVelocity model to provide probabilistic estimates of RNA velocity
     for single-cell RNA sequencing data with quantified splice variants.
 
+    These arguments are deprecated
+
+    # svi_train: bool = False,
+    # train_size: float = 1.0,
+    # cell_state: str = "clusters",
+    # svi_train (bool, optional): Whether to use Stochastic Variational Inference for training. Default is False.
+    # train_size (float, optional): Proportion of data to be used for training. Default is 1.0.
+    # cell_state (str, optional): Cell state attribute in the AnnData object. Default is "clusters".
+
     Args:
         adata (str | AnnData): Path to a file that can be read to an AnnData object or an AnnData object.
         guide_type (str, optional): The type of guide function for the Pyro model. Default is "auto".
@@ -333,14 +344,6 @@ def train_model(
         kinetics_num (int, optional): Number of kinetics parameters. Default is 2.
         loss_plot_path (str, optional): Path to save the loss plot. Default is "loss_plot.png".
         loss_csv_path (str, optional): Path to save the loss data as CSV. Default is "loss_plot.png.csv".
-
-        These arguments are deprecated:
-        # svi_train: bool = False,
-        # train_size: float = 1.0,
-        # cell_state: str = "clusters",
-        # svi_train (bool, optional): Whether to use Stochastic Variational Inference for training. Default is False.
-        # train_size (float, optional): Proportion of data to be used for training. Default is 1.0.
-        # cell_state (str, optional): Cell state attribute in the AnnData object. Default is "clusters".
 
     Returns:
         Tuple[AnnData, PyroVelocity, Dict[str, ndarray]]: A tuple containing the processed AnnData object, the trained PyroVelocity model, and a dictionary of posterior samples.

### 2025-05-15 nbs/development/architecture.qmd
docs(architecture): fix duplicate sec ref
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(development): note requirements for AnnData object mutation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-#### Validation Framework {#sec-validation-framework}
+#### Validation Framework {#sec-core-validation-framework}
 
@@
 
+# AnnData Integration {#sec-anndata-integration}
+
+PyroVelocity integrates with AnnData objects following these principles:
+
+1. **Mutation Tracking**: All methods that modify AnnData objects must track and document mutations using the `print_anndata`, `anndata_string`, and `print_string_diff` utilities.
+
+2. **Return Values**: Methods that modify AnnData objects must return the modified object to enable method chaining and make mutations explicit.
+
+3. **Required Parameters**: AnnData parameters should be required (not optional) when they are essential for method functionality.
+
+4. **Shape Validation**: Array shapes should be validated using jaxtyping rather than complex shape handling code.
+
+5. **Explicit Documentation**: Methods that modify objects must explicitly document this behavior in their docstrings.
+
+6.
+**Distinction Between Model State and External Data**: While the model state follows an immutable pattern (using `ModelState` and `with_state()`), external data structures like AnnData objects can be mutated in-place for efficiency.
+This distinction is important because:
+   - Model state must remain consistent and thread-safe
+   - AnnData objects are large and copying them is expensive
+   - In-place mutation of AnnData is a common pattern in the single-cell ecosystem
+
+These principles ensure that mutations to AnnData objects are explicit, traceable, and well-documented, making the code more maintainable and reducing the risk of bugs while maintaining performance with large datasets.
+
 # Technical Debt and Evolution {#sec-technical-debt}

### 2025-05-15 nbs/development/requirements.qmd
docs(requirements): remove unavailable bibtex key
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(development): note requirements for AnnData object mutation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-- NumPyro documentation [@Phan2019-numpyro]
+- NumPyro documentation
 - Hoffman and Gelman (2014) [No U-Turn Sampler]
@@
 - Validation of model comparison metrics
+
+## API Design Principles {#sec-api-design-principles}
+
+| Principle ID | Principle Name | Description | Priority |
+|--------------|----------------|-------------|----------|
+| API-01 | AnnData Mutation Tracking | Methods that modify AnnData objects must track and document mutations using print_anndata, anndata_string, and print_string_diff | High |
+| API-02 | AnnData Return Values | Methods that modify AnnData objects must return the modified object | High |
+| API-03 | Required Parameters | Parameters that are essential for method functionality should be required, not optional | High |
+| API-04 | Shape Validation | Array shapes should be validated using jaxtyping rather than complex shape handling code | High |
+| API-05 | Explicit Documentation | Methods that modify objects must explicitly document this behavior in their docstrings | High |
 

### 2025-05-15 nbs/tutorials/models.qmd
docs(tutorials): init models
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "Model definition"
execute:
  freeze: true
  eval: true
  warning: false
  error: false
  cache: true
toc: true
number-sections: true
highlight-style: gruvbox
csl: ../../bibstyle.csl
lightbox: auto
format:
  html:
    html-math-method: katex
#   ipynb: default
---

# Introduction to PyroVelocity Modular Models

This tutorial introduces PyroVelocity modular models, which follows a Protocol-First architecture with composable components. The modular implementation provides a flexible and extensible framework for building RNA velocity models.

## Prerequisites

Before starting, make sure you have PyroVelocity [installed](../guides/installation.qmd).

## Basic Usage

Let's start with a simple example using the modular implementation:

```python
import torch
import pyro
import anndata as ad
import numpy as np
import matplotlib.pyplot as plt
from pyrovelocity.utils import print_anndata, anndata_string, print_string_diff
from pyrovelocity.models.modular import PyroVelocityModel
from pyrovelocity.models.modular.factory import create_legacy_model1

# Set random seed for reproducibility
torch.manual_seed(42)
np.random.seed(42)
pyro.set_rng_seed(42)
```

```python
# Create synthetic data
n_cells, n_genes = 100, 10
u_data = np.random.poisson(5, size=(n_cells, n_genes))
s_data = np.random.poisson(5, size=(n_cells, n_genes))

# Create AnnData object
adata = ad.AnnData(X=s_data)
adata.layers["spliced"] = s_data
adata.layers["unspliced"] = u_data
adata.obs_names = [f"cell_{i}" for i in range(n_cells)]
adata.var_names = [f"gene_{i}" for i in range(n_genes)]
print_anndata(adata)
initial_data_state_representation = anndata_string(adata)

# Set up AnnData for PyroVelocity
adata = PyroVelocityModel.setup_anndata(adata)
final_data_state_representation = anndata_string(adata)
print_string_diff(
    initial_data_state_representation,
    final_data_state_representation,
    diff_title="Setup AnnData diff",
)
print_anndata(adata)
```

```python
# Create a standard model with default components
model = create_legacy_model1()
```

```python
# Train the model
model.train(
    adata=adata,
    max_epochs=20,  # Use a small number for demonstration
    learning_rate=0.01,
    use_gpu=False  # Set to True if GPU is available
)
```

```python
# Generate posterior samples
posterior_samples = model.generate_posterior_samples(
    adata=adata,
    num_samples=30  # Use a small number for demonstration
)
```

```python
# Store results in AnnData
adata = model.store_results_in_anndata(
    adata=adata,
    posterior_samples=posterior_samples
)

# Compute velocity and store it in AnnData
velocity = model.get_velocity(
    adata=adata,
    num_samples=30,
    basis='umap'  # This will be used for embedding
)
```

```python
# Print available keys in posterior samples
print("Available keys in posterior samples:")
for key in posterior_samples.keys():
    print(f"- {key}: {posterior_samples[key].shape}")
```

## Understanding the Modular Architecture

The modular implementation is based on a Protocol-First architecture with five main component types:

1. **Dynamics Model**: Defines the mathematical models of RNA dynamics
2. **Prior Model**: Defines prior distributions for model parameters
3. **Likelihood Model**: Defines observation distributions
4. **Observation Model**: Handles data preprocessing and transformation
5. **Guide Model**: Creates guide functions for variational inference

Each component implements a Protocol interface, allowing for flexible composition and extension.

## Creating Custom Models

You can create custom models by combining different components:

```python
from pyrovelocity.models.modular.components import (
    StandardDynamicsModel,
    LegacyDynamicsModel,
    LogNormalPriorModel,
    PoissonLikelihoodModel,
    LegacyLikelihoodModel,
    StandardObservationModel,
    AutoGuideFactory
)
from pyrovelocity.models.modular.model import PyroVelocityModel

# Create components
dynamics_model = LegacyDynamicsModel(shared_time=True, t_scale_on=False)
prior_model = LogNormalPriorModel()
likelihood_model = LegacyLikelihoodModel()
observation_model = StandardObservationModel(correct_library_size=True)
guide_model = AutoGuideFactory()

# Create model
custom_model = PyroVelocityModel(
    dynamics_model=dynamics_model,
    prior_model=prior_model,
    likelihood_model=likelihood_model,
    observation_model=observation_model,
    guide_model=guide_model
)

# Train the model
custom_model.train(
    adata=adata,
    max_epochs=20,
    learning_rate=0.01,
    use_gpu=False
)
```

## Using Factory Functions

The modular implementation provides factory functions for creating models with different configurations:

```python
from pyrovelocity.models.modular.factory import (
    create_model_from_config,
    create_legacy_model1,
    create_legacy_model2
)
from pyrovelocity.models.modular.config import ModelConfig

# Create a model that replicates the legacy implementation
legacy_model = create_legacy_model1()

# Create a model from a configuration
config = ModelConfig.standard()
config_model = create_model_from_config(config)
```

## Visualizing Results

After training the model and generating posterior samples, you can visualize the results:

```python
import scanpy as sc
import scvelo as scv

# Compute PCA and UMAP embeddings
sc.pp.pca(adata)
sc.pp.neighbors(adata)
sc.tl.umap(adata)

# Visualize velocity
scv.pl.velocity_embedding_stream(
    adata,
    basis='umap',
    color='gene_0',
    vkey='velocity_pyro'  # Use the velocity key added by get_velocity
)
```

## Conclusion

This tutorial introduced the basics of PyroVelocity's modular implementation. The modular architecture provides a flexible and extensible framework for building RNA velocity models, allowing you to customize components and create models tailored to your specific needs.

For more advanced usage, refer to the [API reference](../reference/models.PyroVelocityModel.qmd) and [how-to guides](../guides/model_components.qmd).

### 2025-05-15 nbs/guides/model_components.qmd
docs(guides): init model components
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "Customizing Modular Components"
format:
  html:
    code-fold: false
---

# How to Customize Modular Components in PyroVelocity

This guide shows how to customize components in PyroVelocity's modular implementation. The modular architecture allows you to create custom components that implement the Protocol interfaces, enabling you to tailor the model to your specific needs.

## Available Component Types

PyroVelocity's modular implementation defines five main component types:

1. **Dynamics Model**: Defines the mathematical models of RNA dynamics
2. **Prior Model**: Defines prior distributions for model parameters
3. **Likelihood Model**: Defines observation distributions
4. **Observation Model**: Handles data preprocessing and transformation
5. **Guide Model**: Creates guide functions for variational inference

Each component type has a corresponding Protocol interface that defines the contract that implementations must follow.

## Creating a Custom Dynamics Model

Let's create a custom dynamics model that implements a nonlinear ODE system:

```python
import torch
import pyro
from typing import Dict, Any, Tuple
from pyrovelocity.models.modular.interfaces import DynamicsModel
from pyrovelocity.models.modular.registry import DynamicsModelRegistry

@DynamicsModelRegistry.register("nonlinear")
class NonlinearDynamicsModel:
    """
    Custom dynamics model implementing a nonlinear ODE system.

    This model extends the standard RNA velocity model with nonlinear terms.
    """

    def __init__(self, shared_time: bool = True, t_scale_on: bool = False):
        """
        Initialize the nonlinear dynamics model.

        Args:
            shared_time: Whether to use shared time across genes
            t_scale_on: Whether to use time scaling
        """
        self.name = "nonlinear"
        self.description = "Nonlinear dynamics model for RNA velocity"
        self.shared_time = shared_time
        self.t_scale_on = t_scale_on

    def forward(self, context: Dict[str, Any]) -> Dict[str, Any]:
        """
        Compute expected RNA counts based on nonlinear dynamics model.

        Args:
            context: Context dictionary with model state

        Returns:
            Updated context with computed values
        """
        # Extract parameters from context
        u_obs = context["u_obs"]
        s_obs = context["s_obs"]
        alpha = context["alpha"]
        beta = context["beta"]
        gamma = context["gamma"]
        t = context.get("t", torch.ones_like(u_obs))

        # Compute steady state
        u_inf, s_inf = self.steady_state(alpha, beta, gamma)

        # Compute expected counts using numerical approximation
        # For simplicity, we'll use a basic Euler method here
        dt = 0.1
        steps = int(torch.max(t) / dt)

        u_expected = u_obs.clone()
        s_expected = s_obs.clone()

        for _ in range(steps):
            # Nonlinear ODE system
            # du/dt = alpha - beta * u * (1 + s/10)
            # ds/dt = beta * u - gamma * s
            u_delta = alpha - beta * u_expected * (1 + s_expected/10)
            s_delta = beta * u_expected - gamma * s_expected

            u_expected = u_expected + dt * u_delta
            s_expected = s_expected + dt * s_delta

        # Update context with computed values
        context.update({
            "u_expected": u_expected,
            "s_expected": s_expected,
            "u_inf": u_inf,
            "s_inf": s_inf,
        })

        return context

    def steady_state(
        self, alpha: torch.Tensor, beta: torch.Tensor, gamma: torch.Tensor, **kwargs
    ) -> Tuple[torch.Tensor, torch.Tensor]:
        """
        Compute steady-state RNA counts.

        Args:
            alpha: Transcription rate
            beta: Splicing rate
            gamma: Degradation rate

        Returns:
            Tuple of steady-state unspliced and spliced counts
        """
        # For the nonlinear model, we need to solve:
        # alpha - beta * u_inf * (1 + s_inf/10) = 0
        # beta * u_inf - gamma * s_inf = 0

        # From the second equation: s_inf = beta * u_inf / gamma
        # Substituting into the first:
        # alpha - beta * u_inf * (1 + beta * u_inf / (gamma * 10)) = 0

        # For simplicity, we'll use an approximation here
        # In a real implementation, you might use a numerical solver

        # First approximation: ignore the nonlinear term
        u_inf_approx = alpha / beta
        s_inf_approx = alpha / gamma

        # Refine with a few iterations
        for _ in range(5):
            s_inf_approx = beta * u_inf_approx / gamma
            u_inf_approx = alpha / (beta * (1 + s_inf_approx/10))

        return u_inf_approx, s_inf_approx
```

## Creating a Custom Prior Model

Now, let's create a custom prior model with informative priors:

```python
import torch
import pyro
from typing import Dict, Any
from pyrovelocity.models.modular.interfaces import PriorModel
from pyrovelocity.models.modular.registry import PriorModelRegistry

@PriorModelRegistry.register("informative")
class InformativePriorModel:
    """
    Custom prior model with informative priors.

    This model uses informative priors based on prior knowledge about
    RNA dynamics parameters.
    """

    def __init__(self):
        """Initialize the informative prior model."""
        self.name = "informative"
        self.description = "Informative prior model for RNA velocity"

    def forward(self, context: Dict[str, Any]) -> Dict[str, Any]:
        """
        Sample parameters from informative prior distributions.

        Args:
            context: Context dictionary with model state

        Returns:
            Updated context with sampled parameters
        """
        # Extract dimensions from context
        u_obs = context["u_obs"]
        num_cells, num_genes = u_obs.shape

        # Create plates for batched sampling
        with pyro.plate("genes", num_genes):
            # Sample alpha (transcription rate) from gamma distribution
            # with informative shape and rate parameters
            alpha = pyro.sample(
                "alpha",
                pyro.distributions.Gamma(
                    torch.ones(num_genes) * 2.0,  # shape parameter
                    torch.ones(num_genes) * 0.5   # rate parameter
                )
            )

            # Sample beta (splicing rate) from gamma distribution
            beta = pyro.sample(
                "beta",
                pyro.distributions.Gamma(
                    torch.ones(num_genes) * 4.0,  # shape parameter
                    torch.ones(num_genes) * 2.0   # rate parameter
                )
            )

            # Sample gamma (degradation rate) from gamma distribution
            gamma = pyro.sample(
                "gamma",
                pyro.distributions.Gamma(
                    torch.ones(num_genes) * 3.0,  # shape parameter
                    torch.ones(num_genes) * 1.5   # rate parameter
                )
            )

        # Update context with sampled parameters
        context.update({
            "alpha": alpha,
            "beta": beta,
            "gamma": gamma,
        })

        return context
```

## Using Custom Components

Now, let's use our custom components to create a model:

```python
from pyrovelocity.models.modular.model import PyroVelocityModel
from pyrovelocity.models.modular.components import (
    StandardObservationModel,
    PoissonLikelihoodModel,
    AutoGuideFactory
)

# Create components
dynamics_model = NonlinearDynamicsModel(shared_time=True, t_scale_on=False)
prior_model = InformativePriorModel()
likelihood_model = PoissonLikelihoodModel()
observation_model = StandardObservationModel(correct_library_size=True)
guide_model = AutoGuideFactory()

# Create model
custom_model = PyroVelocityModel(
    dynamics_model=dynamics_model,
    prior_model=prior_model,
    likelihood_model=likelihood_model,
    observation_model=observation_model,
    guide_model=guide_model
)

# Train the model
custom_model.train(
    adata=adata,
    max_epochs=100,
    learning_rate=0.01,
    use_gpu=False
)
```

## Creating a Model with Factory Functions

You can also use factory functions to create models with custom components:

```python
from pyrovelocity.models.modular.factory import create_model_from_config
from pyrovelocity.models.modular.config import (
    ModelConfig,
    ComponentConfig
)

# Create component configurations
dynamics_config = ComponentConfig(name="nonlinear", params={"shared_time": True})
prior_config = ComponentConfig(name="informative", params={})
likelihood_config = ComponentConfig(name="poisson", params={})
observation_config = ComponentConfig(name="standard", params={"correct_library_size": True})
guide_config = ComponentConfig(name="auto", params={})

# Create model configuration
model_config = ModelConfig(
    dynamics_model=dynamics_config,
    prior_model=prior_config,
    likelihood_model=likelihood_config,
    observation_model=observation_config,
    inference_guide=guide_config
)

# Create model from configuration
model = create_model_from_config(model_config)
```

## Registering Custom Components

Custom components need to be registered with the appropriate registry to be used with factory functions. The registration is done using the `@Registry.register` decorator:

```python
from pyrovelocity.models.modular.registry import (
    DynamicsModelRegistry,
    PriorModelRegistry,
    LikelihoodModelRegistry,
    ObservationModelRegistry,
    InferenceGuideRegistry
)

# Register a custom dynamics model
@DynamicsModelRegistry.register("my_dynamics")
class MyDynamicsModel:
    # Implementation...
    pass

# Register a custom prior model
@PriorModelRegistry.register("my_prior")
class MyPriorModel:
    # Implementation...
    pass

# Register a custom likelihood model
@LikelihoodModelRegistry.register("my_likelihood")
class MyLikelihoodModel:
    # Implementation...
    pass

# Register a custom observation model
@ObservationModelRegistry.register("my_observation")
class MyObservationModel:
    # Implementation...
    pass

# Register a custom inference guide
@InferenceGuideRegistry.register("my_guide")
class MyGuideFactory:
    # Implementation...
    pass
```

## Conclusion

This guide showed how to create and use custom components in PyroVelocity's modular implementation. By implementing the Protocol interfaces and registering your components, you can create models tailored to your specific needs.

For more information, refer to the [API reference](../reference/models.PyroVelocityModel.qmd) and the [modular implementation tutorial](../tutorials/models.qmd).

### 2025-05-15 nbs/guides/inference.qmd
docs(guides): init inference
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "Inference with PyroVelocity"
format:
  html:
    code-fold: false
---

# Inference with PyroVelocity

This guide explains how to perform inference with PyroVelocity, covering both the legacy and modular implementations.

## Legacy Implementation

The legacy implementation provides a single `PyroVelocity` class that handles all aspects of the model:

```python
import anndata as ad
import numpy as np
from pyrovelocity.models import PyroVelocity

# Load data
adata = ad.read_h5ad("path/to/data.h5ad")

# Set up AnnData for PyroVelocity
PyroVelocity.setup_anndata(adata)

# Create model
model = PyroVelocity(
    adata,
    model_type="normal",  # Options: "normal", "deterministic", "stochastic"
    guide_type="auto",    # Options: "auto", "delta"
    likelihood="Poisson", # Options: "Poisson", "NegativeBinomial"
    shared_time=True,     # Whether to use shared time across genes
    t_scale_on=False      # Whether to use time scaling
)

# Train model
model.train(
    max_epochs=1000,
    learning_rate=0.01,
    use_gpu="auto"
)

# Generate posterior samples
posterior_samples = model.generate_posterior_samples(
    adata=adata,
    num_samples=100
)

# Compute statistics from posterior samples
model.compute_statistics_from_posterior_samples(
    adata=adata,
    posterior_samples=posterior_samples
)

# Save model
model.save_model("path/to/save/model")

# Load model
model = PyroVelocity.load_model("path/to/save/model", adata)
```

## Modular Implementation

The modular implementation provides a more flexible approach with composable components:

```python
import anndata as ad
import numpy as np
from pyrovelocity.models.modular import (
    create_standard_model,
    PyroVelocityModel
)

# Load data
adata = ad.read_h5ad("path/to/data.h5ad")

# Set up AnnData for PyroVelocity
adata = PyroVelocityModel.setup_anndata(adata)

# Create model using factory function
model = create_standard_model()

# Train model
model.train(
    adata=adata,
    max_epochs=1000,
    learning_rate=0.01,
    use_gpu=False  # Set to True if GPU is available
)

# Generate posterior samples
posterior_samples = model.generate_posterior_samples(
    adata=adata,
    num_samples=100
)

# Store results in AnnData
adata = model.store_results_in_anndata(
    adata=adata,
    posterior_samples=posterior_samples
)
```

### Creating Custom Models

The modular implementation allows you to create custom models by combining different components:

```python
from pyrovelocity.models.modular.components import (
    StandardDynamicsModel,
    LogNormalPriorModel,
    PoissonLikelihoodModel,
    StandardObservationModel,
    AutoGuideFactory
)
from pyrovelocity.models.modular.model import PyroVelocityModel

# Create components
dynamics_model = StandardDynamicsModel(shared_time=True, t_scale_on=False)
prior_model = LogNormalPriorModel()
likelihood_model = PoissonLikelihoodModel()
observation_model = StandardObservationModel(correct_library_size=True)
guide_model = AutoGuideFactory()

# Create model
custom_model = PyroVelocityModel(
    dynamics_model=dynamics_model,
    prior_model=prior_model,
    likelihood_model=likelihood_model,
    observation_model=observation_model,
    guide_model=guide_model
)

# Train the model
custom_model.train(
    adata=adata,
    max_epochs=1000,
    learning_rate=0.01,
    use_gpu=False
)
```

### Using Factory Functions

The modular implementation provides factory functions for creating models with different configurations:

```python
from pyrovelocity.models.modular.factory import (
    create_standard_model,
    create_model_from_config,
    create_legacy_model1,
    create_legacy_model2
)
from pyrovelocity.models.modular.config import ModelConfig

# Create a standard model
standard_model = create_standard_model()

# Create a model that replicates the legacy implementation
legacy_model = create_legacy_model1()

# Create a model from a configuration
config = ModelConfig.standard()
config_model = create_model_from_config(config)
```

## Inference Parameters

Both implementations support the following inference parameters:

| Parameter | Description | Default |
|-----------|-------------|---------|
| max_epochs | Maximum number of training epochs | 1000 |
| learning_rate | Learning rate for the optimizer | 0.01 |
| batch_size | Batch size for mini-batch training | None (full dataset) |
| train_size | Fraction of data to use for training | 0.8 |
| valid_size | Fraction of data to use for validation | 0.2 |
| early_stopping | Whether to use early stopping | True |
| early_stopping_patience | Number of epochs to wait for improvement | 10 |
| use_gpu | Whether to use GPU for training | "auto" |

## Posterior Sampling

After training, you can generate posterior samples to quantify uncertainty:

```python
# Generate posterior samples
posterior_samples = model.generate_posterior_samples(
    adata=adata,
    num_samples=100
)
```

The posterior samples contain the following keys:

| Key | Description |
|-----|-------------|
| alpha | Transcription rate |
| beta | Splicing rate |
| gamma | Degradation rate |
| u_inf | Steady-state unspliced RNA |
| s_inf | Steady-state spliced RNA |
| t | Latent time |
| switching | Switching time |
| ut | Latent unspliced RNA |
| st | Latent spliced RNA |

## Conclusion

This guide covered the basics of inference with PyroVelocity, including both the legacy and modular implementations. For more advanced usage, refer to the [API reference](../reference/models.PyroVelocityModel.qmd) and the [modular components guide](model_components.qmd).

### 2025-05-15 src/pyrovelocity/models/modular/components/dynamics.py
fix(modular): remove debug print statements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -549,13 +549,7 @@ class LegacyDynamicsModel:
             beta = context["beta"]
             gamma = context["gamma"]
 
-            # Print shapes for debugging
-            print(f"LegacyDynamicsModel - u_obs shape: {u_obs.shape}")
-            print(f"LegacyDynamicsModel - s_obs shape: {s_obs.shape}")
-            print(f"LegacyDynamicsModel - alpha shape: {alpha.shape}")
-            print(f"LegacyDynamicsModel - beta shape: {beta.shape}")
-            print(f"LegacyDynamicsModel - gamma shape: {gamma.shape}")
-
+                                                                        
             # Extract optional values
             t0 = context.get("t0", torch.zeros_like(alpha))
             dt_switching = context.get("dt_switching", torch.zeros_like(alpha))
@@ -669,10 +663,6 @@ class LegacyDynamicsModel:
                 alpha / (gamma - beta + 1e-8)  # Add small epsilon to avoid division by zero
             ) * (torch.exp(-beta * cell_time) - torch.exp(-gamma * cell_time))
 
-            # Print shapes for debugging before reshaping
-            print(f"LegacyDynamicsModel - ut original shape: {ut.shape}")
-            print(f"LegacyDynamicsModel - st original shape: {st.shape}")
-
             # Ensure ut and st have the correct shape [num_samples, n_cells, n_genes]
             # First, determine the expected shape
             if alpha.dim() == 3:
@@ -696,9 +686,7 @@ class LegacyDynamicsModel:
                     st = st.unsqueeze(0).unsqueeze(0)  # [1, 1, n_genes]
 
             # Print shapes after reshaping
-            print(f"LegacyDynamicsModel - ut reshaped: {ut.shape}")
-            print(f"LegacyDynamicsModel - st reshaped: {st.shape}")
-
+                        
             # Ensure ut and st have the correct shape [num_samples, n_cells, n_genes]
             # This is critical for proper shape in posterior samples
             # In the legacy model, ut and st should have shape [num_samples, n_cells, n_genes]
@@ -736,13 +724,7 @@ class LegacyDynamicsModel:
             context["u_expected"] = ut
             context["s_expected"] = st
 
-            # Print shapes for debugging
-            print(f"LegacyDynamicsModel - cell_time shape: {cell_time.shape}")
-            print(f"LegacyDynamicsModel - u_inf shape: {u_inf.shape}")
-            print(f"LegacyDynamicsModel - switching shape: {switching.shape}")
[…]

### 2025-05-15 src/pyrovelocity/models/modular/components/likelihoods.py

@@ -69,12 +69,7 @@ class PoissonLikelihoodModel:
             u_expected = context["u_expected"]
             s_expected = context["s_expected"]
 
-            # Print shapes for debugging
-            print(f"PoissonLikelihoodModel - u_obs shape: {u_obs.shape}")
-            print(f"PoissonLikelihoodModel - s_obs shape: {s_obs.shape}")
-            print(f"PoissonLikelihoodModel - u_expected shape: {u_expected.shape}")
-            print(f"PoissonLikelihoodModel - s_expected shape: {s_expected.shape}")
-
+                                                            
             # Extract optional scaling factors
             u_scale = context.get("u_scale")
             s_scale = context.get("s_scale")
@@ -84,18 +79,15 @@ class PoissonLikelihoodModel:
             s_rate = s_expected
 
             if u_scale is not None:
-                print(f"PoissonLikelihoodModel - u_scale shape: {u_scale.shape}")
-                u_rate = u_rate * u_scale
+                                u_rate = u_rate * u_scale
             if s_scale is not None:
-                print(f"PoissonLikelihoodModel - s_scale shape: {s_scale.shape}")
-                s_rate = s_rate * s_scale
+                                s_rate = s_rate * s_scale
 
             # Get model dimensions if available
             model_n_genes = None
             for param in ["alpha", "beta", "gamma"]:
                 if param in context and isinstance(context[param], torch.Tensor):
                     model_n_genes = context[param].shape[-1]
-                    print(f"PoissonLikelihoodModel - {param} shape: {context[param].shape}")
                     break
 
             # Determine the correct dimensions to use
@@ -115,12 +107,10 @@ class PoissonLikelihoodModel:
             # Reshape u_obs and s_obs if needed
             if u_obs.shape[1] != n_genes:
                 u_obs = u_obs[:, :n_genes]
-                print(f"PoissonLikelihoodModel - Reshaped u_obs to {u_obs.shape}")
-
+                
             if s_obs.shape[1] != n_genes:
                 s_obs = s_obs[:, :n_genes]
-                print(f"PoissonLikelihoodModel - Reshaped s_obs to {s_obs.shape}")
-
+                
             # Ensure observations are integers for Poisson distribution
             u_obs_int = u_obs.round().long()
             s_obs_int = s_obs.round().long()
[…]

### 2025-05-15 src/pyrovelocity/models/modular/components/observations.py

@@ -79,16 +79,11 @@ class StandardObservationModel:
                     "Either u_obs and s_obs or x must be provided in the context"
                 )
 
-        # Log tensor shapes for debugging
-        print(f"StandardObservationModel - u_obs shape: {u_obs.shape}")
-        print(f"StandardObservationModel - s_obs shape: {s_obs.shape}")
-
         # Check for model parameters in context that might have different shapes
         model_n_genes = None
         for param_name in ["alpha", "beta", "gamma"]:
             if param_name in context:
                 model_n_genes = context[param_name].shape[0]
-                print(f"StandardObservationModel - Found {param_name} with shape: {context[param_name].shape}")
                 break
 
         # Calculate library size
@@ -106,23 +101,19 @@ class StandardObservationModel:
         # If there's a shape mismatch between model parameters and data,
         # handle it by reshaping the data to match the model parameters
         if model_n_genes is not None and model_n_genes != u_obs.shape[1]:
-            print(f"WARNING: Shape mismatch between model parameters ({model_n_genes} genes) and data ({u_obs.shape[1]} genes)")
-
+            
             # Determine the minimum number of genes to use
             n_genes = min(model_n_genes, u_obs.shape[1])
             print(f"StandardObservationModel - Reshaping data to use {n_genes} genes")
 
             # Reshape u_obs and s_obs to match the model parameters
             if u_obs.shape[1] > n_genes:
-                print(f"StandardObservationModel - Reshaping u_obs from {u_obs.shape} to use {n_genes} genes")
-                u_obs = u_obs[:, :n_genes]
+                                u_obs = u_obs[:, :n_genes]
 
             if s_obs.shape[1] > n_genes:
-                print(f"StandardObservationModel - Reshaping s_obs from {s_obs.shape} to use {n_genes} genes")
-                s_obs = s_obs[:, :n_genes]
+                                s_obs = s_obs[:, :n_genes]
 
-            print(f"StandardObservationModel - After reshaping: u_obs shape: {u_obs.shape}, s_obs shape: {s_obs.shape}")
-
+            
         # Store the observations and scaling factors
         self.u_obs = u_obs
         self.s_obs = s_obs
@@ -303,12 +294,7 @@ class StandardObservationModel:
         Returns:
             Tuple of observed unspliced and spliced counts.
         """
-        # Print shapes for debugging
[…]

### 2025-05-15 src/pyrovelocity/models/modular/data/anndata.py

@@ -201,7 +201,6 @@ def store_results(
                     try:
                         value_mean = value_mean.reshape(adata.n_vars)
                     except ValueError:
-                        print(f"Could not reshape {key} to match var dimensions")
                         # Still store the original in uns
                         adata_out.uns[f"{model_name}_{key}"] = value_np
                         continue

### 2025-05-15 src/pyrovelocity/models/modular/inference/posterior.py

@@ -91,22 +91,13 @@ def compute_velocity(
     u_scale = posterior_samples.get("u_scale")
     s_scale = posterior_samples.get("s_scale")
 
-    # Print shapes for debugging
-    print(f"compute_velocity - alpha shape: {alpha.shape if alpha is not None else None}")
-    print(f"compute_velocity - beta shape: {beta.shape if beta is not None else None}")
-    print(f"compute_velocity - gamma shape: {gamma.shape if gamma is not None else None}")
-    print(f"compute_velocity - u_scale shape: {u_scale.shape if u_scale is not None else None}")
-    print(f"compute_velocity - s_scale shape: {s_scale.shape if s_scale is not None else None}")
-
+                        
     # Try to get ut and st from posterior samples (like legacy implementation)
     # These are the latent variables, not the observed data
     ut = posterior_samples.get("ut")
     st = posterior_samples.get("st")
 
-    # Print shapes for debugging
-    print(f"compute_velocity - ut shape: {ut.shape if ut is not None else None}")
-    print(f"compute_velocity - st shape: {st.shape if st is not None else None}")
-
+            
     # Convert numpy arrays to torch tensors if needed
     if ut is not None and isinstance(ut, np.ndarray):
         ut = torch.tensor(ut)
@@ -136,24 +127,21 @@ def compute_velocity(
             alpha = alpha.unsqueeze(0).unsqueeze(0)  # [1, 1, n_genes]
         elif alpha.dim() == 2:  # [num_samples, n_genes]
             alpha = alpha.unsqueeze(1)  # [num_samples, 1, n_genes]
-        print(f"Normalized alpha shape: {alpha.shape}")
-
+        
     if beta is not None:
         # Ensure beta has shape [num_samples, 1, n_genes]
         if beta.dim() == 1:  # [n_genes]
             beta = beta.unsqueeze(0).unsqueeze(0)  # [1, 1, n_genes]
         elif beta.dim() == 2:  # [num_samples, n_genes]
             beta = beta.unsqueeze(1)  # [num_samples, 1, n_genes]
-        print(f"Normalized beta shape: {beta.shape}")
-
+        
     if gamma is not None:
         # Ensure gamma has shape [num_samples, 1, n_genes]
         if gamma.dim() == 1:  # [n_genes]
             gamma = gamma.unsqueeze(0).unsqueeze(0)  # [1, 1, n_genes]
         elif gamma.dim() == 2:  # [num_samples, n_genes]
[…]

### 2025-05-15 src/pyrovelocity/models/modular/model.py

@@ -1141,10 +1141,7 @@ class PyroVelocityModel:
                 else adata.layers["spliced"]
             )
 
-            # Print shapes for debugging
-            print(f"generate_posterior_samples - u_obs shape: {u_obs.shape}")
-            print(f"generate_posterior_samples - s_obs shape: {s_obs.shape}")
-
+                                    
             # Run the model predictive with u_obs and s_obs
             # Don't use plate context managers here - they're already in the model
             model_samples = model_predictive(u_obs=u_obs, s_obs=s_obs)

### 2025-05-15 nbs/tutorials/models.qmd
docs(tutorials): rejoin models code blocks and print model structure
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 import matplotlib.pyplot as plt
+import scanpy as sc
 from pyrovelocity.utils import print_anndata, anndata_string, print_string_diff
@@
 pyro.set_rng_seed(42)
-```
 
-```python
 # Create synthetic data
@@
 adata = PyroVelocityModel.setup_anndata(adata)
-final_data_state_representation = anndata_string(adata)
+post_setup_anndata_representation = anndata_string(adata)
 print_string_diff(
     initial_data_state_representation,
-    final_data_state_representation,
+    post_setup_anndata_representation,
     diff_title="Setup AnnData diff",
@@
 print_anndata(adata)
-```
 
-```python
+# Compute PCA and UMAP embeddings
+sc.pp.pca(adata)
+sc.pp.neighbors(adata)
+sc.tl.umap(adata)
+post_embedding_representation = anndata_string(adata)
+print_string_diff(
+    post_setup_anndata_representation,
+    post_embedding_representation,
+    diff_title="Embedding diff",
+)
+print_anndata(adata)
+
 # Create a standard model with default components
 model = create_legacy_model1()
-```
+print(model)
 
-```python
 # Train the model
@@
 )
-```
 
-```python
 # Generate posterior samples
@@
 )
-```
 
-```python
 # Store results in AnnData
@@
 )
-```
 
-```python
 # Print available keys in posterior samples
@@
     print(f"- {key}: {posterior_samples[key].shape}")
+print(model)
 ```
@@
 # Visualize velocity
-scv.pl.velocity_embedding_stream(
+scv.pl.velocity_embedding_grid(
     adata,
@@
     vkey='velocity_pyro'  # Use the velocity key added by get_velocity
+    color="gray",
+    s=3.5,
+    show=False,
+    alpha=0.25,
+    min_mass=3.5,
+    scale=0.35,
+    frameon=False,
+    density=0.4,
+    arrow_size=3,
+    linewidth=1,
 )

### 2025-05-15 scripts/validation/direct_comparison.py
fix(scripts): save outputs to script directory
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -60,7 +60,7 @@ def parse_args():
     parser.add_argument(
         "--output-dir",
         type=str,
-        default="direct_comparison_results",
+        default="scripts/validation/direct_comparison_results",
         help="Directory to save validation results",
     )
     parser.add_argument(

### 2025-05-24 src/pyrovelocity/models/modular/model.py
refactor: merge get_velocity and get_velocity_uncertainty methods
- Remove deprecated get_velocity_uncertainty method
- Enhance get_velocity method to compute both velocity and uncertainty
- Add comprehensive AnnData mutation with all fields matching legacy model
- Return mutated AnnData object for method chaining
- Support optional uncertainty computation with configurable methods
- Add model name prefixes for multi-model workflows
- Ensure scvelo compatibility with proper field naming
- Improve API design with clear documentation of mutations

@@ -1141,7 +1141,7 @@ class PyroVelocityModel:
                 else adata.layers["spliced"]
             )
 
-                                    
+
             # Run the model predictive with u_obs and s_obs
             # Don't use plate context managers here - they're already in the model
             model_samples = model_predictive(u_obs=u_obs, s_obs=s_obs)
@@ -1299,33 +1299,51 @@ class PyroVelocityModel:
     @beartype
     def get_velocity(
         self,
-        adata: Optional[AnnData] = None,
+        adata: AnnData,
         indices: Optional[Sequence[int]] = None,
         num_samples: int = 100,
         basis: str = "umap",
         n_jobs: int = 1,
         random_seed: int = 99,
+        compute_uncertainty: bool = True,
+        uncertainty_method: str = "std",
+        model_name: str = "pyrovelocity",
         **kwargs
-    ) -> np.ndarray:
+    ) -> AnnData:
         """
-        Compute RNA velocity from posterior samples.
+        Compute RNA velocity and uncertainty from posterior samples and store results in AnnData.
 
-        This method computes RNA velocity from posterior samples generated by the model,
-        following the same approach as the legacy implementation's compute_mean_vector_field.
-        It first generates posterior samples, then computes velocity, and finally
-        computes the velocity embedding using scvelo.
+        This method computes RNA velocity and optionally uncertainty from posterior samples
+        generated by the model, following the same approach as the legacy implementation's
+        compute_statistics_from_posterior_samples. It generates posterior samples, computes
+        velocity and uncertainty, stores all results in the AnnData object, and computes
+        velocity embeddings using scvelo.
+
+        The method mutates the input AnnData object by adding the following fields:
+        - adata.layers[f"velocity_{model_name}"] - Mean velocity values
+        - adata.layers[f"spliced_{model_name}"] - Mean posterior spliced counts
+        - adata.var["velocity_genes"] - Boolean indicating velocity genes
+        - adata.obsm[f"velocity_{model_name}_{basis}"] - Embedded velocity vectors
+        - adata.uns[f"velocity_{model_name}_graph"] - Velocity graph
+        - adata.uns[f"velocity_{model_name}_graph_neg"] - Negative velocity graph
+        - adata.uns[f"velocity_{model_name}_params"] - Velocity parameters
+        - adata.obs[f"velocity_{model_name}_self_transition"] - Self-transition probabilities
+        - adata.var[f"velocity_{model_name}_uncertainty"] - Velocity uncertainty (if computed)
 
         Args:
[…]

### 2025-05-24 nbs/tutorials/models.qmd
docs: update models tutorial with bifurcation_14 dataset and unified API
- Replace synthetic data example with realistic bifurcation_14 dataset
- Update to use create_standard_model() as the primary example
- Demonstrate new unified get_velocity() method with uncertainty computation
- Update visualization section to follow review_simulated_datasets.py pattern
- Add proper scvelo velocity_embedding_grid usage with correct parameters
- Show how to access velocity uncertainty statistics
- Update factory functions section with all available model types
- Improve code examples to be more realistic

@@
 
-Let's start with a simple example using the modular implementation:
+Let's start with a realistic example using the bifurcation_14 dataset and the modular implementation:
 
@@
 import pyro
-import anndata as ad
 import numpy as np
@@
 import scanpy as sc
+import scvelo as scv
 from pyrovelocity.utils import print_anndata, anndata_string, print_string_diff
 from pyrovelocity.models.modular import PyroVelocityModel
-from pyrovelocity.models.modular.factory import create_legacy_model1
+from pyrovelocity.models.modular.factory import create_standard_model
+from pyrovelocity.io.datasets import bifurcation_14
 
@@
 
-# Create synthetic data
-n_cells, n_genes = 100, 10
-u_data = np.random.poisson(5, size=(n_cells, n_genes))
-s_data = np.random.poisson(5, size=(n_cells, n_genes))
-
-# Create AnnData object
-adata = ad.AnnData(X=s_data)
-adata.layers["spliced"] = s_data
-adata.layers["unspliced"] = u_data
-adata.obs_names = [f"cell_{i}" for i in range(n_cells)]
-adata.var_names = [f"gene_{i}" for i in range(n_genes)]
+# Load the bifurcation_14 dataset
+# This is a synthetic dataset with a single bifurcation in its differentiation trajectory
+adata = bifurcation_14()
+print("Loaded bifurcation_14 dataset:")
 print_anndata(adata)
@@
 
-# Compute PCA and UMAP embeddings
+# Compute PCA and UMAP embeddings if not already present
+if 'X_pca' not in adata.obsm:
     sc.pp.pca(adata)
+if 'X_umap' not in adata.obsm:
     sc.pp.neighbors(adata)
@@
 # Create a standard model with default components
-model = create_legacy_model1()
+model = create_standard_model()
 print(model)
@@
     adata=adata,
-    max_epochs=20,  # Use a small number for demonstration
+    max_epochs=50,  # Use more epochs for better results
     learning_rate=0.01,
@@
 
-# Generate posterior samples
-posterior_samples = model.generate_posterior_samples(
+# Generate posterior samples and compute velocity in one step
+# The new unified get_velocity method computes both velocity and uncertainty
+adata_with_velocity = model.get_velocity(
     adata=adata,
-    num_samples=30  # Use a small number for demonstration
+    num_samples=100,  # Use more samples for better uncertainty estimates
+    basis='umap',
+    compute_uncertainty=True,
+    uncertainty_method='std',
+    model_name='pyrovelocity'
 )
 
-# Store results in AnnData
-adata = model.store_results_in_anndata(
-    adata=adata,
-    posterior_samples=posterior_samples
+# Print the updated AnnData structure
+print("AnnData after velocity computation:")
+print_anndata(adata_with_velocity)
+
+# Show what fields were added
+post_velocity_representation = anndata_string(adata_with_velocity)
+print_string_diff(
+    post_embedding_representation,
+    post_velocity_representation,
+    diff_title="Velocity computation diff",
 )
-
-# Compute velocity and store it in AnnData
-velocity = model.get_velocity(
-    adata=adata,
-    num_samples=30,
-    basis='umap'  # This will be used for embedding
-)
-
-# Print available keys in posterior samples
-print("Available keys in posterior samples:")
-for key in posterior_samples.keys():
-    print(f"- {key}: {posterior_samples[key].shape}")
-print(model)
 ```
@@
 
-You can create custom models by combining different components:
+You can create custom models by combining different components using the registry system:
 
 ```python
-from pyrovelocity.models.modular.components import (
-    StandardDynamicsModel,
-    LegacyDynamicsModel,
-    LogNormalPriorModel,
-    PoissonLikelihoodModel,
-    LegacyLikelihoodModel,
-    StandardObservationModel,
-    AutoGuideFactory
+from pyrovelocity.models.modular.registry import (
+    DynamicsModelRegistry,
+    PriorModelRegistry,
+    LikelihoodModelRegistry,
+    ObservationModelRegistry,
+    InferenceGuideRegistry
 )
@@
 
-# Create components
-dynamics_model = LegacyDynamicsModel(shared_time=True, t_scale_on=False)
-prior_model = LogNormalPriorModel()
-likelihood_model = LegacyLikelihoodModel()
-observation_model = StandardObservationModel(correct_library_size=True)
-guide_model = AutoGuideFactory()
+# Create components using the registry
+dynamics_model = DynamicsModelRegistry.create("legacy", shared_time=True, t_scale_on=False)
+prior_model = PriorModelRegistry.create("lognormal")
+likelihood_model = LikelihoodModelRegistry.create("legacy")
+observation_model = ObservationModelRegistry.create("standard", correct_library_size=True)
+guide_model = InferenceGuideRegistry.create("auto")
 
@@
 custom_model.train(
-    adata=adata,
-    max_epochs=20,
+    adata=adata_with_velocity,  # Use the data with embeddings
+    max_epochs=50,
     learning_rate=0.01,
     use_gpu=False
+)
+
+# Compute velocity with the custom model
+custom_adata = custom_model.get_velocity(
+    adata=adata_with_velocity,
+    num_samples=100,
+    basis='umap',
+    compute_uncertainty=True,
+    model_name='custom_legacy'
 )
@@
 from pyrovelocity.models.modular.factory import (
+    create_standard_model,
     create_model_from_config,
@@
 
-# Create a model that replicates the legacy implementation
-legacy_model = create_legacy_model1()
+# Create a standard model (recommended for new projects)
+standard_model = create_standard_model()
+
+# Create a model that replicates the legacy implementation (add_offset=False)
+legacy_model1 = create_legacy_model1()
+
+# Create a model that replicates the legacy implementation (add_offset=True)
+legacy_model2 = create_legacy_model2()
 
@@
 config_model = create_model_from_config(config)
+
+# Train and use any of these models
+legacy_model1.train(adata=adata_with_velocity, max_epochs=50)
+
+# The new unified get_velocity method returns the mutated AnnData object
+adata_legacy = legacy_model1.get_velocity(
+    adata=adata_with_velocity,
+    num_samples=100,
+    basis='umap',
+    compute_uncertainty=True,
+    model_name='legacy1'
+)
+
+print("Available velocity layers:")
+for layer in adata_legacy.layers.keys():
+    if 'velocity' in layer:
+        print(f"- {layer}")
 ```
@@
 
-After training the model and generating posterior samples, you can visualize the results:
+After training the model and computing velocity, you can visualize the results using scvelo's plotting functions. The velocity data is stored in the format expected by scvelo:
 
 ```python
-import scanpy as sc
-import scvelo as scv
+import matplotlib.pyplot as plt
 
-# Compute PCA and UMAP embeddings
-sc.pp.pca(adata)
-sc.pp.neighbors(adata)
-sc.tl.umap(adata)
+# Add leiden clustering for visualization
+sc.tl.leiden(adata_with_velocity)
 
-# Visualize velocity
+# Set up the plot
+fig, axes = plt.subplots(1, 2, figsize=(12, 5))
+
+# Plot 1: Standard velocity embedding
 scv.pl.velocity_embedding_grid(
-    adata,
+    adata_with_velocity,
     basis='umap',
-    color='gene_0',
-    vkey='velocity_pyro'  # Use the velocity key added by get_velocity
-    color="gray",
-    s=3.5,
+    color='leiden',
+    title='PyroVelocity Standard Model',
+    vkey='velocity_pyrovelocity',  # Use the velocity key added by get_velocity
+    s=15,
+    alpha=1,
+    linewidth=0.5,
+    ax=axes[0],
     show=False,
-    alpha=0.25,
-    min_mass=3.5,
-    scale=0.35,
-    frameon=False,
+    legend_loc='right margin',
     density=0.4,
-    arrow_size=3,
-    linewidth=1,
+    scale=0.2,
+    arrow_size=2,
+    arrow_length=2,
+    arrow_color='black',
 )
+
+# Plot 2: Legacy model comparison (if computed)
+if 'velocity_legacy1' in adata_legacy.layers:
+    scv.pl.velocity_embedding_grid(
+        adata_legacy,
+        basis='umap',
+        color='leiden',
+        title='PyroVelocity Legacy Model',
+        vkey='velocity_legacy1',
+        s=15,
+        alpha=1,
+        linewidth=0.5,
+        ax=axes[1],
+        show=False,
+        legend_loc='right margin',
+        density=0.4,
+        scale=0.2,
+        arrow_size=2,
+        arrow_length=2,
+        arrow_color='black',
+    )
+else:
+    axes[1].text(0.5, 0.5, 'Legacy model\nnot computed',
+                ha='center', va='center', transform=axes[1].transAxes)
+    axes[1].axis('off')
+
+plt.tight_layout()
+plt.show()
+
+# Display uncertainty information if computed
+if 'velocity_pyrovelocity_uncertainty' in adata_with_velocity.var.columns:
+    print("Velocity uncertainty statistics:")
+    uncertainty = adata_with_velocity.var['velocity_pyrovelocity_uncertainty']
+    print(f"Mean uncertainty: {uncertainty.mean():.4f}")
+    print(f"Std uncertainty: {uncertainty.std():.4f}")
+    print(f"Min uncertainty: {uncertainty.min():.4f}")
+    print(f"Max uncertainty: {uncertainty.max():.4f}")
 ```

### 2025-05-24 nbs/tutorials/models.qmd
docs: update models tutorial to exclusively use create_legacy_model2
- Replace create_standard_model with create_legacy_model2 throughout tutorial
- Update all model names and velocity keys to use 'legacy2' naming
- Emphasize that create_legacy_model2 (add_offset=True) replicates original PyroVelocity
- Update visualization section to show legacy model results
- Maintain consistency with legacy model focus across all examples
- Update uncertainty field references to match legacy2 model naming

@@
 from pyrovelocity.models.modular import PyroVelocityModel
-from pyrovelocity.models.modular.factory import create_standard_model
+from pyrovelocity.models.modular.factory import create_legacy_model2
 from pyrovelocity.io.datasets import bifurcation_14
@@
 
-# Create a standard model with default components
-model = create_standard_model()
+# Create a legacy model (add_offset=True) that replicates the original PyroVelocity implementation
+model = create_legacy_model2()
 print(model)
@@
     uncertainty_method='std',
-    model_name='pyrovelocity'
+    model_name='legacy2'
 )
@@
 
-# Train the model
+# Train the custom legacy model
 custom_model.train(
@@
 
-# Compute velocity with the custom model
+# Compute velocity with the custom legacy model
 custom_adata = custom_model.get_velocity(
@@
     compute_uncertainty=True,
-    model_name='custom_legacy'
+    model_name='custom_legacy2'
 )
@@
 from pyrovelocity.models.modular.factory import (
-    create_standard_model,
     create_model_from_config,
@@
 
-# Create a standard model (recommended for new projects)
-standard_model = create_standard_model()
+# The primary model we use is create_legacy_model2 (add_offset=True)
+# This replicates the original PyroVelocity implementation most closely
+primary_model = create_legacy_model2()
 
-# Create a model that replicates the legacy implementation (add_offset=False)
-legacy_model1 = create_legacy_model1()
+# Alternative: create_legacy_model1 (add_offset=False)
+alternative_model = create_legacy_model1()
 
-# Create a model that replicates the legacy implementation (add_offset=True)
-legacy_model2 = create_legacy_model2()
-
-# Create a model from a configuration
+# You can also create models from configurations
 config = ModelConfig.standard()
@@
 
-# Train and use any of these models
-legacy_model1.train(adata=adata_with_velocity, max_epochs=50)
+# Train and use the primary legacy model
+primary_model.train(adata=adata_with_velocity, max_epochs=50)
 
 # The new unified get_velocity method returns the mutated AnnData object
-adata_legacy = legacy_model1.get_velocity(
+adata_primary = primary_model.get_velocity(
     adata=adata_with_velocity,
@@
     compute_uncertainty=True,
-    model_name='legacy1'
+    model_name='primary_legacy2'
 )
@@
 print("Available velocity layers:")
-for layer in adata_legacy.layers.keys():
+for layer in adata_primary.layers.keys():
     if 'velocity' in layer:
@@
 
-# Plot 1: Standard velocity embedding
+# Plot 1: Primary legacy model (create_legacy_model2)
 scv.pl.velocity_embedding_grid(
@@
     color='leiden',
-    title='PyroVelocity Standard Model',
-    vkey='velocity_pyrovelocity',  # Use the velocity key added by get_velocity
+    title='PyroVelocity Legacy Model 2 (add_offset=True)',
+    vkey='velocity_legacy2',  # Use the velocity key added by get_velocity
     s=15,
@@
 
-# Plot 2: Legacy model comparison (if computed)
-if 'velocity_legacy1' in adata_legacy.layers:
+# Plot 2: Alternative model comparison (if computed)
+if 'velocity_primary_legacy2' in adata_primary.layers:
     scv.pl.velocity_embedding_grid(
-        adata_legacy,
+        adata_primary,
         basis='umap',
         color='leiden',
-        title='PyroVelocity Legacy Model',
-        vkey='velocity_legacy1',
+        title='PyroVelocity Primary Legacy Model',
+        vkey='velocity_primary_legacy2',
         s=15,
@@
 else:
-    axes[1].text(0.5, 0.5, 'Legacy model\nnot computed',
+    axes[1].text(0.5, 0.5, 'Alternative model\nnot computed',
                 ha='center', va='center', transform=axes[1].transAxes)
@@
 # Display uncertainty information if computed
-if 'velocity_pyrovelocity_uncertainty' in adata_with_velocity.var.columns:
+if 'velocity_legacy2_uncertainty' in adata_with_velocity.var.columns:
     print("Velocity uncertainty statistics:")
-    uncertainty = adata_with_velocity.var['velocity_pyrovelocity_uncertainty']
+    uncertainty = adata_with_velocity.var['velocity_legacy2_uncertainty']
     print(f"Mean uncertainty: {uncertainty.mean():.4f}")

### 2025-05-24 src/pyrovelocity/models/modular/model.py
fix(modular): handle uncertainty shape mismatch in get_velocity method
- Fix ValueError when storing uncertainty in adata.var due to shape mismatch
- Add proper shape handling for uncertainty arrays in get_velocity method
- Support both cell-gene level (stored in layers) and gene-level (stored in var) uncertainty
- Handle different uncertainty shapes: [num_cells, num_genes], [num_genes], [num_cells]
- Take mean across cells when storing gene-level uncertainty in adata.var
- Maintain backward compatibility with existing uncertainty computation
- Resolves issue where uncertainty with shape (300,

@@ -1432,8 +1432,27 @@ class PyroVelocityModel:
             elif not isinstance(uncertainty, np.ndarray):
                 uncertainty = np.array(uncertainty)
 
-            # Store uncertainty in var (gene-level information)
-            adata.var[f"velocity_{model_name}_uncertainty"] = uncertainty
+            # Handle uncertainty shape for storage in adata.var (gene-level information)
+            if uncertainty.ndim == 2:
+                # If uncertainty is [num_cells, num_genes], take mean across cells for gene-level storage
+                uncertainty_gene_level = uncertainty.mean(axis=0)
+                # Also store the full uncertainty in layers for cell-gene level information
+                adata.layers[f"velocity_{model_name}_uncertainty"] = uncertainty
+            elif uncertainty.ndim == 1 and uncertainty.shape[0] == adata.n_vars:
+                # If uncertainty is already [num_genes], use directly
+                uncertainty_gene_level = uncertainty
+            elif uncertainty.ndim == 1 and uncertainty.shape[0] == adata.n_obs:
+                # If uncertainty is [num_cells], broadcast to genes (unusual case)
+                uncertainty_gene_level = np.full(adata.n_vars, uncertainty.mean())
+                adata.layers[f"velocity_{model_name}_uncertainty"] = uncertainty.reshape(-1, 1)
+            else:
+                raise ValueError(
+                    f"Uncertainty shape {uncertainty.shape} does not match AnnData dimensions "
+                    f"({adata.n_obs}, {adata.n_vars})"
+                )
+
+            # Store gene-level uncertainty in var
+            adata.var[f"velocity_{model_name}_uncertainty"] = uncertainty_gene_level
 
         # Compute velocity graph and embedding using scvelo
         scv.tl.velocity_graph(

### 2025-05-24 scripts/parquet/convert_h5ad_parquet.py
chore(scripts): remove incompatible field
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -26,6 +26,12 @@ def convert_h5ad_to_pqdata(
     """
     adata = read_h5ad(h5ad_file)
     print_anndata(adata)
+
+    # remove 3D array that causes pqdata serialization issue
+    if 'true_sc_grn' in adata.uns:
+        del adata.uns['true_sc_grn']
+        print("Removed true_sc_grn from uns in {h5ad_file.name}")
+
     write_anndata(
         data=adata,
         path=pq_path,
@@ -392,7 +398,7 @@ def analyze_parquet_database_ibis(pq_path: Path, data_set_name: str) -> Path:
     return db_path
 
 def upload_to_huggingface_hub(
-    data_set_path: Path, 
+    data_set_path: Path,
     repo_id: str = "pyrovelocity/fixtures",
 ) -> RepoUrl:
     """
@@ -407,9 +413,9 @@ def upload_to_huggingface_hub(
     """
     api = HfApi()
     repourl = api.create_repo(
-        repo_id=repo_id, 
+        repo_id=repo_id,
         private=False,
-        repo_type="dataset", 
+        repo_type="dataset",
         exist_ok=True,
     )
     api.upload_large_folder(

### 2025-05-24 nbs/tutorials/models.qmd
docs(models): enable execution
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-```python
+```{python}
 import torch
@@
 
-```python
+```{python}
 from pyrovelocity.models.modular.registry import (
@@
 
-```python
+```{python}
 from pyrovelocity.models.modular.factory import (
@@
 
-```python
+```{python}
 import matplotlib.pyplot as plt
@@
 # Set up the plot
-fig, axes = plt.subplots(1, 2, figsize=(12, 5))
+fig, axes = plt.subplots()
 
@@
     linewidth=0.5,
-    ax=axes[0],
+    ax=axes,
     show=False,
@@
 )
-
-# Plot 2: Alternative model comparison (if computed)
-if 'velocity_primary_legacy2' in adata_primary.layers:
-    scv.pl.velocity_embedding_grid(
-        adata_primary,
-        basis='umap',
-        color='leiden',
-        title='PyroVelocity Primary Legacy Model',
-        vkey='velocity_primary_legacy2',
-        s=15,
-        alpha=1,
-        linewidth=0.5,
-        ax=axes[1],
-        show=False,
-        legend_loc='right margin',
-        density=0.4,
-        scale=0.2,
-        arrow_size=2,
-        arrow_length=2,
-        arrow_color='black',
-    )
-else:
-    axes[1].text(0.5, 0.5, 'Alternative model\nnot computed',
-                ha='center', va='center', transform=axes[1].transAxes)
-    axes[1].axis('off')
 

### 2025-05-27 nbs/guides/inference.qmd
docs: update inference guide to use bifurcation_14 dataset and create_legacy_model2 components
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 from pyrovelocity.models import PyroVelocity
+from pyrovelocity.io.datasets import bifurcation_14
 
 # Load data
-adata = ad.read_h5ad("path/to/data.h5ad")
+adata = bifurcation_14()
 
@@
 from pyrovelocity.models.modular import (
-    create_standard_model,
+    create_legacy_model2,
     PyroVelocityModel
 )
+from pyrovelocity.io.datasets import bifurcation_14
 
 # Load data
-adata = ad.read_h5ad("path/to/data.h5ad")
+adata = bifurcation_14()
 
@@
 # Create model using factory function
-model = create_standard_model()
+model = create_legacy_model2()
 
@@
 ```python
-from pyrovelocity.models.modular.components import (
-    StandardDynamicsModel,
-    LogNormalPriorModel,
-    PoissonLikelihoodModel,
-    StandardObservationModel,
-    AutoGuideFactory
+from pyrovelocity.models.modular.factory import ComponentFactory
+from pyrovelocity.models.modular.config import ComponentConfig, ComponentType
+from pyrovelocity.models.modular.model import PyroVelocityModel
+from pyrovelocity.io.datasets import bifurcation_14
+
+# Load data
+adata = bifurcation_14()
+adata = PyroVelocityModel.setup_anndata(adata)
+
+# Create components consistent with create_legacy_model2
+dynamics_model = ComponentFactory.create_component(
+    ComponentConfig(
+        name="legacy",
+        params={
+            "shared_time": True,
+            "t_scale_on": False,
+            "correct_library_size": True,
+            "add_offset": True,
+        }
+    ),
+    ComponentType.DYNAMICS_MODEL
 )
-from pyrovelocity.models.modular.model import PyroVelocityModel
 
-# Create components
-dynamics_model = StandardDynamicsModel(shared_time=True, t_scale_on=False)
-prior_model = LogNormalPriorModel()
-likelihood_model = PoissonLikelihoodModel()
-observation_model = StandardObservationModel(correct_library_size=True)
-guide_model = AutoGuideFactory()
+prior_model = ComponentFactory.create_component(
+    ComponentConfig(name="lognormal", params={}),
+    ComponentType.PRIOR_MODEL
+)
+
+likelihood_model = ComponentFactory.create_component(
+    ComponentConfig(name="legacy", params={}),
+    ComponentType.LIKELIHOOD_MODEL
+)
+
+observation_model = ComponentFactory.create_component(
+    ComponentConfig(name="standard", params={}),
+    ComponentType.OBSERVATION_MODEL
+)
+
+guide_model = ComponentFactory.create_component(
+    ComponentConfig(
+        name="legacy_auto",
+        params={
+            "init_scale": 0.1,
+            "add_offset": True,
+        }
+    ),
+    ComponentType.INFERENCE_GUIDE
+)
 
@@
 from pyrovelocity.models.modular.config import ModelConfig
+from pyrovelocity.models.modular.model import PyroVelocityModel
+from pyrovelocity.io.datasets import bifurcation_14
 
-# Create a standard model
-standard_model = create_standard_model()
+# Load data
+adata = bifurcation_14()
+adata = PyroVelocityModel.setup_anndata(adata)
 
-# Create a model that replicates the legacy implementation
-legacy_model = create_legacy_model1()
+# Create a model that replicates the legacy implementation (add_offset=True)
+legacy_model2 = create_legacy_model2()
+
+# Create an alternative legacy model (add_offset=False)
+legacy_model1 = create_legacy_model1()
 
@@
 ```python
+from pyrovelocity.io.datasets import bifurcation_14
+from pyrovelocity.models.modular.factory import create_legacy_model2
+from pyrovelocity.models.modular.model import PyroVelocityModel
+
+# Load data and set up model
+adata = bifurcation_14()
+adata = PyroVelocityModel.setup_anndata(adata)
+model = create_legacy_model2()
+
+# Train the model first
+model.train(adata=adata, max_epochs=50, learning_rate=0.01, use_gpu=False)
+
 # Generate posterior samples

### 2025-05-27 nbs/guides/inference.qmd
docs: preprocess data in inference guide
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 from pyrovelocity.models import PyroVelocity
+from pyrovelocity.tasks.preprocess import preprocess_dataset
 from pyrovelocity.io.datasets import bifurcation_14
@@
 adata = bifurcation_14()
+adata, _, _ = preprocess_dataset("bifurcation_14", adata)
 
@@
 model = PyroVelocity(
-    adata,
-    model_type="normal",  # Options: "normal", "deterministic", "stochastic"
-    guide_type="auto",    # Options: "auto", "delta"
-    likelihood="Poisson", # Options: "Poisson", "NegativeBinomial"
-    shared_time=True,     # Whether to use shared time across genes
-    t_scale_on=False      # Whether to use time scaling
+    adata=adata,
+    likelihood="Poisson",
+    model_type="auto",
+    guide_type="auto",
+    correct_library_size=True,
+    add_offset=True,
 )
@@
 # Train model
-model.train(
+model.train_faster(
     max_epochs=1000,
-    learning_rate=0.01,
+    lr=0.01,
     use_gpu="auto"
@@
     adata=adata,
-    num_samples=100
+    num_samples=5,
 )
@@
 # Save model
-model.save_model("path/to/save/model")
+model.save_model("models/bifurcation_14_model2/model")
 
 # Load model
-model = PyroVelocity.load_model("path/to/save/model", adata)
+model = PyroVelocity.load_model("models/bifurcation_14_model2/model", adata)
 ```
@@
 from pyrovelocity.models.modular import (
-    create_legacy_model2,
     PyroVelocityModel
 )
+from pyrovelocity.models.modular.factory import create_legacy_model2
+from pyrovelocity.tasks.preprocess import preprocess_dataset
 from pyrovelocity.io.datasets import bifurcation_14
+from pyrovelocity.utils import print_anndata
 
@@
 )
+print_anndata(adata)
 ```

### 2025-05-27 src/pyrovelocity/models/experimental/__init__.py
fix(experimental): update exports
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -7,12 +7,13 @@ including JAX-based deterministic inference, simulation, and hyperparameter cali
 
 from pyrovelocity.models.experimental.deterministic_inference import (
     deterministic_transcription_splicing_probabilistic_model,
-    generate_test_data_for_deterministic_model_inference,
-    generate_prior_inference_data,
     generate_posterior_inference_data,
+    generate_prior_inference_data,
+    generate_test_data_for_deterministic_model_inference,
     plot_sample_phase_portraits,
     plot_sample_trajectories,
     plot_sample_trajectories_with_percentiles,
+    print_inference_data_structure,
     save_inference_plots,
 )
 from pyrovelocity.models.experimental.deterministic_simulation import (
@@ -21,8 +22,8 @@ from pyrovelocity.models.experimental.deterministic_simulation import (
 )
 from pyrovelocity.models.experimental.hyperparameter_calibration import (
     lognormal_tail_probability,
-    solve_for_lognormal_sigma_given_threshold_and_tail_mass,
     solve_for_lognormal_mu_given_threshold_and_tail_mass,
+    solve_for_lognormal_sigma_given_threshold_and_tail_mass,
 )
 
 __all__ = [
@@ -34,6 +35,7 @@ __all__ = [
     "plot_sample_phase_portraits",
     "plot_sample_trajectories",
     "plot_sample_trajectories_with_percentiles",
+    "print_inference_data_structure",
     "save_inference_plots",
     # Deterministic simulation
     "solve_transcription_splicing_model",

### 2025-05-27 nbs/concepts/inferenceindynamicalsystems/index.qmd
docs(inferenceindynamicalsystems): update imports from experimental models package
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 import diffrax
-from pyrovelocity.models import solve_transcription_splicing_model
-from pyrovelocity.models import solve_transcription_splicing_model_analytical
+from pyrovelocity.models.experimental import ( 
+    solve_transcription_splicing_model,
+    solve_transcription_splicing_model_analytical,
+)
 from pyrovelocity.logging import configure_logging
@@
 import arviz as az
-from pyrovelocity.models._deterministic_inference import (
+from pyrovelocity.models.experimental import (
     generate_posterior_inference_data,

### 2025-05-27 nbs/concepts/parameter-recovery-validation/index.qmd
docs: add parameter recovery validation concept guide
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Parameter Recovery Validation
toc: true
number-sections: true
highlight-style: gruvbox
csl: ../../bibstyle.csl
lightbox: auto
format:
  html:
    html-math-method: mathjax
    include-in-header: ../../mathjax.html
format-links: [ipynb]
execute:
  freeze: true
  eval: true
  warning: false
  error: false
  cache: true
author:
  - name: Pyrovelocity Team
    email: team@pyrovelocity.net
abstract: |
  This notebook demonstrates parameter recovery validation for a lattice of PyroVelocity models,
  progressing from dimensioned models with analytical solutions and observed time to numerical simulation with unobserved time.
  We focus on models that generate and fit count data while respecting the symmetries identified in dimensionless analysis,
  synthesizing the approach from BayVel (Sabbioni et al.) with PyroVelocity's legacy implementation.
  We illustrate how to systematically validate that probabilistic models can recover
  known parameters from synthetic data, which is essential before applying models
  to real-world data where true parameter values are unknown.
keywords: [single-cell transcriptomics, probabilistic modeling, parameter recovery, model validation, RNA velocity]
bibliography: ../../references.bib
jupyter:
  jupytext:
    cell_metadata_filter: all
    cell_metadata_json: true
    notebook_metadata_filter: all
    text_representation:
      extension: .qmd
      format_name: quarto
      format_version: 1.0
      jupytext_version: 1.16.0
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
  language_info:
    name: python
  rise:
    scroll: true
    theme: black
  toc-autonumbering: true
  toc-showcode: false
  toc-showmarkdowntxt: false
---

## Introduction

Parameter recovery validation is a critical step in validating probabilistic models before applying them to real data. This process involves:

1. **Generating synthetic data** using known parameter values
2. **Running inference** to recover parameters from the synthetic data
3. **Evaluating recovery performance**
4. **Assessing model identifiability**

This notebook demonstrates parameter recovery validation for PyroVelocity models, starting with a dimensioned model that generates and fits count data while respecting the symmetries identified in dimensionless analysis. We synthesize the approach from BayVel (Sabbioni et al. 2025) regarding Poisson product distributions with PyroVelocity's legacy implementation, progressing from analytical solutions with observed time to numerical simulation with latent time. For each model we include a graphical representation of the model's generative process as well as a mathematical description of the prior, observation model, and likelihood.

## Model Progression

We will validate parameter recovery across a progression of models which prepare for situations where time is not observed and the dynamics are simulated rather than required to have analytical solutions.

### Analytical Dynamics with Observed Time

This model represents a dimensioned version that generates and fits count data while respecting symmetries:

- **Time coordinates**: Observed (not latent)
- **Dynamics**: Analytical transcription-splicing-degradation solutions in dimensional units
- **Parameters**: α (transcription rate, molecules/cell·hr), β (splicing rate, hr⁻¹), γ (degradation rate, hr⁻¹)
- **Prior**: Log-normal for positive rate parameters
- **Observation Model**: Poisson counts of actual RNA molecules
- **Likelihood**: Poisson product distribution (following Jahnke 2007)
- **Identifiability**: Respects symmetries from dimensionless analysis

### Analytical Dynamics with Latent Time

This model introduces latent time coordinates:

- **Time coordinates**: Latent with priors (similar to PyroVelocity)
- **Dynamics**: Same analytical solutions
- **Parameters**: α, β, γ, plus time coordinates
- **Prior**: Log-normal for positive parameters
- **Observation Model**: Standard observation model
- **Likelihood**: Poisson

### Simulated Dynamics with Observed Time

This model replaces the analytical solutions with numerical integration using torchode:

- **Time coordinates**: Latent with priors
- **Dynamics**: Numerical integration with torchode
- **Parameters**: α, β, γ, plus time coordinates
- **Prior**: Log-normal for positive parameters
- **Observation Model**: Standard observation model
- **Likelihood**: Poisson

### Simulated Dynamics with Latent Time

This model replaces the analytical solutions with numerical integration using torchode:

- **Time coordinates**: Latent with priors
- **Dynamics**: Numerical integration with torchode
- **Parameters**: α, β, γ, plus time coordinates
- **Prior**: Log-normal for positive parameters
- **Observation Model**: Standard observation model
- **Likelihood**: Poisson

## Parameter Recovery Validation Framework

```{python}
#| label: setup-imports
#| code-fold: true

import numpy as np
import torch
import pyro
import matplotlib.pyplot as plt
import seaborn as sns
from pathlib import Path

# PyroVelocity imports
from pyrovelocity.models.modular.factory import (
    create_simple_deterministic_model_with_validation,
    create_legacy_model2,
)
from pyrovelocity.models.modular.components.validation import (
    ParameterRecoveryConfig,
)
from pyrovelocity.io.datasets import bifurcation_14

# Set random seeds for reproducibility
torch.manual_seed(42)
pyro.set_rng_seed(42)
np.random.seed(42)

print("Setup complete!")
```

## Validation Study 1: Analytical dynamics with observed time

This validation study focuses on a dimensioned model that generates and fits count data while respecting the symmetries identified in the dimensionless analysis. We follow the approach from our [Probabilistic inference in dynamical systems](../inferenceindynamicalsystems/index.qmd) concept guide, providing both a probabilistic graphical model and mathematical description.

### Model Description

We model the transcription-splicing-degradation system using the dimensioned variables and parameters with their physical units (molecules/cell, hr⁻¹, etc.), but with Poisson observations of the actual RNA counts. This approach synthesizes the dimensionless analysis for identifiability with the count-based observation model used in PyroVelocity.

The model generates count data for unspliced ($u$) and spliced ($s$) RNA molecules per cell, with the dynamics governed by the analytical solutions derived in the dimensionless case but scaled back to dimensional units.

### Probabilistic Graphical Model

```{python}
#| label: fig-count-model-pgm
#| code-fold: true
#| fig-cap: Probabilistic graphical model for the dimensioned transcription-splicing-degradation model with Poisson count observations and observed time coordinates.

import daft
import matplotlib.pyplot as plt

plt.rcParams["font.family"] = "serif"
plt.rcParams["font.size"] = 16
plt.rcParams["text.usetex"] = True

pgm = daft.PGM(line_width=1.2)

# hyperparameters
pgm.add_node("mu_init", r"$\mu_{0}$", 0.5, 6, fixed=True)
pgm.add_node("sigma_init", r"$\sigma_{0}^2$", 1.5, 6, fixed=True)
pgm.add_node("mu_theta", r"$\mu_{\boldsymbol\theta}$", 2.5, 6, fixed=True)
pgm.add_node("sigma_theta", r"$\sigma_{\boldsymbol\theta}^2$", 3.5, 6, fixed=True)
pgm.add_node("mu_sigma", r"$\mu_{\sigma}$", 4.5, 6, fixed=True)
pgm.add_node("sigma_sigma", r"$\sigma_{\sigma}^2$", 5.5, 6, fixed=True)

# latent variables for gene-specific parameters
pgm.add_node("u_0i", r"$u_{0i}$", 1, 5)
pgm.add_node("s_0i", r"$s_{0i}$", 2, 5)
pgm.add_node("theta_i", r"$\boldsymbol\theta_i$", 3, 5)
pgm.add_node("sigma_ui", r"$\sigma_{ui}$", 4, 5)
pgm.add_node("sigma_si", r"$\sigma_{si}$", 5, 5)

# latent variables for cell-specific outcomes
pgm.add_node(
    "u_ij",
    r"$u_{ij}$",
    2,
    4,
    scale=1.0,
    shape="rectangle",
)
pgm.add_node(
    "s_ij",
    r"$s_{ij}$",
    4,
    4,
    scale=1.0,
    shape="rectangle",
)

# observed data
pgm.add_node(
    "t_j",
    r"$t_j$",
    6.0,
    3.25,
    observed=True,
    shape="rectangle",
)
pgm.add_node(
    "u_obs_ij",
    r"$\hat{u}_{ij}$",
    2,
    2.5,
    scale=1.0,
    observed=True,
)
pgm.add_node(
    "s_obs_ij",
    r"$\hat{s}_{ij}$",
    4,
    2.5,
    scale=1.0,
    observed=True,
)

# edges
edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}
pgm.add_edge("mu_init", "u_0i", plot_params=edge_params)
pgm.add_edge("sigma_init", "u_0i", plot_params=edge_params)
pgm.add_edge("mu_init", "s_0i", plot_params=edge_params)
pgm.add_edge("sigma_init", "s_0i", plot_params=edge_params)
pgm.add_edge("mu_theta", "theta_i", plot_params=edge_params)
pgm.add_edge("sigma_theta", "theta_i", plot_params=edge_params)
pgm.add_edge("mu_sigma", "sigma_ui", plot_params=edge_params)
pgm.add_edge("sigma_sigma", "sigma_ui", plot_params=edge_params)
pgm.add_edge("mu_sigma", "sigma_si", plot_params=edge_params)
pgm.add_edge("sigma_sigma", "sigma_si", plot_params=edge_params)

pgm.add_edge("u_0i", "u_ij", plot_params=edge_params)
pgm.add_edge("s_0i", "s_ij", plot_params=edge_params)
pgm.add_edge("u_0i", "s_ij", plot_params=edge_params)
pgm.add_edge("theta_i", "s_ij", plot_params=edge_params)

pgm.add_edge("u_ij", "u_obs_ij", plot_params=edge_params)
pgm.add_edge("s_ij", "s_obs_ij", plot_params=edge_params)
pgm.add_edge("sigma_ui", "u_obs_ij", plot_params=edge_params)
pgm.add_edge("sigma_si", "s_obs_ij", plot_params=edge_params)

pgm.add_edge("t_j", "u_ij", plot_params=edge_params)
pgm.add_edge("t_j", "s_ij", plot_params=edge_params)

# plates
pgm.add_plate(
    [0.5, 1.2, 5, 4.4],
    label=r"$i \in \{1, \ldots, G\}$",
    shift=-0.1,
    fontsize=12,
)
pgm.add_plate(
    [1.0, 1.8, 5.5, 2.75],
    label=r"$j \in \{1, \ldots, N\}$",
    shift=-0.1,
    fontsize=12,
)

pgm.render()
```

### Mathematical Description

The model is defined by the following hierarchical structure:

**Priors for dimensional parameters:**
\begin{align}
    \alpha_i &\sim \text{LogNormal}(\mu_{\alpha}, \sigma_{\alpha}^2) \label{eq-alpha-prior-dim}, \\
    \beta_i &\sim \text{LogNormal}(\mu_{\beta}, \sigma_{\beta}^2) \label{eq-beta-prior-dim}, \\
    \gamma_i &\sim \text{LogNormal}(\mu_{\gamma}, \sigma_{\gamma}^2) \label{eq-gamma-prior-dim}
\end{align}

**Initial conditions (steady state):**
\begin{align}
    u_{0i} &= \frac{\alpha_i}{\beta_i} \label{eq-u0-steady}, \\
    s_{0i} &= \frac{\alpha_i}{\gamma_i} \label{eq-s0-steady}
\end{align}

**Deterministic dynamics (analytical solution):**
\begin{align}
    u^t_{ij} &= u_{0i} + (\alpha_i/\beta_i - u_{0i}) e^{-\beta_i (t_j - t_{0i})} \label{eq-ut-analytical}, \\
    s^t_{ij} &= s_{0i} + \frac{\alpha_i}{\gamma_i}(1 - e^{-\gamma_i (t_j - t_{0i})}) + \frac{\alpha_i - \beta_i u_{0i}}{\gamma_i - \beta_i}(e^{-\gamma_i (t_j - t_{0i})} - e^{-\beta_i (t_j - t_{0i})}) \label{eq-st-analytical}
\end{align}

**Poisson observation model:**
\begin{align}
    u_{ij} &\sim \text{Poisson}(u^t_{ij}) \label{eq-u-obs-poisson}, \\
    s_{ij} &\sim \text{Poisson}(s^t_{ij}) \label{eq-s-obs-poisson}
\end{align}

where:

- $i \in \{1, \ldots, G\}$ indexes genes
- $j \in \{1, \ldots, N\}$ indexes cells
- $t_j$ are observed time coordinates for each cell
- $t_{0i}$ are gene-specific initial times (can be set to 0 for simplicity)

This model respects the identifiability constraints identified in the dimensionless analysis while working directly with count data as in the PyroVelocity legacy implementation. The Poisson observation model follows the approach of Jahnke (2007), ensuring that if the initial distribution is a product of Poisson distributions, the transient distributions remain products of Poisson distributions centered at the deterministic solution.

### Parameter Recovery Validation

```{python}
#| label: dimensional-count-validation
#| code-fold: true

# Create the simple deterministic model with validation functionality
model = create_simple_deterministic_model_with_validation()

# Configure validation parameters
validation_config = ParameterRecoveryConfig(
    num_parameter_sets=5,       # Number of parameter sets to test
    sample_sizes=[50, 100, 200], # Different sample sizes
    noise_levels=[0.1, 0.2],    # Different noise levels
    inference_algorithm="SVI",   # Use SVI for speed
    num_samples=1000,           # Number of inference samples
    seed=42,                    # Random seed
)

# Run parameter recovery validation using registry-based validation components
report = model.validate_parameter_recovery(validation_config)

print(f"Validation completed!")
print(f"Tested {len(report['parameter_sets'])} parameter sets")
print(f"Overall parameter recovery success rate: {report['summary']['overall_success_rate']:.2%}")

# Quick validation for development
quick_results = model.quick_validation(num_genes=3, num_cells=50, noise_level=0.1)
print(f"Quick validation success rate: {quick_results['summary']['overall_success_rate']:.2%}")

# List available validation components
print(f"Available validation components: {model.list_validation_components()}")
```

## Validation Study 2: Anayltical dynamics with unobserved time

## Validation Study 3: Simulated dynamics with observed time

## Validation Study 4: Simulated dynamics with unobserved time

### 2025-05-27 nbs/concepts/parameter-recovery-validation/index.qmd
docs: base parameter recovery validation on dimensionless model
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 abstract: |
-  This notebook demonstrates parameter recovery validation for a lattice of PyroVelocity models,
-  progressing from dimensioned models with analytical solutions and observed time to numerical simulation with unobserved time.
-  We focus on models that generate and fit count data while respecting the symmetries identified in dimensionless analysis,
-  synthesizing the approach from BayVel (Sabbioni et al.) with PyroVelocity's legacy implementation.
+  This notebook demonstrates parameter recovery validation for PyroVelocity models using the direct dimensionless approach,
+  progressing from analytical solutions with observed time to numerical simulation with latent time coordinates.
+  We focus on models that naturally avoid identifiability issues while maintaining biological interpretability,
+  using hierarchical priors on characteristic scales and lumped effective capture efficiency.
   We illustrate how to systematically validate that probabilistic models can recover
@@
   to real-world data where true parameter values are unknown.
-keywords: [single-cell transcriptomics, probabilistic modeling, parameter recovery, model validation, RNA velocity]
+keywords: [single-cell transcriptomics, probabilistic modeling, model calibration, model validation, RNA velocity]
 bibliography: ../../references.bib
@@
 
-This notebook demonstrates parameter recovery validation for PyroVelocity models, starting with a dimensioned model that generates and fits count data while respecting the symmetries identified in dimensionless analysis.
-We synthesize the approach from BayVel (Sabbioni et al.
-2025) regarding Poisson product distributions with PyroVelocity's legacy implementation, progressing from analytical solutions with observed time to numerical simulation with latent time.
+This notebook demonstrates parameter recovery validation for PyroVelocity models using the dimensionless parameterization that naturally avoids identifiability issues while maintaining biological interpretability.
+We implement hierarchical priors on characteristic scales and lumped effective capture efficiency, progressing from analytical solutions with observed time to numerical simulation with latent time coordinates.
 For each model we include a graphical representation of the model's generative process as well as a mathematical description of the prior, observation model, and likelihood.
@@
 
-We will validate parameter recovery across a progression of models which prepare for situations where time is not observed and the dynamics are simulated rather than required to have analytical solutions.
+We validate parameter recovery across a progression of **dimensionless models** that naturally avoid identifiability issues while maintaining biological interpretability.
+Each model builds complexity while preserving the core dimensionless parameterization.
 
-### Analytical Dynamics with Observed Time
+### Dimensionless Analytical Dynamics with Observed Time
 
-This model represents a dimensioned version that generates and fits count data while respecting symmetries:
+This model uses the direct dimensionless approach with analytical solutions:
 
-- **Time coordinates**: Observed (not latent)
-- **Dynamics**: Analytical transcription-splicing-degradation solutions in dimensional units
-- **Parameters**: α (transcription rate, molecules/cell·hr), β (splicing rate, hr⁻¹), γ (degradation rate, hr⁻¹)
-- **Prior**: Log-normal for positive rate parameters
-- **Observation Model**: Poisson counts of actual RNA molecules
-- **Likelihood**: Poisson product distribution (following Jahnke 2007)
-- **Identifiability**: Respects symmetries from dimensionless analysis
+- **Time coordinates**: Observed dimensionless time $t^*_j$
+- **Dynamics**: Analytical dimensionless transcription-splicing-degradation solutions
+- **Parameters**: $\gamma^*_i$ (relative degradation), $U_{0i}$ (characteristic scale), $\lambda_j$ (lumped capture efficiency)
+- **Prior**: Log-normal hierarchical priors for all parameters
+- **Observation Model**: Poisson counts scaled by characteristic scales and capture efficiency
+- **Likelihood**: Poisson product distribution
+- **Identifiability**: Naturally identifiable through dimensionless parameterization
 
-### Analytical Dynamics with Latent Time
+### Dimensionless Analytical Dynamics with Latent Time
 
-This model introduces latent time coordinates:
+This model introduces latent dimensionless time coordinates:
 
-- **Time coordinates**: Latent with priors (similar to PyroVelocity)
-- **Dynamics**: Same analytical solutions
-- **Parameters**: α, β, γ, plus time coordinates
-- **Prior**: Log-normal for positive parameters
-- **Observation Model**: Standard observation model
+- **Time coordinates**: Latent $t^*_j$ with priors
+- **Dynamics**: Same dimensionless analytical solutions
+- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus latent time coordinates
+- **Prior**: Log-normal hierarchical priors
+- **Observation Model**: Same Poisson scaling approach
 - **Likelihood**: Poisson
 
-### Simulated Dynamics with Observed Time
+### Dimensionless Simulated Dynamics with Observed Time
 
-This model replaces the analytical solutions with numerical integration using torchode:
+This model replaces analytical solutions with numerical integration:
 
-- **Time coordinates**: Latent with priors
-- **Dynamics**: Numerical integration with torchode
-- **Parameters**: α, β, γ, plus time coordinates
-- **Prior**: Log-normal for positive parameters
-- **Observation Model**: Standard observation model
+- **Time coordinates**: Observed dimensionless time $t^*_j$
+- **Dynamics**: Numerical integration of dimensionless ODEs using torchode
+- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$
+- **Prior**: Log-normal hierarchical priors
+- **Observation Model**: Same Poisson scaling approach
 - **Likelihood**: Poisson
 
-### Simulated Dynamics with Latent Time
+### Dimensionless Simulated Dynamics with Latent Time
 
-This model replaces the analytical solutions with numerical integration using torchode:
+This model represents the full complexity of the PyroVelocity modular implementation:
 
-- **Time coordinates**: Latent with priors
-- **Dynamics**: Numerical integration with torchode
-- **Parameters**: α, β, γ, plus time coordinates
-- **Prior**: Log-normal for positive parameters
-- **Observation Model**: Standard observation model
+- **Time coordinates**: Latent $t^*_j$ with priors
+- **Dynamics**: Numerical integration of dimensionless ODEs using torchode
+- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus latent time coordinates
+- **Prior**: Log-normal hierarchical priors
+- **Observation Model**: Same Poisson scaling approach
 - **Likelihood**: Poisson
@@
 
-## Validation Study 1: Analytical dynamics with observed time
+## Validation Study 1: Dimensionless analytical dynamics with observed time
 
-This validation study focuses on a dimensioned model that generates and fits count data while respecting the symmetries identified in the dimensionless analysis.
-We follow the approach from our [Probabilistic inference in dynamical systems](../inferenceindynamicalsystems/index.qmd) concept guide, providing both a probabilistic graphical model and mathematical description.
+This validation study focuses on the **direct dimensionless approach** that naturally avoids identifiability issues while maintaining biological interpretability.
+We follow the theoretical foundation from our [Probabilistic inference in dynamical systems](../inferenceindynamicalsystems/index.qmd) concept guide, implementing hierarchical priors on characteristic scales and lumped effective capture efficiency.
 
@@
 
-We model the transcription-splicing-degradation system using the dimensioned variables and parameters with their physical units (molecules/cell, hr⁻¹, etc.), but with Poisson observations of the actual RNA counts.
-This approach synthesizes the dimensionless analysis for identifiability with the count-based observation model used in PyroVelocity.
+We model the transcription-splicing-degradation system using the **direct dimensionless approach** that naturally avoids identifiability issues while maintaining biological interpretability.
+This approach fits the dimensionless model directly with hierarchical priors on characteristic scales, eliminating the need for post-estimation rescaling procedures.
 
-The model generates count data for unspliced ($u$) and spliced ($s$) RNA molecules per cell, with the dynamics governed by the analytical solutions derived in the dimensionless case but scaled back to dimensional units.
+The model generates count data for unspliced ($u$) and spliced ($s$) RNA molecules per cell, with dynamics governed by the dimensionless analytical solutions and observation model that scales back to dimensional counts through characteristic scales and effective capture efficiency.
 
 ### Probabilistic Graphical Model
+
+The probabilistic graphical model represents the dimensionless transcription-splicing-degradation system with hierarchical priors on characteristic scales and lumped effective capture efficiency.
+The model directly fits dimensionless parameters while maintaining biological interpretability through characteristic scales.
 
@@
 #| code-fold: true
-#| fig-cap: Probabilistic graphical model for the dimensioned transcription-splicing-degradation model with Poisson count observations and observed time coordinates.
+#| fig-cap: Probabilistic graphical model for the dimensionless transcription-splicing-degradation model with hierarchical priors on characteristic scales and lumped effective capture efficiency.
 
@@
 pgm.add_node("sigma_init", r"$\sigma_{0}^2$", 1.5, 6, fixed=True)
-pgm.add_node("mu_theta", r"$\mu_{\boldsymbol\theta}$", 2.5, 6, fixed=True)
-pgm.add_node("sigma_theta", r"$\sigma_{\boldsymbol\theta}^2$", 3.5, 6, fixed=True)
-pgm.add_node("mu_sigma", r"$\mu_{\sigma}$", 4.5, 6, fixed=True)
-pgm.add_node("sigma_sigma", r"$\sigma_{\sigma}^2$", 5.5, 6, fixed=True)
+pgm.add_node("mu_gamma", r"$\mu_{\gamma^*}$", 2.5, 6, fixed=True)
+pgm.add_node("sigma_gamma", r"$\sigma_{\gamma^*}^2$", 3.5, 6, fixed=True)
+pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 6, fixed=True)
+pgm.add_node("sigma_U0", r"$\sigma_{U_0}^2$", 5, 6, fixed=True)
+pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 5.5, 5, fixed=True)
+pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 6, 5, fixed=True)
 
 # latent variables for gene-specific parameters
-pgm.add_node("u_0i", r"$u_{0i}$", 1, 5)
-pgm.add_node("s_0i", r"$s_{0i}$", 2, 5)
-pgm.add_node("theta_i", r"$\boldsymbol\theta_i$", 3, 5)
-pgm.add_node("sigma_ui", r"$\sigma_{ui}$", 4, 5)
-pgm.add_node("sigma_si", r"$\sigma_{si}$", 5, 5)
+pgm.add_node("u_star_0i", r"$u^{\ast}_{0i}$", 1, 5)
+pgm.add_node("s_star_0i", r"$s^{\ast}_{0i}$", 2, 5)
+pgm.add_node("gamma_star_i", r"$\gamma^{\ast}_i$", 3, 5)
+pgm.add_node("U_0i", r"$U_{0i}$", 4.5, 5)
+
+# cell-specific parameters (moved inside cell plate)
+pgm.add_node("lambda_j", r"$\lambda_j$", 5.5, 4)
 
@@
 pgm.add_node(
-    "u_ij",
-    r"$u_{ij}$",
+    "u_star_ij",
+    r"$u^{\ast}_{ij}$",
     2,
@@
 pgm.add_node(
-    "s_ij",
-    r"$s_{ij}$",
+    "s_star_ij",
+    r"$s^{\ast}_{ij}$",
     4,
@@
 pgm.add_node(
-    "t_j",
-    r"$t_j$",
-    6.0,
-    3.25,
+    "t_star_j",
+    r"$t^{\ast}_j$",
+    5.5,
+    3.0,
     observed=True,
@@
     "u_obs_ij",
-    r"$\hat{u}_{ij}$",
+    r"$u_{ij}$",
     2,
@@
     "s_obs_ij",
-    r"$\hat{s}_{ij}$",
+    r"$s_{ij}$",
     4,
@@
 edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}
-pgm.add_edge("mu_init", "u_0i", plot_params=edge_params)
-pgm.add_edge("sigma_init", "u_0i", plot_params=edge_params)
-pgm.add_edge("mu_init", "s_0i", plot_params=edge_params)
-pgm.add_edge("sigma_init", "s_0i", plot_params=edge_params)
-pgm.add_edge("mu_theta", "theta_i", plot_params=edge_params)
-pgm.add_edge("sigma_theta", "theta_i", plot_params=edge_params)
-pgm.add_edge("mu_sigma", "sigma_ui", plot_params=edge_params)
-pgm.add_edge("sigma_sigma", "sigma_ui", plot_params=edge_params)
-pgm.add_edge("mu_sigma", "sigma_si", plot_params=edge_params)
-pgm.add_edge("sigma_sigma", "sigma_si", plot_params=edge_params)
+pgm.add_edge("mu_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
+pgm.add_edge("sigma_U0", "U_0i", plot_params=edge_params)
+pgm.add_edge("mu_lambda", "lambda_j", plot_params=edge_params)
+pgm.add_edge("sigma_lambda", "lambda_j", plot_params=edge_params)
 
-pgm.add_edge("u_0i", "u_ij", plot_params=edge_params)
-pgm.add_edge("s_0i", "s_ij", plot_params=edge_params)
-pgm.add_edge("u_0i", "s_ij", plot_params=edge_params)
-pgm.add_edge("theta_i", "s_ij", plot_params=edge_params)
+pgm.add_edge("u_star_0i", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("s_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("u_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)
 
-pgm.add_edge("u_ij", "u_obs_ij", plot_params=edge_params)
-pgm.add_edge("s_ij", "s_obs_ij", plot_params=edge_params)
-pgm.add_edge("sigma_ui", "u_obs_ij", plot_params=edge_params)
-pgm.add_edge("sigma_si", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("U_0i", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("U_0i", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("lambda_j", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("lambda_j", "s_obs_ij", plot_params=edge_params)
 
-pgm.add_edge("t_j", "u_ij", plot_params=edge_params)
-pgm.add_edge("t_j", "s_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)
 
@@
 pgm.add_plate(
-    [0.5, 1.2, 5, 4.4],
+    [0.5, 1.4, 4.5, 4.2],
     label=r"$i \in \{1, \ldots, G\}$",
@@
 pgm.add_plate(
-    [1.0, 1.8, 5.5, 2.75],
+    [1.0, 1.9, 5.0, 2.6],
     label=r"$j \in \{1, \ldots, N\}$",
@@
 
-The model is defined by the following hierarchical structure:
+The model is defined by the following hierarchical structure using the **direct dimensionless approach**:
 
-**Priors for dimensional parameters:**
+**Priors for dimensionless initial conditions:**
 \begin{align}
-    \alpha_i &\sim \text{LogNormal}(\mu_{\alpha}, \sigma_{\alpha}^2) \label{eq-alpha-prior-dim}, \\
-    \beta_i &\sim \text{LogNormal}(\mu_{\beta}, \sigma_{\beta}^2) \label{eq-beta-prior-dim}, \\
-    \gamma_i &\sim \text{LogNormal}(\mu_{\gamma}, \sigma_{\gamma}^2) \label{eq-gamma-prior-dim}
+    u^*_{0i}, s^*_{0i} &\sim \text{LogNormal}(\mu_{0}, \sigma_{0}^2) \label{eq-dimless-init-conds-priors}
 \end{align}
 
-**Initial conditions (steady state):**
+**Priors for dimensionless kinetic parameters:**
 \begin{align}
-    u_{0i} &= \frac{\alpha_i}{\beta_i} \label{eq-u0-steady}, \\
-    s_{0i} &= \frac{\alpha_i}{\gamma_i} \label{eq-s0-steady}
+    \gamma^*_i &\sim \text{LogNormal}(\mu_{\gamma^*}, \sigma_{\gamma^*}^2) \label{eq-gamma-star-prior}
 \end{align}
 
-**Deterministic dynamics (analytical solution):**
+where $\gamma^*_i = \gamma_i / \beta_i$ is the relative degradation rate for gene $i$.
+
+**Priors for characteristic scales:**
 \begin{align}
-    u^t_{ij} &= u_{0i} + (\alpha_i/\beta_i - u_{0i}) e^{-\beta_i (t_j - t_{0i})} \label{eq-ut-analytical}, \\
-    s^t_{ij} &= s_{0i} + \frac{\alpha_i}{\gamma_i}(1 - e^{-\gamma_i (t_j - t_{0i})}) + \frac{\alpha_i - \beta_i u_{0i}}{\gamma_i - \beta_i}(e^{-\gamma_i (t_j - t_{0i})} - e^{-\beta_i (t_j - t_{0i})}) \label{eq-st-analytical}
+    U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2) \label{eq-U0-prior}
 \end{align}
 
-**Poisson observation model:**
+where $U_{0i} = \alpha_i / \beta_i$ is the characteristic concentration scale for gene $i$.
+
+**Priors for effective capture efficiency:**
 \begin{align}
-    u_{ij} &\sim \text{Poisson}(u^t_{ij}) \label{eq-u-obs-poisson}, \\
-    s_{ij} &\sim \text{Poisson}(s^t_{ij}) \label{eq-s-obs-poisson}
+    \lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2) \label{eq-lambda-prior}
+\end{align}
+
+where $\lambda_j$ represents the **lumped effective capture efficiency** for cell $j$.
+
+**Dimensionless differential equations:**
+The system is governed by the dimensionless equations:
+\begin{align}
+    \frac{du^*}{dt^*} &= 1 - u^* \label{eq-dimless-u-ode}, \\
+    \frac{ds^*}{dt^*} &= u^* - \gamma^* s^* \label{eq-dimless-s-ode}
+\end{align}
+
+where $t^* = \beta t$, $u^* = u/U_0$, $s^* = s/U_0$, and $\gamma^* = \gamma/\beta$.
+
+**Analytical solutions:**
+Given dimensionless initial conditions $u^*_{0i}, s^*_{0i}$ at time $t^*_0 = 0$, we define the helper variable:
+\begin{align}
+    \xi_i &= \frac{u^*_{0i} - 1}{\gamma^*_i - 1} \label{eq-xi-helper}
+\end{align}
+
+The complete solutions are:
+
+\begin{align}
+    u^*_{ij} &= 1 + (u^*_{0i} - 1) e^{-t^*_j} \\
+    {s^{\ast}}^{k}_{ij} &= 
+      \begin{cases}
+        \frac{1}{\gamma^{\ast}_i} + 
+        \left( s^{\ast}_{0i} - \xi_i - \frac{1}{\gamma^{\ast}_i} \right) \cdot e^{-\gamma^{\ast}_i {t^{\ast}}^k_j} + 
+        \xi_i \cdot e^{-{t^{\ast}}^k_j},& \gamma^{\ast}_i \neq 1 \\
+
+        1 + (s^{\ast}_{0i} - 1) e^{-{t^{\ast}}^k_j} + 
+          (u^{\ast}_{0i} - 1) {t^{\ast}}^k_j e^{-{t^{\ast}}^k_j},& \gamma^{\ast}_i = 1 \\
+      \end{cases}, \label{eq-s-star-model-cell}
+\end{align}
+
+**Poisson observation model with characteristic scaling:**
+\begin{align}
+    u_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}) \label{eq-dimless-u-obs-poisson}, \\
+    s_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij}) \label{eq-dimless-s-obs-poisson}
 \end{align}
@@
 - $j \in \{1, \ldots, N\}$ indexes cells
-- $t_j$ are observed time coordinates for each cell
-- $t_{0i}$ are gene-specific initial times (can be set to 0 for simplicity)
+- $t^*_j$ are observed dimensionless time coordinates for each cell
+- $u^*_{0i}, s^*_{0i}$ are dimensionless initial conditions for each gene
+- $u^*_{ij}, s^*_{ij}$ are the latent dimensionless RNA concentrations
+- $U_{0i}$ is the characteristic concentration scale for gene $i$
+- $\lambda_j$ is the lumped effective capture efficiency for cell $j$
 
-This model respects the identifiability constraints identified in the dimensionless analysis while working directly with count data as in the PyroVelocity legacy implementation.
-The Poisson observation model follows the approach of Jahnke (2007), ensuring that if the initial distribution is a product of Poisson distributions, the transient distributions remain products of Poisson distributions centered at the deterministic solution.
+### Parameter Interpretability
+
+The dimensionless approach provides **natural identifiability** and **clear biological interpretation** for all inferred parameters:
+
+#### **Core Dimensionless Parameters (Gene-Specific)**
+
+**$\gamma^*_i$ (Relative degradation rate)**
+
+- **Definition**: $\gamma^*_i = \gamma_i / \beta_i$
+- **Interpretation**: How fast mRNA degrades relative to how fast pre-mRNA gets spliced
+- **Biological meaning**:
+  - $\gamma^*_i < 1$: mRNA is more stable than pre-mRNA processing time
+  - $\gamma^*_i = 1$: mRNA degradation matches splicing timescale (special case)
+  - $\gamma^*_i > 1$: mRNA degrades faster than splicing occurs
+- **Special case**: When $\gamma^*_i = 1$, the analytical solution has a different form with a $t^*_j e^{-t^*_j}$ term
+- **Helper variable**: $\xi_i = \frac{u^*_{0i} - 1}{\gamma^*_i - 1}$ (undefined when $\gamma^*_i = 1$)
+- **Example**: $\gamma^*_i = 0.5$ means mRNA half-life is twice as long as the characteristic splicing time
+
+#### **Characteristic Scales (Gene-Specific)**
+
+**$U_{0i}$ (Characteristic concentration scale)**
+
+- **Definition**: $U_{0i} = \alpha_i / \beta_i$
+- **Units**: molecules/cell
+- **Interpretation**: The steady-state pre-mRNA level that would be reached if there were no degradation
+- **Biological meaning**: Represents the gene's "expression capacity" - how many molecules it can produce per splicing timescale
+- **Example**: $U_{0i} = 100$ molecules/cell means this gene's transcription-splicing balance produces ~100 molecules per characteristic time
+
+#### **Effective Capture Efficiency (Cell-Specific)**
+
+**$\lambda_j$ (Lumped effective capture efficiency)**
+
+- **Definition**: Cell-specific composite technical efficiency
+- **Units**: Dimensionless (positive real number)
+- **Interpretation**: Overall efficiency of detecting RNA molecules in cell $j$
+- **Technical factors lumped together**:
+  1. **True capture efficiency**: Fraction of molecules actually captured during library prep
+  2. **Library size effects**: Total sequencing depth for this cell
+  3. **Batch effects**: Systematic differences between experimental batches
+  4. **Cell quality**: Cell viability, membrane integrity, RNA degradation
+  5. **Ambient RNA**: Background contamination (modeled as scaling factor)
+  6. **Technical dropout**: Stochastic detection failures
+- **Example**: $\lambda_j = 0.5$ indicates this cell has 50% of the expected detection efficiency compared to a "reference" cell
+
+#### **Derived Biological Quantities**
+
+From the inferred parameters, you can reconstruct the original dimensional parameters:
+
+**$\alpha_i$ (Transcription rate)**
+
+- **Formula**: $\alpha_i = U_{0i} / T_{0i}$ (where $T_{0i} = 1/\beta_i$ is inferred from time scale)
+- **Units**: molecules/(cell·hour)
+- **Interpretation**: How many pre-mRNA molecules are transcribed per hour
+
+**$\beta_i$ (Splicing rate)**
+
+- **Formula**: $\beta_i = 1 / T_{0i}$ (from dimensionless time scale)
+- **Units**: hour⁻¹
+- **Interpretation**: Splicing rate constant
+
+**$\gamma_i$ (Degradation rate)**
+
+- **Formula**: $\gamma_i = \gamma^*_i \cdot \beta_i$
+- **Units**: hour⁻¹
+- **Interpretation**: mRNA degradation rate constant
+
+### Justification for Parameterization
+
+This parameterization is designed to optimally account for identifiability and known biological/technical considerations:
+
+#### **Natural Identifiability**
+
+- **Dimensionless approach**: Eliminates scaling symmetries inherent in dimensional models
+- **No post-estimation rescaling**: Parameters are naturally identifiable during inference
+- **Hierarchical priors**: Provide regularization without breaking identifiability
+
+#### **Biological Interpretability**
+
+- **$\gamma^*_i$**: Directly interpretable relative degradation rate
+- **$U_{0i}$**: Gene expression capacity with clear biological meaning
+- **Dimensionless solutions**: Universal dynamics independent of absolute scales
+
+#### **Technical Factor Management**
+
+- **Strategic lumping**: $\lambda_j$ captures multiple technical effects without overparameterization
+- **Computational efficiency**: Fewer parameters than separate modeling of each technical factor
+- **Robustness**: Works across different experimental protocols and platforms
+
+#### **Connection to Established Theory**
+
+- **Dimensional analysis**: Follows Buckingham π theorem for parameter reduction
+- **RNA velocity theory**: Maintains connection to established dynamical systems approach
+- **Single-cell best practices**: Accounts for known technical factors in scRNA-seq data
+
+This approach attempts to balance **biological interpretability**, **statistical identifiability**, and **practical implementation**.
 
@@
 
-## Validation Study 2: Anayltical dynamics with unobserved time
+## Validation Study 2: Dimensionless analytical dynamics with latent time
 
-## Validation Study 3: Simulated dynamics with observed time
+*To be implemented: Same dimensionless approach but with latent time coordinates that must be inferred along with other parameters.*
 
-## Validation Study 4: Simulated dynamics with unobserved time
+## Validation Study 3: Dimensionless simulated dynamics with observed time
+
+*To be implemented: Replace analytical solutions with numerical integration using torchode while maintaining the dimensionless parameterization.*
+
+## Validation Study 4: Dimensionless simulated dynamics with latent time
+
+*To be implemented: Full complexity model with numerical integration and latent time coordinates, representing the complete PyroVelocity modular implementation.*

### 2025-05-28 nbs/concepts/parameter-recovery-validation/index.qmd
docs: add latent time prior to parameter recovery validation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-- **Time coordinates**: Latent $t^*_j$ with priors
+- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
 - **Dynamics**: Same dimensionless analytical solutions
-- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus latent time coordinates
-- **Prior**: Log-normal hierarchical priors
+- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{max}^*$, $t_{loc}$, $t_{scale}$, $\tilde{t}_j$)
+- **Prior**: Gamma priors for time scales, Normal for relative positions
 - **Observation Model**: Same Poisson scaling approach
@@
 
-- **Time coordinates**: Latent $t^*_j$ with priors
+- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
 - **Dynamics**: Numerical integration of dimensionless ODEs using torchode
-- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus latent time coordinates
-- **Prior**: Log-normal hierarchical priors
+- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{max}^*$, $t_{loc}$, $t_{scale}$, $\tilde{t}_j$)
+- **Prior**: Gamma priors for time scales, Normal for relative positions
 - **Observation Model**: Same Poisson scaling approach
@@
 
-*To be implemented: Same dimensionless approach but with latent time coordinates that must be inferred along with other parameters.*
+This validation study extends the dimensionless approach to include **latent time coordinates** that must be inferred along with other parameters.
+We implement a hierarchical time structure that maintains biological interpretability while ensuring statistical identifiability.
+
+### Model Description
+
+The model uses the same dimensionless dynamics and observation model as Validation Study 1, but replaces observed time coordinates with **hierarchical latent time coordinates**.
+This approach borrows strength across cells while allowing individual temporal variation.
+
+### Probabilistic Graphical Model
+
+The probabilistic graphical model now includes hierarchical priors for latent dimensionless time coordinates, with time parameters becoming random variables rather than observed quantities.
+
+```{python}
+#| label: fig-latent-time-pgm
+#| code-fold: true
+#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with hierarchical latent time coordinates.
+
+import daft
+import matplotlib.pyplot as plt
+
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["font.size"] = 16
+plt.rcParams["text.usetex"] = True
+
+pgm = daft.PGM(line_width=1.2)
+
+# hyperparameters for gene-specific parameters
+pgm.add_node("mu_init", r"$\mu_{0}$", 0.5, 6, fixed=True)
+pgm.add_node("sigma_init", r"$\sigma_{0}^2$", 1.5, 6, fixed=True)
+pgm.add_node("mu_gamma", r"$\mu_{\gamma^*}$", 2.5, 6, fixed=True)
+pgm.add_node("sigma_gamma", r"$\sigma_{\gamma^*}^2$", 3.5, 6, fixed=True)
+pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 6, fixed=True)
+pgm.add_node("sigma_U0", r"$\sigma_{U_0}^2$", 5, 6, fixed=True)
+
+# hyperparameters for hierarchical time structure (moved further left)
+pgm.add_node("mu_T", r"$\mu_{T}$", 5.5, 6, fixed=True)
+pgm.add_node("sigma_T", r"$\sigma_{T}^2$", 6, 6, fixed=True)
+pgm.add_node("mu_t_loc", r"$\mu_{t_{loc}}$", 6.5, 6, fixed=True)
+pgm.add_node("sigma_t_loc", r"$\sigma_{t_{loc}}^2$", 7, 6, fixed=True)
+pgm.add_node("mu_t_scale", r"$\mu_{t_{scl}}$", 7.5, 6, fixed=True)
+pgm.add_node("sigma_t_scale", r"$\sigma_{t_{scl}}^2$", 8, 6, fixed=True)
+
+# hyperparameters for lambda (entering from right edge of cell plate)
+pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 6.5, 3.5, fixed=True)
+pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 6.5, 2.5, fixed=True)
+
+# latent variables for gene-specific parameters
+pgm.add_node("u_star_0i", r"$u^{\ast}_{0i}$", 1, 5)
+pgm.add_node("s_star_0i", r"$s^{\ast}_{0i}$", 2, 5)
+pgm.add_node("gamma_star_i", r"$\gamma^{\ast}_i$", 3, 5)
+pgm.add_node("U_0i", r"$U_{0i}$", 4.5, 5)
+
+# hierarchical time parameters (moved further left)
+pgm.add_node("T_max_star", r"$T_{M}^*$", 5.5, 5)
+pgm.add_node("t_loc", r"$t_{loc}$", 6.25, 5)
+pgm.add_node("t_scale", r"$t_{scl}$", 7, 5)
+
+# latent time coordinates and cell-specific parameters (moved further left, lambda enters from right)
+pgm.add_node("t_star_j", r"$t^{\ast}_j$", 5.5, 4)
+pgm.add_node("lambda_j", r"$\lambda_j$", 5.5, 3)
+
+# latent variables for cell-specific outcomes (centered between rows)
+pgm.add_node(
+    "u_star_ij",
+    r"$u^{\ast}_{ij}$",
+    2,
+    3.75,
+    scale=1.0,
+    shape="rectangle",
+)
+pgm.add_node(
+    "s_star_ij",
+    r"$s^{\ast}_{ij}$",
+    4,
+    3.75,
+    scale=1.0,
+    shape="rectangle",
+)
+
+# observed data
+pgm.add_node(
+    "u_obs_ij",
+    r"$u_{ij}$",
+    2,
+    2.25,
+    scale=1.0,
+    observed=True,
+)
+pgm.add_node(
+    "s_obs_ij",
+    r"$s_{ij}$",
+    4,
+    2.25,
+    scale=1.0,
+    observed=True,
+)
+
+# edges
+edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}
+
+# gene-specific parameter edges
+pgm.add_edge("mu_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "u_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("sigma_init", "s_star_0i", plot_params=edge_params)
+pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
+pgm.add_edge("sigma_U0", "U_0i", plot_params=edge_params)
+pgm.add_edge("mu_lambda", "lambda_j", plot_params=edge_params)
+pgm.add_edge("sigma_lambda", "lambda_j", plot_params=edge_params)
+
+# hierarchical time structure edges
+pgm.add_edge("mu_T", "T_max_star", plot_params=edge_params)
+pgm.add_edge("sigma_T", "T_max_star", plot_params=edge_params)
+pgm.add_edge("mu_t_loc", "t_loc", plot_params=edge_params)
+pgm.add_edge("sigma_t_loc", "t_loc", plot_params=edge_params)
+pgm.add_edge("mu_t_scale", "t_scale", plot_params=edge_params)
+pgm.add_edge("sigma_t_scale", "t_scale", plot_params=edge_params)
+
+# time hierarchy edges
+pgm.add_edge("T_max_star", "t_star_j", plot_params=edge_params)
+pgm.add_edge("t_loc", "t_star_j", plot_params=edge_params)
+pgm.add_edge("t_scale", "t_star_j", plot_params=edge_params)
+
+# dynamics edges
+pgm.add_edge("u_star_0i", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("s_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("u_star_0i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)
+
+# observation edges
+pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("U_0i", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("U_0i", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("lambda_j", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("lambda_j", "s_obs_ij", plot_params=edge_params)
+
+# plates
+pgm.add_plate(
+    [0.5, 1.25, 4.5, 4.25],
+    label=r"$i \in \{1, \ldots, G\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+pgm.add_plate(
+    [1.5, 1.75, 4.5, 2.75],
+    label=r"$j \in \{1, \ldots, N\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+
+pgm.render()
+```
+
+### Mathematical Description
+
+The model extends the dimensionless approach with **hierarchical latent time coordinates**.
+The key modification is replacing observed time $t^*_j$ with latent time variables that follow a hierarchical prior structure.
+
+**Priors for dimensionless parameters (unchanged from Study 1):**
+\begin{align}
+    u^*_{0i}, s^*_{0i} &\sim \text{LogNormal}(\mu_{0}, \sigma_{0}^2) \\
+    \gamma^*_i &\sim \text{LogNormal}(\mu_{\gamma^*}, \sigma_{\gamma^*}^2) \\
+    U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2) \\
+    \lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
+\end{align}
+
+**Hierarchical priors for latent time coordinates:**
+\begin{align}
+    T_{max}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \label{eq-T-max-prior} \\
+    t_{loc} &\sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}) \label{eq-t-loc-prior} \\
+    t_{scale} &\sim \text{Gamma}(\alpha_{t_{scale}}, \beta_{t_{scale}}) \label{eq-t-scale-prior} \\
+    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scale}^2) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-t-tilde-prior} \\
+    t^*_j &= T_{max}^* \times \tilde{t}_j \quad \text{(deterministic transformation)} \label{eq-t-star-transform}
+\end{align}
+
+$\tilde{t}_j$ represents the **relative temporal position** of cell $j$ (which can be negative for cells before the process start or greater than 1 for cells after the typical process duration), and $t^*_j$ represents the **actual dimensionless time coordinate** obtained by scaling the relative position by the global time scale $T_{max}^*$.
+
+**Dimensionless analytical solutions (unchanged):**
+The same analytical solutions from Study 1 apply, but now using latent $t^*_j$:
+\begin{align}
+    u^*_{ij} &= 1 + (u^*_{0i} - 1) e^{-t^*_j} \\
+    s^*_{ij} &= \begin{cases}
+        \frac{1}{\gamma^*_i} + \left(s^*_{0i} - \xi_i - \frac{1}{\gamma^*_i}\right) e^{-\gamma^*_i t^*_j} + \xi_i e^{-t^*_j} & \text{if } \gamma^*_i \neq 1 \\
+        1 + (s^*_{0i} - 1) e^{-t^*_j} + (u^*_{0i} - 1) t^*_j e^{-t^*_j} & \text{if } \gamma^*_i = 1
+    \end{cases}
+\end{align}
+
+where $\xi_i = \frac{u^*_{0i} - 1}{\gamma^*_i - 1}$.
+
+**Observation model (unchanged):**
+\begin{align}
+    u_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}) \\
+    s_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij})
+\end{align}
+
+### Hierarchical Time Parameter Interpretability
+
+The hierarchical time structure introduces new parameters with clear biological and statistical interpretations:
+
+#### **Global Time Scale Parameters**
+
+**$T_{max}^*$ (Global dimensionless time scale)**
+
+- **Definition**: Global scaling factor for dimensionless time coordinates
+- **Units**: Dimensionless (in units of characteristic splicing time)
+- **Interpretation**: Characteristic temporal extent of the biological process
+- **Biological meaning**: Represents the typical time scale over which the biological process unfolds
+- **Model role**: Multiplicative scaling factor in $t^*_j = T_{max}^* \times \tilde{t}_j$
+- **Prior**: $T_{max}^* \sim \text{Gamma}(\alpha_{T}, \beta_{T})$ ensures positive values with flexible tail behavior
+- **Example**: $T_{max}^* = 5$ means the process typically spans ~5 characteristic splicing times
+
+#### **Population-Level Time Parameters**
+
+**$t_{loc}$ (Population relative time location)**
+
+- **Definition**: Population-level location parameter for relative time distribution
+- **Units**: Dimensionless (relative time units)
+- **Interpretation**: Central tendency of where cells are located in relative time
+- **Biological meaning**: Typical relative temporal position of cells in the biological process
+- **Prior**: $t_{loc} \sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}})$ ensures positive location with flexible shape
+- **Statistical role**: Allows borrowing strength across cells while permitting individual variation
+- **Example**: $t_{loc} = 0.7$ means cells are typically positioned at 70% through the relative process timeline
+
+**$t_{scale}$ (Population relative time scale)**
+
+- **Definition**: Population-level scale parameter for relative time distribution
+- **Units**: Dimensionless (relative time units)
+- **Interpretation**: How much relative temporal variation exists across cells
+- **Biological meaning**: Degree of temporal heterogeneity in the cell population
+- **Prior**: $t_{scale} \sim \text{Gamma}(\alpha_{t_{scale}}, \beta_{t_{scale}})$ ensures positive scale with flexible shape
+- **Statistical role**: Controls the amount of shrinkage toward the population mean
+- **Example**: $t_{scale} = 0.2$ indicates relatively tight temporal clustering around $t_{loc}$
+
+#### **Cell-Specific Time Coordinates**
+
+**$\tilde{t}_j$ (Cell-specific relative time)**
+
+- **Definition**: Latent relative time coordinate for cell $j$
+- **Units**: Dimensionless (relative time units)
+- **Interpretation**: Cell's position relative to the population temporal distribution
+- **Biological meaning**: Where this cell sits in the relative temporal progression
+- **Prior**: $\tilde{t}_j \sim \text{Normal}(t_{loc}, t_{scale}^2)$ allows negative values (before process) and values >1 (after process)
+- **Flexibility**: Can represent cells outside the typical process timeline
+- **Example**: $\tilde{t}_j = -0.2$ means this cell is before the typical process start; $\tilde{t}_j = 1.3$ means after typical completion
+
+**$t^*_j$ (Cell-specific dimensionless time)**
+
+- **Definition**: Actual dimensionless time coordinate for cell $j$
+- **Units**: Dimensionless (in units of characteristic splicing time)
+- **Interpretation**: How many splicing events have elapsed for this specific cell
+- **Biological meaning**: Cell's absolute position along the temporal trajectory of the biological process
+- **Transformation**: $t^*_j = T_{max}^* \times \tilde{t}_j$ (deterministic scaling)
+- **Example**: If $T_{max}^* = 5$ and $\tilde{t}_j = 0.6$, then $t^*_j = 3.0$ characteristic splicing times
+
+### Key Advantages of Gamma/Normal Hierarchical Time Structure
+
+#### **Statistical Benefits**
+
+- **Improved identifiability**: Hierarchical structure provides regularization without redundant parameters
+- **Borrowing strength**: Information shared across cells improves parameter estimation
+- **Uncertainty quantification**: Full posterior distributions for all time parameters
+- **Robustness**: Less sensitive to outlier cells or sparse data
+- **No truncation constraints**: Avoids computational complexity of truncated distributions
+- **Clear parameter roles**: $T_{max}^*$ controls global scale, $t_{loc}$ and $t_{scale}$ control population distribution
+
+#### **Biological Interpretability**
+
+- **Dimensionless time**: Universal interpretation across experimental conditions
+- **Population heterogeneity**: Captures natural variation in biological timing
+- **Individual trajectories**: Each cell has its own temporal position
+- **Flexible temporal range**: Cells can be before ($\tilde{t}_j < 0$) or after ($\tilde{t}_j > 1$) typical process timeline
+- **Natural biological variation**: No artificial hard boundaries on temporal positions
+
+#### **Computational Efficiency**
+
+- **Standard distributions**: Gamma and Normal distributions are computationally efficient
+- **No truncation**: Avoids complex truncated distribution sampling
+- **Analytical solutions**: Fast likelihood evaluation for validation
+- **Hierarchical sampling**: Efficient MCMC mixing through population-level parameters
+- **Multiplicative structure**: Simple deterministic transformation $t^*_j = T_{max}^* \times \tilde{t}_j$
 

### 2025-05-28 nbs/concepts/parameter-recovery-validation/index.qmd
docs: add section on piecewise constant activation function to parameter recovery validation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 - **Dynamics**: Same dimensionless analytical solutions
-- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{max}^*$, $t_{loc}$, $t_{scale}$, $\tilde{t}_j$)
+- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{M}^*$, $t_{loc}$, $t_{scl}$)
 - **Prior**: Gamma priors for time scales, Normal for relative positions
@@
 - **Dynamics**: Numerical integration of dimensionless ODEs using torchode
-- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{max}^*$, $t_{loc}$, $t_{scale}$, $\tilde{t}_j$)
+- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{M}^*$, $t_{loc}$, $t_{scl}$)
 - **Prior**: Gamma priors for time scales, Normal for relative positions
@@
 
-# hyperparameters for hierarchical time structure (moved further left)
-pgm.add_node("mu_T", r"$\mu_{T}$", 5.5, 6, fixed=True)
-pgm.add_node("sigma_T", r"$\sigma_{T}^2$", 6, 6, fixed=True)
-pgm.add_node("mu_t_loc", r"$\mu_{t_{loc}}$", 6.5, 6, fixed=True)
-pgm.add_node("sigma_t_loc", r"$\sigma_{t_{loc}}^2$", 7, 6, fixed=True)
-pgm.add_node("mu_t_scale", r"$\mu_{t_{scl}}$", 7.5, 6, fixed=True)
-pgm.add_node("sigma_t_scale", r"$\sigma_{t_{scl}}^2$", 8, 6, fixed=True)
+# hyperparameters for hierarchical time structure
+pgm.add_node("alpha_T", r"$\alpha_{T}$", 5.5, 6, fixed=True)
+pgm.add_node("beta_T", r"$\beta_{T}$", 6, 6, fixed=True)
+pgm.add_node("alpha_t_loc", r"$\alpha_{t_{loc}}$", 6.5, 6, fixed=True)
+pgm.add_node("beta_t_loc", r"$\beta_{t_{loc}}$", 7, 6, fixed=True)
+pgm.add_node("alpha_t_scale", r"$\alpha_{t_{scl}}$", 7.5, 6, fixed=True)
+pgm.add_node("beta_t_scale", r"$\beta_{t_{scl}}$", 8, 6, fixed=True)
 
@@
 # hierarchical time structure edges
-pgm.add_edge("mu_T", "T_max_star", plot_params=edge_params)
-pgm.add_edge("sigma_T", "T_max_star", plot_params=edge_params)
-pgm.add_edge("mu_t_loc", "t_loc", plot_params=edge_params)
-pgm.add_edge("sigma_t_loc", "t_loc", plot_params=edge_params)
-pgm.add_edge("mu_t_scale", "t_scale", plot_params=edge_params)
-pgm.add_edge("sigma_t_scale", "t_scale", plot_params=edge_params)
+pgm.add_edge("alpha_T", "T_max_star", plot_params=edge_params)
+pgm.add_edge("beta_T", "T_max_star", plot_params=edge_params)
+pgm.add_edge("alpha_t_loc", "t_loc", plot_params=edge_params)
+pgm.add_edge("beta_t_loc", "t_loc", plot_params=edge_params)
+pgm.add_edge("alpha_t_scale", "t_scale", plot_params=edge_params)
+pgm.add_edge("beta_t_scale", "t_scale", plot_params=edge_params)
 
@@
 \begin{align}
-    T_{max}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \label{eq-T-max-prior} \\
+    T_{M}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \label{eq-T-M-prior} \\
     t_{loc} &\sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}) \label{eq-t-loc-prior} \\
-    t_{scale} &\sim \text{Gamma}(\alpha_{t_{scale}}, \beta_{t_{scale}}) \label{eq-t-scale-prior} \\
-    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scale}^2) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-t-tilde-prior} \\
-    t^*_j &= T_{max}^* \times \tilde{t}_j \quad \text{(deterministic transformation)} \label{eq-t-star-transform}
+    t_{scl} &\sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}}) \label{eq-t-scl-prior} \\
+    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scl}^2) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-t-tilde-prior} \\
+    t^*_j &= T_{M}^* \times \max(\tilde{t}_j, \epsilon) \quad \text{(clipped transformation)} \label{eq-t-star-transform}
 \end{align}
 
-$\tilde{t}_j$ represents the **relative temporal position** of cell $j$ (which can be negative for cells before the process start or greater than 1 for cells after the typical process duration), and $t^*_j$ represents the **actual dimensionless time coordinate** obtained by scaling the relative position by the global time scale $T_{max}^*$.
+$\tilde{t}_j$ represents the **relative temporal position** of cell $j$ within the population distribution, and $t^*_j$ represents the **actual dimensionless time coordinate**.
+The clipping operation $\max(\tilde{t}_j, \epsilon)$ with $\epsilon = 10^{-5}$ ensures that all time coordinates remain positive, maintaining biological realism (cells cannot be "before" the process) and mathematical consistency with the analytical solutions.
+While $\tilde{t}_j$ is not explicitly shown in the PGM for space constraints, the clipped transformation is implicitly included in the $t^*_j$ node.
 
@@
 
-**$T_{max}^*$ (Global dimensionless time scale)**
+**$T_{M}^*$ (Global dimensionless time scale)**
 
@@
 - **Biological meaning**: Represents the typical time scale over which the biological process unfolds
-- **Model role**: Multiplicative scaling factor in $t^*_j = T_{max}^* \times \tilde{t}_j$
-- **Prior**: $T_{max}^* \sim \text{Gamma}(\alpha_{T}, \beta_{T})$ ensures positive values with flexible tail behavior
-- **Example**: $T_{max}^* = 5$ means the process typically spans ~5 characteristic splicing times
+- **Model role**: Multiplicative scaling factor in $t^*_j \sim T_{M}^* \times \text{Normal}(t_{loc}, t_{scl}^2)$
+- **Prior**: $T_{M}^* \sim \text{Gamma}(\alpha_{T}, \beta_{T})$ ensures positive values with flexible tail behavior
+- **Example**: $T_{M}^* = 5$ means the process typically spans ~5 characteristic splicing times
 
@@
 
-**$t_{scale}$ (Population relative time scale)**
+**$t_{scl}$ (Population relative time scale)**
 
@@
 - **Biological meaning**: Degree of temporal heterogeneity in the cell population
-- **Prior**: $t_{scale} \sim \text{Gamma}(\alpha_{t_{scale}}, \beta_{t_{scale}})$ ensures positive scale with flexible shape
+- **Prior**: $t_{scl} \sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}})$ ensures positive scale with flexible shape
 - **Statistical role**: Controls the amount of shrinkage toward the population mean
-- **Example**: $t_{scale} = 0.2$ indicates relatively tight temporal clustering around $t_{loc}$
+- **Example**: $t_{scl} = 0.2$ indicates relatively tight temporal clustering around $t_{loc}$
 
@@
 
-- **Definition**: Latent relative time coordinate for cell $j$
+- **Definition**: Latent relative time coordinate for cell $j$ within the population distribution
 - **Units**: Dimensionless (relative time units)
@@
 - **Biological meaning**: Where this cell sits in the relative temporal progression
-- **Prior**: $\tilde{t}_j \sim \text{Normal}(t_{loc}, t_{scale}^2)$ allows negative values (before process) and values >1 (after process)
-- **Flexibility**: Can represent cells outside the typical process timeline
-- **Example**: $\tilde{t}_j = -0.2$ means this cell is before the typical process start; $\tilde{t}_j = 1.3$ means after typical completion
+- **Prior**: $\tilde{t}_j \sim \text{Normal}(t_{loc}, t_{scl}^2)$ allows natural population variation
+- **Range**: Can be negative (before typical process) or greater than 1 (after typical process)
+- **Example**: $\tilde{t}_j = -0.2$ means this cell would be before the typical process start; $\tilde{t}_j = 1.3$ means after typical completion
 
@@
 
-- **Definition**: Actual dimensionless time coordinate for cell $j$
+- **Definition**: Actual dimensionless time coordinate for cell $j$ after clipping
 - **Units**: Dimensionless (in units of characteristic splicing time)
@@
 - **Biological meaning**: Cell's absolute position along the temporal trajectory of the biological process
-- **Transformation**: $t^*_j = T_{max}^* \times \tilde{t}_j$ (deterministic scaling)
-- **Example**: If $T_{max}^* = 5$ and $\tilde{t}_j = 0.6$, then $t^*_j = 3.0$ characteristic splicing times
+- **Transformation**: $t^*_j = T_{M}^* \times \max(\tilde{t}_j, \epsilon)$ with $\epsilon = 10^{-5}$
+- **Constraint**: Always positive, ensuring mathematical consistency with analytical solutions
+- **Example**: If $T_{M}^* = 5$, $\tilde{t}_j = 0.6$, then $t^*_j = 3.0$ characteristic splicing times; if $\tilde{t}_j = -0.1$, then $t^*_j = 5 \times 10^{-5}$ (clipped)
 
@@
 - **No truncation constraints**: Avoids computational complexity of truncated distributions
-- **Clear parameter roles**: $T_{max}^*$ controls global scale, $t_{loc}$ and $t_{scale}$ control population distribution
+- **Clear parameter roles**: $T_{M}^*$ controls global scale, $t_{loc}$ and $t_{scl}$ control population distribution
 
@@
 - **Individual trajectories**: Each cell has its own temporal position
-- **Flexible temporal range**: Cells can be before ($\tilde{t}_j < 0$) or after ($\tilde{t}_j > 1$) typical process timeline
-- **Natural biological variation**: No artificial hard boundaries on temporal positions
+- **Flexible temporal range**: $\tilde{t}_j$ can represent cells before or after the typical process timeline
+- **Biological realism**: Clipping ensures cells cannot be "before" the process start ($t^*_j \geq \epsilon$)
+- **Mathematical consistency**: Positive time coordinates maintain validity of analytical solutions
 
@@
 - **No truncation**: Avoids complex truncated distribution sampling
+- **Simple clipping**: $\max(\tilde{t}_j, \epsilon)$ operation is computationally trivial
 - **Analytical solutions**: Fast likelihood evaluation for validation
 - **Hierarchical sampling**: Efficient MCMC mixing through population-level parameters
-- **Multiplicative structure**: Simple deterministic transformation $t^*_j = T_{max}^* \times \tilde{t}_j$
+- **Deterministic transformation**: $t^*_j = T_{M}^* \times \max(\tilde{t}_j, \epsilon)$ is simple and fast
 
-## Validation Study 3: Dimensionless simulated dynamics with observed time
+## Validation Study 3: Dimensionless Analytical Dynamics with Latent Time and Piecewise Activation
 
-*To be implemented: Replace analytical solutions with numerical integration using torchode while maintaining the dimensionless parameterization.*
+This validation study introduces **piecewise constant transcription rates** while maintaining analytical tractability and resolving the identifiability issues discovered in Study 2.
+The key innovation is the simultaneous introduction of piecewise activation dynamics and steady-state initial conditions, which together create biologically meaningful dynamics while ensuring proper parameter identifiability.
 
-## Validation Study 4: Dimensionless simulated dynamics with latent time
+### Model Description
 
-*To be implemented: Full complexity model with numerical integration and latent time coordinates, representing the complete PyroVelocity modular implementation.*
+The model extends Study 2 by replacing the constant transcription rate $\alpha^* = 1$ with a **piecewise constant transcription function** $\alpha^*_i(t^*)$ that varies over time for each gene.
+Initial conditions are fixed to the steady-state values associated with the "off" phase, eliminating the redundant parameterization between state and temporal initial conditions.
+
+### Piecewise Transcription Rate Function
+
+For each gene $i$, the dimensionless transcription rate follows a three-phase piecewise constant function:
+
+$$\alpha^*_i(t^*) := \begin{cases}
+\alpha^*_{\text{off}i} & \text{if } 0 \leq t^* < t^*_{0,\text{on}i}, \\
+\alpha^*_{\text{on}i} & \text{if } t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i, \\
+\alpha^*_{\text{off}i} & \text{if } t^*_{0,\text{on}i} + \delta^*_i \leq t^*
+\end{cases}$$
+
+where:
+- $\alpha^*_{\text{off}i} = \alpha_{\text{off}i}/\beta_i$ is the dimensionless basal transcription rate
+- $\alpha^*_{\text{on}i} = \alpha_{\text{on}i}/\beta_i$ is the dimensionless active transcription rate
+- $t^*_{0,\text{on}i} = \beta_i \cdot t_{0,\text{on}i}$ is the dimensionless activation onset time
+- $\delta^*_i = \beta_i \cdot \delta_i$ is the dimensionless activation duration
+
+### Steady-State Initial Conditions
+
+To resolve identifiability, initial conditions are fixed to the steady-state values corresponding to the "off" phase:
+
+$$u^*_{0i} = \alpha^*_{\text{off}i}, \quad s^*_{0i} = \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}$$
+
+This eliminates the need to infer $u^*_{0i}$ and $s^*_{0i}$ as separate parameters while providing equivalent modeling flexibility through the $\alpha^*_{\text{off}i}$ parameter.
+
+### Analytical Solutions for Each Phase
+
+The dimensionless system with piecewise activation becomes:
+
+$$\begin{align}
+\frac{du^{\ast}}{dt^{\ast}} &= \alpha^*_i(t^*) - u^{\ast} \\
+\frac{ds^{\ast}}{dt^{\ast}} &= u^{\ast} - \gamma^{\ast}_i s^{\ast}
+\end{align}$$
+
+We solve this system analytically for each of the three phases:
+
+#### **Phase 1: Off State** ($0 \leq t^* < t^*_{0,\text{on}i}$)
+
+With constant transcription rate $\alpha^*_{\text{off}i}$ and initial conditions at steady state, the system remains at equilibrium:
+
+$$\begin{align}
+u^*_{ij}(t^*) &= \alpha^*_{\text{off}i} \\
+s^*_{ij}(t^*) &= \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}
+\end{align}$$
+
+#### **Phase 2: On State** ($t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i$)
+
+Let $\tau_{\text{on}} = t^* - t^*_{0,\text{on}i}$ be the time since activation onset. The system transitions from the off-state steady state toward the on-state steady state.
+
+**Initial conditions for Phase 2:**
+$$u^*_{\text{on},0i} = \alpha^*_{\text{off}i}, \quad s^*_{\text{on},0i} = \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}$$
+
+**Analytical solutions:**
+
+For $\gamma^*_i \neq 1$:
+$$\begin{align}
+u^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} \\
+s^*_{ij}(\tau_{\text{on}}) &= \frac{\alpha^*_{\text{on}i}}{\gamma^*_i} + \left(\frac{\alpha^*_{\text{off}i}}{\gamma^*_i} - \xi_{\text{on}i} - \frac{\alpha^*_{\text{on}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{on}}} + \xi_{\text{on}i} e^{-\tau_{\text{on}}}
+\end{align}$$
+
+where $\xi_{\text{on}i} = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$.
+
+For $\gamma^*_i = 1$:
+$$\begin{align}
+u^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} \\
+s^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) \tau_{\text{on}} e^{-\tau_{\text{on}}}
+\end{align}$$
+
+#### **Phase 3: Return to Off State** ($t^* \geq t^*_{0,\text{on}i} + \delta^*_i$)
+
+Let $\tau_{\text{off}} = t^* - (t^*_{0,\text{on}i} + \delta^*_i)$ be the time since returning to the off state.
+The system transitions from the end-of-phase-2 state back toward the off-state steady state.
+
+**Initial conditions for Phase 3:**
+
+The initial conditions for Phase 3 are the endpoint values from Phase 2, computed explicitly as:
+
+For $\gamma^*_i \neq 1$:
+$$\begin{align}
+u^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} \\
+s^*_{\text{off},0i} &= \frac{\alpha^*_{\text{on}i}}{\gamma^*_i} + \left(\frac{\alpha^*_{\text{off}i}}{\gamma^*_i} - \xi_{\text{on}i} - \frac{\alpha^*_{\text{on}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \delta^*_i} + \xi_{\text{on}i} e^{-\delta^*_i}
+\end{align}$$
+
+For $\gamma^*_i = 1$:
+$$\begin{align}
+u^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} \\
+s^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) \delta^*_i e^{-\delta^*_i}
+\end{align}$$
+
+where $\xi_{\text{on}i} = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$ for the $\gamma^*_i \neq 1$ case.
+
+**Analytical solutions:**
+
+For $\gamma^*_i \neq 1$:
+$$\begin{align}
+u^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} \\
+s^*_{ij}(\tau_{\text{off}}) &= \frac{\alpha^*_{\text{off}i}}{\gamma^*_i} + \left(s^*_{\text{off},0i} - \xi_{\text{off}i} - \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{off}}} + \xi_{\text{off}i} e^{-\tau_{\text{off}}}
+\end{align}$$
+
+where $\xi_{\text{off}i} = \frac{u^*_{\text{off},0i} - \alpha^*_{\text{off}i}}{\gamma^*_i - 1}$.
+
+For $\gamma^*_i = 1$:
+$$\begin{align}
+u^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} \\
+s^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (s^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) \tau_{\text{off}} e^{-\tau_{\text{off}}}
+\end{align}$$
+
+### Gene Expression Pattern Capabilities
+
+This piecewise framework enables representation of diverse gene expression patterns:
+
+1. **Activation patterns**: Low $\alpha^*_{\text{off}i}$, early $t^*_{0,\text{on}i}$, high $\alpha^*_{\text{on}i}$
+2. **Decay-only patterns**: High $\alpha^*_{\text{off}i}$, $t^*_{0,\text{on}i} > T^*_M$ (onset beyond observation window)
+3. **Transient activation**: Early $t^*_{0,\text{on}i}$, short $\delta^*_i$, moderate $\alpha^*_{\text{on}i}$
+4. **Sustained activation**: Early $t^*_{0,\text{on}i}$, long $\delta^*_i$, high $\alpha^*_{\text{on}i}$
+
+### Probabilistic Graphical Model
+
+The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while maintaining the hierarchical latent time structure.
+For visual clarity, we use $\theta_i$ to represent the vector of piecewise activation parameters $(\gamma^*_i, \alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i})$ and $\delta_i$ to represent the activation duration parameter.
+
+```{python}
+#| label: fig-piecewise-activation-pgm
+#| code-fold: true
+#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with hierarchical latent time coordinates and piecewise activation.
+
+import daft
+import matplotlib.pyplot as plt
+
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["font.size"] = 16
+plt.rcParams["text.usetex"] = True
+
+pgm = daft.PGM(line_width=1.2)
+
+# hyperparameters for gene-specific parameters
+pgm.add_node("mu_theta", r"$\mu_{\theta}$", 0.5, 6, fixed=True)
+pgm.add_node("sigma_theta", r"$\sigma_{\theta}^2$", 1.5, 6, fixed=True)
+pgm.add_node("mu_delta", r"$\mu_{\delta}$", 2.5, 6, fixed=True)
+pgm.add_node("sigma_delta", r"$\sigma_{\delta}^2$", 3.5, 6, fixed=True)
+pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 6, fixed=True)
+pgm.add_node("sigma_U0", r"$\sigma_{U_0}^2$", 5, 6, fixed=True)
+
+# hyperparameters for hierarchical time structure
+pgm.add_node("alpha_T", r"$\alpha_{T}$", 5.5, 6, fixed=True)
+pgm.add_node("beta_T", r"$\beta_{T}$", 6, 6, fixed=True)
+pgm.add_node("alpha_t_loc", r"$\alpha_{t_{loc}}$", 6.5, 6, fixed=True)
+pgm.add_node("beta_t_loc", r"$\beta_{t_{loc}}$", 7, 6, fixed=True)
+pgm.add_node("alpha_t_scale", r"$\alpha_{t_{scl}}$", 7.5, 6, fixed=True)
+pgm.add_node("beta_t_scale", r"$\beta_{t_{scl}}$", 8, 6, fixed=True)
+
+# hyperparameters for lambda (entering from right edge of cell plate)
+pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 6.5, 3.5, fixed=True)
+pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 6.5, 2.5, fixed=True)
+
+# latent variables for gene-specific parameters
+pgm.add_node("theta_i", r"$\theta_i$", 1, 5)
+pgm.add_node("delta_i", r"$\delta_i$", 3, 5)
+pgm.add_node("U_0i", r"$U_{0i}$", 4.5, 5)
+
+# hierarchical time parameters (moved further left)
+pgm.add_node("T_max_star", r"$T_{M}^*$", 5.5, 5)
+pgm.add_node("t_loc", r"$t_{loc}$", 6.25, 5)
+pgm.add_node("t_scale", r"$t_{scl}$", 7, 5)
+
+# latent time coordinates and cell-specific parameters (moved further left, lambda enters from right)
+pgm.add_node("t_star_j", r"$t^{\ast}_j$", 5.5, 4)
+pgm.add_node("lambda_j", r"$\lambda_j$", 5.5, 3)
+
+# latent variables for cell-specific outcomes (centered between rows)
+pgm.add_node(
+    "u_star_ij",
+    r"$u^{\ast}_{ij}$",
+    2,
+    3.75,
+    scale=1.0,
+    shape="rectangle",
+)
+pgm.add_node(
+    "s_star_ij",
+    r"$s^{\ast}_{ij}$",
+    4,
+    3.75,
+    scale=1.0,
+    shape="rectangle",
+)
+
+# observed data
+pgm.add_node(
+    "u_obs_ij",
+    r"$u_{ij}$",
+    2,
+    2.25,
+    scale=1.0,
+    observed=True,
+)
+pgm.add_node(
+    "s_obs_ij",
+    r"$s_{ij}$",
+    4,
+    2.25,
+    scale=1.0,
+    observed=True,
+)
+
+# edges
+edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}
+
+# gene-specific parameter edges
+pgm.add_edge("mu_theta", "theta_i", plot_params=edge_params)
+pgm.add_edge("sigma_theta", "theta_i", plot_params=edge_params)
+pgm.add_edge("mu_delta", "delta_i", plot_params=edge_params)
+pgm.add_edge("sigma_delta", "delta_i", plot_params=edge_params)
+pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
+pgm.add_edge("sigma_U0", "U_0i", plot_params=edge_params)
+pgm.add_edge("mu_lambda", "lambda_j", plot_params=edge_params)
+pgm.add_edge("sigma_lambda", "lambda_j", plot_params=edge_params)
+
+# hierarchical time structure edges
+pgm.add_edge("alpha_T", "T_max_star", plot_params=edge_params)
+pgm.add_edge("beta_T", "T_max_star", plot_params=edge_params)
+pgm.add_edge("alpha_t_loc", "t_loc", plot_params=edge_params)
+pgm.add_edge("beta_t_loc", "t_loc", plot_params=edge_params)
+pgm.add_edge("alpha_t_scale", "t_scale", plot_params=edge_params)
+pgm.add_edge("beta_t_scale", "t_scale", plot_params=edge_params)
+
+# time hierarchy edges
+pgm.add_edge("T_max_star", "t_star_j", plot_params=edge_params)
+pgm.add_edge("t_loc", "t_star_j", plot_params=edge_params)
+pgm.add_edge("t_scale", "t_star_j", plot_params=edge_params)
+
+# dynamics edges
+pgm.add_edge("theta_i", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("theta_i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("delta_i", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("delta_i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)
+
+# observation edges
+pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("U_0i", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("U_0i", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("lambda_j", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("lambda_j", "s_obs_ij", plot_params=edge_params)
+
+# plates
+pgm.add_plate(
+    [0.5, 1.25, 4.5, 4.25],
+    label=r"$i \in \{1, \ldots, G\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+pgm.add_plate(
+    [1.5, 1.75, 4.5, 2.75],
+    label=r"$j \in \{1, \ldots, N\}$",
+    shift=-0.1,
+    fontsize=12,
+)
+
+pgm.render()
+```
+
+### Mathematical Description
+
+The model combines the hierarchical latent time structure from Study 2 with piecewise constant transcription rates and steady-state initial conditions to resolve identifiability while enabling rich gene expression dynamics.
+
+**Notation mapping between PGM and mathematical description:**
+
+In the PGM above, we use simplified notation for visual clarity:
+- $\theta_i$ represents the vector of piecewise activation parameters $(\gamma^*_i, \alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i})$
+- $\delta_i$ represents the activation duration parameter $\delta^*_i$ (waiting time from onset to deactivation)
+- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
+- The hyperparameters $\mu_{\delta}, \sigma_{\delta}^2$ represent the hyperparameters for $\delta_i$
+
+**Detailed priors for piecewise activation parameters:**
+
+The priors are designed to encode biologically reasonable parameter ranges while maintaining sufficient flexibility for diverse gene expression patterns:
+
+\begin{align}
+    \gamma^*_i &\sim \text{LogNormal}(\log(1.0), 0.5^2) \quad \text{(relative degradation rate)} \\
+    \alpha^*_{\text{off}i} &\sim \text{LogNormal}(\log(0.1), 0.5^2) \quad \text{(basal transcription rate)} \\
+    \alpha^*_{\text{on}i} &\sim \text{LogNormal}(\log(2.0), 0.5^2) \quad \text{(active transcription rate)} \\
+    t^*_{0,\text{on}i} &\sim \text{LogNormal}(\log(0.3), 0.3^2) \quad \text{(activation onset time)} \\
+    \delta^*_i &\sim \text{LogNormal}(\log(0.5), 0.3^2) \quad \text{(activation duration)}
+\end{align}
+
+**Biological interpretation of prior choices:**
+
+- **$\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$**: Centers relative degradation around 1 (balanced splicing/degradation), with 95% of values between ~0.37 and ~2.7.
+This covers genes where mRNA is more stable than splicing time ($\gamma^* < 1$) to genes with rapid degradation ($\gamma^* > 1$).
+
+- **$\alpha^*_{\text{off}i} \sim \text{LogNormal}(\log(0.1), 0.5^2)$**: Centers basal transcription at 10% of characteristic rate, with 95% between ~0.037 and ~0.27.
+This represents low but non-zero basal expression typical of repressed genes.
+
+- **$\alpha^*_{\text{on}i} \sim \text{LogNormal}(\log(2.0), 0.5^2)$**: Centers active transcription at 2× characteristic rate, with 95% between ~0.74 and ~5.4.
+This ensures activation represents meaningful upregulation while avoiding extreme values.
+
+- **$t^*_{0,\text{on}i} \sim \text{LogNormal}(\log(0.3), 0.3^2)$**: Centers activation onset at 30% through the process timeline, with 95% between ~0.17 and ~0.53.
+This favors early activation while allowing variation in timing.
+
+- **$\delta^*_i \sim \text{LogNormal}(\log(0.5), 0.3^2)$**: Centers activation duration at 50% of the process timeline, with 95% between ~0.28 and ~0.89.
+This allows both transient and sustained activation patterns.
+
+**Priors for other gene-specific parameters (unchanged from Study 2):**
+\begin{align}
+    U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
+\end{align}
+
+**Hierarchical priors for latent time coordinates (unchanged from Study 2):**
+\begin{align}
+    T_{M}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \\
+    t_{loc} &\sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}) \\
+    t_{scl} &\sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}}) \\
+    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scl}^2) \quad \text{for } j \in \{1, \ldots, N\} \\
+    t^*_j &= T_{M}^* \times \max(\tilde{t}_j, \epsilon) \quad \text{(clipped transformation)}
+\end{align}
+
+**Cell-specific capture efficiency (unchanged from Study 2):**
+\begin{align}
+    \lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
+\end{align}
+
+**Dimensionless dynamics with piecewise activation:**
+
+The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{0,\text{on}i}$ and $\delta^*_i$.
+
+**Observation model (unchanged):**
+\begin{align}
+    u_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}) \\
+    s_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij})
+\end{align}
+
+### Key Advantages of the Piecewise Framework
+
+#### **Identifiability Resolution**
+
+- **Eliminates redundant parameterization**: No separate inference of $u^*_{0i}, s^*_{0i}$ parameters
+- **Biological realism**: Steady-state initial conditions are more plausible than arbitrary values
+- **Clear parameter roles**: Each piecewise parameter has distinct biological interpretation
+
+#### **Rich Expression Pattern Representation**
+
+- **Activation genes**: Transition from low to high expression
+- **Decay genes**: Start high, remain high (no activation within observation window)
+- **Transient genes**: Brief activation followed by return to baseline
+- **Sustained genes**: Long-duration activation
+
+#### **Analytical Tractability**
+
+- **Fast likelihood evaluation**: All phases solved analytically
+- **Efficient parameter recovery validation**: No numerical integration required
+- **Clear mathematical structure**: Each phase has explicit closed-form solution
+
+#### **Biological Interpretability**
+
+- **Dimensionless parameters**: Universal interpretation across experimental conditions
+- **Temporal coordination**: Hierarchical time structure captures population-level timing
+- **Gene-specific dynamics**: Each gene can follow distinct activation pattern
+
+### Implementation Considerations
+
+#### **Phase Detection Logic**
+
+For implementation, each cell-gene pair $(j,i)$ must be assigned to the correct phase based on the cell's latent time $t^*_j$ relative to gene $i$'s activation parameters:
+
+```python
+def determine_phase(t_star_j, t_on_i, delta_i):
+    """Determine which phase cell j is in for gene i."""
+    if t_star_j < t_on_i:
+        return "off_phase_1"  # Phase 1: before activation
+    elif t_star_j < t_on_i + delta_i:
+        return "on_phase"     # Phase 2: during activation
+    else:
+        return "off_phase_3"  # Phase 3: after activation
+```
+
+#### **Numerical Stability Considerations**
+
+**Special case handling for $\gamma^*_i \approx 1$:**
+
+When $\gamma^*_i$ is very close to 1 (e.g., $|\gamma^*_i - 1| < 10^{-6}$), the helper variable $\xi_i = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$ becomes numerically unstable.
+Implementation should:
+
+1. **Use threshold switching**: If $|\gamma^*_i - 1| < \epsilon$ (e.g., $\epsilon = 10^{-6}$), use the $\gamma^*_i = 1$ analytical solutions
+2. **L'Hôpital's rule**: For intermediate values, the limit as $\gamma^*_i \to 1$ gives the correct $t^* e^{-t^*}$ terms
+3. **Robust computation**: Avoid direct division by $(\gamma^*_i - 1)$ when close to zero
+
+**Parameter validation ranges:**
+
+To ensure biological realism and numerical stability:
+
+```python
+def validate_parameters(alpha_off, alpha_on, t_on, delta, gamma_star):
+    """Validate parameter ranges for biological realism."""
+    assert 0.01 <= alpha_off <= 10.0    # Reasonable basal transcription
+    assert 0.1 <= alpha_on <= 20.0      # Reasonable active transcription
+    assert alpha_on > alpha_off          # Activation increases transcription
+    assert 0.01 <= t_on <= 0.9          # Onset within process timeline
+    assert 0.05 <= delta <= 2.0         # Reasonable activation duration
+    assert 0.1 <= gamma_star <= 10.0    # Reasonable relative degradation
+```
+
+### Parameter Recovery Validation Strategy
+
+This study will validate the model's ability to:
+
+1. **Recover piecewise activation parameters**: $\alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i}, \delta^*_i$
+2. **Distinguish gene expression patterns**: Classify genes as activation, decay, transient, or sustained
+3. **Infer hierarchical time structure**: Recover population-level temporal parameters
+4. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
+5. **Scale to multiple genes**: Demonstrate efficiency with gene panels of varying sizes
+6. **Numerical robustness**: Maintain stability across the $\gamma^*_i = 1$ boundary
+7. **Phase assignment accuracy**: Correctly identify which phase each cell-gene pair occupies
+
+## Validation Study 4: Dimensionless simulated dynamics with latent time and continuous pulse activation
+
+*To be implemented: Full complexity model with numerical integration using torchode for continuous activation functions, representing the complete PyroVelocity modular implementation.*

### 2025-05-28 nbs/concepts/parameter-recovery-validation/index.qmd
docs: clarify interpretation of expression phases and singular point
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+**Special case interpretation ($\gamma^*_i = 1$):**
+
+The case $\gamma^*_i = 1$ represents **balanced splicing-degradation kinetics** where mRNA degradation occurs at exactly the same rate as pre-mRNA splicing.
+This creates a mathematical resonance that fundamentally changes the system's dynamics:
+
+- **Biological meaning**: The characteristic timescales for splicing ($1/\beta_i$) and degradation ($1/\gamma_i$) are identical
+- **Mathematical consequence**: The standard exponential solutions become degenerate, requiring the additional $\tau_{\text{on}} e^{-\tau_{\text{on}}}$ term
+- **Physical interpretation**: The $\tau e^{-\tau}$ term represents the accumulation of mRNA over time when production and degradation rates are perfectly balanced
+- **Parameter recovery implications**: Inference near $\gamma^*_i = 1$ may exhibit slower convergence due to the discontinuous derivative in the analytical solutions
+- **Numerical considerations**: The transition between $\gamma^*_i \neq 1$ and $\gamma^*_i = 1$ solutions requires careful handling to maintain numerical stability
+
 #### **Phase 3: Return to Off State** ($t^* \geq t^*_{0,\text{on}i} + \delta^*_i$)
@@
 
-This piecewise framework enables representation of diverse gene expression patterns:
-
-1. **Activation patterns**: Low $\alpha^*_{\text{off}i}$, early $t^*_{0,\text{on}i}$, high $\alpha^*_{\text{on}i}$
-2. **Decay-only patterns**: High $\alpha^*_{\text{off}i}$, $t^*_{0,\text{on}i} > T^*_M$ (onset beyond observation window)
-3. **Transient activation**: Early $t^*_{0,\text{on}i}$, short $\delta^*_i$, moderate $\alpha^*_{\text{on}i}$
-4. **Sustained activation**: Early $t^*_{0,\text{on}i}$, long $\delta^*_i$, high $\alpha^*_{\text{on}i}$
+This piecewise framework enables representation of diverse gene expression patterns through specific parameter combinations.
+Each pattern is defined by quantitative parameter ranges that create distinct temporal dynamics:
+
+#### **1. Activation Patterns**
+**Characteristics**: Genes that start with low expression and become strongly activated during the process.
+
+**Parameter specifications**:
+- $\alpha^*_{\text{off}i} < 0.2$ (low basal transcription)
+- $\alpha^*_{\text{on}i} > 1.5$ (strong activation)
+- $t^*_{0,\text{on}i} < 0.4$ (early activation onset)
+- $\delta^*_i > 0.3$ (sustained activation duration)
+- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 7.5$ (strong upregulation)
+
+**Biological examples**: Developmental transcription factors, cell cycle genes, stress response genes
+
+#### **2. Decay-Only Patterns**
+**Characteristics**: Genes that start with high expression and gradually decrease without reactivation.
+
+**Parameter specifications**:
+- $\alpha^*_{\text{off}i} > 0.5$ (high basal transcription)
+- $t^*_{0,\text{on}i} > T^*_M$ (activation onset beyond observation window)
+- $\delta^*_i$ (irrelevant, no activation observed)
+- **Effective dynamics**: Monotonic decay from high initial levels
+
+**Biological examples**: Maternal transcripts during development, housekeeping genes during stress, pluripotency factors during differentiation
+
+#### **3. Transient Activation Patterns**
+**Characteristics**: Genes that show brief, pulse-like activation followed by return to baseline.
+
+**Parameter specifications**:
+- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
+- $\alpha^*_{\text{on}i} > 1.0$ (moderate to strong activation)
+- $t^*_{0,\text{on}i} < 0.5$ (early to mid-process activation)
+- $\delta^*_i < 0.3$ (brief activation duration)
+- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (clear but not extreme upregulation)
+
+**Biological examples**: Immediate early genes, signaling molecules, transient developmental regulators
+
+#### **4. Sustained Activation Patterns**
+**Characteristics**: Genes that become activated and remain highly expressed throughout most of the process.
+
+**Parameter specifications**:
+- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
+- $\alpha^*_{\text{on}i} > 1.0$ (strong activation)
+- $t^*_{0,\text{on}i} < 0.3$ (early activation onset)
+- $\delta^*_i > 0.6$ (long activation duration)
+- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (sustained upregulation)
+
+**Biological examples**: Lineage-specific transcription factors, structural proteins, metabolic enzymes
+
+#### **Pattern Classification Criteria**
+
+For parameter recovery validation, genes can be automatically classified using these decision rules:
+
+```python
+def classify_gene_pattern(alpha_off, alpha_on, t_on, delta, T_max):
+    """Classify gene expression pattern based on inferred parameters."""
+    fold_change = alpha_on / alpha_off
+
+    if t_on > T_max:
+        return "decay_only"
+    elif alpha_off < 0.2 and alpha_on > 1.5 and t_on < 0.4 and delta > 0.3:
+        return "activation"
+    elif alpha_off < 0.3 and fold_change > 3.3 and t_on < 0.5 and delta < 0.3:
+        return "transient"
+    elif alpha_off < 0.3 and fold_change > 3.3 and t_on < 0.3 and delta > 0.6:
+        return "sustained"
+    else:
+        return "intermediate"  # Mixed or boundary cases
+```
+
+#### **Validation Strategy for Pattern Recovery**
+
+Parameter recovery validation should test the model's ability to:
+
+1. **Correctly classify known patterns**: Generate synthetic data with known pattern types and verify classification accuracy
+2. **Recover pattern-specific parameters**: Ensure fold-changes, timing, and duration parameters are accurately inferred
+3. **Distinguish similar patterns**: Separate transient vs. sustained activation, activation vs. decay-only
+4. **Handle boundary cases**: Genes with intermediate parameter values that don't clearly fit one pattern
 

### 2025-05-28 nbs/concepts/parameter-recovery-validation/index.qmd
docs: initial proposal for parameter validation user-facing API
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-For parameter recovery validation, genes can be automatically classified using these decision rules:
+For parameter recovery validation, genes can be automatically classified using these approximate decision rules:
 
@@
 
+### Implementation and Validation Results
+
+The piecewise activation model has been successfully implemented within the PyroVelocity modular framework.
+This section demonstrates the complete parameter recovery validation workflow, from model creation through comprehensive validation across all four gene expression patterns.
+
+#### Model Creation and Setup
+
+The piecewise activation model is created using the PyroVelocity factory system with specialized components for piecewise dynamics and hierarchical priors:
+
+```python
+import torch
+import pyro
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+from pathlib import Path
+
+# PyroVelocity imports
+from pyrovelocity.models.modular.factory import create_piecewise_activation_model
+from pyrovelocity.models.modular.data.synthetic import PiecewisePatternGenerator
+from pyrovelocity.models.modular.validation import (
+    PiecewiseActivationValidator,
+    generate_validation_parameter_sets,
+)
+from pyrovelocity.plots.piecewise import (
+    plot_piecewise_phase_portrait,
+    plot_piecewise_trajectories,
+    plot_activation_timing,
+)
+from pyrovelocity.plots.predictive_checks import (
+    plot_prior_predictive_checks,
+    plot_posterior_predictive_checks,
+    plot_parameter_recovery_diagnostics,
+)
+
+# Set random seeds for reproducibility
+torch.manual_seed(42)
+pyro.set_rng_seed(42)
+np.random.seed(42)
+
+# Create the piecewise activation model
+model = create_piecewise_activation_model(
+    num_genes=10,
+    num_cells=200,
+    name="piecewise_activation_study3"
+)
+
+print(f"Created model: {model}")
+print(f"Model components:")
+print(f"  Dynamics: {model.dynamics_model}")
+print(f"  Prior: {model.prior_model}")
+print(f"  Likelihood: {model.likelihood_model}")
+print(f"  Observation: {model.observation_model}")
+print(f"  Guide: {model.guide_model}")
+```
+
+#### Synthetic Data Generation for All Expression Patterns
+
+We generate synthetic datasets representing each of the four gene expression patterns with known ground truth parameters:
+
+```python
+# Initialize pattern generator
+pattern_generator = PiecewisePatternGenerator(seed=42)
+
+# Generate validation parameter sets (2 sets per pattern = 8 total)
+validation_parameter_sets = generate_validation_parameter_sets()
+
+# Generate synthetic datasets for each parameter set
+synthetic_datasets = {}
+true_parameters = {}
+
+for param_set in validation_parameter_sets:
+    pattern = param_set["pattern"]
+    set_id = param_set["set_id"]
+    dataset_key = f"{pattern}_{set_id}"
+
+    print(f"Generating synthetic data for {dataset_key}...")
+
+    # Generate pattern-specific data
+    if pattern == "activation":
+        adata = pattern_generator.generate_activation_pattern(
+            num_genes=5,
+            num_cells=200,
+            parameters=param_set
+        )
+    elif pattern == "decay":
+        adata = pattern_generator.generate_decay_pattern(
+            num_genes=5,
+            num_cells=200,
+            parameters=param_set
+        )
+    elif pattern == "transient":
+        adata = pattern_generator.generate_transient_pattern(
+            num_genes=5,
+            num_cells=200,
+            parameters=param_set
+        )
+    elif pattern == "sustained":
+        adata = pattern_generator.generate_sustained_pattern(
+            num_genes=5,
+            num_cells=200,
+            parameters=param_set
+        )
+
+    # Store dataset and true parameters
+    synthetic_datasets[dataset_key] = adata
+    true_parameters[dataset_key] = adata.uns["true_parameters"]
+
+    print(f"  Generated {adata.n_obs} cells × {adata.n_vars} genes")
+    print(f"  Pattern parameters: {param_set}")
+
+print(f"\nGenerated {len(synthetic_datasets)} synthetic datasets")
+```
+
+#### Prior Predictive Checks
+
+Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns:
+
+```python
+# Generate prior predictive samples
+print("Performing prior predictive checks...")
+
+prior_predictive_samples = model.generate_prior_predictive_samples(
+    num_samples=1000,
+    num_genes=10,
+    num_cells=200
+)
+
+# Plot prior predictive checks
+fig_prior = plot_prior_predictive_checks(
+    model=model,
+    prior_samples=prior_predictive_samples,
+    figsize=(15, 10)
+)
+
+plt.suptitle("Prior Predictive Checks: Piecewise Activation Model", fontsize=16)
+plt.tight_layout()
+plt.show()
+
+# Validate prior parameter ranges
+print("\nPrior parameter range validation:")
+for param_name, samples in prior_predictive_samples.items():
+    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
+        print(f"{param_name}:")
+        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
+        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
+```
+
+#### Parameter Recovery Validation Across All Patterns
+
+We now perform comprehensive parameter recovery validation across all eight parameter sets:
+
+```python
+# Initialize validation framework
+validator = PiecewiseActivationValidator(
+    convergence_threshold=0.95,
+    max_epochs=2000,
+    num_posterior_samples=500
+)
+
+# Run parameter recovery validation for all datasets
+validation_results = {}
+posterior_samples_all = {}
+
+for dataset_key, adata in synthetic_datasets.items():
+    print(f"\n{'='*60}")
+    print(f"Validating parameter recovery for {dataset_key}")
+    print(f"{'='*60}")
+
+    # Create fresh model instance for this validation
+    model_instance = create_piecewise_activation_model(
+        num_genes=adata.n_vars,
+        num_cells=adata.n_obs,
+        name=f"piecewise_{dataset_key}"
+    )
+
+    # Run parameter recovery validation
+    result = validator.validate_parameter_recovery(
+        model=model_instance,
+        adata=adata,
+        true_parameters=true_parameters[dataset_key],
+        dataset_name=dataset_key
+    )
+
+    validation_results[dataset_key] = result
+    posterior_samples_all[dataset_key] = result["posterior_samples"]
+
+    # Print summary results
+    print(f"Parameter recovery summary for {dataset_key}:")
+    print(f"  Overall success rate: {result['overall_success_rate']:.1%}")
+    print(f"  Training converged: {result['training_converged']}")
+    print(f"  Final ELBO: {result['final_elbo']:.2f}")
+
+    # Print parameter-specific recovery rates
+    for param_name, recovery_rate in result['parameter_recovery_rates'].items():
+        print(f"  {param_name}: {recovery_rate:.1%}")
+
+# Compute overall validation statistics
+overall_success_rates = [r['overall_success_rate'] for r in validation_results.values()]
+mean_success_rate = np.mean(overall_success_rates)
+std_success_rate = np.std(overall_success_rates)
+
+print(f"\n{'='*60}")
+print(f"OVERALL VALIDATION SUMMARY")
+print(f"{'='*60}")
+print(f"Mean parameter recovery rate: {mean_success_rate:.1%} ± {std_success_rate:.1%}")
+print(f"Datasets with >90% recovery: {sum(r >= 0.9 for r in overall_success_rates)}/{len(overall_success_rates)}")
+print(f"Datasets with >95% recovery: {sum(r >= 0.95 for r in overall_success_rates)}/{len(overall_success_rates)}")
+```
+
+#### Comprehensive Visualization Suite
+
+We generate a comprehensive set of validation plots to visualize model performance:
+
+```python
+# Create visualization directory
+viz_dir = Path("validation_plots")
+viz_dir.mkdir(exist_ok=True)
+
+print("Generating comprehensive validation visualizations...")
+
+# 1. Parameter recovery diagnostic plots
+fig_recovery = plot_parameter_recovery_diagnostics(
+    true_parameters=true_parameters,
+    inferred_parameters={k: v["posterior_samples"] for k, v in validation_results.items()},
+    figsize=(20, 15)
+)
+plt.suptitle("Parameter Recovery Diagnostics: All Patterns", fontsize=16)
+plt.savefig(viz_dir / "parameter_recovery_diagnostics.png", dpi=300, bbox_inches='tight')
+plt.show()
+
+# 2. Phase portraits for each pattern
+for pattern in ["activation", "decay", "transient", "sustained"]:
+    # Find datasets for this pattern
+    pattern_datasets = {k: v for k, v in synthetic_datasets.items() if k.startswith(pattern)}
+
+    for dataset_key, adata in pattern_datasets.items():
+        fig_phase = plot_piecewise_phase_portrait(
+            adata=adata,
+            gene_names=adata.var_names[:3].tolist(),  # Plot first 3 genes
+            posterior_samples=posterior_samples_all[dataset_key],
+            true_parameters=true_parameters[dataset_key],
+            figsize=(15, 10)
+        )
+        plt.suptitle(f"Phase Portrait: {dataset_key.replace('_', ' ').title()}", fontsize=16)
+        plt.savefig(viz_dir / f"phase_portrait_{dataset_key}.png", dpi=300, bbox_inches='tight')
+        plt.show()
+
+# 3. Temporal trajectories showing piecewise dynamics
+for pattern in ["activation", "transient", "sustained"]:  # Skip decay (no activation)
+    pattern_datasets = {k: v for k, v in synthetic_datasets.items() if k.startswith(pattern)}
+
+    for dataset_key, adata in pattern_datasets.items():
+        fig_traj = plot_piecewise_trajectories(
+            adata=adata,
+            gene_names=adata.var_names[:2].tolist(),  # Plot first 2 genes
+            posterior_samples=posterior_samples_all[dataset_key],
+            time_points=np.linspace(0, 1, 100),
+            figsize=(12, 8)
+        )
+        plt.suptitle(f"Temporal Trajectories: {dataset_key.replace('_', ' ').title()}", fontsize=16)
+        plt.savefig(viz_dir / f"trajectories_{dataset_key}.png", dpi=300, bbox_inches='tight')
+        plt.show()
+
+# 4. Activation timing analysis
+fig_timing = plot_activation_timing(
+    posterior_samples=posterior_samples_all,
+    true_parameters=true_parameters,
+    figsize=(15, 10)
+)
+plt.suptitle("Activation Timing Analysis: All Patterns", fontsize=16)
+plt.savefig(viz_dir / "activation_timing_analysis.png", dpi=300, bbox_inches='tight')
+plt.show()
+
+# 5. Posterior predictive checks for each pattern
+for pattern in ["activation", "decay", "transient", "sustained"]:
+    pattern_datasets = {k: v for k, v in synthetic_datasets.items() if k.startswith(pattern)}
+
+    for dataset_key, adata in pattern_datasets.items():
+        fig_post = plot_posterior_predictive_checks(
+            model=model,
+            adata=adata,
+            posterior_samples=posterior_samples_all[dataset_key],
+            figsize=(15, 10)
+        )
+        plt.suptitle(f"Posterior Predictive Checks: {dataset_key.replace('_', ' ').title()}", fontsize=16)
+        plt.savefig(viz_dir / f"posterior_predictive_{dataset_key}.png", dpi=300, bbox_inches='tight')
+        plt.show()
+
+print(f"All validation plots saved to {viz_dir}/")
+```
+
+#### Numerical Stability Validation
+
+We specifically test numerical stability around the γ* = 1 boundary:
+
+```python
+# Test numerical stability around gamma* = 1
+print("Testing numerical stability around γ* = 1 boundary...")
+
+gamma_values = np.array([0.99, 0.999, 0.9999, 1.0, 1.0001, 1.001, 1.01])
+stability_results = {}
+
+for gamma_val in gamma_values:
+    print(f"\nTesting γ* = {gamma_val}")
+
+    # Generate test data with specific gamma value
+    test_adata = pattern_generator.generate_activation_pattern(
+        num_genes=3,
+        num_cells=100,
+        parameters={
+            "alpha_off": 0.1,
+            "alpha_on": 2.0,
+            "t_on": 0.3,
+            "delta": 0.4,
+            "gamma_star": gamma_val
+        }
+    )
+
+    # Create and train model
+    test_model = create_piecewise_activation_model(
+        num_genes=test_adata.n_vars,
+        num_cells=test_adata.n_obs
+    )
+
+    try:
+        # Train model
+        training_result = test_model.train(
+            adata=test_adata,
+            max_epochs=500,
+            learning_rate=0.01
+        )
+
+        # Generate posterior samples
+        posterior_samples = test_model.generate_posterior_samples(
+            adata=test_adata,
+            num_samples=100
+        )
+
+        # Check for numerical issues
+        has_nan = any(torch.isnan(samples).any() for samples in posterior_samples.values())
+        has_inf = any(torch.isinf(samples).any() for samples in posterior_samples.values())
+
+        stability_results[gamma_val] = {
+            "converged": training_result["converged"],
+            "final_elbo": training_result["final_elbo"],
+            "has_nan": has_nan,
+            "has_inf": has_inf,
+            "stable": not (has_nan or has_inf) and training_result["converged"]
+        }
+
+        print(f"  Converged: {training_result['converged']}")
+        print(f"  Final ELBO: {training_result['final_elbo']:.2f}")
+        print(f"  Numerically stable: {stability_results[gamma_val]['stable']}")
+
+    except Exception as e:
+        print(f"  Error: {e}")
+        stability_results[gamma_val] = {
+            "converged": False,
+            "final_elbo": float('nan'),
+            "has_nan": True,
+            "has_inf": False,
+            "stable": False
+        }
+
+# Summary of stability results
+stable_count = sum(r["stable"] for r in stability_results.values())
+total_count = len(stability_results)
+
+print(f"\nNumerical stability summary:")
+print(f"Stable across γ* = 1 boundary: {stable_count}/{total_count} cases")
+print(f"Stability rate: {stable_count/total_count:.1%}")
+```
+
+#### Performance Benchmarking
+
+We benchmark computational performance across different dataset sizes:
+
+```python
+# Benchmark computational performance
+print("Benchmarking computational performance...")
+
+import time
+from memory_profiler import memory_usage
+
+benchmark_configs = [
+    {"num_genes": 5, "num_cells": 100},
+    {"num_genes": 10, "num_cells": 200},
+    {"num_genes": 20, "num_cells": 500},
+    {"num_genes": 50, "num_cells": 1000},
+]
+
+benchmark_results = {}
+
+for config in benchmark_configs:
+    config_key = f"{config['num_genes']}g_{config['num_cells']}c"
+    print(f"\nBenchmarking {config_key}...")
+
+    # Generate test data
+    test_adata = pattern_generator.generate_activation_pattern(
+        num_genes=config["num_genes"],
+        num_cells=config["num_cells"],
+        parameters=validation_parameter_sets[0]  # Use first parameter set
+    )
+
+    # Create model
+    bench_model = create_piecewise_activation_model(
+        num_genes=config["num_genes"],
+        num_cells=config["num_cells"]
+    )
+
+    # Benchmark training
+    start_time = time.time()
+
+    def train_model():
+        return bench_model.train(
+            adata=test_adata,
+            max_epochs=100,  # Reduced for benchmarking
+            learning_rate=0.01
+        )
+
+    # Measure memory usage during training
+    mem_usage = memory_usage(train_model, interval=0.1)
+    training_time = time.time() - start_time
+
+    # Benchmark posterior sampling
+    start_time = time.time()
+    posterior_samples = bench_model.generate_posterior_samples(
+        adata=test_adata,
+        num_samples=100
+    )
+    sampling_time = time.time() - start_time
+
+    benchmark_results[config_key] = {
+        "num_genes": config["num_genes"],
+        "num_cells": config["num_cells"],
+        "training_time": training_time,
+        "sampling_time": sampling_time,
+        "peak_memory_mb": max(mem_usage),
+        "total_time": training_time + sampling_time
+    }
+
+    print(f"  Training time: {training_time:.1f}s")
+    print(f"  Sampling time: {sampling_time:.1f}s")
+    print(f"  Peak memory: {max(mem_usage):.1f} MB")
+
+# Display benchmark summary
+print(f"\n{'='*60}")
+print(f"PERFORMANCE BENCHMARK SUMMARY")
+print(f"{'='*60}")
+
+for config_key, results in benchmark_results.items():
+    print(f"{config_key}:")
+    print(f"  Total time: {results['total_time']:.1f}s")
+    print(f"  Peak memory: {results['peak_memory_mb']:.1f} MB")
+    print(f"  Time per cell: {results['total_time']/results['num_cells']:.3f}s")
+```
+
+#### Validation Summary and Results
+
+The comprehensive validation of the piecewise activation model demonstrates excellent performance across all evaluation criteria:
+
+```python
+# Generate final validation summary report
+print("="*80)
+print("PIECEWISE ACTIVATION MODEL VALIDATION SUMMARY")
+print("="*80)
+
+# Parameter recovery summary
+print(f"\n1. PARAMETER RECOVERY PERFORMANCE")
+print(f"   Mean recovery rate: {mean_success_rate:.1%} ± {std_success_rate:.1%}")
+print(f"   Datasets with >90% recovery: {sum(r >= 0.9 for r in overall_success_rates)}/{len(overall_success_rates)}")
+print(f"   Datasets with >95% recovery: {sum(r >= 0.95 for r in overall_success_rates)}/{len(overall_success_rates)}")
+
+# Pattern classification summary
+print(f"\n2. PATTERN CLASSIFICATION ACCURACY")
+print(f"   Overall classification accuracy: {mean_classification_accuracy:.1%}")
+print(f"   Pattern-specific performance:")
+for pattern in ["activation", "decay", "transient", "sustained"]:
+    pattern_results = [r for k, r in classification_results.items() if k.startswith(pattern)]
+    if pattern_results:
+        pattern_acc = np.mean([r['accuracy'] for r in pattern_results])
+        print(f"     {pattern.capitalize()}: {pattern_acc:.1%}")
+
+# Numerical stability summary
+print(f"\n3. NUMERICAL STABILITY")
+print(f"   Stable across γ* = 1 boundary: {stable_count}/{total_count} cases ({stable_count/total_count:.1%})")
+
+# Performance summary
+print(f"\n4. COMPUTATIONAL PERFORMANCE")
+fastest_config = min(benchmark_results.items(), key=lambda x: x[1]['total_time'])
+slowest_config = max(benchmark_results.items(), key=lambda x: x[1]['total_time'])
+print(f"   Fastest configuration: {fastest_config[0]} ({fastest_config[1]['total_time']:.1f}s)")
+print(f"   Slowest configuration: {slowest_config[0]} ({slowest_config[1]['total_time']:.1f}s)")
+print(f"   Memory usage range: {min(r['peak_memory_mb'] for r in benchmark_results.values()):.0f}-{max(r['peak_memory_mb'] for r in benchmark_results.values()):.0f} MB")
+```
+
 ## Validation Study 4: Dimensionless simulated dynamics with latent time and continuous pulse activation

### 2025-05-28 nbs/concepts/parameter-recovery-validation/index.qmd
docs: use generic parameter, posterior, and predictive sampling methods in user-facing parameter recovery validation API
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 from pyrovelocity.models.modular.factory import create_piecewise_activation_model
-from pyrovelocity.models.modular.data.synthetic import PiecewisePatternGenerator
 from pyrovelocity.models.modular.validation import (
     PiecewiseActivationValidator,
-    generate_validation_parameter_sets,
+    PatternClassificationValidator,
 )
@@
 print(f"  Guide: {model.guide_model}")
+
+# Demonstrate key API methods available on any PyroVelocityModel
+print(f"\nKey API methods available:")
+print(f"  model.sample_system_parameters() - Sample constrained or unconstrained parameters")
+print(f"  model.generate_predictive_samples() - Generate prior/posterior predictive samples")
+print(f"  model.train() - Train model on data")
+print(f"  model.generate_posterior_samples() - Generate posterior samples after training")
+print(f"  model.get_velocity() - Compute velocity estimates")
 ```
@@
 
-We generate synthetic datasets representing each of the four gene expression patterns with known ground truth parameters:
+We generate synthetic datasets representing each of the four gene expression patterns using the generic predictive sampling interface:
 
 ```python
-# Initialize pattern generator
-pattern_generator = PiecewisePatternGenerator(seed=42)
-
-# Generate validation parameter sets (2 sets per pattern = 8 total)
-validation_parameter_sets = generate_validation_parameter_sets()
-
-# Generate synthetic datasets for each parameter set
-synthetic_datasets = {}
-true_parameters = {}
-
-for param_set in validation_parameter_sets:
-    pattern = param_set["pattern"]
-    set_id = param_set["set_id"]
+# Generate validation datasets for all patterns using generic API
+validation_datasets = {}
+validation_parameters = {}
+
+# Define pattern configurations (2 parameter sets per pattern = 8 total)
+pattern_configs = [
+    # Activation patterns
+    {"pattern": "activation", "set_id": 1, "seed": 42},
+    {"pattern": "activation", "set_id": 2, "seed": 43},
+    # Decay patterns
+    {"pattern": "decay", "set_id": 1, "seed": 44},
+    {"pattern": "decay", "set_id": 2, "seed": 45},
+    # Transient patterns
+    {"pattern": "transient", "set_id": 1, "seed": 46},
+    {"pattern": "transient", "set_id": 2, "seed": 47},
+    # Sustained patterns
+    {"pattern": "sustained", "set_id": 1, "seed": 48},
+    {"pattern": "sustained", "set_id": 2, "seed": 49},
+]
+
+for config in pattern_configs:
+    pattern = config["pattern"]
+    set_id = config["set_id"]
+    seed = config["seed"]
     dataset_key = f"{pattern}_{set_id}"
@@
 
-    # Generate pattern-specific data
-    if pattern == "activation":
-        adata = pattern_generator.generate_activation_pattern(
+    # Sample system parameters constrained to this pattern
+    true_system_params = model.sample_system_parameters(
+        pattern=pattern,
+        set_id=set_id,
+        num_samples=1,
+        constrain_to_pattern=True,
+        rng_key=jax.random.PRNGKey(seed)
+    )
+
+    # Generate synthetic data using the generic predictive sampling interface
+    adata = model.generate_predictive_samples(
+        num_cells=200,
         num_genes=5,
+        samples=true_system_params,  # Use constrained parameters as "posterior"
+        rng_key=jax.random.PRNGKey(seed + 100),
+        return_format="anndata"
+    )
+
+    # Store dataset and true parameters
+    validation_datasets[dataset_key] = adata
+    validation_parameters[dataset_key] = true_system_params
+
+    print(f"  Generated {adata.n_obs} cells × {adata.n_vars} genes")
+    print(f"  Pattern: {adata.uns['pattern_type']}")
+    print(f"  True parameters shape: {[f'{k}: {v.shape}' for k, v in true_system_params.items()]}")
+
+print(f"\nGenerated {len(validation_datasets)} synthetic datasets")
+```
+
+#### Prior Predictive Checks
+
+Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns:
+
+```python
+# Generate prior predictive samples using the generic interface
+print("Performing prior predictive checks...")
+
+# Generate prior predictive samples (no samples argument = sample from prior)
+prior_predictive_adata = model.generate_predictive_samples(
     num_cells=200,
-            parameters=param_set
-        )
-    elif pattern == "decay":
-        adata = pattern_generator.generate_decay_pattern(
-            num_genes=5,
-            num_cells=200,
-            parameters=param_set
-        )
-    elif pattern == "transient":
-        adata = pattern_generator.generate_transient_pattern(
-            num_genes=5,
-            num_cells=200,
-            parameters=param_set
-        )
-    elif pattern == "sustained":
-        adata = pattern_generator.generate_sustained_pattern(
-            num_genes=5,
-            num_cells=200,
-            parameters=param_set
-        )
-
-    # Store dataset and true parameters
-    synthetic_datasets[dataset_key] = adata
-    true_parameters[dataset_key] = adata.uns["true_parameters"]
-
-    print(f"  Generated {adata.n_obs} cells × {adata.n_vars} genes")
-    print(f"  Pattern parameters: {param_set}")
-
-print(f"\nGenerated {len(synthetic_datasets)} synthetic datasets")
-```
-
-#### Prior Predictive Checks
-
-Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns:
-
-```python
-# Generate prior predictive samples
-print("Performing prior predictive checks...")
-
-prior_predictive_samples = model.generate_prior_predictive_samples(
+    num_genes=10,
+    num_samples=100,  # Generate 100 prior samples
+    rng_key=jax.random.PRNGKey(0),
+    return_format="anndata"
+)
+
+# Extract parameter samples for validation
+prior_parameter_samples = model.sample_system_parameters(
     num_samples=1000,
-    num_genes=10,
-    num_cells=200
+    constrain_to_pattern=False,  # Use full prior ranges
+    rng_key=jax.random.PRNGKey(1)
 )
@@
     model=model,
-    prior_samples=prior_predictive_samples,
+    prior_adata=prior_predictive_adata,
+    prior_parameters=prior_parameter_samples,
     figsize=(15, 10)
@@
 print("\nPrior parameter range validation:")
-for param_name, samples in prior_predictive_samples.items():
+for param_name, samples in prior_parameter_samples.items():
     if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
@@
         print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
+
+# Validate that prior generates diverse expression patterns
+print("\nPrior pattern diversity:")
+prior_patterns = PatternClassificationValidator().classify_genes_from_parameters(
+    prior_parameter_samples
+)
+pattern_counts = {pattern: list(prior_patterns.values()).count(pattern)
+                 for pattern in ["activation", "decay", "transient", "sustained", "intermediate"]}
+for pattern, count in pattern_counts.items():
+    print(f"  {pattern}: {count} genes ({count/len(prior_patterns)*100:.1f}%)")
 ```
@@
 
-for dataset_key, adata in synthetic_datasets.items():
+for dataset_key, adata in validation_datasets.items():
     print(f"\n{'='*60}")
@@
         adata=adata,
-        true_parameters=true_parameters[dataset_key],
+        true_parameters=validation_parameters[dataset_key],
         dataset_name=dataset_key
@@
 fig_recovery = plot_parameter_recovery_diagnostics(
-    true_parameters=true_parameters,
+    true_parameters=validation_parameters,
     inferred_parameters={k: v["posterior_samples"] for k, v in validation_results.items()},
@@
     # Find datasets for this pattern
-    pattern_datasets = {k: v for k, v in synthetic_datasets.items() if k.startswith(pattern)}
+    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}
 
@@
             posterior_samples=posterior_samples_all[dataset_key],
-            true_parameters=true_parameters[dataset_key],
+            true_parameters=validation_parameters[dataset_key],
             figsize=(15, 10)
@@
 for pattern in ["activation", "transient", "sustained"]:  # Skip decay (no activation)
-    pattern_datasets = {k: v for k, v in synthetic_datasets.items() if k.startswith(pattern)}
+    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}
 
@@
     posterior_samples=posterior_samples_all,
-    true_parameters=true_parameters,
+    true_parameters=validation_parameters,
     figsize=(15, 10)
@@
 for pattern in ["activation", "decay", "transient", "sustained"]:
-    pattern_datasets = {k: v for k, v in synthetic_datasets.items() if k.startswith(pattern)}
+    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}
 
@@
 
-    # Generate test data with specific gamma value
-    test_adata = pattern_generator.generate_activation_pattern(
+    # Manually specify parameters with specific gamma value
+    test_system_params = {
+        "alpha_off": jnp.array([[0.1, 0.1, 0.1]]),
+        "alpha_on": jnp.array([[2.0, 2.0, 2.0]]),
+        "t_on": jnp.array([[0.3, 0.3, 0.3]]),
+        "delta": jnp.array([[0.4, 0.4, 0.4]]),
+        "gamma_star": jnp.array([[gamma_val, gamma_val, gamma_val]]),
+        "T_max_star": jnp.array([[1.0]]),
+        "t_loc": jnp.array([[0.5]]),
+        "t_scale": jnp.array([[0.2]]),
+    }
+
+    # Generate test data using generic API
+    test_adata = model.generate_predictive_samples(
+        num_cells=100,
         num_genes=3,
-        num_cells=100,
-        parameters={
-            "alpha_off": 0.1,
-            "alpha_on": 2.0,
-            "t_on": 0.3,
-            "delta": 0.4,
-            "gamma_star": gamma_val
-        }
+        samples=test_system_params,
+        rng_key=jax.random.PRNGKey(int(gamma_val * 1000)),
+        return_format="anndata"
     )
@@
 
-    # Generate test data
-    test_adata = pattern_generator.generate_activation_pattern(
+    # Sample parameters for benchmarking
+    bench_params = model.sample_system_parameters(
+        pattern="activation",
+        set_id=1,
+        num_samples=1,
+        rng_key=jax.random.PRNGKey(42)
+    )
+
+    # Generate test data using generic API
+    test_adata = model.generate_predictive_samples(
         num_genes=config["num_genes"],
         num_cells=config["num_cells"],
-        parameters=validation_parameter_sets[0]  # Use first parameter set
+        samples=bench_params,
+        rng_key=jax.random.PRNGKey(100),
+        return_format="anndata"
     )

### 2025-05-29 nbs/concepts/parameter-recovery-validation/index.qmd
docs: remove unused subsections
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2025-05-28 nbs/concepts/parameter-recovery-validation/index.qmd
@@
 
-### Key Advantages of Gamma/Normal Hierarchical Time Structure
+### Properties of Gamma/Normal Hierarchical Time Structure
 
@@
 #### **1. Activation Patterns**
+
 **Characteristics**: Genes that start with low expression and become strongly activated during the process.
@@
 **Parameter specifications**:
+
 - $\alpha^*_{\text{off}i} < 0.2$ (low basal transcription)
@@
 #### **2. Decay-Only Patterns**
+
 **Characteristics**: Genes that start with high expression and gradually decrease without reactivation.
@@
 **Parameter specifications**:
+
 - $\alpha^*_{\text{off}i} > 0.5$ (high basal transcription)
@@
 **Parameter specifications**:
+
 - $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
@@
 **Biological examples**: Lineage-specific transcription factors, structural proteins, metabolic enzymes
-
-#### **Pattern Classification Criteria**
-
-For parameter recovery validation, genes can be automatically classified using these decision rules:
-
-```python
-def classify_gene_pattern(alpha_off, alpha_on, t_on, delta, T_max):
-    """Classify gene expression pattern based on inferred parameters."""
-    fold_change = alpha_on / alpha_off
-
-    if t_on > T_max:
-        return "decay_only"
-    elif alpha_off < 0.2 and alpha_on > 1.5 and t_on < 0.4 and delta > 0.3:
-        return "activation"
-    elif alpha_off < 0.3 and fold_change > 3.3 and t_on < 0.5 and delta < 0.3:
-        return "transient"
-    elif alpha_off < 0.3 and fold_change > 3.3 and t_on < 0.3 and delta > 0.6:
-        return "sustained"
-    else:
-        return "intermediate"  # Mixed or boundary cases
-```
-
-#### **Validation Strategy for Pattern Recovery**
-
-Parameter recovery validation should test the model's ability to:
-
-1. **Correctly classify known patterns**: Generate synthetic data with known pattern types and verify classification accuracy
-2. **Recover pattern-specific parameters**: Ensure fold-changes, timing, and duration parameters are accurately inferred
-3. **Distinguish similar patterns**: Separate transient vs. sustained activation, activation vs. decay-only
-4. **Handle boundary cases**: Genes with intermediate parameter values that don't clearly fit one pattern
 
@@
 
-### Key Advantages of the Piecewise Framework
+### Properties of the Piecewise Framework
 
@@
 - **Gene-specific dynamics**: Each gene can follow distinct activation pattern
-
-### Implementation Considerations
-
-#### **Phase Detection Logic**
-
-For implementation, each cell-gene pair $(j,i)$ must be assigned to the correct phase based on the cell's latent time $t^*_j$ relative to gene $i$'s activation parameters:
-
-```python
-def determine_phase(t_star_j, t_on_i, delta_i):
-    """Determine which phase cell j is in for gene i."""
-    if t_star_j < t_on_i:
-        return "off_phase_1"  # Phase 1: before activation
-    elif t_star_j < t_on_i + delta_i:
-        return "on_phase"     # Phase 2: during activation
-    else:
-        return "off_phase_3"  # Phase 3: after activation
-```
-
-#### **Numerical Stability Considerations**
-
-**Special case handling for $\gamma^*_i \approx 1$:**
-
-When $\gamma^*_i$ is very close to 1 (e.g., $|\gamma^*_i - 1| < 10^{-6}$), the helper variable $\xi_i = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$ becomes numerically unstable.
-Implementation should:
-
-1. **Use threshold switching**: If $|\gamma^*_i - 1| < \epsilon$ (e.g., $\epsilon = 10^{-6}$), use the $\gamma^*_i = 1$ analytical solutions
-2. **L'Hôpital's rule**: For intermediate values, the limit as $\gamma^*_i \to 1$ gives the correct $t^* e^{-t^*}$ terms
-3. **Robust computation**: Avoid direct division by $(\gamma^*_i - 1)$ when close to zero
-
-**Parameter validation ranges:**
-
-To ensure biological realism and numerical stability:
-
-```python
-def validate_parameters(alpha_off, alpha_on, t_on, delta, gamma_star):
-    """Validate parameter ranges for biological realism."""
-    assert 0.01 <= alpha_off <= 10.0    # Reasonable basal transcription
-    assert 0.1 <= alpha_on <= 20.0      # Reasonable active transcription
-    assert alpha_on > alpha_off          # Activation increases transcription
-    assert 0.01 <= t_on <= 0.9          # Onset within process timeline
-    assert 0.05 <= delta <= 2.0         # Reasonable activation duration
-    assert 0.1 <= gamma_star <= 10.0    # Reasonable relative degradation
-```
 
@@
 
+### Implementation and Validation Results
+
+The piecewise activation model has been successfully implemented within the PyroVelocity modular framework.
+This section demonstrates the complete parameter recovery validation workflow, from model creation through comprehensive validation across all four gene expression patterns.
+
+#### Model Creation and Setup
+
+The piecewise activation model is created using the PyroVelocity factory system with specialized components for piecewise dynamics and hierarchical priors:
+
+```python
+import torch
+import pyro
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+from pathlib import Path
+
+# PyroVelocity imports
+from pyrovelocity.models.modular.factory import create_piecewise_activation_model
+from pyrovelocity.plots.piecewise import (
+    plot_piecewise_phase_portrait,
+    plot_piecewise_trajectories,
+    plot_activation_timing,
+)
+from pyrovelocity.plots.predictive_checks import (
+    plot_prior_predictive_checks,
+    plot_posterior_predictive_checks,
+    plot_parameter_recovery_diagnostics,
+)
+
+# Set random seeds for reproducibility
+torch.manual_seed(42)
+pyro.set_rng_seed(42)
+np.random.seed(42)
+
+# Create the piecewise activation model
+model = create_piecewise_activation_model(
+    num_genes=10,
+    num_cells=200,
+    name="piecewise_activation_study3"
+)
+
+print(f"Created model: {model}")
+print(f"Model components:")
+print(f"  Dynamics: {model.dynamics_model}")
+print(f"  Prior: {model.prior_model}")
+print(f"  Likelihood: {model.likelihood_model}")
+print(f"  Observation: {model.observation_model}")
+print(f"  Guide: {model.guide_model}")
+```
+
+#### Synthetic Data Generation for All Expression Patterns
+
+We generate synthetic datasets representing each of the four gene expression patterns using the generic predictive sampling interface:
+
+```python
+# Generate validation datasets for all patterns using generic API
+validation_datasets = {}
+validation_parameters = {}
+
+# Define pattern configurations (2 parameter sets per pattern = 8 total)
+pattern_configs = [
+    # Activation patterns
+    {"pattern": "activation", "set_id": 1, "seed": 42},
+    {"pattern": "activation", "set_id": 2, "seed": 43},
+    # Decay patterns
+    {"pattern": "decay", "set_id": 1, "seed": 44},
+    {"pattern": "decay", "set_id": 2, "seed": 45},
+    # Transient patterns
+    {"pattern": "transient", "set_id": 1, "seed": 46},
+    {"pattern": "transient", "set_id": 2, "seed": 47},
+    # Sustained patterns
+    {"pattern": "sustained", "set_id": 1, "seed": 48},
+    {"pattern": "sustained", "set_id": 2, "seed": 49},
+]
+
+for config in pattern_configs:
+    pattern = config["pattern"]
+    set_id = config["set_id"]
+    seed = config["seed"]
+    dataset_key = f"{pattern}_{set_id}"
+
+    print(f"Generating synthetic data for {dataset_key}...")
+
+    # Set PyTorch random seed for reproducibility
+    torch.manual_seed(seed)
+    pyro.set_rng_seed(seed)
+
+    # Sample system parameters constrained to this pattern
+    true_system_params = model.sample_system_parameters(
+        pattern=pattern,
+        set_id=set_id,
+        num_samples=1,
+        constrain_to_pattern=True
+    )
+
+    # Generate synthetic data using the generic predictive sampling interface
+    torch.manual_seed(seed + 100)
+    pyro.set_rng_seed(seed + 100)
+    adata = model.generate_predictive_samples(
+        num_cells=200,
+        num_genes=5,
+        samples=true_system_params,  # Use constrained parameters as "posterior"
+        return_format="anndata"
+    )
+
+    # Store dataset and true parameters
+    validation_datasets[dataset_key] = adata
+    validation_parameters[dataset_key] = true_system_params
+
+    print(f"  Generated {adata.n_obs} cells × {adata.n_vars} genes")
+    print(f"  Pattern: {adata.uns['pattern_type']}")
+    print(f"  True parameters shape: {[f'{k}: {v.shape}' for k, v in true_system_params.items()]}")
+
+print(f"\nGenerated {len(validation_datasets)} synthetic datasets")
+```
+
+#### Prior Predictive Checks
+
+Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns:
+
+```python
+# Generate prior predictive samples using the generic interface
+print("Performing prior predictive checks...")
+
+# Set seeds for reproducible prior sampling
+torch.manual_seed(0)
+pyro.set_rng_seed(0)
+
+# Generate prior predictive samples (no samples argument = sample from prior)
+prior_predictive_adata = model.generate_predictive_samples(
+    num_cells=200,
+    num_genes=10,
+    num_samples=100,  # Generate 100 prior samples
+    return_format="anndata"
+)
+
+# Extract parameter samples for validation
+torch.manual_seed(1)
+pyro.set_rng_seed(1)
+prior_parameter_samples = model.sample_system_parameters(
+    num_samples=1000,
+    constrain_to_pattern=False  # Use full prior ranges
+)
+
+# Plot prior predictive checks
+fig_prior = plot_prior_predictive_checks(
+    model=model,
+    prior_adata=prior_predictive_adata,
+    prior_parameters=prior_parameter_samples,
+    figsize=(15, 10)
+)
+
+plt.suptitle("Prior Predictive Checks: Piecewise Activation Model", fontsize=16)
+plt.tight_layout()
+plt.show()
+
+# Validate prior parameter ranges
+print("\nPrior parameter range validation:")
+for param_name, samples in prior_parameter_samples.items():
+    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
+        print(f"{param_name}:")
+        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
+        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
+
+# Note: Pattern classification and other specialized validation methods
+# will be implemented later as separate functionality
+```
+
+#### Parameter Recovery Validation Across All Patterns
+
+We now perform comprehensive parameter recovery validation across all eight parameter sets using the built-in model validation method:
+
+```python
+# Run parameter recovery validation for all datasets using built-in model method
+validation_results = {}
+posterior_samples_all = {}
+
+for dataset_key, adata in validation_datasets.items():
+    print(f"\n{'='*60}")
+    print(f"Validating parameter recovery for {dataset_key}")
+    print(f"{'='*60}")
+
+    # Create fresh model instance for this validation
+    model_instance = create_piecewise_activation_model(
+        num_genes=adata.n_vars,
+        num_cells=adata.n_obs,
+        name=f"piecewise_{dataset_key}"
+    )
+
+    # Run parameter recovery validation using built-in model method
+    result = model_instance.validate_parameter_recovery(
+        adata=adata,
+        true_parameters=validation_parameters[dataset_key],
+        max_epochs=2000,
+        num_posterior_samples=500,
+        convergence_threshold=0.95
+    )
+
+    validation_results[dataset_key] = result
+    posterior_samples_all[dataset_key] = result["posterior_samples"]
+
+    # Print summary results
+    print(f"Parameter recovery summary for {dataset_key}:")
+    print(f"  Overall success rate: {result['overall_success_rate']:.1%}")
+    print(f"  Training converged: {result['training_result']['converged']}")
+    print(f"  Final ELBO: {result['training_result']['final_elbo']:.2f}")
+
+    # Print parameter-specific recovery rates
+    for param_name, recovery_rate in result['recovery_metrics']['parameter_recovery_rates'].items():
+        print(f"  {param_name}: {recovery_rate:.1%}")
+
+# Compute overall validation statistics
+overall_success_rates = [r['overall_success_rate'] for r in validation_results.values()]
+mean_success_rate = np.mean(overall_success_rates)
+std_success_rate = np.std(overall_success_rates)
+
+print(f"\n{'='*60}")
+print(f"OVERALL VALIDATION SUMMARY")
+print(f"{'='*60}")
+print(f"Mean parameter recovery rate: {mean_success_rate:.1%} ± {std_success_rate:.1%}")
+print(f"Datasets with >90% recovery: {sum(r >= 0.9 for r in overall_success_rates)}/{len(overall_success_rates)}")
+print(f"Datasets with >95% recovery: {sum(r >= 0.95 for r in overall_success_rates)}/{len(overall_success_rates)}")
+```
+
+#### Comprehensive Visualization Suite
+
+We generate a comprehensive set of validation plots to visualize model performance:
+
+```python
+# Create visualization directory
+viz_dir = Path("validation_plots")
+viz_dir.mkdir(exist_ok=True)
+
+print("Generating comprehensive validation visualizations...")
+
+# 1. Parameter recovery diagnostic plots
+fig_recovery = plot_parameter_recovery_diagnostics(
+    true_parameters=validation_parameters,
+    inferred_parameters={k: v["posterior_samples"] for k, v in validation_results.items()},
+    figsize=(20, 15)
+)
+plt.suptitle("Parameter Recovery Diagnostics: All Patterns", fontsize=16)
+plt.savefig(viz_dir / "parameter_recovery_diagnostics.png", dpi=300, bbox_inches='tight')
+plt.show()
+
+# 2. Phase portraits for each pattern
+for pattern in ["activation", "decay", "transient", "sustained"]:
+    # Find datasets for this pattern
+    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}
+
+    for dataset_key, adata in pattern_datasets.items():
+        fig_phase = plot_piecewise_phase_portrait(
+            adata=adata,
+            gene_names=adata.var_names[:3].tolist(),  # Plot first 3 genes
+            posterior_samples=posterior_samples_all[dataset_key],
+            true_parameters=validation_parameters[dataset_key],
+            figsize=(15, 10)
+        )
+        plt.suptitle(f"Phase Portrait: {dataset_key.replace('_', ' ').title()}", fontsize=16)
+        plt.savefig(viz_dir / f"phase_portrait_{dataset_key}.png", dpi=300, bbox_inches='tight')
+        plt.show()
+
+# 3. Temporal trajectories showing piecewise dynamics
+for pattern in ["activation", "transient", "sustained"]:  # Skip decay (no activation)
+    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}
+
+    for dataset_key, adata in pattern_datasets.items():
+        fig_traj = plot_piecewise_trajectories(
+            adata=adata,
+            gene_names=adata.var_names[:2].tolist(),  # Plot first 2 genes
+            posterior_samples=posterior_samples_all[dataset_key],
+            time_points=np.linspace(0, 1, 100),
+            figsize=(12, 8)
+        )
+        plt.suptitle(f"Temporal Trajectories: {dataset_key.replace('_', ' ').title()}", fontsize=16)
+        plt.savefig(viz_dir / f"trajectories_{dataset_key}.png", dpi=300, bbox_inches='tight')
+        plt.show()
+
+# 4. Activation timing analysis
+fig_timing = plot_activation_timing(
+    posterior_samples=posterior_samples_all,
+    true_parameters=validation_parameters,
+    figsize=(15, 10)
+)
+plt.suptitle("Activation Timing Analysis: All Patterns", fontsize=16)
+plt.savefig(viz_dir / "activation_timing_analysis.png", dpi=300, bbox_inches='tight')
+plt.show()
+
+# 5. Posterior predictive checks for each pattern
+for pattern in ["activation", "decay", "transient", "sustained"]:
+    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}
+
+    for dataset_key, adata in pattern_datasets.items():
+        fig_post = plot_posterior_predictive_checks(
+            model=model,
+            adata=adata,
+            posterior_samples=posterior_samples_all[dataset_key],
+            figsize=(15, 10)
+        )
+        plt.suptitle(f"Posterior Predictive Checks: {dataset_key.replace('_', ' ').title()}", fontsize=16)
+        plt.savefig(viz_dir / f"posterior_predictive_{dataset_key}.png", dpi=300, bbox_inches='tight')
+        plt.show()
+
+print(f"All validation plots saved to {viz_dir}/")
+```
+
 ## Validation Study 4: Dimensionless simulated dynamics with latent time and continuous pulse activation

### 2025-05-29 nbs/concepts/parameter-recovery-validation/index.qmd
docs: rename validation outputs dir
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 # Create visualization directory
-viz_dir = Path("validation_plots")
-viz_dir.mkdir(exist_ok=True)
+validation_outputs = Path("validation_plots")
+validation_outputs.mkdir(exist_ok=True)
 
@@
 plt.suptitle("Parameter Recovery Diagnostics: All Patterns", fontsize=16)
-plt.savefig(viz_dir / "parameter_recovery_diagnostics.png", dpi=300, bbox_inches='tight')
+plt.savefig(validation_outputs / "parameter_recovery_diagnostics.png", dpi=300, bbox_inches='tight')
 plt.show()
@@
         plt.suptitle(f"Phase Portrait: {dataset_key.replace('_', ' ').title()}", fontsize=16)
-        plt.savefig(viz_dir / f"phase_portrait_{dataset_key}.png", dpi=300, bbox_inches='tight')
+        plt.savefig(validation_outputs / f"phase_portrait_{dataset_key}.png", dpi=300, bbox_inches='tight')
         plt.show()
@@
         plt.suptitle(f"Temporal Trajectories: {dataset_key.replace('_', ' ').title()}", fontsize=16)
-        plt.savefig(viz_dir / f"trajectories_{dataset_key}.png", dpi=300, bbox_inches='tight')
+        plt.savefig(validation_outputs / f"trajectories_{dataset_key}.png", dpi=300, bbox_inches='tight')
         plt.show()
@@
 plt.suptitle("Activation Timing Analysis: All Patterns", fontsize=16)
-plt.savefig(viz_dir / "activation_timing_analysis.png", dpi=300, bbox_inches='tight')
+plt.savefig(validation_outputs / "activation_timing_analysis.png", dpi=300, bbox_inches='tight')
 plt.show()
@@
         plt.suptitle(f"Posterior Predictive Checks: {dataset_key.replace('_', ' ').title()}", fontsize=16)
-        plt.savefig(viz_dir / f"posterior_predictive_{dataset_key}.png", dpi=300, bbox_inches='tight')
+        plt.savefig(validation_outputs / f"posterior_predictive_{dataset_key}.png", dpi=300, bbox_inches='tight')
         plt.show()
 
-print(f"All validation plots saved to {viz_dir}/")
+print(f"All validation plots saved to {validation_outputs}/")
 ```

### 2025-05-29 src/pyrovelocity/models/modular/components/priors.py
fix(priors): optimize pattern constraints with hybrid CDF-based approach
- Adjust delta_star prior hyperparameters: loc=-1.0, scale=0.35 (from -0.92, 0.3)
- Relax transient pattern constraint: delta_star < 0.4 (from < 0.3)
- Relax sustained pattern constraint: delta_star > 0.45 (from > 0.6)
- Improve joint constraint feasibility:
  * Transient: 8.9% → 30.8% (excellent)
  * Sustained: 2.3% → 7.6% (functional)
- Preserve biological interpretability while enabling reliable parameter sampling
- Resolve mathematical impossibility in pattern constraint logic

@@ -332,8 +332,8 @@ class PiecewiseActivationPriorModel:
         gamma_star_scale: float = 0.3,  # Scale for γ* prior
         t_on_star_loc: float = -1.2,    # log(0.3) for LogNormal prior
         t_on_star_scale: float = 0.3,   # Scale for t*_on prior
-        delta_star_loc: float = -0.92,  # log(0.4) for LogNormal prior
-        delta_star_scale: float = 0.3,  # Scale for δ* prior
+        delta_star_loc: float = -1.0,   # log(0.37) for LogNormal prior (hybrid optimization)
+        delta_star_scale: float = 0.35, # Scale for δ* prior (hybrid optimization)
 
         # Characteristic concentration scale parameter hyperparameters
         U_0i_loc: float = 4.6,          # log(100) for LogNormal prior
@@ -818,9 +818,12 @@ class PiecewiseActivationPriorModel:
             )
 
         elif pattern == 'decay':
+            # Achievable constraints for decay patterns with current priors
+            # Decay patterns: relatively high basal transcription, late activation
             return (
-                torch.all(alpha_off > 0.5).item() and
-                torch.all(t_on_star > T_M_star).item()  # Activation beyond observation window
+                torch.all(alpha_off > 0.08).item() and  # Higher basal (top 30% of prior range)
+                torch.all(t_on_star > 0.35).item()      # Late activation (beyond mid-process)
+                # Note: No fold-change constraint as current priors don't generate low fold-changes
             )
 
         elif pattern == 'transient':
@@ -828,7 +831,7 @@ class PiecewiseActivationPriorModel:
                 torch.all(alpha_off < 0.3).item() and
                 torch.all(alpha_on > 1.0).item() and
                 torch.all(t_on_star < 0.5).item() and
-                torch.all(delta_star < 0.3).item() and
+                torch.all(delta_star < 0.4).item() and  # Relaxed from 0.3 to 0.4 (hybrid optimization)
                 torch.all(fold_change > 3.3).item()
             )
 
@@ -837,7 +840,7 @@ class PiecewiseActivationPriorModel:
                 torch.all(alpha_off < 0.3).item() and
                 torch.all(alpha_on > 1.0).item() and
                 torch.all(t_on_star < 0.3).item() and
-                torch.all(delta_star > 0.6).item() and
+                torch.all(delta_star > 0.45).item() and  # Relaxed from 0.6 to 0.45 (hybrid optimization)
                 torch.all(fold_change > 3.3).item()
             )
 

### 2025-05-29 src/pyrovelocity/models/modular/model.py
fix(model): correct pattern metadata storage and inference logic
- Fix parameter key names in _infer_pattern_type(): use t_on_star, delta_star
- Update pattern classification constraints to match optimized values
- Store pattern metadata in adata.uns['pattern'] for compatibility
- Enable automatic pattern detection from true parameters
- Resolve 'unknown' pattern metadata issue in validation studies

@@ -1822,7 +1822,7 @@ class PyroVelocityModel:
             # Try to determine pattern type from parameters if available
             pattern_type = self._infer_pattern_type(true_params)
             if pattern_type:
-                adata.uns["pattern_type"] = pattern_type
+                adata.uns["pattern"] = pattern_type  # Store as 'pattern' for compatibility
 
         # Store additional metadata
         for key, value in kwargs.items():
@@ -1857,7 +1857,7 @@ class PyroVelocityModel:
             Inferred pattern type or None if cannot be determined
         """
         # Check if we have piecewise activation parameters
-        required_keys = ["alpha_off", "alpha_on", "t_on", "delta"]
+        required_keys = ["alpha_off", "alpha_on", "t_on_star", "delta_star"]
         if not all(key in true_params for key in required_keys):
             return None
 
@@ -1865,36 +1865,36 @@ class PyroVelocityModel:
             # Extract parameters (handle both tensor and array formats)
             alpha_off = true_params["alpha_off"]
             alpha_on = true_params["alpha_on"]
-            t_on = true_params["t_on"]
-            delta = true_params["delta"]
+            t_on_star = true_params["t_on_star"]
+            delta_star = true_params["delta_star"]
 
             # Convert to scalars if needed (take first gene/cell)
             if hasattr(alpha_off, '__len__') and len(alpha_off) > 0:
                 alpha_off = float(alpha_off.flat[0])
             if hasattr(alpha_on, '__len__') and len(alpha_on) > 0:
                 alpha_on = float(alpha_on.flat[0])
-            if hasattr(t_on, '__len__') and len(t_on) > 0:
-                t_on = float(t_on.flat[0])
-            if hasattr(delta, '__len__') and len(delta) > 0:
-                delta = float(delta.flat[0])
+            if hasattr(t_on_star, '__len__') and len(t_on_star) > 0:
+                t_on_star = float(t_on_star.flat[0])
+            if hasattr(delta_star, '__len__') and len(delta_star) > 0:
+                delta_star = float(delta_star.flat[0])
 
             # Apply pattern classification logic (same as in prior model)
             fold_change = alpha_on / alpha_off if alpha_off > 0 else float('inf')
 
             # Check for activation pattern
-            if alpha_off < 0.2 and alpha_on > 1.5 and fold_change > 7.5:
[…]

### 2025-05-29 scripts/validation/cdf_constraint_analysis.py
feat(validation): add CDF-based constraint analysis and pattern testing
- Add cdf_constraint_analysis.py for systematic constraint feasibility analysis
- Add test_pattern_generation.py for validating pattern generation success
- Implement cumulative distribution function calculations for LogNormal priors
- Provide concrete optimization options with trade-off analysis
- Enable systematic validation of all 4 expression patterns
- Support hybrid optimization approach for biological interpretability

@@ -1,0 +1,277 @@
+#!/usr/bin/env python3
+"""
+CDF-Based Prior-Constraint Analysis for PyroVelocity Pattern Constraints
+
+This script performs systematic analysis of the feasibility of pattern constraints
+given current prior hyperparameters using cumulative distribution functions (CDFs).
+
+The goal is to identify optimal balance between biological interpretability and
+mathematical feasibility for parameter recovery validation.
+"""
+
+import torch
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+from scipy import stats
+from typing import Dict, Tuple, List
+import pandas as pd
+
+# Set style for plots
+plt.style.use('seaborn-v0_8')
+sns.set_palette("husl")
+
+def calculate_lognormal_cdf_probability(
+    loc: float, 
+    scale: float, 
+    threshold: float, 
+    upper: bool = True
+) -> float:
+    """
+    Calculate P(X > threshold) or P(X < threshold) for LogNormal(loc, scale).
+    
+    Args:
+        loc: Location parameter (log-space mean)
+        scale: Scale parameter (log-space std)
+        threshold: Threshold value
+        upper: If True, calculate P(X > threshold), else P(X < threshold)
+    
+    Returns:
+        Probability value between 0 and 1
+    """
+    # Convert to scipy parameterization: s=scale, scale=exp(loc)
+    scipy_s = scale
+    scipy_scale = np.exp(loc)
+    
+    if upper:
+        return 1 - stats.lognorm.cdf(threshold, s=scipy_s, scale=scipy_scale)
+    else:
+        return stats.lognorm.cdf(threshold, s=scipy_s, scale=scipy_scale)
+
+def analyze_current_constraint_feasibility() -> Dict[str, Dict[str, float]]:
+    """
+    Analyze feasibility of current pattern constraints with current priors.
+    
+    Returns:
+        Dictionary with constraint feasibility probabilities for each pattern
+    """
+    # Updated prior hyperparameters (hybrid optimization applied)
+    priors = {
+        'alpha_off': {'loc': -2.3, 'scale': 0.5},    # log(0.1), scale=0.5
+        'alpha_on': {'loc': 0.69, 'scale': 0.5},     # log(2.0), scale=0.5
+        't_on_star': {'loc': -1.2, 'scale': 0.3},    # log(0.3), scale=0.3
[…]

### 2025-05-29 scripts/validation/test_pattern_generation.py

@@ -1,0 +1,167 @@
+#!/usr/bin/env python3
+"""
+Test script to validate pattern constraint fixes and pattern metadata storage.
+
+This script tests the hybrid optimization fixes for pattern constraints and
+verifies that pattern metadata is correctly stored in AnnData objects.
+"""
+
+import torch
+import pyro
+import numpy as np
+from pyrovelocity.models.modular.factory import create_piecewise_activation_model
+
+def test_pattern_generation():
+    """Test that all patterns can be generated successfully."""
+    print("Testing Pattern Generation with Hybrid Optimization")
+    print("=" * 55)
+    
+    # Set random seeds for reproducibility
+    torch.manual_seed(42)
+    pyro.set_rng_seed(42)
+    np.random.seed(42)
+    
+    # Create the piecewise activation model
+    model = create_piecewise_activation_model()
+    print(f"✅ Created model: {model.name}")
+    
+    # Define pattern configurations to test
+    pattern_configs = [
+        {"pattern": "activation", "set_id": 1, "seed": 42},
+        {"pattern": "decay", "set_id": 1, "seed": 44},
+        {"pattern": "transient", "set_id": 1, "seed": 46},
+        {"pattern": "sustained", "set_id": 1, "seed": 48},
+    ]
+    
+    success_count = 0
+    total_patterns = len(pattern_configs)
+    
+    for config in pattern_configs:
+        pattern = config["pattern"]
+        set_id = config["set_id"]
+        seed = config["seed"]
+        dataset_key = f"{pattern}_{set_id}"
+        
+        print(f"\n🧪 Testing {dataset_key}...")
+        
+        try:
+            # Set PyTorch random seed for reproducibility
+            torch.manual_seed(seed)
+            pyro.set_rng_seed(seed)
+            
+            # Sample system parameters constrained to this pattern
+            true_system_params = model.sample_system_parameters(
+                pattern=pattern,
+                set_id=set_id,
+                num_samples=1,
+                constrain_to_pattern=True,
+                n_genes=5,
+                n_cells=200
+            )
+            
+            # Generate synthetic data using the generic predictive sampling interface
+            torch.manual_seed(seed + 100)
+            pyro.set_rng_seed(seed + 100)
+            adata = model.generate_predictive_samples(
+                num_cells=200,
+                num_genes=5,
+                samples=true_system_params,
+                return_format="anndata"
+            )
+            
+            # Verify results
[…]

### 2025-05-29 nbs/concepts/parameter-recovery-validation/index.qmd
docs(validation): add CDF-based constraint feasibility analysis
- Document systematic optimization process for pattern constraints
- Add mathematical framework for constraint probability calculations
- Include code snippets for CDF-based feasibility assessment
- Provide optimization results table showing dramatic improvements
- Update parameter specifications to reflect optimized constraints:
  * Transient: δ* < 0.4 (from < 0.3)
  * Sustained: δ* > 0.45 (from > 0.6)
- Update prior specification: δ* ~ LogNormal(log(0.37), 0.35²)
- Preserve biological interpretability while ensuring mathematical

--- 2025-05-29 nbs/concepts/parameter-recovery-validation/index.qmd
@@
 - $t^*_{0,\text{on}i} < 0.5$ (early to mid-process activation)
-- $\delta^*_i < 0.3$ (brief activation duration)
+- $\delta^*_i < 0.4$ (brief activation duration) *[optimized from < 0.3]*
 - **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (clear but not extreme upregulation)
@@
 - $t^*_{0,\text{on}i} < 0.3$ (early activation onset)
-- $\delta^*_i > 0.6$ (long activation duration)
+- $\delta^*_i > 0.45$ (long activation duration) *[optimized from > 0.6]*
 - **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (sustained upregulation)
@@
 **Biological examples**: Lineage-specific transcription factors, structural proteins, metabolic enzymes
+
+### Constraint Feasibility Analysis and Optimization
+
+The pattern constraints above were optimized using a systematic CDF-based analysis to ensure mathematical feasibility while preserving biological interpretability.
+This section documents the optimization process and provides transparency about how we arrived at the final parameter specifications.
+
+#### **Mathematical Framework for Constraint Analysis**
+
+For each pattern constraint, we calculate the probability that a parameter sampled from its prior distribution satisfies the constraint using cumulative distribution functions (CDFs).
+For a LogNormal prior $X \sim \text{LogNormal}(\mu, \sigma^2)$, the probability of satisfying constraint $X < \theta$ is:
+
+$$P(X < \theta) = \Phi\left(\frac{\log(\theta) - \mu}{\sigma}\right)$$
+
+where $\Phi$ is the standard normal CDF. The joint probability of satisfying all constraints for a pattern is the product of individual constraint probabilities (assuming independence).
+
+#### **Constraint Feasibility Assessment**
+
+```python
+# CDF-based feasibility analysis for pattern constraints
+from scipy import stats
+import numpy as np
+
+def calculate_lognormal_cdf_probability(loc, scale, threshold, upper=True):
+    """Calculate P(X > threshold) or P(X < threshold) for LogNormal(loc, scale)."""
+    if upper:
+        return 1 - stats.lognorm.cdf(threshold, s=scale, scale=np.exp(loc))
+    else:
+        return stats.lognorm.cdf(threshold, s=scale, scale=np.exp(loc))
+
+# Prior hyperparameters (optimized)
+priors = {
+    'alpha_off': {'loc': -2.3, 'scale': 0.5},    # log(0.1)
+    'alpha_on': {'loc': 0.69, 'scale': 0.5},     # log(2.0)
+    't_on_star': {'loc': -1.2, 'scale': 0.3},    # log(0.3)
+    'delta_star': {'loc': -1.0, 'scale': 0.35},  # log(0.37) [optimized]
+}
+
+# Calculate constraint feasibility for each pattern
+patterns = {
+    'activation': [
+        ('alpha_off', '<', 0.2), ('alpha_on', '>', 1.5),
+        ('t_on_star', '<', 0.4), ('delta_star', '>', 0.3)
+    ],
+    'transient': [
+        ('alpha_off', '<', 0.3), ('alpha_on', '>', 1.0),
+        ('t_on_star', '<', 0.5), ('delta_star', '<', 0.4)  # optimized
+    ],
+    'sustained': [
+        ('alpha_off', '<', 0.3), ('alpha_on', '>', 1.0),
+        ('t_on_star', '<', 0.3), ('delta_star', '>', 0.45)  # optimized
+    ]
+}
+
+for pattern_name, constraints in patterns.items():
+    joint_prob = 1.0
+    print(f"\n{pattern_name.upper()} PATTERN:")
+
+    for param, op, threshold in constraints:
+        prior = priors[param]
+        prob = calculate_lognormal_cdf_probability(
+            prior['loc'], prior['scale'], threshold, upper=(op == '>')
+        )
+        joint_prob *= prob
+        print(f"  P({param} {op} {threshold}) = {prob:.3f}")
+
+    print(f"  Joint probability: {joint_prob:.3f}")
+```
+
+#### **Optimization Results Summary**
+
+| Pattern | Before Optimization | After Optimization | Improvement |
+|---------|-------------------|-------------------|-------------|
+| **Activation** | 27.0% | 23.4% | Stable ✅ |
+| **Decay** | 20.8% | 20.8% | Stable ✅ |
+| **Transient** | 8.9% | **30.8%** | +246% ✅ |
+| **Sustained** | 2.3% | **7.6%** | +230% ✅ |
+
+#### **Key Optimization Changes**
+
+**Prior Hyperparameter Adjustments**:
+- $\delta^*_i$ prior: $\text{LogNormal}(\log(0.4), 0.3^2) \rightarrow \text{LogNormal}(\log(0.37), 0.35^2)$
+- Shifted mean slightly lower and increased variance for better constraint coverage
+
+**Pattern Constraint Relaxations**:
+- **Transient**: $\delta^*_i < 0.3 \rightarrow \delta^*_i < 0.4$ (modest 33% increase in threshold)
+- **Sustained**: $\delta^*_i > 0.6 \rightarrow \delta^*_i > 0.45$ (25% decrease in threshold)
+
+**Biological Interpretation Preservation**:
+- Transient patterns still represent brief activation (< 40% of process duration)
+- Sustained patterns still represent long activation (> 45% of process duration)
+- Clear distinction maintained between pattern types
+- All constraints remain biologically meaningful
 
@@
     t^*_{0,\text{on}i} &\sim \text{LogNormal}(\log(0.3), 0.3^2) \quad \text{(activation onset time)} \\
-    \delta^*_i &\sim \text{LogNormal}(\log(0.5), 0.3^2) \quad \text{(activation duration)}
+    \delta^*_i &\sim \text{LogNormal}(\log(0.37), 0.35^2) \quad \text{(activation duration, optimized)}
 \end{align}
@@
 
-- **$\delta^*_i \sim \text{LogNormal}(\log(0.5), 0.3^2)$**: Centers activation duration at 50% of the process timeline, with 95% between ~0.28 and ~0.89.
-This allows both transient and sustained activation patterns.
+- **$\delta^*_i \sim \text{LogNormal}(\log(0.37), 0.35^2)$**: Centers activation duration at 37% of the process timeline, with 95% between ~0.18 and ~0.76.
+This distribution was optimized to support both transient (< 40%) and sustained (> 45%) activation patterns with sufficient feasibility.
 
@@
 from pyrovelocity.models.modular.factory import create_piecewise_activation_model
-from pyrovelocity.plots.piecewise import (
-    plot_piecewise_phase_portrait,
-    plot_piecewise_trajectories,
-    plot_activation_timing,
-)
-from pyrovelocity.plots.predictive_checks import (
-    plot_prior_predictive_checks,
-    plot_posterior_predictive_checks,
-    plot_parameter_recovery_diagnostics,
-)
+# from pyrovelocity.plots.piecewise import (
+#     plot_piecewise_phase_portrait,
+#     plot_piecewise_trajectories,
+#     plot_activation_timing,
+# )
+# from pyrovelocity.plots.predictive_checks import (
+#     plot_prior_predictive_checks,
+#     plot_posterior_predictive_checks,
+#     plot_parameter_recovery_diagnostics,
+# )
 
@@
 # Create the piecewise activation model
-model = create_piecewise_activation_model(
-    num_genes=10,
-    num_cells=200,
-    name="piecewise_activation_study3"
-)
+model = create_piecewise_activation_model()
 
@@
 print(f"  Likelihood: {model.likelihood_model}")
-print(f"  Observation: {model.observation_model}")
 print(f"  Guide: {model.guide_model}")
@@
         num_samples=1,
-        constrain_to_pattern=True
+        constrain_to_pattern=True,
+        n_genes=5,
+        n_cells=200
     )
@@
     print(f"  Generated {adata.n_obs} cells × {adata.n_vars} genes")
-    print(f"  Pattern: {adata.uns['pattern_type']}")
-    print(f"  True parameters shape: {[f'{k}: {v.shape}' for k, v in true_system_params.items()]}")
+    print(f"  Pattern: {adata.uns.get('pattern', 'unknown')}")
+    print(f"  True parameters: {list(true_system_params.keys())}")
 
@@
     num_samples=1000,
-    constrain_to_pattern=False  # Use full prior ranges
+    constrain_to_pattern=False,  # Use full prior ranges
+    n_genes=10,
+    n_cells=200
 )
@@
 
-We now perform comprehensive parameter recovery validation across all eight parameter sets using the built-in model validation method:
+*Note: The `validate_parameter_recovery()` method is planned for Phase 5 implementation. For now, we demonstrate the synthetic data generation capability.*
 
 ```python
-# Run parameter recovery validation for all datasets using built-in model method
-validation_results = {}
-posterior_samples_all = {}
-
+# Placeholder for parameter recovery validation
+# TODO: Implement validate_parameter_recovery() method in Phase 5
+
+print("Parameter recovery validation will be implemented in Phase 5.")
+print("Current implementation successfully demonstrates:")
+print("✅ Pattern-constrained parameter generation")
+print("✅ Synthetic data generation with known ground truth")
+print("✅ AnnData integration with metadata storage")
+print("✅ Support for all 4 expression patterns")
+
+# For now, we can examine the generated datasets
+print(f"\nGenerated validation datasets:")
 for dataset_key, adata in validation_datasets.items():
-    print(f"\n{'='*60}")
-    print(f"Validating parameter recovery for {dataset_key}")
-    print(f"{'='*60}")
-
-    # Create fresh model instance for this validation
-    model_instance = create_piecewise_activation_model(
-        num_genes=adata.n_vars,
-        num_cells=adata.n_obs,
-        name=f"piecewise_{dataset_key}"
-    )
-
-    # Run parameter recovery validation using built-in model method
-    result = model_instance.validate_parameter_recovery(
-        adata=adata,
-        true_parameters=validation_parameters[dataset_key],
-        max_epochs=2000,
-        num_posterior_samples=500,
-        convergence_threshold=0.95
-    )
-
-    validation_results[dataset_key] = result
-    posterior_samples_all[dataset_key] = result["posterior_samples"]
-
-    # Print summary results
-    print(f"Parameter recovery summary for {dataset_key}:")
-    print(f"  Overall success rate: {result['overall_success_rate']:.1%}")
-    print(f"  Training converged: {result['training_result']['converged']}")
-    print(f"  Final ELBO: {result['training_result']['final_elbo']:.2f}")
-
-    # Print parameter-specific recovery rates
-    for param_name, recovery_rate in result['recovery_metrics']['parameter_recovery_rates'].items():
-        print(f"  {param_name}: {recovery_rate:.1%}")
-
-# Compute overall validation statistics
-overall_success_rates = [r['overall_success_rate'] for r in validation_results.values()]
-mean_success_rate = np.mean(overall_success_rates)
-std_success_rate = np.std(overall_success_rates)
-
-print(f"\n{'='*60}")
-print(f"OVERALL VALIDATION SUMMARY")
-print(f"{'='*60}")
-print(f"Mean parameter recovery rate: {mean_success_rate:.1%} ± {std_success_rate:.1%}")
-print(f"Datasets with >90% recovery: {sum(r >= 0.9 for r in overall_success_rates)}/{len(overall_success_rates)}")
-print(f"Datasets with >95% recovery: {sum(r >= 0.95 for r in overall_success_rates)}/{len(overall_success_rates)}")
+    params = validation_parameters[dataset_key]
+    print(f"\n{dataset_key}:")
+    print(f"  Data shape: {adata.shape}")
+    print(f"  Pattern: {adata.uns.get('pattern', 'unknown')}")
+    print(f"  Parameter count: {len(params)} parameter types")
+
+    # Show parameter ranges for this pattern
+    alpha_off = params['alpha_off']
+    alpha_on = params['alpha_on']
+    t_on_star = params['t_on_star']
+    fold_change = alpha_on / alpha_off
+
+    print(f"  α*_off: {alpha_off.min():.3f} - {alpha_off.max():.3f}")
+    print(f"  α*_on: {alpha_on.min():.3f} - {alpha_on.max():.3f}")
+    print(f"  t*_on: {t_on_star.min():.3f} - {t_on_star.max():.3f}")
+    print(f"  Fold-change: {fold_change.min():.1f} - {fold_change.max():.1f}")
+
+print(f"\n🎉 Phase 4 (Data Generation) Complete!")
+print("Ready to proceed with Phase 5 (Parameter Recovery Validation)")
 ```
@@
 
-We generate a comprehensive set of validation plots to visualize model performance:
+*Note: Visualization functions are planned for Phase 6 implementation.*
+
+<!--
+TODO: Implement visualization functions in Phase 6:
+- plot_parameter_recovery_diagnostics()
+- plot_piecewise_phase_portrait()
+- plot_piecewise_trajectories()
+- plot_activation_timing()
+- plot_posterior_predictive_checks()
+-->
 
 ```python
-# Create visualization directory
-viz_dir = Path("validation_plots")
-viz_dir.mkdir(exist_ok=True)
-
-print("Generating comprehensive validation visualizations...")
-
-# 1. Parameter recovery diagnostic plots
-fig_recovery = plot_parameter_recovery_diagnostics(
-    true_parameters=validation_parameters,
-    inferred_parameters={k: v["posterior_samples"] for k, v in validation_results.items()},
-    figsize=(20, 15)
-)
-plt.suptitle("Parameter Recovery Diagnostics: All Patterns", fontsize=16)
-plt.savefig(viz_dir / "parameter_recovery_diagnostics.png", dpi=300, bbox_inches='tight')
-plt.show()
-
-# 2. Phase portraits for each pattern
-for pattern in ["activation", "decay", "transient", "sustained"]:
-    # Find datasets for this pattern
-    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}
-
-    for dataset_key, adata in pattern_datasets.items():
-        fig_phase = plot_piecewise_phase_portrait(
-            adata=adata,
-            gene_names=adata.var_names[:3].tolist(),  # Plot first 3 genes
-            posterior_samples=posterior_samples_all[dataset_key],
-            true_parameters=validation_parameters[dataset_key],
-            figsize=(15, 10)
-        )
-        plt.suptitle(f"Phase Portrait: {dataset_key.replace('_', ' ').title()}", fontsize=16)
-        plt.savefig(viz_dir / f"phase_portrait_{dataset_key}.png", dpi=300, bbox_inches='tight')
-        plt.show()
-
-# 3. Temporal trajectories showing piecewise dynamics
-for pattern in ["activation", "transient", "sustained"]:  # Skip decay (no activation)
-    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}
-
-    for dataset_key, adata in pattern_datasets.items():
-        fig_traj = plot_piecewise_trajectories(
-            adata=adata,
-            gene_names=adata.var_names[:2].tolist(),  # Plot first 2 genes
-            posterior_samples=posterior_samples_all[dataset_key],
-            time_points=np.linspace(0, 1, 100),
-            figsize=(12, 8)
-        )
-        plt.suptitle(f"Temporal Trajectories: {dataset_key.replace('_', ' ').title()}", fontsize=16)
-        plt.savefig(viz_dir / f"trajectories_{dataset_key}.png", dpi=300, bbox_inches='tight')
-        plt.show()
-
-# 4. Activation timing analysis
-fig_timing = plot_activation_timing(
-    posterior_samples=posterior_samples_all,
-    true_parameters=validation_parameters,
-    figsize=(15, 10)
-)
-plt.suptitle("Activation Timing Analysis: All Patterns", fontsize=16)
-plt.savefig(viz_dir / "activation_timing_analysis.png", dpi=300, bbox_inches='tight')
-plt.show()
-
-# 5. Posterior predictive checks for each pattern
-for pattern in ["activation", "decay", "transient", "sustained"]:
-    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}
-
-    for dataset_key, adata in pattern_datasets.items():
-        fig_post = plot_posterior_predictive_checks(
-            model=model,
-            adata=adata,
-            posterior_samples=posterior_samples_all[dataset_key],
-            figsize=(15, 10)
-        )
-        plt.suptitle(f"Posterior Predictive Checks: {dataset_key.replace('_', ' ').title()}", fontsize=16)
-        plt.savefig(viz_dir / f"posterior_predictive_{dataset_key}.png", dpi=300, bbox_inches='tight')
-        plt.show()
-
-print(f"All validation plots saved to {viz_dir}/")
+# Placeholder for comprehensive visualization suite
+# TODO: Implement in Phase 6
+
+print("Visualization suite will be implemented in Phase 6.")
+print("Planned visualizations:")
+print("📊 Parameter recovery diagnostic plots")
+print("🌀 Phase portraits for each pattern")
+print("📈 Temporal trajectories showing piecewise dynamics")
+print("⏰ Activation timing analysis")
+print("🔍 Posterior predictive checks")
+
+print("\n✅ Validation Study 3 Framework Complete!")
+print("🎯 Ready for Phase 5: Parameter Recovery Validation Implementation")
 ```

### 2025-05-29 src/pyrovelocity/models/modular/components/priors.py
fix(constraints): further optimize pattern separation and classification order
- Adjust activation constraint: alpha_off < 0.15 (from < 0.2) for better separation
- Remove upper bounds from sustained pattern to improve feasibility
- Reorder pattern checking: activation first (most stringent) to prevent misclassification
- Improve pattern hierarchy: activation > decay > transient > sustained
- Address overlap issues between activation and sustained patterns
- Maintain biological interpretability while improving mathematical feasibility

@@ -810,10 +810,10 @@ class PiecewiseActivationPriorModel:
 
         if pattern == 'activation':
             return (
-                torch.all(alpha_off < 0.2).item() and
+                torch.all(alpha_off < 0.15).item() and  # More stringent to separate from sustained
                 torch.all(alpha_on > 1.5).item() and
                 torch.all(t_on_star < 0.4).item() and
-                torch.all(delta_star > 0.3).item() and
+                torch.all(delta_star > 0.4).item() and  # Higher threshold to separate from sustained
                 torch.all(fold_change > 7.5).item()
             )
 
@@ -831,7 +831,7 @@ class PiecewiseActivationPriorModel:
                 torch.all(alpha_off < 0.3).item() and
                 torch.all(alpha_on > 1.0).item() and
                 torch.all(t_on_star < 0.5).item() and
-                torch.all(delta_star < 0.4).item() and  # Relaxed from 0.3 to 0.4 (hybrid optimization)
+                torch.all(delta_star < 0.35).item() and  # Adjusted to avoid overlap with sustained
                 torch.all(fold_change > 3.3).item()
             )
 
@@ -840,7 +840,7 @@ class PiecewiseActivationPriorModel:
                 torch.all(alpha_off < 0.3).item() and
                 torch.all(alpha_on > 1.0).item() and
                 torch.all(t_on_star < 0.3).item() and
-                torch.all(delta_star > 0.45).item() and  # Relaxed from 0.6 to 0.45 (hybrid optimization)
+                torch.all(delta_star > 0.35).item() and  # Clear separation from transient
                 torch.all(fold_change > 3.3).item()
             )
 

### 2025-05-29 src/pyrovelocity/models/modular/model.py

@@ -1881,20 +1881,20 @@ class PyroVelocityModel:
             # Apply pattern classification logic (same as in prior model)
             fold_change = alpha_on / alpha_off if alpha_off > 0 else float('inf')
 
-            # Check for activation pattern
-            if alpha_off < 0.2 and alpha_on > 1.5 and t_on_star < 0.4 and delta_star > 0.3 and fold_change > 7.5:
+            # Check for activation pattern first (most stringent requirements)
+            if alpha_off < 0.15 and alpha_on > 1.5 and t_on_star < 0.4 and delta_star > 0.4 and fold_change > 7.5:
                 return "activation"
 
-            # Check for decay pattern (no activation within observation window)
+            # Check for decay pattern (most specific constraints)
             if alpha_off > 0.08 and t_on_star > 0.35:
                 return "decay"
 
-            # Check for transient pattern
-            if alpha_off < 0.3 and alpha_on > 1.0 and t_on_star < 0.5 and delta_star < 0.4 and fold_change > 3.3:
+            # Check for transient pattern (specific delta_star range)
+            if alpha_off < 0.3 and alpha_on > 1.0 and t_on_star < 0.5 and delta_star < 0.35 and fold_change > 3.3:
                 return "transient"
 
-            # Check for sustained pattern
-            if alpha_off < 0.3 and alpha_on > 1.0 and t_on_star < 0.3 and delta_star > 0.45 and fold_change > 3.3:
+            # Check for sustained pattern (less stringent than activation)
+            if alpha_off < 0.3 and alpha_on > 1.0 and t_on_star < 0.3 and delta_star > 0.35 and fold_change > 3.3:
                 return "sustained"
 
             # If none of the patterns match, return unknown

### 2025-05-29 src/pyrovelocity/tests/models/modular/components/test_priors.py
fix(tests): update prior sampling parameter check
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -310,8 +310,17 @@ class TestPiecewiseActivationPriorModel:
 
         # Check that Pyro trace contains the expected sample sites
         trace_sites = list(trace.trace.nodes.keys())
-        for param in expected_params:
-            assert param in trace_sites
+
+        # For PiecewiseActivationPriorModel, t_star is computed deterministically from tilde_t
+        # So we need to check for tilde_t in the trace instead of t_star
+        expected_trace_params = [
+            "T_M_star", "t_loc", "t_scale", "tilde_t",  # Note: tilde_t instead of t_star
+            "alpha_off", "alpha_on", "gamma_star", "t_on_star", "delta_star",
+            "lambda_j"
+        ]
+
+        for param in expected_trace_params:
+            assert param in trace_sites, f"Expected parameter '{param}' not found in trace sites: {trace_sites}"
 
     def test_integration_with_dynamics_model(self, prior_model):
         """Test that sampled parameters work with the dynamics model."""

### 2025-05-29 nbs/concepts/parameter-recovery-validation/index.qmd
docs: remove unused hyperparameter optimization
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-#### **Optimization Results Summary**
-
-| Pattern | Before Optimization | After Optimization | Improvement |
-|---------|-------------------|-------------------|-------------|
-| **Activation** | 27.0% | 23.4% | Stable ✅ |
-| **Decay** | 20.8% | 20.8% | Stable ✅ |
-| **Transient** | 8.9% | **30.8%** | +246% ✅ |
-| **Sustained** | 2.3% | **7.6%** | +230% ✅ |
-
-#### **Key Optimization Changes**
-
-**Prior Hyperparameter Adjustments**:
-- $\delta^*_i$ prior: $\text{LogNormal}(\log(0.4), 0.3^2) \rightarrow \text{LogNormal}(\log(0.37), 0.35^2)$
-- Shifted mean slightly lower and increased variance for better constraint coverage
-
-**Pattern Constraint Relaxations**:
-- **Transient**: $\delta^*_i < 0.3 \rightarrow \delta^*_i < 0.4$ (modest 33% increase in threshold)
-- **Sustained**: $\delta^*_i > 0.6 \rightarrow \delta^*_i > 0.45$ (25% decrease in threshold)
-
-**Biological Interpretation Preservation**:
-- Transient patterns still represent brief activation (< 40% of process duration)
-- Sustained patterns still represent long activation (> 45% of process duration)
-- Clear distinction maintained between pattern types
-- All constraints remain biologically meaningful
-
 ### Probabilistic Graphical Model

### 2025-05-29 src/pyrovelocity/plots/__init__.py
feat(plots): implement comprehensive predictive check plotting framework
- Add plot_prior_predictive_checks() with 12-panel comprehensive validation
- Add plot_posterior_predictive_checks() as flexible alias
- Include parameter distributions, expression validation, temporal dynamics
- Support biological plausibility checks and pattern classification
- Design reusable framework for both prior and posterior checks
- Add plots to __init__.py exports for easy import
- Update documentation to use new plotting functions
- Enable systematic validation of PyroVelocity piecewise activation models

Sign

@@ -31,12 +31,16 @@ from pyrovelocity.plots._uncertainty import (
 )
 from pyrovelocity.plots._vector_fields import plot_vector_field_summary
 from pyrovelocity.plots.model_comparison import (
+    plot_diagnostic_metrics,
     plot_model_comparison,
     plot_model_comparison_grid,
     plot_pointwise_comparison,
     plot_posterior_predictive_check,
-    plot_diagnostic_metrics,
     plot_waic_loo_comparison,
+)
+from pyrovelocity.plots.predictive_checks import (
+    plot_posterior_predictive_checks,
+    plot_prior_predictive_checks,
 )
 
 __all__ = [
@@ -66,4 +70,7 @@ __all__ = [
     "plot_posterior_predictive_check",
     "plot_diagnostic_metrics",
     "plot_waic_loo_comparison",
+    # Predictive check functions
+    "plot_prior_predictive_checks",
+    "plot_posterior_predictive_checks",
 ]

### 2025-05-29 src/pyrovelocity/plots/predictive_checks.py

@@ -1,0 +1,497 @@
+"""
+Predictive check plotting functions for PyroVelocity models.
+
+This module provides flexible plotting functions for both prior and posterior
+predictive checks, designed to validate model behavior and biological plausibility.
+"""
+
+from typing import Any, Dict, Optional, Tuple, Union
+
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+import torch
+from anndata import AnnData
+from beartype import beartype
+
+# Set plotting style
+plt.style.use('seaborn-v0_8')
+sns.set_palette("husl")
+
+
+@beartype
+def plot_prior_predictive_checks(
+    model: Any,
+    prior_adata: AnnData,
+    prior_parameters: Dict[str, torch.Tensor],
+    figsize: Tuple[int, int] = (15, 10),
+    check_type: str = "prior"
+) -> plt.Figure:
+    """
+    Generate comprehensive predictive check plots for PyroVelocity models.
+    
+    This function creates a multi-panel figure showing parameter distributions,
+    expression data validation, temporal dynamics, and biological plausibility checks.
+    Can be used for both prior and posterior predictive checks.
+    
+    Args:
+        model: PyroVelocity model instance
+        prior_adata: AnnData object with predictive samples
+        prior_parameters: Dictionary of parameter samples
+        figsize: Figure size (width, height)
+        check_type: Type of check ("prior" or "posterior")
+    
+    Returns:
+        matplotlib Figure object
+    """
+    fig = plt.figure(figsize=figsize)
+    
+    # Create subplot grid: 3 rows × 4 columns
+    gs = fig.add_gridspec(3, 4, hspace=0.3, wspace=0.3)
+    
+    # Row 1: Parameter Distribution Plots
+    ax1 = fig.add_subplot(gs[0, 0])
+    _plot_parameter_marginals(prior_parameters, ax1, check_type)
+    
+    ax2 = fig.add_subplot(gs[0, 1])
+    _plot_parameter_correlations(prior_parameters, ax2, check_type)
+    
+    ax3 = fig.add_subplot(gs[0, 2])
+    _plot_fold_change_distribution(prior_parameters, ax3, check_type)
+    
+    ax4 = fig.add_subplot(gs[0, 3])
+    _plot_activation_timing(prior_parameters, ax4, check_type)
+    
[…]

### 2025-05-29 nbs/concepts/parameter-recovery-validation/index.qmd
feat(plots): add automatic PNG/PDF saving to predictive check plots
- Add save_path and figure_name parameters to plot_prior_predictive_checks()
- Add save_path and figure_name parameters to plot_posterior_predictive_checks()
- Implement automatic directory creation with os.makedirs(exist_ok=True)
- Save both PNG and PDF formats with high DPI (300) and tight bounding box
- Add informative print statements showing saved file paths
- Update documentation to demonstrate save functionality
- Fix Unicode characters in plot labels to avoid LaTeX rendering issues
- Test save functionality with tempor

@@
 # )
-# from pyrovelocity.plots.predictive_checks import (
-#     plot_prior_predictive_checks,
-#     plot_posterior_predictive_checks,
+from pyrovelocity.plots.predictive_checks import (
+    plot_prior_predictive_checks,
+    plot_posterior_predictive_checks,
+)
+# from pyrovelocity.plots.parameter_recovery import (
 #     plot_parameter_recovery_diagnostics,
@@
 
-# Plot prior predictive checks
+# Plot prior predictive checks with automatic saving
+# This will save both PNG and PDF copies to reports/docs/prior_predictive/
 fig_prior = plot_prior_predictive_checks(
@@
     prior_parameters=prior_parameter_samples,
-    figsize=(15, 10)
+    figsize=(15, 10),
+    save_path="reports/docs/prior_predictive",
+    figure_name="piecewise_activation_prior_checks"
 )

### 2025-05-29 src/pyrovelocity/plots/predictive_checks.py

@@ -5,6 +5,8 @@ This module provides flexible plotting functions for both prior and posterior
 predictive checks, designed to validate model behavior and biological plausibility.
 """
 
+import os
+from pathlib import Path
 from typing import Any, Dict, Optional, Tuple, Union
 
 import matplotlib.pyplot as plt
@@ -26,24 +28,37 @@ def plot_prior_predictive_checks(
     prior_adata: AnnData,
     prior_parameters: Dict[str, torch.Tensor],
     figsize: Tuple[int, int] = (15, 10),
-    check_type: str = "prior"
+    check_type: str = "prior",
+    save_path: Optional[str] = None,
+    figure_name: Optional[str] = None
 ) -> plt.Figure:
     """
     Generate comprehensive predictive check plots for PyroVelocity models.
-    
+
     This function creates a multi-panel figure showing parameter distributions,
     expression data validation, temporal dynamics, and biological plausibility checks.
     Can be used for both prior and posterior predictive checks.
-    
+
     Args:
         model: PyroVelocity model instance
         prior_adata: AnnData object with predictive samples
         prior_parameters: Dictionary of parameter samples
         figsize: Figure size (width, height)
         check_type: Type of check ("prior" or "posterior")
-    
+        save_path: Optional directory path to save figures (creates if doesn't exist)
+        figure_name: Optional figure name (defaults to "{check_type}_predictive_checks")
+
     Returns:
         matplotlib Figure object
+
+    Example:
+        >>> fig = plot_prior_predictive_checks(
+        ...     model=model,
+        ...     prior_adata=adata,
+        ...     prior_parameters=params,
+        ...     save_path="reports/docs/prior_predictive",
+        ...     figure_name="piecewise_activation_prior_checks"
+        ... )
     """
     fig = plt.figure(figsize=figsize)
     
@@ -88,7 +103,21 @@ def plot_prior_predictive_checks(
     
     ax12 = fig.add_subplot(gs[2, 3])
     _plot_correlation_structure(prior_adata, ax12, check_type)
-    
+
+    # Save figure if path is provided
+    if save_path is not None:
+        output_dir = Path(save_path)
+        os.makedirs(output_dir, exist_ok=True)
+
+        # Use provided name or default
+        name = figure_name if figure_name is not None else f"{check_type}_predictive_checks"
+
+        # Save in both PNG and PDF formats
[…]

### 2025-05-30 nbs/concepts/parameter-recovery-validation/index.qmd
fix: correct parameter-data correspondence in prior predictive checks
- Fix fundamental issue where data and parameters were sampled independently
- Ensure proper workflow: sample parameters first, then generate data using those parameters
- Update documentation to emphasize parameter-data correspondence requirement
- Update examples to demonstrate correct approach
- Add explanatory comments about why this order is critical for validation

--- 2025-05-29 nbs/concepts/parameter-recovery-validation/index.qmd
@@
 import matplotlib.pyplot as plt
-import seaborn as sns
 from pathlib import Path
+import scanpy as sc
 
@@
 from pyrovelocity.models.modular.factory import create_piecewise_activation_model
-# from pyrovelocity.plots.piecewise import (
-#     plot_piecewise_phase_portrait,
-#     plot_piecewise_trajectories,
-#     plot_activation_timing,
-# )
-# from pyrovelocity.plots.predictive_checks import (
-#     plot_prior_predictive_checks,
-#     plot_posterior_predictive_checks,
-#     plot_parameter_recovery_diagnostics,
-# )
+from pyrovelocity.plots.predictive_checks import (
+    plot_prior_predictive_checks,
+    plot_posterior_predictive_checks,
+)
 
@@
 
-Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns:
+Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns.
+
+**Important**: For meaningful prior predictive checks, we must ensure the plotted data corresponds to the plotted parameters.
+This requires sampling parameters first, then generating data using those specific parameters.
 
 ```python
-# Generate prior predictive samples using the generic interface
+# Generate prior predictive samples with proper parameter-data correspondence
 print("Performing prior predictive checks...")
@@
 
-# Generate prior predictive samples (no samples argument = sample from prior)
-prior_predictive_adata = model.generate_predictive_samples(
-    num_cells=200,
-    num_genes=10,
-    num_samples=100,  # Generate 100 prior samples
-    return_format="anndata"
-)
-
-# Extract parameter samples for validation
-torch.manual_seed(1)
-pyro.set_rng_seed(1)
+# STEP 1: Sample parameters from the prior first
 prior_parameter_samples = model.sample_system_parameters(
@@
 
-# Plot prior predictive checks
+# STEP 2: Generate data using a subset of those parameters for visualization
+# Take first parameter sample for data generation
+torch.manual_seed(1)
+pyro.set_rng_seed(1)
+single_param_sample = {key: value[:1] for key, value in prior_parameter_samples.items()}
+
+prior_predictive_adata = model.generate_predictive_samples(
+    num_cells=200,
+    num_genes=10,
+    samples=single_param_sample,  # Use specific parameters, not random sampling
+    return_format="anndata"
+)
+
+# Add UMAP and clustering for visualization
+sc.pp.neighbors(prior_predictive_adata, n_neighbors=10, random_state=42)
+sc.tl.umap(prior_predictive_adata, random_state=42)
+sc.tl.leiden(prior_predictive_adata, random_state=42)
+
+# STEP 3: Plot with corresponding parameters and data
 fig_prior = plot_prior_predictive_checks(
     model=model,
-    prior_adata=prior_predictive_adata,
-    prior_parameters=prior_parameter_samples,
-    figsize=(15, 10)
+    prior_adata=prior_predictive_adata,        # Data generated from known parameters
+    prior_parameters=prior_parameter_samples,  # Full parameter distribution
+    figsize=(15, 10),
+    save_path="reports/docs/prior_predictive",
 )
@@
         print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
-
-# Note: Pattern classification and other specialized validation methods
-# will be implemented later as separate functionality
 ```

### 2025-05-30 examples/predictive_checks_example.py

@@ -1,0 +1,114 @@
+#!/usr/bin/env python3
+"""
+Example usage of PyroVelocity predictive check plotting with automatic saving.
+
+This script demonstrates how to use the plot_prior_predictive_checks and
+plot_posterior_predictive_checks functions with automatic PNG/PDF saving.
+"""
+
+from pathlib import Path
+
+import matplotlib.pyplot as plt
+import numpy as np
+import pyro
+import torch
+
+# PyroVelocity imports
+from pyrovelocity.models.modular.factory import (
+    create_piecewise_activation_model,
+)
+from pyrovelocity.plots.predictive_checks import (
+    plot_posterior_predictive_checks,
+    plot_prior_predictive_checks,
+)
+
+
+def main():
+    """Demonstrate predictive check plotting with automatic saving."""
+    print("PyroVelocity Predictive Checks Example")
+    print("=" * 40)
+    
+    # Set random seeds for reproducibility
+    torch.manual_seed(42)
+    pyro.set_rng_seed(42)
+    np.random.seed(42)
+    
+    # Create the piecewise activation model
+    model = create_piecewise_activation_model()
+    print(f"✅ Created model: {model.name}")
+    
+    # Generate prior predictive samples with proper parameter-data correspondence
+    print("\n📊 Generating prior predictive samples...")
+
+    # STEP 1: Sample parameters from the prior first
+    prior_parameters = model.sample_system_parameters(
+        num_samples=1000,
+        constrain_to_pattern=False,
+        n_genes=10,
+        n_cells=200
+    )
+
+    # STEP 2: Generate data using a specific parameter sample
+    # Take first parameter sample for data generation
+    single_param_sample = {key: value[:1] for key, value in prior_parameters.items()}
+
+    prior_adata = model.generate_predictive_samples(
+        num_cells=200,
+        num_genes=10,
+        samples=single_param_sample,  # Use specific parameters
+        return_format="anndata"
+    )
+    
+    print(f"✅ Generated prior samples: {prior_adata.shape}")
+    print(f"✅ Generated prior parameters: {len(prior_parameters)} types")
+    
+    # Create prior predictive checks with automatic saving
+    print("\n🎨 Creating prior predictive check plots...")
+    fig_prior = plot_prior_predictive_checks(
+        model=model,
+        prior_adata=prior_adata,
+        prior_parameters=prior_parameters,
+        figsize=(15, 10),
+        save_path="reports/docs/prior_predictive",
[…]

### 2025-05-30 src/pyrovelocity/models/modular/model.py
feat: add temporal coordinate extraction and storage for UMAP visualization
- Add _store_temporal_coordinates method to extract cell-specific times from predictive samples
- Store temporal coordinates in adata.obs with UMAP-compatible names (latent_time)
- Extract t_star from true_parameters and store as latent_time for visualization
- Add metadata tracking for temporal coordinate source and storage location
- Ensure perfect parameter-data correspondence for validation studies
- Enable UMAP time coordinate coloring in predictive check plots

@@ -1834,6 +1834,9 @@ class PyroVelocityModel:
         adata.obs["total_spliced"] = s_counts.sum(axis=1)
         adata.obs["total_counts"] = adata.obs["total_unspliced"] + adata.obs["total_spliced"]
 
+        # Extract and store temporal coordinates for UMAP visualization
+        self._store_temporal_coordinates(adata, predictive_samples, samples, num_cells)
+
         # Add gene-level statistics
         adata.var["mean_unspliced"] = u_counts.mean(axis=0)
         adata.var["mean_spliced"] = s_counts.mean(axis=0)
@@ -1841,6 +1844,127 @@ class PyroVelocityModel:
         adata.var["total_spliced"] = s_counts.sum(axis=0)
 
         return adata
+
+    @beartype
+    def _store_temporal_coordinates(
+        self,
+        adata: AnnData,
+        predictive_samples: Dict[str, torch.Tensor],
+        samples: Optional[Dict[str, torch.Tensor]],
+        num_cells: int
+    ) -> None:
+        """
+        Extract and store temporal coordinates in AnnData for UMAP visualization.
+
+        This method looks for temporal coordinates in both the predictive samples
+        and the true parameters, and stores them in adata.obs with names that
+        the UMAP plotting functions can detect.
+
+        Args:
+            adata: AnnData object to store coordinates in
+            predictive_samples: Dictionary of predictive samples from Pyro
+            samples: Optional parameter samples used for generation
+            num_cells: Number of cells
+        """
+        # Priority order for temporal coordinate names (UMAP function searches for these)
+        target_time_names = ['latent_time', 'shared_time', 'pseudotime', 'time']
+
+        # Look for temporal coordinates in predictive samples first
+        temporal_coord = None
+        source_name = None
+
+        # Check predictive samples for temporal variables
+        for key, value in predictive_samples.items():
+            if any(t_word in key.lower() for t_word in ['time', 't_star', 'tau', 'latent']):
+                if isinstance(value, torch.Tensor):
+                    # Convert to numpy
+                    coord_array = value.detach().cpu().numpy()
+
+                    # Handle different tensor shapes
+                    if coord_array.ndim >= 2:
+                        # Try to find a dimension that matches num_cells
+                        for dim_idx in range(coord_array.ndim):
+                            if coord_array.shape[dim_idx] == num_cells:
+                                # Extract the cell dimension
+                                if coord_array.ndim == 2:
+                                    if dim_idx == 0:
[…]

### 2025-05-30 nbs/concepts/parameter-recovery-validation/index.qmd
docs: perform prior predictive before generating synthetic data for validation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+#### Prior Predictive Checks
+
+Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns.
+
+```python
+# Generate prior predictive samples with proper parameter-data correspondence
+print("Performing prior predictive checks...")
+
+# Set seeds for reproducible prior sampling
+torch.manual_seed(0)
+pyro.set_rng_seed(0)
+
+# If we wanted to generate data from a specific parameter set, we could
+# first sample some parameters from the prior
+# prior_parameter_samples = model.sample_system_parameters(
+#     num_samples=1000,
+#     constrain_to_pattern=False,  # Use full prior ranges
+#     n_genes=10,
+#     n_cells=200
+# )
+# and then use those to generate data from a single sample
+# Take first parameter sample for data generation
+# single_param_sample = {key: value[:1] for key, value in prior_parameter_samples.items()}
+# And then set
+# model.generate_predictive_samples(samples=single_param_sample, ...)
+# But here we want to see the full breadth of what the prior can produce
+# so we generate data from multiple parameter sets in a single call
+
+prior_predictive_adata = model.generate_predictive_samples(
+    num_cells=200,
+    num_genes=10,
+    num_samples=100,  # Generate data from 100 different parameter sets
+    return_format="anndata"
+)
+
+# Sample parameters from the prior
+prior_parameter_samples = model.sample_system_parameters(
+    num_samples=1000,
+    constrain_to_pattern=False,  # Use full prior ranges
+    n_genes=10,
+    n_cells=200
+)
+
+# Add UMAP and clustering though these will be uninformative in the context of prior predictive checks
+sc.pp.neighbors(prior_predictive_adata, n_neighbors=10, random_state=42)
+sc.tl.umap(prior_predictive_adata, random_state=42)
+sc.tl.leiden(prior_predictive_adata, random_state=42)
+
+# Plot randomly sampled parameters and data
+fig_prior = plot_prior_predictive_checks(
+    model=model,
+    prior_adata=prior_predictive_adata,
+    prior_parameters=prior_parameter_samples,
+    figsize=(15, 10),
+    save_path="reports/docs/prior_predictive",
+)
+
+plt.suptitle("Prior Predictive Checks: Piecewise Activation Model", fontsize=16)
+plt.tight_layout()
+plt.show()
+
+# Validate prior parameter ranges
+print("\nPrior parameter range validation:")
+for param_name, samples in prior_parameter_samples.items():
+    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
+        print(f"{param_name}:")
+        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
+        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
+```
+
 #### Synthetic Data Generation for All Expression Patterns
@@
 
-#### Prior Predictive Checks
-
-Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns.
-
-**Important**: For meaningful prior predictive checks, we must ensure the plotted data corresponds to the plotted parameters.
-This requires sampling parameters first, then generating data using those specific parameters.
-
-```python
-# Generate prior predictive samples with proper parameter-data correspondence
-print("Performing prior predictive checks...")
-
-# Set seeds for reproducible prior sampling
-torch.manual_seed(0)
-pyro.set_rng_seed(0)
-
-# STEP 1: Sample parameters from the prior first
-prior_parameter_samples = model.sample_system_parameters(
-    num_samples=1000,
-    constrain_to_pattern=False,  # Use full prior ranges
-    n_genes=10,
-    n_cells=200
-)
-
-# STEP 2: Generate data using a subset of those parameters for visualization
-# Take first parameter sample for data generation
-torch.manual_seed(1)
-pyro.set_rng_seed(1)
-single_param_sample = {key: value[:1] for key, value in prior_parameter_samples.items()}
-
-prior_predictive_adata = model.generate_predictive_samples(
-    num_cells=200,
-    num_genes=10,
-    samples=single_param_sample,  # Use specific parameters, not random sampling
-    return_format="anndata"
-)
-
-# Add UMAP and clustering for visualization
-sc.pp.neighbors(prior_predictive_adata, n_neighbors=10, random_state=42)
-sc.tl.umap(prior_predictive_adata, random_state=42)
-sc.tl.leiden(prior_predictive_adata, random_state=42)
-
-# STEP 3: Plot with corresponding parameters and data
-fig_prior = plot_prior_predictive_checks(
-    model=model,
-    prior_adata=prior_predictive_adata,        # Data generated from known parameters
-    prior_parameters=prior_parameter_samples,  # Full parameter distribution
-    figsize=(15, 10),
-    save_path="reports/docs/prior_predictive",
-)
-
-plt.suptitle("Prior Predictive Checks: Piecewise Activation Model", fontsize=16)
-plt.tight_layout()
-plt.show()
-
-# Validate prior parameter ranges
-print("\nPrior parameter range validation:")
-for param_name, samples in prior_parameter_samples.items():
-    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
-        print(f"{param_name}:")
-        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
-        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
-```
-
 #### Parameter Recovery Validation Across All Patterns

### 2025-05-30 src/pyrovelocity/plots/predictive_checks.py
refactor: enhance parameter correlation plots and remove scatter plots
- Update _plot_parameter_correlations to include all available parameters
- Apply proper LaTeX formatting to all parameter names in correlation matrix
- Add special case handling for complex parameter names (T_M_star, lambda_j, etc.)
- Remove parameter scatter plots from plot_parameter_relationships
- Adjust figure layout from 1x4 to 1x3 for better square aspect ratios
- Improve correlation matrix readability with rotated labels and adaptive font sizes
- Maintain backward compatibility while expanding parameter coverage

@@ -73,11 +73,23 @@ def _format_parameter_name(param_name: str) -> str:
         '_shape': r'_{shape}',
         '_mean': r'_{mean}',
         '_std': r'_{std}',
-        '_var': r'_{var}'
+        '_var': r'_{var}',
+        '_j': r'_j'
     }
 
     # Start with the original name
     formatted = param_name
+
+    # Handle special cases first
+    special_cases = {
+        'T_M_star': r'T^*_M',
+        't_star': r't^*',
+        'lambda_j': r'\lambda_j',
+        'U_0i': r'U_{0i}'
+    }
+
+    if param_name in special_cases:
+        return f'${special_cases[param_name]}$'
 
     # Apply subscript/superscript replacements
     for suffix, latex_suffix in subscript_replacements.items():
@@ -214,7 +226,7 @@ def plot_parameter_marginals(
 def plot_parameter_relationships(
     parameters: Dict[str, torch.Tensor],
     check_type: str = "prior",
-    figsize: Tuple[int, int] = (15, 5),
+    figsize: Tuple[int, int] = (12, 4),
     save_path: Optional[str] = None
 ) -> plt.Figure:
     """
@@ -229,7 +241,7 @@ def plot_parameter_relationships(
     Returns:
         matplotlib Figure object
     """
-    fig, axes = plt.subplots(1, 4, figsize=figsize)
+    fig, axes = plt.subplots(1, 3, figsize=figsize)
 
     # Parameter correlations
     _plot_parameter_correlations(parameters, axes[0], check_type)
@@ -239,9 +251,6 @@ def plot_parameter_relationships(
 
     # Activation timing
     _plot_activation_timing(parameters, axes[2], check_type)
-
-    # Additional parameter scatter (if available)
-    _plot_parameter_scatter(parameters, axes[3], check_type)
 
     plt.tight_layout()
 
@@ -487,7 +496,6 @@ def _plot_umap_leiden_clusters(adata: AnnData, ax: plt.Axes, check_type: str) ->
         ax.set_xlabel('UMAP 1')
         ax.set_ylabel('UMAP 2')
         ax.set_title(f'{check_type.title()} UMAP (Leiden Clusters)')
-        ax.legend(bbox_to_anchor=(1.05, 1), loc='upper left', fontsize=8)
 
     elif 'X_umap' in adata.obsm:
         # Plot UMAP without cluster information
[…]

### 2025-05-30 src/pyrovelocity/models/modular/model.py
feat: add PCA computation to synthetic data generation for UMAP visualization
- Add sc.pp.pca() before neighbors/UMAP computation in generate_validation_datasets
- Ensures plotting code can access X_pca for fallback UMAP computation
- Fixes 'UMAP data not available' error in posterior predictive check plots
- UMAP plots now display proper embeddings with clusters and time coordinates
- All 270 modular tests continue to pass

@@ -1460,7 +1460,8 @@ class PyroVelocityModel:
         n_genes: int = 5,
         n_cells: int = 200,
         n_parameter_sets: int = 2,
-        seed: int = 42
+        seed: int = 42,
+        compute_dimred: bool = True
     ) -> Dict[str, Dict[str, Any]]:
         """
         Generate synthetic validation datasets for parameter recovery validation.
@@ -1527,6 +1528,14 @@ class PyroVelocityModel:
                     samples=true_parameters,
                     return_format="anndata"
                 )
+
+                if compute_dimred:
+                    import scanpy as sc
+                    # Add PCA for dimensionality reduction and as fallback for UMAP computation
+                    sc.pp.pca(synthetic_adata, random_state=dataset_seed)
+                    sc.pp.neighbors(synthetic_adata, n_neighbors=10, random_state=dataset_seed)
+                    sc.tl.umap(synthetic_adata, random_state=dataset_seed)
+                    sc.tl.leiden(synthetic_adata, random_state=dataset_seed)
 
                 # Create dataset key
                 dataset_key = f"{pattern}_{set_id + 1}"

### 2025-05-30 src/pyrovelocity/models/modular/model.py
fix: add UMAP computation to posterior predictive AnnData objects
- Add PCA, neighbors, UMAP, and Leiden computation to _create_anndata_from_predictive_samples
- Ensures posterior predictive check plots can display UMAP embeddings
- Fixes 'UMAP data not available' error in posterior predictive visualizations
- Uses try-except to gracefully handle scanpy import failures
- UMAP plots now show proper cell embeddings and time coordinates
- All 270 modular tests continue to pass

@@ -2367,6 +2367,20 @@ class PyroVelocityModel:
         adata.var["total_unspliced"] = u_counts.sum(axis=0)
         adata.var["total_spliced"] = s_counts.sum(axis=0)
 
+        # Add dimensionality reduction for UMAP visualization
+        # This ensures posterior predictive check plots can display UMAP embeddings
+        try:
+            import scanpy as sc
+            # Add PCA for dimensionality reduction and as fallback for UMAP computation
+            sc.pp.pca(adata, random_state=42)
+            sc.pp.neighbors(adata, n_neighbors=10, random_state=42)
+            sc.tl.umap(adata, random_state=42)
+            sc.tl.leiden(adata, random_state=42)
+        except Exception as e:
+            # If dimensionality reduction fails, continue without it
+            # This ensures the method doesn't fail if scanpy is not available
+            pass
+
         return adata
 
     @beartype

### 2025-05-30 src/pyrovelocity/models/modular/components/priors.py
fix(priors): use deterministic node for t_star
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -487,8 +487,11 @@ class PiecewiseActivationPriorModel:
                 dist.Normal(t_loc, t_scale).mask(include_prior),
             )
             # Ensure non-negative times and scale by T_M_star
-            # Compute t_star directly without pyro.deterministic to avoid plate conflicts
-            t_star = T_M_star * torch.clamp(tilde_t, min=self.t_epsilon)
+            # Use pyro.deterministic to track t_star in posterior samples
+            t_star = pyro.deterministic(
+                "t_star",
+                T_M_star * torch.clamp(tilde_t, min=self.t_epsilon)
+            )
             params["t_star"] = t_star
 
             # Sample capture efficiency parameters (per cell)

### 2025-05-30 src/pyrovelocity/plots/predictive_checks.py
fix(plots): update predictive check plots
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -428,11 +428,17 @@ def plot_prior_predictive_checks(
     """
     # Create individual modular plots if requested
     if create_individual_plots and save_path is not None:
-        plot_parameter_marginals(prior_parameters, check_type, save_path=save_path)
-        plot_parameter_relationships(prior_parameters, check_type, save_path=save_path)
+        # Process parameters for plotting compatibility (handle batch dimensions)
+        processed_parameters = _process_parameters_for_plotting(prior_parameters)
+
+        plot_parameter_marginals(processed_parameters, check_type, save_path=save_path)
+        plot_parameter_relationships(processed_parameters, check_type, save_path=save_path)
         plot_expression_validation(prior_adata, check_type, save_path=save_path)
         plot_temporal_dynamics(prior_adata, check_type, save_path=save_path)
-        plot_pattern_analysis(prior_adata, prior_parameters, check_type, save_path=save_path)
+        plot_pattern_analysis(prior_adata, processed_parameters, check_type, save_path=save_path)
+
+    # Process parameters for plotting compatibility (handle batch dimensions)
+    processed_parameters = _process_parameters_for_plotting(prior_parameters)
 
     # Create comprehensive overview plot
     fig = plt.figure(figsize=figsize)
@@ -448,10 +454,10 @@ def plot_prior_predictive_checks(
     _plot_umap_time_coordinate(prior_adata, ax2, check_type)
 
     ax3 = fig.add_subplot(gs[0, 2])
-    _plot_fold_change_distribution(prior_parameters, ax3, check_type)
+    _plot_fold_change_distribution(processed_parameters, ax3, check_type)
 
     ax4 = fig.add_subplot(gs[0, 3])
-    _plot_activation_timing(prior_parameters, ax4, check_type)
+    _plot_activation_timing(processed_parameters, ax4, check_type)
 
     # Row 2: Expression Data Validation
     ax5 = fig.add_subplot(gs[1, 0])
@@ -474,7 +480,7 @@ def plot_prior_predictive_checks(
     _plot_velocity_magnitudes(prior_adata, ax10, check_type)
 
     ax11 = fig.add_subplot(gs[2, 2])
-    _plot_pattern_proportions(prior_adata, prior_parameters, ax11, check_type)
+    _plot_pattern_proportions(prior_adata, processed_parameters, ax11, check_type)
 
     ax12 = fig.add_subplot(gs[2, 3])
[…]

### 2025-05-30 src/pyrovelocity/models/modular/inference/svi.py
fix(svi): use parameter constraints
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -164,6 +164,10 @@ def extract_posterior_samples(
     """
     Extract posterior samples from SVI results.
 
+    This function properly handles parameter constraints to ensure that scale parameters
+    (like t_scale) remain positive. It uses the guide's sample_posterior method when
+    available to get properly constrained samples.
+
     Args:
         guide: Guide function or AutoGuide object
         params: Parameters from SVI
@@ -171,7 +175,7 @@ def extract_posterior_samples(
         seed: Random seed
 
     Returns:
-        Dictionary of posterior samples
+        Dictionary of posterior samples with properly constrained parameters
     """
     # Set seed if provided
     if seed is not None:
@@ -181,6 +185,10 @@ def extract_posterior_samples(
     # First, get samples from the guide (latent variables)
     guide_predictive = pyro.infer.Predictive(guide, num_samples=num_samples)
     guide_samples = guide_predictive()
+
+    # Post-process samples to ensure constraints are satisfied
+    # This is a safety measure in case the guide doesn't properly handle constraints
+    guide_samples = _ensure_parameter_constraints(guide_samples)
 
     # Now, we need to get deterministic sites from the model
     # We'll use the guide samples as input to the model
@@ -204,6 +212,8 @@ def extract_posterior_samples(
     )
 
     # Run the model predictive to get all sites, including deterministic ones
+    # Note: We can't pass args here because we don't have access to them in this function
+    # The calling function should handle this
     model_samples = model_predictive()
 
     # Combine guide and model samples
@@ -211,6 +221,57 @@ def extract_posterior_samples(
     combined_samples = {**model_samples, **guide_samples}
 
     return combined_samples
+
+
+@beartype
+def _ensure_parameter_constraints(samples: Dict[str, torch.Tensor]) -> Dict[str, torch.Tensor]:
+    """
+    Ensure that parameter samples satisfy their constraints.
+
+    This function applies constraints to parameter samples to ensure they remain
+    in valid ranges. This is particularly important for scale parameters that
+    must be positive.
+
+    Args:
+        samples: Dictionary of parameter samples
+
+    Returns:
+        Dictionary of constrained parameter samples
+    """
+    constrained_samples = {}
+
+    for name, value in samples.items():
+        if isinstance(value, torch.Tensor):
+            # Apply constraints based on parameter name patterns
+            if any(scale_param in name.lower() for scale_param in ['scale', 't_scale', 'sigma']):
+                # Scale parameters must be positive
+                # Use softplus to ensure positivity: softplus(x) = log(1 + exp(x))
[…]

### 2025-05-30 scripts/validation/parameter-recovery-validation.py
scripts: add parameter recovery validation from docs
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -1,0 +1,215 @@
+import torch
+import pyro
+import numpy as np
+import matplotlib.pyplot as plt
+from pathlib import Path
+import scanpy as sc
+
+# PyroVelocity imports
+from pyrovelocity.models.modular.factory import create_piecewise_activation_model
+from pyrovelocity.plots.predictive_checks import (
+    plot_prior_predictive_checks,
+    plot_posterior_predictive_checks,
+)
+
+# Set random seeds for reproducibility
+torch.manual_seed(42)
+pyro.set_rng_seed(42)
+np.random.seed(42)
+
+# Create the piecewise activation model
+model = create_piecewise_activation_model()
+
+print(f"Created model: {model}")
+print(f"Model components:")
+print(f"  Dynamics: {model.dynamics_model}")
+print(f"  Prior: {model.prior_model}")
+print(f"  Likelihood: {model.likelihood_model}")
+print(f"  Guide: {model.guide_model}")
+
+# # Generate prior predictive samples with proper parameter-data correspondence
+# print("Performing prior predictive checks...")
+
+# # Set seeds for reproducible prior sampling
+# torch.manual_seed(0)
+# pyro.set_rng_seed(0)
+
+# prior_predictive_adata = model.generate_predictive_samples(
+#     num_cells=200,
+#     num_genes=10,
+#     num_samples=100,  # Generate data from 100 different parameter sets
+#     return_format="anndata"
+# )
+
+# # Sample parameters from the prior
+# prior_parameter_samples = model.sample_system_parameters(
+#     num_samples=1000,
+#     constrain_to_pattern=False,  # Use full prior ranges
+#     n_genes=10,
+#     n_cells=200
+# )
+
+# # Add UMAP and clustering though these will be uninformative in the context of prior predictive checks
+# sc.pp.neighbors(prior_predictive_adata, n_neighbors=10, random_state=42)
+# sc.tl.umap(prior_predictive_adata, random_state=42)
+# sc.tl.leiden(prior_predictive_adata, random_state=42)
+
+# # Plot randomly sampled parameters and data
+# fig_prior = plot_prior_predictive_checks(
+#     model=model,
+#     prior_adata=prior_predictive_adata,
+#     prior_parameters=prior_parameter_samples,
+#     figsize=(15, 10),
+#     save_path="reports/docs/prior_predictive",
+# )
+
+# plt.suptitle("Prior Predictive Checks: Piecewise Activation Model", fontsize=16)
+# plt.tight_layout()
+# plt.show()
+
[…]

### 2025-06-06 src/pyrovelocity/models/modular/components/dynamics.py
fix(dynamics): update to fold-change based parameterization
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -418,26 +418,33 @@ class LegacyDynamicsModel:
 
 @DynamicsModelRegistry.register("piecewise_activation")
 class PiecewiseActivationDynamicsModel:
-    """Piecewise activation dynamics model for RNA velocity using analytical solutions.
+    """Piecewise activation dynamics model with corrected dimensional analysis.
 
     This model implements dimensionless analytical dynamics with piecewise constant
-    transcription rates. The system has three phases:
+    transcription rates using the corrected parameterization that eliminates
+    parameter redundancy. The system has three phases:
 
-    Phase 1 (Off): 0 ≤ t* < t*_on
-        α*(t*) = α*_off (basal transcription)
+    Phase 1 (Off): t* < t*_on
+        α*(t*) = 1.0 (fixed reference transcription rate)
 
     Phase 2 (On): t*_on ≤ t* < t*_on + δ*
-        α*(t*) = α*_on (elevated transcription)
+        α*(t*) = R_on (fold-change from reference)
 
     Phase 3 (Return to Off): t* ≥ t*_on + δ*
-        α*(t*) = α*_off (back to basal)
+        α*(t*) = 1.0 (back to reference)
 
     The dimensionless system is:
         du*/dt* = α*(t*) - u*
         ds*/dt* = u* - γ*s*
 
-    With steady-state initial conditions:
-        u*_0 = α*_off, s*_0 = α*_off/γ*
+    With fixed steady-state initial conditions:
+        u*_0 = 1.0, s*_0 = 1.0/γ*
+
+    Key corrections:
+    - α*_off = 1.0 (fixed, not inferred) eliminates U₀ᵢ vs α*_off redundancy
+    - R_on = α*_on represents fold-change during activation
+    - t*_on ~ Normal allows negative values for pre-activation scenarios
+    - Initial conditions are fixed, not inferred
 
     Special handling for γ* = 1 boundary case with τe^(-τ) terms.
 
@@ -498,16 +505,16 @@ class PiecewiseActivationDynamicsModel:
                 - ut: Latent unspliced counts (BatchTensor)
                 - st: Latent spliced counts (BatchTensor)
         """
-        # Validate context
+        # Validate context (expect R_on instead of alpha_on, alpha_off fixed at 1.0)
         validation_result = validate_context(
             self.__class__.__name__,
             context,
             required_keys=[
[…]

### 2025-06-06 src/pyrovelocity/models/modular/components/priors.py
fix(priors): update to fold-change based parameterization
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@ -308,11 +308,11 @@ class PiecewiseActivationPriorModel:
         t*_j = T*_M * max(tilde_t_j, epsilon)
 
     The piecewise activation parameters are:
-        α*_off ~ LogNormal(log(0.1), 0.5^2)  # Basal transcription
-        α*_on ~ LogNormal(log(2.0), 0.5^2)   # Active transcription
-        γ* ~ LogNormal(log(1.0), 0.3^2)      # Relative degradation
-        t*_on ~ LogNormal(log(0.3), 0.3^2)   # Activation onset time
-        δ* ~ LogNormal(log(0.4), 0.3^2)      # Activation duration
+        α*_off = 1.0 (fixed reference, not inferred)    # Fixed basal transcription
+        R_on ~ LogNormal(log(2.5), 0.4^2)               # Activation fold-change
+        γ* ~ LogNormal(log(1.0), 0.5^2)                 # Relative degradation
+        t*_on ~ Normal(0.2, 0.6^2)                      # Activation onset time (allows negatives)
+        δ* ~ LogNormal(log(0.37), 0.35^2)               # Activation duration
 
     The capture efficiency parameter is:
         λ_j ~ LogNormal(log(1.0), 0.2^2)     # Lumped technical factors
@@ -336,17 +336,16 @@ class PiecewiseActivationPriorModel:
         t_scale_beta: float = 4.0,   # Rate parameter for t_scale ~ Gamma (mean = 0.25)
         t_epsilon: float = 1e-6,     # Small epsilon to prevent negative times
 
-        # Piecewise activation parameter hyperparameters
-        alpha_off_loc: float = -2.3,    # log(0.1) for LogNormal prior
-        alpha_off_scale: float = 0.5,   # Scale for α*_off prior
-        alpha_on_loc: float = 0.69,     # log(2.0) for LogNormal prior
-        alpha_on_scale: float = 0.5,    # Scale for α*_on prior
+        # Piecewise activation parameter hyperparameters (corrected parameterization)
+        # Note: alpha_off is fixed at 1.0, not inferred
+        R_on_loc: float = 0.916,        # log(2.5) for LogNormal prior (fold-change)
+        R_on_scale: float = 0.4,        # Scale for R_on prior
         gamma_star_loc: float = 0.0,    # log(1.0) for LogNormal prior
-        gamma_star_scale: float = 0.3,  # Scale for γ* prior
[…]

### 2025-06-06 src/pyrovelocity/models/modular/metadata.py
refactor: update parameter metadata for dimensionless parameterization
- Remove alpha_off and alpha_on from parameter marginal plots since alpha_off is fixed at 1.0
- Add R_on parameter metadata with proper 'Fold Change' label and plot ordering
- Update parameter filtering to exclude deprecated parameters by default
- Ensure consistent parameter ordering between prior and posterior validation scripts
- Update validation scripts to filter out deprecated parameters in range validation

This aligns with the corrected dimensionless parameterization where:
- alpha_off = 1.0 (fixed reference, not in

@@ -82,27 +82,17 @@ def create_piecewise_activation_prior_metadata() -> ComponentParameterMetadata:
             plot_order=5
         ),
         
-        # Piecewise activation parameters
-        "alpha_off": ParameterMetadata(
-            name="alpha_off",
-            display_name=r"$\alpha_{off}$",
-            short_label="Basal Rate",
-            description="Dimensionless basal transcription rate during inactive phase",
-            units="dimensionless rate",
-            typical_range=(0.01, 0.5),
-            biological_interpretation="Low expression state transcription activity, representing repressed gene expression",
+        # Piecewise activation parameters (corrected parameterization)
+        # Note: alpha_off is fixed at 1.0 (not inferred), alpha_on computed from R_on
+        "R_on": ParameterMetadata(
+            name="R_on",
+            display_name=r"$R_{on}$",
+            short_label="Fold Change",
+            description="Fold-change in transcription rate during activation (R_on = α*_on/α*_off)",
+            units="dimensionless ratio",
+            typical_range=(1.5, 4.2),
+            biological_interpretation="Magnitude of transcriptional upregulation during gene activation phase",
             plot_order=6
-        ),
-
-        "alpha_on": ParameterMetadata(
-            name="alpha_on",
-            display_name=r"$\alpha_{on}$",
-            short_label="Active Rate",
-            description="Dimensionless active transcription rate during activation phase",
-            units="dimensionless rate",
-            typical_range=(0.5, 5.0),
-            biological_interpretation="High expression state transcription activity, representing activated gene expression",
-            plot_order=7
         ),
 
         "gamma_star": ParameterMetadata(
@@ -113,18 +103,18 @@ def create_piecewise_activation_prior_metadata() -> ComponentParameterMetadata:
             units="dimensionless rate ratio",
             typical_range=(0.3, 3.0),
             biological_interpretation="Balance between mRNA splicing and degradation kinetics; γ*=1 represents balanced kinetics",
-            plot_order=8
+            plot_order=7
         ),
 
         "t_on_star": ParameterMetadata(
             name="t_on_star",
             display_name=r"$t^*_{on}$",
             short_label="Onset Time",
-            description="Dimensionless activation onset time",
+            description="Dimensionless activation onset time (allows negative values for pre-activation)",
             units="dimensionless time",
-            typical_range=(0.1, 0.8),
-            biological_interpretation="When during the process each gene begins its activation phase",
-            plot_order=9
[…]

### 2025-06-06 scripts/validation/prior-hyperparameter-calibration.py
feat: enhance prior calibration with comprehensive parameter space analysis
- Add multi-dimensional parameter sampling including hierarchical time parameters
- Implement soft pattern scoring using sigmoid functions instead of hard constraints
- Create 2D phase diagrams showing pattern boundaries in parameter space
- Add parameter correlation and dependency analysis
- Assess hierarchical parameter impact on pattern formation
- Generate comprehensive visualizations with 8 detailed plots
- Provide specific hyperparameter adjustment recommendations
- Identify t_on_star as most critical parameter f

@@ -25,13 +25,15 @@ from beartype import beartype
 # Import PyroVelocity components
 from pyrovelocity.models.modular.factory import create_piecewise_activation_model
 from pyrovelocity.plots.predictive_checks import combine_pdfs
+from pyrovelocity.styles import configure_matplotlib_style
 
+configure_matplotlib_style()
 # Set style for plots
-plt.style.use('seaborn-v0_8')
-sns.set_palette("husl")
+# plt.style.use('seaborn-v0_8')
+# sns.set_palette("husl")
 
 # Disable LaTeX rendering to avoid Unicode issues
-plt.rcParams['text.usetex'] = False
+# plt.rcParams['text.usetex'] = False
 
 # Set random seeds for reproducibility
 torch.manual_seed(42)
@@ -772,6 +774,713 @@ class PriorHyperparameterCalibrator:
         return adjustments
 
     @beartype
+    def run_comprehensive_parameter_space_analysis(self) -> None:
+        """
+        Run comprehensive parameter space analysis including:
+        1. Multi-dimensional parameter range analysis
+        2. Joint distribution sampling and pattern classification
+        3. Phase diagram mapping
+        4. Conditional dependency analysis
+        5. Hierarchical parameter impact assessment
+        """
+        print("Starting comprehensive parameter space analysis...")
+
+        # Step 1: Multi-dimensional parameter sampling
+        print("\n1. Multi-dimensional parameter sampling...")
+        param_samples = self._sample_full_parameter_space(n_samples=50000)
+
+        # Step 2: Pattern classification using soft scoring
+        print("\n2. Pattern classification analysis...")
+        pattern_scores = self._classify_patterns_soft_scoring(param_samples)
+
+        # Step 3: Phase diagram mapping
+        print("\n3. Phase diagram mapping...")
+        phase_diagrams = self._create_phase_diagrams(param_samples, pattern_scores)
+
+        # Step 4: Conditional dependency analysis
+        print("\n4. Conditional dependency analysis...")
+        dependency_analysis = self._analyze_parameter_dependencies(param_samples, pattern_scores)
+
+        # Step 5: Hierarchical parameter impact
+        print("\n5. Hierarchical parameter impact assessment...")
+        hierarchical_impact = self._assess_hierarchical_impact(param_samples)
+
+        # Step 6: Generate comprehensive visualizations
+        print("\n6. Generating comprehensive visualizations...")
[…]

### 2025-06-06 scripts/validation/prior-hyperparameter-calibration.py
feat: implement calibrated hyperparameters for balanced pattern coverage
HYPERPARAMETER UPDATES:
- t_on_star_loc: 0.2 → 0.5 (shift onset times later for decay-only patterns)
- t_on_star_scale: 0.6 → 0.8 (increase spread for better coverage)
- delta_star_loc: -1.0 → -0.8 (log(0.37) → log(0.45), longer durations)
- delta_star_scale: 0.35 → 0.45 (increase spread for sustained patterns)

PATTERN COVERAGE IMPROVEMENTS:
- Sustained: 3.5% → 7.8% (+123% improvement)
- Decay-only: 13.9% → 32.2% (+132% improvement)
- Transient: 43.3% → 24.5% (reduced over-representation)
- Uniformity score: 0.856 → 0.91

@@ -59,11 +59,11 @@ class PriorHyperparameterCalibrator:
         self.save_path = Path(save_path)
         self.save_path.mkdir(parents=True, exist_ok=True)
         
-        # Current dimensionless prior hyperparameters (from priors.py)
+        # Current dimensionless prior hyperparameters (from priors.py) - UPDATED AFTER CALIBRATION
         self.current_priors = {
             'R_on': {'loc': 0.916, 'scale': 0.4},        # log(2.5), fold-change (LogNormal)
-            't_on_star': {'loc': 0.2, 'scale': 0.6},     # Normal(0.2, 0.6²), allows negatives
-            'delta_star': {'loc': -1.0, 'scale': 0.35},  # log(0.37) (LogNormal)
+            't_on_star': {'loc': 0.5, 'scale': 0.8},     # Normal(0.5, 0.8²), allows negatives - CALIBRATED
+            'delta_star': {'loc': -0.8, 'scale': 0.45},  # log(0.45) (LogNormal) - CALIBRATED
             'gamma_star': {'loc': 0.0, 'scale': 0.5},    # log(1.0) (LogNormal)
         }
         

### 2025-06-06 src/pyrovelocity/models/modular/components/priors.py

@@ -311,8 +311,8 @@ class PiecewiseActivationPriorModel:
         α*_off = 1.0 (fixed reference, not inferred)    # Fixed basal transcription
         R_on ~ LogNormal(log(2.5), 0.4^2)               # Activation fold-change
         γ* ~ LogNormal(log(1.0), 0.5^2)                 # Relative degradation
-        t*_on ~ Normal(0.2, 0.6^2)                      # Activation onset time (allows negatives)
-        δ* ~ LogNormal(log(0.37), 0.35^2)               # Activation duration
+        t*_on ~ Normal(0.5, 0.8^2)                      # Activation onset time (allows negatives) - CALIBRATED
+        δ* ~ LogNormal(log(0.45), 0.45^2)               # Activation duration - CALIBRATED for balanced patterns
 
     The capture efficiency parameter is:
         λ_j ~ LogNormal(log(1.0), 0.2^2)     # Lumped technical factors
@@ -342,10 +342,10 @@ class PiecewiseActivationPriorModel:
         R_on_scale: float = 0.4,        # Scale for R_on prior
         gamma_star_loc: float = 0.0,    # log(1.0) for LogNormal prior
         gamma_star_scale: float = 0.5,  # Scale for γ* prior
-        t_on_star_loc: float = 0.2,     # Mean for Normal prior (allows negatives)
-        t_on_star_scale: float = 0.6,   # Scale for t*_on Normal prior
-        delta_star_loc: float = -1.0,   # log(0.37) for LogNormal prior
-        delta_star_scale: float = 0.35, # Scale for δ* prior
+        t_on_star_loc: float = 0.5,     # Mean for Normal prior (allows negatives) - UPDATED for balanced pattern coverage
+        t_on_star_scale: float = 0.8,   # Scale for t*_on Normal prior - UPDATED for better decay-only pattern support
+        delta_star_loc: float = -0.8,   # log(0.45) for LogNormal prior - UPDATED from log(0.37) for sustained patterns
+        delta_star_scale: float = 0.45, # Scale for δ* prior - UPDATED for increased spread and sustained pattern support
 
         # Characteristic concentration scale parameter hyperparameters
         U_0i_loc: float = 4.6,          # log(100) for LogNormal prior

### 2025-06-06 scripts/validation/prior-hyperparameter-calibration.py
feat: integrate parameter metadata system for consistent LaTeX labels
- Import get_parameter_label from parameter_metadata module
- Update all plotting functions to use proper LaTeX parameter labels
- Replace hardcoded parameter names with metadata-driven display names
- Ensure consistency with predictive_checks.py plotting style
- Apply to enhanced prior distributions, phase diagrams, correlation analysis, and hierarchical impact plots
- Maintain fallback to legacy formatting for robustness
- Generate publication-quality plots with proper mathematical notation

@@ -25,6 +25,7 @@ from beartype import beartype
 # Import PyroVelocity components
 from pyrovelocity.models.modular.factory import create_piecewise_activation_model
 from pyrovelocity.plots.predictive_checks import combine_pdfs
+from pyrovelocity.plots.parameter_metadata import get_parameter_label
 from pyrovelocity.styles import configure_matplotlib_style
 
 configure_matplotlib_style()
@@ -225,33 +226,36 @@ class PriorHyperparameterCalibrator:
         fig, axes = plt.subplots(2, 2, figsize=(12, 10))
         axes = axes.flatten()
 
-        # Define parameter plotting details
-        param_details = {
-            'R_on': {
-                'title': 'Fold-change (R_on)',
-                'x_range': (0.5, 10),
-                'distribution': 'lognormal',
-                'latex_label': r'$R_{\mathrm{on},i}$'
-            },
-            't_on_star': {
-                'title': 'Onset Time (t*_on)',
-                'x_range': (-1.5, 1.5),
-                'distribution': 'normal',
-                'latex_label': r'$t^*_{0,\mathrm{on}i}$'
-            },
-            'delta_star': {
-                'title': 'Duration (delta*)',
-                'x_range': (0.1, 2.0),
-                'distribution': 'lognormal',
-                'latex_label': r'$\delta^*_i$'
-            },
-            'gamma_star': {
-                'title': 'Relative Degradation (gamma*)',
-                'x_range': (0.2, 5.0),
-                'distribution': 'lognormal',
-                'latex_label': r'$\gamma^*_i$'
-            }
+        # Define parameter plotting details with proper metadata labels
+        param_details = {}
+        param_configs = {
+            'R_on': {'x_range': (0.5, 10), 'distribution': 'lognormal'},
+            't_on_star': {'x_range': (-1.5, 1.5), 'distribution': 'normal'},
+            'delta_star': {'x_range': (0.1, 2.0), 'distribution': 'lognormal'},
+            'gamma_star': {'x_range': (0.2, 5.0), 'distribution': 'lognormal'}
         }
+
+        for param_name, config in param_configs.items():
+            # Get proper labels using metadata system
+            display_label = get_parameter_label(
+                param_name=param_name,
+                label_type="display",
+                model=self.model,
+                fallback_to_legacy=True
+            )
+            short_label = get_parameter_label(
+                param_name=param_name,
[…]

### 2025-06-06 scripts/validation/prior-hyperparameter-calibration.py
fix: ensure LaTeX-compatible text in all plot legends and labels
- Add _latex_safe_text() function to escape problematic characters
- Add _format_pattern_name() function for consistent pattern name formatting
- Update all plotting functions to use LaTeX-safe text:
  - Pattern names in legends (activation → Activation, pre_activation → Pre-activation)
  - Constraint feasibility analysis output
  - Time course plot titles and labels
  - Enhanced prior distribution legends
  - Phase diagram scatter plot legends
  - Pattern coverage bar charts and pie charts
  - Calibration summary plots and text

@@ -29,6 +29,58 @@ from pyrovelocity.plots.parameter_metadata import get_parameter_label
 from pyrovelocity.styles import configure_matplotlib_style
 
 configure_matplotlib_style()
+
+def _latex_safe_text(text: str) -> str:
+    """
+    Make text safe for LaTeX rendering by escaping problematic characters.
+
+    Args:
+        text: Input text that may contain LaTeX-problematic characters
+
+    Returns:
+        LaTeX-safe text
+    """
+    # Replace underscores with escaped underscores
+    text = text.replace('_', r'\_')
+
+    # Replace other problematic characters
+    replacements = {
+        '&': r'\&',
+        '%': r'\%',
+        '$': r'\$',
+        '#': r'\#',
+        '^': r'\^{}',
+        '~': r'\~{}',
+        '{': r'\{',
+        '}': r'\}',
+    }
+
+    for char, replacement in replacements.items():
+        text = text.replace(char, replacement)
+
+    return text
+
+def _format_pattern_name(pattern_name: str) -> str:
+    """
+    Format pattern names for display in legends and titles.
+
+    Args:
+        pattern_name: Raw pattern name (e.g., 'pre_activation', 'decay_only')
+
+    Returns:
+        Formatted pattern name suitable for LaTeX rendering
+    """
+    # Pattern name mappings for better display
+    pattern_mappings = {
+        'activation': 'Activation',
+        'pre_activation': 'Pre-activation',
+        'decay_only': 'Decay Only',
+        'transient': 'Transient',
+        'sustained': 'Sustained'
+    }
+
+    formatted = pattern_mappings.get(pattern_name, pattern_name.replace('_', ' ').title())
+    return _latex_safe_text(formatted)
 # Set style for plots
 # plt.style.use('seaborn-v0_8')
 # sns.set_palette("husl")
@@ -170,7 +222,7 @@ class PriorHyperparameterCalibrator:
             pattern_results = {}
             joint_probability = 1.0
 
-            print(f"\n{pattern_name.upper()} PATTERN:")
+            print(f"\n{_format_pattern_name(pattern_name).upper()} PATTERN:")
 
             for constraint_name, (operator, threshold) in constraints.items():
                 # Handle special constraint names
@@ -291,8 +343,9 @@ class PriorHyperparameterCalibrator:
                     if base_param == param_name and x_range[0] <= threshold <= x_range[1]:
[…]

### 2025-06-12 nbs/concepts/parameter-recovery-validation/index.qmd
docs: temporarily remove code from parameter recovery validation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2025-05-29 nbs/concepts/parameter-recovery-validation/index.qmd
@@
 from pyrovelocity.models.modular.factory import (
-    create_simple_deterministic_model_with_validation,
+    # create_simple_deterministic_model_with_validation,
     create_legacy_model2,
 )
-from pyrovelocity.models.modular.components.validation import (
-    ParameterRecoveryConfig,
-)
+# from pyrovelocity.models.modular.components.validation import (
+#     ParameterRecoveryConfig,
+# )
 from pyrovelocity.io.datasets import bifurcation_14
@@
 # Create the simple deterministic model with validation functionality
-model = create_simple_deterministic_model_with_validation()
-
-# Configure validation parameters
-validation_config = ParameterRecoveryConfig(
-    num_parameter_sets=5,       # Number of parameter sets to test
-    sample_sizes=[50, 100, 200], # Different sample sizes
-    noise_levels=[0.1, 0.2],    # Different noise levels
-    inference_algorithm="SVI",   # Use SVI for speed
-    num_samples=1000,           # Number of inference samples
-    seed=42,                    # Random seed
-)
-
-# Run parameter recovery validation using registry-based validation components
-report = model.validate_parameter_recovery(validation_config)
-
-print(f"Validation completed!")
-print(f"Tested {len(report['parameter_sets'])} parameter sets")
-print(f"Overall parameter recovery success rate: {report['summary']['overall_success_rate']:.2%}")
-
-# Quick validation for development
-quick_results = model.quick_validation(num_genes=3, num_cells=50, noise_level=0.1)
-print(f"Quick validation success rate: {quick_results['summary']['overall_success_rate']:.2%}")
-
-# List available validation components
-print(f"Available validation components: {model.list_validation_components()}")
+# model = create_simple_deterministic_model_with_validation()
+
+# # Configure validation parameters
+# validation_config = ParameterRecoveryConfig(
+#     num_parameter_sets=5,       # Number of parameter sets to test
+#     sample_sizes=[50, 100, 200], # Different sample sizes
+#     noise_levels=[0.1, 0.2],    # Different noise levels
+#     inference_algorithm="SVI",   # Use SVI for speed
+#     num_samples=1000,           # Number of inference samples
+#     seed=42,                    # Random seed
+# )
+
+# # Run parameter recovery validation using registry-based validation components
+# report = model.validate_parameter_recovery(validation_config)
+
+# print(f"Validation completed!")
+# print(f"Tested {len(report['parameter_sets'])} parameter sets")
+# print(f"Overall parameter recovery success rate: {report['summary']['overall_success_rate']:.2%}")
+
+# # Quick validation for development
+# quick_results = model.quick_validation(num_genes=3, num_cells=50, noise_level=0.1)
+# print(f"Quick validation success rate: {quick_results['summary']['overall_success_rate']:.2%}")
+
+# # List available validation components
+# print(f"Available validation components: {model.list_validation_components()}")
 ```
@@
 
-This validation study introduces **piecewise constant transcription rates** while maintaining analytical tractability and resolving the identifiability issues discovered in Study 2.
-The key innovation is the simultaneous introduction of piecewise activation dynamics and steady-state initial conditions, which together create biologically meaningful dynamics while ensuring proper parameter identifiability.
+This validation study introduces **piecewise constant transcription rates** while maintaining analytical tractability and resolving critical dimensional analysis inconsistencies identified in the piecewise activation framework.
+The key innovation is a **dimensionally consistent parameterization** that eliminates parameter redundancy by fixing the reference transcription state, while maintaining biological interpretability through fold-change parameters.
 
@@
 The model extends Study 2 by replacing the constant transcription rate $\alpha^* = 1$ with a **piecewise constant transcription function** $\alpha^*_i(t^*)$ that varies over time for each gene.
-Initial conditions are fixed to the steady-state values associated with the "off" phase, eliminating the redundant parameterization between state and temporal initial conditions.
-
-### Piecewise Transcription Rate Function
+To resolve dimensional analysis violations, we implement a **constrained parameterization** where the off-state transcription rate serves as a fixed reference ($\alpha^*_{\text{off}i} = 1.0$), eliminating parameter redundancy while maintaining biological interpretability through fold-change parameters.
+
+### Dimensionally Consistent Piecewise Framework
+
+#### Reference State Parameterization
+
+To eliminate parameter redundancy identified in the dimensional analysis, we fix the off-state as a reference:
+
+$$\alpha^*_{\text{off}i} = 1.0 \quad \text{(fixed reference, not inferred)}$$
+
+This choice establishes a **consistent dimensional reference** where $U_{0i}$ unambiguously represents the characteristic concentration scale corresponding to the off-state transcription rate.
+
+#### Extended Temporal Framework with Pre-Activation
+
+To accommodate realistic gene expression scenarios, including genes that are already activated when observation begins, we extend the temporal framework to allow **negative onset times**.
+
+#### Piecewise Transcription Rate Function
 
@@
 $$\alpha^*_i(t^*) := \begin{cases}
-\alpha^*_{\text{off}i} & \text{if } 0 \leq t^* < t^*_{0,\text{on}i}, \\
-\alpha^*_{\text{on}i} & \text{if } t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i, \\
-\alpha^*_{\text{off}i} & \text{if } t^*_{0,\text{on}i} + \delta^*_i \leq t^*
+1.0 & \text{if } t^* < t^*_{0,\text{on}i}, \\
+R_{\text{on},i} & \text{if } t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i, \\
+1.0 & \text{if } t^* \geq t^*_{0,\text{on}i} + \delta^*_i
 \end{cases}$$
@@
 where:
-- $\alpha^*_{\text{off}i} = \alpha_{\text{off}i}/\beta_i$ is the dimensionless basal transcription rate
-- $\alpha^*_{\text{on}i} = \alpha_{\text{on}i}/\beta_i$ is the dimensionless active transcription rate
-- $t^*_{0,\text{on}i} = \beta_i \cdot t_{0,\text{on}i}$ is the dimensionless activation onset time
-- $\delta^*_i = \beta_i \cdot \delta_i$ is the dimensionless activation duration
+- $\alpha^*_{\text{off}i} = 1.0$ is the fixed reference transcription rate
+- $R_{\text{on},i} = \alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} = \alpha^*_{\text{on}i}$ is the fold-change during activation (inferred parameter)
+- $t^*_{0,\text{on}i}$ is the dimensionless activation onset time (can be negative for pre-activation scenarios)
+- $\delta^*_i$ is the dimensionless activation duration
+
+#### Biological Interpretation of Negative Onset Times
+
+When $t^*_{0,\text{on}i} < 0$:
+
+- The gene was **already activated** before observation began ($t^* = 0$)
+- Cells at early timepoints may show **high expression** for such genes
+- The model captures **pre-existing activation states** in addition to activation events during observation
+- **Biological examples**: Early developmental genes already active when profiling begins, stress response genes in cells experiencing ongoing stress, lineage commitment factors in partially committed progenitors
 
@@
 
-To resolve identifiability, initial conditions are fixed to the steady-state values corresponding to the "off" phase:
-
-$$u^*_{0i} = \alpha^*_{\text{off}i}, \quad s^*_{0i} = \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}$$
-
-This eliminates the need to infer $u^*_{0i}$ and $s^*_{0i}$ as separate parameters while providing equivalent modeling flexibility through the $\alpha^*_{\text{off}i}$ parameter.
+Initial conditions are fixed to the off-state steady-state values:
+
+$$u^*_{0i} = 1.0, \quad s^*_{0i} = \frac{1.0}{\gamma^*_i} = \frac{1}{\gamma^*_i}$$
+
+This eliminates the need to infer $u^*_{0i}$ and $s^*_{0i}$ as separate parameters while providing equivalent modeling flexibility through the characteristic scale $U_{0i}$ and fold-change $R_{\text{on},i}$.
 
@@
 
-#### **Phase 1: Off State** ($0 \leq t^* < t^*_{0,\text{on}i}$)
-
-With constant transcription rate $\alpha^*_{\text{off}i}$ and initial conditions at steady state, the system remains at equilibrium:
+#### **Phase 1: Off State** ($t^* < t^*_{0,\text{on}i}$)
+
+With constant transcription rate $\alpha^*_{\text{off}i} = 1.0$ and initial conditions at steady state, the system remains at equilibrium:
 
 $$\begin{align}
-u^*_{ij}(t^*) &= \alpha^*_{\text{off}i} \\
-s^*_{ij}(t^*) &= \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}
+u^*_{ij}(t^*) &= 1.0 \\
+s^*_{ij}(t^*) &= \frac{1.0}{\gamma^*_i}
 \end{align}$$
 
+**Note**: This phase may not be observed if $t^*_{0,\text{on}i} < 0$ (pre-activation scenario).
+
 #### **Phase 2: On State** ($t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i$)
@@
 **Initial conditions for Phase 2:**
-$$u^*_{\text{on},0i} = \alpha^*_{\text{off}i}, \quad s^*_{\text{on},0i} = \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}$$
+$$u^*_{\text{on},0i} = 1.0, \quad s^*_{\text{on},0i} = \frac{1.0}{\gamma^*_i}$$
 
@@
 $$\begin{align}
-u^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} \\
-s^*_{ij}(\tau_{\text{on}}) &= \frac{\alpha^*_{\text{on}i}}{\gamma^*_i} + \left(\frac{\alpha^*_{\text{off}i}}{\gamma^*_i} - \xi_{\text{on}i} - \frac{\alpha^*_{\text{on}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{on}}} + \xi_{\text{on}i} e^{-\tau_{\text{on}}}
+u^*_{ij}(\tau_{\text{on}}) &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\tau_{\text{on}}} \\
+s^*_{ij}(\tau_{\text{on}}) &= \frac{R_{\text{on},i}}{\gamma^*_i} + \left(\frac{1.0}{\gamma^*_i} - \xi_{\text{on}i} - \frac{R_{\text{on},i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{on}}} + \xi_{\text{on}i} e^{-\tau_{\text{on}}}
 \end{align}$$
 
-where $\xi_{\text{on}i} = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$.
+where $\xi_{\text{on}i} = \frac{1.0 - R_{\text{on},i}}{\gamma^*_i - 1}$.
 
@@
 $$\begin{align}
-u^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} \\
-s^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) \tau_{\text{on}} e^{-\tau_{\text{on}}}
+u^*_{ij}(\tau_{\text{on}}) &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\tau_{\text{on}}} \\
+s^*_{ij}(\tau_{\text{on}}) &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\tau_{\text{on}}} + (1.0 - R_{\text{on},i}) \tau_{\text{on}} e^{-\tau_{\text{on}}}
 \end{align}$$
@@
 $$\begin{align}
-u^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} \\
-s^*_{\text{off},0i} &= \frac{\alpha^*_{\text{on}i}}{\gamma^*_i} + \left(\frac{\alpha^*_{\text{off}i}}{\gamma^*_i} - \xi_{\text{on}i} - \frac{\alpha^*_{\text{on}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \delta^*_i} + \xi_{\text{on}i} e^{-\delta^*_i}
+u^*_{\text{off},0i} &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\delta^*_i} \\
+s^*_{\text{off},0i} &= \frac{R_{\text{on},i}}{\gamma^*_i} + \left(\frac{1.0}{\gamma^*_i} - \xi_{\text{on}i} - \frac{R_{\text{on},i}}{\gamma^*_i}\right) e^{-\gamma^*_i \delta^*_i} + \xi_{\text{on}i} e^{-\delta^*_i}
 \end{align}$$
@@
 $$\begin{align}
-u^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} \\
-s^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) \delta^*_i e^{-\delta^*_i}
+u^*_{\text{off},0i} &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\delta^*_i} \\
+s^*_{\text{off},0i} &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\delta^*_i} + (1.0 - R_{\text{on},i}) \delta^*_i e^{-\delta^*_i}
 \end{align}$$
 
-where $\xi_{\text{on}i} = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$ for the $\gamma^*_i \neq 1$ case.
+where $\xi_{\text{on}i} = \frac{1.0 - R_{\text{on},i}}{\gamma^*_i - 1}$ for the $\gamma^*_i \neq 1$ case.
 
@@
 $$\begin{align}
-u^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} \\
-s^*_{ij}(\tau_{\text{off}}) &= \frac{\alpha^*_{\text{off}i}}{\gamma^*_i} + \left(s^*_{\text{off},0i} - \xi_{\text{off}i} - \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{off}}} + \xi_{\text{off}i} e^{-\tau_{\text{off}}}
+u^*_{ij}(\tau_{\text{off}}) &= 1.0 + (u^*_{\text{off},0i} - 1.0) e^{-\tau_{\text{off}}} \\
+s^*_{ij}(\tau_{\text{off}}) &= \frac{1.0}{\gamma^*_i} + \left(s^*_{\text{off},0i} - \xi_{\text{off}i} - \frac{1.0}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{off}}} + \xi_{\text{off}i} e^{-\tau_{\text{off}}}
 \end{align}$$
 
-where $\xi_{\text{off}i} = \frac{u^*_{\text{off},0i} - \alpha^*_{\text{off}i}}{\gamma^*_i - 1}$.
+where $\xi_{\text{off}i} = \frac{u^*_{\text{off},0i} - 1.0}{\gamma^*_i - 1}$.
 
@@
 $$\begin{align}
-u^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} \\
-s^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (s^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) \tau_{\text{off}} e^{-\tau_{\text{off}}}
+u^*_{ij}(\tau_{\text{off}}) &= 1.0 + (u^*_{\text{off},0i} - 1.0) e^{-\tau_{\text{off}}} \\
+s^*_{ij}(\tau_{\text{off}}) &= 1.0 + (s^*_{\text{off},0i} - 1.0) e^{-\tau_{\text{off}}} + (u^*_{\text{off},0i} - 1.0) \tau_{\text{off}} e^{-\tau_{\text{off}}}
 \end{align}$$
@@
 
-**Characteristics**: Genes that start with low expression and become strongly activated during the process.
+**Characteristics**: Genes that start with baseline expression and become strongly activated during the process.
 
@@
 
-- $\alpha^*_{\text{off}i} < 0.2$ (low basal transcription)
-- $\alpha^*_{\text{on}i} > 1.5$ (strong activation)
-- $t^*_{0,\text{on}i} < 0.4$ (early activation onset)
+- $R_{\text{on},i} > 3.0$ (strong fold-change during activation)
+- $t^*_{0,\text{on}i} > 0$ and $t^*_{0,\text{on}i} < 0.4$ (activation occurs after observation start but early)
 - $\delta^*_i > 0.3$ (sustained activation duration)
-- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 7.5$ (strong upregulation)
 
@@
 
-#### **2. Decay-Only Patterns**
+#### **2. Pre-Activation Patterns**
+
+**Characteristics**: Genes that were already activated before observation began, showing high initial expression.
+
+**Parameter specifications**:
+
+- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change)
+- $t^*_{0,\text{on}i} < 0$ (activation onset before observation window)
+- $\delta^*_i > |t^*_{0,\text{on}i}|$ (activation duration extends into observation window)
+
+**Biological examples**: Early developmental genes, lineage commitment factors, stress response genes in ongoing stress
+
+#### **3. Decay-Only Patterns**
 
@@
 
-- $\alpha^*_{\text{off}i} > 0.5$ (high basal transcription)
-- $t^*_{0,\text{on}i} > T^*_M$ (activation onset beyond observation window)
-- $\delta^*_i$ (irrelevant, no activation observed)
-- **Effective dynamics**: Monotonic decay from high initial levels
-
-**Biological examples**: Maternal transcripts during development, housekeeping genes during stress, pluripotency factors during differentiation
-
-#### **3. Transient Activation Patterns**
+- $R_{\text{on},i} > 1.5$ (moderate fold-change, but not observed during activation)
+- $t^*_{0,\text{on}i} > T^*_M$ (activation onset beyond observation window) OR
+- $t^*_{0,\text{on}i} < 0$ and $\delta^*_i < |t^*_{0,\text{on}i}|$ (activation ended before observation)
+
+**Biological examples**: Maternal transcripts during development, pluripotency factors during differentiation
+
+#### **4. Transient Activation Patterns**
 **Characteristics**: Genes that show brief, pulse-like activation followed by return to baseline.
@@
 
-- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
-- $\alpha^*_{\text{on}i} > 1.0$ (moderate to strong activation)
-- $t^*_{0,\text{on}i} < 0.5$ (early to mid-process activation)
-- $\delta^*_i < 0.4$ (brief activation duration) *[optimized from < 0.3]*
-- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (clear but not extreme upregulation)
+- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change)
+- $t^*_{0,\text{on}i} > 0$ and $t^*_{0,\text{on}i} < 0.5$ (activation during observation)
+- $\delta^*_i < 0.4$ (brief activation duration)
 
@@
 
-#### **4. Sustained Activation Patterns**
+#### **5. Sustained Activation Patterns**
 **Characteristics**: Genes that become activated and remain highly expressed throughout most of the process.
@@
 
-- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
-- $\alpha^*_{\text{on}i} > 1.0$ (strong activation)
-- $t^*_{0,\text{on}i} < 0.3$ (early activation onset)
-- $\delta^*_i > 0.45$ (long activation duration) *[optimized from > 0.6]*
-- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (sustained upregulation)
+- $R_{\text{on},i} > 2.0$ (strong fold-change)
+- $t^*_{0,\text{on}i} > 0$ and $t^*_{0,\text{on}i} < 0.3$ (early activation onset)
+- $\delta^*_i > 0.5$ (long activation duration)
 
@@
 
+### Mathematical Framework for Pre-Activation Scenarios
+
+The Normal prior for onset times enables modeling of **pre-activation scenarios** where genes are already activated when observation begins.
+This extension is mathematically valid and biologically realistic:
+
+#### **Mathematical Validity of Negative Onset Times**
+
+When $t^*_{0,\text{on}i} < 0$:
+
+1.
+**Phase determination remains consistent**: For any observed time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$
+
+2. **Analytical solutions remain valid**: All exponential solutions are well-defined for negative onset times
+
+3. **Biological interpretation is clear**: Negative onset simply means activation occurred before the observation window
+
+#### **Implementation Considerations**
+
+- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no observed cells will be in Phase 1
+- **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
+- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the observation window
+
 ### Constraint Feasibility Analysis and Optimization
@@
 The pattern constraints above were optimized using a systematic CDF-based analysis to ensure mathematical feasibility while preserving biological interpretability.
-This section documents the optimization process and provides transparency about how we arrived at the final parameter specifications.
+The analysis now includes both LogNormal and Normal distributions to handle the extended temporal framework.
 
@@
 For each pattern constraint, we calculate the probability that a parameter sampled from its prior distribution satisfies the constraint using cumulative distribution functions (CDFs).
-For a LogNormal prior $X \sim \text{LogNormal}(\mu, \sigma^2)$, the probability of satisfying constraint $X < \theta$ is:
-
-$$P(X < \theta) = \Phi\left(\frac{\log(\theta) - \mu}{\sigma}\right)$$
+
+For **LogNormal parameters** ($R_{\text{on},i}$, $\delta^*_i$), the probability of satisfying constraint $X > \theta$ is:
+
+$$P(X > \theta) = 1 - \Phi\left(\frac{\log(\theta) - \mu}{\sigma}\right)$$
+
+For **Normal parameters** ($t^*_{0,\text{on}i}$), the probability of satisfying constraint $X > \theta$ is:
+
+$$P(X > \theta) = 1 - \Phi\left(\frac{\theta - \mu}{\sigma}\right)$$
 
@@
 
-# Prior hyperparameters (optimized)
+def calculate_normal_cdf_probability(loc, scale, threshold, upper=True):
+    """Calculate P(X > threshold) or P(X < threshold) for Normal(loc, scale)."""
+    if upper:
+        return 1 - stats.norm.cdf(threshold, loc=loc, scale=scale)
+    else:
+        return stats.norm.cdf(threshold, loc=loc, scale=scale)
+
+# Prior hyperparameters (updated for constrained parameterization with Normal onset times)
 priors = {
-    'alpha_off': {'loc': -2.3, 'scale': 0.5},    # log(0.1)
-    'alpha_on': {'loc': 0.69, 'scale': 0.5},     # log(2.0)
-    't_on_star': {'loc': -1.2, 'scale': 0.3},    # log(0.3)
-    'delta_star': {'loc': -1.0, 'scale': 0.35},  # log(0.37) [optimized]
+    'R_on': {'loc': 0.916, 'scale': 0.4},        # log(2.5), fold-change (LogNormal)
+    't_on_star': {'loc': 0.5, 'scale': 0.8},     # Normal(0.5, 0.8²), allows negatives [calibrated]
+    'delta_star': {'loc': -0.8, 'scale': 0.45},  # log(0.45) [calibrated for balanced patterns] (LogNormal)
 }
@@
     'activation': [
-        ('alpha_off', '<', 0.2), ('alpha_on', '>', 1.5),
+        ('R_on', '>', 3.0), ('t_on_star', '>', 0.0),
         ('t_on_star', '<', 0.4), ('delta_star', '>', 0.3)
     ],
+    'pre_activation': [
+        ('R_on', '>', 2.0), ('t_on_star', '<', 0.0),
+        ('delta_star', '>', 0.3)  # Simplified constraint
+    ],
     'transient': [
-        ('alpha_off', '<', 0.3), ('alpha_on', '>', 1.0),
-        ('t_on_star', '<', 0.5), ('delta_star', '<', 0.4)  # optimized
+        ('R_on', '>', 2.0), ('t_on_star', '>', 0.0),
+        ('t_on_star', '<', 0.5), ('delta_star', '<', 0.4)
     ],
     'sustained': [
-        ('alpha_off', '<', 0.3), ('alpha_on', '>', 1.0),
-        ('t_on_star', '<', 0.3), ('delta_star', '>', 0.45)  # optimized
+        ('R_on', '>', 2.0), ('t_on_star', '>', 0.0),
+        ('t_on_star', '<', 0.3), ('delta_star', '>', 0.5)
     ]
@@
         prior = priors[param]
+
+        if param == 't_on_star':  # Normal distribution
+            prob = calculate_normal_cdf_probability(
+                prior['loc'], prior['scale'], threshold, upper=(op == '>')
+            )
+        else:  # LogNormal distribution
             prob = calculate_lognormal_cdf_probability(
@@
             )
+
         joint_prob *= prob
@@
 The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while maintaining the hierarchical latent time structure.
-For visual clarity, we use $\theta_i$ to represent the vector of piecewise activation parameters $(\gamma^*_i, \alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i})$ and $\delta_i$ to represent the activation duration parameter.
+For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of relative temporal gene parameters $(\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i)$.
+Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.
 
@@
 pgm.add_node("sigma_theta", r"$\sigma_{\theta}^2$", 1.5, 6, fixed=True)
-pgm.add_node("mu_delta", r"$\mu_{\delta}$", 2.5, 6, fixed=True)
-pgm.add_node("sigma_delta", r"$\sigma_{\delta}^2$", 3.5, 6, fixed=True)
+pgm.add_node("mu_Delta", r"$\mu_{\Delta}$", 2.5, 6, fixed=True)
+pgm.add_node("sigma_Delta", r"$\sigma_{\Delta}^2$", 3.5, 6, fixed=True)
 pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 6, fixed=True)
@@
 pgm.add_node("theta_i", r"$\theta_i$", 1, 5)
-pgm.add_node("delta_i", r"$\delta_i$", 3, 5)
+pgm.add_node("Delta_i", r"$\Delta_i$", 3, 5)
 pgm.add_node("U_0i", r"$U_{0i}$", 4.5, 5)
@@
 pgm.add_edge("sigma_theta", "theta_i", plot_params=edge_params)
-pgm.add_edge("mu_delta", "delta_i", plot_params=edge_params)
-pgm.add_edge("sigma_delta", "delta_i", plot_params=edge_params)
+pgm.add_edge("mu_Delta", "Delta_i", plot_params=edge_params)
+pgm.add_edge("sigma_Delta", "Delta_i", plot_params=edge_params)
 pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
@@
 pgm.add_edge("t_scale", "t_star_j", plot_params=edge_params)
+# temporal scaling edge for gene parameters
+pgm.add_edge("T_max_star", "Delta_i", plot_params=edge_params)
 
@@
 pgm.add_edge("theta_i", "s_star_ij", plot_params=edge_params)
-pgm.add_edge("delta_i", "u_star_ij", plot_params=edge_params)
-pgm.add_edge("delta_i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("Delta_i", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("Delta_i", "s_star_ij", plot_params=edge_params)
 pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
@@
 In the PGM above, we use simplified notation for visual clarity:
-- $\theta_i$ represents the vector of piecewise activation parameters $(\gamma^*_i, \alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i})$
-- $\delta_i$ represents the activation duration parameter $\delta^*_i$ (waiting time from onset to deactivation)
+- $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
+- $\Delta_i$ represents the vector of **relative temporal** gene parameters $(\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i)$
 - The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
-- The hyperparameters $\mu_{\delta}, \sigma_{\delta}^2$ represent the hyperparameters for $\delta_i$
-
-**Detailed priors for piecewise activation parameters:**
-
-The priors are designed to encode biologically reasonable parameter ranges while maintaining sufficient flexibility for diverse gene expression patterns:
-
+- The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
+- The absolute timing parameters $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ and $\delta^*_i = T_{M}^* \times \tilde{\delta}^*_i$ are computed deterministically
+- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
+- **Key edge**: $T_{M}^* \rightarrow \Delta_i$ represents the scaling relationship for all temporal parameters
+
+**Detailed priors for constrained piecewise activation parameters:**
+
+The priors are designed to encode biologically reasonable parameter ranges while eliminating dimensional redundancy and cleanly separating temporal from non-temporal parameters:
+
+**Non-temporal gene-specific parameters:**
 \begin{align}
     \gamma^*_i &\sim \text{LogNormal}(\log(1.0), 0.5^2) \quad \text{(relative degradation rate)} \\
-    \alpha^*_{\text{off}i} &\sim \text{LogNormal}(\log(0.1), 0.5^2) \quad \text{(basal transcription rate)} \\
-    \alpha^*_{\text{on}i} &\sim \text{LogNormal}(\log(2.0), 0.5^2) \quad \text{(active transcription rate)} \\
-    t^*_{0,\text{on}i} &\sim \text{LogNormal}(\log(0.3), 0.3^2) \quad \text{(activation onset time)} \\
-    \delta^*_i &\sim \text{LogNormal}(\log(0.37), 0.35^2) \quad \text{(activation duration, optimized)}
+    R_{\text{on},i} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \quad \text{(activation fold-change)}
+\end{align}
+
+**Relative temporal gene-specific parameters:**
+\begin{align}
+    \tilde{t}^*_{0,\text{on}i} &\sim \text{Normal}(0.5, 0.8^2) \quad \text{(relative activation onset time, allows negatives)} \\
+    \tilde{\delta}^*_i &\sim \text{LogNormal}(\log(0.45), 0.45^2) \quad \text{(relative activation duration)}
+\end{align}
+
+**Absolute temporal parameters (deterministic transformations):**
+\begin{align}
+    t^*_{0,\text{on}i} &= T_{M}^* \times \tilde{t}^*_{0,\text{on}i} \quad \text{(absolute activation onset time)} \\
+    \delta^*_i &= T_{M}^* \times \tilde{\delta}^*_i \quad \text{(absolute activation duration)}
 \end{align}
@@
 
+**Non-temporal parameters:**
 - **$\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$**: Centers relative degradation around 1 (balanced splicing/degradation), with 95% of values between ~0.37 and ~2.7.
-This covers genes where mRNA is more stable than splicing time ($\gamma^* < 1$) to genes with rapid degradation ($\gamma^* > 1$).
-
-- **$\alpha^*_{\text{off}i} \sim \text{LogNormal}(\log(0.1), 0.5^2)$**: Centers basal transcription at 10% of characteristic rate, with 95% between ~0.037 and ~0.27.
-This represents low but non-zero basal expression typical of repressed genes.
-
-- **$\alpha^*_{\text{on}i} \sim \text{LogNormal}(\log(2.0), 0.5^2)$**: Centers active transcription at 2× characteristic rate, with 95% between ~0.74 and ~5.4.
-This ensures activation represents meaningful upregulation while avoiding extreme values.
-
-- **$t^*_{0,\text{on}i} \sim \text{LogNormal}(\log(0.3), 0.3^2)$**: Centers activation onset at 30% through the process timeline, with 95% between ~0.17 and ~0.53.
-This favors early activation while allowing variation in timing.
-
-- **$\delta^*_i \sim \text{LogNormal}(\log(0.37), 0.35^2)$**: Centers activation duration at 37% of the process timeline, with 95% between ~0.18 and ~0.76.
-This distribution was optimized to support both transient (< 40%) and sustained (> 45%) activation patterns with sufficient feasibility.
+
+- **$R_{\text{on},i} \sim \text{LogNormal}(\log(2.5), 0.4^2)$**: Centers fold-change at 2.5× during activation, with 95% between ~1.5 and ~4.2.
+
+**Relative temporal parameters:**
+- **$\tilde{t}^*_{0,\text{on}i} \sim \text{Normal}(0.5, 0.8^2)$**: Centers relative activation onset at 50% through the process timeline, with 95% between approximately -1.1 and +2.1.
+**Negative values** enable pre-activation scenarios.
+
+- **$\tilde{\delta}^*_i \sim \text{LogNormal}(\log(0.45), 0.45^2)$**: Centers relative activation duration at 45% of the process timeline, with 95% between ~0.18 and ~1.13.
+
+**Temporal scaling transformation:**
+- **Absolute onset**: $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ scales relative onset to absolute time
+- **Absolute duration**: $\delta^*_i = T_{M}^* \times \tilde{\delta}^*_i$ scales relative duration to absolute time
+- **Dimensional consistency**: All temporal comparisons now use consistently scaled absolute times
+
+**Parameter interpretation benefits:**
+- **Separation of concerns**: Non-temporal kinetics ($\theta_i$) vs temporal patterns ($\Delta_i$)
+- **Scale invariance**: Relative temporal patterns are independent of absolute time scale
+- **Biological realism**: Global time scale $T_{M}^*$ sets the pace for the entire process
 
@@
 
-#### **Identifiability Resolution**
-
-- **Eliminates redundant parameterization**: No separate inference of $u^*_{0i}, s^*_{0i}$ parameters
+#### **Resolved Dimensional Analysis Issues**
+
+- **Eliminates parameter redundancy**: No ambiguity between $U_{0i}$ and $\alpha^*_{\text{off}i}$ since $\alpha^*_{\text{off}i} = 1.0$ (fixed)
+- **Clear dimensional structure**: $U_{0i}$ unambiguously represents off-state concentration scale
+- **Proper parameter hierarchy**: Each inferred parameter has a unique biological role
+- **No separate inference of initial conditions**: $u^*_{0i}, s^*_{0i}$ fixed to steady-state values
 - **Biological realism**: Steady-state initial conditions are more plausible than arbitrary values
-- **Clear parameter roles**: Each piecewise parameter has distinct biological interpretation
+
+#### **Enhanced Temporal Modeling**
+
+- **Pre-activation scenarios**: Models genes already activated when observation begins
+- **Flexible onset timing**: Normal prior allows negative onset times naturally
+- **Biological realism**: Captures realistic experimental scenarios better
+- **Maintains population structure**: Hierarchical time modeling preserved
 
@@
 
-- **Activation genes**: Transition from low to high expression
+- **Activation genes**: Transition from baseline to high expression during observation
+- **Pre-activation genes**: Already activated before observation, showing high initial expression
 - **Decay genes**: Start high, remain high (no activation within observation window)
@@
 
-This study will validate the model's ability to:
-
-1. **Recover piecewise activation parameters**: $\alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i}, \delta^*_i$
-2. **Distinguish gene expression patterns**: Classify genes as activation, decay, transient, or sustained
-3. **Infer hierarchical time structure**: Recover population-level temporal parameters
-4. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
-5. **Scale to multiple genes**: Demonstrate efficiency with gene panels of varying sizes
-6. **Numerical robustness**: Maintain stability across the $\gamma^*_i = 1$ boundary
-7. **Phase assignment accuracy**: Correctly identify which phase each cell-gene pair occupies
-
-### Implementation and Validation Results
-
-The piecewise activation model has been successfully implemented within the PyroVelocity modular framework.
-This section demonstrates the complete parameter recovery validation workflow, from model creation through comprehensive validation across all four gene expression patterns.
-
-#### Model Creation and Setup
-
-The piecewise activation model is created using the PyroVelocity factory system with specialized components for piecewise dynamics and hierarchical priors:
-
-```python
-import torch
-import pyro
-import numpy as np
-import matplotlib.pyplot as plt
-import seaborn as sns
-from pathlib import Path
-
-# PyroVelocity imports
-from pyrovelocity.models.modular.factory import create_piecewise_activation_model
-# from pyrovelocity.plots.piecewise import (
-#     plot_piecewise_phase_portrait,
-#     plot_piecewise_trajectories,
-#     plot_activation_timing,
-# )
-from pyrovelocity.plots.predictive_checks import (
-    plot_prior_predictive_checks,
-    plot_posterior_predictive_checks,
-)
-# from pyrovelocity.plots.parameter_recovery import (
-#     plot_parameter_recovery_diagnostics,
-# )
-
-# Set random seeds for reproducibility
-torch.manual_seed(42)
-pyro.set_rng_seed(42)
-np.random.seed(42)
-
-# Create the piecewise activation model
-model = create_piecewise_activation_model()
-
-print(f"Created model: {model}")
-print(f"Model components:")
-print(f"  Dynamics: {model.dynamics_model}")
-print(f"  Prior: {model.prior_model}")
-print(f"  Likelihood: {model.likelihood_model}")
-print(f"  Guide: {model.guide_model}")
-```
-
-#### Synthetic Data Generation for All Expression Patterns
-
-We generate synthetic datasets representing each of the four gene expression patterns using the generic predictive sampling interface:
-
-```python
-# Generate validation datasets for all patterns using generic API
-validation_datasets = {}
-validation_parameters = {}
-
-# Define pattern configurations (2 parameter sets per pattern = 8 total)
-pattern_configs = [
-    # Activation patterns
-    {"pattern": "activation", "set_id": 1, "seed": 42},
-    {"pattern": "activation", "set_id": 2, "seed": 43},
-    # Decay patterns
-    {"pattern": "decay", "set_id": 1, "seed": 44},
-    {"pattern": "decay", "set_id": 2, "seed": 45},
-    # Transient patterns
-    {"pattern": "transient", "set_id": 1, "seed": 46},
-    {"pattern": "transient", "set_id": 2, "seed": 47},
-    # Sustained patterns
-    {"pattern": "sustained", "set_id": 1, "seed": 48},
-    {"pattern": "sustained", "set_id": 2, "seed": 49},
-]
-
-for config in pattern_configs:
-    pattern = config["pattern"]
-    set_id = config["set_id"]
-    seed = config["seed"]
-    dataset_key = f"{pattern}_{set_id}"
-
-    print(f"Generating synthetic data for {dataset_key}...")
-
-    # Set PyTorch random seed for reproducibility
-    torch.manual_seed(seed)
-    pyro.set_rng_seed(seed)
-
-    # Sample system parameters constrained to this pattern
-    true_system_params = model.sample_system_parameters(
-        pattern=pattern,
-        set_id=set_id,
-        num_samples=1,
-        constrain_to_pattern=True,
-        n_genes=5,
-        n_cells=200
-    )
-
-    # Generate synthetic data using the generic predictive sampling interface
-    torch.manual_seed(seed + 100)
-    pyro.set_rng_seed(seed + 100)
-    adata = model.generate_predictive_samples(
-        num_cells=200,
-        num_genes=5,
-        samples=true_system_params,  # Use constrained parameters as "posterior"
-        return_format="anndata"
-    )
-
-    # Store dataset and true parameters
-    validation_datasets[dataset_key] = adata
-    validation_parameters[dataset_key] = true_system_params
-
-    print(f"  Generated {adata.n_obs} cells × {adata.n_vars} genes")
-    print(f"  Pattern: {adata.uns.get('pattern', 'unknown')}")
-    print(f"  True parameters: {list(true_system_params.keys())}")
-
-print(f"\nGenerated {len(validation_datasets)} synthetic datasets")
-```
-
-#### Prior Predictive Checks
-
-Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns:
-
-```python
-# Generate prior predictive samples using the generic interface
-print("Performing prior predictive checks...")
-
-# Set seeds for reproducible prior sampling
-torch.manual_seed(0)
-pyro.set_rng_seed(0)
-
-# Generate prior predictive samples (no samples argument = sample from prior)
-prior_predictive_adata = model.generate_predictive_samples(
-    num_cells=200,
-    num_genes=10,
-    num_samples=100,  # Generate 100 prior samples
-    return_format="anndata"
-)
-
-# Extract parameter samples for validation
-torch.manual_seed(1)
-pyro.set_rng_seed(1)
-prior_parameter_samples = model.sample_system_parameters(
-    num_samples=1000,
-    constrain_to_pattern=False,  # Use full prior ranges
-    n_genes=10,
-    n_cells=200
-)
-
-# Plot prior predictive checks with automatic saving
-# This will save both PNG and PDF copies to reports/docs/prior_predictive/
-fig_prior = plot_prior_predictive_checks(
-    model=model,
-    prior_adata=prior_predictive_adata,
-    prior_parameters=prior_parameter_samples,
-    figsize=(15, 10),
-    save_path="reports/docs/prior_predictive",
-    figure_name="piecewise_activation_prior_checks"
-)
-
-plt.suptitle("Prior Predictive Checks: Piecewise Activation Model", fontsize=16)
-plt.tight_layout()
-plt.show()
-
-# Validate prior parameter ranges
-print("\nPrior parameter range validation:")
-for param_name, samples in prior_parameter_samples.items():
-    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
-        print(f"{param_name}:")
-        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
-        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
-
-# Note: Pattern classification and other specialized validation methods
-# will be implemented later as separate functionality
-```
-
-#### Parameter Recovery Validation Across All Patterns
-
-*Note: The `validate_parameter_recovery()` method is planned for Phase 5 implementation. For now, we demonstrate the synthetic data generation capability.*
-
-```python
-# Placeholder for parameter recovery validation
-# TODO: Implement validate_parameter_recovery() method in Phase 5
-
-print("Parameter recovery validation will be implemented in Phase 5.")
-print("Current implementation successfully demonstrates:")
-print("✅ Pattern-constrained parameter generation")
-print("✅ Synthetic data generation with known ground truth")
-print("✅ AnnData integration with metadata storage")
-print("✅ Support for all 4 expression patterns")
-
-# For now, we can examine the generated datasets
-print(f"\nGenerated validation datasets:")
-for dataset_key, adata in validation_datasets.items():
-    params = validation_parameters[dataset_key]
-    print(f"\n{dataset_key}:")
-    print(f"  Data shape: {adata.shape}")
-    print(f"  Pattern: {adata.uns.get('pattern', 'unknown')}")
-    print(f"  Parameter count: {len(params)} parameter types")
-
-    # Show parameter ranges for this pattern
-    alpha_off = params['alpha_off']
-    alpha_on = params['alpha_on']
-    t_on_star = params['t_on_star']
-    fold_change = alpha_on / alpha_off
-
-    print(f"  α*_off: {alpha_off.min():.3f} - {alpha_off.max():.3f}")
-    print(f"  α*_on: {alpha_on.min():.3f} - {alpha_on.max():.3f}")
-    print(f"  t*_on: {t_on_star.min():.3f} - {t_on_star.max():.3f}")
-    print(f"  Fold-change: {fold_change.min():.1f} - {fold_change.max():.1f}")
-
-print(f"\n🎉 Phase 4 (Data Generation) Complete!")
-print("Ready to proceed with Phase 5 (Parameter Recovery Validation)")
-```
-
-#### Comprehensive Visualization Suite
-
-*Note: Visualization functions are planned for Phase 6 implementation.*
-
-<!--
-TODO: Implement visualization functions in Phase 6:
-- plot_parameter_recovery_diagnostics()
-- plot_piecewise_phase_portrait()
-- plot_piecewise_trajectories()
-- plot_activation_timing()
-- plot_posterior_predictive_checks()
--->
-
-```python
-# Placeholder for comprehensive visualization suite
-# TODO: Implement in Phase 6
-
-print("Visualization suite will be implemented in Phase 6.")
-print("Planned visualizations:")
-print("📊 Parameter recovery diagnostic plots")
-print("🌀 Phase portraits for each pattern")
-print("📈 Temporal trajectories showing piecewise dynamics")
-print("⏰ Activation timing analysis")
-print("🔍 Posterior predictive checks")
-
-print("\n✅ Validation Study 3 Framework Complete!")
-print("🎯 Ready for Phase 5: Parameter Recovery Validation Implementation")
-```
-
-## Validation Study 4: Dimensionless simulated dynamics with latent time and continuous pulse activation
-
-*To be implemented: Full complexity model with numerical integration using torchode for continuous activation functions, representing the complete PyroVelocity modular implementation.*
+This study will validate the constrained model's ability to:
+
+1. **Recover constrained piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{0,\text{on}i}$, $\delta^*_i$
+2. **Handle pre-activation scenarios**: Correctly infer negative onset times and their biological implications
+3. **Distinguish enhanced expression patterns**: Classify genes across all five patterns including pre-activation
+4. **Maintain hierarchical time inference**: Recover population-level temporal parameters
+5. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
+6. **Scale efficiently**: Demonstrate computational advantages of reduced parameter space
+7. **Validate fold-change interpretation**: Ensure $R_{\text{on},i}$ estimates match true activation strength
+8. **Assess dimensional consistency**: Verify elimination of parameter redundancy issues
+9. **Validate temporal parameter scaling**: Ensure $T_{M}^*$ scaling correctly propagates to absolute timing parameters
+10. **Assess parameter separation**: Verify clean separation between kinetic and temporal parameter inference

### 2025-06-12 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): add dimless latent time piecewise activation model description to model progression and rename sections
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-### Dimensionless Simulated Dynamics with Observed Time
-
-This model replaces analytical solutions with numerical integration:
-
-- **Time coordinates**: Observed dimensionless time $t^*_j$
-- **Dynamics**: Numerical integration of dimensionless ODEs using torchode
-- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$
-- **Prior**: Log-normal hierarchical priors
-- **Observation Model**: Same Poisson scaling approach
-- **Likelihood**: Poisson
-
-### Dimensionless Simulated Dynamics with Latent Time
-
-This model represents the full complexity of the PyroVelocity modular implementation:
-
-- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
-- **Dynamics**: Numerical integration of dimensionless ODEs using torchode
-- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{M}^*$, $t_{loc}$, $t_{scl}$)
-- **Prior**: Gamma priors for time scales, Normal for relative positions
-- **Observation Model**: Same Poisson scaling approach
+### Dimensionless Analytical Dynamics with Latent Time and Piecewise Activation
+
+This model extends the latent time framework with piecewise constant transcription rates:
+
+- **Time coordinates**: Same hierarchical latent $t^*_j$ structure as previous model
+- **Dynamics**: Piecewise constant transcription with three phases (off, on, return-to-off)
+- **Parameters**: $\gamma^*_i$, $R_{\text{on},i}$ (fold-change), $t^*_{0,\text{on}i}$ (onset), $\delta^*_i$ (duration), plus hierarchical time structure
+- **Key innovation**: Fixed reference state ($\alpha^*_{\text{off}i} = 1.0$) eliminates parameter redundancy
+- **Enhanced patterns**: Supports pre-activation scenarios with negative onset times
+- **Analytical solutions**: Closed-form solutions for all three phases, including $\gamma^*_i = 1$ special case
+- **Biological realism**: Steady-state initial conditions and rich expression pattern representation
+- **Observation Model**: Same Poisson scaling with characteristic scales and capture efficiency
 - **Likelihood**: Poisson
@@
 from pyrovelocity.models.modular.factory import (
-    # create_simple_deterministic_model_with_validation,
+    create_piecewise_activation_model,
     create_legacy_model2,
@@
 
-## Validation Study 1: Dimensionless analytical dynamics with observed time
+## Dimensionless analytical dynamics with observed time {#sec-dimless-observed-time}
 
@@
 
-## Validation Study 2: Dimensionless analytical dynamics with latent time
+## Dimensionless analytical dynamics with latent time {#sec-dimless-latent-time}
 
@@
 
-The model uses the same dimensionless dynamics and observation model as Validation Study 1, but replaces observed time coordinates with **hierarchical latent time coordinates**.
+The model uses the same dimensionless dynamics and observation model as the model described in @sec-dimless-observed-time, but replaces observed time coordinates with **hierarchical latent time coordinates**.
 This approach borrows strength across cells while allowing individual temporal variation.
@@
 
-## Validation Study 3: Dimensionless Analytical Dynamics with Latent Time and Piecewise Activation
+## Dimensionless Analytical Dynamics with Latent Time and Piecewise Activation {#sec-dimless-latent-piecewise}
 
@@
 where:
+
 - $\alpha^*_{\text{off}i} = 1.0$ is the fixed reference transcription rate
@@
 In the PGM above, we use simplified notation for visual clarity:
+
 - $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$

### 2025-06-12 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): update expression pattern descriptions
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-### Gene Expression Pattern Capabilities
+### Gene Expression Pattern Classification
 
 This piecewise framework enables representation of diverse gene expression patterns through specific parameter combinations.
-Each pattern is defined by quantitative parameter ranges that create distinct temporal dynamics:
-
-#### **1. Activation Patterns**
-
-**Characteristics**: Genes that start with baseline expression and become strongly activated during the process.
+Each pattern is defined by quantitative parameter ranges applied to **relative temporal parameters** that create distinct temporal dynamics when scaled by the global time parameter $T_M^*$.
+
+**Parameter Framework**: The classification uses relative temporal parameters ($\tilde{t}^*_{0,\text{on}i}$, $\tilde{\delta}^*_i$) that are scaled by the global maximum time $T_M^*$ to produce absolute temporal parameters:
+
+$$\begin{align}
+t^*_{0,\text{on}i} &= T_M^* \times \tilde{t}^*_{0,\text{on}i} \quad \text{(absolute activation onset time)} \\
+\delta^*_i &= T_M^* \times \tilde{\delta}^*_i \quad \text{(absolute activation duration)}
+\end{align}$$
+
+This hierarchical structure separates global temporal scaling from gene-specific temporal patterns, enabling consistent pattern classification across different experimental timescales.
+
+#### **Pre-Activation Patterns**
+
+**Characteristics**: Genes that were already activated before observation began, showing decay-only dynamics from an initially activated steady-state.
 
@@
 
-- $R_{\text{on},i} > 3.0$ (strong fold-change during activation)
-- $t^*_{0,\text{on}i} > 0$ and $t^*_{0,\text{on}i} < 0.4$ (activation occurs after observation start but early)
-- $\delta^*_i > 0.3$ (sustained activation duration)
-
-**Biological examples**: Developmental transcription factors, cell cycle genes, stress response genes
-
-#### **2. Pre-Activation Patterns**
-
-**Characteristics**: Genes that were already activated before observation began, showing high initial expression.
+- $\tilde{t}^*_{0,\text{on}i} < 0.0$ (relative activation onset before observation window)
+- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change during activation)
+
+**Biological interpretation**: These genes underwent activation before the experimental observation window.
+During the observed period, they exhibit decay-only dynamics as they return toward baseline expression levels.
+The negative relative onset time indicates that activation occurred in the "past" relative to the observation start.
+
+**Biological examples**: Early developmental genes, maternal transcripts during development, lineage commitment factors established before observation, stress response genes in ongoing stress conditions
+
+#### **Transient Activation Patterns**
+
+**Characteristics**: Genes that show complete activation-decay cycles within the observation window, exhibiting brief, pulse-like activation followed by return toward baseline.
 
@@
 
-- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change)
-- $t^*_{0,\text{on}i} < 0$ (activation onset before observation window)
-- $\delta^*_i > |t^*_{0,\text{on}i}|$ (activation duration extends into observation window)
-
-**Biological examples**: Early developmental genes, lineage commitment factors, stress response genes in ongoing stress
-
-#### **3. Decay-Only Patterns**
-
-**Characteristics**: Genes that start with high expression and gradually decrease without reactivation.
+- $\tilde{t}^*_{0,\text{on}i} > 0.0$ (relative activation within observation window)
+- $\tilde{t}^*_{0,\text{on}i} < 0.5$ (activation early enough to complete cycle within 50% of timeline)
+- $\tilde{\delta}^*_i < 0.4$ (short activation duration, 40% of timeline or less)
+- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)
+
+**Biological interpretation**: These genes activate early enough and for a short enough duration that the complete activation-decay cycle is observable within the experimental window.
+The constraints ensure that both the activation and subsequent decay phases are captured.
+
+**Biological examples**: Immediate early genes, signaling molecules, transient developmental regulators, cell cycle checkpoint genes, acute stress response genes
+
+#### **Sustained Activation Patterns**
+
+**Characteristics**: Genes that become activated early and remain highly expressed throughout most of the observation period, showing net increase over the observation window.
 
@@
 
-- $R_{\text{on},i} > 1.5$ (moderate fold-change, but not observed during activation)
-- $t^*_{0,\text{on}i} > T^*_M$ (activation onset beyond observation window) OR
-- $t^*_{0,\text{on}i} < 0$ and $\delta^*_i < |t^*_{0,\text{on}i}|$ (activation ended before observation)
-
-**Biological examples**: Maternal transcripts during development, pluripotency factors during differentiation
-
-#### **4. Transient Activation Patterns**
-**Characteristics**: Genes that show brief, pulse-like activation followed by return to baseline.
-
-**Parameter specifications**:
-
-- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change)
-- $t^*_{0,\text{on}i} > 0$ and $t^*_{0,\text{on}i} < 0.5$ (activation during observation)
-- $\delta^*_i < 0.4$ (brief activation duration)
-
-**Biological examples**: Immediate early genes, signaling molecules, transient developmental regulators
-
-#### **5. Sustained Activation Patterns**
-**Characteristics**: Genes that become activated and remain highly expressed throughout most of the process.
-
-**Parameter specifications**:
-
-- $R_{\text{on},i} > 2.0$ (strong fold-change)
-- $t^*_{0,\text{on}i} > 0$ and $t^*_{0,\text{on}i} < 0.3$ (early activation onset)
-- $\delta^*_i > 0.5$ (long activation duration)
-
-**Biological examples**: Lineage-specific transcription factors, structural proteins, metabolic enzymes
+- $\tilde{t}^*_{0,\text{on}i} > 0.0$ (relative activation within observation window)
+- $\tilde{t}^*_{0,\text{on}i} < 0.3$ (early activation onset within 30% of timeline)
+- $\tilde{\delta}^*_i > 0.5$ (long activation duration, more than 50% of timeline)
+- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)
+
+**Biological interpretation**: These genes activate early in the observation window and maintain high expression levels throughout most of the observed period.
+The long activation duration relative to the observation window means that decay back to baseline is either minimal or occurs beyond the observation period.
+
+**Biological examples**: Lineage-specific transcription factors, structural proteins, metabolic enzymes, terminal differentiation markers, long-term stress adaptation genes
 
@@
 
-```python
+```{python}
+#| code-fold: true
+#| code-summary: "Show/Hide Code"
+#| eval: false
+
 # CDF-based feasibility analysis for pattern constraints
@@
 
-# Prior hyperparameters (updated for constrained parameterization with Normal onset times)
+# Prior hyperparameters (updated for relative temporal parameterization)
 priors = {
     'R_on': {'loc': 0.916, 'scale': 0.4},              # log(2.5), fold-change (LogNormal)
-    't_on_star': {'loc': 0.5, 'scale': 0.8},     # Normal(0.5, 0.8²), allows negatives [calibrated]
-    'delta_star': {'loc': -0.8, 'scale': 0.45},  # log(0.45) [calibrated for balanced patterns] (LogNormal)
+    'tilde_t_on_star': {'loc': 0.5, 'scale': 0.8},     # Normal(0.5, 0.8²), relative onset (allows negatives)
+    'tilde_delta_star': {'loc': -0.8, 'scale': 0.45},  # log(0.45), relative duration (LogNormal)
 }
 
-# Calculate constraint feasibility for each pattern
+# Calculate constraint feasibility for each pattern (updated to match script implementation)
 patterns = {
-    'activation': [
-        ('R_on', '>', 3.0), ('t_on_star', '>', 0.0),
-        ('t_on_star', '<', 0.4), ('delta_star', '>', 0.3)
-    ],
     'pre_activation': [
-        ('R_on', '>', 2.0), ('t_on_star', '<', 0.0),
-        ('delta_star', '>', 0.3)  # Simplified constraint
+        ('tilde_t_on_star', '<', 0.0),  # Activation before observation window
+        ('R_on', '>', 2.0),             # Moderate to strong fold change
     ],
     'transient': [
-        ('R_on', '>', 2.0), ('t_on_star', '>', 0.0),
-        ('t_on_star', '<', 0.5), ('delta_star', '<', 0.4)
+        ('tilde_t_on_star', '>', 0.0),  # Activation within observation window
+        ('tilde_t_on_star', '<', 0.5),  # Early enough to complete cycle (50% of timeline)
+        ('tilde_delta_star', '<', 0.4), # Short pulse duration (40% of timeline)
+        ('R_on', '>', 2.0),             # Sufficient fold change to observe
     ],
     'sustained': [
-        ('R_on', '>', 2.0), ('t_on_star', '>', 0.0),
-        ('t_on_star', '<', 0.3), ('delta_star', '>', 0.5)
+        ('tilde_t_on_star', '>', 0.0),  # Activation within observation window
+        ('tilde_t_on_star', '<', 0.3),  # Early activation onset (30% of timeline)
+        ('tilde_delta_star', '>', 0.5), # Long activation duration (50% of timeline)
+        ('R_on', '>', 2.0),             # Strong fold change
     ]
@@
 
-        if param == 't_on_star':  # Normal distribution
+        if param == 'tilde_t_on_star':  # Normal distribution
             prob = calculate_normal_cdf_probability(
@@
 - $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
-- **Key edge**: $T_{M}^* \rightarrow \Delta_i$ represents the scaling relationship for all temporal parameters
+- $T_{M}^* \rightarrow \Delta_i$ represents the scaling relationship for all temporal parameters
 
@@
 
+**Pattern classification framework**: The three expression patterns are defined by constraints on the relative temporal parameters:
+
+- **Pre-activation**: $\tilde{t}^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
+- **Transient**: $0.0 < \tilde{t}^*_{0,\text{on}i} < 0.5$, $\tilde{\delta}^*_i < 0.4$, and $R_{\text{on},i} > 2.0$
+- **Sustained**: $0.0 < \tilde{t}^*_{0,\text{on}i} < 0.3$, $\tilde{\delta}^*_i > 0.5$, and $R_{\text{on},i} > 2.0$
+
 **Biological interpretation of prior choices:**
@@
 **Non-temporal parameters:**
+
 - **$\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$**: Centers relative degradation around 1 (balanced splicing/degradation), with 95% of values between ~0.37 and ~2.7.
@@
 **Relative temporal parameters:**
+
 - **$\tilde{t}^*_{0,\text{on}i} \sim \text{Normal}(0.5, 0.8^2)$**: Centers relative activation onset at 50% through the process timeline, with 95% between approximately -1.1 and +2.1.
-**Negative values** enable pre-activation scenarios.
+**Negative values** enable pre-activation scenarios where genes were already activated before observation began.
 
 - **$\tilde{\delta}^*_i \sim \text{LogNormal}(\log(0.45), 0.45^2)$**: Centers relative activation duration at 45% of the process timeline, with 95% between ~0.18 and ~1.13.
+This enables both transient (short duration) and sustained (long duration) activation patterns.
+
+**Pattern classification implications:**
+
+- **Pre-activation** ($\tilde{t}^*_{0,\text{on}i} < 0$): ~26.6% probability under the prior
+- **Transient** ($0 < \tilde{t}^*_{0,\text{on}i} < 0.5$ and $\tilde{\delta}^*_i < 0.4$): Balanced probability for observable complete cycles
+- **Sustained** ($0 < \tilde{t}^*_{0,\text{on}i} < 0.3$ and $\tilde{\delta}^*_i > 0.5$): Balanced probability for long-duration activation
 
 **Temporal scaling transformation:**
+
 - **Absolute onset**: $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ scales relative onset to absolute time
@@
 **Parameter interpretation benefits:**
+
 - **Separation of concerns**: Non-temporal kinetics ($\theta_i$) vs temporal patterns ($\Delta_i$)
@@
 
-- **Activation genes**: Transition from baseline to high expression during observation
-- **Pre-activation genes**: Already activated before observation, showing high initial expression
-- **Decay genes**: Start high, remain high (no activation within observation window)
-- **Transient genes**: Brief activation followed by return to baseline
-- **Sustained genes**: Long-duration activation
+The framework captures three core expression patterns that encompass the full range of biologically relevant temporal dynamics:
+
+- **Pre-activation genes**: Already activated before observation, showing decay-only dynamics from activated steady-state
+- **Transient genes**: Complete activation-decay cycles within observation window, exhibiting pulse-like dynamics
+- **Sustained genes**: Early activation with long duration, showing net increase over observation period
+
+**Pattern coverage**: These three patterns effectively capture the essential temporal dynamics while eliminating redundant classifications.
+The relative temporal parameterization ensures consistent pattern identification across different experimental timescales.
 
@@
 
-1. **Recover constrained piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{0,\text{on}i}$, $\delta^*_i$
-2. **Handle pre-activation scenarios**: Correctly infer negative onset times and their biological implications
-3. **Distinguish enhanced expression patterns**: Classify genes across all five patterns including pre-activation
-4. **Maintain hierarchical time inference**: Recover population-level temporal parameters
+1. **Recover constrained piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $\tilde{t}^*_{0,\text{on}i}$, $\tilde{\delta}^*_i$
+2. **Handle pre-activation scenarios**: Correctly infer negative relative onset times and their biological implications
+3. **Distinguish expression patterns**: Classify genes across the three core patterns (pre-activation, transient, sustained)
+4. **Maintain hierarchical time inference**: Recover population-level temporal parameters ($T_M^*$, $t_{loc}$, $t_{scl}$)
 5. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
@@
 7. **Validate fold-change interpretation**: Ensure $R_{\text{on},i}$ estimates match true activation strength
-8. **Assess dimensional consistency**: Verify elimination of parameter redundancy issues
-9. **Validate temporal parameter scaling**: Ensure $T_{M}^*$ scaling correctly propagates to absolute timing parameters
-10. **Assess parameter separation**: Verify clean separation between kinetic and temporal parameter inference
+8. **Assess dimensional consistency**: Verify elimination of parameter redundancy through fixed $\alpha^*_{\text{off}i} = 1.0$
+9. **Validate temporal parameter scaling**: Ensure $T_{M}^*$ scaling correctly propagates from relative to absolute timing parameters
+10. **Assess parameter separation**: Verify clean separation between kinetic ($\gamma^*_i$, $R_{\text{on},i}$) and temporal ($\tilde{t}^*_{0,\text{on}i}$, $\tilde{\delta}^*_i$) parameter inference
+11. **Validate relative parameter framework**: Confirm that relative temporal parameters enable consistent pattern classification across different experimental timescales
+12. **Assess pattern coverage**: Verify that the three-pattern framework adequately captures the diversity of gene expression dynamics in real biological systems
+
+### Parameter Recovery Validation Implementation
+
+The following example demonstrates the user-facing API for performing parameter recovery validation with the piecewise activation model:
+
+```{python}
+#| label: piecewise-activation-validation
+#| code-fold: true
+#| code-summary: "Show/Hide Code"
+#| eval: false
+
+# Create the piecewise activation model
+model = create_piecewise_activation_model()
+
+print(f"Created piecewise activation model: {model}")
+print(f"Model components:")
+print(f"  Dynamics: {model.dynamics_model}")
+print(f"  Prior: {model.prior_model}")
+print(f"  Likelihood: {model.likelihood_model}")
+print(f"  Guide: {model.guide_model}")
+
+# Generate prior predictive data with known true parameters
+print(f"\nGenerating prior predictive data for validation...")
+prior_predictive_adata = model.generate_predictive_samples(
+    num_cells=200,
+    num_genes=100,
+    num_samples=1,
+    return_format="anndata"
+)
+
+print(f"Prior predictive data generated:")
+print(f"  Cells: {prior_predictive_adata.n_obs}")
+print(f"  Genes: {prior_predictive_adata.n_vars}")
+print(f"  Layers: {list(prior_predictive_adata.layers.keys())}")
+
+# Extract true parameters for validation
+true_parameters = prior_predictive_adata.uns.get('true_parameters', {})
+print(f"\nTrue parameters available for validation:")
+for key in sorted(true_parameters.keys()):
+    if key.startswith(('R_on', 'tilde_t_on', 'tilde_delta', 't_on', 'delta', 'gamma_star', 'T_M_star')):
+        param_value = true_parameters[key]
+        if hasattr(param_value, 'shape') and param_value.size > 1:
+            print(f"  {key}: shape {param_value.shape}, range [{param_value.min():.3f}, {param_value.max():.3f}]")
+        else:
+            print(f"  {key}: {param_value:.3f}")
+
+# Analyze pattern distribution in generated data
+print(f"\nPattern analysis of generated genes:")
+if 'tilde_t_on_star' in true_parameters and 'tilde_delta_star' in true_parameters:
+    tilde_t_on = true_parameters['tilde_t_on_star']
+    tilde_delta = true_parameters['tilde_delta_star']
+    R_on = true_parameters['R_on']
+
+    # Count genes in each pattern
+    pre_activation = (tilde_t_on < 0.0) & (R_on > 2.0)
+    transient = (tilde_t_on > 0.0) & (tilde_t_on < 0.5) & (tilde_delta < 0.4) & (R_on > 2.0)
+    sustained = (tilde_t_on > 0.0) & (tilde_t_on < 0.3) & (tilde_delta > 0.5) & (R_on > 2.0)
+
+    print(f"  Pre-activation: {pre_activation.sum()} genes ({pre_activation.mean()*100:.1f}%)")
+    print(f"  Transient: {transient.sum()} genes ({transient.mean()*100:.1f}%)")
+    print(f"  Sustained: {sustained.sum()} genes ({sustained.mean()*100:.1f}%)")
+    print(f"  Other patterns: {(~(pre_activation | transient | sustained)).sum()} genes")
+
+# Train model on prior predictive data (parameter recovery)
+print(f"\nTraining model for parameter recovery...")
+trained_model = model.train(
+    adata=prior_predictive_adata,
+    max_epochs=1000,  # Reduced for demonstration
+    learning_rate=0.01,
+    early_stopping=True,
+    early_stopping_patience=10,
+    use_gpu="auto",
+    seed=42
+)
+
+# Generate posterior samples from trained model
+print(f"\nGenerating posterior samples...")
+posterior_samples = trained_model.generate_posterior_samples(
+    adata=prior_predictive_adata,
+    num_samples=30,
+    seed=42,
+    return_tensors=True
+)
+
+print(f"Posterior samples generated:")
+for key in sorted(posterior_samples.keys()):
+    if key.startswith(('R_on', 'tilde_t_on', 'tilde_delta', 't_on', 'delta', 'gamma_star', 'T_M_star')):
+        param_shape = posterior_samples[key].shape if hasattr(posterior_samples[key], 'shape') else len(posterior_samples[key])
+        print(f"  {key}: {param_shape}")
+
+print(f"\nParameter recovery validation completed!")
+print(f"This demonstrates the complete workflow from:")
+print(f"  1. Model creation with create_piecewise_activation_model()")
+print(f"  2. Prior predictive data generation with known parameters")
+print(f"  3. Model training on synthetic data")
+print(f"  4. Posterior sampling for parameter recovery assessment")
+print(f"  5. Comparison of true vs recovered parameters")
+print(f"  6. Pattern classification analysis with 3-pattern framework")
+```
+
+This validation workflow demonstrates the key capabilities of the piecewise activation model:
+
+- **Analytical tractability**: Fast likelihood evaluation enables efficient parameter recovery validation
+- **Rich parameter space**: Includes fold-change parameters, relative temporal parameters, and hierarchical time structure
+- **Pre-activation modeling**: Supports negative onset times for genes already activated before observation
+- **Dimensional consistency**: Eliminates parameter redundancy through fixed reference state parameterization
+- **Biological interpretability**: All parameters have clear biological meaning and interpretable ranges
+- **Coarse-grained expression pattern classification**: Classification of temporal expression patterns covering pre-activation, transient, and sustained that should all be able to be generated by the prior and prior predictive in order to ensure the ability to represent the expected collection of expression patterns in the underlying data
+- **Relative temporal parameterization**: Enables interpretable parameter values and consistent pattern classification across different experimental timescales and even distinct data sets
+
+The complete validation scripts referenced in the introduction (`prior-predictive-check.py` and `posterior-predictive-check.py`) provide comprehensive parameter recovery validation including visualization and statistical assessment of recovery performance.

### 2025-06-12 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): add posterior predictive description
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-**Observation model (unchanged):**
+**Observation model:**
 \begin{align}
@@
 \end{align}
+
+### Posterior Predictive Check Framework
+
+The posterior predictive checking procedure provides a rigorous mathematical framework for validating the piecewise activation model's ability to recover known parameters from synthetic data.
+This Bayesian validation approach systematically assesses parameter recovery performance by comparing posterior predictive samples to the original synthetic observations.
+
+#### **Mathematical Formulation of Posterior Predictive Checking**
+
+**Prior Predictive Distribution**
+
+The prior predictive distribution represents the model's predictions before observing any data:
+
+$$p(\tilde{y}) = \int p(\tilde{y} | \theta) p(\theta) d\theta$$
+
+where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i})$, hierarchical time parameters $(T_{M}^*, t_{loc}, t_{scl}, t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.
+
+**Posterior Predictive Distribution**
+
+After observing synthetic data $y_{obs}$, the posterior predictive distribution incorporates learned parameter constraints:
+
+$$p(\tilde{y} | y_{obs}) = \int p(\tilde{y} | \theta) p(\theta | y_{obs}) d\theta$$
+
+This distribution quantifies the model's uncertainty about future observations given what it has learned from the synthetic data.
+
+**Hierarchical Uncertainty Decomposition**
+
+The piecewise activation model exhibits a complex uncertainty hierarchy that propagates from parameters through latent concentrations to final observations:
+
+1. **Parameter uncertainty**: $\text{Var}[\theta | y_{obs}]$ from the posterior distribution over all model parameters
+2. **Temporal uncertainty**: Uncertainty in latent time coordinates $t^*_j$ and their hierarchical structure
+3. **Kinetic uncertainty**: Uncertainty in gene-specific kinetic parameters $\gamma^*_i$ and fold-changes $R_{\text{on},i}$
+4. **Concentration uncertainty**: Uncertainty in predicted latent concentrations $u^*_{ij}(\theta), s^*_{ij}(\theta)$ arising from parameter uncertainty
+5. **Observation uncertainty**: Poisson noise in final count observations
+
+The total predictive uncertainty decomposes as:
+
+$$\text{Var}[\tilde{y} | y_{obs}] = \mathbb{E}_\theta[\text{Var}[y|\theta]] + \text{Var}_\theta[\mathbb{E}[y|\theta]]$$
+
+where the first term represents irreducible observation noise (Poisson variance) and the second term represents reducible uncertainty due to parameter estimation.
+
+#### **Model-Specific Implementation for Piecewise Activation**
+
+**Complete Parameter Space**
+
+The piecewise activation model's parameter space $\theta$ includes:
+
+- **Non-temporal gene parameters**: $\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$, $R_{\text{on},i} \sim \text{LogNormal}(\log(2.5), 0.4^2)$
+- **Relative temporal gene parameters**: $\tilde{t}^*_{0,\text{on}i} \sim \text{Normal}(0.5, 0.8^2)$, $\tilde{\delta}^*_i \sim \text{LogNormal}(\log(0.45), 0.45^2)$
+- **Characteristic scales**: $U_{0i} \sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)$
+- **Hierarchical time parameters**: $T_{M}^* \sim \text{Gamma}(\alpha_{T}, \beta_{T})$, $t_{loc} \sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}})$, $t_{scl} \sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}})$
+- **Cell-specific times**: $\tilde{t}_j \sim \text{Normal}(t_{loc}, t_{scl}^2)$, $t^*_j = T_{M}^* \times \max(\tilde{t}_j, \epsilon)$
+- **Capture efficiencies**: $\lambda_j \sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)$
+
+**Dimensionless Dynamics with Piecewise Activation**
+
+The latent concentrations are computed using analytical solutions that depend on which phase each cell-gene pair occupies:
+
+$$u^*_{ij}(\theta), s^*_{ij}(\theta) = f_{\text{piecewise}}(t^*_j, \gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i)$$
+
+where $f_{\text{piecewise}}$ represents the appropriate analytical solution (Phase 1, 2, or 3) determined by comparing $t^*_j$ to the gene-specific activation timing parameters.
+
+**Poisson Observation Model with Scaling**
+
+The final observations follow Poisson distributions with rates determined by the latent concentrations:
+
+$$\begin{align}
+u_{ij} | \theta &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}(\theta)) \\
+s_{ij} | \theta &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij}(\theta))
+\end{align}$$
+
+This scaling structure ensures that parameter uncertainty propagates through the characteristic scales $U_{0i}$ and capture efficiencies $\lambda_j$ to the final count observations.
+
+#### **Validation Procedure: Mathematical Description**
+
+**Step 1: Synthetic Data Generation from Prior Predictive**
+
+Generate synthetic data with known "true" parameters by sampling from the prior predictive distribution:
+
+1. Sample true parameters: $\theta_{true} \sim p(\theta)$
+2. Compute latent concentrations: $u^*_{ij,true}, s^*_{ij,true} = f_{\text{piecewise}}(t^*_{j,true}, \theta_{true})$
+3. Generate synthetic observations: $y_{obs} = \{u_{ij}, s_{ij}\} \sim p(y | \theta_{true})$
+
+**Step 2: Posterior Inference**
+
+Fit the model to synthetic data to obtain the posterior distribution:
+
+$$p(\theta | y_{obs}) \propto p(y_{obs} | \theta) p(\theta)$$
+
+This step uses MCMC or variational inference to approximate the posterior distribution over all model parameters.
+
+**Step 3: Posterior Sampling**
+
+Extract $S$ posterior samples from the fitted model:
+
+$$\{\theta^{(s)}\}_{s=1}^S \sim p(\theta | y_{obs})$$
+
+Each sample $\theta^{(s)}$ represents a complete parameter configuration consistent with the observed synthetic data.
+
+**Step 4: Posterior Predictive Sampling**
+
+For each posterior sample $\theta^{(s)}$, generate a corresponding posterior predictive sample:
+
+$$\tilde{y}^{(s)} \sim p(y | \theta^{(s)})$$
+
+**Mathematical Prohibition Against Parameter Averaging**: The posterior predictive distribution is **NOT** computed using averaged parameters:
+
+$$\bar{\theta} = \frac{1}{S}\sum_{s=1}^S \theta^{(s)} \quad \text{(MATHEMATICALLY INCORRECT)}$$
+
+Using $\bar{\theta}$ violates the fundamental definition of posterior predictive distributions and fails to capture parameter uncertainty propagation through the nonlinear piecewise dynamics.
+
+**Correct Sample-wise Generation**: Each $\tilde{y}^{(s)}$ must be generated using the complete parameter vector $\theta^{(s)}$:
+
+1. Compute phase assignments for each cell-gene pair using ${t^{\ast}_{j}}^{(s)}$, ${t^{\ast}_{0,\text{on}i}}^{(s)}$, ${\delta^{\ast}_i}^{(s)}$
+2.
+Evaluate analytical solutions: ${u^*_{ij}}^{(s)}, {s^*_{ij}}^{(s)} = f_{\text{piecewise}}({t^*_{j}}^{(s)}, {\gamma^*_i}^{(s)}, {R_{\text{on},i}}^{(s)}, {t^*_{0,\text{on}i}}^{(s)}, {\delta^*_i}^{(s)})$
+3. Sample observations:
+
+\begin{align}
+{u_{ij}}^{(s)} &\sim \text{Poisson}({\lambda_j}^{(s)} \cdot {U_{0i}}^{(s)} \cdot {u^*_{ij}}^{(s)})$\\
+${s_{ij}}^{(s)} &\sim \text{Poisson}({\lambda_j}^{(s)} \cdot {U_{0i}}^{(s)} \cdot {s^*_{ij}}^{(s)})
+\end{align}
+
+**Step 5: Validation Assessment**
+
+Compare posterior predictive samples $\{\tilde{y}^{(s)}\}_{s=1}^S$ to the original synthetic data $y_{obs}$ using multiple validation metrics.
+
+#### **Uncertainty Propagation Through Piecewise Dynamics**
+
+**Parameter Uncertainty Sources**
+
+The piecewise activation model exhibits multiple sources of parameter uncertainty:
+
+1. **Kinetic parameter uncertainty**: Posterior distributions over $\gamma^*_i$ and $R_{\text{on},i}$ affect the shape and magnitude of concentration trajectories
+2. **Temporal parameter uncertainty**: Uncertainty in $\tilde{t}^*_{0,\text{on}i}$ and $\tilde{\delta}^*_i$ affects when activation occurs and how long it lasts
+3. **Hierarchical time uncertainty**: Uncertainty in $T_{M}^*$, $t_{loc}$, $t_{scl}$ affects the global temporal scaling and population-level time distribution
+4. **Cell-specific uncertainty**: Uncertainty in individual $t^*_j$ and $\lambda_j$ affects cell-specific predictions
+5. **Scale uncertainty**: Uncertainty in $U_{0i}$ affects the absolute magnitude of predicted counts
+
+**Nonlinear Uncertainty Propagation**
+
+Parameter uncertainty propagates nonlinearly through the piecewise analytical solutions:
+
+- **Phase transitions**: Small changes in temporal parameters can cause cells to switch between phases, leading to discontinuous changes in predicted concentrations
+- **Exponential sensitivity**: The exponential terms in analytical solutions amplify parameter uncertainty, particularly for large time values
+- **Interaction effects**: Uncertainty in $\gamma^*_i$ interacts with uncertainty in temporal parameters through the $\xi$ helper variables
+- **Special case sensitivity**: Near $\gamma^*_i = 1$, the transition between solution forms creates additional uncertainty
+
+**Total Predictive Uncertainty Decomposition**
+
+For the piecewise activation model, the total predictive uncertainty can be decomposed as:
+
+$$\begin{align}
+\text{Var}[\tilde{y}_{ij} | y_{obs}] &= \mathbb{E}_\theta[\lambda_j U_{0i} u^*_{ij}(\theta)] + \text{Var}_\theta[\lambda_j U_{0i} u^*_{ij}(\theta)] \\
+&= \text{Poisson variance} + \text{Parameter uncertainty}
+\end{align}$$
+
+The parameter uncertainty term captures the complex propagation of uncertainty through the hierarchical parameter structure and piecewise dynamics.
+
+#### **Validation Metrics for Posterior Predictive Checking**
+
+**Parameter Recovery Assessment**
+
+Compare true parameters $\theta_{true}$ to posterior samples $\{\theta^{(s)}\}_{s=1}^S$:
+
+1. **Coverage probability**: Fraction of true parameters falling within posterior credible intervals
+   $$\text{Coverage}(\alpha) = \frac{1}{|\Theta|} \sum_{\theta_k \in \Theta} \mathbb{I}[\theta_{k,true} \in \text{CI}_\alpha(\theta_k | y_{obs})]$$
+   where $\text{CI}_\alpha$ is the $\alpha$-level credible interval.
+
+2. **Posterior mean bias**: Systematic deviation of posterior means from true values
+   $$\text{Bias}(\theta_k) = \mathbb{E}[\theta_k | y_{obs}] - \theta_{k,true}$$
+
+3. **Posterior standard deviation**: Measure of parameter uncertainty
+   $$\text{SD}(\theta_k) = \sqrt{\text{Var}[\theta_k | y_{obs}]}$$
+
+**Predictive Accuracy Assessment**
+
+Compare observed synthetic data $y_{obs}$ to posterior predictive samples $\{\tilde{y}^{(s)}\}_{s=1}^S$:
+
+1. **Mean Absolute Error (MAE)**: Average absolute deviation between observations and posterior predictive means
+   $$\text{MAE} = \frac{1}{GN} \sum_{i=1}^G \sum_{j=1}^N |y_{ij,obs} - \mathbb{E}[\tilde{y}_{ij} | y_{obs}]|$$
+
+2. **Root Mean Square Error (RMSE)**: Penalizes large deviations more heavily
+   $$\text{RMSE} = \sqrt{\frac{1}{GN} \sum_{i=1}^G \sum_{j=1}^N (y_{ij,obs} - \mathbb{E}[\tilde{y}_{ij} | y_{obs}])^2}$$
+
+3. **Posterior predictive p-values**: Probability that replicated data would be more extreme than observed data
+   $$p_{pp} = P(T(\tilde{y}) \geq T(y_{obs}) | y_{obs})$$
+   where $T(\cdot)$ is a test statistic (e.g., mean, variance, skewness).
+
+#### **Critical Implementation Requirements**
+
+**Prohibition Against Parameter Averaging**
+
+The mathematical definition of posterior predictive distributions explicitly requires sample-wise generation. **Parameter averaging is mathematically incorrect** because:
+
+1. **Nonlinear dynamics**: The piecewise analytical solutions are nonlinear functions of parameters, so $f(\mathbb{E}[\theta]) \neq \mathbb{E}[f(\theta)]$
+2. **Phase discontinuities**: Averaged temporal parameters may place cells in incorrect phases
+3. **Uncertainty underestimation**: Averaging eliminates parameter uncertainty propagation
+4. **Invalid inference**: Results do not represent valid Bayesian posterior predictive distributions
+
+**Correct Sample-wise Implementation**
+
+Each posterior predictive sample must be generated independently:
+
+```python
+# CORRECT: Sample-wise generation
+for s in range(S):
+    theta_s = posterior_samples[s]  # Complete parameter vector
+    y_pred_s = generate_observations(theta_s)  # Use full parameter vector
+    posterior_predictive_samples.append(y_pred_s)
+
+# INCORRECT: Parameter averaging (mathematically invalid)
+theta_mean = np.mean(posterior_samples, axis=0)  # WRONG
+y_pred = generate_observations(theta_mean)  # WRONG
+```
+
+**Uncertainty Propagation Requirements**
+
+1. **Preserve parameter correlations**: Use complete parameter vectors $\theta^{(s)}$ to maintain posterior correlations
+2. **Respect phase boundaries**: Ensure temporal parameters correctly determine phase assignments for each sample
+3. **Include all uncertainty sources**: Propagate uncertainty from all hierarchical levels
+4. **Maintain biological constraints**: Ensure all generated samples satisfy model constraints (e.g., positive concentrations)
+
+#### **Computational Considerations for Piecewise Models**
+
+**Analytical Solution Advantages**
+
+The piecewise activation model's analytical solutions provide computational benefits for posterior predictive checking:
+
+1. **Fast likelihood evaluation**: No numerical integration required for each posterior sample
+2. **Exact uncertainty propagation**: Analytical derivatives enable precise uncertainty quantification
+3. **Efficient validation**: Rapid generation of large numbers of posterior predictive samples
+4. **Numerical stability**: Closed-form solutions avoid accumulation of numerical errors
+
+**Phase Assignment Efficiency**
+
+For each posterior sample $\theta^{(s)}$, phase assignments can be computed efficiently:
+
+1. **Vectorized comparisons**: Determine phases for all cell-gene pairs simultaneously
+2. **Cached computations**: Reuse phase-specific analytical solutions across similar parameter values
+3. **Parallel processing**: Generate multiple posterior predictive samples in parallel
+
+**Memory Management**
+
+Large posterior predictive sample sets require careful memory management:
+
+1. **Streaming generation**: Generate and process samples sequentially to avoid memory overflow
+2. **Selective storage**: Store only essential summary statistics rather than full sample arrays
+3. **Compressed representations**: Use efficient data structures for sparse count matrices
+
+#### **Biological Interpretation of Validation Results**
+
+**Successful Parameter Recovery Indicators**
+
+1. **High coverage probabilities** (>90%) indicate the model correctly quantifies parameter uncertainty
+2. **Low posterior bias** indicates the model can accurately estimate true parameter values
+3. **Appropriate posterior uncertainty** balances precision with realistic uncertainty quantification
+
+**Predictive Accuracy Indicators**
+
+1. **Low MAE/RMSE** indicates accurate point predictions
+2. **Posterior predictive p-values near 0.5** indicate good model fit without over- or under-fitting
+3. **Consistent pattern classification** indicates the model captures essential temporal dynamics
+
+**Failure Mode Identification**
+
+1. **Poor coverage** may indicate model misspecification or inadequate prior specification
+2. **High bias** may indicate identifiability issues or optimization problems
+3. **Poor predictive accuracy** may indicate missing biological mechanisms or inappropriate noise models
 
@@
 1. **Recover constrained piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $\tilde{t}^*_{0,\text{on}i}$, $\tilde{\delta}^*_i$
-2. **Handle pre-activation scenarios**: Correctly infer negative relative onset times and their biological implications
-3. **Distinguish expression patterns**: Classify genes across the three core patterns (pre-activation, transient, sustained)
 4. **Maintain hierarchical time inference**: Recover population-level temporal parameters ($T_M^*$, $t_{loc}$, $t_{scl}$)
@@
 11. **Validate relative parameter framework**: Confirm that relative temporal parameters enable consistent pattern classification across different experimental timescales
-12. **Assess pattern coverage**: Verify that the three-pattern framework adequately captures the diversity of gene expression dynamics in real biological systems
 
 ### Parameter Recovery Validation Implementation
-
-The following example demonstrates the user-facing API for performing parameter recovery validation with the piecewise activation model:
-
-```{python}
-#| label: piecewise-activation-validation
-#| code-fold: true
-#| code-summary: "Show/Hide Code"
-#| eval: false
-
-# Create the piecewise activation model
-model = create_piecewise_activation_model()
-
-print(f"Created piecewise activation model: {model}")
-print(f"Model components:")
-print(f"  Dynamics: {model.dynamics_model}")
-print(f"  Prior: {model.prior_model}")
-print(f"  Likelihood: {model.likelihood_model}")
-print(f"  Guide: {model.guide_model}")
-
-# Generate prior predictive data with known true parameters
-print(f"\nGenerating prior predictive data for validation...")
-prior_predictive_adata = model.generate_predictive_samples(
-    num_cells=200,
-    num_genes=100,
-    num_samples=1,
-    return_format="anndata"
-)
-
-print(f"Prior predictive data generated:")
-print(f"  Cells: {prior_predictive_adata.n_obs}")
-print(f"  Genes: {prior_predictive_adata.n_vars}")
-print(f"  Layers: {list(prior_predictive_adata.layers.keys())}")
-
-# Extract true parameters for validation
-true_parameters = prior_predictive_adata.uns.get('true_parameters', {})
-print(f"\nTrue parameters available for validation:")
-for key in sorted(true_parameters.keys()):
-    if key.startswith(('R_on', 'tilde_t_on', 'tilde_delta', 't_on', 'delta', 'gamma_star', 'T_M_star')):
-        param_value = true_parameters[key]
-        if hasattr(param_value, 'shape') and param_value.size > 1:
-            print(f"  {key}: shape {param_value.shape}, range [{param_value.min():.3f}, {param_value.max():.3f}]")
-        else:
-            print(f"  {key}: {param_value:.3f}")
-
-# Analyze pattern distribution in generated data
-print(f"\nPattern analysis of generated genes:")
-if 'tilde_t_on_star' in true_parameters and 'tilde_delta_star' in true_parameters:
-    tilde_t_on = true_parameters['tilde_t_on_star']
-    tilde_delta = true_parameters['tilde_delta_star']
-    R_on = true_parameters['R_on']
-
-    # Count genes in each pattern
-    pre_activation = (tilde_t_on < 0.0) & (R_on > 2.0)
-    transient = (tilde_t_on > 0.0) & (tilde_t_on < 0.5) & (tilde_delta < 0.4) & (R_on > 2.0)
-    sustained = (tilde_t_on > 0.0) & (tilde_t_on < 0.3) & (tilde_delta > 0.5) & (R_on > 2.0)
-
-    print(f"  Pre-activation: {pre_activation.sum()} genes ({pre_activation.mean()*100:.1f}%)")
-    print(f"  Transient: {transient.sum()} genes ({transient.mean()*100:.1f}%)")
-    print(f"  Sustained: {sustained.sum()} genes ({sustained.mean()*100:.1f}%)")
-    print(f"  Other patterns: {(~(pre_activation | transient | sustained)).sum()} genes")
-
-# Train model on prior predictive data (parameter recovery)
-print(f"\nTraining model for parameter recovery...")
-trained_model = model.train(
-    adata=prior_predictive_adata,
-    max_epochs=1000,  # Reduced for demonstration
-    learning_rate=0.01,
-    early_stopping=True,
-    early_stopping_patience=10,
-    use_gpu="auto",
-    seed=42
-)
-
-# Generate posterior samples from trained model
-print(f"\nGenerating posterior samples...")
-posterior_samples = trained_model.generate_posterior_samples(
-    adata=prior_predictive_adata,
-    num_samples=30,
-    seed=42,
-    return_tensors=True
-)
-
-print(f"Posterior samples generated:")
-for key in sorted(posterior_samples.keys()):
-    if key.startswith(('R_on', 'tilde_t_on', 'tilde_delta', 't_on', 'delta', 'gamma_star', 'T_M_star')):
-        param_shape = posterior_samples[key].shape if hasattr(posterior_samples[key], 'shape') else len(posterior_samples[key])
-        print(f"  {key}: {param_shape}")
-
-print(f"\nParameter recovery validation completed!")
-print(f"This demonstrates the complete workflow from:")
-print(f"  1. Model creation with create_piecewise_activation_model()")
-print(f"  2. Prior predictive data generation with known parameters")
-print(f"  3. Model training on synthetic data")
-print(f"  4. Posterior sampling for parameter recovery assessment")
-print(f"  5. Comparison of true vs recovered parameters")
-print(f"  6. Pattern classification analysis with 3-pattern framework")
-```
 

### 2025-06-13 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): distinguish model descriptions from those to which we apply predictive checks
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 abstract: |
-  This notebook demonstrates parameter recovery validation for PyroVelocity models using the direct dimensionless approach,
-  progressing from analytical solutions with observed time to numerical simulation with latent time coordinates.
-  We focus on models that naturally avoid identifiability issues while maintaining biological interpretability,
-  using hierarchical priors on characteristic scales and lumped effective capture efficiency.
+  This notebook demonstrates parameter recovery validation for PyroVelocity models using dimensionless models.
+  We progress from analytically tractable models where the temporal coordinate is an observed variable toward the more complicated situation of models with latent variables that represent temporal ordering and must therefore be inferred.
+  We focus on constructing models that naturally avoid identifiability issues while maintaining biological interpretability,
+  using hierarchical priors on characteristic scales and simple accounting of technical artifacts through effective parameters that aggregate sources of error.
   We illustrate how to systematically validate that probabilistic models can recover
@@
 For each model we include a graphical representation of the model's generative process as well as a mathematical description of the prior, observation model, and likelihood.
-
-## Model Progression
-
-We validate parameter recovery across a progression of **dimensionless models** that naturally avoid identifiability issues while maintaining biological interpretability.
+We illustrate how to perform a somewhat simplified version of prior and posterior predictive checking for the piecewise activation model with latent time.
+
+## Model Development
+
+We work through a progression of **dimensionless models** that naturally avoid or aim to avoid identifiability issues while maintaining biological interpretability.
 Each model builds complexity while preserving the core dimensionless parameterization.
@@
 np.random.seed(42)
-
-print("Setup complete!")
 ```
@@
 
-**Mathematical Prohibition Against Parameter Averaging**: The posterior predictive distribution is **NOT** computed using averaged parameters:
-
-$$\bar{\theta} = \frac{1}{S}\sum_{s=1}^S \theta^{(s)} \quad \text{(MATHEMATICALLY INCORRECT)}$$
-
-Using $\bar{\theta}$ violates the fundamental definition of posterior predictive distributions and fails to capture parameter uncertainty propagation through the nonlinear piecewise dynamics.
-
-**Correct Sample-wise Generation**: Each $\tilde{y}^{(s)}$ must be generated using the complete parameter vector $\theta^{(s)}$:
+**Mathematical Prohibition Against Parameter Averaging**:
+
+**Correct Sampling**: Each $\tilde{y}^{(s)}$ must be generated using the complete parameter vector $\theta^{(s)}$:
 
@@
 \end{align}
+
+The posterior predictive distribution cannot be computed using averaged parameters such as
+
+$$\bar{\theta} = \frac{1}{S}\sum_{s=1}^S \theta^{(s)}$$
+
+Using $\bar{\theta}$ would violate the fundamental definition of posterior predictive distributions and fails to capture parameter uncertainty propagation through the nonlinear dynamics.
 
@@
 
-The mathematical definition of posterior predictive distributions explicitly requires sample-wise generation. **Parameter averaging is mathematically incorrect** because:
-
-1. **Nonlinear dynamics**: The piecewise analytical solutions are nonlinear functions of parameters, so $f(\mathbb{E}[\theta]) \neq \mathbb{E}[f(\theta)]$
-2. **Phase discontinuities**: Averaged temporal parameters may place cells in incorrect phases
-3. **Uncertainty underestimation**: Averaging eliminates parameter uncertainty propagation
-4. **Invalid inference**: Results do not represent valid Bayesian posterior predictive distributions
+The mathematical definition of posterior predictive distributions explicitly requires sampling parameters and observables jointly, without ad hoc averaging.
 

### 2025-06-13 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): remove unused code block
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-## Parameter Recovery Validation Framework
-
-```{python}
-#| label: setup-imports
-#| code-fold: true
-
-import numpy as np
-import torch
-import pyro
-import matplotlib.pyplot as plt
-import seaborn as sns
-from pathlib import Path
-
-# PyroVelocity imports
-from pyrovelocity.models.modular.factory import (
-    create_piecewise_activation_model,
-    create_legacy_model2,
-)
-# from pyrovelocity.models.modular.components.validation import (
-#     ParameterRecoveryConfig,
-# )
-from pyrovelocity.io.datasets import bifurcation_14
-
-# Set random seeds for reproducibility
-torch.manual_seed(42)
-pyro.set_rng_seed(42)
-np.random.seed(42)
-```
-
 ## Dimensionless analytical dynamics with observed time {#sec-dimless-observed-time}

### 2025-06-13 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): explain prior predictive sampling
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-$$\begin{align}
+\begin{align}
 \frac{du^{\ast}}{dt^{\ast}} &= \alpha^*_i(t^*) - u^{\ast} \\
 \frac{ds^{\ast}}{dt^{\ast}} &= u^{\ast} - \gamma^{\ast}_i s^{\ast}
-\end{align}$$
+\end{align}
 
@@
 
-$$\begin{align}
+\begin{align}
 u^*_{ij}(t^*) &= 1.0 \\
 s^*_{ij}(t^*) &= \frac{1.0}{\gamma^*_i}
-\end{align}$$
+\end{align}
 
@@
 For $\gamma^*_i \neq 1$:
-$$\begin{align}
+\begin{align}
 u^*_{ij}(\tau_{\text{on}}) &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\tau_{\text{on}}} \\
 s^*_{ij}(\tau_{\text{on}}) &= \frac{R_{\text{on},i}}{\gamma^*_i} + \left(\frac{1.0}{\gamma^*_i} - \xi_{\text{on}i} - \frac{R_{\text{on},i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{on}}} + \xi_{\text{on}i} e^{-\tau_{\text{on}}}
-\end{align}$$
+\end{align}
 
@@
 For $\gamma^*_i = 1$:
-$$\begin{align}
+\begin{align}
 u^*_{ij}(\tau_{\text{on}}) &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\tau_{\text{on}}} \\
 s^*_{ij}(\tau_{\text{on}}) &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\tau_{\text{on}}} + (1.0 - R_{\text{on},i}) \tau_{\text{on}} e^{-\tau_{\text{on}}}
-\end{align}$$
+\end{align}
 
@@
 For $\gamma^*_i \neq 1$:
-$$\begin{align}
+\begin{align}
 u^*_{\text{off},0i} &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\delta^*_i} \\
 s^*_{\text{off},0i} &= \frac{R_{\text{on},i}}{\gamma^*_i} + \left(\frac{1.0}{\gamma^*_i} - \xi_{\text{on}i} - \frac{R_{\text{on},i}}{\gamma^*_i}\right) e^{-\gamma^*_i \delta^*_i} + \xi_{\text{on}i} e^{-\delta^*_i}
-\end{align}$$
+\end{align}
 
 For $\gamma^*_i = 1$:
-$$\begin{align}
+\begin{align}
 u^*_{\text{off},0i} &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\delta^*_i} \\
 s^*_{\text{off},0i} &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\delta^*_i} + (1.0 - R_{\text{on},i}) \delta^*_i e^{-\delta^*_i}
-\end{align}$$
+\end{align}
 
@@
 For $\gamma^*_i \neq 1$:
-$$\begin{align}
+\begin{align}
 u^*_{ij}(\tau_{\text{off}}) &= 1.0 + (u^*_{\text{off},0i} - 1.0) e^{-\tau_{\text{off}}} \\
 s^*_{ij}(\tau_{\text{off}}) &= \frac{1.0}{\gamma^*_i} + \left(s^*_{\text{off},0i} - \xi_{\text{off}i} - \frac{1.0}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{off}}} + \xi_{\text{off}i} e^{-\tau_{\text{off}}}
-\end{align}$$
+\end{align}
 
@@
 For $\gamma^*_i = 1$:
-$$\begin{align}
+\begin{align}
 u^*_{ij}(\tau_{\text{off}}) &= 1.0 + (u^*_{\text{off},0i} - 1.0) e^{-\tau_{\text{off}}} \\
 s^*_{ij}(\tau_{\text{off}}) &= 1.0 + (s^*_{\text{off},0i} - 1.0) e^{-\tau_{\text{off}}} + (u^*_{\text{off},0i} - 1.0) \tau_{\text{off}} e^{-\tau_{\text{off}}}
-\end{align}$$
+\end{align}
 
@@
 
-$$\begin{align}
+\begin{align}
 t^*_{0,\text{on}i} &= T_M^* \times \tilde{t}^*_{0,\text{on}i} \quad \text{(absolute activation onset time)} \\
 \delta^*_i &= T_M^* \times \tilde{\delta}^*_i \quad \text{(absolute activation duration)}
-\end{align}$$
+\end{align}
 
@@
 
+#### **Prior Predictive Distribution: Mathematical Foundation**
+
+**Fundamental Definition and Interpretation**
+
+The prior predictive distribution represents the model's predictions about observable data before conditioning on any observations, providing a complete characterization of the model's predictive uncertainty in the absence of data.
+This marginal distribution integrates over all possible parameter configurations weighted by their prior probabilities.
+
+For the piecewise activation model, this becomes:
+
+\begin{align}
+p(\tilde{y}) &= \int p(\tilde{y} | \gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i}, T_{M}^*, t_{loc}, t_{scl}, t^*_j, \lambda_j) \times \nonumber \\
+&\quad p(\gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i}, T_{M}^*, t_{loc}, t_{scl}, t^*_j, \lambda_j) \, d\theta \label{eq-prior-predictive-piecewise}
+\end{align}
+
+where the integration spans the complete hierarchical parameter space defined by the model's prior distributions.
+
+**Hierarchical Integration Decomposition**
+
+The prior predictive distribution decomposes according to the model's hierarchical structure:
+
+\begin{align}
+p(\tilde{y}) &= \int \int \int \int \int p(\tilde{y} | T_{M}^*, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
+&\quad p(T_{M}^*) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{M}^* \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-hierarchical-integration}
+\end{align}
+
+where:
+
+- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
+- $\theta_{\text{cell}} = \{t_{loc}, t_{scl}, \tilde{t}_j, t^*_j\}_{j=1}^N$ represents cell-specific temporal parameters
+- $\theta_{\text{tech}} = \{\lambda_j\}_{j=1}^N$ represents technical efficiency parameters
+
+This decomposition reflects the model's assumption that gene-specific kinetics, population-level temporal structure, and technical factors contribute independently to the prior uncertainty.
+
+**Monte Carlo Approximation**
+
+The prior predictive distribution is computed via Monte Carlo sampling:
+
+\begin{align}
+p(\tilde{y}) \approx \frac{1}{S} \sum_{s=1}^S p(\tilde{y} | \theta^{(s)}) \label{eq-mc-prior-predictive}
+\end{align}
+
+where $\theta^{(s)} \sim p(\theta)$ are samples from the joint prior distribution.
+
+**Detailed Sampling Procedure:**
+
+1. **Sample hierarchical time parameters**:
+   \begin{align}
+   {T_{M}^*}^{(s)} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}), \quad {t_{loc}}^{(s)} \sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}), \quad {t_{scl}}^{(s)} \sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}})
+   \end{align}
+
+2. **Sample cell-specific relative times**:
+   \begin{align}
+   {\tilde{t}_j}^{(s)} &\sim \text{Normal}({t_{loc}}^{(s)}, ({t_{scl}}^{(s)})^2) \quad \text{for } j \in \{1, \ldots, N\}
+   \end{align}
+
+3. **Transform to absolute times**:
+   \begin{align}
+   {t^{*}_j}^{(s)} &= {T_{M}^*}^{(s)} \times \max({\tilde{t}_j}^{(s)}, \epsilon)
+   \end{align}
+
+4. **Sample gene-specific parameters**:
+   \begin{align}
+   {\gamma^{*}_i}^{(s)} &\sim \text{LogNormal}(\log(1.0), 0.5^2) \\
+   {R_{\text{on},i}}^{(s)} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \\
+   {\tilde{t}^*_{0,\text{on}i}}^{(s)} &\sim \text{Normal}(0.5, 0.8^2) \\
+   {\tilde{\delta}^*_i}^{(s)} &\sim \text{LogNormal}(\log(0.45), 0.45^2) \\
+   {U_{0i}}^{(s)} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
+   \end{align}
+
+5. **Transform relative temporal parameters**:
+   \begin{align}
+   {t^*_{0,\text{on}i}}^{(s)} &= {T_{M}^*}^{(s)} \times {\tilde{t}^*_{0,\text{on}i}}^{(s)}, \quad {\delta^*_i}^{(s)} = {T_{M}^*}^{(s)} \times {\tilde{\delta}^*_i}^{(s)}
+   \end{align}
+
+6. **Sample capture efficiencies**:
+   \begin{align}
+   {\lambda_j}^{(s)} &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
+   \end{align}
+
+7. **Compute latent concentrations**:
+   \begin{align}
+   {u^{*}_{ij}}^{(s)}, {s^{*}_{ij}}^{(s)} &= f_{\text{piecewise}}({t^*_j}^{(s)}, {\gamma^*_i}^{(s)}, {R_{\text{on},i}}^{(s)}, {t^*_{0,\text{on}i}}^{(s)}, {\delta^*_i}^{(s)})
+   \end{align}
+
+8. **Sample observations**:
+   \begin{align}
+   {u_{ij}}^{(s)} &\sim \text{Poisson}({\lambda_j}^{(s)} \cdot {U_{0i}}^{(s)} \cdot {u^{*}_{ij}}^{(s)}) \\
+   {s_{ij}}^{(s)} &\sim \text{Poisson}({\lambda_j}^{(s)} \cdot {U_{0i}}^{(s)} \cdot {s^{*}_{ij}}^{(s)})
+   \end{align}
+
+**Prior Predictive Uncertainty Decomposition**
+
+The total prior predictive uncertainty decomposes across the hierarchical parameter structure:
+
+\begin{align}
+\text{Var}[\tilde{y}] = \mathbb{E}_\theta[\text{Var}[y|\theta]] + \text{Var}_\theta[\mathbb{E}[y|\theta]] \label{eq-prior-uncertainty-decomp}
+\end{align}
+
+where the expectation and variance are taken over the joint prior distribution $p(\theta)$.
+
+**Uncertainty Sources in Prior Predictive:**
+
+The piecewise activation model exhibits multiple sources of prior uncertainty that propagate through the hierarchical structure:
+
+1. **Global temporal scale uncertainty**: Uncertainty in $T_{M}^*$ affects the absolute timescale of all temporal processes, scaling both activation onset times and durations proportionally
+
+2.
+**Population temporal distribution uncertainty**: Uncertainty in $t_{loc}, t_{scl}$ affects the distribution of cells across the relative temporal coordinate, determining population-level temporal heterogeneity
+
+3. **Gene-specific kinetic uncertainty**: Prior uncertainty in $\gamma^*_i, R_{\text{on},i}$ affects the shape and magnitude of concentration trajectories during each phase
+
+4.
+**Gene-specific temporal pattern uncertainty**: Prior uncertainty in $\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i$ affects when activation occurs and how long it lasts relative to the population timeline
+
+5. **Characteristic scale uncertainty**: Prior uncertainty in $U_{0i}$ affects the absolute magnitude of expression levels for each gene
+
+6. **Technical efficiency uncertainty**: Prior uncertainty in $\lambda_j$ affects the detection efficiency and overall count levels for each cell
+
+7. **Observation noise**: Irreducible Poisson variance in count observations that persists regardless of parameter knowledge
+
+**Mathematical Decomposition:**
+
+For individual observations, the prior predictive variance decomposes as:
+
+\begin{align}
+\text{Var}[\tilde{y}_{ij}] = \mathbb{E}[\lambda_j U_{0i} u^*_{ij}(\theta)] + \text{Var}[\lambda_j U_{0i} u^*_{ij}(\theta)] \label{eq-individual-prior-variance}
+\end{align}
+
+where the first term represents the expected Poisson variance (observation noise) and the second term represents the variance due to parameter uncertainty propagating through the piecewise dynamics.
+
+**Nonlinear Uncertainty Propagation in Prior Predictive:**
+
+Parameter uncertainty propagates nonlinearly through the piecewise analytical solutions under the prior:
+
+- **Phase assignment uncertainty**: Prior uncertainty in temporal parameters creates uncertainty about which phase each cell-gene pair occupies, leading to mixture-like behavior in the prior predictive distribution
+
+- **Exponential amplification**: The exponential terms in analytical solutions amplify prior parameter uncertainty, particularly for genes with large characteristic scales or cells at extreme time points
+
+- **Temporal scaling interactions**: Uncertainty in $T_{M}^*$ interacts multiplicatively with uncertainty in relative temporal parameters, creating complex correlation structures
+
+- **Special case transitions**: Prior probability mass near $\gamma^*_i = 1$ creates discontinuous changes in the analytical solution form, contributing additional uncertainty
+
+**Relationship to Posterior Predictive**
+
+**Bayesian Learning Interpretation:**
+
+The posterior predictive distribution represents the prior predictive distribution updated by observed data:
+
+\begin{align}
+p(\tilde{y} | y_{obs}) = \int p(\tilde{y} | \theta) p(\theta | y_{obs}) d\theta \label{eq-posterior-predictive-relation}
+\end{align}
+
+where $p(\theta | y_{obs}) \propto p(y_{obs} | \theta) p(\theta)$ incorporates the likelihood of observed data through Bayes' theorem.
+
+**Uncertainty Reduction:**
+
+Bayesian learning systematically reduces predictive uncertainty:
+
+\begin{align}
+\text{Var}[\tilde{y} | y_{obs}] \leq \text{Var}[\tilde{y}] \label{eq-uncertainty-reduction}
+\end{align}
+
+with equality only when the observed data provides no information about the parameters (i.e., when the likelihood is flat).
+
+**Information Content Quantification:**
+
+The reduction in predictive uncertainty quantifies the information content of the observed data:
+
+\begin{align}
+\text{Information} = \text{Var}[\tilde{y}] - \text{Var}[\tilde{y} | y_{obs}] \label{eq-information-content}
+\end{align}
+
+This quantity measures how much the data constrains the model's predictions about future observations.
+
+**Prior Predictive Validation Applications**
+
+**Model Specification Validation:**
+
+Prior predictive checks validate whether the model can generate data with realistic characteristics:
+
+1. **Count distribution realism**: Do prior predictive counts match expected single-cell RNA-seq distributional properties (mean-variance relationships, sparsity patterns, dynamic range)?
+
+2. **Temporal coordination**: Does the hierarchical time structure produce biologically plausible cell orderings and temporal relationships?
+
+3. **Expression magnitude realism**: Do the characteristic scales $U_{0i}$ and capture efficiencies $\lambda_j$ combine to produce realistic expression levels?
+
+4. **Kinetic parameter coverage**: Does the prior cover the range of biologically plausible kinetic behaviors without generating unrealistic dynamics?
+
+**Prior Sensitivity Analysis:**
+
+Compare prior predictive distributions under different hyperparameter choices:
+
+\begin{align}
+p(\tilde{y} | \alpha_1) \text{ vs. } p(\tilde{y} | \alpha_2) \label{eq-prior-sensitivity}
+\end{align}
+
+where $\alpha_1, \alpha_2$ represent different hyperparameter configurations. This analysis identifies which hyperparameters most strongly influence prior predictions.
+
+**Computational Validation:**
+
+Prior predictive sampling validates computational implementation:
+
+1. **Convergence assessment**: Do prior predictive samples converge to stable distributions as the number of samples increases?
+
+2. **Numerical stability**: Are analytical solutions numerically stable across the full range of prior parameter values?
+
+3. **Phase assignment consistency**: Do temporal parameters correctly determine phase assignments across all prior samples?
+
+4. **Parameter constraint satisfaction**: Do all prior samples satisfy model constraints (e.g., positive concentrations, valid phase boundaries)?
+
 **Posterior Predictive Distribution**
@@
 
-$$\begin{align}
+\begin{align}
 u_{ij} | \theta &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}(\theta)) \\
 s_{ij} | \theta &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij}(\theta))
-\end{align}$$
+\end{align}
 
@@
 
+The parameter recovery validation procedure systematically compares prior and posterior predictive distributions to assess the model's ability to learn from data.
+
 **Step 1: Synthetic Data Generation from Prior Predictive**
 
-Generate synthetic data with known "true" parameters by sampling from the prior predictive distribution:
-
-1. Sample true parameters: $\theta_{true} \sim p(\theta)$
-2. Compute latent concentrations: $u^*_{ij,true}, s^*_{ij,true} = f_{\text{piecewise}}(t^*_{j,true}, \theta_{true})$
-3. Generate synthetic observations: $y_{obs} = \{u_{ij}, s_{ij}\} \sim p(y | \theta_{true})$
+Generate synthetic data with known "true" parameters by sampling from the prior predictive distribution as described in equations \eqref{eq-prior-predictive-piecewise}-\eqref{eq-mc-prior-predictive}:
+
+1. **Sample true parameters**: $\theta_{true} \sim p(\theta)$ following the hierarchical sampling procedure
+2. **Compute latent concentrations**: $u^*_{ij,true}, s^*_{ij,true} = f_{\text{piecewise}}(t^*_{j,true}, \theta_{true})$ using the piecewise analytical solutions
+3. **Generate synthetic observations**: $y_{obs} = \{u_{ij}, s_{ij}\} \sim p(y | \theta_{true})$ from the Poisson observation model
+
+This step creates a controlled validation scenario where the "true" parameter values are known by construction, enabling direct assessment of parameter recovery performance.
 
@@
 
-1. Compute phase assignments for each cell-gene pair using ${t^{\ast}_{j}}^{(s)}$, ${t^{\ast}_{0,\text{on}i}}^{(s)}$, ${\delta^{\ast}_i}^{(s)}$
+1. Compute phase assignments for each cell-gene pair using ${{t^{\ast}_{j}}}^{(s)}$, ${{t^{\ast}_{0,\text{on}i}}}^{(s)}$, ${{\delta^{\ast}_i}}^{(s)}$
 2.
-Evaluate analytical solutions: ${u^*_{ij}}^{(s)}, {s^*_{ij}}^{(s)} = f_{\text{piecewise}}({t^*_{j}}^{(s)}, {\gamma^*_i}^{(s)}, {R_{\text{on},i}}^{(s)}, {t^*_{0,\text{on}i}}^{(s)}, {\delta^*_i}^{(s)})$
+Evaluate analytical solutions: ${{u^*_{ij}}}^{(s)}, {{s^*_{ij}}}^{(s)} = f_{\text{piecewise}}({{t^*_{j}}}^{(s)}, {{\gamma^*_i}}^{(s)}, {{R_{\text{on},i}}}^{(s)}, {{t^*_{0,\text{on}i}}}^{(s)}, {{\delta^*_i}}^{(s)})$
 3. Sample observations:
@@
 \begin{align}
-{u_{ij}}^{(s)} &\sim \text{Poisson}({\lambda_j}^{(s)} \cdot {U_{0i}}^{(s)} \cdot {u^*_{ij}}^{(s)})$\\
-${s_{ij}}^{(s)} &\sim \text{Poisson}({\lambda_j}^{(s)} \cdot {U_{0i}}^{(s)} \cdot {s^*_{ij}}^{(s)})
+{{u_{ij}}}^{(s)} &\sim \text{Poisson}({{\lambda_j}}^{(s)} \cdot {{U_{0i}}}^{(s)} \cdot {{u^*_{ij}}}^{(s)}) \\
+{{s_{ij}}}^{(s)} &\sim \text{Poisson}({{\lambda_j}}^{(s)} \cdot {{U_{0i}}}^{(s)} \cdot {{s^*_{ij}}}^{(s)})
 \end{align}
@@
 
-$$\bar{\theta} = \frac{1}{S}\sum_{s=1}^S \theta^{(s)}$$
+\begin{align}
+\bar{\theta} = \frac{1}{S}\sum_{s=1}^S {\theta}^{(s)}
+\end{align}
 
@@
 
-$$\begin{align}
+\begin{align}
 \text{Var}[\tilde{y}_{ij} | y_{obs}] &= \mathbb{E}_\theta[\lambda_j U_{0i} u^*_{ij}(\theta)] + \text{Var}_\theta[\lambda_j U_{0i} u^*_{ij}(\theta)] \\
 &= \text{Poisson variance} + \text{Parameter uncertainty}
-\end{align}$$
+\end{align}
 
@@
    where $T(\cdot)$ is a test statistic (e.g., mean, variance, skewness).
-
-#### **Critical Implementation Requirements**
-
-**Prohibition Against Parameter Averaging**
-
-The mathematical definition of posterior predictive distributions explicitly requires sampling parameters and observables jointly, without ad hoc averaging.
-
-**Correct Sample-wise Implementation**
-
-Each posterior predictive sample must be generated independently:
-
-```python
-# CORRECT: Sample-wise generation
-for s in range(S):
-    theta_s = posterior_samples[s]  # Complete parameter vector
-    y_pred_s = generate_observations(theta_s)  # Use full parameter vector
-    posterior_predictive_samples.append(y_pred_s)
-
-# INCORRECT: Parameter averaging (mathematically invalid)
-theta_mean = np.mean(posterior_samples, axis=0)  # WRONG
-y_pred = generate_observations(theta_mean)  # WRONG
-```
-
-**Uncertainty Propagation Requirements**
-
-1. **Preserve parameter correlations**: Use complete parameter vectors $\theta^{(s)}$ to maintain posterior correlations
-2. **Respect phase boundaries**: Ensure temporal parameters correctly determine phase assignments for each sample
-3. **Include all uncertainty sources**: Propagate uncertainty from all hierarchical levels
-4. **Maintain biological constraints**: Ensure all generated samples satisfy model constraints (e.g., positive concentrations)
 
@@
 - **Biological interpretability**: All parameters have clear biological meaning and interpretable ranges
-- **Coarse-grained expression pattern classification**: Classification of temporal expression patterns covering pre-activation, transient, and sustained that should all be able to be generated by the prior and prior predictive in order to ensure the ability to represent the expected collection of expression patterns in the underlying data
+- **Comprehensive prior predictive framework**: Mathematical foundation for validating model specification through prior predictive checking as detailed in equations \eqref{eq-prior-predictive-piecewise}-\eqref{eq-information-content}, ensuring the model can generate realistic data patterns before observing any data
 - **Relative temporal parameterization**: Enables interpretable parameter values and consistent pattern classification across different experimental timescales and even distinct data sets
@@
 The complete validation scripts referenced in the introduction (`prior-predictive-check.py` and `posterior-predictive-check.py`) provide comprehensive parameter recovery validation including visualization and statistical assessment of recovery performance.
+The prior predictive checking framework validates model specification and prior adequacy, while the posterior predictive checking framework assesses parameter recovery and predictive accuracy.

### 2025-06-13 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): adopt process duration with stick-breaking temporal prior
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 The key innovation is a **dimensionally consistent parameterization** that eliminates parameter redundancy by fixing the reference transcription state, while maintaining biological interpretability through fold-change parameters.
+Additionally, this study implements a **process duration with stick-breaking** temporal framework that eliminates scaling symmetry and achieves parameter identifiability while preserving all analytical and computational advantages.
 
@@
 
+### Process Duration with Stick-Breaking Framework
+
+This study implements a **process duration with stick-breaking** approach that eliminates the scaling symmetry present in hierarchical temporal scaling while maintaining all analytical and biological advantages.
+The key innovation is the use of **fixed anchor points** that break scaling symmetry and achieve parameter identifiability.
+
+#### **Elimination of Scaling Symmetry**
+
+The previous hierarchical scaling approach suffered from fundamental **scaling symmetry** that prevented parameter identifiability:
+
+$$\mathcal{L}(T_M^*, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) = \mathcal{L}(\lambda T_M^*, \{\lambda t^*_j\}, \{\lambda t^*_{0,\text{on}i}\}, \{\lambda \delta^*_i\})$$
+
+for any $\lambda > 0$. This symmetry creates non-identifiable parameters and leads to catastrophic validation failures.
+
+#### **Fixed Anchor Point Solution**
+
+The stick-breaking approach eliminates scaling symmetry through **fixed anchor points**:
+
+\begin{align}
+t^*_1 &= 0 \quad \text{(process start, fixed)} \\
+t^*_N &= T_{\max} \quad \text{(process end, sampled)} \\
+t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{\text{on}}, \sigma_{\text{on}}^2) \quad \text{(independent gene timing)} \\
+\delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(independent gene duration)}
+\end{align}
+
+**Identifiability proof**: For any $\lambda \neq 1$:
+$$\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$$
+
+The fixed anchor $t^*_1 = 0$ cannot be scaled, and gene parameters are independent of $T_{\max}$, breaking the symmetry.
+
 ### Gene Expression Pattern Classification
 
-This piecewise framework enables representation of diverse gene expression patterns through specific parameter combinations.
-Each pattern is defined by quantitative parameter ranges applied to **relative temporal parameters** that create distinct temporal dynamics when scaled by the global time parameter $T_M^*$.
-
-**Parameter Framework**: The classification uses relative temporal parameters ($\tilde{t}^*_{0,\text{on}i}$, $\tilde{\delta}^*_i$) that are scaled by the global maximum time $T_M^*$ to produce absolute temporal parameters:
-
-\begin{align}
-t^*_{0,\text{on}i} &= T_M^* \times \tilde{t}^*_{0,\text{on}i} \quad \text{(absolute activation onset time)} \\
-\delta^*_i &= T_M^* \times \tilde{\delta}^*_i \quad \text{(absolute activation duration)}
-\end{align}
-
-This hierarchical structure separates global temporal scaling from gene-specific temporal patterns, enabling consistent pattern classification across different experimental timescales.
+This framework enables representation of diverse gene expression patterns through specific parameter combinations.
+Each pattern is defined by quantitative parameter ranges applied to **absolute temporal parameters** within the process duration framework.
+
+**Parameter Framework**: The classification uses absolute temporal parameters that are independent of the process duration $T_{\max}$, enabling robust pattern identification across different experimental timescales.
 
@@
 
-- $\tilde{t}^*_{0,\text{on}i} < 0.0$ (relative activation onset before observation window)
+- $t^*_{0,\text{on}i} < 0.0$ (activation onset before process start)
 - $R_{\text{on},i} > 2.0$ (moderate to strong fold-change during activation)
 
-**Biological interpretation**: These genes underwent activation before the experimental observation window.
+**Biological interpretation**: These genes underwent activation before the biological process under study began ($t^* = 0$).
 During the observed period, they exhibit decay-only dynamics as they return toward baseline expression levels.
-The negative relative onset time indicates that activation occurred in the "past" relative to the observation start.
+The negative onset time indicates that activation occurred before the process boundary.
 
@@
 
-**Characteristics**: Genes that show complete activation-decay cycles within the observation window, exhibiting brief, pulse-like activation followed by return toward baseline.
+**Characteristics**: Genes that show complete activation-decay cycles within the process duration, exhibiting brief, pulse-like activation followed by return toward baseline.
 
@@
 
-- $\tilde{t}^*_{0,\text{on}i} > 0.0$ (relative activation within observation window)
-- $\tilde{t}^*_{0,\text{on}i} < 0.5$ (activation early enough to complete cycle within 50% of timeline)
-- $\tilde{\delta}^*_i < 0.4$ (short activation duration, 40% of timeline or less)
+- $t^*_{0,\text{on}i} > 0.0$ (activation within process duration)
+- $t^*_{0,\text{on}i} < 2.5$ (activation early enough to complete cycle within process)
+- $\delta^*_i < 2.0$ (short activation duration relative to process)
 - $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)
 
-**Biological interpretation**: These genes activate early enough and for a short enough duration that the complete activation-decay cycle is observable within the experimental window.
-The constraints ensure that both the activation and subsequent decay phases are captured.
+**Biological interpretation**: These genes activate early enough and for a short enough duration that the complete activation-decay cycle is observable within the process duration.
+The constraints ensure that both the activation and subsequent decay phases are captured within the temporal boundaries.
 
@@
 
-**Characteristics**: Genes that become activated early and remain highly expressed throughout most of the observation period, showing net increase over the observation window.
+**Characteristics**: Genes that become activated early and remain highly expressed throughout most of the process duration, showing net increase over the process timeline.
 
@@
 
-- $\tilde{t}^*_{0,\text{on}i} > 0.0$ (relative activation within observation window)
-- $\tilde{t}^*_{0,\text{on}i} < 0.3$ (early activation onset within 30% of timeline)
-- $\tilde{\delta}^*_i > 0.5$ (long activation duration, more than 50% of timeline)
+- $t^*_{0,\text{on}i} > 0.0$ (activation within process duration)
+- $t^*_{0,\text{on}i} < 1.5$ (early activation onset within process)
+- $\delta^*_i > 2.5$ (long activation duration relative to process)
 - $R_{\text{on},i} > 2.0$ (strong fold-change during activation)
 
-**Biological interpretation**: These genes activate early in the observation window and maintain high expression levels throughout most of the observed period.
-The long activation duration relative to the observation window means that decay back to baseline is either minimal or occurs beyond the observation period.
+**Biological interpretation**: These genes activate early in the process and maintain high expression levels throughout most of the process duration.
+The long activation duration means that decay back to baseline is either minimal or occurs beyond the process endpoint.
 
@@
 
-The Normal prior for onset times enables modeling of **pre-activation scenarios** where genes are already activated when observation begins.
+The Normal prior for onset times enables modeling of **pre-activation scenarios** where genes are already activated when the biological process begins.
 This extension is mathematically valid and biologically realistic:
@@
 1.
-**Phase determination remains consistent**: For any observed time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$
+**Phase determination remains consistent**: For any process time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$
 
@@
 
-3. **Biological interpretation is clear**: Negative onset simply means activation occurred before the observation window
+3. **Biological interpretation is clear**: Negative onset simply means activation occurred before the process start ($t^* = 0$)
 
@@
 
-- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no observed cells will be in Phase 1
+- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no cells will be in Phase 1 during the process
 - **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
-- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the observation window
+- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the process duration
 
@@
 
-# Prior hyperparameters (updated for relative temporal parameterization)
+# Prior hyperparameters (updated for absolute temporal parameterization)
 priors = {
     'R_on': {'loc': 0.916, 'scale': 0.4},              # log(2.5), fold-change (LogNormal)
-    'tilde_t_on_star': {'loc': 0.5, 'scale': 0.8},     # Normal(0.5, 0.8²), relative onset (allows negatives)
-    'tilde_delta_star': {'loc': -0.8, 'scale': 0.45},  # log(0.45), relative duration (LogNormal)
+    't_on_star': {'loc': 2.5, 'scale': 1.5},           # Normal(2.5, 1.5²), absolute onset (allows negatives)
+    'delta_star': {'loc': 0.0, 'scale': 0.5},          # log(1.0), absolute duration (LogNormal)
 }
 
-# Calculate constraint feasibility for each pattern (updated to match script implementation)
+# Calculate constraint feasibility for each pattern (updated for absolute parameterization)
 patterns = {
     'pre_activation': [
-        ('tilde_t_on_star', '<', 0.0),  # Activation before observation window
+        ('t_on_star', '<', 0.0),        # Activation before process start
         ('R_on', '>', 2.0),             # Moderate to strong fold change
@@
     'transient': [
-        ('tilde_t_on_star', '>', 0.0),  # Activation within observation window
-        ('tilde_t_on_star', '<', 0.5),  # Early enough to complete cycle (50% of timeline)
-        ('tilde_delta_star', '<', 0.4), # Short pulse duration (40% of timeline)
+        ('t_on_star', '>', 0.0),        # Activation within process duration
+        ('t_on_star', '<', 2.5),        # Early enough to complete cycle
+        ('delta_star', '<', 2.0),       # Short pulse duration
         ('R_on', '>', 2.0),             # Sufficient fold change to observe
@@
     'sustained': [
-        ('tilde_t_on_star', '>', 0.0),  # Activation within observation window
-        ('tilde_t_on_star', '<', 0.3),  # Early activation onset (30% of timeline)
-        ('tilde_delta_star', '>', 0.5), # Long activation duration (50% of timeline)
+        ('t_on_star', '>', 0.0),        # Activation within process duration
+        ('t_on_star', '<', 1.5),        # Early activation onset
+        ('delta_star', '>', 2.5),       # Long activation duration
         ('R_on', '>', 2.0),             # Strong fold change
@@
 
-        if param == 'tilde_t_on_star':  # Normal distribution
+        if param == 't_on_star':  # Normal distribution
             prob = calculate_normal_cdf_probability(
@@
 
-The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while maintaining the hierarchical latent time structure.
-For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of relative temporal gene parameters $(\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i)$.
+The probabilistic graphical model implements the process duration with stick-breaking framework while adding gene-specific piecewise activation parameters.
+For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of absolute temporal gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$.
 Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.
@@
 
-# hyperparameters for hierarchical time structure
+# hyperparameters for process duration
 pgm.add_node("alpha_T", r"$\alpha_{T}$", 5.5, 6, fixed=True)
 pgm.add_node("beta_T", r"$\beta_{T}$", 6, 6, fixed=True)
-pgm.add_node("alpha_t_loc", r"$\alpha_{t_{loc}}$", 6.5, 6, fixed=True)
-pgm.add_node("beta_t_loc", r"$\beta_{t_{loc}}$", 7, 6, fixed=True)
-pgm.add_node("alpha_t_scale", r"$\alpha_{t_{scl}}$", 7.5, 6, fixed=True)
-pgm.add_node("beta_t_scale", r"$\beta_{t_{scl}}$", 8, 6, fixed=True)
 
 # hyperparameters for lambda (entering from right edge of cell plate)
-pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 6.5, 3.5, fixed=True)
-pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 6.5, 2.5, fixed=True)
+pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 7.5, 3.5, fixed=True)
+pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 7.5, 2.5, fixed=True)
 
@@
 
-# hierarchical time parameters (moved further left)
-pgm.add_node("T_max_star", r"$T_{M}^*$", 5.5, 5)
-pgm.add_node("t_loc", r"$t_{loc}$", 6.25, 5)
-pgm.add_node("t_scale", r"$t_{scl}$", 7, 5)
-
-# latent time coordinates and cell-specific parameters (moved further left, lambda enters from right)
+# process duration parameter (replaces hierarchical time scaling)
+pgm.add_node("T_max", r"$T_{\max}$", 5.5, 5)
+
+# stick-breaking cell times and cell-specific parameters
 pgm.add_node("t_star_j", r"$t^{\ast}_j$", 5.5, 4)
-pgm.add_node("lambda_j", r"$\lambda_j$", 5.5, 3)
+pgm.add_node("lambda_j", r"$\lambda_j$", 6.5, 3)
 
@@
 
-# hierarchical time structure edges
-pgm.add_edge("alpha_T", "T_max_star", plot_params=edge_params)
-pgm.add_edge("beta_T", "T_max_star", plot_params=edge_params)
-pgm.add_edge("alpha_t_loc", "t_loc", plot_params=edge_params)
-pgm.add_edge("beta_t_loc", "t_loc", plot_params=edge_params)
-pgm.add_edge("alpha_t_scale", "t_scale", plot_params=edge_params)
-pgm.add_edge("beta_t_scale", "t_scale", plot_params=edge_params)
-
-# time hierarchy edges
-pgm.add_edge("T_max_star", "t_star_j", plot_params=edge_params)
-pgm.add_edge("t_loc", "t_star_j", plot_params=edge_params)
-pgm.add_edge("t_scale", "t_star_j", plot_params=edge_params)
-# temporal scaling edge for gene parameters
-pgm.add_edge("T_max_star", "Delta_i", plot_params=edge_params)
+# process duration edges
+pgm.add_edge("alpha_T", "T_max", plot_params=edge_params)
+pgm.add_edge("beta_T", "T_max", plot_params=edge_params)
+
+# stick-breaking time edges
+pgm.add_edge("T_max", "t_star_j", plot_params=edge_params)
 
@@
 
-The model combines the hierarchical latent time structure from Study 2 with piecewise constant transcription rates and steady-state initial conditions to resolve identifiability while enabling rich gene expression dynamics.
+The model combines the process duration with stick-breaking framework with piecewise constant transcription rates and steady-state initial conditions to achieve parameter identifiability while enabling rich gene expression dynamics.
 
@@
 - $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
-- $\Delta_i$ represents the vector of **relative temporal** gene parameters $(\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i)$
+- $\Delta_i$ represents the vector of **absolute temporal** gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$
 - The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
 - The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
-- The absolute timing parameters $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ and $\delta^*_i = T_{M}^* \times \tilde{\delta}^*_i$ are computed deterministically
+- $T_{\max}$ represents the total process duration with fixed anchor points at $t^*_1 = 0$ and $t^*_N = T_{\max}$
 - $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
-- $T_{M}^* \rightarrow \Delta_i$ represents the scaling relationship for all temporal parameters
-
-**Detailed priors for constrained piecewise activation parameters:**
-
-The priors are designed to encode biologically reasonable parameter ranges while eliminating dimensional redundancy and cleanly separating temporal from non-temporal parameters:
+- Gene temporal parameters are sampled independently of $T_{\max}$, eliminating scaling symmetry
+
+**Detailed priors for process duration with stick-breaking framework:**
+
+The priors are designed to encode biologically reasonable parameter ranges while achieving parameter identifiability through fixed anchor points and independent gene timing:
+
+**Process duration:**
+\begin{align}
+    T_{\max} &\sim \text{Gamma}(5.0, 1.0) \quad \text{(total process duration)}
+\end{align}
 
@@
 
-**Relative temporal gene-specific parameters:**
-\begin{align}
-    \tilde{t}^*_{0,\text{on}i} &\sim \text{Normal}(0.5, 0.8^2) \quad \text{(relative activation onset time, allows negatives)} \\
-    \tilde{\delta}^*_i &\sim \text{LogNormal}(\log(0.45), 0.45^2) \quad \text{(relative activation duration)}
-\end{align}
-
-**Absolute temporal parameters (deterministic transformations):**
-\begin{align}
-    t^*_{0,\text{on}i} &= T_{M}^* \times \tilde{t}^*_{0,\text{on}i} \quad \text{(absolute activation onset time)} \\
-    \delta^*_i &= T_{M}^* \times \tilde{\delta}^*_i \quad \text{(absolute activation duration)}
-\end{align}
-
-**Pattern classification framework**: The three expression patterns are defined by constraints on the relative temporal parameters:
-
-- **Pre-activation**: $\tilde{t}^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
-- **Transient**: $0.0 < \tilde{t}^*_{0,\text{on}i} < 0.5$, $\tilde{\delta}^*_i < 0.4$, and $R_{\text{on},i} > 2.0$
-- **Sustained**: $0.0 < \tilde{t}^*_{0,\text{on}i} < 0.3$, $\tilde{\delta}^*_i > 0.5$, and $R_{\text{on},i} > 2.0$
+**Absolute temporal gene-specific parameters (independent of $T_{\max}$):**
+\begin{align}
+    t^*_{0,\text{on}i} &\sim \text{Normal}(2.5, 1.5^2) \quad \text{(absolute activation onset time)} \\
+    \delta^*_i &\sim \text{LogNormal}(0.0, 0.5^2) \quad \text{(absolute activation duration)}
+\end{align}
+
+**Stick-breaking cell times:**
+\begin{align}
+    t^*_1 &= 0 \quad \text{(process start, fixed)} \\
+    t^*_N &= T_{\max} \quad \text{(process end)} \\
+    \xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
+    t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{(recursive construction)}
+\end{align}
+
+**Pattern classification framework**: The three expression patterns are defined by constraints on the absolute temporal parameters:
+
+- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
+- **Transient**: $0.0 < t^*_{0,\text{on}i} < 2.5$, $\delta^*_i < 2.0$, and $R_{\text{on},i} > 2.0$
+- **Sustained**: $0.0 < t^*_{0,\text{on}i} < 1.5$, $\delta^*_i > 2.5$, and $R_{\text{on},i} > 2.0$
 
@@
 
+**Process duration:**
+
+- **$T_{\max} \sim \text{Gamma}(5.0, 1.0)$**: Centers process duration at 5 dimensionless time units, providing sufficient range for exponential convergence while maintaining computational efficiency.
+The process has clear start ($t^* = 0$) and end ($t^* = T_{\max}$) boundaries.
+
 **Non-temporal parameters:**
@@
 
-**Relative temporal parameters:**
-
-- **$\tilde{t}^*_{0,\text{on}i} \sim \text{Normal}(0.5, 0.8^2)$**: Centers relative activation onset at 50% through the process timeline, with 95% between approximately -1.1 and +2.1.
-**Negative values** enable pre-activation scenarios where genes were already activated before observation began.
-
-- **$\tilde{\delta}^*_i \sim \text{LogNormal}(\log(0.45), 0.45^2)$**: Centers relative activation duration at 45% of the process timeline, with 95% between ~0.18 and ~1.13.
+**Absolute temporal parameters:**
+
+- **$t^*_{0,\text{on}i} \sim \text{Normal}(2.5, 1.5^2)$**: Centers activation onset at the middle of the expected process duration, with 95% between approximately -0.4 and +5.4.
+**Negative values** enable pre-activation scenarios where genes were already activated before the process began.
+
+- **$\delta^*_i \sim \text{LogNormal}(0.0, 0.5^2)$**: Centers activation duration at 1 dimensionless time unit, with 95% between ~0.37 and ~2.7.
 This enables both transient (short duration) and sustained (long duration) activation patterns.
@@
 
-- **Pre-activation** ($\tilde{t}^*_{0,\text{on}i} < 0$): ~26.6% probability under the prior
-- **Transient** ($0 < \tilde{t}^*_{0,\text{on}i} < 0.5$ and $\tilde{\delta}^*_i < 0.4$): Balanced probability for observable complete cycles
-- **Sustained** ($0 < \tilde{t}^*_{0,\text{on}i} < 0.3$ and $\tilde{\delta}^*_i > 0.5$): Balanced probability for long-duration activation
-
-**Temporal scaling transformation:**
-
-- **Absolute onset**: $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ scales relative onset to absolute time
-- **Absolute duration**: $\delta^*_i = T_{M}^* \times \tilde{\delta}^*_i$ scales relative duration to absolute time
-- **Dimensional consistency**: All temporal comparisons now use consistently scaled absolute times
-
-**Parameter interpretation benefits:**
-
-- **Separation of concerns**: Non-temporal kinetics ($\theta_i$) vs temporal patterns ($\Delta_i$)
-- **Scale invariance**: Relative temporal patterns are independent of absolute time scale
-- **Biological realism**: Global time scale $T_{M}^*$ sets the pace for the entire process
+- **Pre-activation** ($t^*_{0,\text{on}i} < 0$): ~4.8% probability under the prior
+- **Transient** ($0 < t^*_{0,\text{on}i} < 2.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles
+- **Sustained** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation
+
+**Parameter independence benefits:**
+
+- **Identifiability**: Gene timing parameters are independent of process duration, eliminating scaling symmetry
+- **Biological realism**: Process boundaries ($0$ to $T_{\max}$) provide natural temporal reference frame
+- **Cross-scale applicability**: Same parameter values work across different experimental timescales
+- **Separation of concerns**: Process duration ($T_{\max}$) vs gene-specific timing (onset/duration) vs kinetics (degradation/fold-change)
 
@@
 
-**Hierarchical priors for latent time coordinates (unchanged from Study 2):**
-\begin{align}
-    T_{M}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \\
-    t_{loc} &\sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}) \\
-    t_{scl} &\sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}}) \\
-    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scl}^2) \quad \text{for } j \in \{1, \ldots, N\} \\
-    t^*_j &= T_{M}^* \times \max(\tilde{t}_j, \epsilon) \quad \text{(clipped transformation)}
-\end{align}
+**Stick-breaking cell time coordinates:**
+
+The cell times are sampled using a stick-breaking process that ensures proper ordering and eliminates scaling symmetry:
+
+\begin{align}
+    t^*_1 &= 0 \quad \text{(process start, fixed)} \\
+    t^*_N &= T_{\max} \quad \text{(process end)} \\
+    \xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
+    t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{(recursive construction)}
+\end{align}
+
+This construction automatically ensures $0 = t^*_1 < t^*_2 < \cdots < t^*_N = T_{\max}$ and eliminates the need for clipping operations.
 
@@
 The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{0,\text{on}i}$ and $\delta^*_i$.
+The stick-breaking cell times provide the temporal coordinates for evaluating the piecewise analytical solutions.
 
@@
 
-where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i})$, hierarchical time parameters $(T_{M}^*, t_{loc}, t_{scl}, t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.
+where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i})$, process duration parameter $(T_{\max})$, stick-breaking cell times $(t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.
 
@@
 \begin{align}
-p(\tilde{y}) &= \int p(\tilde{y} | \gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i}, T_{M}^*, t_{loc}, t_{scl}, t^*_j, \lambda_j) \times \nonumber \\
-&\quad p(\gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i}, T_{M}^*, t_{loc}, t_{scl}, t^*_j, \lambda_j) \, d\theta \label{eq-prior-predictive-piecewise}
+p(\tilde{y}) &= \int p(\tilde{y} | \gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}, T_{\max}, t^*_j, \lambda_j) \times \nonumber \\
+&\quad p(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}, T_{\max}, t^*_j, \lambda_j) \, d\theta \label{eq-prior-predictive-piecewise}
 \end{align}
@@
 \begin{align}
-p(\tilde{y}) &= \int \int \int \int \int p(\tilde{y} | T_{M}^*, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
-&\quad p(T_{M}^*) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{M}^* \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-hierarchical-integration}
+p(\tilde{y}) &= \int \int \int \int p(\tilde{y} | T_{\max}, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
+&\quad p(T_{\max}) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{\max} \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-hierarchical-integration}
 \end{align}
@@
 
-- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
-- $\theta_{\text{cell}} = \{t_{loc}, t_{scl}, \tilde{t}_j, t^*_j\}_{j=1}^N$ represents cell-specific temporal parameters
+- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
+- $\theta_{\text{cell}} = \{t^*_j\}_{j=1}^N$ represents stick-breaking cell times
 - $\theta_{\text{tech}} = \{\lambda_j\}_{j=1}^N$ represents technical efficiency parameters
 
-This decomposition reflects the model's assumption that gene-specific kinetics, population-level temporal structure, and technical factors contribute independently to the prior uncertainty.
+This decomposition reflects the model's assumption that gene-specific kinetics, process duration, stick-breaking cell times, and technical factors contribute independently to the prior uncertainty.
 
@@
 
-1. **Sample hierarchical time parameters**:
-   \begin{align}
-   {T_{M}^*}^{(s)} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}), \quad {t_{loc}}^{(s)} \sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}), \quad {t_{scl}}^{(s)} \sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}})
-   \end{align}
-
-2. **Sample cell-specific relative times**:
-   \begin{align}
-   {\tilde{t}_j}^{(s)} &\sim \text{Normal}({t_{loc}}^{(s)}, ({t_{scl}}^{(s)})^2) \quad \text{for } j \in \{1, \ldots, N\}
-   \end{align}
-
-3. **Transform to absolute times**:
-   \begin{align}
-   {t^{*}_j}^{(s)} &= {T_{M}^*}^{(s)} \times \max({\tilde{t}_j}^{(s)}, \epsilon)
-   \end{align}
-
-4. **Sample gene-specific parameters**:
+1. **Sample process duration**:
+   \begin{align}
+   {T_{\max}}^{(s)} &\sim \text{Gamma}(5.0, 1.0)
+   \end{align}
+
+2. **Sample stick-breaking variables and construct cell times**:
+   \begin{align}
+   {\xi_j}^{(s)} &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
+   {t^*_1}^{(s)} &= 0 \\
+   {t^*_N}^{(s)} &= {T_{\max}}^{(s)} \\
+   {t^*_j}^{(s)} &= {t^*_{j-1}}^{(s)} + {\xi_{j-1}}^{(s)}({T_{\max}}^{(s)} - {t^*_{j-1}}^{(s)})
+   \end{align}
+
+3. **Sample gene-specific parameters**:
    \begin{align}
@@
    {R_{\text{on},i}}^{(s)} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \\
-   {\tilde{t}^*_{0,\text{on}i}}^{(s)} &\sim \text{Normal}(0.5, 0.8^2) \\
-   {\tilde{\delta}^*_i}^{(s)} &\sim \text{LogNormal}(\log(0.45), 0.45^2) \\
+   {t^*_{0,\text{on}i}}^{(s)} &\sim \text{Normal}(2.5, 1.5^2) \\
+   {\delta^*_i}^{(s)} &\sim \text{LogNormal}(0.0, 0.5^2) \\
    {U_{0i}}^{(s)} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
@@
 
-5. **Transform relative temporal parameters**:
-   \begin{align}
-   {t^*_{0,\text{on}i}}^{(s)} &= {T_{M}^*}^{(s)} \times {\tilde{t}^*_{0,\text{on}i}}^{(s)}, \quad {\delta^*_i}^{(s)} = {T_{M}^*}^{(s)} \times {\tilde{\delta}^*_i}^{(s)}
-   \end{align}
-
-6. **Sample capture efficiencies**:
+4. **Sample capture efficiencies**:
    \begin{align}
@@
 
-7. **Compute latent concentrations**:
+5. **Compute latent concentrations**:
    \begin{align}
@@
 
-8. **Sample observations**:
+6. **Sample observations**:
    \begin{align}
@@
 
-1. **Global temporal scale uncertainty**: Uncertainty in $T_{M}^*$ affects the absolute timescale of all temporal processes, scaling both activation onset times and durations proportionally
-
-2.
-**Population temporal distribution uncertainty**: Uncertainty in $t_{loc}, t_{scl}$ affects the distribution of cells across the relative temporal coordinate, determining population-level temporal heterogeneity
+1. **Process duration uncertainty**: Uncertainty in $T_{\max}$ affects the total duration of the biological process, determining the temporal scale for all dynamics
+
+2. **Stick-breaking cell time uncertainty**: Uncertainty in the stick-breaking construction affects the distribution of cells across the process timeline
 
@@
 
-4.
-**Gene-specific temporal pattern uncertainty**: Prior uncertainty in $\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i$ affects when activation occurs and how long it lasts relative to the population timeline
+4. **Gene-specific temporal pattern uncertainty**: Prior uncertainty in $t^*_{0,\text{on}i}, \delta^*_i$ affects when activation occurs and how long it lasts in absolute time units
 
@@
 
-- **Temporal scaling interactions**: Uncertainty in $T_{M}^*$ interacts multiplicatively with uncertainty in relative temporal parameters, creating complex correlation structures
+- **Process duration interactions**: Uncertainty in $T_{\max}$ affects the temporal boundaries but does not automatically scale gene parameters, creating asymmetric uncertainty propagation
 
@@
 - **Non-temporal gene parameters**: $\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$, $R_{\text{on},i} \sim \text{LogNormal}(\log(2.5), 0.4^2)$
-- **Relative temporal gene parameters**: $\tilde{t}^*_{0,\text{on}i} \sim \text{Normal}(0.5, 0.8^2)$, $\tilde{\delta}^*_i \sim \text{LogNormal}(\log(0.45), 0.45^2)$
+- **Absolute temporal gene parameters**: $t^*_{0,\text{on}i} \sim \text{Normal}(2.5, 1.5^2)$, $\delta^*_i \sim \text{LogNormal}(0.0, 0.5^2)$
 - **Characteristic scales**: $U_{0i} \sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)$
-- **Hierarchical time parameters**: $T_{M}^* \sim \text{Gamma}(\alpha_{T}, \beta_{T})$, $t_{loc} \sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}})$, $t_{scl} \sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}})$
-- **Cell-specific times**: $\tilde{t}_j \sim \text{Normal}(t_{loc}, t_{scl}^2)$, $t^*_j = T_{M}^* \times \max(\tilde{t}_j, \epsilon)$
+- **Process duration**: $T_{\max} \sim \text{Gamma}(5.0, 1.0)$
+- **Stick-breaking cell times**: $t^*_1 = 0$, $t^*_N = T_{\max}$, $\xi_j \sim \text{Beta}(1, N-j)$, $t^*_j = t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1})$
 - **Capture efficiencies**: $\lambda_j \sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)$
@@
 1. **Kinetic parameter uncertainty**: Posterior distributions over $\gamma^*_i$ and $R_{\text{on},i}$ affect the shape and magnitude of concentration trajectories
-2. **Temporal parameter uncertainty**: Uncertainty in $\tilde{t}^*_{0,\text{on}i}$ and $\tilde{\delta}^*_i$ affects when activation occurs and how long it lasts
-3. **Hierarchical time uncertainty**: Uncertainty in $T_{M}^*$, $t_{loc}$, $t_{scl}$ affects the global temporal scaling and population-level time distribution
+2. **Temporal parameter uncertainty**: Uncertainty in $t^*_{0,\text{on}i}$ and $\delta^*_i$ affects when activation occurs and how long it lasts
+3. **Process duration uncertainty**: Uncertainty in $T_{\max}$ affects the total process duration and temporal boundaries
 4. **Cell-specific uncertainty**: Uncertainty in individual $t^*_j$ and $\lambda_j$ affects cell-specific predictions
@@
 
-### Properties of the Piecewise Framework
+### Properties of the Process Duration with Stick-Breaking Framework
+
+#### **Resolved Identifiability Issues**
+
+- **Eliminates scaling symmetry**: Fixed anchor points ($t^*_1 = 0$, $t^*_N = T_{\max}$) break scaling symmetry that plagued hierarchical scaling approaches
+- **Parameter identifiability**: Gene timing parameters are independent of process duration, ensuring unique parameter estimates
+- **Mathematical rigor**: Formal proof that likelihood is not invariant under parameter scaling
+- **Computational stability**: Eliminates non-identifiable parameter combinations that cause inference failures
 
@@
 
-- **Pre-activation scenarios**: Models genes already activated when observation begins
+- **Process boundaries**: Clear start ($t^* = 0$) and end ($t^* = T_{\max}$) of biological process
+- **Pre-activation scenarios**: Models genes already activated when process begins
 - **Flexible onset timing**: Normal prior allows negative onset times naturally
-- **Biological realism**: Captures realistic experimental scenarios better
-- **Maintains population structure**: Hierarchical time modeling preserved
+- **Automatic ordering**: Stick-breaking ensures proper temporal ordering without constraints
+- **Cross-scale applicability**: Same framework works across different experimental timescales
 
@@
 **Pattern coverage**: These three patterns effectively capture the essential temporal dynamics while eliminating redundant classifications.
-The relative temporal parameterization ensures consistent pattern identification across different experimental timescales.
+The absolute temporal parameterization with process boundaries ensures consistent pattern identification across different experimental timescales.
 
@@
 - **Dimensionless parameters**: Universal interpretation across experimental conditions
-- **Temporal coordination**: Hierarchical time structure captures population-level timing
-- **Gene-specific dynamics**: Each gene can follow distinct activation pattern
+- **Process boundaries**: Natural temporal reference frame from start to completion
+- **Temporal progression**: Intuitive progress metric ($t^*_j / T_{\max}$) for cellular advancement
+- **Gene-specific dynamics**: Each gene can follow distinct activation pattern independent of process duration
 
@@
 
-This study will validate the constrained model's ability to:
-
-1. **Recover constrained piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $\tilde{t}^*_{0,\text{on}i}$, $\tilde{\delta}^*_i$
-4. **Maintain hierarchical time inference**: Recover population-level temporal parameters ($T_M^*$, $t_{loc}$, $t_{scl}$)
+This study will validate the process duration with stick-breaking model's ability to:
+
+1. **Recover identifiable parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{0,\text{on}i}$, $\delta^*_i$, $T_{\max}$
+2. **Demonstrate scaling symmetry elimination**: Verify that different $T_{\max}$ values produce distinguishable likelihoods
+3. **Validate anchor point constraints**: Confirm that $t^*_1 = 0$ and $t^*_N = T_{\max}$ are properly enforced
+4. **Assess stick-breaking cell time inference**: Recover proper temporal ordering and distribution
 5. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
-6. **Scale efficiently**: Demonstrate computational advantages of reduced parameter space
+6. **Scale efficiently**: Demonstrate computational advantages of analytical solutions
 7. **Validate fold-change interpretation**: Ensure $R_{\text{on},i}$ estimates match true activation strength
 8. **Assess dimensional consistency**: Verify elimination of parameter redundancy through fixed $\alpha^*_{\text{off}i} = 1.0$
-9. **Validate temporal parameter scaling**: Ensure $T_{M}^*$ scaling correctly propagates from relative to absolute timing parameters
-10. **Assess parameter separation**: Verify clean separation between kinetic ($\gamma^*_i$, $R_{\text{on},i}$) and temporal ($\tilde{t}^*_{0,\text{on}i}$, $\tilde{\delta}^*_i$) parameter inference
-11. **Validate relative parameter framework**: Confirm that relative temporal parameters enable consistent pattern classification across different experimental timescales
+9. **Validate parameter independence**: Ensure gene timing parameters are independent of process duration
+10. **Assess parameter separation**: Verify clean separation between kinetic ($\gamma^*_i$, $R_{\text{on},i}$) and temporal ($t^*_{0,\text{on}i}$, $\delta^*_i$) parameter inference
+11. **Validate process boundary framework**: Confirm that absolute temporal parameters enable consistent pattern classification across different experimental timescales
+12. **Cross-scale validation**: Test parameter recovery across different process durations to verify identifiability
 
@@
 
-This validation workflow demonstrates the key capabilities of the piecewise activation model:
-
+This validation workflow demonstrates the key capabilities of the process duration with stick-breaking model:
+
+- **Parameter identifiability**: Eliminates scaling symmetry through fixed anchor points and independent gene timing
 - **Analytical tractability**: Fast likelihood evaluation enables efficient parameter recovery validation
-- **Rich parameter space**: Includes fold-change parameters, relative temporal parameters, and hierarchical time structure
-- **Pre-activation modeling**: Supports negative onset times for genes already activated before observation
+- **Rich parameter space**: Includes fold-change parameters, absolute temporal parameters, and process duration
+- **Pre-activation modeling**: Supports negative onset times for genes already activated before process start
 - **Dimensional consistency**: Eliminates parameter redundancy through fixed reference state parameterization
-- **Biological interpretability**: All parameters have clear biological meaning and interpretable ranges
+- **Biological interpretability**: All parameters have clear biological meaning with process boundary interpretation
 - **Comprehensive prior predictive framework**: Mathematical foundation for validating model specification through prior predictive checking as detailed in equations \eqref{eq-prior-predictive-piecewise}-\eqref{eq-information-content}, ensuring the model can generate realistic data patterns before observing any data
-- **Relative temporal parameterization**: Enables interpretable parameter values and consistent pattern classification across different experimental timescales and even distinct data sets
+- **Process boundary parameterization**: Enables interpretable parameter values and consistent pattern classification across different experimental timescales through natural temporal reference frame
 

### 2025-06-13 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): fix process duration with stick-breaking temporal prior description
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-**Identifiability proof**: For any $\lambda \neq 1$:
+**Theorem (Elimination of Scaling Symmetry):**
+
+For any scaling factor $\lambda \neq 1$, the likelihood functions are not equal:
 $$\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$$
 
-The fixed anchor $t^*_1 = 0$ cannot be scaled, and gene parameters are independent of $T_{\max}$, breaking the symmetry.
+**Proof:**
+1. **Fixed anchor constraint**: The scaled system requires $t^*_1 = \lambda \cdot 0 = 0$ and $t^*_N = \lambda T_{\max}$
+2. **Process duration independence**: Gene parameters $t^*_{0,\text{on}i}, \delta^*_i$ are sampled independently of $T_{\max}$ and do not scale
+3. **Asymmetric scaling**: Scaling $T_{\max} \rightarrow \lambda T_{\max}$ and $t^*_j \rightarrow \lambda t^*_j$ changes the relative positioning of gene events within the process timeline
+4. **Likelihood dependence**: The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$, which change under asymmetric scaling
+
+Therefore, the likelihood is not invariant under parameter scaling, eliminating the scaling symmetry. □
 
@@
 
-**Pattern classification framework**: The three expression patterns are defined by constraints on the absolute temporal parameters:
-
-- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
-- **Transient**: $0.0 < t^*_{0,\text{on}i} < 2.5$, $\delta^*_i < 2.0$, and $R_{\text{on},i} > 2.0$
-- **Sustained**: $0.0 < t^*_{0,\text{on}i} < 1.5$, $\delta^*_i > 2.5$, and $R_{\text{on},i} > 2.0$
+**Pattern Classification with Absolute Parameters:**
+
+Using the expected process duration $\mathbb{E}[T_{\max}] = 5.0$, the pattern constraints become:
+
+- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$ (gene activated before process begins)
+- **Transient**: $0.0 < t^*_{0,\text{on}i} < 2.5$ and $\delta^*_i < 2.0$ and $R_{\text{on},i} > 2.0$ (complete cycle within process)
+- **Sustained**: $0.0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$ and $R_{\text{on},i} > 2.0$ (long-duration activation)
+
+These absolute thresholds preserve the same biological pattern meanings as the original relative formulation.
 
@@
 
+**Cross-System Biological Interpretation:**
+
+The same $T_{\max} = 5.0$ represents different real-world durations depending on the characteristic splicing rate $\beta$:
+
+- **Immediate-early response** ($\beta = 5 \text{ hr}^{-1}$): Process duration = $5.0/5 = 1$ hour
+- **Cell cycle progression** ($\beta = 1 \text{ hr}^{-1}$): Process duration = $5.0/1 = 5$ hours
+- **Developmental transition** ($\beta = 0.1 \text{ hr}^{-1}$): Process duration = $5.0/0.1 = 50$ hours
+
+Gene timing parameters maintain consistent biological interpretation across scales.
+
 **Priors for other gene-specific parameters (unchanged from Study 2):**
@@
 
-**Stick-breaking cell time coordinates:**
-
-The cell times are sampled using a stick-breaking process that ensures proper ordering and eliminates scaling symmetry:
-
-\begin{align}
-    t^*_1 &= 0 \quad \text{(process start, fixed)} \\
-    t^*_N &= T_{\max} \quad \text{(process end)} \\
-    \xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
-    t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{(recursive construction)}
-\end{align}
-
-This construction automatically ensures $0 = t^*_1 < t^*_2 < \cdots < t^*_N = T_{\max}$ and eliminates the need for clipping operations.
+**Stick-Breaking Cell Time Construction:**
+
+The cell times $\{t^*_j\}_{j=1}^N$ are constructed using a stick-breaking process with fixed anchor points:
+
+\begin{align}
+t^*_1 &= 0 \quad \text{(fixed process start)} \label{eq-anchor-start} \\
+t^*_N &= T_{\max} \quad \text{(fixed process end)} \label{eq-anchor-end} \\
+\xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \label{eq-stick-breaking-weights} \\
+t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{for } j \in \{2, \ldots, N-1\} \label{eq-stick-breaking-construction}
+\end{align}
+
+This construction automatically ensures:
+- **Ordered constraint**: $0 = t^*_1 < t^*_2 < \cdots < t^*_N = T_{\max}$
+- **Fixed boundaries**: Process start and end are anchored
+- **Efficient sampling**: $O(N)$ complexity with no constraint violations
 
@@
 
-2. **Sample stick-breaking variables and construct cell times**:
+2. **Sample stick-breaking cell times**:
    \begin{align}
    {\xi_j}^{(s)} &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
-   {t^*_1}^{(s)} &= 0 \\
-   {t^*_N}^{(s)} &= {T_{\max}}^{(s)} \\
-   {t^*_j}^{(s)} &= {t^*_{j-1}}^{(s)} + {\xi_{j-1}}^{(s)}({T_{\max}}^{(s)} - {t^*_{j-1}}^{(s)})
+   {t^*_1}^{(s)} &= 0, \quad {t^*_N}^{(s)} = {T_{\max}}^{(s)} \\
+   {t^*_j}^{(s)} &= {t^*_{j-1}}^{(s)} + {\xi_{j-1}}^{(s)}({T_{\max}}^{(s)} - {t^*_{j-1}}^{(s)}) \quad \text{for } j \in \{2, \ldots, N-1\}
    \end{align}

### 2025-06-14 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): add branching mixture model description
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 The prior predictive checking framework validates model specification and prior adequacy, while the posterior predictive checking framework assesses parameter recovery and predictive accuracy.
+
+## Dimensionless Analytical Dynamics with Branching Trajectories and Piecewise Activation {#sec-dimless-branching-piecewise}
+
+This validation study extends the stick-breaking temporal coordinate model to accommodate **branching developmental trajectories** through a mathematically rigorous mixture model framework.
+The key innovation is a **branch mixture extension** that maintains all analytical tractability and identifiability properties while enabling modeling of cell fate decisions, lineage bifurcations, and heterogeneous developmental pathways within a unified probabilistic framework.
+
+### Model Description
+
+The branching mixture model extends the process duration with stick-breaking framework by introducing **discrete branch assignments** that partition cells into distinct developmental trajectories.
+Each branch maintains its own temporal coordinate system and gene expression parameters while sharing the same underlying analytical solutions and dimensional consistency principles.
+This approach enables modeling of complex developmental processes where cells follow different trajectories with branch-specific timing and expression patterns.
+
+The model generates count data for unspliced ($u$) and spliced ($s$) RNA molecules per cell, with dynamics governed by **branch-specific piecewise analytical solutions** and observation model that scales back to dimensional counts through characteristic scales and effective capture efficiency.
+Each cell is assigned to exactly one branch, and all parameters are conditionally independent given branch assignments.
+
+### Probabilistic Graphical Model
+
+The probabilistic graphical model represents the branching extension with discrete branch assignments, branch-specific temporal coordinates, and branch-specific gene expression parameters.
+The model maintains the same dimensional consistency and analytical tractability as the single-branch framework while enabling rich heterogeneity across developmental trajectories.
+
+```{python}
+#| label: fig-branching-model-pgm
+#| code-fold: true
+#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with branching trajectories and piecewise activation.
+Branch assignments partition cells into distinct developmental pathways with branch-specific temporal coordinates and gene expression parameters.
+
+import daft
+import matplotlib.pyplot as plt
+
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["font.size"] = 14
+plt.rcParams["text.usetex"] = True
+plt.rcParams["text.latex.preamble"] = r"\usepackage{amsmath}"
+
+pgm = daft.PGM(line_width=1.2)
+
+# Branch structure hyperparameters
+pgm.add_node("alpha_K", r"$\alpha_K$", 0.5, 7, fixed=True)
+pgm.add_node("alpha_pi", r"$\alpha_\pi$", 1.5, 7, fixed=True)
+
+# Gene parameter hyperparameters (shared across branches)
+pgm.add_node("mu_gamma", r"$\mu_{\gamma^*}$", 3, 7, fixed=True)
+pgm.add_node("sigma_gamma", r"$\sigma_{\gamma^*}^2$", 3.5, 7, fixed=True)
+pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 7, fixed=True)
+pgm.add_node("sigma_U0", r"$\sigma_{U_0}^2$", 5, 7, fixed=True)
+
+# Branch-specific hyperparameters
+pgm.add_node("alpha_T", r"$\alpha_T$", 6, 7, fixed=True)
+pgm.add_node("beta_T", r"$\beta_T$", 6.5, 7, fixed=True)
+pgm.add_node("mu_on_k", r"$\mu_{on,k}$", 7.5, 7, fixed=True)
+pgm.add_node("sigma_on", r"$\sigma_{on}^2$", 8, 7, fixed=True)
+pgm.add_node("mu_delta_k", r"$\mu_{\delta,k}$", 8.5, 7, fixed=True)
+pgm.add_node("sigma_delta", r"$\sigma_{\delta}^2$", 9, 7, fixed=True)
+pgm.add_node("mu_R_k", r"$\mu_{R,k}$", 9.5, 7, fixed=True)
+pgm.add_node("sigma_R", r"$\sigma_R^2$", 10, 7, fixed=True)
+
+# Branch structure
+pgm.add_node("K", r"$K$", 1, 6)
+pgm.add_node("pi", r"$\boldsymbol{\pi}$", 2, 6)
+
+# Branch-specific parameters
+pgm.add_node("T_max_k", r"$T_{\max,k}$", 6.25, 6)
+
+# Gene-specific parameters (shared)
+pgm.add_node("gamma_star_i", r"$\gamma^*_i$", 3.25, 5.5)
+pgm.add_node("U_0i", r"$U_{0i}$", 4.5, 5.5)
+
+# Branch-gene-specific parameters
+pgm.add_node("t_on_ik", r"${t^*_{on}}_{ik}$", 7.5, 5.5)
+pgm.add_node("delta_ik", r"$\delta^*_{ik}$", 8.5, 5.5)
+pgm.add_node("R_on_ik", r"$R_{on,ik}$", 9.5, 5.5)
+
+# Cell assignments and parameters
+pgm.add_node("Z_j", r"$Z_j$", 2, 5)
+pgm.add_node("t_star_j", r"${t^*}_j$", 6.25, 5)
+pgm.add_node("lambda_j", r"$\lambda_j$", 5, 4.5)
+
+# Latent concentrations
+pgm.add_node("u_star_ij", r"$u^*_{ij}$", 3.25, 4, scale=1.0, shape="rectangle")
+pgm.add_node("s_star_ij", r"$s^*_{ij}$", 4.5, 4, scale=1.0, shape="rectangle")
+
+# Observed counts
+pgm.add_node("u_obs_ij", r"$u_{ij}$", 3.25, 3, scale=1.0, observed=True)
+pgm.add_node("s_obs_ij", r"$s_{ij}$", 4.5, 3, scale=1.0, observed=True)
+
+# edges
+edge_params = {"head_length": 0.25, "head_width": 0.2, "lw": 0.6}
+
+# Branch structure edges
+pgm.add_edge("alpha_K", "K", plot_params=edge_params)
+pgm.add_edge("alpha_pi", "pi", plot_params=edge_params)
+pgm.add_edge("K", "pi", plot_params=edge_params)
+pgm.add_edge("pi", "Z_j", plot_params=edge_params)
+
+# Branch-specific parameter edges
+pgm.add_edge("alpha_T", "T_max_k", plot_params=edge_params)
+pgm.add_edge("beta_T", "T_max_k", plot_params=edge_params)
+pgm.add_edge("mu_on_k", "t_on_ik", plot_params=edge_params)
+pgm.add_edge("sigma_on", "t_on_ik", plot_params=edge_params)
+pgm.add_edge("mu_delta_k", "delta_ik", plot_params=edge_params)
+pgm.add_edge("sigma_delta", "delta_ik", plot_params=edge_params)
+pgm.add_edge("mu_R_k", "R_on_ik", plot_params=edge_params)
+pgm.add_edge("sigma_R", "R_on_ik", plot_params=edge_params)
+
+# Gene parameter edges
+pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
+pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
+pgm.add_edge("sigma_U0", "U_0i", plot_params=edge_params)
+
+# Temporal coordinate edges
+pgm.add_edge("Z_j", "t_star_j", plot_params=edge_params)
+pgm.add_edge("T_max_k", "t_star_j", plot_params=edge_params)
+
+# Dynamics edges
+pgm.add_edge("gamma_star_i", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("t_on_ik", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("t_on_ik", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("delta_ik", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("delta_ik", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("R_on_ik", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("R_on_ik", "s_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
+pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)
+
+# Observation edges
+pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("U_0i", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("U_0i", "s_obs_ij", plot_params=edge_params)
+pgm.add_edge("lambda_j", "u_obs_ij", plot_params=edge_params)
+pgm.add_edge("lambda_j", "s_obs_ij", plot_params=edge_params)
+
+# plates
+pgm.add_plate([0.75, 6.25, 1.75, 1.25], label=r"$k \in \{1, \ldots, K\}$", shift=-0.1, fontsize=10)
+pgm.add_plate([2.75, 4.75, 7.25, 1.25], label=r"$i \in \{1, \ldots, G\}$", shift=-0.1, fontsize=10)
+pgm.add_plate([1.5, 2.5, 4, 3], label=r"$j \in \{1, \ldots, N\}$", shift=-0.1, fontsize=10)
+
+pgm.render()
+```
+
+### Mathematical Description
+
+The branching mixture model extends the stick-breaking framework with **discrete branch assignments** and **branch-specific parameters**.
+The model maintains analytical tractability and identifiability while enabling rich heterogeneity across developmental trajectories.
+
+**Branch structure:**
+\begin{align}
+K &\sim \text{Poisson}(\alpha_K) + 1 \label{eq-branch-count} \\
+\boldsymbol{\pi} &\sim \text{Dirichlet}\left(\frac{\alpha_\pi}{K} \mathbf{1}_K\right) \label{eq-branch-probabilities} \\
+Z_j &\sim \text{Categorical}(\boldsymbol{\pi}) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-branch-assignments}
+\end{align}
+
+where $K$ is the number of branches, $\boldsymbol{\pi} = (\pi_1, \ldots, \pi_K)$ are branch probabilities, and $Z_j \in \{1, \ldots, K\}$ is the branch assignment for cell $j$.
+
+**Branch-specific temporal coordinates:**
+\begin{align}
+T_{\max,k} &\sim \text{Gamma}(\alpha_T, \beta_T) \quad \text{for } k \in \{1, \ldots, K\} \label{eq-branch-duration} \\
+{t^*}_j \mid Z_j = k &= \text{stick\_breaking}(0, T_{\max,k}) \label{eq-branch-stick-breaking}
+\end{align}
+
+Each branch $k$ has its own process duration $T_{\max,k}$ and stick-breaking temporal coordinate system with fixed anchors at $0$ and $T_{\max,k}$.
+
+**Gene-specific parameters (shared across branches):**
+\begin{align}
+\gamma^*_i &\sim \text{LogNormal}(\mu_{\gamma^*}, \sigma_{\gamma^*}^2) \label{eq-shared-gamma} \\
+U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2) \label{eq-shared-U0}
+\end{align}
+
+**Branch-gene-specific parameters:**
+\begin{align}
+{t^*_{on}}_{ik} &\sim \text{Normal}(\mu_{on,k}, \sigma_{on}^2) \quad \text{for } i \in \{1, \ldots, G\}, k \in \{1, \ldots, K\} \label{eq-branch-onset} \\
+\delta^*_{ik} &\sim \text{LogNormal}(\mu_{\delta,k}, \sigma_{\delta}^2) \label{eq-branch-duration-gene} \\
+R_{on,ik} &\sim \text{LogNormal}(\mu_{R,k}, \sigma_R^2) \label{eq-branch-fold-change}
+\end{align}
+
+**Cell-specific capture efficiency:**
+\begin{align}
+\lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2) \label{eq-branch-lambda}
+\end{align}
+
+**Branch-specific piecewise analytical solutions:**
+
+For cell $j$ assigned to branch $k$ (i.e., $Z_j = k$), the dimensionless concentrations follow the same piecewise analytical solutions as the single-branch model, but using branch-specific parameters:
+
+\begin{align}
+\alpha^*_{ik}(t^*) &:= \begin{cases}
+1.0 & \text{if } t^* < {t^*_{on}}_{ik} \\
+R_{on,ik} & \text{if } {t^*_{on}}_{ik} \leq t^* < {t^*_{on}}_{ik} + \delta^*_{ik} \\
+1.0 & \text{if } t^* \geq {t^*_{on}}_{ik} + \delta^*_{ik}
+\end{cases} \label{eq-branch-piecewise-transcription}
+\end{align}
+
+The analytical solutions for each phase remain identical to the single-branch case, but now use branch-gene-specific parameters ${t^*_{on}}_{ik}$, $\delta^*_{ik}$, and $R_{on,ik}$.
+
+**Mixture likelihood:**
+\begin{align}
+\mathcal{L}(\text{data}) &= \prod_{j=1}^N \sum_{k=1}^K \pi_k \cdot \mathcal{L}_k(\text{data}_j \mid \text{branch } k) \label{eq-mixture-likelihood}
+\end{align}
+
+where $\mathcal{L}_k(\text{data}_j \mid \text{branch } k)$ is the likelihood for cell $j$ under branch $k$ using branch-specific parameters.
+
+**Observation model (unchanged):**
+\begin{align}
+u_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}) \label{eq-branch-u-obs} \\
+s_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij}) \label{eq-branch-s-obs}
+\end{align}
+
+where $u^*_{ij}$ and $s^*_{ij}$ are computed using the branch-specific piecewise analytical solutions for the branch to which cell $j$ is assigned.
+
+### Branch-Specific Parameter Structure
+
+The branching mixture model introduces a **hierarchical parameter organization** that balances biological interpretability with computational efficiency.
+Parameters are strategically partitioned into shared and branch-specific categories based on biological considerations and identifiability requirements.
+
+#### **Shared Parameters Across Branches**
+
+**$\gamma^*_i$ (Relative degradation rate)**
+
+- **Rationale**: mRNA degradation kinetics are primarily determined by sequence-specific features (3' UTR elements, miRNA binding sites, RNA-binding protein motifs) that are intrinsic to the gene and independent of developmental trajectory
+- **Biological justification**: Degradation machinery (exonucleases, deadenylation complexes) operates consistently across cell types within the same organism
+- **Mathematical advantage**: Reduces parameter space from $G \times K$ to $G$ parameters while maintaining biological realism
+- **Identifiability**: Shared degradation rates enable borrowing strength across branches for better parameter estimation
+
+**$U_{0i}$ (Characteristic concentration scale)**
+
+- **Rationale**: Represents the gene's fundamental expression capacity, which reflects promoter strength, chromatin accessibility, and transcriptional machinery availability
+- **Biological justification**: While absolute expression levels vary across branches, the characteristic scale represents the gene's intrinsic transcriptional potential
+- **Mathematical advantage**: Separates scaling effects from temporal dynamics, maintaining dimensional consistency across branches
+- **Identifiability**: Shared scaling parameters prevent confounding between branch-specific temporal effects and expression magnitude
+
+#### **Branch-Specific Parameters**
+
+**$T_{\max,k}$ (Branch-specific process duration)**
+
+- **Definition**: Maximum dimensionless time for developmental process in branch $k$
+- **Biological interpretation**: Different developmental trajectories may proceed at different rates, with some lineages requiring longer maturation times
+- **Examples**:
+  - Neuronal differentiation (slow, $T_{\max} \sim 10$) vs. myeloid differentiation (fast, $T_{\max} \sim 3$)
+  - Stress response (rapid, $T_{\max} \sim 2$) vs. metabolic adaptation (gradual, $T_{\max} \sim 8$)
+- **Mathematical role**: Provides branch-specific temporal scaling while maintaining stick-breaking identifiability
+
+**${t^*_{on}}_{ik}$ (Branch-gene-specific onset time)**
+
+- **Definition**: Dimensionless activation onset time for gene $i$ in branch $k$
+- **Biological interpretation**: Different developmental trajectories activate the same genes at different relative timepoints
+- **Examples**:
+  - Transcription factor cascades: Early TFs activate first in committed lineages, later in multipotent cells
+  - Metabolic genes: Glycolytic genes activate early in proliferative branches, late in quiescent branches
+  - Stress response: Immediate early genes activate quickly in acute stress, slowly in chronic adaptation
+- **Prior structure**: $\text{Normal}(\mu_{on,k}, \sigma_{on}^2)$ allows branch-specific timing while sharing variance across genes
+
+**$\delta^*_{ik}$ (Branch-gene-specific activation duration)**
+
+- **Definition**: Dimensionless activation duration for gene $i$ in branch $k$
+- **Biological interpretation**: Same genes may have different activation durations across developmental trajectories
+- **Examples**:
+  - Cell cycle genes: Brief activation in rapidly dividing cells, prolonged in slowly cycling cells
+  - Differentiation markers: Transient in intermediate states, sustained in terminal differentiation
+  - Signaling molecules: Pulse-like in signaling cells, sustained in responding cells
+- **Prior structure**: $\text{LogNormal}(\mu_{\delta,k}, \sigma_{\delta}^2)$ ensures positive durations with branch-specific means
+
+**$R_{on,ik}$ (Branch-gene-specific fold-change)**
+
+- **Definition**: Fold-change in transcription rate during activation for gene $i$ in branch $k$
+- **Biological interpretation**: Same genes may be activated to different extents across developmental trajectories
+- **Examples**:
+  - Lineage markers: Strong activation in committed lineages, weak in multipotent cells
+  - Metabolic enzymes: High fold-change in metabolically active branches, low in quiescent branches
+  - Stress genes: Massive activation in stressed cells, moderate in adapted cells
+- **Prior structure**: $\text{LogNormal}(\mu_{R,k}, \sigma_R^2)$ ensures positive fold-changes with branch-specific means
+
+#### **Hyperparameter Interpretation**
+
+**Branch-specific location parameters** ($\mu_{on,k}$, $\mu_{\delta,k}$, $\mu_{R,k}$):
+
+- **Biological meaning**: Characteristic timing and expression patterns for each developmental trajectory
+- **Identifiability**: Enable learning of branch-specific expression programs while sharing variance parameters
+- **Flexibility**: Allow each branch to have distinct temporal and expression characteristics
+
+**Shared variance parameters** ($\sigma_{on}^2$, $\sigma_{\delta}^2$, $\sigma_R^2$):
+
+- **Statistical efficiency**: Borrowing strength across branches improves parameter estimation
+- **Biological realism**: Gene-to-gene variation in timing and expression is similar across developmental contexts
+- **Computational advantage**: Reduces parameter space while maintaining model flexibility
+
+### Stick-Breaking Within Branches
+
+The branching mixture model implements **independent stick-breaking processes** within each branch, maintaining all identifiability and computational advantages of the single-branch framework while enabling branch-specific temporal dynamics.
+
+#### **Branch-Specific Stick-Breaking Construction**
+
+For each branch $k \in \{1, \ldots, K\}$, cells assigned to that branch ($\{j : Z_j = k\}$) follow an independent stick-breaking process:
+
+\begin{align}
+\text{Let } \mathcal{N}_k &= |\{j : Z_j = k\}| \quad \text{(number of cells in branch } k\text{)} \\
+\text{For cells } j_1, j_2, \ldots, j_{\mathcal{N}_k} &\text{ where } Z_{j_\ell} = k: \\
+V_{k,\ell} &\sim \text{Beta}(1, \mathcal{N}_k - \ell) \quad \text{for } \ell \in \{1, \ldots, \mathcal{N}_k - 1\} \label{eq-branch-beta-variables} \\
+\tilde{t}_{k,1} &= 0 \quad \text{(first cell at branch start)} \\
+\tilde{t}_{k,\ell} &= \tilde{t}_{k,\ell-1} + V_{k,\ell-1} \prod_{m=1}^{\ell-2} (1 - V_{k,m}) \quad \text{for } \ell \in \{2, \ldots, \mathcal{N}_k\} \label{eq-branch-stick-construction} \\
+{t^*}_{j_\ell} &= T_{\max,k} \times \tilde{t}_{k,\ell} \quad \text{(branch-specific scaling)} \label{eq-branch-time-scaling}
+\end{align}
+
+#### **Mathematical Properties of Branch-Specific Stick-Breaking**
+
+**Ordering preservation within branches:**
+\begin{align}
+0 = {t^*}_{j_1} < {t^*}_{j_2} < \cdots < {t^*}_{j_{\mathcal{N}_k}} = T_{\max,k}
+\end{align}
+
+**Independence across branches:**
+\begin{align}
+\{V_{k,\ell}\}_{\ell=1}^{\mathcal{N}_k-1} \perp \{V_{k',\ell'}\}_{\ell'=1}^{\mathcal{N}_{k'}-1} \quad \text{for } k \neq k'
+\end{align}
+
+**Identifiability within each branch:**
+
+Each branch maintains the same identifiability properties as the single-branch model:
+- Fixed anchors: ${t^*}_{j_1} = 0$ and ${t^*}_{j_{\mathcal{N}_k}} = T_{\max,k}$
+- No scaling symmetry within branch due to fixed endpoints
+- Gene parameters ${t^*_{on}}_{ik}$, $\delta^*_{ik}$ are independent of $T_{\max,k}$
+
+#### **Computational Complexity Analysis**
+
+**Time complexity:**
+- **Single-branch stick-breaking**: $O(N \log N)$ for $N$ cells
+- **Multi-branch stick-breaking**: $O\left(\sum_{k=1}^K \mathcal{N}_k \log \mathcal{N}_k\right) = O(N \log N)$ since $\sum_{k=1}^K \mathcal{N}_k = N$
+- **No asymptotic increase**: Branching does not increase computational complexity
+
+**Memory complexity:**
+- **Additional storage**: $O(K)$ for branch-specific parameters
+- **Cell assignments**: $O(N)$ for storing $Z_j$ values
+- **Total increase**: $O(K + N)$, linear in problem size
+
+**Parallel computation opportunities:**
+- **Independent branch processing**: Stick-breaking for each branch can be computed in parallel
+- **Gene parameter updates**: Branch-gene parameters can be updated independently across branches
+- **Likelihood evaluation**: Branch-specific likelihoods can be computed in parallel
+
+#### **Implementation Considerations**
+
+**Branch assignment updates:**
+
+During MCMC sampling, branch assignments $Z_j$ are updated using standard categorical sampling:
+\begin{align}
+P(Z_j = k \mid \text{rest}) \propto \pi_k \cdot \mathcal{L}_k(\text{data}_j \mid \text{branch } k)
+\end{align}
+
+**Stick-breaking recomputation:**
+
+When branch assignments change, only the affected branches need stick-breaking recomputation:
+- **Efficient updates**: Only recompute stick-breaking for branches that gained or lost cells
+- **Unchanged branches**: Branches with stable membership retain their temporal coordinates
+- **Computational savings**: Avoid full recomputation when only a few cells change branches
+
+**Numerical stability:**
+
+- **Beta variable bounds**: Ensure $V_{k,\ell} \in (0, 1)$ to prevent degenerate stick-breaking
+- **Minimum branch size**: Require $\mathcal{N}_k \geq 2$ for meaningful stick-breaking (at least start and end points)
+- **Empty branch handling**: Branches with $\mathcal{N}_k = 0$ are automatically excluded from likelihood computation
+
+### Identifiability Analysis
+
+The branching mixture model preserves all identifiability properties of the single-branch framework while introducing additional structure that enhances parameter recovery through **conditional independence** and **hierarchical regularization**.
+
+#### **Preservation of Single-Branch Identifiability**
+
+**Theorem (Branch-Conditional Identifiability):**
+
+Given branch assignments $\{Z_j\}_{j=1}^N$, each branch $k$ maintains the same identifiability properties as the single-branch model.
+
+**Proof:**
+
+1. **Fixed anchors within branch**: For cells in branch $k$, we have ${t^*}_{j_1} = 0$ and ${t^*}_{j_{\mathcal{N}_k}} = T_{\max,k}$
+
+2. **Independent gene parameters**: ${t^*_{on}}_{ik}$ and $\delta^*_{ik}$ are sampled independently of $T_{\max,k}$
+
+3. **No scaling symmetry**: The same argument as in the single-branch case applies within each branch
+
+4. **Analytical solutions**: The piecewise analytical solutions within each branch are identical to the single-branch case
+
+Therefore, conditional on branch assignments, each branch is identifiable. □
+
+#### **Enhanced Identifiability Through Mixture Structure**
+
+**Cross-branch parameter constraints:**
+
+The mixture structure introduces additional identifiability through **cross-branch comparisons**:
+
+\begin{align}
+\mathcal{L}_{\text{mixture}} &= \prod_{j=1}^N \sum_{k=1}^K \pi_k \cdot \mathcal{L}_k(\text{data}_j \mid \text{branch } k) \\
+&\neq \prod_{j=1}^N \sum_{k=1}^K \pi_k \cdot \mathcal{L}_k(\text{data}_j \mid \text{scaled branch } k)
+\end{align}
+
+**Theorem (Mixture Identifiability Enhancement):**
+
+The mixture likelihood is not invariant under branch-specific parameter scaling, providing additional identifiability constraints beyond single-branch identifiability.
+
+**Proof:**
+
+1. **Asymmetric scaling**: Scaling parameters in branch $k$ but not branch $k'$ changes the relative likelihood contributions
+
+2. **Branch assignment sensitivity**: Cells near branch boundaries become more or less likely to be assigned to scaled branches
+
+3. **Cross-branch competition**: The mixture structure creates competition between branches that prevents arbitrary scaling
+
+4. **Hierarchical constraints**: Shared hyperparameters ($\sigma_{on}^2$, $\sigma_{\delta}^2$, $\sigma_R^2$) provide additional regularization
+
+Therefore, the mixture structure enhances identifiability beyond the single-branch case. □
+
+#### **Identifiability of Branch Structure Parameters**
+
+**Number of branches ($K$):**
+
+- **Identifiable through**: Model comparison using information criteria (WAIC, LOO-CV)
+- **Practical bounds**: $K \leq N/2$ to ensure sufficient cells per branch for stick-breaking
+- **Biological constraints**: $K$ should reflect known developmental trajectories
+
+**Branch probabilities ($\boldsymbol{\pi}$):**
+
+- **Identifiable through**: Observed frequencies of branch assignments
+- **Dirichlet prior**: Provides regularization while allowing flexible branch sizes
+- **Constraint**: $\sum_{k=1}^K \pi_k = 1$ ensures proper probability distribution
+
+**Branch assignments ($\{Z_j\}$):**
+
+- **Identifiable through**: Likelihood comparison across branches for each cell
+- **Uncertainty quantification**: Posterior probabilities $P(Z_j = k \mid \text{data})$ provide assignment confidence
+- **Soft clustering**: Cells can have uncertain branch assignments, reflecting biological reality
+
+#### **Practical Identifiability Considerations**
+
+**Minimum branch size requirements:**
+
+- **Stick-breaking constraint**: Each branch needs $\mathcal{N}_k \geq 2$ for meaningful temporal ordering
+- **Parameter estimation**: Larger branches provide better parameter estimates
+- **Recommended minimum**: $\mathcal{N}_k \geq 10$ for stable parameter recovery
+
+**Gene parameter identifiability:**
+
+- **Shared parameters**: $\gamma^*_i$ and $U_{0i}$ benefit from pooling across branches
+- **Branch-specific parameters**: Require sufficient cells and temporal coverage within each branch
+- **Hierarchical regularization**: Branch-specific means with shared variances improve identifiability
+
+**Temporal coverage requirements:**
+
+- **Within-branch coverage**: Each branch should span sufficient temporal range for parameter recovery
+- **Cross-branch comparison**: Overlapping temporal ranges enable branch discrimination
+- **Gene activation coverage**: Each branch should contain cells before, during, and after gene activation events
+
+### Biological Interpretation
+
+The branching mixture model provides a **mathematically rigorous framework** for modeling complex developmental processes where cells follow distinct trajectories with different temporal dynamics and gene expression programs.
+
+#### **Developmental Biology Applications**
+
+**Cell fate decisions and lineage bifurcations:**
+
+The model naturally captures **binary and multi-way cell fate decisions** where progenitor cells commit to distinct developmental lineages:
+
+- **Hematopoiesis**: Multipotent stem cells differentiating into myeloid vs. lymphoid lineages
+- **Neural development**: Neural progenitors becoming neurons vs. glia
+- **Embryonic development**: Pluripotent cells committing to different germ layers
+- **Tissue regeneration**: Stem cells responding to injury with different repair programs
+
+**Branch-specific temporal dynamics:**
+
+Different developmental trajectories often proceed at **different rates** and with **different timing**:
+
+- **Fast vs. slow differentiation**: Some lineages mature rapidly (hours), others slowly (days)
+- **Sequential vs. parallel activation**: Genes may activate in different orders across branches
+- **Transient vs. sustained expression**: Same genes may show different temporal patterns across lineages
+
+**Examples of branch-specific gene expression patterns:**
+
+1. **Transcription factor cascades**:
+   - **Branch 1 (Neural)**: Early activation of neuronal TFs, sustained expression of synaptic genes
+   - **Branch 2 (Glial)**: Later activation of glial TFs, transient expression of migration genes
+
+2. **Metabolic reprogramming**:
+   - **Branch 1 (Proliferative)**: Early glycolytic gene activation, sustained cell cycle gene expression
+   - **Branch 2 (Quiescent)**: Early oxidative metabolism genes, transient stress response activation
+
+3. **Stress response pathways**:
+   - **Branch 1 (Adaptation)**: Gradual activation of protective genes, sustained expression
+   - **Branch 2 (Apoptosis)**: Rapid activation of death genes, transient survival gene expression
+
+#### **Experimental Design Connections**
+
+**Single-cell trajectory inference:**
+
+The model provides a **probabilistic foundation** for trajectory inference methods:
+
+- **Pseudotime estimation**: Branch-specific temporal coordinates provide principled pseudotime
+- **Lineage tracing**: Branch assignments correspond to inferred lineage relationships
+- **Trajectory topology**: Number of branches and their relationships reflect developmental structure
+
+**Lineage tracing validation:**
+
+The model can be **validated against experimental lineage tracing**:
+
+- **Clonal analysis**: Compare inferred branch assignments to known clonal relationships
+- **Barcode sequencing**: Use genetic barcodes to validate branch structure
+- **Time-course experiments**: Compare inferred temporal coordinates to known sampling times
+
+**Perturbation experiments:**
+
+The model enables **quantitative analysis of developmental perturbations**:
+
+- **Knockout effects**: Compare branch structure and gene expression patterns in wild-type vs. knockout
+- **Drug treatments**: Assess how chemical perturbations alter developmental trajectories
+- **Environmental stress**: Model how external conditions affect lineage commitment and timing
+
+#### **Connection to Established Developmental Biology Concepts**
+
+**Waddington's epigenetic landscape:**
+
+The branching model provides a **quantitative implementation** of Waddington's landscape metaphor:
+
+- **Branches as valleys**: Each developmental trajectory corresponds to a valley in the landscape
+- **Branch probabilities as basin sizes**: $\pi_k$ reflects the relative attractiveness of each trajectory
+- **Temporal coordinates as progress**: ${t^*}_j$ represents progress down the developmental valley
+- **Gene expression as landscape features**: Branch-specific parameters define the molecular landscape
+
+**Canalization and developmental constraints:**
+
+The model captures **developmental canalization** through parameter structure:
+
+- **Shared degradation rates**: Reflect conserved post-transcriptional machinery
+- **Branch-specific timing**: Captures lineage-specific developmental programs
+- **Hierarchical priors**: Represent evolutionary constraints on developmental variation
+
+**Developmental plasticity and robustness:**
+
+The mixture structure enables modeling of **developmental plasticity**:
+
+- **Branch assignment uncertainty**: Reflects cellular plasticity and fate choice flexibility
+- **Parameter variation**: Captures individual cell variation within developmental programs
+- **Environmental responsiveness**: Branch probabilities can depend on external conditions
+
+#### **Clinical and Therapeutic Applications**
+
+**Disease modeling:**
+
+The framework can model **pathological developmental processes**:
+
+- **Cancer progression**: Tumor cells following different evolutionary trajectories
+- **Developmental disorders**: Disrupted lineage specification and timing
+- **Aging**: Age-related changes in stem cell differentiation capacity
+
+**Drug discovery:**
+
+The model enables **quantitative assessment of therapeutic interventions**:
+
+- **Target identification**: Genes with branch-specific expression patterns as therapeutic targets
+- **Timing optimization**: Optimal timing of interventions based on developmental stage
+- **Combination therapy**: Targeting multiple branches simultaneously
+
+**Regenerative medicine:**
+
+The framework supports **rational design of regenerative therapies**:
+
+- **Directed differentiation**: Optimizing protocols to bias cells toward desired branches
+- **Reprogramming efficiency**: Quantifying success of cellular reprogramming approaches
+- **Tissue engineering**: Designing scaffolds that support specific developmental trajectories
+
+### Computational Considerations
+
+The branching mixture model maintains **computational efficiency** while significantly expanding modeling capabilities through careful algorithmic design and mathematical structure preservation.
+
+#### **Scalability Analysis**
+
+**Time complexity scaling:**
+
+\begin{align}
+\text{Single-branch model:} &\quad O(N \log N + NG) \\
+\text{Branching mixture model:} &\quad O(K \cdot N \log N + KNG + NK)
+\end{align}
+
+where:
+
+- $N$ = number of cells
+- $G$ = number of genes
+- $K$ = number of branches
+
+**Asymptotic behavior:**
+
+- **Linear in $K$**: Computational cost scales linearly with number of branches
+- **No interaction terms**: $K$, $N$, and $G$ scaling factors are multiplicative, not exponential
+- **Practical bounds**: For typical values ($K \leq 10$, $N \leq 10^4$, $G \leq 10^3$), computational cost remains manageable
+
+**Memory complexity:**
+\begin{align}
+\text{Additional memory:} &\quad O(KG + KN + K) \\
+&= O(K(G + N + 1))
+\end{align}
+
+**Storage breakdown:**
+
+- **Branch-gene parameters**: $O(KG)$ for ${t^*_{on}}_{ik}$, $\delta^*_{ik}$, $R_{on,ik}$
+- **Branch assignments**: $O(N)$ for $Z_j$ values
+- **Branch-specific temporal coordinates**: $O(KN)$ for stick-breaking variables
+- **Branch probabilities**: $O(K)$ for $\boldsymbol{\pi}$
+
+#### **Algorithmic Optimizations**
+
+**Parallel computation strategies:**
+
+1. **Branch-parallel likelihood evaluation**:
+   ```python
+   # Compute likelihood for each branch in parallel
+   branch_likelihoods = parallel_map(
+       lambda k: compute_branch_likelihood(data, branch_k_params),
+       branches
+   )
+   ```
+
+2. **Gene-parallel parameter updates**:
+   ```python
+   # Update gene parameters across branches in parallel
+   for k in branches:
+       gene_params[k] = parallel_map(
+           lambda i: update_gene_params(gene_i, branch_k),
+           genes
+       )
+   ```
+
+3. **Cell-parallel branch assignment updates**:
+   ```python
+   # Update branch assignments for cells in parallel
+   new_assignments = parallel_map(
+       lambda j: sample_branch_assignment(cell_j, branch_likelihoods),
+       cells
+   )
+   ```
+
+**Efficient MCMC sampling:**
+
+**Block updates for branch structure:**
+
+- **Joint branch probability update**: Sample $\boldsymbol{\pi}$ using conjugate Dirichlet posterior
+- **Efficient branch assignment**: Use categorical sampling with precomputed likelihoods
+- **Adaptive branch proposals**: Propose branch reassignments based on likelihood ratios
+
+**Hierarchical parameter updates:**
+
+- **Shared parameter pooling**: Update $\gamma^*_i$ and $U_{0i}$ using data from all branches
+- **Branch-specific parameter blocks**: Update all parameters for branch $k$ jointly
+- **Hyperparameter updates**: Use conjugate priors where possible for efficient sampling
+
+**Stick-breaking optimizations:**
+
+- **Incremental updates**: Only recompute stick-breaking for branches with changed membership
+- **Cached computations**: Store intermediate stick-breaking results to avoid recomputation
+- **Vectorized operations**: Use efficient linear algebra for Beta variable sampling
+
+#### **Implementation Considerations**
+
+**Numerical stability:**
+
+**Beta variable sampling:**
+```python
+# Ensure numerical stability in stick-breaking
+def sample_beta_variables(branch_size):
+    # Avoid extreme values that cause numerical issues
+    alpha, beta = 1.0, max(branch_size - 1, 1.0)
+    return np.clip(np.random.beta(alpha, beta), 1e-8, 1-1e-8)
+```
+
+**Likelihood computation:**
+```python
+# Use log-space arithmetic to prevent overflow/underflow
+def compute_mixture_likelihood(data, branch_params, branch_probs):
+    log_likelihoods = []
+    for k, (params, prob) in enumerate(zip(branch_params, branch_probs)):
+        log_lik = compute_log_likelihood(data, params) + np.log(prob)
+        log_likelihoods.append(log_lik)
+    return logsumexp(log_likelihoods)
+```
+
+**Memory management:**
+
+**Efficient data structures:**
+- **Sparse representations**: Use sparse matrices for gene expression data
+- **Branch indexing**: Maintain efficient mappings from cells to branches
+- **Parameter caching**: Cache frequently accessed parameter combinations
+
+**Memory-efficient sampling:**
+```python
+# Process branches sequentially to limit memory usage
+def memory_efficient_sampling(data, max_memory_per_branch):
+    for branch_k in branches:
+        # Load only data for current branch
+        branch_data = data[branch_assignments == k]
+        # Process branch with limited memory footprint
+        updated_params = sample_branch_parameters(branch_data)
+        # Store results and free memory
+        store_and_cleanup(updated_params)
+```
+
+#### **Validation and Diagnostics**
+
+**Convergence diagnostics:**
+
+**Branch-specific convergence:**
+
+- **Within-branch parameter convergence**: Monitor $\hat{R}$ statistics for each branch separately
+- **Cross-branch parameter comparison**: Ensure branch-specific parameters are distinguishable
+- **Branch assignment stability**: Track branch assignment changes across MCMC iterations
+
+**Model selection criteria:**
+
+**Information criteria for branch number:**
+\begin{align}
+\text{WAIC}_K &= -2 \sum_{j=1}^N \log \mathbb{E}[p(y_j \mid \theta)] + 2 p_{\text{eff}} \\
+\text{LOO-CV}_K &= \sum_{j=1}^N \log p(y_j \mid y_{-j})
+\end{align}
+
+**Cross-validation strategies:**
+
+- **Leave-one-branch-out**: Assess model performance when excluding entire branches
+- **Temporal cross-validation**: Validate on held-out time points within each branch
+- **Gene-wise validation**: Test parameter recovery for held-out genes
+
+**Parameter recovery validation:**
+
+**Branch-specific validation metrics:**
+
+- **Branch assignment accuracy**: Compare inferred vs. true branch assignments
+- **Within-branch parameter recovery**: Assess parameter recovery within each branch
+- **Cross-branch parameter discrimination**: Verify that branch-specific parameters are distinguishable
+
+**Computational benchmarking:**
+```python
+# Benchmark computational performance
+def benchmark_branching_model(n_cells, n_genes, n_branches):
+    start_time = time.time()
+
+    # Generate synthetic data
+    data = generate_branching_data(n_cells, n_genes, n_branches)
+
+    # Fit model
+    model = BranchingPyroVelocityModel()
+    results = model.fit(data, num_samples=1000)
+
+    # Record performance metrics
+    total_time = time.time() - start_time
+    memory_usage = get_peak_memory_usage()
+
+    return {
+        'total_time': total_time,
+        'memory_usage': memory_usage,
+        'time_per_sample': total_time / 1000,
+        'scalability_metrics': compute_scalability_metrics(results)
+    }
+```
+
+This comprehensive computational framework ensures that the branching mixture model remains **practically implementable** while providing **rigorous mathematical foundations** for modeling complex developmental processes with multiple trajectories.

### 2025-06-14 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): improve branching mixture model description
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+**Hyperpriors for branch structure:**
+\begin{align}
+\alpha_K &\sim \text{Gamma}(2.0, 1.0) \label{eq-alpha-K-hyperprior} \\
+\alpha_\pi &\sim \text{Gamma}(1.0, 1.0) \label{eq-alpha-pi-hyperprior}
+\end{align}
+
 **Branch structure:**
@@
 
+**Hyperpriors for temporal parameters:**
+\begin{align}
+\alpha_T &\sim \text{Gamma}(5.0, 1.0) \label{eq-alpha-T-hyperprior} \\
+\beta_T &\sim \text{Gamma}(1.0, 1.0) \label{eq-beta-T-hyperprior}
+\end{align}
+
 **Branch-specific temporal coordinates:**
 \begin{align}
-T_{\max,k} &\sim \text{Gamma}(\alpha_T, \beta_T) \quad \text{for } k \in \{1, \ldots, K\} \label{eq-branch-duration} \\
-{t^*}_j \mid Z_j = k &= \text{stick\_breaking}(0, T_{\max,k}) \label{eq-branch-stick-breaking}
-\end{align}
-
-Each branch $k$ has its own process duration $T_{\max,k}$ and stick-breaking temporal coordinate system with fixed anchors at $0$ and $T_{\max,k}$.
+T_{\max,k} &\sim \text{Gamma}(\alpha_T, \beta_T) \quad \text{for } k \in \{1, \ldots, K\} \label{eq-branch-duration}
+\end{align}
+
+**Rigorous stick-breaking specification within branches:**
+
+For each branch $k \in \{1, \ldots, K\}$, let $\mathcal{C}_k = \{j : Z_j = k\}$ denote the set of cells assigned to branch $k$, and let $\mathcal{N}_k = |\mathcal{C}_k|$ be the number of cells in branch $k$.
+Order the cells within branch $k$ as $j_{k,1}, j_{k,2}, \ldots, j_{k,\mathcal{N}_k}$ (ordering can be arbitrary).
+The stick-breaking process for branch $k$ is defined as:
+
+\begin{align}
+V_{k,\ell} &\sim \text{Beta}(1, \mathcal{N}_k - \ell) \quad \text{for } \ell \in \{1, \ldots, \mathcal{N}_k - 1\} \label{eq-branch-beta-variables-rigorous} \\
+\tilde{t}_{k,1} &= 0 \label{eq-branch-first-time} \\
+\tilde{t}_{k,\ell} &= \tilde{t}_{k,\ell-1} + V_{k,\ell-1} \prod_{m=1}^{\ell-2} (1 - V_{k,m}) \quad \text{for } \ell \in \{2, \ldots, \mathcal{N}_k\} \label{eq-branch-stick-construction-rigorous} \\
+\tilde{t}_{k,\mathcal{N}_k} &= 1 \quad \text{(normalization constraint)} \label{eq-branch-last-time} \\
+{t^*}_{j_{k,\ell}} &= T_{\max,k} \times \tilde{t}_{k,\ell} \quad \text{for } \ell \in \{1, \ldots, \mathcal{N}_k\} \label{eq-branch-time-scaling-rigorous}
+\end{align}
+
+This construction ensures that for each branch $k$:
+\begin{align}
+0 = {t^*}_{j_{k,1}} < {t^*}_{j_{k,2}} < \cdots < {t^*}_{j_{k,\mathcal{N}_k}} = T_{\max,k} \label{eq-branch-ordering}
+\end{align}
+
+**Hyperpriors for shared gene parameters:**
+\begin{align}
+\mu_{\gamma^*} &\sim \text{Normal}(0.0, 1.0^2) \label{eq-mu-gamma-hyperprior} \\
+\sigma_{\gamma^*}^2 &\sim \text{InverseGamma}(2.0, 0.5^2) \label{eq-sigma-gamma-hyperprior} \\
+\mu_{U_0} &\sim \text{Normal}(4.0, 1.0^2) \label{eq-mu-U0-hyperprior} \\
+\sigma_{U_0}^2 &\sim \text{InverseGamma}(2.0, 0.5^2) \label{eq-sigma-U0-hyperprior} \\
+\mu_{\lambda} &\sim \text{Normal}(0.0, 1.0^2) \label{eq-mu-lambda-hyperprior} \\
+\sigma_{\lambda}^2 &\sim \text{InverseGamma}(2.0, 0.5^2) \label{eq-sigma-lambda-hyperprior}
+\end{align}
 
@@
 \begin{align}
-\gamma^*_i &\sim \text{LogNormal}(\mu_{\gamma^*}, \sigma_{\gamma^*}^2) \label{eq-shared-gamma} \\
-U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2) \label{eq-shared-U0}
+\gamma^*_i &\sim \text{LogNormal}(\mu_{\gamma^*}, \sigma_{\gamma^*}^2) \quad \text{for } i \in \{1, \ldots, G\} \label{eq-shared-gamma} \\
+U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2) \quad \text{for } i \in \{1, \ldots, G\} \label{eq-shared-U0}
+\end{align}
+
+**Hyperpriors for branch-specific gene parameters:**
+\begin{align}
+\mu_{on,k} &\sim \text{Normal}(2.5, 1.5^2) \quad \text{for } k \in \{1, \ldots, K\} \label{eq-mu-on-k-hyperprior} \\
+\sigma_{on}^2 &\sim \text{InverseGamma}(2.0, 1.0^2) \label{eq-sigma-on-hyperprior} \\
+\mu_{\delta,k} &\sim \text{Normal}(0.0, 0.5^2) \quad \text{for } k \in \{1, \ldots, K\} \label{eq-mu-delta-k-hyperprior} \\
+\sigma_{\delta}^2 &\sim \text{InverseGamma}(2.0, 0.25^2) \label{eq-sigma-delta-hyperprior} \\
+\mu_{R,k} &\sim \text{Normal}(\log(2.5), 0.4^2) \quad \text{for } k \in \{1, \ldots, K\} \label{eq-mu-R-k-hyperprior} \\
+\sigma_R^2 &\sim \text{InverseGamma}(2.0, 0.2^2) \label{eq-sigma-R-hyperprior}
 \end{align}
@@
 {t^*_{on}}_{ik} &\sim \text{Normal}(\mu_{on,k}, \sigma_{on}^2) \quad \text{for } i \in \{1, \ldots, G\}, k \in \{1, \ldots, K\} \label{eq-branch-onset} \\
-\delta^*_{ik} &\sim \text{LogNormal}(\mu_{\delta,k}, \sigma_{\delta}^2) \label{eq-branch-duration-gene} \\
-R_{on,ik} &\sim \text{LogNormal}(\mu_{R,k}, \sigma_R^2) \label{eq-branch-fold-change}
+\delta^*_{ik} &\sim \text{LogNormal}(\mu_{\delta,k}, \sigma_{\delta}^2) \quad \text{for } i \in \{1, \ldots, G\}, k \in \{1, \ldots, K\} \label{eq-branch-duration-gene} \\
+R_{on,ik} &\sim \text{LogNormal}(\mu_{R,k}, \sigma_R^2) \quad \text{for } i \in \{1, \ldots, G\}, k \in \{1, \ldots, K\} \label{eq-branch-fold-change}
 \end{align}
@@
 \begin{align}
-\lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2) \label{eq-branch-lambda}
+\lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-branch-lambda}
 \end{align}
@@
 
-For cell $j$ assigned to branch $k$ (i.e., $Z_j = k$), the dimensionless concentrations follow the same piecewise analytical solutions as the single-branch model, but using branch-specific parameters:
+For cell $j$ assigned to branch $k$ (i.e., $Z_j = k$), the dimensionless transcription rate for gene $i$ is:
 
@@
 
-The analytical solutions for each phase remain identical to the single-branch case, but now use branch-gene-specific parameters ${t^*_{on}}_{ik}$, $\delta^*_{ik}$, and $R_{on,ik}$.
-
-**Mixture likelihood:**
-\begin{align}
-\mathcal{L}(\text{data}) &= \prod_{j=1}^N \sum_{k=1}^K \pi_k \cdot \mathcal{L}_k(\text{data}_j \mid \text{branch } k) \label{eq-mixture-likelihood}
-\end{align}
-
-where $\mathcal{L}_k(\text{data}_j \mid \text{branch } k)$ is the likelihood for cell $j$ under branch $k$ using branch-specific parameters.
-
-**Observation model (unchanged):**
-\begin{align}
-u_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}) \label{eq-branch-u-obs} \\
-s_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij}) \label{eq-branch-s-obs}
-\end{align}
-
-where $u^*_{ij}$ and $s^*_{ij}$ are computed using the branch-specific piecewise analytical solutions for the branch to which cell $j$ is assigned.
+The dimensionless concentrations $u^*_{ij}$ and $s^*_{ij}$ are computed using the same piecewise analytical solutions as the single-branch model, but with branch-gene-specific parameters ${t^*_{on}}_{ik}$, $\delta^*_{ik}$, and $R_{on,ik}$, and cell-specific temporal coordinate ${t^*}_j$ determined by the stick-breaking process for branch $Z_j$.
+
+**Conditional independence structure:**
+
+Given branch assignments $\{Z_j\}_{j=1}^N$ and all parameters, the observations are conditionally independent:
+\begin{align}
+p(\{u_{ij}, s_{ij}\}_{i,j} \mid \{Z_j\}, \text{all parameters}) &= \prod_{j=1}^N \prod_{i=1}^G p(u_{ij}, s_{ij} \mid Z_j, \text{parameters}) \label{eq-conditional-independence}
+\end{align}
+
+**Observation model:**
+\begin{align}
+u_{ij} \mid Z_j = k &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}(k)) \label{eq-branch-u-obs} \\
+s_{ij} \mid Z_j = k &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij}(k)) \label{eq-branch-s-obs}
+\end{align}
+
+where $u^*_{ij}(k)$ and $s^*_{ij}(k)$ denote the dimensionless concentrations computed using branch $k$'s parameters and temporal coordinates.
+
+**Marginal likelihood (integrating over branch assignments):**
+\begin{align}
+p(\{u_{ij}, s_{ij}\}_{i,j} \mid \text{parameters}) &= \prod_{j=1}^N \sum_{k=1}^K \pi_k \prod_{i=1}^G p(u_{ij}, s_{ij} \mid Z_j = k, \text{parameters}) \label{eq-marginal-likelihood}
+\end{align}
+
+**Complete data likelihood (given branch assignments):**
+\begin{align}
+p(\{u_{ij}, s_{ij}\}_{i,j}, \{Z_j\}_{j=1}^N \mid \text{parameters}) &= \prod_{j=1}^N \pi_{Z_j} \prod_{i=1}^G p(u_{ij}, s_{ij} \mid Z_j, \text{parameters}) \label{eq-complete-likelihood}
+\end{align}
 
@@
 
-For each branch $k \in \{1, \ldots, K\}$, cells assigned to that branch ($\{j : Z_j = k\}$) follow an independent stick-breaking process:
-
-\begin{align}
-\text{Let } \mathcal{N}_k &= |\{j : Z_j = k\}| \quad \text{(number of cells in branch } k\text{)} \\
-\text{For cells } j_1, j_2, \ldots, j_{\mathcal{N}_k} &\text{ where } Z_{j_\ell} = k: \\
-V_{k,\ell} &\sim \text{Beta}(1, \mathcal{N}_k - \ell) \quad \text{for } \ell \in \{1, \ldots, \mathcal{N}_k - 1\} \label{eq-branch-beta-variables} \\
-\tilde{t}_{k,1} &= 0 \quad \text{(first cell at branch start)} \\
-\tilde{t}_{k,\ell} &= \tilde{t}_{k,\ell-1} + V_{k,\ell-1} \prod_{m=1}^{\ell-2} (1 - V_{k,m}) \quad \text{for } \ell \in \{2, \ldots, \mathcal{N}_k\} \label{eq-branch-stick-construction} \\
-{t^*}_{j_\ell} &= T_{\max,k} \times \tilde{t}_{k,\ell} \quad \text{(branch-specific scaling)} \label{eq-branch-time-scaling}
-\end{align}
+The rigorous stick-breaking specification is provided in the Mathematical Description section above (equations \eqref{eq-branch-beta-variables-rigorous}-\eqref{eq-branch-ordering}).
+Each branch $k$ implements an independent stick-breaking process with the following key properties:
+
+- **Independent Beta variables**: $V_{k,\ell} \sim \text{Beta}(1, \mathcal{N}_k - \ell)$ for each position $\ell$ within branch $k$
+- **Recursive construction**: Temporal coordinates built sequentially using stick-breaking formula
+- **Fixed anchors**: First cell at $t^* = 0$, last cell at $t^* = T_{\max,k}$
+- **Proper ordering**: Ensures $0 = {t^*}_{j_{k,1}} < {t^*}_{j_{k,2}} < \cdots < {t^*}_{j_{k,\mathcal{N}_k}} = T_{\max,k}$
 
@@
 **Ordering preservation within branches:**
-\begin{align}
-0 = {t^*}_{j_1} < {t^*}_{j_2} < \cdots < {t^*}_{j_{\mathcal{N}_k}} = T_{\max,k}
+
+From equation \eqref{eq-branch-ordering}, each branch maintains strict temporal ordering:
+\begin{align}
+0 = {t^*}_{j_{k,1}} < {t^*}_{j_{k,2}} < \cdots < {t^*}_{j_{k,\mathcal{N}_k}} = T_{\max,k} \quad \text{for all } k \in \{1, \ldots, K\} \label{eq-branch-ordering-property}
 \end{align}
@@
 **Independence across branches:**
-\begin{align}
-\{V_{k,\ell}\}_{\ell=1}^{\mathcal{N}_k-1} \perp \{V_{k',\ell'}\}_{\ell'=1}^{\mathcal{N}_{k'}-1} \quad \text{for } k \neq k'
+
+The Beta variables for different branches are mutually independent:
+\begin{align}
+\{V_{k,\ell}\}_{\ell=1}^{\mathcal{N}_k-1} \perp \{V_{k',\ell'}\}_{\ell'=1}^{\mathcal{N}_{k'}-1} \quad \text{for } k \neq k' \label{eq-branch-independence}
+\end{align}
+
+**Conditional independence of temporal coordinates:**
+
+Given branch assignments $\{Z_j\}_{j=1}^N$ and process durations $\{T_{\max,k}\}_{k=1}^K$, the temporal coordinates are conditionally independent across branches:
+\begin{align}
+p(\{{t^*}_j\}_{j=1}^N \mid \{Z_j\}, \{T_{\max,k}\}) &= \prod_{k=1}^K p(\{{t^*}_j : j \in \mathcal{C}_k\} \mid T_{\max,k}) \label{eq-conditional-independence-time}
 \end{align}
@@
 Each branch maintains the same identifiability properties as the single-branch model:
-- Fixed anchors: ${t^*}_{j_1} = 0$ and ${t^*}_{j_{\mathcal{N}_k}} = T_{\max,k}$
-- No scaling symmetry within branch due to fixed endpoints
-- Gene parameters ${t^*_{on}}_{ik}$, $\delta^*_{ik}$ are independent of $T_{\max,k}$
+- **Fixed anchors**: ${t^*}_{j_{k,1}} = 0$ and ${t^*}_{j_{k,\mathcal{N}_k}} = T_{\max,k}$ eliminate scaling symmetry
+- **Independent gene parameters**: ${t^*_{on}}_{ik}$, $\delta^*_{ik}$ are sampled independently of $T_{\max,k}$ (equations \eqref{eq-branch-onset}-\eqref{eq-branch-fold-change})
+- **Analytical tractability**: Same piecewise solutions apply within each branch
 
@@
 
-During MCMC sampling, branch assignments $Z_j$ are updated using standard categorical sampling:
-\begin{align}
-P(Z_j = k \mid \text{rest}) \propto \pi_k \cdot \mathcal{L}_k(\text{data}_j \mid \text{branch } k)
-\end{align}
+During MCMC sampling, branch assignments $Z_j$ are updated using the conditional posterior distribution:
+\begin{align}
+P(Z_j = k \mid \{u_{ij}, s_{ij}\}_{i=1}^G, \text{all other parameters}) &\propto \pi_k \prod_{i=1}^G p(u_{ij}, s_{ij} \mid Z_j = k, \text{parameters}) \label{eq-branch-assignment-posterior}
+\end{align}
+
+where the likelihood term is computed using branch $k$'s temporal coordinates and gene-specific parameters as specified in equations \eqref{eq-branch-u-obs}-\eqref{eq-branch-s-obs}.
 
@@
 
-1. **Branch-parallel likelihood evaluation**:
-   ```python
-   # Compute likelihood for each branch in parallel
-   branch_likelihoods = parallel_map(
-       lambda k: compute_branch_likelihood(data, branch_k_params),
-       branches
-   )
-   ```
-
-2. **Gene-parallel parameter updates**:
-   ```python
-   # Update gene parameters across branches in parallel
-   for k in branches:
-       gene_params[k] = parallel_map(
-           lambda i: update_gene_params(gene_i, branch_k),
-           genes
-       )
-   ```
-
-3. **Cell-parallel branch assignment updates**:
-   ```python
-   # Update branch assignments for cells in parallel
-   new_assignments = parallel_map(
-       lambda j: sample_branch_assignment(cell_j, branch_likelihoods),
-       cells
-   )
-   ```
+1.
+**Branch-parallel likelihood evaluation**: The marginal likelihood in equation \eqref{eq-marginal-likelihood} can be computed in parallel across branches since each branch's contribution is independent given the parameters.
+
+2.
+**Gene-parallel parameter updates**: Branch-gene-specific parameters (equations \eqref{eq-branch-onset}-\eqref{eq-branch-fold-change}) can be updated in parallel across genes within each branch due to conditional independence.
+
+3.
+**Cell-parallel branch assignment updates**: The posterior probabilities in equation \eqref{eq-branch-assignment-posterior} can be computed in parallel across cells since assignments are conditionally independent given all other parameters.
 
@@
 
-**Beta variable sampling:**
-```python
-# Ensure numerical stability in stick-breaking
-def sample_beta_variables(branch_size):
-    # Avoid extreme values that cause numerical issues
-    alpha, beta = 1.0, max(branch_size - 1, 1.0)
-    return np.clip(np.random.beta(alpha, beta), 1e-8, 1-1e-8)
-```
-
-**Likelihood computation:**
-```python
-# Use log-space arithmetic to prevent overflow/underflow
-def compute_mixture_likelihood(data, branch_params, branch_probs):
-    log_likelihoods = []
-    for k, (params, prob) in enumerate(zip(branch_params, branch_probs)):
-        log_lik = compute_log_likelihood(data, params) + np.log(prob)
-        log_likelihoods.append(log_lik)
-    return logsumexp(log_likelihoods)
-```
+**Beta variable sampling**: The Beta variables $V_{k,\ell} \sim \text{Beta}(1, \mathcal{N}_k - \ell)$ in equation \eqref{eq-branch-beta-variables-rigorous} should be bounded away from 0 and 1 to prevent degenerate stick-breaking constructions.
+
+**Likelihood computation**: The marginal likelihood in equation \eqref{eq-marginal-likelihood} should be computed in log-space to prevent numerical overflow/underflow, particularly when $K$ is large or when branch probabilities $\pi_k$ are small.
 
@@
 **Efficient data structures:**
-- **Sparse representations**: Use sparse matrices for gene expression data
-- **Branch indexing**: Maintain efficient mappings from cells to branches
-- **Parameter caching**: Cache frequently accessed parameter combinations
-
-**Memory-efficient sampling:**
-```python
-# Process branches sequentially to limit memory usage
-def memory_efficient_sampling(data, max_memory_per_branch):
-    for branch_k in branches:
-        # Load only data for current branch
-        branch_data = data[branch_assignments == k]
-        # Process branch with limited memory footprint
-        updated_params = sample_branch_parameters(branch_data)
-        # Store results and free memory
-        store_and_cleanup(updated_params)
-```
+- **Sparse representations**: Gene expression matrices can be stored sparsely since many genes may have zero counts
+- **Branch indexing**: Maintain efficient mappings between cells and their branch assignments $Z_j$
+- **Parameter caching**: Cache frequently accessed combinations of branch-gene parameters
+
+**Memory-efficient processing**: For large datasets, branches can be processed sequentially rather than simultaneously to limit memory usage, taking advantage of the conditional independence structure in equations \eqref{eq-conditional-independence} and \eqref{eq-conditional-independence-time}.
 
@@
 **Computational benchmarking:**
-```python
-# Benchmark computational performance
-def benchmark_branching_model(n_cells, n_genes, n_branches):
-    start_time = time.time()
-
-    # Generate synthetic data
-    data = generate_branching_data(n_cells, n_genes, n_branches)
-
-    # Fit model
-    model = BranchingPyroVelocityModel()
-    results = model.fit(data, num_samples=1000)
-
-    # Record performance metrics
-    total_time = time.time() - start_time
-    memory_usage = get_peak_memory_usage()
-
-    return {
-        'total_time': total_time,
-        'memory_usage': memory_usage,
-        'time_per_sample': total_time / 1000,
-        'scalability_metrics': compute_scalability_metrics(results)
-    }
-```
+
+Performance evaluation should focus on the scaling properties derived from the mathematical structure:
+- **Time complexity**: $O(K \cdot N \log N + KNG + NK)$ as specified in equations \eqref{eq-branch-beta-variables-rigorous}-\eqref{eq-branch-time-scaling-rigorous}
+- **Memory complexity**: $O(K(G + N + 1))$ for storing branch-specific parameters and assignments
+- **Convergence assessment**: Monitor convergence of branch assignments $Z_j$ and branch-specific parameters using standard MCMC diagnostics
 

### 2025-06-14 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): fix branching mixture model description
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 \tilde{t}_{k,1} &= 0 \label{eq-branch-first-time} \\
-\tilde{t}_{k,\ell} &= \tilde{t}_{k,\ell-1} + V_{k,\ell-1} \prod_{m=1}^{\ell-2} (1 - V_{k,m}) \quad \text{for } \ell \in \{2, \ldots, \mathcal{N}_k\} \label{eq-branch-stick-construction-rigorous} \\
+\tilde{t}_{k,\ell} &= \tilde{t}_{k,\ell-1} + V_{k,\ell-1} \times (1 - \tilde{t}_{k,\ell-1}) \quad \text{for } \ell \in \{2, \ldots, \mathcal{N}_k-1\} \label{eq-branch-stick-construction-rigorous} \\
 \tilde{t}_{k,\mathcal{N}_k} &= 1 \quad \text{(normalization constraint)} \label{eq-branch-last-time} \\

### 2025-06-14 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): justify introduction of stick-breaking procedure
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 **Proof:**
+
 1. **Fixed anchor constraint**: The scaled system requires $t^*_1 = \lambda \cdot 0 = 0$ and $t^*_N = \lambda T_{\max}$
@@
 Therefore, the likelihood is not invariant under parameter scaling, eliminating the scaling symmetry. □
+
+#### **Mathematical and Biological Foundations of Stick-Breaking for Temporal Coordinates**
+
+The stick-breaking construction for temporal coordinates addresses fundamental challenges in modeling cellular developmental progression and provides the mathematically natural solution to the constrained latent temporal assignment problem.
+
+**1. The Fundamental Problem Being Solved**
+
+**Physical vs. Biological Time Mismatch:**
+
+Single-cell RNA sequencing captures all cells simultaneously at a fixed experimental time $t_{\text{experiment}}$, yet each cell exists at a different stage of the underlying biological process.
+This creates a fundamental mismatch:
+
+- **Physical observation**: All cells measured at $t_{\text{experiment}}$
+- **Biological reality**: Each cell $j$ at developmental stage $t^*_j$ within process duration $[0, T_{\max}]$
+
+**Constrained Latent Temporal Assignment Problem:**
+
+The inference challenge becomes assigning temporal coordinates $\{t^*_j\}_{j=1}^N$ that satisfy multiple biological constraints:
+
+\begin{align}
+\mathcal{C}_{\text{boundary}} &: t^*_1 = 0, \quad t^*_N = T_{\max} \quad \text{(process boundaries)} \\
+\mathcal{C}_{\text{ordering}} &: t^*_1 \leq t^*_2 \leq \cdots \leq t^*_N \quad \text{(developmental progression)} \\
+\mathcal{C}_{\text{continuity}} &: \|x_i - x_j\| \text{ small} \Rightarrow |t^*_i - t^*_j| \text{ small} \quad \text{(state-time proximity)}
+\end{align}
+
+where $x_i, x_j$ represent gene expression states and $\mathcal{C}_{\text{continuity}}$ enforces the differential equation requirement that cells close in expression space should be close in developmental time.
+
+**2. Mathematical Object Definition**
+
+**Constrained Order Statistics Space:**
+
+Stick-breaking provides a probability distribution over the constrained order statistics space:
+
+$$\mathcal{T}_N^{T_{\max}} = \{(t^*_1, t^*_2, \ldots, t^*_N) \in \mathbb{R}^N : 0 = t^*_1 \leq t^*_2 \leq \cdots \leq t^*_N = T_{\max}\}$$
+
+This space is isomorphic to the $(N-2)$-simplex with fixed boundary conditions, enabling efficient sampling and probability computation.
+
+**Probability Density with Jacobian Transformation:**
+
+The stick-breaking construction transforms independent Beta random variables to ordered temporal coordinates:
+
+\begin{align}
+\xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
+t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{(recursive construction)}
+\end{align}
+
+The joint density on $\mathcal{T}_N^{T_{\max}}$ becomes:
+
+$$p(t^*_2, \ldots, t^*_{N-1} | T_{\max}) = \frac{(N-2)!}{T_{\max}^{N-2}} \prod_{j=2}^{N-1} (N-j) \quad \text{for } (t^*_1, \ldots, t^*_N) \in \mathcal{T}_N^{T_{\max}}$$
+
+This uniform density over the constrained space reflects maximum entropy subject to biological constraints.
+
+**3. State Space Proximity Preservation**
+
+**Ordering Constraint and Biological Continuity:**
+
+The ordering constraint $t^*_1 \leq t^*_2 \leq \cdots \leq t^*_N$ directly enforces the biological assumption that cells close in expression state should be close in developmental time.
+This constraint ensures that the temporal manifold respects the underlying differential equation dynamics.
+
+**Variable Spacing Property:**
+
+The stick-breaking construction exhibits adaptive spacing that accommodates heterogeneous cell density:
+
+$$\mathbb{E}[t^*_{j+1} - t^*_j | T_{\max}] = \frac{T_{\max}}{N+1} \cdot \frac{1}{N-j}$$
+
+This variable spacing allows:
+- **Dense regions**: Smaller temporal intervals where many cells cluster in expression space
+- **Sparse regions**: Larger temporal intervals where few cells exist
+- **Adaptive resolution**: Automatic adjustment to local cell density
+
+**State Space Distance Integration:**
+
+The Beta parameters can be adapted to incorporate state space distance information:
+
+$$\xi_j \sim \text{Beta}(\alpha_j, \beta_j) \quad \text{where } \alpha_j, \beta_j \propto f(\|x_{j+1} - x_j\|)$$
+
+enabling data-driven temporal spacing that respects expression similarity.
+
+**4. Information-Theoretic Optimality**
+
+**Maximum Entropy Principle:**
+
+Stick-breaking with uniform Beta$(1, N-j)$ parameters is the **maximum entropy distribution** subject to:
+- Fixed endpoints: $t^*_1 = 0$, $t^*_N = T_{\max}$
+- Ordering constraints: $t^*_j \leq t^*_{j+1}$
+- Exchangeability: Symmetric treatment of intermediate cells
+
+**Minimal Assumptions:**
+
+This maximum entropy property ensures **minimal assumptions** beyond biological requirements:
+- No artificial temporal clustering
+- No preferred developmental stages
+- Maximal uncertainty preservation where data provides no constraints
+
+**Information Preservation:**
+
+The construction preserves **maximal information** about temporal relationships by avoiding arbitrary distributional assumptions that could bias inference toward specific temporal patterns.
+
+**5. Why Alternative Approaches Fail**
+
+**Random Temporal Assignment:**
+
+Random assignment of temporal coordinates $t^*_j \sim \text{Uniform}(0, T_{\max})$ violates differential equation continuity assumptions:
+- Cells with similar expression states could be assigned vastly different developmental times
+- Discontinuous jumps in the temporal manifold create unrealistic dynamics
+- Loss of biological interpretability for developmental progression
+
+**Truncated/Constrained Distributions:**
+
+Direct sampling from truncated distributions over ordered statistics creates computational problems:
+- Complex normalization constants that scale exponentially with $N$
+- Inefficient rejection sampling with low acceptance rates
+- Numerical instability for large cell populations
+
+**Uniform Distributions:**
+
+Uniform spacing $t^*_j = j \cdot T_{\max}/N$ truncates the dynamical range needed for meaningful exponential convergence:
+- Fixed spacing cannot adapt to heterogeneous cell density
+- Insufficient resolution in regions of rapid expression change
+- Loss of stochastic modeling of temporal uncertainty
+
+**6. Connection to Differential Equation Framework**
+
+**Bridging Observation and Inference:**
+
+Stick-breaking provides the essential bridge for inverting the observation process:
+
+\begin{align}
+\text{Differential Equation:} \quad &t^* \rightarrow \text{Gene Expression} \rightarrow \text{Counts} \\
+\text{Inference Problem:} \quad &\text{Counts} \rightarrow \text{Gene Expression} \rightarrow t^*
+\end{align}
+
+The stick-breaking construction enables tractable Bayesian inference over the biologically meaningful temporal manifold by:
+
+**Preserving Dynamical Structure:**
+
+The ordered constraint ensures that inferred temporal coordinates respect the underlying dynamical system structure, maintaining consistency with the forward differential equation model.
+
+**Enabling Posterior Inference:**
+
+The explicit probability density over $\mathcal{T}_N^{T_{\max}}$ enables efficient MCMC sampling and variational inference for temporal coordinate estimation.
+
+**Maintaining Analytical Tractability:**
+
+The construction preserves the analytical advantages of the piecewise solutions while providing a principled probabilistic framework for temporal uncertainty quantification.
+
+**Validation Connection:**
+
+The effectiveness of this approach is demonstrated in the parameter recovery validation studies (see §\ref{sec-validation-strategy}), where stick-breaking temporal coordinates enable accurate recovery of both kinetic and temporal parameters across diverse biological scenarios.
 

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): simplify temporal prior
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2025-06-14 nbs/concepts/parameter-recovery-validation/index.qmd
@@
 The key innovation is a **dimensionally consistent parameterization** that eliminates parameter redundancy by fixing the reference transcription state, while maintaining biological interpretability through fold-change parameters.
-Additionally, this study implements a **process duration with stick-breaking** temporal framework that eliminates scaling symmetry and achieves parameter identifiability while preserving all analytical and computational advantages.
 
@@
 
-### Process Duration with Stick-Breaking Framework
-
-This study implements a **process duration with stick-breaking** approach that eliminates the scaling symmetry present in hierarchical temporal scaling while maintaining all analytical and biological advantages.
-The key innovation is the use of **fixed anchor points** that break scaling symmetry and achieve parameter identifiability.
-
-#### **Elimination of Scaling Symmetry**
-
-The previous hierarchical scaling approach suffered from fundamental **scaling symmetry** that prevented parameter identifiability:
-
-$$\mathcal{L}(T_M^*, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) = \mathcal{L}(\lambda T_M^*, \{\lambda t^*_j\}, \{\lambda t^*_{0,\text{on}i}\}, \{\lambda \delta^*_i\})$$
-
-for any $\lambda > 0$. This symmetry creates non-identifiable parameters and leads to catastrophic validation failures.
-
-#### **Fixed Anchor Point Solution**
-
-The stick-breaking approach eliminates scaling symmetry through **fixed anchor points**:
-
-\begin{align}
-t^*_1 &= 0 \quad \text{(process start, fixed)} \\
-t^*_N &= T_{\max} \quad \text{(process end, sampled)} \\
-t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{\text{on}}, \sigma_{\text{on}}^2) \quad \text{(independent gene timing)} \\
-\delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(independent gene duration)}
-\end{align}
-
-**Theorem (Elimination of Scaling Symmetry):**
-
-For any scaling factor $\lambda \neq 1$, the likelihood functions are not equal:
-$$\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$$
-
-**Proof:**
-1. **Fixed anchor constraint**: The scaled system requires $t^*_1 = \lambda \cdot 0 = 0$ and $t^*_N = \lambda T_{\max}$
-2. **Process duration independence**: Gene parameters $t^*_{0,\text{on}i}, \delta^*_i$ are sampled independently of $T_{\max}$ and do not scale
-3. **Asymmetric scaling**: Scaling $T_{\max} \rightarrow \lambda T_{\max}$ and $t^*_j \rightarrow \lambda t^*_j$ changes the relative positioning of gene events within the process timeline
-4. **Likelihood dependence**: The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$, which change under asymmetric scaling
-
-Therefore, the likelihood is not invariant under parameter scaling, eliminating the scaling symmetry. □
-
 ### Gene Expression Pattern Classification
 
-This framework enables representation of diverse gene expression patterns through specific parameter combinations.
-Each pattern is defined by quantitative parameter ranges applied to **absolute temporal parameters** within the process duration framework.
-
-**Parameter Framework**: The classification uses absolute temporal parameters that are independent of the process duration $T_{\max}$, enabling robust pattern identification across different experimental timescales.
+This piecewise framework enables representation of diverse gene expression patterns through specific parameter combinations.
+Each pattern is defined by quantitative parameter ranges applied to **independent absolute temporal parameters** that eliminate scaling symmetry while maintaining biological interpretability.
+
+**Independent Parameter Framework**: The classification uses independent absolute temporal parameters that are sampled directly without hierarchical scaling relationships:
+
+\begin{align}
+T_{\max} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
+t^*_j &\sim \text{Uniform}(0, T_{\max}) \quad \text{(cell temporal positions)} \\
+t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
+\delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(absolute activation duration)}
+\end{align}
+
+This independent structure eliminates scaling symmetry by removing multiplicative relationships between temporal parameters, ensuring unique parameter identifiability while maintaining biological interpretability.
 
@@
 
-- $t^*_{0,\text{on}i} < 0.0$ (activation onset before process start)
+- $t^*_{0,\text{on}i} < 0.0$ (absolute activation onset before observation window)
 - $R_{\text{on},i} > 2.0$ (moderate to strong fold-change during activation)
 
-**Biological interpretation**: These genes underwent activation before the biological process under study began ($t^* = 0$).
+**Biological interpretation**: These genes underwent activation before the experimental observation window.
 During the observed period, they exhibit decay-only dynamics as they return toward baseline expression levels.
-The negative onset time indicates that activation occurred before the process boundary.
+The negative absolute onset time indicates that activation occurred in the "past" relative to the observation start.
 
@@
 
-**Characteristics**: Genes that show complete activation-decay cycles within the process duration, exhibiting brief, pulse-like activation followed by return toward baseline.
+**Characteristics**: Genes that show complete activation-decay cycles within the observation window, exhibiting brief, pulse-like activation followed by return toward baseline.
 
@@
 
-- $t^*_{0,\text{on}i} > 0.0$ (activation within process duration)
-- $t^*_{0,\text{on}i} < 2.5$ (activation early enough to complete cycle within process)
-- $\delta^*_i < 2.0$ (short activation duration relative to process)
+- $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ (early activation within process duration)
+- $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$ (short activation duration relative to process)
 - $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)
 
-**Biological interpretation**: These genes activate early enough and for a short enough duration that the complete activation-decay cycle is observable within the process duration.
-The constraints ensure that both the activation and subsequent decay phases are captured within the temporal boundaries.
+**Biological interpretation**: These genes activate early enough and for a short enough duration that the complete activation-decay cycle is observable within the experimental window.
+The constraints ensure that both the activation and subsequent decay phases are captured.
 
@@
 
-**Characteristics**: Genes that become activated early and remain highly expressed throughout most of the process duration, showing net increase over the process timeline.
+**Characteristics**: Genes that become activated early and remain highly expressed throughout most of the observation period, showing net increase over the observation window.
 
@@
 
-- $t^*_{0,\text{on}i} > 0.0$ (activation within process duration)
-- $t^*_{0,\text{on}i} < 1.5$ (early activation onset within process)
-- $\delta^*_i > 2.5$ (long activation duration relative to process)
+- $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ (early activation within process duration)
+- $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$ (long activation duration relative to process)
 - $R_{\text{on},i} > 2.0$ (strong fold-change during activation)
 
-**Biological interpretation**: These genes activate early in the process and maintain high expression levels throughout most of the process duration.
-The long activation duration means that decay back to baseline is either minimal or occurs beyond the process endpoint.
+**Biological interpretation**: These genes activate early in the observation window and maintain high expression levels throughout most of the observed period.
+The long activation duration relative to the observation window means that decay back to baseline is either minimal or occurs beyond the observation period.
 
@@
 
+### Biological Interpretation of Timescale Prior Hyperparameters
+
+The process duration $T_{\max}$ determines the temporal scope of observable dynamics and requires careful hyperparameter selection based on biological expectations and experimental context.
+
+#### **Dynamical Range and Steady-State Achievement**
+
+In the dimensionless framework where the reference rate is 1, the process duration $T_{\max}$ determines what fraction of steady-state dynamics will be observable:
+
+$$\text{Steady-state approach} = 1 - e^{-T_{\max}}$$
+
+**Key thresholds:**
+
+- $T_{\max} = 1$: Observes 63% approach to steady state
+- $T_{\max} = 3$: Observes 95% approach to steady state
+- $T_{\max} = 5$: Observes >99% approach to steady state
+
+#### **Hyperparameter Strategy and Biological Prior Knowledge**
+
+**Conservative Prior (Expect Complete Dynamics):**
+$$T_{\max} \sim \text{Gamma}(5.0, 1.0) \quad \text{(mean = 5, variance = 5)}$$
+
+- **Biological assumption**: Process likely spans ~5 characteristic timescales
+- **Dynamical expectation**: Will observe near-complete approach to steady state
+- **Data burden**: Data must provide evidence if process is shorter than expected
+- **Use case**: When biological knowledge suggests complete developmental programs
+
+**Permissive Prior (Let Data Determine Range):**
+$$T_{\max} \sim \text{Gamma}(1.0, 0.2) \quad \text{(mean = 5, variance = 25)}$$
+
+- **Biological assumption**: Uncertain about process duration, wide range possible
+- **Dynamical expectation**: Could observe early transients or complete dynamics
+- **Data burden**: Data determines whether short or long timescales are needed
+- **Use case**: When process duration is unknown, exploratory analysis
+
+**Skeptical Prior (Assume Short Processes):**
+$$T_{\max} \sim \text{Gamma}(1.0, 1.0) \quad \text{(mean = 1, variance = 1, heavy tail)}$$
+
+- **Biological assumption**: Most cellular processes are rapid (early dynamics only)
+- **Dynamical expectation**: Primarily observe transient phase (63% approach)
+- **Data burden**: Data must strongly support longer timescales if present
+- **Use case**: When expecting rapid responses, stress conditions, acute perturbations
+
+#### **Prior Selection Guidelines**
+
+**Choose hyperparameters based on experimental context:**
+
+1. **Developmental processes**: Use conservative prior (Gamma(5,1)) - expect complete programs
+2. **Drug responses**: Use skeptical prior (Gamma(1,1)) - expect rapid initial responses
+3. **Unknown systems**: Use permissive prior (Gamma(1,0.2)) - let data decide
+4. **Time-course validation**: Match prior to known experimental duration
+
+**Sensitivity analysis recommendation:**
+Test model robustness across different prior specifications to ensure conclusions are not prior-dependent.
+
+#### **Connection to Gene Parameter Priors**
+
+With independent parameterization, gene timing parameters should be calibrated to $T_{\max}$ expectations:
+
+\begin{align}
+t^*_{0,\text{on}i} &\sim \text{Normal}(0.3 \mathbb{E}[T_{\max}], \sigma_{on}^2) \quad \text{(expect activation at 30% of process)} \\
+\delta^*_i &\sim \text{LogNormal}(\log(0.4 \mathbb{E}[T_{\max}]), \sigma_{\delta}^2) \quad \text{(expect 40% duration activation)}
+\end{align}
+
+This ensures gene parameters are scaled appropriately to the expected process duration while maintaining independence and supporting the three piecewise activation patterns (pre-activation, transient, sustained).
+
+### Identifiability Proof for Independent Parameterization
+
+**Theorem:** The piecewise activation likelihood with independent parameterization is not invariant under parameter scaling.
+
+**Proof:** For any scaling factor $\lambda \neq 1$:
+
+1. **Gene parameters are sampled independently**: $t^*_{0,\text{on}i} \not\propto T_{\max}$ and $\delta^*_i \not\propto T_{\max}$
+
+2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
+   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{0,\text{on}i}$
+   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{0,\text{on}i}$ (unchanged)
+
+3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$**:
+   - Phase boundaries change when cell times are scaled but gene times are not
+   - Different phase assignments lead to different analytical solution forms
+
+4. **Therefore**: $\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$ □
+
+**Corollary:** Parameter recovery is uniquely identifiable because no two distinct parameter configurations can produce identical likelihoods.
+
 ### Mathematical Framework for Pre-Activation Scenarios
 
-The Normal prior for onset times enables modeling of **pre-activation scenarios** where genes are already activated when the biological process begins.
+The Normal prior for onset times enables modeling of **pre-activation scenarios** where genes are already activated when observation begins.
 This extension is mathematically valid and biologically realistic:
@@
 1.
-**Phase determination remains consistent**: For any process time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$
+**Phase determination remains consistent**: For any observed time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$
 
@@
 
-3. **Biological interpretation is clear**: Negative onset simply means activation occurred before the process start ($t^* = 0$)
+3. **Biological interpretation is clear**: Negative onset simply means activation occurred before the observation window
 
@@
 
-- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no cells will be in Phase 1 during the process
+- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no observed cells will be in Phase 1
 - **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
-- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the process duration
-
-### Constraint Feasibility Analysis and Optimization
-
-The pattern constraints above were optimized using a systematic CDF-based analysis to ensure mathematical feasibility while preserving biological interpretability.
-The analysis now includes both LogNormal and Normal distributions to handle the extended temporal framework.
+- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the observation window
+
+### Constraint Feasibility Analysis for Independent Parameterization
+
+The pattern constraints were optimized using a systematic CDF-based analysis to ensure mathematical feasibility while preserving biological interpretability under the independent parameterization framework.
 
@@
 
-# Prior hyperparameters (updated for absolute temporal parameterization)
+# Prior hyperparameters (updated for independent absolute parameterization)
+# Assuming T_max ~ Gamma(5.0, 1.0) with E[T_max] = 5.0
+E_T_max = 5.0
 priors = {
     'R_on': {'loc': 0.916, 'scale': 0.4},                           # log(2.5), fold-change (LogNormal)
-    't_on_star': {'loc': 2.5, 'scale': 1.5},           # Normal(2.5, 1.5²), absolute onset (allows negatives)
-    'delta_star': {'loc': 0.0, 'scale': 0.5},          # log(1.0), absolute duration (LogNormal)
+    't_on_star': {'loc': 0.3 * E_T_max, 'scale': 0.8},             # Normal(1.5, 0.8²), absolute onset
+    'delta_star': {'loc': np.log(0.4 * E_T_max), 'scale': 0.45},   # log(2.0), absolute duration (LogNormal)
 }
 
-# Calculate constraint feasibility for each pattern (updated for absolute parameterization)
+# Calculate constraint feasibility for each pattern (updated for independent parameterization)
 patterns = {
     'pre_activation': [
-        ('t_on_star', '<', 0.0),        # Activation before process start
+        ('t_on_star', '<', 0.0),        # Activation before observation window
         ('R_on', '>', 2.0),             # Moderate to strong fold change
@@
     'transient': [
-        ('t_on_star', '>', 0.0),        # Activation within process duration
-        ('t_on_star', '<', 2.5),        # Early enough to complete cycle
-        ('delta_star', '<', 2.0),       # Short pulse duration
+        ('t_on_star', '>', 0.0),                    # Activation within observation window
+        ('t_on_star', '<', 0.3 * E_T_max),         # Early enough to complete cycle
+        ('delta_star', '<', 0.4 * E_T_max),        # Short pulse duration
         ('R_on', '>', 2.0),                        # Sufficient fold change to observe
@@
     'sustained': [
-        ('t_on_star', '>', 0.0),        # Activation within process duration
-        ('t_on_star', '<', 1.5),        # Early activation onset
-        ('delta_star', '>', 2.5),       # Long activation duration
+        ('t_on_star', '>', 0.0),                    # Activation within observation window
+        ('t_on_star', '<', 0.3 * E_T_max),         # Early activation onset
+        ('delta_star', '>', 0.5 * E_T_max),        # Long activation duration
         ('R_on', '>', 2.0),                        # Strong fold change
@@
 
-The probabilistic graphical model implements the process duration with stick-breaking framework while adding gene-specific piecewise activation parameters.
-For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of absolute temporal gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$.
+The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while using independent temporal parameterization to eliminate scaling symmetry.
+For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of independent temporal gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$.
 Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.
@@
 
-# hyperparameters for process duration
+# hyperparameters for independent time structure
 pgm.add_node("alpha_T", r"$\alpha_{T}$", 5.5, 6, fixed=True)
@@
 # hyperparameters for lambda (entering from right edge of cell plate)
-pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 7.5, 3.5, fixed=True)
-pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 7.5, 2.5, fixed=True)
+pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 6.5, 3.5, fixed=True)
+pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 6.5, 2.5, fixed=True)
 
@@
 
-# process duration parameter (replaces hierarchical time scaling)
+# independent time parameters
 pgm.add_node("T_max", r"$T_{\max}$", 5.5, 5)
 
-# stick-breaking cell times and cell-specific parameters
+# latent time coordinates and cell-specific parameters
 pgm.add_node("t_star_j", r"$t^{\ast}_j$", 5.5, 4)
-pgm.add_node("lambda_j", r"$\lambda_j$", 6.5, 3)
+pgm.add_node("lambda_j", r"$\lambda_j$", 5.5, 3)
 
@@
 
-# process duration edges
+# independent time structure edges
 pgm.add_edge("alpha_T", "T_max", plot_params=edge_params)
@@
 
-# stick-breaking time edges
+# independent time edges (no hierarchical scaling)
 pgm.add_edge("T_max", "t_star_j", plot_params=edge_params)
@@
 
-The model combines the process duration with stick-breaking framework with piecewise constant transcription rates and steady-state initial conditions to achieve parameter identifiability while enabling rich gene expression dynamics.
+The model combines independent temporal parameterization with piecewise constant transcription rates and steady-state initial conditions to eliminate scaling symmetry while enabling rich gene expression dynamics.
 
@@
 - $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
-- $\Delta_i$ represents the vector of **absolute temporal** gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$
+- $\Delta_i$ represents the vector of **independent temporal** gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$
 - The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
 - The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
-- $T_{\max}$ represents the total process duration with fixed anchor points at $t^*_1 = 0$ and $t^*_N = T_{\max}$
 - $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
-- Gene temporal parameters are sampled independently of $T_{\max}$, eliminating scaling symmetry
-
-**Detailed priors for process duration with stick-breaking framework:**
-
-The priors are designed to encode biologically reasonable parameter ranges while achieving parameter identifiability through fixed anchor points and independent gene timing:
-
-**Process duration:**
-\begin{align}
-    T_{\max} &\sim \text{Gamma}(5.0, 1.0) \quad \text{(total process duration)}
-\end{align}
+- $T_{\max}$ determines the process duration and constrains cell temporal positions via $t^*_j \sim \text{Uniform}(0, T_{\max})$
+
+**Detailed priors for independent piecewise activation parameters:**
+
+The priors are designed to encode biologically reasonable parameter ranges while eliminating scaling symmetry through independent parameterization:
 
@@
 
-**Absolute temporal gene-specific parameters (independent of $T_{\max}$):**
-\begin{align}
-    t^*_{0,\text{on}i} &\sim \text{Normal}(2.5, 1.5^2) \quad \text{(absolute activation onset time)} \\
-    \delta^*_i &\sim \text{LogNormal}(0.0, 0.5^2) \quad \text{(absolute activation duration)}
-\end{align}
-
-**Stick-breaking cell times:**
-\begin{align}
-    t^*_1 &= 0 \quad \text{(process start, fixed)} \\
-    t^*_N &= T_{\max} \quad \text{(process end)} \\
-    \xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
-    t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{(recursive construction)}
-\end{align}
-
-**Pattern Classification with Absolute Parameters:**
-
-Using the expected process duration $\mathbb{E}[T_{\max}] = 5.0$, the pattern constraints become:
-
-- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$ (gene activated before process begins)
-- **Transient**: $0.0 < t^*_{0,\text{on}i} < 2.5$ and $\delta^*_i < 2.0$ and $R_{\text{on},i} > 2.0$ (complete cycle within process)
-- **Sustained**: $0.0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$ and $R_{\text{on},i} > 2.0$ (long-duration activation)
-
-These absolute thresholds preserve the same biological pattern meanings as the original relative formulation.
-
-**Biological interpretation of prior choices:**
-
-**Process duration:**
-
-- **$T_{\max} \sim \text{Gamma}(5.0, 1.0)$**: Centers process duration at 5 dimensionless time units, providing sufficient range for exponential convergence while maintaining computational efficiency.
-The process has clear start ($t^* = 0$) and end ($t^* = T_{\max}$) boundaries.
+**Independent temporal parameters:**
+\begin{align}
+    T_{\max} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
+    t^*_j &\sim \text{Uniform}(0, T_{\max}) \quad \text{(cell temporal positions)} \\
+    t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
+    \delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(absolute activation duration)}
+\end{align}
+
+**Pattern classification framework**: The three expression patterns are defined by constraints on the independent temporal parameters:
+
+- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
+- **Transient**: $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$, $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$, and $R_{\text{on},i} > 2.0$
+- **Sustained**: $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$, $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$, and $R_{\text{on},i} > 2.0$
+
+**Biological interpretation of independent prior choices:**
 
@@
 
-**Absolute temporal parameters:**
-
-- **$t^*_{0,\text{on}i} \sim \text{Normal}(2.5, 1.5^2)$**: Centers activation onset at the middle of the expected process duration, with 95% between approximately -0.4 and +5.4.
-**Negative values** enable pre-activation scenarios where genes were already activated before the process began.
-
-- **$\delta^*_i \sim \text{LogNormal}(0.0, 0.5^2)$**: Centers activation duration at 1 dimensionless time unit, with 95% between ~0.37 and ~2.7.
+**Independent temporal parameters:**
+
+- **$T_{\max} \sim \text{Gamma}(\alpha_{T}, \beta_{T})$**: Process duration with hyperparameters chosen based on experimental context (conservative: Gamma(5,1), permissive: Gamma(1,0.2), skeptical: Gamma(1,1)).
+
+- **$t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around $0.3 \mathbb{E}[T_{\max}]$ to expect activation at 30% of process duration.
+**Negative values** enable pre-activation scenarios where genes were already activated before observation began.
+
+- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around $0.4 \mathbb{E}[T_{\max}]$ to expect 40% duration activation.
 This enables both transient (short duration) and sustained (long duration) activation patterns.
@@
 
-- **Pre-activation** ($t^*_{0,\text{on}i} < 0$): ~4.8% probability under the prior
-- **Transient** ($0 < t^*_{0,\text{on}i} < 2.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles
-- **Sustained** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation
-
-**Parameter independence benefits:**
-
-- **Identifiability**: Gene timing parameters are independent of process duration, eliminating scaling symmetry
-- **Biological realism**: Process boundaries ($0$ to $T_{\max}$) provide natural temporal reference frame
-- **Cross-scale applicability**: Same parameter values work across different experimental timescales
-- **Separation of concerns**: Process duration ($T_{\max}$) vs gene-specific timing (onset/duration) vs kinetics (degradation/fold-change)
-
-**Cross-System Biological Interpretation:**
-
-The same $T_{\max} = 5.0$ represents different real-world durations depending on the characteristic splicing rate $\beta$:
-
-- **Immediate-early response** ($\beta = 5 \text{ hr}^{-1}$): Process duration = $5.0/5 = 1$ hour
-- **Cell cycle progression** ($\beta = 1 \text{ hr}^{-1}$): Process duration = $5.0/1 = 5$ hours
-- **Developmental transition** ($\beta = 0.1 \text{ hr}^{-1}$): Process duration = $5.0/0.1 = 50$ hours
-
-Gene timing parameters maintain consistent biological interpretation across scales.
-
-**Priors for other gene-specific parameters (unchanged from Study 2):**
+- **Pre-activation** ($t^*_{0,\text{on}i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
+- **Transient** ($0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ and $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$): Balanced probability for observable complete cycles
+- **Sustained** ($0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ and $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$): Balanced probability for long-duration activation
+
+**Independent parameterization benefits:**
+
+- **Eliminates scaling symmetry**: No multiplicative relationships between temporal parameters
+- **Unique identifiability**: Each parameter configuration produces a unique likelihood
+- **Biological interpretability**: All parameters have direct biological meaning
+- **Flexible prior specification**: Gene timing parameters can be calibrated to expected process duration
+
+**Priors for other gene-specific parameters:**
 \begin{align}
@@
 
-**Stick-Breaking Cell Time Construction:**
-
-The cell times $\{t^*_j\}_{j=1}^N$ are constructed using a stick-breaking process with fixed anchor points:
-
-\begin{align}
-t^*_1 &= 0 \quad \text{(fixed process start)} \label{eq-anchor-start} \\
-t^*_N &= T_{\max} \quad \text{(fixed process end)} \label{eq-anchor-end} \\
-\xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \label{eq-stick-breaking-weights} \\
-t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{for } j \in \{2, \ldots, N-1\} \label{eq-stick-breaking-construction}
-\end{align}
-
-This construction automatically ensures:
-- **Ordered constraint**: $0 = t^*_1 < t^*_2 < \cdots < t^*_N = T_{\max}$
-- **Fixed boundaries**: Process start and end are anchored
-- **Efficient sampling**: $O(N)$ complexity with no constraint violations
-
-**Cell-specific capture efficiency (unchanged from Study 2):**
+**Cell-specific capture efficiency:**
 \begin{align}
@@
 The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{0,\text{on}i}$ and $\delta^*_i$.
-The stick-breaking cell times provide the temporal coordinates for evaluating the piecewise analytical solutions.
+All temporal parameters are now independent, eliminating scaling symmetry.
 
@@
 
-where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i})$, process duration parameter $(T_{\max})$, stick-breaking cell times $(t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.
+where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i})$, independent time parameters $(T_{\max}, t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.
 
@@
 p(\tilde{y}) &= \int \int \int \int p(\tilde{y} | T_{\max}, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
-&\quad p(T_{\max}) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{\max} \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-hierarchical-integration}
+&\quad p(T_{\max}) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{\max} \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-independent-integration}
 \end{align}
@@
 - $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
-- $\theta_{\text{cell}} = \{t^*_j\}_{j=1}^N$ represents stick-breaking cell times
+- $\theta_{\text{cell}} = \{t^*_j\}_{j=1}^N$ represents cell-specific temporal parameters
 - $\theta_{\text{tech}} = \{\lambda_j\}_{j=1}^N$ represents technical efficiency parameters
 
-This decomposition reflects the model's assumption that gene-specific kinetics, process duration, stick-breaking cell times, and technical factors contribute independently to the prior uncertainty.
+This decomposition reflects the model's assumption that gene-specific kinetics, cell-specific temporal positions, and technical factors contribute independently to the prior uncertainty, with no hierarchical scaling relationships.
 
@@
    \begin{align}
-   {T_{\max}}^{(s)} &\sim \text{Gamma}(5.0, 1.0)
-   \end{align}
-
-2. **Sample stick-breaking cell times**:
-   \begin{align}
-   {\xi_j}^{(s)} &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
-   {t^*_1}^{(s)} &= 0, \quad {t^*_N}^{(s)} = {T_{\max}}^{(s)} \\
-   {t^*_j}^{(s)} &= {t^*_{j-1}}^{(s)} + {\xi_{j-1}}^{(s)}({T_{\max}}^{(s)} - {t^*_{j-1}}^{(s)}) \quad \text{for } j \in \{2, \ldots, N-1\}
+   {T_{\max}}^{(s)} &\sim \text{Gamma}(\alpha_{T}, \beta_{T})
+   \end{align}
+
+2. **Sample cell-specific temporal positions**:
+   \begin{align}
+   {t^{*}_j}^{(s)} &\sim \text{Uniform}(0, {T_{\max}}^{(s)}) \quad \text{for } j \in \{1, \ldots, N\}
    \end{align}
@@
    {R_{\text{on},i}}^{(s)} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \\
-   {t^*_{0,\text{on}i}}^{(s)} &\sim \text{Normal}(2.5, 1.5^2) \\
-   {\delta^*_i}^{(s)} &\sim \text{LogNormal}(0.0, 0.5^2) \\
+   {t^*_{0,\text{on}i}}^{(s)} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \\
+   {\delta^*_i}^{(s)} &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \\
    {U_{0i}}^{(s)} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
@@
 
-The piecewise activation model exhibits multiple sources of prior uncertainty that propagate through the hierarchical structure:
-
-1. **Process duration uncertainty**: Uncertainty in $T_{\max}$ affects the total duration of the biological process, determining the temporal scale for all dynamics
-
-2. **Stick-breaking cell time uncertainty**: Uncertainty in the stick-breaking construction affects the distribution of cells across the process timeline
+The piecewise activation model with independent parameterization exhibits multiple sources of prior uncertainty:
+
+1. **Process duration uncertainty**: Uncertainty in $T_{\max}$ affects the temporal scope of the biological process and constrains cell temporal positions
+
+2. **Cell temporal position uncertainty**: Uncertainty in individual $t^*_j$ values affects where each cell is positioned along the temporal trajectory
 
@@
 
-4. **Gene-specific temporal pattern uncertainty**: Prior uncertainty in $t^*_{0,\text{on}i}, \delta^*_i$ affects when activation occurs and how long it lasts in absolute time units
+4. **Gene-specific temporal pattern uncertainty**: Prior uncertainty in $t^*_{0,\text{on}i}, \delta^*_i$ affects when activation occurs and how long it lasts in absolute time
 
@@
 
-- **Process duration interactions**: Uncertainty in $T_{\max}$ affects the temporal boundaries but does not automatically scale gene parameters, creating asymmetric uncertainty propagation
+- **Independent parameter interactions**: Uncertainty in gene timing parameters interacts with uncertainty in cell temporal positions through phase boundary comparisons, but without multiplicative scaling relationships
 
@@
 - **Non-temporal gene parameters**: $\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$, $R_{\text{on},i} \sim \text{LogNormal}(\log(2.5), 0.4^2)$
-- **Absolute temporal gene parameters**: $t^*_{0,\text{on}i} \sim \text{Normal}(2.5, 1.5^2)$, $\delta^*_i \sim \text{LogNormal}(0.0, 0.5^2)$
+- **Independent temporal gene parameters**: $t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$, $\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$
 - **Characteristic scales**: $U_{0i} \sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)$
-- **Process duration**: $T_{\max} \sim \text{Gamma}(5.0, 1.0)$
-- **Stick-breaking cell times**: $t^*_1 = 0$, $t^*_N = T_{\max}$, $\xi_j \sim \text{Beta}(1, N-j)$, $t^*_j = t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1})$
+- **Process duration**: $T_{\max} \sim \text{Gamma}(\alpha_{T}, \beta_{T})$
+- **Cell-specific times**: $t^*_j \sim \text{Uniform}(0, T_{\max})$
 - **Capture efficiencies**: $\lambda_j \sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)$
@@
 
-The piecewise activation model exhibits multiple sources of parameter uncertainty:
+The piecewise activation model with independent parameterization exhibits multiple sources of parameter uncertainty:
 
 1. **Kinetic parameter uncertainty**: Posterior distributions over $\gamma^*_i$ and $R_{\text{on},i}$ affect the shape and magnitude of concentration trajectories
-2. **Temporal parameter uncertainty**: Uncertainty in $t^*_{0,\text{on}i}$ and $\delta^*_i$ affects when activation occurs and how long it lasts
-3. **Process duration uncertainty**: Uncertainty in $T_{\max}$ affects the total process duration and temporal boundaries
+2. **Temporal parameter uncertainty**: Uncertainty in $t^*_{0,\text{on}i}$ and $\delta^*_i$ affects when activation occurs and how long it lasts in absolute time
+3. **Process duration uncertainty**: Uncertainty in $T_{\max}$ affects the temporal scope and constrains cell temporal positions
 4. **Cell-specific uncertainty**: Uncertainty in individual $t^*_j$ and $\lambda_j$ affects cell-specific predictions
@@
 
-### Properties of the Process Duration with Stick-Breaking Framework
-
-#### **Resolved Identifiability Issues**
-
-- **Eliminates scaling symmetry**: Fixed anchor points ($t^*_1 = 0$, $t^*_N = T_{\max}$) break scaling symmetry that plagued hierarchical scaling approaches
-- **Parameter identifiability**: Gene timing parameters are independent of process duration, ensuring unique parameter estimates
-- **Mathematical rigor**: Formal proof that likelihood is not invariant under parameter scaling
-- **Computational stability**: Eliminates non-identifiable parameter combinations that cause inference failures
+### Properties of the Piecewise Framework
 
@@
 
-- **Process boundaries**: Clear start ($t^* = 0$) and end ($t^* = T_{\max}$) of biological process
-- **Pre-activation scenarios**: Models genes already activated when process begins
+- **Pre-activation scenarios**: Models genes already activated when observation begins
 - **Flexible onset timing**: Normal prior allows negative onset times naturally
-- **Automatic ordering**: Stick-breaking ensures proper temporal ordering without constraints
-- **Cross-scale applicability**: Same framework works across different experimental timescales
+- **Biological realism**: Captures realistic experimental scenarios better
+- **Maintains population structure**: Hierarchical time modeling preserved
 
@@
 **Pattern coverage**: These three patterns effectively capture the essential temporal dynamics while eliminating redundant classifications.
-The absolute temporal parameterization with process boundaries ensures consistent pattern identification across different experimental timescales.
+The relative temporal parameterization ensures consistent pattern identification across different experimental timescales.
 
@@
 - **Dimensionless parameters**: Universal interpretation across experimental conditions
-- **Process boundaries**: Natural temporal reference frame from start to completion
-- **Temporal progression**: Intuitive progress metric ($t^*_j / T_{\max}$) for cellular advancement
-- **Gene-specific dynamics**: Each gene can follow distinct activation pattern independent of process duration
+- **Temporal coordination**: Hierarchical time structure captures population-level timing
+- **Gene-specific dynamics**: Each gene can follow distinct activation pattern
 
@@
 
-This study will validate the process duration with stick-breaking model's ability to:
-
-1. **Recover identifiable parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{0,\text{on}i}$, $\delta^*_i$, $T_{\max}$
-2. **Demonstrate scaling symmetry elimination**: Verify that different $T_{\max}$ values produce distinguishable likelihoods
-3. **Validate anchor point constraints**: Confirm that $t^*_1 = 0$ and $t^*_N = T_{\max}$ are properly enforced
-4. **Assess stick-breaking cell time inference**: Recover proper temporal ordering and distribution
-5. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
-6. **Scale efficiently**: Demonstrate computational advantages of analytical solutions
-7. **Validate fold-change interpretation**: Ensure $R_{\text{on},i}$ estimates match true activation strength
-8. **Assess dimensional consistency**: Verify elimination of parameter redundancy through fixed $\alpha^*_{\text{off}i} = 1.0$
-9. **Validate parameter independence**: Ensure gene timing parameters are independent of process duration
-10. **Assess parameter separation**: Verify clean separation between kinetic ($\gamma^*_i$, $R_{\text{on},i}$) and temporal ($t^*_{0,\text{on}i}$, $\delta^*_i$) parameter inference
-11. **Validate process boundary framework**: Confirm that absolute temporal parameters enable consistent pattern classification across different experimental timescales
-12. **Cross-scale validation**: Test parameter recovery across different process durations to verify identifiability
+This study will validate the independent parameterization model's ability to:
+
+1. **Recover independent piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{0,\text{on}i}$, $\delta^*_i$
+2. **Maintain process duration inference**: Recover $T_{\max}$ and cell temporal positions $t^*_j$
+3. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
+4. **Scale efficiently**: Demonstrate computational advantages of eliminated scaling symmetry
+5. **Validate fold-change interpretation**: Ensure $R_{\text{on},i}$ estimates match true activation strength
+6. **Assess dimensional consistency**: Verify elimination of parameter redundancy through fixed $\alpha^*_{\text{off}i} = 1.0$
+7. **Validate parameter identifiability**: Confirm elimination of scaling symmetry through independent parameterization
+8. **Assess parameter separation**: Verify clean separation between kinetic ($\gamma^*_i$, $R_{\text{on},i}$) and temporal ($t^*_{0,\text{on}i}$, $\delta^*_i$) parameter inference
+9. **Validate independent parameter framework**: Confirm that independent temporal parameters enable reliable parameter recovery across different experimental contexts
 
@@
 
-This validation workflow demonstrates the key capabilities of the process duration with stick-breaking model:
-
-- **Parameter identifiability**: Eliminates scaling symmetry through fixed anchor points and independent gene timing
+This validation workflow demonstrates the key capabilities of the piecewise activation model with independent parameterization:
+
 - **Analytical tractability**: Fast likelihood evaluation enables efficient parameter recovery validation
-- **Rich parameter space**: Includes fold-change parameters, absolute temporal parameters, and process duration
-- **Pre-activation modeling**: Supports negative onset times for genes already activated before process start
+- **Rich parameter space**: Includes fold-change parameters, independent temporal parameters, and process duration
+- **Pre-activation modeling**: Supports negative onset times for genes already activated before observation
 - **Dimensional consistency**: Eliminates parameter redundancy through fixed reference state parameterization
-- **Biological interpretability**: All parameters have clear biological meaning with process boundary interpretation
+- **Scaling symmetry elimination**: Independent parameterization ensures unique parameter identifiability
+- **Biological interpretability**: All parameters have clear biological meaning and interpretable ranges
 - **Comprehensive prior predictive framework**: Mathematical foundation for validating model specification through prior predictive checking as detailed in equations \eqref{eq-prior-predictive-piecewise}-\eqref{eq-information-content}, ensuring the model can generate realistic data patterns before observing any data
-- **Process boundary parameterization**: Enables interpretable parameter values and consistent pattern classification across different experimental timescales through natural temporal reference frame
+- **Independent temporal parameterization**: Enables reliable parameter recovery and eliminates catastrophic parameter recovery failures caused by scaling symmetry
 
@@
 The prior predictive checking framework validates model specification and prior adequacy, while the posterior predictive checking framework assesses parameter recovery and predictive accuracy.
-
-## Dimensionless Analytical Dynamics with Branching Trajectories and Piecewise Activation {#sec-dimless-branching-piecewise}
-
-This validation study extends the stick-breaking temporal coordinate model to accommodate **branching developmental trajectories** through a mathematically rigorous mixture model framework.
-The key innovation is a **branch mixture extension** that maintains all analytical tractability and identifiability properties while enabling modeling of cell fate decisions, lineage bifurcations, and heterogeneous developmental pathways within a unified probabilistic framework.
-
-### Model Description
-
-The branching mixture model extends the process duration with stick-breaking framework by introducing **discrete branch assignments** that partition cells into distinct developmental trajectories.
-Each branch maintains its own temporal coordinate system and gene expression parameters while sharing the same underlying analytical solutions and dimensional consistency principles.
-This approach enables modeling of complex developmental processes where cells follow different trajectories with branch-specific timing and expression patterns.
-
-The model generates count data for unspliced ($u$) and spliced ($s$) RNA molecules per cell, with dynamics governed by **branch-specific piecewise analytical solutions** and observation model that scales back to dimensional counts through characteristic scales and effective capture efficiency.
-Each cell is assigned to exactly one branch, and all parameters are conditionally independent given branch assignments.
-
-### Probabilistic Graphical Model
-
-The probabilistic graphical model represents the branching extension with discrete branch assignments, branch-specific temporal coordinates, and branch-specific gene expression parameters.
-The model maintains the same dimensional consistency and analytical tractability as the single-branch framework while enabling rich heterogeneity across developmental trajectories.
-
-```{python}
-#| label: fig-branching-model-pgm
-#| code-fold: true
-#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with branching trajectories and piecewise activation.
-Branch assignments partition cells into distinct developmental pathways with branch-specific temporal coordinates and gene expression parameters.
-
-import daft
-import matplotlib.pyplot as plt
-
-plt.rcParams["font.family"] = "serif"
-plt.rcParams["font.size"] = 14
-plt.rcParams["text.usetex"] = True
-plt.rcParams["text.latex.preamble"] = r"\usepackage{amsmath}"
-
-pgm = daft.PGM(line_width=1.2)
-
-# Branch structure hyperparameters
-pgm.add_node("alpha_K", r"$\alpha_K$", 0.5, 7, fixed=True)
-pgm.add_node("alpha_pi", r"$\alpha_\pi$", 1.5, 7, fixed=True)
-
-# Gene parameter hyperparameters (shared across branches)
-pgm.add_node("mu_gamma", r"$\mu_{\gamma^*}$", 3, 7, fixed=True)
-pgm.add_node("sigma_gamma", r"$\sigma_{\gamma^*}^2$", 3.5, 7, fixed=True)
-pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 7, fixed=True)
-pgm.add_node("sigma_U0", r"$\sigma_{U_0}^2$", 5, 7, fixed=True)
-
-# Branch-specific hyperparameters
-pgm.add_node("alpha_T", r"$\alpha_T$", 6, 7, fixed=True)
-pgm.add_node("beta_T", r"$\beta_T$", 6.5, 7, fixed=True)
-pgm.add_node("mu_on_k", r"$\mu_{on,k}$", 7.5, 7, fixed=True)
-pgm.add_node("sigma_on", r"$\sigma_{on}^2$", 8, 7, fixed=True)
-pgm.add_node("mu_delta_k", r"$\mu_{\delta,k}$", 8.5, 7, fixed=True)
-pgm.add_node("sigma_delta", r"$\sigma_{\delta}^2$", 9, 7, fixed=True)
-pgm.add_node("mu_R_k", r"$\mu_{R,k}$", 9.5, 7, fixed=True)
-pgm.add_node("sigma_R", r"$\sigma_R^2$", 10, 7, fixed=True)
-
-# Branch structure
-pgm.add_node("K", r"$K$", 1, 6)
-pgm.add_node("pi", r"$\boldsymbol{\pi}$", 2, 6)
-
-# Branch-specific parameters
-pgm.add_node("T_max_k", r"$T_{\max,k}$", 6.25, 6)
-
-# Gene-specific parameters (shared)
-pgm.add_node("gamma_star_i", r"$\gamma^*_i$", 3.25, 5.5)
-pgm.add_node("U_0i", r"$U_{0i}$", 4.5, 5.5)
-
-# Branch-gene-specific parameters
-pgm.add_node("t_on_ik", r"${t^*_{on}}_{ik}$", 7.5, 5.5)
-pgm.add_node("delta_ik", r"$\delta^*_{ik}$", 8.5, 5.5)
-pgm.add_node("R_on_ik", r"$R_{on,ik}$", 9.5, 5.5)
-
-# Cell assignments and parameters
-pgm.add_node("Z_j", r"$Z_j$", 2, 5)
-pgm.add_node("t_star_j", r"${t^*}_j$", 6.25, 5)
-pgm.add_node("lambda_j", r"$\lambda_j$", 5, 4.5)
-
-# Latent concentrations
-pgm.add_node("u_star_ij", r"$u^*_{ij}$", 3.25, 4, scale=1.0, shape="rectangle")
-pgm.add_node("s_star_ij", r"$s^*_{ij}$", 4.5, 4, scale=1.0, shape="rectangle")
-
-# Observed counts
-pgm.add_node("u_obs_ij", r"$u_{ij}$", 3.25, 3, scale=1.0, observed=True)
-pgm.add_node("s_obs_ij", r"$s_{ij}$", 4.5, 3, scale=1.0, observed=True)
-
-# edges
-edge_params = {"head_length": 0.25, "head_width": 0.2, "lw": 0.6}
-
-# Branch structure edges
-pgm.add_edge("alpha_K", "K", plot_params=edge_params)
-pgm.add_edge("alpha_pi", "pi", plot_params=edge_params)
-pgm.add_edge("K", "pi", plot_params=edge_params)
-pgm.add_edge("pi", "Z_j", plot_params=edge_params)
-
-# Branch-specific parameter edges
-pgm.add_edge("alpha_T", "T_max_k", plot_params=edge_params)
-pgm.add_edge("beta_T", "T_max_k", plot_params=edge_params)
-pgm.add_edge("mu_on_k", "t_on_ik", plot_params=edge_params)
-pgm.add_edge("sigma_on", "t_on_ik", plot_params=edge_params)
-pgm.add_edge("mu_delta_k", "delta_ik", plot_params=edge_params)
-pgm.add_edge("sigma_delta", "delta_ik", plot_params=edge_params)
-pgm.add_edge("mu_R_k", "R_on_ik", plot_params=edge_params)
-pgm.add_edge("sigma_R", "R_on_ik", plot_params=edge_params)
-
-# Gene parameter edges
-pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
-pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
-pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
-pgm.add_edge("sigma_U0", "U_0i", plot_params=edge_params)
-
-# Temporal coordinate edges
-pgm.add_edge("Z_j", "t_star_j", plot_params=edge_params)
-pgm.add_edge("T_max_k", "t_star_j", plot_params=edge_params)
-
-# Dynamics edges
-pgm.add_edge("gamma_star_i", "u_star_ij", plot_params=edge_params)
-pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)
-pgm.add_edge("t_on_ik", "u_star_ij", plot_params=edge_params)
-pgm.add_edge("t_on_ik", "s_star_ij", plot_params=edge_params)
-pgm.add_edge("delta_ik", "u_star_ij", plot_params=edge_params)
-pgm.add_edge("delta_ik", "s_star_ij", plot_params=edge_params)
-pgm.add_edge("R_on_ik", "u_star_ij", plot_params=edge_params)
-pgm.add_edge("R_on_ik", "s_star_ij", plot_params=edge_params)
-pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
-pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)
-
-# Observation edges
-pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
-pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
-pgm.add_edge("U_0i", "u_obs_ij", plot_params=edge_params)
-pgm.add_edge("U_0i", "s_obs_ij", plot_params=edge_params)
-pgm.add_edge("lambda_j", "u_obs_ij", plot_params=edge_params)
-pgm.add_edge("lambda_j", "s_obs_ij", plot_params=edge_params)
-
-# plates
-pgm.add_plate([0.75, 6.25, 1.75, 1.25], label=r"$k \in \{1, \ldots, K\}$", shift=-0.1, fontsize=10)
-pgm.add_plate([2.75, 4.75, 7.25, 1.25], label=r"$i \in \{1, \ldots, G\}$", shift=-0.1, fontsize=10)
-pgm.add_plate([1.5, 2.5, 4, 3], label=r"$j \in \{1, \ldots, N\}$", shift=-0.1, fontsize=10)
-
-pgm.render()
-```
-
-### Mathematical Description
-
-The branching mixture model extends the stick-breaking framework with **discrete branch assignments** and **branch-specific parameters**.
-The model maintains analytical tractability and identifiability while enabling rich heterogeneity across developmental trajectories.
-
-**Hyperpriors for branch structure:**
-\begin{align}
-\alpha_K &\sim \text{Gamma}(2.0, 1.0) \label{eq-alpha-K-hyperprior} \\
-\alpha_\pi &\sim \text{Gamma}(1.0, 1.0) \label{eq-alpha-pi-hyperprior}
-\end{align}
-
-**Branch structure:**
-\begin{align}
-K &\sim \text{Poisson}(\alpha_K) + 1 \label{eq-branch-count} \\
-\boldsymbol{\pi} &\sim \text{Dirichlet}\left(\frac{\alpha_\pi}{K} \mathbf{1}_K\right) \label{eq-branch-probabilities} \\
-Z_j &\sim \text{Categorical}(\boldsymbol{\pi}) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-branch-assignments}
-\end{align}
-
-where $K$ is the number of branches, $\boldsymbol{\pi} = (\pi_1, \ldots, \pi_K)$ are branch probabilities, and $Z_j \in \{1, \ldots, K\}$ is the branch assignment for cell $j$.
-
-**Hyperpriors for temporal parameters:**
-\begin{align}
-\alpha_T &\sim \text{Gamma}(5.0, 1.0) \label{eq-alpha-T-hyperprior} \\
-\beta_T &\sim \text{Gamma}(1.0, 1.0) \label{eq-beta-T-hyperprior}
-\end{align}
-
-**Branch-specific temporal coordinates:**
-\begin{align}
-T_{\max,k} &\sim \text{Gamma}(\alpha_T, \beta_T) \quad \text{for } k \in \{1, \ldots, K\} \label{eq-branch-duration}
-\end{align}
-
-**Rigorous stick-breaking specification within branches:**
-
-For each branch $k \in \{1, \ldots, K\}$, let $\mathcal{C}_k = \{j : Z_j = k\}$ denote the set of cells assigned to branch $k$, and let $\mathcal{N}_k = |\mathcal{C}_k|$ be the number of cells in branch $k$.
-Order the cells within branch $k$ as $j_{k,1}, j_{k,2}, \ldots, j_{k,\mathcal{N}_k}$ (ordering can be arbitrary).
-The stick-breaking process for branch $k$ is defined as:
-
-\begin{align}
-V_{k,\ell} &\sim \text{Beta}(1, \mathcal{N}_k - \ell) \quad \text{for } \ell \in \{1, \ldots, \mathcal{N}_k - 1\} \label{eq-branch-beta-variables-rigorous} \\
-\tilde{t}_{k,1} &= 0 \label{eq-branch-first-time} \\
-\tilde{t}_{k,\ell} &= \tilde{t}_{k,\ell-1} + V_{k,\ell-1} \prod_{m=1}^{\ell-2} (1 - V_{k,m}) \quad \text{for } \ell \in \{2, \ldots, \mathcal{N}_k\} \label{eq-branch-stick-construction-rigorous} \\
-\tilde{t}_{k,\mathcal{N}_k} &= 1 \quad \text{(normalization constraint)} \label{eq-branch-last-time} \\
-{t^*}_{j_{k,\ell}} &= T_{\max,k} \times \tilde{t}_{k,\ell} \quad \text{for } \ell \in \{1, \ldots, \mathcal{N}_k\} \label{eq-branch-time-scaling-rigorous}
-\end{align}
-
-This construction ensures that for each branch $k$:
-\begin{align}
-0 = {t^*}_{j_{k,1}} < {t^*}_{j_{k,2}} < \cdots < {t^*}_{j_{k,\mathcal{N}_k}} = T_{\max,k} \label{eq-branch-ordering}
-\end{align}
-
-**Hyperpriors for shared gene parameters:**
-\begin{align}
-\mu_{\gamma^*} &\sim \text{Normal}(0.0, 1.0^2) \label{eq-mu-gamma-hyperprior} \\
-\sigma_{\gamma^*}^2 &\sim \text{InverseGamma}(2.0, 0.5^2) \label{eq-sigma-gamma-hyperprior} \\
-\mu_{U_0} &\sim \text{Normal}(4.0, 1.0^2) \label{eq-mu-U0-hyperprior} \\
-\sigma_{U_0}^2 &\sim \text{InverseGamma}(2.0, 0.5^2) \label{eq-sigma-U0-hyperprior} \\
-\mu_{\lambda} &\sim \text{Normal}(0.0, 1.0^2) \label{eq-mu-lambda-hyperprior} \\
-\sigma_{\lambda}^2 &\sim \text{InverseGamma}(2.0, 0.5^2) \label{eq-sigma-lambda-hyperprior}
-\end{align}
-
-**Gene-specific parameters (shared across branches):**
-\begin{align}
-\gamma^*_i &\sim \text{LogNormal}(\mu_{\gamma^*}, \sigma_{\gamma^*}^2) \quad \text{for } i \in \{1, \ldots, G\} \label{eq-shared-gamma} \\
-U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2) \quad \text{for } i \in \{1, \ldots, G\} \label{eq-shared-U0}
-\end{align}
-
-**Hyperpriors for branch-specific gene parameters:**
-\begin{align}
-\mu_{on,k} &\sim \text{Normal}(2.5, 1.5^2) \quad \text{for } k \in \{1, \ldots, K\} \label{eq-mu-on-k-hyperprior} \\
-\sigma_{on}^2 &\sim \text{InverseGamma}(2.0, 1.0^2) \label{eq-sigma-on-hyperprior} \\
-\mu_{\delta,k} &\sim \text{Normal}(0.0, 0.5^2) \quad \text{for } k \in \{1, \ldots, K\} \label{eq-mu-delta-k-hyperprior} \\
-\sigma_{\delta}^2 &\sim \text{InverseGamma}(2.0, 0.25^2) \label{eq-sigma-delta-hyperprior} \\
-\mu_{R,k} &\sim \text{Normal}(\log(2.5), 0.4^2) \quad \text{for } k \in \{1, \ldots, K\} \label{eq-mu-R-k-hyperprior} \\
-\sigma_R^2 &\sim \text{InverseGamma}(2.0, 0.2^2) \label{eq-sigma-R-hyperprior}
-\end{align}
-
-**Branch-gene-specific parameters:**
-\begin{align}
-{t^*_{on}}_{ik} &\sim \text{Normal}(\mu_{on,k}, \sigma_{on}^2) \quad \text{for } i \in \{1, \ldots, G\}, k \in \{1, \ldots, K\} \label{eq-branch-onset} \\
-\delta^*_{ik} &\sim \text{LogNormal}(\mu_{\delta,k}, \sigma_{\delta}^2) \quad \text{for } i \in \{1, \ldots, G\}, k \in \{1, \ldots, K\} \label{eq-branch-duration-gene} \\
-R_{on,ik} &\sim \text{LogNormal}(\mu_{R,k}, \sigma_R^2) \quad \text{for } i \in \{1, \ldots, G\}, k \in \{1, \ldots, K\} \label{eq-branch-fold-change}
-\end{align}
-
-**Cell-specific capture efficiency:**
-\begin{align}
-\lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-branch-lambda}
-\end{align}
-
-**Branch-specific piecewise analytical solutions:**
-
-For cell $j$ assigned to branch $k$ (i.e., $Z_j = k$), the dimensionless transcription rate for gene $i$ is:
-
-\begin{align}
-\alpha^*_{ik}(t^*) &:= \begin{cases}
-1.0 & \text{if } t^* < {t^*_{on}}_{ik} \\
-R_{on,ik} & \text{if } {t^*_{on}}_{ik} \leq t^* < {t^*_{on}}_{ik} + \delta^*_{ik} \\
-1.0 & \text{if } t^* \geq {t^*_{on}}_{ik} + \delta^*_{ik}
-\end{cases} \label{eq-branch-piecewise-transcription}
-\end{align}
-
-The dimensionless concentrations $u^*_{ij}$ and $s^*_{ij}$ are computed using the same piecewise analytical solutions as the single-branch model, but with branch-gene-specific parameters ${t^*_{on}}_{ik}$, $\delta^*_{ik}$, and $R_{on,ik}$, and cell-specific temporal coordinate ${t^*}_j$ determined by the stick-breaking process for branch $Z_j$.
-
-**Conditional independence structure:**
-
-Given branch assignments $\{Z_j\}_{j=1}^N$ and all parameters, the observations are conditionally independent:
-\begin{align}
-p(\{u_{ij}, s_{ij}\}_{i,j} \mid \{Z_j\}, \text{all parameters}) &= \prod_{j=1}^N \prod_{i=1}^G p(u_{ij}, s_{ij} \mid Z_j, \text{parameters}) \label{eq-conditional-independence}
-\end{align}
-
-**Observation model:**
-\begin{align}
-u_{ij} \mid Z_j = k &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}(k)) \label{eq-branch-u-obs} \\
-s_{ij} \mid Z_j = k &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij}(k)) \label{eq-branch-s-obs}
-\end{align}
-
-where $u^*_{ij}(k)$ and $s^*_{ij}(k)$ denote the dimensionless concentrations computed using branch $k$'s parameters and temporal coordinates.
-
-**Marginal likelihood (integrating over branch assignments):**
-\begin{align}
-p(\{u_{ij}, s_{ij}\}_{i,j} \mid \text{parameters}) &= \prod_{j=1}^N \sum_{k=1}^K \pi_k \prod_{i=1}^G p(u_{ij}, s_{ij} \mid Z_j = k, \text{parameters}) \label{eq-marginal-likelihood}
-\end{align}
-
-**Complete data likelihood (given branch assignments):**
-\begin{align}
-p(\{u_{ij}, s_{ij}\}_{i,j}, \{Z_j\}_{j=1}^N \mid \text{parameters}) &= \prod_{j=1}^N \pi_{Z_j} \prod_{i=1}^G p(u_{ij}, s_{ij} \mid Z_j, \text{parameters}) \label{eq-complete-likelihood}
-\end{align}
-
-### Branch-Specific Parameter Structure
-
-The branching mixture model introduces a **hierarchical parameter organization** that balances biological interpretability with computational efficiency.
-Parameters are strategically partitioned into shared and branch-specific categories based on biological considerations and identifiability requirements.
-
-#### **Shared Parameters Across Branches**
-
-**$\gamma^*_i$ (Relative degradation rate)**
-
-- **Rationale**: mRNA degradation kinetics are primarily determined by sequence-specific features (3' UTR elements, miRNA binding sites, RNA-binding protein motifs) that are intrinsic to the gene and independent of developmental trajectory
-- **Biological justification**: Degradation machinery (exonucleases, deadenylation complexes) operates consistently across cell types within the same organism
-- **Mathematical advantage**: Reduces parameter space from $G \times K$ to $G$ parameters while maintaining biological realism
-- **Identifiability**: Shared degradation rates enable borrowing strength across branches for better parameter estimation
-
-**$U_{0i}$ (Characteristic concentration scale)**
-
-- **Rationale**: Represents the gene's fundamental expression capacity, which reflects promoter strength, chromatin accessibility, and transcriptional machinery availability
-- **Biological justification**: While absolute expression levels vary across branches, the characteristic scale represents the gene's intrinsic transcriptional potential
-- **Mathematical advantage**: Separates scaling effects from temporal dynamics, maintaining dimensional consistency across branches
-- **Identifiability**: Shared scaling parameters prevent confounding between branch-specific temporal effects and expression magnitude
-
-#### **Branch-Specific Parameters**
-
-**$T_{\max,k}$ (Branch-specific process duration)**
-
-- **Definition**: Maximum dimensionless time for developmental process in branch $k$
-- **Biological interpretation**: Different developmental trajectories may proceed at different rates, with some lineages requiring longer maturation times
-- **Examples**:
-  - Neuronal differentiation (slow, $T_{\max} \sim 10$) vs. myeloid differentiation (fast, $T_{\max} \sim 3$)
-  - Stress response (rapid, $T_{\max} \sim 2$) vs. metabolic adaptation (gradual, $T_{\max} \sim 8$)
-- **Mathematical role**: Provides branch-specific temporal scaling while maintaining stick-breaking identifiability
-
-**${t^*_{on}}_{ik}$ (Branch-gene-specific onset time)**
-
-- **Definition**: Dimensionless activation onset time for gene $i$ in branch $k$
-- **Biological interpretation**: Different developmental trajectories activate the same genes at different relative timepoints
-- **Examples**:
-  - Transcription factor cascades: Early TFs activate first in committed lineages, later in multipotent cells
-  - Metabolic genes: Glycolytic genes activate early in proliferative branches, late in quiescent branches
-  - Stress response: Immediate early genes activate quickly in acute stress, slowly in chronic adaptation
-- **Prior structure**: $\text{Normal}(\mu_{on,k}, \sigma_{on}^2)$ allows branch-specific timing while sharing variance across genes
-
-**$\delta^*_{ik}$ (Branch-gene-specific activation duration)**
-
-- **Definition**: Dimensionless activation duration for gene $i$ in branch $k$
-- **Biological interpretation**: Same genes may have different activation durations across developmental trajectories
-- **Examples**:
-  - Cell cycle genes: Brief activation in rapidly dividing cells, prolonged in slowly cycling cells
-  - Differentiation markers: Transient in intermediate states, sustained in terminal differentiation
-  - Signaling molecules: Pulse-like in signaling cells, sustained in responding cells
-- **Prior structure**: $\text{LogNormal}(\mu_{\delta,k}, \sigma_{\delta}^2)$ ensures positive durations with branch-specific means
-
-**$R_{on,ik}$ (Branch-gene-specific fold-change)**
-
-- **Definition**: Fold-change in transcription rate during activation for gene $i$ in branch $k$
-- **Biological interpretation**: Same genes may be activated to different extents across developmental trajectories
-- **Examples**:
-  - Lineage markers: Strong activation in committed lineages, weak in multipotent cells
-  - Metabolic enzymes: High fold-change in metabolically active branches, low in quiescent branches
-  - Stress genes: Massive activation in stressed cells, moderate in adapted cells
-- **Prior structure**: $\text{LogNormal}(\mu_{R,k}, \sigma_R^2)$ ensures positive fold-changes with branch-specific means
-
-#### **Hyperparameter Interpretation**
-
-**Branch-specific location parameters** ($\mu_{on,k}$, $\mu_{\delta,k}$, $\mu_{R,k}$):
-
-- **Biological meaning**: Characteristic timing and expression patterns for each developmental trajectory
-- **Identifiability**: Enable learning of branch-specific expression programs while sharing variance parameters
-- **Flexibility**: Allow each branch to have distinct temporal and expression characteristics
-
-**Shared variance parameters** ($\sigma_{on}^2$, $\sigma_{\delta}^2$, $\sigma_R^2$):
-
-- **Statistical efficiency**: Borrowing strength across branches improves parameter estimation
-- **Biological realism**: Gene-to-gene variation in timing and expression is similar across developmental contexts
-- **Computational advantage**: Reduces parameter space while maintaining model flexibility
-
-### Stick-Breaking Within Branches
-
-The branching mixture model implements **independent stick-breaking processes** within each branch, maintaining all identifiability and computational advantages of the single-branch framework while enabling branch-specific temporal dynamics.
-
-#### **Branch-Specific Stick-Breaking Construction**
-
-The rigorous stick-breaking specification is provided in the Mathematical Description section above (equations \eqref{eq-branch-beta-variables-rigorous}-\eqref{eq-branch-ordering}).
-Each branch $k$ implements an independent stick-breaking process with the following key properties:
-
-- **Independent Beta variables**: $V_{k,\ell} \sim \text{Beta}(1, \mathcal{N}_k - \ell)$ for each position $\ell$ within branch $k$
-- **Recursive construction**: Temporal coordinates built sequentially using stick-breaking formula
-- **Fixed anchors**: First cell at $t^* = 0$, last cell at $t^* = T_{\max,k}$
-- **Proper ordering**: Ensures $0 = {t^*}_{j_{k,1}} < {t^*}_{j_{k,2}} < \cdots < {t^*}_{j_{k,\mathcal{N}_k}} = T_{\max,k}$
-
-#### **Mathematical Properties of Branch-Specific Stick-Breaking**
-
-**Ordering preservation within branches:**
-
-From equation \eqref{eq-branch-ordering}, each branch maintains strict temporal ordering:
-\begin{align}
-0 = {t^*}_{j_{k,1}} < {t^*}_{j_{k,2}} < \cdots < {t^*}_{j_{k,\mathcal{N}_k}} = T_{\max,k} \quad \text{for all } k \in \{1, \ldots, K\} \label{eq-branch-ordering-property}
-\end{align}
-
-**Independence across branches:**
-
-The Beta variables for different branches are mutually independent:
-\begin{align}
-\{V_{k,\ell}\}_{\ell=1}^{\mathcal{N}_k-1} \perp \{V_{k',\ell'}\}_{\ell'=1}^{\mathcal{N}_{k'}-1} \quad \text{for } k \neq k' \label{eq-branch-independence}
-\end{align}
-
-**Conditional independence of temporal coordinates:**
-
-Given branch assignments $\{Z_j\}_{j=1}^N$ and process durations $\{T_{\max,k}\}_{k=1}^K$, the temporal coordinates are conditionally independent across branches:
-\begin{align}
-p(\{{t^*}_j\}_{j=1}^N \mid \{Z_j\}, \{T_{\max,k}\}) &= \prod_{k=1}^K p(\{{t^*}_j : j \in \mathcal{C}_k\} \mid T_{\max,k}) \label{eq-conditional-independence-time}
-\end{align}
-
-**Identifiability within each branch:**
-
-Each branch maintains the same identifiability properties as the single-branch model:
-- **Fixed anchors**: ${t^*}_{j_{k,1}} = 0$ and ${t^*}_{j_{k,\mathcal{N}_k}} = T_{\max,k}$ eliminate scaling symmetry
-- **Independent gene parameters**: ${t^*_{on}}_{ik}$, $\delta^*_{ik}$ are sampled independently of $T_{\max,k}$ (equations \eqref{eq-branch-onset}-\eqref{eq-branch-fold-change})
-- **Analytical tractability**: Same piecewise solutions apply within each branch
-
-#### **Computational Complexity Analysis**
-
-**Time complexity:**
-- **Single-branch stick-breaking**: $O(N \log N)$ for $N$ cells
-- **Multi-branch stick-breaking**: $O\left(\sum_{k=1}^K \mathcal{N}_k \log \mathcal{N}_k\right) = O(N \log N)$ since $\sum_{k=1}^K \mathcal{N}_k = N$
-- **No asymptotic increase**: Branching does not increase computational complexity
-
-**Memory complexity:**
-- **Additional storage**: $O(K)$ for branch-specific parameters
-- **Cell assignments**: $O(N)$ for storing $Z_j$ values
-- **Total increase**: $O(K + N)$, linear in problem size
-
-**Parallel computation opportunities:**
-- **Independent branch processing**: Stick-breaking for each branch can be computed in parallel
-- **Gene parameter updates**: Branch-gene parameters can be updated independently across branches
-- **Likelihood evaluation**: Branch-specific likelihoods can be computed in parallel
-
-#### **Implementation Considerations**
-
-**Branch assignment updates:**
-
-During MCMC sampling, branch assignments $Z_j$ are updated using the conditional posterior distribution:
-\begin{align}
-P(Z_j = k \mid \{u_{ij}, s_{ij}\}_{i=1}^G, \text{all other parameters}) &\propto \pi_k \prod_{i=1}^G p(u_{ij}, s_{ij} \mid Z_j = k, \text{parameters}) \label{eq-branch-assignment-posterior}
-\end{align}
-
-where the likelihood term is computed using branch $k$'s temporal coordinates and gene-specific parameters as specified in equations \eqref{eq-branch-u-obs}-\eqref{eq-branch-s-obs}.
-
-**Stick-breaking recomputation:**
-
-When branch assignments change, only the affected branches need stick-breaking recomputation:
-- **Efficient updates**: Only recompute stick-breaking for branches that gained or lost cells
-- **Unchanged branches**: Branches with stable membership retain their temporal coordinates
-- **Computational savings**: Avoid full recomputation when only a few cells change branches
-
-**Numerical stability:**
-
-- **Beta variable bounds**: Ensure $V_{k,\ell} \in (0, 1)$ to prevent degenerate stick-breaking
-- **Minimum branch size**: Require $\mathcal{N}_k \geq 2$ for meaningful stick-breaking (at least start and end points)
-- **Empty branch handling**: Branches with $\mathcal{N}_k = 0$ are automatically excluded from likelihood computation
-
-### Identifiability Analysis
-
-The branching mixture model preserves all identifiability properties of the single-branch framework while introducing additional structure that enhances parameter recovery through **conditional independence** and **hierarchical regularization**.
-
-#### **Preservation of Single-Branch Identifiability**
-
-**Theorem (Branch-Conditional Identifiability):**
-
-Given branch assignments $\{Z_j\}_{j=1}^N$, each branch $k$ maintains the same identifiability properties as the single-branch model.
-
-**Proof:**
-
-1. **Fixed anchors within branch**: For cells in branch $k$, we have ${t^*}_{j_1} = 0$ and ${t^*}_{j_{\mathcal{N}_k}} = T_{\max,k}$
-
-2. **Independent gene parameters**: ${t^*_{on}}_{ik}$ and $\delta^*_{ik}$ are sampled independently of $T_{\max,k}$
-
-3. **No scaling symmetry**: The same argument as in the single-branch case applies within each branch
-
-4. **Analytical solutions**: The piecewise analytical solutions within each branch are identical to the single-branch case
-
-Therefore, conditional on branch assignments, each branch is identifiable. □
-
-#### **Enhanced Identifiability Through Mixture Structure**
-
-**Cross-branch parameter constraints:**
-
-The mixture structure introduces additional identifiability through **cross-branch comparisons**:
-
-\begin{align}
-\mathcal{L}_{\text{mixture}} &= \prod_{j=1}^N \sum_{k=1}^K \pi_k \cdot \mathcal{L}_k(\text{data}_j \mid \text{branch } k) \\
-&\neq \prod_{j=1}^N \sum_{k=1}^K \pi_k \cdot \mathcal{L}_k(\text{data}_j \mid \text{scaled branch } k)
-\end{align}
-
-**Theorem (Mixture Identifiability Enhancement):**
-
-The mixture likelihood is not invariant under branch-specific parameter scaling, providing additional identifiability constraints beyond single-branch identifiability.
-
-**Proof:**
-
-1. **Asymmetric scaling**: Scaling parameters in branch $k$ but not branch $k'$ changes the relative likelihood contributions
-
-2. **Branch assignment sensitivity**: Cells near branch boundaries become more or less likely to be assigned to scaled branches
-
-3. **Cross-branch competition**: The mixture structure creates competition between branches that prevents arbitrary scaling
-
-4. **Hierarchical constraints**: Shared hyperparameters ($\sigma_{on}^2$, $\sigma_{\delta}^2$, $\sigma_R^2$) provide additional regularization
-
-Therefore, the mixture structure enhances identifiability beyond the single-branch case. □
-
-#### **Identifiability of Branch Structure Parameters**
-
-**Number of branches ($K$):**
-
-- **Identifiable through**: Model comparison using information criteria (WAIC, LOO-CV)
-- **Practical bounds**: $K \leq N/2$ to ensure sufficient cells per branch for stick-breaking
-- **Biological constraints**: $K$ should reflect known developmental trajectories
-
-**Branch probabilities ($\boldsymbol{\pi}$):**
-
-- **Identifiable through**: Observed frequencies of branch assignments
-- **Dirichlet prior**: Provides regularization while allowing flexible branch sizes
-- **Constraint**: $\sum_{k=1}^K \pi_k = 1$ ensures proper probability distribution
-
-**Branch assignments ($\{Z_j\}$):**
-
-- **Identifiable through**: Likelihood comparison across branches for each cell
-- **Uncertainty quantification**: Posterior probabilities $P(Z_j = k \mid \text{data})$ provide assignment confidence
-- **Soft clustering**: Cells can have uncertain branch assignments, reflecting biological reality
-
-#### **Practical Identifiability Considerations**
-
-**Minimum branch size requirements:**
-
-- **Stick-breaking constraint**: Each branch needs $\mathcal{N}_k \geq 2$ for meaningful temporal ordering
-- **Parameter estimation**: Larger branches provide better parameter estimates
-- **Recommended minimum**: $\mathcal{N}_k \geq 10$ for stable parameter recovery
-
-**Gene parameter identifiability:**
-
-- **Shared parameters**: $\gamma^*_i$ and $U_{0i}$ benefit from pooling across branches
-- **Branch-specific parameters**: Require sufficient cells and temporal coverage within each branch
-- **Hierarchical regularization**: Branch-specific means with shared variances improve identifiability
-
-**Temporal coverage requirements:**
-
-- **Within-branch coverage**: Each branch should span sufficient temporal range for parameter recovery
-- **Cross-branch comparison**: Overlapping temporal ranges enable branch discrimination
-- **Gene activation coverage**: Each branch should contain cells before, during, and after gene activation events
-
-### Biological Interpretation
-
-The branching mixture model provides a **mathematically rigorous framework** for modeling complex developmental processes where cells follow distinct trajectories with different temporal dynamics and gene expression programs.
-
-#### **Developmental Biology Applications**
-
-**Cell fate decisions and lineage bifurcations:**
-
-The model naturally captures **binary and multi-way cell fate decisions** where progenitor cells commit to distinct developmental lineages:
-
-- **Hematopoiesis**: Multipotent stem cells differentiating into myeloid vs. lymphoid lineages
-- **Neural development**: Neural progenitors becoming neurons vs. glia
-- **Embryonic development**: Pluripotent cells committing to different germ layers
-- **Tissue regeneration**: Stem cells responding to injury with different repair programs
-
-**Branch-specific temporal dynamics:**
-
-Different developmental trajectories often proceed at **different rates** and with **different timing**:
-
-- **Fast vs. slow differentiation**: Some lineages mature rapidly (hours), others slowly (days)
-- **Sequential vs. parallel activation**: Genes may activate in different orders across branches
-- **Transient vs. sustained expression**: Same genes may show different temporal patterns across lineages
-
-**Examples of branch-specific gene expression patterns:**
-
-1. **Transcription factor cascades**:
-   - **Branch 1 (Neural)**: Early activation of neuronal TFs, sustained expression of synaptic genes
-   - **Branch 2 (Glial)**: Later activation of glial TFs, transient expression of migration genes
-
-2. **Metabolic reprogramming**:
-   - **Branch 1 (Proliferative)**: Early glycolytic gene activation, sustained cell cycle gene expression
-   - **Branch 2 (Quiescent)**: Early oxidative metabolism genes, transient stress response activation
-
-3. **Stress response pathways**:
-   - **Branch 1 (Adaptation)**: Gradual activation of protective genes, sustained expression
-   - **Branch 2 (Apoptosis)**: Rapid activation of death genes, transient survival gene expression
-
-#### **Experimental Design Connections**
-
-**Single-cell trajectory inference:**
-
-The model provides a **probabilistic foundation** for trajectory inference methods:
-
-- **Pseudotime estimation**: Branch-specific temporal coordinates provide principled pseudotime
-- **Lineage tracing**: Branch assignments correspond to inferred lineage relationships
-- **Trajectory topology**: Number of branches and their relationships reflect developmental structure
-
-**Lineage tracing validation:**
-
-The model can be **validated against experimental lineage tracing**:
-
-- **Clonal analysis**: Compare inferred branch assignments to known clonal relationships
-- **Barcode sequencing**: Use genetic barcodes to validate branch structure
-- **Time-course experiments**: Compare inferred temporal coordinates to known sampling times
-
-**Perturbation experiments:**
-
-The model enables **quantitative analysis of developmental perturbations**:
-
-- **Knockout effects**: Compare branch structure and gene expression patterns in wild-type vs. knockout
-- **Drug treatments**: Assess how chemical perturbations alter developmental trajectories
-- **Environmental stress**: Model how external conditions affect lineage commitment and timing
-
-#### **Connection to Established Developmental Biology Concepts**
-
-**Waddington's epigenetic landscape:**
-
-The branching model provides a **quantitative implementation** of Waddington's landscape metaphor:
-
-- **Branches as valleys**: Each developmental trajectory corresponds to a valley in the landscape
-- **Branch probabilities as basin sizes**: $\pi_k$ reflects the relative attractiveness of each trajectory
-- **Temporal coordinates as progress**: ${t^*}_j$ represents progress down the developmental valley
-- **Gene expression as landscape features**: Branch-specific parameters define the molecular landscape
-
-**Canalization and developmental constraints:**
-
-The model captures **developmental canalization** through parameter structure:
-
-- **Shared degradation rates**: Reflect conserved post-transcriptional machinery
-- **Branch-specific timing**: Captures lineage-specific developmental programs
-- **Hierarchical priors**: Represent evolutionary constraints on developmental variation
-
-**Developmental plasticity and robustness:**
-
-The mixture structure enables modeling of **developmental plasticity**:
-
-- **Branch assignment uncertainty**: Reflects cellular plasticity and fate choice flexibility
-- **Parameter variation**: Captures individual cell variation within developmental programs
-- **Environmental responsiveness**: Branch probabilities can depend on external conditions
-
-#### **Clinical and Therapeutic Applications**
-
-**Disease modeling:**
-
-The framework can model **pathological developmental processes**:
-
-- **Cancer progression**: Tumor cells following different evolutionary trajectories
-- **Developmental disorders**: Disrupted lineage specification and timing
-- **Aging**: Age-related changes in stem cell differentiation capacity
-
-**Drug discovery:**
-
-The model enables **quantitative assessment of therapeutic interventions**:
-
-- **Target identification**: Genes with branch-specific expression patterns as therapeutic targets
-- **Timing optimization**: Optimal timing of interventions based on developmental stage
-- **Combination therapy**: Targeting multiple branches simultaneously
-
-**Regenerative medicine:**
-
-The framework supports **rational design of regenerative therapies**:
-
-- **Directed differentiation**: Optimizing protocols to bias cells toward desired branches
-- **Reprogramming efficiency**: Quantifying success of cellular reprogramming approaches
-- **Tissue engineering**: Designing scaffolds that support specific developmental trajectories
-
-### Computational Considerations
-
-The branching mixture model maintains **computational efficiency** while significantly expanding modeling capabilities through careful algorithmic design and mathematical structure preservation.
-
-#### **Scalability Analysis**
-
-**Time complexity scaling:**
-
-\begin{align}
-\text{Single-branch model:} &\quad O(N \log N + NG) \\
-\text{Branching mixture model:} &\quad O(K \cdot N \log N + KNG + NK)
-\end{align}
-
-where:
-
-- $N$ = number of cells
-- $G$ = number of genes
-- $K$ = number of branches
-
-**Asymptotic behavior:**
-
-- **Linear in $K$**: Computational cost scales linearly with number of branches
-- **No interaction terms**: $K$, $N$, and $G$ scaling factors are multiplicative, not exponential
-- **Practical bounds**: For typical values ($K \leq 10$, $N \leq 10^4$, $G \leq 10^3$), computational cost remains manageable
-
-**Memory complexity:**
-\begin{align}
-\text{Additional memory:} &\quad O(KG + KN + K) \\
-&= O(K(G + N + 1))
-\end{align}
-
-**Storage breakdown:**
-
-- **Branch-gene parameters**: $O(KG)$ for ${t^*_{on}}_{ik}$, $\delta^*_{ik}$, $R_{on,ik}$
-- **Branch assignments**: $O(N)$ for $Z_j$ values
-- **Branch-specific temporal coordinates**: $O(KN)$ for stick-breaking variables
-- **Branch probabilities**: $O(K)$ for $\boldsymbol{\pi}$
-
-#### **Algorithmic Optimizations**
-
-**Parallel computation strategies:**
-
-1.
-**Branch-parallel likelihood evaluation**: The marginal likelihood in equation \eqref{eq-marginal-likelihood} can be computed in parallel across branches since each branch's contribution is independent given the parameters.
-
-2.
-**Gene-parallel parameter updates**: Branch-gene-specific parameters (equations \eqref{eq-branch-onset}-\eqref{eq-branch-fold-change}) can be updated in parallel across genes within each branch due to conditional independence.
-
-3.
-**Cell-parallel branch assignment updates**: The posterior probabilities in equation \eqref{eq-branch-assignment-posterior} can be computed in parallel across cells since assignments are conditionally independent given all other parameters.
-
-**Efficient MCMC sampling:**
-
-**Block updates for branch structure:**
-
-- **Joint branch probability update**: Sample $\boldsymbol{\pi}$ using conjugate Dirichlet posterior
-- **Efficient branch assignment**: Use categorical sampling with precomputed likelihoods
-- **Adaptive branch proposals**: Propose branch reassignments based on likelihood ratios
-
-**Hierarchical parameter updates:**
-
-- **Shared parameter pooling**: Update $\gamma^*_i$ and $U_{0i}$ using data from all branches
-- **Branch-specific parameter blocks**: Update all parameters for branch $k$ jointly
-- **Hyperparameter updates**: Use conjugate priors where possible for efficient sampling
-
-**Stick-breaking optimizations:**
-
-- **Incremental updates**: Only recompute stick-breaking for branches with changed membership
-- **Cached computations**: Store intermediate stick-breaking results to avoid recomputation
-- **Vectorized operations**: Use efficient linear algebra for Beta variable sampling
-
-#### **Implementation Considerations**
-
-**Numerical stability:**
-
-**Beta variable sampling**: The Beta variables $V_{k,\ell} \sim \text{Beta}(1, \mathcal{N}_k - \ell)$ in equation \eqref{eq-branch-beta-variables-rigorous} should be bounded away from 0 and 1 to prevent degenerate stick-breaking constructions.
-
-**Likelihood computation**: The marginal likelihood in equation \eqref{eq-marginal-likelihood} should be computed in log-space to prevent numerical overflow/underflow, particularly when $K$ is large or when branch probabilities $\pi_k$ are small.
-
-**Memory management:**
-
-**Efficient data structures:**
-- **Sparse representations**: Gene expression matrices can be stored sparsely since many genes may have zero counts
-- **Branch indexing**: Maintain efficient mappings between cells and their branch assignments $Z_j$
-- **Parameter caching**: Cache frequently accessed combinations of branch-gene parameters
-
-**Memory-efficient processing**: For large datasets, branches can be processed sequentially rather than simultaneously to limit memory usage, taking advantage of the conditional independence structure in equations \eqref{eq-conditional-independence} and \eqref{eq-conditional-independence-time}.
-
-#### **Validation and Diagnostics**
-
-**Convergence diagnostics:**
-
-**Branch-specific convergence:**
-
-- **Within-branch parameter convergence**: Monitor $\hat{R}$ statistics for each branch separately
-- **Cross-branch parameter comparison**: Ensure branch-specific parameters are distinguishable
-- **Branch assignment stability**: Track branch assignment changes across MCMC iterations
-
-**Model selection criteria:**
-
-**Information criteria for branch number:**
-\begin{align}
-\text{WAIC}_K &= -2 \sum_{j=1}^N \log \mathbb{E}[p(y_j \mid \theta)] + 2 p_{\text{eff}} \\
-\text{LOO-CV}_K &= \sum_{j=1}^N \log p(y_j \mid y_{-j})
-\end{align}
-
-**Cross-validation strategies:**
-
-- **Leave-one-branch-out**: Assess model performance when excluding entire branches
-- **Temporal cross-validation**: Validate on held-out time points within each branch
-- **Gene-wise validation**: Test parameter recovery for held-out genes
-
-**Parameter recovery validation:**
-
-**Branch-specific validation metrics:**
-
-- **Branch assignment accuracy**: Compare inferred vs. true branch assignments
-- **Within-branch parameter recovery**: Assess parameter recovery within each branch
-- **Cross-branch parameter discrimination**: Verify that branch-specific parameters are distinguishable
-
-**Computational benchmarking:**
-
-Performance evaluation should focus on the scaling properties derived from the mathematical structure:
-- **Time complexity**: $O(K \cdot N \log N + KNG + NK)$ as specified in equations \eqref{eq-branch-beta-variables-rigorous}-\eqref{eq-branch-time-scaling-rigorous}
-- **Memory complexity**: $O(K(G + N + 1))$ for storing branch-specific parameters and assignments
-- **Convergence assessment**: Monitor convergence of branch assignments $Z_j$ and branch-specific parameters using standard MCMC diagnostics
-
-This comprehensive computational framework ensures that the branching mixture model remains **practically implementable** while providing **rigorous mathematical foundations** for modeling complex developmental processes with multiple trajectories.

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): remove extra latent time section
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-### Dimensionless Analytical Dynamics with Latent Time
-
-This model introduces latent dimensionless time coordinates:
-
-- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
-- **Dynamics**: Same dimensionless analytical solutions
-- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{M}^*$, $t_{loc}$, $t_{scl}$)
-- **Prior**: Gamma priors for time scales, Normal for relative positions
-- **Observation Model**: Same Poisson scaling approach
-- **Likelihood**: Poisson
-
 ### Dimensionless Analytical Dynamics with Latent Time and Piecewise Activation
@@
 
-### Parameter Recovery Validation
-
-```{python}
-#| label: dimensional-count-validation
-#| code-fold: true
-
-# Create the simple deterministic model with validation functionality
-# model = create_simple_deterministic_model_with_validation()
-
-# # Configure validation parameters
-# validation_config = ParameterRecoveryConfig(
-#     num_parameter_sets=5,       # Number of parameter sets to test
-#     sample_sizes=[50, 100, 200], # Different sample sizes
-#     noise_levels=[0.1, 0.2],    # Different noise levels
-#     inference_algorithm="SVI",   # Use SVI for speed
-#     num_samples=1000,           # Number of inference samples
-#     seed=42,                    # Random seed
-# )
-
-# # Run parameter recovery validation using registry-based validation components
-# report = model.validate_parameter_recovery(validation_config)
-
-# print(f"Validation completed!")
-# print(f"Tested {len(report['parameter_sets'])} parameter sets")
-# print(f"Overall parameter recovery success rate: {report['summary']['overall_success_rate']:.2%}")
-
-# # Quick validation for development
-# quick_results = model.quick_validation(num_genes=3, num_cells=50, noise_level=0.1)
-# print(f"Quick validation success rate: {quick_results['summary']['overall_success_rate']:.2%}")
-
-# # List available validation components
-# print(f"Available validation components: {model.list_validation_components()}")
-```
-
-## Dimensionless analytical dynamics with latent time {#sec-dimless-latent-time}
-
-This validation study extends the dimensionless approach to include **latent time coordinates** that must be inferred along with other parameters.
-We implement a hierarchical time structure that maintains biological interpretability while ensuring statistical identifiability.
-
-### Model Description
-
-The model uses the same dimensionless dynamics and observation model as the model described in @sec-dimless-observed-time, but replaces observed time coordinates with **hierarchical latent time coordinates**.
-This approach borrows strength across cells while allowing individual temporal variation.
-
-### Probabilistic Graphical Model
-
-The probabilistic graphical model now includes hierarchical priors for latent dimensionless time coordinates, with time parameters becoming random variables rather than observed quantities.
-
-```{python}
-#| label: fig-latent-time-pgm
-#| code-fold: true
-#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with hierarchical latent time coordinates.
-
-import daft
-import matplotlib.pyplot as plt
-
-plt.rcParams["font.family"] = "serif"
-plt.rcParams["font.size"] = 16
-plt.rcParams["text.usetex"] = True
-
-pgm = daft.PGM(line_width=1.2)
-
-# hyperparameters for gene-specific parameters
-pgm.add_node("mu_init", r"$\mu_{0}$", 0.5, 6, fixed=True)
-pgm.add_node("sigma_init", r"$\sigma_{0}^2$", 1.5, 6, fixed=True)
-pgm.add_node("mu_gamma", r"$\mu_{\gamma^*}$", 2.5, 6, fixed=True)
-pgm.add_node("sigma_gamma", r"$\sigma_{\gamma^*}^2$", 3.5, 6, fixed=True)
-pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 6, fixed=True)
-pgm.add_node("sigma_U0", r"$\sigma_{U_0}^2$", 5, 6, fixed=True)
-
-# hyperparameters for hierarchical time structure
-pgm.add_node("alpha_T", r"$\alpha_{T}$", 5.5, 6, fixed=True)
-pgm.add_node("beta_T", r"$\beta_{T}$", 6, 6, fixed=True)
-pgm.add_node("alpha_t_loc", r"$\alpha_{t_{loc}}$", 6.5, 6, fixed=True)
-pgm.add_node("beta_t_loc", r"$\beta_{t_{loc}}$", 7, 6, fixed=True)
-pgm.add_node("alpha_t_scale", r"$\alpha_{t_{scl}}$", 7.5, 6, fixed=True)
-pgm.add_node("beta_t_scale", r"$\beta_{t_{scl}}$", 8, 6, fixed=True)
-
-# hyperparameters for lambda (entering from right edge of cell plate)
-pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 6.5, 3.5, fixed=True)
-pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 6.5, 2.5, fixed=True)
-
-# latent variables for gene-specific parameters
-pgm.add_node("u_star_0i", r"$u^{\ast}_{0i}$", 1, 5)
-pgm.add_node("s_star_0i", r"$s^{\ast}_{0i}$", 2, 5)
-pgm.add_node("gamma_star_i", r"$\gamma^{\ast}_i$", 3, 5)
-pgm.add_node("U_0i", r"$U_{0i}$", 4.5, 5)
-
-# hierarchical time parameters (moved further left)
-pgm.add_node("T_max_star", r"$T_{M}^*$", 5.5, 5)
-pgm.add_node("t_loc", r"$t_{loc}$", 6.25, 5)
-pgm.add_node("t_scale", r"$t_{scl}$", 7, 5)
-
-# latent time coordinates and cell-specific parameters (moved further left, lambda enters from right)
-pgm.add_node("t_star_j", r"$t^{\ast}_j$", 5.5, 4)
-pgm.add_node("lambda_j", r"$\lambda_j$", 5.5, 3)
-
-# latent variables for cell-specific outcomes (centered between rows)
-pgm.add_node(
-    "u_star_ij",
-    r"$u^{\ast}_{ij}$",
-    2,
-    3.75,
-    scale=1.0,
-    shape="rectangle",
-)
-pgm.add_node(
-    "s_star_ij",
-    r"$s^{\ast}_{ij}$",
-    4,
-    3.75,
-    scale=1.0,
-    shape="rectangle",
-)
-
-# observed data
-pgm.add_node(
-    "u_obs_ij",
-    r"$u_{ij}$",
-    2,
-    2.25,
-    scale=1.0,
-    observed=True,
-)
-pgm.add_node(
-    "s_obs_ij",
-    r"$s_{ij}$",
-    4,
-    2.25,
-    scale=1.0,
-    observed=True,
-)
-
-# edges
-edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}
-
-# gene-specific parameter edges
-pgm.add_edge("mu_init", "u_star_0i", plot_params=edge_params)
-pgm.add_edge("sigma_init", "u_star_0i", plot_params=edge_params)
-pgm.add_edge("mu_init", "s_star_0i", plot_params=edge_params)
-pgm.add_edge("sigma_init", "s_star_0i", plot_params=edge_params)
-pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
-pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
-pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
-pgm.add_edge("sigma_U0", "U_0i", plot_params=edge_params)
-pgm.add_edge("mu_lambda", "lambda_j", plot_params=edge_params)
-pgm.add_edge("sigma_lambda", "lambda_j", plot_params=edge_params)
-
-# hierarchical time structure edges
-pgm.add_edge("alpha_T", "T_max_star", plot_params=edge_params)
-pgm.add_edge("beta_T", "T_max_star", plot_params=edge_params)
-pgm.add_edge("alpha_t_loc", "t_loc", plot_params=edge_params)
-pgm.add_edge("beta_t_loc", "t_loc", plot_params=edge_params)
-pgm.add_edge("alpha_t_scale", "t_scale", plot_params=edge_params)
-pgm.add_edge("beta_t_scale", "t_scale", plot_params=edge_params)
-
-# time hierarchy edges
-pgm.add_edge("T_max_star", "t_star_j", plot_params=edge_params)
-pgm.add_edge("t_loc", "t_star_j", plot_params=edge_params)
-pgm.add_edge("t_scale", "t_star_j", plot_params=edge_params)
-
-# dynamics edges
-pgm.add_edge("u_star_0i", "u_star_ij", plot_params=edge_params)
-pgm.add_edge("s_star_0i", "s_star_ij", plot_params=edge_params)
-pgm.add_edge("u_star_0i", "s_star_ij", plot_params=edge_params)
-pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)
-pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
-pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)
-
-# observation edges
-pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
-pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
-pgm.add_edge("U_0i", "u_obs_ij", plot_params=edge_params)
-pgm.add_edge("U_0i", "s_obs_ij", plot_params=edge_params)
-pgm.add_edge("lambda_j", "u_obs_ij", plot_params=edge_params)
-pgm.add_edge("lambda_j", "s_obs_ij", plot_params=edge_params)
-
-# plates
-pgm.add_plate(
-    [0.5, 1.25, 4.5, 4.25],
-    label=r"$i \in \{1, \ldots, G\}$",
-    shift=-0.1,
-    fontsize=12,
-)
-pgm.add_plate(
-    [1.5, 1.75, 4.5, 2.75],
-    label=r"$j \in \{1, \ldots, N\}$",
-    shift=-0.1,
-    fontsize=12,
-)
-
-pgm.render()
-```
-
-### Mathematical Description
-
-The model extends the dimensionless approach with **hierarchical latent time coordinates**.
-The key modification is replacing observed time $t^*_j$ with latent time variables that follow a hierarchical prior structure.
-
-**Priors for dimensionless parameters (unchanged from Study 1):**
-\begin{align}
-    u^*_{0i}, s^*_{0i} &\sim \text{LogNormal}(\mu_{0}, \sigma_{0}^2) \\
-    \gamma^*_i &\sim \text{LogNormal}(\mu_{\gamma^*}, \sigma_{\gamma^*}^2) \\
-    U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2) \\
-    \lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
-\end{align}
-
-**Hierarchical priors for latent time coordinates:**
-\begin{align}
-    T_{M}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \label{eq-T-M-prior} \\
-    t_{loc} &\sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}) \label{eq-t-loc-prior} \\
-    t_{scl} &\sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}}) \label{eq-t-scl-prior} \\
-    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scl}^2) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-t-tilde-prior} \\
-    t^*_j &= T_{M}^* \times \max(\tilde{t}_j, \epsilon) \quad \text{(clipped transformation)} \label{eq-t-star-transform}
-\end{align}
-
-$\tilde{t}_j$ represents the **relative temporal position** of cell $j$ within the population distribution, and $t^*_j$ represents the **actual dimensionless time coordinate**.
-The clipping operation $\max(\tilde{t}_j, \epsilon)$ with $\epsilon = 10^{-5}$ ensures that all time coordinates remain positive, maintaining biological realism (cells cannot be "before" the process) and mathematical consistency with the analytical solutions.
-While $\tilde{t}_j$ is not explicitly shown in the PGM for space constraints, the clipped transformation is implicitly included in the $t^*_j$ node.
-
-**Dimensionless analytical solutions (unchanged):**
-The same analytical solutions from Study 1 apply, but now using latent $t^*_j$:
-\begin{align}
-    u^*_{ij} &= 1 + (u^*_{0i} - 1) e^{-t^*_j} \\
-    s^*_{ij} &= \begin{cases}
-        \frac{1}{\gamma^*_i} + \left(s^*_{0i} - \xi_i - \frac{1}{\gamma^*_i}\right) e^{-\gamma^*_i t^*_j} + \xi_i e^{-t^*_j} & \text{if } \gamma^*_i \neq 1 \\
-        1 + (s^*_{0i} - 1) e^{-t^*_j} + (u^*_{0i} - 1) t^*_j e^{-t^*_j} & \text{if } \gamma^*_i = 1
-    \end{cases}
-\end{align}
-
-where $\xi_i = \frac{u^*_{0i} - 1}{\gamma^*_i - 1}$.
-
-**Observation model (unchanged):**
-\begin{align}
-    u_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}) \\
-    s_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij})
-\end{align}
-
-### Hierarchical Time Parameter Interpretability
-
-The hierarchical time structure introduces new parameters with clear biological and statistical interpretations:
-
-#### **Global Time Scale Parameters**
-
-**$T_{M}^*$ (Global dimensionless time scale)**
-
-- **Definition**: Global scaling factor for dimensionless time coordinates
-- **Units**: Dimensionless (in units of characteristic splicing time)
-- **Interpretation**: Characteristic temporal extent of the biological process
-- **Biological meaning**: Represents the typical time scale over which the biological process unfolds
-- **Model role**: Multiplicative scaling factor in $t^*_j \sim T_{M}^* \times \text{Normal}(t_{loc}, t_{scl}^2)$
-- **Prior**: $T_{M}^* \sim \text{Gamma}(\alpha_{T}, \beta_{T})$ ensures positive values with flexible tail behavior
-- **Example**: $T_{M}^* = 5$ means the process typically spans ~5 characteristic splicing times
-
-#### **Population-Level Time Parameters**
-
-**$t_{loc}$ (Population relative time location)**
-
-- **Definition**: Population-level location parameter for relative time distribution
-- **Units**: Dimensionless (relative time units)
-- **Interpretation**: Central tendency of where cells are located in relative time
-- **Biological meaning**: Typical relative temporal position of cells in the biological process
-- **Prior**: $t_{loc} \sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}})$ ensures positive location with flexible shape
-- **Statistical role**: Allows borrowing strength across cells while permitting individual variation
-- **Example**: $t_{loc} = 0.7$ means cells are typically positioned at 70% through the relative process timeline
-
-**$t_{scl}$ (Population relative time scale)**
-
-- **Definition**: Population-level scale parameter for relative time distribution
-- **Units**: Dimensionless (relative time units)
-- **Interpretation**: How much relative temporal variation exists across cells
-- **Biological meaning**: Degree of temporal heterogeneity in the cell population
-- **Prior**: $t_{scl} \sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}})$ ensures positive scale with flexible shape
-- **Statistical role**: Controls the amount of shrinkage toward the population mean
-- **Example**: $t_{scl} = 0.2$ indicates relatively tight temporal clustering around $t_{loc}$
-
-#### **Cell-Specific Time Coordinates**
-
-**$\tilde{t}_j$ (Cell-specific relative time)**
-
-- **Definition**: Latent relative time coordinate for cell $j$ within the population distribution
-- **Units**: Dimensionless (relative time units)
-- **Interpretation**: Cell's position relative to the population temporal distribution
-- **Biological meaning**: Where this cell sits in the relative temporal progression
-- **Prior**: $\tilde{t}_j \sim \text{Normal}(t_{loc}, t_{scl}^2)$ allows natural population variation
-- **Range**: Can be negative (before typical process) or greater than 1 (after typical process)
-- **Example**: $\tilde{t}_j = -0.2$ means this cell would be before the typical process start; $\tilde{t}_j = 1.3$ means after typical completion
-
-**$t^*_j$ (Cell-specific dimensionless time)**
-
-- **Definition**: Actual dimensionless time coordinate for cell $j$ after clipping
-- **Units**: Dimensionless (in units of characteristic splicing time)
-- **Interpretation**: How many splicing events have elapsed for this specific cell
-- **Biological meaning**: Cell's absolute position along the temporal trajectory of the biological process
-- **Transformation**: $t^*_j = T_{M}^* \times \max(\tilde{t}_j, \epsilon)$ with $\epsilon = 10^{-5}$
-- **Constraint**: Always positive, ensuring mathematical consistency with analytical solutions
-- **Example**: If $T_{M}^* = 5$, $\tilde{t}_j = 0.6$, then $t^*_j = 3.0$ characteristic splicing times; if $\tilde{t}_j = -0.1$, then $t^*_j = 5 \times 10^{-5}$ (clipped)
-
-### Properties of Gamma/Normal Hierarchical Time Structure
-
-#### **Statistical Benefits**
-
-- **Improved identifiability**: Hierarchical structure provides regularization without redundant parameters
-- **Borrowing strength**: Information shared across cells improves parameter estimation
-- **Uncertainty quantification**: Full posterior distributions for all time parameters
-- **Robustness**: Less sensitive to outlier cells or sparse data
-- **No truncation constraints**: Avoids computational complexity of truncated distributions
-- **Clear parameter roles**: $T_{M}^*$ controls global scale, $t_{loc}$ and $t_{scl}$ control population distribution
-
-#### **Biological Interpretability**
-
-- **Dimensionless time**: Universal interpretation across experimental conditions
-- **Population heterogeneity**: Captures natural variation in biological timing
-- **Individual trajectories**: Each cell has its own temporal position
-- **Flexible temporal range**: $\tilde{t}_j$ can represent cells before or after the typical process timeline
-- **Biological realism**: Clipping ensures cells cannot be "before" the process start ($t^*_j \geq \epsilon$)
-- **Mathematical consistency**: Positive time coordinates maintain validity of analytical solutions
-
-#### **Computational Efficiency**
-
-- **Standard distributions**: Gamma and Normal distributions are computationally efficient
-- **No truncation**: Avoids complex truncated distribution sampling
-- **Simple clipping**: $\max(\tilde{t}_j, \epsilon)$ operation is computationally trivial
-- **Analytical solutions**: Fast likelihood evaluation for validation
-- **Hierarchical sampling**: Efficient MCMC mixing through population-level parameters
-- **Deterministic transformation**: $t^*_j = T_{M}^* \times \max(\tilde{t}_j, \epsilon)$ is simple and fast
-
 ## Dimensionless Analytical Dynamics with Latent Time and Piecewise Activation {#sec-dimless-latent-piecewise}

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): remove references to hierarchical temporal prior
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-- **Time coordinates**: Same hierarchical latent $t^*_j$ structure as previous model
+- **Time coordinates**: Independent latent $t^*_j$ coordinates sampled uniformly within process duration $T_{\max}$
 - **Dynamics**: Piecewise constant transcription with three phases (off, on, return-to-off)
-- **Parameters**: $\gamma^*_i$, $R_{\text{on},i}$ (fold-change), $t^*_{0,\text{on}i}$ (onset), $\delta^*_i$ (duration), plus hierarchical time structure
+- **Parameters**: $\gamma^*_i$, $R_{\text{on},i}$ (fold-change), $t^*_{0,\text{on}i}$ (onset), $\delta^*_i$ (duration), with independent temporal structure
 - **Key innovation**: Fixed reference state ($\alpha^*_{\text{off}i} = 1.0$) eliminates parameter redundancy
@@
 
-- $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ (early activation within process duration)
-- $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$ (short activation duration relative to process)
+- $0 < t^*_{0,\text{on}i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
+- $\delta^*_i < 2.0$ (short activation duration relative to process, assuming conservative prior Gamma(5,1))
 - $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)
@@
 
-- $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ (early activation within process duration)
-- $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$ (long activation duration relative to process)
+- $0 < t^*_{0,\text{on}i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
+- $\delta^*_i > 2.5$ (long activation duration relative to process, assuming conservative prior Gamma(5,1))
 - $R_{\text{on},i} > 2.0$ (strong fold-change during activation)
@@
 #| code-fold: true
-#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with hierarchical latent time coordinates and piecewise activation.
+#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with independent latent time coordinates and piecewise activation.
 
@@
 
-- **$t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around $0.3 \mathbb{E}[T_{\max}]$ to expect activation at 30% of process duration.
+- **$t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around 1.5 to expect activation at 30% of process duration (assuming conservative prior Gamma(5,1) with mean 5).
 **Negative values** enable pre-activation scenarios where genes were already activated before observation began.
 
-- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around $0.4 \mathbb{E}[T_{\max}]$ to expect 40% duration activation.
+- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around 2.0 to expect 40% duration activation (assuming conservative prior Gamma(5,1) with mean 5).
 This enables both transient (short duration) and sustained (long duration) activation patterns.
@@
 - **Pre-activation** ($t^*_{0,\text{on}i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
-- **Transient** ($0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ and $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$): Balanced probability for observable complete cycles
-- **Sustained** ($0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ and $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$): Balanced probability for long-duration activation
+- **Transient** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles (assuming conservative prior Gamma(5,1))
+- **Sustained** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation (assuming conservative prior Gamma(5,1))
 
@@
 
-2. **Temporal coordination**: Does the hierarchical time structure produce biologically plausible cell orderings and temporal relationships?
+2. **Temporal coordination**: Does the independent time structure produce biologically plausible cell orderings and temporal relationships?
 
@@
 1. **Parameter uncertainty**: $\text{Var}[\theta | y_{obs}]$ from the posterior distribution over all model parameters
-2. **Temporal uncertainty**: Uncertainty in latent time coordinates $t^*_j$ and their hierarchical structure
+2. **Temporal uncertainty**: Uncertainty in latent time coordinates $t^*_j$ and process duration $T_{\max}$
 3. **Kinetic uncertainty**: Uncertainty in gene-specific kinetic parameters $\gamma^*_i$ and fold-changes $R_{\text{on},i}$
@@
 - **Biological realism**: Captures realistic experimental scenarios better
-- **Maintains population structure**: Hierarchical time modeling preserved
+- **Maintains temporal structure**: Independent time modeling with process duration constraints
 
@@
 - **Dimensionless parameters**: Universal interpretation across experimental conditions
-- **Temporal coordination**: Hierarchical time structure captures population-level timing
+- **Temporal coordination**: Independent time structure with process duration constraints captures temporal relationships
 - **Gene-specific dynamics**: Each gene can follow distinct activation pattern

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): use T_M instead of max
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 \begin{align}
-T_{\max} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
-t^*_j &\sim \text{Uniform}(0, T_{\max}) \quad \text{(cell temporal positions)} \\
-t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
+T_M &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
+t^*_j &\sim \text{Uniform}(0, T_M) \quad \text{(cell temporal positions)} \\
+t^*_{on,i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
 \delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(absolute activation duration)}
@@
 
-- $t^*_{0,\text{on}i} < 0.0$ (absolute activation onset before observation window)
+- $t^*_{on,i} < 0.0$ (absolute activation onset before observation window)
 - $R_{\text{on},i} > 2.0$ (moderate to strong fold-change during activation)
@@
 
-- $0 < t^*_{0,\text{on}i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
+- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
 - $\delta^*_i < 2.0$ (short activation duration relative to process, assuming conservative prior Gamma(5,1))
@@
 
-- $0 < t^*_{0,\text{on}i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
+- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
 - $\delta^*_i > 2.5$ (long activation duration relative to process, assuming conservative prior Gamma(5,1))
@@
 
-The process duration $T_{\max}$ determines the temporal scope of observable dynamics and requires careful hyperparameter selection based on biological expectations and experimental context.
+The process duration $T_M$ determines the temporal scope of observable dynamics and requires careful hyperparameter selection based on biological expectations and experimental context.
 
@@
 
-In the dimensionless framework where the reference rate is 1, the process duration $T_{\max}$ determines what fraction of steady-state dynamics will be observable:
-
-$$\text{Steady-state approach} = 1 - e^{-T_{\max}}$$
+In the dimensionless framework where the reference rate is 1, the process duration $T_M$ determines what fraction of steady-state dynamics will be observable:
+
+$$\text{Steady-state approach} = 1 - e^{-T_M}$$
 
@@
 
-- $T_{\max} = 1$: Observes 63% approach to steady state
-- $T_{\max} = 3$: Observes 95% approach to steady state
-- $T_{\max} = 5$: Observes >99% approach to steady state
+- $T_M = 1$: Observes 63% approach to steady state
+- $T_M = 3$: Observes 95% approach to steady state
+- $T_M = 5$: Observes >99% approach to steady state
 
@@
 **Conservative Prior (Expect Complete Dynamics):**
-$$T_{\max} \sim \text{Gamma}(5.0, 1.0) \quad \text{(mean = 5, variance = 5)}$$
+$$T_M \sim \text{Gamma}(5.0, 1.0) \quad \text{(mean = 5, variance = 5)}$$
 
@@
 **Permissive Prior (Let Data Determine Range):**
-$$T_{\max} \sim \text{Gamma}(1.0, 0.2) \quad \text{(mean = 5, variance = 25)}$$
+$$T_M \sim \text{Gamma}(1.0, 0.2) \quad \text{(mean = 5, variance = 25)}$$
 
@@
 **Skeptical Prior (Assume Short Processes):**
-$$T_{\max} \sim \text{Gamma}(1.0, 1.0) \quad \text{(mean = 1, variance = 1, heavy tail)}$$
+$$T_M \sim \text{Gamma}(1.0, 1.0) \quad \text{(mean = 1, variance = 1, heavy tail)}$$
 
@@
 
-With independent parameterization, gene timing parameters should be calibrated to $T_{\max}$ expectations:
-
-\begin{align}
-t^*_{0,\text{on}i} &\sim \text{Normal}(0.3 \mathbb{E}[T_{\max}], \sigma_{on}^2) \quad \text{(expect activation at 30% of process)} \\
-\delta^*_i &\sim \text{LogNormal}(\log(0.4 \mathbb{E}[T_{\max}]), \sigma_{\delta}^2) \quad \text{(expect 40% duration activation)}
+With independent parameterization, gene timing parameters should be calibrated to $T_M$ expectations:
+
+\begin{align}
+t^*_{on,i} &\sim \text{Normal}(0.3 \times 5, \sigma_{on}^2) = \text{Normal}(1.5, \sigma_{on}^2) \quad \text{(expect activation at 30% of process)} \\
+\delta^*_i &\sim \text{LogNormal}(\log(0.4 \times 5), \sigma_{\delta}^2) = \text{LogNormal}(\log(2.0), \sigma_{\delta}^2) \quad \text{(expect 40% duration activation)}
 \end{align}
@@
 
-1. **Gene parameters are sampled independently**: $t^*_{0,\text{on}i} \not\propto T_{\max}$ and $\delta^*_i \not\propto T_{\max}$
+1. **Gene parameters are sampled independently**: $t^*_{on,i} \not\propto T_M$ and $\delta^*_i \not\propto T_M$
 
 2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
-   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{0,\text{on}i}$
-   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{0,\text{on}i}$ (unchanged)
-
-3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$**:
+   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{on,i}$
+   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{on,i}$ (unchanged)
+
+3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{on,i}$**:
    - Phase boundaries change when cell times are scaled but gene times are not
@@
 
-4. **Therefore**: $\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$ □
+4. **Therefore**: $\mathcal{L}(T_M, \{t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_M, \{\lambda t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\})$ □
 
@@
 
-When $t^*_{0,\text{on}i} < 0$:
-
-1.
-**Phase determination remains consistent**: For any observed time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$
+When $t^*_{on,i} < 0$:
+
+1. **Phase determination remains consistent**: For any observed time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{on,i}$ and $t^*_{on,i} + \delta^*_i$
 
@@
 
-- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no observed cells will be in Phase 1
+- **Phase 1 may be unobserved**: If $t^*_{on,i} < 0$, no observed cells will be in Phase 1
 - **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
-- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the observation window
+- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{on,i}|$ ensures activation extends into the observation window
 
@@
 
-For **Normal parameters** ($t^*_{0,\text{on}i}$), the probability of satisfying constraint $X > \theta$ is:
+For **Normal parameters** ($t^*_{on,i}$), the probability of satisfying constraint $X > \theta$ is:
 
@@
 # Prior hyperparameters (updated for independent absolute parameterization)
-# Assuming T_max ~ Gamma(5.0, 1.0) with E[T_max] = 5.0
-E_T_max = 5.0
+# Assuming T_M ~ Gamma(5.0, 1.0) with E[T_M] = 5.0
+E_T_M = 5.0
 priors = {
     'R_on': {'loc': 0.916, 'scale': 0.4},                           # log(2.5), fold-change (LogNormal)
-    't_on_star': {'loc': 0.3 * E_T_max, 'scale': 0.8},             # Normal(1.5, 0.8²), absolute onset
-    'delta_star': {'loc': np.log(0.4 * E_T_max), 'scale': 0.45},   # log(2.0), absolute duration (LogNormal)
+    't_on_star': {'loc': 1.5, 'scale': 0.8},                       # Normal(1.5, 0.8²), absolute onset
+    'delta_star': {'loc': np.log(2.0), 'scale': 0.45},             # log(2.0), absolute duration (LogNormal)
 }
@@
         ('t_on_star', '>', 0.0),                    # Activation within observation window
-        ('t_on_star', '<', 0.3 * E_T_max),         # Early enough to complete cycle
-        ('delta_star', '<', 0.4 * E_T_max),        # Short pulse duration
+        ('t_on_star', '<', 1.5),                   # Early enough to complete cycle
+        ('delta_star', '<', 2.0),                  # Short pulse duration
         ('R_on', '>', 2.0),                        # Sufficient fold change to observe
@@
         ('t_on_star', '>', 0.0),                    # Activation within observation window
-        ('t_on_star', '<', 0.3 * E_T_max),         # Early activation onset
-        ('delta_star', '>', 0.5 * E_T_max),        # Long activation duration
+        ('t_on_star', '<', 1.5),                   # Early activation onset
+        ('delta_star', '>', 2.5),                  # Long activation duration
         ('R_on', '>', 2.0),                        # Strong fold change
@@
 # independent time parameters
-pgm.add_node("T_max", r"$T_{\max}$", 5.5, 5)
+pgm.add_node("T_M", r"$T_M$", 5.5, 5)
 
@@
 # independent time structure edges
-pgm.add_edge("alpha_T", "T_max", plot_params=edge_params)
-pgm.add_edge("beta_T", "T_max", plot_params=edge_params)
+pgm.add_edge("alpha_T", "T_M", plot_params=edge_params)
+pgm.add_edge("beta_T", "T_M", plot_params=edge_params)
 
 # independent time edges (no hierarchical scaling)
-pgm.add_edge("T_max", "t_star_j", plot_params=edge_params)
+pgm.add_edge("T_M", "t_star_j", plot_params=edge_params)
 
@@
 - $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
-- $\Delta_i$ represents the vector of **independent temporal** gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$
+- $\Delta_i$ represents the vector of **independent temporal** gene parameters $(t^*_{on,i}, \delta^*_i)$
 - The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
@@
 - $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
-- $T_{\max}$ determines the process duration and constrains cell temporal positions via $t^*_j \sim \text{Uniform}(0, T_{\max})$
+- $T_M$ determines the process duration and constrains cell temporal positions via $t^*_j \sim \text{Uniform}(0, T_M)$
 
@@
 \begin{align}
-    T_{\max} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
-    t^*_j &\sim \text{Uniform}(0, T_{\max}) \quad \text{(cell temporal positions)} \\
-    t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
+    T_M &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
+    t^*_j &\sim \text{Uniform}(0, T_M) \quad \text{(cell temporal positions)} \\
+    t^*_{on,i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
     \delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(absolute activation duration)}
@@
 
-- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
-- **Transient**: $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$, $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$, and $R_{\text{on},i} > 2.0$
-- **Sustained**: $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$, $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$, and $R_{\text{on},i} > 2.0$
+- **Pre-activation**: $t^*_{on,i} < 0.0$ and $R_{\text{on},i} > 2.0$
+- **Transient**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i < 2.0$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))
+- **Sustained**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i > 2.5$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))
 
@@
 
-- **$T_{\max} \sim \text{Gamma}(\alpha_{T}, \beta_{T})$**: Process duration with hyperparameters chosen based on experimental context (conservative: Gamma(5,1), permissive: Gamma(1,0.2), skeptical: Gamma(1,1)).
-
-- **$t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around 1.5 to expect activation at 30% of process duration (assuming conservative prior Gamma(5,1) with mean 5).
+- **$T_M \sim \text{Gamma}(\alpha_{T}, \beta_{T})$**: Process duration with hyperparameters chosen based on experimental context (conservative: Gamma(5,1), permissive: Gamma(1,0.2), skeptical: Gamma(1,1)).
+
+- **$t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around 1.5 to expect activation at 30% of process duration (assuming conservative prior Gamma(5,1) with mean 5).
 **Negative values** enable pre-activation scenarios where genes were already activated before observation began.
@@
 
-- **Pre-activation** ($t^*_{0,\text{on}i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
-- **Transient** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles (assuming conservative prior Gamma(5,1))
-- **Sustained** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation (assuming conservative prior Gamma(5,1))
+- **Pre-activation** ($t^*_{on,i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
+- **Transient** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles (assuming conservative prior Gamma(5,1))
+- **Sustained** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation (assuming conservative prior Gamma(5,1))
 
@@
 
-The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{0,\text{on}i}$ and $\delta^*_i$.
+The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{on,i}$ and $\delta^*_i$.
 All temporal parameters are now independent, eliminating scaling symmetry.
@@
 
-where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i})$, independent time parameters $(T_{\max}, t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.
+where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{on,i}, \delta^*_i, U_{0i})$, independent time parameters $(T_M, t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.
 
@@
 \begin{align}
-p(\tilde{y}) &= \int p(\tilde{y} | \gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}, T_{\max}, t^*_j, \lambda_j) \times \nonumber \\
-&\quad p(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}, T_{\max}, t^*_j, \lambda_j) \, d\theta \label{eq-prior-predictive-piecewise}
+p(\tilde{y}) &= \int p(\tilde{y} | \gamma^*_i, R_{\text{on},i}, t^*_{on,i}, \delta^*_i, U_{0i}, T_M, t^*_j, \lambda_j) \times \nonumber \\
+&\quad p(\gamma^*_i, R_{\text{on},i}, t^*_{on,i}, \delta^*_i, U_{0i}, T_M, t^*_j, \lambda_j) \, d\theta \label{eq-prior-predictive-piecewise}
 \end{align}
@@
 \begin{align}
-p(\tilde{y}) &= \int \int \int \int p(\tilde{y} | T_{\max}, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
-&\quad p(T_{\max}) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{\max} \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-independent-integration}
+p(\tilde{y}) &= \int \int \int \int p(\tilde{y} | T_M, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
+&\quad p(T_M) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_M \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-independent-integration}
 \end{align}
@@
 
-- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
+- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, t^*_{on,i}, \delta^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
 - $\theta_{\text{cell}} = \{t^*_j\}_{j=1}^N$ represents cell-specific temporal parameters
@@
    \begin{align}
-   {T_{\max}}^{(s)} &\sim \text{Gamma}(\alpha_{T}, \beta_{T})
+   {T_M}^{(s)} &\sim \text{Gamma}(\alpha_{T}, \beta_{T})
    \end{align}
@@
    \begin{align}
-   {t^{*}_j}^{(s)} &\sim \text{Uniform}(0, {T_{\max}}^{(s)}) \quad \text{for } j \in \{1, \ldots, N\}
+   {t^{*}_j}^{(s)} &\sim \text{Uniform}(0, {T_M}^{(s)}) \quad \text{for } j \in \{1, \ldots, N\}
    \end{align}
@@
    {R_{\text{on},i}}^{(s)} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \\
-   {t^*_{0,\text{on}i}}^{(s)} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \\
+   {t^*_{on,i}}^{(s)} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \\
    {\delta^*_i}^{(s)} &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \\
@@
    \begin{align}
-   {u^{*}_{ij}}^{(s)}, {s^{*}_{ij}}^{(s)} &= f_{\text{piecewise}}({t^*_j}^{(s)}, {\gamma^*_i}^{(s)}, {R_{\text{on},i}}^{(s)}, {t^*_{0,\text{on}i}}^{(s)}, {\delta^*_i}^{(s)})
+   {u^{*}_{ij}}^{(s)}, {s^{*}_{ij}}^{(s)} &= f_{\text{piecewise}}({t^*_j}^{(s)}, {\gamma^*_i}^{(s)}, {R_{\text{on},i}}^{(s)}, {t^*_{on,i}}^{(s)}, {\delta^*_i}^{(s)})
    \end{align}
@@
 
-1. **Process duration uncertainty**: Uncertainty in $T_{\max}$ affects the temporal scope of the biological process and constrains cell temporal positions
+1. **Process duration uncertainty**: Uncertainty in $T_M$ affects the temporal scope of the biological process and constrains cell temporal positions
 
@@
 
-4. **Gene-specific temporal pattern uncertainty**: Prior uncertainty in $t^*_{0,\text{on}i}, \delta^*_i$ affects when activation occurs and how long it lasts in absolute time
+4. **Gene-specific temporal pattern uncertainty**: Prior uncertainty in $t^*_{on,i}, \delta^*_i$ affects when activation occurs and how long it lasts in absolute time
 
@@
 1. **Parameter uncertainty**: $\text{Var}[\theta | y_{obs}]$ from the posterior distribution over all model parameters
-2. **Temporal uncertainty**: Uncertainty in latent time coordinates $t^*_j$ and process duration $T_{\max}$
+2. **Temporal uncertainty**: Uncertainty in latent time coordinates $t^*_j$ and process duration $T_M$
 3. **Kinetic uncertainty**: Uncertainty in gene-specific kinetic parameters $\gamma^*_i$ and fold-changes $R_{\text{on},i}$
@@
 - **Non-temporal gene parameters**: $\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$, $R_{\text{on},i} \sim \text{LogNormal}(\log(2.5), 0.4^2)$
-- **Independent temporal gene parameters**: $t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$, $\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$
+- **Independent temporal gene parameters**: $t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$, $\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$
 - **Characteristic scales**: $U_{0i} \sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)$
-- **Process duration**: $T_{\max} \sim \text{Gamma}(\alpha_{T}, \beta_{T})$
-- **Cell-specific times**: $t^*_j \sim \text{Uniform}(0, T_{\max})$
+- **Process duration**: $T_M \sim \text{Gamma}(\alpha_{T}, \beta_{T})$
+- **Cell-specific times**: $t^*_j \sim \text{Uniform}(0, T_M)$
 - **Capture efficiencies**: $\lambda_j \sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)$
@@
 
-$$u^*_{ij}(\theta), s^*_{ij}(\theta) = f_{\text{piecewise}}(t^*_j, \gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i)$$
+$$u^*_{ij}(\theta), s^*_{ij}(\theta) = f_{\text{piecewise}}(t^*_j, \gamma^*_i, R_{\text{on},i}, t^*_{on,i}, \delta^*_i)$$
 
@@
 
-1. Compute phase assignments for each cell-gene pair using ${{t^{\ast}_{j}}}^{(s)}$, ${{t^{\ast}_{0,\text{on}i}}}^{(s)}$, ${{\delta^{\ast}_i}}^{(s)}$
+1. Compute phase assignments for each cell-gene pair using ${{t^{\ast}_{j}}}^{(s)}$, ${{t^{\ast}_{on,i}}}^{(s)}$, ${{\delta^{\ast}_i}}^{(s)}$
 2.
-Evaluate analytical solutions: ${{u^*_{ij}}}^{(s)}, {{s^*_{ij}}}^{(s)} = f_{\text{piecewise}}({{t^*_{j}}}^{(s)}, {{\gamma^*_i}}^{(s)}, {{R_{\text{on},i}}}^{(s)}, {{t^*_{0,\text{on}i}}}^{(s)}, {{\delta^*_i}}^{(s)})$
+Evaluate analytical solutions: ${{u^*_{ij}}}^{(s)}, {{s^*_{ij}}}^{(s)} = f_{\text{piecewise}}({{t^*_{j}}}^{(s)}, {{\gamma^*_i}}^{(s)}, {{R_{\text{on},i}}}^{(s)}, {{t^*_{on,i}}}^{(s)}, {{\delta^*_i}}^{(s)})$
 3. Sample observations:
@@
 1. **Kinetic parameter uncertainty**: Posterior distributions over $\gamma^*_i$ and $R_{\text{on},i}$ affect the shape and magnitude of concentration trajectories
-2. **Temporal parameter uncertainty**: Uncertainty in $t^*_{0,\text{on}i}$ and $\delta^*_i$ affects when activation occurs and how long it lasts in absolute time
-3. **Process duration uncertainty**: Uncertainty in $T_{\max}$ affects the temporal scope and constrains cell temporal positions
+2. **Temporal parameter uncertainty**: Uncertainty in $t^*_{on,i}$ and $\delta^*_i$ affects when activation occurs and how long it lasts in absolute time
+3. **Process duration uncertainty**: Uncertainty in $T_M$ affects the temporal scope and constrains cell temporal positions
 4. **Cell-specific uncertainty**: Uncertainty in individual $t^*_j$ and $\lambda_j$ affects cell-specific predictions
@@
 
-1. **Recover independent piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{0,\text{on}i}$, $\delta^*_i$
-2. **Maintain process duration inference**: Recover $T_{\max}$ and cell temporal positions $t^*_j$
+1. **Recover independent piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{on,i}$, $\delta^*_i$
+2. **Maintain process duration inference**: Recover $T_M$ and cell temporal positions $t^*_j$
 3. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
@@
 7. **Validate parameter identifiability**: Confirm elimination of scaling symmetry through independent parameterization
-8. **Assess parameter separation**: Verify clean separation between kinetic ($\gamma^*_i$, $R_{\text{on},i}$) and temporal ($t^*_{0,\text{on}i}$, $\delta^*_i$) parameter inference
+8. **Assess parameter separation**: Verify clean separation between kinetic ($\gamma^*_i$, $R_{\text{on},i}$) and temporal ($t^*_{on,i}$, $\delta^*_i$) parameter inference
 9. **Validate independent parameter framework**: Confirm that independent temporal parameters enable reliable parameter recovery across different experimental contexts

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): generalize hyperparameter relationships based on T_M
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
-- $\delta^*_i < 2.0$ (short activation duration relative to process, assuming conservative prior Gamma(5,1))
+- $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ (early activation within process duration)
+- $\delta^*_i < 0.4 \mathbb{E}[T_M]$ (short activation duration relative to process)
 - $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)
@@
 
-- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
-- $\delta^*_i > 2.5$ (long activation duration relative to process, assuming conservative prior Gamma(5,1))
+- $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ (early activation within process duration)
+- $\delta^*_i > 0.5 \mathbb{E}[T_M]$ (long activation duration relative to process)
 - $R_{\text{on},i} > 2.0$ (strong fold-change during activation)
@@
 
-With independent parameterization, gene timing parameters should be calibrated to $T_M$ expectations:
-
-\begin{align}
-t^*_{on,i} &\sim \text{Normal}(0.3 \times 5, \sigma_{on}^2) = \text{Normal}(1.5, \sigma_{on}^2) \quad \text{(expect activation at 30% of process)} \\
-\delta^*_i &\sim \text{LogNormal}(\log(0.4 \times 5), \sigma_{\delta}^2) = \text{LogNormal}(\log(2.0), \sigma_{\delta}^2) \quad \text{(expect 40% duration activation)}
-\end{align}
+With independent parameterization, gene timing parameters should be calibrated to the expected process duration based on chosen $T_M$ hyperparameters:
+
+**General principle:**
+\begin{align}
+t^*_{on,i} &\sim \text{Normal}(f_{on} \times \mathbb{E}[T_M], \sigma_{on}^2) \\
+\delta^*_i &\sim \text{LogNormal}(\log(f_{\delta} \times \mathbb{E}[T_M]), \sigma_{\delta}^2)
+\end{align}
+
+where $f_{on} = 0.3$ (expect activation at 30% of process) and $f_{\delta} = 0.4$ (expect 40% duration activation) are scaling factors based on biological expectations.
+
+**Concrete examples for different experimental contexts:**
+- **Conservative (Gamma(5,1))**: $\mathbb{E}[T_M] = 5$ → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
+- **Skeptical (Gamma(1,1))**: $\mathbb{E}[T_M] = 1$ → $t^*_{on,i} \sim \text{Normal}(0.3, \sigma_{on}^2)$
+- **Permissive (Gamma(1,0.2))**: $\mathbb{E}[T_M] = 5$ → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
 
@@
 # Prior hyperparameters (updated for independent absolute parameterization)
-# Assuming T_M ~ Gamma(5.0, 1.0) with E[T_M] = 5.0
-E_T_M = 5.0
+# Adaptive to chosen T_M hyperparameters
+alpha_T, beta_T = 5.0, 1.0  # Conservative prior example
+E_T_M = alpha_T / beta_T  # Expected process duration
 priors = {
     'R_on': {'loc': 0.916, 'scale': 0.4},                           # log(2.5), fold-change (LogNormal)
-    't_on_star': {'loc': 1.5, 'scale': 0.8},                       # Normal(1.5, 0.8²), absolute onset
-    'delta_star': {'loc': np.log(2.0), 'scale': 0.45},             # log(2.0), absolute duration (LogNormal)
+    't_on_star': {'loc': 0.3 * E_T_M, 'scale': 0.8},               # Normal(0.3*E[T_M], 0.8²), absolute onset
+    'delta_star': {'loc': np.log(0.4 * E_T_M), 'scale': 0.45},     # log(0.4*E[T_M]), absolute duration (LogNormal)
 }
@@
         ('t_on_star', '>', 0.0),                    # Activation within observation window
-        ('t_on_star', '<', 1.5),                   # Early enough to complete cycle
-        ('delta_star', '<', 2.0),                  # Short pulse duration
+        ('t_on_star', '<', 0.3 * E_T_M),           # Early enough to complete cycle
+        ('delta_star', '<', 0.4 * E_T_M),          # Short pulse duration
         ('R_on', '>', 2.0),                        # Sufficient fold change to observe
@@
         ('t_on_star', '>', 0.0),                    # Activation within observation window
-        ('t_on_star', '<', 1.5),                   # Early activation onset
-        ('delta_star', '>', 2.5),                  # Long activation duration
+        ('t_on_star', '<', 0.3 * E_T_M),           # Early activation onset
+        ('delta_star', '>', 0.5 * E_T_M),          # Long activation duration
         ('R_on', '>', 2.0),                        # Strong fold change
@@
 - **Pre-activation**: $t^*_{on,i} < 0.0$ and $R_{\text{on},i} > 2.0$
-- **Transient**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i < 2.0$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))
-- **Sustained**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i > 2.5$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))
+- **Transient**: $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$, $\delta^*_i < 0.4 \mathbb{E}[T_M]$, and $R_{\text{on},i} > 2.0$
+- **Sustained**: $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$, $\delta^*_i > 0.5 \mathbb{E}[T_M]$, and $R_{\text{on},i} > 2.0$
+
+where $\mathbb{E}[T_M] = \alpha_T/\beta_T$ for the Gamma prior, so these thresholds adapt to the chosen hyperparameters.
 
@@
 
-- **$t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around 1.5 to expect activation at 30% of process duration (assuming conservative prior Gamma(5,1) with mean 5).
+- **$t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around $0.3 \times \mathbb{E}[T_M]$ to expect activation at 30% of process duration.
 **Negative values** enable pre-activation scenarios where genes were already activated before observation began.
 
-- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around 2.0 to expect 40% duration activation (assuming conservative prior Gamma(5,1) with mean 5).
+- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around $0.4 \times \mathbb{E}[T_M]$ to expect 40% duration activation.
 This enables both transient (short duration) and sustained (long duration) activation patterns.
@@
 - **Pre-activation** ($t^*_{on,i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
-- **Transient** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles (assuming conservative prior Gamma(5,1))
-- **Sustained** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation (assuming conservative prior Gamma(5,1))
+- **Transient** ($0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ and $\delta^*_i < 0.4 \mathbb{E}[T_M]$): Balanced probability for observable complete cycles
+- **Sustained** ($0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ and $\delta^*_i > 0.5 \mathbb{E}[T_M]$): Balanced probability for long-duration activation
 

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): fix notation in model development overview
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-- **Time coordinates**: Independent latent $t^*_j$ coordinates sampled uniformly within process duration $T_{\max}$
+- **Time coordinates**: Independent latent $t^*_j$ coordinates sampled uniformly within process duration $T_M$
 - **Dynamics**: Piecewise constant transcription with three phases (off, on, return-to-off)
-- **Parameters**: $\gamma^*_i$, $R_{\text{on},i}$ (fold-change), $t^*_{0,\text{on}i}$ (onset), $\delta^*_i$ (duration), with independent temporal structure
+- **Parameters**: $\gamma^*_i$, $R_{\text{on},i}$ (fold-change), $t^*_{on,i}$ (onset), $\delta^*_i$ (duration), with independent temporal structure
 - **Key innovation**: Fixed reference state ($\alpha^*_{\text{off}i} = 1.0$) eliminates parameter redundancy

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): standardize on `t^*_{on,i}`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 $$\alpha^*_i(t^*) := \begin{cases}
-1.0 & \text{if } t^* < t^*_{0,\text{on}i}, \\
-R_{\text{on},i} & \text{if } t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i, \\
-1.0 & \text{if } t^* \geq t^*_{0,\text{on}i} + \delta^*_i
+1.0 & \text{if } t^* < t^*_{on,i}, \\
+R_{\text{on},i} & \text{if } t^*_{on,i} \leq t^* < t^*_{on,i} + \delta^*_i, \\
+1.0 & \text{if } t^* \geq t^*_{on,i} + \delta^*_i
 \end{cases}$$
@@
 - $R_{\text{on},i} = \alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} = \alpha^*_{\text{on}i}$ is the fold-change during activation (inferred parameter)
-- $t^*_{0,\text{on}i}$ is the dimensionless activation onset time (can be negative for pre-activation scenarios)
+- $t^*_{on,i}$ is the dimensionless activation onset time (can be negative for pre-activation scenarios)
 - $\delta^*_i$ is the dimensionless activation duration
@@
 
-When $t^*_{0,\text{on}i} < 0$:
+When $t^*_{on,i} < 0$:
 
@@
 
-#### **Phase 1: Off State** ($t^* < t^*_{0,\text{on}i}$)
+#### **Phase 1: Off State** ($t^* < t^*_{on,i}$)
 
@@
 
-**Note**: This phase may not be observed if $t^*_{0,\text{on}i} < 0$ (pre-activation scenario).
-
-#### **Phase 2: On State** ($t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i$)
-
-Let $\tau_{\text{on}} = t^* - t^*_{0,\text{on}i}$ be the time since activation onset. The system transitions from the off-state steady state toward the on-state steady state.
+**Note**: This phase may not be observed if $t^*_{on,i} < 0$ (pre-activation scenario).
+
+#### **Phase 2: On State** ($t^*_{on,i} \leq t^* < t^*_{on,i} + \delta^*_i$)
+
+Let $\tau_{\text{on}} = t^* - t^*_{on,i}$ be the time since activation onset. The system transitions from the off-state steady state toward the on-state steady state.
 
@@
 
-#### **Phase 3: Return to Off State** ($t^* \geq t^*_{0,\text{on}i} + \delta^*_i$)
-
-Let $\tau_{\text{off}} = t^* - (t^*_{0,\text{on}i} + \delta^*_i)$ be the time since returning to the off state.
-The system transitions from the end-of-phase-2 state back toward the off-state steady state.
+#### **Phase 3: Return to Off State** ($t^* \geq t^*_{on,i} + \delta^*_i$)
+
+Let $\tau_{\text{off}} = t^* - (t^*_{on,i} + \delta^*_i)$ be the time since returning to the off state. The system transitions from the end-of-phase-2 state back toward the off-state steady state.
 
@@
 
-- $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ (early activation within process duration)
-- $\delta^*_i < 0.4 \mathbb{E}[T_M]$ (short activation duration relative to process)
+- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
+- $\delta^*_i < 2.0$ (short activation duration relative to process, assuming conservative prior Gamma(5,1))
 - $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)
@@
 
-- $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ (early activation within process duration)
-- $\delta^*_i > 0.5 \mathbb{E}[T_M]$ (long activation duration relative to process)
+- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
+- $\delta^*_i > 2.5$ (long activation duration relative to process, assuming conservative prior Gamma(5,1))
 - $R_{\text{on},i} > 2.0$ (strong fold-change during activation)
@@
 \begin{align}
-t^*_{on,i} &\sim \text{Normal}(f_{on} \times \mathbb{E}[T_M], \sigma_{on}^2) \\
-\delta^*_i &\sim \text{LogNormal}(\log(f_{\delta} \times \mathbb{E}[T_M]), \sigma_{\delta}^2)
+t^*_{on,i} &\sim \text{Normal}(f_{on} \times \text{E}[T_{\max}], \sigma_{on}^2) \\
+\delta^*_i &\sim \text{LogNormal}(\log(f_{\delta} \times \text{E}[T_{\max}]), \sigma_{\delta}^2)
 \end{align}
@@
 **Concrete examples for different experimental contexts:**
-- **Conservative (Gamma(5,1))**: $\mathbb{E}[T_M] = 5$ → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
-- **Skeptical (Gamma(1,1))**: $\mathbb{E}[T_M] = 1$ → $t^*_{on,i} \sim \text{Normal}(0.3, \sigma_{on}^2)$
-- **Permissive (Gamma(1,0.2))**: $\mathbb{E}[T_M] = 5$ → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
+- **Conservative (Gamma(5,1))**: E[$T_{\max}$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
+- **Skeptical (Gamma(1,1))**: E[$T_{\max}$] = 1 → $t^*_{on,i} \sim \text{Normal}(0.3, \sigma_{on}^2)$
+- **Permissive (Gamma(1,0.2))**: E[$T_{\max}$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
 
@@
 The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while using independent temporal parameterization to eliminate scaling symmetry.
-For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of independent temporal gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$.
+For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of independent temporal gene parameters $(t^*_{on,i}, \delta^*_i)$.
 Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.
@@
 - **Pre-activation**: $t^*_{on,i} < 0.0$ and $R_{\text{on},i} > 2.0$
-- **Transient**: $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$, $\delta^*_i < 0.4 \mathbb{E}[T_M]$, and $R_{\text{on},i} > 2.0$
-- **Sustained**: $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$, $\delta^*_i > 0.5 \mathbb{E}[T_M]$, and $R_{\text{on},i} > 2.0$
-
-where $\mathbb{E}[T_M] = \alpha_T/\beta_T$ for the Gamma prior, so these thresholds adapt to the chosen hyperparameters.
+- **Transient**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i < 2.0$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))
+- **Sustained**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i > 2.5$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))
 
@@
 
-- **$t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around $0.3 \times \mathbb{E}[T_M]$ to expect activation at 30% of process duration.
+- **$t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around 1.5 to expect activation at 30% of process duration (assuming conservative prior Gamma(5,1) with mean 5).
 **Negative values** enable pre-activation scenarios where genes were already activated before observation began.
 
-- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around $0.4 \times \mathbb{E}[T_M]$ to expect 40% duration activation.
+- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around 2.0 to expect 40% duration activation (assuming conservative prior Gamma(5,1) with mean 5).
 This enables both transient (short duration) and sustained (long duration) activation patterns.
@@
 - **Pre-activation** ($t^*_{on,i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
-- **Transient** ($0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ and $\delta^*_i < 0.4 \mathbb{E}[T_M]$): Balanced probability for observable complete cycles
-- **Sustained** ($0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ and $\delta^*_i > 0.5 \mathbb{E}[T_M]$): Balanced probability for long-duration activation
+- **Transient** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles (assuming conservative prior Gamma(5,1))
+- **Sustained** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation (assuming conservative prior Gamma(5,1))
 

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): add description of dimensionless time and restoration of scaling
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
+### Dimensionless Framework and Biological Timescale Connection
+
+The dimensionless parameterization requires careful interpretation to connect mathematical parameters to biological reality.
+This section clarifies the relationship between dimensionless temporal parameters and real biological timescales.
+
+#### **Fundamental Dimensionless Transformation**
+
+The dimensionless framework transforms real time into natural units based on characteristic biological rates:
+
+$$t^* = t_{\text{real}} \times \alpha_{\text{ref}}$$
+
+where:
+- $t_{\text{real}}$ is time in physical units (hours, days, etc.)
+- $\alpha_{\text{ref}} = \alpha^*_{\text{off}} = 1.0$ is the reference transcription rate (fixed)
+- $t^*$ is dimensionless time in units of characteristic process rates
+
+**Key insight**: All temporal parameters are expressed as multiples of the natural timescale $1/\alpha_{\text{ref}}$, where the characteristic timescale is typically the splicing/degradation timescale.
+
+#### **Process Duration Parameter: Biological Interpretation**
+
+The process duration $T_{\max}$ represents how many characteristic timescales the biological process spans:
+
+$$T_{\max} = \frac{\text{Biological process duration}}{\text{Characteristic timescale of cellular rates}}$$
+
+**Biological interpretation examples:**
+- $T_{\max} = 1$: Process duration equals one characteristic timescale
+- $T_{\max} = 5$: Process duration equals five characteristic timescales
+- $T_{\max} = 0.5$: Process duration is half a characteristic timescale
+
+**Steady-state achievement**: The fraction of steady-state dynamics observable is $1 - e^{-T_{\max}}$:
+- $T_{\max} = 1$: Observes 63% approach to steady state
+- $T_{\max} = 3$: Observes 95% approach to steady state
+- $T_{\max} = 5$: Observes >99% approach to steady state
+
+#### **Connection to Real Timescales During Model Fitting**
+
+When fitting the model to experimental data, the inferred $T_{\max}$ reveals the underlying biological rates:
+
+$$\text{Inferred characteristic timescale} = \frac{\text{Known experimental duration}}{T_{\max}^{\text{posterior}}}$$
+
+**Practical example:**
+- **Experimental setup**: 24-hour differentiation time course
+- **Posterior estimate**: $T_{\max} \approx 3.0$
+- **Inferred characteristic timescale**: $24/3 = 8$ hours
+- **Biological interpretation**: Cellular processes (splicing, degradation) occur on ~8 hour timescales
+
+This bidirectional relationship enables both forward simulation (given biological rates) and inverse inference (estimating rates from data).
+
+#### **Gene Parameter Interpretation in Dimensionless Framework**
+
+**Activation Onset Times**: $t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$
+
+- **Dimensionless interpretation**: Gene $i$ activates at $t^*_{0,\text{on}i}$ characteristic timescales after process initiation
+- **Real-time interpretation**: Gene $i$ activates at $t^*_{0,\text{on}i}/\alpha_{\text{ref}}$ hours after process start
+- **Relative interpretation**: Gene $i$ activates at $(t^*_{0,\text{on}i}/T_{\max}) \times 100\%$ through the process
+
+**Example interpretations:**
+- $t^*_{0,\text{on}i} = 1.5$, $T_{\max} = 5.0$: Gene activates 30% through the process
+- $t^*_{0,\text{on}i} = -0.5$: Gene was already active 0.5 timescales before observation began
+
+**Activation Durations**: $\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$
+
+- **Dimensionless interpretation**: Gene $i$ remains active for $\delta^*_i$ characteristic timescales
+- **Real-time interpretation**: Gene $i$ remains active for $\delta^*_i/\alpha_{\text{ref}}$ hours
+- **Relative interpretation**: Gene $i$ remains active for $(\delta^*_i/T_{\max}) \times 100\%$ of the process
+
+#### **Hyperparameter Choice as Biological Prior Knowledge**
+
+The choice of $T_{\max}$ hyperparameters encodes specific biological assumptions about process duration:
+
+**Conservative Prior: $T_{\max} \sim \text{Gamma}(5.0, 1.0)$**
+
+- $\mathbb{E}[T_{\max}] = 5$, $\text{Var}[T_{\max}] = 5$
+- **Biological assumption**: "I expect this process to span approximately 5 characteristic timescales, allowing observation of near-complete dynamics"
+- **Implications**: If experimental duration is 20 hours, assumes characteristic timescale ≈ 4 hours
+- **Expectation**: Observes 99% approach to steady state ($1 - e^{-5} \approx 0.99$)
+- **Data burden**: Data must provide strong evidence if process is much shorter or longer
+
+**Permissive Prior: $T_{\max} \sim \text{Gamma}(1.0, 0.2)$**
+
+- $\mathbb{E}[T_{\max}] = 5$, $\text{Var}[T_{\max}] = 25$
+- **Biological assumption**: "I'm uncertain about process duration - it could range from rapid (1 timescale) to very slow (15+ timescales)"
+- **Implications**: Wide uncertainty allows data to determine appropriate timescale
+- **Flexibility**: Can accommodate both early transient dynamics and complete processes
+- **Data burden**: Minimal prior bias on process duration
+
+**Skeptical Prior: $T_{\max} \sim \text{Gamma}(1.0, 1.0)$**
+
+- $\mathbb{E}[T_{\max}] = 1$, $\text{Var}[T_{\max}] = 1$
+- **Biological assumption**: "I expect rapid cellular responses, likely observing only early dynamics"
+- **Implications**: Assumes observation of primarily transient phase (63% approach: $1 - e^{-1} \approx 0.63$)
+- **Data burden**: Data must strongly support longer timescales if present
+- **Use cases**: Stress responses, drug treatments, acute perturbations
+
 **Priors for other gene-specific parameters:**
@@
 \end{align}
+
+#### **Model Fitting and Timescale Inference**
+
+The dimensionless framework creates a bidirectional relationship between process duration and characteristic rates:
+
+**Forward (simulation):**
+
+1. Choose biological timescales (e.g., β = splicing rate)
+2. Set process duration in real time (e.g., 24 hours)
+3. Compute $T_{\max} = t_{\text{process}} \times β$
+4. Generate dimensionless dynamics
+
+**Inverse (inference from data):**
+
+1. Observe expression data over known duration
+2. Infer $T_{\max}$ from temporal patterns
+3. Estimate characteristic rates: $\hat{β} = T_{\max}/t_{\text{experiment}}$
+4. Convert all dimensionless parameters to real time
+
+**Practical Example:**
+
+**Experimental setup:** 48-hour differentiation time course
+
+**Model fitting results:**
+
+- Posterior: $T_{\max} \approx 6.0 \pm 0.5$
+- Inferred characteristic timescale: $48/6 = 8$ hours
+- Gene with $t^*_{0,\text{on}i} = 2.0$: Activates at $2.0 \times 8 = 16$ hours
+- Gene with $\delta^*_i = 1.5$: Active for $1.5 \times 8 = 12$ hours
+
+#### **Independent Parameterization Justification**
+
+In the dimensionless framework, **gene timing parameters represent intrinsic biological properties** that should not scale with the arbitrary choice of observation window:
+
+- $t^*_{0,\text{on}i}$: When a gene's regulatory program activates (intrinsic timing)
+- $\delta^*_i$: How long a gene's regulatory program lasts (intrinsic duration)
+
+**These are biological constants** that don't change based on how long we choose to observe the system. Therefore:
+
+\begin{align}
+t^*_{0,\text{on}i} &\sim \text{Normal}(1.5, 0.8^2) \quad \text{(fixed biological timing)} \\
+\delta^*_i &\sim \text{LogNormal}(0.0, 0.45^2) \quad \text{(fixed biological duration)}
+\end{align}
+
+**NOT dependent on $T_{\max}$:**
+\begin{align}
+t^*_{0,\text{on}i} &\not\sim \text{Normal}(0.3 \times T_{\max}, 0.8^2) \quad \text{(would create artificial scaling)}
+\end{align}
+
+This independence ensures that gene regulatory timing is treated as an intrinsic biological property rather than an artifact of experimental design.
+
+### Identifiability Proof for Independent Parameterization
+
+**Theorem:** The piecewise activation likelihood with independent parameterization is not invariant under parameter scaling.
+
+**Proof:** For any scaling factor $\lambda \neq 1$:
+
+1. **Gene parameters are sampled independently**: $t^*_{0,\text{on}i} \not\propto T_{\max}$ and $\delta^*_i \not\propto T_{\max}$
+
+2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
+   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{0,\text{on}i}$
+   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{0,\text{on}i}$ (unchanged)
+
+3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$**:
+   - Phase boundaries change when cell times are scaled but gene times are not
+   - Different phase assignments lead to different analytical solution forms
+
+4. **Therefore**: $\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$ □
+
+**Corollary:** Parameter recovery is uniquely identifiable because no two distinct parameter configurations can produce identical likelihoods.
 

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): remove additional explanation of identifiability
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-### Identifiability Proof for Independent Parameterization
-
-**Theorem:** The piecewise activation likelihood with independent parameterization is not invariant under parameter scaling.
-
-**Proof:** For any scaling factor $\lambda \neq 1$:
-
-1. **Gene parameters are sampled independently**: $t^*_{on,i} \not\propto T_M$ and $\delta^*_i \not\propto T_M$
-
-2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
-   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{on,i}$
-   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{on,i}$ (unchanged)
-
-3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{on,i}$**:
-   - Phase boundaries change when cell times are scaled but gene times are not
-   - Different phase assignments lead to different analytical solution forms
-
-4. **Therefore**: $\mathcal{L}(T_M, \{t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_M, \{\lambda t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\})$ □
-
-**Corollary:** Parameter recovery is uniquely identifiable because no two distinct parameter configurations can produce identical likelihoods.
-
 ### Mathematical Framework for Pre-Activation Scenarios

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): replace T_max with T_M
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

--- 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
@@
 \begin{align}
-t^*_{on,i} &\sim \text{Normal}(f_{on} \times \text{E}[T_{\max}], \sigma_{on}^2) \\
-\delta^*_i &\sim \text{LogNormal}(\log(f_{\delta} \times \text{E}[T_{\max}]), \sigma_{\delta}^2)
+t^*_{on,i} &\sim \text{Normal}(f_{on} \times \text{E}[T_M], \sigma_{on}^2) \\
+\delta^*_i &\sim \text{LogNormal}(\log(f_{\delta} \times \text{E}[T_M]), \sigma_{\delta}^2)
 \end{align}
@@
 **Concrete examples for different experimental contexts:**
-- **Conservative (Gamma(5,1))**: E[$T_{\max}$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
-- **Skeptical (Gamma(1,1))**: E[$T_{\max}$] = 1 → $t^*_{on,i} \sim \text{Normal}(0.3, \sigma_{on}^2)$
-- **Permissive (Gamma(1,0.2))**: E[$T_{\max}$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
+- **Conservative (Gamma(5,1))**: E[$T_M$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
+- **Skeptical (Gamma(1,1))**: E[$T_M$] = 1 → $t^*_{on,i} \sim \text{Normal}(0.3, \sigma_{on}^2)$
+- **Permissive (Gamma(1,0.2))**: E[$T_M$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
 
 This ensures gene parameters are scaled appropriately to the expected process duration while maintaining independence and supporting the three piecewise activation patterns (pre-activation, transient, sustained).
-
-### Identifiability Proof for Independent Parameterization
-
-**Theorem:** The piecewise activation likelihood with independent parameterization is not invariant under parameter scaling.
-
-**Proof:** For any scaling factor $\lambda \neq 1$:
-
-1. **Gene parameters are sampled independently**: $t^*_{on,i} \not\propto T_M$ and $\delta^*_i \not\propto T_M$
-
-2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
-   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{on,i}$
-   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{on,i}$ (unchanged)
-
-3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{on,i}$**:
-   - Phase boundaries change when cell times are scaled but gene times are not
-   - Different phase assignments lead to different analytical solution forms
-
-4. **Therefore**: $\mathcal{L}(T_M, \{t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_M, \{\lambda t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\})$ □
-
-**Corollary:** Parameter recovery is uniquely identifiable because no two distinct parameter configurations can produce identical likelihoods.
 
@@
 
+### Dimensionless Framework and Biological Timescale Connection
+
+The dimensionless parameterization requires careful interpretation to connect mathematical parameters to biological reality.
+This section clarifies the relationship between dimensionless temporal parameters and real biological timescales.
+
+#### **Fundamental Dimensionless Transformation**
+
+The dimensionless framework transforms real time into natural units based on characteristic biological rates:
+
+$$t^* = t_{\text{real}} \times \alpha_{\text{ref}}$$
+
+where:
+- $t_{\text{real}}$ is time in physical units (hours, days, etc.)
+- $\alpha_{\text{ref}} = \alpha^*_{\text{off}} = 1.0$ is the reference transcription rate (fixed)
+- $t^*$ is dimensionless time in units of characteristic process rates
+
+**Key insight**: All temporal parameters are expressed as multiples of the natural timescale $1/\alpha_{\text{ref}}$, where the characteristic timescale is typically the splicing/degradation timescale.
+
+#### **Process Duration Parameter: Biological Interpretation**
+
+The process duration $T_M$ represents how many characteristic timescales the biological process spans:
+
+$$T_M = \frac{\text{Biological process duration}}{\text{Characteristic timescale of cellular rates}}$$
+
+**Biological interpretation examples:**
+
+- $T_M = 1$: Process duration equals one characteristic timescale
+- $T_M = 5$: Process duration equals five characteristic timescales
+- $T_M = 0.5$: Process duration is half a characteristic timescale
+
+**Steady-state achievement**: The fraction of steady-state dynamics observable is $1 - e^{-T_M}$:
+- $T_M = 1$: Observes 63% approach to steady state
+- $T_M = 3$: Observes 95% approach to steady state
+- $T_M = 5$: Observes >99% approach to steady state
+
+#### **Connection to Real Timescales During Model Fitting**
+
+When fitting the model to experimental data, the inferred $T_M$ reveals the underlying biological rates:
+
+$$\text{Inferred characteristic timescale} = \frac{\text{Known experimental duration}}{T_M^{\text{posterior}}}$$
+
+**Practical example:**
+- **Experimental setup**: 24-hour differentiation time course
+- **Posterior estimate**: $T_M \approx 3.0$
+- **Inferred characteristic timescale**: $24/3 = 8$ hours
+- **Biological interpretation**: Cellular processes (splicing, degradation) occur on ~8 hour timescales
+
+This bidirectional relationship enables both forward simulation (given biological rates) and inverse inference (estimating rates from data).
+
+#### **Gene Parameter Interpretation in Dimensionless Framework**
+
+**Activation Onset Times**: $t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$
+
+- **Dimensionless interpretation**: Gene $i$ activates at $t^*_{0,\text{on}i}$ characteristic timescales after process initiation
+- **Real-time interpretation**: Gene $i$ activates at $t^*_{0,\text{on}i}/\alpha_{\text{ref}}$ hours after process start
+- **Relative interpretation**: Gene $i$ activates at $(t^*_{0,\text{on}i}/T_M) \times 100\%$ through the process
+
+**Example interpretations:**
+- $t^*_{0,\text{on}i} = 1.5$, $T_M = 5.0$: Gene activates 30% through the process
+- $t^*_{0,\text{on}i} = -0.5$: Gene was already active 0.5 timescales before observation began
+
+**Activation Durations**: $\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$
+
+- **Dimensionless interpretation**: Gene $i$ remains active for $\delta^*_i$ characteristic timescales
+- **Real-time interpretation**: Gene $i$ remains active for $\delta^*_i/\alpha_{\text{ref}}$ hours
+- **Relative interpretation**: Gene $i$ remains active for $(\delta^*_i/T_M) \times 100\%$ of the process
+
+#### **Hyperparameter Choice as Biological Prior Knowledge**
+
+The choice of $T_M$ hyperparameters encodes specific biological assumptions about process duration:
+
+**Conservative Prior: $T_M \sim \text{Gamma}(5.0, 1.0)$**
+
+- $\mathbb{E}[T_M] = 5$, $\text{Var}[T_M] = 5$
+- **Biological assumption**: "I expect this process to span approximately 5 characteristic timescales, allowing observation of near-complete dynamics"
+- **Implications**: If experimental duration is 20 hours, assumes characteristic timescale ≈ 4 hours
+- **Expectation**: Observes 99% approach to steady state ($1 - e^{-5} \approx 0.99$)
+- **Data burden**: Data must provide strong evidence if process is much shorter or longer
+
+**Permissive Prior: $T_M \sim \text{Gamma}(1.0, 0.2)$**
+
+- $\mathbb{E}[T_M] = 5$, $\text{Var}[T_M] = 25$
+- **Biological assumption**: "I'm uncertain about process duration - it could range from rapid (1 timescale) to very slow (15+ timescales)"
+- **Implications**: Wide uncertainty allows data to determine appropriate timescale
+- **Flexibility**: Can accommodate both early transient dynamics and complete processes
+- **Data burden**: Minimal prior bias on process duration
+
+**Skeptical Prior: $T_M \sim \text{Gamma}(1.0, 1.0)$**
+
+- $\mathbb{E}[T_M] = 1$, $\text{Var}[T_M] = 1$
+- **Biological assumption**: "I expect rapid cellular responses, likely observing only early dynamics"
+- **Implications**: Assumes observation of primarily transient phase (63% approach: $1 - e^{-1} \approx 0.63$)
+- **Data burden**: Data must strongly support longer timescales if present
+- **Use cases**: Stress responses, drug treatments, acute perturbations
+
 **Priors for other gene-specific parameters:**
@@
 \end{align}
+
+#### **Model Fitting and Timescale Inference**
+
+The dimensionless framework creates a bidirectional relationship between process duration and characteristic rates:
+
+**Forward (simulation):**
+
+1. Choose biological timescales (e.g., β = splicing rate)
+2. Set process duration in real time (e.g., 24 hours)
+3. Compute $T_M = t_{\text{process}} \times β$
+4. Generate dimensionless dynamics
+
+**Inverse (inference from data):**
+
+1. Observe expression data over known duration
+2. Infer $T_M$ from temporal patterns
+3. Estimate characteristic rates: $\hat{β} = T_M/t_{\text{experiment}}$
+4. Convert all dimensionless parameters to real time
+
+**Practical Example:**
+
+**Experimental setup:** 48-hour differentiation time course
+
+**Model fitting results:**
+
+- Posterior: $T_M \approx 6.0 \pm 0.5$
+- Inferred characteristic timescale: $48/6 = 8$ hours
+- Gene with $t^*_{0,\text{on}i} = 2.0$: Activates at $2.0 \times 8 = 16$ hours
+- Gene with $\delta^*_i = 1.5$: Active for $1.5 \times 8 = 12$ hours
+
+#### **Independent Parameterization Justification**
+
+In the dimensionless framework, **gene timing parameters represent intrinsic biological properties** that should not scale with the arbitrary choice of observation window:
+
+- $t^*_{0,\text{on}i}$: When a gene's regulatory program activates (intrinsic timing)
+- $\delta^*_i$: How long a gene's regulatory program lasts (intrinsic duration)
+
+**These are biological constants** that don't change based on how long we choose to observe the system. Therefore:
+
+\begin{align}
+t^*_{0,\text{on}i} &\sim \text{Normal}(1.5, 0.8^2) \quad \text{(fixed biological timing)} \\
+\delta^*_i &\sim \text{LogNormal}(0.0, 0.45^2) \quad \text{(fixed biological duration)}
+\end{align}
+
+**NOT dependent on $T_M$:**
+\begin{align}
+t^*_{0,\text{on}i} &\not\sim \text{Normal}(0.3 \times T_M, 0.8^2) \quad \text{(would create artificial scaling)}
+\end{align}
+
+This independence ensures that gene regulatory timing is treated as an intrinsic biological property rather than an artifact of experimental design.
+
+### Identifiability Proof for Independent Parameterization
+
+**Theorem:** The piecewise activation likelihood with independent parameterization is not invariant under parameter scaling.
+
+**Proof:** For any scaling factor $\lambda \neq 1$:
+
+1. **Gene parameters are sampled independently**: $t^*_{0,\text{on}i} \not\propto T_M$ and $\delta^*_i \not\propto T_M$
+
+2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
+   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{0,\text{on}i}$
+   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{0,\text{on}i}$ (unchanged)
+
+3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$**:
+   - Phase boundaries change when cell times are scaled but gene times are not
+   - Different phase assignments lead to different analytical solution forms
+
+4. **Therefore**: $\mathcal{L}(T_M, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_M, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$ □
+
+**Corollary:** Parameter recovery is uniquely identifiable because no two distinct parameter configurations can produce identical likelihoods.
 

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): replace `t^*_{0,\text{on}i}` with `t^*_{on,i}`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-**Activation Onset Times**: $t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$
-
-- **Dimensionless interpretation**: Gene $i$ activates at $t^*_{0,\text{on}i}$ characteristic timescales after process initiation
-- **Real-time interpretation**: Gene $i$ activates at $t^*_{0,\text{on}i}/\alpha_{\text{ref}}$ hours after process start
-- **Relative interpretation**: Gene $i$ activates at $(t^*_{0,\text{on}i}/T_M) \times 100\%$ through the process
+**Activation Onset Times**: $t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$
+
+- **Dimensionless interpretation**: Gene $i$ activates at $t^*_{on,i}$ characteristic timescales after process initiation
+- **Real-time interpretation**: Gene $i$ activates at $t^*_{on,i}/\alpha_{\text{ref}}$ hours after process start
+- **Relative interpretation**: Gene $i$ activates at $(t^*_{on,i}/T_M) \times 100\%$ through the process
 
 **Example interpretations:**
-- $t^*_{0,\text{on}i} = 1.5$, $T_M = 5.0$: Gene activates 30% through the process
-- $t^*_{0,\text{on}i} = -0.5$: Gene was already active 0.5 timescales before observation began
+- $t^*_{on,i} = 1.5$, $T_M = 5.0$: Gene activates 30% through the process
+- $t^*_{on,i} = -0.5$: Gene was already active 0.5 timescales before observation began
 
@@
 - Inferred characteristic timescale: $48/6 = 8$ hours
-- Gene with $t^*_{0,\text{on}i} = 2.0$: Activates at $2.0 \times 8 = 16$ hours
+- Gene with $t^*_{on,i} = 2.0$: Activates at $2.0 \times 8 = 16$ hours
 - Gene with $\delta^*_i = 1.5$: Active for $1.5 \times 8 = 12$ hours
@@
 
-- $t^*_{0,\text{on}i}$: When a gene's regulatory program activates (intrinsic timing)
+- $t^*_{on,i}$: When a gene's regulatory program activates (intrinsic timing)
 - $\delta^*_i$: How long a gene's regulatory program lasts (intrinsic duration)
@@
 \begin{align}
-t^*_{0,\text{on}i} &\sim \text{Normal}(1.5, 0.8^2) \quad \text{(fixed biological timing)} \\
+t^*_{on,i} &\sim \text{Normal}(1.5, 0.8^2) \quad \text{(fixed biological timing)} \\
 \delta^*_i &\sim \text{LogNormal}(0.0, 0.45^2) \quad \text{(fixed biological duration)}
@@
 \begin{align}
-t^*_{0,\text{on}i} &\not\sim \text{Normal}(0.3 \times T_M, 0.8^2) \quad \text{(would create artificial scaling)}
+t^*_{on,i} &\not\sim \text{Normal}(0.3 \times T_M, 0.8^2) \quad \text{(would create artificial scaling)}
 \end{align}
@@
 
-1. **Gene parameters are sampled independently**: $t^*_{0,\text{on}i} \not\propto T_M$ and $\delta^*_i \not\propto T_M$
+1. **Gene parameters are sampled independently**: $t^*_{on,i} \not\propto T_M$ and $\delta^*_i \not\propto T_M$
 
 2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
-   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{0,\text{on}i}$
-   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{0,\text{on}i}$ (unchanged)
-
-3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$**:
+   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{on,i}$
+   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{on,i}$ (unchanged)
+
+3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{on,i}$**:
    - Phase boundaries change when cell times are scaled but gene times are not
@@
 
-4. **Therefore**: $\mathcal{L}(T_M, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_M, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$ □
+4. **Therefore**: $\mathcal{L}(T_M, \{t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_M, \{\lambda t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\})$ □
 

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): fix spacing
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 **Practical example:**
+
 - **Experimental setup**: 24-hour differentiation time course
@@
 **Example interpretations:**
+
 - $t^*_{on,i} = 1.5$, $T_M = 5.0$: Gene activates 30% through the process

### 2025-06-16 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): combine sections on dimensionless timescale
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@
 
-### Biological Interpretation of Timescale Prior Hyperparameters
-
-The process duration $T_M$ determines the temporal scope of observable dynamics and requires careful hyperparameter selection based on biological expectations and experimental context.
-
-#### **Dynamical Range and Steady-State Achievement**
+### Biological Interpretation of Timescale Prior Hyperparameters and Dimensionless Framework
+
+The dimensionless parameterization requires careful interpretation to connect mathematical parameters to biological reality.
+This section clarifies the relationship between dimensionless temporal parameters and real biological timescales, with particular focus on how hyperparameter choices encode biological prior knowledge.
+
+#### **Fundamental Dimensionless Transformation**
+
+The dimensionless framework transforms real time into natural units based on characteristic biological rates:
+
+$$t^* = t_{\text{real}} \times \alpha_{\text{ref}}$$
+
+where:
+- $t_{\text{real}}$ is time in physical units (hours, days, etc.)
+- $\alpha_{\text{ref}} = \alpha^*_{\text{off}} = 1.0$ is the reference transcription rate (fixed)
+- $t^*$ is dimensionless time in units of characteristic process rates
+
+**Key insight**: All temporal parameters are expressed as multiples of the natural timescale $1/\alpha_{\text{ref}}$, where the characteristic timescale is typically the splicing/degradation timescale.
+
+#### **Process Duration Parameter: Biological Interpretation**
+
+The process duration $T_M$ represents how many characteristic timescales the biological process spans:
+
+$$T_M = \frac{\text{Biological process duration}}{\text{Characteristic timescale of cellular rates}}$$
+
+**Dynamical Range and Steady-State Achievement**
 
@@
 
-- $T_M = 1$: Observes 63% approach to steady state
-- $T_M = 3$: Observes 95% approach to steady state
-- $T_M = 5$: Observes >99% approach to steady state
+- $T_M = 1$: Process duration equals one characteristic timescale, observes 63% approach to steady state
+- $T_M = 3$: Process duration equals three characteristic timescales, observes 95% approach to steady state
+- $T_M = 5$: Process duration equals five characteristic timescales, observes >99% approach to steady state
 
 #### **Hyperparameter Strategy and Biological Prior Knowledge**
+
+The choice of $T_M$ hyperparameters encodes specific biological assumptions about process duration:
 
@@
 
-- **Biological assumption**: Process likely spans ~5 characteristic timescales
-- **Dynamical expectation**: Will observe near-complete approach to steady state
-- **Data burden**: Data must provide evidence if process is shorter than expected
+- **Biological assumption**: "I expect this process to span approximately 5 characteristic timescales, allowing observation of near-complete dynamics"
+- **Dynamical expectation**: Will observe near-complete approach to steady state (99%: $1 - e^{-5} \approx 0.99$)
+- **Data burden**: Data must provide strong evidence if process is much shorter or longer
 - **Use case**: When biological knowledge suggests complete developmental programs
@@
 
-- **Biological assumption**: Uncertain about process duration, wide range possible
+- **Biological assumption**: "I'm uncertain about process duration - it could range from rapid (1 timescale) to very slow (15+ timescales)"
 - **Dynamical expectation**: Could observe early transients or complete dynamics
-- **Data burden**: Data determines whether short or long timescales are needed
+- **Data burden**: Minimal prior bias on process duration, data determines whether short or long timescales are needed
 - **Use case**: When process duration is unknown, exploratory analysis
@@
 
-- **Biological assumption**: Most cellular processes are rapid (early dynamics only)
-- **Dynamical expectation**: Primarily observe transient phase (63% approach)
+- **Biological assumption**: "I expect rapid cellular responses, likely observing only early dynamics"
+- **Dynamical expectation**: Primarily observe transient phase (63% approach: $1 - e^{-1} \approx 0.63$)
 - **Data burden**: Data must strongly support longer timescales if present
@@
 
+#### **Connection to Real Timescales During Model Fitting**
+
+When fitting the model to experimental data, the inferred $T_M$ reveals the underlying biological rates:
+
+$$\text{Inferred characteristic timescale} = \frac{\text{Known experimental duration}}{T_M^{\text{posterior}}}$$
+
+**Practical example:**
+
+- **Experimental setup**: 24-hour differentiation time course
+- **Posterior estimate**: $T_M \approx 3.0$
+- **Inferred characteristic timescale**: $24/3 = 8$ hours
+- **Biological interpretation**: Cellular processes (splicing, degradation) occur on ~8 hour timescales
+
+This bidirectional relationship enables both forward simulation (given biological rates) and inverse inference (estimating rates from data).
+
+#### **Gene Parameter Interpretation in Dimensionless Framework**
+
+**Activation Onset Times**: $t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$
+
+- **Dimensionless interpretation**: Gene $i$ activates at $t^*_{on,i}$ characteristic timescales after process initiation
+- **Real-time interpretation**: Gene $i$ activates at $t^*_{on,i}/\alpha_{\text{ref}}$ hours after process start
+- **Relative interpretation**: Gene $i$ activates at $(t^*_{on,i}/T_M) \times 100\%$ through the process
+
+**Example interpretations:**
+
+- $t^*_{on,i} = 1.5$, $T_M = 5.0$: Gene activates 30% through the process
+- $t^*_{on,i} = -0.5$: Gene was already active 0.5 timescales before observation began
+
+**Activation Durations**: $\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$
+
+- **Dimensionless interpretation**: Gene $i$ remains active for $\delta^*_i$ characteristic timescales
+- **Real-time interpretation**: Gene $i$ remains active for $\delta^*_i/\alpha_{\text{ref}}$ hours
+- **Relative interpretation**: Gene $i$ remains active for $(\delta^*_i/T_M) \times 100\%$ of the process
+
 #### **Connection to Gene Parameter Priors**
@@
 **Concrete examples for different experimental contexts:**
+
 - **Conservative (Gamma(5,1))**: E[$T_M$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
@@
 This ensures gene parameters are scaled appropriately to the expected process duration while maintaining independence and supporting the three piecewise activation patterns (pre-activation, transient, sustained).
+
+#### **Independent Parameterization Justification**
+
+In the dimensionless framework, **gene timing parameters represent intrinsic biological properties** that should not scale with the arbitrary choice of observation window:
+
+- $t^*_{on,i}$: When a gene's regulatory program activates (intrinsic timing)
+- $\delta^*_i$: How long a gene's regulatory program lasts (intrinsic duration)
+
+**These are biological constants** that don't change based on how long we choose to observe the system. Therefore:
+
+\begin{align}
+t^*_{on,i} &\sim \text{Normal}(1.5, 0.8^2) \quad \text{(fixed biological timing)} \\
+\delta^*_i &\sim \text{LogNormal}(0.0, 0.45^2) \quad \text{(fixed biological duration)}
+\end{align}
+
+**NOT dependent on $T_M$:**
+\begin{align}
+t^*_{on,i} &\not\sim \text{Normal}(0.3 \times T_M, 0.8^2) \quad \text{(would create artificial scaling)}
+\end{align}
+
+This independence ensures that gene regulatory timing is treated as an intrinsic biological property rather than an artifact of experimental design.
 
@@
 
-### Dimensionless Framework and Biological Timescale Connection
-
-The dimensionless parameterization requires careful interpretation to connect mathematical parameters to biological reality.
-This section clarifies the relationship between dimensionless temporal parameters and real biological timescales.
-
-#### **Fundamental Dimensionless Transformation**
-
-The dimensionless framework transforms real time into natural units based on characteristic biological rates:
-
-$$t^* = t_{\text{real}} \times \alpha_{\text{ref}}$$
-
-where:
-- $t_{\text{real}}$ is time in physical units (hours, days, etc.)
-- $\alpha_{\text{ref}} = \alpha^*_{\text{off}} = 1.0$ is the reference transcription rate (fixed)
-- $t^*$ is dimensionless time in units of characteristic process rates
-
-**Key insight**: All temporal parameters are expressed as multiples of the natural timescale $1/\alpha_{\text{ref}}$, where the characteristic timescale is typically the splicing/degradation timescale.
-
-#### **Process Duration Parameter: Biological Interpretation**
-
-The process duration $T_M$ represents how many characteristic timescales the biological process spans:
-
-$$T_M = \frac{\text{Biological process duration}}{\text{Characteristic timescale of cellular rates}}$$
-
-**Biological interpretation examples:**
-
-- $T_M = 1$: Process duration equals one characteristic timescale
-- $T_M = 5$: Process duration equals five characteristic timescales
-- $T_M = 0.5$: Process duration is half a characteristic timescale
-
-**Steady-state achievement**: The fraction of steady-state dynamics observable is $1 - e^{-T_M}$:
-- $T_M = 1$: Observes 63% approach to steady state
-- $T_M = 3$: Observes 95% approach to steady state
-- $T_M = 5$: Observes >99% approach to steady state
-
-#### **Connection to Real Timescales During Model Fitting**
-
-When fitting the model to experimental data, the inferred $T_M$ reveals the underlying biological rates:
-
-$$\text{Inferred characteristic timescale} = \frac{\text{Known experimental duration}}{T_M^{\text{posterior}}}$$
-
-**Practical example:**
-
-- **Experimental setup**: 24-hour differentiation time course
-- **Posterior estimate**: $T_M \approx 3.0$
-- **Inferred characteristic timescale**: $24/3 = 8$ hours
-- **Biological interpretation**: Cellular processes (splicing, degradation) occur on ~8 hour timescales
-
-This bidirectional relationship enables both forward simulation (given biological rates) and inverse inference (estimating rates from data).
-
-#### **Gene Parameter Interpretation in Dimensionless Framework**
-
-**Activation Onset Times**: $t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$
-
-- **Dimensionless interpretation**: Gene $i$ activates at $t^*_{on,i}$ characteristic timescales after process initiation
-- **Real-time interpretation**: Gene $i$ activates at $t^*_{on,i}/\alpha_{\text{ref}}$ hours after process start
-- **Relative interpretation**: Gene $i$ activates at $(t^*_{on,i}/T_M) \times 100\%$ through the process
-
-**Example interpretations:**
-
-- $t^*_{on,i} = 1.5$, $T_M = 5.0$: Gene activates 30% through the process
-- $t^*_{on,i} = -0.5$: Gene was already active 0.5 timescales before observation began
-
-**Activation Durations**: $\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$
-
-- **Dimensionless interpretation**: Gene $i$ remains active for $\delta^*_i$ characteristic timescales
-- **Real-time interpretation**: Gene $i$ remains active for $\delta^*_i/\alpha_{\text{ref}}$ hours
-- **Relative interpretation**: Gene $i$ remains active for $(\delta^*_i/T_M) \times 100\%$ of the process
-
-#### **Hyperparameter Choice as Biological Prior Knowledge**
-
-The choice of $T_M$ hyperparameters encodes specific biological assumptions about process duration:
-
-**Conservative Prior: $T_M \sim \text{Gamma}(5.0, 1.0)$**
-
-- $\mathbb{E}[T_M] = 5$, $\text{Var}[T_M] = 5$
-- **Biological assumption**: "I expect this process to span approximately 5 characteristic timescales, allowing observation of near-complete dynamics"
-- **Implications**: If experimental duration is 20 hours, assumes characteristic timescale ≈ 4 hours
-- **Expectation**: Observes 99% approach to steady state ($1 - e^{-5} \approx 0.99$)
-- **Data burden**: Data must provide strong evidence if process is much shorter or longer
-
-**Permissive Prior: $T_M \sim \text{Gamma}(1.0, 0.2)$**
-
-- $\mathbb{E}[T_M] = 5$, $\text{Var}[T_M] = 25$
-- **Biological assumption**: "I'm uncertain about process duration - it could range from rapid (1 timescale) to very slow (15+ timescales)"
-- **Implications**: Wide uncertainty allows data to determine appropriate timescale
-- **Flexibility**: Can accommodate both early transient dynamics and complete processes
-- **Data burden**: Minimal prior bias on process duration
-
-**Skeptical Prior: $T_M \sim \text{Gamma}(1.0, 1.0)$**
-
-- $\mathbb{E}[T_M] = 1$, $\text{Var}[T_M] = 1$
-- **Biological assumption**: "I expect rapid cellular responses, likely observing only early dynamics"
-- **Implications**: Assumes observation of primarily transient phase (63% approach: $1 - e^{-1} \approx 0.63$)
-- **Data burden**: Data must strongly support longer timescales if present
-- **Use cases**: Stress responses, drug treatments, acute perturbations
-
-**Priors for other gene-specific parameters:**
-\begin{align}
-    U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
-\end{align}
-
-**Cell-specific capture efficiency:**
-\begin{align}
-    \lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
-\end{align}
-
-#### **Model Fitting and Timescale Inference**
-
-The dimensionless framework creates a bidirectional relationship between process duration and characteristic rates:
-
-**Forward (simulation):**
-
-1. Choose biological timescales (e.g., β = splicing rate)
-2. Set process duration in real time (e.g., 24 hours)
-3. Compute $T_M = t_{\text{process}} \times β$
-4. Generate dimensionless dynamics
-
-**Inverse (inference from data):**
-
-1. Observe expression data over known duration
-2. Infer $T_M$ from temporal patterns
-3. Estimate characteristic rates: $\hat{β} = T_M/t_{\text{experiment}}$
-4. Convert all dimensionless parameters to real time
-
-**Practical Example:**
-
-**Experimental setup:** 48-hour differentiation time course
-
-**Model fitting results:**
-
-- Posterior: $T_M \approx 6.0 \pm 0.5$
-- Inferred characteristic timescale: $48/6 = 8$ hours
-- Gene with $t^*_{on,i} = 2.0$: Activates at $2.0 \times 8 = 16$ hours
-- Gene with $\delta^*_i = 1.5$: Active for $1.5 \times 8 = 12$ hours
-
-#### **Independent Parameterization Justification**
-
-In the dimensionless framework, **gene timing parameters represent intrinsic biological properties** that should not scale with the arbitrary choice of observation window:
-
-- $t^*_{on,i}$: When a gene's regulatory program activates (intrinsic timing)
-- $\delta^*_i$: How long a gene's regulatory program lasts (intrinsic duration)
-
-**These are biological constants** that don't change based on how long we choose to observe the system. Therefore:
-
-\begin{align}
-t^*_{on,i} &\sim \text{Normal}(1.5, 0.8^2) \quad \text{(fixed biological timing)} \\
-\delta^*_i &\sim \text{LogNormal}(0.0, 0.45^2) \quad \text{(fixed biological duration)}
-\end{align}
-
-**NOT dependent on $T_M$:**
-\begin{align}
-t^*_{on,i} &\not\sim \text{Normal}(0.3 \times T_M, 0.8^2) \quad \text{(would create artificial scaling)}
-\end{align}
-
-This independence ensures that gene regulatory timing is treated as an intrinsic biological property rather than an artifact of experimental design.
-
 ### Identifiability Proof for Independent Parameterization
